### 2009-01-05 crystallography/paper.tex
initial commit with a skeleton crystallography paper based on the GECCO MRPSO paper.

\documentclass[letterpaper]{sig-alternate}

% GECCO Submission Number: 
% GECCO Camera-Ready Number:

\usepackage{cite}
\usepackage{graphicx}
\usepackage{psfrag}
\usepackage{subfigure}
\usepackage{listings}
\usepackage{amsmath,amssymb}
\usepackage{array}
\usepackage{algorithm}

\lstloadlanguages{Python}
\floatname{algorithm}{Function}
%\lstset{numbers=left, numberstyle=\tiny}

\newlength{\figwidth}
\newlength{\graphwidth}
\newlength{\figheight}
\newlength{\figsmallheight}
\setlength{\figwidth}{0.46\textwidth}
\setlength{\graphwidth}{0.46\textwidth}
\setlength{\figheight}{0.26\textheight}
\setlength{\figsmallheight}{0.8\figheight}

\providecommand{\Norm}[2][2]{\ensuremath{\|#2\|_{#1}}}
\providecommand{\Est}[1]{\ensuremath{\widehat{#1}}}
\renewcommand{\Vec}[1]{\mathbf{#1}}
\providecommand{\Dims}{\ensuremath{D}}
\providecommand{\PosBest}{\ensuremath{{\Pos^\star}}}
\providecommand{\PosElement}{\ensuremath{x}}
\providecommand{\Pos}{\ensuremath{\Vec{\PosElement}}}
\providecommand{\PosMatrix}{\ensuremath{\Vec{X}}}
\providecommand{\Func}{\ensuremath{\mathcal{F}}}
\providecommand{\Val}{\ensuremath{y}}
\providecommand{\ValVector}{\ensuremath{\Vec{y}}}
\providecommand{\ValBest}{\ensuremath{{\Val^\star}}}
\providecommand{\Optfunc}{\ensuremath{\Omega}}
\providecommand{\Samplefunc}{\ensuremath{\Delta}}
\providecommand{\Density}{\ensuremath{\mathrm{\pi}}}
\providecommand{\Prob}{\ensuremath{\mathrm{p}}}
\providecommand{\Scale}{\ensuremath{\alpha}}
\providecommand{\Exponent}{\ensuremath{\beta}}
\providecommand{\Normal}{\ensuremath{\mathcal{N}}}
\providecommand{\Particle}{\ensuremath{\Vec{s}}}
\providecommand{\ParticleMatrix}{\ensuremath{\Vec{S}}}
\providecommand{\AdditionalState}{\ensuremath{\Vec{\phi}}}

%% PSO Stuff
\DeclareMathOperator{\URand}{U}
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\DeclareMathOperator{\Diversity}{diversity}
\providecommand{\Vmax}{\ensuremath{V_\text{max}}}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\posmean}{\ensuremath{\bar{\ppos}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\pacc}{\ensuremath{\ddot{\ppos}}}
\providecommand{\gbest}{\ensuremath{\Vec{g}}}
\providecommand{\greallybest}{\ensuremath{\Vec{g}^\star}}
\providecommand{\pbest}{\ensuremath{\Vec{p}}}
\providecommand{\bestmean}{\ensuremath{\bar{\pbest}}}
\providecommand{\momentum}{\ensuremath{\omega}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\coeff}{\ensuremath{\phi}}
\providecommand{\sttrue}{\ensuremath{\Vec{\theta}}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}
\providecommand{\transition}{\ensuremath{\mathcal{F}}}
\providecommand{\observation}{\ensuremath{\mathcal{H}}}

%% Other notation
\providecommand{\FeasibleDiagonal}{\ensuremath{L}}
\providecommand{\Radius}{\ensuremath{r}}
\providecommand{\Adapt}{\ensuremath{\gamma}}
\providecommand{\BounceCount}{\ensuremath{b}}
\providecommand{\Swarm}{\ensuremath{S}}
\providecommand{\ArpsoLow}{\ensuremath{\eta^-}}
\providecommand{\ArpsoHigh}{\ensuremath{\eta^+}}
\providecommand{\Direction}{\ensuremath{s}}

% MapReduce notation:
\providecommand{\List}[1]{\ensuremath{\mathsf{list}(#1)}}
\providecommand{\Pair}[2]{\ensuremath{(#1, #2)}}
\providecommand{\NaturalSet}{\ensuremath{\mathbb{N}}}
\providecommand{\IntegerSet}{\ensuremath{\mathbb{Z}}}
\providecommand{\StringsSet}{\ensuremath{\mathsf{Strings}}}
\providecommand{\StateStringsSet}{\ensuremath{\mathsf{StateStrings}}}
\providecommand{\InputKeySet}{\ensuremath{K_1}}
\providecommand{\OutputKeySet}{\ensuremath{K_2}}
\providecommand{\InputValSet}{\ensuremath{V_1}}
\providecommand{\OutputValSet}{\ensuremath{V_2}}

% Make figures easier to place on the page with the text
\renewcommand{\dblfloatpagefraction}{0.70}
\renewcommand{\dbltopfraction}{0.8}
\renewcommand{\textfraction}{0.1}

% Names of things
\providecommand{\RadiusSEPSOLong}{Contracting Radius SEPSO}
\providecommand{\DistanceSEPSOLong}{Contracting Radius, Increasing Bounce SEPSO}
\providecommand{\RadiusSEPSO}{CRS}
\providecommand{\DistanceSEPSO}{CRIBS}

%\widowpenalty=10000
%\clubpenalty=10000
\begin{document}

\hyphenation{}

\conferenceinfo{GECCO'07,} {July 7--11, 2007, London, England.}
\CopyrightYear{2007}
%\crdata{1-59593-186-4/06/0007}

\title{MRPSO: MapReduce Particle Swarm Optimization}
\numberofauthors{3}
%\author{%
%\alignauthor Andrew McNabb\\
%    \affaddr{Brigham Young University}\\
%    \affaddr{Computer Science Dept.}\\
%    \affaddr{3361 TMCB, Provo, UT 84602}\\
%    \email{a@cs.byu.edu}
%\alignauthor Christopher Monson\\
%    \affaddr{Google, Inc.}\\
%    \affaddr{4720 Forbes Ave., Lower Level}\\
%    \affaddr{Pittsburgh, PA 15213}\\
%    \email{c@cs.byu.edu}
%\alignauthor Kevin Seppi\\
%    \affaddr{Brigham Young University}\\
%    \affaddr{Computer Science Dept.}\\
%    \affaddr{3361 TMCB, Provo, UT 84602}\\
%    \email{k@cs.byu.edu}
%}
\date{}

\author{Author names omitted for review}

\maketitle

\begin{abstract}
TODO.
\end{abstract}

\section*{Track Category}
FIXME. Ant Colony Optimization, Swarm Intelligence, and Artificial Immune Systems

\category{FIXME. G.1.6}{Numerical Analysis}{Optimization}[\\ nonlinear programming, unconstrained optimization]

\terms{FIXME. Algorithms}

\keywords{FIXME. Swarm intelligence, Parallelization, Optimization}

\section{Introduction}

TODO.

\section{Particle Swarm Optimization}
\label{sec:pso}

TODO: **Matt, you probably want to put this in your own words, but if we
really run out of time, you don't really have to.**

Particle Swarm Optimization simulates the motion of particles in the domain of
a function.  These particles search for the optimum by evaluating the function
as they move.  During each iteration of the algorithm, the position and
velocity of each particle are updated.  Each particle is pulled toward the
best position it has sampled (personal best) and the best position of any
particle in its neighborhood (global best).  This attraction is weak enough to
allow exploration but strong enough to encourage exploitation of good
locations and to guarantee convergence.

Each particle's position and velocity are initialized to random values based
on a function-specific feasible region.  During each iteration of constricted
PSO, the following equations update a particle's position $\ppos$ and velocity
$\pvel$ with respect to personal best $\pbest$ and global best $\gbest$:
\begin{align}
\label{eq:velupdate}
    \pvel_{t+1} &=
        \constriction \left[ \pvel_t + 
            \coeff_1\URand()\otimes(\pbest - \ppos_{t}) +
            \coeff_2\URand()\otimes(\gbest - \ppos_{t})
        \right] \\
\label{eq:posupdate}
    \ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \(\coeff_1 = \coeff_2 = 2.05\), \(\URand()\) is a vector of samples
drawn from a standard uniform distribution, and \( \otimes \) represents
element-wise multiplication.  The constriction constant \( \constriction \)
is:
\begin{align}
    \notag
    \constriction &= \frac{2 \kappa}{|2 - \coeff - \sqrt{\coeff^2 - 4\coeff}|}
\end{align}
where
$\coeff = \coeff_1 + \coeff_2$ and $\kappa$ is usually
$1.0$.~\citep{clerc-tec02}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MRPSO

\section{MapReduce}
\label{sec:mapreduce}

TODO: **Matt, you probably want to put this in your own words, but if we
really run out of time, you don't really have to.  This should also be trimmed
down.**

MapReduce is a functional programming model that is well suited to parallel
computation.  In the model, a program consists of a map function and a reduce
function which meet a few simple requirements.  These functions are only
concerned with the high-level task of processing input and returning output.
For formulating a problem in terms of a map function and a reduce function, a
programmer is rewarded with automatic parallelization, with communication,
load balancing, fault tolerance, and all other details managed by the
MapReduce implementation.

In MapReduce, all data are in the form of keys with associated values.  For
example, in a program that counts the frequency of occurrences for various
words, the key would be a word and the value would be its frequency.

A MapReduce operation takes place in two main stages.  In the first stage, the
map function is called once for each input record.  At each call, it may
produce any number of output records.  In the second stage, this intermediate
output is sorted and grouped by key, and the reduce function is called once
for each key.  The reduce function is given all associated values for the
key and outputs a new list of values (often ``reduced'' in length from the
original list of values).

The following notation and example are based on the original
presentation~\citep{dean-osdi04}.

\subsection{Map Function}

A map function is defined as a function that takes a single key-value pair and
outputs a list of new key-value pairs.  The input key may be of a different
type than the output keys, and the input value may be of a different type than
the output values:
\begin{align}
\mathsf{map}:\:&(\InputKeySet, \InputValSet) \rightarrow
    \List{(\OutputKeySet, \OutputValSet)}
\end{align}

Since the map function only takes a single record, all map operations are
independent of each other and fully parallelizable.

\subsection{Reduce Function}

A reduce function is a function that reads a key and a corresponding list of
values and outputs a new list of values for that key.  The input and output
values are of the same type.  Mathematically, this would be written:
\begin{align}
\mathsf{reduce}:\:&\Pair{\OutputKeySet}{\List{\OutputValSet}} \rightarrow
\List{\OutputValSet}
\end{align}

A reduce operation may depend on the output from any number of map calls, so
no reduce operation can begin until all map operations have completed.
However, the reduce operations are independent of each other and may be run in
parallel.

Although the formal definition of map and reduce functions would indicate
building up a list of outputs and then returning the list at the end, it is
more convenient in practice to emit one element of the list at a time and
return nothing.  Conceptually, these emitted elements still constitute a list.

\subsection{Example: WordCount}

The classic MapReduce example is WordCount, a program which counts the number
of occurrences of each word in a document or set of documents.  For this
program, the input and output sets are:
\begin{align}
\InputKeySet:\: &\NaturalSet
\label{eq:wc-inputkey}
\\
\InputValSet:\: & \text{set of all strings}
\label{eq:wc-inputval}
\\
\OutputKeySet:\: &\text{set of all strings}
\label{eq:wc-outputkey}
\\
\OutputValSet:\: &\NaturalSet
\label{eq:wc-outputval}
\end{align}

In WordCount, the input value is a line of text.  The input key is ignored but
arbitrarily set to be the line number for the input value.  The output key is
a word, and the output value is its count.

\begin{algorithm}[tbp]
\caption{WordCount Map}
\label{alg:wc-map}
\begin{lstlisting}[language=Python]
def mapper(key, value):
    for word in value.split():
        emit((word, 1))
\end{lstlisting}
\end{algorithm}

\begin{algorithm}[tbp]
\caption{WordCount Reduce}
\label{alg:wc-reduce}
\begin{lstlisting}[language=Python]
def reducer(key, value_list):
    total = sum(value_list)
    emit(total)
\end{lstlisting}
\end{algorithm}

The map function, shown as Function~\ref{alg:wc-map}, splits the input line
into individual words.  For each word, it emits the key-value pair formed by
the word and the value 1.

The reduce function, shown as Function~\ref{alg:wc-reduce}, takes a word and
list of counts, performs a sum reduction, and emits the result.  This is the
only element emitted, so the output of the reduce function is a list of size
1.

These map and reduce functions are deceptively simple.  The problem itself is
inherently difficult---implementing a scalable distributed word count system
with fault-tolerance and load-balancing is not easy.  However all of the
complexity is found in the surrounding MapReduce infrastructure rather than in
the map and reduce functions.  Note that the reduce function does not even
output a key, since the MapReduce system already knows what key it passed in.

The data given to map and reduce functions, as in this example, are generally
as fine-grained as possible.  This ensures that the implementation can split
up and distribute tasks.  The MapReduce system consolidates the intermediate
output from all of the map tasks.  These records are sorted and grouped by
key before being sent to the reduce tasks.

If the map tasks emit a large number of records, the sort phase can take a
long time.  MapReduce addresses this potential problem by introducing the
concept of a combiner function.  If a combiner is available, the MapReduce
system will locally sort the output from several map calls on the same machine
and perform a ``local reduce'' using the combiner function.  This reduces the
amount of data that must be sent over the network for the main sort leading to
the reduce phase. In WordCount, the reduce function would work as a combiner
without any modifications.

\subsection{Benefits of MapReduce}

Although not all algorithms can be efficiently formulated in terms of map and
reduce functions, MapReduce provides many benefits over other parallel
processing models.  In this model, a program consists of only a map function
and a reduce function.  Everything else is common to all programs.  The
infrastructure provided by a MapReduce implementation manages all of the
details of communication, load balancing, fault tolerance, resource
allocation, job startup, and file distribution.  This runtime system is
written and maintained by parallel programming specialists, who can ensure
that the system is robust and optimized, while those who write mappers and
reducers can focus on the problem at hand without worrying about
implementation details.

A MapReduce system determines task granularity at runtime and
distributes tasks to compute nodes as processors become available.  If some
nodes are faster than others, they will be given more tasks, and if a node
fails, the system automatically reassigns the interrupted task.

\subsection{MapReduce Implementations}

Google has described its MapReduce implementation in published papers and
slides, but it has not released the system to the public.  Presumably the
implementation is highly optimized because Google uses it to produce its web
index.

The Apache Lucene project has developed Hadoop, an Java-based open-source
clone of Google's closed MapReduce platform.  The platform is relatively new
but rapidly maturing.  At this time, Hadoop overhead is significant but not
overwhelming, and is expected to decrease with further development.

\section{MapReduce PSO (MRPSO)}
\label{sec:mrpso}

%(1, ``1,2,3;1.3,-0.5;-0.8,0.4;1.94;0.9,0.7;1.3;-0.2,1.0;1.04")
\begin{figure*}
\begin{align*}
(3, ``1,2,3,4;1.7639,2.5271;52.558,50.444;9.4976;1.7639,2.5271;9.4976;-1.0151,-2.0254;5.1325")
\end{align*}
\caption{A particle as a key-value pair}
\label{fig:sample-particle}
\end{figure*}

\begin{algorithm*}[tbp]
\caption{MRPSO Map}
\label{alg:pso-map}
\begin{lstlisting}[language=Python]
def mapper(key, value):
    particle = Particle(value)

    # Update the particle:
    new_position, new_velocity = pso_motion(particle)
    y = evaluate_function(new_position)
    particle.update(new_position, new_velocity, y)

    # Emit a message for each dependent particle:
    message = particle.make_message()
    for dependent_id in particle.dependent_list:
        if dependent_id == key:
            particle.gbest_candidate(particle.pbest_position, particle.pbest_value)
        else:
            emit((dependent_id, repr(message)))

    # Emit the updated particle without changing its id:
    emit((key, repr(particle)))
\end{lstlisting}
\end{algorithm*}

\begin{algorithm*}[tbp]
\caption{MRPSO Reduce}
\label{alg:pso-reduce}
\begin{lstlisting}[language=Python]
def reducer(key, value_list):
    particle = None
    best = None

    # Of all of the inputs, find the record with the best gbest_value:
    for value in value_list:
        record = Particle(value)
        if (best is None) or (record.gbest_value <= best.gbest_value):
            best = record
        if not record.is_message():
            particle = record

    # Update the gbest of the particle and emit:
    if particle is not None:
        particle.gbest_candidate(best.gbest_position, best_value)
        emit(repr(particle))
    else:
        emit(repr(best))
\end{lstlisting}
\end{algorithm*}

% The first sentence here is still broken:

In an iteration of Particle Swarm Optimizations, each particle in the swarm
moves to a new position, updates its velocity, evaluates the function at the
new point, updates its personal best if this value is the best seen so far,
and updates its global best after comparison with its neighbors.  Except for
updating its global best, each particle updates independently of the rest of
the swarm.

Due to the limited communication among particles, updating a swarm can be
formulated as a MapReduce operation.  As a particle is mapped, it receives a
new position, velocity, value, and personal best.  In the reduce phase, it
incorporates information from other particles in the swarm to update its
global best.  The MRPSO implementation conforms to the MapReduce model while
performing the same calculations as standard Particle Swarm Optimization.

\subsection{Particle Representation and Messages}

In MRPSO, the input and output sets are:
\begin{align}
\InputKeySet:\: &\NaturalSet
\label{eq:pso-inputkey}
\\
\InputValSet:\: & \text{set of all strings}
\label{eq:pso-inputval}
\\
\OutputKeySet:\: &\NaturalSet
\label{eq:pso-outputkey}
\\
\OutputValSet:\: &\text{set of all strings}
\label{eq:pso-outputval}
\end{align}

Each particle is identified by a numerical \verb$id$ key, and particle state
is represented as a string.  The state of a particle consists of its
dependents list (neighbors' \verb$id$s), position, velocity, value, personal
best position, personal best value, global best position, and global best
value.  The state string is a semicolon-separated list of fields.  If a field
is vector valued, its individual elements are comma-separated.  The state
string is of the form:
\begin{align*}
\text{deps};\text{pos};\text{vel};\text{val};\text{pbpos};\text{pbval};\text{gbpos};\text{gbval}
\end{align*}

A typical particle is shown in Figure~\ref{fig:sample-particle}.  This
particle is exploring the function $f(\Vec{x}) = x_1^2 + x_2^2$.  Its
components are interpreted as follows:

\begin{tabbing}
$3$ \hspace{72pt} \= particle id \\
$1,2,3,4$ \> dependents (neighbors) \\
$1.7639,2.5271$ \> current position $(x_1, x_2)$ \\
$52.558,50.444$ \> velocity $(x_1, x_2)$ \\
$9.4976$ \> value of $f(\Vec{x})$ at the current position \\
\> \hspace{16pt} ($1.7639^2 + 2.5271^2$) \\
$1.7639,2.5271$ \> personal best position $(x_1, x_2)$ \\
$9.4976$ \> personal best value \\
$-1.0151,-2.0254$ \> global best position $(x_1, x_2)$ \\
$5.1325$ \> global best value \\
\end{tabbing}

MRPSO also creates messages, which are like particles except that they have
empty dependents lists.  A message is sent from one particle to another as
part of the MapReduce operation.  In the reduce phase, the recipient reads the
personal best from the message and updates its global best accordingly.

\subsection{MRPSO Map Function}

The MRPSO map function, shown as Function~\ref{alg:pso-map}, is called once
for each particle in the population.  The key is the \verb$id$ of the
particle, and the value is its state string representation.  The PSO mapper
finds the new position and velocity of the particle and evaluates the function
at the new point.  It then calls the update method of the particle with this
information.  In addition to modifying the particle's state to reflect the new
position, velocity, and value, this method replaces the personal best if a
more fit position has been found.

The key to implementing PSO in MapReduce is communication between particles.
Each particle maintains a dependents list containing the \verb$id$s of all
neighbors that need information from the particle to update their own global
bests.  After the map function updates the state of a particle, it emits
messages to all dependent particles.  When a message is emitted, its
corresponding key is the \verb$id$ of the destination particle, and its value
is the string representation, which includes the position, velocity, value,
and personal best of the source particle.  The global best of the message is
also set to the personal best.

If the particle is a dependent of itself, as is usually the case, the map
function updates the global best of the particle if the personal best is an
improvement.  Finally, the map function emits the updated
particle and terminates.

\subsection{MRPSO Reduce Function}

The MRPSO reduce function, shown as Function~\ref{alg:pso-reduce}, receives a
key and a list of all associated values.  The key is the \verb$id$ of a
particular particle in the population.  The values list contains the newly
updated particle and a message from each neighbor.  The PSO reducer combines
information from all of these messages to update the global best position and
global best value of the particle.  The reducer emits only the updated
particle.

Function~\ref{alg:pso-reduce} also works as a combiner.  If no particle is
found in the input value list, the function combines the list by emitting only
the best message.  This message is then sent to the reducer.

\subsection{Map and Reduce in Context}

When a particle is emitted by a reducer, it is fully updated.  In the map
phase, it updated its velocity, moved to a new position, evaluated the
function at that position, and updated its personal best.  In the reduce
phase, it updated its global best after receiving messages from all of its
neighbors in the swarm.  A map phase followed by a reduce phase performs an
iteration of the swarm that is exactly equivalent to an iteration in
single-processor PSO.

Observe that the MRPSO implementation does not explicitly deal with
communication across nodes, load balancing, or node failures.  The MapReduce
formulation of the problem allows the work to be divided in small enough
pieces that the MapReduce system can balance work across processors and deal
with failed nodes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography

\bibliographystyle{plain}
\bibliography{%
\bib{mapreduce/dean-osdi04},%
\bib{pso/clerc-tec02},%
\bib{pso/kennedy-icnn95},%
\bib{pso/bratton-sis07},%
\bib{aml/mcnabb-cec07}}

\end{document}
% vim: lbr et sw=4 sts=4

### 2009-01-05 crystallography/paper.tex
removed some great features so paper.tex will actually compile

@@
 
-\usepackage{cite}
 \usepackage{graphicx}
@@
 \setlength{\figsmallheight}{0.8\figheight}
+
+% AML bibliography references
+\newcommand{\bib}[1]{../../papers/#1}
 
@@
 
-% Names of things
-\providecommand{\RadiusSEPSOLong}{Contracting Radius SEPSO}
-\providecommand{\DistanceSEPSOLong}{Contracting Radius, Increasing Bounce SEPSO}
-\providecommand{\RadiusSEPSO}{CRS}
-\providecommand{\DistanceSEPSO}{CRIBS}
-
 %\widowpenalty=10000
@@
 
-TODO: **Matt, you probably want to put this in your own words, but if we
-really run out of time, you don't really have to.**
+TODO: Matt, you probably want to put this in your own words, but if we
+really run out of time, you don't really have to.
 
@@
 \end{align}
-where \(\coeff_1 = \coeff_2 = 2.05\), \(\URand()\) is a vector of samples
-drawn from a standard uniform distribution, and \( \otimes \) represents
-element-wise multiplication.  The constriction constant \( \constriction \)
+where $\coeff_1 = \coeff_2 = 2.05$, $\URand()$ is a vector of samples
+drawn from a standard uniform distribution, and $\otimes$ represents
+element-wise multiplication.  The constriction constant $\constriction$
 is:
@@
 $\coeff = \coeff_1 + \coeff_2$ and $\kappa$ is usually
-$1.0$.~\citep{clerc-tec02}
+$1.0$~\cite{clerc-tec02}.
 
@@
 
-TODO: **Matt, you probably want to put this in your own words, but if we
+TODO: Matt, you probably want to put this in your own words, but if we
 really run out of time, you don't really have to.  This should also be trimmed
-down.**
+down.
 
@@
 The following notation and example are based on the original
-presentation~\citep{dean-osdi04}.
+presentation~\cite{dean-osdi04}.
 

### 2009-01-06 crystallography/paper.tex
Modified date on paper.tex, added a few papers
Started on the crystallography description

@@
 
-\conferenceinfo{GECCO'07,} {July 7--11, 2007, London, England.}
-\CopyrightYear{2007}
+\conferenceinfo{GECCO'09,} {July 8--12, 2009, Montr\'eal, Canada.}
+\CopyrightYear{2009}
 %\crdata{1-59593-186-4/06/0007}
@@
 \title{MRPSO: MapReduce Particle Swarm Optimization}
-\numberofauthors{3}
+\numberofauthors{4}
 %\author{%
+%\alignauthor Matthew Gardner\\
+%    \affaddr{Brigham Young University}\\
+%    \affaddr{Computer Science Dept.}\\
+%    \affaddr{3361 TMCB, Provo, UT 84602}\\
+%    \email{mjg82@cs.byu.edu}
 %\alignauthor Andrew McNabb\\
@@
 %    \email{a@cs.byu.edu}
-%\alignauthor Christopher Monson\\
-%    \affaddr{Google, Inc.}\\
-%    \affaddr{4720 Forbes Ave., Lower Level}\\
-%    \affaddr{Pittsburgh, PA 15213}\\
-%    \email{c@cs.byu.edu}
+%\alignauthor Branton Campbell\\
+%    \affaddr{Brigham Young University}\\
+%    \affaddr{Physics Dept.}\\
+%    \affaddr{N261 ESC, Provo, UT 84602}\\
+%    \email{branton_campbell@byu.edu}
 %\alignauthor Kevin Seppi\\
@@
 \section*{Track Category}
-FIXME. Ant Colony Optimization, Swarm Intelligence, and Artificial Immune Systems
+FIXME. Ant Colony Optimization, Swarm Intelligence
 
@@
 \section{Introduction}
+\label{sec:intro}
 
 TODO.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Crystallography
+
+\section{Crystallography}
+\label{sec:crystallography}
+
+FIXME.
+TODO: Where to put the butterfly problem definition? Should I describe 
+simulated annealing or gradient descent, at least briefly?
+
+The point of crystallography is to determine the structure of a crystal.  This
+is done by shining x-rays through the crystal and observing the patterns that
+come out.  Those patterns are then compared to the output of a model of the 
+crystal -- if the two match, the model can be considered correct.  The model
+may consist of the positions of all of the atoms in the crystal, the 
+distortion coefficients of the materials, the length and width of nanopolar
+domains, or various other parameters.  The task of finding the parameters to 
+the model that best fit the data is well suited to optimization techniques.  
+The function to be optimized is an error function -- the average difference
+between the model output and the data, for every data point.
+
+There are commercial packages that readily solve some crystallography problems,
+such as (ask Dr. Campbell), that uses simulated annealing for specific types 
+of problems.  Gradient descent is also a commonly used method for solving these
+problems.  Particle swarm optimization, however (to our knowledge?), has not 
+been used.
+
+We encountered a problem for which gradient descent proved uneffective.
+
+To solve crystallography problems, one must fit the parameters of the model
+to the data.  The more data is used, the more accurate the result will be, but
+with large amounts of data the evaluation time can be inhibitively large.  In
+our first attempts at solving this problem, the function evaluation took close
+to an hour.  We then trimmed the amount of data we were using, and the time for
+a function evaluation decreased to about four minutes.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% PSO
 

### 2009-01-06 crystallography/paper.tex
Some initial work on an intro

@@
 
-\title{MRPSO: MapReduce Particle Swarm Optimization}
+\title{Particle Swarm Optimization for Crystallography}
 \numberofauthors{4}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Introduction
+
 \section{Introduction}
@@
 TODO.
+Particle swarm optimization is most often used with functions that are quickly
+evaluated.  Many real-life problems take much longer.  What does one do when 
+presented with a function that takes several minutes or hours to evaluate?
+
+In this paper we present a crystallography problem whose function evaluation
+takes several minutes to evaluate.  We discuss issues in solving this problem
+specifically, and in optimizing expensive functions in general with particle 
+swarm optimization.
 

### 2009-01-07 crystallography/paper.tex
Some initial work on an intro
Reworked introduction, started on results

@@
 \begin{abstract}
-TODO.
+FIXME.
+
+In this paper we present a crystallography problem whose function evaluation
+takes several minutes to evaluate.  We discuss issues in solving this problem
+specifically, and in optimizing expensive functions in general with particle 
+swarm optimization.
 \end{abstract}
@@
 
-TODO.
-Particle swarm optimization is most often used with functions that are quickly
-evaluated.  Many real-life problems take much longer.  What does one do when 
-presented with a function that takes several minutes or hours to evaluate?
+FIXME.
 
-In this paper we present a crystallography problem whose function evaluation
-takes several minutes to evaluate.  We discuss issues in solving this problem
-specifically, and in optimizing expensive functions in general with particle 
-swarm optimization.
+Crystallography is the science of determining the structure of a crystal.  This
+is done by shining x-rays through the crystal and observing the patterns that
+come out, generally consisting of intensities of light at various points in
+space.  Those patterns are then compared to the output of a model of the
+crystal -- the predicted intensities of light at the same points in space.  If
+the two match, the model can be considered correct.  The model may consist of
+the positions of all of the atoms in the crystal, the distortion coefficients
+of the material, the length and width of nanopolar domains, or various other
+parameters.  The task of finding the parameters to the model that best fit the
+data is well suited to optimization techniques.  The function to be optimized
+is an error function -- the average difference between the model output and the
+data, for every data point.
+
+There are commercial packages that readily solve some crystallography problems,
+such as (ask Dr. Campbell), that uses simulated annealing for specific types of
+problems.  Gradient descent is also a commonly used method for solving these
+problems.  Particle swarm optimization (PSO), described in a later section, has
+also been used, but only in limited contexts \cite{liu-anc05}.
+
+To solve crystallography problems, one must fit the parameters of the model to
+the data.  The more data is used, the more accurate the result will be, but
+with large amounts of data the error function evaluation time can be
+inhibitively large, easily reaching minutes or even hours.  In such cases,
+performing any kind of optimization on a single machine is infeasible; parallel
+techniques are needed.
+
+We present a crystallography problem whose error function evaluation takes
+several minutes. We then present our solution to the problem, using MRPSO, a
+parallel particle swarm optimization implementation \cite{mcnabb-cec07}.  The
+issues we encountered and solved dealing with this crystallography problem are
+applicable to any problem whose function evaluation takes large amounts of
+time.
+
+Section~\ref{sec:butterfly} introduces the crystallography problem we solved.
+Section~\ref{sec:pso} explains the particle swarm optimization algorithm, and
+section~\ref{sec:mapreduce} explains MapReduce and PSO in the context of
+MapReduce.  Section~\ref{sec:results} then describes our solution to the
+problem.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Crystallography
-
-\section{Crystallography}
-\label{sec:crystallography}
-
-FIXME.
-TODO: Where to put the butterfly problem definition? Should I describe 
-simulated annealing or gradient descent, at least briefly?
-
-The point of crystallography is to determine the structure of a crystal.  This
-is done by shining x-rays through the crystal and observing the patterns that
-come out.  Those patterns are then compared to the output of a model of the 
-crystal -- if the two match, the model can be considered correct.  The model
-may consist of the positions of all of the atoms in the crystal, the 
-distortion coefficients of the materials, the length and width of nanopolar
-domains, or various other parameters.  The task of finding the parameters to 
-the model that best fit the data is well suited to optimization techniques.  
-The function to be optimized is an error function -- the average difference
-between the model output and the data, for every data point.
-
-There are commercial packages that readily solve some crystallography problems,
-such as (ask Dr. Campbell), that uses simulated annealing for specific types 
-of problems.  Gradient descent is also a commonly used method for solving these
-problems.  Particle swarm optimization, however (to our knowledge?), has not 
-been used.
-
-We encountered a problem for which gradient descent proved uneffective.
-
-To solve crystallography problems, one must fit the parameters of the model
-to the data.  The more data is used, the more accurate the result will be, but
-with large amounts of data the evaluation time can be inhibitively large.  In
-our first attempts at solving this problem, the function evaluation took close
-to an hour.  We then trimmed the amount of data we were using, and the time for
-a function evaluation decreased to about four minutes.
+% Butterfly Problem
+\section{Crystallography Problem}
+\label{sec:butterfly}
+TODO.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Results
+\section{Results (Maybe Issues with Optimization?)}
+\label{sec:results}
+TODO.
+
+In the course of solving this problem, it was discovered that a few of the 
+variables were correlated.  In initial attempts at optimization, the algorithm
+never converged -- the function value reached a certain point and stayed close
+to that point while a couple of the parameters fluctuated greatly in a 
+correlated fashion.  Figure (make figure) shows a range of values that two of
+parameters took on, without changing the function value more than (check number)
+.5, or this \%.  Upon noticing this empirical correlation, the function was
+more closely examined, and a theoretical correlation was also discovered.  
+Given that the parameters were correlated and the function had too many free
+variables, one of the parameters was fixed at a reasonable value, and the 
+optimization algorithm successfully and quickly solved the problem.
+
+Work has been done on the use of particle swarm optimization in parallel with 
+large numbers of particles \cite{mcnabb-gecco09}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Bibliography
@@
     \bib{pso/bratton-sis07},%
-    \bib{aml/mcnabb-cec07}}
+    \bib{aml/mcnabb-cec07},%
+    \bib{pso/crystallography/liu-anc05}}
 

### 2009-02-26 crystallography/paper.tex
Started on honors thesis, added some other things.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Results
-\section{Results (Maybe Issues with Optimization?)}
-\label{sec:results}
+% Issues with Optimization
+\section{Issues with Optimization}
+\label{sec:issues}
 TODO.
@@
 correlated fashion.  Figure (make figure) shows a range of values that two of
-parameters took on, without changing the function value more than (check number)
-.5, or this \%.  Upon noticing this empirical correlation, the function was
-more closely examined, and a theoretical correlation was also discovered.  
+parameters took on, without changing the function value more than (check
+number) .5, or this \%.  Upon noticing this empirical correlation, the function
+was more closely examined, and a theoretical correlation was also discovered.
 Given that the parameters were correlated and the function had too many free
@@
 
-Work has been done on the use of particle swarm optimization in parallel with 
-large numbers of particles \cite{mcnabb-gecco09}.
+The topology that was used in the PSO algorithm was the gbest topology.
+Experiments were done with lbest to determine which was more efficient, and it
+was discovered that for this problem gbest was better.  Averages over three
+runs are shown in Figure (make figure) for the two topologies.
+
+Work has been done showing that when running PSO in parallel with large swarms
+more complicated topologies can do just as well as gbest while decreasing the
+amount of communication required~\cite{mcnabb-gecco09}.  However, as the
+function evaluation for this problem takes several minutes, the amount of
+communication time for gbest is negligible when compared to the function
+evaluation time.  The communication overhead for each iteration was discovered
+to be about 8 seconds for the MRPSO implementation used.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Results
+\section{Results}
+\label{sec:results}
+TODO.
 

### 2009-02-26 honors_thesis/paper.tex
Started on honors thesis, added some other things.

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{Honors Thesis}
\author{Matthew Gardner}
\date{\today}

\begin{document}
\maketitle

\section{Statement of Intent}

\section{Background and Significance}

\section{Methodology/Procedures}

\section{Preliminary Prospectus of Finished Thesis}

\section{Preliminary Research}

\section{Qualifications of Investigator}

\section{Qualifications of Faculty Advisor}

\section{Schedule}

\section{Expenses/Budget}

\section{Closure}

\section{References}

\section{Appendices}

\end{document}

### 2009-02-27 honors_thesis/paper.tex
Working on the paper

@@
 
+\section{Background and Significance}
+
+Optimization problems are ubiquitous in our modern world.  Businesses such as
+Google need to decide where to place ads, scientists need to fit models to
+data, and airlines need to schedule their flights.  All of those problems are
+optimization problems, and optimization techniques can be used to find
+solutions.  Optimization algorithms are derived from all sorts of natural
+phenomena.  Some methods look at the landscape of the function as they are
+searching and follow hills to the best values, as in the gradient ascent
+method.  Other methods take ideas from biology or metallurgy, like genetic
+algorithms and simulated annealing.  Particle swarm optimization is a recently
+developed optimization technique that draws on ideas from the sociology of
+flocking birds.  All of these methods, however, are fundamentally sequential in
+nature---they need to be run in a specific order, and because of that, they are
+typically only run on one machine.  The problems people want to solve are
+getting bigger, and larger problems need to run on multiple machines in
+parallel.  Parallel implementations of these algorithms have been attempted,
+but they do not often perform as well as sequential versions for the amount of
+computation performed.  The work I propose focuses on one of those algorithms,
+particle swarm optimization, and improving its performance in a large-scale,
+parallel world.
+
 \section{Statement of Intent}
 
-\section{Background and Significance}
+I will study the feasibility and potential performance gain of speculative
+execution in particle swarm optimization.  By speculative execution I mean the
+use of additional computational resources, presumably extra processors running
+in parallel, to perform two or more iterations of particle swarm optimization
+at a time.
 
-\section{Methodology/Procedures}
+\section{Methodology}
+
+The methodology of this research will procede as follows:
+
+Compare to original How to control for external factors (nice in CS, though not
+in other sciences) Once you've run experiments, what are you looking for?
+Anything you can say about what code you're using, and how you intend to
+compare results.  People are concerned about how you're going to draw
+conclusions.  It matters what kind of statistical tests you are going to do.
+We're going to run a number of trials because it's a randomized algorithm.
+Test against multiple functions, with multiple topologies - some synthetic
+functions, and some real-world applications.  Run with standard constriction
+coefficient.  How do you compare your results to things other people have done?
+
+\begin{itemize}
+
+\item
+
+\end{itemize}
 
@@
 
+I am a Computer Science major.  My main focuses in my undergraduate career have
+been optimization, natural language processing, and machine learning.  The
+optimization emphasis has come mainly in my work with Dr. Kevin Seppi, for whom
+I am a research assistant.  I have spent the last year working with him,
+working mainly on particle swarm optimization topics.  We worked together with
+Dr. Branton Campbell, a physics professor, to solve a crystallography problem
+for him using a parallel implementation of PSO.  Those results will be
+submitted to Nature Materials.  Recently we, along with a graduate student who
+works in the lab, submitted a paper to the Congress on Evolutionary
+Computation, a major venue for optimization techniques.  The paper was on
+topologies and communication in large particle swarms, and formed the
+beginnings of the work I propose.
+
 \section{Qualifications of Faculty Advisor}
+
+Dr. Kevin Seppi is the head of the Applied Machine Learning Laboratory at BYU.
+Three of his last five graduate students have done their graduate work in
+particle swarm optimization, and he has authored or co-authored over 10 papers
+on the topic.  I took CS 236 from him in the fall of 2007 and decided I wanted
+to do research in the area he was researching, so I joined his lab.  I too
+became interested in PSO and have worked in that area for the past year.
 
@@
 \section{Expenses/Budget}
+
+This project will not require any outside funding.  The only expenditure
+will be my own time researching, conducting experiments, and writing.
 

### 2009-02-27 honors_thesis/paper.tex
Working on proposal

@@
 
-\section{Methodology}
+\section{Procedures}
 
-The methodology of this research will procede as follows:
+The bulk of the work for this project will be writing code and running
+experiments.  Research has been done to see what has been done in the field, to
+know what the results need to be compared to.  In implementing speculative
+execution, care will be taken that the algorithm behaves exactly like the
+original, standard particle swarm optimization (PSO) algorithm in every aspect
+except the speed-up.  The same constants will be used as are commonly published
+in the literature.
 
-Compare to original How to control for external factors (nice in CS, though not
-in other sciences) Once you've run experiments, what are you looking for?
-Anything you can say about what code you're using, and how you intend to
-compare results.  People are concerned about how you're going to draw
-conclusions.  It matters what kind of statistical tests you are going to do.
-We're going to run a number of trials because it's a randomized algorithm.
-Test against multiple functions, with multiple topologies - some synthetic
-functions, and some real-world applications.  Run with standard constriction
-coefficient.  How do you compare your results to things other people have done?
-
-\begin{itemize}
-
-\item
-
-\end{itemize}
+In running experiments, many trials will be run and averaged, as there is a
+random component in this algorithm.  Statistics will be used for all
+comparisons to the original PSO algorithm.  The success criterion will be a
+better performance of the algorithm as measured by the lowest function
+evaluation seen over a number of iterations for several benchmark and
+real-world problems.  Speculative execution takes more work than regular PSO
+with the same number of particles, so the two algorithms will be compared at
+points where they use equal amounts of processing power.
 
 \section{Preliminary Prospectus of Finished Thesis}
+
+\begin{enumerate}
+\item Introduction
+\item Related Work
+\item Particle Swarm Optimization
+\item Speculative Execution
+\item Speculative Execution in PSO
+\item Experiments
+\item Results and Conclusions
+\item Future Work
+\item References
+\end{enumerate}
 

### 2009-03-09 honors_thesis/paper.tex
Some revisions.

@@
 \usepackage{graphicx}
+\usepackage{cite}
+\usepackage{amsmath}
 
-\title{Honors Thesis}
+\DeclareMathOperator{\URand}{U}
+\providecommand{\ppos}{\ensuremath{\Vec{x}}}
+\providecommand{\pvel}{\ensuremath{\Vec{v}}}
+\providecommand{\gbest}{\ensuremath{\Vec{g}}}
+\providecommand{\pbest}{\ensuremath{\Vec{p}}}
+\providecommand{\constriction}{\ensuremath{\chi}}
+\providecommand{\coeff}{\ensuremath{\phi}}
+\newcommand{\bib}[1]{../../bib/#1}
+
+\topmargin .5in
+
+\title{Honors Thesis Proposal}
 \author{Matthew Gardner}
@@
 in parallel, to perform two or more iterations of particle swarm optimization
-at a time.
+at a time, speeding up the algorithm significantly.
 
@@
 The bulk of the work for this project will be writing code and running
-experiments.  Research has been done to see what has been done in the field, to
-know what the results need to be compared to.  In implementing speculative
-execution, care will be taken that the algorithm behaves exactly like the
-original, standard particle swarm optimization (PSO) algorithm in every aspect
-except the speed-up.  The same constants will be used as are commonly published
-in the literature.
+experiments.  I have already read many conference papers on particle swarm
+optimization to know the current state of the field, so I will know what the
+results need to be compared to.  In implementing speculative execution, I will
+take care that the algorithm behaves exactly like the original, standard
+particle swarm optimization (PSO) algorithm in every aspect except the
+speed-up.  The same constants will be used as are commonly published in the
+literature.
 
 In running experiments, many trials will be run and averaged, as there is a
-random component in this algorithm.  Statistics will be used for all
+random component in this algorithm.  I will use statistical tests for all
 comparisons to the original PSO algorithm.  The success criterion will be a
@@
 evaluation seen over a number of iterations for several benchmark and
-real-world problems.  Speculative execution takes more work than regular PSO
-with the same number of particles, so the two algorithms will be compared at
-points where they use equal amounts of processing power.
+real-world problems.  Speculative execution takes more computation than regular
+PSO with the same number of particles, so the two algorithms will be compared
+at points where they use equal amounts of processing power.
 
 \section{Preliminary Prospectus of Finished Thesis}
+
+The finished thesis will be presented as a conference paper, including the
+following sections:
 
@@
 \section{Preliminary Research}
+
+\subsection{Particle Swarm Optimization}
+
+Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
+Eberhart.  It tries to intelligently search a multi-dimensional space by
+mimicking the swarming and flocking behavior of birds and other
+animals.~\cite{kennedy-icnn95} It is a sociological algorithm that depends on
+interaction between particles to quickly and consistently find the optimal
+solution to a problem.  The algorithm keeps track of a number of potential
+solutions, called particles, which move somewhat randomly through the search
+space.  Particles remember the best place they have been, or solution they have
+evaluated, and are attracted back to that place, as well as to the best
+solution other particles have seen.  Specifically, the formulas for updating
+the position $\ppos_t$ and velocity $\pvel_t$ of a particle at iteration $t$ are
+as follows:
+\begin{align}
+\label{eq:velupdate}
+	\pvel_{t+1} &=
+		\constriction \left[ \pvel_t +
+			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
+			\coeff_2\URand()\otimes(\gbest - \ppos_t)
+		\right] \\
+\label{eq:posupdate}
+	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
+\end{align}
+where \( \URand() \) is a vector of random numbers drawn from a uniform
+distribution, the \( \otimes \) operator is an element-wise vector
+multiplication, $\pbest$ is the best position the current particle has seen,
+and $\gbest$ is the best position any of the other particles have seen.  \(
+\coeff_1 \), \( \coeff_2 \), and \( \constriction \) are parameters with
+prescribed values required to ensure convergence (2.05, 2.05, and .73,
+respectively).~\cite{clerc-tec02}~\cite{poli-aea08}
+
+The sociology in this algorithm defines which other particles form the
+``neighborhood'' of a given particle---the other particles whose best solution
+it sees.  There are many ways to define a particle's neighborhood, varying from
+the entire rest of the swarm to just one other particle.
+
+The way a neighborhood is defined can have drastic effects on the performance
+of the algorithm; the more neighbors each particle has, the faster information
+spreads throughout the particles, and the quicker the algorithm converges.  For
+some problems it is possible that the algorithm converges too quickly and gets
+caught in a local optimum; it stops on a hill when there is a mountain next to
+it.  For those kinds of problems, less communication, or smaller neighborhoods,
+is often preferable.
+
+\subsection{Speculative Execution}
+
+Speculative execution in a program is the execution of code that may or may not
+end up actually being needed.  Modern processors routinely do this when a
+conditional branch is encountered; they try to predict which branch will
+actually be needed, and continue their execution.  If it turns out that the
+prediction was incorrect, the work is discarded.  But if the prediction was
+right, the program can execute much faster than if it had waited on the branch.
+
+In particle swarm optimization, speculative execution can be performed because
+the next position a particle moves to doesn't depend on the value of the
+function at its current position.  That means you can just compute all possible
+next positions (depending on whether or not the gbest or pbest were updated---\gbest 
+and \pbest in equation \ref{eq:velupdate} got updated), and evaluate each position 
+in parallel.  That computes two iterations at the same time.
 
@@
 to do research in the area he was researching, so I joined his lab.  I too
-became interested in PSO and have worked in that area for the past year.
+became interested in PSO and have worked in that area with him for the past
+year.
 
 \section{Schedule}
+
+\begin{tabular}{|l|l|}
+\hline
+Run preliminary tests to ensure feasibility & February 28, 2009 \\
+\hline
+Finish writing and debugging code & March 30, 2009 \\
+\hline
+Finish gathering data and statistics & June 30, 2009 \\
+\hline
+Complete analysis of data and begin writing thesis & September 1, 2009 \\
+\hline
+Submit completed copy of thesis to advisor & November 15, 2009 \\
+\hline
+Submit paper detailing results to CEC 2010 & December 20, 2009 \\
+\hline
+Final thesis and portfolio submitted & February 1, 2010 \\
+\hline
+Thesis defense completed & March 1, 2010 \\
+\hline
+Final four thesis copies submitted & March 8, 2010 \\
+\hline
+\end{tabular}
 
@@
 
-\section{References}
+Thank you for reviewing my proposal.  I look forward to exploring speculative
+execution in particle swarm optimization.
 
-\section{Appendices}
+\bibliographystyle{plain}
+\bibliography{%
+\bib{pso/kennedy-icnn95},%
+\bib{pso/clerc-tec02},%
+\bib{pso/poli-aea08}}
 

### 2009-03-13 honors_thesis_proposal/paper.tex
Renamed honors_thesis to honors_thesis_proposal

--- 2009-03-09 honors_thesis/paper.tex
@@
 optimization problems, and optimization techniques can be used to find
-solutions.  Optimization algorithms are derived from all sorts of natural
+solutions.  Optimization algorithms have been derived from all sorts of natural
 phenomena.  Some methods look at the landscape of the function as they are
@@
 particle swarm optimization (PSO) algorithm in every aspect except the
-speed-up.  The same constants will be used as are commonly published in the
-literature.
+speed-up.  My work can then be accurately compared to other work in the field.
 
@@
 better performance of the algorithm as measured by the lowest function
-evaluation seen over a number of iterations for several benchmark and
-real-world problems.  Speculative execution takes more computation than regular
+evaluation seen over a number of function evaluations for several benchmark and
+real-world problems.  Speculative execution takes more computation than parallel
 PSO with the same number of particles, so the two algorithms will be compared
-at points where they use equal amounts of processing power.
+at points where they use equal amounts of function evaluations.
 
@@
 
-The finished thesis will be presented as a conference paper, including the
-following sections:
+The finished thesis will be presented in the format of a peer-reviewed
+conference paper, including the following sections:
 
@@
 conditional branch is encountered; they try to predict which branch will
-actually be needed, and continue their execution.  If it turns out that the
+actually be needed and continue their execution.  If it turns out that the
 prediction was incorrect, the work is discarded.  But if the prediction was
@@
 
-In particle swarm optimization, speculative execution can be performed because
-the next position a particle moves to doesn't depend on the value of the
-function at its current position.  That means you can just compute all possible
-next positions (depending on whether or not the gbest or pbest were updated---\gbest 
-and \pbest in equation \ref{eq:velupdate} got updated), and evaluate each position 
-in parallel.  That computes two iterations at the same time.
+In a lot of optimization algorithms, speculative execution is impossible,
+because the next iteration of the algorithm depends on the evaluation of the
+function during the previous iteration.  In simulated annealing, for instance,
+the probability of accepting a new position depends on the difference in
+function value between the new position and the current one.  In gradient
+descent, the gradient of the function must be determined (which includes
+actually evaluating it) to know how far to move before sampling the function
+again.  But particle swarm optimization is rather unique in that it doesn't
+require the evaluation of the function to know what the next position of each
+particle will be.  It simply depends on \emph{which} particle ended up having
+the best position and whether or not the current particle found a better spot
+than it had seen before.  That means you can just compute all possible next
+positions---all combinations of which particle is the new gbest and whether or
+not pbest was updated (\gbest and \pbest from \eqref{eq:velupdate})---and
+evaluate each position in parallel.  That computes two iterations at the same
+time.  It uses a lot of extra work to get the second iteration, but for some
+functions more iterations are worth more than extra particles.
 
@@
 for him using a parallel implementation of PSO.  Those results will be
-submitted to Nature Materials.  Recently we, along with a graduate student who
-works in the lab, submitted a paper to the Congress on Evolutionary
-Computation, a major venue for optimization techniques.  The paper was on
-topologies and communication in large particle swarms, and formed the
+submitted to Nature Materials.  Recently we, along with Andrew McNabb, a
+graduate student who works in the lab, submitted a paper to the Congress on
+Evolutionary Computation, a major venue for optimization techniques.  The paper
+was on topologies and communication in large particle swarms, and formed the
 beginnings of the work I propose.

### 2009-03-19 honors_thesis_proposal/paper.tex
Some updates to my thesis proposal

@@
 
-I will study the feasibility and potential performance gain of speculative
-execution in particle swarm optimization.  By speculative execution I mean the
-use of additional computational resources, presumably extra processors running
-in parallel, to perform two or more iterations of particle swarm optimization
-at a time, speeding up the algorithm significantly.
+I will study the feasibility and potential performance gain of a technique
+which we call speculative execution in particle swarm optimization.  By
+speculative execution I mean the use of additional computational resources,
+presumably extra processors running in parallel, to perform two or more
+iterations of particle swarm optimization at a time, speeding up the algorithm
+significantly.
 
@@
 evaluation seen over a number of function evaluations for several benchmark and
-real-world problems.  Speculative execution takes more computation than parallel
+real-world problems.  Speculative execution requires more function evaluations 
+per particle than a na\"ive parallel PSO, %TODO
 PSO with the same number of particles, so the two algorithms will be compared
-at points where they use equal amounts of function evaluations.
+at points where they use equal amounts of function evaluations per iteration.
 
@@
 distribution, the \( \otimes \) operator is an element-wise vector
-multiplication, $\pbest$ is the best position the current particle has seen,
-and $\gbest$ is the best position any of the other particles have seen.  \(
-\coeff_1 \), \( \coeff_2 \), and \( \constriction \) are parameters with
-prescribed values required to ensure convergence (2.05, 2.05, and .73,
+multiplication, $\pbest$ (called pbest) is the best position the current
+particle has seen, and $\gbest$ (called gbest) is the best position any of the
+other particles have seen.  \( \coeff_1 \), \( \coeff_2 \), and \(
+\constriction \) are parameters with prescribed values required to ensure
+convergence (2.05, 2.05, and .73,
 respectively).~\cite{clerc-tec02}~\cite{poli-aea08}
@@
 
-In a lot of optimization algorithms, speculative execution is impossible,
-because the next iteration of the algorithm depends on the evaluation of the
-function during the previous iteration.  In simulated annealing, for instance,
-the probability of accepting a new position depends on the difference in
-function value between the new position and the current one.  In gradient
-descent, the gradient of the function must be determined (which includes
-actually evaluating it) to know how far to move before sampling the function
-again.  But particle swarm optimization is rather unique in that it doesn't
-require the evaluation of the function to know what the next position of each
-particle will be.  It simply depends on \emph{which} particle ended up having
-the best position and whether or not the current particle found a better spot
-than it had seen before.  That means you can just compute all possible next
-positions---all combinations of which particle is the new gbest and whether or
-not pbest was updated (\gbest and \pbest from \eqref{eq:velupdate})---and
-evaluate each position in parallel.  That computes two iterations at the same
-time.  It uses a lot of extra work to get the second iteration, but for some
-functions more iterations are worth more than extra particles.
+In many optimization algorithms speculative execution is impossible, because
+the next iteration of the algorithm depends on the evaluation of the function
+during the previous iteration.  In simulated annealing, for instance, the
+probability of accepting a new position depends on the difference in function
+value between the new position and the current one.  In gradient descent, the
+gradient of the function must be determined (which includes actually evaluating
+it) to know how far to move before sampling the function again.  But particle
+swarm optimization is rather unique in that it does not require the evaluation
+of the function to know what the next position of each particle will be.  It
+simply depends on \emph{which} particle ended up having the best position and
+whether or not the current particle found a better spot than it had seen
+before.  That means you can just compute all possible next positions---all
+combinations of which particle is the new gbest and whether or not pbest was
+updated (\gbest\ and \pbest\ from \eqref{eq:velupdate})---and evaluate each
+position in parallel.  That computes two iterations at the same time.  It uses
+a lot of extra work to get the second iteration, but for some functions
+performing more iterations makes more progress than adding extra particles.
 
@@
 \hline
-Finish writing and debugging code & March 30, 2009 \\
+Finish writing and debugging code & April 30, 2009 \\
 \hline
@@
 
-This project will not require any outside funding.  The only expenditure
-will be my own time researching, conducting experiments, and writing.
+This project will not require any outside funding.  The only expenditure will
+be my own time researching, conducting experiments, and writing.  I have
+received a grant from the Office of Research and Creative Activities to perform
+this research, for which I am grateful.
 

### 2009-04-12 honors_thesis/proposal/paper.tex
Started working on a preliminary thesis for English 316.  Will be changed a lot for the actual thesis.

--- 2009-03-19 honors_thesis_proposal/paper.tex
@@
 evaluation seen over a number of function evaluations for several benchmark and
-real-world problems.  Speculative execution requires more function evaluations 
-per particle than a na\"ive parallel PSO, %TODO
-PSO with the same number of particles, so the two algorithms will be compared
-at points where they use equal amounts of function evaluations per iteration.
+real-world problems.  Performing speculative execution in PSO requires more
+function evaluations per particle than regular PSO, so to be fair the two
+algorithms will be compared with configurations where each uses the same number
+of functions evaluations per iteration, not the same number of particles.
 

### 2009-04-12 honors_thesis/thesis/paper.tex
Started working on a preliminary thesis for English 316.  Will be changed a lot for the actual thesis.

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{amsmath}

\DeclareMathOperator{\URand}{U}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\gbest}{\ensuremath{\Vec{g}}}
\providecommand{\pbest}{\ensuremath{\Vec{p}}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\coeff}{\ensuremath{\phi}}
\newcommand{\bib}[1]{../../../bib/#1}

\topmargin .5in

\title{Improving Particle Swarm Optimization \\ in a Parallel World}
\author{Matthew Gardner}
\date{\today}

\begin{document}
\maketitle

\section{Introduction}
\label{sec:intro}

Optimization problems are ubiquitous in our modern world.  Businesses such as
Google need to decide where to place ads, scientists need to fit models to
data, and airlines need to schedule their flights.  All of those problems are
optimization problems, and optimization techniques can be used to find
solutions.  Optimization algorithms have been derived from all sorts of natural
phenomena.  Some methods look at the landscape of the function as they are
searching and follow hills to the best values, as in the gradient ascent
method.  Other methods take ideas from biology or metallurgy, like genetic
algorithms and simulated annealing.  Particle swarm optimization is a recently
developed optimization technique that draws on ideas from the sociology of
flocking birds.

All of these methods, however, are fundamentally sequential in nature---they
need to be run in a specific order, and because of that, they are typically
only run on one machine.  The problems people want to solve are getting bigger,
and larger problems need to run on multiple machines in parallel.  Parallel
implementations of these algorithms have been attempted, but they do not often
perform as well as sequential versions for the amount of computation performed.
This work focuses on one of those algorithms, particle swarm optimization, and
improving its performance in a large-scale, parallel environment.  The work I
describe in Section~\ref{sec:topology} was done in collaboration with other
students I work with.  The work in Section~\ref{sec:specex} I did by myself.

The rest of this paper is outlined as follows.  In Section~\ref{sec:pso}, I
formally describe the particle swarm optimization algorithm, after giving a
brief description of optimization in general.  In Section~\ref{sec:topology},
I describe the sociological aspect of particle swarm optimization and
improvements that we have made for large swarms.  Section~\ref{sec:specex} then
introduces the idea of speculative execution in general and shows how it can be
done in particle swarm optimization.  Section~\ref{sec:results} shows results
of experiments run, and in Section~\ref{sec:conclusion} I draw some
conclusions.

\section{Particle Swarm Optimization}
\label{sec:pso}

\subsection{Optimization in General}

The goal of optimization is to find the parameters to a function that either
maximize or minimize the function value.  To take an example from above, Google
needs to find the placements of website advertisements that generate the most
clicks for those websites.  In this example, the placements of advertisements
are the parameters to the function, and the number of clicks generated is the
function to be optimized.  A more mathematical example can be seen in
Figure~\ref{fig:cosine}.  The function shown has a point $x$ for which $f(x)$
is the highest.  Finding the precise location of that point is not a trivial
task, especially for more complicated functions, and it is that problem which
optimization techniques try to solve.

\begin{figure}
  \centering
  \includegraphics{cosine.eps}
  \caption{A complicated function: $\cos(x)\sin(10x)e^{-|x|}$}
  \label{fig:cosine}
\end{figure}

\subsection{Particle Swarm Optimization Specifically}

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart.  It tries to intelligently search a multi-dimensional space by
mimicking the swarming and flocking behavior of birds and other
animals~\cite{kennedy-icnn95}. It is a sociological algorithm that depends on
interaction between particles to quickly and consistently find the optimal
solution to a problem.  The algorithm keeps track of a number of potential
solutions, called particles, which move somewhat randomly through the search
space at each iteration.  When the algorithm is done, the best position that
any of the particles ever found is returned.  Particles remember the best place
they have been, or solution they have evaluated, and when they move they are
attracted back to that place, as well as to the best solution other particles
have seen.  Specifically, the formulas for updating the position $\ppos_t$ and
velocity $\pvel_t$ of a particle at iteration $t$ are as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \left[ \pvel_t +
			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
			\coeff_2\URand()\otimes(\gbest - \ppos_t)
		\right] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \( \URand() \) is a vector of random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called pbest) is the best position the current
particle has seen, and $\gbest$ (called gbest) is the best position any of the
other particles have seen.  \( \coeff_1 \), \( \coeff_2 \), and \(
\constriction \) are parameters with prescribed values required to ensure
convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}~\cite{poli-aea08}.

The sociology in this algorithm defines which other particles form the
``neighborhood'' of a given particle---the other particles whose best solution
it sees (again, that is $\gbest$ in the equations above).  If particle 1 is a
neighbor of particle 2, then particle 1 will be influenced by good values that
particle 2 sees.  But if particle 3 is not a neighbor of particle 1, particle 1
will be completely unaware of good values that particle 3 finds.

There are many ways to define a particle's neighborhood, varying from the
entire rest of the swarm to just one other particle.

The way a neighborhood is defined can have drastic effects on the performance
of the algorithm; the more neighbors each particle has, the faster information
spreads throughout the swarm, and the quicker the algorithm converges.  For
some problems it is possible that the algorithm converges too quickly and gets
caught in a local optimum; it stops on a hill when there is a mountain next to
it.  For those kinds of problems, less communication, or smaller neighborhoods,
is often preferable.

\section{Topology in PSO}
\label{sec:topology}

Much work has been done on the sociological aspect of particle swarm
optimization, trying to determine the best neighborhood structure to use for
various classes of problems.  That work focused on swarms of tens of particles
to several hundred particles on a single machine.  Conclusions were reached
about which neighborhood structures worked best, but only in this limited
context~\cite{bratton-sis07}.  With the advent of parallel processing and the
potential to have swarms of thousands or hundreds of thousands of particles,
some of the earlier work needs rethinking.  This section focuses on
neighborhood structures in very large swarms, especially when the algorithm is
run in parallel on multiple machines.

\subsection{Defining Topology}

Many different neighborhood structures have been proposed and
tested~\cite{kennedy-cec02}.  In the literature, neighborhood structures are
referred to as topologies.  Two of them are more common---the complete or gbest
topology, where all particles communicate with all other particles, and the
ring or lbest topology, where each particle only communicates with its two
immediate neighbors~\cite{bratton-sis07}.

To formalize the description of topologies, we can consider the swarm as a
graph $\{V,E\}$, where the set of particles forms the nodes $V$ on the graph
and the (directed) edges $E$ between nodes represent communication between
particles.  If (1,2) is an edge in the graph, then particle 1 influences
particle 2's gbest, or particle 1 is in particle 2's neighborhood.  The
neighborhood $N_i$ of particle $i$ can be written formally as 
\[N_i = \{p_j|(j,i)\in E\}\]

Neighborhoods can also be thought of in terms of which other particles a
particular particle sends its information to, instead of which ones it gets
information from.  This conceptualization makes the communication a little bit
more explicit (and more directly relates to how to actually run the algorithm)
and will be used to describe topologies throughout the rest of this paper.
$N_i$ then becomes
\[N_i = \{p_j|(i,j)\in E\}\]

The two topologies mentioned previously, complete and ring, would then have the
following neighborhoods for each particle $i$ in a swarm of $n$ particles:
\[\mathrm{Complete:}\ N_i = \{1,2,3,\dots,n\}\]
\[\mathrm{Ring:}\ N_i = \{i-1,i,i+1\}\]

For the benefit of more visual readers, the two topologies are shown in graph
form in Figure~\ref{fig:topgraphs}; notice how many more lines are in the 
complete topology than in the ring topology.

\begin{figure}
  \centering
  \includegraphics{gbest_and_lbest.eps}
  \caption{The complete and ring topologies in graph form.  Figure taken from
  tracer.uc3m.es/tws/pso/neighborhood.html.}
  \label{fig:topgraphs}
\end{figure}

\subsection{Which Topology to Use}

There is a theorem about optimization called ``No Free Lunch''
\cite{wolpert-tec97}. In essence it says that no optimization algorithm can do
better than any other optimization algorithm across all functions.  If you do
better on some functions with a given algorithm that is because you do worse on
other functions.  This applies directly to the question of which topology in
PSO is the ``best,'' because it proves that there really is no ``best''
topology across all functions.  When selecting a topology you must have a
specific type of function in mind, and pick the topology that works best for
that type of function.

An interesting question along those lines is, ``Are the problems that most
people are interested in solving inherently similar?''  If that is the case,
one could design an algorithm that performs better on the problems people care
about, taking a performance hit only on functions that don't matter anyway.
But, because of No Free Lunch, any argument claiming that a particular
algorithm or topology is better than another must make that claim only for a
specific set of functions.

When deciding which topology to use, then, making a wise decision requires
having a certain kind of function in mind.

The complete topology is often preferable for many problems, because it allows
information to be shared quickly throughout the swarm and helps the algorithm
converge more quickly.  However, too much shared information can cause
premature convergence, as mentioned previously, so some researchers are wary of
the complete topology.  In the absence of knowledge about the function to be
optimized, researchers recommended, the ring topology should be used.  It might
take longer, but at least it won't get stuck, they say \cite{bratton-sis07}.

\begin{figure}
  \centering
  \includegraphics{iters_rast.eps}
  \caption{PSO performance on the function Rastrigin with varying numbers of
  particles.}
  \label{fig:rast}
\end{figure}

For a swarm of 50 particles it is accurate that complete often doesn't perform
well, and most researchers were satisfied to stop with a swarm that size.  We
have shown, however, that increasing the number of particles in a swarm will
help the algorithm avoid premature convergence, making complete the best
topology even for somewhat complicated functions~\cite{mcnabb-cec09}.
Figure~\ref{fig:rast} shows the performance of PSO on a commonly used benchmark
function, the Rastrigin function, with varying numbers of particles.  Notice
that with 50 particles the algorithm does get stuck pretty early, not finding
significantly better values after 250 iterations.  This is precisely what many
researchers in the field were saying.  With more particles, however, the
algorithm finds much better values and continues to find better values well
past 250 iterations.

\subsection{Communication Overhead}

In a parallel system, another problem with the complete topology is that
communication between processors is expensive relative to the amount of time it
takes to evaluate the functions many people work with.  A quick look at
Figure~\ref{fig:topgraphs} shows the incredibly large number of messages that
need to be sent between particles for a relatively small swarm.  Communicating
all of the information from every particle at every iteration can take orders
of magnitude more time than doing a simple evaluation and render
parallelization of the algorithm worthless.

We developed a modification to the algorithm along with a new topology that
approximates the complete topology with much less communication.  If each
particle at each iteration randomly chooses a small set of particles to send
its information to, eventually information will be communicated throughout the
whole swarm.  We call this the Random topology, and it is formally defined as
follows:
\[\mathrm{Random:}\ N_i = \{i,U_1,U_2,\dots,U_k\}\]
where $U_j$ is a random number from 1 to $n-1$, and $k$ is the number of 
neighbors to send information to at each iteration.

In the standard particle swarm optimization algorithm, the best neighbor that a
particle is attracted to (the \gbest in the formula above) is taken from its
current neighborhood.  With the random topology we proposed, particles can get
confused, being attracted to a different neighbor at every iteration, and
performance significantly decreases.  We modified the algorithm slightly such
that each particle remembers the best neighbor it has ever had and is attracted
to that position instead of its current best neighbor.  Using this modification
and a random topology, we saw performance that was indistinguishable from
complete with only 5\% of the communication, and the speed up is significant
even in a serial implementation~\cite{mcnabb-cec09}.

Figure~\ref{fig:rastrand} shows the performance of Random versus Complete on
the function Rastrigin for various swarm sizes.  The x-axis shows the number of
particles in the swarm, and the y-axis shows the best function value found
after 500 iterations.  It can be seen from the graph that Random does a very
good job at approximating the Complete topology while required only a small
fraction of the communication.

\begin{figure}
  \centering
  \includegraphics{randstar.eps}
  \caption{Random with various amounts of communication versus the Complete
  topology on the function Rastrigin.}
  \label{fig:rastrand}
\end{figure}

\subsection{Subswarms}

We have run all of the experiments I've talked about so far with swarms of up
to 4000 particles.  All of them were run with a serial implementation of
particle swarm optimization on a single machine, though the results can be used
in designing parallel algorithms.  With more than 4000 particles the algorithm
becomes very computationally expensive on a single machine.  We have
implemented a parallel version of particle swarm optimization using Google's
MapReduce framework for parallel computation~\cite{dean-osdi04}.  Soon we will
have it working well enough to run swarms of hundreds of thousands of particles
on hundreds of machines at a time, and we will further explore the issues
involved with parallel particle swarm optimization.

Many new possibilities for topologies arise when the algorithm is run in
parallel with large swarms.  We have proposed, but not yet fully tested, a
topology of subswarms, where each of a hundred machines would evaluate 500 or
1000 particles and send only its best value to all other machines, drastically
reducing the amount of inter-processor communication.  In our formal notation,
the subswarms topology can be described as

\[\mathrm{Subswarms:}\ N_i = \{j|j\%k = i\%k\}\]
where $k$ is the number of subswarms in the swarm, and \% is the mod operator.
For a small example, in a swarm of 12 particles with 3 subswarms, the sets of
neighbors would be \{1, 4, 7, 10\}, \{2, 5, 8, 11\}, and \{3, 6, 9, 12\}.

Our preliminary results show that this approach is very
promising~\cite{mcnabb-cec09}.  Figure~\ref{fig:subswarms} shows the results
of using the Subswarms topology with 400 particles in each swarm, versus just
using 400 particles in a Complete topology.  Subswarms clearly outperforms
Complete in this case.

\begin{figure}
  \centering
  \includegraphics{islands_iters_rast.eps}
  \caption{Subswarms versus Complete topologies on the function Rastrigin.}
  \label{fig:subswarms}
\end{figure}

\section{A Different Speed up---Speculative Execution}
\label{sec:specex}

\subsection{Speculative Exectuion in General}

Speculative execution in a program is the execution of code that may or may not
end up actually being needed.  Modern processors routinely do this when a
conditional branch is encountered---they try to predict which branch will
actually be needed and continue their execution.  If it turns out that the
prediction was incorrect, the work is discarded.  But if the prediction was
right, the program can execute much faster than if it had waited on the branch.  
\subsection{Speculative Execution in Particle Swarm Optimization}

In many optimization algorithms speculative execution is impossible, because
the next iteration of the algorithm depends on the evaluation of the function
during the previous iteration.  In simulated annealing, for instance, the
probability of accepting a new position depends on the difference in function
value between the new position and the current one.  In gradient descent, the
gradient of the function must be determined (which includes actually evaluating
it) to know how far to move before sampling the function again.

Particle swarm optimization is rather unique in that it does not require the
evaluation of the function to know what the next position of each particle will
be.  It simply depends on \emph{which} particle ended up having the best
position and whether or not the current particle found a better spot than it
had seen before.  That means you can just compute all possible next
positions---all combinations of which particle is the new gbest and whether or
not pbest was updated (\gbest\ and \pbest\ from equation
\eqref{eq:velupdate})---and evaluate each position in parallel.  That computes
two iterations at the same time.

\subsection{Is is worth it?}

It uses a lot of extra work to get the second iteration, but for some functions
performing more iterations makes more progress than adding extra particles.

\section{Conclusion}
\label{sec:conclusion}

Thank you for reviewing my proposal.  I look forward to exploring speculative
execution in particle swarm optimization.

\bibliographystyle{plain}
\bibliography{%
\bib{pso/kennedy-icnn95},%
\bib{pso/clerc-tec02},%
\bib{pso/poli-aea08},%
\bib{pso/topology/kennedy-cec02},%
\bib{pso/bratton-sis07},%
\bib{aml/mcnabb-cec09},%
\bib{mapreduce/dean-osdi04},%
\bib{wolpert-tec97}}

\end{document}

### 2009-04-12 honors_thesis/thesis/islands_iters_rast.py

@@ -1,0 +1,39 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 500
+DATADIR = '../../../psodata/data/'
+K400_FILE = DATADIR + 'star/400particles/rastrigin.out20batches'
+ISLANDS10_400_FILE = DATADIR + '10communicating_islands/4000particles/rastrigin.out20batches'
+
+plot = Plot()
+
+data = PSOData(open(K400_FILE))
+points = []
+for i in xrange(1, ITERATIONS + 1):
+    points.append((i, data.average(i)))
+plot.append(Points(points, title='Complete', style='lines'))
+
+data = PSOData(open(ISLANDS10_400_FILE))
+points = []
+for i in xrange(1, ITERATIONS + 1):
+    points.append((i, data.average(i)))
+plot.append(Points(points, title='10 Subswarms of 400 particles', style='lines'))
+
+plot.xmin = 100
+plot.xmax = ITERATIONS
+#plot.xlogscale = 10
+plot.ymin = 3
+plot.ymax = 50
+plot.ylogscale = 10
+#plot.xtics = dict((x, x) for x in SWARMSIZES if x != 400)
+plot.ytics = dict((x, x) for x in (5, 10, 20, 50, 100, 200))
+plot.xlabel = 'Iterations'
+plot.ylabel = 'Best Function Value'
+plot.show()
+plot.write('islands_iters_rast.gpi')
+
+# vim: et sw=4 sts=4

### 2009-04-12 honors_thesis/thesis/iters_rast.py

@@ -1,0 +1,37 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 500
+# How many high and low outliers to drop:
+TRIM = 2
+DATADIR = '/aml/home/mjg82/psodata/data/star/'
+SWARMSIZES = [50, 100, 500, 1000, 4000]
+
+pattern = DATADIR + '%sparticles/rastrigin.out20batches'
+
+plot = Plot()
+for swarmsize in SWARMSIZES:
+    points = []
+    data = PSOData(open(pattern % swarmsize))
+    for i in xrange(1, ITERATIONS + 1):
+        points.append((i, data.average(i)))
+    plot.append(Points(points, title='%s particles' % swarmsize, style='lines'))
+
+
+plot.xmin = 1
+plot.xmax = ITERATIONS
+#plot.xlogscale = 10
+plot.ymin = 3
+plot.ymax = 300
+plot.ylogscale = 10
+#plot.xtics = dict((x, x) for x in SWARMSIZES if x != 400)
+plot.ytics = dict((x, x) for x in (5, 10, 20, 50, 100, 200))
+plot.xlabel = 'Iterations'
+plot.ylabel = 'Best Function Value'
+plot.show()
+plot.write('iters_rast.gpi')
+
+# vim: et sw=4 sts=4

### 2009-04-12 honors_thesis/thesis/randstar.py

@@ -1,0 +1,55 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points
+
+ITERATIONS = 500
+# How many high and low outliers to drop:
+TRIM = 2
+DATADIR = '../../../psodata/data/'
+SWARMSIZES = [50, 100, 200, 300, 400, 500, 1000, 2000, 4000]
+#SWARMSIZES = [50, 100, 200, 300, 400, 500, 1000, 2000]
+
+star_pattern = DATADIR + 'star/%sparticles/rastrigin.out20batches'
+rand20_pattern = DATADIR + 'random20/%sparticles/rastrigin.out20batches'
+rand5_pattern = DATADIR + 'random5/%sparticles/rastrigin.out20batches'
+rand2_pattern = DATADIR + 'random2particles/%sparticles/rastrigin.out20batches'
+
+star_points = []
+star_bars = []
+rand20_points = []
+rand5_points = []
+rand2_points = []
+for swarmsize in SWARMSIZES:
+    star_data = PSOData(open(star_pattern % swarmsize))
+    star_points.append((swarmsize, star_data.average(ITERATIONS)))
+    low, med, high = star_data.statistics(ITERATIONS, TRIM)
+    print 'star:', swarmsize, low, med, high
+    star_bars.append((swarmsize, med, low, high))
+
+    rand20_data = PSOData(open(rand20_pattern % swarmsize))
+    rand20_points.append((swarmsize, rand20_data.average(ITERATIONS)))
+
+    rand5_data = PSOData(open(rand5_pattern % swarmsize))
+    rand5_points.append((swarmsize, rand5_data.average(ITERATIONS)))
+
+    rand2_data = PSOData(open(rand2_pattern % swarmsize))
+    rand2_points.append((swarmsize, rand2_data.average(ITERATIONS)))
+
+plot = Plot(xmin=45, xmax=4400, xlogscale=2, ymin=4, ymax=50, ylogscale=10)
+plot.xtics = dict((x, x) for x in SWARMSIZES if x != 400)
+plot.ytics = dict((x, x) for x in (5, 10, 20, 50))
+plot.xlabel = 'Swarm Size'
+plot.ylabel = 'Value after %s iterations' % ITERATIONS
+plot.append(Points(star_points, style='lines', title='Complete'))
+plot.append(Points(rand20_points, style='lines', 
+    title='Random, 20% communication'))
[…]

### 2009-04-12 honors_thesis/thesis/paper.tex
A draft of my paper for English 316 is done.

@@
 introduces the idea of speculative execution in general and shows how it can be
-done in particle swarm optimization.  Section~\ref{sec:results} shows results
-of experiments run, and in Section~\ref{sec:conclusion} I draw some
-conclusions.
+done in particle swarm optimization.  I then conclude in Section~\ref{sec:conclusion}.
 
@@
 conditional branch is encountered---they try to predict which branch will
-actually be needed and continue their execution.  If it turns out that the
+end up being needed and continue their execution.  If it turns out that the
 prediction was incorrect, the work is discarded.  But if the prediction was
 right, the program can execute much faster than if it had waited on the branch.
+
 \subsection{Speculative Execution in Particle Swarm Optimization}
@@
 the next iteration of the algorithm depends on the evaluation of the function
-during the previous iteration.  In simulated annealing, for instance, the
-probability of accepting a new position depends on the difference in function
-value between the new position and the current one.  In gradient descent, the
-gradient of the function must be determined (which includes actually evaluating
-it) to know how far to move before sampling the function again.
+during the previous iteration.  To know which positions to speculatively
+evaluate you have to already have evaluated the current position, and then it's
+not speculative anymore.  In simulated annealing, for instance, the probability
+of accepting a new position depends on the difference in function value between
+the new position and the current one.  In gradient descent, the gradient of the
+function must be determined (which includes actually evaluating it) to know how
+far to move before sampling the function again.  It may be possible to modify
+these algorithms to allow for speculative execution, but their standard
+executions cannot be done speculatively.
 
@@
 
-It uses a lot of extra work to get the second iteration, but for some functions
-performing more iterations makes more progress than adding extra particles.
+Speculative execution uses a lot of extra work to compute two iterations at the
+same time, but for some functions performing more iterations makes more
+progress than adding extra particles.  The No Free Lunch theorem again shows
+its head here, though, because if you do better with one algorithm on a certain
+function or set of functions, that's because you do worse on other functions.
+There are, however, a set of functions that do better with more iterations than
+with more particles.
+
+Figure~\ref{fig:sphere} shows an example of this: the function Sphere.  The
+graph shows the best function value found over a number of function
+evaluations.  This is different than previous graphs that have shown function
+value against a number of iterations.  At each iteration, a swarm of 50
+particles performs 50 function evaluations, where a swarm of 4000 particles
+performs 4000 evaluations.  Figure~\ref{fig:sphere} shows that the extra 3950
+function evaluations spent on the extra particles in a 4000 particle swarm
+really don't do any good.  It would be more worth it to just do more iterations
+of a 50 particle swarm than to have a swarm of 4000 particles.
+
+The benefit of adding more particles comes in when you have hundreds or
+thousands of processors in a parallel cluster.  You can't use 4000 machines to
+run a swarm of 50 particles---3950 of the processors would be idle the whole
+time.  Adding more particles to the swarm then improves performance without
+adding any extra time (ignoring inter-processor communication cost for the
+moment).  But for some functions, a better idea than using the extra 3950
+processors to add more particles to the swarm is to use the processors
+speculatively so that two iterations can be done at the same time.
+Figure~\ref{fig:sphere} shows just one example of a function that would be
+better optimized by performing speculative execution than by adding extra
+particles.
+
+\begin{figure}
+  \centering
+  \includegraphics{evals_sphere.eps}
+  \caption{Best value found on the function Sphere versus the number of 
+  function evaluations with varying swarm sizes.}
+  \label{fig:sphere}
+\end{figure}
 
@@
 
-Thank you for reviewing my proposal.  I look forward to exploring speculative
-execution in particle swarm optimization.
+In a world where people are swimming in processing power, many old algorithms
+need to be rethought.  Parallel computing architectures allow for new
+techniques that were previously impossible.  In the realm of optimization,
+particle swarm optimization is uniquely capable of reaping great benefits from
+parallel processing.  I have reviewed just a few ways in which PSO can be
+improved to harness the increased power of parallel clusters.  As more
+researchers tackle the problems of modifying and adapting algorithms to better
+utilize the resources that are now available, our ability to solve problems
+with computers will increase dramatically.
 

### 2009-10-02 honors_thesis/thesis/paper.tex
.
First draft of my honors thesis is complete

--- 2009-04-12 honors_thesis/thesis/paper.tex
@@
-\documentclass[onecolumn, 12pt]{article}
-\usepackage{fullpage}
+\documentclass[conference,letterpaper]{IEEEtran}
+
+\usepackage{cite}
 \usepackage{graphicx}
-\usepackage{cite}
-\usepackage{amsmath}
+\usepackage{amsmath,amssymb}
+\usepackage{algorithm}
+\usepackage{algorithmic}
 
+\IEEEoverridecommandlockouts
+
+% AML bibliography references
+\newcommand{\bib}[1]{/aml/home/mjg82/bib/#1}
+
+% Better references, I think
+\renewcommand{\sec}[1]{Section~\ref{sec:#1}}
+\newcommand{\fig}[1]{Figure~\ref{fig:#1}}
+\newcommand{\alg}[1]{Algorithm~\ref{alg:#1}}
+
+% Algorithmic changes
+\renewcommand{\algorithmicforall}{\textbf{for each}}
+
+%% PSO Stuff
 \DeclareMathOperator{\URand}{U}
+\DeclareMathOperator*{\argmin}{arg\;min}
+\DeclareMathOperator*{\argmax}{arg\;max}
+\DeclareMathOperator*{\arginf}{arg\;inf}
+\DeclareMathOperator*{\argsup}{arg\;sup}
 \providecommand{\ppos}{\ensuremath{\Vec{x}}}
 \providecommand{\pvel}{\ensuremath{\Vec{v}}}
-\providecommand{\gbest}{\ensuremath{\Vec{g}}}
-\providecommand{\pbest}{\ensuremath{\Vec{p}}}
+\providecommand{\nbest}{\ensuremath{\Vec{b}_N}}
+\providecommand{\pbest}{\ensuremath{\Vec{b}_P}}
 \providecommand{\constriction}{\ensuremath{\chi}}
 \providecommand{\coeff}{\ensuremath{\phi}}
-\newcommand{\bib}[1]{../../../bib/#1}
+\providecommand{\obs}{\ensuremath{\Vec{\xi}}}
 
-\topmargin .5in
+%SpecExPSO Stuff
+\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
+\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
+\providecommand{\p}{\ensuremath{p}}
+\providecommand{\pset}{\ensuremath{\mathbf{p}}}
+\providecommand{\s}{\ensuremath{s}}
+\providecommand{\sset}{\ensuremath{\mathbf{s}}}
+\providecommand{\n}{\ensuremath{n}}
+\providecommand{\nset}{\ensuremath{\mathbf{n}}}
 
-\title{Improving Particle Swarm Optimization \\ in a Parallel World}
-\author{Matthew Gardner}
-\date{\today}
+\title{\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization%
+\thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department
+of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602
+(phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}%
+}
+
+\date{}
+
+\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
 
@@
 
+\begin{abstract}
+
+Particle swarm optimization (PSO) has previously been parallelized only by 
+adding more particles to the swarm or by parallelizing the evaluation of the
+objective function.  However, some functions are better optimized with more
+iterations and fewer particles.  Accordingly, we take inspiration from 
+speculative execution commonly performed in modern processors and propose what
+we call speculative evaluation in PSO.  Future positions of the particles are
+speculated and evaluated in parallel with current positions, performing two
+iterations of PSO at once.
+
+We also propose another way of making use of these speculative particles,
+keeping the best position found instead of the position that PSO actually would
+have taken.  We show results that for a number of functions, speculative
+evaluation can have dramatic improvements over adding additional particles to
+the swarm.
+
+\end{abstract}
+
 \section{Introduction}
@@
 
-Optimization problems are ubiquitous in our modern world.  Businesses such as
-Google need to decide where to place ads, scientists need to fit models to
-data, and airlines need to schedule their flights.  All of those problems are
-optimization problems, and optimization techniques can be used to find
-solutions.  Optimization algorithms have been derived from all sorts of natural
-phenomena.  Some methods look at the landscape of the function as they are
-searching and follow hills to the best values, as in the gradient ascent
-method.  Other methods take ideas from biology or metallurgy, like genetic
-algorithms and simulated annealing.  Particle swarm optimization is a recently
-developed optimization technique that draws on ideas from the sociology of
-flocking birds.
+Particle swarm optimization (PSO) has been found to be a highly robust and
+effective algorithm for solving many types of optimization problems.  For much
+of the algorithm's history, PSO was run serially on a single machine.  However,
+the world's computing power is increasingly coming from large clusters of
+processors.  Even in desktop machines, systems with 4 or 8 cores are common
+place.  In order to efficiently utilize these resources for computationally
+intensive problems, algorithms need to run in parallel.
 
-All of these methods, however, are fundamentally sequential in nature---they
-need to be run in a specific order, and because of that, they are typically
-only run on one machine.  The problems people want to solve are getting bigger,
-and larger problems need to run on multiple machines in parallel.  Parallel
-implementations of these algorithms have been attempted, but they do not often
-perform as well as sequential versions for the amount of computation performed.
-This work focuses on one of those algorithms, particle swarm optimization, and
-improving its performance in a large-scale, parallel environment.  The work I
-describe in Section~\ref{sec:topology} was done in collaboration with other
-students I work with.  The work in Section~\ref{sec:specex} I did by myself.
+Within the last few years, researchers have begun to recognize the need to
+develop parallel implementations of PSO, publishing many papers on the subject.
+The methods they have used include various synchronous algorithms
+\cite{belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
+asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05},
+and formulating PSO in Google's MapReduce framework \cite{mcnabb-cec07}.
+Parallelizing the evaluation of the objective function can also be done in some
+cases, though that is not an adaption of the PSO algorithm and thus is not the
+focus of this paper.
 
-The rest of this paper is outlined as follows.  In Section~\ref{sec:pso}, I
-formally describe the particle swarm optimization algorithm, after giving a
-brief description of optimization in general.  In Section~\ref{sec:topology},
-I describe the sociological aspect of particle swarm optimization and
-improvements that we have made for large swarms.  Section~\ref{sec:specex} then
-introduces the idea of speculative execution in general and shows how it can be
-done in particle swarm optimization.  Section~\ref{sec:results} shows results
-of experiments run, and in Section~\ref{sec:conclusion} I draw some
-conclusions.
+All of these parallel techniques use additional processors to increase the
+number of particles in the swarm, either by adding individual particles or by
+adding entire new subswarms.  The number of iterations of PSO that the
+algorithms can perform is thus inherently limited by the time it takes to
+evaluate the objective function--additional processors add more particles, not
+more iterations.  However, there are many functions for which performing more
+iterations with fewer particles is a better use of resources than performing
+fewer iterations with more particles.
+
+To solve this problem, we take inspiration from speculative execution
+techniques commonly used in processors.  Modern processors, when faced with a
+branch on which they must wait, guess which way the branch will go and start
+executing.  If the processor guessed right, execution is much farther ahead
+than if it had just idly waited on the branch.  If it guessed wrong, execution
+is right where it would have been anyway.  This is called speculative
+execution or branch prediction.
+
+The idea of speculative execution can be applied to some types of optimization
+algorithms.  When the next sampling position of the algorithm does not depend
+explicitly on the value of the objective function at the current position,
+future positions can be speculated and evaluated in parallel with the current
+position.  This could not be done in gradient methods, for instance, where the
+step size for sampling depends on the gradient at the current position, but it
+has been done in simulated annealing~\cite{witte-tpds91}.
+
+Because the motion of particles in PSO depends only on the \emph{positions} of
+the particle and its neighbors, not the \emph{value} of the objective function
+at those positions, speculative evaluation is possible.  The relative values of
+the positions of the particle's neighbors do matter, in that the best one is
+used as the neighborhood best position, but it's which position has the highest
+value, not what that value is, that matters.  We then propose a new method of
+parallelizing PSO in which additional processors are used to perform more
+iterations, not add particles to the swarm.
+
+Because we do not have to wait for a particle's position to be evaluated to know
+where it might move to next, we can compute the set of all possible next
+positions and send them to be evaluated along with the particle's current
+position.  Once the current position of each particle has been evaluated, we
+can determine which of the speculatively evaluated positions was correct.  We
+have thus evaluated two iterations at once.  This can easily get unwieldy if
+the set of next possible positions for each particle is large, but a wise
+choice of topology limits that set to a reasonable size.
+
+The original PSO can be reproduced exactly, two iterations at a time, using
+this technique.  However, in strictly using speculative evaluation to reproduce
+the original PSO, a lot of positions are speculatively evaluated whose values
+get ignored after deciding which branch was the correct one.  A better use of
+those speculative evaluations may be to incorporate the information into the
+swarm, no longer exactly reproducing PSO but perhaps producing better results.
+
+\sec{pso} describes the particle swarm optimization algorithm.  \sec{sepso}
+then describes how speculative evaluation can be done in parallel PSO to
+perform two iterations at once.  In \sec{extra}, we discuss a variant of the
+speculative algorithm which uses all of the speculative evaluations performed
+instead of throwing away all but one for each particle.  We then show in
+\sec{results} a comparison between the original PSO algorithm, our speculative
+algorithm, and its variant.
 
@@
 
-\subsection{Optimization in General}
+Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
+Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
+a multi-dimensional space by mimicking the swarming and flocking behavior of
+birds and other animals. It is a social algorithm that depends on interaction
+between particles to quickly and consistently find the optimal solution to some
+objective function.
 
-The goal of optimization is to find the parameters to a function that either
-maximize or minimize the function value.  To take an example from above, Google
-needs to find the placements of website advertisements that generate the most
-clicks for those websites.  In this example, the placements of advertisements
-are the parameters to the function, and the number of clicks generated is the
-function to be optimized.  A more mathematical example can be seen in
-Figure~\ref{fig:cosine}.  The function shown has a point $x$ for which $f(x)$
-is the highest.  Finding the precise location of that point is not a trivial
-task, especially for more complicated functions, and it is that problem which
-optimization techniques try to solve.
+The motion of particles through the search space has three components: an
+inertial component that gives particles momentum as they move, a cognitive
+component where particles remember the best solution they have found and are
+attracted back to that place, and a social component by which particles are
+attracted to the best solution that any of their neighbors have found.
 
-\begin{figure}
-  \centering
-  \includegraphics{cosine.eps}
-  \caption{A complicated function: $\cos(x)\sin(10x)e^{-|x|}$}
-  \label{fig:cosine}
-\end{figure}
-
-\subsection{Particle Swarm Optimization Specifically}
-
-Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
-Eberhart.  It tries to intelligently search a multi-dimensional space by
-mimicking the swarming and flocking behavior of birds and other
-animals~\cite{kennedy-icnn95}. It is a sociological algorithm that depends on
-interaction between particles to quickly and consistently find the optimal
-solution to a problem.  The algorithm keeps track of a number of potential
-solutions, called particles, which move somewhat randomly through the search
-space at each iteration.  When the algorithm is done, the best position that
-any of the particles ever found is returned.  Particles remember the best place
-they have been, or solution they have evaluated, and when they move they are
-attracted back to that place, as well as to the best solution other particles
-have seen.  Specifically, the formulas for updating the position $\ppos_t$ and
-velocity $\pvel_t$ of a particle at iteration $t$ are as follows:
+At each iteration of the algorithm, the position $\ppos_t$ and velocity
+$\pvel_t$ of each particle are updated as follows:
 \begin{align}
@@
 			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
-			\coeff_2\URand()\otimes(\gbest - \ppos_t)
+			\coeff_2\URand()\otimes(\nbest - \ppos_t)
 		\right] \\
@@
 distribution, the \( \otimes \) operator is an element-wise vector
-multiplication, $\pbest$ (called pbest) is the best position the current
-particle has seen, and $\gbest$ (called gbest) is the best position any of the
-other particles have seen.  \( \coeff_1 \), \( \coeff_2 \), and \(
-\constriction \) are parameters with prescribed values required to ensure
-convergence (2.05, 2.05, and .73,
-respectively)~\cite{clerc-tec02}~\cite{poli-aea08}.
+multiplication, $\pbest$ (called personal best) is the best position the
+current particle has seen, and $\nbest$ (called neighborhood best) is the best
+position the neighbors of the current particle have seen.  \( \coeff_1 \), \(
+\coeff_2 \), and \( \constriction \) are parameters with prescribed values
+required to ensure convergence (2.05, 2.05, and .73,
+respectively)~\cite{clerc-tec02}.
 
-The sociology in this algorithm defines which other particles form the
-``neighborhood'' of a given particle---the other particles whose best solution
-it sees (again, that is $\gbest$ in the equations above).  If particle 1 is a
-neighbor of particle 2, then particle 1 will be influenced by good values that
-particle 2 sees.  But if particle 3 is not a neighbor of particle 1, particle 1
-will be completely unaware of good values that particle 3 finds.
+There are many ways of defining the neighbors of a particle.  Different
+neighborhood topologies have varying performance on different kinds of
+functions, as would be expected by the No Free Lunch
+theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
+beyond the scope of this paper.  We use the Ring topology, where each particle
+is neighbors only with two other particles, and a parallel adaptation of the
+Complete topology~\cite{mcnabb-cec09}, where each particle is neighbors with
+all other particles.  These are the two most commonly used topologies in the
+literature and have been shown to work well in practice~\cite{bratton-sis07}.
 
-There are many ways to define a particle's neighborhood, varying from the
-entire rest of the swarm to just one other particle.
+\section{Speculative Evaluation in PSO}
+\label{sec:sepso}
+\subsection{Overview}
 
-The way a neighborhood is defined can have drastic effects on the performance
-of the algorithm; the more neighbors each particle has, the faster information
-spreads throughout the swarm, and the quicker the algorithm converges.  For
-some problems it is possible that the algorithm converges too quickly and gets
-caught in a local optimum; it stops on a hill when there is a mountain next to
-it.  For those kinds of problems, less communication, or smaller neighborhoods,
-is often preferable.
+The PSO algorithm just described can be parallelized by having each additional
+processor add a particle or a set of particles to the swarm.  But for some
+functions, a more desirable use of resources is to speed up the algorithm,
+performing more iterations instead of adding more particles.  Barring
+parallelizing the evaluation of the objective function itself, which may not be
+possible, previous techniques have not addressed this issue.
 
-\section{Topology in PSO}
-\label{sec:topology}
+In order to perform additional iterations by adding additional processors, we
+take inspiration from speculative execution in processors.  If we can guess
+what a particle's next position will be from its current position, we can
+evaluate both in parallel and perform two iterations of PSO at once.  Lacking
+a good way to guess a particle's next position, we can at least enumerate all
+possible next positions of the particle, evaluate all of them at once, and one
+of them will be right.
 
-Much work has been done on the sociological aspect of particle swarm
-optimization, trying to determine the best neighborhood structure to use for
-various classes of problems.  That work focused on swarms of tens of particles
-to several hundred particles on a single machine.  Conclusions were reached
-about which neighborhood structures worked best, but only in this limited
-context~\cite{bratton-sis07}.  With the advent of parallel processing and the
-potential to have swarms of thousands or hundreds of thousands of particles,
-some of the earlier work needs rethinking.  This section focuses on
-neighborhood structures in very large swarms, especially when the algorithm is
-run in parallel on multiple machines.
+While in some algorithms enumerating all possible next sampling positions would
+be impossible or extremely difficult, in PSO all we need to know are the
+current positions of the particle and its neighbors, from which we can easily
+enumerate and evaluate all possible positions of that particle at the next
+iteration.  The defining components of a particle's motion
+(see~\eqref{eq:velupdate}) are $\pvel_t$, $\ppos_t$, $\pbest$, and $\nbest$.
+For a particle at iteration $t-1$, $\pvel_t$ and $\ppos_t$ are defined by
+Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  The particle already
+knows its personal best $\pbest$ at iteration $t-1$, but it could be updated to
+$\ppos_t$ if its position at iteration $t$ has a better value than its current
+personal best.  Likewise, the particle knows its neighborhood best $\nbest$ at
+iteration $t-1$, though it could be updated by any of its neighbors' positions
+at iteration $t$.
 
-\subsection{Defining Topology}
+To enumerate all of the possible next positions of a particle, one must simply
+evaluate the motion equations for all possible combinations of personal best
+and neighborhood best updates that a particle could see.  This produces
+$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
+neighbors the particle has (a particle either updates its personal best or it
+does not, and one of its $n$ neighbors must update its neighborhood best, or
+there is no neighborhood best update--$n+1$ possibilities).  If all of those
+speculative positions are evaluated in parallel along with the original,
+non-speculative position of the particle at time $t-1$, then one of the
+speculative positions has to be the correct one, and two iterations have been
+evaluated at once.
 
-Many different neighborhood structures have been proposed and
-tested~\cite{kennedy-cec02}.  In the literature, neighborhood structures are
-referred to as topologies.  Two of them are more common---the complete or gbest
-topology, where all particles communicate with all other particles, and the
-ring or lbest topology, where each particle only communicates with its two
-immediate neighbors~\cite{bratton-sis07}.
+Because $2(n+1)$ speculative evaluations must be performed for each particle,
+the choice of topology is an important one.  The use of the Complete topology,
+where every particle is a neighbor to every other particle, would require on
+the order of $p^2$ speculative evaluations per iteration, where $p$ is the
+swarm size.  Clearly it is much more desirable to have a sparse topology, where
+$O(np)$ is much smaller than $O(p^2)$.  However, some functions are better
+optimized with the Complete topology and the quick spread of information it
+entails than with sparse topologies.  Accordingly, we use the Random topology
+described in~\cite{mcnabb-cec09}, which has been shown to effectively simulate
+the Complete topology.  In that topology, 2 neighbors are chosen at random from
+the swarm for each particle at each iteration.  Because a particle is also
+always its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report
+the results for using both the Ring topology (where $n$ again is $3$) and the
+Random topology on a number of common benchmark functions.
 
-To formalize the description of topologies, we can consider the swarm as a
-graph $\{V,E\}$, where the set of particles forms the nodes $V$ on the graph
-and the (directed) edges $E$ between nodes represent communication between
-particles.  If (1,2) is an edge in the graph, then particle 1 influences
-particle 2's gbest, or particle 1 is in particle 2's neighborhood.  The
-neighborhood $N_i$ of particle $i$ can be written formally as 
-\[N_i = \{p_j|(j,i)\in E\}\]
+It is not trivial in some parallel architectures to determine which speculative
+position was the correct next position of each particle.  First we discuss the
+relatively easy case of a centralized parallel PSO algorithm.  In such an
+architecture, the master keeps track of all necessary information with only
+trivial message passing needed.  Then we discuss the more complicated case of a
+distributed algorithm, where each particle is on its own and needs to send and
+receive messages to and from other particles.  Finally we discuss the further
+complications of a dynamic topology, where a particle's neighbors change from
+one iteration to another.
 
-Neighborhoods can also be thought of in terms of which other particles a
-particular particle sends its information to, instead of which ones it gets
-information from.  This conceptualization makes the communication a little bit
-more explicit (and more directly relates to how to actually run the algorithm)
-and will be used to describe topologies throughout the rest of this paper.
-$N_i$ then becomes
-\[N_i = \{p_j|(i,j)\in E\}\]
+\subsection{Terminology}
 
-The two topologies mentioned previously, complete and ring, would then have the
-following neighborhoods for each particle $i$ in a swarm of $n$ particles:
-\[\mathrm{Complete:}\ N_i = \{1,2,3,\dots,n\}\]
-\[\mathrm{Ring:}\ N_i = \{i-1,i,i+1\}\]
+To aid in describing our methodology, we introduce a few terms.  A particle's
+set of \emph{speculative children} is the set of all possible next iteration
+states that a particle could have.  We use $\p_t$ to denote a particle at
+iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
+$\sset_{t+1}$.  $\n_t$ is a neighbor of particle $\p_t$.  A set of particles is
+$\pset$, whereas a single particle is simply $\p$.
 
-For the benefit of more visual readers, the two topologies are shown in graph
-form in Figure~\ref{fig:topgraphs}; notice how many more lines are in the 
-complete topology than in the ring topology.
+We separate each iteration of PSO into several steps.  First there is the
+motion step, where a particle updates its position and velocity.  Then a
+particle's position is evaluated, and the particle updates its current value
+and its personal best.  Finally, a particle gets information from its neighbors
+and updates its neighborhood best.
+
+A particle at iteration $t-1$ that has been moved to iteration $t$, but whose
+position has not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its
+position has been evaluated, but it has still not yet received information from
+its neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
+updated its neighborhood best is it a complete particle at iteration $t$.  It is
+then simply denoted as $\p_t$.
+
+\subsection{Centralized Algorithms}
+
+In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
+master, keeps track of all necessary information, and all other machines are
+merely used to evaluate the objective function at various positions as directed
+by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
+a scheme, the master generates the positions to evaluate speculatively and
+decides which one to accept for each particle.
+
+Given a set of particles at iteration $t-1$ (perhaps which have just been
+initialized), the master must move each particle using
+Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
+$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
+set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
+position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
+and $\nbest$ that particle $\noeval{\p}_t$ could end up with.  Each of those
+combinations defines a speculative particle, $\s_t$, that can be moved using
+the motion equations to obtain $\noeval{\s}_{t+1}$.  The result is a set of
+particles $\noeval{\pset}_t$, and for each particle a set of speculative
+children $\noeval{\sset}_{t+1}$, which can all be evaluated at once.
+
+The master then has the slaves evaluate the particles.  Once all particles,
+speculative and original, have been evaluated and the values reported to the
+master, the master sorts out which speculative child of each particle was the
+correct one.
+
+This is done by first updating each particle $\noeval{\p}_t$ to
+$\nonbest{\p}_t$ by associating the value determined by the slave to the
+particle it came from, and updating that particle's personal best if the value
+reported is better than its previous personal best value.  Then each
+$\nonbest{\p}_t$ must be updated to $\p_t$ by updating its neighborhood best
+position and value with information about each of its neighbors.  This is
+simply the original PSO algorithm.
+
+Then, with the set $\pset_t$, the branch that was actually taken by each
+particle can be determined, and the correct speculative child can be chosen.
+The branches depended entirely on whether or not the personal best of $p_t$ was
+updated and which particle, if any, updated the particle's neighborhood best.
+Both of those items are now known, and the child that matches the branch is
+selected.
+
+The parent $\p_t$ must pass its personal best value to the child, as the child
+knows only the position that it guessed, not the function value at that
+position.  It is possible that both $\p_t$ and $\s_{t+1}$ updated their
+personal bests, but $\p_t$'s value was better.  For example, suppose that
+$\p_{t-1}$ has a personal best value of 3.  $\noeval{\p}_t$ was created, and
+$\noeval{\s}_{t+1}$ was moved assuming that $\p_t$ had updated its personal
+best with its position at time $t$.  Then both $\noeval{\p}_t$ and
+$\noeval{\s}_{t+1}$ were evaluated, with values 1 and 2, respectively.
+$\nonbest{\s}_{t+1}$ would think that its current position was its personal
+best, as the value it found--2--was better than its previous personal best
+value of 3.  It needs to receive the personal best value from its parent to
+know that its personal best position $\pbest$ is actually the position of
+$\p_t$, not $\s_{t+1}$.
+
+The parent also needs to pass the value of the neighborhood best that the child
+guessed.  The child only knows the position and needs the value in order to
+make future comparisons between neighborhood best positions.
+
+Upon picking the correct branch for each particle and updating the children's
+personal best and neighborhood best value (from iteration $t$), the result is
+the set $\nonbest{\pset}_{t+1}$, as the particles are now no longer
+speculative.  What remains is to update the neighborhood best of those
+particles from their neighbors (from iteration $t+1$), as above, to obtain
+$\pset_{t+1}$.  That set of particles can subsequently be used to produce the
+sets $\noeval{\pset}_{t+2}$ and $\noeval{\sset}_{t+3}$, and the process repeats
+itself.
+
+The algorithm is outlined in \alg{centralized}.
+
+\begin{algorithm}
+  \caption{Speculative Evaluation in a Centralized PSO}
+  \label{alg:centralized}
+  \begin{algorithmic}[1]
+	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$
+	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
+	  generate $\noeval{\sset}_{t+1}$
+	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
+	\STATE Update personal best for each $\noeval{\p}_t$ and
+	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
+	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
+	  $\pset_t$
+	\FORALL{$\p_t$}
+	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
+	  the branch taken by $\p_t$
+	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
+	  making $\nonbest{\p}_{t+1}$
+	\ENDFOR
+	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
+	  $\pset_{t+1}$
+	\STATE Repeat
+  \end{algorithmic}
+\end{algorithm}
+
+\subsection{Distributed Algorithms}
+
+In a distributed parallel PSO algorithm, individual processors not only perform
+evaluations of particles, but also their movement.  The information for each
+particle is not held by a central machine that directs the algorithm; instead,
+each processor has the information for the particle or particles that it is in
+charge of and must perform the steps of the algorithm for the particles it
+manages.  Messages such as values and positions for the neighborhood best are
+sent between processors.  There may be (in fact there probably must be) some
+machine that collects information from all of the particles and outputs the
+result of the algorithm, though that machine's importance is much less than in
+centralized algorithms.
+
+To perform speculative evaluation in a distributed PSO algorithm, there must be
+some way to have processors evaluate the speculative children of particles,
+without giving the speculative particles the same treatment as actual
+particles.  Perhaps that can be done by assigning each particle a set of
+machines instead of a single machine, and the particle directs its extra
+machines to evaluate its speculative children.  In the MapReduce framework,
+each particle and speculative child has its own Map task, during which the
+particle is evaluated, with the speculative children sending their output back
+to the original particle's Reduce task and any neighbors it has.  The Reduce
+task for each particle gets all of the messages from its neighbors, its
+speculative children, and the speculative neighbors of its speculative
+children, and produces the next particle and children for another Map task.
+
+No matter which framework is used the message passing should be similar, and
+that is what we describe here.
+
+A processor that is controlling a single particle $\p_{t-1}$ must first move
+the particle to $\noeval{\p}_t$ and produce the particle's speculative children
+$\noeval{\sset}_{t+1}$.  This is done in the same way as described above.
+However, in order to produce $\noeval{\sset}_{t+1}$, the processor needs
+information about the particle's neighbors, so there must be some message
+passing to get that information.  Particularly, the information that the
+processor needs is the position of each of the particle's neighbors at
+iteration $t$.
+
+This can be accomplished in two ways, either by performing several rounds of
+very simple message passing, or by performing a single round with a lot more
+messages.
+
+The first method takes three rounds of passing information, two for iteration
+$t$ and one for iteration $t+1$.  First, each particle must be moved from
+iteration $t-1$ to $\noeval{\p}_t$.  Then a round of message passing is
+required so that each particle knows the position of its neighbors at iteration
+$t$, from which $\noeval{\sset}_{t+1}$ can be generated.  After each particle
+evaluates its position and the positions of its speculative children, it passes
+information about the outcome iteration $t$ to its neighbors, so that each
+particle can receive information about its neighborhood best to be updated to
+$\p_t$.  Once that communication is finished, the particle can select the
+speculative child which matched the branch that iteration $t$ actually
+produced.  Then another round of information passing follows, for iteration
+$t+1$, so that $\nonbest{\p}_{t+1}$ can be updated to $\p_{t+1}$.  Two
+iterations have then been completed with only one round of evaluations, and the
+next iteration can start again with the first round of message passing.
+
+Rounds of message passing can be expensive.  Accordingly, we describe a method
+that requires only one round of communication for each pair of iterations.  A
+lot more messages are needed, but that is sometimes more desirable than
+synchronizing all of the machines three times.
+
+This second method only requires one round of passing information because
+information about iteration $t$ and $t+1$ are passed at the same time.  Each
+processor reconstructs the from the messages it receives all of the information
+that it needs about its neighbors.  Messages are passed directly after
+evaluating each particle and its children, so all messages are of the form of
+$\nonbest{\p}_t$ or $\nonbest{\s}_{t+1}$.  The first iteration needs to be
+treated specially, so each particle can produce its initial set of speculative
+children--neighbors need only pass their initial position.  This kind of
+message passing necessitates the careful use of random seeds, so that when each
+processor computes the motion equations for its neighbors it gets the same
+results as its neighboring processors.
+
+With the results of evaluating $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$,
+along with all of the required messages from neighboring particles, the goal is
+to produce $\p_{t+1}$ and output $\noeval{\p}_{t+2}$ and $\noeval{\sset}_{t+3}$
+ready to be evaluated for the next iteration.  We first focus on the messages
+needed to produce $\p_{t+1}$.
+
+Upon evaluation, $\noeval{\p}_t$ becomes $\nonbest{\p}_t$, needing only to get
+its neighborhood best information from its neighbors.  All of its neighbors,
+then, must send it a message, so that from their updated personal best at
+iteration $t$ the particle becomes $\p_t$.
+
+With $\p_t$ we can select the correct speculative child as described above and
+produce $\nonbest{\p}_{t+1}$.  We then need the set of neighbors to
+$\nonbest{\p}_{t+1}$, $\nonbest{\nset}_{t+1}$, so we can update
+$\nonbest{\p}_{t+1}$'s neighborhood best.
+
+To produce each neighbor $\nonbest{\n}_{t+1}$, we need the same information for
+the neighboring particle that we needed to produce the original particle,
+$\nonbest{\p}_{t+1}$--we need the original neighbor particle, its speculative
+children, and its neighbors.  With that information, the set
+$\nonbest{\nset}_{t+1}$ can be obtained by following the same process used to
+obtain $\nonbest{\p}_{t+1}$.  With $\nonbest{\nset}_{t+1}$ and
+$\nonbest{\p}_{t+1}$, we can produce $\p_{t+1}$.
+
+In order to get $\p_{t+1}$, then, a particle needs to receive messages from its
+neighbors, its neighbors' neighbors, its speculative children, and its
+neighbors' speculative children.  \fig{origparticle} shows graphically how this
+information is used to produce $\p_{t+1}$.  The particle $\p_{t+1}$ can be
+passed to some central machine to track the progress of the algorithm, and it
+can be moved to $\noeval{\p}_{t+2}$ in order to start the next iteration.
 
@@
   \centering
-  \includegraphics{gbest_and_lbest.eps}
-  \caption{The complete and ring topologies in graph form.  Figure taken from
-  tracer.uc3m.es/tws/pso/neighborhood.html.}
-  \label{fig:topgraphs}
+  \includegraphics[width=\columnwidth]{messages.eps}
+  \caption{The messages needed to produce $\p_{t+1}$. $nn$ are the neighbors
+  of $p$'s neighbors, and $ns$ are the speculative children of $n$.  Items
+  without an arrow pointing to them are messages the particle needs to receive.}
+  \label{fig:origparticle}
 \end{figure}
 
-\subsection{Which Topology to Use}
+The next goal is to produce the set $\noeval{\sset}_{t+3}$.  As described
+above, the necessary components to produce $\noeval{\sset}_{t+3}$ are
+$\noeval{\p}_{t+2}$ and the neighbors of $\noeval{\p}_{t+2}$,
+$\noeval{\nset}_{t+2}$.  We already have $\noeval{\p}_{t+2}$, so what remains
+is to produce $\noeval{\nset}_{t+2}$.  It is sufficient to obtain
+$\nset_{t+1}$, as each neighbor particle $\n_{t+1}$ can be moved with the
+motion equations to $\noeval{\n}_{t+2}$.
 
-There is a theorem about optimization called ``No Free Lunch''
-\cite{wolpert-tec97}. In essence it says that no optimization algorithm can do
-better than any other optimization algorithm across all functions.  If you do
-better on some functions with a given algorithm that is because you do worse on
-other functions.  This applies directly to the question of which topology in
-PSO is the ``best,'' because it proves that there really is no ``best''
-topology across all functions.  When selecting a topology you must have a
-specific type of function in mind, and pick the topology that works best for
-that type of function.
+We have already described how to use a set of messages to obtain $\p_{t+1}$.
+The process is exactly the same to produce each $\n_{t+1}$, requiring the same
+messages, except for the neighbor particles instead of the particle itself.
 
-An interesting question along those lines is, ``Are the problems that most
-people are interested in solving inherently similar?''  If that is the case,
-one could design an algorithm that performs better on the problems people care
-about, taking a performance hit only on functions that don't matter anyway.
-But, because of No Free Lunch, any argument claiming that a particular
-algorithm or topology is better than another must make that claim only for a
-specific set of functions.
+Having obtained both $\noeval{\p}_{t+2}$ and $\noeval{\sset}_{t+3}$ from the
+messages received, the algorithm then moves to the evaluation phase, and the
+process repeats itself.  The particles are evaluated, send their messages, and
+produce the next set of particles to be evaluated from the messages received.
 
-When deciding which topology to use, then, making a wise decision requires
-having a certain kind of function in mind.
+To perform the entire process, at each message passing round a particle must
+receive messages from its neighbors, its neighbors' neighbors, its neighbors'
+neighbors' neighbors, its speculative children, its neighbors' speculative
+children, and its neighbors' neighbors' speculative children.  With the Ring
+topology, that looks like more messages than it really is, as many of the
+neighbors' neighbors are duplicates.  With the Random topology, however, the
+list of necessary messages could be rather large.
 
-The complete topology is often preferable for many problems, because it allows
-information to be shared quickly throughout the swarm and helps the algorithm
-converge more quickly.  However, too much shared information can cause
-premature convergence, as mentioned previously, so some researchers are wary of
-the complete topology.  In the absence of knowledge about the function to be
-optimized, researchers recommended, the ring topology should be used.  It might
-take longer, but at least it won't get stuck, they say \cite{bratton-sis07}.
+The decision between these two methods of performing speculative evaluation in
+a distributed parallel PSO algorithm depends on which is more expensive: 
+individual messages between processors or rounds of communication.
+
+\subsection{Dynamic Topologies}
+
+The original PSO algorithm was developed with only static topologies in mind. A
+few of the details about the neighborhood best $\nbest$ were left unspecified,
+as they are not issues with static topologies.  However, when neighbors change
+from iteration to iteration a few questions arise.  Is $\nbest$ the best
+position of any of its all-time neighbors, or only its current neighbors?  Do
+particles share their current value, their personal best value, or their
+neighborhood best value to contribute to their neighbors' neighborhood best?
+
+It has been shown that in dynamic topologies using the $\nbest$ only from
+current neighbors leads to poor performance--$\nbest$ should instead be defined
+as the best position of any of the neighbors the particle has ever
+had~\cite{mcnabb-cec09}.  The question of which value to share with neighbors
+is an open one.  Giving more information to a particle's neighbors shares
+information more quickly through the swarm.  We choose to share the personal
+best value as a good middle ground.  Also, sharing the neighborhood best would
+require more message passing in distributed algorithms, as particles would need
+more information in order to generate their speculative children from their
+neighbors.
+
+Performing speculative evaluation in PSO with a dynamic topology also raises a
+few issues of its own.  First, with neighbors changing each iteration, messages
+that processors pass to their neighbors need to be sent to the correct
+neighbors for each iteration.  In centralized algorithms and in distributed
+algorithms with several rounds of message passing (the first method described
+above), nothing needs to be changed.  However, in the case where processors
+reconstruct information about neighboring particles from messages, those
+messages need to be sent to the correct neighbors.  A particle cannot simply
+send messages to its neighbors' neighbors' neighbors--it needs to send messages
+to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.  For
+every neighbor outward information is sent, the iteration also needs to be
+incremented, as information about neighbors' neighbors is used during iteration
+$t+1$, and information about neighbors' neighbors' neighbors is used to
+reconstruct information about iteration $t+2$.  Also, this method of message
+passing again requires the use of random seeds if the topology is random.
+
+Also, an issue common to all methods of performing speculative evaluation is
+that of getting available information from neighbors before generating
+speculative children.  In a static topology, at iteration $t$ a particle
+already has all of the information about the positions of its neighbors during
+iterations $1$ through $t-1$.  If the neighbor finds a better position at 
+iteration $t$, the particle updates its neighborhood best, but if it does not,
+it still has its old neighborhood best from its neighbors for all previous 
+iterations.
+
+In a dynamic topology, a particle might not have information about the previous
+positions of its neighbors at iteration $t$.  That means that its new
+neighborhood best could come not only from its neighbors' positions at
+iteration $t$, but also from their personal best from iteration $t-1$, as
+neighbors' personal bests are what are used to update a particle's neighborhood
+best.  That creates a problem for speculative evaluation--there are potentially
+more than $2(n+1)$ possible next positions, increasing the amount of work that
+must be done to perform the second iteration at the same time as the first.
+
+This is easily fixed by updating $\noeval{\p}_{t+2}$ with the currently
+available information about $\noeval{\nset}_{t+2}$ before producing
+$\noeval{\sset}_{t+3}$.  If a particle $\noeval{\p}_{t+2}$ updates its 
+neighborhood best with the personal bests of $\noeval{\nset}_{t+2}$ before
+calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
+still only $2(n+1)$ possible next positions, and the problem is avoided.
+
+\section{Using All Speculative Evaluations}
+\label{sec:extra}
+
+In performing speculative evaluation as we have described it, $2(n+1)$
+speculative evaluations are done per particle, while all but one of them are
+completely ignored.  It seems reasonable to try to make use of those
+evaluations instead of throwing them away.  Making use of them changes the
+behavior of PSO, instead of reproducing it exactly as the above method
+explains, but the change could be an improvement.
+
+To use the extra speculative evaluations, instead of choosing the speculative
+child that matches that branch that the original PSO would have taken, we take
+the child that has the best value.  The methodology is exactly the same as
+above except for the process of choosing which speculative child to accept.
+The branch that PSO would have taken is ignored, but $\p_t$ must still be
+created so it can give the right personal best and neighborhood best values and
+positions to $\p_{t+1}$.  Then only change needed in \alg{centralized} is in
+step 7, where the $\noeval{\s}_{t+1}$ with the highest value is chosen from
+$\noeval{\sset}_{t+1}$ instead of with the matching branch.
+
+This can be thought of as drawing a number of samples from the next iteration
+and accepting the best one.  Speculative particles that move in good directions
+are kept.  Clearly this modification to PSO would produce more exploitation,
+decreasing the amount of exploration.  With functions that are less deceptive
+we would expect this method to work well, while performance on more deceptive
+functions would probably be hurt.
+
+\section{Results}
+\label{sec:results}
+
+In our experiments we compared our speculative PSO algorithm to the standard
+PSO algorithm.  To have the same number of function evaluations per iteration
+it is necessary to use a smaller swarm size in the speculative algorithm.
+There are $2(n+1)$ speculative evaluations per particle for every two
+iterations, plus one for the original, non-speculative particle.  For each of
+the topologies we used, a particle has three neighbors, including itself.  One
+of the $2(n+1)$ cases can be eliminated, however, as the particle cannot be its
+own neighborhood best update while not also updating its personal best
+(otherwise its personal best would be the neighborhood best, and it would not
+have updated).  So we have $2(n+1)$ total evaluations per particle, giving us
+$8p_s$ evaluations for every two iterations, where $p_s$ is the number of
+particles in the speculative swarm.  A regular swarm uses $2p_r$ evaluations
+per two iterations.  To have the same number of function evaluations per
+iteration, we have that $p_s = \frac{1}{4}p_r$.
+
+We use a Random topology in our speculative algorithm, as the Complete topology
+leads to an explosion of speculative evaluations.  However, if speculative
+evaluation were not being performed, it is likely that the Complete topology
+would be used.  To be fair in our comparisons, for the standard PSO algorithm
+we compare to both the Random topology with two neighbors and to the Complete
+topology.  All functions have 20 dimensions, and all results shown are averages
+over 20 runs.  For ease of labeling, we call our speculative algorithm SEPSO,
+the variant SEPSO Pick Best, and standard PSO just PSO, with the topology
+following if necessary.
+
+First we look at the function Sphere, the simplest of common benchmark
+functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
+has no local optima and is best optimized in terms of function evaluations with
+a small swarm using a Complete topology.  Accordingly, we show results for a
+small Complete swarm in \fig{sphere1}.  In comparing SEPSO with PSO, we can see
+that SEPSO clearly beat PSO with a Random topology.  However, the performance
+between SEPSO and PSO Complete is not as encouraging--the results are
+indistinguishable.  But SEPSO Pick Best handily outperforms all other
+algorithms in this problem.
 
 \begin{figure}
-  \centering
-  \includegraphics{iters_rast.eps}
-  \caption{PSO performance on the function Rastrigin with varying numbers of
-  particles.}
-  \label{fig:rast}
+  \includegraphics[width=.8\columnwidth]{sphere1.eps}
+  \caption{Function Sphere with a swarm that performs 104 evaluations per two
+  iterations.  Speculative algorithms have a swarm size of 13, and the original
+  PSO has a swarm size of 52.}
+  \label{fig:sphere1}
 \end{figure}
 
-For a swarm of 50 particles it is accurate that complete often doesn't perform
-well, and most researchers were satisfied to stop with a swarm that size.  We
-have shown, however, that increasing the number of particles in a swarm will
-help the algorithm avoid premature convergence, making complete the best
-topology even for somewhat complicated functions~\cite{mcnabb-cec09}.
-Figure~\ref{fig:rast} shows the performance of PSO on a commonly used benchmark
-function, the Rastrigin function, with varying numbers of particles.  Notice
-that with 50 particles the algorithm does get stuck pretty early, not finding
-significantly better values after 250 iterations.  This is precisely what many
-researchers in the field were saying.  With more particles, however, the
-algorithm finds much better values and continues to find better values well
-past 250 iterations.
-
-\subsection{Communication Overhead}
-
-In a parallel system, another problem with the complete topology is that
-communication between processors is expensive relative to the amount of time it
-takes to evaluate the functions many people work with.  A quick look at
-Figure~\ref{fig:topgraphs} shows the incredibly large number of messages that
-need to be sent between particles for a relatively small swarm.  Communicating
-all of the information from every particle at every iteration can take orders
-of magnitude more time than doing a simple evaluation and render
-parallelization of the algorithm worthless.
-
-We developed a modification to the algorithm along with a new topology that
-approximates the complete topology with much less communication.  If each
-particle at each iteration randomly chooses a small set of particles to send
-its information to, eventually information will be communicated throughout the
-whole swarm.  We call this the Random topology, and it is formally defined as
-follows:
-\[\mathrm{Random:}\ N_i = \{i,U_1,U_2,\dots,U_k\}\]
-where $U_j$ is a random number from 1 to $n-1$, and $k$ is the number of 
-neighbors to send information to at each iteration.
-
-In the standard particle swarm optimization algorithm, the best neighbor that a
-particle is attracted to (the \gbest in the formula above) is taken from its
-current neighborhood.  With the random topology we proposed, particles can get
-confused, being attracted to a different neighbor at every iteration, and
-performance significantly decreases.  We modified the algorithm slightly such
-that each particle remembers the best neighbor it has ever had and is attracted
-to that position instead of its current best neighbor.  Using this modification
-and a random topology, we saw performance that was indistinguishable from
-complete with only 5\% of the communication, and the speed up is significant
-even in a serial implementation~\cite{mcnabb-cec09}.
-
-Figure~\ref{fig:rastrand} shows the performance of Random versus Complete on
-the function Rastrigin for various swarm sizes.  The x-axis shows the number of
-particles in the swarm, and the y-axis shows the best function value found
-after 500 iterations.  It can be seen from the graph that Random does a very
-good job at approximating the Complete topology while required only a small
-fraction of the communication.
+To further belabor the point that performing speculative evaluation is a much
+better idea with some functions than simply adding particles, in \fig{sphere2}
+we show a graph comparing the performance of SEPSO Pick Best with 13 particles
+and with 50 particles.  The difference between the two in terms of function
+evaluations is staggering.
 
 \begin{figure}
-  \centering
-  \includegraphics{randstar.eps}
-  \caption{Random with various amounts of communication versus the Complete
-  topology on the function Rastrigin.}
-  \label{fig:rastrand}
+  \includegraphics[width=.8\columnwidth]{sphere2.eps}
+  \caption{Function Sphere with SEPSO Pick Best in two different swarm sizes.
+  The difference in performance is incredible and suggests that for this
+  function processors should most certainly be used to increase the number of
+  iterations performed and not add more particles to the swarm.}
+  \label{fig:sphere2}
 \end{figure}
 
-\subsection{Subswarms}
-
-We have run all of the experiments I've talked about so far with swarms of up
-to 4000 particles.  All of them were run with a serial implementation of
-particle swarm optimization on a single machine, though the results can be used
-in designing parallel algorithms.  With more than 4000 particles the algorithm
-becomes very computationally expensive on a single machine.  We have
-implemented a parallel version of particle swarm optimization using Google's
-MapReduce framework for parallel computation~\cite{dean-osdi04}.  Soon we will
-have it working well enough to run swarms of hundreds of thousands of particles
-on hundreds of machines at a time, and we will further explore the issues
-involved with parallel particle swarm optimization.
-
-Many new possibilities for topologies arise when the algorithm is run in
-parallel with large swarms.  We have proposed, but not yet fully tested, a
-topology of subswarms, where each of a hundred machines would evaluate 500 or
-1000 particles and send only its best value to all other machines, drastically
-reducing the amount of inter-processor communication.  In our formal notation,
-the subswarms topology can be described as
-
-\[\mathrm{Subswarms:}\ N_i = \{j|j\%k = i\%k\}\]
-where $k$ is the number of subswarms in the swarm, and \% is the mod operator.
-For a small example, in a swarm of 12 particles with 3 subswarms, the sets of
-neighbors would be \{1, 4, 7, 10\}, \{2, 5, 8, 11\}, and \{3, 6, 9, 12\}.
-
-Our preliminary results show that this approach is very
-promising~\cite{mcnabb-cec09}.  Figure~\ref{fig:subswarms} shows the results
-of using the Subswarms topology with 400 particles in each swarm, versus just
-using 400 particles in a Complete topology.  Subswarms clearly outperforms
-Complete in this case.
+Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
+\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
+\right) + 1$.  It is generally recommended to use the Ring topology when
+optimizing the Griewank function, as Complete is prone to premature convergence
+on a local optima.  We show results in \fig{griewank1} for a swarm size of 50
+particles with SEPSO, which has been recommended as a good standard swarm
+size~\cite{bratton-sis07}.  One can see in the figure that both SEPSO
+algorithms initially outperformed PSO, but got stuck in a local optima and were
+eventually outperformed.
 
 \begin{figure}
-  \centering
-  \includegraphics{islands_iters_rast.eps}
-  \caption{Subswarms versus Complete topologies on the function Rastrigin.}
-  \label{fig:subswarms}
+  \includegraphics[width=.8\columnwidth]{griewank1.eps}
+  \caption{Function Griewank with a swarm that performs 400 evaluations per two
+  iterations.  Speculative algorithms have a swarm size of 50, and the original
+  PSO has a swarm size of 200.}
+  \label{fig:griewank1}
 \end{figure}
 
-\section{A Different Speed up---Speculative Execution}
-\label{sec:specex}
+It has also been shown that with Griewank, swarms that pass a threshold of size
+consistently find the global optimum.  That size is about 400
+particles~\cite{mcnabb-cec09}.  Thus we also show results for a swarm size of
+400 with SEPSO in \fig{griewank2}.  Once the speculative swarm is of a
+sufficient size to not get caught in the local optima, it clearly outperforms
+standard PSO, finding the global optimum in fewer function evaluations.
 
-\subsection{Speculative Exectuion in General}
+\begin{figure}
+  \includegraphics[width=.8\columnwidth]{griewank2.eps}
+  \caption{Function Griewank with a swarm that performs 3200 evaluations per
+  two iterations.  Speculative algorithms have a swarm size of 400, and the
+  original PSO has a swarm size of 1600.  The local optima with the lowest
+  function value is close to $10^{-6}$.  Thus both algorithms consistently find
+  the global optimum without getting stuck.}
+  \label{fig:griewank2}
+\end{figure}
 
-Speculative execution in a program is the execution of code that may or may not
-end up actually being needed.  Modern processors routinely do this when a
-conditional branch is encountered---they try to predict which branch will
-actually be needed and continue their execution.  If it turns out that the
-prediction was incorrect, the work is discarded.  But if the prediction was
-right, the program can execute much faster than if it had waited on the branch.  
-\subsection{Speculative Execution in Particle Swarm Optimization}
+Lastly we look at functions that don't show much or any improvement with this
+method.  The Ackley function is defined as $f(\Vec{x}) = -20\exp\left\{
+-0.2\sqrt{\frac{1}{D}\sum_{i=1}^Dx_i^2}\right\} - \exp\left\{\frac{1}{D}
+\sum_{i=1}^D\cos(2\pi x_i)\right\}+20+e$. \fig{ackley} shows a comparison of
+our speculative algorithms versus the original PSO with swarm sizes of 200 and
+800, respectively.  There is no significant difference between the performance
+of SEPSO and PSO Complete, though the performance of SEPSO Pick Best is
+interesting to note.  It quickly converged on a local optimum and compared to
+the other algorithms had dismal performance on this function.
 
-In many optimization algorithms speculative execution is impossible, because
-the next iteration of the algorithm depends on the evaluation of the function
-during the previous iteration.  In simulated annealing, for instance, the
-probability of accepting a new position depends on the difference in function
-value between the new position and the current one.  In gradient descent, the
-gradient of the function must be determined (which includes actually evaluating
-it) to know how far to move before sampling the function again.
+We also ran experiments with the function Rastrigin, $f(\Vec{x}) = \sum_{i=1}^D
+\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$,  though space does not
+permit us to present a graph of our results.  However, performance of SEPSO and
+PSO Complete was largely indistinguishable, as it was with Ackley.
 
-Particle swarm optimization is rather unique in that it does not require the
-evaluation of the function to know what the next position of each particle will
-be.  It simply depends on \emph{which} particle ended up having the best
-position and whether or not the current particle found a better spot than it
-had seen before.  That means you can just compute all possible next
-positions---all combinations of which particle is the new gbest and whether or
-not pbest was updated (\gbest\ and \pbest\ from equation
-\eqref{eq:velupdate})---and evaluate each position in parallel.  That computes
-two iterations at the same time.
+\begin{figure}
+  \includegraphics[width=.8\columnwidth]{ackley.eps}
+  \caption{Function Ackley with a swarm that performs 400 evaluations per
+  two iterations.  Speculative algorithms have a swarm size of 50, and the
+  original PSO has a swarm size of 200.  Note the particularly horrible
+  performance of SEPSO Pick Best.}
+  \label{fig:ackley}
+\end{figure}
 
-\subsection{Is is worth it?}
-
-It uses a lot of extra work to get the second iteration, but for some functions
-performing more iterations makes more progress than adding extra particles.
-
-\section{Conclusion}
+\section{Conclusions}
 \label{sec:conclusion}
 
-Thank you for reviewing my proposal.  I look forward to exploring speculative
-execution in particle swarm optimization.
+We have described how parallel implementations of particle swarm optimization
+can be modified to allow additional processors to increase the number of
+iterations of the algorithm performed, instead of merely add more particles to
+the swarm.  We have detailed several different implementations of the
+algorithm, depending on the parallel PSO framework being used.  Using this
+algorithm, the original PSO algorithm is exactly reproduced two iterations at a
+time.  This technique requires more function evaluations per iteration than
+regular PSO, but for a number of functions still performs better.
 
-\bibliographystyle{plain}
+We have also described a method for making use of extra speculative evaluations
+that performs incredibly well on some functions and incredibly poorly on other
+functions.
+
+Though space has limited our presentation of results to a very small sample of
+the interesting and relevant applications of our technique, we believe the
+results shown are sufficient motivation for the further exploration of
+speculative evaluation in PSO.
+
+Our results show that performing speculative evaluation instead of increasing
+swarm size is only beneficial for some functions at the swarm sizes we used.
+However, we would venture that for all functions there is some point at which
+increasing the swarm size provides relatively little benefit.  At that point,
+additional processors should be used to speed up the algorithm instead of add
+additional particles.
+
+There are some functions for which almost no exploration needs to be done;
+Sphere is an example of such a function, as would be a function where the
+general area of the optimal point is known, but the exact point needs to be
+found.  For such functions the best use of processors is to have a small swarm
+performing speculative evaluation with our second method, where all speculative
+evaluations are used.
+
+There are other functions for which there is a point where the algorithm has
+done ``enough'' exploration and consistently finds the global optimum.  Adding
+additional particles to the swarm is akin to increasing its capacity for
+exploration, and it has been shown that for a 20-dimensional Griewank function,
+400 particles is ``enough'' exploration~\cite{mcnabb-cec09}.  In fact, the
+class of functions mentioned above could be considered a special case of these
+functions, where ``enough'' exploration can be achieved with only a very small
+number of particles proportional to the dimensionality of the problem.  For
+these functions, one should use processors to increase the swarm size until
+enough exploration has been reached, then use additional processors to increase
+the number of iterations performed.
+
+There are still other functions for which it seems there is never enough
+exploration, such as the Rastrigin function.  There are an infinite number of
+local minima in the Rastrigin function, and it has been shown that up to 4000
+particles there is no point at which ``enough'' exploration has been done.  For
+those functions the correct use of processors depends cannot be easily
+prescribed and depends largely on the desired outcome.  It is likely that the
+best thing to do with those functions is to use all processors to add particles
+to the swarm and do no speculative evaluation.
+
+\section{Future Work}
+\label{sec:future}
+
+This work represents the first attempt of which we are aware at using
+additional processors in parallel with PSO to do something other than either
+add particles to the swarm or parallelize the evaluation of the objective
+function.  We anticipate that many future advancements can be made.  For
+example, our technique required speculatively evaluating every possible next
+position of each particle.  The number of speculative evaluations could
+potentially be greatly reduced by using statistics for which branch of motion
+particles most often follow.  Only the subset of the most probable next
+locations would then be evaluated.  Using an asynchronous PSO variant would
+also be helpful in this case, allowing particles to be at different iterations
+in case the correct branch was not guessed, and needs to be evaluated
+separately.
+
+Another possible advancement would be to reduce communication costs by dropping
+some of the inter-processor messages in distributed frameworks.  The PSO
+algorithm would have to be adapted to work with less information, perhaps not
+requiring that each processor have completely accurate information about what
+neighboring particles are doing.
+
+Still another possibility is to view the set of possible evaluations as an
+infinite tree.  Each level in the tree is an iteration, and the branches on
+each level represent the possible next positions of the particle.  One need not
+simply evaluate all of one level for each particle--perhaps going two or three
+iterations ahead on the most likely path would perform better.
+
+We have mentioned that this technique has been previously used to parallelize
+simulated annealing.  Any other optimization algorithm that only depends on 
+current sampling positions when computing the next position to sample can be
+parallelized with this technique.  In particular, genetic algorithms produce
+future generations by combining individuals from the current generation.  With
+a large population size there would be an unwieldy amount of possible future
+individuals, but the potential exists to modify the algorithm to use some kind
+of speculative evaluation.  A related but quite distinct technique for reducing
+computation in evolutionary algorithms was proposed in~\cite{poli-ai06}.
+
+Large parallel clusters are often required to successfully optimize practical
+modern problems.  To properly use PSO with such clusters, a balance needs to be
+made between using processors to increase the swarm size and using them to
+increase the speed of the algorithm, or perhaps as yet unconsidered techniques.
+This work is a first step in that direction.
+
+\bibliographystyle{unsrt}
 \bibliography{%
+\bib{wolpert-tec97},%
+\bib{merris-01},%
+\bib{pso/clerc-tec02},%
 \bib{pso/kennedy-icnn95},%
-\bib{pso/clerc-tec02},%
-\bib{pso/poli-aea08},%
-\bib{pso/topology/kennedy-cec02},%
 \bib{pso/bratton-sis07},%
-\bib{aml/mcnabb-cec09},%
-\bib{mapreduce/dean-osdi04},%
-\bib{wolpert-tec97}}
+\bib{pso/largeswarms/perez-emc05},%
+\bib{pso/largeswarms/montes-gecco08.bib},%
+\bib{pso/topology/liang-sis05.bib},%
+\bib{pso/parallel/belal-ijicis04},%
+\bib{pso/parallel/chu-sci06},%
+\bib{pso/parallel/jin-aps05},%
+\bib{pso/parallel/koh-ijnme06},%
+\bib{pso/parallel/mostaghim-report06},%
+\bib{pso/parallel/parsopoulos-aia04},%
+\bib{pso/parallel/schutte-ijnme04},%
+\bib{pso/parallel/venter-wcsmo05},%
+\bib{ga/poli-ai06},%
+\bib{sim-anneal/parallel/witte-tpds91.bib},%
+\bib{pso/topology/kennedy-cec02.bib},%
+\bib{pso/topology/mohais-cec04.bib},%
+\bib{pso/topology/jordan-gecco08.bib},%
+\bib{pso/topology/clerc-oep03.bib},%
+\bib{pso/topology/miranda-ijcir08.bib},%
+\bib{pso/topology/mendes-phd04.bib},%
+\bib{aml/mcnabb-cec07},%
+\bib{aml/mcnabb-cec09}}
 

### 2009-10-02 honors_thesis/proposal/paper.tex
Proposal is finished

--- 2009-03-19 honors_thesis_proposal/paper.tex
@@
 
-\title{Honors Thesis Proposal}
+\title{Honors Thesis Proposal\\Speculative Execution in Particle Swarm
+Optimization}
 \author{Matthew Gardner}
@@
 presumably extra processors running in parallel, to perform two or more
-iterations of particle swarm optimization at a time, speeding up the algorithm
-significantly.
+iterations of particle swarm optimization at a time, thus finding better values
+faster.
 
 \section{Procedures}
-
-The bulk of the work for this project will be writing code and running
-experiments.  I have already read many conference papers on particle swarm
-optimization to know the current state of the field, so I will know what the
-results need to be compared to.  In implementing speculative execution, I will
-take care that the algorithm behaves exactly like the original, standard
-particle swarm optimization (PSO) algorithm in every aspect except the
-speed-up.  My work can then be accurately compared to other work in the field.
-
-In running experiments, many trials will be run and averaged, as there is a
-random component in this algorithm.  I will use statistical tests for all
-comparisons to the original PSO algorithm.  The success criterion will be a
-better performance of the algorithm as measured by the lowest function
-evaluation seen over a number of function evaluations for several benchmark and
-real-world problems.  Speculative execution requires more function evaluations 
-per particle than a na\"ive parallel PSO, %TODO
-PSO with the same number of particles, so the two algorithms will be compared
-at points where they use equal amounts of function evaluations per iteration.
-
-\section{Preliminary Prospectus of Finished Thesis}
-
-The finished thesis will be presented in the format of a peer-reviewed
-conference paper, including the following sections:
-
-\begin{enumerate}
-\item Introduction
-\item Related Work
-\item Particle Swarm Optimization
-\item Speculative Execution
-\item Speculative Execution in PSO
-\item Experiments
-\item Results and Conclusions
-\item Future Work
-\item References
-\end{enumerate}
-
-\section{Preliminary Research}
+\label{sec:proc}
 
@@
 \subsection{Speculative Execution}
+\label{sec:specex}
 
@@
 performing more iterations makes more progress than adding extra particles.
+
+\subsection{Planned Experiments}
+
+I will implement speculative execution in particle swarm optimization.  I will
+take care that my implementation can behave exactly like the original particle
+swarm optimization, so a fair comparison can be made.  At each iteration a 
+number of positions are speculatively evaluated, and to make it behave like
+the original PSO all that needs to be done is throw away the particles that
+do not match the actual execution path of the original particles.
+
+After showing that speculative execution reproduces PSO without affecting
+results, I will expand the idea to use the information gathered from
+speculative particles.  The positions that are speculatively evaluated are
+potentially, and probably provably, worthwhile samples from the sampling
+distribution of PSO.  Throwing them away discards useful information that could
+help the algorithm in more ways than just speeding it up.  The end result would
+be an algorithm that not only performs twice as many iterations as standard
+PSO, but takes useful samples and potentially does better while going faster.
+
+I will test this algorithm on standard benchmark functions, including at least
+the Sphere, Rastrigin and Griewank functions.  I will compare the performance
+of the speculative algorithms to the standard PSO on those functions with
+varying swarm sizes.
+
+\subsection{Evaluation}
+
+In running experiments, I will run many trials and average the results, as
+there is a random component in this algorithm.  I will use statistical tests,
+such as a paired permutation test, for all comparisons to the original PSO
+algorithm.  The success criterion will be a better performance of the algorithm
+as measured by the lowest function evaluation seen over a number of function
+evaluations.  Performing speculative execution in PSO requires more function
+evaluations per particle than regular PSO, so to be fair the algorithms will be
+compared with configurations where each uses the same number of function
+evaluations per iteration, not the same number of particles.
+
+\section{Preliminary Prospectus of Finished Thesis}
+
+The finished thesis will be presented in the format of a peer-reviewed
+conference paper, including the following sections:
+
+\begin{enumerate}
+  \item Introduction
+  \item Related Work
+  \item Particle Swarm Optimization
+  \item Speculative Execution
+  \item Speculative Execution in PSO
+  \item Experiments
+  \item Results and Conclusions
+  \item Future Work
+  \item References
+\end{enumerate}
+
+\section{Preliminary Research}
+
+Before going forward with this idea, I ran some basic, preliminary tests that
+show that for some functions, performing more iterations of PSO with fewer
+particles does better than performing fewer iterations with more particles.
+The tests were very preliminary but were enough to establish the feasibility of
+using speculative execution to improve the performance of PSO on some
+functions.
+
+Since then I have finished coding both speculative algorithms mentioned in
+Section~\ref{sec:proc}.  Further, though still preliminary, testing shows again
+that these methods perform better than standard PSO on a number of functions.
+What remains to be done is to do more exhaustive tests, collecting results that
+can be published.  I anticipate that the results that I find as I continue
+testing these algorithms will drive further improvements, but the direction of
+those improvements has yet to be seen.
 
@@
 \hline
-Finish writing and debugging code & April 30, 2009 \\
+Finish writing and debugging code & April 25, 2009 \\
 \hline
-Finish gathering data and statistics & June 30, 2009 \\
+Finish gathering data and statistics & August 31, 2009 \\
 \hline
-Complete analysis of data and begin writing thesis & September 1, 2009 \\
+Complete analysis of data and begin writing thesis & October 1, 2009 \\
 \hline

### 2009-10-02 applications/google_essay.tex
Added a few small essays for scholarship applications

\documentclass[onecolumn, 11pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{The Impact of a Simple Example}
\author{Matthew Gardner}
\date{\today}

\begin{document}
\maketitle
I was in a car barreling down a Mexican highway with my family on a trip to see
the Mayan ruins in Yucatan.  As can only happen in the tropics, a storm came
out of nowhere, and the rain was so hard my aunt pulled to the side of the
road.  As we were waiting for the storm to subside, one of the highway signs
fell over right in front of us, with a large part of it sticking out into the
road.  We saw cars go zooming by, some of them barely swerving in time to miss
the fallen sign.  Someone was going to hit the sign and wreck their car and
die; we were sure of it.  But no one was going to do anything about the sign,
and no government workers were on their way to fix it.  So my brother and I
jumped out of our car, ran through the rain to the sign, and started pulling it
out of the road.  We didn't get it very far.  But a few other people, when they
saw us pulling the sign, stopped to help us get the sign out of the way.  With
their help we were able to get the sign onto the shoulder, where passing cars
weren't in danger of hitting it.

I've reflected a lot on that experience.  No one stopped to move the sign while
we were sitting in our car watching.  They probably thought that someone should
move it, but didn't think of doing it themselves.  It wasn't until they saw
someone doing something about it that they started to stop and help.

An odd coincidence to me was that at the time I was reading in the book 1776,
by David McCullough.  As we drove away from the highway sign, I read about
Thomas Paine writing his pamphlet, "The American Crisis."  He said, "These are
the times that try men's souls: The summer soldier and the sunshine patriot
will, in this crisis, shrink from the service of his country; but he that
stands by it now, deserves the love and thanks of man and woman."  George
Washington's army was deserting him, thinking that the cause of America had
failed.  Thomas Paine's words, and his courage in the face of great opposition,
inspired the men to return to Washington and continue fighting.

I think in all of us there is a pull to be something greater than we are, to do
great things.  To be noble and good.  Most of the time we're too afraid or too
busy or too lazy to do anything about it.  But when we see someone else doing
great things, it gives us courage to step up and be the people we know we
should be.

What is it that gives a man or a woman the courage to be a leader?  I don't
know, except that he or she was probably inspired by someone else first. Ever
since my epiphany on a Mexican highway, I've tried to be the kind of person
that inspires others to be great.

For example, I had a roommate not too long ago that was quick to get upset and
argue with me and my other roommates.  I was often the calming influence in
those arguments, helping everyone to act civilly with each other.  Eventually,
through my example of not getting upset, the atmosphere of the apartment
changed, and there was a lot less arguing.  It's a small thing, I know, but
character is defined much more by small, daily actions than one heroic feat.
\end{document}

### 2009-10-02 honors_thesis/thesis/islands_iters_rast.py
Got rid of some old files

@@ -1,39 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 500
-DATADIR = '../../../psodata/data/'
-K400_FILE = DATADIR + 'star/400particles/rastrigin.out20batches'
-ISLANDS10_400_FILE = DATADIR + '10communicating_islands/4000particles/rastrigin.out20batches'
-
-plot = Plot()
-
-data = PSOData(open(K400_FILE))
-points = []
-for i in xrange(1, ITERATIONS + 1):
-    points.append((i, data.average(i)))
-plot.append(Points(points, title='Complete', style='lines'))
-
-data = PSOData(open(ISLANDS10_400_FILE))
-points = []
-for i in xrange(1, ITERATIONS + 1):
-    points.append((i, data.average(i)))
-plot.append(Points(points, title='10 Subswarms of 400 particles', style='lines'))
-
-plot.xmin = 100
-plot.xmax = ITERATIONS
-#plot.xlogscale = 10
-plot.ymin = 3
-plot.ymax = 50
-plot.ylogscale = 10
-#plot.xtics = dict((x, x) for x in SWARMSIZES if x != 400)
-plot.ytics = dict((x, x) for x in (5, 10, 20, 50, 100, 200))
-plot.xlabel = 'Iterations'
-plot.ylabel = 'Best Function Value'
-plot.show()
-plot.write('islands_iters_rast.gpi')
-
-# vim: et sw=4 sts=4

### 2009-10-02 honors_thesis/thesis/iters_rast.py

@@ -1,37 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 500
-# How many high and low outliers to drop:
-TRIM = 2
-DATADIR = '/aml/home/mjg82/psodata/data/star/'
-SWARMSIZES = [50, 100, 500, 1000, 4000]
-
-pattern = DATADIR + '%sparticles/rastrigin.out20batches'
-
-plot = Plot()
-for swarmsize in SWARMSIZES:
-    points = []
-    data = PSOData(open(pattern % swarmsize))
-    for i in xrange(1, ITERATIONS + 1):
-        points.append((i, data.average(i)))
-    plot.append(Points(points, title='%s particles' % swarmsize, style='lines'))
-
-
-plot.xmin = 1
-plot.xmax = ITERATIONS
-#plot.xlogscale = 10
-plot.ymin = 3
-plot.ymax = 300
-plot.ylogscale = 10
-#plot.xtics = dict((x, x) for x in SWARMSIZES if x != 400)
-plot.ytics = dict((x, x) for x in (5, 10, 20, 50, 100, 200))
-plot.xlabel = 'Iterations'
-plot.ylabel = 'Best Function Value'
-plot.show()
-plot.write('iters_rast.gpi')
-
-# vim: et sw=4 sts=4

### 2009-10-02 honors_thesis/thesis/randstar.py

@@ -1,55 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points
-
-ITERATIONS = 500
-# How many high and low outliers to drop:
-TRIM = 2
-DATADIR = '../../../psodata/data/'
-SWARMSIZES = [50, 100, 200, 300, 400, 500, 1000, 2000, 4000]
-#SWARMSIZES = [50, 100, 200, 300, 400, 500, 1000, 2000]
-
-star_pattern = DATADIR + 'star/%sparticles/rastrigin.out20batches'
-rand20_pattern = DATADIR + 'random20/%sparticles/rastrigin.out20batches'
-rand5_pattern = DATADIR + 'random5/%sparticles/rastrigin.out20batches'
-rand2_pattern = DATADIR + 'random2particles/%sparticles/rastrigin.out20batches'
-
-star_points = []
-star_bars = []
-rand20_points = []
-rand5_points = []
-rand2_points = []
-for swarmsize in SWARMSIZES:
-    star_data = PSOData(open(star_pattern % swarmsize))
-    star_points.append((swarmsize, star_data.average(ITERATIONS)))
-    low, med, high = star_data.statistics(ITERATIONS, TRIM)
-    print 'star:', swarmsize, low, med, high
-    star_bars.append((swarmsize, med, low, high))
-
-    rand20_data = PSOData(open(rand20_pattern % swarmsize))
-    rand20_points.append((swarmsize, rand20_data.average(ITERATIONS)))
-
-    rand5_data = PSOData(open(rand5_pattern % swarmsize))
-    rand5_points.append((swarmsize, rand5_data.average(ITERATIONS)))
-
-    rand2_data = PSOData(open(rand2_pattern % swarmsize))
-    rand2_points.append((swarmsize, rand2_data.average(ITERATIONS)))
-
-plot = Plot(xmin=45, xmax=4400, xlogscale=2, ymin=4, ymax=50, ylogscale=10)
-plot.xtics = dict((x, x) for x in SWARMSIZES if x != 400)
-plot.ytics = dict((x, x) for x in (5, 10, 20, 50))
-plot.xlabel = 'Swarm Size'
-plot.ylabel = 'Value after %s iterations' % ITERATIONS
-plot.append(Points(star_points, style='lines', title='Complete'))
-plot.append(Points(rand20_points, style='lines', 
-    title='Random, 20% communication'))
[…]

### 2009-10-29 honors_thesis/thesis/paper.tex
Revisions to my honors thesis, a bunch of other random stuff

@@
 than if it had just idly waited on the branch.  If it guessed wrong, execution
-is right where it would have been anyway.  This is called speculative
+restarts right where it would have been anyway.  This is called speculative
 execution or branch prediction.
@@
 the particle and its neighbors, not the \emph{value} of the objective function
-at those positions, speculative evaluation is possible.  The relative values of
-the positions of the particle's neighbors do matter, in that the best one is
-used as the neighborhood best position, but it's which position has the highest
-value, not what that value is, that matters.  We then propose a new method of
-parallelizing PSO in which additional processors are used to perform more
-iterations, not add particles to the swarm.
-
-Because we do not have to wait for a particle's position to be evaluated to know
-where it might move to next, we can compute the set of all possible next
-positions and send them to be evaluated along with the particle's current
-position.  Once the current position of each particle has been evaluated, we
-can determine which of the speculatively evaluated positions was correct.  We
+at those positions, speculative evaluation is possible.  We do not have to wait
+for a particle's position to be evaluated to know where it might move to next,
+so we can compute the set of all possible next positions and send them to be
+evaluated along with the particle's current position.  Once the current
+position of each particle has been evaluated, we can determine which of the
+speculatively evaluated positions matches the branch taken by the particle.  We
 have thus evaluated two iterations at once.  This can easily get unwieldy if
@@
 \sec{results} a comparison between the original PSO algorithm, our speculative
-algorithm, and its variant.
+algorithm, and its variant.  In \sec{conclusion} and \sec{future} we conclude
+and discuss future work.
 
@@
 theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
-beyond the scope of this paper.  We use the Ring topology, where each particle
-is neighbors only with two other particles, and a parallel adaptation of the
-Complete topology~\cite{mcnabb-cec09}, where each particle is neighbors with
-all other particles.  These are the two most commonly used topologies in the
-literature and have been shown to work well in practice~\cite{bratton-sis07}.
+beyond the scope of this paper.  The two most common topologies used in the
+literature are the Ring topology and the Complete
+topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
+as a parallel adaptation of the Complete topology called Random that has been
+shown to simulate the behavior of Complete with far less
+communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
+be defined as having $k$ neighbors; we use 2 neighbors for all of our
+experiments.
 
@@
 information about iteration $t$ and $t+1$ are passed at the same time.  Each
-processor reconstructs the from the messages it receives all of the information
+processor reconstructs from the messages it receives all of the information
 that it needs about its neighbors.  Messages are passed directly after
@@
 messages, except for the neighbor particles instead of the particle itself.  
+\fig{specparticles} shows graphically how $\noeval{\sset}_{t+3}$ is produced.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.6\columnwidth]{messages2.eps}
+  \caption{The messages needed to produce $\noeval{\sset}_{t+3}$. $\p_{t+1}$
+  and $\nset_{t+1}$ are both produced as in \fig{origparticle}.}
+  \label{fig:specparticles}
+\end{figure}
 
@@
 topology.  All functions have 20 dimensions, and all results shown are averages
-over 20 runs.  For ease of labeling, we call our speculative algorithm SEPSO,
-the variant SEPSO Pick Best, and standard PSO just PSO, with the topology
-following if necessary.
+over 20 runs.  Except where noted, we use swarms of close to 50 particles, as
+that has been recommended as a general-purpose swarm size~\cite{bratton-sis07}.
+When the speculative swarms have 50 particles, regular PSO swarms have 200.
+For ease of labeling, we call our speculative algorithm SEPSO, the variant
+SEPSO Pick Best, and standard PSO just PSO, with the topology following if
+necessary.
 
@@
 on a local optima.  We show results in \fig{griewank1} for a swarm size of 50
-particles with SEPSO, which has been recommended as a good standard swarm
-size~\cite{bratton-sis07}.  One can see in the figure that both SEPSO
-algorithms initially outperformed PSO, but got stuck in a local optima and were
-eventually outperformed.
+particles with SEPSO.  One can see in the figure that both SEPSO algorithms
+initially outperformed PSO, but got stuck in a local optima and were eventually
+outperformed.
 
@@
 
-Lastly we look at functions that don't show much or any improvement with this
+Lastly we look at functions that do not show much or any improvement with this
 method.  The Ackley function is defined as $f(\Vec{x}) = -20\exp\left\{
@@
 \sum_{i=1}^D\cos(2\pi x_i)\right\}+20+e$. \fig{ackley} shows a comparison of
-our speculative algorithms versus the original PSO with swarm sizes of 200 and
-800, respectively.  There is no significant difference between the performance
+our speculative algorithms versus the original PSO with swarm sizes of 50 and
+200, respectively.  There is no significant difference between the performance
 of SEPSO and PSO Complete, though the performance of SEPSO Pick Best is
@@
 exploration, and it has been shown that for a 20-dimensional Griewank function,
-400 particles is ``enough'' exploration~\cite{mcnabb-cec09}.  In fact, the
-class of functions mentioned above could be considered a special case of these
-functions, where ``enough'' exploration can be achieved with only a very small
-number of particles proportional to the dimensionality of the problem.  For
-these functions, one should use processors to increase the swarm size until
-enough exploration has been reached, then use additional processors to increase
-the number of iterations performed.
+400 particles is ``enough'' exploration~\cite{mcnabb-cec09}.  For these
+functions, one should use processors to increase the swarm size until enough
+exploration has been reached, then use additional processors to increase the
+number of iterations performed.
 
 There are still other functions for which it seems there is never enough
-exploration, such as the Rastrigin function.  There are an infinite number of
-local minima in the Rastrigin function, and it has been shown that up to 4000
-particles there is no point at which ``enough'' exploration has been done.  For
-those functions the correct use of processors depends cannot be easily
-prescribed and depends largely on the desired outcome.  It is likely that the
-best thing to do with those functions is to use all processors to add particles
-to the swarm and do no speculative evaluation.
+exploration, such as the Rastrigin function.  It has been shown that up to 4000
+particles there is no point at which ``enough'' exploration has been
+done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
+cannot be easily prescribed and depends largely on the desired outcome.  It is
+likely that the best thing to do with those functions is to use all processors
+to add particles to the swarm and do no speculative evaluation.
 
@@
 also be helpful in this case, allowing particles to be at different iterations
-in case the correct branch was not guessed, and needs to be evaluated
+in case the correct branch was not guessed and needs to be evaluated
 separately.
@@
 made between using processors to increase the swarm size and using them to
-increase the speed of the algorithm, or perhaps as yet unconsidered techniques.
-This work is a first step in that direction.
+increase the speed of the algorithm, or for perhaps as yet unconsidered
+techniques.  This work is a first step in that direction.
 

### 2009-10-29 orca/finalreport2008.tex
Revisions to my honors thesis, a bunch of other random stuff

--- 2009-04-12 honors_thesis/thesis/paper.tex
@@
 \usepackage{cite}
-\usepackage{amsmath}
 
-\DeclareMathOperator{\URand}{U}
-\providecommand{\ppos}{\ensuremath{\Vec{x}}}
-\providecommand{\pvel}{\ensuremath{\Vec{v}}}
-\providecommand{\gbest}{\ensuremath{\Vec{g}}}
-\providecommand{\pbest}{\ensuremath{\Vec{p}}}
-\providecommand{\constriction}{\ensuremath{\chi}}
-\providecommand{\coeff}{\ensuremath{\phi}}
-\newcommand{\bib}[1]{../../../bib/#1}
+\newcommand{\bib}[1]{/aml/home/mjg82/bib/#1}
 
-\topmargin .5in
-
-\title{Improving Particle Swarm Optimization \\ in a Parallel World}
-\author{Matthew Gardner}
-\date{\today}
+\title{Speculative Evaluation in Particle Swarm Optimization}
+\author{Matthew Gardner\\Mentor: Kevin Seppi, Computer Science}
+\date{}
 
@@
 
-\section{Introduction}
-\label{sec:intro}
-
 Optimization problems are ubiquitous in our modern world.  Businesses such as
 Google need to decide where to place ads, scientists need to fit models to
-data, and airlines need to schedule their flights.  All of those problems are
+data, and airlines need to schedule their flights.  All of these problems are
 optimization problems, and optimization techniques can be used to find
-solutions.  Optimization algorithms have been derived from all sorts of natural
-phenomena.  Some methods look at the landscape of the function as they are
-searching and follow hills to the best values, as in the gradient ascent
-method.  Other methods take ideas from biology or metallurgy, like genetic
-algorithms and simulated annealing.  Particle swarm optimization is a recently
-developed optimization technique that draws on ideas from the sociology of
-flocking birds.
-
-All of these methods, however, are fundamentally sequential in nature---they
-need to be run in a specific order, and because of that, they are typically
-only run on one machine.  The problems people want to solve are getting bigger,
-and larger problems need to run on multiple machines in parallel.  Parallel
-implementations of these algorithms have been attempted, but they do not often
-perform as well as sequential versions for the amount of computation performed.
-This work focuses on one of those algorithms, particle swarm optimization, and
-improving its performance in a large-scale, parallel environment.  The work I
-describe in Section~\ref{sec:topology} was done in collaboration with other
-students I work with.  The work in Section~\ref{sec:specex} I did by myself.
-
-The rest of this paper is outlined as follows.  In Section~\ref{sec:pso}, I
-formally describe the particle swarm optimization algorithm, after giving a
-brief description of optimization in general.  In Section~\ref{sec:topology},
-I describe the sociological aspect of particle swarm optimization and
-improvements that we have made for large swarms.  Section~\ref{sec:specex} then
-introduces the idea of speculative execution in general and shows how it can be
-done in particle swarm optimization.  I then conclude in Section~\ref{sec:conclusion}.
-
-\section{Particle Swarm Optimization}
-\label{sec:pso}
-
-\subsection{Optimization in General}
-
-The goal of optimization is to find the parameters to a function that either
-maximize or minimize the function value.  To take an example from above, Google
-needs to find the placements of website advertisements that generate the most
-clicks for those websites.  In this example, the placements of advertisements
-are the parameters to the function, and the number of clicks generated is the
-function to be optimized.  A more mathematical example can be seen in
-Figure~\ref{fig:cosine}.  The function shown has a point $x$ for which $f(x)$
-is the highest.  Finding the precise location of that point is not a trivial
-task, especially for more complicated functions, and it is that problem which
-optimization techniques try to solve.
-
-\begin{figure}
-  \centering
-  \includegraphics{cosine.eps}
-  \caption{A complicated function: $\cos(x)\sin(10x)e^{-|x|}$}
-  \label{fig:cosine}
-\end{figure}
-
-\subsection{Particle Swarm Optimization Specifically}
+solutions.  Particle swarm optimization (PSO) is a relatively new optimization
+technique that draws on ideas from the sociology of flocking birds.  It was
+originally developed for use on a single machine, and attempts at modifying the
+algorithm to run in parallel are still quite preliminary.  However, the
+problems people want to solve are getting bigger, and larger problems require
+more computation power than is available on any single computer.  To
+effectively solve those problems, optimization algorithms need to be improved
+to efficiently use parallel clusters of machines.  This project addressed that
+issue.
 
 Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
-Eberhart.  It tries to intelligently search a multi-dimensional space by
-mimicking the swarming and flocking behavior of birds and other
-animals~\cite{kennedy-icnn95}. It is a sociological algorithm that depends on
-interaction between particles to quickly and consistently find the optimal
-solution to a problem.  The algorithm keeps track of a number of potential
-solutions, called particles, which move somewhat randomly through the search
-space at each iteration.  When the algorithm is done, the best position that
-any of the particles ever found is returned.  Particles remember the best place
-they have been, or solution they have evaluated, and when they move they are
-attracted back to that place, as well as to the best solution other particles
-have seen.  Specifically, the formulas for updating the position $\ppos_t$ and
-velocity $\pvel_t$ of a particle at iteration $t$ are as follows:
-\begin{align}
-\label{eq:velupdate}
-	\pvel_{t+1} &=
-		\constriction \left[ \pvel_t +
-			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
-			\coeff_2\URand()\otimes(\gbest - \ppos_t)
-		\right] \\
-\label{eq:posupdate}
-	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
-\end{align}
-where \( \URand() \) is a vector of random numbers drawn from a uniform
-distribution, the \( \otimes \) operator is an element-wise vector
-multiplication, $\pbest$ (called pbest) is the best position the current
-particle has seen, and $\gbest$ (called gbest) is the best position any of the
-other particles have seen.  \( \coeff_1 \), \( \coeff_2 \), and \(
-\constriction \) are parameters with prescribed values required to ensure
-convergence (2.05, 2.05, and .73,
-respectively)~\cite{clerc-tec02}~\cite{poli-aea08}.
+Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
+a multi-dimensional space by mimicking the swarming and flocking behavior of
+birds and other animals. It is a social algorithm that depends on interaction
+between particles to quickly and consistently find the optimal solution to some
+objective function.
 
-The sociology in this algorithm defines which other particles form the
-``neighborhood'' of a given particle---the other particles whose best solution
-it sees (again, that is $\gbest$ in the equations above).  If particle 1 is a
-neighbor of particle 2, then particle 1 will be influenced by good values that
-particle 2 sees.  But if particle 3 is not a neighbor of particle 1, particle 1
-will be completely unaware of good values that particle 3 finds.
+The motion of particles through the search space has three components: an
+inertial component that gives particles momentum as they move, a cognitive
+component where particles remember the best solution they have found and are
+attracted back to that place, and a social component by which particles are
+attracted to the best solution that any of their neighbors have found.
 
-There are many ways to define a particle's neighborhood, varying from the
-entire rest of the swarm to just one other particle.
+Within the last few years, researchers have begun to recognize the need to
+develop parallel implementations of PSO.  The methods they have used include
+various synchronous algorithms, asynchronous algorithms, and formulating PSO in
+Google's MapReduce framework.  Parallelizing the evaluation of the objective
+function can also be done in some cases, though that is not an adaption of the
+PSO algorithm and thus is not the focus of this paper.
 
-The way a neighborhood is defined can have drastic effects on the performance
-of the algorithm; the more neighbors each particle has, the faster information
-spreads throughout the swarm, and the quicker the algorithm converges.  For
-some problems it is possible that the algorithm converges too quickly and gets
-caught in a local optimum; it stops on a hill when there is a mountain next to
-it.  For those kinds of problems, less communication, or smaller neighborhoods,
-is often preferable.
+All of these parallel techniques use additional processors to increase the
+number of particles in the swarm, either by adding individual particles or by
+adding entire new subswarms.  The number of iterations of PSO that the
+algorithms can perform is thus inherently limited by the time it takes to
+evaluate the objective function--additional processors add more particles, not
+more iterations.  However, there are many functions for which performing more
+iterations with fewer particles is a better use of resources than performing
+fewer iterations with more particles.
 
-\section{Topology in PSO}
-\label{sec:topology}
+To solve this problem, we take inspiration from speculative execution
+techniques commonly used in processors.  Modern processors, when faced with a
+branch on which they must wait, guess which way the branch will go and start
+executing.  If the processor guessed right, execution is much farther ahead
+than if it had just idly waited on the branch.  If it guessed wrong, execution
+restarts right where it would have been anyway.  This is called speculative
+execution or branch prediction.
 
-Much work has been done on the sociological aspect of particle swarm
-optimization, trying to determine the best neighborhood structure to use for
-various classes of problems.  That work focused on swarms of tens of particles
-to several hundred particles on a single machine.  Conclusions were reached
-about which neighborhood structures worked best, but only in this limited
-context~\cite{bratton-sis07}.  With the advent of parallel processing and the
-potential to have swarms of thousands or hundreds of thousands of particles,
-some of the earlier work needs rethinking.  This section focuses on
-neighborhood structures in very large swarms, especially when the algorithm is
-run in parallel on multiple machines.
+In PSO, the motion of particles depends only on the \emph{positions} of the
+particle and its neighbors and which neighbor's position is best, not the
+\emph{value} of the objective function at those positions.   We do not have to
+wait for a particle's position to be evaluated to know where it might move to
+next, so we can compute the set of all possible next positions and send them to
+be evaluated along with the particle's current position.  Once the current
+position of each particle has been evaluated, we can determine which of the
+speculatively evaluated positions matches the branch actually taken by the
+algorithm.  Because we already evaluated that position, we have performed two
+iterations at once.
 
-\subsection{Defining Topology}
+We implemented this parallel adaptation of the PSO algorithm and found that in
+many cases it performed better than a traditional parallelization of PSO.  This
+work was my honors thesis and was submitted to BYU for that purpose.  We also
+submitted our work to \emph{The Genetic and Evolutionary Computation
+Conference} (GECCO), one of the most prestigious conferences for PSO-related
+research, and we will find out if it was accepted by March 10, 2010.
 
-Many different neighborhood structures have been proposed and
-tested~\cite{kennedy-cec02}.  In the literature, neighborhood structures are
-referred to as topologies.  Two of them are more common---the complete or gbest
-topology, where all particles communicate with all other particles, and the
-ring or lbest topology, where each particle only communicates with its two
-immediate neighbors~\cite{bratton-sis07}.
+This work left us with a lot of open questions.  The most burning question is
+how to make the speculative algorithm more usable.  The algorithm we developed
+for speculative evaluation was somewhat limited in that it required a large
+number of extra processors to perform speculative evaluations.  For example,
+for a swarm of size 50 it required 400 processors.  If you want a swarm of 50
+particles, but only have 200 processors, with the algorithm we developed you
+can not use the extra 150 processors to perform some speculative evaluation;
+you have to have 400, or decrease your swarm size.  A much more desirable trait
+of the algorithm would be to specify a swarm size, then use any additional
+processors, however many there are, to perform what speculative evaluation you
+can.  This would require some adaptation to the algorithm, not requiring that
+particles are at the same iteration at the same time, as there are only enough
+processors for some of the particles to speculate about their future positions.
 
-To formalize the description of topologies, we can consider the swarm as a
-graph $\{V,E\}$, where the set of particles forms the nodes $V$ on the graph
-and the (directed) edges $E$ between nodes represent communication between
-particles.  If (1,2) is an edge in the graph, then particle 1 influences
-particle 2's gbest, or particle 1 is in particle 2's neighborhood.  The
-neighborhood $N_i$ of particle $i$ can be written formally as 
-\[N_i = \{p_j|(j,i)\in E\}\]
-
-Neighborhoods can also be thought of in terms of which other particles a
-particular particle sends its information to, instead of which ones it gets
-information from.  This conceptualization makes the communication a little bit
-more explicit (and more directly relates to how to actually run the algorithm)
-and will be used to describe topologies throughout the rest of this paper.
-$N_i$ then becomes
-\[N_i = \{p_j|(i,j)\in E\}\]
-
-The two topologies mentioned previously, complete and ring, would then have the
-following neighborhoods for each particle $i$ in a swarm of $n$ particles:
-\[\mathrm{Complete:}\ N_i = \{1,2,3,\dots,n\}\]
-\[\mathrm{Ring:}\ N_i = \{i-1,i,i+1\}\]
-
-For the benefit of more visual readers, the two topologies are shown in graph
-form in Figure~\ref{fig:topgraphs}; notice how many more lines are in the 
-complete topology than in the ring topology.
-
-\begin{figure}
-  \centering
-  \includegraphics{gbest_and_lbest.eps}
-  \caption{The complete and ring topologies in graph form.  Figure taken from
-  tracer.uc3m.es/tws/pso/neighborhood.html.}
-  \label{fig:topgraphs}
-\end{figure}
-
-\subsection{Which Topology to Use}
-
-There is a theorem about optimization called ``No Free Lunch''
-\cite{wolpert-tec97}. In essence it says that no optimization algorithm can do
-better than any other optimization algorithm across all functions.  If you do
-better on some functions with a given algorithm that is because you do worse on
-other functions.  This applies directly to the question of which topology in
-PSO is the ``best,'' because it proves that there really is no ``best''
-topology across all functions.  When selecting a topology you must have a
-specific type of function in mind, and pick the topology that works best for
-that type of function.
-
-An interesting question along those lines is, ``Are the problems that most
-people are interested in solving inherently similar?''  If that is the case,
-one could design an algorithm that performs better on the problems people care
-about, taking a performance hit only on functions that don't matter anyway.
-But, because of No Free Lunch, any argument claiming that a particular
-algorithm or topology is better than another must make that claim only for a
-specific set of functions.
-
-When deciding which topology to use, then, making a wise decision requires
-having a certain kind of function in mind.
-
-The complete topology is often preferable for many problems, because it allows
-information to be shared quickly throughout the swarm and helps the algorithm
-converge more quickly.  However, too much shared information can cause
-premature convergence, as mentioned previously, so some researchers are wary of
-the complete topology.  In the absence of knowledge about the function to be
-optimized, researchers recommended, the ring topology should be used.  It might
-take longer, but at least it won't get stuck, they say \cite{bratton-sis07}.
-
-\begin{figure}
-  \centering
-  \includegraphics{iters_rast.eps}
-  \caption{PSO performance on the function Rastrigin with varying numbers of
-  particles.}
-  \label{fig:rast}
-\end{figure}
-
-For a swarm of 50 particles it is accurate that complete often doesn't perform
-well, and most researchers were satisfied to stop with a swarm that size.  We
-have shown, however, that increasing the number of particles in a swarm will
-help the algorithm avoid premature convergence, making complete the best
-topology even for somewhat complicated functions~\cite{mcnabb-cec09}.
-Figure~\ref{fig:rast} shows the performance of PSO on a commonly used benchmark
-function, the Rastrigin function, with varying numbers of particles.  Notice
-that with 50 particles the algorithm does get stuck pretty early, not finding
-significantly better values after 250 iterations.  This is precisely what many
-researchers in the field were saying.  With more particles, however, the
-algorithm finds much better values and continues to find better values well
-past 250 iterations.
-
-\subsection{Communication Overhead}
-
-In a parallel system, another problem with the complete topology is that
-communication between processors is expensive relative to the amount of time it
-takes to evaluate the functions many people work with.  A quick look at
-Figure~\ref{fig:topgraphs} shows the incredibly large number of messages that
-need to be sent between particles for a relatively small swarm.  Communicating
-all of the information from every particle at every iteration can take orders
-of magnitude more time than doing a simple evaluation and render
-parallelization of the algorithm worthless.
-
-We developed a modification to the algorithm along with a new topology that
-approximates the complete topology with much less communication.  If each
-particle at each iteration randomly chooses a small set of particles to send
-its information to, eventually information will be communicated throughout the
-whole swarm.  We call this the Random topology, and it is formally defined as
-follows:
-\[\mathrm{Random:}\ N_i = \{i,U_1,U_2,\dots,U_k\}\]
-where $U_j$ is a random number from 1 to $n-1$, and $k$ is the number of 
-neighbors to send information to at each iteration.
-
-In the standard particle swarm optimization algorithm, the best neighbor that a
-particle is attracted to (the \gbest in the formula above) is taken from its
-current neighborhood.  With the random topology we proposed, particles can get
-confused, being attracted to a different neighbor at every iteration, and
-performance significantly decreases.  We modified the algorithm slightly such
-that each particle remembers the best neighbor it has ever had and is attracted
-to that position instead of its current best neighbor.  Using this modification
-and a random topology, we saw performance that was indistinguishable from
-complete with only 5\% of the communication, and the speed up is significant
-even in a serial implementation~\cite{mcnabb-cec09}.
-
-Figure~\ref{fig:rastrand} shows the performance of Random versus Complete on
-the function Rastrigin for various swarm sizes.  The x-axis shows the number of
-particles in the swarm, and the y-axis shows the best function value found
-after 500 iterations.  It can be seen from the graph that Random does a very
-good job at approximating the Complete topology while required only a small
-fraction of the communication.
-
-\begin{figure}
-  \centering
-  \includegraphics{randstar.eps}
-  \caption{Random with various amounts of communication versus the Complete
-  topology on the function Rastrigin.}
-  \label{fig:rastrand}
-\end{figure}
-
-\subsection{Subswarms}
-
-We have run all of the experiments I've talked about so far with swarms of up
-to 4000 particles.  All of them were run with a serial implementation of
-particle swarm optimization on a single machine, though the results can be used
-in designing parallel algorithms.  With more than 4000 particles the algorithm
-becomes very computationally expensive on a single machine.  We have
-implemented a parallel version of particle swarm optimization using Google's
-MapReduce framework for parallel computation~\cite{dean-osdi04}.  Soon we will
-have it working well enough to run swarms of hundreds of thousands of particles
-on hundreds of machines at a time, and we will further explore the issues
-involved with parallel particle swarm optimization.
-
-Many new possibilities for topologies arise when the algorithm is run in
-parallel with large swarms.  We have proposed, but not yet fully tested, a
-topology of subswarms, where each of a hundred machines would evaluate 500 or
-1000 particles and send only its best value to all other machines, drastically
-reducing the amount of inter-processor communication.  In our formal notation,
-the subswarms topology can be described as
-
-\[\mathrm{Subswarms:}\ N_i = \{j|j\%k = i\%k\}\]
-where $k$ is the number of subswarms in the swarm, and \% is the mod operator.
-For a small example, in a swarm of 12 particles with 3 subswarms, the sets of
-neighbors would be \{1, 4, 7, 10\}, \{2, 5, 8, 11\}, and \{3, 6, 9, 12\}.
-
-Our preliminary results show that this approach is very
-promising~\cite{mcnabb-cec09}.  Figure~\ref{fig:subswarms} shows the results
-of using the Subswarms topology with 400 particles in each swarm, versus just
-using 400 particles in a Complete topology.  Subswarms clearly outperforms
-Complete in this case.
-
-\begin{figure}
-  \centering
-  \includegraphics{islands_iters_rast.eps}
-  \caption{Subswarms versus Complete topologies on the function Rastrigin.}
-  \label{fig:subswarms}
-\end{figure}
-
-\section{A Different Speed up---Speculative Execution}
-\label{sec:specex}
-
-\subsection{Speculative Exectuion in General}
-
-Speculative execution in a program is the execution of code that may or may not
-end up actually being needed.  Modern processors routinely do this when a
-conditional branch is encountered---they try to predict which branch will
-end up being needed and continue their execution.  If it turns out that the
-prediction was incorrect, the work is discarded.  But if the prediction was
-right, the program can execute much faster than if it had waited on the branch.
-
-\subsection{Speculative Execution in Particle Swarm Optimization}
-
-In many optimization algorithms speculative execution is impossible, because
-the next iteration of the algorithm depends on the evaluation of the function
-during the previous iteration.  To know which positions to speculatively
-evaluate you have to already have evaluated the current position, and then it's
-not speculative anymore.  In simulated annealing, for instance, the probability
-of accepting a new position depends on the difference in function value between
-the new position and the current one.  In gradient descent, the gradient of the
-function must be determined (which includes actually evaluating it) to know how
-far to move before sampling the function again.  It may be possible to modify
-these algorithms to allow for speculative execution, but their standard
-executions cannot be done speculatively.
-
-Particle swarm optimization is rather unique in that it does not require the
-evaluation of the function to know what the next position of each particle will
-be.  It simply depends on \emph{which} particle ended up having the best
-position and whether or not the current particle found a better spot than it
-had seen before.  That means you can just compute all possible next
-positions---all combinations of which particle is the new gbest and whether or
-not pbest was updated (\gbest\ and \pbest\ from equation
-\eqref{eq:velupdate})---and evaluate each position in parallel.  That computes
-two iterations at the same time.
-
-\subsection{Is is worth it?}
-
-Speculative execution uses a lot of extra work to compute two iterations at the
-same time, but for some functions performing more iterations makes more
-progress than adding extra particles.  The No Free Lunch theorem again shows
-its head here, though, because if you do better with one algorithm on a certain
-function or set of functions, that's because you do worse on other functions.
-There are, however, a set of functions that do better with more iterations than
-with more particles.
-
-Figure~\ref{fig:sphere} shows an example of this: the function Sphere.  The
-graph shows the best function value found over a number of function
-evaluations.  This is different than previous graphs that have shown function
-value against a number of iterations.  At each iteration, a swarm of 50
-particles performs 50 function evaluations, where a swarm of 4000 particles
-performs 4000 evaluations.  Figure~\ref{fig:sphere} shows that the extra 3950
-function evaluations spent on the extra particles in a 4000 particle swarm
-really don't do any good.  It would be more worth it to just do more iterations
-of a 50 particle swarm than to have a swarm of 4000 particles.
-
-The benefit of adding more particles comes in when you have hundreds or
-thousands of processors in a parallel cluster.  You can't use 4000 machines to
-run a swarm of 50 particles---3950 of the processors would be idle the whole
-time.  Adding more particles to the swarm then improves performance without
-adding any extra time (ignoring inter-processor communication cost for the
-moment).  But for some functions, a better idea than using the extra 3950
-processors to add more particles to the swarm is to use the processors
-speculatively so that two iterations can be done at the same time.
-Figure~\ref{fig:sphere} shows just one example of a function that would be
-better optimized by performing speculative execution than by adding extra
-particles.
-
-\begin{figure}
-  \centering
-  \includegraphics{evals_sphere.eps}
-  \caption{Best value found on the function Sphere versus the number of 
-  function evaluations with varying swarm sizes.}
-  \label{fig:sphere}
-\end{figure}
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-In a world where people are swimming in processing power, many old algorithms
-need to be rethought.  Parallel computing architectures allow for new
-techniques that were previously impossible.  In the realm of optimization,
-particle swarm optimization is uniquely capable of reaping great benefits from
-parallel processing.  I have reviewed just a few ways in which PSO can be
-improved to harness the increased power of parallel clusters.  As more
-researchers tackle the problems of modifying and adapting algorithms to better
-utilize the resources that are now available, our ability to solve problems
-with computers will increase dramatically.
+We plan to pursue this area of research, and are submitting another ORCA grant
+proposal to answer that question.  We would then take the conference paper we
+submitted to GECCO and expand it with the aim of submitting it to
+\emph{Transactions on Evolutionary Computation}, a journal that deals with
+PSO-related research, along with other evolutionary computation topics.
 
@@
 \bibliography{%
+\bib{wolpert-tec97},%
+\bib{merris-01},%
+\bib{pso/clerc-tec02},%
 \bib{pso/kennedy-icnn95},%
-\bib{pso/clerc-tec02},%
-\bib{pso/poli-aea08},%
-\bib{pso/topology/kennedy-cec02},%
 \bib{pso/bratton-sis07},%
-\bib{aml/mcnabb-cec09},%
-\bib{mapreduce/dean-osdi04},%
-\bib{wolpert-tec97}}
+\bib{pso/largeswarms/perez-emc05},%
+\bib{pso/largeswarms/montes-gecco08.bib},%
+\bib{pso/topology/liang-sis05.bib},%
+\bib{pso/parallel/belal-ijicis04},%
+\bib{pso/parallel/chu-sci06},%
+\bib{pso/parallel/jin-aps05},%
+\bib{pso/parallel/koh-ijnme06},%
+\bib{pso/parallel/mostaghim-report06},%
+\bib{pso/parallel/parsopoulos-aia04},%
+\bib{pso/parallel/schutte-ijnme04},%
+\bib{pso/parallel/venter-wcsmo05},%
+\bib{ga/poli-ai06},%
+\bib{sim-anneal/parallel/witte-tpds91.bib},%
+\bib{pso/topology/kennedy-cec02.bib},%
+\bib{pso/topology/mohais-cec04.bib},%
+\bib{pso/topology/jordan-gecco08.bib},%
+\bib{pso/topology/clerc-oep03.bib},%
+\bib{pso/topology/miranda-ijcir08.bib},%
+\bib{pso/topology/mendes-phd04.bib},%
+\bib{aml/mcnabb-cec07},%
+\bib{aml/mcnabb-cec09}}
 

### 2009-10-29 honors_thesis/thesis/sepsodata/plots/ackley.py

@@ -1,0 +1,49 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 500
+SWARMSIZE = 50
+
+SEPSO_FILE = '../sepso_reproduce/ackley/rand50particles'
+PICKBEST_FILE = '../sepso_pickbest/ackley/rand50particles'
+STANDARD_RAND = '../standardpso/ackley/rand200particles'
+STANDARD_RING = '../standardpso/ackley/ring200particles'
+STANDARD_COMPLETE = '../standardpso/ackley/complete200particles'
+
+sepso_data = PSOData(open(SEPSO_FILE))
+pickbest_data = PSOData(open(PICKBEST_FILE))
+rand_data = PSOData(open(STANDARD_RAND))
+complete_data = PSOData(open(STANDARD_COMPLETE))
+sepso_points = []
+pickbest_points = []
+rand_points = []
+complete_points = []
+for i in xrange(1, ITERATIONS):
+    evaluations = i * SWARMSIZE
+    sepso_points.append((evaluations, sepso_data.average(i*2)))
+    pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+    rand_points.append((evaluations, rand_data.average(i)))
+    complete_points.append((evaluations, complete_data.average(i)))
+
+plot = Plot()
+plot.xmin = SWARMSIZE
+plot.xmax = ITERATIONS * SWARMSIZE
+plot.ymin = 10 ** -10
+plot.ymax = 100
+plot.ylogscale = 10
+plot.xlabel = 'Function Evaluations'
+plot.ylabel = 'Best Function Value'
+
+plot.append(Points(sepso_points, title='SEPSO', style='lines'))
+plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
+plot.append(Points(rand_points, title='PSO Random', style='lines'))
+plot.append( Points(complete_points, title='PSO Complete', style='lines'))
+
+plot.show()
+plot.write('ackley.gpi')
+
+
+# vim: et sw=4 sts=4

### 2009-10-29 honors_thesis/thesis/sepsodata/plots/complete_200.py

@@ -1,0 +1,59 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 500
+SWARMSIZE = 200
+
+FUNCTIONS = ['sphere', 'rastrigin', 'rosenbrock', 'griewank', 'ackley']
+
+FUNCTION_RANGE = [(10 ** -25, 10 ** 5), (0, 300), (10, float(10 ** 10)),
+        (10 ** -3, 500), (10 ** -10, 100)]
+
+FUNCTION_LOGSCALE = [10, 0, 10, 10, 10]
+
+SEPSO_FILE = '../sepso_reproduce/function/rand50particles'
+PICKBEST_FILE = '../sepso_pickbest/function/rand50particles'
+STANDARD_RAND = '../standardpso/function/rand200particles'
+STANDARD_COMPLETE = '../standardpso/function/complete200particles'
+
+for f, function in enumerate(FUNCTIONS):
+    sepso_data = PSOData(open(SEPSO_FILE.replace('function', function)))
+    pickbest_data = PSOData(open(PICKBEST_FILE.replace('function', function)))
+    rand_data = PSOData(open(STANDARD_RAND.replace('function', function)))
+    complete_data = PSOData(open(STANDARD_COMPLETE.replace('function',
+            function)))
+    sepso_points = []
+    pickbest_points = []
+    rand_points = []
+    complete_points = []
+    for i in xrange(1, ITERATIONS):
+        evaluations = i * SWARMSIZE
+        sepso_points.append((evaluations, sepso_data.average(i*2)))
+        pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+        rand_points.append((evaluations, rand_data.average(i)))
+        complete_points.append((evaluations, complete_data.average(i)))
+
+    plot = Plot()
+    plot.xmin = SWARMSIZE
+    plot.xmax = ITERATIONS * SWARMSIZE
+    plot.ymin = FUNCTION_RANGE[f][0]
+    plot.ymax = FUNCTION_RANGE[f][1]
+    plot.ylogscale = FUNCTION_LOGSCALE[f]
+    plot.xlabel = 'Function Evaluations'
+    plot.ylabel = 'Best Function Value'
+    plot.title = 'Complete topology, 50 and 200 particles, function: %s' \
+            % function
+
+    plot.append(Points(sepso_points, title='SEPSO', style='lines'))
[…]

### 2009-10-29 honors_thesis/thesis/sepsodata/plots/complete_50.py

@@ -1,0 +1,59 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 500
+SWARMSIZE = 52
+
+FUNCTIONS = ['sphere', 'rastrigin', 'rosenbrock', 'griewank', 'ackley']
+
+FUNCTION_RANGE = [(10 ** -20, 10 ** 5), (0, 300), (10, float(10 ** 10)),
+        (10 ** -3, 500), (10 ** -5, 100)]
+
+FUNCTION_LOGSCALE = [10, 0, 10, 10, 10]
+
+SEPSO_FILE = '../sepso_reproduce/function/rand13particles'
+PICKBEST_FILE = '../sepso_pickbest/function/rand13particles'
+STANDARD_RAND = '../standardpso/function/rand52particles'
+STANDARD_COMPLETE = '../standardpso/function/complete52particles'
+
+for f, function in enumerate(FUNCTIONS):
+    sepso_data = PSOData(open(SEPSO_FILE.replace('function', function)))
+    pickbest_data = PSOData(open(PICKBEST_FILE.replace('function', function)))
+    rand_data = PSOData(open(STANDARD_RAND.replace('function', function)))
+    complete_data = PSOData(open(STANDARD_COMPLETE.replace('function',
+            function)))
+    sepso_points = []
+    pickbest_points = []
+    rand_points = []
+    complete_points = []
+    for i in xrange(1, ITERATIONS):
+        evaluations = i * SWARMSIZE
+        sepso_points.append((evaluations, sepso_data.average(i*2)))
+        pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+        rand_points.append((evaluations, rand_data.average(i)))
+        complete_points.append((evaluations, complete_data.average(i)))
+
+    plot = Plot()
+    plot.xmin = SWARMSIZE
+    plot.xmax = ITERATIONS * SWARMSIZE
+    plot.ymin = FUNCTION_RANGE[f][0]
+    plot.ymax = FUNCTION_RANGE[f][1]
+    plot.ylogscale = FUNCTION_LOGSCALE[f]
+    plot.xlabel = 'Function Evaluations'
+    plot.ylabel = 'Best Function Value'
+    plot.title = 'Complete topology, 13 and 52 particles, function: %s' \
+            % function
+
+    plot.append(Points(sepso_points, title='SEPSO', style='lines'))
[…]

### 2009-10-29 honors_thesis/thesis/sepsodata/plots/griewank.py

@@ -1,0 +1,42 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 1000
+SWARMSIZE = 50
+
+SEPSO_FILE = '../sepso_reproduce/griewank/ring50particles'
+PICKBEST_FILE = '../sepso_pickbest/griewank/ring50particles'
+STANDARD_RAND = '../standardpso/griewank/ring200particles'
+
+sepso_data = PSOData(open(SEPSO_FILE))
+pickbest_data = PSOData(open(PICKBEST_FILE))
+ring_data = PSOData(open(STANDARD_RAND))
+sepso_points = []
+pickbest_points = []
+ring_points = []
+for i in xrange(1, ITERATIONS):
+    evaluations = i * SWARMSIZE
+    sepso_points.append((evaluations, sepso_data.average(i*2)))
+    pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+    ring_points.append((evaluations, ring_data.average(i)))
+
+plot = Plot()
+plot.xmin = SWARMSIZE
+plot.xmax = ITERATIONS * SWARMSIZE
+plot.ymin = 10 ** -9
+plot.ymax = 500
+plot.ylogscale = 10
+plot.xlabel = 'Function Evaluations'
+plot.ylabel = 'Best Function Value'
+
+plot.append(Points(sepso_points, title='SEPSO', style='lines'))
+plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
+plot.append(Points(ring_points, title='PSO Ring', style='lines'))
+
+plot.show()
+plot.write('griewank1.gpi')
+
+# vim: et sw=4 sts=4

### 2009-10-29 honors_thesis/thesis/sepsodata/plots/griewank2.py

@@ -1,0 +1,42 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 1000
+SWARMSIZE = 400
+
+SEPSO_FILE = '../sepso_reproduce/griewank/ring400particles'
+PICKBEST_FILE = '../sepso_pickbest/griewank/ring400particles'
+STANDARD_RAND = '../standardpso/griewank/ring1600particles'
+
+sepso_data = PSOData(open(SEPSO_FILE))
+pickbest_data = PSOData(open(PICKBEST_FILE))
+ring_data = PSOData(open(STANDARD_RAND))
+sepso_points = []
+pickbest_points = []
+ring_points = []
+for i in xrange(1, ITERATIONS):
+    evaluations = i * SWARMSIZE
+    sepso_points.append((evaluations, sepso_data.average(i*2)))
+    #pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+    ring_points.append((evaluations, ring_data.average(i)))
+
+plot = Plot()
+plot.xmin = SWARMSIZE
+plot.xmax = ITERATIONS * SWARMSIZE
+plot.ymin = 10 ** -8
+plot.ymax = 500
+plot.ylogscale = 10
+plot.xlabel = 'Function Evaluations'
+plot.ylabel = 'Best Function Value'
+
+plot.append(Points(sepso_points, title='SEPSO', style='lines'))
+#plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
+plot.append(Points(ring_points, title='PSO Ring', style='lines'))
+
+plot.show()
+plot.write('griewank2.gpi')
+
+# vim: et sw=4 sts=4

### 2009-10-29 honors_thesis/thesis/sepsodata/plots/rastrigin.py

@@ -1,0 +1,52 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 1000
+SWARMSIZE = 200
+
+SEPSO_FILE = '../sepso_reproduce/rastrigin/rand200particles'
+PICKBEST_FILE = '../sepso_pickbest/rastrigin/rand200particles'
+STANDARD_RAND = '../standardpso/rastrigin/rand800particles'
+STANDARD_RING = '../standardpso/rastrigin/ring800particles'
+STANDARD_COMPLETE = '../standardpso/rastrigin/complete800particles'
+
+sepso_data = PSOData(open(SEPSO_FILE))
+#pickbest_data = PSOData(open(PICKBEST_FILE))
+rand_data = PSOData(open(STANDARD_RAND))
+#ring_data = PSOData(open(STANDARD_RING))
+complete_data = PSOData(open(STANDARD_COMPLETE))
+sepso_points = []
+pickbest_points = []
+rand_points = []
+#ring_points = []
+complete_points = []
+for i in xrange(1, ITERATIONS):
+    evaluations = i * SWARMSIZE
+    sepso_points.append((evaluations, sepso_data.average(i*2)))
+    #pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+    rand_points.append((evaluations, rand_data.average(i)))
+    #ring_points.append((evaluations, ring_data.average(i)))
+    complete_points.append((evaluations, complete_data.average(i)))
+
+plot = Plot()
+plot.xmin = SWARMSIZE
+plot.xmax = ITERATIONS * SWARMSIZE
+plot.ymin = 0
+plot.ymax = 100
+plot.xlabel = 'Function Evaluations'
+plot.ylabel = 'Best Function Value'
+
+plot.append(Points(sepso_points, title='SEPSO', style='lines'))
+#plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
+plot.append(Points(rand_points, title='PSO Random', style='lines'))
+#plot.append(Points(ring_points, title='PSO Ring', style='lines'))
+plot.append( Points(complete_points, title='PSO Complete', style='lines'))
+
+plot.show()
+plot.write('rastrigin.gpi')
+
+
+# vim: et sw=4 sts=4

### 2009-10-29 honors_thesis/thesis/sepsodata/plots/ring_200.py

@@ -1,0 +1,50 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 500
+SWARMSIZE = 200
+
+FUNCTIONS = ['sphere', 'rastrigin', 'rosenbrock', 'griewank', 'ackley']
+
+FUNCTION_RANGE = [(10 ** -20, 10 ** 5), (0, 300), (10, float(10 ** 10)),
+        (10 ** -5, 500), (10 ** -10, 100)]
+
+FUNCTION_LOGSCALE = [10, 0, 10, 10, 10]
+
+SEPSO_FILE = '../sepso_reproduce/function/ring50particles'
+PICKBEST_FILE = '../sepso_pickbest/function/ring50particles'
+STANDARD_FILE = '../standardpso/function/ring200particles'
+
+for f, function in enumerate(FUNCTIONS):
+    sepso_data = PSOData(open(SEPSO_FILE.replace('function', function)))
+    pickbest_data = PSOData(open(PICKBEST_FILE.replace('function', function)))
+    standard_data = PSOData(open(STANDARD_FILE.replace('function', function)))
+    sepso_points = []
+    pickbest_points = []
+    standard_points = []
+    for i in xrange(1, ITERATIONS):
+        evaluations = i * SWARMSIZE
+        sepso_points.append((evaluations, sepso_data.average(i*2)))
+        pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+        standard_points.append((evaluations, standard_data.average(i)))
+
+    plot = Plot()
+    plot.xmin = SWARMSIZE
+    plot.xmax = ITERATIONS * SWARMSIZE
+    plot.ymin = FUNCTION_RANGE[f][0]
+    plot.ymax = FUNCTION_RANGE[f][1]
+    plot.ylogscale = FUNCTION_LOGSCALE[f]
+    plot.xlabel = 'Function Evaluations'
+    plot.ylabel = 'Best Function Value'
+    plot.title = 'Ring topology, 50 and 200 particles, function: %s' % function
+
+    plot.append(Points(sepso_points, title='SEPSO', style='lines'))
+    plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
+    plot.append(Points(standard_points, title='Standard PSO', style='lines'))
+
+    plot.show()
+
+# vim: et sw=4 sts=4

### 2009-10-29 honors_thesis/thesis/sepsodata/plots/ring_50.py

@@ -1,0 +1,50 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 500
+SWARMSIZE = 52
+
+FUNCTIONS = ['sphere', 'rastrigin', 'rosenbrock', 'griewank', 'ackley']
+
+FUNCTION_RANGE = [(10 ** -20, 10 ** 5), (0, 300), (100, float(10 ** 10)),
+        (10 ** -2, 500), (10 ** -2, 100)]
+
+FUNCTION_LOGSCALE = [10, 0, 10, 10, 10]
+
+SEPSO_FILE = '../sepso_reproduce/function/ring13particles'
+PICKBEST_FILE = '../sepso_pickbest/function/ring13particles'
+STANDARD_FILE = '../standardpso/function/ring52particles'
+
+for f, function in enumerate(FUNCTIONS):
+    sepso_data = PSOData(open(SEPSO_FILE.replace('function', function)))
+    pickbest_data = PSOData(open(PICKBEST_FILE.replace('function', function)))
+    standard_data = PSOData(open(STANDARD_FILE.replace('function', function)))
+    sepso_points = []
+    pickbest_points = []
+    standard_points = []
+    for i in xrange(1, ITERATIONS):
+        evaluations = i * SWARMSIZE
+        sepso_points.append((evaluations, sepso_data.average(i*2)))
+        pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+        standard_points.append((evaluations, standard_data.average(i)))
+
+    plot = Plot()
+    plot.xmin = SWARMSIZE
+    plot.xmax = ITERATIONS * SWARMSIZE
+    plot.ymin = FUNCTION_RANGE[f][0]
+    plot.ymax = FUNCTION_RANGE[f][1]
+    plot.ylogscale = FUNCTION_LOGSCALE[f]
+    plot.xlabel = 'Function Evaluations'
+    plot.ylabel = 'Best Function Value'
+    plot.title = 'Ring topology, 13 and 52 particles, function: %s' % function
+
+    plot.append(Points(sepso_points, title='SEPSO', style='lines'))
+    plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
+    plot.append(Points(standard_points, title='Standard PSO', style='lines'))
+
+    plot.show()
+
+# vim: et sw=4 sts=4

### 2009-10-29 honors_thesis/thesis/sepsodata/plots/sphere.py

@@ -1,0 +1,78 @@
+#!/usr/bin/env python
+
+from __future__ import division
+from psodata import PSOData
+from evilplot import Plot, Points, RawData
+
+ITERATIONS = 500
+SWARMSIZE = 52
+
+SEPSO_FILE = '../sepso_reproduce/sphere/rand13particles'
+PICKBEST_FILE = '../sepso_pickbest/sphere/rand13particles'
+STANDARD_RAND = '../standardpso/sphere/rand52particles'
+STANDARD_COMPLETE = '../standardpso/sphere/complete52particles'
+
+sepso_data = PSOData(open(SEPSO_FILE))
+pickbest_data = PSOData(open(PICKBEST_FILE))
+rand_data = PSOData(open(STANDARD_RAND))
+complete_data = PSOData(open(STANDARD_COMPLETE))
+sepso_points = []
+pickbest_points = []
+rand_points = []
+complete_points = []
+for i in xrange(1, ITERATIONS):
+    evaluations = i * SWARMSIZE
+    sepso_points.append((evaluations, sepso_data.average(i*2)))
+    pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+    rand_points.append((evaluations, rand_data.average(i)))
+    complete_points.append((evaluations, complete_data.average(i)))
+
+plot = Plot()
+plot.xmin = SWARMSIZE
+plot.xmax = ITERATIONS * SWARMSIZE
+plot.ymin = 10 ** -20
+plot.ymax = 10 ** 5
+plot.ylogscale = 10
+plot.xlabel = 'Function Evaluations'
+plot.ylabel = 'Best Function Value'
+
+plot.append(Points(sepso_points, title='SEPSO', style='lines'))
+plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
+plot.append(Points(rand_points, title='PSO Random', style='lines'))
+plot.append( Points(complete_points, title='PSO Complete', style='lines'))
+
+plot.show()
+plot.write('sphere1.gpi')
+
+ITERATIONS = 250
+SWARMSIZE = 50
+
+PICKBEST_FILE = '../sepso_pickbest/sphere/rand52particles'
+PICKBEST13_FILE = '../sepso_pickbest/sphere/rand13particles2000iters'
+
+pickbest_data = PSOData(open(PICKBEST_FILE))
+pickbest13_data = PSOData(open(PICKBEST13_FILE))
[…]

### 2009-11-19 honors_thesis/thesis/paper.tex
Some changes to my honors thesis based on a comments from Dr. Goodrich.  Not done yet.

@@
 the world's computing power is increasingly coming from large clusters of
-processors.  Even in desktop machines, systems with 4 or 8 cores are common
-place.  In order to efficiently utilize these resources for computationally
-intensive problems, algorithms need to run in parallel.
+processors.  Even in desktop machines, systems with 4 or 8 cores are
+commonplace.  In order to efficiently utilize these resources for
+computationally intensive problems, algorithms need to run in parallel.
 
@@
 algorithms can perform is thus inherently limited by the time it takes to
-evaluate the objective function--additional processors add more particles, not
+evaluate the objective function---additional processors add more particles, not
 more iterations.  However, there are many functions for which performing more
@@
 For a particle at iteration $t-1$, $\pvel_t$ and $\ppos_t$ are defined by
-Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  The particle already
-knows its personal best $\pbest$ at iteration $t-1$, but it could be updated to
-$\ppos_t$ if its position at iteration $t$ has a better value than its current
-personal best.  Likewise, the particle knows its neighborhood best $\nbest$ at
-iteration $t-1$, though it could be updated by any of its neighbors' positions
-at iteration $t$.
+Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  The only remaining
+components are $\pbest$ and $\nbest$, each of which has only a few
+possibilities.  The particle already knows its personal best $\pbest$ at
+iteration $t-1$, but it could be updated to $\ppos_t$ if its position at
+iteration $t$ has a better value than its current personal best.  Likewise, the
+particle knows its neighborhood best $\nbest$ at iteration $t-1$, though it
+could be updated by any of its neighbors' positions at iteration $t$.
 
@@
 $2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
-neighbors the particle has (a particle either updates its personal best or it
-does not, and one of its $n$ neighbors must update its neighborhood best, or
-there is no neighborhood best update--$n+1$ possibilities).  If all of those
-speculative positions are evaluated in parallel along with the original,
+neighbors the particle has, because there are two possibilities for updating a
+particle's personal best (updated and not updated) and $n+1$ possibilities for
+updating its neighborhood best (one of $n$ neighbors, or no update).  If all of
+those speculative positions are evaluated in parallel along with the original,
 non-speculative position of the particle at time $t-1$, then one of the
@@
 states that a particle could have.  We use $\p_t$ to denote a particle at
-iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
-$\sset_{t+1}$.  $\n_t$ is a neighbor of particle $\p_t$.  A set of particles is
-$\pset$, whereas a single particle is simply $\p$.
+iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children.
+$\n_t$ is a neighbor of particle $\p_t$.  A set of particles is $\pset$,
+$\sset$, or $\nset$, whereas a single particle is simply $\p$, $\s$, or $\n$.
 
@@
 $\nonbest{\s}_{t+1}$ would think that its current position was its personal
-best, as the value it found--2--was better than its previous personal best
+best, as the value it found, 2, was better than its previous personal best
 value of 3.  It needs to receive the personal best value from its parent to
@@
 manages.  Messages such as values and positions for the neighborhood best are
-sent between processors.  There may be (in fact there probably must be) some
-machine that collects information from all of the particles and outputs the
-result of the algorithm, though that machine's importance is much less than in
-centralized algorithms.
+sent between processors.  There may be some machine that collects information
+from all of the particles and outputs the result of the algorithm, though that
+machine's importance is much less than in centralized algorithms.
 
@@
 without giving the speculative particles the same treatment as actual
-particles.  Perhaps that can be done by assigning each particle a set of
-machines instead of a single machine, and the particle directs its extra
-machines to evaluate its speculative children.  In the MapReduce framework,
-each particle and speculative child has its own Map task, during which the
-particle is evaluated, with the speculative children sending their output back
-to the original particle's Reduce task and any neighbors it has.  The Reduce
-task for each particle gets all of the messages from its neighbors, its
-speculative children, and the speculative neighbors of its speculative
-children, and produces the next particle and children for another Map task.
-
-No matter which framework is used the message passing should be similar, and
-that is what we describe here.
+particles, as the speculative children only live for one iteration.  Perhaps
+that can be done by assigning each particle a set of machines instead of a
+single machine, and the particle directs its extra machines to evaluate its
+speculative children.  No matter the distributed framework used the message
+passing should be similar, and that is what we describe here.
 
@@
 treated specially, so each particle can produce its initial set of speculative
-children--neighbors need only pass their initial position.  This kind of
+children---neighbors need only pass their initial position.  This kind of
 message passing necessitates the careful use of random seeds, so that when each
@@
 the neighboring particle that we needed to produce the original particle,
-$\nonbest{\p}_{t+1}$--we need the original neighbor particle, its speculative
+$\nonbest{\p}_{t+1}$; we need the original neighbor particle, its speculative
 children, and its neighbors.  With that information, the set
@@
 It has been shown that in dynamic topologies using the $\nbest$ only from
-current neighbors leads to poor performance--$\nbest$ should instead be defined
-as the best position of any of the neighbors the particle has ever
+current neighbors leads to poor performance.  $\nbest$ should instead be
+defined as the best position of any of the neighbors the particle has ever
 had~\cite{mcnabb-cec09}.  The question of which value to share with neighbors
@@
 messages need to be sent to the correct neighbors.  A particle cannot simply
-send messages to its neighbors' neighbors' neighbors--it needs to send messages
-to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.  For
-every neighbor outward information is sent, the iteration also needs to be
+send messages to its neighbors' neighbors' neighbors---it needs to send
+messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
+For every neighbor outward information is sent, the iteration also needs to be
 incremented, as information about neighbors' neighbors is used during iteration
@@
 neighbors' personal bests are what are used to update a particle's neighborhood
-best.  That creates a problem for speculative evaluation--there are potentially
-more than $2(n+1)$ possible next positions, increasing the amount of work that
-must be done to perform the second iteration at the same time as the first.
+best.  That creates a problem for speculative evaluation---there are
+potentially more than $2(n+1)$ possible next positions, increasing the amount
+of work that must be done to perform the second iteration at the same time as
+the first.
 
@@
 created so it can give the right personal best and neighborhood best values and
-positions to $\p_{t+1}$.  Then only change needed in \alg{centralized} is in
+positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
 step 7, where the $\noeval{\s}_{t+1}$ with the highest value is chosen from
@@
 that SEPSO clearly beat PSO with a Random topology.  However, the performance
-between SEPSO and PSO Complete is not as encouraging--the results are
+between SEPSO and PSO Complete is not as encouraging---the results are
 indistinguishable.  But SEPSO Pick Best handily outperforms all other
@@
 
-To further belabor the point that performing speculative evaluation is a much
-better idea with some functions than simply adding particles, in \fig{sphere2}
-we show a graph comparing the performance of SEPSO Pick Best with 13 particles
-and with 50 particles.  The difference between the two in terms of function
-evaluations is staggering.
+To further illustrate the point that performing speculative evaluation is a
+much better idea with some functions than simply adding particles, in
+\fig{sphere2} we show a graph comparing the performance of SEPSO Pick Best with
+13 particles and with 50 particles.  The difference between the two in terms of
+function evaluations is staggering.  However, this result is well known; with
+Sphere and similar functions, the best thing to do is have a small swarm run
+longer.  We show the results here because our method allows the algorithm to
+use spare processors in parallel to perform more iterations at the same time.
+Clearly with these functions the best thing to do is not to add particles to
+the swarm, but to perform speculative evaluation.
 
@@
 can be modified to allow additional processors to increase the number of
-iterations of the algorithm performed, instead of merely add more particles to
-the swarm.  We have detailed several different implementations of the
+iterations of the algorithm performed, instead of merely adding more particles
+to the swarm.  We have detailed several different implementations of the
 algorithm, depending on the parallel PSO framework being used.  Using this
@@
 each level represent the possible next positions of the particle.  One need not
-simply evaluate all of one level for each particle--perhaps going two or three
+simply evaluate all of one level for each particle---perhaps going two or three
 iterations ahead on the most likely path would perform better.

### 2009-11-19 honors_thesis/thesis/sepsodata/plots/psodata.py
Some random things, mostly just because I moved the files

@@ -1,73 +1,0 @@
-
-from __future__ import division
-
-class Batch(object):
-    def __init__(self):
-        self.keys = []
-        self.map = {}
-
-    def add(self, key, value):
-        self.keys.append(key)
-        self.map[key] = value
-
-    def __getitem__(self, key):
-        return self.map[key]
-
-    def __iter__(self):
-        return iter(self.keys)
-
-    def last(self):
-        return self.keys[-1]
-
-    def lastitem(self):
-        key = self.last()
-        return key, self.map[key]
-
-
-class PSOData(object):
-    def __init__(self, infile):
-        self.headers = []
-        self.batches = []
-
-        for line in infile:
-            if line[0:7] == '# Batch':
-                self.batches.append(Batch())
-                continue
-            if line.isspace():
-                continue
-            if (line[0] == '#'):
-                if line.split()[1] not in ('Batch', 'DONE'):
-                    self.headers.append(line)
-                continue
-            iteration, value = line.split()
-            iteration = int(iteration)
-            value = float(value)
-            lastbatch = self.batches[-1]
-            lastbatch.add(iteration, value)
-
-    def __iter__(self):
-        return iter(self.batches)
-
-    def __len__(self):
-        return len(self.batches)
-
-    def average(self, key):
-        """Finds the average for a given key."""
-        values = [batch[key] for batch in self.batches]
-        return sum(values) / len(values)
-
-    def statistics(self, key, trim):
-        """Finds low, median, and high values after trimming outliers.
-        
-        Note that if trim is 2, and we have 20 samples, this is the 10th,
-        50th, and 90th percentiles.
-        """
-        values = [batch[key] for batch in self.batches]
-        values.sort()
-        trimmed = values[trim:-1-trim]
-        midpoint = int(len(values) / 2)
-        med = (values[midpoint] + values[-midpoint]) / 2
-        return trimmed[0], med, trimmed[-1]
-
-
-# vim: et sw=4 sts=4

### 2009-11-19 great_works/music_brahms.tex
Some initial work on writing more responses

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{Mendelssohn's Italian Symphony: A Music Response}
\author{Matthew Gardner}
\date{}

\begin{document}

\textbf{Music Response}

\begin{tabular}{ll}
  Name:&Matthew Gardner \\
  Work:&Mendelssohn's Symphony No. 4 (The Italian Symphony) \\
  Artist:&Felix Mendelssohn \\
  Date:&1833 \\
\end{tabular}

\section*{Historical Context}

\section*{Critical Analysis}

\section*{Personal Reflection}

\end{document}

### 2009-11-20 great_works/theater_eden.tex
Some initial work on writing more responses
Finished the response to Children of Eden

--- 2009-11-19 great_works/music_brahms.tex
@@
 \usepackage{graphicx}
+\usepackage{setspace}
+\doublespacing
 
-\title{Mendelssohn's Italian Symphony: A Music Response}
+\title{Children of Eden: A Theater Response}
 \author{Matthew Gardner}
@@
 
-\textbf{Music Response}
+\textbf{Theater Response}
 
@@
   Name:&Matthew Gardner \\
-  Work:&Mendelssohn's Symphony No. 4 (The Italian Symphony) \\
-  Artist:&Felix Mendelssohn \\
-  Date:&1833 \\
+  Play:&Children of Eden \\
+  Playwright:&Steven Schwartz \\
+  Date:&1991 \\
 \end{tabular}
@@
 
+``Children of Eden'' began as a production for a religious high school theater
+camp.  It was originally titled ``Family Tree'' and was a short play.  In 1990,
+Steven Schwartz, composer and lyricist of the recent hit musical ``Wicked,''
+took the original script and music of ``Family Tree'' and made it into a full
+length musical, renaming it to ``Children of Eden.''  The play depicts the
+early history of the world, starting with God creating the earth, along Adam
+and Eve, going through Adam and Eve's expulsion from the Garden of Eden, and
+ending in act one with Cain killing Abel.  Act two tells the story of Noah.
+The play ends with Noah and his family getting off of the ark and singing
+praise to God.
+
+``Children of Eden'' opened in London in 1991.  It received poor reviews and
+only stayed open for three months.  Since 1991, the show has never made it to
+Broadway, though it has been performed by amateurs and professionals all over
+the country.  Its appeal comes from its religious nature and its treatment of
+themes common to all people, such as love, anger, and family conflicts, as well
+as its ability to accommodate any reasonable cast size.
+
 \section*{Critical Analysis}
+
+Perhaps the central theme of the entire play is the interaction of parents with
+their children, particularly their disobedient children.  In the three separate
+stories told in the show, very similar dialogue is used between the father and
+his children.  Through these parallels, Schwartz explores the proper role of
+parents in the lives of their children, and how a father grows through his
+experience of being a parent.
+
+The first parent-child story told is that of God, called Father, and Adam and
+Eve.  God is the archetypical father, portrayed more as a human father than a
+God, lacking many of the qualities we normally ascribe to him.  He seems to be
+learning how he should be treating his children, and when they disobey him, Eve
+eating the fruit and Adam choosing to follow her, he becomes spiteful and
+petty.  He refuses to ever speak to Adam and Eve again, visiting their children
+while ignoring them.  At this point in the story, Father still has a lot to
+learn about being a father.
+
+The play then moves to Adam and Eve raising their two children, Cain and Abel.
+Cain sees a waterfall on a hill and asks his parents about it, using almost
+exactly the same words as Eve used when she asked Father about the Tree of
+Knowledge.  The parallel dialogue continues throughout the story; Schwartz
+clearly tries to set up a comparison between Eve and Cain, and Father and Adam.
+In this case again, Adam fails as a parent, Cain kills Abel, and sorrow comes
+to the family.
+
+Things look just as bleak when the story of Noah starts.  Noah's son, Japheth,
+has yet to choose a wife, and God forbids Noah from letting Japheth marry a
+woman of the race of Cain.  Japheth ends up choosing the serving girl, who is a
+descendant of Cain, and sneaking her on the ark, upsetting God and the whole
+family.  Again the dialogue between Japheth, the servant girl, and Noah is very
+similar to the parent-child dialogue in the earlier stories.  God, at this
+point the epitomy of a petty grudge-holder, literally turns his back on the
+world, including Noah.  But Noah realizes that his family is all that he has,
+and forgives Japheth and allows him to marry the servant girl.  Finally, a
+parent in the story learned to forgive, and apparently Noah taught God the same
+lesson, for they both sing about how the most important part of love is ``the
+letting go.''
+
+The parallels that Schwartz creates are clearly meant to be a commentary on the
+proper role of a parent in the life of his or her child.  Through the three
+stories that Schwartz uses, the audience can see God, the quintessential father
+figure, learning how to treat his children.  Likewise, we are meant to learn
+through Schwartz's examples how we can be better parents to our children.
 
@@
 
+My wife and I are expecting our first child in three months.  The opening scene
+of the play, portraying God very much as any loving earthly father, made me
+think a lot about how having children teaches us about God.  Perhaps that is
+one of the main reasons that God commands us to live in families and have
+children---it teaches us about how He feels towards us.  I didn't enjoy the
+rest of the play nearly as much, as I don't believe God to be so petty (and my
+wife didn't appreciate the parallel between Eve and Cain).  But when I realized
+that Schwartz was commenting more on how we should treat our children than the
+specifics of a perfect God's interaction with man, I gained more appreciation
+for the story that was told.
+
 \end{document}

### 2009-12-03 honors_thesis/thesis/paper.tex
I'm pretty much done with the GECCO paper.
I significantly changed the results section, I added a number of
figures using xfig that really help out the explanation, and I
split the second distributed method into its own subsection.

@@
 \usepackage{algorithmic}
+\usepackage{psfrag}
 
@@
 \providecommand{\sset}{\ensuremath{\mathbf{s}}}
+\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
 \providecommand{\n}{\ensuremath{n}}
 \providecommand{\nset}{\ensuremath{\mathbf{n}}}
+\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}
 
@@
 $2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
-neighbors the particle has, because there are two possibilities for updating a
-particle's personal best (updated and not updated) and $n+1$ possibilities for
-updating its neighborhood best (one of $n$ neighbors, or no update).  If all of
-those speculative positions are evaluated in parallel along with the original,
-non-speculative position of the particle at time $t-1$, then one of the
-speculative positions has to be the correct one, and two iterations have been
-evaluated at once.
+neighbors the particle has.  If a particle is also its own neighbor, as is most
+often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
+a particle cannot be the source of its $\nbest$ update while not updating its
+$\pbest$.  Table~\ref{tab:evals} lists all possible combinations of updates
+that can be seen by a particle with two neighbors other than itself, which is
+the case in all of the topologies we use.  If all of these speculative
+positions are evaluated in parallel along with the original, non-speculative
+position of the particle at time $t-1$, then one of the speculative positions
+has to be the correct one, and two iterations have been evaluated at once.
+
+\begin{table}
+  \caption{All possible updates for a particle with two neighbors.}
+  \label{tab:evals}
+  \centering
+  \begin{tabular}{c|c}
+	$\pbest$ updated&Source of $\nbest$ update\\
+	\hline
+	No update&No update\\
+	Updated&No update\\
+	No update&Neighbor 1\\
+	No update&Neighbor 2\\
+	Updated&The particle itself\\
+	Updated&Neigbhor 1\\
+	Updated&Neighbor 2\\
+  \end{tabular}
+\end{table}
 
@@
 states that a particle could have.  We use $\p_t$ to denote a particle at
-iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children.
-$\n_t$ is a neighbor of particle $\p_t$.  A set of particles is $\pset$,
-$\sset$, or $\nset$, whereas a single particle is simply $\p$, $\s$, or $\n$.
+iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
+corresponding to one of the rows in Table~\ref{tab:evals}.  $\n_t$ is a
+neighbor of particle $\p_t$.  A set of particles is $\pset$, $\sset$, or
+$\nset$, whereas a single particle is simply $\p$, $\s$, or $\n$.
 
@@
 
+\begin{algorithm}
+  \caption{Speculative Evaluation in a Centralized PSO}
+  \label{alg:centralized}
+  \begin{algorithmic}[1]
+	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$
+	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
+	  generate $\noeval{\sset}_{t+1}$
+	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
+	\STATE Update personal best for each $\noeval{\p}_t$ and
+	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
+	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
+	  $\pset_t$
+	\FORALL{$\p_t$}
+	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
+	  the branch taken by $\p_t$
+	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
+	  making $\nonbest{\p}_{t+1}$
+	\ENDFOR
+	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
+	  $\pset_{t+1}$
+	\STATE Repeat
+  \end{algorithmic}
+\end{algorithm}
+
 In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
@@
 position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
-and $\nbest$ that particle $\noeval{\p}_t$ could end up with.  Each of those
-combinations defines a speculative particle, $\s_t$, that can be moved using
-the motion equations to obtain $\noeval{\s}_{t+1}$.  The result is a set of
-particles $\noeval{\pset}_t$, and for each particle a set of speculative
-children $\noeval{\sset}_{t+1}$, which can all be evaluated at once.
+and $\nbest$ that particle $\noeval{\p}_t$ could end up with (as shown in
+Table~\ref{tab:evals}).  Each of those combinations defines a speculative
+particle, $\s_t$, that can be moved using the motion equations to obtain
+$\noeval{\s}_{t+1}$.  The result is a set of particles $\noeval{\pset}_t$, and
+for each particle a set of speculative children $\noeval{\sset}_{t+1}$, which
+can all be evaluated at once.
 
@@
 
-\begin{algorithm}
-  \caption{Speculative Evaluation in a Centralized PSO}
-  \label{alg:centralized}
-  \begin{algorithmic}[1]
-	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$
-	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
-	  generate $\noeval{\sset}_{t+1}$
-	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
-	\STATE Update personal best for each $\noeval{\p}_t$ and
-	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
-	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
-	  $\pset_t$
-	\FORALL{$\p_t$}
-	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
-	  the branch taken by $\p_t$
-	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
-	  making $\nonbest{\p}_{t+1}$
-	\ENDFOR
-	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
-	  $\pset_{t+1}$
-	\STATE Repeat
-  \end{algorithmic}
-\end{algorithm}
-
 \subsection{Distributed Algorithms}
@@
 
-This can be accomplished in two ways, either by performing several rounds of
-very simple message passing, or by performing a single round with a lot more
-messages.
+To get that information, a round of message passing is required.  Each particle
+sends its position to its neighbors at iteration $t$, so that all particles can
+generate $\noeval{\sset}_{t+1}$.  After each particle evaluates its position
+and the positions of its speculative children, it passes information about the
+outcome iteration $t$ to its neighbors, so that each particle can receive
+information about its neighborhood best to be updated to $\p_t$.  Once that
+communication is finished, the particle can select the speculative child which
+matched the branch that iteration $t$ actually produced.  Then another round of
+information passing follows, for iteration $t+1$, so that $\nonbest{\p}_{t+1}$
+can be updated to $\p_{t+1}$.  Two iterations have then been completed with
+only one round of evaluations, and the next iteration can start again with the
+first round of message passing.
 
-The first method takes three rounds of passing information, two for iteration
-$t$ and one for iteration $t+1$.  First, each particle must be moved from
-iteration $t-1$ to $\noeval{\p}_t$.  Then a round of message passing is
-required so that each particle knows the position of its neighbors at iteration
-$t$, from which $\noeval{\sset}_{t+1}$ can be generated.  After each particle
-evaluates its position and the positions of its speculative children, it passes
-information about the outcome iteration $t$ to its neighbors, so that each
-particle can receive information about its neighborhood best to be updated to
-$\p_t$.  Once that communication is finished, the particle can select the
-speculative child which matched the branch that iteration $t$ actually
-produced.  Then another round of information passing follows, for iteration
-$t+1$, so that $\nonbest{\p}_{t+1}$ can be updated to $\p_{t+1}$.  Two
-iterations have then been completed with only one round of evaluations, and the
-next iteration can start again with the first round of message passing.
+\subsection{Alternate form of message passing}
 
-Rounds of message passing can be expensive.  Accordingly, we describe a method
-that requires only one round of communication for each pair of iterations.  A
-lot more messages are needed, but that is sometimes more desirable than
-synchronizing all of the machines three times.
+In some distributed frameworks, like the MapReduce framework that we used in
+our experiments, rounds of message passing can be expensive.  The method
+just described uses three rounds of message passing for every two iterations.
+Accordingly, we describe a method that requires only one round of communication
+for each pair of iterations.  A lot more messages are needed, but that is
+sometimes more desirable than synchronizing all of the machines three times.
 
@@
 then, must send it a message, so that from their updated personal best at
-iteration $t$ the particle becomes $\p_t$.
+iteration $t$ the particle becomes $\p_t$.  The work done with the messages
+received thus far is just as in regular PSO, and is graphically depicted in 
+\fig{messages1}.
+
+\begin{figure}
+  \centering
+  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
+  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
+  \psfrag{pt}[C][C]{$\p_{t}$}
+  \includegraphics[width=.3\columnwidth]{messages1.eps}
+  \caption{The messages needed to produce $\p_{t}$.  Items without an arrow
+  pointing to them are messages the particle needs to receive.}
+  \label{fig:messages1}
+\end{figure}
 
 With $\p_t$ we can select the correct speculative child as described above and
-produce $\nonbest{\p}_{t+1}$.  We then need the set of neighbors to
-$\nonbest{\p}_{t+1}$, $\nonbest{\nset}_{t+1}$, so we can update
-$\nonbest{\p}_{t+1}$'s neighborhood best.
+produce $\nonbest{\p}_{t+1}$.  Again we show the use of messages thus far
+graphically, in \fig{messages2}.
 
-To produce each neighbor $\nonbest{\n}_{t+1}$, we need the same information for
-the neighboring particle that we needed to produce the original particle,
-$\nonbest{\p}_{t+1}$; we need the original neighbor particle, its speculative
-children, and its neighbors.  With that information, the set
+\begin{figure}
+  \centering
+  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
+  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
+  \psfrag{pt}[C][C]{$\p_{t}$}
+  \psfrag{pt1-n}[C][C]{$\nonbest{\p}_{t+1}$}
+  \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
+  \includegraphics[width=.5\columnwidth]{messages2.eps}
+  \caption{The messages needed to produce $\nonbest{\p}_{t+1}$.  Items without
+  an arrow pointing to them are messages the particle needs to receive.}
+  \label{fig:messages2}
+\end{figure}
+
+We then need the set of neighbors to $\nonbest{\p}_{t+1}$,
+$\nonbest{\nset}_{t+1}$, so we can update $\nonbest{\p}_{t+1}$'s neighborhood
+best.  To produce each neighbor $\nonbest{\n}_{t+1}$, we need the same
+information for the neighboring particle that we needed to produce the original
+particle, $\nonbest{\p}_{t+1}$; we need the original neighbor particle, its
+speculative children, and its neighbors.  With that information, the set
 $\nonbest{\nset}_{t+1}$ can be obtained by following the same process used to
-obtain $\nonbest{\p}_{t+1}$.  With $\nonbest{\nset}_{t+1}$ and
-$\nonbest{\p}_{t+1}$, we can produce $\p_{t+1}$.
+obtain $\nonbest{\p}_{t+1}$.  We graphically show the messages needed to
+produce $\nonbest{\p}_{t+1}$ in \fig{messages3}.  Note that it looks identical
+to \fig{messages2}, just with different sets of particles.
+
+\begin{figure}
+  \centering
+  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
+  \psfrag{nnt-n}[C][C]{$\nonbest{\nnset}_{t}$}
+  \psfrag{nt}[C][C]{$\nset_{t}$}
+  \psfrag{nt1-n}[C][C]{$\nonbest{\nset}_{t+1}$}
+  \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
+  \includegraphics[width=.6\columnwidth]{messages3.eps}
+  \caption{The messages needed to produce $\nonbest{\nset}_{t+1}$.  Items
+  without an arrow pointing to them are messages the particle needs to receive.
+  $\nnset$ is the set of neighbors for each particle $\n$, and $\nsset$ is the
+  set of $\n$'s speculative children.  Note the similarity between this and
+  \fig{messages2}.}
+  \label{fig:messages3}
+\end{figure}
+
+With $\nonbest{\nset}_{t+1}$ and $\nonbest{\p}_{t+1}$, we can produce
+$\p_{t+1}$.  This is shown in \fig{messages4}.  Note that we just combined
+Figures~\ref{fig:messages2}~and~\ref{fig:messages3}, putting them together
+to make $\p_{t+1}$, as all the particle needs is it's neighborhood best to be
+updated.
+
+\begin{figure}
+  \centering
+  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
+  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
+  \psfrag{pt}[C][C]{$\p_{t}$}
+  \psfrag{pt1-n}[C][C]{$\nonbest{\p}_{t+1}$}
+  \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
+  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
+  \psfrag{nnt-n}[C][C]{$\nonbest{\nnset}_{t}$}
+  \psfrag{nt}[C][C]{$\nset_{t}$}
+  \psfrag{nt1-n}[C][C]{$\nonbest{\nset}_{t+1}$}
+  \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
+  \psfrag{pt1}[C][C]{$\p_{t+1}$}
+  \includegraphics[width=.8\columnwidth]{messages4.eps}
+  \caption{The messages needed to produce $\p_{t+1}$.  Items
+  without an arrow pointing to them are messages the particle needs to receive.
+  Note that this is just a combination of \fig{messages2} and \fig{messages3}.}
+  \label{fig:messages4}
+\end{figure}
 
@@
 neighbors, its neighbors' neighbors, its speculative children, and its
-neighbors' speculative children.  \fig{origparticle} shows graphically how this
-information is used to produce $\p_{t+1}$.  The particle $\p_{t+1}$ can be
-passed to some central machine to track the progress of the algorithm, and it
-can be moved to $\noeval{\p}_{t+2}$ in order to start the next iteration.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\columnwidth]{messages.eps}
-  \caption{The messages needed to produce $\p_{t+1}$. $nn$ are the neighbors
-  of $p$'s neighbors, and $ns$ are the speculative children of $n$.  Items
-  without an arrow pointing to them are messages the particle needs to receive.}
-  \label{fig:origparticle}
-\end{figure}
+neighbors' speculative children.  The particle $\p_{t+1}$ can be passed to some
+central machine to track the progress of the algorithm, and it can be moved to
+$\noeval{\p}_{t+2}$ in order to start the next iteration.
 
@@
 messages, except for the neighbor particles instead of the particle itself.  
-\fig{specparticles} shows graphically how $\noeval{\sset}_{t+3}$ is produced.
+\fig{messages5} shows graphically how $\noeval{\sset}_{t+3}$ is produced.
 
@@
   \centering
-  \includegraphics[width=.6\columnwidth]{messages2.eps}
+  \psfrag{pt1}[C][C]{$\p_{t+1}$}
+  \psfrag{nt1}[C][C]{$\nset_{t+1}$}
+  \psfrag{nt2-e}[C][C]{$\noeval{\nset}_{t+2}$}
+  \psfrag{pt2-e}[C][C]{$\noeval{\p}_{t+2}$}
+  \psfrag{st3-e}[C][C]{$\noeval{\sset}_{t+3}$}
+  \includegraphics[width=.5\columnwidth]{messages5.eps}
   \caption{The messages needed to produce $\noeval{\sset}_{t+3}$. $\p_{t+1}$
-  and $\nset_{t+1}$ are both produced as in \fig{origparticle}.}
-  \label{fig:specparticles}
+  and $\nset_{t+1}$ are both produced as in \fig{messages4}.}
+  \label{fig:messages5}
 \end{figure}
@@
 
-The decision between these two methods of performing speculative evaluation in
-a distributed parallel PSO algorithm depends on which is more expensive: 
-individual messages between processors or rounds of communication.
+This may seem like an inordinate amount of work, and in some instances it is.
+However, some parallel frameworks necessitate this type of message passing, so
+we showed how speculative evaluation can be performed in such frameworks.
 
@@
 
-The original PSO algorithm was developed with only static topologies in mind. A
-few of the details about the neighborhood best $\nbest$ were left unspecified,
-as they are not issues with static topologies.  However, when neighbors change
-from iteration to iteration a few questions arise.  Is $\nbest$ the best
-position of any of its all-time neighbors, or only its current neighbors?  Do
-particles share their current value, their personal best value, or their
-neighborhood best value to contribute to their neighbors' neighborhood best?
-
-It has been shown that in dynamic topologies using the $\nbest$ only from
-current neighbors leads to poor performance.  $\nbest$ should instead be
-defined as the best position of any of the neighbors the particle has ever
-had~\cite{mcnabb-cec09}.  The question of which value to share with neighbors
-is an open one.  Giving more information to a particle's neighbors shares
-information more quickly through the swarm.  We choose to share the personal
-best value as a good middle ground.  Also, sharing the neighborhood best would
-require more message passing in distributed algorithms, as particles would need
-more information in order to generate their speculative children from their
-neighbors.
-
-Performing speculative evaluation in PSO with a dynamic topology also raises a
-few issues of its own.  First, with neighbors changing each iteration, messages
+Performing speculative evaluation in PSO with a dynamic topology raises a few
+issues of its own.  First, with neighbors changing each iteration, messages
 that processors pass to their neighbors need to be sent to the correct
@@
 and accepting the best one.  Speculative particles that move in good directions
-are kept.  Clearly this modification to PSO would produce more exploitation,
+are kept.  It seems this modification to PSO would produce more exploitation,
 decreasing the amount of exploration.  With functions that are less deceptive
@@
 In our experiments we compared our speculative PSO algorithm to the standard
-PSO algorithm.  To have the same number of function evaluations per iteration
-it is necessary to use a smaller swarm size in the speculative algorithm.
-There are $2(n+1)$ speculative evaluations per particle for every two
-iterations, plus one for the original, non-speculative particle.  For each of
-the topologies we used, a particle has three neighbors, including itself.  One
-of the $2(n+1)$ cases can be eliminated, however, as the particle cannot be its
-own neighborhood best update while not also updating its personal best
-(otherwise its personal best would be the neighborhood best, and it would not
-have updated).  So we have $2(n+1)$ total evaluations per particle, giving us
-$8p_s$ evaluations for every two iterations, where $p_s$ is the number of
-particles in the speculative swarm.  A regular swarm uses $2p_r$ evaluations
-per two iterations.  To have the same number of function evaluations per
-iteration, we have that $p_s = \frac{1}{4}p_r$.
+PSO algorithm.  To use the same number of processors for each algorithm,
+speculative evaluation needs a smaller swarm size than standard PSO.  For each
+of the topologies we used, a particle has two neighbors other than itself.  As
+shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
+particle.  With one evaluation needed for the original, non-speculative
+particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
+the number of particles in the speculative swarm.  The extra evaluations done
+in speculative evaluation would instead be particles in regular PSO, so we
+compare swarms of size $p_s$ in speculative evaluation with swarms of size
+$8p_s$ in regular PSO.
 
@@
 evaluation were not being performed, it is likely that the Complete topology
-would be used.  To be fair in our comparisons, for the standard PSO algorithm
-we compare to both the Random topology with two neighbors and to the Complete
-topology.  All functions have 20 dimensions, and all results shown are averages
-over 20 runs.  Except where noted, we use swarms of close to 50 particles, as
-that has been recommended as a general-purpose swarm size~\cite{bratton-sis07}.
-When the speculative swarms have 50 particles, regular PSO swarms have 200.
-For ease of labeling, we call our speculative algorithm SEPSO, the variant
-SEPSO Pick Best, and standard PSO just PSO, with the topology following if
-necessary.
+would be used, ignoring the communication cost of Complete over Random.  To be
+fair in our comparisons, for the standard PSO algorithm we compare to both the
+Random topology with two neighbors and to the Complete topology.  All functions
+have 20 dimensions, and all results shown are averages over 20 runs.  We use
+240 processors in our experiments, so speculative algorithms use 30 particles
+and standard PSO uses 240 particles.  For ease of labeling, we call our
+speculative algorithm SEPSO, the variant SEPSO Pick Best, and standard PSO just
+PSO.
 
@@
 small Complete swarm in \fig{sphere1}.  In comparing SEPSO with PSO, we can see
-that SEPSO clearly beat PSO with a Random topology.  However, the performance
-between SEPSO and PSO Complete is not as encouraging---the results are
-indistinguishable.  But SEPSO Pick Best handily outperforms all other
-algorithms in this problem.
+that SEPSO clearly beats PSO with a Random topology.  However, the performance
+between SEPSO and PSO Complete is not as encouraging---PSO Complete performs
+slightly better.  But SEPSO Pick Best handily outperforms all other algorithms
+in this problem.
 
 \begin{figure}
+  \centering
   \includegraphics[width=.8\columnwidth]{sphere1.eps}
-  \caption{Function Sphere with a swarm that performs 104 evaluations per two
-  iterations.  Speculative algorithms have a swarm size of 13, and the original
-  PSO has a swarm size of 52.}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
+  iterations per round, while standard PSO algorithms use 240 processors and do
+  one iteration per round.}
   \label{fig:sphere1}
-\end{figure}
-
-To further illustrate the point that performing speculative evaluation is a
-much better idea with some functions than simply adding particles, in
-\fig{sphere2} we show a graph comparing the performance of SEPSO Pick Best with
-13 particles and with 50 particles.  The difference between the two in terms of
-function evaluations is staggering.  However, this result is well known; with
-Sphere and similar functions, the best thing to do is have a small swarm run
-longer.  We show the results here because our method allows the algorithm to
-use spare processors in parallel to perform more iterations at the same time.
-Clearly with these functions the best thing to do is not to add particles to
-the swarm, but to perform speculative evaluation.
-
-\begin{figure}
-  \includegraphics[width=.8\columnwidth]{sphere2.eps}
-  \caption{Function Sphere with SEPSO Pick Best in two different swarm sizes.
-  The difference in performance is incredible and suggests that for this
-  function processors should most certainly be used to increase the number of
-  iterations performed and not add more particles to the swarm.}
-  \label{fig:sphere2}
 \end{figure}
@@
 optimizing the Griewank function, as Complete is prone to premature convergence
-on a local optima.  We show results in \fig{griewank1} for a swarm size of 50
-particles with SEPSO.  One can see in the figure that both SEPSO algorithms
-initially outperformed PSO, but got stuck in a local optima and were eventually
-outperformed.
+on a local optima.  We show results in \fig{griewank1} for swarms of size 30
+and 240 using the Ring topology. One can see in the figure that both SEPSO
+algorithms initially outperformed PSO, but got stuck in a local optima and were
+eventually outperformed.
 
 \begin{figure}
+  \centering
   \includegraphics[width=.8\columnwidth]{griewank1.eps}
-  \caption{Function Griewank with a swarm that performs 400 evaluations per two
-  iterations.  Speculative algorithms have a swarm size of 50, and the original
-  PSO has a swarm size of 200.}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
+  iterations per round, while standard PSO algorithms use 240 processors and do
+  one iteration per round.}
   \label{fig:griewank1}
@@
 
-It has also been shown that with Griewank, swarms that pass a threshold of size
-consistently find the global optimum.  That size is about 400
-particles~\cite{mcnabb-cec09}.  Thus we also show results for a swarm size of
-400 with SEPSO in \fig{griewank2}.  Once the speculative swarm is of a
-sufficient size to not get caught in the local optima, it clearly outperforms
-standard PSO, finding the global optimum in fewer function evaluations.
+\fig{griewank1} is slightly deceptive, however, as with Griewank some of the
+runs found the global optimum, 0, while some did not.  A more accurate plot for
+this function is one showing what percentage of the runs have found the global
+optimum at each iteration.  \fig{griewank2} shows such a plot.  A swarm of size
+30 performing the PSO algorithm gets stuck a little less than half of the time,
+while 240 particles provides enough exploration to find the global optimum
+every time.  Surprisingly, using the SEPSO Pick Best algorithm with 30
+particles greatly improves the chance of success with Griewank, while also
+finding the global optimum in far fewer rounds of evaluations than standard
+PSO.
 
 \begin{figure}
+  \centering
   \includegraphics[width=.8\columnwidth]{griewank2.eps}
-  \caption{Function Griewank with a swarm that performs 3200 evaluations per
-  two iterations.  Speculative algorithms have a swarm size of 400, and the
-  original PSO has a swarm size of 1600.  The local optima with the lowest
-  function value is close to $10^{-6}$.  Thus both algorithms consistently find
-  the global optimum without getting stuck.}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations.  Instead of showing average function value, we show the
+  percentage of runs that found the global optimum at each iteration.}
   \label{fig:griewank2}
@@
 
-Lastly we look at functions that do not show much or any improvement with this
-method.  The Ackley function is defined as $f(\Vec{x}) = -20\exp\left\{
--0.2\sqrt{\frac{1}{D}\sum_{i=1}^Dx_i^2}\right\} - \exp\left\{\frac{1}{D}
-\sum_{i=1}^D\cos(2\pi x_i)\right\}+20+e$. \fig{ackley} shows a comparison of
-our speculative algorithms versus the original PSO with swarm sizes of 50 and
-200, respectively.  There is no significant difference between the performance
-of SEPSO and PSO Complete, though the performance of SEPSO Pick Best is
-interesting to note.  It quickly converged on a local optimum and compared to
-the other algorithms had dismal performance on this function.
-
-We also ran experiments with the function Rastrigin, $f(\Vec{x}) = \sum_{i=1}^D
-\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$,  though space does not
-permit us to present a graph of our results.  However, performance of SEPSO and
-PSO Complete was largely indistinguishable, as it was with Ackley.
+Lastly we look at a function that does not show much or any improvement with
+this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
+\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
+with Rastrigin, the more particles there are in the swarm, the lower function
+value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
+swarms get caught in local optima.  Because our speculative algorithms have a
+significantly smaller swarm size, they get stuck at higher values while the
+larger swarms performing regular PSO continue to improve the best value found.
+\fig{rastrigin} shows the results graphically.
 
 \begin{figure}
-  \includegraphics[width=.8\columnwidth]{ackley.eps}
-  \caption{Function Ackley with a swarm that performs 400 evaluations per
-  two iterations.  Speculative algorithms have a swarm size of 50, and the
-  original PSO has a swarm size of 200.  Note the particularly horrible
-  performance of SEPSO Pick Best.}
-  \label{fig:ackley}
+  \centering
+  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
+  \caption{Function Rastrigin with a swarm that uses 240 processors per round
+  of evaluations.  Speculative algorithms have a swarm of 30 particles and do
+  two iterations per round, while standard PSO algorithms use 240 processors
+  and do one iteration per round.}
+  \label{fig:rastrigin}
 \end{figure}
@@
 to the swarm.  We have detailed several different implementations of the
-algorithm, depending on the parallel PSO framework being used.  Using this
-algorithm, the original PSO algorithm is exactly reproduced two iterations at a
-time.  This technique requires more function evaluations per iteration than
-regular PSO, but for a number of functions still performs better.
+algorithm, depending on the parallel PSO framework being used.  Using our
+modifications, the original PSO algorithm is exactly reproduced two iterations
+at a time.  This technique requires more function evaluations per iteration
+than regular PSO, but for some functions still performs better when run in a
+parallel environment.
 
 We have also described a method for making use of extra speculative evaluations
-that performs incredibly well on some functions and incredibly poorly on other
-functions.
+that performs incredibly well on some functions.
 
@@
 additional particles to the swarm is akin to increasing its capacity for
-exploration, and it has been shown that for a 20-dimensional Griewank function,
-400 particles is ``enough'' exploration~\cite{mcnabb-cec09}.  For these
-functions, one should use processors to increase the swarm size until enough
-exploration has been reached, then use additional processors to increase the
-number of iterations performed.
+exploration.  We have seen that for a 20-dimensional Griewank function, a
+number of particles greater than 30 but less than 240 has ``enough''
+exploration and finds the optimal value every time.  For these functions, one
+should use processors to increase the swarm size until enough exploration has
+been reached, then use additional processors to increase the number of
+iterations performed.
 

### 2009-12-03 great_works/science_lda.tex
Finished another Great Works response, a science one on LDA

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{setspace}
\doublespacing

\title{}
\author{Matthew Gardner}
\date{}

\begin{document}

\textbf{Physical Science Response}

\begin{tabular}{ll}
  Name:&Matthew Gardner \\
  Work:&Computer Science Colloquium---Latent Dirichlet Allocation \\
  Speaker:&Charles Elkan \\
  Date:&October 1, 2009 \\
\end{tabular}

\section*{Historical Context}

Though still largely in its infancy, the field of natural language processing
in computer science has flowered in recent years.  Natural language processing
is using a computer to understand, analyze, or otherwise process natural spoken
or written language.  Companies like Google use natural language processing
when trying to understand a search query or when trying to decide whether any
particular email is spam or not.  Machine translation systems also fall under
the heading of natural language processing.  As little as 10 years ago,
attempts to build search engines or machine translation systems were faltering
at best.  Today, the ability of computers to analyze large collections of
documents and perform important tasks with the results of the analysis is very
impressive.  However, a lot of work still needs to be done before we can claim
that computers completely understand natural language.

\section*{In-Depth Analysis}

There are some common tasks that are often done with large collections of
documents.  The first is called classification.  When performing
classification, each document has some kind of label, such as ``spam'' or ``not
spam'' if we are dealing with emails.  The task of classification is to use the
labeled data, often called ``training data,'' to learn some way of
distinguishing between the various labels.  With this method, often a
statistical model, the computer can then decide what label to give to a new
document that was not part of the training documents and does not have a label.
In this way, Google can analyze a lot of emails that have been labeled as being
spam and figure out common characteristics, and use those characteristics to
implement a spam filter.

A slightly more complicated task is that of clustering.  The clustering task
is very similar to the classification task, except it is done without labels.
The computer analyzes all of the documents and groups them into some number of
``clusters,'' where each cluster represents what would have been a label in the
classification task.  Thus without any information except that contained in the
documents themselves, the computer can decide which documents are similar to
each other and which are different.  For example, a researcher could take a
large collection of news articles on different topics and use this method to 
determine which articles talk about the same or similar themes.

A limitation of clustering is that it assumes that each document fundamentally
is about one topic, and tries to figure out which topic each document belongs
to.  What about documents that are about more than one topic, such as faith and
repentance?  If there is a lot of overlap between topics that the documents
talk about, the performance of clustering methods degrades rather quickly.  To
get around this issue, a technique called Latent Dirichlet Allocation, or LDA,
was introduced in 2002.  LDA breaks the assumption that each document comes
from a single topic and instead assumes that each word in each document comes
from a particular topic, allowing documents to contain many topics.  Performing
LDA on a collection of documents will produce a set of topics that contain
similar words and information about how often each document uses words from
each topic.  A good example again is a collection of news articles.  Suppose
the articles talked about Iraq, Vietnam, Health Care, the President of the
United States, and Congress.  Likely, there would be a topic containing words
used in politics, one with words used when talking about war, one about
doctors, and so on.  Each of the articles would have some proportion of words
from each topic, and someone looking for articles that talked about both
politics and health care would have an easy time finding them.

Natural language processing has shown to be very useful when trying to analyze
and perform tasks with large collections of documents.  LDA is the most
advanced and most recently developed of these techniques, and it is still being
improved.  This lecture talked about one particular issue with LDA and how it 
can be addressed.  The problem was that of ``burstiness'' in natural languages,
the fact that if a rare word is used once in a document, the probability of
seeing it again is very high.  Charles Elkan solved the problem by using a
different probability distribution when training his LDA model.  He showed
results that seemed to be promising.

\section*{Personal Reflection}

I first learned about LDA in one of my computer science classes.  This lecture
introduced me to some techniques that I hadn't considered before, and I quite
enjoyed the lecture.  I did a project for a class using LDA where I used it to
analyze the set of talks from LDS General Conferences from 1900 to the present.
The results I saw were fascinating---there was a topic about tithing, one about
welfare, one about Relief Society, one about Sunday School, and many more.  And
the proportions of those topics in the talks over time that LDA found matched
very well with real-world events.  For example, the tithing topic spiked in
1900, when Lorenzo Snow was president of the church and focused heavily on
paying a full tithe, and has since leveled off.  The Relief Society topic was
almost non-existent in the talks until the 1970s, when women started speaking
in General Conference.  I gained an appreciation for the technique through the
experiments I did, and I'm planning on doing graduate work in natural language
processing because of how much I enjoyed using LDA and other algorithms.

\end{document}

### 2009-12-05 great_works/film_sweetnothing.tex
Finished another great works response.  Film this time.  It's not very good...

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{setspace}
\doublespacing

\title{}
\author{Matthew Gardner}
\date{}

\begin{document}

\textbf{Film Response}

\begin{tabular}{ll}
  Name:&Matthew Gardner \\
  Film:&\emph{Sweet Nothing in My Ear} \\
  Production Year:&2008 \\
  Language:&English and American Sign Language \\
  Director:&Joseph Sargent \\
  Country of Origin:& USA \\
  Location:&International Cinema \\
  Date Attended:&11/18/2009 \\
\end{tabular}

\section*{Historical Context}

\emph{Sweet Nothing in My Ear} debuted in 2008 as a Hallmark Hall of Fame
movie.  The movie is based on a play by the same name written by Stephen Sachs
in 1998.  It is a story about a married couple, Dan and Laura Miller.  Laura is
deaf, as are her parents, and they are all very committed to deaf culture.
They have a son, Adam, who was born hearing, but gradually lost his hearing
after a few years.  When the father decides he wants to try giving his son a
cochlear implant, it causes marital problems, because the wife doesn't see
herself as handicapped and doesn't see anything wrong with her son.  They
separate, and the movie is shown as a series of flashbacks during a court
battle for custody of the boy.  The film focuses on the struggle of minority
groups and their attempts to define their identity in society.  The director,
Joseph Sargent, directed a movie with a similar theme 35 years ago titled
\emph{Love is Never Silent}, using some of the same actors.

\section*{Critical Analysis}

While the main point the director attempts to discuss is minority groups trying
to find their place in society, \emph{Sweet Nothing in My Ear} also highlights
the pain caused by pride, particularly in family struggles.  Throughout the 
movie we see children caught in the middle of fights between their parents,
damaging everyone involved.  We see this in Laura's parents, in harm caused to
Adam, and between Laura and Dan.

Laura's father, Max, was born deaf.  He was ridiculed as a child and appears to
have developed a complex because of it, thinking that being deaf is better than
being hearing.  He wrote a book on deaf pride and the deaf culture.  In one of
the most climactic scenes in the movie, he tells Laura that she wasn't actually
born deaf, something he had lied about to her all her life.  His pride in his
deafness had caused him to hide the truth from her, and she is deeply hurt when
he finally tells her she was born hearing and only gradually developed
deafness.

Much of the same pride seen in Max is found in his daughter Laura, and the
fights between Laura and her husband Dan leave their son Adam as the victim.
In scene after scene in the movie, we see Adam wondering why his parents are
fighting, wanting both of them to be together again, and wondering why his dad
isn't home with them.  In the climax of the movie, during Thanksgiving dinner
with his parents and Laura's parents, Adam tries to please his dad by speaking
to him.  He says ``Daddy'' in a halting voice that hasn't been used in years.
Dan is jubilant, but Laura and her parents are shocked, feeling betrayed.  The
phone rings, and Dan goes to get it, while Laura and her parents get up to talk
in another room.  In what to me was the most poignant scene in the movie, Adam
is left at the dinner table by himself, wondering what went wrong.  In the
power struggle between his parents, Adam is the one who is hurt the most.

While it was Adam who suffers the most from his parents' fighting, Laura and
Dan aren't left unscathed.  Their pride and unwillingness to come to a
compromise ruins their marriage and leaves them both miserable.  After deciding
without consulting his wife that he would look into getting Adam a cochlear
implant, Dan begins scheduling meetings for the two of them behind Laura's
back.  Neither one is willing to let go of what they want, and their arguments
get progressively more bitter.  It all culminates in a verbal fight they have
one night at home where Dan says to Laura, ``God gave me a hearing son!'' and
Laura asks why Dan thinks she isn't good enough because she can't hear.  In the
end of the movie, they both realize the harm they have caused each other and
their son with their pride, and Dan repentantly comes back to Laura, seeking
her forgiveness.

\emph{Sweet Nothing in My Ear} does a masterful job describing the conflict
between minority and majority groups.  Those conflicts are often caused by
pride, and Joseph Sargent poignantly portayed the damage done by parents in
their power struggles.  Family life can only be happy when parents are willing
to set aside their pride and do what's best for their family.

\section*{Personal Reflection}

I have had some experience with American Sign Language, though not very much
with deaf culture.  I had some friends in high school who took ASL classes, and
I taught a deaf man the gospel on my mission through an interpeter.  I have
also watched my sister teach her young children sign language so they can
communicate before they are able to speak.  It was fascinating to me to see
more of a world I have only had very limited contact with.  It made me
appreciate more the struggles that parents go through in trying to raise their
children, especially when disabilities are involved.  And as I am fairly
recently married and my wife is expected our first child, it made me think
about what it will be like to try to raise a family, and made me grateful that
my wife and I are able to peacefully resolve our conflicts.

\end{document}

### 2009-12-29 great_works/art_walter_wick.tex
Finished another film response, added some that I forgot to earlier

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{setspace}
\doublespacing

\title{}
\author{Matthew Gardner}
\date{}

\begin{document}

\textbf{Art Response}

\begin{tabular}{ll}
  Name:&Matthew Gardner \\
  Work:&Levers, Ramps and Pulleys \\
  Artist:&Walter Wick \\
  Date:&1995 \\
\end{tabular}

\section*{Historical Context}

Walter Wick is the artist and photographer who produced the popular \emph{I
Spy} books.  He was born in Hartford, Connecticut in 1953.  His early career
was in commercial photography, and he quickly ended up doing photo-illustration
for books and creating photographic puzzles and optical illusions.  The \emph{I
Spy} books stem from his work as a photographer.  For many of the pictures used
in his books, he created complex scenes, either of toys or of miniatures or
whatever he needed, and photographed them.  The series of eight books sold over
a million copies.  \emph{Levers, Ramps and Pulleys} was originally published in
\emph{I Spy: School Days}.

\section*{Critical Analysis}

\emph{Levers, Ramps and Pulleys} is a fun photograph of a balloon popper that
Walter Wick constructed.  Hitting on an oft-used theme, Wick creates his
balloon popper as an immensely complicated contraption, where starting one
little thing in motion causes another to move, and so on, until some result is
achieved, in this case popping a balloon.  Wick didn't just make any old
contraption for his balloon popper, however; he clearly understood the
principles of good art and used them in his piece.  The structure of the
popper, the extra items included in the picture, and its playful character all
give the piece an added depth and evoke a strong emotional response from the
viewer.

The contraption starts in the top left corner of the photograph.  Following the
path of the ball after it is set in motion leads one across the page and down,
following clear lines that draw the viewer's eyes from the beginning to the
balloon in the lower right corner.  While what's going on is somewhat
complicated, the general path the motion takes is easy to follow, and the
details give viewers places to pause and delight in Wick's ingenuity.  Also
cleverly placed at the conclusion of the motion is the ``scoreboard,'' showing
a tally of successful vs. failed attempts at popping the balloon.  The way the
photograph is structured shows the careful planning that went into its
production.

There is a lot going on in \emph{Levers, Ramps and Pulleys}.  A large portion
of the materials seen in the photograph are part of the balloon popper, but
there are also a number of toys and other objects that don't appear to have any
purpose in being there.  However, in addition to filling up some blank space
that would otherwise have been in the picture, they bring images to the viewer
of the process that Wick went through in creating his popper.  There are
scissors left open as if recently used to cut the pieces of string found in the
contraption.  The source of the sand used in the balloon popper is readily
found to be the large bucket of sand that is missing a scoop.  The tally in the
corner leaves one wondering if the other pieces were used in the failed
attempts, but were replaced and left aside when Wick found that they failed in
their part of the contraption.  Wick's inclusion of extra items in his piece
greatly enhances the character of the photograph.

Every child plays with toys, and there aren't many that do not line up dominoes
to knock over at some point in their lives.  Wick's use of common children's
toys in his contraption make the piece approachable by younger audiences and
adorable to adults.  For children, the photograph likely makes stirs their
imagination, making them think of all of the fun and interesting contraptions
\emph{they} could build with the toys they have.  In adults, the piece brings
back childhood memories, or perhaps recent memories of playing with their
children.  Whoever the audience, Wick's picture is evocative and fun.

\emph{Levers, Ramps and Pulleys} is a playful work of ingenuity.  Not only is
it entertaining to look at, it demonstrates Wick's understanding of his
audience, young and old, and his grasp of artistic principles.  It is a
masterpiece of modern photography.

\section*{Personal Reflection}

\emph{Levers, Ramps and Pulleys} was really fun for me to see.  I find Wick's
work fascinating.  I always enjoyed building things with Tinker Toys, blocks,
and dominoes, so the piece really resonated with me.  I have never been very
good at analyzing artwork, however, so trying to find something to write about
was difficult for me.  But I found that as I spent time thinking about
different characteristics of the piece, I appreciated it a lot more.  Noticing
the motion in the picture and thinking about the purpose of the extra elements
why the tally was there were very rewarding.  And in the extra reading I did
when preparing to write my response I found a video on the contraption working
on Walter Wick's website, walterwick.com.  That, of course, was the best part
of the whole process.

\begin{center}
  \includegraphics[width=.5\textwidth]{levers.jpg}
\end{center}

\end{document}

### 2009-12-29 great_works/film_bobby_fischer.tex
Finished another film response, added some that I forgot to earlier

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{setspace}
\doublespacing

\title{}
\author{Matthew Gardner}
\date{}

\begin{document}

\textbf{Film Response}

\begin{tabular}{ll}
  Name:&Matthew Gardner \\
  Film:&\emph{Searching for Bobby Fischer}\\
  Production Year:&1993 \\
  Language:&English \\
  Director:&Steven Zaillian \\
  Country of Origin:&USA \\
  Location:&HBLL \\
  Date Attended:&12/28/2009 \\
\end{tabular}

\section*{Historical Context}

\emph{Searching for Bobby Fischer} is based on a true story about chess prodigy
Josh Waitzkin.  Josh becomes interested in chess at age 6 when he starts
playing speed chess in a park.  He gets a professional teacher and enters
tournaments, winning the national junior championship at the end of the movie.
The title references the only American to ever have claimed the World Chess
Championship, Bobby Fischer.  Josh's teacher and others in the movie make
frequent reference to Fischer, and Fischer's attitude towards chess plays a
central role in the plot.  The movie is based on a book written by Josh's
father, who told the story of Josh's chess playing.

\section*{Critical Analysis}

In \emph{Searching for Bobby Fischer}, Zaillian sets up a series of conflicts
between life and chess, having fun and winning.  Throughout the movie, two of
the main characters, Josh's dad Fred and his teacher Bruce, go through a 
transformation from caring only about being the best to caring about people and
enjoying life.  We see this conflict and transformation in a number of scenes,
starting with Fred looking for a teacher for his son and ending with the 
national tournament.

Towards the beginning of the movie, Fred realizes that his son has a gift for
playing chess and seeks a teacher for him.  He finds Bruce, who initially tries
to dissuade Fred by showing him a tournament with a bunch of chess fanatics,
one of whom spends his whole life thinking about nothing but chess, and earns
\$2,000 a year.  Bruce's point is that to win at chess, you have to give your
life to it.  Fred is persuaded, and he takes that attitude throughout most of
the rest of the movie.

Fred enters Josh in tournaments, and Josh quickly becomes the top-rated junior
player in the state of New York.  The conflict raises its head again at an open
house for Josh's school, where his teacher talks to Fred about her concern that
Josh is falling behind in his school work and his friendships over ``this chess
thing.''  Fred is irate and storms out with Josh, offended enough that he moves
Josh to a private school where they have chess classes.  Chess, and his son
winning chess tournaments, has become Fred's obsession far more than Josh's.

The height of the conflict between winning and having fun comes during the
state finals.  Josh doesn't really want to go, because he's afraid he might
lose, but his dad tells him, ``You won't lose.''  He goes to the tournament and
loses, quickly, upon which his dad takes him to a rainy park and berates him
for throwing the match.  Josh's childhood is lost because all his dad cares
about is that he wins.

Shortly after the match, Bruce tries to get Josh to play more like Bobby
Fischer, holding the opponent in contempt.  Josh is an innocent kid who just
wants to have fun, and he doesn't like the idea of turning chess into a war.
Bruce pushes him, and Josh's mom kicks him out.  When Fred sides with Bruce,
she threatens divorce to get Josh out of a situation that would ruin him.  This
seems to be the turning point for Fred.

When Fred realizes that he has gone too far in his desire to see Josh win, he
starts taking things easy.  Josh still plays chess, but only with the people in
the park.  They go to the national championship, and though Josh wants to win,
he and his dad are not stressed out about it anymore.  Fred even takes Josh
fishing for two weeks before the tournament and tells Josh he's not allowed to
even think about chess.  Bruce even comes around in the end, coming to the
championship even though he initially told Fred that he wouldn't.  He
apologizes to Josh and wishes him the best.

In the beginning of the movie Bruce had told Josh that chess was an art to
Bobby Fischer.  It is to Josh too, and it is his love for the game, not his
love for winning, that ends up winning him the national tournament.  Much to
Bruce's chagrin, Josh uses a style that he learned playing speed chess in the
park in his final match---his art.  But he also utilized the analytical skills
taught him by Bruce.  When he sees that he has won the game, Josh offers his
opponent, a chess fanatic, a draw.  The other boy refuses, not realizing that
he has lost, and Josh wins the championship.

Though the movie ostensibly is about chess, the entire plot seems to center
around this struggle between enjoying life and wanting to win.  The director
clearly shows which he thinks is better by having the happy, loving, and
playful Josh win the final tournament over the boy who devotes his entire life
to chess.

\section*{Personal Reflection}

I really enjoyed watching \emph{Searching for Bobby Fischer}, for a number of
reasons.  First, I quite enjoy the game of chess, though I'm no prodigy, so the
subject matter was interesting to me.  Second, it made me think a lot about
parenting and the best way to raise children, something that is particularly
applicable to me at the moment because I will be a father in less than two
months.  And lastly, I have thought a lot about wanting to be the best, to win.
If you want to be the best in the world at anything, you pretty much have to
devote your life to it.  No matter what your innate capacity, someone with
similar capacity is going to be devoting their life to that thing, so in order
to compete with them you have to also devote similar time.  Is being the best
in the world, or even the best in your school, really worth that kind of time?
In almost every case, I don't think it is.  There are far too many things to do
in life to worry about being the best at any particular talent.  So, I always
try to do \emph{my} best, but I don't worry too much about being \emph{the}
best.

\end{document}

### 2009-12-29 great_works/film_sweetnothing.tex
Finished another film response, added some that I forgot to earlier

@@
 children, especially when disabilities are involved.  And as I am fairly
-recently married and my wife is expected our first child, it made me think
+recently married and my wife is expecting our first child, it made me think
 about what it will be like to try to raise a family, and made me grateful that

### 2009-12-29 great_works/theater_pimpernel.tex
Finished another film response, added some that I forgot to earlier

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{setspace}
\doublespacing

\title{}
\author{Matthew Gardner}
\date{}

\begin{document}

\textbf{Theater Response}

\begin{tabular}{ll}
  Name:&Matthew Gardner \\
  Play:&The Scarlet Pimpernel \\
  Playwright:&Frank Wildhorn \\
  Date Written:&1997 \\
  Date Seen:&September 1, 2009 \\
  Location Seen:&Hale Center Theater, Salt Lake \\
\end{tabular}

\section*{Historical Context}

\emph{The Scarlet Pimpernel} was originally a play Baroness Emmuska Orczy,
produced in 1903.  Orczy wrote a novel out of it two years later.  It has since
seen many film and television adaptations and was made into a Broadway musical
in 1997 with some minor changes to the storyline.  It is set in England and
France during the French Revolution.  Sir Percy Blakeney, an English nobleman,
marries a French actress, Marguerite St. Just.  Percy is disgusted by the
brutality of the French Revolution, and decides to begin a personal crusade
against it, calling himself the Scarlet Pimpernel.  Thinking his wife is a
French spy, he hides his identity from her and everyone else except his secret
band by acting like a fop.  He and his friends end up saving the lives of many
Frenchman and manage to frame Citizen Chauvelin, the French agent charged with
finding him, as the Scarlet Pimpernel.

\section*{Critical Analysis}

\emph{The Scarlet Pimpernel} is a very moving story.  The music and backdrop
for one of the opening songs, ``Madame Guillotine'' does an incredible job at
capturing the terror and brutality of the French Revolution.  Shortly
afterward, Percy sings a stirring challenge to those who become his band to not
sit idly by and let this great evil go unhindered.  There are some very powerful
moments in the play.  But woven throughout \emph{The Scarlet Pimpernel} is the
recurring use of disguises, both as a necessary protection to those fighting
evil and as the unfortunate consequence of a failure to communicate.

The play opens with Marguerite singing in her final show before leaving to be
married.  She sings of fairy tales and storybooks, setting up a theme that is
touched on throughout the play on two levels---she is acting, and she is acting
about false identities.  But her acting turns out to be her downfall.  Before
she met Percy, she had been a spy for Chauvelin during the beginnings of the
revolution.  Chauvelin uses that to blackmail her into betraying a friend of
both her and Percy, and when Percy learns of her betrayal he thinks she is
still a spy.  Thus Percy withdraws his love and hides his escapades from her,
leaving her lonely and confused about why her husband changed so dramatically.

Percy and his band also use disguises, though for different reasons.  In order
to protect themselves from the French they have to hide their true identities.
They do so in a very odd manner, pretending to be nincompoops and fops.  This
guise reaches its apex when they parade their fancy frillery in front of the
prince, singing, ``Each species needs a sex that's fated to be highly
decorated.  That is why the Lord created \emph{men}.''  Because they are so
focused on fashion, no one suspects that they really are the band of the
Scarlet Pimpernel.  They effectively use disguises to thwart many executions of
innocent people in France.

Percy's disguise does have its problems, however.  As mentioned earlier, Percy
alienates his wife with his disguises.  In the final song of Act I, ``The
Riddle,'' Percy, Marguerite and Chauvelin sing about the perils of having too
many disguises, not knowing who to trust.  ``And we all have so many faces,''
they sing, ``the real self often erases.  Enticing lies flicker through our
eyes.''  Percy wants to trust his wife, Marguerite is torn between trusting her
husband and going back to Chauvelin, Chauvelin wants to trust Marguerite, but
none of them can see through the disguises to the real motives underneath.

Hidden identities play a central role in \emph{The Scarlet Pimpernel}.  At
least five of the songs in the musical have disguise or concealment as their
focus.  Through this imagery Wildhorn attempts to convey the loss and confusion
that disguises can bring, in both good and bad ways.  Percy and Marguerite
eventually learn the truth about each other and are reconciled, but they lost a
lot of time together because of the barriers they built between each other.
And Chauvelin in the end gets the just reward of his treachery---Percy and his
gang capture him and leave him looking like the Scarlet Pimpernel, to be found
and surely executed by the leaders of the French Revolution.

\section*{Personal Reflection}

I first saw \emph{The Scarlet Pimpernel} when I was about 14, and I have loved
it ever since.  What was particularly striking to me when I saw it at the Hale
Center Theater was the feeling of sheer terror that the cast created during the
beginning, while singing ``Madame Guillotine.''  I was on the second row, and
they brought the ``Reign of Terror'' in the French Revolution to life.  It made
Percy's song about fighting injustice and not letting evil go unchallenged all
the more powerful to me, as it was a very real terror that they were fighting.
Though perhaps on a smaller scale, are we not to do the same in our lives
today?  I also was struck by all of the grief caused to both Percy and
Marguerite because they were quick to assume the worst in each other.  If they
had clearly communicated and resolved their differences, they would have had a
much happier marriage early on.  It is seeing tragedies like the unnecessary
sorrow between Percy and Marguerite that make me try especially hard to be open
and clear in my relationship with my wife.

\end{document}

### 2009-12-29 applications/byu_studies_lda.tex
A proposal for a paper at the BYU studies symposium on the LDA work I did

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{cite}

\title{Using Natural Language Processing Techniques\\To Automatically Analyze
General Conference Talks}
\author{Matthew Gardner and Eric Ringger}
\date{}

\begin{document}
\maketitle

In natural language processing, topic identification is the problem of finding
the underlying semantic meaning in a set of documents.  More practically, it is
the problem of finding what topics are discussed in a corpus and in what
proportion each document contains each of those topics.

Latent Dirichlet allocation (LDA) is a statistical model that is commonly used
to perform automatic topic identification.  It was first proposed in 2002 and
has been used with a number of different corpora since then~\cite{blei-jmlr03}.
In this paper we present the results of using LDA to analyze talks from the
past 100 years of LDS General Conferences.  We found our results to be
intriguing, correlating quite well in a number of instances with real-world
events.  The topics that were found by the algorithm included tithing, the
Sabbath day, Relief Society, the welfare program, and many more.  In looking at
the distribution of these topics over time we find some interesting patterns.
The tithing topic had its highest proportion in General Conference around the
year 1900, when President Lorenzo Snow made tithing a big focus.  The Relief
Society topic was almost completely absent until the 1970s, when women started
speaking in Conference.  The welfare topic had high points around the late
1930s and early 1940s (during the Great Depression) and in the 1970s, when a
separate welfare session was held for a few years.

This project has already been completed.  It is the first attempt of which we
are aware at using latent Dirichlet allocation or similar techniques to try to
extract meaning from General Conference talks.  We believe the results can be
used in a variety of ways, including for automatically discovering connections
between talks.

\bibliographystyle{plain}
\bibliography{/aml/home/mjg82/bib/nlp/lda/blei-jmlr03}

\end{document}

### 2009-12-29 resume_cv/cv.tex
Added an updated resume and cv

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% Simple LaTeX CV Template %%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% NOTE: If you find that it says                                     %%
%%                                                                    %%
%%                           1 of ??                                  %%
%%                                                                    %%
%% at the bottom of your first page, this means that the AUX file     %%
%% was not available when you ran LaTeX on this source. Simply RERUN  %%
%% LaTeX to get the ``??'' replaced with the number of the last page  %%
%% of the document. The AUX file will be generated on the first run   %%
%% of LaTeX and used on the second run to fill in all of the          %%
%% references.                                                        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%% Document Setup %%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Don't like 10pt? Try 11pt or 12pt
\documentclass[10pt]{article}

% This is a helpful package that puts math inside length specifications
\usepackage{calc}

% Simpler bibsection for CV sections
% (thanks to natbib for inspiration)
\makeatletter
\newlength{\bibhang}
\setlength{\bibhang}{1em}
\newlength{\bibsep}
 {\@listi \global\bibsep\itemsep \global\advance\bibsep by\parsep}
\newenvironment{bibsection}
    {\minipage[t]{\linewidth}\list{}{%
        \setlength{\leftmargin}{\bibhang}%
        \setlength{\itemindent}{-\leftmargin}%
        \setlength{\itemsep}{\bibsep}%
        \setlength{\parsep}{\z@}%
        }}
    {\endlist\endminipage}
\makeatother

% Layout: Puts the section titles on left side of page
\reversemarginpar

%
%         PAPER SIZE, PAGE NUMBER, AND DOCUMENT LAYOUT NOTES:
%
% The next \usepackage line changes the layout for CV style section
% headings as marginal notes. It also sets up the paper size as either
% letter or A4. By default, letter was used. If A4 paper is desired,
% comment out the letterpaper lines and uncomment the a4paper lines.
%
% As you can see, the margin widths and section title widths can be
% easily adjusted.
%
% ALSO: Notice that the includefoot option can be commented OUT in order
% to put the PAGE NUMBER *IN* the bottom margin. This will make the
% effective text area larger.
%
% IF YOU WISH TO REMOVE THE ``of LASTPAGE'' next to each page number,
% see the note about the +LP and -LP lines below. Comment out the +LP
% and uncomment the -LP.
%
% IF YOU WISH TO REMOVE PAGE NUMBERS, be sure that the includefoot line
% is uncommented and ALSO uncomment the \pagestyle{empty} a few lines
% below.
%

%% Use these lines for letter-sized paper
\usepackage[paper=letterpaper,
            %includefoot, % Uncomment to put page number above margin
            marginparwidth=1.2in,     % Length of section titles
            marginparsep=.05in,       % Space between titles and text
            margin=1in,               % 1 inch margins
            includemp]{geometry}

%% Use these lines for A4-sized paper
%\usepackage[paper=a4paper,
%            %includefoot, % Uncomment to put page number above margin
%            marginparwidth=30.5mm,    % Length of section titles
%            marginparsep=1.5mm,       % Space between titles and text
%            margin=25mm,              % 25mm margins
%            includemp]{geometry}

%% More layout: Get rid of indenting throughout entire document
\setlength{\parindent}{0in}

%% This gives us fun enumeration environments. compactitem will be nice.
\usepackage{paralist}

%% Reference the last page in the page number
%
% NOTE: comment the +LP line and uncomment the -LP line to have page
%       numbers without the ``of ##'' last page reference)
%
% NOTE: uncomment the \pagestyle{empty} line to get rid of all page
%       numbers (make sure includefoot is commented out above)
%
\usepackage{fancyhdr,lastpage}
\pagestyle{fancy}
%\pagestyle{empty}      % Uncomment this to get rid of page numbers
\fancyhf{}\renewcommand{\headrulewidth}{0pt}
\fancyfootoffset{\marginparsep+\marginparwidth}
\newlength{\footpageshift}
\setlength{\footpageshift}
          {0.5\textwidth+0.5\marginparsep+0.5\marginparwidth-2in}
\lfoot{\hspace{\footpageshift}%
       \parbox{4in}{\, \hfill %
                    \arabic{page} of \protect\pageref*{LastPage} % +LP
%                    \arabic{page}                               % -LP
                    \hfill \,}}

% Finally, give us PDF bookmarks
\usepackage{color,hyperref}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\hypersetup{colorlinks,breaklinks,
            linkcolor=darkblue,urlcolor=darkblue,
            anchorcolor=darkblue,citecolor=darkblue}

%%%%%%%%%%%%%%%%%%%%%%%% End Document Setup %%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%% Helper Commands %%%%%%%%%%%%%%%%%%%%%%%%%%%%

% The title (name) with a horizontal rule under it
%
% Usage: \makeheading{name}
%
% Place at top of document. It should be the first thing.
\newcommand{\makeheading}[1]%
        {\hspace*{-\marginparsep minus \marginparwidth}%
         \begin{minipage}[t]{\textwidth+\marginparwidth+\marginparsep}%
                {\large \bfseries #1}\\[-0.15\baselineskip]%
                 \rule{\columnwidth}{1pt}%
         \end{minipage}}

% The section headings
%
% Usage: \section{section name}
%
% Follow this section IMMEDIATELY with the first line of the section
% text. Do not put whitespace in between. That is, do this:
%
%       \section{My Information}
%       Here is my information.
%
% and NOT this:
%
%       \section{My Information}
%
%       Here is my information.
%
% Otherwise the top of the section header will not line up with the top
% of the section. Of course, using a single comment character (%) on
% empty lines allows for the function of the first example with the
% readability of the second example.
\renewcommand{\section}[2]%
        {\pagebreak[2]\vspace{1.3\baselineskip}%
         \phantomsection\addcontentsline{toc}{section}{#1}%
         \hspace{0in}%
         \marginpar{
         \raggedright \scshape #1}#2}

% An itemize-style list with lots of space between items
\newenvironment{outerlist}[1][\enskip\textbullet]%
        {\begin{itemize}[#1]}{\end{itemize}%
         \vspace{-.6\baselineskip}}

% An environment IDENTICAL to outerlist that has better pre-list spacing
% when used as the first thing in a \section
\newenvironment{lonelist}[1][\enskip\textbullet]%
        {\vspace{-\baselineskip}\begin{list}{#1}{%
        \setlength{\partopsep}{0pt}%
        \setlength{\topsep}{0pt}}}
        {\end{list}\vspace{-.6\baselineskip}}

% An itemize-style list with little space between items
\newenvironment{innerlist}[1][\enskip\textbullet]%
		{\begin{compactitem}[#1]}{\end{compactitem}}

% To add some paragraph space between lines.
% This also tells LaTeX to preferably break a page on one of these gaps
% if there is a needed pagebreak nearby.
\newcommand{\blankline}{\quad\pagebreak[2]}

%

%%%%%%%%%%%%%%%%%%%%%%%% End Helper Commands %%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% Begin CV Document %%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\makeheading{Matthew J. Gardner}

\section{Contact Information}
%
% NOTE: Mind where the & separators and \\ breaks are in the following
%       table.
%
% ALSO: \rcollength is the width of the right column of the table
%       (adjust it to your liking; default is 1.85in).
%
\newlength{\rcollength}\setlength{\rcollength}{1.85in}%
%
\begin{tabular}[t]{@{}p{\textwidth-\rcollength}p{\rcollength}}
795 Walnut Avenue & \textit{Cell:} (801) 995-9460 \\
Provo, UT 84604 USA    & \textit{E-mail:} mjg82@byu.edu \\
\end{tabular}

\section{Research Interests}
%
Particle swarm optimization, natural language processing, Bayesian machine
learning, Markov decision processes, value iteration

\section{Education}
%
{\textbf{Brigham Young University}},
Provo, Utah USA
\begin{outerlist}
\item[] B.S.,
		Computer Science, April 2010
        \begin{innerlist}
        \item Minors in Physics and Mathematics
        \item Thesis Topic: \emph{Speculative Evaluation in Particle Swarm
            Optimization}
        \item Adviser: Professor Kevin Seppi
        \end{innerlist}

\end{outerlist}

\section{Awards}
%
Brigham Young University
\begin{innerlist}
\item National Merit Scholarship, 2004--2010
\item Inducted into the Phi Kappa Phi Honors Society, 2009
\item Barry M. Goldwater Scholarship Nominee, 2009
\item Claricode Medical Essay Scholarship Finalist, 2009
\end{innerlist}

\section{Academic Experience}
{\textbf{Brigham Young University}},
Provo, Utah USA
\begin{outerlist}

\item[] \textit{Undergraduate Researcher}%
        \hfill \textbf{December 2007 to Present}
\begin{innerlist}
\item Worked in the Applied Machine Learning Lab in the Computer Science
  Department under Dr. Kevin Seppi.
\item Reviewed papers for the International Conference on Machine Learning 2009.
\item Research topics included particle swarm optimization and Markov decision
  processes.
\end{innerlist}
\end{outerlist}

\section{Conference Publications} \begin{bibsection}
    \item Andrew McNabb, Matthew Gardner, and Kevin Seppi. An Exploration of
	  Topologies and Communication in Large Particle Swarms.  In:
	  \emph{Congress on Evolutionary Computation, 2009}.

    \item Matthew Gardner, Andrew McNabb, and Kevin Seppi.  Speculative
	  Evaluation in Particle Swarm Optimization.  Submitted to: \emph{The
	  Genetic and Evolutionary Computation Conference, 2010}.  Awaiting
	  acceptance notification.
\end{bibsection}

\section{Professional Experience}
%
\textbf{Google, Inc.},
Irvine, California USA
\begin{outerlist}

\item[] \textit{Software Engineering Intern for the TV Ads Team}%
        \hfill \textbf{May 2009 to August 2009}
\begin{innerlist}
\item Used machine learning techniques to attribute calls received by
  advertisers to specific ads played on TV networks.
\item Gave several presentations to the machine learning lunch group.
\end{innerlist}

\end{outerlist}

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%% End CV Document %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2009-12-29 resume_cv/tpavlic_cv.tex
Added an updated resume and cv

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% Simple LaTeX CV Template %%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% NOTE: If you find that it says                                     %%
%%                                                                    %%
%%                           1 of ??                                  %%
%%                                                                    %%
%% at the bottom of your first page, this means that the AUX file     %%
%% was not available when you ran LaTeX on this source. Simply RERUN  %%
%% LaTeX to get the ``??'' replaced with the number of the last page  %%
%% of the document. The AUX file will be generated on the first run   %%
%% of LaTeX and used on the second run to fill in all of the          %%
%% references.                                                        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%% Document Setup %%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Don't like 10pt? Try 11pt or 12pt
\documentclass[10pt]{article}

% This is a helpful package that puts math inside length specifications
\usepackage{calc}

% Simpler bibsection for CV sections
% (thanks to natbib for inspiration)
\makeatletter
\newlength{\bibhang}
\setlength{\bibhang}{1em}
\newlength{\bibsep}
 {\@listi \global\bibsep\itemsep \global\advance\bibsep by\parsep}
\newenvironment{bibsection}
    {\minipage[t]{\linewidth}\list{}{%
        \setlength{\leftmargin}{\bibhang}%
        \setlength{\itemindent}{-\leftmargin}%
        \setlength{\itemsep}{\bibsep}%
        \setlength{\parsep}{\z@}%
        }}
    {\endlist\endminipage}
\makeatother

% Layout: Puts the section titles on left side of page
\reversemarginpar

%
%         PAPER SIZE, PAGE NUMBER, AND DOCUMENT LAYOUT NOTES:
%
% The next \usepackage line changes the layout for CV style section
% headings as marginal notes. It also sets up the paper size as either
% letter or A4. By default, letter was used. If A4 paper is desired,
% comment out the letterpaper lines and uncomment the a4paper lines.
%
% As you can see, the margin widths and section title widths can be
% easily adjusted.
%
% ALSO: Notice that the includefoot option can be commented OUT in order
% to put the PAGE NUMBER *IN* the bottom margin. This will make the
% effective text area larger.
%
% IF YOU WISH TO REMOVE THE ``of LASTPAGE'' next to each page number,
% see the note about the +LP and -LP lines below. Comment out the +LP
% and uncomment the -LP.
%
% IF YOU WISH TO REMOVE PAGE NUMBERS, be sure that the includefoot line
% is uncommented and ALSO uncomment the \pagestyle{empty} a few lines
% below.
%

%% Use these lines for letter-sized paper
\usepackage[paper=letterpaper,
            %includefoot, % Uncomment to put page number above margin
            marginparwidth=1.2in,     % Length of section titles
            marginparsep=.05in,       % Space between titles and text
            margin=1in,               % 1 inch margins
            includemp]{geometry}

%% Use these lines for A4-sized paper
%\usepackage[paper=a4paper,
%            %includefoot, % Uncomment to put page number above margin
%            marginparwidth=30.5mm,    % Length of section titles
%            marginparsep=1.5mm,       % Space between titles and text
%            margin=25mm,              % 25mm margins
%            includemp]{geometry}

%% More layout: Get rid of indenting throughout entire document
\setlength{\parindent}{0in}

%% This gives us fun enumeration environments. compactitem will be nice.
\usepackage{paralist}

%% Reference the last page in the page number
%
% NOTE: comment the +LP line and uncomment the -LP line to have page
%       numbers without the ``of ##'' last page reference)
%
% NOTE: uncomment the \pagestyle{empty} line to get rid of all page
%       numbers (make sure includefoot is commented out above)
%
\usepackage{fancyhdr,lastpage}
\pagestyle{fancy}
%\pagestyle{empty}      % Uncomment this to get rid of page numbers
\fancyhf{}\renewcommand{\headrulewidth}{0pt}
\fancyfootoffset{\marginparsep+\marginparwidth}
\newlength{\footpageshift}
\setlength{\footpageshift}
          {0.5\textwidth+0.5\marginparsep+0.5\marginparwidth-2in}
\lfoot{\hspace{\footpageshift}%
       \parbox{4in}{\, \hfill %
                    \arabic{page} of \protect\pageref*{LastPage} % +LP
%                    \arabic{page}                               % -LP
                    \hfill \,}}

% Finally, give us PDF bookmarks
\usepackage{color,hyperref}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\hypersetup{colorlinks,breaklinks,
            linkcolor=darkblue,urlcolor=darkblue,
            anchorcolor=darkblue,citecolor=darkblue}

%%%%%%%%%%%%%%%%%%%%%%%% End Document Setup %%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%% Helper Commands %%%%%%%%%%%%%%%%%%%%%%%%%%%%

% The title (name) with a horizontal rule under it
%
% Usage: \makeheading{name}
%
% Place at top of document. It should be the first thing.
\newcommand{\makeheading}[1]%
        {\hspace*{-\marginparsep minus \marginparwidth}%
         \begin{minipage}[t]{\textwidth+\marginparwidth+\marginparsep}%
                {\large \bfseries #1}\\[-0.15\baselineskip]%
                 \rule{\columnwidth}{1pt}%
         \end{minipage}}

% The section headings
%
% Usage: \section{section name}
%
% Follow this section IMMEDIATELY with the first line of the section
% text. Do not put whitespace in between. That is, do this:
%
%       \section{My Information}
%       Here is my information.
%
% and NOT this:
%
%       \section{My Information}
%
%       Here is my information.
%
% Otherwise the top of the section header will not line up with the top
% of the section. Of course, using a single comment character (%) on
% empty lines allows for the function of the first example with the
% readability of the second example.
\renewcommand{\section}[2]%
        {\pagebreak[2]\vspace{1.3\baselineskip}%
         \phantomsection\addcontentsline{toc}{section}{#1}%
         \hspace{0in}%
         \marginpar{
         \raggedright \scshape #1}#2}

% An itemize-style list with lots of space between items
\newenvironment{outerlist}[1][\enskip\textbullet]%
        {\begin{itemize}[#1]}{\end{itemize}%
         \vspace{-.6\baselineskip}}

% An environment IDENTICAL to outerlist that has better pre-list spacing
% when used as the first thing in a \section
\newenvironment{lonelist}[1][\enskip\textbullet]%
        {\vspace{-\baselineskip}\begin{list}{#1}{%
        \setlength{\partopsep}{0pt}%
        \setlength{\topsep}{0pt}}}
        {\end{list}\vspace{-.6\baselineskip}}

% An itemize-style list with little space between items
\newenvironment{innerlist}[1][\enskip\textbullet]%
        {\begin{compactitem}[#1]}{\end{compactitem}}

% To add some paragraph space between lines.
% This also tells LaTeX to preferably break a page on one of these gaps
% if there is a needed pagebreak nearby.
\newcommand{\blankline}{\quad\pagebreak[2]}

%

%%%%%%%%%%%%%%%%%%%%%%%% End Helper Commands %%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% Begin CV Document %%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\makeheading{Matthew J. Gardner}

\section{Contact Information}
%
% NOTE: Mind where the & separators and \\ breaks are in the following
%       table.
%
% ALSO: \rcollength is the width of the right column of the table
%       (adjust it to your liking; default is 1.85in).
%
\newlength{\rcollength}\setlength{\rcollength}{1.85in}%
%
\begin{tabular}[t]{@{}p{\textwidth-\rcollength}p{\rcollength}}
795 Walnut Avenue & \textit{E-mail:} mjg82@byu.edu\\
Provo, UT 84604 USA    & \\
\end{tabular}

\section{Security Clearance}
%
Department of Defense Top Secret SCI with polygraph (expired: 2002)

\section{Citizenship}
%
USA

\section{Research Interests}
%
Control theory, communication theory, behavioral ecology, cooperation
theory, engineering education

\section{Education}
%
\href{http://www.osu.edu/}{\textbf{The Ohio State University}},
Columbus, Ohio USA
\begin{outerlist}

\item[] Ph.D.,
        \href{http://www.ece.osu.edu/}
             {Electrical and Computer Engineering}
             (expected graduation date: June 2010)
        \begin{innerlist}
        \item Thesis Topic: \emph{Dynamics of Distributed Cooperative
            Agents}
        \item Adviser:
              \href{http://www.ece.osu.edu/~passino/}
                   {Professor Kevin M.~Passino}
        \item Area of Study: Control Engineering
        \end{innerlist}

\item[] M.S.,
        \href{http://www.ece.osu.edu/}
             {Electrical and Computer Engineering}, August 2007
        \begin{innerlist}
        \item Thesis Topic: \emph{Optimal Foraging Theory Revisited}
        \item Adviser:
              \href{http://www.ece.osu.edu/~passino/}
                   {Professor Kevin M.~Passino}
        \item Area of Study: Control Engineering
        \end{innerlist}

\item[] B.S.,
        \href{http://www.ece.osu.edu/}
             {Electrical and Computer Engineering}, June 2004
        \begin{innerlist}
        \item \emph{Magna cum Laude}, With Honors in Engineering
        \item Electrical specialization (emphasis on electromagnetics and digital computers)
        \item Minor in \href{http://www.cse.ohio-state.edu/}
                            {Computer and Information Systems}
              (programming and algorithms track)
        \end{innerlist}

\end{outerlist}

\section{Awards}
%
\href{http://www.nsf.gov/}{National Science Foundation}
\begin{innerlist}
\item \href{http://www.nsfgk12.org/}{GK-12 Fellowship}, 2006
\item \href{http://www.nsf.gov/grfp}
           {Graduate Research Fellowship} Honorable Mention, 2005
\end{innerlist}

\blankline

\href{http://www.osu.edu}{The Ohio State University}
\begin{innerlist}
\item \href{http://www.gradsch.osu.edu/Content.aspx?Content=44&itemid=2}
           {Dean's Distinguished University Fellowship}, 2004
\item Electrical and Computer Engineering Bradshaw Scholarship,
        2002--2004
\item Electrical and Computer Engineering Shafstall Scholarship,
        2001--2003
\item University Scholarship, 1999--2003
\end{innerlist}

\section{Academic Experience}
\href{http://www.osu.edu}{\textbf{The Ohio State University}},
Columbus, Ohio USA
\begin{outerlist}

\item[] \textit{Teaching Assistant}%
    \hfill \textbf{September 2007 to August 2009}
    \begin{innerlist}
        \item Instructor for ECE~327: Electronic Devices and Circuits Laboratory I
        \begin{innerlist}
            \item Autumn~2007, Winter~2008 (2 sections), Spring~2008
                (2 sections), Winter~2009 (2 sections), and Summer~2009

            \item Responsible for 1~hour lecture and supervision of
                3~hour laboratory where junior and senior undergraduate
                students design and implement infrared modem and speaker
                driver for analog electronic audio signals

            \item Developed hundreds of pages of supplementary course
                material, including a course web page archived at
                \href
                {http://www.tedpavlic.com/teaching/osu/ece327}
                {\texttt{http://www.tedpavlic.com/teaching/osu/ece327}}
        \end{innerlist}~

        \item Grader for ECE~481 Ethics in Electrical and Computer Engineering
        \begin{innerlist}
            \item Autumn~2007 and Autumn~2008
        \end{innerlist}~

        \item Instructor for ECE~209: Circuits and Electronics
            Laboratory
        \begin{innerlist}
            \item Autumn~2008

            \item Responsible for 0.5~hour lecture and supervision of
                3.5~hour laboratory where sophomore undergraduate
                students learn learn how to use basic laboratory
                equipment to study properties of electronic circuits

            \item Developed supplementary course
                material, including a course web page archived at
                \href
                {http://www.tedpavlic.com/teaching/osu/ece209}
                {\texttt{http://www.tedpavlic.com/teaching/osu/ece209}}
        \end{innerlist}~

        \item Instructor for ECE~557: Control, Signals, and Systems
            Laboratory
        \begin{innerlist}
            \item Summer~2008 (2 sections) and Summer~2009

            \item Responsible for 0.5~hour lecture and supervision of
                3.5~hour laboratory where senior undergraduate students
                combine
                \href{http://www.mathworks.com/products/simulink/}{Simulink},
                with \href{http://www.dspaceinc.com/}{dSPACE} RTI1104
                real-time control hardware and software to do analysis
                and control implementation for linear systems

            \item Developed supplementary course
                material, including a course web page archived at
                \href
                {http://www.tedpavlic.com/teaching/osu/ece557}
                {\texttt{http://www.tedpavlic.com/teaching/osu/ece557}}
        \end{innerlist}~

        \item Lab Instructor for ECE~758: Control Systems Implementation
            Laboratory
        \begin{innerlist}
            \item Spring~2009 (2 sections)

            \item Responsible for 0.5~hour lecture and supervision of
                3.5~hour laboratory where graduate students and senior
                undergraduate students combine
                \href{http://www.mathworks.com/products/simulink/}{Simulink},
                with \href{http://www.dspaceinc.com/}{dSPACE} RTI1104
                real-time control hardware and software to do analysis
                and advanced control implementation for linear and
                non-linear systems

            \item Developed supplementary course
                material, including a course web page archived at
                \href
                {http://www.tedpavlic.com/teaching/osu/ece758}
                {\texttt{http://www.tedpavlic.com/teaching/osu/ece758}}
        \end{innerlist}
    \end{innerlist}

\item[] \textit{Graduate Student}%
        \hfill \textbf{June 2004 to present}
\begin{innerlist}
\item \href{http://www.gradsch.osu.edu/Content.aspx?Content=44&itemid=2}
           {Dean's Distinguished University Fellow}
      (June 2004 to present)
        \begin{innerlist}
        \item[] Includes M.S.~and Ph.D.~research and course work.
        \end{innerlist}
\item \href{http://www.nsfgk12.org/}
           {National Science Foundation GK-12 Fellow}
      (September 2006 to October 2007)
        \begin{innerlist}
        \item[] Developed, implemented, and evaluated daily fourth grade
                science lessons for a local inner-city public school
                class.
        \end{innerlist}
\end{innerlist}

\item[] \textit{Instructor}%
        \hfill \textbf{March 2002 to June 2004}
\begin{innerlist}
\item Member of \href{http://feh.eng.ohio-state.edu/}
                     {Fundamentals of Engineering for Honors}
      instructional team.
\item Special graduate teaching appointment as undergraduate.
\item Lectured weekly laboratory on engineering fundamentals (ENG~H191,
        H192, and~H193).
\item Trained in-class undergraduate teaching assistants in laboratory
        procedure.
\item Graded weekly lab reports and provided laboratory exams.
\end{innerlist}

\item[] \textit{Teaching Assistant}%
        \hfill \textbf{September 2000 to March 2002}
\begin{innerlist}
\item Assisted \href{http://feh.eng.ohio-state.edu/}
                    {Fundamentals of Engineering for Honors}
      instructional team.
\item Provided in-class support to first-year engineering students (ENG
        H191, H192, and H193).
\item Graded daily assignments on programming and drafting.
\end{innerlist}

\item[] \textit{Undergraduate Researcher}%
        \hfill \textbf{September 2000 to March 2002}
\begin{innerlist}
\item Participated in the
        \href{http://www.cse.ohio-state.edu/europa/}{Europa
        Undergraduate Research Forum}, a part of the
        \href{http://www.cse.ohio-state.edu/rsrg/}{Reusable Software
        Research Group}.
\item Worked to improve undergraduate education of component based
        software engineering topics.
\item Researched needed changes to RESOLVE/C++ implementation for
        ANSI/C++ compliance.
\end{innerlist}

\item[] \textit{Grader}%
        \hfill \textbf{September 2001 to December 2001}
\begin{innerlist}
\item Graded daily electromagnetics assignments (ECE~311).
\end{innerlist}

\item[] \textit{Undergraduate Student}%
        \hfill \textbf{September 1999 to June 2004}
\end{outerlist}

\section{Publications} \begin{bibsection}
    \item Pavlic, T.P., and K.M.~Passino. Cooperative task
        processing. \emph{IEEE Transactions of Automatic Control}. 2009.
        Submitted.

    \item Pavlic, T.P., and K.M.~Passino. When rate maximization is
        impulsive. \emph{Behavioral Ecology and Sociobiology}. 2009.
        Accepted pending revisions.

    \item Pavlic, T.P., and K.M.~Passino. The sunk-cost effect and
        other optimal rate-maximizing behaviors. \emph{Acta
        Biotheoretica}. 2009. Submitted.

    \item Pavlic, T.P., and K.M.~Passino. Foraging theory for mobile
        agent speed choice. \emph{Engineering Applications of Artificial
        Intelligence}, 22(3):482--489, April 2009.

    \item Pavlic, T.P. \emph{Optimal Foraging Theory Revisited}.
        Master's thesis, The Ohio State University, Columbus, OH, 2007.
\end{bibsection}

\section{Books in Preparation} \begin{bibsection}
    \item Pavlic, T.P., B.W.~Andrews, K.M.~Passino, and T.A.~Waite.
        \emph{Foraging Theory for Engineering}.
\end{bibsection}

\section{Conference Publications} \begin{bibsection}
    \item Pavlic, T.P., and K.M.~Passino. Cooperative task
        processing. In: \emph{Proceedings of the ICAM 2009 Symposium:
        Emergence in Physical, Biological, and Social Systems IV},
        November 13, 2009. Poster abstract.

    \item Freuler, R.J., M.J.~Hoffmann, T.P.~Pavlic, J.M.~Beams,
        J.P.~Radigan, P.K.~Dutta, J.T.~Demel, and E.D.~Justen.
        Experiences with a Comprehensive Freshman Hands-On Course~--
        Designing, Building, and Testing Small Autonomous Robots. In:
        \emph{Proceedings of the 2003 American Society for Engineering
        Education Annual Conference \& Exposition}, 2003.
\end{bibsection}

\section{Professional Experience}
%
\href{http://www.ni.com/}{\textbf{National Instruments}},
Austin, Texas USA
\begin{outerlist}

\item[] \textit{Hardware R\&D Intern for Multifunction DAQ}%
        \hfill \textbf{June 2003 to September 2003}
\begin{innerlist}
\item Designed final verification testing fixture for use with STC2 MIO
        products.
\item Designed and executed study of the effect of varying burn-in time
        on long-term drift of common industry voltage references.
\end{innerlist}

\item[] \textit{Hardware R\&D Intern for Multifunction DAQ}%
        \hfill \textbf{June 2002 to September 2002}
\begin{innerlist}
\item Designed and performed validation tests on new 16-bit 800 kHz
        NI-6120 SMIO DAQ board.

\item Designed high quality filter/amplifier source for use with NI-5411
        arbitrary function generator.
\end{innerlist}

\end{outerlist}

\blankline

\textbf{\href{http://www.ibm.com/}{IBM} Network Storage},
Research Triangle Park, North Carolina USA
\begin{outerlist}

\item[] \textit{Core Systems Software Developer for FlexNAS}%
        \hfill \textbf{June 2001 to September 2001}
\begin{innerlist}
\item Designed and implemented high-availability, redundant internode
        communications subsystem.
\item Participated in software development of various vital box
        services.
\end{innerlist}

\end{outerlist}

\blankline

\href{http://www.calltech.com/}{\textbf{CallTech Communications}},
Columbus, Ohio USA
\begin{outerlist}

\item[] \textit{Information Technology Systems Engineer}%
        \hfill \textbf{June 1997 to May 2001}
\begin{innerlist}
\item Responsible for the acquisition, setup, maintenance, and
        administration of all Internet hardware and software supporting
        \href{http://www.netwalk.com/}{NetWalk} Internet service
        and web presence provider.
\item Designed and implemented state of the art open source
        high-availability load balancing system supporting thousands of
        virtual servers.
\item Developed software call center support software for clients such
        as CompuServe, AOL, and Priceline.
\end{innerlist}

\end{outerlist}

\blankline

MegaLinx Communications, Dublin, Ohio USA
\begin{outerlist}

\item[] \textit{Web Developer and Support Representative}%
        \hfill \textbf{June 1995 to May 1997}
\begin{innerlist}
\item Produced web content for commercial clients.
\item Assisted in administration of UltraSPARC, x86, 68020, 68030, and
        PowerPC systems running Sun Solaris, Linux, Microsoft DOS,
        Microsoft Windows NT, and Apple Macintosh operating systems.
\item Developed multi-platform open source file sharing solution.
\item Provided technical support for Internet and web presence
        customers.
\end{innerlist}

\end{outerlist}

\section{Service}
%
Director of Computers,
\href{http://ec.osu.edu/}{Engineers' Council},
\href{http://www.osu.edu/}{The Ohio State University}, 2002

\blankline

\href{http://www.osufirst.org/}{OSU FIRST Robotics Team},
\href{http://www.osu.edu}{The Ohio State University}, 2000--2004
\begin{innerlist}
\item Introduced middle school and high school students to science and
        technology by participating with them in national robotics
        competitions.
\item Led 2002 team to regional silver medal
        \href{http://www.firstwiki.org/Engineering_Inspiration_Award}
             {\emph{Engineering Inspiration Award}}.
\item \emph{Lead Team Mentor}, 2002--2004
\item \emph{Component Design Team Lead Mentor}, 2001--2002
\end{innerlist}

\blankline

\href{http://www.linuxvirtualserver.org/}
     {Linux Virtual Server Project}, 1999--2000
\begin{innerlist}
\item Early member of the team that formed the open source project that
        is now an important load balancing solution for the Linux
        software platform.
\end{innerlist}

\blankline

\href{http://www.gcfn.org/}
     {Greater Columbus Free-Net}, 1995--1997
\begin{innerlist}
\item Provided technical support services.
\end{innerlist}

\blankline

CompuTeen Bulletin Board System, 1993--1995
\begin{innerlist}
\item Administrated dial-up bulletin board system.
\item Founded and administrated TeenLiNK, an international electronic
        mail network that spread through the United States, Canada, and
        Australia and delivered mail over a series of electronic dial-up
        drop offs.
\end{innerlist}

\section{Technical Skills}
%
Extensive hardware and software experience in networking and
        information technology

\blankline

\href{http://www.mathworks.com/products/matlab/}{\textsc{Matlab}}
        experience: linear algebra, Fourier transforms,
        nonlinear numerical methods, polynomials, statistics,
        visualization

\blankline

\href{http://www.mathworks.com/products/matlab/}{\textsc{Matlab}}
        toolboxes: communications, control system, filter
        design, genetic algorithm and direct search, signal processing,
        system identification

\blankline

Instrumentation and Control:
        \href{http://www.dspaceinc.com/}{dSPACE} hardware and software,
        \href{http://www.mathworks.com/products/simulink/}{Simulink},
        \href{http://www.ni.com/}{LabVIEW} and other
        \href{http://www.ni.com}{National Instruments}
        control and data acquisition hardware and software

\blankline

Programming: C, C++, Pascal, Perl, PHP, Lisp, UNIX shell scripting, SQL,
        RCS, CVS, SVN, and others

\blankline

Applications: \TeX{}, \LaTeX{}, B\textsc{ib}\TeX{}, Microsoft Office,
        and other common productivity packages for Windows, OS X, and
        Linux platforms

\blankline

Operating Systems: Microsoft Windows XP/2000, Apple OS X, Linux, BSD,
        IRIX, AIX, Solaris, and other UNIX variants

\section{Mathematical Expertise}
%
Linear and Nonlinear Systems Theory

\blankline

Probability, Random Variables, and Stochastic Processes

\blankline

Dynamic Optimization

\blankline

Game Theory

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%% End CV Document %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2009-12-29 honors_thesis/thesis/paper.tex
Some (final?) modifications to the GECCO paper, after my own reading through

@@
 \begin{table}
-  \caption{All possible updates for a particle with two neighbors.}
+  \caption{All possible updates for a particle with two neighbors}
   \label{tab:evals}
@@
 always its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report
-the results for using both the Ring topology (where $n$ again is $3$) and the
-Random topology on a number of common benchmark functions.
+the results for performing speculative evaluation using both the Ring topology
+(where $n$ again is $3$) and the Random topology on a number of common
+benchmark functions.
 
@@
 This is done by first updating each particle $\noeval{\p}_t$ to
-$\nonbest{\p}_t$ by associating the value determined by the slave to the
+$\nonbest{\p}_t$ and each speculative particle $\noeval{\s}_{t+1}$ to
+$\nonbest{\s}_{t+1}$ by associating the value determined by the slave to the
 particle it came from, and updating that particle's personal best if the value
@@
 particle can be determined, and the correct speculative child can be chosen.
-The branches depended entirely on whether or not the personal best of $p_t$ was
+The branches depend entirely on whether or not the personal best of $p_t$ was
 updated and which particle, if any, updated the particle's neighborhood best.
@@
 knows only the position that it guessed, not the function value at that
-position.  It is possible that both $\p_t$ and $\s_{t+1}$ updated their
-personal bests, but $\p_t$'s value was better.  For example, suppose that
-$\p_{t-1}$ has a personal best value of 3.  $\noeval{\p}_t$ was created, and
-$\noeval{\s}_{t+1}$ was moved assuming that $\p_t$ had updated its personal
+position.  It is possible that both $\p_t$ and $\s_{t+1}$ update their
+personal bests, but $\p_t$'s value is better.  For example, suppose that
+$\p_{t-1}$ has a personal best value of 3.  $\noeval{\p}_t$ is created, and
+$\noeval{\s}_{t+1}$ is moved assuming that $\p_t$ has updated its personal
 best with its position at time $t$.  Then both $\noeval{\p}_t$ and
-$\noeval{\s}_{t+1}$ were evaluated, with values 1 and 2, respectively.
-$\nonbest{\s}_{t+1}$ would think that its current position was its personal
-best, as the value it found, 2, was better than its previous personal best
+$\noeval{\s}_{t+1}$ are evaluated, with values 1 and 2, respectively.
+$\nonbest{\s}_{t+1}$ would think that its current position is its personal
+best, as the value it found, 2, is better than its previous personal best
 value of 3.  It needs to receive the personal best value from its parent to
@@
 each processor has the information for the particle or particles that it is in
-charge of and must perform the steps of the algorithm for the particles it
-manages.  Messages such as values and positions for the neighborhood best are
-sent between processors.  There may be some machine that collects information
-from all of the particles and outputs the result of the algorithm, though that
+charge of and must perform the steps of the algorithm for those particles.
+Messages such as values and positions for the neighborhood best are sent
+between processors.  There may be some machine that collects information from
+all of the particles and outputs the result of the algorithm, though that
 machine's importance is much less than in centralized algorithms.
@@
 and the positions of its speculative children, it passes information about the
-outcome iteration $t$ to its neighbors, so that each particle can receive
-information about its neighborhood best to be updated to $\p_t$.  Once that
-communication is finished, the particle can select the speculative child which
-matched the branch that iteration $t$ actually produced.  Then another round of
-information passing follows, for iteration $t+1$, so that $\nonbest{\p}_{t+1}$
-can be updated to $\p_{t+1}$.  Two iterations have then been completed with
-only one round of evaluations, and the next iteration can start again with the
-first round of message passing.
+outcome of iteration $t$ to its neighbors, so that neighboring particles can
+update their neighborhood bests to move from $\nonbest{\p}_t$ to $\p_t$.  Once
+that communication is finished, the particle can select the speculative child
+which matched the branch that iteration $t$ actually produced.  Then another
+round of information passing follows, for iteration $t+1$, so that
+$\nonbest{\p}_{t+1}$ can be updated to $\p_{t+1}$.  Two iterations have then
+been completed with only one round of evaluations, and the next iteration can
+start again with the first round of message passing.
 
@@
 This second method only requires one round of passing information because
-information about iteration $t$ and $t+1$ are passed at the same time.  Each
+information about iterations $t$ and $t+1$ is passed at the same time.  Each
 processor reconstructs from the messages it receives all of the information
@@
 obtain $\nonbest{\p}_{t+1}$.  We graphically show the messages needed to
-produce $\nonbest{\p}_{t+1}$ in \fig{messages3}.  Note that it looks identical
-to \fig{messages2}, just with different sets of particles.
+produce $\nonbest{\nset}_{t+1}$ in \fig{messages3}.  Note that it looks
+identical to \fig{messages2}, just with different sets of particles.
 
@@
 However, some parallel frameworks necessitate this type of message passing, so
-we showed how speculative evaluation can be performed in such frameworks.
+we have described how speculative evaluation can be performed in such
+frameworks.
 
@@
 positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
-step 7, where the $\noeval{\s}_{t+1}$ with the highest value is chosen from
+step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
 $\noeval{\sset}_{t+1}$ instead of with the matching branch.
@@
 algorithms initially outperformed PSO, but got stuck in a local optima and were
-eventually outperformed.
+eventually outperformed themselves.
 
@@
   evaluations.  Instead of showing average function value, we show the
-  percentage of runs that found the global optimum at each iteration.}
+  percentage of runs that have found the global optimum by each iteration.}
   \label{fig:griewank2}
@@
 particles most often follow.  Only the subset of the most probable next
-locations would then be evaluated.  Using an asynchronous PSO variant would
-also be helpful in this case, allowing particles to be at different iterations
-in case the correct branch was not guessed and needs to be evaluated
-separately.
+locations would then be evaluated.  Using an asynchronous PSO variant may also
+be helpful in this case, allowing particles to be at different iterations in
+case the correct branch was not guessed and needs to be evaluated separately.
 

### 2009-12-29 honors_thesis/thesis/sepsodata/plots/griewank.py
Some updates to sepsodata/plots, though I really do not use it anymore...

@@ -20,7 +20,7 @@ ring_points = []
 for i in xrange(1, ITERATIONS):
     evaluations = i * SWARMSIZE
     sepso_points.append((evaluations, sepso_data.average(i*2)))
-    pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+    #pickbest_points.append((evaluations, pickbest_data.average(i*2)))
     ring_points.append((evaluations, ring_data.average(i)))
 
 plot = Plot()
@@ -33,7 +33,7 @@ plot.xlabel = 'Function Evaluations'
 plot.ylabel = 'Best Function Value'
 
 plot.append(Points(sepso_points, title='SEPSO', style='lines'))
-plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
+#plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
 plot.append(Points(ring_points, title='PSO Ring', style='lines'))
 
 plot.show()

### 2009-12-29 honors_thesis/thesis/sepsodata/plots/sphere.py

@@ -23,7 +23,7 @@ complete_points = []
 for i in xrange(1, ITERATIONS):
     evaluations = i * SWARMSIZE
     sepso_points.append((evaluations, sepso_data.average(i*2)))
-    pickbest_points.append((evaluations, pickbest_data.average(i*2)))
+    #pickbest_points.append((evaluations, pickbest_data.average(i*2)))
     rand_points.append((evaluations, rand_data.average(i)))
     complete_points.append((evaluations, complete_data.average(i)))
 
@@ -37,42 +37,11 @@ plot.xlabel = 'Function Evaluations'
 plot.ylabel = 'Best Function Value'
 
 plot.append(Points(sepso_points, title='SEPSO', style='lines'))
-plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
+#plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
 plot.append(Points(rand_points, title='PSO Random', style='lines'))
 plot.append( Points(complete_points, title='PSO Complete', style='lines'))
 
 plot.show()
-plot.write('sphere1.gpi')
-
-ITERATIONS = 250
-SWARMSIZE = 50
-
-PICKBEST_FILE = '../sepso_pickbest/sphere/rand52particles'
-PICKBEST13_FILE = '../sepso_pickbest/sphere/rand13particles2000iters'
-
-pickbest_data = PSOData(open(PICKBEST_FILE))
-pickbest13_data = PSOData(open(PICKBEST13_FILE))
-pickbest_points = []
-pickbest13_points = []
-for i in xrange(1, ITERATIONS):
-    evaluations = i * SWARMSIZE
-    pickbest_points.append((evaluations, pickbest_data.average(i*2)))
-    pickbest13_points.append((evaluations, pickbest13_data.average(i*8)))
-
-plot = Plot()
-plot.xmin = SWARMSIZE
-plot.xmax = ITERATIONS * SWARMSIZE
-plot.ymin = 10 ** -40
-plot.ymax = 10 ** 5
-plot.ylogscale = 10
-plot.xlabel = 'Function Evaluations'
-plot.ylabel = 'Best Function Value'
-
-plot.append(Points(pickbest_points, title='52 Particles', style='lines'))
-plot.append(Points(pickbest13_points, title='13 Particles',
-    style='lines'))
-
-plot.show()
-plot.write('sphere2.gpi')
+plot.write('sphere.gpi')
 
 # vim: et sw=4 sts=4

### 2009-12-29 honors_thesis/journal_paper/paper.tex
Started working on the journal version.  It is coming along

--- 2009-12-29 honors_thesis/thesis/paper.tex
@@
-\documentclass[conference,letterpaper]{IEEEtran}
+\documentclass[journal,letterpaper]{IEEEtran}
 
@@
 The original PSO can be reproduced exactly, two iterations at a time, using
-this technique.  However, in strictly using speculative evaluation to reproduce
-the original PSO, a lot of positions are speculatively evaluated whose values
-get ignored after deciding which branch was the correct one.  A better use of
-those speculative evaluations may be to incorporate the information into the
-swarm, no longer exactly reproducing PSO but perhaps producing better results.
+this technique.  However, if we relax the requirements of the algorithm, no
+longer demanding that it strictly mimic the behavior of the original PSO, we
+can introduce new speculative techniques that often out-perform PSO.
 
@@
 then describes how speculative evaluation can be done in parallel PSO to
-perform two iterations at once.  In \sec{extra}, we discuss a variant of the
-speculative algorithm which uses all of the speculative evaluations performed
-instead of throwing away all but one for each particle.  We then show in
-\sec{results} a comparison between the original PSO algorithm, our speculative
-algorithm, and its variant.  In \sec{conclusion} and \sec{future} we conclude
-and discuss future work.
+perform two iterations at once, then presents a comparison of the speculative
+algorithm with the original PSO.  In \sec{relax}, we discuss various methods of
+improving the performance of speculative evaluation in PSO, all of which break
+the requirement of strictly reproducing the behavior of the original algorithm.
+In \sec{conclusion} and \sec{future} we conclude and discuss future work.
 
@@
 
-\section{Using All Speculative Evaluations}
-\label{sec:extra}
+\subsection{Comparison with Standard PSO}
+
+We ran experiments to compare our speculative PSO algorithm to the standard
+PSO algorithm.  To use the same number of processors for each algorithm,
+speculative evaluation needs a smaller swarm size than standard PSO.  For each
+of the topologies we used, a particle has two neighbors other than itself.  As
+shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
+particle.  With one evaluation needed for the original, non-speculative
+particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
+the number of particles in the speculative swarm.  The extra evaluations done
+in speculative evaluation would instead be particles in regular PSO, so we
+compare swarms of size $p_s$ in speculative evaluation with swarms of size
+$8p_s$ in regular PSO.
+
+We use a Random topology in our speculative algorithm, as the Complete topology
+leads to an explosion of speculative evaluations.  If speculative evaluation
+were not being performed, it is possible that the Complete topology would be
+used.  However, the Complete topology also requires a very large amount of
+interprocessor communication in parallel PSO, so it is still quite possible
+that Random would be used even with standard PSO.  But, to be fair in our
+comparisons, we compare to the standard PSO algorithm using both the Random
+topology with two neighbors and the Complete topology.  All functions have 20
+dimensions, and all results shown are averages over 20 runs.  We use 240
+processors in our experiments, so our speculative algorithm uses 30 particles
+and standard PSO uses 240 particles.  For ease of labeling, we call our
+speculative algorithm SEPSO and standard PSO just PSO, with the topology
+following.
+
+First we look at the function Sphere, the simplest of common benchmark
+functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
+has no local optima and is best optimized in terms of function evaluations with
+a small swarm using a Complete topology.  Accordingly, we show results for a
+Complete swarm in \fig{basic-sphere}.  In comparing SEPSO with PSO, we can see
+that when both are using a Random topology, SEPSO clearly outperforms PSO.
+However, the comparison between SEPSO and PSO Complete is not as
+encouraging---PSO Complete performs slightly better.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere.eps}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
+  two iterations per round, while standard PSO algorithms use 240 processors
+  and do one iteration per round.}
+  \label{fig:basic-sphere}
+\end{figure}
+
+Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
+\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
+\right) + 1$.  It is generally recommended to use the Ring topology when
+optimizing the Griewank function, as Complete is prone to premature convergence
+on a local optima.  We show results in \fig{basic-griewank1} for swarms of size
+30 and 240 using the Ring topology. One can see in the figure that using
+speculative evaluation initially outperforms PSO, but gets stuck in a local
+optima and is eventually outperformed.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank1.eps}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
+  two iterations per round, while standard PSO algorithms use 240 processors
+  and do one iteration per round.}
+  \label{fig:basic-griewank1}
+\end{figure}
+
+\fig{basic-griewank1} is slightly deceptive, however, as with Griewank some of
+the runs found the global optimum, 0, while some did not.  A more accurate plot
+for this function is one showing what percentage of the runs have found the
+global optimum at each iteration.  \fig{basic-griewank2} shows such a plot.  A
+swarm of size 30 performing the PSO algorithm gets stuck a little less than
+half of the time, while 240 particles provides enough exploration to find the
+global optimum every time.  As would be expected, when our speculative
+algorithm finds the global optimum, it finds it about twice as fast as regular
+PSO---it is just performing regular PSO twice as fast with a smaller swarm
+size.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank2.eps}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations.  Instead of showing average function value, we show the
+  percentage of runs that have found the global optimum by each iteration.}
+  \label{fig:basic-griewank2}
+\end{figure}
+
+Lastly we look at a function that does not show any kind of improvement with
+this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
+\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
+with Rastrigin, the more particles there are in the swarm, the lower function
+value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
+swarms get caught in local optima.  Because our speculative algorithm has a
+significantly smaller swarm size, it gets stuck at higher values while the
+larger swarms performing regular PSO continue to improve the best value found.
+\fig{rastrigin} shows the results graphically.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
+  \caption{Function Rastrigin with a swarm that uses 240 processors per round
+  of evaluations.  Our speculative algorithm has a swarm of 30 particles and
+  does two iterations per round, while standard PSO algorithms use 240
+  processors and do one iteration per round.}
+  \label{fig:rastrigin}
+\end{figure}
+
+\section{Relaxing the Requirements}
+\label{sec:relax}
+
+While the idea of speculative evaluation in particle swarm optimization is
+intriguing, the results of the previous section show that it requires too many
+extra evaluations to produce competitive results.  On most functions, the
+larger swarm size that can be used in regular PSO leads to better performance
+than is found by trying to do speculative evaluation.  However, if we keep the
+idea of speculative evaluation while relaxing the requirement of exactly
+reproducing the behavior of the original PSO algorithm, we see some very
+impressive results.
+
+Should I present results for each improvement individually, or all together, or
+both?
+
+\subsection{Pick the Best Child}
 
@@
 speculative evaluations are done per particle, while all but one of them are
-completely ignored.  It seems reasonable to try to make use of those
-evaluations instead of throwing them away.  Making use of them changes the
-behavior of PSO, instead of reproducing it exactly as the above method
-explains, but the change could be an improvement.
+completely ignored.  We can try to make use of those evaluations instead of
+throwing them away.
 
@@
 
+\subsection{Pruning the Speculative Evaluations}
+
+Because the problem facing speculative evaluation is having too many possible
+evaluations to perform, a natural step to take is to eliminate some of the
+evaluations.  If we could reliably predict which branch were going to be taken,
+we could limit ourselves to one speculative evaluation per particle instead of
+$2(n+1)$.  While we cannot with certainty predict the correct branch every
+time, we can look at the statistics of the behavior of particles in general and
+perhaps come up with a good guess, reducing the number of necessary speculative
+evaluations.
+
+\subsubsection{Branch Statistics}
+
+In Table~\ref{tab:evals} we presented all possible branches that a particle
+with two neighbors could take.  Here we lump all of the neighbors together and
+consider the statistics for the five branches shown in Table~\ref{tab:branches}.
+
+\begin{table}
+  \caption{Five Branches to Consider for Statistics}
+  \label{tab:branches}
+  \centering
+  \begin{tabular}{rc|c}
+	&$\pbest$ updated&Source of $\nbest$ update\\
+	\hline
+	1.&No update&No update\\
+	2.&Updated&No update\\
+	3.&Updated&The particle itself\\
+	4.&No update&Some neighbor\\
+	5.&Updated&Some neigbhor\\
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Branch Statistics in PSO}
+  \label{tab:stats}
+  \centering
+  \begin{tabular}{c|c|c|c|c|c|c}
+	Topology&Function&1&2&3&4&5\\
+	\hline
+	Ring&Sphere&0.530&0.093&0.114&0.202&0.062\\
+	&Griewank&0.517&0.084&0.122&0.207&0.070\\
+	&Rastrigin&0.495&0.048&0.146&0.213&0.099\\
+	&Rosenbrock&0.513&0.074&0.129&0.211&0.073\\
+	\hline
+	Random&Sphere&0.667&0.119&0.026&0.156&0.031\\
+	&Griewank&0.690&0.109&0.025&0.149&0.027\\
+	&Rastrigin&0.819&0.055&0.015&0.100&0.010\\
+	&Rosenbrock&0.742&0.077&0.022&0.140&0.018\\
+	\hline
+	Complete&Sphere&0.319&0.092&0.002&0.451&0.135\\
+	&Griewank&0.353&0.084&0.002&0.441&0.119\\
+	&Rastrigin&0.477&0.067&0.002&0.382&0.070\\
+	&Rosenbrock&0.353&0.034&0.003&0.544&0.066\\
+  \end{tabular}
+\end{table}
+
+Row 1 in Table~\ref{tab:branches} corresponds to the phenomon commonly called
+stagnation in the literature, but only with a single particle.  As an
+interesting aside, we found that complete stagnation practically never occurs
+in PSO, though frequently large percentages of the swarm are stagnant.
+Iterations where the entire swarm is stagnant average less than 2 in 1000
+iterations, unless the swarm converges beyond machine precision.  When graphs
+of the best function value flatline, what really has happened is a contraction
+of the particles' velocity, not stagnation in the technical sense.
+
+We now show with what percentage a particle takes each of these branches for
+three different topologies and four different functions.  All of our statistics
+are from swarms of 240 particles.  Brief experimentation showed that other
+swarm sizes had similar statistics.  We ran 750 iterations on all combinations
+of functions and topologies except for the functions Griewank and Rastrigin
+with the Complete topology.  We found that those runs frequently converged past
+machine precision after 500 iterations, and that led to erroneously high values
+for the probability of stagnation.  Instead we ran for only 450 iterations on
+those two combinations.  All of our results were averaged over 20 runs of the
+algorithm; thus the probabilities presented are the averages of 3.6 million
+trials for the branch taken (2.16 million for the two with only 450
+iterations).  Table~\ref{tab:stats} contains the results, with the branch
+number coming from Table~\ref{tab:branches}.  We have already defined the
+functions Sphere, Griewank, and Rastrigin.  Rosenbrock is defined as
+$f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.
+
+The probabilities presented in Table~\ref{tab:stats} are interesting in and of
+themselves and could probably be used to better understand the characteristics
+of various topologies.  It is notable that there is small variation between
+functions in any given topology, but the variation across topologies is far
+greater.  However, our concern is with speculative evaluation.  We are
+interested in predicting the branch that any given particle will take at a
+particular iteration.
+
+From Table~\ref{tab:stats} we can see that with the Random topology, we can
+pick the first branch, corresponding to stagnation, and be right around 70\% of
+the time.  With the Ring topology, we would be right 50\% of the time.
+Branches four and five really correspond to several actual branches, as all of
+the neighbors are lumped together.  The 20\% probability of taking branch four
+with the Ring topology can be split into two branches, as there are only two
+neighbors.  It also turns out that the neighbor that last updated the
+neighborhood best is the most likely to update it next time, so keeping track
+of that information could be fruitful in trying to predict that branch.
+
+The statistics for the Complete topology are less promising (as there are 240
+neighbors that branch four splits into, instead of two), though we do not use
+that topology with speculative evaluation anyway.
+
+Because there are a few branches with very high probabilities, we can have hope
+that cutting out some of the evaluations that have low probability will lead to
+an increase in overall performance.
+
+\subsubsection{What if we guessed wrong?}
+
+When not all of the branches are evaluated, there is some probability of not
+evaluating the branch that was actually taken by the original particle.  We 
+need to decide what to do in this case.
+
+One possibility is to leave the particle as it is, not accepting any of the
+speculative evaluations, because none of the were correct.  This leads to
+particles being at different iterations, as some particles guess correctly
+while others do not.  Thus we lose exact compatibility with the original PSO,
+though this particular relaxation is nothing new; asynchronous adaptations of
+PSO do the same thing.  As an aside, it is equivalent in this case to simply
+increment the iteration number of particles which fail to correctly predict
+their branch.  This keeps the iteration number constant across all particles,
+simplifying the work needed to be done when dynamic topologies are involved.
+
+Another possibility is to pick the best child, as described above.  This
+ignores the fact that the branch might have been wrong; we do not care, because
+we just want the child that had the best value, anyway.  In almost all of the
+experiments that we ran, it turned out that picking the best child performed
+better than keeping the particle back an iteration.
+
+\subsection{More than one iteration ahead}
+
+\subsection{Tokens}
+
 \section{Results}
@@
 
-In our experiments we compared our speculative PSO algorithm to the standard
-PSO algorithm.  To use the same number of processors for each algorithm,
-speculative evaluation needs a smaller swarm size than standard PSO.  For each
-of the topologies we used, a particle has two neighbors other than itself.  As
-shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
-particle.  With one evaluation needed for the original, non-speculative
-particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
-the number of particles in the speculative swarm.  The extra evaluations done
-in speculative evaluation would instead be particles in regular PSO, so we
-compare swarms of size $p_s$ in speculative evaluation with swarms of size
-$8p_s$ in regular PSO.
-
-We use a Random topology in our speculative algorithm, as the Complete topology
-leads to an explosion of speculative evaluations.  However, if speculative
-evaluation were not being performed, it is likely that the Complete topology
-would be used, ignoring the communication cost of Complete over Random.  To be
-fair in our comparisons, for the standard PSO algorithm we compare to both the
-Random topology with two neighbors and to the Complete topology.  All functions
-have 20 dimensions, and all results shown are averages over 20 runs.  We use
-240 processors in our experiments, so speculative algorithms use 30 particles
-and standard PSO uses 240 particles.  For ease of labeling, we call our
-speculative algorithm SEPSO, the variant SEPSO Pick Best, and standard PSO just
-PSO.
-
-First we look at the function Sphere, the simplest of common benchmark
-functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
-has no local optima and is best optimized in terms of function evaluations with
-a small swarm using a Complete topology.  Accordingly, we show results for a
-small Complete swarm in \fig{sphere1}.  In comparing SEPSO with PSO, we can see
-that SEPSO clearly beats PSO with a Random topology.  However, the performance
-between SEPSO and PSO Complete is not as encouraging---PSO Complete performs
-slightly better.  But SEPSO Pick Best handily outperforms all other algorithms
-in this problem.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere1.eps}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
-  iterations per round, while standard PSO algorithms use 240 processors and do
-  one iteration per round.}
-  \label{fig:sphere1}
-\end{figure}
-
-Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
-\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
-\right) + 1$.  It is generally recommended to use the Ring topology when
-optimizing the Griewank function, as Complete is prone to premature convergence
-on a local optima.  We show results in \fig{griewank1} for swarms of size 30
-and 240 using the Ring topology. One can see in the figure that both SEPSO
-algorithms initially outperformed PSO, but got stuck in a local optima and were
-eventually outperformed themselves.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank1.eps}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
-  iterations per round, while standard PSO algorithms use 240 processors and do
-  one iteration per round.}
-  \label{fig:griewank1}
-\end{figure}
-
-\fig{griewank1} is slightly deceptive, however, as with Griewank some of the
-runs found the global optimum, 0, while some did not.  A more accurate plot for
-this function is one showing what percentage of the runs have found the global
-optimum at each iteration.  \fig{griewank2} shows such a plot.  A swarm of size
-30 performing the PSO algorithm gets stuck a little less than half of the time,
-while 240 particles provides enough exploration to find the global optimum
-every time.  Surprisingly, using the SEPSO Pick Best algorithm with 30
-particles greatly improves the chance of success with Griewank, while also
-finding the global optimum in far fewer rounds of evaluations than standard
-PSO.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank2.eps}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.  Instead of showing average function value, we show the
-  percentage of runs that have found the global optimum by each iteration.}
-  \label{fig:griewank2}
-\end{figure}
-
-Lastly we look at a function that does not show much or any improvement with
-this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
-\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
-with Rastrigin, the more particles there are in the swarm, the lower function
-value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
-swarms get caught in local optima.  Because our speculative algorithms have a
-significantly smaller swarm size, they get stuck at higher values while the
-larger swarms performing regular PSO continue to improve the best value found.
-\fig{rastrigin} shows the results graphically.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
-  \caption{Function Rastrigin with a swarm that uses 240 processors per round
-  of evaluations.  Speculative algorithms have a swarm of 30 particles and do
-  two iterations per round, while standard PSO algorithms use 240 processors
-  and do one iteration per round.}
-  \label{fig:rastrigin}
-\end{figure}
-
 \section{Conclusions}
 \label{sec:conclusion}
-
-We have described how parallel implementations of particle swarm optimization
-can be modified to allow additional processors to increase the number of
-iterations of the algorithm performed, instead of merely adding more particles
-to the swarm.  We have detailed several different implementations of the
-algorithm, depending on the parallel PSO framework being used.  Using our
-modifications, the original PSO algorithm is exactly reproduced two iterations
-at a time.  This technique requires more function evaluations per iteration
-than regular PSO, but for some functions still performs better when run in a
-parallel environment.
-
-We have also described a method for making use of extra speculative evaluations
-that performs incredibly well on some functions.
-
-Though space has limited our presentation of results to a very small sample of
-the interesting and relevant applications of our technique, we believe the
-results shown are sufficient motivation for the further exploration of
-speculative evaluation in PSO.
-
-Our results show that performing speculative evaluation instead of increasing
-swarm size is only beneficial for some functions at the swarm sizes we used.
-However, we would venture that for all functions there is some point at which
-increasing the swarm size provides relatively little benefit.  At that point,
-additional processors should be used to speed up the algorithm instead of add
-additional particles.
-
-There are some functions for which almost no exploration needs to be done;
-Sphere is an example of such a function, as would be a function where the
-general area of the optimal point is known, but the exact point needs to be
-found.  For such functions the best use of processors is to have a small swarm
-performing speculative evaluation with our second method, where all speculative
-evaluations are used.
-
-There are other functions for which there is a point where the algorithm has
-done ``enough'' exploration and consistently finds the global optimum.  Adding
-additional particles to the swarm is akin to increasing its capacity for
-exploration.  We have seen that for a 20-dimensional Griewank function, a
-number of particles greater than 30 but less than 240 has ``enough''
-exploration and finds the optimal value every time.  For these functions, one
-should use processors to increase the swarm size until enough exploration has
-been reached, then use additional processors to increase the number of
-iterations performed.
-
-There are still other functions for which it seems there is never enough
-exploration, such as the Rastrigin function.  It has been shown that up to 4000
-particles there is no point at which ``enough'' exploration has been
-done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
-cannot be easily prescribed and depends largely on the desired outcome.  It is
-likely that the best thing to do with those functions is to use all processors
-to add particles to the swarm and do no speculative evaluation.
 
@@
 add particles to the swarm or parallelize the evaluation of the objective
-function.  We anticipate that many future advancements can be made.  For
-example, our technique required speculatively evaluating every possible next
-position of each particle.  The number of speculative evaluations could
-potentially be greatly reduced by using statistics for which branch of motion
-particles most often follow.  Only the subset of the most probable next
-locations would then be evaluated.  Using an asynchronous PSO variant may also
-be helpful in this case, allowing particles to be at different iterations in
-case the correct branch was not guessed and needs to be evaluated separately.
+function.  We anticipate that many future advancements can be made.
 
-Another possible advancement would be to reduce communication costs by dropping
-some of the inter-processor messages in distributed frameworks.  The PSO
-algorithm would have to be adapted to work with less information, perhaps not
-requiring that each processor have completely accurate information about what
-neighboring particles are doing.
+We have speculated (pardon the pun\dots) that our ``pick the best child''
+method simply draws a number of samples from the next iteration of the sampling
+distribution of PSO and keeps one of them.  It would then be expected that the
+sampling distribution of that method is very similar to the distribution of the
+original PSO.  This remains to be proven, or even demonstrated empirically.
 
-Still another possibility is to view the set of possible evaluations as an
-infinite tree.  Each level in the tree is an iteration, and the branches on
-each level represent the possible next positions of the particle.  One need not
-simply evaluate all of one level for each particle---perhaps going two or three
-iterations ahead on the most likely path would perform better.
+We gave a passing glance to some very interesting statistical characteristics
+of PSO.  Our motivation was to use the statistics to improve our speculative
+algorithm, but it seems there are some nuggets in the numbers that could be
+used to improve our understanding of the behavior of PSO in general, and what
+role a topology plays.  A potentially very fruitful avenue of future work would
+be to further analyze the branch statistics in PSO.  There are some particles
+that seem to drive the swarm and others that are stagnant almost the entire
+time.  Such information could be used in a TRIBES-like fashion to determine
+which particles should be removed from the swarm, perhaps starting them close
+to well-performing particles.
+
+We turned what used to be a very simple algorithm with only one major parameter
+to tweak (the topology) into a big hairy mess with countless possibilities for
+speculative evaluations.  Perhaps future research will be able to narrow down
+the endless possibilities into the few best ones, making the algorithm simple
+again.
 

### 2009-12-30 great_works/semester_summary.tex
Wrote a couple more documents for my portfolio.  Just the great works left, I think

\documentclass[onecolumn, 10pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{Semester Summary}
\author{}
\date{}

\begin{document}
\maketitle

\subsubsection*{Fall 2004}
\begin{itemize}
  \item Started out double majoring in physics and music.
  \item Quickly switched to double majoring in physics and computer science.
  \item Started off college with 18 credits.
\end{itemize}
\subsubsection*{Winter 2005}
\begin{itemize}
  \item Took an ambitious 21 credits and managed to keep a 4.0.
  \item Received a mission call to Alabama, speaking Spanish.
  \item Member of the Honors Student Advisory Council.
\end{itemize}
\subsubsection*{Summer 2007}
\begin{itemize}
  \item First classes back from the mission, eased back into physics and math.
  \item Started doing physics research for my physics senior thesis.
\end{itemize}
\subsubsection*{Fall 2007}
\begin{itemize}
  \item Tried 18 credits, but I had forgotten too much math, so I dropped a 
	physics class.
  \item Realized I didn't like physics research, and that I wanted to do
	computer science more than physics.
  \item Dropped the physics major to a minor.
  \item Completed all coursework for both a physics and a math minor.
  \item Started working in a computer science research lab with Dr. Kevin Seppi.
\end{itemize}
\subsubsection*{Winter 2008}
\begin{itemize}
  \item Completed several GE courses.
  \item Continued work with Dr. Seppi, doing particle swarm optimization (PSO).
  \item Used PSO to help a physics professor with his crystallography research.
\end{itemize}
\subsubsection*{Spring/Summer 2008}
\begin{itemize}
  \item Presented the PSO/crystallography research at the BYU Spring Research
	Conference.
  \item Took 4 computer science classes and jumped a year ahead in the program.
  \item Began some experiments that led to a publication in a PSO conference.
\end{itemize}
\subsubsection*{Fall 2008}
\begin{itemize}
  \item Did a computer science capstone project with my brother and his friend.
  \item Took my first natural language processing class, starting an interest
	that I plan to pursue in my masters degree.
  \item Met and married my wife---maintained my 4.0 GPA while engaged.
  \item Co-authored a publication accepted at the Congress on Evolutionary
	Computation, a prestigious venue for PSO research.
\end{itemize}
\subsubsection*{Winter 2009}
\begin{itemize}
  \item Took a graduate-level class in natural language processing, which led
	to a submission to the BYU Studies 50th Anniversary Symposium.
  \item Decided on a topic for my honors thesis.
  \item Wrote most of the code for my honors thesis experiments.
\end{itemize}
\subsubsection*{Spring/Summer 2009}
\begin{itemize}
  \item Did a summer internship in California at Google.
\end{itemize}
\subsubsection*{Fall 2009}
\begin{itemize}
  \item Ran all of the experiments I needed for my honors thesis.
  \item Wrote my honors thesis.
  \item Audited an ancient Greek class that I really enjoyed.
  \item Took another graduate-level CS class, in multi-agent systems and game
	theory.
\end{itemize}
\subsubsection*{Winter 2010 (Projected)}
\begin{itemize}
  \item Defended my thesis.
  \item Finished all of my GEs.
  \item My wife had our first child.
  \item I graduated with honors and a 4.0 GPA!
\end{itemize}

\end{document}

### 2009-12-30 great_works/service.tex
Wrote a couple more documents for my portfolio.  Just the great works left, I think

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{Service Experience}
\author{}
\date{}

\begin{document}
\maketitle

My most memorable service experiences at BYU were in two separate areas.  The
first is volunteer tutoring.  I tutored a number of people over the years, in
physics, math, and computer science.  Tutoring helped my solidify concepts that
I had learned, and it let me share the talents God has given me with those who
needed my help.  One particular example of service in tutoring comes to mind.
My friend James was taking CS 142, the introductory computer science class, and
was struggling quite a bit.  I helped him on a few of the labs earlier in the
semester, but he was having a very difficult time with the last lab.  I found
him in the open labs the evening before the lab was due and told him I could
help him.  The next day he came for help, but it was late and I didn't have
much time.  I helped him for a few minutes then had to go to class.  After
class my wife and I went to the Mormon Tabernacle Choir Christmas Concert with
the Honors program, so I could go help him then.  But I called him and asked if
he needed help, and helped him over the phone for an hour or so, before the
concert.  After the concert I called him again, and he still needed help, so I
talked him through it some more on the way home.  We got back at 11:30 PM, the
lab was due at midnight, and he still hadn't finished.  The bus dropped us off
at the Maeser building, and James was in the Talmage building and was running
out of time.  I ran to the Talmage building and was able to help him finish and
pass off the lab five minutes before it was due.

I tell that story because the appreciation that James showed to me affected me
a lot.  He was incredibly grateful and said some very nice things, and it
reminded me of the value of giving service.  Not only was James very happy that
he was able to finish his assignment, I was happy because I knew that I had
done something good for someone else.  That experience reinforced for me the
value of giving selfless service and made me want to continue serving.

The second area of service that has been very memorable for me has been in
volunteering at the Missionary Training Center.  My wife served a mission in
Slovenia, and there are very few Slovene speakers in Provo.  The MTC needs
volutneers to come every week so that the missionaries heading to Slovenia have
someone to practice with.  My wife went every week before I met her, and since
we met I have gone with her.  At first I didn't understand what was being said,
but then I started to learn Slovene, in large part so I could help out at the
MTC.  For more than a year I have gone to the MTC almost every week to help the
missionaries practice their teaching and language skills.  This experience was
formative for me because now I speak another language, having learned it
through my service.  It also has been wonderful for me in strengthening my
testimony of and knowledge about the gospel of Jesus Christ.

\end{document}

### 2010-01-10 great_works/lit_gorilla.tex
One more down.  Two to go...

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{setspace}
\doublespacing

\title{}
\author{Matthew Gardner}
\date{}

\begin{document}

\textbf{Literature Response}

\begin{tabular}{ll}
  Name:&Matthew Gardner \\
  Work:&\emph{Enoch and the Gorilla} \\
  Author:&Flannery O'Connor \\
  Date Published:& 1971 \\
  Date Read:& January 8, 2010 \\
\end{tabular}

\section*{Historical Context}

Flannery O'Connor was born in 1925 in Georgia.  She was a Catholic living in
the Protestant Bible Belt, and that affected the way that she wrote.  She often
was critical of the secular society around her and felt a return to Catholicism
would be beneficial.  She spent the last 14 years of her life struggling to
fight off lupus, but eventually died of it in 1964, at 39 years of age.
\emph{Enoch and the Gorilla} was published only in 1971, in a collection of
O'Connor's short stories published posthumously under the title \emph{The
Complete Stories}.  However, the story seems to be taken almost directly out of
O'Connor's first major work, a novel called \emph{Wise Blood}, written in 1952.
I was unable to discover why it was published separately as a short
story---perhaps \emph{Wise Blood} started out as several related short stories.
When people were looking through her manuscripts and notes after her death,
they probably found the story on its own and published it with \emph{The
Complete Stories}.

\section*{Critical Analysis}

O'Connor's \emph{Enoch and the Gorilla} seems to be a harsh commentary on the
failures of secular society.  The main character is an 18-year-old boy named
Enoch Emery, who was abandoned by his father and thus forced to live and work
in the city to survive.  The story details a transformation that takes place in
Enoch, and through that transformation reveals the inadequacies of popular
culture.  Through her descriptions of people in the story, the cause of Enoch's
transformation, and its outcome, O'Connor shows secular life to be deficient.

The characters in \emph{Enoch and the Gorilla} reflect O'Connor's negative
views of life in a worldly society.  Every single character in the story is
rude, unkind, or otherwise of miserable character.  From the stingy old
landlady at the beginning of the story that lends Enoch her old, broken
umbrella, to the man in a gorilla suit who tells him to ``go to hell,'' and the
man at the end of the story who wordlessly abandons his girlfriend to be
attacked by a gorilla, all of the people portrayed are seriously flawed, 
including Enoch himself.

Enoch is shown to be lacking confidence in himself and to be quite rude.  He
sees a poster advertising a movie with a gorilla, with a promotion at the
theater featuring the gorilla itself shaking hands with viewers of the movie.
He goes to shake the gorilla's hand, thinking that ``the opportunity to insult
a successful ape came from the hand of Providence.''  While in front of the ape
he gets so nervous that he stammers something largely incoherent, and the ape
turns out to be merely a man in a suit when he gets mad at the boy babbling at
him.  The humiliation Enoch experiences eventually causes a transformation in
him.  He gains self-confidence, and for the first time O'Connor uses the words
``happy'' and ``kind'' in her story, and they are describing Enoch.  The cause
of this transformation, however, was a movie star, an epitome of worldliness.
It seems for a moment that O'Connor is promoting secular living, as it is the
cause of Enoch's newfound happiness.

The outcome of Enoch's transformation tells a different story.  Enoch decides
he wants to be like the gorilla, having people line up to shake his hand.  So
he finds the place the gorilla is, sneaks into the truck, kills the man who was
wearing the suit and takes the suit for himself.  He goes into the woods,
buries his old clothes, puts on the suit, and bounces around growling like a
gorilla.  He literally becomes a beast, both in his murder of the man and in
his insanity in tramping around like a gorilla.  He walks to the highway and
finds a couple sitting on a bench.  He walks up to them to shake their hand,
hoping to find the success he was searching for, but the two run away, leaving
him to stare out into the city, alone again in his misery.  What started out as
a happy transformation turned into another failure, worse than before.

\emph{Enoch and the Gorilla} uses some rather gruesome imagery to provide a
stinging rebuke of secular society.  Even the phrase she uses to describe the
gorilla in the story, ``successful ape,'' hints at how O'Connor feels about
those who are ``successful'' in the secular sense.  The society failed Enoch
and drove him out of it, turning him into a beast in the process and denying
him the happiness he hoped to find.  Likewise O'Connor sees life in a purely
secular world to be meaningless, discouraging, and unable to fulfil our hopes
and our dreams.

\section*{Personal Reflection}

The first time I read \emph{Enoch and the Gorilla}, I was really quite confused
by it.  I had to read some biographical information on Wikipedia about her and
her stories before I began to see any kind of point to the story.  When I read
it again, I could see some meaning, though I really have no way of knowing
whether the meaning I saw was intended by O'Connor or not.  It is for that
reason that I have very mixed feelings about fiction as social commentary.  I
prefer straightforward arguments and commentary, rather than convoluted ones
that are easily misunderstood.  I do appreciate the stories more when I
understand the purpose for which they were written, but most of the time I have
to be told what they mean---it is never readily apparent to me.

\end{document}

### 2010-01-12 honors_thesis/thesis/sepsodata/plots/ackley.py
Removed some old data that I replaced and do not use anymore

@@ -1,49 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 500
-SWARMSIZE = 50
-
-SEPSO_FILE = '../sepso_reproduce/ackley/rand50particles'
-PICKBEST_FILE = '../sepso_pickbest/ackley/rand50particles'
-STANDARD_RAND = '../standardpso/ackley/rand200particles'
-STANDARD_RING = '../standardpso/ackley/ring200particles'
-STANDARD_COMPLETE = '../standardpso/ackley/complete200particles'
-
-sepso_data = PSOData(open(SEPSO_FILE))
-pickbest_data = PSOData(open(PICKBEST_FILE))
-rand_data = PSOData(open(STANDARD_RAND))
-complete_data = PSOData(open(STANDARD_COMPLETE))
-sepso_points = []
-pickbest_points = []
-rand_points = []
-complete_points = []
-for i in xrange(1, ITERATIONS):
-    evaluations = i * SWARMSIZE
-    sepso_points.append((evaluations, sepso_data.average(i*2)))
-    pickbest_points.append((evaluations, pickbest_data.average(i*2)))
-    rand_points.append((evaluations, rand_data.average(i)))
-    complete_points.append((evaluations, complete_data.average(i)))
-
-plot = Plot()
-plot.xmin = SWARMSIZE
-plot.xmax = ITERATIONS * SWARMSIZE
-plot.ymin = 10 ** -10
-plot.ymax = 100
-plot.ylogscale = 10
-plot.xlabel = 'Function Evaluations'
-plot.ylabel = 'Best Function Value'
-
-plot.append(Points(sepso_points, title='SEPSO', style='lines'))
-plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
-plot.append(Points(rand_points, title='PSO Random', style='lines'))
-plot.append( Points(complete_points, title='PSO Complete', style='lines'))
-
-plot.show()
-plot.write('ackley.gpi')
-
-
-# vim: et sw=4 sts=4

### 2010-01-12 honors_thesis/thesis/sepsodata/plots/complete_200.py

@@ -1,59 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 500
-SWARMSIZE = 200
-
-FUNCTIONS = ['sphere', 'rastrigin', 'rosenbrock', 'griewank', 'ackley']
-
-FUNCTION_RANGE = [(10 ** -25, 10 ** 5), (0, 300), (10, float(10 ** 10)),
-        (10 ** -3, 500), (10 ** -10, 100)]
-
-FUNCTION_LOGSCALE = [10, 0, 10, 10, 10]
-
-SEPSO_FILE = '../sepso_reproduce/function/rand50particles'
-PICKBEST_FILE = '../sepso_pickbest/function/rand50particles'
-STANDARD_RAND = '../standardpso/function/rand200particles'
-STANDARD_COMPLETE = '../standardpso/function/complete200particles'
-
-for f, function in enumerate(FUNCTIONS):
-    sepso_data = PSOData(open(SEPSO_FILE.replace('function', function)))
-    pickbest_data = PSOData(open(PICKBEST_FILE.replace('function', function)))
-    rand_data = PSOData(open(STANDARD_RAND.replace('function', function)))
-    complete_data = PSOData(open(STANDARD_COMPLETE.replace('function',
-            function)))
-    sepso_points = []
-    pickbest_points = []
-    rand_points = []
-    complete_points = []
-    for i in xrange(1, ITERATIONS):
-        evaluations = i * SWARMSIZE
-        sepso_points.append((evaluations, sepso_data.average(i*2)))
-        pickbest_points.append((evaluations, pickbest_data.average(i*2)))
-        rand_points.append((evaluations, rand_data.average(i)))
-        complete_points.append((evaluations, complete_data.average(i)))
-
-    plot = Plot()
-    plot.xmin = SWARMSIZE
-    plot.xmax = ITERATIONS * SWARMSIZE
-    plot.ymin = FUNCTION_RANGE[f][0]
-    plot.ymax = FUNCTION_RANGE[f][1]
-    plot.ylogscale = FUNCTION_LOGSCALE[f]
-    plot.xlabel = 'Function Evaluations'
-    plot.ylabel = 'Best Function Value'
-    plot.title = 'Complete topology, 50 and 200 particles, function: %s' \
-            % function
-
-    plot.append(Points(sepso_points, title='SEPSO', style='lines'))
[…]

### 2010-01-12 honors_thesis/thesis/sepsodata/plots/complete_50.py

@@ -1,59 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 500
-SWARMSIZE = 52
-
-FUNCTIONS = ['sphere', 'rastrigin', 'rosenbrock', 'griewank', 'ackley']
-
-FUNCTION_RANGE = [(10 ** -20, 10 ** 5), (0, 300), (10, float(10 ** 10)),
-        (10 ** -3, 500), (10 ** -5, 100)]
-
-FUNCTION_LOGSCALE = [10, 0, 10, 10, 10]
-
-SEPSO_FILE = '../sepso_reproduce/function/rand13particles'
-PICKBEST_FILE = '../sepso_pickbest/function/rand13particles'
-STANDARD_RAND = '../standardpso/function/rand52particles'
-STANDARD_COMPLETE = '../standardpso/function/complete52particles'
-
-for f, function in enumerate(FUNCTIONS):
-    sepso_data = PSOData(open(SEPSO_FILE.replace('function', function)))
-    pickbest_data = PSOData(open(PICKBEST_FILE.replace('function', function)))
-    rand_data = PSOData(open(STANDARD_RAND.replace('function', function)))
-    complete_data = PSOData(open(STANDARD_COMPLETE.replace('function',
-            function)))
-    sepso_points = []
-    pickbest_points = []
-    rand_points = []
-    complete_points = []
-    for i in xrange(1, ITERATIONS):
-        evaluations = i * SWARMSIZE
-        sepso_points.append((evaluations, sepso_data.average(i*2)))
-        pickbest_points.append((evaluations, pickbest_data.average(i*2)))
-        rand_points.append((evaluations, rand_data.average(i)))
-        complete_points.append((evaluations, complete_data.average(i)))
-
-    plot = Plot()
-    plot.xmin = SWARMSIZE
-    plot.xmax = ITERATIONS * SWARMSIZE
-    plot.ymin = FUNCTION_RANGE[f][0]
-    plot.ymax = FUNCTION_RANGE[f][1]
-    plot.ylogscale = FUNCTION_LOGSCALE[f]
-    plot.xlabel = 'Function Evaluations'
-    plot.ylabel = 'Best Function Value'
-    plot.title = 'Complete topology, 13 and 52 particles, function: %s' \
-            % function
-
-    plot.append(Points(sepso_points, title='SEPSO', style='lines'))
[…]

### 2010-01-12 honors_thesis/thesis/sepsodata/plots/griewank.py

@@ -1,42 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 1000
-SWARMSIZE = 50
-
-SEPSO_FILE = '../sepso_reproduce/griewank/ring50particles'
-PICKBEST_FILE = '../sepso_pickbest/griewank/ring50particles'
-STANDARD_RAND = '../standardpso/griewank/ring200particles'
-
-sepso_data = PSOData(open(SEPSO_FILE))
-pickbest_data = PSOData(open(PICKBEST_FILE))
-ring_data = PSOData(open(STANDARD_RAND))
-sepso_points = []
-pickbest_points = []
-ring_points = []
-for i in xrange(1, ITERATIONS):
-    evaluations = i * SWARMSIZE
-    sepso_points.append((evaluations, sepso_data.average(i*2)))
-    #pickbest_points.append((evaluations, pickbest_data.average(i*2)))
-    ring_points.append((evaluations, ring_data.average(i)))
-
-plot = Plot()
-plot.xmin = SWARMSIZE
-plot.xmax = ITERATIONS * SWARMSIZE
-plot.ymin = 10 ** -9
-plot.ymax = 500
-plot.ylogscale = 10
-plot.xlabel = 'Function Evaluations'
-plot.ylabel = 'Best Function Value'
-
-plot.append(Points(sepso_points, title='SEPSO', style='lines'))
-#plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
-plot.append(Points(ring_points, title='PSO Ring', style='lines'))
-
-plot.show()
-plot.write('griewank1.gpi')
-
-# vim: et sw=4 sts=4

### 2010-01-12 honors_thesis/thesis/sepsodata/plots/griewank2.py

@@ -1,42 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 1000
-SWARMSIZE = 400
-
-SEPSO_FILE = '../sepso_reproduce/griewank/ring400particles'
-PICKBEST_FILE = '../sepso_pickbest/griewank/ring400particles'
-STANDARD_RAND = '../standardpso/griewank/ring1600particles'
-
-sepso_data = PSOData(open(SEPSO_FILE))
-pickbest_data = PSOData(open(PICKBEST_FILE))
-ring_data = PSOData(open(STANDARD_RAND))
-sepso_points = []
-pickbest_points = []
-ring_points = []
-for i in xrange(1, ITERATIONS):
-    evaluations = i * SWARMSIZE
-    sepso_points.append((evaluations, sepso_data.average(i*2)))
-    #pickbest_points.append((evaluations, pickbest_data.average(i*2)))
-    ring_points.append((evaluations, ring_data.average(i)))
-
-plot = Plot()
-plot.xmin = SWARMSIZE
-plot.xmax = ITERATIONS * SWARMSIZE
-plot.ymin = 10 ** -8
-plot.ymax = 500
-plot.ylogscale = 10
-plot.xlabel = 'Function Evaluations'
-plot.ylabel = 'Best Function Value'
-
-plot.append(Points(sepso_points, title='SEPSO', style='lines'))
-#plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
-plot.append(Points(ring_points, title='PSO Ring', style='lines'))
-
-plot.show()
-plot.write('griewank2.gpi')
-
-# vim: et sw=4 sts=4

### 2010-01-12 honors_thesis/thesis/sepsodata/plots/rastrigin.py

@@ -1,52 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 1000
-SWARMSIZE = 200
-
-SEPSO_FILE = '../sepso_reproduce/rastrigin/rand200particles'
-PICKBEST_FILE = '../sepso_pickbest/rastrigin/rand200particles'
-STANDARD_RAND = '../standardpso/rastrigin/rand800particles'
-STANDARD_RING = '../standardpso/rastrigin/ring800particles'
-STANDARD_COMPLETE = '../standardpso/rastrigin/complete800particles'
-
-sepso_data = PSOData(open(SEPSO_FILE))
-#pickbest_data = PSOData(open(PICKBEST_FILE))
-rand_data = PSOData(open(STANDARD_RAND))
-#ring_data = PSOData(open(STANDARD_RING))
-complete_data = PSOData(open(STANDARD_COMPLETE))
-sepso_points = []
-pickbest_points = []
-rand_points = []
-#ring_points = []
-complete_points = []
-for i in xrange(1, ITERATIONS):
-    evaluations = i * SWARMSIZE
-    sepso_points.append((evaluations, sepso_data.average(i*2)))
-    #pickbest_points.append((evaluations, pickbest_data.average(i*2)))
-    rand_points.append((evaluations, rand_data.average(i)))
-    #ring_points.append((evaluations, ring_data.average(i)))
-    complete_points.append((evaluations, complete_data.average(i)))
-
-plot = Plot()
-plot.xmin = SWARMSIZE
-plot.xmax = ITERATIONS * SWARMSIZE
-plot.ymin = 0
-plot.ymax = 100
-plot.xlabel = 'Function Evaluations'
-plot.ylabel = 'Best Function Value'
-
-plot.append(Points(sepso_points, title='SEPSO', style='lines'))
-#plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
-plot.append(Points(rand_points, title='PSO Random', style='lines'))
-#plot.append(Points(ring_points, title='PSO Ring', style='lines'))
-plot.append( Points(complete_points, title='PSO Complete', style='lines'))
-
-plot.show()
-plot.write('rastrigin.gpi')
-
-
-# vim: et sw=4 sts=4

### 2010-01-12 honors_thesis/thesis/sepsodata/plots/ring_200.py

@@ -1,50 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 500
-SWARMSIZE = 200
-
-FUNCTIONS = ['sphere', 'rastrigin', 'rosenbrock', 'griewank', 'ackley']
-
-FUNCTION_RANGE = [(10 ** -20, 10 ** 5), (0, 300), (10, float(10 ** 10)),
-        (10 ** -5, 500), (10 ** -10, 100)]
-
-FUNCTION_LOGSCALE = [10, 0, 10, 10, 10]
-
-SEPSO_FILE = '../sepso_reproduce/function/ring50particles'
-PICKBEST_FILE = '../sepso_pickbest/function/ring50particles'
-STANDARD_FILE = '../standardpso/function/ring200particles'
-
-for f, function in enumerate(FUNCTIONS):
-    sepso_data = PSOData(open(SEPSO_FILE.replace('function', function)))
-    pickbest_data = PSOData(open(PICKBEST_FILE.replace('function', function)))
-    standard_data = PSOData(open(STANDARD_FILE.replace('function', function)))
-    sepso_points = []
-    pickbest_points = []
-    standard_points = []
-    for i in xrange(1, ITERATIONS):
-        evaluations = i * SWARMSIZE
-        sepso_points.append((evaluations, sepso_data.average(i*2)))
-        pickbest_points.append((evaluations, pickbest_data.average(i*2)))
-        standard_points.append((evaluations, standard_data.average(i)))
-
-    plot = Plot()
-    plot.xmin = SWARMSIZE
-    plot.xmax = ITERATIONS * SWARMSIZE
-    plot.ymin = FUNCTION_RANGE[f][0]
-    plot.ymax = FUNCTION_RANGE[f][1]
-    plot.ylogscale = FUNCTION_LOGSCALE[f]
-    plot.xlabel = 'Function Evaluations'
-    plot.ylabel = 'Best Function Value'
-    plot.title = 'Ring topology, 50 and 200 particles, function: %s' % function
-
-    plot.append(Points(sepso_points, title='SEPSO', style='lines'))
-    plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
-    plot.append(Points(standard_points, title='Standard PSO', style='lines'))
-
-    plot.show()
-
-# vim: et sw=4 sts=4

### 2010-01-12 honors_thesis/thesis/sepsodata/plots/ring_50.py

@@ -1,50 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 500
-SWARMSIZE = 52
-
-FUNCTIONS = ['sphere', 'rastrigin', 'rosenbrock', 'griewank', 'ackley']
-
-FUNCTION_RANGE = [(10 ** -20, 10 ** 5), (0, 300), (100, float(10 ** 10)),
-        (10 ** -2, 500), (10 ** -2, 100)]
-
-FUNCTION_LOGSCALE = [10, 0, 10, 10, 10]
-
-SEPSO_FILE = '../sepso_reproduce/function/ring13particles'
-PICKBEST_FILE = '../sepso_pickbest/function/ring13particles'
-STANDARD_FILE = '../standardpso/function/ring52particles'
-
-for f, function in enumerate(FUNCTIONS):
-    sepso_data = PSOData(open(SEPSO_FILE.replace('function', function)))
-    pickbest_data = PSOData(open(PICKBEST_FILE.replace('function', function)))
-    standard_data = PSOData(open(STANDARD_FILE.replace('function', function)))
-    sepso_points = []
-    pickbest_points = []
-    standard_points = []
-    for i in xrange(1, ITERATIONS):
-        evaluations = i * SWARMSIZE
-        sepso_points.append((evaluations, sepso_data.average(i*2)))
-        pickbest_points.append((evaluations, pickbest_data.average(i*2)))
-        standard_points.append((evaluations, standard_data.average(i)))
-
-    plot = Plot()
-    plot.xmin = SWARMSIZE
-    plot.xmax = ITERATIONS * SWARMSIZE
-    plot.ymin = FUNCTION_RANGE[f][0]
-    plot.ymax = FUNCTION_RANGE[f][1]
-    plot.ylogscale = FUNCTION_LOGSCALE[f]
-    plot.xlabel = 'Function Evaluations'
-    plot.ylabel = 'Best Function Value'
-    plot.title = 'Ring topology, 13 and 52 particles, function: %s' % function
-
-    plot.append(Points(sepso_points, title='SEPSO', style='lines'))
-    plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
-    plot.append(Points(standard_points, title='Standard PSO', style='lines'))
-
-    plot.show()
-
-# vim: et sw=4 sts=4

### 2010-01-12 honors_thesis/thesis/sepsodata/plots/sphere.py

@@ -1,47 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import division
-from psodata import PSOData
-from evilplot import Plot, Points, RawData
-
-ITERATIONS = 500
-SWARMSIZE = 52
-
-SEPSO_FILE = '../sepso_reproduce/sphere/rand13particles'
-PICKBEST_FILE = '../sepso_pickbest/sphere/rand13particles'
-STANDARD_RAND = '../standardpso/sphere/rand52particles'
-STANDARD_COMPLETE = '../standardpso/sphere/complete52particles'
-
-sepso_data = PSOData(open(SEPSO_FILE))
-pickbest_data = PSOData(open(PICKBEST_FILE))
-rand_data = PSOData(open(STANDARD_RAND))
-complete_data = PSOData(open(STANDARD_COMPLETE))
-sepso_points = []
-pickbest_points = []
-rand_points = []
-complete_points = []
-for i in xrange(1, ITERATIONS):
-    evaluations = i * SWARMSIZE
-    sepso_points.append((evaluations, sepso_data.average(i*2)))
-    #pickbest_points.append((evaluations, pickbest_data.average(i*2)))
-    rand_points.append((evaluations, rand_data.average(i)))
-    complete_points.append((evaluations, complete_data.average(i)))
-
-plot = Plot()
-plot.xmin = SWARMSIZE
-plot.xmax = ITERATIONS * SWARMSIZE
-plot.ymin = 10 ** -20
-plot.ymax = 10 ** 5
-plot.ylogscale = 10
-plot.xlabel = 'Function Evaluations'
-plot.ylabel = 'Best Function Value'
-
-plot.append(Points(sepso_points, title='SEPSO', style='lines'))
-#plot.append(Points(pickbest_points, title='SEPSO Pick Best', style='lines'))
-plot.append(Points(rand_points, title='PSO Random', style='lines'))
-plot.append( Points(complete_points, title='PSO Complete', style='lines'))
-
-plot.show()
-plot.write('sphere.gpi')
-
-# vim: et sw=4 sts=4

### 2010-01-12 honors_thesis/gecco_paper/paper.tex
Added another branch of this stuff in the form of a GECCO paper - I should have found a better way to do this...

--- 2009-12-29 honors_thesis/thesis/paper.tex
@@
-\documentclass[conference,letterpaper]{IEEEtran}
+\documentclass[letterpaper]{sig-alt-release2}
 
@@
 \usepackage{psfrag}
-
-\IEEEoverridecommandlockouts
 
@@
 keeping the best position found instead of the position that PSO actually would
-have taken.  We show results that for a number of functions, speculative
-evaluation can have dramatic improvements over adding additional particles to
-the swarm.
+have taken.  We show that for a number of functions, speculative evaluation
+gives dramatic improvements over adding additional particles to the swarm.
 
@@
 asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05},
-and formulating PSO in Google's MapReduce framework \cite{mcnabb-cec07}.
+and algorithms based on Google's MapReduce framework \cite{mcnabb-cec07}.
 Parallelizing the evaluation of the objective function can also be done in some
-cases, though that is not an adaption of the PSO algorithm and thus is not the
-focus of this paper.
+cases, though that is not an adaption of the PSO algorithm itself and thus is
+not the focus of this paper.
 
@@
 techniques commonly used in processors.  Modern processors, when faced with a
-branch on which they must wait, guess which way the branch will go and start
-executing.  If the processor guessed right, execution is much farther ahead
-than if it had just idly waited on the branch.  If it guessed wrong, execution
-restarts right where it would have been anyway.  This is called speculative
-execution or branch prediction.
+branch on which they must wait (e.g. a memory cache miss), guess which way the
+branch will go and start executing.  If the processor guessed right, execution
+is much farther ahead than if it had just idly waited on the branch.  If it
+guessed wrong, execution restarts right where it would have been anyway.  This
+is called speculative execution or branch prediction.
 
@@
 
-Because the motion of particles in PSO depends only on the \emph{positions} of
-the particle and its neighbors, not the \emph{value} of the objective function
-at those positions, speculative evaluation is possible.  We do not have to wait
-for a particle's position to be evaluated to know where it might move to next,
-so we can compute the set of all possible next positions and send them to be
-evaluated along with the particle's current position.  Once the current
-position of each particle has been evaluated, we can determine which of the
-speculatively evaluated positions matches the branch taken by the particle.  We
-have thus evaluated two iterations at once.  This can easily get unwieldy if
-the set of next possible positions for each particle is large, but a wise
-choice of topology limits that set to a reasonable size.
+Speculative evaluation is possible in PSO because the motion of particles
+depends only on the \emph{positions} of the particle and its best neighbor, not
+the \emph{value} of the objective function at those positions.  We do not have
+to wait for a particle's function evaluation to complete in order to know where
+it might move to next, so we can compute the set of all possible next positions
+and send them to be evaluated along with the particle's current position.  Once
+the current position of each particle has been evaluated, we can determine
+which of the speculatively evaluated positions matches the branch taken by the
+particle.  We have thus evaluated two iterations at once.  This can easily get
+unwieldy if the set of next possible positions for each particle is large, but
+a wise choice of topology limits that set to a reasonable size.
 
@@
 those speculative evaluations may be to incorporate the information into the
-swarm, no longer exactly reproducing PSO but perhaps producing better results.
+swarm, no longer exactly reproducing PSO but in some instances producing better
+results.
 
@@
 birds and other animals. It is a social algorithm that depends on interaction
-between particles to quickly and consistently find the optimal solution to some
-objective function.
+between particles to quickly and consistently find the optimal solution to a
+given objective function.
 
@@
 a particle cannot be the source of its $\nbest$ update while not updating its
-$\pbest$.  Table~\ref{tab:evals} lists all possible combinations of updates
-that can be seen by a particle with two neighbors other than itself, which is
-the case in all of the topologies we use.  If all of these speculative
-positions are evaluated in parallel along with the original, non-speculative
-position of the particle at time $t-1$, then one of the speculative positions
-has to be the correct one, and two iterations have been evaluated at once.
+$\pbest$.  Thus we have $2(n+1)-1$ evaluations.  Table~\ref{tab:evals} lists
+all possible combinations of updates that can be seen by a particle with two
+neighbors other than itself, which is the case in all of the topologies we use.
+If all of these speculative positions are evaluated in parallel along with the
+original, non-speculative position of the particle at time $t-1$, then one of
+the speculative positions has to be the correct one, and two iterations have
+been evaluated at once.
 
 \begin{table}
-  \caption{All possible updates for a particle with two neighbors}
+  \caption{All possible updates for a particle with two neighbors.}
   \label{tab:evals}
   \centering
-  \begin{tabular}{c|c}
+  \begin{tabular}{cc}
 	$\pbest$ updated&Source of $\nbest$ update\\
 	\hline
+	\hline
 	No update&No update\\
+	\hline
+	No update&Neighbor 1\\
+	\hline
+	No update&Neighbor 2\\
+	\hline
 	Updated&No update\\
-	No update&Neighbor 1\\
-	No update&Neighbor 2\\
+	\hline
+	Updated&Neighbor 1\\
+	\hline
+	Updated&Neighbor 2\\
+	\hline
 	Updated&The particle itself\\
-	Updated&Neigbhor 1\\
-	Updated&Neighbor 2\\
+	\hline
   \end{tabular}
@@
 
-Because $2(n+1)$ speculative evaluations must be performed for each particle,
+Because $2(n+1)-1$ speculative evaluations must be performed for each particle,
 the choice of topology is an important one.  The use of the Complete topology,
-where every particle is a neighbor to every other particle, would require on
-the order of $p^2$ speculative evaluations per iteration, where $p$ is the
-swarm size.  Clearly it is much more desirable to have a sparse topology, where
-$O(np)$ is much smaller than $O(p^2)$.  However, some functions are better
-optimized with the Complete topology and the quick spread of information it
-entails than with sparse topologies.  Accordingly, we use the Random topology
-described in~\cite{mcnabb-cec09}, which has been shown to effectively simulate
-the Complete topology.  In that topology, 2 neighbors are chosen at random from
-the swarm for each particle at each iteration.  Because a particle is also
-always its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report
-the results for performing speculative evaluation using both the Ring topology
+where every particle is a neighbor to every other particle, would require
+$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
+Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
+much smaller than $O(p^2)$.  However, some functions are better optimized with
+the Complete topology and the quick spread of information it entails than with
+sparse topologies.  Accordingly, we use the Random topology described
+in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
+Complete topology.  In that topology, 2 neighbors are chosen at random from the
+swarm for each particle at each iteration.  Because a particle is also always
+its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report the
+results for performing speculative evaluation using both the Ring topology
 (where $n$ again is $3$) and the Random topology on a number of common
@@
 corresponding to one of the rows in Table~\ref{tab:evals}.  $\n_t$ is a
-neighbor of particle $\p_t$.  A set of particles is $\pset$, $\sset$, or
-$\nset$, whereas a single particle is simply $\p$, $\s$, or $\n$.
+neighbor of particle $\p_t$.  Sets of particles are given by $\pset$, $\sset$,
+or $\nset$, whereas single particles are simply $\p$, $\s$, or $\n$.
 
@@
 a scheme, the master generates the positions to evaluate speculatively and
-decides which one to accept for each particle.
+decides which one to accept for each particle.  The outline of the procedure is
+given in \alg{centralized}.
 
@@
 position and value with information about each of its neighbors.  This is
-simply the original PSO algorithm.
+simply the original PSO algorithm, and corresponds to steps 1--5 in
+\alg{centralized}.
 
@@
 Both of those items are now known, and the child that matches the branch is
-selected.
+selected (step 7 in \alg{centralized}).
 
@@
 guessed.  The child only knows the position and needs the value in order to
-make future comparisons between neighborhood best positions.
+make future comparisons between neighborhood best positions (step 8).
 
@@
 $\pset_{t+1}$.  That set of particles can subsequently be used to produce the
-sets $\noeval{\pset}_{t+2}$ and $\noeval{\sset}_{t+3}$, and the process repeats
-itself.
-
-The algorithm is outlined in \alg{centralized}.
+sets $\noeval{\pset}_{t+2}$ and $\noeval{\sset}_{t+3}$ (steps 1 and 2 in
+\alg{centralized}), and the process repeats itself.
 
@@
 Messages such as values and positions for the neighborhood best are sent
-between processors.  There may be some machine that collects information from
-all of the particles and outputs the result of the algorithm, though that
+between processors.  There may still be some machine that collects information
+from all of the particles and outputs the result of the algorithm, though that
 machine's importance is much less than in centralized algorithms.
@@
 without giving the speculative particles the same treatment as actual
-particles, as the speculative children only live for one iteration.  Perhaps
-that can be done by assigning each particle a set of machines instead of a
+particles, as the speculative children only live for one iteration.  One way
+that can be done is by assigning each particle a set of machines instead of a
 single machine, and the particle directs its extra machines to evaluate its
-speculative children.  No matter the distributed framework used the message
-passing should be similar, and that is what we describe here.
+speculative children.  However the distributed framework functions, the same
+information needs to be passed between particles, and we describe here the
+messages each particle needs to receive to perform speculative evaluation.
 
@@
 In some distributed frameworks, like the MapReduce framework that we used in
-our experiments, rounds of message passing can be expensive.  The method
-just described uses three rounds of message passing for every two iterations.
-Accordingly, we describe a method that requires only one round of communication
-for each pair of iterations.  A lot more messages are needed, but that is
-sometimes more desirable than synchronizing all of the machines three times.
+our experiments, rounds of message passing can be expensive.  The method just
+described uses three rounds of message passing for every two iterations
+(corresponding to steps 2, 5 and 10 in \alg{centralized}).  Accordingly, we
+describe a method that requires only one round of communication for each pair
+of iterations, which happens at step 5 of \alg{centralized}.  Many more
+messages are needed, but that is sometimes more desirable than synchronizing
+all of the machines three times.
 
@@
 Figures~\ref{fig:messages2}~and~\ref{fig:messages3}, putting them together
-to make $\p_{t+1}$, as all the particle needs is it's neighborhood best to be
+to make $\p_{t+1}$, as all the particle needs is its neighborhood best to be
 updated.
@@
 
-This may seem like an inordinate amount of work, and in some instances it is.
-However, some parallel frameworks necessitate this type of message passing, so
-we have described how speculative evaluation can be performed in such
-frameworks.
+This may seem like an inordinate amount of work, and with some distributed PSO
+frameworks it is.  However, other parallel frameworks necessitate this type of
+message passing, so we have described how speculative evaluation can be
+performed in those circumstances.
 
@@
 reconstruct information about iteration $t+2$.  Also, this method of message
-passing again requires the use of random seeds if the topology is random.
+passing again requires the use of random seeds if the topology is random, so
+that each processor computes the same neighbors for a particle as all other
+processors.
 
@@
 best.  That creates a problem for speculative evaluation---there are
-potentially more than $2(n+1)$ possible next positions, increasing the amount
+potentially more than $2(n+1)-1$ possible next positions, increasing the amount
 of work that must be done to perform the second iteration at the same time as
@@
 calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
-still only $2(n+1)$ possible next positions, and the problem is avoided.
+still only $2(n+1)-1$ possible next positions, and the problem is avoided.
 
@@
 
-In performing speculative evaluation as we have described it, $2(n+1)$
+In performing speculative evaluation as we have described it, $2(n+1)-1$
 speculative evaluations are done per particle, while all but one of them are
@@
 In our experiments we compared our speculative PSO algorithm to the standard
-PSO algorithm.  To use the same number of processors for each algorithm,
-speculative evaluation needs a smaller swarm size than standard PSO.  For each
-of the topologies we used, a particle has two neighbors other than itself.  As
-shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
+PSO algorithm.  At each iteration of the algorithms, we use one processor to
+perform one function evaluation for one particle, be it speculative or
+non-speculative.  The speculative algorithm actually performs two iterations of
+PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
+evaluations.  For benchmark functions with very fast evaluations this may not
+be the best use of parallelism in PSO.  But for many real-world applications,
+the objective function takes on the order of minutes or more to evaluate; in
+such cases our assumption is perfectly reasonable.
+
+To use the same number of processors for each algorithm, speculative evaluation
+needs a smaller swarm size than standard PSO.  For each of the topologies we
+used, a particle has two neighbors other than itself.  As shown in
+Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
 particle.  With one evaluation needed for the original, non-speculative
@@
 
-We use a Random topology in our speculative algorithm, as the Complete topology
-leads to an explosion of speculative evaluations.  However, if speculative
-evaluation were not being performed, it is likely that the Complete topology
-would be used, ignoring the communication cost of Complete over Random.  To be
-fair in our comparisons, for the standard PSO algorithm we compare to both the
-Random topology with two neighbors and to the Complete topology.  All functions
-have 20 dimensions, and all results shown are averages over 20 runs.  We use
-240 processors in our experiments, so speculative algorithms use 30 particles
-and standard PSO uses 240 particles.  For ease of labeling, we call our
-speculative algorithm SEPSO, the variant SEPSO Pick Best, and standard PSO just
-PSO.
+Where the Complete topology would normally be used, we use a Random topology in
+our speculative algorithm, as Complete leads to an explosion of speculative
+evaluations.  However, if speculative evaluation were not being performed, it
+is likely that the Complete topology would be used, ignoring the communication
+cost of Complete over Random.  To be fair in our comparisons, for the standard
+PSO algorithm we compare to both the Random topology with two neighbors and to
+the Complete topology.  All functions have 20 dimensions, and all results shown
+are averages over 20 runs.  We use 240 processors in our experiments, so
+speculative algorithms use 30 particles and standard PSO uses 240 particles.
+For ease of labeling, we call our speculative algorithm SEPSO, the variant
+SEPSO Pick Best, and standard PSO just PSO.
 

### 2010-01-12 applications/byu_statement_of_intent.tex
Grad school application stuff

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{Statement of Intent}
\author{Matthew Gardner}
\date{}

\begin{document}
\maketitle

I started out my post-secondary education double majoring at BYU in physics and
music.  I quickly switched to physics and computer science, then later realized
I was a lot more interested in computer science than in physics.  I ended up
with minors in physics and math because of my early forays into physics, and I
think that has been a great preparation for some of the more rigorous material
I have encountered and will encounter in advanced computer science classes and
research.

During my sophomore year, I started working in the Applied Machine Learning lab
with Dr.  Kevin Seppi, with whom I have worked for the past two years.  I moved
through the CS program quickly, leaving me time to take two graduate-level
courses as an undergraduate.  Because I have been moving so quickly, by the
time I get my bachelors degree I will hopefully have two and possibly three
publications.  I was the second author on a conference paper that was accepted
at the Congress on Evolutionary Computation 2009, and I am the first author on
a paper we are submitting to the Genetic and Evolutionary Computation
Conference 2010.  I am also in the process of expanding my conference paper to
a journal paper that we hope to submit to Transactions on Evolutionary
Computation in February.  While the journal paper certainly will be not
published by the time I graduate,  at least it will have been submitted as an
undergraduate paper.

I am applying to the BYU Masters program because I have really enjoyed my time
here as an undergraduate.  I think Dr. Seppi is a great advisor and want to 
keep working with him for my masters degree.  Because of the extra classes I
have taken, I also will have three classes to apply towards a masters degree
that I probably would have difficulty transferring to another school.  As I
already have an advisor and three of the eight classes done, I hope to finish
my masters degree by April 2011.  My current intended area of focus during my
masters degree is natural language processing, and probably unsupervised
morphology in that field.  Because that is my intended topic, I will probably
also be working closely with Dr. Ringger during my stay here.

\end{document}

### 2010-01-12 great_works/science_aquarium.tex
One more down, only two left!

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{setspace}
\doublespacing

\title{}
\author{Matthew Gardner}
\date{}

\begin{document}

\textbf{Life Science Response}

\begin{tabular}{ll}
  Name:&Matthew Gardner \\
  Site:&Aquarium of the Pacific \\
  Location:&Long Beach, California \\
  Date Established:&1998 \\
  Date Visited:&12/17/2008 \\
\end{tabular}

\section*{Historical Context}

The Aquarium of the Pacific opened to the public in 1998.  It has three major
areas that each highlight a different habitat of the Pacific Ocean---Southern
California and Baja, the Northern Pacific, and the Tropical Pacific.  It also
has smaller exhibits that change periodically.  I grew up in Southern
California and made several trips to the aquarium when I was younger.  More
recently, I visited it with my wife over Christmas break in 2008, and each time
the experience has been fascinating for me.

\section*{The Experience}

The first part of the experience that was enlightening to me was seeing the
incredible diversity of life that exits in the ocean, most of it largely unseen
to us on the land.  There's an odd beauty to a large bloom of jellyfish, or a
motley collection of exotic marine life.  One of my favorites is the
humuhumunukunukuapua'a, or trigger fish, which is the state fish of Hawai'i.
Going to zoos or aquariums always gives me an incredible appreciation for the
beauty and complexity of God's creations.

Along with the beauty of the creatures on display at the aquarium, the rarity
of many of the animals struck me in a couple of ways.  First, there were some
animals that live at the bottom of the ocean, where relatively little
exploration has taken place.  There were some giant crabs that seemed massive
to me, as well as some really strange looking seahorses.  Seeing those animals
made me wonder about what other forms of life exist in the deep ocean that we
know only very little about, like the giant squid, deep sea angler, or
long-nosed chimaera.  Not only is there beauty in God's creations, there's a
breadth that we have only begun to explore.

The rarity of the animals also made me think about how we are impacting our
environment, and whether or not that will adversely affect the habitats of
these creatures.  Some, it seems, are common and widespread enough that any
relatively minor change in the environment wouldn't do them too much harm.
Others seem delicate and exotic, and destroying the one reef where they are
found might see them completely disappear.  If it turns out that the earth does
warm significantly, many of these creatures may some day exist only in pictures
and memories.  That would be a sad day indeed.  However, history is replete
with examples of animals that failed to adapt to their environment and became
extinct---the adaptability of life is one of the wonders of God's creations,
even if it means some species must die out to make room for new ones.

My trip to the Aquarium of the Pacific filled me with both a child-like wonder
at the beauty all around us and a feeling of responsibility to treat our
environment a little nicer.  It was a great experience.

\end{document}

### 2010-01-12 honors_thesis/journal_paper/paper.tex
I think I am finally done with revisions to my honors thesis/GECCO paper

@@
 a particle cannot be the source of its $\nbest$ update while not updating its
-$\pbest$.  Table~\ref{tab:evals} lists all possible combinations of updates
-that can be seen by a particle with two neighbors other than itself, which is
-the case in all of the topologies we use.  If all of these speculative
-positions are evaluated in parallel along with the original, non-speculative
-position of the particle at time $t-1$, then one of the speculative positions
-has to be the correct one, and two iterations have been evaluated at once.
+$\pbest$.  Thus we have $2(n+1)-1$ evaluations.  Table~\ref{tab:evals} lists
+all possible combinations of updates that can be seen by a particle with two
+neighbors other than itself, which is the case in all of the topologies we use.
+If all of these speculative positions are evaluated in parallel along with the
+original, non-speculative position of the particle at time $t-1$, then one of
+the speculative positions has to be the correct one, and two iterations have
+been evaluated at once.
 
@@
   \centering
-  \begin{tabular}{c|c}
+  \begin{tabular}{cc}
 	$\pbest$ updated&Source of $\nbest$ update\\
 	\hline
+	\hline
 	No update&No update\\
+	\hline
+	No update&Neighbor 1\\
+	\hline
+	No update&Neighbor 2\\
+	\hline
 	Updated&No update\\
-	No update&Neighbor 1\\
-	No update&Neighbor 2\\
+	\hline
+	Updated&Neighbor 1\\
+	\hline
+	Updated&Neighbor 2\\
+	\hline
 	Updated&The particle itself\\
-	Updated&Neigbhor 1\\
-	Updated&Neighbor 2\\
+	\hline
   \end{tabular}
@@
 We ran experiments to compare our speculative PSO algorithm to the standard
-PSO algorithm.  To use the same number of processors for each algorithm,
-speculative evaluation needs a smaller swarm size than standard PSO.  For each
-of the topologies we used, a particle has two neighbors other than itself.  As
-shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
+PSO algorithm.  At each iteration of the algorithms, we use one processor to
+perform one function evaluation for one particle, be it speculative or
+non-speculative.  The speculative algorithm actually performs two iterations of
+PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
+evaluations.  For benchmark functions with very fast evaluations this may not
+be the best use of parallelism in PSO.  But for many real-world applications,
+the objective function takes on the order of minutes or more to evaluate; in
+such cases our assumption is perfectly reasonable.
+
+To use the same number of processors for each algorithm, speculative evaluation
+needs a smaller swarm size than standard PSO.  For each of the topologies we
+used, a particle has two neighbors other than itself.  As shown in
+Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
 particle.  With one evaluation needed for the original, non-speculative
@@
 
-We use a Random topology in our speculative algorithm, as the Complete topology
-leads to an explosion of speculative evaluations.  If speculative evaluation
-were not being performed, it is possible that the Complete topology would be
-used.  However, the Complete topology also requires a very large amount of
-interprocessor communication in parallel PSO, so it is still quite possible
-that Random would be used even with standard PSO.  But, to be fair in our
-comparisons, we compare to the standard PSO algorithm using both the Random
-topology with two neighbors and the Complete topology.  All functions have 20
-dimensions, and all results shown are averages over 20 runs.  We use 240
-processors in our experiments, so our speculative algorithm uses 30 particles
-and standard PSO uses 240 particles.  For ease of labeling, we call our
-speculative algorithm SEPSO and standard PSO just PSO, with the topology
-following.
+Where the Complete topology would normally be used, we use a Random topology in
+our speculative algorithm, as Complete leads to an explosion of speculative
+evaluations.  If speculative evaluation were not being performed, it is
+possible that the Complete topology would be used.  However, the Complete
+topology also requires a very large amount of interprocessor communication in
+parallel PSO, so it is still quite possible that Random would be used even with
+standard PSO.  But, to be fair in our comparisons, we compare to the standard
+PSO algorithm using both the Random topology with two neighbors and the
+Complete topology.  All functions have 20 dimensions, and all results shown are
+averages over 20 runs.  We use 240 processors in our experiments, so our
+speculative algorithm uses 30 particles and standard PSO uses 240 particles.
+For ease of labeling, we call our speculative algorithm SEPSO and standard PSO
+just PSO, with the topology following.
 
@@
 we could limit ourselves to one speculative evaluation per particle instead of
-$2(n+1)$.  While we cannot with certainty predict the correct branch every
+$2(n+1)-1$.  While we cannot with certainty predict the correct branch every
 time, we can look at the statistics of the behavior of particles in general and
@@
 
-The statistics for the Complete topology are less promising (as there are 240
-neighbors that branch four splits into, instead of two), though we do not use
-that topology with speculative evaluation anyway.
-
-Because there are a few branches with very high probabilities, we can have hope
-that cutting out some of the evaluations that have low probability will lead to
-an increase in overall performance.
+The statistics for the Complete topology are less promising, as there are 240
+neighbors that branch four splits into, instead of two.  Pruning does, however,
+allow for the possibility of using the Complete topology in speculative
+evaluation while avoiding the explosion in the number of evaluations it would
+otherwise entail.  Though pruning makes its use possible, we still found the
+communication costs in our parallel environment prohibitive, so we do not
+present any results for the Complete topology here.
+
+With the other topologies, however, there are a few branches with very high
+probabilities, so we can have hope that cutting out some of the evaluations
+that have low probability will lead to an increase in overall performance.
 
@@
 
+We need not simply produce speculative children for the next iteration.  We can
+view all possible speculative evaluations for a particle as an infinite tree
+with branching factor $2(n+1)-1$.  As we have already seen that doing one full
+level of the tree produces too many extra evaluations to be profitable, it is
+incredibly unlikely that doing two full iterations would produce decent
+results.  But, if the idea of speculating more than one iteration ahead is
+combined with wisely pruning the possible evaluations based on branch
+statistics, we can use just a few extra evaluations to go two or three
+iterations ahead on the most likely branches.
+
+The question of which branches to take in this infinite tree is an intriguing
+one that we can only begin to explore here.  If the branch corresponding to
+stagnation has a 75\% chance of being taken, as in the Random topology with
+most functions, we could speculate three iterations ahead on that branch and
+still have a 42\% chance of predicting correctly.  However, intuition would say
+that perhaps it is better to hope that the particle is productive instead of
+stagnant, so a branch where the particle updates its personal best might be
+more fruitful to try.  In our experiments we try just one of the countless
+possibilities, but that one turns out to work extremely well.  More work is
+needed to compare the different branching possibilities on various functions.
+
 \subsection{Tokens}
+
+One issue remains.  What if we really want a swarm of 50 particles, but we have
+230 processors to use?  The methods we have thus far presented do the same
+thing for every particle, and thus would require some multiple of 50
+processors---we would have 30 extra processors with nothing to do.  It might
+very well be the best idea to simply adjust the swarm size slightly, but we
+present another method that allows particles to behave differently, some doing
+more evaluations than others.
+
+If we use the concept of tokens commonly used in multirobot stuff (see, for
+example, \dots), we can specify a number of tokens available for particles to
+use, where each token corresponds to a speculative evaluation.  Whichever
+method is used to which evaluations to do, they are prioritized so that
+particles with only one or two tokens perform the most valuable speculative
+evaluations.  Particles that perform well keep their tokens, particles that
+perform poorly send one token (if they have one) to another random particle in
+the swarm.
+
+We choose to define ``perform poorly'' as a failure to update the personal best
+when the original particle's position is evaluated.  Clearly there are other
+possible methods of deciding when to give up tokens, but many of the methods
+that sound the best are difficult in distributed environments, like giving
+tokens to that particles that most recently found the globally best positions.
+The use of a central machine to decide which particles get tokens would help to
+facilitate such methods.
+
+There are thus two distinct uses for tokens.  First, they could be used to try
+to evenly spread extra evaluations due to an incapatible number of particles
+and processors.  In cursory experimentation, we found that when only going one
+iteration ahead keeping the number of evaluations per particle as even as
+possible worked much better than trying to prioritize particles and let some
+guess correctly more often because they have more tokens. (horrible explanation\dots)
+
+The other, more compelling reason to use tokens is when speculating many
+iterations ahead.  Particles that are going in good directions, as evidenced by
+relatively frequent updates to the particle's personal best, keep their tokens
+and are able to speculate up to three or four iterations.  Particles that
+aren't doing as well lose their tokens and probably contribute to the
+exploration of the algorithm, for which speculative evaluation doesn't help
+much.
+
+As in the previous section, we can only offer a very preliminary exploration of
+the possibilities of using tokens to improve the allocation of speculative
+evaluations to particles.
 

### 2010-01-14 great_works/music_mendelssohn.tex
Started work on the last two responses

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{setspace}
\doublespacing

\title{Mendelssohn's Italian Symphony: A Music Response}
\author{Matthew Gardner}
\date{}

\begin{document}

\textbf{Music Response}

\begin{tabular}{ll}
  Name:&Matthew Gardner \\
  Work:&Mendelssohn's Symphony No. 4 (The Italian Symphony) \\
  Artist:&Felix Mendelssohn \\
  Date:&1833 \\
\end{tabular}

\section*{Historical Context}

Felix Mendelssohn was an early Romantic composer who lived from 1809 to 1847.
He went on a tour to Europe from 1829 to 1831, and a number of his compositions
originated on that tour, including his fourth symphony, called ``The Italian
Symphony.''  He wrote the movements of the symphony in response to various
experiences he had in Italy; the second, slow movement is based on a religious
ceremony Mendelssohn viewed in Naples.

The Italian Symphony's original performance was conducted by Mendelssohn in
London, and it was very well received.  However, Mendelssohn was not entirely
happy with it and revised it in 1837, intending even further revisions.  The
symphony was not even published until 1851, after Mendelssohn's death.

\section*{Critical Analysis}

\section*{Personal Reflection}

\end{document}

### 2010-01-14 honors_thesis/thesis/paper.tex

@@
-\documentclass[conference,letterpaper]{IEEEtran}
+\documentclass[oneside,honors]{honors}
 
@@
 \usepackage{psfrag}
-
-\IEEEoverridecommandlockouts
+\pagestyle{plain}
 
@@
 
-\title{\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization%
-\thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department
-of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602
-(phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}%
-}
+\TitleTop{Speculative Evaluation in}
+\TitleBottom{Particle Swarm Optimization}
+\Year{2010}
+\Month{April}
+\degree{Bachelor of Science}
 
-\date{}
-
-\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-
+\Author{Matthew Gardner}
+\Advisor{Kevin Seppi}
+\Abstract{
 Particle swarm optimization (PSO) has previously been parallelized only by 
@@
 keeping the best position found instead of the position that PSO actually would
-have taken.  We show results that for a number of functions, speculative
-evaluation can have dramatic improvements over adding additional particles to
-the swarm.
+have taken.  We show that for a number of functions, speculative evaluation
+gives dramatic improvements over adding additional particles to the swarm.
+}
 
-\end{abstract}
+\begin{document}
+\frontmatter
+\makepreliminarypages
+
+\doublespace
+\tableofcontents
+\clearemptydoublepage
+
+\singlespace
+\listoffigures
+\clearemptydoublepage
+
+\doublespace
+
+\mainmatter
 
@@
 
-There are some functions for which almost no exploration needs to be done;
+There are some functions for which very little exploration needs to be done;
 Sphere is an example of such a function, as would be a function where the

### 2010-01-14 great_works/music_mendelssohn.tex
One more down.  Only one to go!  Finally...

@@
 
+The first movement of ``The Italian Symphony'' conjures images of trotting on a
+horse through the Italian countryside on a beautiful day.  The tempo and
+rhythms in the piece aren't too fast or energetic, making the ride a leisurely
+one.  Before long the music rises to a small climax, as if some big city has
+been seen and entered.  After the climax, the music calms a bit and switches to
+a number of different instruments all taking turns with new melodies only
+slightly related to the theme.  I imagine entering the city and talking with
+all sorts of different people.  Occasionally the original theme crops up
+throughout this part of the piece, as if we are telling those we are speaking
+with about our wonderful journey.
+
+When the solo instruments all quiet down, the theme comes back strong again, as
+if to say that we are leaving the city and starting out into the countryside
+again.  However, the music quickly turns into a minor key, playing a darker
+variation on the original theme.  Something bad has happened in our journey.
+Mendelssohn spends some time here developing the struggle, having some major
+lines battling with the growing dark mood, but just when it seems that the
+journey has gone sour the darkness recedes with a decrescendo to almost
+nothing.  A few halting sounds are made in the original major key, which grow
+steadily into the theme for our journey.
+
+The movement then starts it conclusion with a flute playing a quick moving line
+almost like a dance, interwoven with parts of the original theme.  I imagine a
+celebration after escaping whatever caused the dark mood.  There is even a
+little bit of the minor melody encroaching on the end, again as if telling a
+story about what had happened on the road.  The piece ends with a climax on
+the original theme.
+
+The other three movements all contain imagery just as vivid as the first,
+though I won't go into as much detail with them.  Knowing that the second
+movement was written in response to a religious procession greatly affected the
+way I listened to the piece.  The tone was a sort of minor, perhaps eluding to
+medieval modes that were used in Catholic services.  It was very solemn, and
+there were staccato strings in the background with the sense of walking.  There
+were some happy parts in the movement, however, showing that even though their
+religion was solemn, it did preach of joy.  There was no exciting climax at the
+end of the second movement.
+
+The third movement was also rather mellow, and it seemed almost contemplative
+to me.  Perhaps after watching the religious procession Mendelssohn tried to
+portray going off under a tree to think about what was just witnessed.  There
+were some small variations on the theme of the movement, but all of it was very
+calm and smooth.
+
+With the two previous movements being soft and mellow, the opening notes of the
+fourth movement are jarring.  The final movement is a combination of Italian
+dances, and the first one is exciting and energetic.  It seems like Mendelssohn
+varies between two main dances, one that is fast and sensual and one that is
+stately and cultured.  Towards the end he gives an effective image of a dance
+at court with the violins and the violas taking turns with fast, cheerful runs.
+You can really picture the women dancing with the violins, and the men dancing
+with the violas.  The short crescendo to the end of the piece gives a nice
+ending to a tour through Italy.
+
 \section*{Personal Reflection}
 
+I started out my college experience as a music major, accepted into the program
+with a scholarship.  I was intending to study music composition.  But because I
+switched my major to computer science, I had not had a deep musical experience
+in quite some time.  Reading about this piece and listening for the experiences
+that may have inspired Mendelssohn to write what he did was a great experience
+for me.  It reminded me of why I love music and wanted to be a composer.  There
+are not many pieces of music that I can clearly put imagery to, but
+Mendelssohn's Italian Symphony was one of them.
+
 \end{document}

### 2010-01-14 great_works/music_brahms.tex
Adding double spacing
Started work on the last two responses
Working on the last response

@@
 \usepackage{graphicx}
+\usepackage{setspace}
+\doublespacing
 
-\title{Mendelssohn's Italian Symphony: A Music Response}
+\title{}
 \author{Matthew Gardner}
@@
   Name:&Matthew Gardner \\
-  Work:&Mendelssohn's Symphony No. 4 (The Italian Symphony) \\
-  Artist:&Felix Mendelssohn \\
-  Date:&1833 \\
+  Work:&Piano Concerto No. 1 in D Minor, Op. 15 \\
+  Artist:&Johannes Brahms \\
+  Date:&1858 \\
 \end{tabular}
@@
 \section*{Historical Context}
+
+Johannes Brahms lived from 1833 to 1897 and was one of the central figures in
+the Romantic period of music.  He was heavily influenced by earlier periods,
+especially fellow German composers Johann Sebastian Bach and Ludwig van
+Beethoven.
 

### 2010-01-14 great_works/music_brahms.tex
Some more work on Brahms

@@
 especially fellow German composers Johann Sebastian Bach and Ludwig van
-Beethoven.
+Beethoven.  Brahms was a virtuoso pianist and many of his works were piano
+pieces.  Quite often he gave the initial performance of his piano works, 
+including his first piano concerto.
+
+Brahms started work on his first piano concerto in 1854, when he was 20 years
+old.  It was originally intended to be a sonata for two pianos, then turned
+into a symphony.  When he found the symphony to be unacceptable, he settled on
+making it a concerto.  He finished the work in 1859 and debuted the work
+himself.  The initially performance was incredibly poorly received---Brahms was
+hissed at by the audience.  However, over time, the work has come to be
+considered a masterpiece.
 

### 2010-01-14 great_works/music_brahms.tex
And I am done!  At least, I have all of the responses written.  I will probably need to revise some...

@@
 making it a concerto.  He finished the work in 1859 and debuted the work
-himself.  The initially performance was incredibly poorly received---Brahms was
-hissed at by the audience.  However, over time, the work has come to be
-considered a masterpiece.
+himself.  As it initially started out as a symphony, it is very long---at about
+50 minutes, it is one of the longest concertos ever written.  The initial
+performance was incredibly poorly received---Brahms was hissed at by the
+audience.  However, over time, the work has come to be considered a
+masterpiece.
 
@@
 
+Movement 1:
+
+The first movement begins with a long orchestral introduction, which is not
+surprising, given Brahms' original intent of writing a symphonic work.  After
+the introduction, the piano continues to develop the themes introduced in the
+introduction.  As the piece progresses, there is a lot of back and forth
+between the piano and the orchestra.  The orchestra takes a central role in the
+piece, developing themes and often playing without the piano, instead of merely
+acting as accompaniment for the main performer, as is often the case in a
+concerto.
+
+Some pieces of music capture a piece of artwork, or a scene, and conjure images
+of events to the listener; Mendelssohn is famous for his vivid compositions.
+Other pieces of music are more cathartic, simply taking you through emotions
+instead of bringing actual images to your mind.  To me, this piece was more of
+the second type.  There were moments of intense conflict followed by sweet
+resolution, and swells that rise and ebb and take you along with them.  The
+movement and dynamic contrast throughout the piece elicit a powerful emotional
+response.
+
+There are several parts throughout the first movement, and the entire piece,
+that feature either the horn or the timpani.  Those instruments are notoriously
+difficult to write for, and Brahms does a masterful job as using those
+instruments.
+
+The first movement is almost 25 minutes long, leaving a lot of time for development of the original theme.  I was amazed at how well Brahms kept me engaged
+throughout the piece.  There was enough variation in the theme that the music
+always seemed new and interesting to me, but it was similar enough to what came
+before that I was always reminded of the original theme.
+
+The second movement was slower and calmer, as is typical in multi-movement
+pieces like symphonies and concertos.  It gave a stark contrast with the first,
+energetic movement.  Throughout the piece, the music was peaceful, though the
+occasional discord leaves the listener with something to think about and
+retains the listener's interest.  The movement does not have very much dynamic
+contrast, most of the piece being soft, though Brahms compensates for that with
+beautiful melodic lines, particularly in the woodwinds.  The ending of the
+movement decrescendos down to almost nothing, with a lone timpani playing the
+final notes---another example of Brahms' fine use of the timpani.
+
+The final movement is a rondo, which are characterized by an oft-repeated
+central theme.  This rondo is no different, and the energetic theme is
+introduced right at the beginning in the piano.  The orchestra immediately
+echoes the theme.  Throughout the rest of the piece, it is almost always the
+orchestra that repeats the theme, while the piano plays complex variations on
+it.  The piece ends in the typical fashion with a large crescendo to a final 
+climax.
+
 \section*{Personal Reflection}
 
+Brahms' first piano concerto was a great experience for me a listen to,
+especially as I listened to it again to find things I missed the first time.
+The more I listen to music the more I notice in it, and it gives me a greater
+appreciation for the skill of both the composer and the performers.  I found
+the first movement of the concerto incredibly intriguing, especially the amount
+of variation that Brahms' could come up with while still basing the piece on a
+single theme (a point I mentioned earlier).  Because the piece was so long,
+however, I found myself losing interest during the middle of the second
+movement.  Perhaps that is why Brahms' original audience hissed at him during
+the performance.
+
 \end{document}

### 2010-01-25 honors_thesis/journal_paper/paper.tex
Some last updates to gecco before submitting, and other minor stuff

@@
   \includegraphics[width=.3\columnwidth]{messages1.eps}
-  \caption{The messages needed to produce $\p_{t}$.  Items without an arrow
-  pointing to them are messages the particle needs to receive.}
+  \caption{The production of $\p_{t}$ from the original particle
+  $\nonbest{\p}_{t}$ and the messages $\nonbest{\nset}_{t}$.}
   \label{fig:messages1}
@@
   \includegraphics[width=.5\columnwidth]{messages2.eps}
-  \caption{The messages needed to produce $\nonbest{\p}_{t+1}$.  Items without
-  an arrow pointing to them are messages the particle needs to receive.}
+  \caption{The production of $\nonbest{\p}_{t+1}$ from the original particle 
+  $\nonbest{\p}_{t}$, messages $\nonbest{\nset}_{t}$ and
+  $\nonbest{\sset}_{t+1}$, and intermediate particles.}
   \label{fig:messages2}
@@
   \includegraphics[width=.6\columnwidth]{messages3.eps}
-  \caption{The messages needed to produce $\nonbest{\nset}_{t+1}$.  Items
-  without an arrow pointing to them are messages the particle needs to receive.
-  $\nnset$ is the set of neighbors for each particle $\n$, and $\nsset$ is the
-  set of $\n$'s speculative children.  Note the similarity between this and
-  \fig{messages2}.}
+  \caption{The production of each $\nonbest{\n}_{t+1}$ from the original
+  particle $\nonbest{\n}_{t}$, messages $\nonbest{\nnset}_{t}$ and
+  $\nonbest{\nsset}_{t+1}$, and intermediate particles.  $\nnset$ is the set of
+  neighbors for each particle $\n$, and $\nsset$ is the set of $\n$'s
+  speculative children.  Note the similarity between this and \fig{messages2}.}
   \label{fig:messages3}
@@
   \includegraphics[width=.8\columnwidth]{messages4.eps}
-  \caption{The messages needed to produce $\p_{t+1}$.  Items
-  without an arrow pointing to them are messages the particle needs to receive.
-  Note that this is just a combination of \fig{messages2} and \fig{messages3}.}
+  \caption{The production of $\p_{t+1}$ from the original particle 
+  $\nonbest{\p}_{t}$, messages $\nonbest{\nset}_{t}$, $\nonbest{\sset}_{t+1}$,
+  $\nonbest{\nnset}_{t}$, and $\nonbest{\nsset}_{t+1}$, and intermediate
+  particles.  Note that this is just a combination of \fig{messages2} and
+  \fig{messages3}.}
   \label{fig:messages4}
@@
   \includegraphics[width=.5\columnwidth]{messages5.eps}
-  \caption{The messages needed to produce $\noeval{\sset}_{t+3}$. $\p_{t+1}$
-  and $\nset_{t+1}$ are both produced as in \fig{messages4}.}
+  \caption{The production of $\noeval{\p}_{t+2}$ and $\noeval{\sset}_{t+3}$
+  from $\p_{t+1}$ and $\n_{t+1}$, each of which are produced as in
+  \fig{messages4}.}
   \label{fig:messages5}

### 2010-01-25 honors_thesis/thesis/paper.tex
Some last updates to gecco before submitting, and other minor stuff

@@
 \Advisor{Kevin Seppi}
+\HonorsRep{}
 \Abstract{

### 2010-01-25 great_works/faculty_references.tex
A "faculty references" document for my honors portfolio

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{Faculty References}
\author{}
\date{}
\setlength{\parskip}{20pt}

\begin{document}
\maketitle

\begin{tabular}{ll}
  \textbf{Name:}&\textbf{Kevin Seppi} \\
  \emph{Phone:}&\emph{422-4619} \\
  \emph{Office:}&\emph{3330 TMCB} \\
  Classes taken:&CS 236 \\
\end{tabular}

Dr. Seppi is my honors thesis advisor, and I have worked with him as a research
assistant for the past two years.

\begin{tabular}{ll}
  \textbf{Name:}&\textbf{Mike Goodrich} \\
  \emph{Phone:}&\emph{422-6468} \\
  \emph{Office:}&\emph{2220 TMCB} \\
  Classes taken:&CS 252, CS 470, CS 670 \\
\end{tabular}

Dr. Goodrich is my honors thesis referee.

\begin{tabular}{ll}
  \textbf{Name:}&\textbf{Eric Ringger} \\
  \emph{Phone:}&\emph{422-7615} \\
  \emph{Office:}&\emph{3368 TMCB} \\
  Classes taken:&CS 479, CS 679 \\
\end{tabular}

I submitted some work I did with Dr. Ringger to the BYU Studies Symposium, and
some of my work with Dr. Seppi has been in collaboration with Dr. Ringger.

\end{document}

### 2010-02-25 byu_studies/poster_package/poster.tex
Started working on a poster for BYU Studies

\documentclass[portrait,final]{baposter}
%\documentclass[a4shrink,portrait,final]{baposter}
% Usa a4shrink for an a4 sized paper.

\tracingstats=2

\usepackage{times}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{relsize}
\usepackage{multirow}
\usepackage{bm}

\usepackage{graphicx}
\usepackage{multicol}

\usepackage{pgfbaselayers}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}

\usepackage{helvet}
%\usepackage{bookman}
\usepackage{palatino}

\newcommand{\captionfont}{\footnotesize}

\selectcolormodel{cmyk}

\graphicspath{{images/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Some math symbols used in the text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Format 
\newcommand{\Matrix}[1]{\begin{bmatrix} #1 \end{bmatrix}}
\newcommand{\Vector}[1]{\Matrix{#1}}
\newcommand*{\SET}[1]  {\ensuremath{\mathcal{#1}}}
\newcommand*{\MAT}[1]  {\ensuremath{\mathbf{#1}}}
\newcommand*{\VEC}[1]  {\ensuremath{\bm{#1}}}
\newcommand*{\CONST}[1]{\ensuremath{\mathit{#1}}}
\newcommand*{\norm}[1]{\mathopen\| #1 \mathclose\|}% use instead of $\|x\|$
\newcommand*{\abs}[1]{\mathopen| #1 \mathclose|}% use instead of $\|x\|$
\newcommand*{\absLR}[1]{\left| #1 \right|}% use instead of $\|x\|$

\def\norm#1{\mathopen\| #1 \mathclose\|}% use instead of $\|x\|$
\newcommand{\normLR}[1]{\left\| #1 \right\|}% use instead of $\|x\|$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Multicol Settings
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setlength{\columnsep}{0.7em}
\setlength{\columnseprule}{0mm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Save space in lists. Use this after the opening of the list
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\compresslist}{%
\setlength{\itemsep}{1pt}%
\setlength{\parskip}{0pt}%
\setlength{\parsep}{0pt}%
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Begin of Document
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Here starts the poster
%%%---------------------------------------------------------------------------
%%% Format it to your taste with the options
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Define some colors
\definecolor{silver}{cmyk}{0,0,0,0.3}
\definecolor{yellow}{cmyk}{0,0,0.9,0.0}
\definecolor{reddishyellow}{cmyk}{0,0.22,1.0,0.0}
\definecolor{black}{cmyk}{0,0,0.0,1.0}
\definecolor{darkYellow}{cmyk}{0,0,1.0,0.5}
\definecolor{darkSilver}{cmyk}{0,0,0,0.1}

\definecolor{lightyellow}{cmyk}{0,0,0.3,0.0}
\definecolor{lighteryellow}{cmyk}{0,0,0.1,0.0}
\definecolor{lighteryellow}{cmyk}{0,0,0.1,0.0}
\definecolor{lightestyellow}{cmyk}{0,0,0.05,0.0}

%%
\typeout{Poster Starts}
\background{
  \begin{tikzpicture}[remember picture,overlay]%
    \draw (current page.north west)+(-2em,2em) node[anchor=north west] {\includegraphics[height=1.1\textheight]{silhouettes_background}};
  \end{tikzpicture}%
}

\newlength{\leftimgwidth}
\begin{poster}%
  % Poster Options
  {
  % Show grid to help with alignment
  grid=no,
  % Column spacing
  colspacing=1em,
  % Color style
  bgColorOne=lighteryellow,
  bgColorTwo=lightestyellow,
  borderColor=reddishyellow,
  headerColorOne=yellow,
  headerColorTwo=reddishyellow,
  headerFontColor=black,
  boxColorOne=lightyellow,
  boxColorTwo=lighteryellow,
  % Format of textbox
  textborder=roundedleft,
  % Format of text header
  eyecatcher=no,
  headerborder=open,
  headerheight=0.08\textheight,
  headershape=roundedright,
  headershade=plain,
  headerfont=\Large\textsf, %Sans Serif
  boxshade=plain,
%  background=shade-tb,
  background=plain,
  linewidth=2pt
  }
  % Eye Catcher
  {\includegraphics[width=10em]{D1077}} % No eye catcher for this poster. (eyecatcher=no above). If an eye catcher is present, the title is centered between eye-catcher and logo.
  % Title
  {\sf %Sans Serif
  %\bf% Serif
  Expression Invariant Face Recognition with a 3DMM}
  % Authors
  {\sf %Sans Serif
  % Serif
  \vspace{1em}\{Brian.Amberg, Reinhard.Knothe, Thomas.Vetter\}@unibas.ch
  }
  % University logo
  {% The makebox allows the title to flow into the logo, this is a hack because of the L shaped logo.
    \makebox[8em][r]{%
      \begin{minipage}{16em}
        \hfill
        \includegraphics[height=2em]{msrlogo}
        \includegraphics[height=7.0em]{logo}
      \end{minipage}
    }
  }

  \tikzstyle{light shaded}=[top color=baposterBGtwo!30!white,bottom color=baposterBGone!30!white,shading=axis,shading angle=30]

  % Width of left inset image
     \setlength{\leftimgwidth}{0.78em+8.0em}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Now define the boxes that make up the poster
%%%---------------------------------------------------------------------------
%%% Each box has a name and can be placed absolutely or relatively.
%%% The only inconvenience is that you can only specify a relative position 
%%% towards an already declared box. So if you have a box attached to the 
%%% bottom, one to the top and a third one which should be in between, you 
%%% have to specify the top and bottom boxes before you specify the middle 
%%% box.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %
    % A coloured circle useful as a bullet with an adjustably strong filling
    \newcommand{\colouredcircle}[1]{%
      \tikz{\useasboundingbox (-0.2em,-0.32em) rectangle(0.2em,0.32em); \draw[draw=black,fill=baposterBGone!80!black!#1!white,line width=0.03em] (0,0) circle(0.18em);}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Contribution}{name=contribution,column=0,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
   {}We introduce a method for expression invariant face recognition. A
   generative 3D Morphable Model (3DMM) is used to separate identity and
   expression components. The expression removal results in increased
   recognition performance, even on difficult datasets, without a decrease in
   performance on expression-less datasets.

   It is applicable to any kind of input data, and was evaluated here on
   textureless range scans.
  \vspace{0.3em}
 }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Model}{name=model,column=0,below=contribution}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    The Model was learnt from 275 subjects. We used one neutral expression scan
    per identity and 150 expression scans of a subset of the subjects.

    The identity model is a linear model build from the neutral scans.
    \begin{align}
      \VEC f&=\VEC\mu + \MAT M_n\VEC\alpha_n\qquad.
    \end{align}
    For each of the 150 expression scans, we calculated an expression vector as
    the difference between the expression scan and the corresponding neutral
    scan of that subject.  This data is already mode-centered, if we regard the
    neutral expression as the natural mode of expression data. From these offset
    vectors an additional expression matrix $\MAT M_e$ was calculated, such that the complete linear Model is
    \begin{align}
      \VEC f&=\VEC\mu + \MAT M_n\VEC\alpha_n + \MAT M_e\VEC\alpha_e
    \end{align}
    The assumption here is, that the face and expression space are linearly
    independent, such that each face is represented by a unique set of
    coefficients.  
  \vspace{0.3em}
  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Fitting}{name=fitting,column=0,below=model}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    A Robust Nonrigid ICP method was used to fit the model to the data.
    Robustness was achieved by iteratively reweighting the correspondences and
    using hard compatability test for the closest points.

    Fitting was initialized by a simple nose detector and proceeded fully
    automatic.
  \vspace{0.3em}
  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Distance Measure}{name=measure,column=0,below=fitting}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
   The Mahalanobis angle between the identity coefficients $\VEC{\alpha_{n}}$
   was used for classification.
  \vspace{0.3em}
  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Expression Neutralization}{name=results neutralization,column=1,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{tikzpicture}[x=0.3333\linewidth,y=-0.42\linewidth]
    \path [use as bounding box] (-0.5,-0.5) rectangle(2.5,1.7);
    \path
    (0,0) node{\includegraphics[width=0.42\linewidth]{D1077}}
    (1,0) node{\includegraphics[width=0.47\linewidth]{D1077_fit_expression}}
    (2,0) node{\includegraphics[width=0.47\linewidth]{D1077_fit}}

    (0,1) node{\includegraphics[width=0.42\linewidth]{D1360}}
    (1,1) node{\includegraphics[width=0.47\linewidth]{D1360_fit_expression}}
    (2,1) node{\includegraphics[width=0.47\linewidth]{D1360_fit}}

    (0,1.6) node {\smaller a) Target}
    (1,1.6) node {\smaller b) Fit}
    (2,1.6) node {\smaller c) Normalized};
  \end{tikzpicture}
  \vspace{0.5em}

  Expression normalisation for two scans of the same individual.  
  The robust fitting gives a good estimate (b) of the true face surface given
  the noisy measurement (a). It fills in holes and removes artifacts using
  prior knowledge from the face model. The pose and expression normalized faces
  (c) are used for face recognition.
  \vspace{0.5em}
  }
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Funding}{name=funding,column=1,span=2,above=bottom}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \smaller 
  \hspace{1em}This work was supported in part by Microsoft Research through the European PhD Scholarship Programme.
  }
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Results}{name=results,column=1,span=2,below=results neutralization,above=funding}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      \begin{multicols}{2}
        The method was evaluated on the GavabDB expression dataset which
        contains 427 Scans, with 3 neutral scans and 4 expression scans per ID.
        To test the impact of expression invariance on neutral data we used the
        UND Dataset from the Face Recognition Great Vendor Test, which contains
        953 neutral scans with one to eight scans per subject.
      \end{multicols}\vspace{-1em}
      \mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\
      \begin{tabular}{cc}
        \hspace{-0.5em}\scalebox{0.735}{\input{shrec_mncg}} &
        \hspace{0.5em}\scalebox{0.735}{\input{und_mncg}}
      \end{tabular}\\
%      \begin{multicols}{2}
        {Expression neutralization improves results on the expression dataset
        without decreasing the accuracy on the neutral testset. Plotted is the
        ratio of correct answers to  the number of possible correct answers.
        %Note the different scales for the two graphs.
        %Our approach has a high accuracy on the neutral (UND) dataset.
        }
%      \end{multicols}\vspace{-1em}
      \\\mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\
      \begin{tabular}{cc}
        \hspace{-0.5em}\scalebox{0.735}{\input{shrec_pr}} &
        \hspace{0.5em}\scalebox{0.735}{\input{und_pr}}
      \end{tabular}\\
%      \begin{multicols}{2}
        {Plotted are precision and recall for different retrieval depths. The lower
        precision of the UND database is due to the fact that some queries have no
        correct answers.}
%      \end{multicols}\vspace{-1em}
      \\\mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\
      \begin{tabular}{cc}
        \hspace{-0.5em}\scalebox{0.735}{\input{shrec_far_frr}} &
        \hspace{0.5em}\scalebox{0.735}{\input{und_far_frr}}
      \end{tabular}\\
%      \begin{multicols}{2}
        {Impostor detection is reliable, as the minimum distance to a match
        is smaller than the minimum distance to a nonmatch. }
%      \end{multicols}
  \\
  \vspace{0.5em}
  }
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Robustness}{name=robustness,column=2,row=0,above=results,span=1}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{tikzpicture}[x=0.3333\linewidth,y=-0.42\linewidth]
    \path [use as bounding box] (-0.5,-0.5) rectangle(2.5,1.7);
    \path
    (0,0) node{\includegraphics[width=0.42\linewidth]{D1160}}
    (1,0) node{\includegraphics[width=0.42\linewidth]{D1425}}
    (2,0) node{\includegraphics[width=0.42\linewidth]{D1205}}

    (0,1) node{\includegraphics[width=0.28\linewidth]{D1160_fit_expression}}
    (1,1) node{\includegraphics[width=0.28\linewidth]{D1425_fit_expression}}
    (2,1) node{\includegraphics[width=0.28\linewidth]{D1205_fit_expression}}

    (1,0.5) node {\smaller a) Target}
    (1,1.6) node {\smaller b) Robust Reconstruction};
  \end{tikzpicture}
  \vspace{0.5em}

  The reconstruction (b) is robust against scans (a) with artifacts, noise, and
  holes.

  This is achieved by a robust iteratively reweighted ICP algorithm and outlier
  rejection based on angle comparisions between corresponding points.
  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{References}{name=references,column=0,above=bottom}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \smaller
    \vspace{-0.4em}
    \bibliographystyle{ieee}
    \renewcommand{\section}[2]{\vskip 0.05em}
      \begin{thebibliography}{1}\itemsep=-0.01em
      \setlength{\baselineskip}{0.4em}
      \bibitem{amberg07:nonrigid}
        B.~Amberg, S.~Romdhani, T. Vetter.
        \newblock {O}ptimal {S}tep {N}onrigid {ICP} {A}lgorithms for {S}urface {R}egistration
        \newblock In {\em Computer Vision and Pattern Recognition 2007}
      \bibitem{amberg08:recognition}
        B.~Amberg, R.~Knothe, T. Vetter.
        \newblock Expression Invariant Face Recognition with a 3D Morphable Model
        \newblock In {\em Automated Face and Gesture Recognition 2008}
      \end{thebibliography}
  }
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Open Questions}{name=questions,column=0,span=1,below=measure,above=references}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    While the expression and identity space are linearly independent, there is
    some expression left in the identity model. This is because a ``neutral''
    face is interpreted differently by the subjects.

    We investigate learning ``pure'' separated models from our the available ``impure'' data.
  }

\end{poster}%
\end{document}

### 2010-02-25 honors_thesis/journal_paper/paper.tex
A lot of work on the journal paper

@@
 communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
-be defined as having $k$ neighbors; we use 2 neighbors for all of our
+be defined as having $n$ neighbors; we use 2 neighbors for all of our
 experiments.
@@
 a particle cannot be the source of its $\nbest$ update while not updating its
-$\pbest$.  Thus we have $2(n+1)-1$ evaluations.  Table~\ref{tab:evals} lists
-all possible combinations of updates that can be seen by a particle with two
-neighbors other than itself, which is the case in all of the topologies we use.
-If all of these speculative positions are evaluated in parallel along with the
-original, non-speculative position of the particle at time $t-1$, then one of
-the speculative positions has to be the correct one, and two iterations have
-been evaluated at once.
+$\pbest$.  Thus we have $2(n+1)-1$, or $2n+1$, evaluations.
+Table~\ref{tab:evals} lists all possible combinations of updates that can be
+seen by a particle with two neighbors other than itself, which is the case in
+all of the topologies we use.  If all of these speculative positions are
+evaluated in parallel along with the original, non-speculative position of the
+particle at time $t-1$, then one of the speculative positions has to be the
+correct one, and two iterations have been evaluated at once.
 
@@
 
-Because $2(n+1)-1$ speculative evaluations must be performed for each particle,
-the choice of topology is an important one.  The use of the Complete topology,
+Because $2n+1$ speculative evaluations must be performed for each particle, the
+choice of topology is an important one.  The use of the Complete topology,
 where every particle is a neighbor to every other particle, would require
@@
 \subsection{Centralized Algorithms}
+
+In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
+master, keeps track of all necessary information, and all other machines are
+merely used to evaluate the objective function at various positions as directed
+by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
+a scheme, the master generates the positions to evaluate speculatively and
+decides which one to accept for each particle.  The outline of the procedure is
+given in \alg{centralized}.
 
@@
 
-In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
-master, keeps track of all necessary information, and all other machines are
-merely used to evaluate the objective function at various positions as directed
-by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
-a scheme, the master generates the positions to evaluate speculatively and
-decides which one to accept for each particle.  The outline of the procedure is
-given in \alg{centralized}.
-
 Given a set of particles at iteration $t-1$ (perhaps which have just been
@@
 \subsection{Distributed Algorithms}
+
+\label{sec:distributed}
 
@@
 best.  That creates a problem for speculative evaluation---there are
-potentially more than $2(n+1)-1$ possible next positions, increasing the amount
-of work that must be done to perform the second iteration at the same time as
-the first.
+potentially more than $2n+1$ possible next positions, increasing the amount of
+work that must be done to perform the second iteration at the same time as the
+first.
 
@@
 calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
-still only $2(n+1)-1$ possible next positions, and the problem is avoided.
+still only $2n+1$ possible next positions, and the problem is avoided.
 
 \subsection{Comparison with Standard PSO}
+\label{sec:results}
 
@@
   \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
-  two iterations per round, while standard PSO algorithms use 240 processors
-  and do one iteration per round.}
+  evaluations.}
   \label{fig:basic-griewank1}
@@
 
+Because our speculative algorithm got stuck with such a small swarm size, we
+ran a set of experiments on Griewank with 800 processors, giving swarms of size
+100 and 800 instead of 30 and 240.  \fig{basic-griewank3} shows the results.
+\fig{basic-griewank3} is very similar to \fig{basic-griewank2}, except that
+speculative evaluation finds the global optimum essentially 100\% of the time.
+Thus with a few more processors, our algorithm finds the optimum twice as fast
+as the original PSO without a significant loss of accuracy.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank3.eps}
+  \caption{Function Griewank with a swarm that uses 800 processors per round of
+  evaluations.}
+  \label{fig:basic-griewank3}
+\end{figure}
+
 Lastly we look at a function that does not show any kind of improvement with
@@
   \caption{Function Rastrigin with a swarm that uses 240 processors per round
-  of evaluations.  Our speculative algorithm has a swarm of 30 particles and
-  does two iterations per round, while standard PSO algorithms use 240
-  processors and do one iteration per round.}
+  of evaluations.}
   \label{fig:rastrigin}
@@
 While the idea of speculative evaluation in particle swarm optimization is
-intriguing, the results of the previous section show that it requires too many
-extra evaluations to produce competitive results.  On most functions, the
-larger swarm size that can be used in regular PSO leads to better performance
-than is found by trying to do speculative evaluation.  However, if we keep the
-idea of speculative evaluation while relaxing the requirement of exactly
-reproducing the behavior of the original PSO algorithm, we see some very
-impressive results.
-
-Should I present results for each improvement individually, or all together, or
-both?
+intriguing, the results of the previous section show that in many cases it
+requires too many extra evaluations to produce competitive results.  On most
+functions, the larger swarm size that can be used in regular PSO leads to
+better performance than is found by trying to do speculative evaluation.
+However, if we keep the idea of speculative evaluation while relaxing the
+requirement of exactly reproducing the behavior of the original PSO algorithm,
+we see some very impressive results.
 
@@
 
-In performing speculative evaluation as we have described it, $2(n+1)-1$
+In performing speculative evaluation as we have described it, $2n+1$
 speculative evaluations are done per particle, while all but one of them are
@@
 
+We show in \fig{sphere-pickbest} the results of using this modification on the
+function Sphere.  For ease of labeling, we call the modified algorithm SEPSO
+Pick Best.  As can be seen, where SEPSO did not manage to perform as well as
+PSO Complete, SEPSO Pick Best handily outperforms it with the same number of
+processors.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere2.eps}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing SEPSO Pick Best with previous results.}
+  \label{fig:sphere-pickbest}
+\end{figure}
+
+Sphere is a function for which very little exploration needs to be done, so
+always keeping the best speculative position found makes intuitive sense.  A
+less intuitive result occurs when using SEPSO Pick Best with the function
+Griewank.  Griewank is a highly deceptive function prone to premature
+convergence.  Our intuition on the behavior of picking the best speculative
+child is that it would tend to prematurely converge more often than the
+original PSO.  However, as seen in \fig{griewank-pickbest}, SEPSO Pick Best
+increases the chance of finding the optimum while also reducing the number of
+iterations required to converge.  Perhaps SEPSO Pick Best encourages local
+exploitation, while the sparse communication of the Ring topology still allows
+for global exploration.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank4.eps}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing SEPSO Pick Best with previous results.}
+  \label{fig:griewank-pickbest}
+\end{figure}
+
 \subsection{Pruning the Speculative Evaluations}
@@
 we could limit ourselves to one speculative evaluation per particle instead of
-$2(n+1)-1$.  While we cannot with certainty predict the correct branch every
-time, we can look at the statistics of the behavior of particles in general and
+$2n+1$.  While we cannot with certainty predict the correct branch every time,
+we can look at the statistics of the behavior of particles in general and
 perhaps come up with a good guess, reducing the number of necessary speculative
@@
 	4.&No update&Some neighbor\\
-	5.&Updated&Some neigbhor\\
+	5.&Updated&Some neighbor\\
   \end{tabular}
@@
 
-Row 1 in Table~\ref{tab:branches} corresponds to the phenomon commonly called
+Row 1 in Table~\ref{tab:branches} corresponds to the phenomenon commonly called
 stagnation in the literature, but only with a single particle.  As an
@@
 iterations, unless the swarm converges beyond machine precision.  When graphs
-of the best function value flatline, what really has happened is a contraction
+of the best function value flat-line, what really has happened is a contraction
 of the particles' velocity, not stagnation in the technical sense.
@@
 probabilities, so we can have hope that cutting out some of the evaluations
-that have low probability will lead to an increase in overall performance.
+that have low probability will lead to an increase in overall performance.  In
+order to implement this kind of pruning, the only change that needs to be made
+to \alg{centralized} is in Step 2, where the speculative children are
+generated.  Simply generate a subset of the children that would normally be
+generated.
 
@@
 One possibility is to leave the particle as it is, not accepting any of the
-speculative evaluations, because none of the were correct.  This leads to
+speculative evaluations, because none of them were correct.  This leads to
 particles being at different iterations, as some particles guess correctly
@@
 their branch.  This keeps the iteration number constant across all particles,
-simplifying the work needed to be done when dynamic topologies are involved.
+simplifying the work needed to be done in determining neighbors when dynamic
+topologies are involved.
 
@@
 
+There are a large number of ways to prune speculative evaluations.  We
+experimented with several, but present results only for one possible pruning.
+The pruning we present uses only the two most probable branches in the Random
+topology, as seen in~\ref{tab:stats}, the two branches where the $\nbest$ value
+was not updated.  Those branches are also convenient in that no messages are
+needed from neighbors in order to produce the positions of the speculative
+particles---in distributed frameworks using several rounds of communication
+(see \sec{distributed}), one of the rounds of communication can be dropped
+entirely.
+
+Because we are now only doing two speculative evaluations per particle, with
+240 processors we can have a swarm of size 80, instead of the 30 particles in
+previous experiments.
+
+In \fig{sphere-pruned}, we show the results of this simple pruning on the
+function Sphere.  As can be seen, the larger number of particles that pruning
+allows results in better results when picking the best speculative child.  
+When rejecting all speculative children, which we with tongue in cheek call
+Social Promotion, no real improvement is seen.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere3.eps}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing pruning with previous results.}
+  \label{fig:sphere-pruned}
+\end{figure}
+
+When looking at the function Griewank, presented in \fig{griewank-pruned}, we
+see different results.  Our intuition about picking the best child is more
+correct in this case; while adding 50 more particles to the swarm, Pick Best
+with pruning still gets stuck more often than the original Pick Best.  However,
+Social Promotion does well in this case, finding the optimum every time and
+faster than the original PSO, without the need to increase the number of
+processors.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank5.eps}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations, comparing pruning with previous results.}
+  \label{fig:griewank-pruned}
+\end{figure}
+
 \subsection{More than one iteration ahead}
@@
 view all possible speculative evaluations for a particle as an infinite tree
-with branching factor $2(n+1)-1$.  As we have already seen that doing one full
+with branching factor $2n+1$.  As we have already seen that doing one full
 level of the tree produces too many extra evaluations to be profitable, it is
@@
 combined with wisely pruning the possible evaluations based on branch
-statistics, we can use just a few extra evaluations to go two or three
+statistics, we can use just a few extra evaluations to go two or more
 iterations ahead on the most likely branches.
+
+When speculating more than one iteration ahead, the idea of picking the child
+that matches the correct branch, and holding back the particle if no child
+matches the branch, does not make very much sense, as correct branches can only
+be determined for the first iteration.  Thus when speculating more than one
+iteration ahead we always pick the child that has the highest value.
 
@@
 
+The combination of branches we tried uses seven speculative evaluations per
+particle, matching the swarm size of the original speculative algorithm.  The
+seven evaluations we used corresponded to several iterations of branches 1 and
+2 in Table~\ref{tab:branches}.  Two of the evaluations were just one iteration
+ahead, four were two iterations ahead, and one was three iterations ahead.  The
+evaluations that were one iteration ahead were branches 1 and 2; those that
+went two iterations ahead were formed by taking either branch 1 or 2 and then
+branch 1 or 2; and the evaluation going three iterations ahead followed branch
+1 on all three iterations.
+
+The results when optimizing the function Sphere, presented in
+\fig{sphere-manyiters}, are rather astounding.  When 240 processors are
+available for use, performing speculative evaluation several iterations ahead
+finds values an average of 46 orders of magnitude better than using the
+processors to perform the standard PSO algorithm with 240 particles.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere4.eps}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing speculating several iterations ahead with previous
+  results.}
+  \label{fig:sphere-manyiters}
+\end{figure}
+
+With the function Griewank, the premature convergence problems inherent in
+picking the best child are exacerbated when speculating several iterations
+ahead.  When the algorithm converges, it converges quicker than any other 
+method we tried.  However, it also got stuck and failed to find the optimum
+more than any other method.  In order for this method to be consistently useful
+on functions like Griewank, it seems that more processors are required.  The
+results are shown in \fig{griewank-manyiters}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank6.eps}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations, comparing speculating several iterations ahead with previous
+  results.}
+  \label{fig:griewank-manyiters}
+\end{figure}
+
 \subsection{Tokens}
@@
 use, where each token corresponds to a speculative evaluation.  Whichever
-method is used to which evaluations to do, they are prioritized so that
+method is used to decide which evaluations to do, they are prioritized so that
 particles with only one or two tokens perform the most valuable speculative
-evaluations.  Particles that perform well keep their tokens, particles that
+evaluations.  Particles that perform well keep their tokens, and particles that
 perform poorly send one token (if they have one) to another random particle in
@@
 to evenly spread extra evaluations due to an incapatible number of particles
-and processors.  In cursory experimentation, we found that when only going one
-iteration ahead keeping the number of evaluations per particle as even as
-possible worked much better than trying to prioritize particles and let some
-guess correctly more often because they have more tokens. (horrible explanation\dots)
-
-The other, more compelling reason to use tokens is when speculating many
-iterations ahead.  Particles that are going in good directions, as evidenced by
-relatively frequent updates to the particle's personal best, keep their tokens
-and are able to speculate up to three or four iterations.  Particles that
-aren't doing as well lose their tokens and probably contribute to the
-exploration of the algorithm, for which speculative evaluation doesn't help
-much.
+and processors.  In cursory experimentation, we found that this was not a very
+effective use of tokens.  In virtually every case we tried, it was better to
+simply change the number of particles to match the number of processors, and 
+spread the speculative evaluations evenly throughout the swarm.
+
+The other, more compelling reason to use tokens is to intelligently and
+dynamically allocate speculative evaluations to particles that are performing
+well.  This really only makes sense when speculating several iterations ahead;
+after one or two speculative evaluations, the added benefit from also
+evaluating the low-probability branches is negligible.  But by intelligently
+allocating speculative evaluations, particles that are going in good directions
+move faster by speculating farther with extra tokens, while particles that are
+not doing as well lose their tokens.  Those particles likely contribute to the
+exploration of the algorithm, and when they find good locations, they start
+keeping their tokens, switching roles to become productive particles
+themselves.  This approach is very similar in concept to other dynamic swarms,
+like the TRIBES method proposed by Clerc (cite).
 
@@
 the possibilities of using tokens to improve the allocation of speculative
-evaluations to particles.
-
-\section{Results}
-\label{sec:results}
+evaluations to particles.  We used a distributed parallel PSO framework, and in
+this setting determining which particles are performing the best globally is
+difficult without extensive communication.  Thus we only present results for
+the one method of token allocation described above.
+
+The only function for which we saw a benefit when using this technique was
+Griewank.  As seen in \fig{griewank-token}, using tokens increased the chance
+of finding the optimum to 100\%, while still using far fewer iterations than
+standard PSO.  We assume that the relatively poor performance of the technique
+on other functions is due to the lack of better token allocation.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank7.eps}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations, comparing token-based allocation with previous results.}
+  \label{fig:griewank-token}
+\end{figure}
 
@@
 \label{sec:conclusion}
+
+We have described a new technique for using processors in parallel PSO to
+improve the performance of the algorithm.  To our knowledge, this is the first
+time extra processors have been used to do anything in PSO besides increase the
+swarm size.  In an increasingly parallel world, such advancements will prove to
+be crucial in the continued effectiveness of PSO.
+
+We have detailed how to perform speculative evaluation in PSO in several
+different parallel frameworks.  Using this methodology, the behavior of the
+original PSO algorithm can either be exactly reproduced, two iterations at a
+time, or the behavior can be modified in order to improve performance.  We have
+shown results that conclusively demonstrate the superiority of our techniques
+over naively adding particles to the swarm when extra processors are available.
+
+We have shown several different ways to modify the behavior of PSO when doing
+speculative evaluation, and, as expected given the No Free Lunch Theorem,
+different modifications show varied performance with different functions.
+While our exploration of these techniques has been cursory in places, we
+believe the results shown are sufficient to demonstrate their promise.
 
@@
 
-We have speculated (pardon the pun\dots) that our ``pick the best child''
-method simply draws a number of samples from the next iteration of the sampling
-distribution of PSO and keeps one of them.  It would then be expected that the
-sampling distribution of that method is very similar to the distribution of the
-original PSO.  This remains to be proven, or even demonstrated empirically.
+We have speculated that our ``pick the best child'' method simply draws a
+number of samples from the next iteration of the sampling distribution of PSO
+and keeps one of them.  It would then be expected that the sampling
+distribution of that method is very similar to the distribution of the original
+PSO.  This remains to be proven, or even demonstrated empirically.
 
@@
 role a topology plays.  A potentially very fruitful avenue of future work would
-be to further analyze the branch statistics in PSO.  There are some particles
-that seem to drive the swarm and others that are stagnant almost the entire
-time.  Such information could be used in a TRIBES-like fashion to determine
-which particles should be removed from the swarm, perhaps starting them close
-to well-performing particles.
+be to further analyze the branch statistics in PSO across many functions and
+topologies, categorizing both the topologies and the functions and discovering
+what it is that makes certain topologies work well with certain functions.
+There are also some particles that seem to drive the swarm and others that are
+stagnant almost the entire time.  Such information could be used in a
+TRIBES-like fashion to determine which particles should be removed from the
+swarm, perhaps starting them close to well-performing particles.
 
@@
 again.
+
+We gave only a very little attention to the possibilities of token-based
+allocation of speculative evaluations to particles.  We imagine that the
+benefits of intelligently and dynamically deciding which particles get extra
+evaluations could produce very good results.  Future work could explore the use
+of a centralized allocator for extra processors or ideas in the multi-robot
+task allocation literature, such as auctions, to better assign processors to
+particles for speculative evaluations.  While such methods involve extra
+communication, with long-running function evaluations the extra overhead of
+intelligently allocating resources would likely be negligible.
 

### 2010-02-25 honors_thesis/thesis/paper.tex
Final draft of thesis submitted to be bound

@@
 \Advisor{Kevin Seppi}
-\HonorsRep{}
+\HonorsRep{Bruce Collings}
 \Abstract{
@@
 literature are the Ring topology and the Complete
-topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
-as a parallel adaptation of the Complete topology called Random that has been
-shown to simulate the behavior of Complete with far less
-communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
-be defined as having $k$ neighbors; we use 2 neighbors for all of our
-experiments.
+topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
+neighbor to either side of it; in the Complete topology, every particle is a
+neighbor to every other particle.  In all topologies, a particle is also a
+neighbor to itself, in that its own position is considered when updating the
+particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
+particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
+topology, each particle has $p$ neighbors.  In this paper we use these
+topologies, as well as a parallel adaptation of the Complete topology, called
+Random, that has been shown to simulate the behavior of Complete with far less
+communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
+randomly picks two other particles to share information with at each iteration,
+along with itself.  Thus in both the Ring and the Random topologies, all
+particles have 3 neighbors.
 
@@
 $2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
-neighbors the particle has.  If a particle is also its own neighbor, as is most
-often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
-a particle cannot be the source of its $\nbest$ update while not updating its
-$\pbest$.  Thus we have $2(n+1)-1$ evaluations.  Table~\ref{tab:evals} lists
-all possible combinations of updates that can be seen by a particle with two
+neighbors the particle has.  Because a particle is also its own neighbor, one
+of the $2(n+1)$ speculative positions can be eliminated, as a particle cannot
+be the source of its $\nbest$ update while not updating its $\pbest$.  Thus we
+have $2(n+1)-1$, or $2n+1$, evaluations.  Table~\ref{tab:evals} lists all
+possible combinations of updates that can be seen by a particle with two
 neighbors other than itself, which is the case in all of the topologies we use.
@@
 in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
-Complete topology.  In that topology, 2 neighbors are chosen at random from the
-swarm for each particle at each iteration.  Because a particle is also always
-its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report the
-results for performing speculative evaluation using both the Ring topology
-(where $n$ again is $3$) and the Random topology on a number of common
-benchmark functions.
+Complete topology.  In~\sec{results} we report the results for performing
+speculative evaluation using both the Ring topology and the Random topology on
+a number of common benchmark functions.
 
@@
 position was the correct next position of each particle.  First we discuss the
-relatively easy case of a centralized parallel PSO algorithm.  In such an
-architecture, the master keeps track of all necessary information with only
-trivial message passing needed.  Then we discuss the more complicated case of a
-distributed algorithm, where each particle is on its own and needs to send and
-receive messages to and from other particles.  Finally we discuss the further
-complications of a dynamic topology, where a particle's neighbors change from
-one iteration to another.
+relatively easy case of a centralized parallel PSO algorithm with a master
+computer and many slaves.  In such an architecture, the master keeps track of
+all necessary information with only trivial message passing needed.  Then we
+discuss the more complicated case of a distributed algorithm, where each
+particle is on its own and needs to send and receive messages to and from other
+particles.  Finally we discuss the further complications of a dynamic topology,
+where a particle's neighbors change from one iteration to another.
 
@@
 
+In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
+master, keeps track of all necessary information, and all other machines are
+merely used to evaluate the objective function at various positions as directed
+by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
+a scheme, the master generates the positions to evaluate speculatively and
+decides which one to accept for each particle.  The outline of the procedure is
+given in \alg{centralized}.
+
 \begin{algorithm}
@@
 \end{algorithm}
-
-In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
-master, keeps track of all necessary information, and all other machines are
-merely used to evaluate the objective function at various positions as directed
-by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
-a scheme, the master generates the positions to evaluate speculatively and
-decides which one to accept for each particle.  The outline of the procedure is
-given in \alg{centralized}.
 
@@
 
-Upon picking the correct branch for each particle and updating the children's
+Upon picking the correct branch for each particle and updating the child's
 personal best and neighborhood best value (from iteration $t$), the result is
@@
 The process is exactly the same to produce each $\n_{t+1}$, requiring the same
-messages, except for the neighbor particles instead of the particle itself.  
+messages, only for the neighbor particles instead of the particle itself.
 \fig{messages5} shows graphically how $\noeval{\sset}_{t+3}$ is produced.
@@
 $t+1$, and information about neighbors' neighbors' neighbors is used to
-reconstruct information about iteration $t+2$.  Also, this method of message
-passing again requires the use of random seeds if the topology is random, so
-that each processor computes the same neighbors for a particle as all other
-processors.
+reconstruct information about iteration $t+2$.  This method of message passing
+again requires the use of random seeds if the topology is random, so that each
+processor computes the same neighbors for a particle as all other processors.
 
-Also, an issue common to all methods of performing speculative evaluation is
-that of getting available information from neighbors before generating
-speculative children.  In a static topology, at iteration $t$ a particle
-already has all of the information about the positions of its neighbors during
-iterations $1$ through $t-1$.  If the neighbor finds a better position at 
-iteration $t$, the particle updates its neighborhood best, but if it does not,
-it still has its old neighborhood best from its neighbors for all previous 
-iterations.
+An issue common to all methods of performing speculative evaluation is that of
+getting available information from neighbors before generating speculative
+children.  In a static topology, at iteration $t$ a particle already has all of
+the information about the positions of its neighbors during iterations $1$
+through $t-1$.  If the neighbor finds a better position at iteration $t$, the
+particle updates its neighborhood best, but if it does not, it still has its
+old neighborhood best from its neighbors for all previous iterations.
 
@@
 best.  That creates a problem for speculative evaluation---there are
-potentially more than $2(n+1)-1$ possible next positions, increasing the amount
+potentially more than $2n+1$ possible next positions, increasing the amount
 of work that must be done to perform the second iteration at the same time as
@@
 calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
-still only $2(n+1)-1$ possible next positions, and the problem is avoided.
+still only $2n+1$ possible next positions, and the problem is avoided.
 
@@
 the objective function takes on the order of minutes or more to evaluate; in
-such cases our assumption is perfectly reasonable.
+such cases our choice of framework is perfectly reasonable.
 
@@
 optimizing the Griewank function, as Complete is prone to premature convergence
-on a local optima.  We show results in \fig{griewank1} for swarms of size 30
-and 240 using the Ring topology. One can see in the figure that both SEPSO
-algorithms initially outperformed PSO, but got stuck in a local optima and were
-eventually outperformed themselves.
+on a local optimum.  Griewank has a global optimum with a value of 0, and
+sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
+showing average function value at each iteration, a more accurate plot for
+Griewank shows the percent of runs that have found the global optimum by each
+iteration.  We show results in \fig{griewank1} for swarms of size 30 and 240
+using the Ring topology.  One can see in the figure that both SEPSO algorithms
+find the optimum faster than the original PSO.  However, because the swarm size
+is so small, SEPSO gets stuck a little less than half of the time.
+Surprisingly, SEPSO Pick Best improves the chance of success with Griewank,
+while also finding the optimum faster.
 
@@
   \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
-  iterations per round, while standard PSO algorithms use 240 processors and do
-  one iteration per round.}
+  evaluations, showing percent successful by each iteration instead of average
+  function value.}
   \label{fig:griewank1}
@@
 
-\fig{griewank1} is slightly deceptive, however, as with Griewank some of the
-runs found the global optimum, 0, while some did not.  A more accurate plot for
-this function is one showing what percentage of the runs have found the global
-optimum at each iteration.  \fig{griewank2} shows such a plot.  A swarm of size
-30 performing the PSO algorithm gets stuck a little less than half of the time,
-while 240 particles provides enough exploration to find the global optimum
-every time.  Surprisingly, using the SEPSO Pick Best algorithm with 30
-particles greatly improves the chance of success with Griewank, while also
-finding the global optimum in far fewer rounds of evaluations than standard
-PSO.
+Because our speculative algorithm got stuck with such a small swarm size, we
+ran a set of experiments on Griewank with 800 processors, giving swarms of size
+100 and 800 instead of 30 and 240.  \fig{griewank2} shows the results.
+\fig{griewank2} is very similar to \fig{griewank1}, except the speculative
+algorithms find the global optimum essentially 100\% of the time.  Thus with a
+few more processors, our algorithms find the optimum two to three times faster
+than the original PSO without a significant loss of accuracy.
 
@@
   \includegraphics[width=.8\columnwidth]{griewank2.eps}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.  Instead of showing average function value, we show the
-  percentage of runs that have found the global optimum by each iteration.}
+  \caption{Function Griewank with a swarm that uses 800 processors per round of
+  evaluations.}
   \label{fig:griewank2}
@@
   \caption{Function Rastrigin with a swarm that uses 240 processors per round
-  of evaluations.  Speculative algorithms have a swarm of 30 particles and do
-  two iterations per round, while standard PSO algorithms use 240 processors
-  and do one iteration per round.}
+  of evaluations.}
   \label{fig:rastrigin}
@@
 than regular PSO, but for some functions still performs better when run in a
-parallel environment.
-
-We have also described a method for making use of extra speculative evaluations
-that performs incredibly well on some functions.
+parallel environment.  We have also described a method for making use of extra
+speculative evaluations that performs incredibly well on some functions.
 
@@
 There are some functions for which very little exploration needs to be done;
-Sphere is an example of such a function, as would be a function where the
-general area of the optimal point is known, but the exact point needs to be
-found.  For such functions the best use of processors is to have a small swarm
-performing speculative evaluation with our second method, where all speculative
-evaluations are used.
+Sphere is an example of such a function.  For such functions the best use of
+processors is to have a small swarm performing speculative evaluation with our
+second method, where all speculative evaluations are used.
 
@@
 additional particles to the swarm is akin to increasing its capacity for
-exploration.  We have seen that for a 20-dimensional Griewank function, a
-number of particles greater than 30 but less than 240 has ``enough''
-exploration and finds the optimal value every time.  For these functions, one
+exploration.  We have seen that for a 20-dimensional Griewank function, 30
+particles is not ``enough'' exploration, but 100 is.  For these functions, one
 should use processors to increase the swarm size until enough exploration has
@@
 done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
-cannot be easily prescribed and depends largely on the desired outcome.  It is
-likely that the best thing to do with those functions is to use all processors
-to add particles to the swarm and do no speculative evaluation.
+cannot be easily prescribed and depends largely on the desired outcome and how
+many processors are available.  It is likely that the best thing to do with
+those functions is to use all processors to add particles to the swarm and do
+no speculative evaluation.
 

### 2010-02-25 byu_studies/poster_package/poster_landscape.tex
Some work on changing their poster document into mine

--- 2010-02-25 byu_studies/poster_package/poster.tex
@@
-\documentclass[portrait,final]{baposter}
-%\documentclass[a4shrink,portrait,final]{baposter}
-% Usa a4shrink for an a4 sized paper.
-
-\tracingstats=2
+\documentclass[landscape,final]{baposter}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Define some colors
+\typeout{Poster Starts}
+\background{
+  \begin{tikzpicture}[remember picture,overlay]%
+    \draw (current page.north west)+(-2em,-0em) node[anchor=north west] {\hspace{-2em}\includegraphics[height=1.1\textheight]{silhouettes_background}};
+  \end{tikzpicture}%
+}
 \definecolor{silver}{cmyk}{0,0,0,0.3}
 \definecolor{yellow}{cmyk}{0,0,0.9,0.0}
+\definecolor{blue}{cmyk}{0.9,0,0,0.0}
 \definecolor{reddishyellow}{cmyk}{0,0.22,1.0,0.0}
+\definecolor{reddishblue}{cmyk}{1,0.22,0,0.0}
 \definecolor{black}{cmyk}{0,0,0.0,1.0}
@@
 \definecolor{lighteryellow}{cmyk}{0,0,0.1,0.0}
-\definecolor{lighteryellow}{cmyk}{0,0,0.1,0.0}
 \definecolor{lightestyellow}{cmyk}{0,0,0.05,0.0}
-
-%%
-\typeout{Poster Starts}
-\background{
-  \begin{tikzpicture}[remember picture,overlay]%
-    \draw (current page.north west)+(-2em,2em) node[anchor=north west] {\includegraphics[height=1.1\textheight]{silhouettes_background}};
-  \end{tikzpicture}%
-}
-
-\newlength{\leftimgwidth}
-\begin{poster}%
-  % Poster Options
-  {
+\definecolor{lightblue}{cmyk}{.3,0,0,0.0}
+\definecolor{lighterblue}{cmyk}{.1,0,0,0.0}
+\definecolor{lightestblue}{cmyk}{.05,0,0,0.0}
+\begin{poster}{
   % Show grid to help with alignment
@@
   % Color style
-  bgColorOne=lighteryellow,
-  bgColorTwo=lightestyellow,
-  borderColor=reddishyellow,
-  headerColorOne=yellow,
-  headerColorTwo=reddishyellow,
+  bgColorOne=lighterblue,
+  bgColorTwo=lightestblue,
+  borderColor=reddishblue,
+  headerColorOne=blue,
+  headerColorTwo=reddishblue,
   headerFontColor=black,
-  boxColorOne=lightyellow,
-  boxColorTwo=lighteryellow,
+  boxColorOne=lightblue,
+  boxColorTwo=lighterblue,
   % Format of textbox
@@
   % Eye Catcher
-  {\includegraphics[width=10em]{D1077}} % No eye catcher for this poster. (eyecatcher=no above). If an eye catcher is present, the title is centered between eye-catcher and logo.
+  {} % No eye catcher for this poster. If an eye catcher is present, the title is centered between eye-catcher and logo.
   % Title
@@
   %\bf% Serif
-  Expression Invariant Face Recognition with a 3DMM}
+  Automatic Topic Discovery in 100 Years of General Conference Talks}
   % Authors
@@
   % Serif
-  \vspace{1em}\{Brian.Amberg, Reinhard.Knothe, Thomas.Vetter\}@unibas.ch
+  Matthew Gardner
+  (mjg82@byu.edu)
+  and
+  Eric Ringger
+  (ringger@cs.byu.edu)
   }
   % University logo
-  {% The makebox allows the title to flow into the logo, this is a hack because of the L shaped logo.
-    \makebox[8em][r]{%
-      \begin{minipage}{16em}
+  {{\begin{minipage}{16em}
     \hfill
-        \includegraphics[height=2em]{msrlogo}
-        \includegraphics[height=7.0em]{logo}
-      \end{minipage}
-    }
+    \includegraphics[height=5.5em]{logo}
+  \end{minipage}}
   }
@@
   % Width of left inset image
+     \newlength{\leftimgwidth}
      \setlength{\leftimgwidth}{0.78em+8.0em}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-   {}We introduce a method for expression invariant face recognition. A
-   generative 3D Morphable Model (3DMM) is used to separate identity and
-   expression components. The expression removal results in increased
-   recognition performance, even on difficult datasets, without a decrease in
-   performance on expression-less datasets.
-
-   It is applicable to any kind of input data, and was evaluated here on
-   textureless range scans.
-  \vspace{0.3em}
+   {}We use state of the art natural language processing techniques to gain new
+   insights from 100 years of General Conference Talks.  Using computers to
+   sift through a corpus that would be overwhelmingly large to process
+   manually, we can discover relationships between talks and trends over time
+   that help us to better appreciate the trove of knowledge in these inspired
+   discourses.
  }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Model}{name=model,column=0,below=contribution}{
+  \headerbox{Methodology}{name=methodology,column=0,below=contribution}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    The Model was learnt from 275 subjects. We used one neutral expression scan
-    per identity and 150 expression scans of a subset of the subjects.
+	Document clustering is the task of determining which documents are similar
+	to each other in any given corpus.  The fundamental assumption made by
+	clustering techniques is that there are several topics used in the corpus,
+	and each document is generated by a single (unknown) topic.  Given a
+	clustering model and a set of documents, statistical methods can be used to
+	determine the best clustering of documents for that corpus.
 
-    The identity model is a linear model build from the neutral scans.
-    \begin{align}
-      \VEC f&=\VEC\mu + \MAT M_n\VEC\alpha_n\qquad.
-    \end{align}
-    For each of the 150 expression scans, we calculated an expression vector as
-    the difference between the expression scan and the corresponding neutral
-    scan of that subject.  This data is already mode-centered, if we regard the
-    neutral expression as the natural mode of expression data. From these offset
-    vectors an additional expression matrix $\MAT M_e$ was calculated, such that the complete linear Model is
-    \begin{align}
-      \VEC f&=\VEC\mu + \MAT M_n\VEC\alpha_n + \MAT M_e\VEC\alpha_e
-    \end{align}
-    The assumption here is, that the face and expression space are linearly
-    independent, such that each face is represented by a unique set of
-    coefficients.  
-  \vspace{0.3em}
-  }
+	When documents frequently cover multiple topics, however, the performance
+	of document clustering methods degenerates.  This is often the case with
+	religious discourse, as talks frequently combine faith and repentance, or
+	faith and tithing, making the talks seem similar in regards to their use of
+	words describing faith, but different in the other topics they cover.  To
+	solve this problem, Latent Dirichlet Allocation (LDA) was proposed.  LDA
+	breaks the assumption of document clustering that each document is
+	generated from a single topic, and instead assumes that each word in each
+	document comes from a particular topic.  Thus documents can have multiple
+	topics, creating a better model for many corpora of interest today.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Fitting}{name=fitting,column=0,below=model}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    A Robust Nonrigid ICP method was used to fit the model to the data.
-    Robustness was achieved by iteratively reweighting the correspondences and
-    using hard compatability test for the closest points.
+	LDA can be used both to discover the topics that are spoken of throughout a
+	corpus and to determine which documents in the corpus use which topics most
+	frequently.  When documents have dates and authors associated with them,
+	other interesting attributes of the corpus, such as the relative
+	proportions of topics over time, can also be deduced.  We used LDA to
+	automatically discover topics found in General Conference talks from
+	1900--2005.  We obtained the talks from the internet site
+	www.ldslibrary.org.  There is some noise in the data because of parsing
+	error, but overall the quality is high.
 
-    Fitting was initialized by a simple nose detector and proceeded fully
-    automatic.
-  \vspace{0.3em}
-  }
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Distance Measure}{name=measure,column=0,below=fitting}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-   The Mahalanobis angle between the identity coefficients $\VEC{\alpha_{n}}$
-   was used for classification.
-  \vspace{0.3em}
   }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \begin{tikzpicture}[x=0.3333\linewidth,y=-0.42\linewidth]
-    \path [use as bounding box] (-0.5,-0.5) rectangle(2.5,1.7);
-    \path
-    (0,0) node{\includegraphics[width=0.42\linewidth]{D1077}}
-    (1,0) node{\includegraphics[width=0.47\linewidth]{D1077_fit_expression}}
-    (2,0) node{\includegraphics[width=0.47\linewidth]{D1077_fit}}
-
-    (0,1) node{\includegraphics[width=0.42\linewidth]{D1360}}
-    (1,1) node{\includegraphics[width=0.47\linewidth]{D1360_fit_expression}}
-    (2,1) node{\includegraphics[width=0.47\linewidth]{D1360_fit}}
-
-    (0,1.6) node {\smaller a) Target}
-    (1,1.6) node {\smaller b) Fit}
-    (2,1.6) node {\smaller c) Normalized};
-  \end{tikzpicture}
-  \vspace{0.5em}
-
+  \begin{tabular}{@{}c@{ }c@{ }c@{ }c@{}@{ }@{ }c@{ }c@{ }c@{ }c@{ }}
+    \includegraphics[height=0.42\linewidth]{16_1_tgt}&
+    \includegraphics[height=0.42\linewidth]{16_1_expression}&
+    \includegraphics[height=0.42\linewidth]{16_1_neutral}\\[-0.8em]
+    \smaller a) Target & \smaller b) Fit & \smaller c) Normalized\\[0.8em]
+    \includegraphics[height=0.42\linewidth]{16_6_tgt}&
+    \includegraphics[height=0.42\linewidth]{16_6_expression}&
+    \includegraphics[height=0.42\linewidth]{16_6_neutral}\\[-0.8em]
+    \smaller a) Target & \smaller b) Fit & \smaller c) Normalized
+  \end{tabular}
   Expression normalisation for two scans of the same individual.  
@@
   (c) are used for face recognition.
-  \vspace{0.5em}
   }
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Funding}{name=funding,column=1,span=2,above=bottom}{
+  \headerbox{Robustness}{name=robustness,column=1,below=results neutralization,span=1,above=bottom}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \smaller 
-  \hspace{1em}This work was supported in part by Microsoft Research through the European PhD Scholarship Programme.
+  \begin{tabular}{@{}c@{ }c@{ }c@{ }c@{}}
+    \includegraphics[height=0.42\linewidth]{56_4_tgt}&
+    \includegraphics[height=0.42\linewidth]{23_2_tgt}&
+    \includegraphics[height=0.42\linewidth]{5_6_tgt}\\[-0.8em]
+                       & \smaller a) Targets & \\[0.8em]
+    \includegraphics[height=0.42\linewidth]{56_4_expression}&
+    \includegraphics[height=0.42\linewidth]{23_2_expression}& 
+    \includegraphics[height=0.42\linewidth]{5_6_expression}\\[-0.8em]
+                    & \smaller b) Fits & 
+  \end{tabular}
+  The reconstruction (b) is robust against scans (a) with artifacts, noise, and
+  holes.
   }
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Results}{name=results,column=1,span=2,below=results neutralization,above=funding}{
+  \headerbox{Results}{name=results,column=2,span=2,row=0}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
       \begin{tabular}{cc}
-        \hspace{-0.5em}\scalebox{0.735}{\input{shrec_mncg}} &
-        \hspace{0.5em}\scalebox{0.735}{\input{und_mncg}}
+        \hspace{0.5em}\scalebox{0.74}{\input{shrec_MNCG}} &
+        \hspace{0.5em}\scalebox{0.74}{\input{und_MNCG}}
       \end{tabular}\\
@@
       \begin{tabular}{cc}
-        \hspace{-0.5em}\scalebox{0.735}{\input{shrec_pr}} &
-        \hspace{0.5em}\scalebox{0.735}{\input{und_pr}}
+        \hspace{0.5em}\scalebox{0.74}{\input{shrec_PR}} &
+        \hspace{0.5em}\scalebox{0.74}{\input{und_PR}}
       \end{tabular}\\
@@
       \begin{tabular}{cc}
-        \hspace{-0.5em}\scalebox{0.735}{\input{shrec_far_frr}} &
-        \hspace{0.5em}\scalebox{0.735}{\input{und_far_frr}}
+        \hspace{0.5em}\scalebox{0.74}{\input{shrec_FARFRR}} &
+        \hspace{0.5em}\scalebox{0.74}{\input{und_FARFRR}}
       \end{tabular}\\
@@
 \\
-  \vspace{0.5em}
-  }
+  }%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Robustness}{name=robustness,column=2,row=0,above=results,span=1}{
+  \headerbox{Open Questions}{name=questions,column=2,span=1,above=bottom,below=results}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \begin{tikzpicture}[x=0.3333\linewidth,y=-0.42\linewidth]
-    \path [use as bounding box] (-0.5,-0.5) rectangle(2.5,1.7);
-    \path
-    (0,0) node{\includegraphics[width=0.42\linewidth]{D1160}}
-    (1,0) node{\includegraphics[width=0.42\linewidth]{D1425}}
-    (2,0) node{\includegraphics[width=0.42\linewidth]{D1205}}
-
-    (0,1) node{\includegraphics[width=0.28\linewidth]{D1160_fit_expression}}
-    (1,1) node{\includegraphics[width=0.28\linewidth]{D1425_fit_expression}}
-    (2,1) node{\includegraphics[width=0.28\linewidth]{D1205_fit_expression}}
-
-    (1,0.5) node {\smaller a) Target}
-    (1,1.6) node {\smaller b) Robust Reconstruction};
-  \end{tikzpicture}
-  \vspace{0.5em}
-
-  The reconstruction (b) is robust against scans (a) with artifacts, noise, and
-  holes.
-
-  This is achieved by a robust iteratively reweighted ICP algorithm and outlier
-  rejection based on angle comparisions between corresponding points.
-  }
-
+    While the expression and identity space are linearly independent, there is
+    some expression left in the identity model. This is because a ``neutral''
+    face is interpreted differently by the subjects. We investigate the
+    possibilty to build an identity/expression separated model without using
+    the data labelling, based on a measure of independence.
+  }%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{References}{name=references,column=0,above=bottom}{
+  \headerbox{Funding}{name=funding,column=3,span=1,above=bottom}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \smaller 
+  This work was supported in part by Microsoft Research through the European PhD Scholarship Programme.
+  }%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \headerbox{References}{name=references,column=3,above=funding,below=results}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
         \newblock {O}ptimal {S}tep {N}onrigid {ICP} {A}lgorithms for {S}urface {R}egistration
-        \newblock In {\em Computer Vision and Pattern Recognition 2007}
+        \newblock In {\em CVPR 2007}
       \bibitem{amberg08:recognition}
@@
         \newblock Expression Invariant Face Recognition with a 3D Morphable Model
-        \newblock In {\em Automated Face and Gesture Recognition 2008}
+        \newblock In {\em AFGR 2008}
       \end{thebibliography}
-  }
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Open Questions}{name=questions,column=0,span=1,below=measure,above=references}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    While the expression and identity space are linearly independent, there is
-    some expression left in the identity model. This is because a ``neutral''
-    face is interpreted differently by the subjects.
-
-    We investigate learning ``pure'' separated models from our the available ``impure'' data.
-  }
-
+  }%
 \end{poster}%
+%
 \end{document}

### 2010-02-26 byu_studies/poster.tex
First pass of the poster done, still a ways to go

--- 2010-02-25 byu_studies/poster_package/poster_landscape.tex
@@
 \typeout{Poster Starts}
-\background{
-  \begin{tikzpicture}[remember picture,overlay]%
-    \draw (current page.north west)+(-2em,-0em) node[anchor=north west] {\hspace{-2em}\includegraphics[height=1.1\textheight]{silhouettes_background}};
-  \end{tikzpicture}%
-}
-\definecolor{silver}{cmyk}{0,0,0,0.3}
-\definecolor{yellow}{cmyk}{0,0,0.9,0.0}
 \definecolor{blue}{cmyk}{0.9,0,0,0.0}
-\definecolor{reddishyellow}{cmyk}{0,0.22,1.0,0.0}
 \definecolor{reddishblue}{cmyk}{1,0.22,0,0.0}
 \definecolor{black}{cmyk}{0,0,0.0,1.0}
-\definecolor{darkYellow}{cmyk}{0,0,1.0,0.5}
-\definecolor{darkSilver}{cmyk}{0,0,0,0.1}
 
-\definecolor{lightyellow}{cmyk}{0,0,0.3,0.0}
-\definecolor{lighteryellow}{cmyk}{0,0,0.1,0.0}
-\definecolor{lightestyellow}{cmyk}{0,0,0.05,0.0}
 \definecolor{lightblue}{cmyk}{.3,0,0,0.0}
-\definecolor{lighterblue}{cmyk}{.1,0,0,0.0}
-\definecolor{lightestblue}{cmyk}{.05,0,0,0.0}
+\definecolor{lighterblue}{cmyk}{.2,0,0,0.0}
+\definecolor{lightestblue}{cmyk}{.1,0,0,0.0}
 \begin{poster}{
@@
   boxshade=plain,
-%  background=shade-tb,
-  background=plain,
+  background=shade-tb,
+  %background=plain,
   linewidth=2pt
@@
     \hfill
-    \includegraphics[height=5.5em]{logo}
+    %\includegraphics[height=5.5em]{logo}
   \end{minipage}}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-   {}We use state of the art natural language processing techniques to gain new
+   {}
+
+   We use state of the art natural language processing techniques to gain new
    insights from 100 years of General Conference Talks.  Using computers to
@@
    discourses.
+
  }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Methodology}{name=methodology,column=0,below=contribution}{
+  \headerbox{Background and Methodology}
+  {name=methodology,column=0,below=contribution}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Expression Neutralization}{name=results neutralization,column=1,row=0}{
+  \headerbox{Results}
+  {name=results word distributions,column=1,span=2,row=0}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \begin{tabular}{@{}c@{ }c@{ }c@{ }c@{}@{ }@{ }c@{ }c@{ }c@{ }c@{ }}
-    \includegraphics[height=0.42\linewidth]{16_1_tgt}&
-    \includegraphics[height=0.42\linewidth]{16_1_expression}&
-    \includegraphics[height=0.42\linewidth]{16_1_neutral}\\[-0.8em]
-    \smaller a) Target & \smaller b) Fit & \smaller c) Normalized\\[0.8em]
-    \includegraphics[height=0.42\linewidth]{16_6_tgt}&
-    \includegraphics[height=0.42\linewidth]{16_6_expression}&
-    \includegraphics[height=0.42\linewidth]{16_6_neutral}\\[-0.8em]
-    \smaller a) Target & \smaller b) Fit & \smaller c) Normalized
+
+	The main result of using LDA is a list of which words are most likely to be
+	used in each topic.  In the algorithm, the topics just have numbers; the
+	computer merely looks for statistical relationship between words.  The
+	labels are thus our own.  We present here a few of the topics we found.
+
+	\small
+	\begin{center}
+	  \begin{tabular}{ccccc}
+		Topic 19: Tithing&Topic 33: Welfare&Topic 52: Women&Topic 37: Families
+		&Topic 7: Government\\
+		\hline
+		tithing&welfare&women&children&government \\
+		pay&poor&society&home&united \\
+		money&church&relief&family&liberty \\
+		debt&fast&sisters&parents&nation \\
+		means&help&hinckley&teach&states \\
+		out&services&each&child&country \\
+		people&program&b&families&freedom \\
+		one&plan&gordon&homes&land \\
+		paid&care&lives&love&free \\
+		poor&needs&service&mother&men \\
+		tithes&food&sister&taught&constitution \\
+		get&need&forward&fathers&people \\
+		more&family&strength&responsibility&america \\
 	  \end{tabular}
-  Expression normalisation for two scans of the same individual.  
-  The robust fitting gives a good estimate (b) of the true face surface given
-  the noisy measurement (a). It fills in holes and removes artifacts using
-  prior knowledge from the face model. The pose and expression normalized faces
-  (c) are used for face recognition.
-  }
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Robustness}{name=robustness,column=1,below=results neutralization,span=1,above=bottom}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \begin{tabular}{@{}c@{ }c@{ }c@{ }c@{}}
-    \includegraphics[height=0.42\linewidth]{56_4_tgt}&
-    \includegraphics[height=0.42\linewidth]{23_2_tgt}&
-    \includegraphics[height=0.42\linewidth]{5_6_tgt}\\[-0.8em]
-                       & \smaller a) Targets & \\[0.8em]
-    \includegraphics[height=0.42\linewidth]{56_4_expression}&
-    \includegraphics[height=0.42\linewidth]{23_2_expression}& 
-    \includegraphics[height=0.42\linewidth]{5_6_expression}\\[-0.8em]
-                    & \smaller b) Fits & 
-  \end{tabular}
-  The reconstruction (b) is robust against scans (a) with artifacts, noise, and
-  holes.
-  }
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Results}{name=results,column=2,span=2,row=0}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-      \begin{multicols}{2}
-        The method was evaluated on the GavabDB expression dataset which
-        contains 427 Scans, with 3 neutral scans and 4 expression scans per ID.
-        To test the impact of expression invariance on neutral data we used the
-        UND Dataset from the Face Recognition Great Vendor Test, which contains
-        953 neutral scans with one to eight scans per subject.
-      \end{multicols}\vspace{-1em}
+	\end{center}
+
+	\normalsize
 	\mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\
-      \begin{tabular}{cc}
-        \hspace{0.5em}\scalebox{0.74}{\input{shrec_MNCG}} &
-        \hspace{0.5em}\scalebox{0.74}{\input{und_MNCG}}
-      \end{tabular}\\
-%      \begin{multicols}{2}
-        {Expression neutralization improves results on the expression dataset
-        without decreasing the accuracy on the neutral testset. Plotted is the
-        ratio of correct answers to  the number of possible correct answers.
-        %Note the different scales for the two graphs.
-        %Our approach has a high accuracy on the neutral (UND) dataset.
-        }
-%      \end{multicols}\vspace{-1em}
-      \\\mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\
-      \begin{tabular}{cc}
-        \hspace{0.5em}\scalebox{0.74}{\input{shrec_PR}} &
-        \hspace{0.5em}\scalebox{0.74}{\input{und_PR}}
-      \end{tabular}\\
-%      \begin{multicols}{2}
-        {Plotted are precision and recall for different retrieval depths. The lower
-        precision of the UND database is due to the fact that some queries have no
-        correct answers.}
-%      \end{multicols}\vspace{-1em}
-      \\\mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\
-      \begin{tabular}{cc}
-        \hspace{0.5em}\scalebox{0.74}{\input{shrec_FARFRR}} &
-        \hspace{0.5em}\scalebox{0.74}{\input{und_FARFRR}}
-      \end{tabular}\\
-%      \begin{multicols}{2}
-        {Impostor detection is reliable, as the minimum distance to a match
-        is smaller than the minimum distance to a nonmatch. }
-%      \end{multicols}
-\\
+	Because each talk has a date associated with it, we can use LDA to analyze
+	the distribution of these topics over time, seeing at when certain topics
+	were most discussed.  In doing this, we see clear correlations with
+	real-world occurrences.
+
+	\begin{center}
+	  \includegraphics[height=.34\linewidth]{topic19.eps}
+	  \includegraphics[height=.34\linewidth]{topic33.eps}
+
+	  \includegraphics[height=.34\linewidth]{topic52.eps}
+	  \includegraphics[height=.34\linewidth]{topic56.eps}
+	\end{center}
+
   }%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Open Questions}{name=questions,column=2,span=1,above=bottom,below=results}{
+  \headerbox{Future Work}{name=future work,column=3,span=1}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    While the expression and identity space are linearly independent, there is
-    some expression left in the identity model. This is because a ``neutral''
-    face is interpreted differently by the subjects. We investigate the
-    possibilty to build an identity/expression separated model without using
-    the data labelling, based on a measure of independence.
+
+	We have used natural language processing techniques on a novel data
+	source---LDS General Conference talks.  However, our exploration of the
+	results has been quite preliminary.  Our research lays the groundwork for a
+	more thorough analysis of religious discourse in the LDS Church.  These
+	techniques allow us to quickly answer questions such as, ``Which topics are
+	constant over time, and which come and go?'' and ``Which years had the
+	highest diversity of topics, and which were more narrow-focused?''
+
+	We also have only considered General Conference talks.  A fruitful avenue
+	of future work would be to compare the results of our techniques on many
+	different corpora of LDS religious and scholarly texts, such as the Journal
+	of Discourses, the Book of Mormon, the Doctrine and Covenants, the
+	collected works of Hugh Nibley, or the past 50 years of BYU Studies.  Such
+	analysis could broaden our understanding of and appreciation for our
+	textual history.
+
   }%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Funding}{name=funding,column=3,span=1,above=bottom}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \smaller 
-  This work was supported in part by Microsoft Research through the European PhD Scholarship Programme.
-  }%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{References}{name=references,column=3,above=funding,below=results}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    \smaller
-    \vspace{-0.4em}
-    \bibliographystyle{ieee}
-    \renewcommand{\section}[2]{\vskip 0.05em}
-      \begin{thebibliography}{1}\itemsep=-0.01em
-      \setlength{\baselineskip}{0.4em}
-      \bibitem{amberg07:nonrigid}
-        B.~Amberg, S.~Romdhani, T. Vetter.
-        \newblock {O}ptimal {S}tep {N}onrigid {ICP} {A}lgorithms for {S}urface {R}egistration
-        \newblock In {\em CVPR 2007}
-      \bibitem{amberg08:recognition}
-        B.~Amberg, R.~Knothe, T. Vetter.
-        \newblock Expression Invariant Face Recognition with a 3D Morphable Model
-        \newblock In {\em AFGR 2008}
-      \end{thebibliography}
-  }%
+
 \end{poster}%

### 2010-02-27 byu_studies/poster.tex
Some more work on the poster.  Getting closer

@@
 \usepackage{bm}
+\usepackage{psfrag}
 
@@
     \hfill
-    %\includegraphics[height=5.5em]{logo}
+    \includegraphics[height=5.5em]{byunlp.eps}
   \end{minipage}}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Background and Methodology}
-  {name=methodology,column=0,below=contribution}{
+  \headerbox{Topic Discovery with LDA}
+  {name=lda,column=0,below=contribution}{
+
+	  We automatically discover topics using Latent Dirichlet Allocation (LDA).
+	  The LDA algorithm works as follows:
+	  \begin{itemize}
+		\item There is some number of topics in the corpus (we picked 100).
+		\item Assume each word in each document comes from one topic.
+		\item Use statistical relationships among words in the corpus to group 
+		  the words into topics.
+	  \end{itemize}
+
+  }
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	Document clustering is the task of determining which documents are similar
-	to each other in any given corpus.  The fundamental assumption made by
-	clustering techniques is that there are several topics used in the corpus,
-	and each document is generated by a single (unknown) topic.  Given a
-	clustering model and a set of documents, statistical methods can be used to
-	determine the best clustering of documents for that corpus.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \headerbox{LDA Graphical Model}
+  {name=methodology,column=0,below=lda}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-	When documents frequently cover multiple topics, however, the performance
-	of document clustering methods degenerates.  This is often the case with
-	religious discourse, as talks frequently combine faith and repentance, or
-	faith and tithing, making the talks seem similar in regards to their use of
-	words describing faith, but different in the other topics they cover.  To
-	solve this problem, Latent Dirichlet Allocation (LDA) was proposed.  LDA
-	breaks the assumption of document clustering that each document is
-	generated from a single topic, and instead assumes that each word in each
-	document comes from a particular topic.  Thus documents can have multiple
-	topics, creating a better model for many corpora of interest today.
-
-	LDA can be used both to discover the topics that are spoken of throughout a
-	corpus and to determine which documents in the corpus use which topics most
-	frequently.  When documents have dates and authors associated with them,
-	other interesting attributes of the corpus, such as the relative
-	proportions of topics over time, can also be deduced.  We used LDA to
-	automatically discover topics found in General Conference talks from
-	1900--2005.  We obtained the talks from the internet site
-	www.ldslibrary.org.  There is some noise in the data because of parsing
-	error, but overall the quality is high.
+	\begin{centering}
+	  \psfrag{b}[C][C]{\huge$\beta$}
+	  \psfrag{a}[C][C]{\huge$\alpha$}
+	  \psfrag{t}[C][C]{\huge$\theta$}
+	  \psfrag{f}[C][C]{\huge$\phi$}
+	  \psfrag{w}[C][C]{\huge$w$}
+	  \psfrag{z}[C][C]{\huge$z$}
+	  \psfrag{N}[C][C]{\Large$N$}
+	  \psfrag{M}[C][C]{\Large$M$}
+	  \psfrag{K}[C][C]{\Large$K$}
+	  \includegraphics[height=.82\linewidth]{lda_model.eps}
+	  \begin{align*} 
+		w_i|z_i,\phi^{(z_i)} &\sim \mathrm{Categorical}(\phi^{(z_i)}) \\
+		\phi &\sim \mathrm{Dirichlet}(\beta) \\
+		z_i|\theta^{(d_i)} &\sim \mathrm{Categorical}(\theta^{(d_i)}) \\ 
+		\theta &\sim \mathrm{Dirichlet}(\alpha) 
+	  \end{align*}
+	\end{centering}
 

### 2010-03-04 byu_studies/poster.tex
Finished the poster for BYU Studies

@@
 \graphicspath{{images/}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%% Some math symbols used in the text
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Format 
-\newcommand{\Matrix}[1]{\begin{bmatrix} #1 \end{bmatrix}}
-\newcommand{\Vector}[1]{\Matrix{#1}}
-\newcommand*{\SET}[1]  {\ensuremath{\mathcal{#1}}}
-\newcommand*{\MAT}[1]  {\ensuremath{\mathbf{#1}}}
-\newcommand*{\VEC}[1]  {\ensuremath{\bm{#1}}}
-\newcommand*{\CONST}[1]{\ensuremath{\mathit{#1}}}
-\newcommand*{\norm}[1]{\mathopen\| #1 \mathclose\|}% use instead of $\|x\|$
-\newcommand*{\abs}[1]{\mathopen| #1 \mathclose|}% use instead of $\|x\|$
-\newcommand*{\absLR}[1]{\left| #1 \right|}% use instead of $\|x\|$
-
-\def\norm#1{\mathopen\| #1 \mathclose\|}% use instead of $\|x\|$
-\newcommand{\normLR}[1]{\left\| #1 \right\|}% use instead of $\|x\|$
 
@@
   % Format of text header
-  eyecatcher=no,
+  eyecatcher=yes,
   headerborder=open,
@@
   % Eye Catcher
-  {} % No eye catcher for this poster. If an eye catcher is present, the title is centered between eye-catcher and logo.
+  {{\begin{minipage}{7em}
+    \includegraphics[height=5em]{amllogo.eps}
+  \end{minipage}}
+  }
   % Title
@@
   % University logo
-  {{\begin{minipage}{16em}
+  {{\begin{minipage}{7em}
     \hfill
-    \includegraphics[height=5.5em]{byunlp.eps}
+    \includegraphics[height=5em]{byunlp.eps}
   \end{minipage}}
@@
 	\begin{centering}
-	  \psfrag{b}[C][C]{\huge$\beta$}
-	  \psfrag{a}[C][C]{\huge$\alpha$}
-	  \psfrag{t}[C][C]{\huge$\theta$}
-	  \psfrag{f}[C][C]{\huge$\phi$}
-	  \psfrag{w}[C][C]{\huge$w$}
-	  \psfrag{z}[C][C]{\huge$z$}
-	  \psfrag{N}[C][C]{\Large$N$}
-	  \psfrag{M}[C][C]{\Large$M$}
-	  \psfrag{K}[C][C]{\Large$K$}
-	  \includegraphics[height=.82\linewidth]{lda_model.eps}
+	  \psfrag{b}[C][C]{\Large$\beta$}
+	  \psfrag{a}[C][C]{\Large$\alpha$}
+	  \psfrag{t}[C][C]{\Large$\theta$}
+	  \psfrag{f}[C][C]{\Large$\phi$}
+	  \psfrag{w}[C][C]{\Large$w$}
+	  \psfrag{z}[C][C]{\Large$z$}
+	  \psfrag{N}[C][C]{\large$N$}
+	  \psfrag{M}[C][C]{\large$M$}
+	  \psfrag{K}[C][C]{\large$K$}
+	  \includegraphics[height=.485\linewidth]{lda_model.eps}
 	  \begin{align*} 
+		M &= \mathrm{Number\ of\ documents} \\
+		N &= \mathrm{Number\ of\ words\ in\ a\ document} \\
+		K &= \mathrm{Number\ of\ topics} \\
+		w &= \mathrm{Word\ in\ a\ document} \\
+		z &= \mathrm{Topic\ corresponding\ to\ word\ } w \\
 		w_i|z_i,\phi^{(z_i)} &\sim \mathrm{Categorical}(\phi^{(z_i)}) \\
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Results}
-  {name=results word distributions,column=1,span=2,row=0}{
+  \headerbox{Results: Topic--Word Distributions}
+  {name=words,column=1,span=2,row=0}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 	used in each topic.  In the algorithm, the topics just have numbers; the
-	computer merely looks for statistical relationship between words.  The
-	labels are thus our own.  We present here a few of the topics we found.
+	computer merely looks for statistical relationships between words.  Even
+	so, the topics are easily distinguishable.  We present here a few of the
+	topics we found.
 
-	\small
 	\begin{center}
-	  \begin{tabular}{ccccc}
-		Topic 19: Tithing&Topic 33: Welfare&Topic 52: Women&Topic 37: Families
-		&Topic 7: Government\\
+	  \begin{tabular}{cccccccc}
+		Topic 19&Topic 33&Topic 52&Topic 37&Topic 7&Topic 26&Topic 28&Topic 22\\
 		\hline
-		tithing&welfare&women&children&government \\
-		pay&poor&society&home&united \\
-		money&church&relief&family&liberty \\
-		debt&fast&sisters&parents&nation \\
-		means&help&hinckley&teach&states \\
-		out&services&each&child&country \\
-		people&program&b&families&freedom \\
-		one&plan&gordon&homes&land \\
-		paid&care&lives&love&free \\
-		poor&needs&service&mother&men \\
-		tithes&food&sister&taught&constitution \\
-		get&need&forward&fathers&people \\
-		more&family&strength&responsibility&america \\
+		tithing&welfare&women&children&government&see&years&true \\
+		pay&poor&society&home&united&d&one&honest \\
+		money&church&relief&family&liberty&c&two&good \\
+		debt&fast&sisters&parents&nation&1&hundred&integrity \\
+		means&help&hinckley&teach&states&2&five&virtue \\
+		out&services&each&child&country&3&three&character \\
+		people&program&b&families&freedom&s&thousand&honesty \\
+		one&plan&gordon&homes&land&ne&twenty&without \\
+		paid&care&lives&love&free&4&ago&one \\
+		poor&needs&service&mother&men&5&four&men \\
+		tithes&food&sister&taught&constitution&alma&year&honor \\
+		get&need&forward&fathers&people&6&six&respect \\
+		more&family&strength&responsibility&america&7&over&god \\
 	  \end{tabular}
@@
 
-	\normalsize
-	\mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\
+  }
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \headerbox{Results: Topics over Time}
+  {name=results topics over time,column=1,span=2,below=words}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 	Because each talk has a date associated with it, we can use LDA to analyze
-	the distribution of these topics over time, seeing at when certain topics
-	were most discussed.  In doing this, we see clear correlations with
+	the distribution of these topics over time, seeing at what points certain
+	topics were most discussed.  In doing this, we see clear correlations with
 	real-world occurrences.
@@
 	\begin{center}
-	  \includegraphics[height=.34\linewidth]{topic19.eps}
-	  \includegraphics[height=.34\linewidth]{topic33.eps}
+	  \includegraphics[height=.305\linewidth]{topic19.eps}
+	  \includegraphics[height=.305\linewidth]{topic33.eps}
 
-	  \includegraphics[height=.34\linewidth]{topic52.eps}
-	  \includegraphics[height=.34\linewidth]{topic56.eps}
+	  \includegraphics[height=.305\linewidth]{topic52.eps}
+	  \includegraphics[height=.305\linewidth]{topic56.eps}
 	\end{center}
 
-  }%
+  }
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Future Work}{name=future work,column=3,span=1}{
+  \headerbox{Results: Topics by Speaker}
+  {name=speaker,column=3,span=1}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+	Because each talk has a speaker, we can look at topics to see who was most
+	likely to use them in their talks.  Again we see here that our automatic
+	analysis is validated by real-world occurrences.
+
+	\begin{center}
+	  \begin{tabular}{cc}
+		Relief Society Topic&Sunday School Topic\\
+		\hline
+		Sister Barbara B. Smith&Karl G. Maeser \\
+		Mary Ellen W. Smoot&George Reynolds \\
+		Mary Ellen Smoot&E. C. Phillips \\
+		Elaine L. Jack&Josiah H. Burrows \\
+		Marian R. Boyer&George Goddard \\
+		Virginia U. Jensen&Joseph J. Jackson\\
+		Wayne M. Hancock&L. John Nuttall \\
+		Shirley W. Thomas&Sunday School Union \\
+		Bonnie D. Parkin&John D. Peters \\
+		Kathleen H. Hughes&Joseph Jackson \\
+		Sheri L. Dew&George Q. Cannon \\
+		Richard H. Winkel&Heber J. Grant \\
+		Aileen H. Clyde&Joseph Summerhays \\
+	  \end{tabular}
+	\end{center}
+
+  }
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \headerbox{Future Work}{name=future work,column=3,span=1,below=speaker}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 	source---LDS General Conference talks.  However, our exploration of the
-	results has been quite preliminary.  Our research lays the groundwork for a
-	more thorough analysis of religious discourse in the LDS Church.  These
-	techniques allow us to quickly answer questions such as, ``Which topics are
-	constant over time, and which come and go?'' and ``Which years had the
-	highest diversity of topics, and which were more narrow-focused?''
+	results has been quite preliminary.  Future areas of research could 
+	include:
+	\begin{itemize}
+	  \item A more thorough analysis of religious discourse in the LDS Church,
+		answering questions such as ``Which topics are constant over time, and
+		which come and go?''
+	  \item Explore other religious and scholarly LDS texts, including the
+		Journal of Discourses, the Book of Mormon, the Doctrine and Covenants,
+		the collected works of Hugh Nibley, or the past 50 years of BYU
+		Studies.
+	  \item Not only explore other data sets in isolation, but compare them;
+		see how the past 100 years of General Conference differs from the
+		previous 100 years in the Journal of Discourses.
+	  \item The use of other natural language processing techniques to analyze
+		the data.  Possibilities include Topical N-grams, an extension of LDA
+		that returns strings of words in the topic instead of isolated words.
+	\end{itemize}
 
-	We also have only considered General Conference talks.  A fruitful avenue
-	of future work would be to compare the results of our techniques on many
-	different corpora of LDS religious and scholarly texts, such as the Journal
-	of Discourses, the Book of Mormon, the Doctrine and Covenants, the
-	collected works of Hugh Nibley, or the past 50 years of BYU Studies.  Such
-	analysis could broaden our understanding of and appreciation for our
-	textual history.
+  }
 
-  }%
+\end{poster}
 
-\end{poster}%
-%
 \end{document}

### 2010-03-14 honors_thesis/journal_paper/paper.tex
mathy version added with hacked text

@@
 %% PSO Stuff
-\DeclareMathOperator{\URand}{U}
 \DeclareMathOperator*{\argmin}{arg\;min}
@@
 \DeclareMathOperator*{\argsup}{arg\;sup}
+\providecommand{\pers}{\ensuremath{c}}
+\providecommand{\neigh}{\ensuremath{s}}
+\providecommand{\nURand}{\ensuremath{U^\neigh}}
+\providecommand{\pURand}{\ensuremath{U^\pers}}
 \providecommand{\ppos}{\ensuremath{\Vec{x}}}
 \providecommand{\pvel}{\ensuremath{\Vec{v}}}
-\providecommand{\nbest}{\ensuremath{\Vec{b}_N}}
-\providecommand{\pbest}{\ensuremath{\Vec{b}_P}}
+\providecommand{\nbest}{\ensuremath{\Vec{x}^{\neigh*}}}
+\providecommand{\pbest}{\ensuremath{\Vec{x}^{\pers*}}}
 \providecommand{\constriction}{\ensuremath{\chi}}
-\providecommand{\coeff}{\ensuremath{\phi}}
+\providecommand{\ncoeff}{\ensuremath{\phi^s}}
+\providecommand{\pcoeff}{\ensuremath{\phi^P}}
 \providecommand{\obs}{\ensuremath{\Vec{\xi}}}
+\providecommand{\ofunc}{\ensuremath{f}}
+\providecommand{\swarm}{\ensuremath{swarm}}
 
 %SpecExPSO Stuff
+\providecommand{\indic}{\ensuremath{I}}
+\providecommand{\specvel}{\ensuremath{V}}
+\providecommand{\specpos}{\ensuremath{X}}
+\providecommand{\ln}{\ensuremath{ln}}
+\providecommand{\rn}{\ensuremath{rn}}
+\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
+\providecommand{\casedef}{\ensuremath{(\pbest:\nbest)}}
+\providecommand{\casexn}{\ensuremath{(x:\neigh*)}}
+\providecommand{\casexx}{\ensuremath{(x:x)}}
+\providecommand{\casexl}{\ensuremath{(x:\ln)}}
+\providecommand{\casexr}{\ensuremath{(x:\rn)}}
+\providecommand{\casepn}{\ensuremath{(\pers*:\neigh*)}}
+\providecommand{\casepl}{\ensuremath{(\pers*:\ln)}}
+\providecommand{\casepr}{\ensuremath{(\pers*:\rn)}}
 \providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
@@
 the world's computing power is increasingly coming from large clusters of
-processors.  Even in desktop machines, systems with 4 or 8 cores are
-commonplace.  In order to efficiently utilize these resources for
+processors.  i
+%Even in desktop machines, systems with 4 or 8 cores are
+%commonplace.  
+In order to efficiently utilize these resources for
 computationally intensive problems, algorithms need to run in parallel.
@@
 
-All of these parallel techniques use additional processors to increase the
+The purpose of this paper is to consider pso parallelization strategies for
+clusters of thousands of processors and functions for which a single
+evaluation will take on the order of hours.
+Our purpose is to explore the question ``since pso is often run with
+a swarm of size of 100, what would one do if 500 or a few thousand processors were available?''
+
+Previous parallel techniques use additional processors to increase the
 number of particles in the swarm, either by adding individual particles or by
-adding entire new subswarms.  The number of iterations of PSO that the
+adding entire new sub-swarms but in all cases increasing the number of particles.  The number of iterations of PSO that the
 algorithms can perform is thus inherently limited by the time it takes to
 evaluate the objective function---additional processors add more particles, not
-more iterations.  However, there are many functions for which performing more
+more iterations.
+for many functions there comes a point of diminishing returns with respect to adding particles.
+>>>>>>>>> TODO WARNING WARNING MADE UP NUMBERS GET REAL ONES!! <<<<<<<<<<<<<<
+For the well-know benchmark function Sphere and as shown in Figure XXXXa increasing the swarm size from 20 to 120 has a significant effect on the 
+speed of convergence. In this case, the swam goes below $1*10^5$ in x iteration with 20 particles, but 
+achieves this same value in  y iterations with a swarms size of 120, the additional 100 particles.
+However consider the change going from 2000 particles to 2100 particles. Here the additional 100 particles
+reduce the number of iterations to achieve $1*10^5$ from x iterations to y.
+Other benchmark function behave similarly. ans shown in fig xxxxb for ?? and xxxxc ?? (in all case 20 dimensions, average of 50 runs).
+Certainly these benchmark functions do not prove that this will be the case for all functions,
+but it does provide some evidence that for at least some functions, adding additional particles does not seem to have much benefit
+after some point, some time earlier fig xx some time later. This phenomena has been commented on by others
+in the form of recommended swarms sizes for various types of functions cite standard swarm paper.
+\begin{quote}
+Skip this??
+
+However, there are many functions for which performing more
 iterations with fewer particles is a better use of resources than performing
 fewer iterations with more particles.  
+\end{quote}
 
@@
 has been done in simulated annealing~\cite{witte-tpds91}.
+
+\begin{quote}
+Skip this??
 
@@
 
+But the following does need some intro
+\end{quote}
+
 The original PSO can be reproduced exactly, two iterations at a time, using
@@
 At each iteration of the algorithm, the position $\ppos_t$ and velocity
-$\pvel_t$ of each particle are updated as follows:
+$\pvel_t$ of each particle $i$ are updated as follows:
 \begin{align}
 \label{eq:velupdate}
-	\pvel_{t+1} &=
+	\pvel_{i,t+1} &=
 		\constriction \left[ \pvel_t +
-			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
-			\coeff_2\URand()\otimes(\nbest - \ppos_t)
+			\pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) +
+			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
 		\right] \\
 \label{eq:posupdate}
-	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
+	\ppos_{i,t+1} &= \ppos_{i,t} + \pvel_{i,t+1}
 \end{align}
-where \( \URand() \) is a vector of random numbers drawn from a uniform
+where \( \nURand_{i,t} \) and \( \nURand_{i,t} \) are vectors of i.i.d. random numbers drawn from a uniform
 distribution, the \( \otimes \) operator is an element-wise vector
@@
 current particle has seen, and $\nbest$ (called neighborhood best) is the best
-position the neighbors of the current particle have seen.  \( \coeff_1 \), \(
-\coeff_2 \), and \( \constriction \) are parameters with prescribed values
+position the neighbors of the current particle have seen.  \( \ncoeff \), \(
+\pcoeff \), and \( \constriction \) are parameters with prescribed values
 required to ensure convergence (2.05, 2.05, and .73,
-respectively)~\cite{clerc-tec02}.
+respectively)~\cite{clerc-tec02}. Note also that we have made the particle identifier $i$
+and the time step $t$ or $t+1$ explicit in the subscript and moved other qualifiers,
+such as the personal $P$ and personal best $P*$ qualifiers to the superscript position.
+Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.
 
@@
 topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
-as a parallel adaptation of the Complete topology called Random that has been
+as a parallel surrogate for the Complete topology called Random that has been
 shown to simulate the behavior of Complete with far less
@@
 experiments.
+%TODO except as noted??
 
@@
 possible next positions of the particle, evaluate all of them at once, and one
-of them will be right.
+of them would have been computed by normal PSO.
+
+We do this as described here.
+
+First consider formula for the \emph{next} iteration of PSO, which is identical to 
+Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time subscripts are incremented.
+As noted above and for simplicity in this explanation, we restrict our discussion here to
+the case of a Ring topology with two neighbors which we will call the
+``right neighbor'' and ``left neighbor'' neighbors identified for each
+particle $i$ as $\rn(i)$ and $\ln(i)$ respectively.
+
+Next note that the computation of $\pbest$ and $\nbest$ is not seen in
+the update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so
+we will make this mathematically explicit.
+This update is usually implemented as a pair of independent ``if``-statements, but that is
+a bit awkward for our purpose here so we instead use a set of indicator variables,
+one each representing each possible update to the \emph{pair} $\casedef$.
+These indicator variables allow us to easily identify the cases in which updates will
+be needed for $\nbest$, $\pbest$ or both.
+
+For example, we can write $\casexn$ to identify the case where $\pbest$ will
+be updated with the better position $\ppos_{i,t}$, shown in the first positions,
+but a no new neighborhood best was found so
+$\nbest$ remains in the second position.
+
+For any topology with 2 neighbors there are 7 possible update cases listed in
+Table~\ref{tab:evals}.
+This list assumes, as is typically the case, that each particle is also
+part of its own neighborhood (can cause an update to $\nbest$). The list also
+omits the case where $\nbest$ is updated by $\ppos_{i,t}$ but $\pbest$ was not updated, as this is
+logically impossible. $\caseset$ represents the set of all cases.
+
+For minimization, we can define the indicator variable representing the case where for particle $i$ at time $t$
+$\pbest$ will
+be updated with the better position $x_{i,t}$ but
+$\nbest$ will not updated by:
+\begin{align}
+\nonumber
+	\indic_{i,t}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
+\nonumber
+	& 1 \ if \  \ofunc(\ppos_{i,t+1}) > \ofunc(\pbest_{i,t}) \\
+\nonumber
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{i,t}) \\
+\nonumber
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\ln(i),t}) \\
+\nonumber
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\rn(i),t}) \\
+\label{eq:deficasexn}
+	& 0 \ otherwise
+\end{align}
+
+For compactness in our equations we represent this without the parameters since
+they can be inferred from the subscripts on $\indic_{i,t}^{\casexn}$.
+The indicators variables for the other update cases shown in 
+Table~\ref{tab:evals}
+($\casepn$, $\casepl$, $\casepr$, etc.)
+can be defined in similar fashion but obviously using the
+corresponding boolean expression.
+
+For each cases identified in
+Table~\ref{tab:evals}
+and identified by a corresponding indicator variable defined above,
+we define also a velocity update equation; once again using $\casexn$ as an example:
+\begin{align}
+\nonumber
+	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
+\label{eq:defvocasexn}
+		& \constriction \left[ \pvel_t +
+			\pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) +
+			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
+		\right]
+\end{align}
+
+Again we will drop the parameters for compactness and since
+they can be inferred from the subscripts on $\specvel_{i,t}^{\casexn}$.
+Since this equation is restricted to a specific case, $\casexn$ for this example,
+it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
+$\pbest_{i,t+1}=\ppos_{i,t+1}$ and $\nbest_{i,t+1}=\nbest_{i,t+1}$ we can re-write
+$\specvel_{i,t}^{\casexn}$ as:
+\begin{align}
+\nonumber
+	\specvel_{i,t+1}^{\casexn} & = \\
+\label{eq:defvcasexn}
+		& \constriction \left[ \pvel_t +
+			\pcoeff\pURand_{i,t}\otimes(\ppos_{i,t} - \ppos_{i,t}) +
+			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
+		\right]
+\end{align}
+
+For each cases identified in
+Table~\ref{tab:evals}
+and identified by a corresponding indicator variable defined above,
+we define also a position update equation with its compact form; once again using $\casexn$ as an example:
+\begin{align}
+\label{eq:defpcasexn}
+	\specpos_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
+\nonumber
+	\specpos_{i,t+1}^{\casexn} & =  \ppos_{i,t} + \specvel_{i,t+1}^{\casexn}
+\end{align}
+
+With this notation we can re-write the ``next iteration'' form of the PSO motion equations as:
+\begin{align}
+\label{eq:vel2update}
+	\pvel_{i,t+2} &= \sum_{c \in \caseset} \ \specvel_{i,t+1}^{c} \ \indic_{i,t}^{c} \\
+\label{eq:pos2update}
+	\ppos_{i,t+2} &= \sum_{c \in \caseset} \ \specpos_{i,t+1}^{c} \ \indic_{i,t}^{c}
+\end{align}
+
+The value of the objective function at $\ppos_{i,t+2}$ is given by:
+\begin{align}
+\label{eq:pos2value}
+	\ofunc (\ppos_{i,t+2}) = \sum_{c \in \caseset} \ \ofunc(\specpos_{i,t+1}^{c}) \ \indic_{i,t}^{c}
+\end{align}
+or, with $\specpos$ and $\indic$ written with the parameters which were omitted above:
+\begin{align}
+\nonumber
+\ofunc & (\ppos_{i,t+2}) = \sum_{c \in \caseset} \\
+\nonumber
+& \ofunc(\specpos_{i,t+1}^{c}(\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})) \\
+\label{eq:pos2valuelong}
+& \indic_{i,t}^{c}(\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))
+\end{align}
+
+None of the $\specpos_{i,t+1}^{c} \ \ c \in\ \caseset$ depend upon $f(i,x_t)$
+or any other objective function evaluation at $t+1$.
+Thus PSO has been refactored such that the algorithm can begin computing
+all of the objective function evaluations needed in iteration $t+1$ before $f(x_t)$ is computed.
+
+This refactoring comes a significant expense, instead of just computing
+$\ofunc(\ppos_{i,t+1})$ and then $\ofunc(\ppos_{i,t+2})$ in the usual way, this refactored PSO
+can compute $\ofunc(\ppos_{i,t+1})$ and all 
+of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time.
+Once all of the evaluations $f(\ppos_{i,t+1})$ are completed only one
+of the indicator functions $\indic_{i,t}^{c} \ \ c \in \caseset$ will have been set to one;
+the rest will be ignored, and might just as well never have been computed.
+We call these evaluated points $\specpos_{i,t+1}^{c}$ ``speculative children''
+because they may or may not be needed.
+
+The potential value of these speculative children lies in the phase used above
+``this refactored PSO
+can compute $\ofunc(\ppos_{i,t+1})$ and all 
+of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
+Remember that our intent is to accelerate parallel PSO and that we
+are interested in objective functions that have at least non-trivial computational
+complexity.
+Thus we assume that computing
+$\ofunc(\ppos_{i,t+1})$ and all 
+of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ is the computationally difficult part.
+
+Suppose that unlimited processors were available and that the evaluation of 
+the objective function took 1 hour. If for each particle $i$ 
+the computations
+$\ofunc(\ppos_{i,t+1})$ and all 
+of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ were sent to separate
+processors and completed in parallel.
+In this environment all of the function evaluations needed for 2 iterations of PSO
+could be completed in just one hour where as the conventional implementation would
+require two hours for those same 2 iterations of PSO.
+
+the same time''. 
+in the indicator functions $indic_{i,t}^{c} \ \ c \in \caseset$
+given the values $\ofunc(\ppos_{i,t+1}) i \in \swarm$ 
+is inconsequential relative to the computational cost of obtaining those values 
+$\ofunc(\ppos_{i,t+1})$ and all
 
@@
   \centering
-  \begin{tabular}{cc}
-	$\pbest$ updated&Source of $\nbest$ update\\
+  \begin{tabular}{lcc}
+	Identifier&$\pbest$ updated&Source of $\nbest$ update\\
 	\hline
 	\hline
-	No update&No update\\
+	$\casepn$&No update&No update\\
 	\hline
-	No update&Neighbor 1\\
+	$\casepl$&No update&Left Neighbor\\
 	\hline
-	No update&Neighbor 2\\
+	$\casepr$&No update&Right Neighbor\\
 	\hline
-	Updated&No update\\
+	$\casexn$&Updated&No update\\
 	\hline
-	Updated&Neighbor 1\\
+	$\casexl$&Updated&Left Neighbor\\
 	\hline
-	Updated&Neighbor 2\\
+	$\casexr$&Updated&Right Neighbor\\
 	\hline
-	Updated&The particle itself\\
+	$\casexx$&Updated&The particle itself\\
 	\hline

### 2010-03-15 honors_thesis/journal_paper/paper.tex
Moved discussion to appendix, and other fixes from honors thesis

--- 2010-02-25 honors_thesis/journal_paper/paper.tex
@@
 
-We also propose another way of making use of these speculative particles,
-keeping the best position found instead of the position that PSO actually would
-have taken.  We show that for a number of functions, speculative evaluation
-gives dramatic improvements over adding additional particles to the swarm.
+We show that while speculative evaluation in PSO is an intriguing idea, its
+usefulness is limited unless some slight modifications are made to the PSO
+algorithm.  When we relax the requirement that speculative evaluation exactly
+reproduce the behavior of the original PSO, we can make improvements such as
+using all of the extra speculative particles, instead of throwing away all but
+one of them, or speculating several iterations ahead instead of just one.
+Using these algorithmic modifications, we see dramatic improvements on a number
+of functions over simply adding additional particles to the swarm.
 
@@
 branch on which they must wait (e.g. a memory cache miss), guess which way the
-branch will go and start executing.  If the processor guessed right, execution
+branch will go and start executing.  If the processor guesses right, execution
 is much farther ahead than if it had just idly waited on the branch.  If it
-guessed wrong, execution restarts right where it would have been anyway.  This
+guesses wrong, execution restarts right where it would have been anyway.  This
 is called speculative execution or branch prediction.
@@
 literature are the Ring topology and the Complete
-topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
-as a parallel adaptation of the Complete topology called Random that has been
-shown to simulate the behavior of Complete with far less
-communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
-be defined as having $n$ neighbors; we use 2 neighbors for all of our
-experiments.
+topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
+neighbor to either side of it; in the Complete topology, every particle is a
+neighbor to every other particle.  In all topologies, a particle is also a
+neighbor to itself, in that its own position is considered when updating the
+particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
+particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
+topology, each particle has $p$ neighbors.  In this paper we use these
+topologies, as well as a parallel adaptation of the Complete topology, called
+Random, that has been shown to simulate the behavior of Complete with far less
+communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
+randomly picks two other particles to share information with at each iteration,
+along with itself.  Thus in both the Ring and the Random topologies, all
+particles have 3 neighbors.
 
@@
 $2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
-neighbors the particle has.  If a particle is also its own neighbor, as is most
-often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
-a particle cannot be the source of its $\nbest$ update while not updating its
-$\pbest$.  Thus we have $2(n+1)-1$, or $2n+1$, evaluations.
-Table~\ref{tab:evals} lists all possible combinations of updates that can be
-seen by a particle with two neighbors other than itself, which is the case in
-all of the topologies we use.  If all of these speculative positions are
-evaluated in parallel along with the original, non-speculative position of the
-particle at time $t-1$, then one of the speculative positions has to be the
-correct one, and two iterations have been evaluated at once.
+neighbors the particle has.  Because a particle is also its own neighbor, one
+of the $2(n+1)$ speculative positions can be eliminated, as a particle cannot
+be the source of its $\nbest$ update while not updating its $\pbest$.  Thus we
+have $2(n+1)-1$, or $2n+1$, evaluations.  Table~\ref{tab:evals} lists all
+possible combinations of updates that can be seen by a particle with two
+neighbors other than itself, which is the case in all of the topologies we use.
+If all of these speculative positions are evaluated in parallel along with the
+original, non-speculative position of the particle at time $t-1$, then one of
+the speculative positions has to be the correct one, and two iterations have
+been evaluated at once.
 
@@
 
-\subsection{Alternate form of message passing}
-
 In some distributed frameworks, like the MapReduce framework that we used in
@@
 described uses three rounds of message passing for every two iterations
-(corresponding to steps 2, 5 and 10 in \alg{centralized}).  Accordingly, we
-describe a method that requires only one round of communication for each pair
-of iterations, which happens at step 5 of \alg{centralized}.  Many more
-messages are needed, but that is sometimes more desirable than synchronizing
-all of the machines three times.
+(corresponding to steps 2, 5 and 10 in \alg{centralized}).  It is possible to
+perform speculative evaluation in PSO with only one round of communication per
+two iterations.  However, the methodology is tedius and is not the focus of
+this paper, so we defer its description to the Appendix.
+
+\subsection{Dynamic Topologies}
+
+Performing speculative evaluation in PSO with a dynamic topology raises a
+sticky issue of its own.  In a static topology, at iteration $t$ a particle
+already has all of the information about the positions of its neighbors during
+iterations $1$ through $t-1$.  If the neighbor finds a better position at
+iteration $t$, the particle updates its neighborhood best, but if it does not,
+it still has its old neighborhood best from its neighbors for all previous
+iterations.
+
+In a dynamic topology, a particle might not have information about the previous
+positions of its neighbors at iteration $t$.  That means that its new
+neighborhood best could come not only from its neighbors' positions at
+iteration $t$, but also from their personal best from iteration $t-1$, as
+neighbors' personal bests are what are used to update a particle's neighborhood
+best.  That creates a problem for speculative evaluation---there are
+potentially more than $2n+1$ possible next positions, increasing the amount of
+work that must be done to perform the second iteration at the same time as the
+first.
+
+This is easily fixed by updating each new particle $\noeval{\p}_{t+2}$ with the
+currently available information about its neighbors $\noeval{\nset}_{t+2}$
+before producing its children $\noeval{\sset}_{t+3}$.  If a particle
+$\noeval{\p}_{t+2}$ updates its neighborhood best with the personal bests of
+$\noeval{\nset}_{t+2}$ before calculating the next possible positions for
+$\noeval{\sset}_{t+3}$, there are still only $2n+1$ possible next positions,
+and the problem is avoided.
+
+\subsection{Comparison with Standard PSO}
+\label{sec:results}
+
+We ran experiments to compare our speculative PSO algorithm to the standard
+PSO algorithm.  At each iteration of the algorithms, we use one processor to
+perform one function evaluation for one particle, be it speculative or
+non-speculative.  The speculative algorithm actually performs two iterations of
+PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
+evaluations.  For benchmark functions with very fast evaluations this may not
+be the best use of parallelism in PSO.  But for many real-world applications,
+the objective function takes on the order of minutes or more to evaluate; in
+such cases our assumption is perfectly reasonable.
+
+To use the same number of processors for each algorithm, speculative evaluation
+needs a smaller swarm size than standard PSO.  For each of the topologies we
+used, a particle has two neighbors other than itself.  As shown in
+Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
+particle.  With one evaluation needed for the original, non-speculative
+particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
+the number of particles in the speculative swarm.  The extra evaluations done
+in speculative evaluation would instead be particles in regular PSO, so we
+compare swarms of size $p_s$ in speculative evaluation with swarms of size
+$8p_s$ in regular PSO.
+
+Where the Complete topology would normally be used, we use a Random topology in
+our speculative algorithm, as Complete leads to an explosion of speculative
+evaluations.  If speculative evaluation were not being performed, it is
+possible that the Complete topology would be used.  However, the Complete
+topology also requires a very large amount of interprocessor communication in
+parallel PSO, so it is still quite possible that Random would be used even with
+standard PSO.  But, to be fair in our comparisons, we compare to the standard
+PSO algorithm using both the Random topology with two neighbors and the
+Complete topology.  All functions have 20 dimensions, and all results shown are
+averages over 20 runs.  We use 240 processors in our experiments, so our
+speculative algorithm uses 30 particles and standard PSO uses 240 particles.
+For ease of labeling, we call our speculative algorithm SEPSO and standard PSO
+just PSO, with the topology following.
+
+First we look at the function Sphere, the simplest of common benchmark
+functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
+has no local optima and is best optimized in terms of function evaluations with
+a small swarm using a Complete topology.  Accordingly, we show results for a
+Complete swarm in \fig{basic-sphere}.  In comparing SEPSO with PSO, we can see
+that when both are using a Random topology, SEPSO clearly outperforms PSO.
+However, the comparison between SEPSO and PSO Complete is not as
+encouraging---PSO Complete performs slightly better.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere.eps}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
+  two iterations per round, while standard PSO algorithms use 240 processors
+  and do one iteration per round.}
+  \label{fig:basic-sphere}
+\end{figure}
+
+Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
+\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
+\right) + 1$.  It is generally recommended to use the Ring topology when
+optimizing the Griewank function, as Complete is prone to premature convergence
+on a local optima.  We show results in \fig{basic-griewank1} for swarms of size
+30 and 240 using the Ring topology. One can see in the figure that using
+speculative evaluation initially outperforms PSO, but gets stuck in a local
+optima and is eventually outperformed.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank1.eps}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations.}
+  \label{fig:basic-griewank1}
+\end{figure}
+
+\fig{basic-griewank1} is slightly deceptive, however, as with Griewank some of
+the runs found the global optimum, 0, while some did not.  A more accurate plot
+for this function is one showing what percentage of the runs have found the
+global optimum at each iteration.  \fig{basic-griewank2} shows such a plot.  A
+swarm of size 30 performing the PSO algorithm gets stuck a little less than
+half of the time, while 240 particles provides enough exploration to find the
+global optimum every time.  As would be expected, when our speculative
+algorithm finds the global optimum, it finds it about twice as fast as regular
+PSO---it is just performing regular PSO twice as fast with a smaller swarm
+size.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank2.eps}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations.  Instead of showing average function value, we show the
+  percentage of runs that have found the global optimum by each iteration.}
+  \label{fig:basic-griewank2}
+\end{figure}
+
+Because our speculative algorithm got stuck with such a small swarm size, we
+ran a set of experiments on Griewank with 800 processors, giving swarms of size
+100 and 800 instead of 30 and 240.  \fig{basic-griewank3} shows the results.
+\fig{basic-griewank3} is very similar to \fig{basic-griewank2}, except that
+speculative evaluation finds the global optimum essentially 100\% of the time.
+Thus with a few more processors, our algorithm finds the optimum twice as fast
+as the original PSO without a significant loss of accuracy.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank3.eps}
+  \caption{Function Griewank with a swarm that uses 800 processors per round of
+  evaluations.}
+  \label{fig:basic-griewank3}
+\end{figure}
+
+Lastly we look at a function that does not show any kind of improvement with
+this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
+\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
+with Rastrigin, the more particles there are in the swarm, the lower function
+value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
+swarms get caught in local optima.  Because our speculative algorithm has a
+significantly smaller swarm size, it gets stuck at higher values while the
+larger swarms performing regular PSO continue to improve the best value found.
+\fig{rastrigin} shows the results graphically.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
+  \caption{Function Rastrigin with a swarm that uses 240 processors per round
+  of evaluations.}
+  \label{fig:rastrigin}
+\end{figure}
+
+\section{Relaxing the Requirements}
+\label{sec:relax}
+
+While the idea of speculative evaluation in particle swarm optimization is
+intriguing, the results of the previous section show that in many cases it
+requires too many extra evaluations to produce competitive results.  On most
+functions, the larger swarm size that can be used in regular PSO leads to
+better performance than is found by trying to do speculative evaluation.
+However, if we keep the idea of speculative evaluation while relaxing the
+requirement of exactly reproducing the behavior of the original PSO algorithm,
+we see some very impressive results.
+
+\subsection{Pick the Best Child}
+\label{sec:pickbest}
+
+In performing speculative evaluation as we have described it, $2n+1$
+speculative evaluations are done per particle, while all but one of them are
+completely ignored.  We can try to make use of those evaluations instead of
+throwing them away.
+
+To use the extra speculative evaluations, instead of choosing the speculative
+child that matches that branch that the original PSO would have taken, we take
+the child that has the best value.  The methodology is exactly the same as
+above except for the process of choosing which speculative child to accept.
+The branch that PSO would have taken is ignored, but $\p_t$ must still be
+created so it can give the right personal best and neighborhood best values and
+positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
+step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
+$\noeval{\sset}_{t+1}$ instead of with the matching branch.
+
+This can be thought of as drawing a number of samples from the next iteration
+and accepting the best one.  Speculative particles that move in good directions
+are kept.  It seems this modification to PSO would produce more exploitation,
+decreasing the amount of exploration.  With functions that are less deceptive
+we would expect this method to work well, while performance on more deceptive
+functions would probably be hurt.
+
+We show in \fig{sphere-pickbest} the results of using this modification on the
+function Sphere.  For ease of labeling, we call the modified algorithm SEPSO
+Pick Best.  As can be seen, where SEPSO did not manage to perform as well as
+PSO Complete, SEPSO Pick Best handily outperforms it with the same number of
+processors.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere2.eps}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing SEPSO Pick Best with previous results.}
+  \label{fig:sphere-pickbest}
+\end{figure}
+
+Sphere is a function for which very little exploration needs to be done, so
+always keeping the best speculative position found makes intuitive sense.  A
+less intuitive result occurs when using SEPSO Pick Best with the function
+Griewank.  Griewank is a highly deceptive function prone to premature
+convergence.  Our intuition on the behavior of picking the best speculative
+child is that it would tend to prematurely converge more often than the
+original PSO.  However, as seen in \fig{griewank-pickbest}, SEPSO Pick Best
+increases the chance of finding the optimum while also reducing the number of
+iterations required to converge.  Perhaps SEPSO Pick Best encourages local
+exploitation, while the sparse communication of the Ring topology still allows
+for global exploration.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank4.eps}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing SEPSO Pick Best with previous results.}
+  \label{fig:griewank-pickbest}
+\end{figure}
+
+\subsection{Pruning the Speculative Evaluations}
+
+Because the problem facing speculative evaluation is having too many possible
+evaluations to perform, a natural step to take is to eliminate some of the
+evaluations.  If we could reliably predict which branch were going to be taken,
+we could limit ourselves to one speculative evaluation per particle instead of
+$2n+1$.  While we cannot with certainty predict the correct branch every time,
+we can look at the statistics of the behavior of particles in general and
+perhaps come up with a good guess, reducing the number of necessary speculative
+evaluations.
+
+\subsubsection{Branch Statistics}
+
+In Table~\ref{tab:evals} we presented all possible branches that a particle
+with two neighbors could take.  Here we lump all of the neighbors together and
+consider the statistics for the five branches shown in Table~\ref{tab:branches}.
+
+\begin{table}[h]
+  \caption{Five Branches to Consider for Statistics}
+  \label{tab:branches}
+  \centering
+  \begin{tabular}{rc|c}
+	&$\pbest$ updated&Source of $\nbest$ update\\
+	\hline
+	1.&No update&No update\\
+	2.&Updated&No update\\
+	3.&Updated&The particle itself\\
+	4.&No update&Some neighbor\\
+	5.&Updated&Some neighbor\\
+  \end{tabular}
+\end{table}
+
+\begin{table}[h]
+  \caption{Branch Statistics in PSO}
+  \label{tab:stats}
+  \centering
+  \begin{tabular}{c|c|c|c|c|c|c}
+	Topology&Function&1&2&3&4&5\\
+	\hline
+	\hline
+	Ring&Sphere&53.0\%&9.3\%&11.4\%&20.2\%&6.2\%\\
+	&Griewank&51.7\%&8.4\%&12.2\%&20.7\%&7.0\%\\
+	&Rastrigin&49.5\%&4.8\%&14.6\%&21.3\%&9.9\%\\
+	&Rosenbrock&51.3\%&7.4\%&12.9\%&21.1\%&7.3\%\\
+	\cline{2-7}
+	&Average&{51.3\%}&{7.5\%}&{12.8\%}&{20.8\%}&
+	{7.6\%}\\
+	\hline
+	\hline
+	Random&Sphere&66.7\%&11.9\%&2.6\%&15.6\%&3.1\%\\
+	&Griewank&69.0\%&10.9\%&2.5\%&14.9\%&2.7\%\\
+	&Rastrigin&81.9\%&5.5\%&1.5\%&10.0\%&1.0\%\\
+	&Rosenbrock&74.2\%&7.7\%&2.2\%&14.0\%&1.8\%\\
+	\cline{2-7}
+	&Average&{73.0\%}&{9.0\%}&{2.2\%}&{13.6\%}&
+	{2.2\%}\\
+	\hline
+	\hline
+	Complete&Sphere&31.9\%&9.2\%&0.2\%&45.1\%&13.5\%\\
+	&Griewank&35.3\%&8.4\%&0.2\%&44.1\%&11.9\%\\
+	&Rastrigin&47.7\%&6.7\%&0.2\%&38.2\%&7.0\%\\
+	&Rosenbrock&35.3\%&3.4\%&0.3\%&54.4\%&6.6\%\\
+	\cline{2-7}
+	&Average&{37.6\%}&{6.9\%}&{.2\%}&{45.5\%}&
+	{9.8\%}\\
+	\hline
+	\multicolumn{7}{l}{The numbers in the column headings refer to branch
+	numbers in Table~\ref{tab:branches}.}
+  \end{tabular}
+\end{table}
+
+Row 1 in Table~\ref{tab:branches} corresponds to the phenomenon commonly called
+stagnation in the literature, but only with a single particle.  As an
+interesting aside, we found that complete stagnation practically never occurs
+in PSO, though frequently large percentages of the swarm are stagnant.
+Iterations where the entire swarm is stagnant average less than 2 in 1000
+iterations, unless the swarm converges beyond machine precision.  When graphs
+of the best function value flat-line, what really has happened is a contraction
+of the particles' velocity, not stagnation in the technical sense.
+
+We now show with what percentage a particle takes each of these branches for
+three different topologies and four different functions.  All of our statistics
+are from swarms of 240 particles.  Brief experimentation showed that other
+swarm sizes had similar statistics.  We ran 750 iterations on all combinations
+of functions and topologies except for the functions Griewank and Rastrigin
+with the Complete topology.  We found that those runs frequently converged past
+machine precision after 500 iterations, and that led to erroneously high values
+for the probability of stagnation.  Instead we ran for only 450 iterations on
+those two combinations.  All of our results were averaged over 20 runs of the
+algorithm; thus the probabilities presented are the averages of 3.6 million
+trials for the branch taken (2.16 million for the two with only 450
+iterations).  Table~\ref{tab:stats} contains the results, with the branch
+number coming from Table~\ref{tab:branches}.  We have already defined the
+functions Sphere, Griewank, and Rastrigin.  Rosenbrock is defined as
+$f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.
+
+The probabilities presented in Table~\ref{tab:stats} are interesting in and of
+themselves and could probably be used to better understand the characteristics
+of various topologies.  It is notable that there is small variation between
+functions in any given topology, but the variation across topologies is far
+greater.  However, our concern is with speculative evaluation.  We are
+interested in predicting the branch that any given particle will take at a
+particular iteration.
+
+From Table~\ref{tab:stats} we can see that with the Random topology, we can
+pick the first branch, corresponding to stagnation, and be right around 70\% of
+the time.  With the Ring topology, we would be right 50\% of the time.
+Branches four and five really correspond to several actual branches, as all of
+the neighbors are lumped together.  The 20\% probability of taking branch four
+with the Ring topology can be split into two branches, as there are only two
+neighbors.  It also turns out that the neighbor that last updated the
+neighborhood best is the most likely to update it next time, so keeping track
+of that information could be fruitful in trying to predict that branch.
+
+The statistics for the Complete topology are less promising, as there are 240
+neighbors that branch four splits into, instead of two.  Pruning does, however,
+allow for the possibility of using the Complete topology in speculative
+evaluation while avoiding the explosion in the number of evaluations it would
+otherwise entail.  Though pruning makes its use possible, we still found the
+communication costs in our parallel environment prohibitive, so we do not
+present any results for the Complete topology here.
+
+With the other topologies, however, there are a few branches with very high
+probabilities, so we can have hope that cutting out some of the evaluations
+that have low probability will lead to an increase in overall performance.  In
+order to implement this kind of pruning, the only change that needs to be made
+to \alg{centralized} is in Step 2, where the speculative children are
+generated.  Simply generate a subset of the children that would normally be
+generated.
+
+\subsubsection{What if we guessed wrong?}
+
+When not all of the branches are evaluated, there is some probability of not
+evaluating the branch that was actually taken by the original particle.  We 
+need to decide what to do in this case.
+
+One possibility is to leave the particle as it is, not accepting any of the
+speculative evaluations, because none of them were correct.  This leads to
+particles being at different iterations, as some particles guess correctly
+while others do not.  Thus we lose exact compatibility with the original PSO,
+though this particular relaxation is nothing new; asynchronous adaptations of
+PSO do the same thing.  As an aside, it is equivalent in this case to simply
+increment the iteration number of particles which fail to correctly predict
+their branch.  This keeps the iteration number constant across all particles,
+simplifying the work needed to be done in determining neighbors when dynamic
+topologies are involved.
+
+Another possibility is to pick the best child, as described in \sec{pickbest}.
+This ignores the fact that the branch might have been wrong; we do not care,
+because we just want the child that had the best value, anyway.  In almost all
+of the experiments that we ran, it turned out that picking the best child
+performed better than keeping the particle back an iteration.
+
+There are a large number of ways to prune speculative evaluations.  We
+experimented with several, but present results only for one possible pruning.
+The pruning we present uses only the two most probable branches in the Random
+topology, as seen in~\ref{tab:stats}, the two branches where the $\nbest$ value
+was not updated.  Those branches are also convenient in that no messages are
+needed from neighbors in order to produce the positions of the speculative
+particles---in distributed frameworks using several rounds of communication
+(see \sec{distributed}), one of the rounds of communication can be dropped
+entirely.
+
+Because we are now only doing two speculative evaluations per particle, with
+240 processors we can have a swarm of size 80, instead of the 30 particles in
+previous experiments.
+
+In \fig{sphere-pruned}, we show the results of this simple pruning on the
+function Sphere.  As can be seen, the larger number of particles that pruning
+allows results in better results when picking the best speculative child.  
+When rejecting all speculative children, which we with tongue in cheek call
+Social Promotion, no real improvement is seen.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere3.eps}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing pruning with previous results.}
+  \label{fig:sphere-pruned}
+\end{figure}
+
+When looking at the function Griewank, presented in \fig{griewank-pruned}, we
+see different results.  Our intuition about picking the best child is more
+correct in this case; while adding 50 more particles to the swarm, Pick Best
+with pruning still gets stuck more often than the original Pick Best.  However,
+Social Promotion does well in this case, finding the optimum every time and
+faster than the original PSO, without the need to increase the number of
+processors.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank5.eps}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations, comparing pruning with previous results.}
+  \label{fig:griewank-pruned}
+\end{figure}
+
+\subsection{More than one iteration ahead}
+
+We need not simply produce speculative children for the next iteration.  We can
+view all possible speculative evaluations for a particle as an infinite tree
+with branching factor $2n+1$.  As we have already seen that doing one full
+level of the tree produces too many extra evaluations to be profitable, it is
+incredibly unlikely that doing two full iterations would produce decent
+results.  But, if the idea of speculating more than one iteration ahead is
+combined with wisely pruning the possible evaluations based on branch
+statistics, we can use just a few extra evaluations to go two or more
+iterations ahead on the most likely branches.
+
+When speculating more than one iteration ahead, the idea of picking the child
+that matches the correct branch, and holding back the particle if no child
+matches the branch, does not make very much sense, as correct branches can only
+be determined for the first iteration.  Thus when speculating more than one
+iteration ahead we always pick the child that has the highest value.
+
+The question of which branches to take in this infinite tree is an intriguing
+one that we can only begin to explore here.  If the branch corresponding to
+stagnation has a 75\% chance of being taken, as in the Random topology with
+most functions, we could speculate three iterations ahead on that branch and
+still have a 42\% chance of predicting correctly.  However, intuition would say
+that perhaps it is better to hope that the particle is productive instead of
+stagnant, so a branch where the particle updates its personal best might be
+more fruitful to try.  In our experiments we try just one of the countless
+possibilities, but that one turns out to work extremely well.  More work is
+needed to compare the different branching possibilities on various functions.
+
+The combination of branches we tried uses seven speculative evaluations per
+particle, matching the swarm size of the original speculative algorithm.  The
+seven evaluations we used corresponded to several iterations of branches 1 and
+2 in Table~\ref{tab:branches}.  Two of the evaluations were just one iteration
+ahead, four were two iterations ahead, and one was three iterations ahead.  The
+evaluations that were one iteration ahead were branches 1 and 2; those that
+went two iterations ahead were formed by taking either branch 1 or 2 and then
+branch 1 or 2; and the evaluation going three iterations ahead followed branch
+1 on all three iterations.
+
+The results when optimizing the function Sphere, presented in
+\fig{sphere-manyiters}, are rather astounding.  When 240 processors are
+available for use, performing speculative evaluation several iterations ahead
+finds values an average of 46 orders of magnitude better than using the
+processors to perform the standard PSO algorithm with 240 particles.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere4.eps}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing speculating several iterations ahead with previous
+  results.}
+  \label{fig:sphere-manyiters}
+\end{figure}
+
+With the function Griewank, the premature convergence problems inherent in
+picking the best child are exacerbated when speculating several iterations
+ahead.  When the algorithm converges, it converges quicker than any other 
+method we tried.  However, it also got stuck and failed to find the optimum
+more than any other method.  In order for this method to be consistently useful
+on functions like Griewank, it seems that more processors are required.  The
+results are shown in \fig{griewank-manyiters}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank6.eps}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations, comparing speculating several iterations ahead with previous
+  results.}
+  \label{fig:griewank-manyiters}
+\end{figure}
+
+\subsection{Tokens}
+
+One issue remains.  What if we really want a swarm of 50 particles, but we have
+230 processors to use?  The methods we have thus far presented do the same
+thing for every particle, and thus would require some multiple of 50
+processors---we would have 30 extra processors with nothing to do.  It might
+very well be the best idea to simply adjust the swarm size slightly, but we
+present another method that allows particles to behave differently, some doing
+more evaluations than others.
+
+If we use the concept of tokens commonly used in multirobot stuff (see, for
+example, \dots), we can specify a number of tokens available for particles to
+use, where each token corresponds to a speculative evaluation.  Whichever
+method is used to decide which evaluations to do, they are prioritized so that
+particles with only one or two tokens perform the most valuable speculative
+evaluations.  Particles that perform well keep their tokens, and particles that
+perform poorly send one token (if they have one) to another random particle in
+the swarm.
+
+We choose to define ``perform poorly'' as a failure to update the personal best
+when the original particle's position is evaluated.  Clearly there are other
+possible methods of deciding when to give up tokens, but many of the methods
+that sound the best are difficult in distributed environments, like giving
+tokens to that particles that most recently found the globally best positions.
+The use of a central machine to decide which particles get tokens would help to
+facilitate such methods.
+
+There are thus two distinct uses for tokens.  First, they could be used to try
+to evenly spread extra evaluations due to an incapatible number of particles
+and processors.  In cursory experimentation, we found that this was not a very
+effective use of tokens.  In virtually every case we tried, it was better to
+simply change the number of particles to match the number of processors, and 
+spread the speculative evaluations evenly throughout the swarm.
+
+The other, more compelling reason to use tokens is to intelligently and
+dynamically allocate speculative evaluations to particles that are performing
+well.  This really only makes sense when speculating several iterations ahead;
+after one or two speculative evaluations, the added benefit from also
+evaluating the low-probability branches is negligible.  But by intelligently
+allocating speculative evaluations, particles that are going in good directions
+move faster by speculating farther with extra tokens, while particles that are
+not doing as well lose their tokens.  Those particles likely contribute to the
+exploration of the algorithm, and when they find good locations, they start
+keeping their tokens, switching roles to become productive particles
+themselves.  This approach is very similar in concept to other dynamic swarms,
+like the TRIBES method proposed by Clerc (cite).
+
+As in the previous section, we can only offer a very preliminary exploration of
+the possibilities of using tokens to improve the allocation of speculative
+evaluations to particles.  We used a distributed parallel PSO framework, and in
+this setting determining which particles are performing the best globally is
+difficult without extensive communication.  Thus we only present results for
+the one method of token allocation described above.
+
+The only function for which we saw a benefit when using this technique was
+Griewank.  As seen in \fig{griewank-token}, using tokens increased the chance
+of finding the optimum to 100\%, while still using far fewer iterations than
+standard PSO.  We assume that the relatively poor performance of the technique
+on other functions is due to the lack of better token allocation.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank7.eps}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations, comparing token-based allocation with previous results.}
+  \label{fig:griewank-token}
+\end{figure}
+
+\section{Conclusions}
+\label{sec:conclusion}
+
+We have described a new technique for using processors in parallel PSO to
+improve the performance of the algorithm.  To our knowledge, this is the first
+time extra processors have been used to do anything in PSO besides increase the
+swarm size.  In an increasingly parallel world, such advancements will prove to
+be crucial in the continued effectiveness of PSO.
+
+We have detailed how to perform speculative evaluation in PSO in several
+different parallel frameworks.  Using this methodology, the behavior of the
+original PSO algorithm can either be exactly reproduced, two iterations at a
+time, or the behavior can be modified in order to improve performance.  We have
+shown results that conclusively demonstrate the superiority of our techniques
+over naively adding particles to the swarm when extra processors are available.
+
+We have shown several different ways to modify the behavior of PSO when doing
+speculative evaluation, and, as expected given the No Free Lunch Theorem,
+different modifications show varied performance with different functions.
+While our exploration of these techniques has been cursory in places, we
+believe the results shown are sufficient to demonstrate their promise.
+
+\section{Future Work}
+\label{sec:future}
+
+This work represents the first attempt of which we are aware at using
+additional processors in parallel with PSO to do something other than either
+add particles to the swarm or parallelize the evaluation of the objective
+function.  We anticipate that many future advancements can be made.
+
+We have speculated that our ``pick the best child'' method simply draws a
+number of samples from the next iteration of the sampling distribution of PSO
+and keeps one of them.  It would then be expected that the sampling
+distribution of that method is very similar to the distribution of the original
+PSO.  This remains to be proven, or even demonstrated empirically.
+
+We gave a passing glance to some very interesting statistical characteristics
+of PSO.  Our motivation was to use the statistics to improve our speculative
+algorithm, but it seems there are some nuggets in the numbers that could be
+used to improve our understanding of the behavior of PSO in general, and what
+role a topology plays.  A potentially very fruitful avenue of future work would
+be to further analyze the branch statistics in PSO across many functions and
+topologies, categorizing both the topologies and the functions and discovering
+what it is that makes certain topologies work well with certain functions.
+There are also some particles that seem to drive the swarm and others that are
+stagnant almost the entire time.  Such information could be used in a
+TRIBES-like fashion to determine which particles should be removed from the
+swarm, perhaps starting them close to well-performing particles.
+
+We turned what used to be a very simple algorithm with only one major parameter
+to tweak (the topology) into a big hairy mess with countless possibilities for
+speculative evaluations.  Perhaps future research will be able to narrow down
+the endless possibilities into the few best ones, making the algorithm simple
+again.
+
+We gave only a very little attention to the possibilities of token-based
+allocation of speculative evaluations to particles.  We imagine that the
+benefits of intelligently and dynamically deciding which particles get extra
+evaluations could produce very good results.  Future work could explore the use
+of a centralized allocator for extra processors or ideas in the multi-robot
+task allocation literature, such as auctions, to better assign processors to
+particles for speculative evaluations.  While such methods involve extra
+communication, with long-running function evaluations the extra overhead of
+intelligently allocating resources would likely be negligible.
+
+We have mentioned that this technique has been previously used to parallelize
+simulated annealing.  Any other optimization algorithm that only depends on 
+current sampling positions when computing the next position to sample can be
+parallelized with this technique.  In particular, genetic algorithms produce
+future generations by combining individuals from the current generation.  With
+a large population size there would be an unwieldy amount of possible future
+individuals, but the potential exists to modify the algorithm to use some kind
+of speculative evaluation.  A related but quite distinct technique for reducing
+computation in evolutionary algorithms was proposed in~\cite{poli-ai06}.
+
+Large parallel clusters are often required to successfully optimize practical
+modern problems.  To properly use PSO with such clusters, a balance needs to be
+made between using processors to increase the swarm size and using them to
+increase the speed of the algorithm, or for perhaps as yet unconsidered
+techniques.  This work is a first step in that direction.
+
+\appendix[Alternate form of message passing]
+
+Here we describe a method that requires only one round of communication for
+each pair of iterations, which happens at step 5 of \alg{centralized}.  Many
+more messages are needed, but that is sometimes more desirable than
+synchronizing all of the machines three times.
 
@@
 
+One more issue arises when dealing with dynamic topologies.  With neighbors
+changing each iteration, messages that processors pass to their neighbors need
+to be sent to the correct neighbors for each iteration.  In centralized
+algorithms and in distributed algorithms with several rounds of message passing
+(the first method described above), nothing needs to be changed.  However, in
+the case where processors reconstruct information about neighboring particles
+from messages, those messages need to be sent to the correct neighbors.  A
+particle cannot simply send messages to its neighbors' neighbors'
+neighbors---it needs to send messages to its iteration $t$ neighbors' iteration
+$t+1$ neighbors, and so on.  For every neighbor outward information is sent,
+the iteration also needs to be incremented, as information about neighbors'
+neighbors is used during iteration $t+1$, and information about neighbors'
+neighbors' neighbors is used to reconstruct information about iteration $t+2$.
+Also, this method of message passing again requires the use of random seeds if
+the topology is random, so that each processor computes the same neighbors for
+a particle as all other processors.
+
 This may seem like an inordinate amount of work, and with some distributed PSO
@@
 performed in those circumstances.
-
-\subsection{Dynamic Topologies}
-
-Performing speculative evaluation in PSO with a dynamic topology raises a few
-issues of its own.  First, with neighbors changing each iteration, messages
-that processors pass to their neighbors need to be sent to the correct
-neighbors for each iteration.  In centralized algorithms and in distributed
-algorithms with several rounds of message passing (the first method described
-above), nothing needs to be changed.  However, in the case where processors
-reconstruct information about neighboring particles from messages, those
-messages need to be sent to the correct neighbors.  A particle cannot simply
-send messages to its neighbors' neighbors' neighbors---it needs to send
-messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
-For every neighbor outward information is sent, the iteration also needs to be
-incremented, as information about neighbors' neighbors is used during iteration
-$t+1$, and information about neighbors' neighbors' neighbors is used to
-reconstruct information about iteration $t+2$.  Also, this method of message
-passing again requires the use of random seeds if the topology is random, so
-that each processor computes the same neighbors for a particle as all other
-processors.
-
-Also, an issue common to all methods of performing speculative evaluation is
-that of getting available information from neighbors before generating
-speculative children.  In a static topology, at iteration $t$ a particle
-already has all of the information about the positions of its neighbors during
-iterations $1$ through $t-1$.  If the neighbor finds a better position at 
-iteration $t$, the particle updates its neighborhood best, but if it does not,
-it still has its old neighborhood best from its neighbors for all previous 
-iterations.
-
-In a dynamic topology, a particle might not have information about the previous
-positions of its neighbors at iteration $t$.  That means that its new
-neighborhood best could come not only from its neighbors' positions at
-iteration $t$, but also from their personal best from iteration $t-1$, as
-neighbors' personal bests are what are used to update a particle's neighborhood
-best.  That creates a problem for speculative evaluation---there are
-potentially more than $2n+1$ possible next positions, increasing the amount of
-work that must be done to perform the second iteration at the same time as the
-first.
-
-This is easily fixed by updating $\noeval{\p}_{t+2}$ with the currently
-available information about $\noeval{\nset}_{t+2}$ before producing
-$\noeval{\sset}_{t+3}$.  If a particle $\noeval{\p}_{t+2}$ updates its
-neighborhood best with the personal bests of $\noeval{\nset}_{t+2}$ before
-calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
-still only $2n+1$ possible next positions, and the problem is avoided.
-
-\subsection{Comparison with Standard PSO}
-\label{sec:results}
-
-We ran experiments to compare our speculative PSO algorithm to the standard
-PSO algorithm.  At each iteration of the algorithms, we use one processor to
-perform one function evaluation for one particle, be it speculative or
-non-speculative.  The speculative algorithm actually performs two iterations of
-PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
-evaluations.  For benchmark functions with very fast evaluations this may not
-be the best use of parallelism in PSO.  But for many real-world applications,
-the objective function takes on the order of minutes or more to evaluate; in
-such cases our assumption is perfectly reasonable.
-
-To use the same number of processors for each algorithm, speculative evaluation
-needs a smaller swarm size than standard PSO.  For each of the topologies we
-used, a particle has two neighbors other than itself.  As shown in
-Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
-particle.  With one evaluation needed for the original, non-speculative
-particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
-the number of particles in the speculative swarm.  The extra evaluations done
-in speculative evaluation would instead be particles in regular PSO, so we
-compare swarms of size $p_s$ in speculative evaluation with swarms of size
-$8p_s$ in regular PSO.
-
-Where the Complete topology would normally be used, we use a Random topology in
-our speculative algorithm, as Complete leads to an explosion of speculative
-evaluations.  If speculative evaluation were not being performed, it is
-possible that the Complete topology would be used.  However, the Complete
-topology also requires a very large amount of interprocessor communication in
-parallel PSO, so it is still quite possible that Random would be used even with
-standard PSO.  But, to be fair in our comparisons, we compare to the standard
-PSO algorithm using both the Random topology with two neighbors and the
-Complete topology.  All functions have 20 dimensions, and all results shown are
-averages over 20 runs.  We use 240 processors in our experiments, so our
-speculative algorithm uses 30 particles and standard PSO uses 240 particles.
-For ease of labeling, we call our speculative algorithm SEPSO and standard PSO
-just PSO, with the topology following.
-
-First we look at the function Sphere, the simplest of common benchmark
-functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
-has no local optima and is best optimized in terms of function evaluations with
-a small swarm using a Complete topology.  Accordingly, we show results for a
-Complete swarm in \fig{basic-sphere}.  In comparing SEPSO with PSO, we can see
-that when both are using a Random topology, SEPSO clearly outperforms PSO.
-However, the comparison between SEPSO and PSO Complete is not as
-encouraging---PSO Complete performs slightly better.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere.eps}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
-  two iterations per round, while standard PSO algorithms use 240 processors
-  and do one iteration per round.}
-  \label{fig:basic-sphere}
-\end{figure}
-
-Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
-\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
-\right) + 1$.  It is generally recommended to use the Ring topology when
-optimizing the Griewank function, as Complete is prone to premature convergence
-on a local optima.  We show results in \fig{basic-griewank1} for swarms of size
-30 and 240 using the Ring topology. One can see in the figure that using
-speculative evaluation initially outperforms PSO, but gets stuck in a local
-optima and is eventually outperformed.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank1.eps}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.}
-  \label{fig:basic-griewank1}
-\end{figure}
-
-\fig{basic-griewank1} is slightly deceptive, however, as with Griewank some of
-the runs found the global optimum, 0, while some did not.  A more accurate plot
-for this function is one showing what percentage of the runs have found the
-global optimum at each iteration.  \fig{basic-griewank2} shows such a plot.  A
-swarm of size 30 performing the PSO algorithm gets stuck a little less than
-half of the time, while 240 particles provides enough exploration to find the
-global optimum every time.  As would be expected, when our speculative
-algorithm finds the global optimum, it finds it about twice as fast as regular
-PSO---it is just performing regular PSO twice as fast with a smaller swarm
-size.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank2.eps}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.  Instead of showing average function value, we show the
-  percentage of runs that have found the global optimum by each iteration.}
-  \label{fig:basic-griewank2}
-\end{figure}
-
-Because our speculative algorithm got stuck with such a small swarm size, we
-ran a set of experiments on Griewank with 800 processors, giving swarms of size
-100 and 800 instead of 30 and 240.  \fig{basic-griewank3} shows the results.
-\fig{basic-griewank3} is very similar to \fig{basic-griewank2}, except that
-speculative evaluation finds the global optimum essentially 100\% of the time.
-Thus with a few more processors, our algorithm finds the optimum twice as fast
-as the original PSO without a significant loss of accuracy.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank3.eps}
-  \caption{Function Griewank with a swarm that uses 800 processors per round of
-  evaluations.}
-  \label{fig:basic-griewank3}
-\end{figure}
-
-Lastly we look at a function that does not show any kind of improvement with
-this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
-\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
-with Rastrigin, the more particles there are in the swarm, the lower function
-value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
-swarms get caught in local optima.  Because our speculative algorithm has a
-significantly smaller swarm size, it gets stuck at higher values while the
-larger swarms performing regular PSO continue to improve the best value found.
-\fig{rastrigin} shows the results graphically.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
-  \caption{Function Rastrigin with a swarm that uses 240 processors per round
-  of evaluations.}
-  \label{fig:rastrigin}
-\end{figure}
-
-\section{Relaxing the Requirements}
-\label{sec:relax}
-
-While the idea of speculative evaluation in particle swarm optimization is
-intriguing, the results of the previous section show that in many cases it
-requires too many extra evaluations to produce competitive results.  On most
-functions, the larger swarm size that can be used in regular PSO leads to
-better performance than is found by trying to do speculative evaluation.
-However, if we keep the idea of speculative evaluation while relaxing the
-requirement of exactly reproducing the behavior of the original PSO algorithm,
-we see some very impressive results.
-
-\subsection{Pick the Best Child}
-
-In performing speculative evaluation as we have described it, $2n+1$
-speculative evaluations are done per particle, while all but one of them are
-completely ignored.  We can try to make use of those evaluations instead of
-throwing them away.
-
-To use the extra speculative evaluations, instead of choosing the speculative
-child that matches that branch that the original PSO would have taken, we take
-the child that has the best value.  The methodology is exactly the same as
-above except for the process of choosing which speculative child to accept.
-The branch that PSO would have taken is ignored, but $\p_t$ must still be
-created so it can give the right personal best and neighborhood best values and
-positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
-step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
-$\noeval{\sset}_{t+1}$ instead of with the matching branch.
-
-This can be thought of as drawing a number of samples from the next iteration
-and accepting the best one.  Speculative particles that move in good directions
-are kept.  It seems this modification to PSO would produce more exploitation,
-decreasing the amount of exploration.  With functions that are less deceptive
-we would expect this method to work well, while performance on more deceptive
-functions would probably be hurt.
-
-We show in \fig{sphere-pickbest} the results of using this modification on the
-function Sphere.  For ease of labeling, we call the modified algorithm SEPSO
-Pick Best.  As can be seen, where SEPSO did not manage to perform as well as
-PSO Complete, SEPSO Pick Best handily outperforms it with the same number of
-processors.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere2.eps}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing SEPSO Pick Best with previous results.}
-  \label{fig:sphere-pickbest}
-\end{figure}
-
-Sphere is a function for which very little exploration needs to be done, so
-always keeping the best speculative position found makes intuitive sense.  A
-less intuitive result occurs when using SEPSO Pick Best with the function
-Griewank.  Griewank is a highly deceptive function prone to premature
-convergence.  Our intuition on the behavior of picking the best speculative
-child is that it would tend to prematurely converge more often than the
-original PSO.  However, as seen in \fig{griewank-pickbest}, SEPSO Pick Best
-increases the chance of finding the optimum while also reducing the number of
-iterations required to converge.  Perhaps SEPSO Pick Best encourages local
-exploitation, while the sparse communication of the Ring topology still allows
-for global exploration.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank4.eps}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing SEPSO Pick Best with previous results.}
-  \label{fig:griewank-pickbest}
-\end{figure}
-
-\subsection{Pruning the Speculative Evaluations}
-
-Because the problem facing speculative evaluation is having too many possible
-evaluations to perform, a natural step to take is to eliminate some of the
-evaluations.  If we could reliably predict which branch were going to be taken,
-we could limit ourselves to one speculative evaluation per particle instead of
-$2n+1$.  While we cannot with certainty predict the correct branch every time,
-we can look at the statistics of the behavior of particles in general and
-perhaps come up with a good guess, reducing the number of necessary speculative
-evaluations.
-
-\subsubsection{Branch Statistics}
-
-In Table~\ref{tab:evals} we presented all possible branches that a particle
-with two neighbors could take.  Here we lump all of the neighbors together and
-consider the statistics for the five branches shown in Table~\ref{tab:branches}.
-
-\begin{table}
-  \caption{Five Branches to Consider for Statistics}
-  \label{tab:branches}
-  \centering
-  \begin{tabular}{rc|c}
-	&$\pbest$ updated&Source of $\nbest$ update\\
-	\hline
-	1.&No update&No update\\
-	2.&Updated&No update\\
-	3.&Updated&The particle itself\\
-	4.&No update&Some neighbor\\
-	5.&Updated&Some neighbor\\
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \caption{Branch Statistics in PSO}
-  \label{tab:stats}
-  \centering
-  \begin{tabular}{c|c|c|c|c|c|c}
-	Topology&Function&1&2&3&4&5\\
-	\hline
-	Ring&Sphere&0.530&0.093&0.114&0.202&0.062\\
-	&Griewank&0.517&0.084&0.122&0.207&0.070\\
-	&Rastrigin&0.495&0.048&0.146&0.213&0.099\\
-	&Rosenbrock&0.513&0.074&0.129&0.211&0.073\\
-	\hline
-	Random&Sphere&0.667&0.119&0.026&0.156&0.031\\
-	&Griewank&0.690&0.109&0.025&0.149&0.027\\
-	&Rastrigin&0.819&0.055&0.015&0.100&0.010\\
-	&Rosenbrock&0.742&0.077&0.022&0.140&0.018\\
-	\hline
-	Complete&Sphere&0.319&0.092&0.002&0.451&0.135\\
-	&Griewank&0.353&0.084&0.002&0.441&0.119\\
-	&Rastrigin&0.477&0.067&0.002&0.382&0.070\\
-	&Rosenbrock&0.353&0.034&0.003&0.544&0.066\\
-  \end{tabular}
-\end{table}
-
-Row 1 in Table~\ref{tab:branches} corresponds to the phenomenon commonly called
-stagnation in the literature, but only with a single particle.  As an
-interesting aside, we found that complete stagnation practically never occurs
-in PSO, though frequently large percentages of the swarm are stagnant.
-Iterations where the entire swarm is stagnant average less than 2 in 1000
-iterations, unless the swarm converges beyond machine precision.  When graphs
-of the best function value flat-line, what really has happened is a contraction
-of the particles' velocity, not stagnation in the technical sense.
-
-We now show with what percentage a particle takes each of these branches for
-three different topologies and four different functions.  All of our statistics
-are from swarms of 240 particles.  Brief experimentation showed that other
-swarm sizes had similar statistics.  We ran 750 iterations on all combinations
-of functions and topologies except for the functions Griewank and Rastrigin
-with the Complete topology.  We found that those runs frequently converged past
-machine precision after 500 iterations, and that led to erroneously high values
-for the probability of stagnation.  Instead we ran for only 450 iterations on
-those two combinations.  All of our results were averaged over 20 runs of the
-algorithm; thus the probabilities presented are the averages of 3.6 million
-trials for the branch taken (2.16 million for the two with only 450
-iterations).  Table~\ref{tab:stats} contains the results, with the branch
-number coming from Table~\ref{tab:branches}.  We have already defined the
-functions Sphere, Griewank, and Rastrigin.  Rosenbrock is defined as
-$f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.
-
-The probabilities presented in Table~\ref{tab:stats} are interesting in and of
-themselves and could probably be used to better understand the characteristics
-of various topologies.  It is notable that there is small variation between
-functions in any given topology, but the variation across topologies is far
-greater.  However, our concern is with speculative evaluation.  We are
-interested in predicting the branch that any given particle will take at a
-particular iteration.
-
-From Table~\ref{tab:stats} we can see that with the Random topology, we can
-pick the first branch, corresponding to stagnation, and be right around 70\% of
-the time.  With the Ring topology, we would be right 50\% of the time.
-Branches four and five really correspond to several actual branches, as all of
-the neighbors are lumped together.  The 20\% probability of taking branch four
-with the Ring topology can be split into two branches, as there are only two
-neighbors.  It also turns out that the neighbor that last updated the
-neighborhood best is the most likely to update it next time, so keeping track
-of that information could be fruitful in trying to predict that branch.
-
-The statistics for the Complete topology are less promising, as there are 240
-neighbors that branch four splits into, instead of two.  Pruning does, however,
-allow for the possibility of using the Complete topology in speculative
-evaluation while avoiding the explosion in the number of evaluations it would
-otherwise entail.  Though pruning makes its use possible, we still found the
-communication costs in our parallel environment prohibitive, so we do not
-present any results for the Complete topology here.
-
-With the other topologies, however, there are a few branches with very high
-probabilities, so we can have hope that cutting out some of the evaluations
-that have low probability will lead to an increase in overall performance.  In
-order to implement this kind of pruning, the only change that needs to be made
-to \alg{centralized} is in Step 2, where the speculative children are
-generated.  Simply generate a subset of the children that would normally be
-generated.
-
-\subsubsection{What if we guessed wrong?}
-
-When not all of the branches are evaluated, there is some probability of not
-evaluating the branch that was actually taken by the original particle.  We 
-need to decide what to do in this case.
-
-One possibility is to leave the particle as it is, not accepting any of the
-speculative evaluations, because none of them were correct.  This leads to
-particles being at different iterations, as some particles guess correctly
-while others do not.  Thus we lose exact compatibility with the original PSO,
-though this particular relaxation is nothing new; asynchronous adaptations of
-PSO do the same thing.  As an aside, it is equivalent in this case to simply
-increment the iteration number of particles which fail to correctly predict
-their branch.  This keeps the iteration number constant across all particles,
-simplifying the work needed to be done in determining neighbors when dynamic
-topologies are involved.
-
-Another possibility is to pick the best child, as described above.  This
-ignores the fact that the branch might have been wrong; we do not care, because
-we just want the child that had the best value, anyway.  In almost all of the
-experiments that we ran, it turned out that picking the best child performed
-better than keeping the particle back an iteration.
-
-There are a large number of ways to prune speculative evaluations.  We
-experimented with several, but present results only for one possible pruning.
-The pruning we present uses only the two most probable branches in the Random
-topology, as seen in~\ref{tab:stats}, the two branches where the $\nbest$ value
-was not updated.  Those branches are also convenient in that no messages are
-needed from neighbors in order to produce the positions of the speculative
-particles---in distributed frameworks using several rounds of communication
-(see \sec{distributed}), one of the rounds of communication can be dropped
-entirely.
-
-Because we are now only doing two speculative evaluations per particle, with
-240 processors we can have a swarm of size 80, instead of the 30 particles in
-previous experiments.
-
-In \fig{sphere-pruned}, we show the results of this simple pruning on the
-function Sphere.  As can be seen, the larger number of particles that pruning
-allows results in better results when picking the best speculative child.  
-When rejecting all speculative children, which we with tongue in cheek call
-Social Promotion, no real improvement is seen.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere3.eps}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing pruning with previous results.}
-  \label{fig:sphere-pruned}
-\end{figure}
-
-When looking at the function Griewank, presented in \fig{griewank-pruned}, we
-see different results.  Our intuition about picking the best child is more
-correct in this case; while adding 50 more particles to the swarm, Pick Best
-with pruning still gets stuck more often than the original Pick Best.  However,
-Social Promotion does well in this case, finding the optimum every time and
-faster than the original PSO, without the need to increase the number of
-processors.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank5.eps}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing pruning with previous results.}
-  \label{fig:griewank-pruned}
-\end{figure}
-
-\subsection{More than one iteration ahead}
-
-We need not simply produce speculative children for the next iteration.  We can
-view all possible speculative evaluations for a particle as an infinite tree
-with branching factor $2n+1$.  As we have already seen that doing one full
-level of the tree produces too many extra evaluations to be profitable, it is
-incredibly unlikely that doing two full iterations would produce decent
-results.  But, if the idea of speculating more than one iteration ahead is
-combined with wisely pruning the possible evaluations based on branch
-statistics, we can use just a few extra evaluations to go two or more
-iterations ahead on the most likely branches.
-
-When speculating more than one iteration ahead, the idea of picking the child
-that matches the correct branch, and holding back the particle if no child
-matches the branch, does not make very much sense, as correct branches can only
-be determined for the first iteration.  Thus when speculating more than one
-iteration ahead we always pick the child that has the highest value.
-
-The question of which branches to take in this infinite tree is an intriguing
-one that we can only begin to explore here.  If the branch corresponding to
-stagnation has a 75\% chance of being taken, as in the Random topology with
-most functions, we could speculate three iterations ahead on that branch and
-still have a 42\% chance of predicting correctly.  However, intuition would say
-that perhaps it is better to hope that the particle is productive instead of
-stagnant, so a branch where the particle updates its personal best might be
-more fruitful to try.  In our experiments we try just one of the countless
-possibilities, but that one turns out to work extremely well.  More work is
-needed to compare the different branching possibilities on various functions.
-
-The combination of branches we tried uses seven speculative evaluations per
-particle, matching the swarm size of the original speculative algorithm.  The
-seven evaluations we used corresponded to several iterations of branches 1 and
-2 in Table~\ref{tab:branches}.  Two of the evaluations were just one iteration
-ahead, four were two iterations ahead, and one was three iterations ahead.  The
-evaluations that were one iteration ahead were branches 1 and 2; those that
-went two iterations ahead were formed by taking either branch 1 or 2 and then
-branch 1 or 2; and the evaluation going three iterations ahead followed branch
-1 on all three iterations.
-
-The results when optimizing the function Sphere, presented in
-\fig{sphere-manyiters}, are rather astounding.  When 240 processors are
-available for use, performing speculative evaluation several iterations ahead
-finds values an average of 46 orders of magnitude better than using the
-processors to perform the standard PSO algorithm with 240 particles.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere4.eps}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing speculating several iterations ahead with previous
-  results.}
-  \label{fig:sphere-manyiters}
-\end{figure}
-
-With the function Griewank, the premature convergence problems inherent in
-picking the best child are exacerbated when speculating several iterations
-ahead.  When the algorithm converges, it converges quicker than any other 
-method we tried.  However, it also got stuck and failed to find the optimum
-more than any other method.  In order for this method to be consistently useful
-on functions like Griewank, it seems that more processors are required.  The
-results are shown in \fig{griewank-manyiters}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank6.eps}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing speculating several iterations ahead with previous
-  results.}
-  \label{fig:griewank-manyiters}
-\end{figure}
-
-\subsection{Tokens}
-
-One issue remains.  What if we really want a swarm of 50 particles, but we have
-230 processors to use?  The methods we have thus far presented do the same
-thing for every particle, and thus would require some multiple of 50
-processors---we would have 30 extra processors with nothing to do.  It might
-very well be the best idea to simply adjust the swarm size slightly, but we
-present another method that allows particles to behave differently, some doing
-more evaluations than others.
-
-If we use the concept of tokens commonly used in multirobot stuff (see, for
-example, \dots), we can specify a number of tokens available for particles to
-use, where each token corresponds to a speculative evaluation.  Whichever
-method is used to decide which evaluations to do, they are prioritized so that
-particles with only one or two tokens perform the most valuable speculative
-evaluations.  Particles that perform well keep their tokens, and particles that
-perform poorly send one token (if they have one) to another random particle in
-the swarm.
-
-We choose to define ``perform poorly'' as a failure to update the personal best
-when the original particle's position is evaluated.  Clearly there are other
-possible methods of deciding when to give up tokens, but many of the methods
-that sound the best are difficult in distributed environments, like giving
-tokens to that particles that most recently found the globally best positions.
-The use of a central machine to decide which particles get tokens would help to
-facilitate such methods.
-
-There are thus two distinct uses for tokens.  First, they could be used to try
-to evenly spread extra evaluations due to an incapatible number of particles
-and processors.  In cursory experimentation, we found that this was not a very
-effective use of tokens.  In virtually every case we tried, it was better to
-simply change the number of particles to match the number of processors, and 
-spread the speculative evaluations evenly throughout the swarm.
-
-The other, more compelling reason to use tokens is to intelligently and
-dynamically allocate speculative evaluations to particles that are performing
-well.  This really only makes sense when speculating several iterations ahead;
-after one or two speculative evaluations, the added benefit from also
-evaluating the low-probability branches is negligible.  But by intelligently
-allocating speculative evaluations, particles that are going in good directions
-move faster by speculating farther with extra tokens, while particles that are
-not doing as well lose their tokens.  Those particles likely contribute to the
-exploration of the algorithm, and when they find good locations, they start
-keeping their tokens, switching roles to become productive particles
-themselves.  This approach is very similar in concept to other dynamic swarms,
-like the TRIBES method proposed by Clerc (cite).
-
-As in the previous section, we can only offer a very preliminary exploration of
-the possibilities of using tokens to improve the allocation of speculative
-evaluations to particles.  We used a distributed parallel PSO framework, and in
-this setting determining which particles are performing the best globally is
-difficult without extensive communication.  Thus we only present results for
-the one method of token allocation described above.
-
-The only function for which we saw a benefit when using this technique was
-Griewank.  As seen in \fig{griewank-token}, using tokens increased the chance
-of finding the optimum to 100\%, while still using far fewer iterations than
-standard PSO.  We assume that the relatively poor performance of the technique
-on other functions is due to the lack of better token allocation.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank7.eps}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing token-based allocation with previous results.}
-  \label{fig:griewank-token}
-\end{figure}
-
-\section{Conclusions}
-\label{sec:conclusion}
-
-We have described a new technique for using processors in parallel PSO to
-improve the performance of the algorithm.  To our knowledge, this is the first
-time extra processors have been used to do anything in PSO besides increase the
-swarm size.  In an increasingly parallel world, such advancements will prove to
-be crucial in the continued effectiveness of PSO.
-
-We have detailed how to perform speculative evaluation in PSO in several
-different parallel frameworks.  Using this methodology, the behavior of the
-original PSO algorithm can either be exactly reproduced, two iterations at a
-time, or the behavior can be modified in order to improve performance.  We have
-shown results that conclusively demonstrate the superiority of our techniques
-over naively adding particles to the swarm when extra processors are available.
-
-We have shown several different ways to modify the behavior of PSO when doing
-speculative evaluation, and, as expected given the No Free Lunch Theorem,
-different modifications show varied performance with different functions.
-While our exploration of these techniques has been cursory in places, we
-believe the results shown are sufficient to demonstrate their promise.
-
-\section{Future Work}
-\label{sec:future}
-
-This work represents the first attempt of which we are aware at using
-additional processors in parallel with PSO to do something other than either
-add particles to the swarm or parallelize the evaluation of the objective
-function.  We anticipate that many future advancements can be made.
-
-We have speculated that our ``pick the best child'' method simply draws a
-number of samples from the next iteration of the sampling distribution of PSO
-and keeps one of them.  It would then be expected that the sampling
-distribution of that method is very similar to the distribution of the original
-PSO.  This remains to be proven, or even demonstrated empirically.
-
-We gave a passing glance to some very interesting statistical characteristics
-of PSO.  Our motivation was to use the statistics to improve our speculative
-algorithm, but it seems there are some nuggets in the numbers that could be
-used to improve our understanding of the behavior of PSO in general, and what
-role a topology plays.  A potentially very fruitful avenue of future work would
-be to further analyze the branch statistics in PSO across many functions and
-topologies, categorizing both the topologies and the functions and discovering
-what it is that makes certain topologies work well with certain functions.
-There are also some particles that seem to drive the swarm and others that are
-stagnant almost the entire time.  Such information could be used in a
-TRIBES-like fashion to determine which particles should be removed from the
-swarm, perhaps starting them close to well-performing particles.
-
-We turned what used to be a very simple algorithm with only one major parameter
-to tweak (the topology) into a big hairy mess with countless possibilities for
-speculative evaluations.  Perhaps future research will be able to narrow down
-the endless possibilities into the few best ones, making the algorithm simple
-again.
-
-We gave only a very little attention to the possibilities of token-based
-allocation of speculative evaluations to particles.  We imagine that the
-benefits of intelligently and dynamically deciding which particles get extra
-evaluations could produce very good results.  Future work could explore the use
-of a centralized allocator for extra processors or ideas in the multi-robot
-task allocation literature, such as auctions, to better assign processors to
-particles for speculative evaluations.  While such methods involve extra
-communication, with long-running function evaluations the extra overhead of
-intelligently allocating resources would likely be negligible.
-
-We have mentioned that this technique has been previously used to parallelize
-simulated annealing.  Any other optimization algorithm that only depends on 
-current sampling positions when computing the next position to sample can be
-parallelized with this technique.  In particular, genetic algorithms produce
-future generations by combining individuals from the current generation.  With
-a large population size there would be an unwieldy amount of possible future
-individuals, but the potential exists to modify the algorithm to use some kind
-of speculative evaluation.  A related but quite distinct technique for reducing
-computation in evolutionary algorithms was proposed in~\cite{poli-ai06}.
-
-Large parallel clusters are often required to successfully optimize practical
-modern problems.  To properly use PSO with such clusters, a balance needs to be
-made between using processors to increase the swarm size and using them to
-increase the speed of the algorithm, or for perhaps as yet unconsidered
-techniques.  This work is a first step in that direction.
 

### 2010-03-15 honors_thesis/journal_paper/paper.tex

--- 2010-03-14 honors_thesis/journal_paper/paper.tex
@@
 for many functions there comes a point of diminishing returns with respect to adding particles.
->>>>>>>>> TODO WARNING WARNING MADE UP NUMBERS GET REAL ONES!! <<<<<<<<<<<<<<
+!!!! TODO WARNING WARNING MADE UP NUMBERS GET REAL ONES!! !!!!!
 For the well-know benchmark function Sphere and as shown in Figure XXXXa increasing the swarm size from 20 to 120 has a significant effect on the 
@@
 literature are the Ring topology and the Complete
-topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
-as a parallel surrogate for the Complete topology called Random that has been
-shown to simulate the behavior of Complete with far less
-communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
-be defined as having $n$ neighbors; we use 2 neighbors for all of our
-experiments.
-%TODO except as noted??
+topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
+neighbor to either side of it; in the Complete topology, every particle is a
+neighbor to every other particle.  In all topologies, a particle is also a
+neighbor to itself, in that its own position is considered when updating the
+particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
+particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
+topology, each particle has $p$ neighbors.  In this paper we use these
+topologies, as well as a parallel adaptation of the Complete topology, called
+Random, that has been shown to simulate the behavior of Complete with far less
+communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
+randomly picks two other particles to share information with at each iteration,
+along with itself.  Thus in both the Ring and the Random topologies, all
+particles have 3 neighbors.
 

### 2010-03-16 honors_thesis/journal_paper/paper.tex
Fixed some paragraphs, and changed the notation a bit
Made a lot of changes to the journal paper, from comments I got.

--- 2010-03-15 honors_thesis/journal_paper/paper.tex
@@
 
+%Other math
+\providecommand{\prob}{\ensuremath{\mathrm{Pr}}}
+
 \title{\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization%
@@
 longer demanding that it strictly mimic the behavior of the original PSO, we
-can introduce new speculative techniques that often out-perform PSO.
+can introduce new speculative techniques that often out-perform PSO.  Some
+modifications that we explore are picking the speculative position with the
+best value instead of the position that matches the branch that PSO took;
+pruning some of the speculative evaluations to increase the number of particles
+in the swarm; and speculating several iterations ahead instead of just one.
 
@@
 current particle has seen, and $\nbest$ (called neighborhood best) is the best
-position the neighbors of the current particle have seen.  \( \coeff_1 \), \(
-\coeff_2 \), and \( \constriction \) are parameters with prescribed values
-required to ensure convergence (2.05, 2.05, and .73,
+position the neighbors of the current particle have seen~\cite{bratton-sis07}.
+\( \coeff_1 \), \( \coeff_2 \), and \( \constriction \) are parameters with
+prescribed values required to ensure convergence (2.05, 2.05, and .73,
 respectively)~\cite{clerc-tec02}.
@@
 perform speculative evaluation in PSO with only one round of communication per
-two iterations.  However, the methodology is tedius and is not the focus of
+two iterations.  However, the methodology is tedious and is not the focus of
 this paper, so we defer its description to the Appendix.
@@
 
-We ran experiments to compare our speculative PSO algorithm to the standard
-PSO algorithm.  At each iteration of the algorithms, we use one processor to
+We ran experiments to compare our speculative PSO algorithm to the standard PSO
+algorithm.  At each iteration of the algorithms, we use one processor to
 perform one function evaluation for one particle, be it speculative or
-non-speculative.  The speculative algorithm actually performs two iterations of
-PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
-evaluations.  For benchmark functions with very fast evaluations this may not
-be the best use of parallelism in PSO.  But for many real-world applications,
-the objective function takes on the order of minutes or more to evaluate; in
-such cases our assumption is perfectly reasonable.
-
-To use the same number of processors for each algorithm, speculative evaluation
-needs a smaller swarm size than standard PSO.  For each of the topologies we
-used, a particle has two neighbors other than itself.  As shown in
-Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
-particle.  With one evaluation needed for the original, non-speculative
-particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
-the number of particles in the speculative swarm.  The extra evaluations done
-in speculative evaluation would instead be particles in regular PSO, so we
-compare swarms of size $p_s$ in speculative evaluation with swarms of size
-$8p_s$ in regular PSO.
+non-speculative.  For benchmark functions with very fast evaluations this may
+not be the best use of parallelism in PSO.  But for many real-world
+applications, the objective function takes on the order of minutes or more to
+evaluate; in such cases our choice of framework is reasonable.
+
+The speculative algorithm actually performs two iterations of PSO at each
+``iteration,'' so we instead call each ``iteration'' a round of evaluations.
+
+To use the same number of processors (and hence evaluations) for each
+algorithm, speculative evaluation needs a smaller swarm size than standard PSO.
+For each of the topologies we used, a particle has three neighbors including
+itself.  As shown in Table~\ref{tab:evals}, this results in $7$ speculative
+evaluations per particle.  With one evaluation needed for the original,
+non-speculative particle, we have $8p_s$ evaluations for every two iterations,
+where $p_s$ is the number of particles in the speculative swarm.  The extra
+evaluations done in speculative evaluation would instead be particles in
+regular PSO, so we compare swarms of size $p_s$ in speculative evaluation with
+swarms of size $8p_s$ in regular PSO.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{sphere.eps}
+  \includegraphics[width=.8\columnwidth]{sphere}
   \caption{Function Sphere with a swarm that uses 240 processors per round of
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank1.eps}
+  \includegraphics[width=.8\columnwidth]{griewank1}
   \caption{Function Griewank with a swarm that uses 240 processors per round of
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank2.eps}
+  \includegraphics[width=.8\columnwidth]{griewank2}
   \caption{Function Griewank with a swarm that uses 240 processors per round of
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank3.eps}
+  \includegraphics[width=.8\columnwidth]{griewank3}
   \caption{Function Griewank with a swarm that uses 800 processors per round of
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
+  \includegraphics[width=.8\columnwidth]{rastrigin}
   \caption{Function Rastrigin with a swarm that uses 240 processors per round
@@
 we see some very impressive results.
+
+We outline three main improvements to speculative evaluation.  First, in
+\sec{pickbest} we describe a method that uses all of the information found in
+doing speculative evaluations.  Then Sections~\ref{sec:pruning}
+through~\ref{sec:wrong} present a technique that reduces the number of
+speculative evaluations that need to be done for each particle.  Finally,
+\sec{manyiters} shows a method for speculating several iterations ahead,
+instead of just one.  \sec{other} presents a few other ideas that seem
+promising but need further development.
 
@@
 We show in \fig{sphere-pickbest} the results of using this modification on the
-function Sphere.  For ease of labeling, we call the modified algorithm SEPSO
-Pick Best.  As can be seen, where SEPSO did not manage to perform as well as
-PSO Complete, SEPSO Pick Best handily outperforms it with the same number of
-processors.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere2.eps}
+function Sphere.  For ease of labeling, we call the modified algorithm Pick
+Best.  As can be seen, where SEPSO did not manage to perform as well as PSO
+Complete, Pick Best handily outperforms it with the same number of processors.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere2}
   \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing SEPSO Pick Best with previous results.}
+  evaluations, comparing Pick Best with previous results.}
   \label{fig:sphere-pickbest}
@@
 always keeping the best speculative position found makes intuitive sense.  A
-less intuitive result occurs when using SEPSO Pick Best with the function
-Griewank.  Griewank is a highly deceptive function prone to premature
-convergence.  Our intuition on the behavior of picking the best speculative
-child is that it would tend to prematurely converge more often than the
-original PSO.  However, as seen in \fig{griewank-pickbest}, SEPSO Pick Best
-increases the chance of finding the optimum while also reducing the number of
-iterations required to converge.  Perhaps SEPSO Pick Best encourages local
-exploitation, while the sparse communication of the Ring topology still allows
-for global exploration.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank4.eps}
+less intuitive result occurs when using Pick Best with the function Griewank.
+Griewank is a highly deceptive function prone to premature convergence.  Our
+intuition on the behavior of picking the best speculative child is that it
+would tend to prematurely converge more often than the original PSO.  However,
+as seen in \fig{griewank-pickbest}, Pick Best increases the chance of finding
+the optimum while also reducing the number of iterations required to converge.
+Perhaps Pick Best encourages local exploitation, while the sparse communication
+of the Ring topology still allows for global exploration.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank4}
   \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing SEPSO Pick Best with previous results.}
+  evaluations, comparing Pick Best with previous results.}
   \label{fig:griewank-pickbest}
@@
 \subsection{Pruning the Speculative Evaluations}
+\label{sec:pruning}
 
@@
 we could limit ourselves to one speculative evaluation per particle instead of
-$2n+1$.  While we cannot with certainty predict the correct branch every time,
-we can look at the statistics of the behavior of particles in general and
-perhaps come up with a good guess, reducing the number of necessary speculative
-evaluations.
-
-\subsubsection{Branch Statistics}
+$2n+1$.  As not all of the branches are evaluated in any given iteration, we
+call this technique pruning.
+
+A naive approach to pruning is to keep track of the last branch taken by each
+particle and speculate on that branch.  This turns out to be correct far less
+than half of the time on average.
+
+A more principled approach would be to use the statistical behavior of PSO to
+find probabilities of taking any particular branch.  While we cannot with
+certainty predict which branch a particle will take every time, if we can use
+statistics to narrow down the $2(n+1)$ possible evaluations to a few likely
+candidates, we can decrease the amount of computation required to do
+speculative evaluation and improve our performance.
+
+\subsection{Branch Statistics}
 
@@
 
-\begin{table}[h]
+\begin{table}[ht]
   \caption{Five Branches to Consider for Statistics}
@@
 
-\begin{table}[h]
+\begin{table}[ht]
   \caption{Branch Statistics in PSO}
@@
 
-We now show with what percentage a particle takes each of these branches for
-three different topologies and four different functions.  All of our statistics
-are from swarms of 240 particles.  Brief experimentation showed that other
-swarm sizes had similar statistics.  We ran 750 iterations on all combinations
-of functions and topologies except for the functions Griewank and Rastrigin
-with the Complete topology.  We found that those runs frequently converged past
-machine precision after 500 iterations, and that led to erroneously high values
-for the probability of stagnation.  Instead we ran for only 450 iterations on
-those two combinations.  All of our results were averaged over 20 runs of the
-algorithm; thus the probabilities presented are the averages of 3.6 million
-trials for the branch taken (2.16 million for the two with only 450
-iterations).  Table~\ref{tab:stats} contains the results, with the branch
-number coming from Table~\ref{tab:branches}.  We have already defined the
-functions Sphere, Griewank, and Rastrigin.  Rosenbrock is defined as
-$f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.
+We seek to find the probability of taking any given branch, given whatever
+information is needed: $\prob(B_t|\cdot)$.  In finding these probabilities, we
+do not attempt to derive any distribution from the PSO equations, we simply
+look at empirical distributions.  However, even with empirical distributions,
+the problem with this approach is that it is not clear what information
+influences the probability of taking a branch.  The failure of the naive
+approach seems to say that the previous branch, $B_{t+1}$, is not incredibly
+useful.  We look at two factors that we believe have a significant influence on
+$\prob(B_t)$: topology and function.  Thus we are looking at $\prob(B_t|T,F)$.
+
+We show in Table~\ref{tab:stats} with what percentage a particle takes each of
+these branches for three different topologies and four different functions.
+All of our statistics are from swarms of 240 particles.  Brief experimentation
+showed that other swarm sizes had similar statistics.  We ran 750 iterations on
+all combinations of functions and topologies except for the functions Griewank
+and Rastrigin with the Complete topology.  We found that those runs frequently
+converged past machine precision after 500 iterations, and that led to
+erroneously high values for the probability of stagnation.  Instead we ran for
+only 450 iterations on those two combinations.  All of our results were
+averaged over 20 runs of the algorithm; thus the probabilities presented are
+the averages of 3.6 million trials for the branch taken (2.16 million for the
+two with only 450 iterations).  Table~\ref{tab:stats} contains the results,
+with the branch number coming from Table~\ref{tab:branches}.  We have already
+defined the functions Sphere, Griewank, and Rastrigin.  Rosenbrock is defined
+as $f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.
 
@@
 interested in predicting the branch that any given particle will take at a
-particular iteration.
+particular iteration.  For our purposes, it appears that given a topology, the
+probability of selecting a branch and the function are independent, or
+$\prob(B_t|T,F) \approx \prob(B_t|T)$.
 
@@
 to \alg{centralized} is in Step 2, where the speculative children are
-generated.  Simply generate a subset of the children that would normally be
-generated.
-
-\subsubsection{What if we guessed wrong?}
+generated.  Instead of generating all possible speculative children, generate
+the subset of the children that is desired.
+
+\subsection{What if we guessed wrong?}
+\label{sec:wrong}
 
@@
 simplifying the work needed to be done in determining neighbors when dynamic
-topologies are involved.
+topologies are involved.  We call this technique Social Promotion.
 
@@
 allows results in better results when picking the best speculative child.  
-When rejecting all speculative children, which we with tongue in cheek call
-Social Promotion, no real improvement is seen.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere3.eps}
+When rejecting all speculative children no real improvement is seen.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere3}
   \caption{Function Sphere with a swarm that uses 240 processors per round of
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank5.eps}
+  \includegraphics[width=.8\columnwidth]{griewank5}
   \caption{Function Griewank with a swarm that uses 240 processors per round of
@@
 \subsection{More than one iteration ahead}
+\label{sec:manyiters}
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{sphere4.eps}
+  \includegraphics[width=.8\columnwidth]{sphere4}
   \caption{Function Sphere with a swarm that uses 240 processors per round of
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank6.eps}
+  \includegraphics[width=.8\columnwidth]{griewank6}
   \caption{Function Griewank with a swarm that uses 240 processors per round of
@@
 
-\subsection{Tokens}
+\subsection{Other possibilities}
+\label{sec:other}
 
@@
 
-If we use the concept of tokens commonly used in multirobot stuff (see, for
+If we use the concept of tokens commonly used in multi-robot stuff (see, for
 example, \dots), we can specify a number of tokens available for particles to
@@
 There are thus two distinct uses for tokens.  First, they could be used to try
-to evenly spread extra evaluations due to an incapatible number of particles
+to evenly spread extra evaluations due to an incompatible number of particles
 and processors.  In cursory experimentation, we found that this was not a very
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank7.eps}
+  \includegraphics[width=.8\columnwidth]{griewank7}
   \caption{Function Griewank with a swarm that uses 240 processors per round of
@@
 swarm size.  In an increasingly parallel world, such advancements will prove to
-be crucial in the continued effectiveness of PSO.
+be crucial to the continued effectiveness of PSO.
 
@@
 original PSO algorithm can either be exactly reproduced, two iterations at a
-time, or the behavior can be modified in order to improve performance.  We have
-shown results that conclusively demonstrate the superiority of our techniques
-over naively adding particles to the swarm when extra processors are available.
-
-We have shown several different ways to modify the behavior of PSO when doing
-speculative evaluation, and, as expected given the No Free Lunch Theorem,
-different modifications show varied performance with different functions.
-While our exploration of these techniques has been cursory in places, we
-believe the results shown are sufficient to demonstrate their promise.
+time, or the behavior can be modified in order to improve performance.  While
+exactly reproducing PSO often uses too many extra processors to be useful, when
+we allow ourselves some freedom with the algorithm we see great improvements
+over previous methods.  We have shown results that conclusively demonstrate the
+superiority of our techniques for several functions over naively adding
+particles to the swarm when extra processors are available.
+
+Though our methods show great improvements on some functions, they do not work
+for all functions.  As is commonly known, in PSO there is a trade-off between
+exploration and exploitation.  Some functions need only minimal exploration,
+and some never seem to have enough.  Increasing the swarm size is a natural way
+to increase exploration in a parallel environment.  However, once ``enough''
+exploration has been reached for any particular function, adding additional
+particles adds only incremental benefits.  A better use of the additional
+processors, as we have shown, is to perform some amount of speculative
+evaluation.
+
+Sphere is a function for which only a very small amount of exploration needs to
+be done, and our speculative methods show improvements on Sphere of over 40
+orders of magnitude.  Such functions are perfect for our methods.
+
+Rastrigin, on the other hand, is a function for which increasing the swarm size
+up to 4000 particles still shows improvements in the performance of the
+algorithm.  With such functions, the smaller swarm size required by speculative
+evaluation is not able to produce enough exploration to perform better than
+standard PSO at the swarm sizes we were able to experiment with.
+
+Griewank is a function in between Sphere and Rastrigin.  It is highly deceptive
+and is prone to premature convergence, but by adding particles to the swarm a
+point is reached where ``enough'' exploration is done, and the algorithm finds
+the optimum essentially all of the time.  For such functions, the best approach
+seems to be to increase the swarm size until ``enough'' exploration is reached,
+then use extra processors to perform speculative evaluation and increase the
+number of iterations performed.  Sphere and Rastrigin can be thought of as
+special cases of these types of functions; Sphere simply needs a very small
+swarm size to produce ``enough'' exploration, and Rastrigin requires a very
+large swarm.  We expect that for all functions there is a swarm size for which
+additional particles are less useful than additional iterations.
+
+Large parallel clusters are often required to successfully optimize practical
+modern problems.  To properly use PSO with such clusters, a balance needs to be
+made between using processors to increase the swarm size and using them to
+increase the speed of the algorithm.  This work is a first step in that
+direction.
 
@@
 
-We have speculated that our ``pick the best child'' method simply draws a
-number of samples from the next iteration of the sampling distribution of PSO
-and keeps one of them.  It would then be expected that the sampling
-distribution of that method is very similar to the distribution of the original
-PSO.  This remains to be proven, or even demonstrated empirically.
+We have speculated that our Pick Best method simply draws a number of samples
+from the next iteration of the sampling distribution of PSO and keeps one of
+them.  It would then be expected that the sampling distribution of that method
+is very similar to the distribution of the original PSO.  This remains to be
+proven, or even demonstrated empirically.
 
@@
 We turned what used to be a very simple algorithm with only one major parameter
-to tweak (the topology) into a big hairy mess with countless possibilities for
+to tweak (the topology) into a convoluted mess with countless possibilities for
 speculative evaluations.  Perhaps future research will be able to narrow down
 the endless possibilities into the few best ones, making the algorithm simple
-again.
+again.  We have presented some small amount of exploration of the possible
+alternatives, but our discussion is by no means exhaustive or conclusive.
 
@@
 computation in evolutionary algorithms was proposed in~\cite{poli-ai06}.
-
-Large parallel clusters are often required to successfully optimize practical
-modern problems.  To properly use PSO with such clusters, a balance needs to be
-made between using processors to increase the swarm size and using them to
-increase the speed of the algorithm, or for perhaps as yet unconsidered
-techniques.  This work is a first step in that direction.
 
@@
   \psfrag{pt}[C][C]{$\p_{t}$}
-  \includegraphics[width=.3\columnwidth]{messages1.eps}
+  \includegraphics[width=.3\columnwidth]{messages1}
   \caption{The production of $\p_{t}$ from the original particle
@@
   \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
-  \includegraphics[width=.5\columnwidth]{messages2.eps}
+  \includegraphics[width=.5\columnwidth]{messages2}
   \caption{The production of $\nonbest{\p}_{t+1}$ from the original particle 
@@
   \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
-  \includegraphics[width=.6\columnwidth]{messages3.eps}
+  \includegraphics[width=.6\columnwidth]{messages3}
   \caption{The production of each $\nonbest{\n}_{t+1}$ from the original
@@
   \psfrag{pt1}[C][C]{$\p_{t+1}$}
-  \includegraphics[width=.8\columnwidth]{messages4.eps}
+  \includegraphics[width=.8\columnwidth]{messages4}
   \caption{The production of $\p_{t+1}$ from the original particle 
@@
   \psfrag{st3-e}[C][C]{$\noeval{\sset}_{t+3}$}
-  \includegraphics[width=.5\columnwidth]{messages5.eps}
+  \includegraphics[width=.5\columnwidth]{messages5}
   \caption{The production of $\noeval{\p}_{t+2}$ and $\noeval{\sset}_{t+3}$

### 2010-03-16 honors_thesis/journal_paper/paper.tex

@@
 \DeclareMathOperator*{\argsup}{arg\;sup}
+\providecommand{\pers}{\ensuremath{P}}
+\providecommand{\neigh}{\ensuremath{N}}
+\providecommand{\nURand}{\ensuremath{U^\neigh}}
+\providecommand{\pURand}{\ensuremath{U^\pers}}
 \providecommand{\ppos}{\ensuremath{\Vec{x}}}
 \providecommand{\pvel}{\ensuremath{\Vec{v}}}
-\providecommand{\nbest}{\ensuremath{\Vec{b}_N}}
-\providecommand{\pbest}{\ensuremath{\Vec{b}_P}}
+\providecommand{\nbest}{\ensuremath{\Vec{b}^\neigh}}
+\providecommand{\pbest}{\ensuremath{\Vec{b}^\pers}}
 \providecommand{\constriction}{\ensuremath{\chi}}
-\providecommand{\coeff}{\ensuremath{\phi}}
+\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
+\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
 \providecommand{\obs}{\ensuremath{\Vec{\xi}}}
+\providecommand{\ofunc}{\ensuremath{f}}
+\providecommand{\swarm}{\ensuremath{swarm}}
 
 %SpecExPSO Stuff
+\providecommand{\indic}{\ensuremath{I}}
+\providecommand{\specvel}{\ensuremath{V}}
+\providecommand{\specpos}{\ensuremath{X}}
+\providecommand{\ln}{\ensuremath{ln}}
+\providecommand{\rn}{\ensuremath{rn}}
+\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
+\providecommand{\casedef}{\ensuremath{(\pbest:\nbest)}}
+\providecommand{\casexn}{\ensuremath{(x:\neigh*)}}
+\providecommand{\casexx}{\ensuremath{(x:x)}}
+\providecommand{\casexl}{\ensuremath{(x:\ln)}}
+\providecommand{\casexr}{\ensuremath{(x:\rn)}}
+\providecommand{\casepn}{\ensuremath{(\pers*:\neigh*)}}
+\providecommand{\casepl}{\ensuremath{(\pers*:\ln)}}
+\providecommand{\casepr}{\ensuremath{(\pers*:\rn)}}
 \providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
@@
 
-All of these parallel techniques use additional processors to increase the
+The purpose of this paper is to consider pso parallelization strategies for
+clusters of thousands of processors and functions for which a single
+evaluation will take on the order of hours.
+Our purpose is to explore the question ``since pso is often run with
+a swarm of size of 100, what would one do if 500 or a few thousand processors were available?''
+
+Previous parallel techniques use additional processors to increase the
 number of particles in the swarm, either by adding individual particles or by
-adding entire new subswarms.  The number of iterations of PSO that the
+adding entire new sub-swarms but in all cases increasing the number of particles.  The number of iterations of PSO that the
 algorithms can perform is thus inherently limited by the time it takes to
 evaluate the objective function---additional processors add more particles, not
-more iterations.  However, there are many functions for which performing more
+more iterations.
+for many functions there comes a point of diminishing returns with respect to adding particles.
+!!!! TODO WARNING WARNING MADE UP NUMBERS GET REAL ONES!! !!!!!
+For the well-know benchmark function Sphere and as shown in Figure XXXXa increasing the swarm size from 20 to 120 has a significant effect on the 
+speed of convergence. In this case, the swam goes below $1*10^5$ in x iteration with 20 particles, but 
+achieves this same value in  y iterations with a swarms size of 120, the additional 100 particles.
+However consider the change going from 2000 particles to 2100 particles. Here the additional 100 particles
+reduce the number of iterations to achieve $1*10^5$ from x iterations to y.
+Other benchmark function behave similarly. ans shown in fig xxxxb for ?? and xxxxc ?? (in all case 20 dimensions, average of 50 runs).
+Certainly these benchmark functions do not prove that this will be the case for all functions,
+but it does provide some evidence that for at least some functions, adding additional particles does not seem to have much benefit
+after some point, some time earlier fig xx some time later. This phenomena has been commented on by others
+in the form of recommended swarms sizes for various types of functions cite standard swarm paper.
+\begin{quote}
+Skip this??
+
+However, there are many functions for which performing more
 iterations with fewer particles is a better use of resources than performing
 fewer iterations with more particles.  
+\end{quote}
 
@@
 At each iteration of the algorithm, the position $\ppos_t$ and velocity
-$\pvel_t$ of each particle are updated as follows:
+$\pvel_t$ of each particle $i$ are updated as follows:
 \begin{align}
 \label{eq:velupdate}
-	\pvel_{t+1} &=
+	\pvel_{i,t+1} &=
 		\constriction \left[ \pvel_t +
-			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
-			\coeff_2\URand()\otimes(\nbest - \ppos_t)
+			\pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) +
+			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
 		\right] \\
 \label{eq:posupdate}
-	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
+	\ppos_{i,t+1} &= \ppos_{i,t} + \pvel_{i,t+1}
 \end{align}
-where \( \URand() \) is a vector of random numbers drawn from a uniform
-distribution, the \( \otimes \) operator is an element-wise vector
-multiplication, $\pbest$ (called personal best) is the best position the
-current particle has seen, and $\nbest$ (called neighborhood best) is the best
-position the neighbors of the current particle have seen~\cite{bratton-sis07}.
-\( \coeff_1 \), \( \coeff_2 \), and \( \constriction \) are parameters with
-prescribed values required to ensure convergence (2.05, 2.05, and .73,
-respectively)~\cite{clerc-tec02}.
+where \( \nURand_{i,t} \) and \( \nURand_{i,t} \) are vectors of i.i.d. random
+numbers drawn from a uniform distribution, the \( \otimes \) operator is an
+element-wise vector multiplication, $\pbest$ (called personal best) is the best
+position the current particle has seen, and $\nbest$ (called neighborhood best)
+is the best position the neighbors of the current particle have
+seen~\cite{bratton-sis07}.  \( \ncoeff \), \( \pcoeff \), and \( \constriction
+\) are parameters with prescribed values required to ensure convergence (2.05,
+2.05, and .73, respectively)~\cite{clerc-tec02}. Note also that we have made
+the particle identifier $i$ and the time step $t$ or $t+1$ explicit in the
+subscript and moved other qualifiers, such as the personal $P$ and personal
+best $P*$ qualifiers to the superscript position.  Superscripts can be miss
+read as exponents, but they are qualifiers throughout this paper.
 
@@
 possible next positions of the particle, evaluate all of them at once, and one
-of them will be right.
+of them would have been computed by normal PSO.
+
+We do this as described here.
+
+First consider formula for the \emph{next} iteration of PSO, which is identical
+to Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time
+subscripts are incremented.  As noted above and for simplicity in this
+explanation, we restrict our discussion here to the case of a Ring topology
+with two neighbors which we will call the ``right neighbor'' and ``left
+neighbor'' neighbors identified for each particle $i$ as $\rn(i)$ and $\ln(i)$
+respectively.
+
+Next note that the computation of $\pbest$ and $\nbest$ is not seen in the
+update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so we will make
+this mathematically explicit.  This update is usually implemented as a pair of
+independent ``if``-statements, but that is a bit awkward for our purpose here
+so we instead use a set of indicator variables, one each representing each
+possible update to the \emph{pair} $\casedef$.  These indicator variables allow
+us to easily identify the cases in which updates will be needed for $\nbest$,
+$\pbest$ or both.
+
+For example, we can write $\casexn$ to identify the case where $\pbest$ will be
+updated with the better position $\ppos_{i,t}$, shown in the first positions,
+but a no new neighborhood best was found so $\nbest$ remains in the second
+position.
+
+For any topology with 2 neighbors there are 7 possible update cases listed in
+Table~\ref{tab:evals}.  This list assumes, as is typically the case, that each
+particle is also part of its own neighborhood (can cause an update to
+$\nbest$). The list also omits the case where $\nbest$ is updated by
+$\ppos_{i,t}$ but $\pbest$ was not updated, as this is logically impossible.
+$\caseset$ represents the set of all cases.
+
+For minimization, we can define the indicator variable representing the case
+where for particle $i$ at time $t$ $\pbest$ will be updated with the better
+position $x_{i,t}$ but $\nbest$ will not updated by:
+\begin{align}
+\nonumber
+	\indic_{i,t}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
+\nonumber
+	& 1 \ if \  \ofunc(\ppos_{i,t+1}) > \ofunc(\pbest_{i,t}) \\
+\nonumber
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{i,t}) \\
+\nonumber
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\ln(i),t}) \\
+\nonumber
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\rn(i),t}) \\
+\label{eq:deficasexn}
+	& 0 \ otherwise
+\end{align}
+
+For compactness in our equations we represent this without the parameters since
+they can be inferred from the subscripts on $\indic_{i,t}^{\casexn}$.  The
+indicators variables for the other update cases shown in Table~\ref{tab:evals}
+($\casepn$, $\casepl$, $\casepr$, etc.) can be defined in similar fashion but
+obviously using the corresponding boolean expression.
+
+For each cases identified in Table~\ref{tab:evals} and identified by a
+corresponding indicator variable defined above, we define also a velocity
+update equation; once again using $\casexn$ as an example:
+\begin{align}
+\nonumber
+	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
+\label{eq:defvocasexn}
+		& \constriction \left[ \pvel_t +
+			\pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) +
+			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
+		\right]
+\end{align}
+
+Again we will drop the parameters for compactness and since
+they can be inferred from the subscripts on $\specvel_{i,t}^{\casexn}$.
+Since this equation is restricted to a specific case, $\casexn$ for this example,
+it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
+$\pbest_{i,t+1}=\ppos_{i,t+1}$ and $\nbest_{i,t+1}=\nbest_{i,t+1}$ we can re-write
+$\specvel_{i,t}^{\casexn}$ as:
+\begin{align}
+\nonumber
+	\specvel_{i,t+1}^{\casexn} & = \\
+\label{eq:defvcasexn}
+		& \constriction \left[ \pvel_t +
+			\pcoeff\pURand_{i,t}\otimes(\ppos_{i,t} - \ppos_{i,t}) +
+			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
+		\right]
+\end{align}
+
+For each cases identified in
+Table~\ref{tab:evals}
+and identified by a corresponding indicator variable defined above,
+we define also a position update equation with its compact form; once again using $\casexn$ as an example:
+\begin{align}
+\label{eq:defpcasexn}
+	\specpos_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
+\nonumber
+	\specpos_{i,t+1}^{\casexn} & =  \ppos_{i,t} + \specvel_{i,t+1}^{\casexn}
+\end{align}
+
+With this notation we can re-write the ``next iteration'' form of the PSO motion equations as:
+\begin{align}
+\label{eq:vel2update}
+	\pvel_{i,t+2} &= \sum_{c \in \caseset} \ \specvel_{i,t+1}^{c} \ \indic_{i,t}^{c} \\
+\label{eq:pos2update}
+	\ppos_{i,t+2} &= \sum_{c \in \caseset} \ \specpos_{i,t+1}^{c} \ \indic_{i,t}^{c}
+\end{align}
+
+The value of the objective function at $\ppos_{i,t+2}$ is given by:
+\begin{align}
+\label{eq:pos2value}
+	\ofunc (\ppos_{i,t+2}) = \sum_{c \in \caseset} \ \ofunc(\specpos_{i,t+1}^{c}) \ \indic_{i,t}^{c}
+\end{align}
+or, with $\specpos$ and $\indic$ written with the parameters which were omitted above:
+\begin{align}
+\nonumber
+\ofunc & (\ppos_{i,t+2}) = \sum_{c \in \caseset} \\
+\nonumber
+& \ofunc(\specpos_{i,t+1}^{c}(\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})) \\
+\label{eq:pos2valuelong}
+& \indic_{i,t}^{c}(\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))
+\end{align}
+
+None of the $\specpos_{i,t+1}^{c} \ \ c \in\ \caseset$ depend upon $f(i,x_t)$
+or any other objective function evaluation at $t+1$.
+Thus PSO has been refactored such that the algorithm can begin computing
+all of the objective function evaluations needed in iteration $t+1$ before $f(x_t)$ is computed.
+
+This refactoring comes a significant expense, instead of just computing
+$\ofunc(\ppos_{i,t+1})$ and then $\ofunc(\ppos_{i,t+2})$ in the usual way, this refactored PSO
+can compute $\ofunc(\ppos_{i,t+1})$ and all 
+of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time.
+Once all of the evaluations $f(\ppos_{i,t+1})$ are completed only one
+of the indicator functions $\indic_{i,t}^{c} \ \ c \in \caseset$ will have been set to one;
+the rest will be ignored, and might just as well never have been computed.
+We call these evaluated points $\specpos_{i,t+1}^{c}$ ``speculative children''
+because they may or may not be needed.
+
+The potential value of these speculative children lies in the phase used above
+``this refactored PSO
+can compute $\ofunc(\ppos_{i,t+1})$ and all 
+of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
+Remember that our intent is to accelerate parallel PSO and that we
+are interested in objective functions that have at least non-trivial computational
+complexity.
+Thus we assume that computing
+$\ofunc(\ppos_{i,t+1})$ and all 
+of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ is the computationally difficult part.
+
+Suppose that unlimited processors were available and that the evaluation of 
+the objective function took 1 hour. If for each particle $i$ 
+the computations
+$\ofunc(\ppos_{i,t+1})$ and all 
+of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ were sent to separate
+processors and completed in parallel.
+In this environment all of the function evaluations needed for 2 iterations of PSO
+could be completed in just one hour where as the conventional implementation would
+require two hours for those same 2 iterations of PSO.
+
+the same time''. 
+in the indicator functions $indic_{i,t}^{c} \ \ c \in \caseset$
+given the values $\ofunc(\ppos_{i,t+1}) i \in \swarm$ 
+is inconsequential relative to the computational cost of obtaining those values 
+$\ofunc(\ppos_{i,t+1})$ and all
 

### 2010-03-18 honors_thesis/journal_paper/paper.tex
Halfway through the algorithm

--- 2010-03-15 honors_thesis/journal_paper/paper.tex
@@
 \DeclareMathOperator*{\argsup}{arg\;sup}
-\providecommand{\pers}{\ensuremath{c}}
-\providecommand{\neigh}{\ensuremath{s}}
+\providecommand{\pers}{\ensuremath{P}}
+\providecommand{\neigh}{\ensuremath{N}}
 \providecommand{\nURand}{\ensuremath{U^\neigh}}
@@
 \providecommand{\ppos}{\ensuremath{\Vec{x}}}
+\providecommand{\fppos}{\ensuremath{\Vec{fx}}}
 \providecommand{\pvel}{\ensuremath{\Vec{v}}}
-\providecommand{\nbest}{\ensuremath{\Vec{x}^{\neigh*}}}
-\providecommand{\pbest}{\ensuremath{\Vec{x}^{\pers*}}}
+\providecommand{\nbest}{\ensuremath{\Vec{b}^\neigh}}
+\providecommand{\fnbest}{\ensuremath{\Vec{fb}^\neigh}}
+\providecommand{\pbest}{\ensuremath{\Vec{b}^\pers}}
+\providecommand{\fpbest}{\ensuremath{\Vec{fb}^\pers}}
 \providecommand{\constriction}{\ensuremath{\chi}}
-\providecommand{\ncoeff}{\ensuremath{\phi^s}}
-\providecommand{\pcoeff}{\ensuremath{\phi^P}}
+\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
+\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
 \providecommand{\obs}{\ensuremath{\Vec{\xi}}}
@@
 \providecommand{\specpos}{\ensuremath{X}}
+\providecommand{\specval}{\ensuremath{FX}}
 \providecommand{\ln}{\ensuremath{ln}}
@@
 
+This previous parallel techniques use additional processors to increase the
+number of particles in the swarm, either by adding individual particles or by
+adding entire new sub-swarms but in all cases increasing the number of particles.  The number of iterations of PSO that the
+algorithms can perform is thus inherently limited by the time it takes to
+evaluate the objective function---additional processors add more particles, not
+more iterations.
+
 The purpose of this paper is to consider pso parallelization strategies for
@@
 
-Previous parallel techniques use additional processors to increase the
-number of particles in the swarm, either by adding individual particles or by
-adding entire new sub-swarms but in all cases increasing the number of particles.  The number of iterations of PSO that the
-algorithms can perform is thus inherently limited by the time it takes to
-evaluate the objective function---additional processors add more particles, not
-more iterations.
-for many functions there comes a point of diminishing returns with respect to adding particles.
+For many functions there comes a point of diminishing returns with respect to adding particles.
 !!!! TODO WARNING WARNING MADE UP NUMBERS GET REAL ONES!! !!!!!
-For the well-know benchmark function Sphere and as shown in Figure XXXXa increasing the swarm size from 20 to 120 has a significant effect on the 
+For the well-know benchmark function Sphere and as shown in \fig{basic-sphere} increasing the swarm size from 20 to 120 has a significant effect on the 
 speed of convergence. In this case, the swam goes below $1*10^5$ in x iteration with 20 particles, but 
@@
 reduce the number of iterations to achieve $1*10^5$ from x iterations to y.
-Other benchmark function behave similarly. ans shown in fig xxxxb for ?? and xxxxc ?? (in all case 20 dimensions, average of 50 runs).
+The effect is easier to see if we plot the value obtained as a function of swarm size, as show in 
+fig{sideways-sphere} increasing the swarm size from 20 to 120 has a significant effect on the 
+Other benchmark function behave similarly. As shown in fig xxxxb for ?? and xxxxc ?? (in all case 20 dimensions, average of 50 runs).
 Certainly these benchmark functions do not prove that this will be the case for all functions,
-but it does provide some evidence that for at least some functions, adding additional particles does not seem to have much benefit
-after some point, some time earlier fig xx some time later. This phenomena has been commented on by others
-in the form of recommended swarms sizes for various types of functions cite standard swarm paper.
-\begin{quote}
-Skip this??
-
-However, there are many functions for which performing more
-iterations with fewer particles is a better use of resources than performing
-fewer iterations with more particles.  
-\end{quote}
+but it does provide some evidence that for at least some functions,
+additional particles does not seem to have much benefit after some point,
+some time earlier, some time later. This issue has been commented on by others
+in the form of recommended swarms sizes for various types of functions (cite standard swarm paper).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{evals_sphere.eps}
+  \caption{Function Sphere with a swarm with various swarm sizes.}
+  \label{fig:evals-sphere}
+\end{figure}
 
@@
 position.  This could not be done in gradient methods, for instance, where the
-step size for sampling depends on the gradient at the current position, but it
+step size for sampling depends on the gradient and therefor the position and \emph{value}
+at the current position, but it
 has been done in simulated annealing~\cite{witte-tpds91}.
 
-\begin{quote}
-Skip this??
-
-Speculative evaluation is possible in PSO because the motion of particles
-depends only on the \emph{positions} of the particle and its best neighbor, not
-the \emph{value} of the objective function at those positions.  We do not have
-to wait for a particle's function evaluation to complete in order to know where
-it might move to next, so we can compute the set of all possible next positions
-and send them to be evaluated along with the particle's current position.  Once
-the current position of each particle has been evaluated, we can determine
-which of the speculatively evaluated positions matches the branch taken by the
-particle.  We have thus evaluated two iterations at once.  This can easily get
-unwieldy if the set of next possible positions for each particle is large, but
-a wise choice of topology limits that set to a reasonable size.
-
-But the following does need some intro
-\end{quote}
+% Use this later?
+% Skip this??
+% 
+% Speculative evaluation is possible in PSO because the motion of particles
+% depends only on the \emph{positions} of the particle and its best neighbor, not
+% the \emph{value} of the objective function at those positions.  We do not have
+% to wait for a particle's function evaluation to complete in order to know where
+% it might move to next, so we can compute the set of all possible next positions
+% and send them to be evaluated along with the particle's current position.  Once
+% the current position of each particle has been evaluated, we can determine
+% which of the speculatively evaluated positions matches the branch taken by the
+% particle.  We have thus evaluated two iterations at once.  This can easily get
+% unwieldy if the set of next possible positions for each particle is large, but
+% a wise choice of topology limits that set to a reasonable size.
+% 
+% But the following does need some intro
 
@@
 \end{align}
-where \( \nURand_{i,t} \) and \( \nURand_{i,t} \) are vectors of i.i.d. random numbers drawn from a uniform
+where \( \nURand_{i,t} \) and \( \nURand_{i,t} \) are vectors of independent
+and identically distributed random numbers drawn from a uniform
 distribution, the \( \otimes \) operator is an element-wise vector
@@
 and the time step $t$ or $t+1$ explicit in the subscript and moved other qualifiers,
-such as the personal $P$ and personal best $P*$ qualifiers to the superscript position.
+such as the personal $\pers$ and personal best $\pers*$ qualifiers to the superscript position.
 Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.
@@
 functions, as would be expected by the No Free Lunch
-theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
-beyond the scope of this paper.  The two most common topologies used in the
+theorem~\cite{wolpert-tec97}.
+The two most common topologies used in the
 literature are the Ring topology and the Complete
@@
 particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
-topology, each particle has $p$ neighbors.  In this paper we use these
-topologies, as well as a parallel adaptation of the Complete topology, called
-Random, that has been shown to simulate the behavior of Complete with far less
-communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
-randomly picks two other particles to share information with at each iteration,
-along with itself.  Thus in both the Ring and the Random topologies, all
-particles have 3 neighbors.
+topology, each particle has $p$ neighbors.
+% TOD USE THIS LATER!
+% In this paper we use these
+% topologies, as well as a parallel adaptation of the Complete topology, called
+% Random, that has been shown to simulate the behavior of Complete with far less
+% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
+% randomly picks two other particles to share information with at each iteration,
+% along with itself.  Thus in both the Ring and the Random topologies, all
+% particles have 3 neighbors.
 
@@
 The PSO algorithm just described can be parallelized by having each additional
-processor add a particle or a set of particles to the swarm.  But for some
+processor add a particle or a set of particles to the swarm.
+But as we have seen in the introduction, for some
 functions, a more desirable use of resources is to speed up the algorithm,
-performing more iterations instead of adding more particles.  Barring
-parallelizing the evaluation of the objective function itself, which may not be
-possible, previous techniques have not addressed this issue.
-
+performing more iterations instead of adding more particles (for a fixed number of function evaluations),
+previous techniques have not addressed this issue.
 In order to perform additional iterations by adding additional processors, we
-take inspiration from speculative execution in processors.  If we can guess
-what a particle's next position will be from its current position, we can
-evaluate both in parallel and perform two iterations of PSO at once.  Lacking
-a good way to guess a particle's next position, we can at least enumerate all
-possible next positions of the particle, evaluate all of them at once, and one
-of them would have been computed by normal PSO.
-
-We do this as described here.
-
-First consider formula for the \emph{next} iteration of PSO, which is identical to 
-Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time subscripts are incremented.
-As noted above and for simplicity in this explanation, we restrict our discussion here to
-the case of a Ring topology with two neighbors which we will call the
-``right neighbor'' and ``left neighbor'' neighbors identified for each
-particle $i$ as $\rn(i)$ and $\ln(i)$ respectively.
-
-Next note that the computation of $\pbest$ and $\nbest$ is not seen in
-the update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so
-we will make this mathematically explicit.
-This update is usually implemented as a pair of independent ``if``-statements, but that is
-a bit awkward for our purpose here so we instead use a set of indicator variables,
-one each representing each possible update to the \emph{pair} $\casedef$.
-These indicator variables allow us to easily identify the cases in which updates will
-be needed for $\nbest$, $\pbest$ or both.
-
-For example, we can write $\casexn$ to identify the case where $\pbest$ will
-be updated with the better position $\ppos_{i,t}$, shown in the first positions,
-but a no new neighborhood best was found so
-$\nbest$ remains in the second position.
+take inspiration from speculative execution in processors.
+We do this by refactoring the standard PSO equations as described here.
+
+%First consider formula for the \emph{next} iteration of PSO, which is identical
+%to Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time
+%subscripts are incremented.  As noted above and
+For simplicity in this
+explanation, we restrict our discussion here to the case of a Ring topology
+with two neighbors which we will call the ``right neighbor'' and ``left
+neighbor'' neighbors identified for each particle $i$ as $\rn(i)$ and $\ln(i)$
+respectively.
+
+Next note that the computation of $\pbest$ and $\nbest$ is not seen in the
+update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so we will make
+this mathematically explicit.  This update is usually implemented as a pair of
+independent ``if``-statements, but that is a bit awkward for our purpose here
+so we instead use a set of indicator variables, one each representing each
+possible update to the \emph{pair} $\casedef$.  These indicator variables allow
+us to easily identify the cases in which updates will be needed for $\nbest$,
+$\pbest$ or both.
+
+For example, we can write $\casexn$ to identify the case where $\pbest$ will be
+updated with the better position $\ppos_{i,t}$, shown in the first positions,
+but a no new neighborhood best was found so $\nbest$ remains in the second
+position.
 
 For any topology with 2 neighbors there are 7 possible update cases listed in
-Table~\ref{tab:evals}.
-This list assumes, as is typically the case, that each particle is also
-part of its own neighborhood (can cause an update to $\nbest$). The list also
-omits the case where $\nbest$ is updated by $\ppos_{i,t}$ but $\pbest$ was not updated, as this is
-logically impossible. $\caseset$ represents the set of all cases.
-
-For minimization, we can define the indicator variable representing the case where for particle $i$ at time $t$
-$\pbest$ will
-be updated with the better position $x_{i,t}$ but
-$\nbest$ will not updated by:
+Table~\ref{tab:evals}.  This list assumes, as is typically the case, that each
+particle is also part of its own neighborhood (can cause an update to
+$\nbest$). The list also omits the case where $\nbest$ is updated by
+$\ppos_{i,t}$ but $\pbest$ was not updated, as this is logically impossible.
+$\caseset$ represents the set of all cases.
+
+For minimization, we can define this example indicator variable representing the case
+where for particle $i$ will need to update $\pbest$ with the better
+position $x_{i,t}$ for the next iteration ($t+1$), but $\nbest$ will not updated by:
 \begin{align}
 \nonumber
-	\indic_{i,t}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
+	\indic_{i,t+1}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
 \nonumber
@@
 \nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{i,t}) \\
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{i,t+1}) \\
 \nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\ln(i),t}) \\
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\ln(i),t+1}) \\
 \nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\rn(i),t}) \\
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\rn(i),t+1}) \\
 \label{eq:deficasexn}
@@
 For compactness in our equations we represent this without the parameters since
-they can be inferred from the subscripts on $\indic_{i,t}^{\casexn}$.
-The indicators variables for the other update cases shown in 
-Table~\ref{tab:evals}
-($\casepn$, $\casepl$, $\casepr$, etc.)
-can be defined in similar fashion but obviously using the
-corresponding boolean expression.
-
-For each cases identified in
-Table~\ref{tab:evals}
-and identified by a corresponding indicator variable defined above,
-we define also a velocity update equation; once again using $\casexn$ as an example:
+they can be inferred from the subscripts on $\indic_{i,t}^{\casexn}$.  The
+indicator variables for the other update cases shown in Table~\ref{tab:evals}
+($\casepn$, $\casepl$, $\casepr$, etc.) can be defined in similar fashion but
+obviously using the corresponding boolean expression.
+
+For each cases identified in Table~\ref{tab:evals} and identified by a
+corresponding indicator variable defined above, we also provide notation for the \emph{next} velocity
+update equation; once again using $\casexn$ as an example:
 \begin{align}
 \nonumber
-	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
+	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1}, \ppos_{\ln(i),t+1}, \ppos_{\rn(i),t+1},
+        \pbest_{i,t} , \nbest_{i,t}, \pURand_{i,t}, \nURand_{i,t})= \\
 \label{eq:defvocasexn}
 		& \constriction \left[ \pvel_t +
-			\pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) +
-			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
+			\pcoeff\pURand_{i,t+1}\otimes(\pbest_{i,t+1} - \ppos_{i,t}+1) +
+			\ncoeff\nURand_{i,t+1}\otimes(\nbest_{i,t+1} - \ppos_{i,t+1})
 		\right]
@@
 they can be inferred from the subscripts on $\specvel_{i,t}^{\casexn}$.
-Since this equation is restricted to a specific case, $\casexn$ for this example,
+The update equation is taken exactly from the statdard update equations~\eqref{eq:velupdate}
+and \eqref{eq:posupdate}, but
+since it is restricted to a specific case, $\casexn$ for this example,
 it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
 $\pbest_{i,t+1}=\ppos_{i,t+1}$ and $\nbest_{i,t+1}=\nbest_{i,t+1}$ we can re-write
-$\specvel_{i,t}^{\casexn}$ as:
+$\specvel_{i,t+1}^{\casexn}$ as:
 \begin{align}
@@
 		& \constriction \left[ \pvel_t +
-			\pcoeff\pURand_{i,t}\otimes(\ppos_{i,t} - \ppos_{i,t}) +
-			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
+			\pcoeff\pURand_{i,t+1}\otimes(\ppos_{i,t+1} - \ppos_{i,t+1}) +
+			\ncoeff\nURand_{i,t+1}\otimes(\nbest_{i,t+1} - \ppos_{i,t+1})
 		\right]
@@
 
+Lastly, we also define notation for the value of the objective function at these points,
+once again using $\casexn$ as an example:
+\begin{align}
+\label{eq:deffcasexn}
+	\specval_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
+\nonumber
+	\specval_{i,t+1}^{\casexn} & =  \ofunc(\specval_{i,t+1}^{\casexn}
+\end{align}
+
 With this notation we can re-write the ``next iteration'' form of the PSO motion equations as:
@@
 \label{eq:vel2update}
-	\pvel_{i,t+2} &= \sum_{c \in \caseset} \ \specvel_{i,t+1}^{c} \ \indic_{i,t}^{c} \\
+	\pvel_{i,t+1} &= \sum_{c \in \caseset} \ \specvel_{i,t+1}^{c} \ \indic_{i,t+1}^{c} \\
 \label{eq:pos2update}
-	\ppos_{i,t+2} &= \sum_{c \in \caseset} \ \specpos_{i,t+1}^{c} \ \indic_{i,t}^{c}
+	\ppos_{i,t+1} &= \sum_{c \in \caseset} \ \specpos_{i,t+1}^{c} \ \indic_{i,t+1}^{c}
 \end{align}
 
-The value of the objective function at $\ppos_{i,t+2}$ is given by:
+The value of the objective function at $\ppos_{i,t+1}$ is given by:
 \begin{align}
-\label{eq:pos2value}
-	\ofunc (\ppos_{i,t+2}) = \sum_{c \in \caseset} \ \ofunc(\specpos_{i,t+1}^{c}) \ \indic_{i,t}^{c}
+\label{eq:val2update}
+	\ofunc (\ppos_{i,t+1}) = \sum_{c \in \caseset} \ \specval_{i,t+1}^{c} \ \indic_{i,t+1}^{c}
 \end{align}
@@
 \nonumber
-\ofunc & (\ppos_{i,t+2}) = \sum_{c \in \caseset} \\
+\ofunc & (\ppos_{i,t+1}) = \sum_{c \in \caseset} \\
 \nonumber
-& \ofunc(\specpos_{i,t+1}^{c}(\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})) \\
-\label{eq:pos2valuelong}
+& \ofunc(\specpos_{i,t+1}^{c}(\ppos_{i,t+1},\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1},
+\pbest_{i,t},\nbest_{i,t},\pURand_{i,t}, \nURand_{i,t})) \\
+\label{eq:val2updatelong}
 & \indic_{i,t}^{c}(\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))
@@
 or any other objective function evaluation at $t+1$.
-Thus PSO has been refactored such that the algorithm can begin computing
+Thus PSO has been re-factored such that the algorithm can begin computing
 all of the objective function evaluations needed in iteration $t+1$ before $f(x_t)$ is computed.
@@
 The potential value of these speculative children lies in the phase used above
-``this refactored PSO
+``this re-factored PSO
 can compute $\ofunc(\ppos_{i,t+1})$ and all 
@@
 Suppose that unlimited processors were available and that the evaluation of 
-the objective function took 1 hour. If for each particle $i$ 
+the objective function took 1 hour. Suppose also that for each particle $i$ 
 the computations
@@
 processors and completed in parallel.
+\emph{Given} the values 
+of $\ofunc(\ppos_{i,t+1})$ and all 
+$\ofunc(\ppos_{i,t+1}) i \in \swarm$ 
+the computation of the indicator functions $indic_{i,t}^{c} \ \ c \in \caseset$ and to complete iteration t+1
+is inconsequential relative to the computational cost of obtaining the function values. 
 In this environment all of the function evaluations needed for 2 iterations of PSO
@@
 
-the same time''. 
-in the indicator functions $indic_{i,t}^{c} \ \ c \in \caseset$
-given the values $\ofunc(\ppos_{i,t+1}) i \in \swarm$ 
-is inconsequential relative to the computational cost of obtaining those values 
-$\ofunc(\ppos_{i,t+1})$ and all
+While the algebra show that the computation of 
+equations~\eqref{eq:vel2update} and \eqref{eq:pos2update} are identical to the 
+equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} (at iteration $t+2$),
+There are some issues better seen in an algorithmic setting, in particular the use of the random
+vectors, 
+$\pURand_{i,t}$ and $\nURand_{i,t}$. As with the algebra each transformation of the algorithm is
+provably equivalent (except that the algorithms will no longer be able to do an odd number of iteration,
+since it will do two at a time). At this point we are moving work around and adding work, but the
+resultant algorithm, produces numerically identical results.
+
+\begin{algorithm} [h!]
+  \caption{Standard PSO (minimizing)}
+  \label{alg:standard}
+  \begin{algorithmic}[1]
+    \STATE t=0
+    \STATE Initialize the swarm including
+    \FOR{$i \in \swarm$}
+      \STATE {$\ppos_{i,t} \leftarrow Random position$}
+      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
+      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
+      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
+    \ENDFOR
+    \REPEAT
+        \STATE $t=t+1$
+        \FOR{$i \in \swarm$}
+          \STATE {$\fppos_{i.t} \leftarrow \ofunc(\ppos_{i,t})$}
+        \ENDFOR
+        \FOR{$i \in \swarm$}
+          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
+          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
+          \label{alg1:startifs}
+          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
+          \STATE $\fpbest_{i,t}\leftarrow \fpbest_{i,t-1}$
+          \IF {$\fppos_{i,t}<\fpbest_{i,t-1}$}
+            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
+            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
+          \ENDIF
+          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
+          \STATE $\fnbest_{i,t}\leftarrow \fnbest_{i,t-1}$
+          \IF {$\fppos_{i,t}<\fnbest_{i,t}$}
+            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
+            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
+          \ENDIF
+          \IF {$\fppos_{\ln(i),t}<\fnbest_{i,t}$}
+            \STATE $\pbest_{i,t}\leftarrow \ppos_{\ln(i),t}$
+            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\ln(i),t}$
+          \ENDIF
+          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\nbest_{\rn(i),t})$}
+            \STATE $\pbest_{i,t}\leftarrow \ppos_{\rn(i),t}$
+            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\rn(i),t}$
+          \label{alg1:endifs}
+          \ENDIF
+          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
+          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
+        \ENDFOR
+    \UNTIL {$t >$ max iteration or convergence criteria met}
+  \end{algorithmic}
+\end{algorithm}
+
+This algorithm can be transformed, will retaining it exact behavior, by
+``unrolling'' one iteration of the ``Repeat'' loop, re-writing the ``if'''s in lines
+\ref{alg1:startifs} through
+\ref{alg1:endifs} to use the indicator functions $\indic_{i,t+1}$, and
+
+\begin{algorithm} [h!]
+  \caption{Unrolled PSO}
+  \label{alg:unrolled}
+  \begin{algorithmic}[1]
+    \STATE t=0
+    \STATE Initialize the swarm including
+    \FOR{$i \in \swarm$}
+      \STATE {$\ppos_{i,t} \leftarrow Random position$}
+      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
+      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
+      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
+    \ENDFOR
+    \REPEAT
+        \STATE $t=t+1$
+        \FOR{$i \in \swarm$}
+          \label{alg2:fatitert}
+          \STATE DISPATCH {$\fppos_{i.t} \leftarrow \ofunc(\ppos_{i,t})$} to a seperate processor
+        \ENDFOR
+        \FOR{$i \in \swarm$}
+          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
+          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
+          \FOR{$c \in \caseset$}
+            \STATE Compute $\specvel_{i,t+1}^c$ and $\specpos_{i,t+1}^c$.
+	    Examples shown in equations~\eqref{eq:defvcasexn} and~\eqref{eq:defpcasexn}
+            \label{alg2:fatitertp1}
+            \STATE DISPATCH computation of $\specval_{i,t+1}^c$ to a seperate processor.
+	    Example shown in equation~\eqref{eq:deffcasexn}.
+	    \COMMENT Note that none of the computations, $\specvel_{i,t+1}^c$,
+	    $\specpos_{i,t+1}^c$ and $\specval_{i,t+1}^c$ depend on $\fppos_{i.t}$.
+          \ENDFOR
+        \ENDFOR
+        \STATE WAIT for all function evaluations dispatched above to complete
+        \FOR{$i \in \swarm$}
+          \FOR{$c \in \caseset$}
+            \STATE Compute $\indic_{i,t+1}^c$
+	           Examples shown in equation~\eqref{eq:deficasexn} and~\eqref{eq:defpcasexn}
+	  \ENDFOR 
+          \STATE Compute $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:vel2update} 
+          \STATE Compute $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:pos2update}
+          \STATE Compute $\fppos_{i,t+1}$ as shown in equation~\eqref{eq:val2update}
+        \ENDFOR
+
+        \STATE $t=t+1$
+        \FOR{$i \in \swarm$}
+          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
+          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
+          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
+          \STATE $\fpbest_{i,t}\leftarrow \fpbest_{i,t-1}$
+          \IF {$\fppos_{i,t}<\fpbest_{i,t-1}$}
+            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
+            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
+          \ENDIF
+          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
+          \STATE $\fnbest_{i,t}\leftarrow \fnbest_{i,t-1}$
+          \IF {$\fppos_{i,t}<\fnbest_{i,t}$}
+            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
+            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
+          \ENDIF
+          \IF {$\fppos_{\ln(i),t}<\fnbest_{i,t}$}
+            \STATE $\pbest_{i,t}\leftarrow \ppos_{\ln(i),t}$
+            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\ln(i),t}$
+          \ENDIF
+          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\nbest_{\rn(i),t})$}
+            \STATE $\pbest_{i,t}\leftarrow \ppos_{\rn(i),t}$
+            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\rn(i),t}$
+          \ENDIF
+          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
+          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
+        \ENDFOR
+    \UNTIL {$t >$ max iteration or convergence criteria met}
+  \end{algorithmic}
+\end{algorithm}
 

### 2010-03-18 honors_thesis/journal_paper/paper.tex
the commit that f9f984298 ('start alg') should have been

@@
 \providecommand{\specpos}{\ensuremath{X}}
-\providecommand{\specval}{\ensuremath{FX}}
 \providecommand{\ln}{\ensuremath{ln}}
@@
 
-Lastly, we also define notation for the value of the objective function at these points,
-once again using $\casexn$ as an example:
-\begin{align}
-\label{eq:deffcasexn}
-	\specval_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
-\nonumber
-	\specval_{i,t+1}^{\casexn} & =  \ofunc(\specval_{i,t+1}^{\casexn}
-\end{align}
-
 With this notation we can re-write the ``next iteration'' form of the PSO motion equations as:
@@
 \begin{align}
-\label{eq:val2update}
-	\ofunc (\ppos_{i,t+1}) = \sum_{c \in \caseset} \ \specval_{i,t+1}^{c} \ \indic_{i,t+1}^{c}
+\label{eq:pos2value}
+	\ofunc (\ppos_{i,t+1}) = \sum_{c \in \caseset} \ \ofunc(\specpos_{i,t+1}^{c}) \ \indic_{i,t+1}^{c}
 \end{align}
@@
 \pbest_{i,t},\nbest_{i,t},\pURand_{i,t}, \nURand_{i,t})) \\
-\label{eq:val2updatelong}
+\label{eq:pos2valuelong}
 & \indic_{i,t}^{c}(\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))
@@
         \FOR{$i \in \swarm$}
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
           \label{alg1:startifs}
@@
             \STATE $\fpbest_{i,t}\leftarrow \fppos_{\rn(i),t}$
-          \label{alg1:endifs}
           \ENDIF
+          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
+          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
           \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
@@
 
-This algorithm can be transformed, will retaining it exact behavior, by
-``unrolling'' one iteration of the ``Repeat'' loop, re-writing the ``if'''s in lines
-\ref{alg1:startifs} through
-\ref{alg1:endifs} to use the indicator functions $\indic_{i,t+1}$, and
-
-\begin{algorithm} [h!]
-  \caption{Unrolled PSO}
-  \label{alg:unrolled}
-  \begin{algorithmic}[1]
-    \STATE t=0
-    \STATE Initialize the swarm including
-    \FOR{$i \in \swarm$}
-      \STATE {$\ppos_{i,t} \leftarrow Random position$}
-      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
-      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
-      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
-    \ENDFOR
-    \REPEAT
-        \STATE $t=t+1$
-        \FOR{$i \in \swarm$}
-          \label{alg2:fatitert}
-          \STATE DISPATCH {$\fppos_{i.t} \leftarrow \ofunc(\ppos_{i,t})$} to a seperate processor
-        \ENDFOR
-        \FOR{$i \in \swarm$}
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \FOR{$c \in \caseset$}
-            \STATE Compute $\specvel_{i,t+1}^c$ and $\specpos_{i,t+1}^c$.
-	    Examples shown in equations~\eqref{eq:defvcasexn} and~\eqref{eq:defpcasexn}
-            \label{alg2:fatitertp1}
-            \STATE DISPATCH computation of $\specval_{i,t+1}^c$ to a seperate processor.
-	    Example shown in equation~\eqref{eq:deffcasexn}.
-	    \COMMENT Note that none of the computations, $\specvel_{i,t+1}^c$,
-	    $\specpos_{i,t+1}^c$ and $\specval_{i,t+1}^c$ depend on $\fppos_{i.t}$.
-          \ENDFOR
-        \ENDFOR
-        \STATE WAIT for all function evaluations dispatched above to complete
-        \FOR{$i \in \swarm$}
-          \FOR{$c \in \caseset$}
-            \STATE Compute $\indic_{i,t+1}^c$
-	           Examples shown in equation~\eqref{eq:deficasexn} and~\eqref{eq:defpcasexn}
-	  \ENDFOR 
-          \STATE Compute $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:vel2update} 
-          \STATE Compute $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:pos2update}
-          \STATE Compute $\fppos_{i,t+1}$ as shown in equation~\eqref{eq:val2update}
-        \ENDFOR
-
-        \STATE $t=t+1$
-        \FOR{$i \in \swarm$}
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
-          \STATE $\fpbest_{i,t}\leftarrow \fpbest_{i,t-1}$
-          \IF {$\fppos_{i,t}<\fpbest_{i,t-1}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
-          \ENDIF
-          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
-          \STATE $\fnbest_{i,t}\leftarrow \fnbest_{i,t-1}$
-          \IF {$\fppos_{i,t}<\fnbest_{i,t}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
-          \ENDIF
-          \IF {$\fppos_{\ln(i),t}<\fnbest_{i,t}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{\ln(i),t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\ln(i),t}$
-          \ENDIF
-          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\nbest_{\rn(i),t})$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{\rn(i),t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\rn(i),t}$
-          \ENDIF
-          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
-          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
-        \ENDFOR
-    \UNTIL {$t >$ max iteration or convergence criteria met}
-  \end{algorithmic}
-\end{algorithm}
+stuff \ref{alg1:startifs} more
 

### 2010-03-18 honors_thesis/journal_paper/paper.tex
Fixed the permissions on paper.tex

--- 2010-03-18 honors_thesis/journal_paper/paper.tex
@@
 current particle has seen, and $\nbest$ (called neighborhood best) is the best
-position the neighbors of the current particle have seen.  \( \ncoeff \), \(
+position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
 \pcoeff \), and \( \constriction \) are parameters with prescribed values
@@
 We show in \fig{sphere-pickbest} the results of using this modification on the
-function Sphere.  For ease of labeling, we call the modified algorithm SEPSO
-Pick Best.  As can be seen, where SEPSO did not manage to perform as well as
-PSO Complete, SEPSO Pick Best handily outperforms it with the same number of
-processors.
+function Sphere.  For ease of labeling, we call the modified algorithm Pick
+Best.  As can be seen, where SEPSO did not manage to perform as well as PSO
+Complete, Pick Best handily outperforms it with the same number of processors.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{sphere2.eps}
+  \includegraphics[width=.8\columnwidth]{sphere2}
   \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing SEPSO Pick Best with previous results.}
+  evaluations, comparing Pick Best with previous results.}
   \label{fig:sphere-pickbest}
@@
 always keeping the best speculative position found makes intuitive sense.  A
-less intuitive result occurs when using SEPSO Pick Best with the function
-Griewank.  Griewank is a highly deceptive function prone to premature
-convergence.  Our intuition on the behavior of picking the best speculative
-child is that it would tend to prematurely converge more often than the
-original PSO.  However, as seen in \fig{griewank-pickbest}, SEPSO Pick Best
-increases the chance of finding the optimum while also reducing the number of
-iterations required to converge.  Perhaps SEPSO Pick Best encourages local
-exploitation, while the sparse communication of the Ring topology still allows
-for global exploration.
+less intuitive result occurs when using Pick Best with the function Griewank.
+Griewank is a highly deceptive function prone to premature convergence.  Our
+intuition on the behavior of picking the best speculative child is that it
+would tend to prematurely converge more often than the original PSO.  However,
+as seen in \fig{griewank-pickbest}, Pick Best increases the chance of finding
+the optimum while also reducing the number of iterations required to converge.
+Perhaps Pick Best encourages local exploitation, while the sparse communication
+of the Ring topology still allows for global exploration.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank4.eps}
+  \includegraphics[width=.8\columnwidth]{griewank4}
   \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing SEPSO Pick Best with previous results.}
+  evaluations, comparing Pick Best with previous results.}
   \label{fig:griewank-pickbest}
@@
 allows results in better results when picking the best speculative child.  
-When rejecting all speculative children, which we with tongue in cheek call
-Social Promotion, no real improvement is seen.
+When rejecting all speculative children no real improvement is seen.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{sphere3.eps}
+  \includegraphics[width=.8\columnwidth]{sphere3}
   \caption{Function Sphere with a swarm that uses 240 processors per round of

### 2010-03-22 honors_thesis/journal_paper/paper.tex
started notation and alg updates

@@
 \providecommand{\neigh}{\ensuremath{N}}
+\providecommand{\leftind}{\ensuremath{L}}
+\providecommand{\rightind}{\ensuremath{R}}
 \providecommand{\nURand}{\ensuremath{U^\neigh}}
@@
 \providecommand{\specval}{\ensuremath{FX}}
-\providecommand{\ln}{\ensuremath{ln}}
-\providecommand{\rn}{\ensuremath{rn}}
+\providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
+\providecommand{\rightn}{\ensuremath{\Vec{x}^\rightind}}
 \providecommand{\caseset}{\ensuremath{\mathcal{C}}}
 \providecommand{\casedef}{\ensuremath{(\pbest:\nbest)}}
-\providecommand{\casexn}{\ensuremath{(x:\neigh*)}}
+\providecommand{\casexn}{\ensuremath{(x:\neigh)}}
 \providecommand{\casexx}{\ensuremath{(x:x)}}
-\providecommand{\casexl}{\ensuremath{(x:\ln)}}
-\providecommand{\casexr}{\ensuremath{(x:\rn)}}
-\providecommand{\casepn}{\ensuremath{(\pers*:\neigh*)}}
-\providecommand{\casepl}{\ensuremath{(\pers*:\ln)}}
-\providecommand{\casepr}{\ensuremath{(\pers*:\rn)}}
+\providecommand{\casexl}{\ensuremath{(x:\leftind)}}
+\providecommand{\casexr}{\ensuremath{(x:\rightind)}}
+\providecommand{\casepn}{\ensuremath{(\pers:\neigh)}}
+\providecommand{\casepl}{\ensuremath{(\pers:\leftind)}}
+\providecommand{\casepr}{\ensuremath{(\pers:\rightind)}}
 \providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
@@
 \begin{align}
+\nonumber
+	\pvel_{i,t+1} &=
+		\constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) \\
 \label{eq:velupdate}
-	\pvel_{i,t+1} &=
-		\constriction \left[ \pvel_t +
-			\pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) +
-			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
-		\right] \\
+			& \quad \quad + \ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
+		\bigr] \\
 \label{eq:posupdate}
@@
 with two neighbors which we will call the ``right neighbor'' and ``left
-neighbor'' neighbors identified for each particle $i$ as $\rn(i)$ and $\ln(i)$
+neighbor'' neighbors identified for each particle $i$ as $\rightn$ and $\leftn$
 respectively.
@@
 \nonumber
-	\indic_{i,t+1}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
+	\indic_{i,t+1}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\leftn_{i,t+1}),\ofunc(\rightn_{i,t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
 \nonumber
@@
 \nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\ln(i),t+1}) \\
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\leftn_{i,t+1}) \\
 \nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\rn(i),t+1}) \\
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\rightn_{i,t+1}) \\
 \label{eq:deficasexn}
@@
 \nonumber
-	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1}, \ppos_{\ln(i),t+1}, \ppos_{\rn(i),t+1},
+	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1}, \leftn_{i,t+1}, \rightn_{i,t+1},
         \pbest_{i,t} , \nbest_{i,t}, \pURand_{i,t}, \nURand_{i,t})= \\
@@
 \label{eq:defpcasexn}
-	\specpos_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
+	\specpos_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\leftn_{i,t+1},\rightn_{i,t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
 \nonumber
@@
 \label{eq:deffcasexn}
-	\specval_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
+	\specval_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\leftn_{i,t+1},\rightn_{i,t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
 \nonumber
@@
 \nonumber
-& \ofunc(\specpos_{i,t+1}^{c}(\ppos_{i,t+1},\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1},
+& \ofunc(\specpos_{i,t+1}^{c}(\ppos_{i,t+1},\leftn_{i,t+1},\rightn_{i,t+1},
 \pbest_{i,t},\nbest_{i,t},\pURand_{i,t}, \nURand_{i,t})) \\
 \label{eq:val2updatelong}
-& \indic_{i,t}^{c}(\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))
+& \indic_{i,t}^{c}(\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\leftn_{i,t+1}),\ofunc(\rightn_{i,t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))
 \end{align}
@@
 
+First we define a helper function to update the personal and neighborhood bests, $\pbest$ and $\nbest$.
+This algorithm is shown in \alg{alg:bests}.
+
+\begin{algorithm}
+  \caption{``UpdateBests'': Update personal and neighborhood best values (minimizing)}
+  \label{alg:bests}
+  \begin{algorithmic}[1]
+          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
+          \STATE $\ofunc(\pbest_{i,t})\leftarrow \ofunc(\pbest_{i,t-1})$
+          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\pbest_{i,t-1})$}
+            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
+            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
+          \ENDIF
+          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
+          \STATE $\ofunc(\nbest_{i,t})\leftarrow \ofunc(\nbest_{i,t-1})$
+          \IF {$\ofunc(\fppos_{i,t})<\ofunc(\nbest_{i,t})$}
+            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
+            \STATE $\ofunc(\pbest_{i,t})\leftarrow \ofunc(\ppos_{i,t})$
+          \ENDIF
+          \IF {$\ofunc(\leftn_{i,t})<\ofunc(nbest_{i,t})$}
+            \STATE $\pbest_{i,t}\leftarrow \leftn_{i,t}$
+            \STATE $\ofunc(pbest_{i,t})\leftarrow \ofunc(\leftn_{i,t})$
+          \ENDIF
+          \IF {$\ofunc(\rightn_{i,t})<\ofunc(\nbest_{i,t})$}
+            \STATE $\pbest_{i,t}\leftarrow \rightn_{i,t}$
+            \STATE $\ofunc(\pbest_{i,t})\leftarrow \ofunc(\rightn_{i,t}$
+          \label{alg1:endifs}
+          \ENDIF
+          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
+          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
+  \end{algorithmic}
+\end{algorithm}
+
 \begin{algorithm} [h!]
@@
     \STATE t=0
-    \STATE Initialize the swarm including
-    \FOR{$i \in \swarm$}
-      \STATE {$\ppos_{i,t} \leftarrow Random position$}
-      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
-      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
-      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
-    \ENDFOR
+    \STATE Initialize the swarm
+%    \FOR{$i \in \swarm$}
+%      \STATE {$\ppos_{i,t} \leftarrow Random position$}
+%      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
+%      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
+%      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
+%    \ENDFOR
     \REPEAT
@@
           \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \label{alg1:startifs}
-          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
-          \STATE $\fpbest_{i,t}\leftarrow \fpbest_{i,t-1}$
-          \IF {$\fppos_{i,t}<\fpbest_{i,t-1}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
-          \ENDIF
-          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
-          \STATE $\fnbest_{i,t}\leftarrow \fnbest_{i,t-1}$
-          \IF {$\fppos_{i,t}<\fnbest_{i,t}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
-          \ENDIF
-          \IF {$\fppos_{\ln(i),t}<\fnbest_{i,t}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{\ln(i),t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\ln(i),t}$
-          \ENDIF
-          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\nbest_{\rn(i),t})$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{\rn(i),t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\rn(i),t}$
-          \label{alg1:endifs}
-          \ENDIF
+          \STATE UpdateBests
           \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
@@
         \ENDFOR
-    \UNTIL {$t >$ max iteration or convergence criteria met}
+    \UNTIL {$t > max \ iterations$ or convergence criteria met}
   \end{algorithmic}
@@
     \STATE Initialize the swarm including
-    \FOR{$i \in \swarm$}
-      \STATE {$\ppos_{i,t} \leftarrow Random position$}
-      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
-      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
-      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
-    \ENDFOR
+%    \FOR{$i \in \swarm$}
+%      \STATE {$\ppos_{i,t} \leftarrow Random position$}
+%      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
+%      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
+%      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
+%    \ENDFOR
     \REPEAT
@@
           \label{alg2:fatitert}
-          \STATE DISPATCH {$\fppos_{i.t} \leftarrow \ofunc(\ppos_{i,t})$} to a seperate processor
+          \STATE DISPATCH the computation of {$\ofunc(\ppos_{i.t})$} to a seperate processor
         \ENDFOR
@@
             \label{alg2:fatitertp1}
-            \STATE DISPATCH computation of $\specval_{i,t+1}^c$ to a seperate processor.
+            \STATE DISPATCH computation of $\ofunc(\specpos_{i,t+1}^c)$ to a seperate processor.
 	    Example shown in equation~\eqref{eq:deffcasexn}.
-	    \COMMENT Note that none of the computations, $\specvel_{i,t+1}^c$,
-	    $\specpos_{i,t+1}^c$ and $\specval_{i,t+1}^c$ depend on $\fppos_{i.t}$.
+	    // Note that NONE of the computations, $\specvel_{i,t+1}^c$,
+	    $\specpos_{i,t+1}^c$ and $\ofunc(\specpos_{i,t+1}^c)$ depend on $\ofunc(\ppos_{i.t})$.
           \ENDFOR
@@
           \STATE Compute $\fppos_{i,t+1}$ as shown in equation~\eqref{eq:val2update}
+	  \STATE UpdateBests
         \ENDFOR
@@
           \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
-          \STATE $\fpbest_{i,t}\leftarrow \fpbest_{i,t-1}$
-          \IF {$\fppos_{i,t}<\fpbest_{i,t-1}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
-          \ENDIF
-          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
-          \STATE $\fnbest_{i,t}\leftarrow \fnbest_{i,t-1}$
-          \IF {$\fppos_{i,t}<\fnbest_{i,t}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
-          \ENDIF
-          \IF {$\fppos_{\ln(i),t}<\fnbest_{i,t}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{\ln(i),t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\ln(i),t}$
-          \ENDIF
-          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\nbest_{\rn(i),t})$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{\rn(i),t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\rn(i),t}$
-          \ENDIF
+          \STATE UpdateBests
           \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
@@
         \ENDFOR
-    \UNTIL {$t >$ max iteration or convergence criteria met}
+    \UNTIL {$t > max \ iteration/2$ or convergence criteria met}
   \end{algorithmic}

### 2010-03-22 honors_thesis/journal_paper/paper.tex

@@
 and the time step $t$ or $t+1$ explicit in the subscript and moved other qualifiers,
-such as the personal $\pers$ and personal best $\pers*$ qualifiers to the superscript position.
+such as the personal $\pers$ qualifier to the superscript position.
 Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.

### 2010-03-23 honors_thesis/journal_paper/paper.tex
Finished making changes based on feedback.  Removed token section

--- 2010-03-18 honors_thesis/journal_paper/paper.tex
@@
 %% PSO Stuff
+\DeclareMathOperator{\URand}{U}
 \DeclareMathOperator*{\argmin}{arg\;min}
@@
 \DeclareMathOperator*{\argsup}{arg\;sup}
-\providecommand{\pers}{\ensuremath{P}}
-\providecommand{\neigh}{\ensuremath{N}}
-\providecommand{\nURand}{\ensuremath{U^\neigh}}
-\providecommand{\pURand}{\ensuremath{U^\pers}}
 \providecommand{\ppos}{\ensuremath{\Vec{x}}}
-\providecommand{\fppos}{\ensuremath{\Vec{fx}}}
 \providecommand{\pvel}{\ensuremath{\Vec{v}}}
-\providecommand{\nbest}{\ensuremath{\Vec{b}^\neigh}}
-\providecommand{\fnbest}{\ensuremath{\Vec{fb}^\neigh}}
-\providecommand{\pbest}{\ensuremath{\Vec{b}^\pers}}
-\providecommand{\fpbest}{\ensuremath{\Vec{fb}^\pers}}
+\providecommand{\nbest}{\ensuremath{\Vec{b}_N}}
+\providecommand{\pbest}{\ensuremath{\Vec{b}_P}}
 \providecommand{\constriction}{\ensuremath{\chi}}
-\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
-\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
+\providecommand{\coeff}{\ensuremath{\phi}}
 \providecommand{\obs}{\ensuremath{\Vec{\xi}}}
-\providecommand{\ofunc}{\ensuremath{f}}
-\providecommand{\swarm}{\ensuremath{swarm}}
 
 %SpecExPSO Stuff
-\providecommand{\indic}{\ensuremath{I}}
-\providecommand{\specvel}{\ensuremath{V}}
-\providecommand{\specpos}{\ensuremath{X}}
-\providecommand{\specval}{\ensuremath{FX}}
-\providecommand{\ln}{\ensuremath{ln}}
-\providecommand{\rn}{\ensuremath{rn}}
-\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
-\providecommand{\casedef}{\ensuremath{(\pbest:\nbest)}}
-\providecommand{\casexn}{\ensuremath{(x:\neigh*)}}
-\providecommand{\casexx}{\ensuremath{(x:x)}}
-\providecommand{\casexl}{\ensuremath{(x:\ln)}}
-\providecommand{\casexr}{\ensuremath{(x:\rn)}}
-\providecommand{\casepn}{\ensuremath{(\pers*:\neigh*)}}
-\providecommand{\casepl}{\ensuremath{(\pers*:\ln)}}
-\providecommand{\casepr}{\ensuremath{(\pers*:\rn)}}
 \providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
@@
 the world's computing power is increasingly coming from large clusters of
-processors.  i
-%Even in desktop machines, systems with 4 or 8 cores are
-%commonplace.  
-In order to efficiently utilize these resources for
+processors.  Even in desktop machines, systems with 4 or 8 cores are
+commonplace.  In order to efficiently utilize these resources for
 computationally intensive problems, algorithms need to run in parallel.
@@
 
-This previous parallel techniques use additional processors to increase the
+All of these parallel techniques use additional processors to increase the
 number of particles in the swarm, either by adding individual particles or by
-adding entire new sub-swarms but in all cases increasing the number of particles.  The number of iterations of PSO that the
+adding entire new subswarms.  The number of iterations of PSO that the
 algorithms can perform is thus inherently limited by the time it takes to
 evaluate the objective function---additional processors add more particles, not
-more iterations.
-
-The purpose of this paper is to consider pso parallelization strategies for
-clusters of thousands of processors and functions for which a single
-evaluation will take on the order of hours.
-Our purpose is to explore the question ``since pso is often run with
-a swarm of size of 100, what would one do if 500 or a few thousand processors were available?''
-
-For many functions there comes a point of diminishing returns with respect to adding particles.
-!!!! TODO WARNING WARNING MADE UP NUMBERS GET REAL ONES!! !!!!!
-For the well-know benchmark function Sphere and as shown in \fig{basic-sphere} increasing the swarm size from 20 to 120 has a significant effect on the 
-speed of convergence. In this case, the swam goes below $1*10^5$ in x iteration with 20 particles, but 
-achieves this same value in  y iterations with a swarms size of 120, the additional 100 particles.
-However consider the change going from 2000 particles to 2100 particles. Here the additional 100 particles
-reduce the number of iterations to achieve $1*10^5$ from x iterations to y.
-The effect is easier to see if we plot the value obtained as a function of swarm size, as show in 
-fig{sideways-sphere} increasing the swarm size from 20 to 120 has a significant effect on the 
-Other benchmark function behave similarly. As shown in fig xxxxb for ?? and xxxxc ?? (in all case 20 dimensions, average of 50 runs).
-Certainly these benchmark functions do not prove that this will be the case for all functions,
-but it does provide some evidence that for at least some functions,
-additional particles does not seem to have much benefit after some point,
-some time earlier, some time later. This issue has been commented on by others
-in the form of recommended swarms sizes for various types of functions (cite standard swarm paper).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{evals_sphere.eps}
-  \caption{Function Sphere with a swarm with various swarm sizes.}
-  \label{fig:evals-sphere}
-\end{figure}
+more iterations.  However, there are many functions for which performing more
+iterations with fewer particles is a better use of resources than performing
+fewer iterations with more particles.
 
@@
 position.  This could not be done in gradient methods, for instance, where the
-step size for sampling depends on the gradient and therefor the position and \emph{value}
-at the current position, but it
+step size for sampling depends on the gradient at the current position, but it
 has been done in simulated annealing~\cite{witte-tpds91}.
 
-% Use this later?
-% Skip this??
-% 
-% Speculative evaluation is possible in PSO because the motion of particles
-% depends only on the \emph{positions} of the particle and its best neighbor, not
-% the \emph{value} of the objective function at those positions.  We do not have
-% to wait for a particle's function evaluation to complete in order to know where
-% it might move to next, so we can compute the set of all possible next positions
-% and send them to be evaluated along with the particle's current position.  Once
-% the current position of each particle has been evaluated, we can determine
-% which of the speculatively evaluated positions matches the branch taken by the
-% particle.  We have thus evaluated two iterations at once.  This can easily get
-% unwieldy if the set of next possible positions for each particle is large, but
-% a wise choice of topology limits that set to a reasonable size.
-% 
-% But the following does need some intro
+Speculative evaluation is possible in PSO because the motion of particles
+depends only on the \emph{positions} of the particle and its best neighbor, not
+the \emph{value} of the objective function at those positions.  We do not have
+to wait for a particle's function evaluation to complete in order to know where
+it might move to next, so we can compute the set of all possible next positions
+and send them to be evaluated along with the particle's current position.  Once
+the current position of each particle has been evaluated, we can determine
+which of the speculatively evaluated positions matches the branch taken by the
+particle.  We have thus evaluated two iterations at once.  This can easily get
+unwieldy if the set of next possible positions for each particle is large, but
+a wise choice of topology limits that set to a reasonable size.
 
@@
 At each iteration of the algorithm, the position $\ppos_t$ and velocity
-$\pvel_t$ of each particle $i$ are updated as follows:
+$\pvel_t$ of each particle are updated as follows:
 \begin{align}
 \label{eq:velupdate}
-	\pvel_{i,t+1} &=
+	\pvel_{t+1} &=
 		\constriction \left[ \pvel_t +
-			\pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) +
-			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
+			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
+			\coeff_2\URand()\otimes(\nbest - \ppos_t)
 		\right] \\
 \label{eq:posupdate}
-	\ppos_{i,t+1} &= \ppos_{i,t} + \pvel_{i,t+1}
+	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
 \end{align}
-where \( \nURand_{i,t} \) and \( \nURand_{i,t} \) are vectors of independent
-and identically distributed random numbers drawn from a uniform
+where \( \URand() \) is a vector of random numbers drawn from a uniform
 distribution, the \( \otimes \) operator is an element-wise vector
@@
 current particle has seen, and $\nbest$ (called neighborhood best) is the best
-position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
-\pcoeff \), and \( \constriction \) are parameters with prescribed values
-required to ensure convergence (2.05, 2.05, and .73,
-respectively)~\cite{clerc-tec02}. Note also that we have made the particle identifier $i$
-and the time step $t$ or $t+1$ explicit in the subscript and moved other qualifiers,
-such as the personal $\pers$ and personal best $\pers*$ qualifiers to the superscript position.
-Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.
+position the neighbors of the current particle have seen~\cite{bratton-sis07}.
+\( \coeff_1 \), \( \coeff_2 \), and \( \constriction \) are parameters with
+prescribed values required to ensure convergence (2.05, 2.05, and .73,
+respectively)~\cite{clerc-tec02}.
 
@@
 functions, as would be expected by the No Free Lunch
-theorem~\cite{wolpert-tec97}.
-The two most common topologies used in the
+theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
+beyond the scope of this paper.  The two most common topologies used in the
 literature are the Ring topology and the Complete
@@
 particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
-topology, each particle has $p$ neighbors.
-% TOD USE THIS LATER!
-% In this paper we use these
-% topologies, as well as a parallel adaptation of the Complete topology, called
-% Random, that has been shown to simulate the behavior of Complete with far less
-% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
-% randomly picks two other particles to share information with at each iteration,
-% along with itself.  Thus in both the Ring and the Random topologies, all
-% particles have 3 neighbors.
+topology, each particle has $p$ neighbors.  In this paper we use these
+topologies, as well as a parallel adaptation of the Complete topology, called
+Random, that has been shown to simulate the behavior of Complete with far less
+communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
+randomly picks two other particles to share information with at each iteration,
+along with itself.  Thus in both the Ring and the Random topologies, all
+particles have 3 neighbors.
 
@@
 The PSO algorithm just described can be parallelized by having each additional
-processor add a particle or a set of particles to the swarm.
-But as we have seen in the introduction, for some
+processor add a particle or a set of particles to the swarm.  But for some
 functions, a more desirable use of resources is to speed up the algorithm,
-performing more iterations instead of adding more particles (for a fixed number of function evaluations),
-previous techniques have not addressed this issue.
+performing more iterations instead of adding more particles.  Barring
+parallelizing the evaluation of the objective function itself, which may not be
+possible, previous techniques have not addressed this issue.
+
 In order to perform additional iterations by adding additional processors, we
-take inspiration from speculative execution in processors.
-We do this by refactoring the standard PSO equations as described here.
-
-%First consider formula for the \emph{next} iteration of PSO, which is identical
-%to Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time
-%subscripts are incremented.  As noted above and
-For simplicity in this
-explanation, we restrict our discussion here to the case of a Ring topology
-with two neighbors which we will call the ``right neighbor'' and ``left
-neighbor'' neighbors identified for each particle $i$ as $\rn(i)$ and $\ln(i)$
-respectively.
-
-Next note that the computation of $\pbest$ and $\nbest$ is not seen in the
-update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so we will make
-this mathematically explicit.  This update is usually implemented as a pair of
-independent ``if``-statements, but that is a bit awkward for our purpose here
-so we instead use a set of indicator variables, one each representing each
-possible update to the \emph{pair} $\casedef$.  These indicator variables allow
-us to easily identify the cases in which updates will be needed for $\nbest$,
-$\pbest$ or both.
-
-For example, we can write $\casexn$ to identify the case where $\pbest$ will be
-updated with the better position $\ppos_{i,t}$, shown in the first positions,
-but a no new neighborhood best was found so $\nbest$ remains in the second
-position.
-
-For any topology with 2 neighbors there are 7 possible update cases listed in
-Table~\ref{tab:evals}.  This list assumes, as is typically the case, that each
-particle is also part of its own neighborhood (can cause an update to
-$\nbest$). The list also omits the case where $\nbest$ is updated by
-$\ppos_{i,t}$ but $\pbest$ was not updated, as this is logically impossible.
-$\caseset$ represents the set of all cases.
-
-For minimization, we can define this example indicator variable representing the case
-where for particle $i$ will need to update $\pbest$ with the better
-position $x_{i,t}$ for the next iteration ($t+1$), but $\nbest$ will not updated by:
-\begin{align}
-\nonumber
-	\indic_{i,t+1}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
-\nonumber
-	& 1 \ if \  \ofunc(\ppos_{i,t+1}) > \ofunc(\pbest_{i,t}) \\
-\nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{i,t+1}) \\
-\nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\ln(i),t+1}) \\
-\nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\rn(i),t+1}) \\
-\label{eq:deficasexn}
-	& 0 \ otherwise
-\end{align}
-
-For compactness in our equations we represent this without the parameters since
-they can be inferred from the subscripts on $\indic_{i,t}^{\casexn}$.  The
-indicator variables for the other update cases shown in Table~\ref{tab:evals}
-($\casepn$, $\casepl$, $\casepr$, etc.) can be defined in similar fashion but
-obviously using the corresponding boolean expression.
-
-For each cases identified in Table~\ref{tab:evals} and identified by a
-corresponding indicator variable defined above, we also provide notation for the \emph{next} velocity
-update equation; once again using $\casexn$ as an example:
-\begin{align}
-\nonumber
-	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1}, \ppos_{\ln(i),t+1}, \ppos_{\rn(i),t+1},
-        \pbest_{i,t} , \nbest_{i,t}, \pURand_{i,t}, \nURand_{i,t})= \\
-\label{eq:defvocasexn}
-		& \constriction \left[ \pvel_t +
-			\pcoeff\pURand_{i,t+1}\otimes(\pbest_{i,t+1} - \ppos_{i,t}+1) +
-			\ncoeff\nURand_{i,t+1}\otimes(\nbest_{i,t+1} - \ppos_{i,t+1})
-		\right]
-\end{align}
-
-Again we will drop the parameters for compactness and since
-they can be inferred from the subscripts on $\specvel_{i,t}^{\casexn}$.
-The update equation is taken exactly from the statdard update equations~\eqref{eq:velupdate}
-and \eqref{eq:posupdate}, but
-since it is restricted to a specific case, $\casexn$ for this example,
-it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
-$\pbest_{i,t+1}=\ppos_{i,t+1}$ and $\nbest_{i,t+1}=\nbest_{i,t+1}$ we can re-write
-$\specvel_{i,t+1}^{\casexn}$ as:
-\begin{align}
-\nonumber
-	\specvel_{i,t+1}^{\casexn} & = \\
-\label{eq:defvcasexn}
-		& \constriction \left[ \pvel_t +
-			\pcoeff\pURand_{i,t+1}\otimes(\ppos_{i,t+1} - \ppos_{i,t+1}) +
-			\ncoeff\nURand_{i,t+1}\otimes(\nbest_{i,t+1} - \ppos_{i,t+1})
-		\right]
-\end{align}
-
-For each cases identified in
-Table~\ref{tab:evals}
-and identified by a corresponding indicator variable defined above,
-we define also a position update equation with its compact form; once again using $\casexn$ as an example:
-\begin{align}
-\label{eq:defpcasexn}
-	\specpos_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
-\nonumber
-	\specpos_{i,t+1}^{\casexn} & =  \ppos_{i,t} + \specvel_{i,t+1}^{\casexn}
-\end{align}
-
-Lastly, we also define notation for the value of the objective function at these points,
-once again using $\casexn$ as an example:
-\begin{align}
-\label{eq:deffcasexn}
-	\specval_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
-\nonumber
-	\specval_{i,t+1}^{\casexn} & =  \ofunc(\specval_{i,t+1}^{\casexn}
-\end{align}
-
-With this notation we can re-write the ``next iteration'' form of the PSO motion equations as:
-\begin{align}
-\label{eq:vel2update}
-	\pvel_{i,t+1} &= \sum_{c \in \caseset} \ \specvel_{i,t+1}^{c} \ \indic_{i,t+1}^{c} \\
-\label{eq:pos2update}
-	\ppos_{i,t+1} &= \sum_{c \in \caseset} \ \specpos_{i,t+1}^{c} \ \indic_{i,t+1}^{c}
-\end{align}
-
-The value of the objective function at $\ppos_{i,t+1}$ is given by:
-\begin{align}
-\label{eq:val2update}
-	\ofunc (\ppos_{i,t+1}) = \sum_{c \in \caseset} \ \specval_{i,t+1}^{c} \ \indic_{i,t+1}^{c}
-\end{align}
-or, with $\specpos$ and $\indic$ written with the parameters which were omitted above:
-\begin{align}
-\nonumber
-\ofunc & (\ppos_{i,t+1}) = \sum_{c \in \caseset} \\
-\nonumber
-& \ofunc(\specpos_{i,t+1}^{c}(\ppos_{i,t+1},\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1},
-\pbest_{i,t},\nbest_{i,t},\pURand_{i,t}, \nURand_{i,t})) \\
-\label{eq:val2updatelong}
-& \indic_{i,t}^{c}(\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))
-\end{align}
-
-None of the $\specpos_{i,t+1}^{c} \ \ c \in\ \caseset$ depend upon $f(i,x_t)$
-or any other objective function evaluation at $t+1$.
-Thus PSO has been re-factored such that the algorithm can begin computing
-all of the objective function evaluations needed in iteration $t+1$ before $f(x_t)$ is computed.
-
-This refactoring comes a significant expense, instead of just computing
-$\ofunc(\ppos_{i,t+1})$ and then $\ofunc(\ppos_{i,t+2})$ in the usual way, this refactored PSO
-can compute $\ofunc(\ppos_{i,t+1})$ and all 
-of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time.
-Once all of the evaluations $f(\ppos_{i,t+1})$ are completed only one
-of the indicator functions $\indic_{i,t}^{c} \ \ c \in \caseset$ will have been set to one;
-the rest will be ignored, and might just as well never have been computed.
-We call these evaluated points $\specpos_{i,t+1}^{c}$ ``speculative children''
-because they may or may not be needed.
-
-The potential value of these speculative children lies in the phase used above
-``this re-factored PSO
-can compute $\ofunc(\ppos_{i,t+1})$ and all 
-of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
-Remember that our intent is to accelerate parallel PSO and that we
-are interested in objective functions that have at least non-trivial computational
-complexity.
-Thus we assume that computing
-$\ofunc(\ppos_{i,t+1})$ and all 
-of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ is the computationally difficult part.
-
-Suppose that unlimited processors were available and that the evaluation of 
-the objective function took 1 hour. Suppose also that for each particle $i$ 
-the computations
-$\ofunc(\ppos_{i,t+1})$ and all 
-of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ were sent to separate
-processors and completed in parallel.
-\emph{Given} the values 
-of $\ofunc(\ppos_{i,t+1})$ and all 
-$\ofunc(\ppos_{i,t+1}) i \in \swarm$ 
-the computation of the indicator functions $indic_{i,t}^{c} \ \ c \in \caseset$ and to complete iteration t+1
-is inconsequential relative to the computational cost of obtaining the function values. 
-In this environment all of the function evaluations needed for 2 iterations of PSO
-could be completed in just one hour where as the conventional implementation would
-require two hours for those same 2 iterations of PSO.
-
-While the algebra show that the computation of 
-equations~\eqref{eq:vel2update} and \eqref{eq:pos2update} are identical to the 
-equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} (at iteration $t+2$),
-There are some issues better seen in an algorithmic setting, in particular the use of the random
-vectors, 
-$\pURand_{i,t}$ and $\nURand_{i,t}$. As with the algebra each transformation of the algorithm is
-provably equivalent (except that the algorithms will no longer be able to do an odd number of iteration,
-since it will do two at a time). At this point we are moving work around and adding work, but the
-resultant algorithm, produces numerically identical results.
-
-\begin{algorithm} [h!]
-  \caption{Standard PSO (minimizing)}
-  \label{alg:standard}
-  \begin{algorithmic}[1]
-    \STATE t=0
-    \STATE Initialize the swarm including
-    \FOR{$i \in \swarm$}
-      \STATE {$\ppos_{i,t} \leftarrow Random position$}
-      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
-      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
-      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
-    \ENDFOR
-    \REPEAT
-        \STATE $t=t+1$
-        \FOR{$i \in \swarm$}
-          \STATE {$\fppos_{i.t} \leftarrow \ofunc(\ppos_{i,t})$}
-        \ENDFOR
-        \FOR{$i \in \swarm$}
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \label{alg1:startifs}
-          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
-          \STATE $\fpbest_{i,t}\leftarrow \fpbest_{i,t-1}$
-          \IF {$\fppos_{i,t}<\fpbest_{i,t-1}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
-          \ENDIF
-          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
-          \STATE $\fnbest_{i,t}\leftarrow \fnbest_{i,t-1}$
-          \IF {$\fppos_{i,t}<\fnbest_{i,t}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
-          \ENDIF
-          \IF {$\fppos_{\ln(i),t}<\fnbest_{i,t}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{\ln(i),t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\ln(i),t}$
-          \ENDIF
-          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\nbest_{\rn(i),t})$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{\rn(i),t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\rn(i),t}$
-          \label{alg1:endifs}
-          \ENDIF
-          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
-          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
-        \ENDFOR
-    \UNTIL {$t >$ max iteration or convergence criteria met}
-  \end{algorithmic}
-\end{algorithm}
-
-This algorithm can be transformed, will retaining it exact behavior, by
-``unrolling'' one iteration of the ``Repeat'' loop, re-writing the ``if'''s in lines
-\ref{alg1:startifs} through
-\ref{alg1:endifs} to use the indicator functions $\indic_{i,t+1}$, and
-
-\begin{algorithm} [h!]
-  \caption{Unrolled PSO}
-  \label{alg:unrolled}
-  \begin{algorithmic}[1]
-    \STATE t=0
-    \STATE Initialize the swarm including
-    \FOR{$i \in \swarm$}
-      \STATE {$\ppos_{i,t} \leftarrow Random position$}
-      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
-      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
-      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
-    \ENDFOR
-    \REPEAT
-        \STATE $t=t+1$
-        \FOR{$i \in \swarm$}
-          \label{alg2:fatitert}
-          \STATE DISPATCH {$\fppos_{i.t} \leftarrow \ofunc(\ppos_{i,t})$} to a seperate processor
-        \ENDFOR
-        \FOR{$i \in \swarm$}
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \FOR{$c \in \caseset$}
-            \STATE Compute $\specvel_{i,t+1}^c$ and $\specpos_{i,t+1}^c$.
-	    Examples shown in equations~\eqref{eq:defvcasexn} and~\eqref{eq:defpcasexn}
-            \label{alg2:fatitertp1}
-            \STATE DISPATCH computation of $\specval_{i,t+1}^c$ to a seperate processor.
-	    Example shown in equation~\eqref{eq:deffcasexn}.
-	    \COMMENT Note that none of the computations, $\specvel_{i,t+1}^c$,
-	    $\specpos_{i,t+1}^c$ and $\specval_{i,t+1}^c$ depend on $\fppos_{i.t}$.
-          \ENDFOR
-        \ENDFOR
-        \STATE WAIT for all function evaluations dispatched above to complete
-        \FOR{$i \in \swarm$}
-          \FOR{$c \in \caseset$}
-            \STATE Compute $\indic_{i,t+1}^c$
-	           Examples shown in equation~\eqref{eq:deficasexn} and~\eqref{eq:defpcasexn}
-	  \ENDFOR 
-          \STATE Compute $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:vel2update} 
-          \STATE Compute $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:pos2update}
-          \STATE Compute $\fppos_{i,t+1}$ as shown in equation~\eqref{eq:val2update}
-        \ENDFOR
-
-        \STATE $t=t+1$
-        \FOR{$i \in \swarm$}
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
-          \STATE $\fpbest_{i,t}\leftarrow \fpbest_{i,t-1}$
-          \IF {$\fppos_{i,t}<\fpbest_{i,t-1}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
-          \ENDIF
-          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
-          \STATE $\fnbest_{i,t}\leftarrow \fnbest_{i,t-1}$
-          \IF {$\fppos_{i,t}<\fnbest_{i,t}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
-          \ENDIF
-          \IF {$\fppos_{\ln(i),t}<\fnbest_{i,t}$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{\ln(i),t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\ln(i),t}$
-          \ENDIF
-          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\nbest_{\rn(i),t})$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{\rn(i),t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\rn(i),t}$
-          \ENDIF
-          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
-          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
-        \ENDFOR
-    \UNTIL {$t >$ max iteration or convergence criteria met}
-  \end{algorithmic}
-\end{algorithm}
+take inspiration from speculative execution in processors.  If we can guess
+what a particle's next position will be from its current position, we can
+evaluate both in parallel and perform two iterations of PSO at once.  Lacking
+a good way to guess a particle's next position, we can at least enumerate all
+possible next positions of the particle, evaluate all of them at once, and one
+of them will be right.
 
@@
 possible combinations of updates that can be seen by a particle with two
-neighbors other than itself, which is the case in all of the topologies we use.
-If all of these speculative positions are evaluated in parallel along with the
-original, non-speculative position of the particle at time $t-1$, then one of
-the speculative positions has to be the correct one, and two iterations have
-been evaluated at once.
+neighbors in addition to itself, which is the case in all of the topologies we
+use.  If all of these speculative positions are evaluated in parallel along
+with the original, non-speculative position of the particle at time $t-1$, then
+one of the speculative positions has to be the correct one, and two iterations
+have been evaluated at once.
 
@@
   \centering
-  \begin{tabular}{lcc}
-	Identifier&$\pbest$ updated&Source of $\nbest$ update\\
-	\hline
-	\hline
-	$\casepn$&No update&No update\\
-	\hline
-	$\casepl$&No update&Left Neighbor\\
-	\hline
-	$\casepr$&No update&Right Neighbor\\
-	\hline
-	$\casexn$&Updated&No update\\
-	\hline
-	$\casexl$&Updated&Left Neighbor\\
-	\hline
-	$\casexr$&Updated&Right Neighbor\\
-	\hline
-	$\casexx$&Updated&The particle itself\\
+  \begin{tabular}{cc}
+	$\pbest$ updated&Source of $\nbest$ update\\
+	\hline
+	\hline
+	No update&No update\\
+	\hline
+	No update&Neighbor 1\\
+	\hline
+	No update&Neighbor 2\\
+	\hline
+	Updated&No update\\
+	\hline
+	Updated&Neighbor 1\\
+	\hline
+	Updated&Neighbor 2\\
+	\hline
+	Updated&The particle itself\\
 	\hline
@@
 
-A particle at iteration $t-1$ that has been moved to iteration $t$, but whose
-position has not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its
-position has been evaluated, but it has still not yet received information from
-its neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
-updated its neighborhood best is it a complete particle at iteration $t$.  It is
-then simply denoted as $\p_t$.
+A particle at iteration $t-1$ that has been moved to iteration $t$ using
+Equations~\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has
+not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has
+been evaluated, but it has still not yet received information from its
+neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
+updated its neighborhood best is it a complete particle at iteration $t$.  It
+is then simply denoted as $\p_t$.
 
@@
 by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
-a scheme, the master generates the positions to evaluate speculatively and
-decides which one to accept for each particle.  The outline of the procedure is
-given in \alg{centralized}.
+an architecture, the master generates the positions to evaluate speculatively
+and decides which one to accept for each particle.  The outline of the
+procedure is given in \alg{centralized}.
 
@@
   \begin{algorithmic}[1]
-	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$
+	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
+	  Equations~\eqref{eq:velupdate}~and~\eqref{eq:posupdate}
 	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
@@
 	  $\pset_{t+1}$
-	\STATE Repeat
+	\STATE Repeat from Step 1 until finished
   \end{algorithmic}
@@
 speculative and original, have been evaluated and the values reported to the
-master, the master sorts out which speculative child of each particle was the
+master, the master determines which speculative child of each particle was the
 correct one.
@@
 knows only the position that it guessed, not the function value at that
-position.  It is possible that both $\p_t$ and $\s_{t+1}$ update their
-personal bests, but $\p_t$'s value is better.  For example, suppose that
-$\p_{t-1}$ has a personal best value of 3.  $\noeval{\p}_t$ is created, and
-$\noeval{\s}_{t+1}$ is moved assuming that $\p_t$ has updated its personal
-best with its position at time $t$.  Then both $\noeval{\p}_t$ and
-$\noeval{\s}_{t+1}$ are evaluated, with values 1 and 2, respectively.
-$\nonbest{\s}_{t+1}$ would think that its current position is its personal
-best, as the value it found, 2, is better than its previous personal best
-value of 3.  It needs to receive the personal best value from its parent to
-know that its personal best position $\pbest$ is actually the position of
-$\p_t$, not $\s_{t+1}$.
+position.  It is possible that both $\p_t$ and $\s_{t+1}$ update their personal
+bests, but $\p_t$'s value is better.  For example, suppose that $\p_{t-1}$ has
+a personal best value of 3, and the we are seeking to minimize the function.
+$\noeval{\p}_t$ is created, and $\noeval{\s}_{t+1}$ is moved assuming that
+$\p_t$ has updated its personal best with its position at time $t$.  Then both
+$\noeval{\p}_t$ and $\noeval{\s}_{t+1}$ are evaluated, with values 1 and 2,
+respectively.  $\nonbest{\s}_{t+1}$ would think that its current position is
+its personal best, as the value it found, 2, is better than its previous
+personal best value of 3.  It needs to receive the personal best value from its
+parent to know that its personal best position $\pbest$ is actually the
+position of $\p_t$, not $\s_{t+1}$.
 
@@
 perform one function evaluation for one particle, be it speculative or
-non-speculative.  For benchmark functions with very fast evaluations this may
-not be the best use of parallelism in PSO.  But for many real-world
-applications, the objective function takes on the order of minutes or more to
-evaluate; in such cases our choice of framework is reasonable.
+non-speculative.  We acknowledge that for benchmark functions with very fast
+evaluations this may not be the best use of parallelism in PSO.  But for many
+real-world applications, the objective function takes on the order of minutes
+or more to evaluate; in such cases our choice of framework is reasonable.
 
@@
 To use the same number of processors (and hence evaluations) for each
-algorithm, speculative evaluation needs a smaller swarm size than standard PSO.
-For each of the topologies we used, a particle has three neighbors including
-itself.  As shown in Table~\ref{tab:evals}, this results in $7$ speculative
-evaluations per particle.  With one evaluation needed for the original,
-non-speculative particle, we have $8p_s$ evaluations for every two iterations,
-where $p_s$ is the number of particles in the speculative swarm.  The extra
-evaluations done in speculative evaluation would instead be particles in
-regular PSO, so we compare swarms of size $p_s$ in speculative evaluation with
-swarms of size $8p_s$ in regular PSO.
+algorithm, speculative evaluation must use a smaller swarm size than standard
+PSO.  For each of the topologies we used, a particle has three neighbors
+including itself.  As shown in Table~\ref{tab:evals}, this results in $7$
+speculative evaluations per particle.  With one evaluation needed for the
+original, non-speculative particle, we have $8p_s$ evaluations for every two
+iterations, where $p_s$ is the number of particles in the speculative swarm.
+The extra evaluations done in speculative evaluation would instead be particles
+in regular PSO, so we compare swarms of size $p_s$ in speculative evaluation
+with swarms of size $8p_s$ in regular PSO.
 
@@
 Complete topology.  All functions have 20 dimensions, and all results shown are
-averages over 20 runs.  We use 240 processors in our experiments, so our
-speculative algorithm uses 30 particles and standard PSO uses 240 particles.
-For ease of labeling, we call our speculative algorithm SEPSO and standard PSO
-just PSO, with the topology following.
+averages over 20 runs.  We use 240 processors in our experiments (except where
+otherwise noted), so our speculative algorithm uses 30 particles and standard
+PSO uses 240 particles.  For ease of labeling, we call our speculative
+algorithm SEPSO and standard PSO just PSO, with the topology following.
 
@@
 optimizing the Griewank function, as Complete is prone to premature convergence
-on a local optima.  We show results in \fig{basic-griewank1} for swarms of size
-30 and 240 using the Ring topology. One can see in the figure that using
-speculative evaluation initially outperforms PSO, but gets stuck in a local
-optima and is eventually outperformed.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank1}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.}
-  \label{fig:basic-griewank1}
-\end{figure}
-
-\fig{basic-griewank1} is slightly deceptive, however, as with Griewank some of
-the runs found the global optimum, 0, while some did not.  A more accurate plot
-for this function is one showing what percentage of the runs have found the
-global optimum at each iteration.  \fig{basic-griewank2} shows such a plot.  A
-swarm of size 30 performing the PSO algorithm gets stuck a little less than
-half of the time, while 240 particles provides enough exploration to find the
-global optimum every time.  As would be expected, when our speculative
-algorithm finds the global optimum, it finds it about twice as fast as regular
-PSO---it is just performing regular PSO twice as fast with a smaller swarm
-size.
+on a local optimum.  Griewank has a global optimum with a value of 0, and
+sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
+showing average function value at each iteration, a more accurate plot for
+Griewank shows the percent of runs that have found the global optimum by each
+iteration.  We show results in \fig{basic-griewank2} for swarms of size 30 and
+240 using the Ring topology.  One can see in the figure that SEPSO algorithm
+finds the optimum faster than the original PSO.  However, because the swarm
+size is so small, SEPSO gets stuck a little less than half of the time.
 
@@
 requirement of exactly reproducing the behavior of the original PSO algorithm,
-we see some very impressive results.
+we see some impressive results.
 
@@
 stagnation in the literature, but only with a single particle.  As an
-interesting aside, we found that complete stagnation practically never occurs
-in PSO, though frequently large percentages of the swarm are stagnant.
-Iterations where the entire swarm is stagnant average less than 2 in 1000
-iterations, unless the swarm converges beyond machine precision.  When graphs
-of the best function value flat-line, what really has happened is a contraction
-of the particles' velocity, not stagnation in the technical sense.
+interesting aside, we found that complete stagnation (where the entire swarm is
+stagnant) practically never occurs in PSO, though frequently large percentages
+of the swarm are stagnant.  Iterations where the entire swarm is stagnant
+average less than 2 in 1000 iterations, unless the swarm converges beyond
+machine precision.  When graphs of the best function value flat-line, what
+really has happened is a contraction of the particles' velocity, not stagnation
+in the technical sense.
 
@@
 Another possibility is to pick the best child, as described in \sec{pickbest}.
-This ignores the fact that the branch might have been wrong; we do not care,
-because we just want the child that had the best value, anyway.  In almost all
-of the experiments that we ran, it turned out that picking the best child
+This ignores the fact that the branch might have been wrong; it does not
+matter, because we simply accept the child that had the best value.  In almost
+all of the experiments that we ran, it turned out that picking the best child
 performed better than keeping the particle back an iteration.
@@
 
-Because we are now only doing two speculative evaluations per particle, with
-240 processors we can have a swarm of size 80, instead of the 30 particles in
-previous experiments.
+Because we are now only doing two speculative evaluations per particle (along
+with evaluating the original particle), with 240 processors we can have a swarm
+of size 80, instead of the 30 particles in previous experiments.
 
@@
 Social Promotion does well in this case, finding the optimum every time and
-faster than the original PSO, without the need to increase the number of
-processors.
+faster than the original PSO.
 
@@
 
-When speculating more than one iteration ahead, the idea of picking the child
-that matches the correct branch, and holding back the particle if no child
-matches the branch, does not make very much sense, as correct branches can only
-be determined for the first iteration.  Thus when speculating more than one
-iteration ahead we always pick the child that has the highest value.
+When speculating more than one iteration ahead, the idea of Social Promotion
+cannot feasibly be implemented, as we can only determine correct branches for
+the first iteration.  Thus when speculating more than one iteration ahead we
+always pick the child that has the highest value.
 
@@
 more fruitful to try.  In our experiments we try just one of the countless
-possibilities, but that one turns out to work extremely well.  More work is
-needed to compare the different branching possibilities on various functions.
+possibilities, but that one turns out to work very well.  More work is needed
+to compare the different branching possibilities on various functions.
 
@@
 picking the best child are exacerbated when speculating several iterations
-ahead.  When the algorithm converges, it converges quicker than any other 
-method we tried.  However, it also got stuck and failed to find the optimum
-more than any other method.  In order for this method to be consistently useful
-on functions like Griewank, it seems that more processors are required.  The
-results are shown in \fig{griewank-manyiters}.
+ahead.  When the algorithm finds the optimum, it finds it quicker than any
+other method we tried.  However, it also got stuck and failed to find the
+optimum more than any other method.  In order for this method to be
+consistently useful on functions like Griewank, it seems that more processors
+are required.  The results are shown in \fig{griewank-manyiters}.
 
@@
   \label{fig:griewank-manyiters}
-\end{figure}
-
-\subsection{Other possibilities}
-\label{sec:other}
-
-One issue remains.  What if we really want a swarm of 50 particles, but we have
-230 processors to use?  The methods we have thus far presented do the same
-thing for every particle, and thus would require some multiple of 50
-processors---we would have 30 extra processors with nothing to do.  It might
-very well be the best idea to simply adjust the swarm size slightly, but we
-present another method that allows particles to behave differently, some doing
-more evaluations than others.
-
-If we use the concept of tokens commonly used in multi-robot stuff (see, for
-example, \dots), we can specify a number of tokens available for particles to
-use, where each token corresponds to a speculative evaluation.  Whichever
-method is used to decide which evaluations to do, they are prioritized so that
-particles with only one or two tokens perform the most valuable speculative
-evaluations.  Particles that perform well keep their tokens, and particles that
-perform poorly send one token (if they have one) to another random particle in
-the swarm.
-
-We choose to define ``perform poorly'' as a failure to update the personal best
-when the original particle's position is evaluated.  Clearly there are other
-possible methods of deciding when to give up tokens, but many of the methods
-that sound the best are difficult in distributed environments, like giving
-tokens to that particles that most recently found the globally best positions.
-The use of a central machine to decide which particles get tokens would help to
-facilitate such methods.
-
-There are thus two distinct uses for tokens.  First, they could be used to try
-to evenly spread extra evaluations due to an incompatible number of particles
-and processors.  In cursory experimentation, we found that this was not a very
-effective use of tokens.  In virtually every case we tried, it was better to
-simply change the number of particles to match the number of processors, and 
-spread the speculative evaluations evenly throughout the swarm.
-
-The other, more compelling reason to use tokens is to intelligently and
-dynamically allocate speculative evaluations to particles that are performing
-well.  This really only makes sense when speculating several iterations ahead;
-after one or two speculative evaluations, the added benefit from also
-evaluating the low-probability branches is negligible.  But by intelligently
-allocating speculative evaluations, particles that are going in good directions
-move faster by speculating farther with extra tokens, while particles that are
-not doing as well lose their tokens.  Those particles likely contribute to the
-exploration of the algorithm, and when they find good locations, they start
-keeping their tokens, switching roles to become productive particles
-themselves.  This approach is very similar in concept to other dynamic swarms,
-like the TRIBES method proposed by Clerc (cite).
-
-As in the previous section, we can only offer a very preliminary exploration of
-the possibilities of using tokens to improve the allocation of speculative
-evaluations to particles.  We used a distributed parallel PSO framework, and in
-this setting determining which particles are performing the best globally is
-difficult without extensive communication.  Thus we only present results for
-the one method of token allocation described above.
-
-The only function for which we saw a benefit when using this technique was
-Griewank.  As seen in \fig{griewank-token}, using tokens increased the chance
-of finding the optimum to 100\%, while still using far fewer iterations than
-standard PSO.  We assume that the relatively poor performance of the technique
-on other functions is due to the lack of better token allocation.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank7}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing token-based allocation with previous results.}
-  \label{fig:griewank-token}
 \end{figure}
@@
 
-We gave only a very little attention to the possibilities of token-based
-allocation of speculative evaluations to particles.  We imagine that the
-benefits of intelligently and dynamically deciding which particles get extra
-evaluations could produce very good results.  Future work could explore the use
-of a centralized allocator for extra processors or ideas in the multi-robot
-task allocation literature, such as auctions, to better assign processors to
-particles for speculative evaluations.  While such methods involve extra
-communication, with long-running function evaluations the extra overhead of
-intelligently allocating resources would likely be negligible.
+Our methods for determining which speculative evaluations to perform were
+independent of the particle; all particles performed the same number and type
+of evaluations.  Another way to allocate speculative evaluations is to somehow
+use the performance of each particle to determine how many and which extra
+evaluations it can have.  We imagine that the benefits of intelligently and
+dynamically deciding which particles get extra evaluations could produce very
+good results.  Future work could explore the use of a centralized allocator for
+extra processors or ideas in the multi-robot task allocation literature, such
+as auctions, to better assign processors to particles for speculative
+evaluations.  While such methods involve extra communication, with long-running
+function evaluations the extra overhead of intelligently allocating resources
+would likely be negligible.
 
@@
 changing each iteration, messages that processors pass to their neighbors need
-to be sent to the correct neighbors for each iteration.  In centralized
-algorithms and in distributed algorithms with several rounds of message passing
-(the first method described above), nothing needs to be changed.  However, in
-the case where processors reconstruct information about neighboring particles
-from messages, those messages need to be sent to the correct neighbors.  A
-particle cannot simply send messages to its neighbors' neighbors'
-neighbors---it needs to send messages to its iteration $t$ neighbors' iteration
-$t+1$ neighbors, and so on.  For every neighbor outward information is sent,
-the iteration also needs to be incremented, as information about neighbors'
-neighbors is used during iteration $t+1$, and information about neighbors'
-neighbors' neighbors is used to reconstruct information about iteration $t+2$.
-Also, this method of message passing again requires the use of random seeds if
-the topology is random, so that each processor computes the same neighbors for
-a particle as all other processors.
+to be sent to the correct neighbors for each iteration.  A particle cannot
+simply send messages to its neighbors' neighbors' neighbors---it needs to send
+messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
+For every neighbor outward information is sent, the iteration also needs to be
+incremented, as information about neighbors' neighbors is used during iteration
+$t+1$, and information about neighbors' neighbors' neighbors is used to
+reconstruct information about iteration $t+2$.  Also, this method of message
+passing again requires the use of random seeds if the topology is random, so
+that each processor computes the same neighbors for a particle as all other
+processors.
 

### 2010-03-28 honors_thesis/journal_paper/paper.tex
half way through fixing formulas SOME MAY BE BROKEN
half way through cleaning up new intro

--- 2010-03-22 honors_thesis/journal_paper/paper.tex
@@
 \providecommand{\indic}{\ensuremath{I}}
-\providecommand{\specvel}{\ensuremath{V}}
-\providecommand{\specpos}{\ensuremath{X}}
-\providecommand{\specval}{\ensuremath{FX}}
+\providecommand{\specvel}{\ensuremath{\vec{V}}}
+\providecommand{\specpos}{\ensuremath{\vec{X}}}
 \providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
@@
 \providecommand{\caseset}{\ensuremath{\mathcal{C}}}
-\providecommand{\casedef}{\ensuremath{(\pbest:\nbest)}}
-\providecommand{\casexn}{\ensuremath{(x:\neigh)}}
-\providecommand{\casexx}{\ensuremath{(x:x)}}
-\providecommand{\casexl}{\ensuremath{(x:\leftind)}}
-\providecommand{\casexr}{\ensuremath{(x:\rightind)}}
-\providecommand{\casepn}{\ensuremath{(\pers:\neigh)}}
-\providecommand{\casepl}{\ensuremath{(\pers:\leftind)}}
-\providecommand{\casepr}{\ensuremath{(\pers:\rightind)}}
+\providecommand{\casedef}{\ensuremath{(\pbest,\nbest)}}
+\providecommand{\casexn}{\ensuremath{(x,\neigh)}}
+\providecommand{\casexx}{\ensuremath{(x,x)}}
+\providecommand{\casexl}{\ensuremath{(x,\leftind)}}
+\providecommand{\casexr}{\ensuremath{(x,\rightind)}}
+\providecommand{\casepn}{\ensuremath{(\pers,\neigh)}}
+\providecommand{\casepl}{\ensuremath{(\pers,\leftind)}}
+\providecommand{\casepr}{\ensuremath{(\pers,\rightind)}}
 \providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
@@
 adding more particles to the swarm or by parallelizing the evaluation of the
-objective function.  However, some functions are better optimized with more
-iterations and fewer particles.  Accordingly, we take inspiration from 
-speculative execution commonly performed in modern processors and propose what
-we call speculative evaluation in PSO.  Future positions of the particles are
-speculated and evaluated in parallel with current positions, performing two
-iterations of PSO at once.
-
-We show that while speculative evaluation in PSO is an intriguing idea, its
+objective function.  However, in many cases this approach does not scale to 
+very large populations sizes (on very large clusters).
+In response to this issue, we take inspiration from 
+speculative execution commonly performed in modern processors and propose
+Speculative Evaluation in PSO (SEPSO).
+
+In SEPSO future positions of the particles are
+speculated and evaluated in parallel with the evaluation of current positions.
+We show that the set of speculative evaluations will always include the position and corresponding
+evaluation that would normally have occurred at the next iteration.
+Thus SEPSO performs two
+iterations of PSO at once, allowing it to achieve better fitness faster.
+Not only can SEPSO can achieve better fitness in the same number of iterations,
+it also reaches better fitness for the same number of total function evaluations
+in large scale parallel environments.
+
+While Speculative Evaluation in PSO is an intriguing idea, its
 usefulness is limited unless some slight modifications are made to the PSO
-algorithm.  When we relax the requirement that speculative evaluation exactly
-reproduce the behavior of the original PSO, we can make improvements such as
-using all of the extra speculative particles, instead of throwing away all but
-one of them, or speculating several iterations ahead instead of just one.
-Using these algorithmic modifications, we see dramatic improvements on a number
-of functions over simply adding additional particles to the swarm.
-
+algorithm.  By relaxing the requirement that speculative evaluation exactly
+reproduce the behavior of the original PSO, we can make better use of the information
+obtained by the speculative evaluations.
+Using these algorithmic modifications, we see dramatic improvements in the
+effectiveness of the algorithm. 
 \end{abstract}
@@
 the world's computing power is increasingly coming from large clusters of
-processors.  i
+processors.
 %Even in desktop machines, systems with 4 or 8 cores are
@@
 In order to efficiently utilize these resources for
-computationally intensive problems, algorithms need to run in parallel.
+computationally intensive problems, PSO needs to run in parallel.
 
@@
 The methods they have used include various synchronous algorithms
-\cite{belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
-asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05},
-and algorithms based on Google's MapReduce framework \cite{mcnabb-cec07}.
+\cite{mcnabb-cec07,belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
+and asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05}.
 Parallelizing the evaluation of the objective function can also be done in some
@@
 
-This previous parallel techniques use additional processors to increase the
+These previous parallel techniques use additional processors to increase the
 number of particles in the swarm, either by adding individual particles or by
-adding entire new sub-swarms but in all cases increasing the number of particles.  The number of iterations of PSO that the
+adding entire new sub-swarms but in all cases
+the number of processors never exceeds the number of particles.
+The number of iterations of PSO that the
 algorithms can perform is thus inherently limited by the time it takes to
@@
 clusters of thousands of processors and functions for which a single
-evaluation will take on the order of hours.
-Our purpose is to explore the question ``since pso is often run with
-a swarm of size of 100, what would one do if 500 or a few thousand processors were available?''
+evaluation will take at least minutes, but probably hours or perhaps days.
+Our purpose is to explore the question ``Since pso is often run with
+a swarm of size of 50-100, what would one do if a few thousand processors were available?''
 
 For many functions there comes a point of diminishing returns with respect to adding particles.
-!!!! TODO WARNING WARNING MADE UP NUMBERS GET REAL ONES!! !!!!!
-For the well-know benchmark function Sphere and as shown in \fig{basic-sphere} increasing the swarm size from 20 to 120 has a significant effect on the 
-speed of convergence. In this case, the swam goes below $1*10^5$ in x iteration with 20 particles, but 
-achieves this same value in  y iterations with a swarms size of 120, the additional 100 particles.
-However consider the change going from 2000 particles to 2100 particles. Here the additional 100 particles
-reduce the number of iterations to achieve $1*10^5$ from x iterations to y.
-The effect is easier to see if we plot the value obtained as a function of swarm size, as show in 
-fig{sideways-sphere} increasing the swarm size from 20 to 120 has a significant effect on the 
-Other benchmark function behave similarly. As shown in fig xxxxb for ?? and xxxxc ?? (in all case 20 dimensions, average of 50 runs).
-Certainly these benchmark functions do not prove that this will be the case for all functions,
-but it does provide some evidence that for at least some functions,
-additional particles does not seem to have much benefit after some point,
-some time earlier, some time later. This issue has been commented on by others
-in the form of recommended swarms sizes for various types of functions (cite standard swarm paper).
+In \fig{basic-sphere} we show the value obtained after 20,000 function evaluations (not iterations)
+as a function of swarm size
+for the well-know benchmark function Sphere (20 dimensions, reporting the average of thirty runs). 
+Increasing the swarm size from 10 to 50 has a significant effect value obtained
+however, increasing the swarm size from 100 to 200 makes the algorithm less efficient,
+reducing the progress the algorithm makes per-evaluation.
+For this reason, research recommends the use a swarm size of 50 for this function
+~\cite{bratton-sis07}.
+Thus, in at least some cases, added particles indefinitely will not yield an efficient implementation.
 
@@
 is much farther ahead than if it had just idly waited on the branch.  If it
-guesses wrong, execution restarts right where it would have been anyway.  This
-is called speculative execution or branch prediction.
+guesses wrong, execution restarts where it would have been anyway.
+This is called speculative execution or branch prediction.
 
@@
 algorithms.  When the next sampling position of the algorithm does not depend
-explicitly on the value of the objective function at the current position,
+on the value of the objective function at the current position,
 future positions can be speculated and evaluated in parallel with the current
-position.  This could not be done in gradient methods, for instance, where the
-step size for sampling depends on the gradient and therefor the position and \emph{value}
-at the current position, but it
-has been done in simulated annealing~\cite{witte-tpds91}.
+position.
+This is trivially true for optimization by random sampling, for example.
+The next random sample does not depend on the previous point nor on the value
+of the objective function at that point in any way.
+It depends only on a random number.
+There is no need to wait to see
+what objective function value was obtained at the previous point,
+evaluation of the object function at the next point can begin as soon as the next random number is drawn.
+In other optimization algorithms, the next position depends on the previous position, but not the
+value obtained there, a random walk for example.
+In this case
+the next position to be evaluated
+depends only on the previous position and a random number.
+In this case, as it was in the case of random sampling,
+the object function evaluation of the next position can be done in parallel
+because it 
+does not depend on the \emph{value} of the objective function at the previous position.
+There are an infinite number of possible next positions, but only one given the draw of the random number.
+
+In still other cases, the next position to be evaluated may depend on the current position and some
+discrete valued (binary, for example) test on the value obtained at the previous position. In such
+cases all of the next possible positions can be evaluated at the same time as the
+evaluation of the current position.
+One of these speculative positions will exactly preserve the behavior of the algorithm
+and the others are wasted.
+This approach has been applied to 
+simulated annealing where the next position depends on (1) the last position, (2) a random number
+and (3) the result of a binary test comparing the value of the objective function at the current state with the value of the objective function at the
+last position.
+In this context the next position can be computed in the case where the binary test is true
+and in the case where it is false.
+in simulated annealing to compute (1) the value of the objective function at the current position,
+(2) the value at the position if the test turns out to be true and (3) the value at the position
+if the test turns out to be false; all in parallel.
+The test can be evaluated once all three of these evaluations of the objective function are complete, thus
+computing all of the objective function evaluations
+needed for two iterations in parallel and all at once~\cite{witte-tpds91}.
+For all but the most computationally-trivial objective functions, this approach saves almost half of the
+``wall-clock'' time.
+By way of contrast, this approach can not be used in gradient methods where the
+step size for sampling typically depends on the gradient and therefor the position and \emph{value}
+at the current position. In this case there are an infinite number of possible speculative
+evaluations.
 
@@
 
-The original PSO can be reproduced exactly, two iterations at a time, using
-this technique.  However, if we relax the requirements of the algorithm, no
-longer demanding that it strictly mimic the behavior of the original PSO, we
-can introduce new speculative techniques that often out-perform PSO.  Some
-modifications that we explore are picking the speculative position with the
-best value instead of the position that matches the branch that PSO took;
-pruning some of the speculative evaluations to increase the number of particles
-in the swarm; and speculating several iterations ahead instead of just one.
-
-\sec{pso} describes the particle swarm optimization algorithm.  \sec{sepso}
-then describes how speculative evaluation can be done in parallel PSO to
+In this paper we show that the results of standard PSO can be reproduced \emph{exactly}, two iterations at a time, using
+a speculative evaluation approach like the ones described above. We prove that this approach
+produces exactly equivalent results, and demonstrate this by implementation.
+The resulting implementation runs efficiently on large clusters where the number of processors is much larger than
+a typical or reasonable number of particles. Producing better results in less ``wall-clock'' time and,
+surprisingly, with fewer total function evaluations.
+
+Furthermore, we show that if we relax the requirements of the algorithm, no
+longer demanding that it strictly reproduce the exact behavior of standard PSO, we
+can introduce new speculative techniques that often out-perform PSO.
+These variants make better use of the extra information obtained from
+the extra exploration made by the speculative function evaluations.
+We also explore the idea that, like branch prediction in processors, we
+need not speculatively evaluate \emph{all} possible future positions,
+We can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
+By pruning the speculation to just paths that are statistically likely to reproduce pso we can also 
+speculating several iterations ahead instead of just one.
+We also consider several recovery strategies for cases where
+the pruned set of speculative evaluations does not contain the evaluation that standard pso would have done.
+
+The balance of this paper is organized as follows:
+\sec{pso} describes the particle swarm optimization algorithm.
+\sec{sepso}
+describes how speculative evaluation can be done in parallel PSO to
 perform two iterations at once, then presents a comparison of the speculative
@@
 birds and other animals. It is a social algorithm that depends on interaction
-between particles to quickly and consistently find the optimal solution to a
+between particles to quickly and consistently approximate the optimal solution to a
 given objective function.
@@
 
-At each iteration of the algorithm, the position $\ppos_t$ and velocity
-$\pvel_t$ of each particle $i$ are updated as follows:
+At each iteration ($t+1$) of the algorithm, the position $\ppos_t$ and velocity
+$\pvel_t$ of each particle are updated as follows:
 \begin{align}
 \nonumber
-	\pvel_{i,t+1} &=
+	\pvel_{t+1} &=
 		\constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) \\
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
 \label{eq:velupdate}
-			& \quad \quad + \ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
+			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
 		\bigr] \\
 \label{eq:posupdate}
-	\ppos_{i,t+1} &= \ppos_{i,t} + \pvel_{i,t+1}
+	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
 \end{align}
-where \( \nURand_{i,t} \) and \( \nURand_{i,t} \) are vectors of independent
+where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent
 and identically distributed random numbers drawn from a uniform
@@
 required to ensure convergence (2.05, 2.05, and .73,
-respectively)~\cite{clerc-tec02}. Note also that we have made the particle identifier $i$
-and the time step $t$ or $t+1$ explicit in the subscript and moved other qualifiers,
+respectively)~\cite{clerc-tec02}. Note also that we use a subscript for the 
+time step $t$ or $t+1$ and have moved all other qualifiers,
 such as the personal $\pers$ qualifier to the superscript position.
@@
 
-There are many ways of defining the neighbors of a particle.  Different
-neighborhood topologies have varying performance on different kinds of
-functions, as would be expected by the No Free Lunch
+There are many ways of defining the neighbors of a particle. 
+Changing the
+neighborhood topology has a significant effect on the performance of the algorithm and 
+this effect varies widely by function type, as would be expected by the No Free Lunch
 theorem~\cite{wolpert-tec97}.
@@
 particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
-particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
+particle has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
 topology, each particle has $p$ neighbors.
@@
 
-The PSO algorithm just described can be parallelized by having each additional
-processor add a particle or a set of particles to the swarm.
+The PSO algorithm can be trivially parallelized by distributing the processing needed for each
+particle on a separate processor.
 But as we have seen in the introduction, for some
-functions, a more desirable use of resources is to speed up the algorithm,
-performing more iterations instead of adding more particles (for a fixed number of function evaluations),
+functions, and for large numbers of processors where just adding particles (processors)
+reaches a point of diminishing returns.
+That is, adding processors does not help us reach any given level of fitness appreciably faster.
+
+To address this issue we present here a very different use of additional processors.
+Performing more iterations instead of adding more particles (for a fixed number of function evaluations),
 previous techniques have not addressed this issue.
@@
 with two neighbors which we will call the ``right neighbor'' and ``left
-neighbor'' neighbors identified for each particle $i$ as $\rightn$ and $\leftn$
+neighbor'' neighbors identified for each particle as $\rightn$ and $\leftn$
 respectively.
@@
 this mathematically explicit.  This update is usually implemented as a pair of
-independent ``if``-statements, but that is a bit awkward for our purpose here
+independent ``if``-statements, but that is a bit awkward for our purposes
 so we instead use a set of indicator variables, one each representing each
@@
 $\nbest$). The list also omits the case where $\nbest$ is updated by
-$\ppos_{i,t}$ but $\pbest$ was not updated, as this is logically impossible.
+$\ppos_{t}$ but $\pbest$ was not updated, as this is logically impossible.
 $\caseset$ represents the set of all cases.
@@
 For minimization, we can define this example indicator variable representing the case
-where for particle $i$ will need to update $\pbest$ with the better
-position $x_{i,t}$ for the next iteration ($t+1$), but $\nbest$ will not updated by:
+where the particle will need to update $\pbest$ with the better
+position $x_{t}$ for the next iteration ($t+1$), but $\nbest$ will not updated by:
 \begin{align}
 \nonumber
-	\indic_{i,t+1}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\leftn_{i,t+1}),\ofunc(\rightn_{i,t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
+	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))= \\
 \nonumber
-	& 1 \ if \  \ofunc(\ppos_{i,t+1}) > \ofunc(\pbest_{i,t}) \\
+	& 1 \ if \  \ofunc(\ppos_{t+1}) > \ofunc(\pbest_{t}) \\
 \nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{i,t+1}) \\
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\ppos_{t+1}) \\
 \nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\leftn_{i,t+1}) \\
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\leftn_{t+1}) \\
 \nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\rightn_{i,t+1}) \\
+	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\rightn_{t+1}) \\
 \label{eq:deficasexn}
@@
 For compactness in our equations we represent this without the parameters since
-they can be inferred from the subscripts on $\indic_{i,t}^{\casexn}$.  The
+they can be inferred from the subscripts on $\indic_{t}^{\casexn}$.  The
 indicator variables for the other update cases shown in Table~\ref{tab:evals}
@@
 \nonumber
-	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1}, \leftn_{i,t+1}, \rightn_{i,t+1},
-        \pbest_{i,t} , \nbest_{i,t}, \pURand_{i,t}, \nURand_{i,t})= \\
+	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t+1}, \leftn_{t+1}, \rightn_{t+1}, \pbest_{t},
+        \nbest_{t}, \pURand_{t}, \nURand_{t}) \\
 \label{eq:defvocasexn}
-		& \constriction \left[ \pvel_t +
-			\pcoeff\pURand_{i,t+1}\otimes(\pbest_{i,t+1} - \ppos_{i,t}+1) +
-			\ncoeff\nURand_{i,t+1}\otimes(\nbest_{i,t+1} - \ppos_{i,t+1})
-		\right]
+		&= \constriction \bigl[ \pvel_{t} +
+			\pcoeff\pURand_{t+1}\otimes(\pbest_{t+1} - \ppos_{t+1}) \\
+		& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t+1} - \ppos_{t+1})
+		\bigr]
 \end{align}
@@
 Again we will drop the parameters for compactness and since
-they can be inferred from the subscripts on $\specvel_{i,t}^{\casexn}$.
-The update equation is taken exactly from the statdard update equations~\eqref{eq:velupdate}
+they can be inferred from the subscripts on $\specvel_{t}^{\casexn}$.
+The update equation is taken exactly from the standard update equations~\eqref{eq:velupdate}
 and \eqref{eq:posupdate}, but
@@
 it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
-$\pbest_{i,t+1}=\ppos_{i,t+1}$ and $\nbest_{i,t+1}=\nbest_{i,t+1}$ we can re-write
-$\specvel_{i,t+1}^{\casexn}$ as:
+$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t+1}$ we can re-write
+$\specvel_{t+1}^{\casexn}$ as:
 \begin{align}
 \nonumber
-	\specvel_{i,t+1}^{\casexn} & = \\
+	\specvel_{t+1}^{\casexn} &= \constriction \bigl[ \pvel_{t} +
+			\pcoeff\pURand_{t+1}\otimes(\ppos_{t+1} - \ppos_{t+1}) \\
 \label{eq:defvcasexn}
-		& \constriction \left[ \pvel_t +
-			\pcoeff\pURand_{i,t+1}\otimes(\ppos_{i,t+1} - \ppos_{i,t+1}) +
-			\ncoeff\nURand_{i,t+1}\otimes(\nbest_{i,t+1} - \ppos_{i,t+1})
-		\right]
+			& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t+1} - \ppos_{t+1})
+		\bigr]
 \end{align}
@@
 \label{eq:defpcasexn}
-	\specpos_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\leftn_{i,t+1},\rightn_{i,t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
+	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t},\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t})= \\
 \nonumber
-	\specpos_{i,t+1}^{\casexn} & =  \ppos_{i,t} + \specvel_{i,t+1}^{\casexn}
+	& = \specpos_{t+1}^{\casexn} =  \ppos_{t} + \specvel_{t+1}^{\casexn}
 \end{align}
 
-Lastly, we also define notation for the value of the objective function at these points,
-once again using $\casexn$ as an example:
+Lastly, we use the normal notation for the value of the objective function at these points,
+once again using the case $\casexn$ as an example:
 \begin{align}
+\nonumber
+	\ofunc(\specpos_{t+1}^{\casexn} & (\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t}))= \\
 \label{eq:deffcasexn}
-	\specval_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\leftn_{i,t+1},\rightn_{i,t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
+	& =  \ofunc(\specpos_{t+1}^{\casexn})
+\end{align}
+
+With this notation we can re-write the ``next iteration'' form of the original PSO motion equations
+given in equations~\eqref{eq:velupdate}
+and \eqref{eq:posupdate} as:
+\begin{align}
 \nonumber
-	\specval_{i,t+1}^{\casexn} & =  \ofunc(\specval_{i,t+1}^{\casexn}
+	\pvel_{t+1} &=
+		\constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
+\nonumber
+			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
+		\bigr] \\
+\nonumber
+	&= \sum_{c \in \caseset} \constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
+\nonumber
+			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
+		\bigr] \indic_{t+1}^{c} \\
+\label{eq:vel2update}
+	&= \sum_{c \in \caseset} \ \specvel_{t+1}^{c} \ \indic_{t+1}^{c}
 \end{align}
-
-With this notation we can re-write the ``next iteration'' form of the PSO motion equations as:
+and
 \begin{align}
-\label{eq:vel2update}
-	\pvel_{i,t+1} &= \sum_{c \in \caseset} \ \specvel_{i,t+1}^{c} \ \indic_{i,t+1}^{c} \\
+\nonumber
+	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
 \label{eq:pos2update}
-	\ppos_{i,t+1} &= \sum_{c \in \caseset} \ \specpos_{i,t+1}^{c} \ \indic_{i,t+1}^{c}
+	&= \sum_{c \in \caseset} \ \specpos_{t+1}^{c} \ \indic_{t+1}^{c}
 \end{align}
 
-The value of the objective function at $\ppos_{i,t+1}$ is given by:
+The value of the objective function at $\ppos_{t+1}$ is given by:
 \begin{align}
 \label{eq:val2update}
-	\ofunc (\ppos_{i,t+1}) = \sum_{c \in \caseset} \ \specval_{i,t+1}^{c} \ \indic_{i,t+1}^{c}
+	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \ofunc(\specpos_{t+1}^{c}) \ \indic_{t+1}^{c}
 \end{align}
@@
 \nonumber
-\ofunc & (\ppos_{i,t+1}) = \sum_{c \in \caseset} \\
+\ofunc & (\ppos_{t+1}) = \sum_{c \in \caseset} \\
 \nonumber
-& \ofunc(\specpos_{i,t+1}^{c}(\ppos_{i,t+1},\leftn_{i,t+1},\rightn_{i,t+1},
-\pbest_{i,t},\nbest_{i,t},\pURand_{i,t}, \nURand_{i,t})) \\
+& \ofunc(\specpos_{t+1}^{c}(\ppos_{t+1},\leftn_{t+1},\rightn_{t+1},
+\pbest_{t},\nbest_{t},\pURand_{t}, \nURand_{t})) \\
 \label{eq:val2updatelong}
-& \indic_{i,t}^{c}(\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\leftn_{i,t+1}),\ofunc(\rightn_{i,t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))
+& \indic_{t}^{c}(\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))
 \end{align}
 
-None of the $\specpos_{i,t+1}^{c} \ \ c \in\ \caseset$ depend upon $f(i,x_t)$
+In this form the important point to notice is that none of the $\ofunc(\specpos_{t+1}^{c}) \ c \in\ \caseset$ depend upon $f(x_t)$
 or any other objective function evaluation at $t+1$.
@@
 This refactoring comes a significant expense, instead of just computing
-$\ofunc(\ppos_{i,t+1})$ and then $\ofunc(\ppos_{i,t+2})$ in the usual way, this refactored PSO
-can compute $\ofunc(\ppos_{i,t+1})$ and all 
-of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time.
-Once all of the evaluations $f(\ppos_{i,t+1})$ are completed only one
-of the indicator functions $\indic_{i,t}^{c} \ \ c \in \caseset$ will have been set to one;
-the rest will be ignored, and might just as well never have been computed.
-We call these evaluated points $\specpos_{i,t+1}^{c}$ ``speculative children''
+$\ofunc(\ppos_{i,t})$ and then $\ofunc(\ppos_{i,t+1})$ in the usual way, this refactored PSO
+computes $\ofunc(\ppos_{t+1})$ and all 
+of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time.
+Once all of the evaluations $f(\ppos_{t+1})$ are completed only one
+of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ will be set to one.
+We will refer to the indicator function which is set to one as the
+\emph{behavior preserving indicator} or $bp$, since it indicates which of
+the computations $\ofunc(\specpos_{t}^{c}) \  c \in \caseset$ preservers the behavior of PSO.
+The rest 
+computations $\ofunc(\specpos_{t}^{c}) \ c \in \caseset-\{bp\}$
+will be ignored, and might just as well never have been computed.
+We call these evaluations of $\ofunc(\specpos_{t+1}^{c})$ ``speculative children''
 because they may or may not be needed.
 
-The potential value of these speculative children lies in the phase used above
+The potential value of these speculative children lies in the phrase used above
 ``this re-factored PSO
-can compute $\ofunc(\ppos_{i,t+1})$ and all 
-of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
+can compute $\ofunc(\ppos_{t+1})$ and all 
+of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
 Remember that our intent is to accelerate parallel PSO and that we
@@
 Thus we assume that computing
-$\ofunc(\ppos_{i,t+1})$ and all 
-of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ is the computationally difficult part.
+$\ofunc(\ppos_{t+1})$ and all 
+of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ is the computationally difficult part of PSO.
 
 Suppose that unlimited processors were available and that the evaluation of 
-the objective function took 1 hour. Suppose also that for each particle $i$ 
+the objective function took 1 hour. Suppose also that for each particle
 the computations
-$\ofunc(\ppos_{i,t+1})$ and all 
-of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ were sent to separate
+$\ofunc(\ppos_{t+1})$ and all 
+of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ were sent to separate
 processors and completed in parallel.
 \emph{Given} the values 
-of $\ofunc(\ppos_{i,t+1})$ and all 
-$\ofunc(\ppos_{i,t+1}) i \in \swarm$ 
-the computation of the indicator functions $indic_{i,t}^{c} \ \ c \in \caseset$ and to complete iteration t+1
+of $\ofunc(\ppos_{t+1})$ and all 
+$\ofunc(\ppos_{t+1})$ 
+the computation of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ and to complete iteration t+1
 is inconsequential relative to the computational cost of obtaining the function values. 
@@
 equations~\eqref{eq:vel2update} and \eqref{eq:pos2update} are identical to the 
-equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} (at iteration $t+2$),
+equations~\eqref{eq:velupdate} and \eqref{eq:posupdate},
 There are some issues better seen in an algorithmic setting, in particular the use of the random
 vectors, 
-$\pURand_{i,t}$ and $\nURand_{i,t}$. As with the algebra each transformation of the algorithm is
-provably equivalent (except that the algorithms will no longer be able to do an odd number of iteration,
-since it will do two at a time). At this point we are moving work around and adding work, but the
+$\pURand_{t}$ and $\nURand_{t}$. As with the algebra given above, each transformation of the algorithm is
+provably equivalent (except that the algorithms will no longer be able to do an odd number of iterations,
+since it will do two at a time). In the algorithmic transformations given below,
+instructions are moved around and added, but the
 resultant algorithm, produces numerically identical results.
 
-First we define a helper function to update the personal and neighborhood bests, $\pbest$ and $\nbest$.
-This algorithm is shown in \alg{alg:bests}.
+We first define a helper function to update the personal and neighborhood bests, $\pbest$ and $\nbest$.
+This algorithm is shown in~\alg{alg:bests}.
 
@@
   \begin{algorithmic}[1]
-          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
-          \STATE $\ofunc(\pbest_{i,t})\leftarrow \ofunc(\pbest_{i,t-1})$
-          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\pbest_{i,t-1})$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
+          \STATE $\pbest_{t}\leftarrow \pbest_{t-1}$
+          \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\pbest_{t-1})$
+          \IF {$\ofunc(\ppos_{t})<\ofunc(\pbest_{t-1})$}
+            \STATE $\pbest_{t}\leftarrow \ppos_{t}$
+            \STATE $\fpbest_{t}\leftarrow \fppos_{t}$
           \ENDIF
-          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
-          \STATE $\ofunc(\nbest_{i,t})\leftarrow \ofunc(\nbest_{i,t-1})$
-          \IF {$\ofunc(\fppos_{i,t})<\ofunc(\nbest_{i,t})$}
-            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
-            \STATE $\ofunc(\pbest_{i,t})\leftarrow \ofunc(\ppos_{i,t})$
+          \STATE $\nbest_{t}\leftarrow \nbest_{t-1}$
+          \STATE $\ofunc(\nbest_{t})\leftarrow \ofunc(\nbest_{t-1})$
+          \IF {$\ofunc(\fppos_{t})<\ofunc(\nbest_{t})$}
+            \STATE $\pbest_{t}\leftarrow \ppos_{t}$
+            \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\ppos_{t})$
           \ENDIF
-          \IF {$\ofunc(\leftn_{i,t})<\ofunc(nbest_{i,t})$}
-            \STATE $\pbest_{i,t}\leftarrow \leftn_{i,t}$
-            \STATE $\ofunc(pbest_{i,t})\leftarrow \ofunc(\leftn_{i,t})$
+          \IF {$\ofunc(\leftn_{t})<\ofunc(nbest_{t})$}
+            \STATE $\pbest_{t}\leftarrow \leftn_{t}$
+            \STATE $\ofunc(pbest_{t})\leftarrow \ofunc(\leftn_{t})$
           \ENDIF
-          \IF {$\ofunc(\rightn_{i,t})<\ofunc(\nbest_{i,t})$}
-            \STATE $\pbest_{i,t}\leftarrow \rightn_{i,t}$
-            \STATE $\ofunc(\pbest_{i,t})\leftarrow \ofunc(\rightn_{i,t}$
+          \IF {$\ofunc(\rightn_{t})<\ofunc(\nbest_{t})$}
+            \STATE $\pbest_{t}\leftarrow \rightn_{t}$
+            \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\rightn_{t}$
           \label{alg1:endifs}
           \ENDIF
-          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
-          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
+          \STATE Update $\pvel_{t+1}$ as shown in equation~\eqref{eq:velupdate} 
+          \STATE Update $\ppos_{t+1}$ as shown in equation~\eqref{eq:posupdate}
   \end{algorithmic}
@@
         \FOR{$i \in \swarm$}
-          \STATE {$\fppos_{i.t} \leftarrow \ofunc(\ppos_{i,t})$}
+          \STATE {$\fppos_{t} \leftarrow \ofunc(\ppos_{t})$}
         \ENDFOR
         \FOR{$i \in \swarm$}
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
+          \STATE $\nURand_{t} \Leftarrow RandomVector()$
+          \STATE $\nURand_{t} \Leftarrow RandomVector()$
           \STATE UpdateBests
-          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
-          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
+          \STATE Update $\pvel_{t+1}$ as shown in equation~\eqref{eq:velupdate} 
+          \STATE Update $\ppos_{t+1}$ as shown in equation~\eqref{eq:posupdate}
         \ENDFOR
@@
 
-This algorithm can be transformed, will retaining it exact behavior, by
-``unrolling'' one iteration of the ``Repeat'' loop, re-writing the ``if'''s in lines
+Using this helper function we can write PSO as shown in~\alg{alg:standard}.
+This algorithm can be transformed, while retaining it exact behavior, by
+``unrolling'' one iteration of the ``Repeat'' loop, redundantly expressing work done in the ``if'''s in lines
 \ref{alg1:startifs} through
-\ref{alg1:endifs} to use the indicator functions $\indic_{i,t+1}$, and
+\ref{alg1:endifs} to use the indicator functions $\indic_{t+1}$ and re-writing the computation of the next velocity,
+positions and objective function value using 
+equations~\eqref{eq:vel2update}, \eqref{eq:pos2update} and \eqref{eq:val2update} 
+as shown in~\alg{alg:unrolled}
 
@@
           \label{alg2:fatitert}
-          \STATE DISPATCH the computation of {$\ofunc(\ppos_{i.t})$} to a seperate processor
+          \STATE DISPATCH the computation of {$\ofunc(\ppos_{t})$} to a seperate processor
         \ENDFOR
         \FOR{$i \in \swarm$}
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
+          \STATE $\nURand_{t} \Leftarrow RandomVector()$
+          \STATE $\nURand_{t} \Leftarrow RandomVector()$
           \FOR{$c \in \caseset$}
-            \STATE Compute $\specvel_{i,t+1}^c$ and $\specpos_{i,t+1}^c$.
+            \STATE Compute $\specvel_{t+1}^c$ and $\specpos_{t+1}^c$.
 	    Examples shown in equations~\eqref{eq:defvcasexn} and~\eqref{eq:defpcasexn}
             \label{alg2:fatitertp1}
-            \STATE DISPATCH computation of $\ofunc(\specpos_{i,t+1}^c)$ to a seperate processor.
+            \STATE DISPATCH computation of $\ofunc(\specpos_{t+1}^c)$ to a seperate processor.
 	    Example shown in equation~\eqref{eq:deffcasexn}.
-	    // Note that NONE of the computations, $\specvel_{i,t+1}^c$,
-	    $\specpos_{i,t+1}^c$ and $\ofunc(\specpos_{i,t+1}^c)$ depend on $\ofunc(\ppos_{i.t})$.
+	    // Note that NONE of the computations, $\specvel_{t+1}^c$,
+	    $\specpos_{t+1}^c$ and $\ofunc(\specpos_{t+1}^c)$ depend on $\ofunc(\ppos_{t})$.
           \ENDFOR
@@
           \FOR{$c \in \caseset$}
-            \STATE Compute $\indic_{i,t+1}^c$
+            \STATE Compute $\indic_{t+1}^c$
 	           Examples shown in equation~\eqref{eq:deficasexn} and~\eqref{eq:defpcasexn}
 	  \ENDFOR 
-          \STATE Compute $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:vel2update} 
-          \STATE Compute $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:pos2update}
-          \STATE Compute $\fppos_{i,t+1}$ as shown in equation~\eqref{eq:val2update}
+            \label{alg2:startcut}
+          \STATE Compute $\pvel_{t+1}$ as shown in equation~\eqref{eq:vel2update} 
+          \STATE Compute $\ppos_{t+1}$ as shown in equation~\eqref{eq:pos2update}
+          \STATE Compute $\fppos_{t+1}$ as shown in equation~\eqref{eq:val2update}
 	  \STATE UpdateBests
+            \label{alg2:endcut}
         \ENDFOR
@@
         \FOR{$i \in \swarm$}
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
+          \STATE $\nURand_{t} \Leftarrow RandomVector()$
+          \STATE $\nURand_{t} \Leftarrow RandomVector()$
           \STATE UpdateBests
-          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
-          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
+          \STATE Update $\pvel_{t+1}$ as shown in equation~\eqref{eq:velupdate} 
+          \STATE Update $\ppos_{t+1}$ as shown in equation~\eqref{eq:posupdate}
         \ENDFOR
@@
 \end{algorithm}
+
+In this form we can easily see that the random values
+$\pURand_{t}$ and $\nURand_{t}$ usually used just for the computations in
+Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} are passed to all of the
+speculative children, but only the behavior preserving computation will
+pass through the  computations in lines
+\ref{alg2:startcut} through
+\ref{alg2:endcut}. The fact that these values are also used by other speculative
+children will not change the behavior preserving case.
 

### 2010-03-29 honors_thesis/journal_paper/paper.tex
finished second pass

@@
 
-The purpose of this paper is to consider pso parallelization strategies for
+The purpose of this paper is to consider PSO parallelization strategies for
 clusters of thousands of processors and functions for which a single
 evaluation will take at least minutes, but probably hours or perhaps days.
-Our purpose is to explore the question ``Since pso is often run with
+Our purpose is to explore the question ``Since PSO is often run with
 a swarm of size of 50-100, what would one do if a few thousand processors were available?''
@@
 In still other cases, the next position to be evaluated may depend on the current position and some
-discrete valued (binary, for example) test on the value obtained at the previous position. In such
-cases all of the next possible positions can be evaluated at the same time as the
-evaluation of the current position.
-One of these speculative positions will exactly preserve the behavior of the algorithm
-and the others are wasted.
-This approach has been applied to 
-simulated annealing where the next position depends on (1) the last position, (2) a random number
-and (3) the result of a binary test comparing the value of the objective function at the current state with the value of the objective function at the
-last position.
-In this context the next position can be computed in the case where the binary test is true
-and in the case where it is false.
-in simulated annealing to compute (1) the value of the objective function at the current position,
+discrete valued (binary, for example) test on the value obtained at the current position.
+For example, consider a greedy random walk, in which the state only advances if a better
+position is found.
+That is, if the current value is better than the previous value,
+use the sum of the current position and the random number as the next position.
+Otherwise, computer the next position using the sum of the previous position and the random number.
+In this case the objective function can be evaluated at three positions:
+(1) the value of the objective function at the current position,
 (2) the value at the position if the test turns out to be true and (3) the value at the position
 if the test turns out to be false; all in parallel.
-The test can be evaluated once all three of these evaluations of the objective function are complete, thus
-computing all of the objective function evaluations
-needed for two iterations in parallel and all at once~\cite{witte-tpds91}.
+In this example it is important to note that we do not need the value of the objectve function at the current
+point to to reduce the infinite number of possible next postions to just two alternatives.
+Given the random sample, the next point detments only on the binary test and the
+random sample is, by definition, idependent of the positions, and the vaues obtained.
+The choice of the next position can be made post-hoc, after the value at the current position is obtained.
 For all but the most computationally-trivial objective functions, this approach saves almost half of the
 ``wall-clock'' time.
-By way of contrast, this approach can not be used in gradient methods where the
-step size for sampling typically depends on the gradient and therefor the position and \emph{value}
-at the current position. In this case there are an infinite number of possible speculative
-evaluations.
+This approach has been applied to simulated annealing~\cite{witte-tpds91} which like the greedy random walk,
+depends on positions and a binary test.
+
+This approach can not be used in gradient methods where the
+step size for sampling typically depends on the gradient and therefor the position and 
+a set of infinitly many posible \emph{values} outcomes
+obtained at the current position (not binary or even finite).
 
@@
 need not speculatively evaluate \emph{all} possible future positions,
-We can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
-By pruning the speculation to just paths that are statistically likely to reproduce pso we can also 
+we can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
+By pruning the speculation to just paths that are statistically likely to reproduce the paths that are equivelent to PSO we can also 
 speculating several iterations ahead instead of just one.
 We also consider several recovery strategies for cases where
-the pruned set of speculative evaluations does not contain the evaluation that standard pso would have done.
+the pruned set of speculative evaluations does not contain the evaluation that standard PSO would have done.
 
@@
 time step $t$ or $t+1$ and have moved all other qualifiers,
-such as the personal $\pers$ qualifier to the superscript position.
+such as the personal qualifier ``$\pers$'' to the superscript position.
 Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.
@@
 topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
-neighbor to either side of it; in the Complete topology, every particle is a
+or more neighbors to either side of it; in the Complete topology, every particle is a
 neighbor to every other particle.  In all topologies, a particle is also a
-neighbor to itself, in that its own position is considered when updating the
-particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
+neighbor to itself in that its own position is considered when updating the
+particle's personal best, $\nbest$.  Thus with $p$ particles, using the Ring topology each
 particle has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
@@
 \label{sec:sepso}
-\subsection{Overview}
 
@@
 But as we have seen in the introduction, for some
-functions, and for large numbers of processors where just adding particles (processors)
+functions, and for large numbers of processors just adding particles
 reaches a point of diminishing returns.
@@
 To address this issue we present here a very different use of additional processors.
-Performing more iterations instead of adding more particles (for a fixed number of function evaluations),
-previous techniques have not addressed this issue.
-In order to perform additional iterations by adding additional processors, we
-take inspiration from speculative execution in processors.
-We do this by refactoring the standard PSO equations as described here.
+We take inspiration from speculative execution in processors and instead of adding particles
+we will employ a speculative evaluation approach like those described in the introduction. 
+This allows us to perform iterations two at a time.
+To do so we must first re-factor PSO such that the determination of the $\pbest$ and $\nbest$,
+which depend upon the value of the objective function, is sepearate from the rest of the computation.
 
@@
 $\ppos_{t}$ but $\pbest$ was not updated, as this is logically impossible.
-$\caseset$ represents the set of all cases.
+$\caseset$ represents the set of all update cases.
 
@@
 they can be inferred from the subscripts on $\specvel_{t}^{\casexn}$.
-The update equation is taken exactly from the standard update equations~\eqref{eq:velupdate}
-and \eqref{eq:posupdate}, but
+The update equation is taken exactly from the standard update equation~\eqref{eq:velupdate}, but
 since it is restricted to a specific case, $\casexn$ for this example,
 it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
-$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t+1}$ we can re-write
+$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t}$ we can re-write
 $\specvel_{t+1}^{\casexn}$ as:
@@
 \label{eq:defvcasexn}
-			& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t+1} - \ppos_{t+1})
+			& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t} - \ppos_{t+1})
 		\bigr]
@@
 \end{align}
-
-Lastly, we use the normal notation for the value of the objective function at these points,
+We call these values, $\specpos^{c}) \ c \in\ \caseset$ (7 of them in this case where each particle has 2 neighbors) as ``speculative children''
+since they will only be useful if the corresponding $\indic^{c} \ c \in \caseset$ has the value 1.
+
+Lastly, we use the normal notation for the value of the objective function for each speculative child,
 once again using the case $\casexn$ as an example:
@@
 
-With this notation we can re-write the ``next iteration'' form of the original PSO motion equations
-given in equations~\eqref{eq:velupdate}
-and \eqref{eq:posupdate} as:
+With this notation we can re-write the original PSO velocity
+equation~\eqref{eq:velupdate} as:
 \begin{align}
@@
 \end{align}
-and
+and the position update 
+equation~\eqref{eq:posupdate} as:
 \begin{align}
@@
 \end{align}
-
 The value of the objective function at $\ppos_{t+1}$ is given by:
@@
 \end{align}
-or, with $\specpos$ and $\indic$ written with the parameters which were omitted above:
+
+Returning our attention to the computtaion of $\specpos$ in \eqref{eq:pos2update}and writing it with the parameters which were omitted above,
+we obtain:
 \begin{align}
 \nonumber
-\ofunc & (\ppos_{t+1}) = \sum_{c \in \caseset} \\
+\ppos_{t+1} &= \sum_{c \in \caseset} \\
 \nonumber
-& \ofunc(\specpos_{t+1}^{c}(\ppos_{t+1},\leftn_{t+1},\rightn_{t+1},
-\pbest_{t},\nbest_{t},\pURand_{t}, \nURand_{t})) \\
+& \specpos_{t+1}^{c}(\ppos_{t+1},\leftn_{t+1},\rightn_{t+1},
+\pbest_{t},\nbest_{t},\pURand_{t}, \nURand_{t}) \\
 \label{eq:val2updatelong}
@@
 \end{align}
-
-In this form the important point to notice is that none of the $\ofunc(\specpos_{t+1}^{c}) \ c \in\ \caseset$ depend upon $f(x_t)$
-or any other objective function evaluation at $t+1$.
+In this form the important point to notice is that none of the $\specpos_{t+1}^{c} \ c \in\ \caseset$ depend upon $f(x_t)$
+or any other objective function evaluation at $t$.
 Thus PSO has been re-factored such that the algorithm can begin computing
-all of the objective function evaluations needed in iteration $t+1$ before $f(x_t)$ is computed.
+all 7 of the objective function evaluations potentially needed in iteration $t+1$ before $f(x_t)$ is computed.
 
@@
 $\ofunc(\ppos_{i,t})$ and then $\ofunc(\ppos_{i,t+1})$ in the usual way, this refactored PSO
-computes $\ofunc(\ppos_{t+1})$ and all 
+computes $\ofunc(\ppos_{t})$ and all 
 of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time.
-Once all of the evaluations $f(\ppos_{t+1})$ are completed only one
+Once the evaluation of $f(\ppos_{t})$ is completed only one
 of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ will be set to one.
@@
 complexity.
-Thus we assume that computing
-$\ofunc(\ppos_{t+1})$ and all 
-of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ is the computationally difficult part of PSO.
-
-Suppose that unlimited processors were available and that the evaluation of 
+That is, we assume that computing
+$\ofunc(\ppos_{t})$ and all 
+of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ are the computationally difficult parts.
+
+For example, suppose that unlimited processors were available and that the evaluation of 
 the objective function took 1 hour. Suppose also that for each particle
 the computations
-$\ofunc(\ppos_{t+1})$ and all 
+$\ofunc(\ppos_{t})$ and all 
 of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ were sent to separate
@@
 \emph{Given} the values 
-of $\ofunc(\ppos_{t+1})$ and all 
-$\ofunc(\ppos_{t+1})$ 
+of $\ofunc(\ppos_{t})$ for all particles
 the computation of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ and to complete iteration t+1
@@
 vectors, 
-$\pURand_{t}$ and $\nURand_{t}$. As with the algebra given above, each transformation of the algorithm is
-provably equivalent (except that the algorithms will no longer be able to do an odd number of iterations,
-since it will do two at a time). In the algorithmic transformations given below,
+$\pURand_{t}$ and $\nURand_{t}$.
+As with the algebra given above, each transformation (used below) of the algorithm is
+provably equivalent; except that the algorithms will no longer be able to do an odd number of iterations,
+since it will do two at a time. In the algorithmic transformations given below,
 instructions are moved around and added, but the
@@
             \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\rightn_{t}$
-          \label{alg1:endifs}
           \ENDIF
-          \STATE Update $\pvel_{t+1}$ as shown in equation~\eqref{eq:velupdate} 
-          \STATE Update $\ppos_{t+1}$ as shown in equation~\eqref{eq:posupdate}
   \end{algorithmic}
@@
         \STATE $t=t+1$
-        \FOR{$i \in \swarm$}
+        \FOR{each particle in the swarm}
           \STATE {$\fppos_{t} \leftarrow \ofunc(\ppos_{t})$}
         \ENDFOR
-        \FOR{$i \in \swarm$}
+        \FOR{each particle in the swarm}
           \STATE $\nURand_{t} \Leftarrow RandomVector()$
@@
 This algorithm can be transformed, while retaining it exact behavior, by
-``unrolling'' one iteration of the ``Repeat'' loop, redundantly expressing work done in the ``if'''s in lines
-\ref{alg1:startifs} through
-\ref{alg1:endifs} to use the indicator functions $\indic_{t+1}$ and re-writing the computation of the next velocity,
+``unrolling'' one iteration of the ``Repeat'' loop, redundantly expressing work done in the ``if'''s
+in~\alg{alg:bests}
+using the indicator functions $\indic_{t+1}$ and re-writing the computation of the next velocity,
 positions and objective function value using 
 equations~\eqref{eq:vel2update}, \eqref{eq:pos2update} and \eqref{eq:val2update} 
-as shown in~\alg{alg:unrolled}
+as shown in the new version of the algorithm as shown in~\alg{alg:unrolled}
 
@@
         \STATE $t=t+1$
-        \FOR{$i \in \swarm$}
+        \FOR{each particle in the swarm}
           \label{alg2:fatitert}
@@
         \ENDFOR
-        \FOR{$i \in \swarm$}
+        \FOR{each particle in the swarm}
           \STATE $\nURand_{t} \Leftarrow RandomVector()$
@@
         \STATE WAIT for all function evaluations dispatched above to complete
-        \FOR{$i \in \swarm$}
+        \FOR{each particle in the swarm}
           \FOR{$c \in \caseset$}
@@
           \STATE Compute $\ppos_{t+1}$ as shown in equation~\eqref{eq:pos2update}
-          \STATE Compute $\fppos_{t+1}$ as shown in equation~\eqref{eq:val2update}
+          \STATE Compute $\ofunc(ppos_{t+1})$ as shown in equation~\eqref{eq:val2update}
 	  \STATE UpdateBests
@@
         \STATE $t=t+1$
-        \FOR{$i \in \swarm$}
+        \FOR{each particle in the swarm}
           \STATE $\nURand_{t} \Leftarrow RandomVector()$

### 2010-03-30 honors_thesis/journal_paper/paper.tex
kevin finished second pass

@@
 
-Particle swarm optimization (PSO) has previously been parallelized only by 
-adding more particles to the swarm or by parallelizing the evaluation of the
-objective function.  However, in many cases this approach does not scale to 
+Particle swarm optimization (PSO) has previously been parallelized primarily by 
+distributing the computation corresponding to particles or sets
+across multiple processors.  However, in many cases this approach is not efficient when scale to 
 very large populations sizes (on very large clusters).
-In response to this issue, we take inspiration from 
-speculative execution commonly performed in modern processors and propose
+In this paper we present a different way of using multiple processors of the 
+computation in PSO, motivated by the ``speculative execution'' in modern computer hardware and which we call
 Speculative Evaluation in PSO (SEPSO).
@@
 speculated and evaluated in parallel with the evaluation of current positions.
-We show that the set of speculative evaluations will always include the position and corresponding
-evaluation that would normally have occurred at the next iteration.
+We show that a set of separate speculative evaluation threads can be constructed
+such that the set will always include the position and corresponding
+evaluation that would normally have occurred at the next iteration of standard PSO.
 Thus SEPSO performs two
@@
 
-While Speculative Evaluation in PSO is an intriguing idea, its
-usefulness is limited unless some slight modifications are made to the PSO
-algorithm.  By relaxing the requirement that speculative evaluation exactly
-reproduce the behavior of the original PSO, we can make better use of the information
+Speculative Evaluation in PSO can be enhanced by diverging from the
+behavior of the standard PSO to make better use of the information
 obtained by the speculative evaluations.
@@
 
-These previous parallel techniques use additional processors to increase the
+These previous parallel techniques 
+distribute the computation needed by the particles in the swarm over the available processors.
+if more processors are available, these prior techniques increase the
 number of particles in the swarm, either by adding individual particles or by
@@
 algorithms can perform is thus inherently limited by the time it takes to
-evaluate the objective function---additional processors add more particles, not
-more iterations.
+evaluate the objective function---additional processors add more particles, but do not
+make the iterations go any faster.
 
@@
 clusters of thousands of processors and functions for which a single
-evaluation will take at least minutes, but probably hours or perhaps days.
+evaluation will take at least tens of seconds, but probably hours or perhaps days.
 Our purpose is to explore the question ``Since PSO is often run with
@@
 however, increasing the swarm size from 100 to 200 makes the algorithm less efficient,
-reducing the progress the algorithm makes per-evaluation.
-For this reason, research recommends the use a swarm size of 50 for this function
+that is, it reduces the progress the algorithm makes per-evaluation.
+For this reason, previous work has recommended the use a swarm size of 50 for this function
 ~\cite{bratton-sis07}.
@@
 branch on which they must wait (e.g. a memory cache miss), guess which way the
-branch will go and start executing.  If the processor guesses right, execution
-is much farther ahead than if it had just idly waited on the branch.  If it
+branch will go and start executing, being careful not to make changes that can not be undone.
+If the processor guesses right, execution
+is much farther ahead than if it had just idly waited on the memory reference.  If it
 guesses wrong, execution restarts where it would have been anyway.
@@
 
-The idea of speculative execution can be applied to some types of optimization
+\begin{quote}
+SKIP SKIP
+
+No really, try reading without this! The idea of speculative execution can be applied to some types of optimization
 algorithms.  When the next sampling position of the algorithm does not depend
@@
 if the test turns out to be false; all in parallel.
-In this example it is important to note that we do not need the value of the objectve function at the current
-point to to reduce the infinite number of possible next postions to just two alternatives.
-Given the random sample, the next point detments only on the binary test and the
-random sample is, by definition, idependent of the positions, and the vaues obtained.
-The choice of the next position can be made post-hoc, after the value at the current position is obtained.
+In this example it is important to note that we do not need the value of the objective function at the current
+point to to reduce the infinite number of possible next positions to just two alternatives.
+Choosing between these two alternatives can be done post-hoc, after the objective function value
+at the current position and at the two alternative are computed.
 For all but the most computationally-trivial objective functions, this approach saves almost half of the
@@
 step size for sampling typically depends on the gradient and therefor the position and 
-a set of infinitly many posible \emph{values} outcomes
+a set of infinitely many possible \emph{values} outcomes
 obtained at the current position (not binary or even finite).
+\end{quote}
 
@@
 In this paper we show that the results of standard PSO can be reproduced \emph{exactly}, two iterations at a time, using
-a speculative evaluation approach like the ones described above. We prove that this approach
-produces exactly equivalent results, and demonstrate this by implementation.
+a speculative approach like the ones described above. We prove that the standard PSO equation can
+be factored such that a small set of speculative positions can be found which will \emph{always}
+include the position computed in the next iteration.
+By computing the value of the objective function for each of the speculative positions at
+the same time the algorithm evaluates the objective function for the current position,
+It is possible to know the objective function values for both the current and the next iteration at the same time.
+We demonstrate this principle by implementation and show that it
+produces exactly the same results as standard PSO, but two iterations at a time.
 The resulting implementation runs efficiently on large clusters where the number of processors is much larger than
@@
 we can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
-By pruning the speculation to just paths that are statistically likely to reproduce the paths that are equivelent to PSO we can also 
+By pruning the speculation to just paths that are statistically likely to reproduce
+the paths that are equivalent to PSO we can also 
 speculating several iterations ahead instead of just one.
@@
 Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.
+We use $p$ to indicate the number of particles in the swarm.
 
@@
 neighborhood topology has a significant effect on the performance of the algorithm and 
-this effect varies widely by function type, as would be expected by the No Free Lunch
-theorem~\cite{wolpert-tec97}.
+this effect varies widely by function type.
+%, as would be expected by the No Free Lunch
+%theorem~\cite{wolpert-tec97}.
 The two most common topologies used in the
@@
 topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
-or more neighbors to either side of it; in the Complete topology, every particle is a
-neighbor to every other particle.  In all topologies, a particle is also a
-neighbor to itself in that its own position is considered when updating the
+(or more) neighbors to either side of it; in the Complete topology, every particle is a
+neighbor to every other particle.  It is generally assumed that, a particle is also a
+neighbor to itself in that its own position and value is considered when updating the
 particle's personal best, $\nbest$.  Thus with $p$ particles, using the Ring topology each
@@
 topology, each particle has $p$ neighbors.
-% TOD USE THIS LATER!
+% TODO USE THIS LATER!
 % In this paper we use these
@@
 
-To address this issue we present here a very different use of additional processors.
+To address this issue we present here a different use of additional processors.
 We take inspiration from speculative execution in processors and instead of adding particles
@@
 This allows us to perform iterations two at a time.
+
 To do so we must first re-factor PSO such that the determination of the $\pbest$ and $\nbest$,
 which depend upon the value of the objective function, is sepearate from the rest of the computation.
-
 %First consider formula for the \emph{next} iteration of PSO, which is identical
@@
 so we instead use a set of indicator variables, one each representing each
-possible update to the \emph{pair} $\casedef$.  These indicator variables allow
+possible update to the pair $\casedef$.  These indicator variables allow
 us to easily identify the cases in which updates will be needed for $\nbest$,
@@
 
-For example, we can write $\casexn$ to identify the case where $\pbest$ will be
+For a topology with 2 neighbors there are 7 possible update cases identified in
+Table~\ref{tab:evals}.
+For example and as indicated in the table, we write $\casexn$ to represent the case where $\pbest$ will be
 updated with the better position $\ppos_{i,t}$, shown in the first positions,
@@
 position.
-
-For any topology with 2 neighbors there are 7 possible update cases listed in
-Table~\ref{tab:evals}.  This list assumes, as is typically the case, that each
-particle is also part of its own neighborhood (can cause an update to
-$\nbest$). The list also omits the case where $\nbest$ is updated by
+%This list assumes, as is typically the case for the ring topology, that each
+%particle is also part of its own neighborhood (can cause an update to
+%$\nbest$).
+The list omits the case where $\nbest$ is updated by
 $\ppos_{t}$ but $\pbest$ was not updated, as this is logically impossible.
-$\caseset$ represents the set of all update cases.
-
-For minimization, we can define this example indicator variable representing the case
+$\caseset$ represents the set of all possible update cases.
+
+For minimization, we can define the indicator variable representing the case $\casexn$,
 where the particle will need to update $\pbest$ with the better
-position $x_{t}$ for the next iteration ($t+1$), but $\nbest$ will not updated by:
+position $x_{t}$, but $\nbest$ will not updated, by:
 \begin{align}
@@
 For each cases identified in Table~\ref{tab:evals} and identified by a
-corresponding indicator variable defined above, we also provide notation for the \emph{next} velocity
+corresponding indicator variable defined above, we also provide notation for the velocity
 update equation; once again using $\casexn$ as an example:
@@
 they can be inferred from the subscripts on $\specvel_{t}^{\casexn}$.
-The update equation is taken exactly from the standard update equation~\eqref{eq:velupdate}, but
-since it is restricted to a specific case, $\casexn$ for this example,
-it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
-$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t}$ we can re-write
+The update equation is taken exactly from the standard update equation~\eqref{eq:velupdate}.
+However, since $\specvel_{t}^{\casexn}$ is restricted to the specific case, $\casexn$,
+it is no longer necessary to use $\pbest$ and $\nbest$, since when restricted to this case
+$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t}$ and we can re-write
 $\specvel_{t+1}^{\casexn}$ as:
@@
 
-For each cases identified in
-Table~\ref{tab:evals}
-and identified by a corresponding indicator variable defined above,
-we define also a position update equation with its compact form; once again using $\casexn$ as an example:
+In the same way we can create notation for the position update
+equation~\eqref{eq:velupdate}; once again using $\casexn$ as an example:
 \begin{align}
@@
 \end{align}
-We call these values, $\specpos^{c}) \ c \in\ \caseset$ (7 of them in this case where each particle has 2 neighbors) as ``speculative children''
+We call these values, $\specpos^{c}) \ c \in\ \caseset$ (7 of them in this case where each particle has 2 neighbors plus itself)
+as ``speculative children''
 since they will only be useful if the corresponding $\indic^{c} \ c \in \caseset$ has the value 1.
@@
 
-Returning our attention to the computtaion of $\specpos$ in \eqref{eq:pos2update}and writing it with the parameters which were omitted above,
+Returning our attention to the computation of $\specpos$ in \eqref{eq:pos2update} and writing it
+with the parameters which were omitted above,
 we obtain:
@@
 \end{align}
-In this form the important point to notice is that none of the $\specpos_{t+1}^{c} \ c \in\ \caseset$ depend upon $f(x_t)$
+In this form the important point to notice is that there are only 7 values (for this ring topology) in the set formed by
+$\specpos_{t+1}^{c} \ c \in\ \caseset$ and that none of them depend upon $f(x_t)$
 or any other objective function evaluation at $t$.
 Thus PSO has been re-factored such that the algorithm can begin computing
-all 7 of the objective function evaluations potentially needed in iteration $t+1$ before $f(x_t)$ is computed.
-
-This refactoring comes a significant expense, instead of just computing
-$\ofunc(\ppos_{i,t})$ and then $\ofunc(\ppos_{i,t+1})$ in the usual way, this refactored PSO
-computes $\ofunc(\ppos_{t})$ and all 
-of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time.
-Once the evaluation of $f(\ppos_{t})$ is completed only one
+all 7 of the objective function evaluations potentially needed in iteration $t+1$ \emph{before} $f(x_t)$ is computed.
+Once the evaluation of $f(\ppos_{t})$ is completed for all particles only one
 of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ will be set to one.
 We will refer to the indicator function which is set to one as the
-\emph{behavior preserving indicator} or $bp$, since it indicates which of
+``behavior preserving indicator'' or $bp$, since it indicates which of
 the computations $\ofunc(\specpos_{t}^{c}) \  c \in \caseset$ preservers the behavior of PSO.
@@
 
+This refactoring comes a significant expense, instead of just computing
+$\ofunc(\ppos_{i,t})$ and then $\ofunc(\ppos_{i,t+1})$ in the usual way, this refactored PSO
+computes $\ofunc(\ppos_{t})$ and all 
+of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$, 8 function evaluations in this case.
 The potential value of these speculative children lies in the phrase used above
@@
 That is, we assume that computing
-$\ofunc(\ppos_{t})$ and all 
-of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ are the computationally difficult parts.
-
-For example, suppose that unlimited processors were available and that the evaluation of 
+$\ofunc(\ppos_{t})$ is the computationally difficult part of one iteration of PSO.
+
+The value of the refactoring can best be seen in a small example.
+Suppose that unlimited processors were available and that the evaluation of 
 the objective function took 1 hour. Suppose also that for each particle
@@
 
-While the algebra show that the computation of 
+While the algebra shows that the computation of 
 equations~\eqref{eq:vel2update} and \eqref{eq:pos2update} are identical to the 
@@
 \ref{alg2:endcut}. The fact that these values are also used by other speculative
-children will not change the behavior preserving case.
+children and later discarded will not change the behavior preserving case.
+
+End of Kevin
 

### 2010-03-30 honors_thesis/journal_paper/paper.tex
Fixed some broken references

@@
 We first define a helper function to update the personal and neighborhood bests, $\pbest$ and $\nbest$.
-This algorithm is shown in~\alg{alg:bests}.
+This algorithm is shown in~\alg{bests}.
 
@@
 
-\begin{algorithm} [h!]
+\begin{algorithm} [ht!]
   \caption{Standard PSO (minimizing)}
@@
 
-Using this helper function we can write PSO as shown in~\alg{alg:standard}.
+Using this helper function we can write PSO as shown in~\alg{standard}.
 This algorithm can be transformed, while retaining it exact behavior, by
 ``unrolling'' one iteration of the ``Repeat'' loop, redundantly expressing work done in the ``if'''s
-in~\alg{alg:bests}
+in~\alg{bests}
 using the indicator functions $\indic_{t+1}$ and re-writing the computation of the next velocity,
@@
 equations~\eqref{eq:vel2update}, \eqref{eq:pos2update} and \eqref{eq:val2update} 
-as shown in the new version of the algorithm as shown in~\alg{alg:unrolled}
-
-\begin{algorithm} [h!]
+as shown in the new version of the algorithm as shown in~\alg{unrolled}
+
+\begin{algorithm} [ht!]
   \caption{Unrolled PSO}
@@
 \sec{manyiters} shows a method for speculating several iterations ahead,
-instead of just one.  \sec{other} presents a few other ideas that seem
-promising but need further development.
+instead of just one.
 

### 2010-04-02 honors_thesis/journal_paper/paper.tex
Major revisions to the journal paper from feedback at the meeting

@@
 \providecommand{\pvel}{\ensuremath{\Vec{v}}}
-\providecommand{\nbest}{\ensuremath{\Vec{b}^\neigh}}
-\providecommand{\fnbest}{\ensuremath{\Vec{fb}^\neigh}}
-\providecommand{\pbest}{\ensuremath{\Vec{b}^\pers}}
-\providecommand{\fpbest}{\ensuremath{\Vec{fb}^\pers}}
+\providecommand{\nbest}{\ensuremath{\Vec{x}^\neigh}}
+\providecommand{\fnbest}{\ensuremath{\Vec{fx}^\neigh}}
+\providecommand{\pbest}{\ensuremath{\Vec{x}^\pers}}
+\providecommand{\fpbest}{\ensuremath{\Vec{fx}^\pers}}
 \providecommand{\constriction}{\ensuremath{\chi}}
@@
 \providecommand{\caseset}{\ensuremath{\mathcal{C}}}
+\providecommand{\casegen}{\ensuremath{c}}
 \providecommand{\casedef}{\ensuremath{(\pbest,\nbest)}}
-\providecommand{\casexn}{\ensuremath{(x,\neigh)}}
-\providecommand{\casexx}{\ensuremath{(x,x)}}
-\providecommand{\casexl}{\ensuremath{(x,\leftind)}}
-\providecommand{\casexr}{\ensuremath{(x,\rightind)}}
-\providecommand{\casepn}{\ensuremath{(\pers,\neigh)}}
-\providecommand{\casepl}{\ensuremath{(\pers,\leftind)}}
-\providecommand{\casepr}{\ensuremath{(\pers,\rightind)}}
+\providecommand{\casexn}{\ensuremath{(S,-)}}
+\providecommand{\casexx}{\ensuremath{(S,S)}}
+\providecommand{\casexl}{\ensuremath{(S,\leftind)}}
+\providecommand{\casexr}{\ensuremath{(S,\rightind)}}
+\providecommand{\casepn}{\ensuremath{(-,-)}}
+\providecommand{\casepl}{\ensuremath{(-,\leftind)}}
+\providecommand{\casepr}{\ensuremath{(-,\rightind)}}
 \providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
@@
 Particle swarm optimization (PSO) has previously been parallelized primarily by
-distributing the computation corresponding to particles or sets
-across multiple processors.  However, in many cases this approach is not efficient when scale to 
-very large populations sizes (on very large clusters).
-In this paper we present a different way of using multiple processors of the 
-computation in PSO, motivated by the ``speculative execution'' in modern computer hardware and which we call
-Speculative Evaluation in PSO (SEPSO).
-
-In SEPSO future positions of the particles are
-speculated and evaluated in parallel with the evaluation of current positions.
-We show that a set of separate speculative evaluation threads can be constructed
-such that the set will always include the position and corresponding
-evaluation that would normally have occurred at the next iteration of standard PSO.
-Thus SEPSO performs two
-iterations of PSO at once, allowing it to achieve better fitness faster.
-Not only can SEPSO can achieve better fitness in the same number of iterations,
-it also reaches better fitness for the same number of total function evaluations
+distributing the computation corresponding to particles across multiple
+processors.  In these approaches, additional processors can only add more
+particles to the swarm.  However, in many cases this is not efficient when
+scaled to very large swarm sizes (on very large clusters).  In this paper we
+present a different way of using additional processors in the computation of
+PSO motivated by speculative execution in modern computer hardware.  We call
+our method Speculative Evaluation in PSO (SEPSO).
+
+In SEPSO future positions of the particles are speculated and evaluated in
+parallel with the evaluation of current positions.  We show that a set of
+speculative evaluation threads can be constructed such that the set will always
+include the position and corresponding evaluation that would normally have
+occurred at the next iteration of standard PSO.  Thus SEPSO performs two
+iterations of PSO at once, allowing it to achieve better fitness faster.  Not
+only can SEPSO can achieve better fitness in the same number of iterations, it
+also reaches better fitness for the same number of total function evaluations
 in large scale parallel environments.
 
-Speculative Evaluation in PSO can be enhanced by diverging from the
-behavior of the standard PSO to make better use of the information
-obtained by the speculative evaluations.
-Using these algorithmic modifications, we see dramatic improvements in the
-effectiveness of the algorithm. 
+Speculative Evaluation in PSO can be enhanced by diverging from the behavior of
+the standard PSO to make better use of the information obtained by the
+speculative evaluations.  Using these algorithmic modifications, we see
+dramatic improvements in the effectiveness of the algorithm.
+
 \end{abstract}
@@
 the world's computing power is increasingly coming from large clusters of
-processors.
-%Even in desktop machines, systems with 4 or 8 cores are
-%commonplace.  
-In order to efficiently utilize these resources for
+processors.  In order to efficiently utilize these resources for
 computationally intensive problems, PSO needs to run in parallel.
@@
 \cite{mcnabb-cec07,belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
-and asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05}.
-Parallelizing the evaluation of the objective function can also be done in some
-cases, though that is not an adaption of the PSO algorithm itself and thus is
-not the focus of this paper.
-
-These previous parallel techniques 
-distribute the computation needed by the particles in the swarm over the available processors.
-if more processors are available, these prior techniques increase the
-number of particles in the swarm, either by adding individual particles or by
-adding entire new sub-swarms but in all cases
-the number of processors never exceeds the number of particles.
-The number of iterations of PSO that the
-algorithms can perform is thus inherently limited by the time it takes to
-evaluate the objective function---additional processors add more particles, but do not
-make the iterations go any faster.
-
-The purpose of this paper is to consider PSO parallelization strategies for
-clusters of thousands of processors and functions for which a single
-evaluation will take at least tens of seconds, but probably hours or perhaps days.
-Our purpose is to explore the question ``Since PSO is often run with
-a swarm of size of 50-100, what would one do if a few thousand processors were available?''
-
-For many functions there comes a point of diminishing returns with respect to adding particles.
-In \fig{basic-sphere} we show the value obtained after 20,000 function evaluations (not iterations)
-as a function of swarm size
-for the well-know benchmark function Sphere (20 dimensions, reporting the average of thirty runs). 
-Increasing the swarm size from 10 to 50 has a significant effect value obtained
-however, increasing the swarm size from 100 to 200 makes the algorithm less efficient,
-that is, it reduces the progress the algorithm makes per-evaluation.
-For this reason, previous work has recommended the use a swarm size of 50 for this function
-~\cite{bratton-sis07}.
-Thus, in at least some cases, added particles indefinitely will not yield an efficient implementation.
+and asynchronous algorithms
+\cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05}.  Parallelizing the
+evaluation of the objective function can also be done in some cases, though
+that is not an adaption of the PSO algorithm itself and thus is not the focus
+of this paper.
+
+These previous parallel techniques distribute the computation needed by the
+particles in the swarm over the available processors.  If more processors are
+available, these techniques increase the number of particles in the swarm,
+either by adding individual particles or by adding entire new sub-swarms, but
+in all cases the number of processors never exceeds the number of particles.
+The number of iterations of PSO that the algorithms can perform is thus
+inherently limited by the time it takes to evaluate the objective
+function---additional processors add more particles, but do not make the
+iterations go any faster.
+
+For many functions there comes a point of diminishing returns with respect to
+adding particles.  In \fig{evals-sphere} we show the value obtained after
+20,000 function evaluations (not iterations) as a function of swarm size for
+the well-know benchmark function Sphere (20 dimensions, reporting the average
+of thirty runs).  Increasing the swarm size from 10 to 50 has a significant
+effect on the value obtained.  However, increasing the swarm size from 100 to
+200 makes the algorithm less efficient, that is, it reduces the progress the
+algorithm makes per evaluation.  For this reason, previous work has recommended
+the use of a swarm size of 50 for this function~\cite{bratton-sis07}.  Thus, in
+at least some cases, adding particles indefinitely will not yield an efficient
+implementation.
 
@@
 
-To solve this problem, we take inspiration from speculative execution
-techniques commonly used in processors.  Modern processors, when faced with a
-branch on which they must wait (e.g. a memory cache miss), guess which way the
-branch will go and start executing, being careful not to make changes that can not be undone.
-If the processor guesses right, execution
-is much farther ahead than if it had just idly waited on the memory reference.  If it
-guesses wrong, execution restarts where it would have been anyway.
-This is called speculative execution or branch prediction.
-
-\begin{quote}
-SKIP SKIP
-
-No really, try reading without this! The idea of speculative execution can be applied to some types of optimization
-algorithms.  When the next sampling position of the algorithm does not depend
-on the value of the objective function at the current position,
-future positions can be speculated and evaluated in parallel with the current
-position.
-This is trivially true for optimization by random sampling, for example.
-The next random sample does not depend on the previous point nor on the value
-of the objective function at that point in any way.
-It depends only on a random number.
-There is no need to wait to see
-what objective function value was obtained at the previous point,
-evaluation of the object function at the next point can begin as soon as the next random number is drawn.
-In other optimization algorithms, the next position depends on the previous position, but not the
-value obtained there, a random walk for example.
-In this case
-the next position to be evaluated
-depends only on the previous position and a random number.
-In this case, as it was in the case of random sampling,
-the object function evaluation of the next position can be done in parallel
-because it 
-does not depend on the \emph{value} of the objective function at the previous position.
-There are an infinite number of possible next positions, but only one given the draw of the random number.
-
-In still other cases, the next position to be evaluated may depend on the current position and some
-discrete valued (binary, for example) test on the value obtained at the current position.
-For example, consider a greedy random walk, in which the state only advances if a better
-position is found.
-That is, if the current value is better than the previous value,
-use the sum of the current position and the random number as the next position.
-Otherwise, computer the next position using the sum of the previous position and the random number.
-In this case the objective function can be evaluated at three positions:
-(1) the value of the objective function at the current position,
-(2) the value at the position if the test turns out to be true and (3) the value at the position
-if the test turns out to be false; all in parallel.
-In this example it is important to note that we do not need the value of the objective function at the current
-point to to reduce the infinite number of possible next positions to just two alternatives.
-Choosing between these two alternatives can be done post-hoc, after the objective function value
-at the current position and at the two alternative are computed.
-For all but the most computationally-trivial objective functions, this approach saves almost half of the
-``wall-clock'' time.
-This approach has been applied to simulated annealing~\cite{witte-tpds91} which like the greedy random walk,
-depends on positions and a binary test.
-
-This approach can not be used in gradient methods where the
-step size for sampling typically depends on the gradient and therefor the position and 
-a set of infinitely many possible \emph{values} outcomes
-obtained at the current position (not binary or even finite).
-\end{quote}
+In this paper we consider PSO parallelization strategies for clusters of
+thousands of processors and functions for which a single evaluation will take
+at least tens of seconds, but probably minutes or perhaps hours.  Our purpose
+is to explore the question of what to do with a thousand processors when 50 or
+100 particles is the recommended swarm size, and simply increasing the swarm
+size encounters diminishing returns.
+
+To solve this problem, we propose a method for performing two iterations of PSO
+at the same in parallel that we call speculative evaluation.  The name comes
+from an analogy to speculative execution techniques commonly used in
+processors.  Modern processors, when faced with a branch on which they must
+wait (e.g. a memory cache miss), guess which way the branch will go and start
+executing, being careful not to make changes that can not be undone.  If the
+processor guesses right, execution is much farther ahead than if it had just
+idly waited on the memory reference.  If it guesses wrong, execution restarts
+where it would have been anyway.  This is called speculative execution or
+branch prediction.
+
+%\begin{quote}
+%SKIP SKIP
+%
+%No really, try reading without this! The idea of speculative execution can be applied to some types of optimization
+%algorithms.  When the next sampling position of the algorithm does not depend
+%on the value of the objective function at the current position,
+%future positions can be speculated and evaluated in parallel with the current
+%position.
+%This is trivially true for optimization by random sampling, for example.
+%The next random sample does not depend on the previous point nor on the value
+%of the objective function at that point in any way.
+%It depends only on a random number.
+%There is no need to wait to see
+%what objective function value was obtained at the previous point,
+%evaluation of the object function at the next point can begin as soon as the next random number is drawn.
+%In other optimization algorithms, the next position depends on the previous position, but not the
+%value obtained there, a random walk for example.
+%In this case
+%the next position to be evaluated
+%depends only on the previous position and a random number.
+%In this case, as it was in the case of random sampling,
+%the object function evaluation of the next position can be done in parallel
+%because it 
+%does not depend on the \emph{value} of the objective function at the previous position.
+%There are an infinite number of possible next positions, but only one given the draw of the random number.
+%
+%In still other cases, the next position to be evaluated may depend on the current position and some
+%discrete valued (binary, for example) test on the value obtained at the current position.
+%For example, consider a greedy random walk, in which the state only advances if a better
+%position is found.
+%That is, if the current value is better than the previous value,
+%use the sum of the current position and the random number as the next position.
+%Otherwise, computer the next position using the sum of the previous position and the random number.
+%In this case the objective function can be evaluated at three positions:
+%(1) the value of the objective function at the current position,
+%(2) the value at the position if the test turns out to be true and (3) the value at the position
+%if the test turns out to be false; all in parallel.
+%In this example it is important to note that we do not need the value of the objective function at the current
+%point to to reduce the infinite number of possible next positions to just two alternatives.
+%Choosing between these two alternatives can be done post-hoc, after the objective function value
+%at the current position and at the two alternative are computed.
+%For all but the most computationally-trivial objective functions, this approach saves almost half of the
+%``wall-clock'' time.
+%This approach has been applied to simulated annealing~\cite{witte-tpds91} which like the greedy random walk,
+%depends on positions and a binary test.
+%
+%This approach can not be used in gradient methods where the
+%step size for sampling typically depends on the gradient and therefor the position and 
+%a set of infinitely many possible \emph{values} outcomes
+%obtained at the current position (not binary or even finite).
+%\end{quote}
 
@@
 
-In this paper we show that the results of standard PSO can be reproduced \emph{exactly}, two iterations at a time, using
-a speculative approach like the ones described above. We prove that the standard PSO equation can
-be factored such that a small set of speculative positions can be found which will \emph{always}
-include the position computed in the next iteration.
-By computing the value of the objective function for each of the speculative positions at
-the same time the algorithm evaluates the objective function for the current position,
-It is possible to know the objective function values for both the current and the next iteration at the same time.
-We demonstrate this principle by implementation and show that it
-produces exactly the same results as standard PSO, but two iterations at a time.
-The resulting implementation runs efficiently on large clusters where the number of processors is much larger than
-a typical or reasonable number of particles. Producing better results in less ``wall-clock'' time and,
-surprisingly, with fewer total function evaluations.
+In this paper we show that the results of standard PSO can be reproduced
+\emph{exactly}, two iterations at a time, using a speculative approach adapted
+from speculative execution. We prove that the standard PSO equation can be
+factored such that a set of speculative positions can be found which will
+\emph{always} include the position computed in the next iteration.  By
+computing the value of the objective function for each of the speculative
+positions at the same time the algorithm evaluates the objective function for
+the current position, it is possible to know the objective function values for
+both the current and the next iteration at the same time.  We demonstrate this
+principle by implementation and show that it produces exactly the same results
+as standard PSO, but two iterations at a time.  The resulting implementation
+runs efficiently on large clusters where the number of processors is much
+larger than a typical or reasonable number of particles, producing better
+results in less ``wall-clock'' time and, surprisingly, with fewer total
+function evaluations.
 
 Furthermore, we show that if we relax the requirements of the algorithm, no
-longer demanding that it strictly reproduce the exact behavior of standard PSO, we
-can introduce new speculative techniques that often out-perform PSO.
-These variants make better use of the extra information obtained from
-the extra exploration made by the speculative function evaluations.
-We also explore the idea that, like branch prediction in processors, we
-need not speculatively evaluate \emph{all} possible future positions,
-we can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
-By pruning the speculation to just paths that are statistically likely to reproduce
-the paths that are equivalent to PSO we can also 
-speculating several iterations ahead instead of just one.
-We also consider several recovery strategies for cases where
-the pruned set of speculative evaluations does not contain the evaluation that standard PSO would have done.
-
-The balance of this paper is organized as follows:
-\sec{pso} describes the particle swarm optimization algorithm.
-\sec{sepso}
-describes how speculative evaluation can be done in parallel PSO to
-perform two iterations at once, then presents a comparison of the speculative
-algorithm with the original PSO.  In \sec{relax}, we discuss various methods of
-improving the performance of speculative evaluation in PSO, all of which break
-the requirement of strictly reproducing the behavior of the original algorithm.
-In \sec{conclusion} and \sec{future} we conclude and discuss future work.
+longer demanding that it strictly reproduce the exact behavior of standard PSO,
+we can introduce new speculative techniques that often out-perform PSO.  These
+variants make better use of the information obtained from the extra exploration
+made by the speculative function evaluations.  We also explore the idea that,
+like branch prediction in processors, we need not speculatively evaluate
+\emph{all} possible future positions, we can accelerate the algorithm even if
+we are just \emph{likely} to have guessed right.  By pruning the speculation to
+just paths that are statistically likely to reproduce the paths that are
+equivalent to PSO we can increase the swarm size without increasing the number
+of speculative evaluations.  We also consider several recovery strategies for
+cases where the pruned set of speculative evaluations does not contain the
+evaluation that standard PSO would have done.  A further improvement we explore
+is speculating several iterations ahead instead of just one, which is made
+possible by pruning the number of speculative evaluations.
+
+The balance of this paper is organized as follows: \sec{pso} describes the
+particle swarm optimization algorithm.  \sec{sepso} describes how speculative
+evaluation can be done in parallel PSO to perform two iterations at once, then
+presents a comparison of the speculative algorithm with the original PSO.  In
+\sec{relax}, we discuss various methods of improving the performance of
+speculative evaluation in PSO, all of which break the requirement of strictly
+reproducing the behavior of the original algorithm.  In \sec{conclusion} and
+\sec{future} we conclude and discuss future work.
 
@@
 birds and other animals. It is a social algorithm that depends on interaction
-between particles to quickly and consistently approximate the optimal solution to a
-given objective function.
+between particles to quickly and consistently approximate the optimal solution
+to a given objective function.
 
@@
 
-At each iteration ($t+1$) of the algorithm, the position $\ppos_t$ and velocity
+At each iteration of the algorithm, the position $\ppos_t$ and velocity
 $\pvel_t$ of each particle are updated as follows:
@@
 \end{align}
-where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent
-and identically distributed random numbers drawn from a uniform
-distribution, the \( \otimes \) operator is an element-wise vector
-multiplication, $\pbest$ (called personal best) is the best position the
-current particle has seen, and $\nbest$ (called neighborhood best) is the best
-position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
-\pcoeff \), and \( \constriction \) are parameters with prescribed values
-required to ensure convergence (2.05, 2.05, and .73,
-respectively)~\cite{clerc-tec02}. Note also that we use a subscript for the 
-time step $t$ or $t+1$ and have moved all other qualifiers,
-such as the personal qualifier ``$\pers$'' to the superscript position.
-Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.
-We use $p$ to indicate the number of particles in the swarm.
-
-There are many ways of defining the neighbors of a particle. 
-Changing the
-neighborhood topology has a significant effect on the performance of the algorithm and 
-this effect varies widely by function type.
-%, as would be expected by the No Free Lunch
-%theorem~\cite{wolpert-tec97}.
-The two most common topologies used in the
-literature are the Ring topology and the Complete
-topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
-(or more) neighbors to either side of it; in the Complete topology, every particle is a
-neighbor to every other particle.  It is generally assumed that, a particle is also a
-neighbor to itself in that its own position and value is considered when updating the
-particle's personal best, $\nbest$.  Thus with $p$ particles, using the Ring topology each
-particle has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
-topology, each particle has $p$ neighbors.
-% TODO USE THIS LATER!
-% In this paper we use these
-% topologies, as well as a parallel adaptation of the Complete topology, called
-% Random, that has been shown to simulate the behavior of Complete with far less
-% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
-% randomly picks two other particles to share information with at each iteration,
-% along with itself.  Thus in both the Ring and the Random topologies, all
-% particles have 3 neighbors.
+where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
+numbers drawn from a standard uniform distribution, the \( \otimes \) operator
+is an element-wise vector multiplication, $\pbest$ (called personal best) is
+the best position the current particle has seen, and $\nbest$ (called
+neighborhood best) is the best position the neighbors of the current particle
+have seen~\cite{bratton-sis07}.  \( \ncoeff \), \( \pcoeff \), and \(
+\constriction \) are parameters with prescribed values required to ensure
+convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-tec02}.
+
+There are many ways of defining the neighbors of a particle.  Changing the
+neighborhood topology has a significant effect on the performance of the
+algorithm and this effect varies widely by function type.  The two most common
+topologies used in the literature are the Ring topology and the Complete
+topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one (or
+more) neighbors to either side of it; in the Complete topology, every particle
+is a neighbor to every other particle.  It is generally assumed that, a
+particle is also a neighbor to itself in that its own position and value is
+considered when updating the particle's personal best, $\nbest$.  Thus with $p$
+particles, using the Ring topology each particle has three neighbors: $i-1$,
+$i$ (itself), and $i+1$.  With the Complete topology, each particle has $p$
+neighbors.
+
+In this paper we use these topologies as well as a parallel adaptation of the
+Complete topology, called Random, that has been shown to simulate the behavior
+of Complete with far less communication~\cite{mcnabb-cec09}.  In the Random
+topology, each particle randomly picks two other particles to share information
+with at each iteration, along with itself.  Thus in both the Ring and the
+Random topologies, all particles have 3 neighbors.
 
@@
 
-The PSO algorithm can be trivially parallelized by distributing the processing needed for each
-particle on a separate processor.
-But as we have seen in the introduction, for some
-functions, and for large numbers of processors just adding particles
-reaches a point of diminishing returns.
-That is, adding processors does not help us reach any given level of fitness appreciably faster.
+The PSO algorithm can be trivially parallelized by distributing the processing
+needed for each particle on across processors.  But as we have seen in the
+introduction, for some functions, and for large numbers of processors, just
+adding particles reaches a point of diminishing returns.  That is, adding
+processors does not help us reach any given level of fitness appreciably
+faster.
 
 To address this issue we present here a different use of additional processors.
-We take inspiration from speculative execution in processors and instead of adding particles
-we will employ a speculative evaluation approach like those described in the introduction. 
-This allows us to perform iterations two at a time.
-
-To do so we must first re-factor PSO such that the determination of the $\pbest$ and $\nbest$,
-which depend upon the value of the objective function, is sepearate from the rest of the computation.
-%First consider formula for the \emph{next} iteration of PSO, which is identical
-%to Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time
-%subscripts are incremented.  As noted above and
-For simplicity in this
-explanation, we restrict our discussion here to the case of a Ring topology
-with two neighbors which we will call the ``right neighbor'' and ``left
-neighbor'' neighbors identified for each particle as $\rightn$ and $\leftn$
-respectively.
-
-Next note that the computation of $\pbest$ and $\nbest$ is not seen in the
-update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so we will make
-this mathematically explicit.  This update is usually implemented as a pair of
-independent ``if``-statements, but that is a bit awkward for our purposes
-so we instead use a set of indicator variables, one each representing each
-possible update to the pair $\casedef$.  These indicator variables allow
-us to easily identify the cases in which updates will be needed for $\nbest$,
-$\pbest$ or both.
-
-For a topology with 2 neighbors there are 7 possible update cases identified in
-Table~\ref{tab:evals}.
-For example and as indicated in the table, we write $\casexn$ to represent the case where $\pbest$ will be
-updated with the better position $\ppos_{i,t}$, shown in the first positions,
-but a no new neighborhood best was found so $\nbest$ remains in the second
-position.
-%This list assumes, as is typically the case for the ring topology, that each
-%particle is also part of its own neighborhood (can cause an update to
-%$\nbest$).
-The list omits the case where $\nbest$ is updated by
-$\ppos_{t}$ but $\pbest$ was not updated, as this is logically impossible.
-$\caseset$ represents the set of all possible update cases.
-
-For minimization, we can define the indicator variable representing the case $\casexn$,
-where the particle will need to update $\pbest$ with the better
-position $x_{t}$, but $\nbest$ will not updated, by:
-\begin{align}
-\nonumber
-	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))= \\
-\nonumber
-	& 1 \ if \  \ofunc(\ppos_{t+1}) > \ofunc(\pbest_{t}) \\
-\nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\ppos_{t+1}) \\
-\nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\leftn_{t+1}) \\
-\nonumber
-	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\rightn_{t+1}) \\
-\label{eq:deficasexn}
-	& 0 \ otherwise
-\end{align}
-
-For compactness in our equations we represent this without the parameters since
-they can be inferred from the subscripts on $\indic_{t}^{\casexn}$.  The
-indicator variables for the other update cases shown in Table~\ref{tab:evals}
-($\casepn$, $\casepl$, $\casepr$, etc.) can be defined in similar fashion but
-obviously using the corresponding boolean expression.
-
-For each cases identified in Table~\ref{tab:evals} and identified by a
-corresponding indicator variable defined above, we also provide notation for the velocity
-update equation; once again using $\casexn$ as an example:
-\begin{align}
-\nonumber
-	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t+1}, \leftn_{t+1}, \rightn_{t+1}, \pbest_{t},
-        \nbest_{t}, \pURand_{t}, \nURand_{t}) \\
-\label{eq:defvocasexn}
-		&= \constriction \bigl[ \pvel_{t} +
-			\pcoeff\pURand_{t+1}\otimes(\pbest_{t+1} - \ppos_{t+1}) \\
-		& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t+1} - \ppos_{t+1})
-		\bigr]
-\end{align}
-
-Again we will drop the parameters for compactness and since
-they can be inferred from the subscripts on $\specvel_{t}^{\casexn}$.
-The update equation is taken exactly from the standard update equation~\eqref{eq:velupdate}.
-However, since $\specvel_{t}^{\casexn}$ is restricted to the specific case, $\casexn$,
-it is no longer necessary to use $\pbest$ and $\nbest$, since when restricted to this case
-$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t}$ and we can re-write
-$\specvel_{t+1}^{\casexn}$ as:
-\begin{align}
-\nonumber
-	\specvel_{t+1}^{\casexn} &= \constriction \bigl[ \pvel_{t} +
-			\pcoeff\pURand_{t+1}\otimes(\ppos_{t+1} - \ppos_{t+1}) \\
-\label{eq:defvcasexn}
-			& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t} - \ppos_{t+1})
-		\bigr]
-\end{align}
-
-In the same way we can create notation for the position update
-equation~\eqref{eq:velupdate}; once again using $\casexn$ as an example:
-\begin{align}
-\label{eq:defpcasexn}
-	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t},\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t})= \\
-\nonumber
-	& = \specpos_{t+1}^{\casexn} =  \ppos_{t} + \specvel_{t+1}^{\casexn}
-\end{align}
-We call these values, $\specpos^{c}) \ c \in\ \caseset$ (7 of them in this case where each particle has 2 neighbors plus itself)
-as ``speculative children''
-since they will only be useful if the corresponding $\indic^{c} \ c \in \caseset$ has the value 1.
-
-Lastly, we use the normal notation for the value of the objective function for each speculative child,
-once again using the case $\casexn$ as an example:
-\begin{align}
-\nonumber
-	\ofunc(\specpos_{t+1}^{\casexn} & (\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t}))= \\
-\label{eq:deffcasexn}
-	& =  \ofunc(\specpos_{t+1}^{\casexn})
-\end{align}
-
-With this notation we can re-write the original PSO velocity
-equation~\eqref{eq:velupdate} as:
-\begin{align}
-\nonumber
-	\pvel_{t+1} &=
-		\constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
-\nonumber
-			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
-		\bigr] \\
-\nonumber
-	&= \sum_{c \in \caseset} \constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
-\nonumber
-			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
-		\bigr] \indic_{t+1}^{c} \\
-\label{eq:vel2update}
-	&= \sum_{c \in \caseset} \ \specvel_{t+1}^{c} \ \indic_{t+1}^{c}
-\end{align}
-and the position update 
-equation~\eqref{eq:posupdate} as:
-\begin{align}
-\nonumber
-	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
-\label{eq:pos2update}
-	&= \sum_{c \in \caseset} \ \specpos_{t+1}^{c} \ \indic_{t+1}^{c}
-\end{align}
-The value of the objective function at $\ppos_{t+1}$ is given by:
-\begin{align}
-\label{eq:val2update}
-	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \ofunc(\specpos_{t+1}^{c}) \ \indic_{t+1}^{c}
-\end{align}
-
-Returning our attention to the computation of $\specpos$ in \eqref{eq:pos2update} and writing it
-with the parameters which were omitted above,
-we obtain:
-\begin{align}
-\nonumber
-\ppos_{t+1} &= \sum_{c \in \caseset} \\
-\nonumber
-& \specpos_{t+1}^{c}(\ppos_{t+1},\leftn_{t+1},\rightn_{t+1},
-\pbest_{t},\nbest_{t},\pURand_{t}, \nURand_{t}) \\
-\label{eq:val2updatelong}
-& \indic_{t}^{c}(\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))
-\end{align}
-In this form the important point to notice is that there are only 7 values (for this ring topology) in the set formed by
-$\specpos_{t+1}^{c} \ c \in\ \caseset$ and that none of them depend upon $f(x_t)$
-or any other objective function evaluation at $t$.
-Thus PSO has been re-factored such that the algorithm can begin computing
-all 7 of the objective function evaluations potentially needed in iteration $t+1$ \emph{before} $f(x_t)$ is computed.
-Once the evaluation of $f(\ppos_{t})$ is completed for all particles only one
-of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ will be set to one.
-We will refer to the indicator function which is set to one as the
-``behavior preserving indicator'' or $bp$, since it indicates which of
-the computations $\ofunc(\specpos_{t}^{c}) \  c \in \caseset$ preservers the behavior of PSO.
-The rest 
-computations $\ofunc(\specpos_{t}^{c}) \ c \in \caseset-\{bp\}$
-will be ignored, and might just as well never have been computed.
-We call these evaluations of $\ofunc(\specpos_{t+1}^{c})$ ``speculative children''
-because they may or may not be needed.
-
-This refactoring comes a significant expense, instead of just computing
-$\ofunc(\ppos_{i,t})$ and then $\ofunc(\ppos_{i,t+1})$ in the usual way, this refactored PSO
-computes $\ofunc(\ppos_{t})$ and all 
-of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$, 8 function evaluations in this case.
-The potential value of these speculative children lies in the phrase used above
-``this re-factored PSO
-can compute $\ofunc(\ppos_{t+1})$ and all 
-of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
-Remember that our intent is to accelerate parallel PSO and that we
-are interested in objective functions that have at least non-trivial computational
-complexity.
-That is, we assume that computing
-$\ofunc(\ppos_{t})$ is the computationally difficult part of one iteration of PSO.
-
-The value of the refactoring can best be seen in a small example.
-Suppose that unlimited processors were available and that the evaluation of 
-the objective function took 1 hour. Suppose also that for each particle
-the computations
-$\ofunc(\ppos_{t})$ and all 
-of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ were sent to separate
-processors and completed in parallel.
-\emph{Given} the values 
-of $\ofunc(\ppos_{t})$ for all particles
-the computation of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ and to complete iteration t+1
-is inconsequential relative to the computational cost of obtaining the function values. 
-In this environment all of the function evaluations needed for 2 iterations of PSO
-could be completed in just one hour where as the conventional implementation would
-require two hours for those same 2 iterations of PSO.
-
-While the algebra shows that the computation of 
-equations~\eqref{eq:vel2update} and \eqref{eq:pos2update} are identical to the 
-equations~\eqref{eq:velupdate} and \eqref{eq:posupdate},
-There are some issues better seen in an algorithmic setting, in particular the use of the random
-vectors, 
-$\pURand_{t}$ and $\nURand_{t}$.
-As with the algebra given above, each transformation (used below) of the algorithm is
-provably equivalent; except that the algorithms will no longer be able to do an odd number of iterations,
-since it will do two at a time. In the algorithmic transformations given below,
-instructions are moved around and added, but the
-resultant algorithm, produces numerically identical results.
-
-We first define a helper function to update the personal and neighborhood bests, $\pbest$ and $\nbest$.
-This algorithm is shown in~\alg{bests}.
-
-\begin{algorithm}
-  \caption{``UpdateBests'': Update personal and neighborhood best values (minimizing)}
-  \label{alg:bests}
-  \begin{algorithmic}[1]
-          \STATE $\pbest_{t}\leftarrow \pbest_{t-1}$
-          \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\pbest_{t-1})$
-          \IF {$\ofunc(\ppos_{t})<\ofunc(\pbest_{t-1})$}
-            \STATE $\pbest_{t}\leftarrow \ppos_{t}$
-            \STATE $\fpbest_{t}\leftarrow \fppos_{t}$
-          \ENDIF
-          \STATE $\nbest_{t}\leftarrow \nbest_{t-1}$
-          \STATE $\ofunc(\nbest_{t})\leftarrow \ofunc(\nbest_{t-1})$
-          \IF {$\ofunc(\fppos_{t})<\ofunc(\nbest_{t})$}
-            \STATE $\pbest_{t}\leftarrow \ppos_{t}$
-            \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\ppos_{t})$
-          \ENDIF
-          \IF {$\ofunc(\leftn_{t})<\ofunc(nbest_{t})$}
-            \STATE $\pbest_{t}\leftarrow \leftn_{t}$
-            \STATE $\ofunc(pbest_{t})\leftarrow \ofunc(\leftn_{t})$
-          \ENDIF
-          \IF {$\ofunc(\rightn_{t})<\ofunc(\nbest_{t})$}
-            \STATE $\pbest_{t}\leftarrow \rightn_{t}$
-            \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\rightn_{t}$
-          \ENDIF
-  \end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm} [ht!]
-  \caption{Standard PSO (minimizing)}
-  \label{alg:standard}
-  \begin{algorithmic}[1]
-    \STATE t=0
-    \STATE Initialize the swarm
-%    \FOR{$i \in \swarm$}
-%      \STATE {$\ppos_{i,t} \leftarrow Random position$}
-%      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
-%      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
-%      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
-%    \ENDFOR
-    \REPEAT
-        \STATE $t=t+1$
-        \FOR{each particle in the swarm}
-          \STATE {$\fppos_{t} \leftarrow \ofunc(\ppos_{t})$}
-        \ENDFOR
-        \FOR{each particle in the swarm}
-          \STATE $\nURand_{t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{t} \Leftarrow RandomVector()$
-          \STATE UpdateBests
-          \STATE Update $\pvel_{t+1}$ as shown in equation~\eqref{eq:velupdate} 
-          \STATE Update $\ppos_{t+1}$ as shown in equation~\eqref{eq:posupdate}
-        \ENDFOR
-    \UNTIL {$t > max \ iterations$ or convergence criteria met}
-  \end{algorithmic}
-\end{algorithm}
-
-Using this helper function we can write PSO as shown in~\alg{standard}.
-This algorithm can be transformed, while retaining it exact behavior, by
-``unrolling'' one iteration of the ``Repeat'' loop, redundantly expressing work done in the ``if'''s
-in~\alg{bests}
-using the indicator functions $\indic_{t+1}$ and re-writing the computation of the next velocity,
-positions and objective function value using 
-equations~\eqref{eq:vel2update}, \eqref{eq:pos2update} and \eqref{eq:val2update} 
-as shown in the new version of the algorithm as shown in~\alg{unrolled}
-
-\begin{algorithm} [ht!]
-  \caption{Unrolled PSO}
-  \label{alg:unrolled}
-  \begin{algorithmic}[1]
-    \STATE t=0
-    \STATE Initialize the swarm including
-%    \FOR{$i \in \swarm$}
-%      \STATE {$\ppos_{i,t} \leftarrow Random position$}
-%      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
-%      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
-%      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
-%    \ENDFOR
-    \REPEAT
-        \STATE $t=t+1$
-        \FOR{each particle in the swarm}
-          \label{alg2:fatitert}
-          \STATE DISPATCH the computation of {$\ofunc(\ppos_{t})$} to a seperate processor
-        \ENDFOR
-        \FOR{each particle in the swarm}
-          \STATE $\nURand_{t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{t} \Leftarrow RandomVector()$
-          \FOR{$c \in \caseset$}
-            \STATE Compute $\specvel_{t+1}^c$ and $\specpos_{t+1}^c$.
-	    Examples shown in equations~\eqref{eq:defvcasexn} and~\eqref{eq:defpcasexn}
-            \label{alg2:fatitertp1}
-            \STATE DISPATCH computation of $\ofunc(\specpos_{t+1}^c)$ to a seperate processor.
-	    Example shown in equation~\eqref{eq:deffcasexn}.
-	    // Note that NONE of the computations, $\specvel_{t+1}^c$,
-	    $\specpos_{t+1}^c$ and $\ofunc(\specpos_{t+1}^c)$ depend on $\ofunc(\ppos_{t})$.
-          \ENDFOR
-        \ENDFOR
-        \STATE WAIT for all function evaluations dispatched above to complete
-        \FOR{each particle in the swarm}
-          \FOR{$c \in \caseset$}
-            \STATE Compute $\indic_{t+1}^c$
-	           Examples shown in equation~\eqref{eq:deficasexn} and~\eqref{eq:defpcasexn}
-	  \ENDFOR 
-            \label{alg2:startcut}
-          \STATE Compute $\pvel_{t+1}$ as shown in equation~\eqref{eq:vel2update} 
-          \STATE Compute $\ppos_{t+1}$ as shown in equation~\eqref{eq:pos2update}
-          \STATE Compute $\ofunc(ppos_{t+1})$ as shown in equation~\eqref{eq:val2update}
-	  \STATE UpdateBests
-            \label{alg2:endcut}
-        \ENDFOR
-
-        \STATE $t=t+1$
-        \FOR{each particle in the swarm}
-          \STATE $\nURand_{t} \Leftarrow RandomVector()$
-          \STATE $\nURand_{t} \Leftarrow RandomVector()$
-          \STATE UpdateBests
-          \STATE Update $\pvel_{t+1}$ as shown in equation~\eqref{eq:velupdate} 
-          \STATE Update $\ppos_{t+1}$ as shown in equation~\eqref{eq:posupdate}
-        \ENDFOR
-    \UNTIL {$t > max \ iteration/2$ or convergence criteria met}
-  \end{algorithmic}
-\end{algorithm}
-
-In this form we can easily see that the random values
-$\pURand_{t}$ and $\nURand_{t}$ usually used just for the computations in
-Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} are passed to all of the
-speculative children, but only the behavior preserving computation will
-pass through the computations in lines
-\ref{alg2:startcut} through
-\ref{alg2:endcut}. The fact that these values are also used by other speculative
-children and later discarded will not change the behavior preserving case.
-
-End of Kevin
-
-While in some algorithms enumerating all possible next sampling positions would
-be impossible or extremely difficult, in PSO all we need to know are the
-current positions of the particle and its neighbors, from which we can easily
-enumerate and evaluate all possible positions of that particle at the next
-iteration.  The defining components of a particle's motion
-(see~\eqref{eq:velupdate}) are $\pvel_t$, $\ppos_t$, $\pbest$, and $\nbest$.
-For a particle at iteration $t-1$, $\pvel_t$ and $\ppos_t$ are defined by
-Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  The only remaining
-components are $\pbest$ and $\nbest$, each of which has only a few
-possibilities.  The particle already knows its personal best $\pbest$ at
-iteration $t-1$, but it could be updated to $\ppos_t$ if its position at
-iteration $t$ has a better value than its current personal best.  Likewise, the
-particle knows its neighborhood best $\nbest$ at iteration $t-1$, though it
-could be updated by any of its neighbors' positions at iteration $t$.
-
-To enumerate all of the possible next positions of a particle, one must simply
-evaluate the motion equations for all possible combinations of personal best
-and neighborhood best updates that a particle could see.  This produces
-$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
-neighbors the particle has.  Because a particle is also its own neighbor, one
-of the $2(n+1)$ speculative positions can be eliminated, as a particle cannot
-be the source of its $\nbest$ update while not updating its $\pbest$.  Thus we
-have $2(n+1)-1$, or $2n+1$, evaluations.  Table~\ref{tab:evals} lists all
-possible combinations of updates that can be seen by a particle with two
-neighbors in addition to itself, which is the case in all of the topologies we
-use.  If all of these speculative positions are evaluated in parallel along
-with the original, non-speculative position of the particle at time $t-1$, then
-one of the speculative positions has to be the correct one, and two iterations
-have been evaluated at once.
+Instead of adding particles we employ a speculative evaluation approach as
+described in the introduction.  This allows us to perform iterations two at a
+time.
+
+To do so we must first refactor PSO such that the determination of the social
+and cognitive components, $\nbest$ and $\pbest$, which depend upon the value of
+the objective function, is separate from the rest of the computation.  For
+simplicity in this explanation, we restrict our discussion here to the case of
+a Ring topology with two neighbors which we will call the ``right neighbor''
+and ``left neighbor,'' whose positions are represented as $\rightn$ and
+$\leftn$ respectively.  Though we will only describe the case of the Ring
+topology, our methods can easily be extended to arbitrary topologies.
+
+Our refactoring hinges on the idea that there is a finite set of possible
+updates to $\nbest$ and $\pbest$ at each iteration.  For the Ring topology we
+are discussing, there are 7 possible update cases, identified in
+Table~\ref{tab:evals}.  We label each case with an identifier referring to the
+source of the update; a hyphen represents no update, $L$ represents an update
+to $\nbest$ coming from the left neighbor, $R$ represents an update to $\nbest$
+coming from the right neighbor, and $S$ represents an update to either $\pbest$
+or $\nbest$ coming from the particle itself.  In the equations that follow, we
+refer to an unspecified update case as $\casegen$, and to the set of cases
+collectively as $\caseset$
 
@@
   \begin{tabular}{lcc}
-	Identifier&$\pbest$ updated&Source of $\nbest$ update\\
+	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
 	\hline
@@
 	\hline
-	$\casexn$&Updated&No update\\
+	$\casexn$&Self&No update\\
 	\hline
-	$\casexl$&Updated&Left Neighbor\\
+	$\casexl$&Self&Left Neighbor\\
 	\hline
-	$\casexr$&Updated&Right Neighbor\\
+	$\casexr$&Self&Right Neighbor\\
 	\hline
-	$\casexx$&Updated&The particle itself\\
+	$\casexx$&Self&Self\\
 	\hline
@@
 
-Because $2n+1$ speculative evaluations must be performed for each particle, the
-choice of topology is an important one.  The use of the Complete topology,
-where every particle is a neighbor to every other particle, would require
-$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
-Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
-much smaller than $O(p^2)$.  However, some functions are better optimized with
-the Complete topology and the quick spread of information it entails than with
-sparse topologies.  Accordingly, we use the Random topology described
-in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
-Complete topology.  In~\sec{results} we report the results for performing
-speculative evaluation using both the Ring topology and the Random topology on
-a number of common benchmark functions.
+Next, note that the computation of $\pbest$ and $\nbest$ (and hence which case
+in $\caseset$ actually occurs) is not seen in the update
+equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  This update is
+usually implemented as a pair of independent if-statements, but that is a bit
+awkward for our purposes.  We will instead make this mathematically explicit by
+introducing a set of indicator functions, each representing a case from
+Table~\ref{tab:evals}.  We can then sum over all of the cases, and the
+indicator function will make all of the terms drop to zero except for the case
+that actually occurs.
+
+In what follows we will examine the indicator function and the position and
+velocity update equations for a specific case, $\casexn$, as presenting
+equations for every case would be tedious.  The extension to cases not
+presented is straightforward.  When we are performing function minimization, we
+can define the indicator function representing the case $\casexn$ by:
+
+\begin{align}
+  \nonumber
+	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
+	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
+  \label{eq:deficasexn}
+	&\begin{cases}
+	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
+	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
+	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
+	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
+	   0 & \text{otherwise}
+	\end{cases}
+\end{align}
+
+For compactness in our equations we represent this without the parameters since
+they can be inferred from the subscripts on $\indic_{t+1}^{\casexn}$.
+
+We now provide notation for the velocity update equation.  Once again we use
+$\casexn$ as an example:
+
+\begin{align}
+\nonumber
+	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
+	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
+\nonumber
+		&= \constriction \bigl[ \pvel_{t} +
+			\pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
+\nonumber
+		& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
+		\ppos_{t}) \bigr] \\
+\nonumber
+		&= \constriction \bigl[ \pvel_{t} +
+			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t}) \\
+\label{eq:defvcasexn}
+			& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
+			\ppos_{t}) \bigr]
+\end{align}
+
+Note that we have substituted the actual values obtained for $\pbest$ and
+$\nbest$ into Equation~\eqref{eq:defvcasexn}.  Since we have restricted
+ourselves to a specific case, we know the values of $\pbest$ and $\nbest$ and
+do not need those terms as parameters to $\specvel_{t+1}^{\casexn}$.  In this
+case, $\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
+position, and $\nbest_{t}=\nbest_{t-1}$, as \nbest was not updated at iteration
+$t$.  Again we will drop the parameters to $\specvel_{t+1}^{\casexn}$ for
+compactness.
+
+In the same way we can create notation for the position update
+equation~\eqref{eq:velupdate}:
+\begin{align}
+\label{eq:defpcasexn}
+	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
+	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
+\nonumber
+	& = \ppos_{t} + \specvel_{t+1}^{\casexn}
+\end{align}
+
+With this notation we can re-write the original PSO velocity
+equation~\eqref{eq:velupdate}, introducing our sum over cases with the
+indicator functions.  The equation becomes:
+\begin{align}
+\nonumber
+	\pvel_{t+1} &=
+		\constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
+\nonumber
+			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
+			\ppos_{t}) \bigr] \\
+\nonumber
+	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
+\nonumber
+			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
+			\ppos_{t}) \bigr]  \\
+\label{eq:vel2update}
+	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
+\end{align}
+
+We can similarly re-write the position update equation~\eqref{eq:posupdate} as:
+\begin{align}
+\nonumber
+	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
+\label{eq:pos2update}
+	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
+\end{align}
+
+The value of the objective function at $\ppos_{t+1}$ is given by:
+\begin{align}
+\label{eq:val2update}
+	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \indic_{t+1}^{c}
+	\ \ofunc(\specpos_{t+1}^{c})
+\end{align}
+
+Returning our attention to the computation of $\ppos_{t+1}$ in
+\eqref{eq:pos2update} and writing it with the parameters which were omitted
+above, we obtain:
+\begin{align}
+\nonumber
+  \ppos_{t+1} &= \sum_{c \in \caseset} \\
+\nonumber
+	& \indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
+	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
+\label{eq:val2updatelong}
+	& \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
+	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t})
+\end{align}
+
+In this form the important point to notice is that there are only 7 values (for
+this Ring topology) in the set formed by $\specpos_{t+1}^{c} \ c \in\ \caseset$
+and that none of them depend upon $f(\ppos_t)$ or any other objective function
+evaluation at iteration $t$ (note also that while there are random numbers in
+the equation, they are assumed fixed once drawn for any particular particle at
+a specific iteration).  Thus PSO has been refactored such that the algorithm
+can begin computing all 7 of the objective function evaluations potentially
+needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is computed.  Once the
+evaluation of $f(\ppos_{t})$ is completed for all particles only one of the
+indicator functions $\indic_{t+1}^{c}$ will be set to 1.  We will refer to
+the indicator function which is set to 1 as the ``behavior preserving
+indicator'' or $bp$, since it indicates which of the computations
+$\ofunc(\specpos_{t+1}^{c})$ preserves the behavior of PSO.  The rest of the
+computations $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset-\{bp\}$ will be
+ignored, and might just as well never have been computed.  We call these
+evaluations of $\ofunc(\specpos_{t+1}^{c})$ ``speculative children'' because
+they may or may not be needed.
+
+This refactoring comes at significant expense; instead of just computing
+$\ofunc(\ppos_{t})$ and then $\ofunc(\ppos_{t+1})$ in the usual way, this
+refactored PSO computes $\ofunc(\ppos_{t})$ and all of the
+$\ofunc(\specpos_{t+1}^{c})$, eight function evaluations in this case.  The
+potential value of these speculative children lies in the fact that we can
+perform the computation needed for both iteration $t$ and iteration $t+1$ at
+the same time.  Remember that our intent is to accelerate parallel PSO and that
+we are interested in objective functions that have at least non-trivial
+computational complexity.  That is, we assume that computing
+$\ofunc(\ppos_{t})$ is the computationally difficult part of one iteration of
+PSO.
+
+The value of the refactoring can best be seen in a small example.  Suppose that
+unlimited processors were available and that the evaluation of the objective
+function took 1 hour. Suppose also that for each particle the computations
+$\ofunc(\ppos_{t})$ and all of the $\ofunc(\specpos_{t+1}^{c})$ were sent to
+separate processors and completed in parallel.  Once the values of
+$\ofunc(\ppos_{t})$ have been computed for all particles, the computation
+needed to determine the values of the indicator functions $\indic_{t+1}^{c}$
+and to complete iteration $t+1$ is inconsequential.  In this environment all of
+the function evaluations needed for 2 iterations of PSO could be completed in
+just one hour where as the conventional implementation would require two hours
+for those same 2 iterations.
+
+However, we do not have an infinite number of processors at our disposal, and
+performing speculative evaluation significantly increases the number of
+function evaluations needed in PSO.  We show in the rest of this paper that in
+many, though not all, instances the extra performance achieved by being able to
+do more iterations at a time outweighs the benefit that would have been gained
+by merely increasing the number of particles in the swarm.
+
+Because the number of speculative evaluations depends on the number of
+neighbors a particle has, the choice of topology is an important one.  The use
+of the Complete topology, where every particle is a neighbor to every other
+particle, would require $O(p^2)$ speculative evaluations per iteration, where
+$p$ is the swarm size.  Clearly it is much more desirable to have a sparse
+topology, where $O(np)$ is much smaller than $O(p^2)$.  However, some functions
+are better optimized with the Complete topology and the quick spread of
+information it entails than with sparse topologies.  Accordingly, we use the
+Random topology described in~\cite{mcnabb-cec09}, which has been shown to
+effectively simulate the Complete topology.  In~\sec{results} we report the
+results for performing speculative evaluation using both the Ring topology and
+the Random topology on a number of common benchmark functions.
 
@@
 a small swarm using a Complete topology.  Accordingly, we show results for a
-Complete swarm in \fig{basic-sphere}.  In comparing SEPSO with PSO, we can see
-that when both are using a Random topology, SEPSO clearly outperforms PSO.
-However, the comparison between SEPSO and PSO Complete is not as
-encouraging---PSO Complete performs slightly better.
+Complete swarm in \fig{basic-sphere}.  We can see that when both SEPSO and PSO
+are using a Random topology, SEPSO clearly outperforms PSO.  However, the
+comparison between SEPSO and PSO Complete is not as encouraging---PSO Complete
+performs slightly better.
 

### 2010-04-05 honors_thesis/journal_paper/paper.tex
Finished revisions of the journal paper.  It is almost ready to submit

@@
 \providecommand{\casepr}{\ensuremath{(-,\rightind)}}
+\providecommand{\casepN}{\ensuremath{(-,N)}}
+\providecommand{\casexN}{\ensuremath{(S,N)}}
 \providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
@@
 To solve this problem, we propose a method for performing two iterations of PSO
-at the same in parallel that we call speculative evaluation.  The name comes
-from an analogy to speculative execution techniques commonly used in
-processors.  Modern processors, when faced with a branch on which they must
-wait (e.g. a memory cache miss), guess which way the branch will go and start
-executing, being careful not to make changes that can not be undone.  If the
-processor guesses right, execution is much farther ahead than if it had just
-idly waited on the memory reference.  If it guesses wrong, execution restarts
-where it would have been anyway.  This is called speculative execution or
-branch prediction.
+at the same time in parallel that we call speculative evaluation.  The name
+comes from an analogy to speculative execution (also known as branch
+prediction) techniques commonly used in processors.  Modern processors, when
+faced with a branch on which they must wait (e.g. a memory cache miss), guess
+which way the branch will go and start executing, being careful not to make
+changes that can not be undone.  If the processor guesses right, execution is
+much farther ahead than if it had just idly waited on the memory reference.  If
+it guesses wrong, execution restarts where it would have been anyway.
 
@@
 \end{align}
-where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
+where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
 numbers drawn from a standard uniform distribution, the \( \otimes \) operator
@@
 topologies used in the literature are the Ring topology and the Complete
-topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one (or
-more) neighbors to either side of it; in the Complete topology, every particle
-is a neighbor to every other particle.  It is generally assumed that, a
-particle is also a neighbor to itself in that its own position and value is
-considered when updating the particle's personal best, $\nbest$.  Thus with $p$
-particles, using the Ring topology each particle has three neighbors: $i-1$,
-$i$ (itself), and $i+1$.  With the Complete topology, each particle has $p$
-neighbors.
+topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
+neighbor to either side of it; in the Complete topology, every particle is a
+neighbor to every other particle.  In all topologies a particle is also a
+neighbor to itself in that its own position and value are considered when
+updating the particle's neighborhood best, $\nbest$.  Thus with $p$ particles,
+using the Ring topology each particle with index $i$ has three neighbors:
+$i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each particle has
+$p$ neighbors.
 
 In this paper we use these topologies as well as a parallel adaptation of the
-Complete topology, called Random, that has been shown to simulate the behavior
-of Complete with far less communication~\cite{mcnabb-cec09}.  In the Random
-topology, each particle randomly picks two other particles to share information
-with at each iteration, along with itself.  Thus in both the Ring and the
-Random topologies, all particles have 3 neighbors.
+Complete topology, called Random, that has been shown to approximate the
+behavior of Complete with far less communication~\cite{mcnabb-cec09}.  In the
+Random topology, each particle randomly picks two other particles to share
+information with at each iteration, along with itself.  Thus in both the Ring
+and the Random topologies, all particles have three neighbors.
 
@@
 To address this issue we present here a different use of additional processors.
-Instead of adding particles we employ a speculative evaluation approach as
-described in the introduction.  This allows us to perform iterations two at a
-time.
+Instead of adding particles we employ a speculative evaluation approach that
+allows us to perform iterations two at a time.
 
@@
 updates to $\nbest$ and $\pbest$ at each iteration.  For the Ring topology we
-are discussing, there are 7 possible update cases, identified in
+are discussing there are 7 possible update cases, identified in
 Table~\ref{tab:evals}.  We label each case with an identifier referring to the
@@
 coming from the right neighbor, and $S$ represents an update to either $\pbest$
-or $\nbest$ coming from the particle itself.  In the equations that follow, we
-refer to an unspecified update case as $\casegen$, and to the set of cases
-collectively as $\caseset$
+or $\nbest$ coming from the particle itself.  As an example, $\casexn$ refers
+to the case that the particle finds a new personal best, but neither it nor its
+neighbors found a position that updated its neighborhood best.  In the
+equations that follow, we refer to an unspecified update case as $\casegen$,
+and to the set of cases collectively as $\caseset$
 
@@
 unlimited processors were available and that the evaluation of the objective
-function took 1 hour. Suppose also that for each particle the computations
+function took one hour. Suppose also that for each particle the computations
 $\ofunc(\ppos_{t})$ and all of the $\ofunc(\specpos_{t+1}^{c})$ were sent to
@@
 and to complete iteration $t+1$ is inconsequential.  In this environment all of
-the function evaluations needed for 2 iterations of PSO could be completed in
+the function evaluations needed for two iterations of PSO could be completed in
 just one hour where as the conventional implementation would require two hours
-for those same 2 iterations.
+for those same two iterations.
 
@@
 
+The number of speculative evaluations needed per particle depends on the number
+of neighbors each particle has.  The number of update cases in a topology where
+each particle has $n$ neighbors is $2(n+1)$; there are two possibilities for
+updates to $\pbest$ (updated by the particle itself and not updated), and $n+1$
+possibilities for updates to $\nbest$ (updated by each neighboring particle and
+not updated).  When the particle is also a neighbor to itself, as is always the
+case in commonly used topologies, one of the cases can be eliminated, as a
+particle cannot be the source of an update to its neighborhood best while not
+also updating its personal best.  Thus we have $2(n+1)-1$, or $2n+1$,
+speculative evaluations per particle.  In a swarm with $p$ particles and $n$
+neighbors per particle, $(2n+1)p$ speculative evaluations are needed.
+
 Because the number of speculative evaluations depends on the number of
@@
 of the Complete topology, where every particle is a neighbor to every other
-particle, would require $O(p^2)$ speculative evaluations per iteration, where
-$p$ is the swarm size.  Clearly it is much more desirable to have a sparse
-topology, where $O(np)$ is much smaller than $O(p^2)$.  However, some functions
-are better optimized with the Complete topology and the quick spread of
-information it entails than with sparse topologies.  Accordingly, we use the
-Random topology described in~\cite{mcnabb-cec09}, which has been shown to
-effectively simulate the Complete topology.  In~\sec{results} we report the
-results for performing speculative evaluation using both the Ring topology and
-the Random topology on a number of common benchmark functions.
+particle, would require $O(p^2)$ speculative evaluations per iteration.
+Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
+much smaller than $O(p^2)$.  However, some functions are better optimized with
+the Complete topology and the quick spread of information it entails than with
+sparse topologies.  Accordingly, we use the Random topology described
+in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
+Complete topology.  In~\sec{results} we report the results for performing
+speculative evaluation using both the Ring topology and the Random topology on
+a number of common benchmark functions.
 
 It is not trivial in some parallel architectures to determine which speculative
-position was the correct next position of each particle.  First we discuss the
-relatively easy case of a centralized parallel PSO algorithm with a master
-computer and many slaves.  In such an architecture, the master keeps track of
-all necessary information with only trivial message passing needed.  Then we
-discuss the more complicated case of a distributed algorithm, where each
-particle is on its own and needs to send and receive messages to and from other
-particles.  Finally we discuss the further complications of a dynamic topology,
-where a particle's neighbors change from one iteration to another.
+position was the correct next position of each particle.  In the rest of this
+section we move from the mathematics of our method to its implementation.
+First we discuss the relatively easy case of a centralized parallel PSO
+algorithm with a master computer and many slaves.  In such an architecture, the
+master keeps track of all necessary information with only trivial message
+passing needed.  Then we discuss the more complicated case of a distributed
+algorithm, where each particle is on its own and needs to send and receive
+messages to and from other particles.  Finally we discuss the further
+complications of a dynamic topology, where a particle's neighbors change from
+one iteration to another.
 
@@
 an architecture, the master generates the positions to evaluate speculatively
-and decides which one to accept for each particle.  The outline of the
-procedure is given in \alg{centralized}.
+as in Equation~\eqref{eq:pos2update}.  After having the slaves evaluate the
+objective function at all necessary positions, the master then decides which
+position to accept for each particle, as in Equation~\eqref{eq:val2updatelong}.
+The outline of the procedure is given in \alg{centralized}.
 
@@
 	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
-	  generate $\noeval{\sset}_{t+1}$
+	  generate $\noeval{\sset}_{t+1}$ according to
+	  Equation~\eqref{eq:pos2update}.
 	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
@@
 	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
-	  the branch taken by $\p_t$
+	  the branch taken by $\p_t$ according to
+	  Equation~\eqref{eq:val2updatelong}.
 	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
@@
 Table~\ref{tab:evals}).  Each of those combinations defines a speculative
-particle, $\s_t$, that can be moved using the motion equations to obtain
-$\noeval{\s}_{t+1}$.  The result is a set of particles $\noeval{\pset}_t$, and
-for each particle a set of speculative children $\noeval{\sset}_{t+1}$, which
-can all be evaluated at once.
+particle, $\s_t$, that can be moved using Equation~\eqref{eq:pos2update} to
+obtain $\noeval{\s}_{t+1}$.  The result is a set of particles
+$\noeval{\pset}_t$, and for each particle a set of speculative children
+$\noeval{\sset}_{t+1}$, which can all be evaluated at once.
 
@@
 
-This is done by first updating each particle $\noeval{\p}_t$ to
-$\nonbest{\p}_t$ and each speculative particle $\noeval{\s}_{t+1}$ to
-$\nonbest{\s}_{t+1}$ by associating the value determined by the slave to the
-particle it came from, and updating that particle's personal best if the value
-reported is better than its previous personal best value.  Then each
-$\nonbest{\p}_t$ must be updated to $\p_t$ by updating its neighborhood best
-position and value with information about each of its neighbors.  This is
-simply the original PSO algorithm, and corresponds to steps 1--5 in
-\alg{centralized}.
+This is done by first updating each particle's personal and neighborhood bests.
+Mathematically, this corresponds to the evaluation of an indicator function
+similar to that found in Equation~\eqref{eq:deficasexn}.  In practice, this is
+done by moving each $\noeval{\p}_t$ to $\nonbest{\p}_t$ and each speculative
+particle $\noeval{\s}_{t+1}$ to $\nonbest{\s}_{t+1}$ by associating the value
+determined by the slave to the particle it came from, and updating that
+particle's personal best if the value reported is better than its previous
+personal best value.  Then each $\nonbest{\p}_t$ must be updated to $\p_t$ by
+updating its neighborhood best position and value with information about each
+of its neighbors.  This is simply the original PSO algorithm, and corresponds
+to steps 1--5 in \alg{centralized}.
 
@@
 Both of those items are now known, and the child that matches the branch is
-selected (step 7 in \alg{centralized}).
+selected as per Equation~\eqref{eq:val2updatelong} (step 7 in
+\alg{centralized}).
 
@@
 Where the Complete topology would normally be used, we use a Random topology in
-our speculative algorithm, as Complete leads to an explosion of speculative
-evaluations.  If speculative evaluation were not being performed, it is
-possible that the Complete topology would be used.  However, the Complete
+our speculative algorithm, as Complete leads to an explosion in the number of
+speculative evaluations.  If speculative evaluation were not being performed,
+it is possible that the Complete topology would be used.  However, the Complete
 topology also requires a very large amount of interprocessor communication in
@@
 iteration.  We show results in \fig{basic-griewank2} for swarms of size 30 and
-240 using the Ring topology.  One can see in the figure that SEPSO algorithm
-finds the optimum faster than the original PSO.  However, because the swarm
-size is so small, SEPSO gets stuck a little less than half of the time.
+240 using the Ring topology.  One can see in the figure that the SEPSO
+algorithm finds the optimum faster than the original PSO.  However, because the
+swarm size is so small, SEPSO gets stuck a little less than half of the time.
 
@@
 To use the extra speculative evaluations, instead of choosing the speculative
-child that matches that branch that the original PSO would have taken, we take
-the child that has the best value.  The methodology is exactly the same as
-above except for the process of choosing which speculative child to accept.
-The branch that PSO would have taken is ignored, but $\p_t$ must still be
-created so it can give the right personal best and neighborhood best values and
-positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
-step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
+child that matches the branch that the original PSO would have taken (the
+``behavior preserving'' branch), we take the child that has the best value.
+The methodology is exactly the same as above except for the process of choosing
+which speculative child to accept.  The only change needed in \alg{centralized}
+is in step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
 $\noeval{\sset}_{t+1}$ instead of with the matching branch.
@@
 with two neighbors could take.  Here we lump all of the neighbors together and
-consider the statistics for the five branches shown in Table~\ref{tab:branches}.
+consider the statistics for the five branches shown in
+Table~\ref{tab:branches}.  In the identifiers, N represents an update to
+$\nbest$ coming from any neighbor.
 
@@
   \centering
-  \begin{tabular}{rc|c}
-	&$\pbest$ updated&Source of $\nbest$ update\\
-	\hline
-	1.&No update&No update\\
-	2.&Updated&No update\\
-	3.&Updated&The particle itself\\
-	4.&No update&Some neighbor\\
-	5.&Updated&Some neighbor\\
+  \begin{tabular}{lcc}
+	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
+	\hline
+	\hline
+	$\casepn$&No update&No update\\
+	\hline
+	$\casexn$&Self&No update\\
+	\hline
+	$\casexx$&Self&Self\\
+	\hline
+	$\casepl$&No update&Some Neighbor\\
+	\hline
+	$\casexl$&Self&Some Neighbor\\
+	\hline
   \end{tabular}
@@
   \centering
+  \scriptsize
   \begin{tabular}{c|c|c|c|c|c|c}
-	Topology&Function&1&2&3&4&5\\
+	Topology&Function&\casepn&\casexn&\casexx&\casepN&\casexN\\
 	\hline
@@
 	\cline{2-7}
-	&Average&{37.6\%}&{6.9\%}&{.2\%}&{45.5\%}&
+	&Average&{37.6\%}&{6.9\%}&{0.2\%}&{45.5\%}&
 	{9.8\%}\\
 	\hline
-	\multicolumn{7}{l}{The numbers in the column headings refer to branch
-	numbers in Table~\ref{tab:branches}.}
   \end{tabular}
@@
 
-Row 1 in Table~\ref{tab:branches} corresponds to the phenomenon commonly called
-stagnation in the literature, but only with a single particle.  As an
-interesting aside, we found that complete stagnation (where the entire swarm is
-stagnant) practically never occurs in PSO, though frequently large percentages
-of the swarm are stagnant.  Iterations where the entire swarm is stagnant
-average less than 2 in 1000 iterations, unless the swarm converges beyond
-machine precision.  When graphs of the best function value flat-line, what
-really has happened is a contraction of the particles' velocity, not stagnation
-in the technical sense.
+Row 1 in Table~\ref{tab:branches}, \casepn, corresponds to the phenomenon
+commonly called stagnation in the literature, but only with a single particle.
+As an interesting aside, we found that complete stagnation (where the entire
+swarm is stagnant) practically never occurs in PSO, though frequently large
+percentages of the swarm are stagnant.  Iterations where the entire swarm is
+stagnant average less than 2 in 1000 iterations, unless the swarm converges
+beyond machine precision.  When graphs of the best function value flat-line,
+what really has happened is a contraction of the particles' velocity, not
+stagnation in the technical sense.
 
 We seek to find the probability of taking any given branch, given whatever
-information is needed: $\prob(B_t|\cdot)$.  In finding these probabilities, we
-do not attempt to derive any distribution from the PSO equations, we simply
-look at empirical distributions.  However, even with empirical distributions,
-the problem with this approach is that it is not clear what information
-influences the probability of taking a branch.  The failure of the naive
-approach seems to say that the previous branch, $B_{t+1}$, is not incredibly
-useful.  We look at two factors that we believe have a significant influence on
-$\prob(B_t)$: topology and function.  Thus we are looking at $\prob(B_t|T,F)$.
+information is needed: $\prob(\caseset_t|\cdot)$.  In finding these
+probabilities, we do not attempt to derive any distribution from the PSO
+equations, we simply look at empirical distributions.  However, even with
+empirical distributions, the problem with this approach is that it is not clear
+what information influences the probability of taking a branch.  The failure of
+the naive approach seems to say that the previous branch, $\caseset_{t-1}$, is
+not incredibly useful.  We look at two factors that we believe have a
+significant influence on $\prob(\caseset_t)$: topology and function.  Thus we
+are looking at $\prob(\caseset_t|T,F)$.
 
@@
 the averages of 3.6 million trials for the branch taken (2.16 million for the
-two with only 450 iterations).  Table~\ref{tab:stats} contains the results,
-with the branch number coming from Table~\ref{tab:branches}.  We have already
-defined the functions Sphere, Griewank, and Rastrigin.  Rosenbrock is defined
-as $f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.
+two with only 450 iterations).  Table~\ref{tab:stats} contains the results.  We
+have already defined the functions Sphere, Griewank, and Rastrigin.  Rosenbrock
+is defined as $f(\Vec{x}) = \sum_{i=1}^D
+\left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.
 
@@
 particular iteration.  For our purposes, it appears that given a topology, the
-probability of selecting a branch and the function are independent, or
-$\prob(B_t|T,F) \approx \prob(B_t|T)$.
+probability of selecting a branch and the function are close to independent, or
+$\prob(\caseset_t|T,F) \approx \prob(\caseset_t|T)$.
 
@@
 the time.  With the Ring topology, we would be right 50\% of the time.
-Branches four and five really correspond to several actual branches, as all of
-the neighbors are lumped together.  The 20\% probability of taking branch four
-with the Ring topology can be split into two branches, as there are only two
-neighbors.  It also turns out that the neighbor that last updated the
-neighborhood best is the most likely to update it next time, so keeping track
-of that information could be fruitful in trying to predict that branch.
+Branches \casepN and \casexN really correspond to several actual branches, as
+all of the neighbors are lumped together.  The 20\% probability of taking
+branch \casepN with the Ring topology can be split into two branches, as there
+are only two neighbors.  It also turns out that the neighbor that last updated
+the neighborhood best is the most likely to update it next time, so keeping
+track of that information could be fruitful in trying to predict that branch.
 
 The statistics for the Complete topology are less promising, as there are 240
-neighbors that branch four splits into, instead of two.  Pruning does, however,
-allow for the possibility of using the Complete topology in speculative
-evaluation while avoiding the explosion in the number of evaluations it would
-otherwise entail.  Though pruning makes its use possible, we still found the
-communication costs in our parallel environment prohibitive, so we do not
-present any results for the Complete topology here.
+neighbors that branch \casepN splits into, instead of two.  Pruning does,
+however, allow for the possibility of using the Complete topology in
+speculative evaluation while avoiding the explosion in the number of
+evaluations it would otherwise entail.  Though pruning makes its use possible,
+we still found the communication costs in our parallel environment prohibitive,
+so we do not present any results for the Complete topology here.
 
@@
 This ignores the fact that the branch might have been wrong; it does not
-matter, because we simply accept the child that had the best value.  In almost
-all of the experiments that we ran, it turned out that picking the best child
+matter, because we simply accept the child that had the best value.  In most of
+the experiments that we ran, it turned out that picking the best child
 performed better than keeping the particle back an iteration.
@@
 The pruning we present uses only the two most probable branches in the Random
-topology, as seen in~\ref{tab:stats}, the two branches where the $\nbest$ value
-was not updated.  Those branches are also convenient in that no messages are
-needed from neighbors in order to produce the positions of the speculative
-particles---in distributed frameworks using several rounds of communication
-(see \sec{distributed}), one of the rounds of communication can be dropped
-entirely.
+topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
+the $\nbest$ value was not updated---\casepn and \casexn.  Those branches are
+also convenient in that no messages are needed from neighbors in order to
+produce the positions of the speculative particles---in distributed frameworks
+using several rounds of communication (see \sec{distributed}), one of the
+rounds of communication can be dropped entirely.
 
@@
 allows results in better results when picking the best speculative child.  
-When rejecting all speculative children no real improvement is seen.
+When rejecting all speculative children no real improvement is seen over the
+original PSO algorithm.
 
@@
 cannot feasibly be implemented, as we can only determine correct branches for
-the first iteration.  Thus when speculating more than one iteration ahead we
-always pick the child that has the highest value.
+the first iteration.  Thus in this case we always pick the child that has the
+highest value.
 
@@
 particle, matching the swarm size of the original speculative algorithm.  The
-seven evaluations we used corresponded to several iterations of branches 1 and
-2 in Table~\ref{tab:branches}.  Two of the evaluations were just one iteration
-ahead, four were two iterations ahead, and one was three iterations ahead.  The
-evaluations that were one iteration ahead were branches 1 and 2; those that
-went two iterations ahead were formed by taking either branch 1 or 2 and then
-branch 1 or 2; and the evaluation going three iterations ahead followed branch
-1 on all three iterations.
+seven evaluations we used corresponded to several iterations of branches
+\casepn and \casexn.  Two of the evaluations were just one iteration ahead,
+four were two iterations ahead, and one was three iterations ahead.  The
+evaluations that were one iteration ahead were branches \casepn and \casexn;
+those that went two iterations ahead were formed by taking either branch
+\casepn or \casexn and then branch \casepn or \casexn; and the evaluation going
+three iterations ahead followed branch \casepn on all three iterations.
 
@@
 ahead.  When the algorithm finds the optimum, it finds it quicker than any
-other method we tried.  However, it also got stuck and failed to find the
+other method we tried.  However, it also gets stuck and fails to find the
 optimum more than any other method.  In order for this method to be
@@
 stagnant almost the entire time.  Such information could be used in a
-TRIBES-like fashion to determine which particles should be removed from the
-swarm, perhaps starting them close to well-performing particles.
+TRIBES-like fashion~\cite{clerc-oep03} to determine which particles should be
+removed from the swarm, perhaps starting them close to well-performing
+particles.
 
@@
 
-We have mentioned that this technique has been previously used to parallelize
-simulated annealing.  Any other optimization algorithm that only depends on 
-current sampling positions when computing the next position to sample can be
-parallelized with this technique.  In particular, genetic algorithms produce
-future generations by combining individuals from the current generation.  With
-a large population size there would be an unwieldy amount of possible future
-individuals, but the potential exists to modify the algorithm to use some kind
-of speculative evaluation.  A related but quite distinct technique for reducing
-computation in evolutionary algorithms was proposed in~\cite{poli-ai06}.
+A technique similar to ours has been previously used to parallelize simulated
+annealing~\cite{witte-tpds91}.  Any other optimization algorithm that only
+depends on current sampling positions when computing the next position to
+sample can be parallelized with this technique.  In particular, genetic
+algorithms produce future generations by combining individuals from the current
+generation.  With a large population size there would be an unwieldy amount of
+possible future individuals, but the potential exists to modify the algorithm
+to use some kind of speculative evaluation.  A related but quite distinct
+technique for reducing computation in evolutionary algorithms was proposed
+in~\cite{poli-ai06}.
 

### 2010-04-05 resume_cv/cv.tex
Added a "graduated" line to my cv
Added a link to the blog post for my Google intern project
Fixed a small typo in my cv
Updated my CV to include the poster presentation I did

@@
 %
-Particle swarm optimization, natural language processing, Bayesian machine
-learning, Markov decision processes, value iteration
+Particle swarm optimization, natural language processing, Latent Dirichlet
+Allocation, Bayesian machine learning, Markov decision processes, value
+iteration
 
@@
         \item Minors in Physics and Mathematics
+        \item Graduated with a 4.0 GPA and University honors (anticipated)
         \item Thesis Topic: \emph{Speculative Evaluation in Particle Swarm
             Optimization}
-        \item Adviser: Professor Kevin Seppi
+        \item Advisor: Professor Kevin Seppi
         \end{innerlist}
@@
 
+	\item Matthew Gardner and Eric Ringger.  Automatic Topic Discovery in 100
+	  Years of General Conference Talks.  Presented at the 50th Anniversary
+	  BYU Studies Symposium, 2010.
+
 	\item Matthew Gardner, Andrew McNabb, and Kevin Seppi.  Speculative
-	  Evaluation in Particle Swarm Optimization.  Submitted to: \emph{The
-	  Genetic and Evolutionary Computation Conference, 2010}.  Awaiting
-	  acceptance notification.
+	  Evaluation in Particle Swarm Optimization.  Submitted to:
+	  \emph{Transactions on Evolutionary Computation}.  Awaiting acceptance
+	  notification.
 \end{bibsection}
@@
 \begin{innerlist}
+
 \item Used machine learning techniques to attribute calls received by
-  advertisers to specific ads played on TV networks.
+  advertisers to specific ads played on TV networks.  Official blog post about
+  my project can be found at:
+  http://google-tmads.blogspot.com/2010/01/track-cost-per-call-data-through-google.html.
 \item Gave several presentations to the machine learning lunch group.

### 2010-04-05 honors_thesis/ppsn_paper/paper.tex
Added another instance of the paper for PPSN

--- 2010-01-12 honors_thesis/gecco_paper/paper.tex
@@
 \title{\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization%
-\thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department
-of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602
-(phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}%
+%\thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department
+%of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602
+%(phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}%
 }
@@
 
-\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
+%\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
+\author{Author names omitted for review}
 
@@
 \end{abstract}
+
+\category{G.1.6}{Numerical Analysis}{Optimization}
+
+\terms{Algorithms}
+
+\keywords{Particle swarm optimization; speculative evaluation; parallel
+optimization}
 
@@
 optimizing the Griewank function, as Complete is prone to premature convergence
-on a local optima.  We show results in \fig{griewank1} for swarms of size 30
-and 240 using the Ring topology. One can see in the figure that both SEPSO
-algorithms initially outperformed PSO, but got stuck in a local optima and were
-eventually outperformed themselves.
+on a local optimum.  Griewank has a global optimum with a value of 0, and
+sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
+showing average function value at each iteration, a more accurate plot for
+Griewank shows the percent of runs that have found the global optimum by each
+iteration.  We show results in \fig{griewank1} for swarms of size 30 and 240
+using the Ring topology.  One can see in the figure that both SEPSO algorithms
+find the optimum faster than the original PSO.  However, because the swarm size
+is so small, SEPSO gets stuck a little less than half of the time.
+Surpisingly, SEPSO Pick Best improves the chance of success with Griewank,
+while also finding the optimum faster.
 
@@
 
-There are some functions for which almost no exploration needs to be done;
-Sphere is an example of such a function, as would be a function where the
-general area of the optimal point is known, but the exact point needs to be
-found.  For such functions the best use of processors is to have a small swarm
-performing speculative evaluation with our second method, where all speculative
-evaluations are used.
+There are some functions for which very little exploration needs to be done;
+Sphere is an example of such a function.  For such functions the best use of
+processors is to have a small swarm performing speculative evaluation with our
+second method, where all speculative evaluations are used.
 

### 2010-04-08 honors_thesis/journal_paper/paper.tex
tweaks in the Introduction and some rewriting in Section II
I wouldn't be shocked if you hate some of these changes, so feel free to
"cherry pick" those that you like.

@@
 faced with a branch on which they must wait (e.g. a memory cache miss), guess
-which way the branch will go and start executing, being careful not to make
-changes that can not be undone.  If the processor guesses right, execution is
-much farther ahead than if it had just idly waited on the memory reference.  If
+which way the branch will go and start executing, ensuring that any changes
+can be undone.  If the processor guesses right, execution is
+much farther ahead than if it had idly waited on the memory reference.  If
 it guesses wrong, execution restarts where it would have been anyway.
@@
 
-The balance of this paper is organized as follows: \sec{pso} describes the
+The balance of this paper is organized as follows. \sec{pso} describes the
 particle swarm optimization algorithm.  \sec{sepso} describes how speculative
@@
 neighborhood best) is the best position the neighbors of the current particle
-have seen~\cite{bratton-sis07}.  \( \ncoeff \), \( \pcoeff \), and \(
-\constriction \) are parameters with prescribed values required to ensure
+have seen~\cite{bratton-sis07}.  The parameters \( \ncoeff \), \( \pcoeff \),
+and \( \constriction \) are given prescribed values required to ensure
 convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-tec02}.
 
-There are many ways of defining the neighbors of a particle.  Changing the
+Changing the
 neighborhood topology has a significant effect on the performance of the
@@
 The PSO algorithm can be trivially parallelized by distributing the processing
-needed for each particle on across processors.  But as we have seen in the
-introduction, for some functions, and for large numbers of processors, just
-adding particles reaches a point of diminishing returns.  That is, adding
-processors does not help us reach any given level of fitness appreciably
-faster.
-
-To address this issue we present here a different use of additional processors.
-Instead of adding particles we employ a speculative evaluation approach that
-allows us to perform iterations two at a time.
-
-To do so we must first refactor PSO such that the determination of the social
-and cognitive components, $\nbest$ and $\pbest$, which depend upon the value of
-the objective function, is separate from the rest of the computation.  For
-simplicity in this explanation, we restrict our discussion here to the case of
-a Ring topology with two neighbors which we will call the ``right neighbor''
-and ``left neighbor,'' whose positions are represented as $\rightn$ and
-$\leftn$ respectively.  Though we will only describe the case of the Ring
-topology, our methods can easily be extended to arbitrary topologies.
-
-Our refactoring hinges on the idea that there is a finite set of possible
-updates to $\nbest$ and $\pbest$ at each iteration.  For the Ring topology we
-are discussing there are 7 possible update cases, identified in
-Table~\ref{tab:evals}.  We label each case with an identifier referring to the
-source of the update; a hyphen represents no update, $L$ represents an update
-to $\nbest$ coming from the left neighbor, $R$ represents an update to $\nbest$
-coming from the right neighbor, and $S$ represents an update to either $\pbest$
-or $\nbest$ coming from the particle itself.  As an example, $\casexn$ refers
-to the case that the particle finds a new personal best, but neither it nor its
-neighbors found a position that updated its neighborhood best.  In the
-equations that follow, we refer to an unspecified update case as $\casegen$,
-and to the set of cases collectively as $\caseset$
+needed for each particle on across processors.  But as we have seen in
+\fig{fig:evals-sphere}, for some functions, and for large numbers of
+processors, just adding particles reaches a point of diminishing returns.
+That is, adding processors does not help us reach any given level of fitness
+appreciably faster.  Instead of adding particles we employ a speculative
+evaluation approach that allows us to perform iterations two at a time.
+
+To do so we must first refactor PSO such that the determination of the value
+of the objective function is separate from the rest of the computation.  For
+simplicity, this discussion will describe the case where PSO is performing
+function minimization using the Ring topology.  In this example, each particle
+has two neighbors, the ``right neighbor'' and ``left neighbor,'' whose
+positions are represented as $\rightn$ and $\leftn$ respectively.  Though we
+will only describe the case of the Ring topology, this method can easily be
+extended to arbitrary topologies.
+
+The refactoring hinges on the idea that once the random coefficients are
+determined, there are only a few possibilities for the new values of updates
+to $\nbest$ and $\pbest$.  For the Ring topology there are 7 possible update
+cases, identified in Table~\ref{tab:evals}.  We label each case with an
+identifier referring to the source of the update: a minus sign ($-$)
+represents no update, $L$ represents an update to $\nbest$ coming from the
+left neighbor, $R$ represents an update to $\nbest$ coming from the right
+neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$ coming
+from the particle itself.  As an example, $\casexn$ refers to the case that
+the particle finds a new personal best, but neither it nor its neighbors found
+a position that updated its neighborhood best.  In the equations that follow,
+we refer to an unspecified update case as $\casegen$, and to the set of cases
+collectively as $\caseset$.
 
@@
 
-Next, note that the computation of $\pbest$ and $\nbest$ (and hence which case
-in $\caseset$ actually occurs) is not seen in the update
-equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  This update is
-usually implemented as a pair of independent if-statements, but that is a bit
-awkward for our purposes.  We will instead make this mathematically explicit by
-introducing a set of indicator functions, each representing a case from
-Table~\ref{tab:evals}.  We can then sum over all of the cases, and the
-indicator function will make all of the terms drop to zero except for the case
-that actually occurs.
-
-In what follows we will examine the indicator function and the position and
-velocity update equations for a specific case, $\casexn$, as presenting
-equations for every case would be tedious.  The extension to cases not
-presented is straightforward.  When we are performing function minimization, we
-can define the indicator function representing the case $\casexn$ by:
+For each case $\casegen \in \caseset$, we introduce an indicator function
+$\indic_{t}^{\casegen}$.  This function is 1 for the realized case from
+Table~\ref{tab:evals} and 0 for the other cases.  We can then sum over all of
+the cases, and the indicator function will make all of the terms drop to zero
+except for the case that actually occurs.  Although this update is usually
+implemented using if-statements, it is convenient to describe it as a
+mathematical expression.  For example, the indicator function for the specific
+case $\casexn$ can be written as follows:
 

### 2010-04-08 honors_thesis/journal_paper/paper.tex
updated references in journal_paper

--- 2010-04-05 honors_thesis/journal_paper/paper.tex
@@
 \IEEEoverridecommandlockouts
-
-% AML bibliography references
-\newcommand{\bib}[1]{/aml/home/mjg82/bib/#1}
 
@@
 The methods they have used include various synchronous algorithms
-\cite{mcnabb-cec07,belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
+\cite{mcnabb-2007-parallel-pso-using-mapreduce,belal-2004-parallel-models-for-pso,chu-2006-intelligent-parallel-pso,jin-2005-pso-antenna-designs,parsopoulos-2004-parallel-vector-evaluated-pso,schutte-2004-parallel-global-optimization-with-pso}
 and asynchronous algorithms
-\cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05}.  Parallelizing the
+\cite{koh-2006-parallel-asynchronous-pso,mostaghim-2006-multi-objective-pso-on-grids,venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
 evaluation of the objective function can also be done in some cases, though
@@
 algorithm makes per evaluation.  For this reason, previous work has recommended
-the use of a swarm size of 50 for this function~\cite{bratton-sis07}.  Thus, in
+the use of a swarm size of 50 for this function~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in
 at least some cases, adding particles indefinitely will not yield an efficient
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{evals_sphere.eps}
+  \includegraphics[width=.8\columnwidth]{evals_sphere}
   \caption{Function Sphere with a swarm with various swarm sizes.}
@@
 %``wall-clock'' time.
-%This approach has been applied to simulated annealing~\cite{witte-tpds91} which like the greedy random walk,
+%This approach has been applied to simulated annealing~\cite{witte-1991-parallel-simulated-annealing-speculative} which like the greedy random walk,
 %depends on positions and a binary test.
@@
 Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
-Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
+Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is used to intelligently search
 a multi-dimensional space by mimicking the swarming and flocking behavior of
@@
 neighborhood best) is the best position the neighbors of the current particle
-have seen~\cite{bratton-sis07}.  \( \ncoeff \), \( \pcoeff \), and \(
+have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  \( \ncoeff \), \( \pcoeff \), and \(
 \constriction \) are parameters with prescribed values required to ensure
-convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-tec02}.
+convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-2002-constricted-pso}.
 
@@
 topologies used in the literature are the Ring topology and the Complete
-topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
+topology~\cite{bratton-2007-defining-a-standard-for-pso}.  In the Ring topology, each particle has one
 neighbor to either side of it; in the Complete topology, every particle is a
@@
 Complete topology, called Random, that has been shown to approximate the
-behavior of Complete with far less communication~\cite{mcnabb-cec09}.  In the
+behavior of Complete with far less communication~\cite{mcnabb-2009-large-particle-swarms}.  In the
 Random topology, each particle randomly picks two other particles to share
@@
 sparse topologies.  Accordingly, we use the Random topology described
-in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
+in~\cite{mcnabb-2009-large-particle-swarms}, which has been shown to effectively simulate the
 Complete topology.  In~\sec{results} we report the results for performing
@@
 merely used to evaluate the objective function at various positions as directed
-by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
+by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform speculative evaluation in such
 an architecture, the master generates the positions to evaluate speculatively
@@
 with Rastrigin, the more particles there are in the swarm, the lower function
-value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
+value it finds, up to at least 4000 particles~\cite{mcnabb-2009-large-particle-swarms}.  Smaller
 swarms get caught in local optima.  Because our speculative algorithm has a
@@
 stagnant almost the entire time.  Such information could be used in a
-TRIBES-like fashion~\cite{clerc-oep03} to determine which particles should be
+TRIBES-like fashion~\cite{clerc-2003-tribes} to determine which particles should be
 removed from the swarm, perhaps starting them close to well-performing
@@
 A technique similar to ours has been previously used to parallelize simulated
-annealing~\cite{witte-tpds91}.  Any other optimization algorithm that only
+annealing~\cite{witte-1991-parallel-simulated-annealing-speculative}.  Any other optimization algorithm that only
 depends on current sampling positions when computing the next position to
@@
 technique for reducing computation in evolutionary algorithms was proposed
-in~\cite{poli-ai06}.
+in~\cite{poli-2006-backward-chaining-evolutionary-algorithms}.
 
@@
 \bibliography{%
-\bib{wolpert-tec97},%
-\bib{merris-01},%
-\bib{pso/clerc-tec02},%
-\bib{pso/kennedy-icnn95},%
-\bib{pso/bratton-sis07},%
-\bib{pso/largeswarms/perez-emc05},%
-\bib{pso/largeswarms/montes-gecco08.bib},%
-\bib{pso/topology/liang-sis05.bib},%
-\bib{pso/parallel/belal-ijicis04},%
-\bib{pso/parallel/chu-sci06},%
-\bib{pso/parallel/jin-aps05},%
-\bib{pso/parallel/koh-ijnme06},%
-\bib{pso/parallel/mostaghim-report06},%
-\bib{pso/parallel/parsopoulos-aia04},%
-\bib{pso/parallel/schutte-ijnme04},%
-\bib{pso/parallel/venter-wcsmo05},%
-\bib{ga/poli-ai06},%
-\bib{sim-anneal/parallel/witte-tpds91.bib},%
-\bib{pso/topology/kennedy-cec02.bib},%
-\bib{pso/topology/mohais-cec04.bib},%
-\bib{pso/topology/jordan-gecco08.bib},%
-\bib{pso/topology/clerc-oep03.bib},%
-\bib{pso/topology/miranda-ijcir08.bib},%
-\bib{pso/topology/mendes-phd04.bib},%
-\bib{aml/mcnabb-cec07},%
-\bib{aml/mcnabb-cec09}}
+../../../bib/aml/bib,%
+../../../bib/pso/general/bib,%
+../../../bib/pso/parallel/bib,%
+../../../bib/pso/topology/bib,%
+../../../bib/sim-anneal/parallel/bib,%
+../../../bib/ga/bib,%
+}
 

### 2010-04-08 honors_thesis/journal_paper/paper.tex

@@
 neighborhood best) is the best position the neighbors of the current particle
-have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  \( \ncoeff \), \( \pcoeff \), and \(
-\constriction \) are parameters with prescribed values required to ensure
+have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \( \ncoeff \), \( \pcoeff \),
+and \( \constriction \) are given prescribed values required to ensure
 convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-2002-constricted-pso}.
 
-There are many ways of defining the neighbors of a particle.  Changing the
+Changing the
 neighborhood topology has a significant effect on the performance of the

### 2010-04-08 honors_thesis/journal_paper/paper.tex
Fixed a typo in paper.tex

--- 2010-04-08 honors_thesis/journal_paper/paper.tex
@@
 faced with a branch on which they must wait (e.g. a memory cache miss), guess
-which way the branch will go and start executing, ensuring that any changes
-can be undone.  If the processor guesses right, execution is
-much farther ahead than if it had idly waited on the memory reference.  If
+which way the branch will go and start executing, being careful not to make
+changes that cannot be undone.  If the processor guesses right, execution is
+much farther ahead than if it had just idly waited on the memory reference.  If
 it guesses wrong, execution restarts where it would have been anyway.
@@
 
-The balance of this paper is organized as follows. \sec{pso} describes the
+The balance of this paper is organized as follows: \sec{pso} describes the
 particle swarm optimization algorithm.  \sec{sepso} describes how speculative
@@
 neighborhood best) is the best position the neighbors of the current particle
-have seen~\cite{bratton-sis07}.  The parameters \( \ncoeff \), \( \pcoeff \),
-and \( \constriction \) are given prescribed values required to ensure
+have seen~\cite{bratton-sis07}.  \( \ncoeff \), \( \pcoeff \), and \(
+\constriction \) are parameters with prescribed values required to ensure
 convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-tec02}.
 
-Changing the
+There are many ways of defining the neighbors of a particle.  Changing the
 neighborhood topology has a significant effect on the performance of the
@@
 The PSO algorithm can be trivially parallelized by distributing the processing
-needed for each particle on across processors.  But as we have seen in
-\fig{fig:evals-sphere}, for some functions, and for large numbers of
-processors, just adding particles reaches a point of diminishing returns.
-That is, adding processors does not help us reach any given level of fitness
-appreciably faster.  Instead of adding particles we employ a speculative
-evaluation approach that allows us to perform iterations two at a time.
-
-To do so we must first refactor PSO such that the determination of the value
-of the objective function is separate from the rest of the computation.  For
-simplicity, this discussion will describe the case where PSO is performing
-function minimization using the Ring topology.  In this example, each particle
-has two neighbors, the ``right neighbor'' and ``left neighbor,'' whose
-positions are represented as $\rightn$ and $\leftn$ respectively.  Though we
-will only describe the case of the Ring topology, this method can easily be
-extended to arbitrary topologies.
-
-The refactoring hinges on the idea that once the random coefficients are
-determined, there are only a few possibilities for the new values of updates
-to $\nbest$ and $\pbest$.  For the Ring topology there are 7 possible update
-cases, identified in Table~\ref{tab:evals}.  We label each case with an
-identifier referring to the source of the update: a minus sign ($-$)
-represents no update, $L$ represents an update to $\nbest$ coming from the
-left neighbor, $R$ represents an update to $\nbest$ coming from the right
-neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$ coming
-from the particle itself.  As an example, $\casexn$ refers to the case that
-the particle finds a new personal best, but neither it nor its neighbors found
-a position that updated its neighborhood best.  In the equations that follow,
-we refer to an unspecified update case as $\casegen$, and to the set of cases
-collectively as $\caseset$.
+needed for each particle on across processors.  But as we have seen in the
+introduction, for some functions, and for large numbers of processors, just
+adding particles reaches a point of diminishing returns.  That is, adding
+processors does not help us reach any given level of fitness appreciably
+faster.
+
+To address this issue we present here a different use of additional processors.
+Instead of adding particles we employ a speculative evaluation approach that
+allows us to perform iterations two at a time.
+
+To do so we must first refactor PSO such that the determination of the social
+and cognitive components, $\nbest$ and $\pbest$, which depend upon the value of
+the objective function, is separate from the rest of the computation.  For
+simplicity in this explanation, we restrict our discussion here to the case of
+a Ring topology with two neighbors which we will call the ``right neighbor''
+and ``left neighbor,'' whose positions are represented as $\rightn$ and
+$\leftn$ respectively.  Though we will only describe the case of the Ring
+topology, our methods can easily be extended to arbitrary topologies.
+
+Our refactoring hinges on the idea that there is a finite set of possible
+updates to $\nbest$ and $\pbest$ at each iteration.  For the Ring topology we
+are discussing there are 7 possible update cases, identified in
+Table~\ref{tab:evals}.  We label each case with an identifier referring to the
+source of the update; a hyphen represents no update, $L$ represents an update
+to $\nbest$ coming from the left neighbor, $R$ represents an update to $\nbest$
+coming from the right neighbor, and $S$ represents an update to either $\pbest$
+or $\nbest$ coming from the particle itself.  As an example, $\casexn$ refers
+to the case that the particle finds a new personal best, but neither it nor its
+neighbors found a position that updated its neighborhood best.  In the
+equations that follow, we refer to an unspecified update case as $\casegen$,
+and to the set of cases collectively as $\caseset$
 
@@
 
-For each case $\casegen \in \caseset$, we introduce an indicator function
-$\indic_{t}^{\casegen}$.  This function is 1 for the realized case from
-Table~\ref{tab:evals} and 0 for the other cases.  We can then sum over all of
-the cases, and the indicator function will make all of the terms drop to zero
-except for the case that actually occurs.  Although this update is usually
-implemented using if-statements, it is convenient to describe it as a
-mathematical expression.  For example, the indicator function for the specific
-case $\casexn$ can be written as follows:
+Next, note that the computation of $\pbest$ and $\nbest$ (and hence which case
+in $\caseset$ actually occurs) is not seen in the update
+equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  This update is
+usually implemented as a pair of independent if-statements, but that is a bit
+awkward for our purposes.  We will instead make this mathematically explicit by
+introducing a set of indicator functions, each representing a case from
+Table~\ref{tab:evals}.  We can then sum over all of the cases, and the
+indicator function will make all of the terms drop to zero except for the case
+that actually occurs.
+
+In what follows we will examine the indicator function and the position and
+velocity update equations for a specific case, $\casexn$, as presenting
+equations for every case would be tedious.  The extension to cases not
+presented is straightforward.  When we are performing function minimization, we
+can define the indicator function representing the case $\casexn$ by:
 

### 2010-04-08 honors_thesis/journal_paper/paper.tex
Some minor fixes to make things work better

--- 2010-04-08 honors_thesis/journal_paper/paper.tex
@@
 which way the branch will go and start executing, being careful not to make
-changes that can not be undone.  If the processor guesses right, execution is
+changes that cannot be undone.  If the processor guesses right, execution is
 much farther ahead than if it had just idly waited on the memory reference.  If
@@
 ../../../bib/sim-anneal/parallel/bib,%
-../../../bib/ga/bib,%
+../../../bib/ga/bib%
 }

### 2010-04-08 honors_thesis/journal_paper/paper.tex
made a one line change to make bibtex happy

@@
 which way the branch will go and start executing, being careful not to make
-changes that cannot be undone.  If the processor guesses right, execution is
+changes that can not be undone.  If the processor guesses right, execution is
 much farther ahead than if it had just idly waited on the memory reference.  If
@@
 ../../../bib/sim-anneal/parallel/bib,%
-../../../bib/ga/bib%
-}
+../../../bib/ga/bib}
 

### 2010-04-08 honors_thesis/journal_paper/paper.tex
Fixed the line breaks in paper.tex after changing to the new bib style

--- 2010-04-08 honors_thesis/journal_paper/paper.tex
@@
 The methods they have used include various synchronous algorithms
-\cite{mcnabb-2007-parallel-pso-using-mapreduce,belal-2004-parallel-models-for-pso,chu-2006-intelligent-parallel-pso,jin-2005-pso-antenna-designs,parsopoulos-2004-parallel-vector-evaluated-pso,schutte-2004-parallel-global-optimization-with-pso}
-and asynchronous algorithms
-\cite{koh-2006-parallel-asynchronous-pso,mostaghim-2006-multi-objective-pso-on-grids,venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
+\cite{mcnabb-2007-parallel-pso-using-mapreduce,
+belal-2004-parallel-models-for-pso, chu-2006-intelligent-parallel-pso,
+jin-2005-pso-antenna-designs, parsopoulos-2004-parallel-vector-evaluated-pso,
+schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
+\cite{koh-2006-parallel-asynchronous-pso,
+mostaghim-2006-multi-objective-pso-on-grids,
+venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
 evaluation of the objective function can also be done in some cases, though
@@
 algorithm makes per evaluation.  For this reason, previous work has recommended
-the use of a swarm size of 50 for this function~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in
-at least some cases, adding particles indefinitely will not yield an efficient
+the use of a swarm size of 50 for this
+function~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least
+some cases, adding particles indefinitely will not yield an efficient
 implementation.
@@
 Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
-Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is used to intelligently search
-a multi-dimensional space by mimicking the swarming and flocking behavior of
-birds and other animals. It is a social algorithm that depends on interaction
-between particles to quickly and consistently approximate the optimal solution
-to a given objective function.
+Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is
+used to intelligently search a multi-dimensional space by mimicking the
+swarming and flocking behavior of birds and other animals. It is a social
+algorithm that depends on interaction between particles to quickly and
+consistently approximate the optimal solution to a given objective function.
 
@@
 neighborhood best) is the best position the neighbors of the current particle
-have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  \( \ncoeff \), \( \pcoeff \), and \(
-\constriction \) are parameters with prescribed values required to ensure
-convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-2002-constricted-pso}.
+have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  \( \ncoeff \), \(
+\pcoeff \), and \( \constriction \) are parameters with prescribed values
+required to ensure convergence (2.05, 2.05, and .73,
+respectively)~\cite{clerc-2002-constricted-pso}.
 
@@
 topologies used in the literature are the Ring topology and the Complete
-topology~\cite{bratton-2007-defining-a-standard-for-pso}.  In the Ring topology, each particle has one
-neighbor to either side of it; in the Complete topology, every particle is a
-neighbor to every other particle.  In all topologies a particle is also a
-neighbor to itself in that its own position and value are considered when
-updating the particle's neighborhood best, $\nbest$.  Thus with $p$ particles,
-using the Ring topology each particle with index $i$ has three neighbors:
-$i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each particle has
-$p$ neighbors.
+topology~\cite{bratton-2007-defining-a-standard-for-pso}.  In the Ring
+topology, each particle has one neighbor to either side of it; in the Complete
+topology, every particle is a neighbor to every other particle.  In all
+topologies a particle is also a neighbor to itself in that its own position and
+value are considered when updating the particle's neighborhood best, $\nbest$.
+Thus with $p$ particles, using the Ring topology each particle with index $i$
+has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
+topology, each particle has $p$ neighbors.
 
@@
 Complete topology, called Random, that has been shown to approximate the
-behavior of Complete with far less communication~\cite{mcnabb-2009-large-particle-swarms}.  In the
-Random topology, each particle randomly picks two other particles to share
-information with at each iteration, along with itself.  Thus in both the Ring
-and the Random topologies, all particles have three neighbors.
+behavior of Complete with far less
+communication~\cite{mcnabb-2009-large-particle-swarms}.  In the Random
+topology, each particle randomly picks two other particles to share information
+with at each iteration, along with itself.  Thus in both the Ring and the
+Random topologies, all particles have three neighbors.
 
@@
 sparse topologies.  Accordingly, we use the Random topology described
-in~\cite{mcnabb-2009-large-particle-swarms}, which has been shown to effectively simulate the
-Complete topology.  In~\sec{results} we report the results for performing
-speculative evaluation using both the Ring topology and the Random topology on
-a number of common benchmark functions.
+in~\cite{mcnabb-2009-large-particle-swarms}, which has been shown to
+effectively simulate the Complete topology.  In~\sec{results} we report the
+results for performing speculative evaluation using both the Ring topology and
+the Random topology on a number of common benchmark functions.
 
@@
 merely used to evaluate the objective function at various positions as directed
-by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform speculative evaluation in such
-an architecture, the master generates the positions to evaluate speculatively
-as in Equation~\eqref{eq:pos2update}.  After having the slaves evaluate the
-objective function at all necessary positions, the master then decides which
-position to accept for each particle, as in Equation~\eqref{eq:val2updatelong}.
-The outline of the procedure is given in \alg{centralized}.
+by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
+speculative evaluation in such an architecture, the master generates the
+positions to evaluate speculatively as in Equation~\eqref{eq:pos2update}.
+After having the slaves evaluate the objective function at all necessary
+positions, the master then decides which position to accept for each particle,
+as in Equation~\eqref{eq:val2updatelong}.  The outline of the procedure is
+given in \alg{centralized}.
 
@@
 with Rastrigin, the more particles there are in the swarm, the lower function
-value it finds, up to at least 4000 particles~\cite{mcnabb-2009-large-particle-swarms}.  Smaller
-swarms get caught in local optima.  Because our speculative algorithm has a
-significantly smaller swarm size, it gets stuck at higher values while the
-larger swarms performing regular PSO continue to improve the best value found.
-\fig{rastrigin} shows the results graphically.
+value it finds, up to at least 4000
+particles~\cite{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
+in local optima.  Because our speculative algorithm has a significantly smaller
+swarm size, it gets stuck at higher values while the larger swarms performing
+regular PSO continue to improve the best value found.  \fig{rastrigin} shows
+the results graphically.
 
@@
 stagnant almost the entire time.  Such information could be used in a
-TRIBES-like fashion~\cite{clerc-2003-tribes} to determine which particles should be
-removed from the swarm, perhaps starting them close to well-performing
-particles.
+TRIBES-like fashion~\cite{clerc-2003-tribes} to determine which particles
+should be removed from the swarm, perhaps starting them close to
+well-performing particles.
 
@@
 A technique similar to ours has been previously used to parallelize simulated
-annealing~\cite{witte-1991-parallel-simulated-annealing-speculative}.  Any other optimization algorithm that only
-depends on current sampling positions when computing the next position to
-sample can be parallelized with this technique.  In particular, genetic
-algorithms produce future generations by combining individuals from the current
-generation.  With a large population size there would be an unwieldy amount of
-possible future individuals, but the potential exists to modify the algorithm
-to use some kind of speculative evaluation.  A related but quite distinct
-technique for reducing computation in evolutionary algorithms was proposed
+annealing~\cite{witte-1991-parallel-simulated-annealing-speculative}.  Any
+other optimization algorithm that only depends on current sampling positions
+when computing the next position to sample can be parallelized with this
+technique.  In particular, genetic algorithms produce future generations by
+combining individuals from the current generation.  With a large population
+size there would be an unwieldy amount of possible future individuals, but the
+potential exists to modify the algorithm to use some kind of speculative
+evaluation.  A related but quite distinct technique for reducing computation in
+evolutionary algorithms was proposed
 in~\cite{poli-2006-backward-chaining-evolutionary-algorithms}.

### 2010-04-08 honors_thesis/journal_paper/paper.tex
more brutality (I hope you like it)

--- 2010-04-08 honors_thesis/journal_paper/paper.tex
@@
 case $\casexn$ can be written as follows:
-
 \begin{align}
@@
 
-For compactness in our equations we represent this without the parameters since
-they can be inferred from the subscripts on $\indic_{t+1}^{\casexn}$.
-
-We now provide notation for the velocity update equation.  Once again we use
-$\casexn$ as an example:
-
+For each case $\casegen \in \caseset$, there is also a corresponding
+velocity update function $\specvel_{t+1}^{\casegen}$.  When the case is
+known, the specific values of $\pbest_t$ and $\nbest_t$ may be substituted
+directly into~\eqref{eq:velupdate}.  For example, in case $\casexn$,
+$\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
+position, and $\nbest_{t}=\nbest_{t-1}$, as $\nbest$ was not updated at
+iteration $t$:
 \begin{align}
@@
 	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
-\nonumber
-		&= \constriction \bigl[ \pvel_{t} +
-			\pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
-\nonumber
-		& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
-		\ppos_{t}) \bigr] \\
 \nonumber
@@
 
-Note that we have substituted the actual values obtained for $\pbest$ and
-$\nbest$ into Equation~\eqref{eq:defvcasexn}.  Since we have restricted
-ourselves to a specific case, we know the values of $\pbest$ and $\nbest$ and
-do not need those terms as parameters to $\specvel_{t+1}^{\casexn}$.  In this
-case, $\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
-position, and $\nbest_{t}=\nbest_{t-1}$, as \nbest was not updated at iteration
-$t$.  Again we will drop the parameters to $\specvel_{t+1}^{\casexn}$ for
-compactness.
-
-In the same way we can create notation for the position update
-equation~\eqref{eq:velupdate}:
+In the same way we can create notation for the position update function by
+substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
+parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
+subscripts on $\indic_{t+1}^{\casegen}$.
 \begin{align}
-\label{eq:defpcasexn}
-	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
+\label{eq:defpcasegen}
+	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
 	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
 \nonumber
-	& = \ppos_{t} + \specvel_{t+1}^{\casexn}
+	& = \ppos_{t} + \specvel_{t+1}^{\casegen}
 \end{align}
@@
 equation~\eqref{eq:velupdate}, introducing our sum over cases with the
-indicator functions.  The equation becomes:
+indicator functions.  Again, we represent the indicator functions and velocity
+functions without the parameters for compactness.  The equation becomes:
 \begin{align}
@@
 
-We can similarly re-write the position update equation~\eqref{eq:posupdate} as:
+Similarly, the position update equation~\eqref{eq:posupdate} becomes:
 \begin{align}
@@
 A particle at iteration $t-1$ that has been moved to iteration $t$ using
-Equations~\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has
+\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has
 not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has
@@
 an architecture, the master generates the positions to evaluate speculatively
-as in Equation~\eqref{eq:pos2update}.  After having the slaves evaluate the
+as in \eqref{eq:pos2update}.  After having the slaves evaluate the
 objective function at all necessary positions, the master then decides which
-position to accept for each particle, as in Equation~\eqref{eq:val2updatelong}.
+position to accept for each particle, as in \eqref{eq:val2updatelong}.
 The outline of the procedure is given in \alg{centralized}.
@@
 	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
-	  Equations~\eqref{eq:velupdate}~and~\eqref{eq:posupdate}
+	  \eqref{eq:velupdate}~and~\eqref{eq:posupdate}
 	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
 	  generate $\noeval{\sset}_{t+1}$ according to
-	  Equation~\eqref{eq:pos2update}.
+	  \eqref{eq:pos2update}.
 	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
@@
 	  the branch taken by $\p_t$ according to
-	  Equation~\eqref{eq:val2updatelong}.
+	  \eqref{eq:val2updatelong}.
 	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
@@
 initialized), the master must move each particle using
-Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
+\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
 $\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
@@
 Table~\ref{tab:evals}).  Each of those combinations defines a speculative
-particle, $\s_t$, that can be moved using Equation~\eqref{eq:pos2update} to
+particle, $\s_t$, that can be moved using \eqref{eq:pos2update} to
 obtain $\noeval{\s}_{t+1}$.  The result is a set of particles
@@
 Mathematically, this corresponds to the evaluation of an indicator function
-similar to that found in Equation~\eqref{eq:deficasexn}.  In practice, this is
+similar to that found in \eqref{eq:deficasexn}.  In practice, this is
 done by moving each $\noeval{\p}_t$ to $\nonbest{\p}_t$ and each speculative
@@
 Both of those items are now known, and the child that matches the branch is
-selected as per Equation~\eqref{eq:val2updatelong} (step 7 in
+selected as per \eqref{eq:val2updatelong} (step 7 in
 \alg{centralized}).

### 2010-04-08 honors_thesis/journal_paper/paper.tex
rewrote my least-favorite paragraph, but is it now better or worse...

@@
 
-In this form the important point to notice is that there are only 7 values (for
-this Ring topology) in the set formed by $\specpos_{t+1}^{c} \ c \in\ \caseset$
-and that none of them depend upon $f(\ppos_t)$ or any other objective function
-evaluation at iteration $t$ (note also that while there are random numbers in
-the equation, they are assumed fixed once drawn for any particular particle at
-a specific iteration).  Thus PSO has been refactored such that the algorithm
-can begin computing all 7 of the objective function evaluations potentially
-needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is computed.  Once the
-evaluation of $f(\ppos_{t})$ is completed for all particles only one of the
-indicator functions $\indic_{t+1}^{c}$ will be set to 1.  We will refer to
-the indicator function which is set to 1 as the ``behavior preserving
-indicator'' or $bp$, since it indicates which of the computations
-$\ofunc(\specpos_{t+1}^{c})$ preserves the behavior of PSO.  The rest of the
-computations $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset-\{bp\}$ will be
-ignored, and might just as well never have been computed.  We call these
-evaluations of $\ofunc(\specpos_{t+1}^{c})$ ``speculative children'' because
-they may or may not be needed.
+In this form the important point to notice is that there are only 7 values
+(for this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
+\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
+objective function evaluation at iteration $t$ (note also that while there are
+random numbers in the equation, they are assumed fixed once drawn for any
+particular particle at a specific iteration).  Thus PSO has been refactored
+such that the algorithm can begin computing all 7 of the objective function
+evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$
+is computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all
+particles only one of the indicator functions $\indic_{t+1}^{\casegen}$ will
+be set to 1.  Although this speculative algorithm computes
+$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even
+those for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will
+be ignored, and might just as well never have been computed.  We call the set
+of computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
+``speculative children'' because only one of them will be needed.
 
@@
 
-The value of the refactoring can best be seen in a small example.  Suppose that
-unlimited processors were available and that the evaluation of the objective
-function took one hour. Suppose also that for each particle the computations
-$\ofunc(\ppos_{t})$ and all of the $\ofunc(\specpos_{t+1}^{c})$ were sent to
-separate processors and completed in parallel.  Once the values of
-$\ofunc(\ppos_{t})$ have been computed for all particles, the computation
+Suppose that unlimited processors were available and that the evaluation of
+the objective function took one hour. Suppose also that for each particle the
+computations $\ofunc(\ppos_{t})$ and all of the $\ofunc(\specpos_{t+1}^{c})$
+were sent to separate processors and completed in parallel.  Once the values
+of $\ofunc(\ppos_{t})$ have been computed for all particles, the computation
 needed to determine the values of the indicator functions $\indic_{t+1}^{c}$
-and to complete iteration $t+1$ is inconsequential.  In this environment all of
-the function evaluations needed for two iterations of PSO could be completed in
-just one hour where as the conventional implementation would require two hours
-for those same two iterations.
+and to complete iteration $t+1$ is inconsequential.  In this environment all
+of the function evaluations needed for two iterations of PSO could be
+completed in just one hour where as the conventional implementation would
+require two hours for those same two iterations.
 

### 2010-04-09 honors_thesis/journal_paper/paper.tex

@@
 neighborhood best) is the best position the neighbors of the current particle
-have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \( \ncoeff \), \( \pcoeff \),
-and \( \constriction \) are given prescribed values required to ensure
-convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-2002-constricted-pso}.
+have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
+\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
+required to ensure convergence (2.05, 2.05, and .73,
+respectively)~\cite{clerc-2002-constricted-pso}.
 
@@
 merely used to evaluate the objective function at various positions as directed
-by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform speculative evaluation in such
-an architecture, the master generates the positions to evaluate speculatively
-as in \eqref{eq:pos2update}.  After having the slaves evaluate the
-objective function at all necessary positions, the master then decides which
-position to accept for each particle, as in \eqref{eq:val2updatelong}.
-The outline of the procedure is given in \alg{centralized}.
+by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
+speculative evaluation in such an architecture, the master generates the
+positions to evaluate speculatively as in \eqref{eq:pos2update}.  After having
+the slaves evaluate the objective function at all necessary positions, the
+master then decides which position to accept for each particle, as in
+\eqref{eq:val2updatelong}.  The outline of the procedure is given in
+\alg{centralized}.
 

### 2010-04-13 honors_thesis/journal_paper/paper.tex
Some fixes to Andrew's fixes; pretty close to done, I think

@@
 
-Changing the
-neighborhood topology has a significant effect on the performance of the
-algorithm and this effect varies widely by function type.  The two most common
-topologies used in the literature are the Ring topology and the Complete
-topology~\cite{bratton-2007-defining-a-standard-for-pso}.  In the Ring
-topology, each particle has one neighbor to either side of it; in the Complete
-topology, every particle is a neighbor to every other particle.  In all
-topologies a particle is also a neighbor to itself in that its own position and
-value are considered when updating the particle's neighborhood best, $\nbest$.
-Thus with $p$ particles, using the Ring topology each particle with index $i$
-has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
-topology, each particle has $p$ neighbors.
+Changing the way neighbors are defined (often called the ``topology'' in the
+literature) has a significant effect on the performance of the algorithm.  In
+the Ring topology, each particle has one neighbor to either side of it; in the
+Complete topology, every particle is a neighbor to every other
+particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
+particle is also a neighbor to itself in that its own position and value are
+considered when updating the particle's neighborhood best, $\nbest$.  Thus with
+$p$ particles, using the Ring topology each particle with index $i$ has three
+neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
+particle has $p$ neighbors.
 
@@
 needed for each particle on across processors.  But as we have seen in
-\fig{fig:evals-sphere}, for some functions, and for large numbers of
+\fig{evals-sphere}, for some functions, and for large numbers of
 processors, just adding particles reaches a point of diminishing returns.
@@
 The refactoring hinges on the idea that once the random coefficients are
-determined, there are only a few possibilities for the new values of updates
-to $\nbest$ and $\pbest$.  For the Ring topology there are 7 possible update
-cases, identified in Table~\ref{tab:evals}.  We label each case with an
-identifier referring to the source of the update: a minus sign ($-$)
-represents no update, $L$ represents an update to $\nbest$ coming from the
-left neighbor, $R$ represents an update to $\nbest$ coming from the right
-neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$ coming
-from the particle itself.  As an example, $\casexn$ refers to the case that
-the particle finds a new personal best, but neither it nor its neighbors found
-a position that updated its neighborhood best.  In the equations that follow,
-we refer to an unspecified update case as $\casegen$, and to the set of cases
-collectively as $\caseset$.
+determined, there are only a few possibile updates to $\nbest$ and $\pbest$.
+For the Ring topology there are 7 possible update cases, identified in
+Table~\ref{tab:evals}.  We label each case with an identifier referring to the
+source of the update: a minus sign ($-$) represents no update, $L$ represents
+an update to $\nbest$ coming from the left neighbor, $R$ represents an update
+to $\nbest$ coming from the right neighbor, and $S$ represents an update to
+either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
+$\casexn$ refers to the case that the particle finds a new personal best, but
+neither it nor its neighbors found a position that updated its neighborhood
+best.  In the equations that follow, we refer to an unspecified update case as
+$\casegen$, and to the set of cases collectively as $\caseset$.
 
@@
 
-For each case $\casegen \in \caseset$, we introduce an indicator function
-$\indic_{t}^{\casegen}$.  This function is 1 for the realized case from
-Table~\ref{tab:evals} and 0 for the other cases.  We can then sum over all of
-the cases, and the indicator function will make all of the terms drop to zero
-except for the case that actually occurs.  Although this update is usually
-implemented using if-statements, it is convenient to describe it as a
-mathematical expression.  For example, the indicator function for the specific
-case $\casexn$ can be written as follows:
+In order to incorporate the determination of which case occurs into the
+position and velocity update equations, we introduce an indicator function
+$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
+$\casegen$ corresponds to the case actually taken by PSO,
+$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
+then sum over all of the cases, and the indicator function will make all of the
+terms drop to zero except for the case that actually occurs.  For example, the
+indicator function for the specific case $\casexn$ can be written as follows:
+
 \begin{align}
@@
 parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
-subscripts on $\indic_{t+1}^{\casegen}$.
+subscripts.
 \begin{align}

### 2010-04-13 honors_thesis/journal_paper/paper.tex
Done with revisions.  I think it's ready to submit

@@
 adding particles.  In \fig{evals-sphere} we show the value obtained after
-20,000 function evaluations (not iterations) as a function of swarm size for
+50,000 function evaluations (not iterations) as a function of swarm size for
 the well-know benchmark function Sphere (20 dimensions, reporting the average
-of thirty runs).  Increasing the swarm size from 10 to 50 has a significant
-effect on the value obtained.  However, increasing the swarm size from 100 to
-200 makes the algorithm less efficient, that is, it reduces the progress the
-algorithm makes per evaluation.  For this reason, previous work has recommended
-the use of a swarm size of 50 for this
-function~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least
-some cases, adding particles indefinitely will not yield an efficient
+of twenty runs).  Increasing the swarm size from 5 to 10 has a significant
+effect on the value obtained.  However, increasing the swarm size from 16 to 30
+makes the algorithm less efficient, that is, it reduces the progress the
+algorithm makes per evaluation.  Other functions show similar trends, though
+often the optimal swarm size is slightly larger.  For this reason, previous
+work has recommended the use of a swarm size of 50 for
+PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
+cases, adding particles indefinitely will not yield an efficient
 implementation.
@@
 
-This refactoring comes at significant expense; instead of just computing
-$\ofunc(\ppos_{t})$ and then $\ofunc(\ppos_{t+1})$ in the usual way, this
-refactored PSO computes $\ofunc(\ppos_{t})$ and all of the
-$\ofunc(\specpos_{t+1}^{c})$, eight function evaluations in this case.  The
-potential value of these speculative children lies in the fact that we can
-perform the computation needed for both iteration $t$ and iteration $t+1$ at
-the same time.  Remember that our intent is to accelerate parallel PSO and that
-we are interested in objective functions that have at least non-trivial
-computational complexity.  That is, we assume that computing
-$\ofunc(\ppos_{t})$ is the computationally difficult part of one iteration of
-PSO.
-
-Suppose that unlimited processors were available and that the evaluation of
-the objective function took one hour. Suppose also that for each particle the
-computations $\ofunc(\ppos_{t})$ and all of the $\ofunc(\specpos_{t+1}^{c})$
-were sent to separate processors and completed in parallel.  Once the values
-of $\ofunc(\ppos_{t})$ have been computed for all particles, the computation
-needed to determine the values of the indicator functions $\indic_{t+1}^{c}$
-and to complete iteration $t+1$ is inconsequential.  In this environment all
-of the function evaluations needed for two iterations of PSO could be
-completed in just one hour where as the conventional implementation would
-require two hours for those same two iterations.
-
-However, we do not have an infinite number of processors at our disposal, and
-performing speculative evaluation significantly increases the number of
-function evaluations needed in PSO.  We show in the rest of this paper that in
-many, though not all, instances the extra performance achieved by being able to
-do more iterations at a time outweighs the benefit that would have been gained
-by merely increasing the number of particles in the swarm.
+To see the value of this refactoring, suppose that 1000 processors are
+available, and that the evaluation of the function to be optimized takes one
+hour.  If we only want a swarm of 100 particles, 900 of the processors would be
+sitting idle for an hour at every iteration, and it would take two hours to run
+two iterations.  If instead we perform speculative evaluation, sending each of
+the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
+$\ofunc(\ppos_{t})$, we would use 800 of the 1000 processors and perform two
+iterations in one hour.
+
+In order to do two iterations at once, we must use 8 times as many processors
+as there are particles in the swarm.  If we were not performing speculative
+evaluation, those processors might very well be used to add particles to the
+swarm.  This then raises the question of whether a swarm of 100 particles doing
+twice as many iterations outperforms a swarm of 800 particles.  We show in the
+rest of this paper that in many, though not all, instances, a smaller swarm
+performing more iterations does in fact outperform a larger swarm.
 

### 2010-04-15 honors_thesis/journal_paper/paper.tex
Ok, a couple of minor things.
Another two-line change
Made a bunch of revisions based on my read through and Andrew's feedback

@@
 parallel with the evaluation of current positions.  We show that a set of
-speculative evaluation threads can be constructed such that the set will always
+speculative evaluations can be constructed such that the set will always
 include the position and corresponding evaluation that would normally have
 occurred at the next iteration of standard PSO.  Thus SEPSO performs two
-iterations of PSO at once, allowing it to achieve better fitness faster.  Not
-only can SEPSO can achieve better fitness in the same number of iterations, it
-also reaches better fitness for the same number of total function evaluations
-in large scale parallel environments.
+iterations of PSO at once, allowing it to achieve better fitness faster.
 
@@
 The methods they have used include various synchronous algorithms
-\cite{mcnabb-2007-parallel-pso-using-mapreduce,
-belal-2004-parallel-models-for-pso, chu-2006-intelligent-parallel-pso,
-jin-2005-pso-antenna-designs, parsopoulos-2004-parallel-vector-evaluated-pso,
+\cite{chu-2006-intelligent-parallel-pso, jin-2005-pso-antenna-designs,
+parsopoulos-2004-parallel-vector-evaluated-pso,
 schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
-\cite{koh-2006-parallel-asynchronous-pso,
-mostaghim-2006-multi-objective-pso-on-grids,
+\cite{mostaghim-2006-multi-objective-pso-on-grids,
 venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
@@
   \includegraphics[width=.8\columnwidth]{evals_sphere}
-  \caption{Function Sphere with a swarm with various swarm sizes.}
+  \caption{Function Sphere with various swarm sizes. Error bars show median and
+  10th and 90th percentiles.}
   \label{fig:evals-sphere}
@@
 it guesses wrong, execution restarts where it would have been anyway.
-
-%\begin{quote}
-%SKIP SKIP
-%
-%No really, try reading without this! The idea of speculative execution can be applied to some types of optimization
-%algorithms.  When the next sampling position of the algorithm does not depend
-%on the value of the objective function at the current position,
-%future positions can be speculated and evaluated in parallel with the current
-%position.
-%This is trivially true for optimization by random sampling, for example.
-%The next random sample does not depend on the previous point nor on the value
-%of the objective function at that point in any way.
-%It depends only on a random number.
-%There is no need to wait to see
-%what objective function value was obtained at the previous point,
-%evaluation of the object function at the next point can begin as soon as the next random number is drawn.
-%In other optimization algorithms, the next position depends on the previous position, but not the
-%value obtained there, a random walk for example.
-%In this case
-%the next position to be evaluated
-%depends only on the previous position and a random number.
-%In this case, as it was in the case of random sampling,
-%the object function evaluation of the next position can be done in parallel
-%because it 
-%does not depend on the \emph{value} of the objective function at the previous position.
-%There are an infinite number of possible next positions, but only one given the draw of the random number.
-%
-%In still other cases, the next position to be evaluated may depend on the current position and some
-%discrete valued (binary, for example) test on the value obtained at the current position.
-%For example, consider a greedy random walk, in which the state only advances if a better
-%position is found.
-%That is, if the current value is better than the previous value,
-%use the sum of the current position and the random number as the next position.
-%Otherwise, computer the next position using the sum of the previous position and the random number.
-%In this case the objective function can be evaluated at three positions:
-%(1) the value of the objective function at the current position,
-%(2) the value at the position if the test turns out to be true and (3) the value at the position
-%if the test turns out to be false; all in parallel.
-%In this example it is important to note that we do not need the value of the objective function at the current
-%point to to reduce the infinite number of possible next positions to just two alternatives.
-%Choosing between these two alternatives can be done post-hoc, after the objective function value
-%at the current position and at the two alternative are computed.
-%For all but the most computationally-trivial objective functions, this approach saves almost half of the
-%``wall-clock'' time.
-%This approach has been applied to simulated annealing~\cite{witte-1991-parallel-simulated-annealing-speculative} which like the greedy random walk,
-%depends on positions and a binary test.
-%
-%This approach can not be used in gradient methods where the
-%step size for sampling typically depends on the gradient and therefor the position and 
-%a set of infinitely many possible \emph{values} outcomes
-%obtained at the current position (not binary or even finite).
-%\end{quote}
-
-% Use this later?
-% Skip this??
-% 
-% Speculative evaluation is possible in PSO because the motion of particles
-% depends only on the \emph{positions} of the particle and its best neighbor, not
-% the \emph{value} of the objective function at those positions.  We do not have
-% to wait for a particle's function evaluation to complete in order to know where
-% it might move to next, so we can compute the set of all possible next positions
-% and send them to be evaluated along with the particle's current position.  Once
-% the current position of each particle has been evaluated, we can determine
-% which of the speculatively evaluated positions matches the branch taken by the
-% particle.  We have thus evaluated two iterations at once.  This can easily get
-% unwieldy if the set of next possible positions for each particle is large, but
-% a wise choice of topology limits that set to a reasonable size.
-% 
-% But the following does need some intro
 
@@
 larger than a typical or reasonable number of particles, producing better
-results in less ``wall-clock'' time and, surprisingly, with fewer total
-function evaluations.
+results in less ``wall-clock'' time.
 
@@
 
-In this form the important point to notice is that there are only 7 values
-(for this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
+In this form the important point to notice is that there are only 7 values (for
+this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
 \caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
@@
 such that the algorithm can begin computing all 7 of the objective function
-evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$
-is computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all
-particles only one of the indicator functions $\indic_{t+1}^{\casegen}$ will
-be set to 1.  Although this speculative algorithm computes
-$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even
-those for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will
-be ignored, and might just as well never have been computed.  We call the set
-of computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
+evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
+computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
+only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
+hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.
+
+Although this speculative algorithm computes
+$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
+for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will be
+ignored, and might just as well never have been computed.  We call the set of
+computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
 ``speculative children'' because only one of them will be needed.
@@
 speculative evaluation in such an architecture, the master generates the
-positions to evaluate speculatively as in \eqref{eq:pos2update}.  After having
+positions to evaluate speculatively as in \eqref{eq:defpcasegen}.  After having
 the slaves evaluate the objective function at all necessary positions, the
@@
 	  generate $\noeval{\sset}_{t+1}$ according to
-	  \eqref{eq:pos2update}.
+	  \eqref{eq:defpcasegen}.
 	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
@@
 Given a set of particles at iteration $t-1$ (perhaps which have just been
-initialized), the master must move each particle using
-\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
-$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
-set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
-position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
-and $\nbest$ that particle $\noeval{\p}_t$ could end up with (as shown in
-Table~\ref{tab:evals}).  Each of those combinations defines a speculative
-particle, $\s_t$, that can be moved using \eqref{eq:pos2update} to
-obtain $\noeval{\s}_{t+1}$.  The result is a set of particles
+initialized), the master must move each particle using \eqref{eq:velupdate} and
+\eqref{eq:posupdate} to obtain the set $\noeval{\pset}_t$.  For each particle
+$\noeval{\p}_t$, the master must then get its set of neighbors
+$\noeval{\nset}_t$ and use their positions, along with the position of
+$\noeval{\p}_t$, to calculate all possible values of
+$\specpos_{t+1}^{\casegen}$, using \eqref{eq:defpcasegen}.  These positions,
+along with the original particle's associated information (such as values for
+$\pbest$ and $\nbest$), define a set of speculative children,
+$\noeval{\sset}_{t+1}$.  The master than has a set of particles
 $\noeval{\pset}_t$, and for each particle a set of speculative children
@@
 master, the master determines which speculative child of each particle was the
-correct one.
-
-This is done by first updating each particle's personal and neighborhood bests.
-Mathematically, this corresponds to the evaluation of an indicator function
-similar to that found in \eqref{eq:deficasexn}.  In practice, this is
-done by moving each $\noeval{\p}_t$ to $\nonbest{\p}_t$ and each speculative
-particle $\noeval{\s}_{t+1}$ to $\nonbest{\s}_{t+1}$ by associating the value
-determined by the slave to the particle it came from, and updating that
-particle's personal best if the value reported is better than its previous
-personal best value.  Then each $\nonbest{\p}_t$ must be updated to $\p_t$ by
-updating its neighborhood best position and value with information about each
-of its neighbors.  This is simply the original PSO algorithm, and corresponds
-to steps 1--5 in \alg{centralized}.
-
-Then, with the set $\pset_t$, the branch that was actually taken by each
-particle can be determined, and the correct speculative child can be chosen.
-The branches depend entirely on whether or not the personal best of $p_t$ was
-updated and which particle, if any, updated the particle's neighborhood best.
-Both of those items are now known, and the child that matches the branch is
-selected as per \eqref{eq:val2updatelong} (step 7 in
-\alg{centralized}).
+correct one.  Mathematically, this corresponds to the evaluation of an
+indicator function similar to that found in \eqref{eq:deficasexn}.  In
+practice, this is done first by updating each particle's $\pbest$, if
+necessary, then by updating the particle's $\nbest$ with information from the
+particle's neighbors.  This is simply the original PSO algorithm, and
+corresponds to steps 1--5 in \alg{centralized}.  Given the updates to $\pbest$
+and $\nbest$, the case from Table~\ref{tab:evals} can be determined, as per
+\eqref{eq:val2updatelong}.  The child with the matching case is kept, and all
+other speculative children are discarded (step 7 in \alg{centralized}.
 
@@
 each processor has the information for the particle or particles that it is in
-charge of and must perform the steps of the algorithm for those particles.
-Messages such as values and positions for the neighborhood best are sent
-between processors.  There may still be some machine that collects information
-from all of the particles and outputs the result of the algorithm, though that
-machine's importance is much less than in centralized algorithms.
+charge of and must perform the steps of the algorithm for those
+particles~\cite{mcnabb-2007-parallel-pso-using-mapreduce}.  Messages such as
+values and positions for the neighborhood best are sent between processors.
+There may still be some machine that collects information from all of the
+particles and outputs the result of the algorithm, though that machine's
+importance is much less than in centralized algorithms.
 
@@
 The speculative algorithm actually performs two iterations of PSO at each
-``iteration,'' so we instead call each ``iteration'' a round of evaluations.
+``iteration,'' so we instead call each ``iteration'' a ``round of
+evaluations''.
 
@@
 speculative evaluations per particle.  With one evaluation needed for the
-original, non-speculative particle, we have $8p_s$ evaluations for every two
-iterations, where $p_s$ is the number of particles in the speculative swarm.
+original, non-speculative particle, we have $8p$ evaluations for every two
+iterations, where $p$ is the number of particles in the speculative swarm.
 The extra evaluations done in speculative evaluation would instead be particles
-in regular PSO, so we compare swarms of size $p_s$ in speculative evaluation
-with swarms of size $8p_s$ in regular PSO.
+in regular PSO, so we compare swarms of size $p$ in speculative evaluation
+with swarms of size $8p$ in regular PSO.
 
@@
 topology also requires a very large amount of interprocessor communication in
-parallel PSO, so it is still quite possible that Random would be used even with
-standard PSO.  But, to be fair in our comparisons, we compare to the standard
-PSO algorithm using both the Random topology with two neighbors and the
-Complete topology.  All functions have 20 dimensions, and all results shown are
-averages over 20 runs.  We use 240 processors in our experiments (except where
-otherwise noted), so our speculative algorithm uses 30 particles and standard
-PSO uses 240 particles.  For ease of labeling, we call our speculative
+distributed PSO, so it is still quite possible that Random would be used even
+with standard PSO.  But, to be fair in our comparisons, we compare to the
+standard PSO algorithm using both the Random topology with two neighbors and
+the Complete topology.  All functions have 20 dimensions, and all results shown
+are averages over 20 runs.  We use 240 processors in our experiments (except
+where otherwise noted), so our speculative algorithm uses 30 particles and
+standard PSO uses 240 particles.  For ease of labeling, we call our speculative
 algorithm SEPSO and standard PSO just PSO, with the topology following.
@@
 optimizing the Griewank function, as Complete is prone to premature convergence
-on a local optimum.  Griewank has a global optimum with a value of 0, and
-sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
-showing average function value at each iteration, a more accurate plot for
-Griewank shows the percent of runs that have found the global optimum by each
-iteration.  We show results in \fig{basic-griewank2} for swarms of size 30 and
-240 using the Ring topology.  One can see in the figure that the SEPSO
-algorithm finds the optimum faster than the original PSO.  However, because the
-swarm size is so small, SEPSO gets stuck a little less than half of the time.
+on a local optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank
+has a global optimum with a value of 0, and sometimes the swarm finds the
+optimum and sometimes it doesn't.  Instead of showing average function value at
+each iteration, a more informative plot for Griewank shows the percent of runs
+that have found the global optimum by each iteration.  We show results in
+\fig{basic-griewank2} for swarms of size 30 and 240 using the Ring topology.
+One can see in the figure that the SEPSO algorithm finds the optimum faster
+than the original PSO.  However, because the swarm size is so small, SEPSO gets
+stuck a little less than half of the time.
 
@@
 
-Because our speculative algorithm got stuck with such a small swarm size, we
+Because the speculative algorithm got stuck with such a small swarm size, we
 ran a set of experiments on Griewank with 800 processors, giving swarms of size
@@
 To use the extra speculative evaluations, instead of choosing the speculative
-child that matches the branch that the original PSO would have taken (the
-``behavior preserving'' branch), we take the child that has the best value.
-The methodology is exactly the same as above except for the process of choosing
-which speculative child to accept.  The only change needed in \alg{centralized}
-is in step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
-$\noeval{\sset}_{t+1}$ instead of with the matching branch.
+child that matches the branch that the original PSO would have taken, we take
+the child that has the best value.  The methodology is exactly the same as
+above except for the process of choosing which speculative child to accept.
+The only change needed in \alg{centralized} is in step 7, where
+$\noeval{\s}_{t+1}$ with the best value is chosen from $\noeval{\sset}_{t+1}$
+instead of with the matching branch.
 
@@
   \includegraphics[width=.8\columnwidth]{griewank4}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
   evaluations, comparing Pick Best with previous results.}
@@
 	\hline
-	$\casepl$&No update&Some Neighbor\\
-	\hline
-	$\casexl$&Self&Some Neighbor\\
+	$\casepN$&No update&Some Neighbor\\
+	\hline
+	$\casexN$&Self&Some Neighbor\\
 	\hline
@@
 Row 1 in Table~\ref{tab:branches}, \casepn, corresponds to the phenomenon
-commonly called stagnation in the literature, but only with a single particle.
+commonly called stagnation in the literature, but with only a single particle.
 As an interesting aside, we found that complete stagnation (where the entire
@@
 the time.  With the Ring topology, we would be right 50\% of the time.
-Branches \casepN and \casexN really correspond to several actual branches, as
+Branches \casepN\ and \casexN\ really correspond to several actual branches, as
 all of the neighbors are lumped together.  The 20\% probability of taking
-branch \casepN with the Ring topology can be split into two branches, as there
+branch \casepN\ with the Ring topology can be split into two branches, as there
 are only two neighbors.  It also turns out that the neighbor that last updated
@@
 The statistics for the Complete topology are less promising, as there are 240
-neighbors that branch \casepN splits into, instead of two.  Pruning does,
+neighbors that branch \casepN\ splits into, instead of two.  Pruning does,
 however, allow for the possibility of using the Complete topology in
@@
 
-\subsection{What if we guessed wrong?}
+\subsection{Recovering from Pruning Too Much}
 \label{sec:wrong}
@@
 When not all of the branches are evaluated, there is some probability of not
-evaluating the branch that was actually taken by the original particle.  We 
-need to decide what to do in this case.
-
-One possibility is to leave the particle as it is, not accepting any of the
-speculative evaluations, because none of them were correct.  This leads to
-particles being at different iterations, as some particles guess correctly
-while others do not.  Thus we lose exact compatibility with the original PSO,
-though this particular relaxation is nothing new; asynchronous adaptations of
-PSO do the same thing.  As an aside, it is equivalent in this case to simply
-increment the iteration number of particles which fail to correctly predict
-their branch.  This keeps the iteration number constant across all particles,
-simplifying the work needed to be done in determining neighbors when dynamic
-topologies are involved.  We call this technique Social Promotion.
+evaluating the branch that was actually taken by the original particle.  There
+are a few possibilities for recovery in this case.  One is to leave the
+particle as it is, not accepting any of the speculative evaluations, because
+none of them were correct.  This leads to particles being at different
+iterations, as some particles guess correctly while others do not.  Thus we
+lose exact compatibility with the original PSO, though this particular
+relaxation is nothing new; asynchronous adaptations of PSO do the same
+thing~\cite{koh-2006-parallel-asynchronous-pso}.  As an aside, it is equivalent
+in this case to simply increment the iteration number of particles which fail
+to correctly predict their branch.  This keeps the iteration number constant
+across all particles, simplifying the work needed to be done in determining
+neighbors when dynamic topologies are involved.  We call this technique Social
+Promotion.
 
@@
 topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
-the $\nbest$ value was not updated---\casepn and \casexn.  Those branches are
+the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
 also convenient in that no messages are needed from neighbors in order to
@@
 function Sphere.  As can be seen, the larger number of particles that pruning
-allows results in better results when picking the best speculative child.  
-When rejecting all speculative children no real improvement is seen over the
-original PSO algorithm.
+allows leads to better results when picking the best speculative child.  When
+using Social Promotion no real improvement is seen over the original PSO
+algorithm.
 
@@
 more fruitful to try.  In our experiments we try just one of the countless
-possibilities, but that one turns out to work very well.  More work is needed
+possibilities, but one that turns out to work very well.  More work is needed
 to compare the different branching possibilities on various functions.
@@
 seven evaluations we used corresponded to several iterations of branches
-\casepn and \casexn.  Two of the evaluations were just one iteration ahead,
+\casepn\ and \casexn.  Two of the evaluations were just one iteration ahead,
 four were two iterations ahead, and one was three iterations ahead.  The
-evaluations that were one iteration ahead were branches \casepn and \casexn;
+evaluations that were one iteration ahead were branches \casepn\ and \casexn;
 those that went two iterations ahead were formed by taking either branch
-\casepn or \casexn and then branch \casepn or \casexn; and the evaluation going
-three iterations ahead followed branch \casepn on all three iterations.
+\casepn\ or \casexn and then branch \casepn\ or \casexn; and the evaluation
+going three iterations ahead followed branch \casepn\ on all three iterations.
 
@@
 algorithm.  With such functions, the smaller swarm size required by speculative
-evaluation is not able to produce enough exploration to perform better than
-standard PSO at the swarm sizes we were able to experiment with.
+evaluation does not produce enough exploration to perform better than standard
+PSO at the swarm sizes we were able to experiment with.
 
@@
 them.  It would then be expected that the sampling distribution of that method
-is very similar to the distribution of the original PSO.  This remains to be
-proven, or even demonstrated empirically.
+is very similar to the distribution of the original
+PSO~\cite{poli-2008-sampling-distribution-of-pso}.  This remains to be proven,
+or even demonstrated empirically.
 

### 2010-04-19 honors_thesis/ppsn_paper/paper.tex
Got the style file for PPSN and made some simple changes to the paper

@@
-\documentclass[letterpaper]{sig-alt-release2}
+\documentclass{llncs}
 
@@
 \DeclareMathOperator*{\argsup}{arg\;sup}
+\providecommand{\pers}{\ensuremath{P}}
+\providecommand{\neigh}{\ensuremath{N}}
+\providecommand{\leftind}{\ensuremath{L}}
+\providecommand{\rightind}{\ensuremath{R}}
+\providecommand{\nURand}{\ensuremath{U^\neigh}}
+\providecommand{\pURand}{\ensuremath{U^\pers}}
 \providecommand{\ppos}{\ensuremath{\Vec{x}}}
+\providecommand{\fppos}{\ensuremath{\Vec{fx}}}
 \providecommand{\pvel}{\ensuremath{\Vec{v}}}
-\providecommand{\nbest}{\ensuremath{\Vec{b}_N}}
-\providecommand{\pbest}{\ensuremath{\Vec{b}_P}}
+\providecommand{\nbest}{\ensuremath{\Vec{x}^\neigh}}
+\providecommand{\fnbest}{\ensuremath{\Vec{fx}^\neigh}}
+\providecommand{\pbest}{\ensuremath{\Vec{x}^\pers}}
+\providecommand{\fpbest}{\ensuremath{\Vec{fx}^\pers}}
 \providecommand{\constriction}{\ensuremath{\chi}}
-\providecommand{\coeff}{\ensuremath{\phi}}
+\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
+\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
 \providecommand{\obs}{\ensuremath{\Vec{\xi}}}
+\providecommand{\ofunc}{\ensuremath{f}}
+\providecommand{\swarm}{\ensuremath{swarm}}
 
 %SpecExPSO Stuff
+\providecommand{\indic}{\ensuremath{I}}
+\providecommand{\specvel}{\ensuremath{\vec{V}}}
+\providecommand{\specpos}{\ensuremath{\vec{X}}}
+\providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
+\providecommand{\rightn}{\ensuremath{\Vec{x}^\rightind}}
+\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
+\providecommand{\casegen}{\ensuremath{c}}
+\providecommand{\casedef}{\ensuremath{(\pbest,\nbest)}}
+\providecommand{\casexn}{\ensuremath{(S,-)}}
+\providecommand{\casexx}{\ensuremath{(S,S)}}
+\providecommand{\casexl}{\ensuremath{(S,\leftind)}}
+\providecommand{\casexr}{\ensuremath{(S,\rightind)}}
+\providecommand{\casepn}{\ensuremath{(-,-)}}
+\providecommand{\casepl}{\ensuremath{(-,\leftind)}}
+\providecommand{\casepr}{\ensuremath{(-,\rightind)}}
+\providecommand{\casepN}{\ensuremath{(-,N)}}
+\providecommand{\casexN}{\ensuremath{(S,N)}}
 \providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
@@
 
-\category{G.1.6}{Numerical Analysis}{Optimization}
-
-\terms{Algorithms}
-
-\keywords{Particle swarm optimization; speculative evaluation; parallel
-optimization}
-
 \section{Introduction}
@@
 the world's computing power is increasingly coming from large clusters of
-processors.  Even in desktop machines, systems with 4 or 8 cores are
-commonplace.  In order to efficiently utilize these resources for
-computationally intensive problems, algorithms need to run in parallel.
+processors.  In order to efficiently utilize these resources for
+computationally intensive problems, PSO needs to run in parallel.
 
@@
 The methods they have used include various synchronous algorithms
-\cite{belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
-asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05},
-and algorithms based on Google's MapReduce framework \cite{mcnabb-cec07}.
-Parallelizing the evaluation of the objective function can also be done in some
-cases, though that is not an adaption of the PSO algorithm itself and thus is
-not the focus of this paper.
+\cite{chu-2006-intelligent-parallel-pso, jin-2005-pso-antenna-designs,
+parsopoulos-2004-parallel-vector-evaluated-pso,
+schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
+\cite{mostaghim-2006-multi-objective-pso-on-grids,
+venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
+evaluation of the objective function can also be done in some cases, though
+that is not an adaption of the PSO algorithm itself and thus is not the focus
+of this paper.
 
-All of these parallel techniques use additional processors to increase the
-number of particles in the swarm, either by adding individual particles or by
-adding entire new subswarms.  The number of iterations of PSO that the
-algorithms can perform is thus inherently limited by the time it takes to
-evaluate the objective function---additional processors add more particles, not
-more iterations.  However, there are many functions for which performing more
-iterations with fewer particles is a better use of resources than performing
-fewer iterations with more particles.
+These previous parallel techniques distribute the computation needed by the
+particles in the swarm over the available processors.  If more processors are
+available, these techniques increase the number of particles in the swarm,
+either by adding individual particles or by adding entire new sub-swarms, but
+in all cases the number of processors never exceeds the number of particles.
+The number of iterations of PSO that the algorithms can perform is thus
+inherently limited by the time it takes to evaluate the objective
+function---additional processors add more particles, but do not make the
+iterations go any faster.
 
-To solve this problem, we take inspiration from speculative execution
-techniques commonly used in processors.  Modern processors, when faced with a
-branch on which they must wait (e.g. a memory cache miss), guess which way the
-branch will go and start executing.  If the processor guessed right, execution
-is much farther ahead than if it had just idly waited on the branch.  If it
-guessed wrong, execution restarts right where it would have been anyway.  This
-is called speculative execution or branch prediction.
+For many functions there comes a point of diminishing returns with respect to
+adding particles.  In \fig{evals-sphere} we show the value obtained after
+50,000 function evaluations (not iterations) as a function of swarm size for
+the well-know benchmark function Sphere (20 dimensions, reporting the average
+of twenty runs).  Increasing the swarm size from 5 to 10 has a significant
+effect on the value obtained.  However, increasing the swarm size from 16 to 30
+makes the algorithm less efficient, that is, it reduces the progress the
+algorithm makes per evaluation.  Other functions show similar trends, though
+often the optimal swarm size is slightly larger.  For this reason, previous
+work has recommended the use of a swarm size of 50 for
+PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
+cases, adding particles indefinitely will not yield an efficient
+implementation.
 
-The idea of speculative execution can be applied to some types of optimization
-algorithms.  When the next sampling position of the algorithm does not depend
-explicitly on the value of the objective function at the current position,
-future positions can be speculated and evaluated in parallel with the current
-position.  This could not be done in gradient methods, for instance, where the
-step size for sampling depends on the gradient at the current position, but it
-has been done in simulated annealing~\cite{witte-tpds91}.
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{evals_sphere}
+  \caption{Function Sphere with various swarm sizes. Error bars show median and
+  10th and 90th percentiles.}
+  \label{fig:evals-sphere}
+\end{figure}
 
-Speculative evaluation is possible in PSO because the motion of particles
-depends only on the \emph{positions} of the particle and its best neighbor, not
-the \emph{value} of the objective function at those positions.  We do not have
-to wait for a particle's function evaluation to complete in order to know where
-it might move to next, so we can compute the set of all possible next positions
-and send them to be evaluated along with the particle's current position.  Once
-the current position of each particle has been evaluated, we can determine
-which of the speculatively evaluated positions matches the branch taken by the
-particle.  We have thus evaluated two iterations at once.  This can easily get
-unwieldy if the set of next possible positions for each particle is large, but
-a wise choice of topology limits that set to a reasonable size.
+In this paper we consider PSO parallelization strategies for clusters of
+thousands of processors and functions for which a single evaluation will take
+at least tens of seconds, but probably minutes or perhaps hours.  Our purpose
+is to explore the question of what to do with a thousand processors when 50 or
+100 particles is the recommended swarm size, and simply increasing the swarm
+size encounters diminishing returns.
 
-The original PSO can be reproduced exactly, two iterations at a time, using
-this technique.  However, in strictly using speculative evaluation to reproduce
-the original PSO, a lot of positions are speculatively evaluated whose values
-get ignored after deciding which branch was the correct one.  A better use of
-those speculative evaluations may be to incorporate the information into the
-swarm, no longer exactly reproducing PSO but in some instances producing better
-results.
+To solve this problem, we propose a method for performing two iterations of PSO
+at the same time in parallel that we call speculative evaluation.  The name
+comes from an analogy to speculative execution (also known as branch
+prediction) techniques commonly used in processors.  Modern processors, when
+faced with a branch on which they must wait (e.g. a memory cache miss), guess
+which way the branch will go and start executing, ensuring that any changes
+can be undone.  If the processor guesses right, execution is
+much farther ahead than if it had idly waited on the memory reference.  If
+it guesses wrong, execution restarts where it would have been anyway.
 
-\sec{pso} describes the particle swarm optimization algorithm.  \sec{sepso}
-then describes how speculative evaluation can be done in parallel PSO to
-perform two iterations at once.  In \sec{extra}, we discuss a variant of the
-speculative algorithm which uses all of the speculative evaluations performed
-instead of throwing away all but one for each particle.  We then show in
-\sec{results} a comparison between the original PSO algorithm, our speculative
-algorithm, and its variant.  In \sec{conclusion} and \sec{future} we conclude
-and discuss future work.
+In this paper we show that the results of standard PSO can be reproduced
+\emph{exactly}, two iterations at a time, using a speculative approach adapted
+from speculative execution. We prove that the standard PSO equation can be
+factored such that a set of speculative positions can be found which will
+\emph{always} include the position computed in the next iteration.  By
+computing the value of the objective function for each of the speculative
+positions at the same time the algorithm evaluates the objective function for
+the current position, it is possible to know the objective function values for
+both the current and the next iteration at the same time.  We demonstrate this
+principle by implementation and show that it produces exactly the same results
+as standard PSO, but two iterations at a time.  The resulting implementation
+runs efficiently on large clusters where the number of processors is much
+larger than a typical or reasonable number of particles, producing better
+results in less ``wall-clock'' time.
+
+The balance of this paper is organized as follows. \sec{pso} describes the
+particle swarm optimization algorithm.  \sec{sepso} describes how speculative
+evaluation can be done in parallel PSO to perform two iterations at once, then
+presents a comparison of the speculative algorithm with the original PSO.  In
+\sec{conclusion} and \sec{future} we conclude and discuss future work.
 
@@
 Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
-Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
-a multi-dimensional space by mimicking the swarming and flocking behavior of
-birds and other animals. It is a social algorithm that depends on interaction
-between particles to quickly and consistently find the optimal solution to a
-given objective function.
+Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is
+used to intelligently search a multi-dimensional space by mimicking the
+swarming and flocking behavior of birds and other animals. It is a social
+algorithm that depends on interaction between particles to quickly and
+consistently approximate the optimal solution to a given objective function.
 
@@
 \begin{align}
+\nonumber
+	\pvel_{t+1} &=
+		\constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
 \label{eq:velupdate}
-	\pvel_{t+1} &=
-		\constriction \left[ \pvel_t +
-			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
-			\coeff_2\URand()\otimes(\nbest - \ppos_t)
-		\right] \\
+			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
+		\bigr] \\
 \label{eq:posupdate}
-	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
+	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
 \end{align}
-where \( \URand() \) is a vector of random numbers drawn from a uniform
-distribution, the \( \otimes \) operator is an element-wise vector
-multiplication, $\pbest$ (called personal best) is the best position the
-current particle has seen, and $\nbest$ (called neighborhood best) is the best
-position the neighbors of the current particle have seen.  \( \coeff_1 \), \(
-\coeff_2 \), and \( \constriction \) are parameters with prescribed values
+where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
+numbers drawn from a standard uniform distribution, the \( \otimes \) operator
+is an element-wise vector multiplication, $\pbest$ (called personal best) is
+the best position the current particle has seen, and $\nbest$ (called
+neighborhood best) is the best position the neighbors of the current particle
+have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
+\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
 required to ensure convergence (2.05, 2.05, and .73,
-respectively)~\cite{clerc-tec02}.
+respectively)~\cite{clerc-2002-constricted-pso}.
 
-There are many ways of defining the neighbors of a particle.  Different
-neighborhood topologies have varying performance on different kinds of
-functions, as would be expected by the No Free Lunch
-theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
-beyond the scope of this paper.  The two most common topologies used in the
-literature are the Ring topology and the Complete
-topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
-as a parallel adaptation of the Complete topology called Random that has been
-shown to simulate the behavior of Complete with far less
-communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
-be defined as having $k$ neighbors; we use 2 neighbors for all of our
-experiments.
+Changing the way neighbors are defined (often called the ``topology'' in the
+literature) has a significant effect on the performance of the algorithm.  In
+the Ring topology, each particle has one neighbor to either side of it; in the
+Complete topology, every particle is a neighbor to every other
+particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
+particle is also a neighbor to itself in that its own position and value are
+considered when updating the particle's neighborhood best, $\nbest$.  Thus with
+$p$ particles, using the Ring topology each particle with index $i$ has three
+neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
+particle has $p$ neighbors.
+
+In this paper we use these topologies as well as a parallel adaptation of the
+Complete topology, called Random, that has been shown to approximate the
+behavior of Complete with far less
+communication~\cite{mcnabb-2009-large-particle-swarms}.  In the Random
+topology, each particle randomly picks two other particles to share information
+with at each iteration, along with itself.  Thus in both the Ring and the
+Random topologies, all particles have three neighbors.
 
@@
 \label{sec:sepso}
-\subsection{Overview}
 
-The PSO algorithm just described can be parallelized by having each additional
-processor add a particle or a set of particles to the swarm.  But for some
-functions, a more desirable use of resources is to speed up the algorithm,
-performing more iterations instead of adding more particles.  Barring
-parallelizing the evaluation of the objective function itself, which may not be
-possible, previous techniques have not addressed this issue.
+The PSO algorithm can be trivially parallelized by distributing the processing
+needed for each particle on across processors.  But as we have seen in
+\fig{evals-sphere}, for some functions, and for large numbers of
+processors, just adding particles reaches a point of diminishing returns.
+That is, adding processors does not help us reach any given level of fitness
+appreciably faster.  Instead of adding particles we employ a speculative
+evaluation approach that allows us to perform iterations two at a time.
 
-In order to perform additional iterations by adding additional processors, we
-take inspiration from speculative execution in processors.  If we can guess
-what a particle's next position will be from its current position, we can
-evaluate both in parallel and perform two iterations of PSO at once.  Lacking
-a good way to guess a particle's next position, we can at least enumerate all
-possible next positions of the particle, evaluate all of them at once, and one
-of them will be right.
+To do so we must first refactor PSO such that the determination of the value
+of the objective function is separate from the rest of the computation.  For
+simplicity, this discussion will describe the case where PSO is performing
+function minimization using the Ring topology.  In this example, each particle
+has two neighbors, the ``right neighbor'' and ``left neighbor,'' whose
+positions are represented as $\rightn$ and $\leftn$ respectively.  Though we
+will only describe the case of the Ring topology, this method can easily be
+extended to arbitrary topologies.
 
-While in some algorithms enumerating all possible next sampling positions would
-be impossible or extremely difficult, in PSO all we need to know are the
-current positions of the particle and its neighbors, from which we can easily
-enumerate and evaluate all possible positions of that particle at the next
-iteration.  The defining components of a particle's motion
-(see~\eqref{eq:velupdate}) are $\pvel_t$, $\ppos_t$, $\pbest$, and $\nbest$.
-For a particle at iteration $t-1$, $\pvel_t$ and $\ppos_t$ are defined by
-Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  The only remaining
-components are $\pbest$ and $\nbest$, each of which has only a few
-possibilities.  The particle already knows its personal best $\pbest$ at
-iteration $t-1$, but it could be updated to $\ppos_t$ if its position at
-iteration $t$ has a better value than its current personal best.  Likewise, the
-particle knows its neighborhood best $\nbest$ at iteration $t-1$, though it
-could be updated by any of its neighbors' positions at iteration $t$.
-
-To enumerate all of the possible next positions of a particle, one must simply
-evaluate the motion equations for all possible combinations of personal best
-and neighborhood best updates that a particle could see.  This produces
-$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
-neighbors the particle has.  If a particle is also its own neighbor, as is most
-often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
-a particle cannot be the source of its $\nbest$ update while not updating its
-$\pbest$.  Thus we have $2(n+1)-1$ evaluations.  Table~\ref{tab:evals} lists
-all possible combinations of updates that can be seen by a particle with two
-neighbors other than itself, which is the case in all of the topologies we use.
-If all of these speculative positions are evaluated in parallel along with the
-original, non-speculative position of the particle at time $t-1$, then one of
-the speculative positions has to be the correct one, and two iterations have
-been evaluated at once.
+The refactoring hinges on the idea that once the random coefficients are
+determined, there are only a few possibile updates to $\nbest$ and $\pbest$.
+For the Ring topology there are 7 possible update cases, identified in
+Table~\ref{tab:evals}.  We label each case with an identifier referring to the
+source of the update: a minus sign ($-$) represents no update, $L$ represents
+an update to $\nbest$ coming from the left neighbor, $R$ represents an update
+to $\nbest$ coming from the right neighbor, and $S$ represents an update to
+either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
+$\casexn$ refers to the case that the particle finds a new personal best, but
+neither it nor its neighbors found a position that updated its neighborhood
+best.  In the equations that follow, we refer to an unspecified update case as
+$\casegen$, and to the set of cases collectively as $\caseset$.
 
 \begin{table}
-  \caption{All possible updates for a particle with two neighbors.}
+  \caption{All possible updates for a particle with two neighbors}
   \label{tab:evals}
   \centering
-  \begin{tabular}{cc}
-	$\pbest$ updated&Source of $\nbest$ update\\
+  \begin{tabular}{lcc}
+	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
 	\hline
 	\hline
-	No update&No update\\
+	$\casepn$&No update&No update\\
 	\hline
-	No update&Neighbor 1\\
+	$\casepl$&No update&Left Neighbor\\
 	\hline
-	No update&Neighbor 2\\
+	$\casepr$&No update&Right Neighbor\\
 	\hline
-	Updated&No update\\
+	$\casexn$&Self&No update\\
 	\hline
-	Updated&Neighbor 1\\
+	$\casexl$&Self&Left Neighbor\\
 	\hline
-	Updated&Neighbor 2\\
+	$\casexr$&Self&Right Neighbor\\
 	\hline
-	Updated&The particle itself\\
+	$\casexx$&Self&Self\\
 	\hline
@@
 
-Because $2(n+1)-1$ speculative evaluations must be performed for each particle,
-the choice of topology is an important one.  The use of the Complete topology,
-where every particle is a neighbor to every other particle, would require
-$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
+In order to incorporate the determination of which case occurs into the
+position and velocity update equations, we introduce an indicator function
+$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
+$\casegen$ corresponds to the case actually taken by PSO,
+$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
+then sum over all of the cases, and the indicator function will make all of the
+terms drop to zero except for the case that actually occurs.  For example, the
+indicator function for the specific case $\casexn$ can be written as follows:
+
+\begin{align}
+  \nonumber
+	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
+	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
+  \label{eq:deficasexn}
+	&\begin{cases}
+	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
+	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
+	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
+	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
+	   0 & \text{otherwise}
+	\end{cases}
+\end{align}
+
+For each case $\casegen \in \caseset$, there is also a corresponding
+velocity update function $\specvel_{t+1}^{\casegen}$.  When the case is
+known, the specific values of $\pbest_t$ and $\nbest_t$ may be substituted
+directly into~\eqref{eq:velupdate}.  For example, in case $\casexn$,
+$\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
+position, and $\nbest_{t}=\nbest_{t-1}$, as $\nbest$ was not updated at
+iteration $t$:
+\begin{align}
+\nonumber
+	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
+	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
+\nonumber
+		&= \constriction \bigl[ \pvel_{t} +
+			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t}) \\
+\label{eq:defvcasexn}
+			& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
+			\ppos_{t}) \bigr]
+\end{align}
+
+In the same way we can create notation for the position update function by
+substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
+parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
+subscripts.
+\begin{align}
+\label{eq:defpcasegen}
+	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
+	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
+\nonumber
+	& = \ppos_{t} + \specvel_{t+1}^{\casegen}
+\end{align}
+
+With this notation we can re-write the original PSO velocity
+equation~\eqref{eq:velupdate}, introducing our sum over cases with the
+indicator functions.  Again, we represent the indicator functions and velocity
+functions without the parameters for compactness.  The equation becomes:
+\begin{align}
+\nonumber
+	\pvel_{t+1} &=
+		\constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
+\nonumber
+			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
+			\ppos_{t}) \bigr] \\
+\nonumber
+	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
+\nonumber
+			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
+			\ppos_{t}) \bigr]  \\
+\label{eq:vel2update}
+	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
+\end{align}
+
+Similarly, the position update equation~\eqref{eq:posupdate} becomes:
+\begin{align}
+\nonumber
+	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
+\label{eq:pos2update}
+	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
+\end{align}
+
+The value of the objective function at $\ppos_{t+1}$ is given by:
+\begin{align}
+\label{eq:val2update}
+	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \indic_{t+1}^{c}
+	\ \ofunc(\specpos_{t+1}^{c})
+\end{align}
+
+Returning our attention to the computation of $\ppos_{t+1}$ in
+\eqref{eq:pos2update} and writing it with the parameters which were omitted
+above, we obtain:
+\begin{align}
+\nonumber
+  \ppos_{t+1} &= \sum_{c \in \caseset} \\
+\nonumber
+	& \indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
+	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
+\label{eq:val2updatelong}
+	& \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
+	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t})
+\end{align}
+
+In this form the important point to notice is that there are only 7 values (for
+this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
+\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
+objective function evaluation at iteration $t$ (note also that while there are
+random numbers in the equation, they are assumed fixed once drawn for any
+particular particle at a specific iteration).  Thus PSO has been refactored
+such that the algorithm can begin computing all 7 of the objective function
+evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
+computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
+only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
+hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.
+
+Although this speculative algorithm computes
+$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
+for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will be
+ignored, and might just as well never have been computed.  We call the set of
+computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
+``speculative children'' because only one of them will be needed.
+
+To see the value of this refactoring, suppose that 1000 processors are
+available, and that the evaluation of the function to be optimized takes one
+hour.  If we only want a swarm of 100 particles, 900 of the processors would be
+sitting idle for an hour at every iteration, and it would take two hours to run
+two iterations.  If instead we perform speculative evaluation, sending each of
+the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
+$\ofunc(\ppos_{t})$, we would use 800 of the 1000 processors and perform two
+iterations in one hour.
+
+In order to do two iterations at once, we must use 8 times as many processors
+as there are particles in the swarm.  If we were not performing speculative
+evaluation, those processors might very well be used to add particles to the
+swarm.  This then raises the question of whether a swarm of 100 particles doing
+twice as many iterations outperforms a swarm of 800 particles.  We show in the
+rest of this paper that in many, though not all, instances, a smaller swarm
+performing more iterations does in fact outperform a larger swarm.
+
+The number of speculative evaluations needed per particle depends on the number
+of neighbors each particle has.  The number of update cases in a topology where
+each particle has $n$ neighbors is $2(n+1)$; there are two possibilities for
+updates to $\pbest$ (updated by the particle itself and not updated), and $n+1$
+possibilities for updates to $\nbest$ (updated by each neighboring particle and
+not updated).  When the particle is also a neighbor to itself, as is always the
+case in commonly used topologies, one of the cases can be eliminated, as a
+particle cannot be the source of an update to its neighborhood best while not
+also updating its personal best.  Thus we have $2(n+1)-1$, or $2n+1$,
+speculative evaluations per particle.  In a swarm with $p$ particles and $n$
+neighbors per particle, $(2n+1)p$ speculative evaluations are needed.
+
+Because the number of speculative evaluations depends on the number of
+neighbors a particle has, the choice of topology is an important one.  The use
+of the Complete topology, where every particle is a neighbor to every other
+particle, would require $O(p^2)$ speculative evaluations per iteration.
 Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
@@
 sparse topologies.  Accordingly, we use the Random topology described
-in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
-Complete topology.  In that topology, 2 neighbors are chosen at random from the
-swarm for each particle at each iteration.  Because a particle is also always
-its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report the
-results for performing speculative evaluation using both the Ring topology
-(where $n$ again is $3$) and the Random topology on a number of common
-benchmark functions.
+in~\cite{mcnabb-2009-large-particle-swarms}, which has been shown to
+effectively simulate the Complete topology.  In~\sec{results} we report the
+results for performing speculative evaluation using both the Ring topology and
+the Random topology on a number of common benchmark functions.
 
 It is not trivial in some parallel architectures to determine which speculative
-position was the correct next position of each particle.  First we discuss the
-relatively easy case of a centralized parallel PSO algorithm.  In such an
-architecture, the master keeps track of all necessary information with only
-trivial message passing needed.  Then we discuss the more complicated case of a
-distributed algorithm, where each particle is on its own and needs to send and
-receive messages to and from other particles.  Finally we discuss the further
+position was the correct next position of each particle.  In the rest of this
+section we move from the mathematics of our method to its implementation.
+First we discuss the relatively easy case of a centralized parallel PSO
+algorithm with a master computer and many slaves.  In such an architecture, the
+master keeps track of all necessary information with only trivial message
+passing needed.  Then we discuss the more complicated case of a distributed
+algorithm, where each particle is on its own and needs to send and receive
+messages to and from other particles.  Finally we discuss the further
 complications of a dynamic topology, where a particle's neighbors change from
@@
 
-A particle at iteration $t-1$ that has been moved to iteration $t$, but whose
-position has not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its
-position has been evaluated, but it has still not yet received information from
-its neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
-updated its neighborhood best is it a complete particle at iteration $t$.  It is
-then simply denoted as $\p_t$.
+A particle at iteration $t-1$ that has been moved to iteration $t$ using
+\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has
+not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has
+been evaluated, but it has still not yet received information from its
+neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
+updated its neighborhood best is it a complete particle at iteration $t$.  It
+is then simply denoted as $\p_t$.
 
 \subsection{Centralized Algorithms}
+
+In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
+master, keeps track of all necessary information, and all other machines are
+merely used to evaluate the objective function at various positions as directed
+by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
+speculative evaluation in such an architecture, the master generates the
+positions to evaluate speculatively as in \eqref{eq:defpcasegen}.  After having
+the slaves evaluate the objective function at all necessary positions, the
+master then decides which position to accept for each particle, as in
+\eqref{eq:val2updatelong}.  The outline of the procedure is given in
+\alg{centralized}.
 
@@
   \begin{algorithmic}[1]
-	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$
+	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
+	  \eqref{eq:velupdate}~and~\eqref{eq:posupdate}
 	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
-	  generate $\noeval{\sset}_{t+1}$
+	  generate $\noeval{\sset}_{t+1}$ according to
+	  \eqref{eq:defpcasegen}.
 	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
@@
 	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
-	  the branch taken by $\p_t$
+	  the branch taken by $\p_t$ according to
+	  \eqref{eq:val2updatelong}.
 	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
@@
 
-In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
-master, keeps track of all necessary information, and all other machines are
-merely used to evaluate the objective function at various positions as directed
-by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
-a scheme, the master generates the positions to evaluate speculatively and
-decides which one to accept for each particle.  The outline of the procedure is
-given in \alg{centralized}.
-
 Given a set of particles at iteration $t-1$ (perhaps which have just been
-initialized), the master must move each particle using
-Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
-$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
-set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
-position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
-and $\nbest$ that particle $\noeval{\p}_t$ could end up with (as shown in
-Table~\ref{tab:evals}).  Each of those combinations defines a speculative
-particle, $\s_t$, that can be moved using the motion equations to obtain
-$\noeval{\s}_{t+1}$.  The result is a set of particles $\noeval{\pset}_t$, and
-for each particle a set of speculative children $\noeval{\sset}_{t+1}$, which
-can all be evaluated at once.
+initialized), the master must move each particle using \eqref{eq:velupdate} and
+\eqref{eq:posupdate} to obtain the set $\noeval{\pset}_t$.  For each particle
+$\noeval{\p}_t$, the master must then get its set of neighbors
+$\noeval{\nset}_t$ and use their positions, along with the position of
+$\noeval{\p}_t$, to calculate all possible values of
+$\specpos_{t+1}^{\casegen}$, using \eqref{eq:defpcasegen}.  These positions,
+along with the original particle's associated information (such as values for
+$\pbest$ and $\nbest$), define a set of speculative children,
+$\noeval{\sset}_{t+1}$.  The master than has a set of particles
+$\noeval{\pset}_t$, and for each particle a set of speculative children
+$\noeval{\sset}_{t+1}$, which can all be evaluated at once.
 
@@
 speculative and original, have been evaluated and the values reported to the
-master, the master sorts out which speculative child of each particle was the
-correct one.
-
-This is done by first updating each particle $\noeval{\p}_t$ to
-$\nonbest{\p}_t$ and each speculative particle $\noeval{\s}_{t+1}$ to
-$\nonbest{\s}_{t+1}$ by associating the value determined by the slave to the
-particle it came from, and updating that particle's personal best if the value
-reported is better than its previous personal best value.  Then each
-$\nonbest{\p}_t$ must be updated to $\p_t$ by updating its neighborhood best
-position and value with information about each of its neighbors.  This is
-simply the original PSO algorithm, and corresponds to steps 1--5 in
-\alg{centralized}.
-
-Then, with the set $\pset_t$, the branch that was actually taken by each
-particle can be determined, and the correct speculative child can be chosen.
-The branches depend entirely on whether or not the personal best of $p_t$ was
-updated and which particle, if any, updated the particle's neighborhood best.
-Both of those items are now known, and the child that matches the branch is
-selected (step 7 in \alg{centralized}).
+master, the master determines which speculative child of each particle was the
+correct one.  Mathematically, this corresponds to the evaluation of an
+indicator function similar to that found in \eqref{eq:deficasexn}.  In
+practice, this is done first by updating each particle's $\pbest$, if
+necessary, then by updating the particle's $\nbest$ with information from the
+particle's neighbors.  This is simply the original PSO algorithm, and
+corresponds to steps 1--5 in \alg{centralized}.  Given the updates to $\pbest$
+and $\nbest$, the case from Table~\ref{tab:evals} can be determined, as per
+\eqref{eq:val2updatelong}.  The child with the matching case is kept, and all
+other speculative children are discarded (step 7 in \alg{centralized}.
 
@@
 
+\label{sec:distributed}
+
 In a distributed parallel PSO algorithm, individual processors not only perform
@@
 described uses three rounds of message passing for every two iterations
-(corresponding to steps 2, 5 and 10 in \alg{centralized}).  Accordingly, we
-describe a method that requires only one round of communication for each pair
-of iterations, which happens at step 5 of \alg{centralized}.  Many more
-messages are needed, but that is sometimes more desirable than synchronizing
-all of the machines three times.
-
-This second method only requires one round of passing information because
-information about iterations $t$ and $t+1$ is passed at the same time.  Each
-processor reconstructs from the messages it receives all of the information
-that it needs about its neighbors.  Messages are passed directly after
-evaluating each particle and its children, so all messages are of the form of
-$\nonbest{\p}_t$ or $\nonbest{\s}_{t+1}$.  The first iteration needs to be
-treated specially, so each particle can produce its initial set of speculative
-children---neighbors need only pass their initial position.  This kind of
-message passing necessitates the careful use of random seeds, so that when each
-processor computes the motion equations for its neighbors it gets the same
-results as its neighboring processors.
-
-With the results of evaluating $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$,
-along with all of the required messages from neighboring particles, the goal is
-to produce $\p_{t+1}$ and output $\noeval{\p}_{t+2}$ and $\noeval{\sset}_{t+3}$
-ready to be evaluated for the next iteration.  We first focus on the messages
-needed to produce $\p_{t+1}$.
-
-Upon evaluation, $\noeval{\p}_t$ becomes $\nonbest{\p}_t$, needing only to get
-its neighborhood best information from its neighbors.  All of its neighbors,
-then, must send it a message, so that from their updated personal best at
-iteration $t$ the particle becomes $\p_t$.  The work done with the messages
-received thus far is just as in regular PSO, and is graphically depicted in 
-\fig{messages1}.
-
-\begin{figure}
-  \centering
-  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
-  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
-  \psfrag{pt}[C][C]{$\p_{t}$}
-  \includegraphics[width=.3\columnwidth]{messages1.eps}
-  \caption{The production of $\p_{t}$ from the original particle
-  $\nonbest{\p}_{t}$ and the messages $\nonbest{\nset}_{t}$.}
-  \label{fig:messages1}
-\end{figure}
-
-With $\p_t$ we can select the correct speculative child as described above and
-produce $\nonbest{\p}_{t+1}$.  Again we show the use of messages thus far
-graphically, in \fig{messages2}.
-
-\begin{figure}
-  \centering
-  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
-  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
-  \psfrag{pt}[C][C]{$\p_{t}$}
-  \psfrag{pt1-n}[C][C]{$\nonbest{\p}_{t+1}$}
-  \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
-  \includegraphics[width=.5\columnwidth]{messages2.eps}
-  \caption{The production of $\nonbest{\p}_{t+1}$ from the original particle 
-  $\nonbest{\p}_{t}$, messages $\nonbest{\nset}_{t}$ and
-  $\nonbest{\sset}_{t+1}$, and intermediate particles.}
-  \label{fig:messages2}
-\end{figure}
-
-We then need the set of neighbors to $\nonbest{\p}_{t+1}$,
-$\nonbest{\nset}_{t+1}$, so we can update $\nonbest{\p}_{t+1}$'s neighborhood
-best.  To produce each neighbor $\nonbest{\n}_{t+1}$, we need the same
-information for the neighboring particle that we needed to produce the original
-particle, $\nonbest{\p}_{t+1}$; we need the original neighbor particle, its
-speculative children, and its neighbors.  With that information, the set
-$\nonbest{\nset}_{t+1}$ can be obtained by following the same process used to
-obtain $\nonbest{\p}_{t+1}$.  We graphically show the messages needed to
-produce $\nonbest{\nset}_{t+1}$ in \fig{messages3}.  Note that it looks
-identical to \fig{messages2}, just with different sets of particles.
-
-\begin{figure}
-  \centering
-  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
-  \psfrag{nnt-n}[C][C]{$\nonbest{\nnset}_{t}$}
-  \psfrag{nt}[C][C]{$\nset_{t}$}
-  \psfrag{nt1-n}[C][C]{$\nonbest{\nset}_{t+1}$}
-  \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
-  \includegraphics[width=.6\columnwidth]{messages3.eps}
-  \caption{The production of each $\nonbest{\n}_{t+1}$ from the original
-  particle $\nonbest{\n}_{t}$, messages $\nonbest{\nnset}_{t}$ and
-  $\nonbest{\nsset}_{t+1}$, and intermediate particles.  $\nnset$ is the set of
-  neighbors for each particle $\n$, and $\nsset$ is the set of $\n$'s
-  speculative children.  Note the similarity between this and \fig{messages2}.}
-  \label{fig:messages3}
-\end{figure}
-
-With $\nonbest{\nset}_{t+1}$ and $\nonbest{\p}_{t+1}$, we can produce
-$\p_{t+1}$.  This is shown in \fig{messages4}.  Note that we just combined
-Figures~\ref{fig:messages2}~and~\ref{fig:messages3}, putting them together
-to make $\p_{t+1}$, as all the particle needs is its neighborhood best to be
-updated.
-
-\begin{figure}
-  \centering
-  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
-  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
-  \psfrag{pt}[C][C]{$\p_{t}$}
-  \psfrag{pt1-n}[C][C]{$\nonbest{\p}_{t+1}$}
-  \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
-  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
-  \psfrag{nnt-n}[C][C]{$\nonbest{\nnset}_{t}$}
-  \psfrag{nt}[C][C]{$\nset_{t}$}
-  \psfrag{nt1-n}[C][C]{$\nonbest{\nset}_{t+1}$}
-  \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
-  \psfrag{pt1}[C][C]{$\p_{t+1}$}
-  \includegraphics[width=.8\columnwidth]{messages4.eps}
-  \caption{The production of $\p_{t+1}$ from the original particle 
-  $\nonbest{\p}_{t}$, messages $\nonbest{\nset}_{t}$, $\nonbest{\sset}_{t+1}$,
-  $\nonbest{\nnset}_{t}$, and $\nonbest{\nsset}_{t+1}$, and intermediate
-  particles.  Note that this is just a combination of \fig{messages2} and
-  \fig{messages3}.}
-  \label{fig:messages4}
-\end{figure}
-
-In order to get $\p_{t+1}$, then, a particle needs to receive messages from its
-neighbors, its neighbors' neighbors, its speculative children, and its
-neighbors' speculative children.  The particle $\p_{t+1}$ can be passed to some
-central machine to track the progress of the algorithm, and it can be moved to
-$\noeval{\p}_{t+2}$ in order to start the next iteration.
-
-The next goal is to produce the set $\noeval{\sset}_{t+3}$.  As described
-above, the necessary components to produce $\noeval{\sset}_{t+3}$ are
-$\noeval{\p}_{t+2}$ and the neighbors of $\noeval{\p}_{t+2}$,
-$\noeval{\nset}_{t+2}$.  We already have $\noeval{\p}_{t+2}$, so what remains
-is to produce $\noeval{\nset}_{t+2}$.  It is sufficient to obtain
-$\nset_{t+1}$, as each neighbor particle $\n_{t+1}$ can be moved with the
-motion equations to $\noeval{\n}_{t+2}$.
-
-We have already described how to use a set of messages to obtain $\p_{t+1}$.
-The process is exactly the same to produce each $\n_{t+1}$, requiring the same
-messages, except for the neighbor particles instead of the particle itself.  
-\fig{messages5} shows graphically how $\noeval{\sset}_{t+3}$ is produced.
-
-\begin{figure}
-  \centering
-  \psfrag{pt1}[C][C]{$\p_{t+1}$}
-  \psfrag{nt1}[C][C]{$\nset_{t+1}$}
-  \psfrag{nt2-e}[C][C]{$\noeval{\nset}_{t+2}$}
-  \psfrag{pt2-e}[C][C]{$\noeval{\p}_{t+2}$}
-  \psfrag{st3-e}[C][C]{$\noeval{\sset}_{t+3}$}
-  \includegraphics[width=.5\columnwidth]{messages5.eps}
-  \caption{The production of $\noeval{\p}_{t+2}$ and $\noeval{\sset}_{t+3}$
-  from $\p_{t+1}$ and $\n_{t+1}$, each of which are produced as in
-  \fig{messages4}.}
-  \label{fig:messages5}
-\end{figure}
-
-Having obtained both $\noeval{\p}_{t+2}$ and $\noeval{\sset}_{t+3}$ from the
-messages received, the algorithm then moves to the evaluation phase, and the
-process repeats itself.  The particles are evaluated, send their messages, and
-produce the next set of particles to be evaluated from the messages received.
-
-To perform the entire process, at each message passing round a particle must
-receive messages from its neighbors, its neighbors' neighbors, its neighbors'
-neighbors' neighbors, its speculative children, its neighbors' speculative
-children, and its neighbors' neighbors' speculative children.  With the Ring
-topology, that looks like more messages than it really is, as many of the
-neighbors' neighbors are duplicates.  With the Random topology, however, the
-list of necessary messages could be rather large.
-
-This may seem like an inordinate amount of work, and with some distributed PSO
-frameworks it is.  However, other parallel frameworks necessitate this type of
-message passing, so we have described how speculative evaluation can be
-performed in those circumstances.
+(corresponding to steps 2, 5 and 10 in \alg{centralized}).  It is possible to
+perform speculative evaluation in PSO with only one round of communication per
+two iterations.  However, the methodology is tedious and is not the focus of
+this paper, so we defer its description to the Appendix.
 
@@
 
-Performing speculative evaluation in PSO with a dynamic topology raises a few
-issues of its own.  First, with neighbors changing each iteration, messages
-that processors pass to their neighbors need to be sent to the correct
-neighbors for each iteration.  In centralized algorithms and in distributed
-algorithms with several rounds of message passing (the first method described
-above), nothing needs to be changed.  However, in the case where processors
-reconstruct information about neighboring particles from messages, those
-messages need to be sent to the correct neighbors.  A particle cannot simply
-send messages to its neighbors' neighbors' neighbors---it needs to send
-messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
-For every neighbor outward information is sent, the iteration also needs to be
-incremented, as information about neighbors' neighbors is used during iteration
-$t+1$, and information about neighbors' neighbors' neighbors is used to
-reconstruct information about iteration $t+2$.  Also, this method of message
-passing again requires the use of random seeds if the topology is random, so
-that each processor computes the same neighbors for a particle as all other
-processors.
-
-Also, an issue common to all methods of performing speculative evaluation is
-that of getting available information from neighbors before generating
-speculative children.  In a static topology, at iteration $t$ a particle
+Performing speculative evaluation in PSO with a dynamic topology raises a
+sticky issue of its own.  In a static topology, at iteration $t$ a particle
 already has all of the information about the positions of its neighbors during
@@
 best.  That creates a problem for speculative evaluation---there are
-potentially more than $2(n+1)-1$ possible next positions, increasing the amount
-of work that must be done to perform the second iteration at the same time as
-the first.
+potentially more than $2n+1$ possible next positions, increasing the amount of
+work that must be done to perform the second iteration at the same time as the
+first.
 
-This is easily fixed by updating $\noeval{\p}_{t+2}$ with the currently
-available information about $\noeval{\nset}_{t+2}$ before producing
-$\noeval{\sset}_{t+3}$.  If a particle $\noeval{\p}_{t+2}$ updates its 
-neighborhood best with the personal bests of $\noeval{\nset}_{t+2}$ before
-calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
-still only $2(n+1)-1$ possible next positions, and the problem is avoided.
+This is easily fixed by updating each new particle $\noeval{\p}_{t+2}$ with the
+currently available information about its neighbors $\noeval{\nset}_{t+2}$
+before producing its children $\noeval{\sset}_{t+3}$.  If a particle
+$\noeval{\p}_{t+2}$ updates its neighborhood best with the personal bests of
+$\noeval{\nset}_{t+2}$ before calculating the next possible positions for
+$\noeval{\sset}_{t+3}$, there are still only $2n+1$ possible next positions,
+and the problem is avoided.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{sphere1.eps}
+  \includegraphics[width=.8\columnwidth]{sphere1}
   \caption{Function Sphere with a swarm that uses 240 processors per round of
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank2.eps}
+  \includegraphics[width=.8\columnwidth]{griewank2}
   \caption{Function Griewank with a swarm that uses 800 processors per round of
@@
 with Rastrigin, the more particles there are in the swarm, the lower function
-value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
-swarms get caught in local optima.  Because our speculative algorithms have a
-significantly smaller swarm size, they get stuck at higher values while the
-larger swarms performing regular PSO continue to improve the best value found.
+value it finds, up to at least 4000
+particles~\cite{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
+in local optima.  Because our speculative algorithms have a significantly
+smaller swarm size, they get stuck at higher values while the larger swarms
+performing regular PSO continue to improve the best value found.
 \fig{rastrigin} shows the results graphically.
@@
 of speculative evaluation.  A related but quite distinct technique for reducing
-computation in evolutionary algorithms was proposed in~\cite{poli-ai06}.
+computation in evolutionary algorithms was proposed
+in~\cite{poli-2006-backward-chaining-evolutionary-algorithms}.
 
@@
 \bibliography{%
-\bib{wolpert-tec97},%
-\bib{merris-01},%
-\bib{pso/clerc-tec02},%
-\bib{pso/kennedy-icnn95},%
-\bib{pso/bratton-sis07},%
-\bib{pso/largeswarms/perez-emc05},%
-\bib{pso/largeswarms/montes-gecco08.bib},%
-\bib{pso/topology/liang-sis05.bib},%
-\bib{pso/parallel/belal-ijicis04},%
-\bib{pso/parallel/chu-sci06},%
-\bib{pso/parallel/jin-aps05},%
-\bib{pso/parallel/koh-ijnme06},%
-\bib{pso/parallel/mostaghim-report06},%
-\bib{pso/parallel/parsopoulos-aia04},%
-\bib{pso/parallel/schutte-ijnme04},%
-\bib{pso/parallel/venter-wcsmo05},%
-\bib{ga/poli-ai06},%
-\bib{sim-anneal/parallel/witte-tpds91.bib},%
-\bib{pso/topology/kennedy-cec02.bib},%
-\bib{pso/topology/mohais-cec04.bib},%
-\bib{pso/topology/jordan-gecco08.bib},%
-\bib{pso/topology/clerc-oep03.bib},%
-\bib{pso/topology/miranda-ijcir08.bib},%
-\bib{pso/topology/mendes-phd04.bib},%
-\bib{aml/mcnabb-cec07},%
-\bib{aml/mcnabb-cec09}}
-
+../../../bib/aml/bib,%
+../../../bib/pso/general/bib,%
+../../../bib/pso/parallel/bib,%
+../../../bib/pso/topology/bib,%
+../../../bib/sim-anneal/parallel/bib,%
+../../../bib/ga/bib}
 \end{document}

### 2010-04-20 honors_thesis/ppsn_paper/paper.tex
first cutting pass

@@
 adding more particles to the swarm or by parallelizing the evaluation of the
-objective function.  However, some functions are better optimized with more
+objective function.  However, some functions are more efficiently optimized with more
 iterations and fewer particles.  Accordingly, we take inspiration from 
 speculative execution commonly performed in modern processors and propose what
-we call speculative evaluation in PSO.  Future positions of the particles are
+we call speculative evaluation in PSO (SEPSO).  Future positions of the particles are
 speculated and evaluated in parallel with current positions, performing two
@@
 either by adding individual particles or by adding entire new sub-swarms, but
-in all cases the number of processors never exceeds the number of particles.
+in all cases the number of processors does not exceed the number of particles.
 The number of iterations of PSO that the algorithms can perform is thus
@@
 algorithm makes per evaluation.  Other functions show similar trends, though
-often the optimal swarm size is slightly larger.  For this reason, previous
+often the optimal swarm size is larger.  For this reason, previous
 work has recommended the use of a swarm size of 50 for
@@
 In this paper we consider PSO parallelization strategies for clusters of
-thousands of processors and functions for which a single evaluation will take
-at least tens of seconds, but probably minutes or perhaps hours.  Our purpose
-is to explore the question of what to do with a thousand processors when 50 or
+hundreds of processors and functions for which a single evaluation will take
+at least seconds, but probably minutes.  Our purpose
+is to explore the question of what to do with a hundred processors when 50 or
 100 particles is the recommended swarm size, and simply increasing the swarm
@@
 comes from an analogy to speculative execution (also known as branch
-prediction) techniques commonly used in processors.  Modern processors, when
+prediction) a technique commonly used in processors.  Modern processors, when
 faced with a branch on which they must wait (e.g. a memory cache miss), guess
@@
 \begin{align}
-\nonumber
+\label{eq:velupdate}
 	\pvel_{t+1} &=
 		\constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
-\label{eq:velupdate}
-			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
+			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
 		\bigr] \\
@@
 
-The refactoring hinges on the idea that once the random coefficients are
-determined, there are only a few possibile updates to $\nbest$ and $\pbest$.
+The refactoring hinges on the idea that once the random coefficients $\pURand_{t}$ and $\pURand_{t}$ are
+determined, there are only a few possible updates to $\nbest$ and $\pbest$.
 For the Ring topology there are 7 possible update cases, identified in
@@
 $\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
-$\casegen$ corresponds to the case actually taken by PSO,
+$\casegen$ corresponds to the case taken by PSO,
 $\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
@@
 	&\begin{cases}
-	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
-	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
-	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
-	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
+	   1 & \text{if} \ [ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) ]
+	    \wedge [ \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) ] \\
+	   &\quad \wedge \quad [ \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) ]
+	    \wedge [ \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) ] \\
 	   0 & \text{otherwise}
@@
 	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
-\nonumber
+\label{eq:defvcasexn}
 		&= \constriction \bigl[ \pvel_{t} +
-			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t}) \\
-\label{eq:defvcasexn}
-			& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
+			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t})
+			+ \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
 			\ppos_{t}) \bigr]
@@
 	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
-	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
-\nonumber
-	& = \ppos_{t} + \specvel_{t+1}^{\casegen}
+	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t})
+	 = \ppos_{t} + \specvel_{t+1}^{\casegen}
 \end{align}
@@
 		\constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
-\nonumber
-			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
+			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
 			\ppos_{t}) \bigr] \\
@@
 	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
-\nonumber
-			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
+			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
 			\ppos_{t}) \bigr]  \\
@@
 \begin{align}
-\nonumber
-	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
 \label{eq:pos2update}
-	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
+	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
+	= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
 \end{align}
@@
 \nonumber
-  \ppos_{t+1} &= \sum_{c \in \caseset} \\
-\nonumber
-	& \indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
+  \ppos_{t+1} &= \sum_{c \in \caseset} \bigl[
+	\indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
 	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
 \label{eq:val2updatelong}
-	& \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
-	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t})
+	& \quad \quad \quad \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
+	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t}) \bigr]
 \end{align}
 
-In this form the important point to notice is that there are only 7 values (for
+In this form the important point to notice is that there are only $7$ values (for
 this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
 \caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
-objective function evaluation at iteration $t$ (note also that while there are
+objective function evaluation at iteration $t$. Note also that while there are
 random numbers in the equation, they are assumed fixed once drawn for any
-particular particle at a specific iteration).  Thus PSO has been refactored
-such that the algorithm can begin computing all 7 of the objective function
+particular particle at a specific iteration.  Thus PSO has been refactored
+such that the algorithm can begin computing all $7$ of the objective function
 evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
@@
 
-Although this speculative algorithm computes
+Although this speculative approach computes
 $\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
@@
 
-To see the value of this refactoring, suppose that 1000 processors are
+To see the value of this refactoring, suppose that $800$ processors are
 available, and that the evaluation of the function to be optimized takes one
-hour.  If we only want a swarm of 100 particles, 900 of the processors would be
+hour.  If we only want a swarm of $100$ particles, $700$ of the processors would be
 sitting idle for an hour at every iteration, and it would take two hours to run
@@
 the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
-$\ofunc(\ppos_{t})$, we would use 800 of the 1000 processors and perform two
+$\ofunc(\ppos_{t})$, we could create a swarm of size 100, each particle with $7$ speculative
+evaluations ($700$ processors dedicated to speculative evaluation), thus using all $800$
+processors and performing two
 iterations in one hour.
 
-In order to do two iterations at once, we must use 8 times as many processors
-as there are particles in the swarm.  If we were not performing speculative
-evaluation, those processors might very well be used to add particles to the
-swarm.  This then raises the question of whether a swarm of 100 particles doing
+In order to do two iterations at once, we use 8 times as many processors
+as there are particles in the swarm.  If these processors were not performing speculative
+evaluation, they might instead be used for function evaluation needed to support a large swarm.
+This raises the question of whether a swarm of 100 particles doing
 twice as many iterations outperforms a swarm of 800 particles.  We show in the
-rest of this paper that in many, though not all, instances, a smaller swarm
+rest of this paper that in many, instances, a smaller swarm
 performing more iterations does in fact outperform a larger swarm.
+We acknowledge that both intuition and prior research~\cite{mcnabb-2009-large-particle-swarms}
+indicate that the optimization of deceptive functions benefits greatly from
+large and even very large swarm sizes. Thus this work will focus on less deceptive functions.
+
+\subsection{Topology}
+\label{sec:topology}
 
@@
 the Complete topology and the quick spread of information it entails than with
-sparse topologies.  Accordingly, we use the Random topology described
-in~\cite{mcnabb-2009-large-particle-swarms}, which has been shown to
-effectively simulate the Complete topology.  In~\sec{results} we report the
-results for performing speculative evaluation using both the Ring topology and
-the Random topology on a number of common benchmark functions.
+sparse topologies.  In such cases, we use the Random topology 
+which has been shown to
+effectively approximate the rapid information flow of the Complete topology.
+In our implementation of the random topology there are to randomly chosen neighbors
+(chosen anew at every iteration)~\cite{mcnabb-2009-large-particle-swarms}.
 
-It is not trivial in some parallel architectures to determine which speculative
-position was the correct next position of each particle.  In the rest of this
+In the rest of this
 section we move from the mathematics of our method to its implementation.
@@
 
-To aid in describing our methodology, we introduce a few terms.  A particle's
-set of \emph{speculative children} is the set of all possible next iteration
-states that a particle could have.  We use $\p_t$ to denote a particle at
+To aid in describing our implementation, we introduce a few terms. We use $\p_t$ to denote a particle at
 iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
@@
 
-\subsection{Centralized Algorithms}
+\subsection{Implementation}
 
@@
 master then decides which position to accept for each particle, as in
-\eqref{eq:val2updatelong}.  The outline of the procedure is given in
+\eqref{eq:val2updatelong}.  The outline of the speculative evaluation in PSO (SEPSO) algorithm is given in
 \alg{centralized}.
@@
 \begin{algorithm}
-  \caption{Speculative Evaluation in a Centralized PSO}
+  \caption{Speculative Evaluation PSO (SEPSO)}
   \label{alg:centralized}
@@
 
-Given a set of particles at iteration $t-1$ (perhaps which have just been
-initialized), the master must move each particle using \eqref{eq:velupdate} and
-\eqref{eq:posupdate} to obtain the set $\noeval{\pset}_t$.  For each particle
-$\noeval{\p}_t$, the master must then get its set of neighbors
-$\noeval{\nset}_t$ and use their positions, along with the position of
-$\noeval{\p}_t$, to calculate all possible values of
-$\specpos_{t+1}^{\casegen}$, using \eqref{eq:defpcasegen}.  These positions,
-along with the original particle's associated information (such as values for
-$\pbest$ and $\nbest$), define a set of speculative children,
-$\noeval{\sset}_{t+1}$.  The master than has a set of particles
-$\noeval{\pset}_t$, and for each particle a set of speculative children
-$\noeval{\sset}_{t+1}$, which can all be evaluated at once.
-
-The master then has the slaves evaluate the particles.  Once all particles,
-speculative and original, have been evaluated and the values reported to the
-master, the master determines which speculative child of each particle was the
-correct one.  Mathematically, this corresponds to the evaluation of an
-indicator function similar to that found in \eqref{eq:deficasexn}.  In
-practice, this is done first by updating each particle's $\pbest$, if
-necessary, then by updating the particle's $\nbest$ with information from the
-particle's neighbors.  This is simply the original PSO algorithm, and
-corresponds to steps 1--5 in \alg{centralized}.  Given the updates to $\pbest$
-and $\nbest$, the case from Table~\ref{tab:evals} can be determined, as per
-\eqref{eq:val2updatelong}.  The child with the matching case is kept, and all
-other speculative children are discarded (step 7 in \alg{centralized}.
-
-The parent $\p_t$ must pass its personal best value to the child, as the child
-knows only the position that it guessed, not the function value at that
-position.  It is possible that both $\p_t$ and $\s_{t+1}$ update their personal
-bests, but $\p_t$'s value is better.  For example, suppose that $\p_{t-1}$ has
-a personal best value of 3, and the we are seeking to minimize the function.
-$\noeval{\p}_t$ is created, and $\noeval{\s}_{t+1}$ is moved assuming that
-$\p_t$ has updated its personal best with its position at time $t$.  Then both
-$\noeval{\p}_t$ and $\noeval{\s}_{t+1}$ are evaluated, with values 1 and 2,
-respectively.  $\nonbest{\s}_{t+1}$ would think that its current position is
-its personal best, as the value it found, 2, is better than its previous
-personal best value of 3.  It needs to receive the personal best value from its
-parent to know that its personal best position $\pbest$ is actually the
-position of $\p_t$, not $\s_{t+1}$.
-
-The parent also needs to pass the value of the neighborhood best that the child
-guessed.  The child only knows the position and needs the value in order to
-make future comparisons between neighborhood best positions (step 8).
-
-Upon picking the correct branch for each particle and updating the child's
-personal best and neighborhood best value (from iteration $t$), the result is
-the set $\nonbest{\pset}_{t+1}$, as the particles are now no longer
-speculative.  What remains is to update the neighborhood best of those
-particles from their neighbors (from iteration $t+1$), as above, to obtain
-$\pset_{t+1}$.  That set of particles can subsequently be used to produce the
-sets $\noeval{\pset}_{t+2}$ and $\noeval{\sset}_{t+3}$ (steps 1 and 2 in
-\alg{centralized}), and the process repeats itself.
-
-\subsection{Distributed Algorithms}
-
-\label{sec:distributed}
-
-In a distributed parallel PSO algorithm, individual processors not only perform
+In a distributed parallel PSO implementation, individual processors not only perform
 evaluations of particles, but also their movement.  The information for each
@@
 charge of and must perform the steps of the algorithm for those
-particles~\cite{mcnabb-2007-parallel-pso-using-mapreduce}.  Messages such as
+particles.  Messages such as
 values and positions for the neighborhood best are sent between processors.
@@
 importance is much less than in centralized algorithms.
-
-To perform speculative evaluation in a distributed PSO algorithm, there must be
-some way to have processors evaluate the speculative children of particles,
-without giving the speculative particles the same treatment as actual
-particles, as the speculative children only live for one iteration.  One way
-that can be done is by assigning each particle a set of machines instead of a
-single machine, and the particle directs its extra machines to evaluate its
-speculative children.  However the distributed framework functions, the same
-information needs to be passed between particles, and we describe here the
-messages each particle needs to receive to perform speculative evaluation.
-
-A processor that is controlling a single particle $\p_{t-1}$ must first move
-the particle to $\noeval{\p}_t$ and produce the particle's speculative children
-$\noeval{\sset}_{t+1}$.  This is done in the same way as described above.
-However, in order to produce $\noeval{\sset}_{t+1}$, the processor needs
-information about the particle's neighbors, so there must be some message
-passing to get that information.  Particularly, the information that the
-processor needs is the position of each of the particle's neighbors at
-iteration $t$.
-
-To get that information, a round of message passing is required.  Each particle
-sends its position to its neighbors at iteration $t$, so that all particles can
-generate $\noeval{\sset}_{t+1}$.  After each particle evaluates its position
-and the positions of its speculative children, it passes information about the
-outcome of iteration $t$ to its neighbors, so that neighboring particles can
-update their neighborhood bests to move from $\nonbest{\p}_t$ to $\p_t$.  Once
-that communication is finished, the particle can select the speculative child
-which matched the branch that iteration $t$ actually produced.  Then another
-round of information passing follows, for iteration $t+1$, so that
-$\nonbest{\p}_{t+1}$ can be updated to $\p_{t+1}$.  Two iterations have then
-been completed with only one round of evaluations, and the next iteration can
-start again with the first round of message passing.
-
-In some distributed frameworks, like the MapReduce framework that we used in
-our experiments, rounds of message passing can be expensive.  The method just
-described uses three rounds of message passing for every two iterations
+In some distributed frameworks rounds of message passing can be expensive.
+A N\:aive implementation of speculative evaluation
+uses three rounds of message passing for every two iterations
 (corresponding to steps 2, 5 and 10 in \alg{centralized}).  It is possible to
 perform speculative evaluation in PSO with only one round of communication per
-two iterations.  However, the methodology is tedious and is not the focus of
-this paper, so we defer its description to the Appendix.
-
-\subsection{Dynamic Topologies}
-
-Performing speculative evaluation in PSO with a dynamic topology raises a
-sticky issue of its own.  In a static topology, at iteration $t$ a particle
-already has all of the information about the positions of its neighbors during
-iterations $1$ through $t-1$.  If the neighbor finds a better position at
-iteration $t$, the particle updates its neighborhood best, but if it does not,
-it still has its old neighborhood best from its neighbors for all previous
-iterations.
-
-In a dynamic topology, a particle might not have information about the previous
-positions of its neighbors at iteration $t$.  That means that its new
-neighborhood best could come not only from its neighbors' positions at
-iteration $t$, but also from their personal best from iteration $t-1$, as
-neighbors' personal bests are what are used to update a particle's neighborhood
-best.  That creates a problem for speculative evaluation---there are
-potentially more than $2n+1$ possible next positions, increasing the amount of
-work that must be done to perform the second iteration at the same time as the
-first.
-
-This is easily fixed by updating each new particle $\noeval{\p}_{t+2}$ with the
-currently available information about its neighbors $\noeval{\nset}_{t+2}$
-before producing its children $\noeval{\sset}_{t+3}$.  If a particle
-$\noeval{\p}_{t+2}$ updates its neighborhood best with the personal bests of
-$\noeval{\nset}_{t+2}$ before calculating the next possible positions for
-$\noeval{\sset}_{t+3}$, there are still only $2n+1$ possible next positions,
-and the problem is avoided.
+two iterations. 
+Dynamic topologies like the random topology discussed above, further complicate the
+needed message passing.
+However, the message passing implementation is not particularly
+interesting, but is tedious thus, in the interest of space, is omitted.
 
@@
 
-In performing speculative evaluation as we have described it, $2(n+1)-1$
+In performing speculative evaluation as we have described it, $2n+1$
 speculative evaluations are done per particle, while all but one of them are
-completely ignored.  It seems reasonable to try to make use of those
-evaluations instead of throwing them away.  Making use of them changes the
+completely ignored.  It seems reasonable to try to make use of the
+information obtained through those evaluations instead of ignoring it
+Making use of this information changes the
 behavior of PSO, instead of reproducing it exactly as the above method
-explains, but the change could be an improvement.
+explains, but the change turns out to be an improvement in our context.
 
-To use the extra speculative evaluations, instead of choosing the speculative
+To make better use of the speculative evaluations, instead of choosing the speculative
 child that matches that branch that the original PSO would have taken, we take
@@
 step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
-$\noeval{\sset}_{t+1}$ instead of with the matching branch.
+$\noeval{\sset}_{t+1}$ instead of with the matching branch. We call this
+varient ``Pick Best''.
 
@@
 and accepting the best one.  Speculative particles that move in good directions
-are kept.  It seems this modification to PSO would produce more exploitation,
-decreasing the amount of exploration.  With functions that are less deceptive
-we would expect this method to work well, while performance on more deceptive
+are kept.  It seems this modification to PSO would contribute local exploration,
+having knowledge of all of the speculative children, but more global exploitation,
+since Pick Best only considers the best of these alternatives.
+With functions that are less deceptive, or only locally deceptive,
+Pick Best works well, while performance on more deceptive
 functions would probably be hurt.
@@
 be the best use of parallelism in PSO.  But for many real-world applications,
-the objective function takes on the order of minutes or more to evaluate; in
+the objective function takes on the order of at least seconds (or more) to evaluate; in
 such cases our assumption is perfectly reasonable.
 
-To use the same number of processors for each algorithm, speculative evaluation
-needs a smaller swarm size than standard PSO.  For each of the topologies we
-used, a particle has two neighbors other than itself.  As shown in
+In each set of experiments we keep the number of processors for each algorithm constant.
+In all of our SEPSO experiments each particle has
+two neighbors in addition to itself.
+As shown in
 Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
 particle.  With one evaluation needed for the original, non-speculative
-particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
-the number of particles in the speculative swarm.  The extra evaluations done
-in speculative evaluation would instead be particles in regular PSO, so we
-compare swarms of size $p_s$ in speculative evaluation with swarms of size
-$8p_s$ in regular PSO.
+particle, we have a total of $8p_s$ evaluations for every two iterations, where $p_s$ is
+the number of particles in the speculative swarm.
+We run these experiments on $8p_s$ processors.
+In the base-line experiments, standard pso the same number of processors and thus enjoys the
+benefit of $8p_s$ particles running on 
+$8p_s$ processors.
+At least in benchmark functions, extra particles only accelerate
+the performance of PSO~\cite{mcnabb-2009-large-particle-swarms}. They generally have diminishing value, as
+seen earlier in \fig{evals-sphere}, but these extra particles give standard PSO the best possible
+context and full access to the same resources as SEPSO.
 
-Where the Complete topology would normally be used, we use a Random topology in
-our speculative algorithm, as Complete leads to an explosion of speculative
-evaluations.  However, if speculative evaluation were not being performed, it
-is likely that the Complete topology would be used, ignoring the communication
-cost of Complete over Random.  To be fair in our comparisons, for the standard
-PSO algorithm we compare to both the Random topology with two neighbors and to
-the Complete topology.  All functions have 20 dimensions, and all results shown
+As discussed in \sec{topology}
+where the Complete topology would normally be used, we use a Random topology in
+our speculative algorithm, as complete leads to an explosion of speculative
+evaluations.
+For the standard PSO baseline we have included experiments with the
+ring and the random topologies, both with two neighbors, as well as for the complete topology.
+It is important to note however, that in many cases the complete topology is not a
+practical alternative in the context of large
+swarms on large cluster where where the messaging complexity is $O(n^2)$ and can
+overwhelm the system.
+
+We follow the experimental set up proposed by Bratton, et.al.~\cite{bratton-2007-defining-a-standard-for-pso}. 
+All functions have 20 dimensions, and all results shown
 are averages over 20 runs.  We use 240 processors in our experiments, so
@@
 For ease of labeling, we call our speculative algorithm SEPSO, the variant
-SEPSO Pick Best, and standard PSO just PSO.
+SEPSO Pick Best, and standard PSO just PSO and identify the topology in a suffix,
+``PSO Ring'' for example.
 
-First we look at the function Sphere, the simplest of common benchmark
-functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
-has no local optima and is best optimized in terms of function evaluations with
-a small swarm using a Complete topology.  Accordingly, we show results for a
-small Complete swarm in \fig{sphere1}.  In comparing SEPSO with PSO, we can see
-that SEPSO clearly beats PSO with a Random topology.  However, the performance
-between SEPSO and PSO Complete is not as encouraging---PSO Complete performs
-slightly better.  But SEPSO Pick Best handily outperforms all other algorithms
+The benchmark function Sphere ($f(\Vec{x}) = \sum_{i=1}^D x_i^2$)
+has no local optima and is most efficiently optimized using
+a small swarm but many iterations.
+In \fig{sphere1}, we can see
+that SEPSO clearly beats PSO with a Random topology or a ring topology.
+
+SEPSO approaches the performance of PSO with the complete topology, even though
+PSO with the complete topology requires $O(n^2)$ communication.
+SEPSO Pick Best handily outperforms all other algorithms
 in this problem.
@@
   iterations per round, while standard PSO algorithms use 240 processors and do
-  one iteration per round.}
+  one iteration per round. Note that PSO Complete requires $O(n^2)$ messaging
+  and may not be practical in many cases.}
   \label{fig:sphere1}
@@
 
-Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
+The benchmark function Griewank, defined by the equation $f(\Vec{x}) =
 \frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
-\right) + 1$.  It is generally recommended to use the Ring topology when
-optimizing the Griewank function, as Complete is prone to premature convergence
+\right) + 1$.  It is generally best solved in pso using the Ring topology,
+as Complete is prone to premature convergence
 on a local optimum.  Griewank has a global optimum with a value of 0, and
-sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
-showing average function value at each iteration, a more accurate plot for
+sometimes the swarm finds the optimum and sometimes it does not.  Instead of
+showing average function value at each iteration, a more enlightening plot for
 Griewank shows the percent of runs that have found the global optimum by each
-iteration.  We show results in \fig{griewank1} for swarms of size 30 and 240
-using the Ring topology.  One can see in the figure that both SEPSO algorithms
-find the optimum faster than the original PSO.  However, because the swarm size
-is so small, SEPSO gets stuck a little less than half of the time.
-Surpisingly, SEPSO Pick Best improves the chance of success with Griewank,
-while also finding the optimum faster.
+iteration~\cite{bratton-2007-defining-a-standard-for-pso}.
 
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank1}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, showing percent successful by each iteration instead of average
-  function value.}
-  \label{fig:griewank1}
-\end{figure}
-
-Because our speculative algorithm got stuck with such a small swarm size, we
-ran a set of experiments on Griewank with 800 processors, giving swarms of size
-100 and 800 instead of 30 and 240.  \fig{griewank2} shows the results.
-\fig{griewank2} is very similar to \fig{griewank1}, except the speculative
-algorithms find the global optimum essentially 100\% of the time.  Thus with a
-few more processors, our algorithms find the optimum two to three times faster
-than the original PSO without a significant loss of accuracy.
+PSO and SEPSO get caught in local minima with small swarm sizes so
+we show results in \fig{griewank1} for swarms of size 100 (SEPSO) and 800 (standard)
+using the Ring topology 
+\fig{griewank2} shows that SEPSO quickly finds
+the global optimum essentially 100\% of the time.
 
@@
 
-Lastly we look at a function that does not show much or any improvement with
-this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
-\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
-with Rastrigin, the more particles there are in the swarm, the lower function
-value it finds, up to at least 4000
-particles~\cite{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
+As noted in the introduction previous work has shown that the optimization of highly
+deceptive functions like Rastrigin
+($f(\Vec{x}) = \sum_{i=1}^D \left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$)
+benefit greatly from the addition of particles. 
+Smaller swarms get caught
 in local optima.  Because our speculative algorithms have a significantly
@@
 performing regular PSO continue to improve the best value found.
-\fig{rastrigin} shows the results graphically.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{rastrigin}
-  \caption{Function Rastrigin with a swarm that uses 240 processors per round
-  of evaluations.}
-  \label{fig:rastrigin}
-\end{figure}
+Our experiments with SEPSO on Rastrigin were predicable lack luster,
+yielding a vaerage value of xxx after x evaluations, as compared to y for 
+standard PSO.
 

### 2010-04-20 honors_thesis/ppsn_paper/paper.tex
Down to 11 pages on the PPSN paper, but still some work to go

@@
 The methods they have used include various synchronous algorithms
-\cite{chu-2006-intelligent-parallel-pso, jin-2005-pso-antenna-designs,
-parsopoulos-2004-parallel-vector-evaluated-pso,
+\cite{chu-2006-intelligent-parallel-pso,
 schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
-\cite{mostaghim-2006-multi-objective-pso-on-grids,
-venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
+\cite{venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
 evaluation of the objective function can also be done in some cases, though
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{evals_sphere}
+  \includegraphics[width=.45\columnwidth]{evals_sphere}
   \caption{Function Sphere with various swarm sizes. Error bars show median and
@@
 presents a comparison of the speculative algorithm with the original PSO.  In
-\sec{conclusion} and \sec{future} we conclude and discuss future work.
+\sec{conclusion} we conclude.
 
@@
 
-The value of the objective function at $\ppos_{t+1}$ is given by:
-\begin{align}
-\label{eq:val2update}
-	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \indic_{t+1}^{c}
-	\ \ofunc(\specpos_{t+1}^{c})
-\end{align}
-
 Returning our attention to the computation of $\ppos_{t+1}$ in
@@
 
-In this form the important point to notice is that there are only $7$ values (for
-this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
+In this form the important point to notice is that there are only $7$ values
+(for this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
 \caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
@@
 random numbers in the equation, they are assumed fixed once drawn for any
-particular particle at a specific iteration.  Thus PSO has been refactored
-such that the algorithm can begin computing all $7$ of the objective function
+particular particle at a specific iteration.  Thus PSO has been refactored such
+that the algorithm can begin computing all $7$ of the objective function
 evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
@@
 available, and that the evaluation of the function to be optimized takes one
-hour.  If we only want a swarm of $100$ particles, $700$ of the processors would be
-sitting idle for an hour at every iteration, and it would take two hours to run
-two iterations.  If instead we perform speculative evaluation, sending each of
-the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
-$\ofunc(\ppos_{t})$, we could create a swarm of size 100, each particle with $7$ speculative
-evaluations ($700$ processors dedicated to speculative evaluation), thus using all $800$
-processors and performing two
-iterations in one hour.
+hour.  If we only want a swarm of $100$ particles, $700$ of the processors
+would be sitting idle for an hour at every iteration, and it would take two
+hours to run two iterations.  If instead we perform speculative evaluation,
+sending each of the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same
+time as $\ofunc(\ppos_{t})$, we could create a swarm of size 100, each particle
+with $7$ speculative evaluations ($700$ processors dedicated to speculative
+evaluation), thus using all $800$ processors and performing two iterations in
+one hour.
 
-In order to do two iterations at once, we use 8 times as many processors
-as there are particles in the swarm.  If these processors were not performing speculative
-evaluation, they might instead be used for function evaluation needed to support a large swarm.
-This raises the question of whether a swarm of 100 particles doing
-twice as many iterations outperforms a swarm of 800 particles.  We show in the
-rest of this paper that in many, instances, a smaller swarm
-performing more iterations does in fact outperform a larger swarm.
-We acknowledge that both intuition and prior research~\cite{mcnabb-2009-large-particle-swarms}
-indicate that the optimization of deceptive functions benefits greatly from
-large and even very large swarm sizes. Thus this work will focus on less deceptive functions.
+In order to do two iterations at once, we use 8 times as many processors as
+there are particles in the swarm.  If these processors were not performing
+speculative evaluation, they might instead be used for function evaluation
+needed to support a large swarm.  This raises the question of whether a swarm
+of 100 particles doing twice as many iterations outperforms a swarm of 800
+particles.  We show in the rest of this paper that in many, instances, a
+smaller swarm performing more iterations does in fact outperform a larger
+swarm.  We acknowledge that both intuition and prior
+research~\cite{mcnabb-2009-large-particle-swarms} indicate that the
+optimization of deceptive functions benefits greatly from large and even very
+large swarm sizes. Thus this work will focus on less deceptive functions.
 
-\subsection{Topology}
-\label{sec:topology}
+\subsection{Implementation}
+\label{sec:implementation}
 
 The number of speculative evaluations needed per particle depends on the number
-of neighbors each particle has.  The number of update cases in a topology where
-each particle has $n$ neighbors is $2(n+1)$; there are two possibilities for
-updates to $\pbest$ (updated by the particle itself and not updated), and $n+1$
-possibilities for updates to $\nbest$ (updated by each neighboring particle and
-not updated).  When the particle is also a neighbor to itself, as is always the
-case in commonly used topologies, one of the cases can be eliminated, as a
-particle cannot be the source of an update to its neighborhood best while not
-also updating its personal best.  Thus we have $2(n+1)-1$, or $2n+1$,
-speculative evaluations per particle.  In a swarm with $p$ particles and $n$
-neighbors per particle, $(2n+1)p$ speculative evaluations are needed.
+of neighbors each particle has.  In a swarm with $p$ particles and $n$
+neighbors per particle, $(2n+1)p$ speculative evaluations are necessary (each
+additional neighbor adds two rows to Table~\ref{tab:evals}).  This dependence
+on the number of neighbors necessitates a wise choice of topology.  The use of
+the Complete topology, where every particle is a neighbor to every other
+particle, would require $O(p^2)$ speculative evaluations per iteration.  It is
+much more desirable to have a sparse topology, where $O(np)$ is much smaller
+than $O(p^2)$.  However, some functions are better optimized with the Complete
+topology and the quick spread of information it entails than with sparse
+topologies.  In such cases, we use the Random topology described in \sec{pso}.
 
-Because the number of speculative evaluations depends on the number of
-neighbors a particle has, the choice of topology is an important one.  The use
-of the Complete topology, where every particle is a neighbor to every other
-particle, would require $O(p^2)$ speculative evaluations per iteration.
-Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
-much smaller than $O(p^2)$.  However, some functions are better optimized with
-the Complete topology and the quick spread of information it entails than with
-sparse topologies.  In such cases, we use the Random topology 
-which has been shown to
-effectively approximate the rapid information flow of the Complete topology.
-In our implementation of the random topology there are to randomly chosen neighbors
-(chosen anew at every iteration)~\cite{mcnabb-2009-large-particle-swarms}.
-
-In the rest of this
-section we move from the mathematics of our method to its implementation.
-First we discuss the relatively easy case of a centralized parallel PSO
-algorithm with a master computer and many slaves.  In such an architecture, the
-master keeps track of all necessary information with only trivial message
-passing needed.  Then we discuss the more complicated case of a distributed
-algorithm, where each particle is on its own and needs to send and receive
-messages to and from other particles.  Finally we discuss the further
-complications of a dynamic topology, where a particle's neighbors change from
-one iteration to another.
-
-\subsection{Terminology}
-
-To aid in describing our implementation, we introduce a few terms. We use $\p_t$ to denote a particle at
-iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
-corresponding to one of the rows in Table~\ref{tab:evals}.  $\n_t$ is a
-neighbor of particle $\p_t$.  Sets of particles are given by $\pset$, $\sset$,
-or $\nset$, whereas single particles are simply $\p$, $\s$, or $\n$.
-
-We separate each iteration of PSO into several steps.  First there is the
-motion step, where a particle updates its position and velocity.  Then a
-particle's position is evaluated, and the particle updates its current value
-and its personal best.  Finally, a particle gets information from its neighbors
-and updates its neighborhood best.
+To aid in describing our implementation, we introduce a few terms. We use
+$\p_t$ to denote a particle at iteration $t$ and $\s_{t+1}$ to denote one of
+$\p_t$'s speculative children, corresponding to one of the rows in
+Table~\ref{tab:evals}.  $\n_t$ is a neighbor of particle $\p_t$.  Sets of
+particles are given by $\pset$, $\sset$, or $\nset$, whereas single particles
+are simply $\p$, $\s$, or $\n$.
 
 A particle at iteration $t-1$ that has been moved to iteration $t$ using
-\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has
-not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has
-been evaluated, but it has still not yet received information from its
-neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
-updated its neighborhood best is it a complete particle at iteration $t$.  It
-is then simply denoted as $\p_t$.
+\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has not yet
+been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has been
+evaluated, but it has still not yet received information from its neighbors, it
+is denoted as $\nonbest{\p}_t$.  Only when the particle has updated its
+neighborhood best is it a complete particle at iteration $t$.  It is then
+simply denoted as $\p_t$.
 
-\subsection{Implementation}
-
-In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
-master, keeps track of all necessary information, and all other machines are
-merely used to evaluate the objective function at various positions as directed
-by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
-speculative evaluation in such an architecture, the master generates the
-positions to evaluate speculatively as in \eqref{eq:defpcasegen}.  After having
-the slaves evaluate the objective function at all necessary positions, the
-master then decides which position to accept for each particle, as in
-\eqref{eq:val2updatelong}.  The outline of the speculative evaluation in PSO (SEPSO) algorithm is given in
+The outline of the speculative evaluation in PSO (SEPSO) algorithm is given in
 \alg{centralized}.
@@
 \end{algorithm}
-
-In a distributed parallel PSO implementation, individual processors not only perform
-evaluations of particles, but also their movement.  The information for each
-particle is not held by a central machine that directs the algorithm; instead,
-each processor has the information for the particle or particles that it is in
-charge of and must perform the steps of the algorithm for those
-particles.  Messages such as
-values and positions for the neighborhood best are sent between processors.
-There may still be some machine that collects information from all of the
-particles and outputs the result of the algorithm, though that machine's
-importance is much less than in centralized algorithms.
-In some distributed frameworks rounds of message passing can be expensive.
-A N\:aive implementation of speculative evaluation
-uses three rounds of message passing for every two iterations
-(corresponding to steps 2, 5 and 10 in \alg{centralized}).  It is possible to
-perform speculative evaluation in PSO with only one round of communication per
-two iterations. 
-Dynamic topologies like the random topology discussed above, further complicate the
-needed message passing.
-However, the message passing implementation is not particularly
-interesting, but is tedious thus, in the interest of space, is omitted.
 
@@
 
-As discussed in \sec{topology}
+As discussed in \sec{implementation},
 where the Complete topology would normally be used, we use a Random topology in
-our speculative algorithm, as complete leads to an explosion of speculative
+our speculative algorithm, as Complete leads to an explosion of speculative
 evaluations.
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{sphere1}
+  \includegraphics[width=.45\columnwidth]{sphere1}
   \caption{Function Sphere with a swarm that uses 240 processors per round of
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank2}
+  \includegraphics[width=.45\columnwidth]{griewank2}
   \caption{Function Griewank with a swarm that uses 800 processors per round of
@@
 particles there is no point at which ``enough'' exploration has been
-done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
-cannot be easily prescribed and depends largely on the desired outcome and how
-many processors are available.  It is likely that the best thing to do with
-those functions is to use all processors to add particles to the swarm and do
-no speculative evaluation.
-
-\section{Future Work}
-\label{sec:future}
-
-This work represents the first attempt of which we are aware at using
-additional processors in parallel with PSO to do something other than either
-add particles to the swarm or parallelize the evaluation of the objective
-function.  We anticipate that many future advancements can be made.  For
-example, our technique required speculatively evaluating every possible next
-position of each particle.  The number of speculative evaluations could
-potentially be greatly reduced by using statistics for which branch of motion
-particles most often follow.  Only the subset of the most probable next
-locations would then be evaluated.  Using an asynchronous PSO variant may also
-be helpful in this case, allowing particles to be at different iterations in
-case the correct branch was not guessed and needs to be evaluated separately.
-
-Another possible advancement would be to reduce communication costs by dropping
-some of the inter-processor messages in distributed frameworks.  The PSO
-algorithm would have to be adapted to work with less information, perhaps not
-requiring that each processor have completely accurate information about what
-neighboring particles are doing.
-
-Still another possibility is to view the set of possible evaluations as an
-infinite tree.  Each level in the tree is an iteration, and the branches on
-each level represent the possible next positions of the particle.  One need not
-simply evaluate all of one level for each particle---perhaps going two or three
-iterations ahead on the most likely path would perform better.
-
-We have mentioned that this technique has been previously used to parallelize
-simulated annealing.  Any other optimization algorithm that only depends on
-current sampling positions when computing the next position to sample can be
-parallelized with this technique.  In particular, genetic algorithms produce
-future generations by combining individuals from the current generation.  With
-a large population size there would be an unwieldy amount of possible future
-individuals, but the potential exists to modify the algorithm to use some kind
-of speculative evaluation.  A related but quite distinct technique for reducing
-computation in evolutionary algorithms was proposed
-in~\cite{poli-2006-backward-chaining-evolutionary-algorithms}.
-
-Large parallel clusters are often required to successfully optimize practical
-modern problems.  To properly use PSO with such clusters, a balance needs to be
-made between using processors to increase the swarm size and using them to
-increase the speed of the algorithm, or for perhaps as yet unconsidered
-techniques.  This work is a first step in that direction.
+done~\cite{mcnabb-2009-large-particle-swarms}.  For those functions the correct
+use of processors cannot be easily prescribed and depends largely on the
+desired outcome and how many processors are available.  It is likely that the
+best thing to do with those functions is to use all processors to add particles
+to the swarm and do no speculative evaluation.
 

### 2010-04-20 honors_thesis/ppsn_paper/paper.tex
fixed some formatting in equations and cut a line
combined equations for compactness

@@
 \newcommand{\alg}[1]{Algorithm~\ref{alg:#1}}
+
+% Mathy help
+\newcommand{\parens}[1]{\!\left(#1\right)}
 
@@
 indicator function for the specific case $\casexn$ can be written as follows:
-
+$f(x)$
 \begin{align}
   \nonumber
-	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
-	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
-  \label{eq:deficasexn}
-	&\begin{cases}
-	   1 & \text{if} \ [ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) ]
-	    \wedge [ \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) ] \\
-	   &\quad \wedge \quad [ \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) ]
-	    \wedge [ \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) ] \\
+	&\indic_{t+1}^{\casexn}\parens{
+	  \ofunc\parens{\ppos_{t}},
+	  \ofunc\parens{\leftn_{t}},
+	  \ofunc\parens{\rightn_{t}},
+	  \ofunc\parens{\pbest_{t-1}},
+	  \ofunc\parens{\nbest_{t-1}}
+	  } \\
+  &\hspace{15pt}= \label{eq:deficasexn}
+	\begin{cases}
+	  1 & \text{if}~
+		\ofunc\parens{\ppos_{t}} < \ofunc\parens{\pbest_{t-1}}
+	    \,\text{and}\,
+		\ofunc\parens{\nbest_{t-1}} < \min\parens{
+		  \ofunc\parens{\ppos_{t}},
+		  \ofunc\parens{\leftn_{t}},
+		  \ofunc\parens{\rightn_{t}}
+		} \\
 	   0 & \text{otherwise}
@@
 \nonumber
-	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
-	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
+	\specvel_{t+1}^{\casexn} & \parens{\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
+	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}} \\
 \label{eq:defvcasexn}
-		&= \constriction \bigl[ \pvel_{t} +
-			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t})
-			+ \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
-			\ppos_{t}) \bigr]
+		&= \constriction \left[ \pvel_{t} +
+			\pcoeff\pURand_{t}\otimes\parens{\ppos_{t} - \ppos_{t}}
+			+ \ncoeff\nURand_{t}\otimes\parens{\nbest_{t-1} -
+			\ppos_{t}} \right]
 \end{align}
@@
 \label{eq:defpcasegen}
-	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
-	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t})
+	\specpos_{t+1}^{\casegen} & \parens{\ppos_{t}, \pvel_{t}, \leftn_{t},
+	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}}
 	 = \ppos_{t} + \specvel_{t+1}^{\casegen}
@@
 indicator functions.  Again, we represent the indicator functions and velocity
-functions without the parameters for compactness.  The equation becomes:
+functions without the parameters for compactness.
+The velocity update from~\eqref{eq:velupdate} becomes:
 \begin{align}
-\nonumber
-	\pvel_{t+1} &=
-		\constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
-			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
-			\ppos_{t}) \bigr] \\
-\nonumber
-	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
-			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
-			\ppos_{t}) \bigr]  \\
 \label{eq:vel2update}
+%\nonumber
+	\pvel_{t+1} 
+%		\constriction \bigl[ \pvel_t
+%			+ \pcoeff\pURand_{t}\otimes\parens{\pbest_{t} - \ppos_{t}}
+%			+ \ncoeff\nURand_{t}\otimes\parens{\nbest_{t} -
+%			\ppos_{t}} \bigr] \\
+%\nonumber
+%	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
+%			+ \pcoeff\pURand_{t}\otimes\parens{\pbest_{t} - \ppos_{t}}
+%			+ \ncoeff\nURand_{t}\otimes\parens{\nbest_{t} -
+%			\ppos_{t}} \bigr]  \\
 	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
@@
 
-Similarly, the position update equation~\eqref{eq:posupdate} becomes:
-\begin{align}
-\label{eq:pos2update}
-	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
-	= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
-\end{align}
-
-Returning our attention to the computation of $\ppos_{t+1}$ in
-\eqref{eq:pos2update} and writing it with the parameters which were omitted
-above, we obtain:
+Similarly, the position update from~\eqref{eq:posupdate}
+can also be rewritten using the indicator functions, now with the parameters reinserted:
 \begin{align}
@@
   \ppos_{t+1} &= \sum_{c \in \caseset} \bigl[
-	\indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
-	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
+	\indic_{t+1}^{c}\parens{
+	  \ofunc\parens{\ppos_{t}},
+	  \ofunc\parens{\leftn_{t}},
+	  \ofunc\parens{\rightn_{t}},
+	  \ofunc\parens{\pbest_{t-1}},
+	  \ofunc\parens{\nbest_{t-1}}
+	  } \\
 \label{eq:val2updatelong}
-	& \quad \quad \quad \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
-	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t}) \bigr]
+	& \quad \quad \quad \specpos_{t+1}^{c}\parens{
+	  \ppos_{t},
+	  \pvel_{t},
+	  \leftn_{t},
+	  \rightn_{t},
+	  \pbest_{t-1},
+	  \nbest_{t-1},
+	  \pURand_{t},
+	  \nURand_{t}
+	  }\bigr]
 \end{align}

### 2010-04-20 honors_thesis/ppsn_paper/paper.tex
removed compactness for compactness

@@
 indicator function for the specific case $\casexn$ can be written as follows:
-$f(x)$
 \begin{align}
@@
 In the same way we can create notation for the position update function by
-substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
-parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
-subscripts.
+substituting into~\eqref{eq:posupdate}:
 \begin{align}
-\label{eq:defpcasegen}
+    \nonumber
 	\specpos_{t+1}^{\casegen} & \parens{\ppos_{t}, \pvel_{t}, \leftn_{t},
 	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}}
-	 = \ppos_{t} + \specvel_{t+1}^{\casegen}
+        \\
+    \label{eq:defpcasegen}
+        &=
+	\ppos_{t} + \specvel_{t+1}^{\casegen}
+	\parens{\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
+	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}}
 \end{align}
@@
 equation~\eqref{eq:velupdate}, introducing our sum over cases with the
-indicator functions.  Again, we represent the indicator functions and velocity
-functions without the parameters for compactness.
-The velocity update from~\eqref{eq:velupdate} becomes:
+indicator functions.
+The velocity~\eqref{eq:velupdate} and position~\eqref{eq:posupdate} equations become:
 \begin{align}
-\label{eq:vel2update}
-%\nonumber
-	\pvel_{t+1} 
-%		\constriction \bigl[ \pvel_t
-%			+ \pcoeff\pURand_{t}\otimes\parens{\pbest_{t} - \ppos_{t}}
-%			+ \ncoeff\nURand_{t}\otimes\parens{\nbest_{t} -
-%			\ppos_{t}} \bigr] \\
-%\nonumber
-%	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
-%			+ \pcoeff\pURand_{t}\otimes\parens{\pbest_{t} - \ppos_{t}}
-%			+ \ncoeff\nURand_{t}\otimes\parens{\nbest_{t} -
-%			\ppos_{t}} \bigr]  \\
-	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
-\end{align}
-
-Similarly, the position update from~\eqref{eq:posupdate}
-can also be rewritten using the indicator functions, now with the parameters reinserted:
-\begin{align}
+\nonumber
+  \pvel_{t+1} &= \sum_{c \in \caseset} \bigl[
+	\indic_{t+1}^{c}\parens{
+	  \ofunc\parens{\ppos_{t}},
+	  \ofunc\parens{\leftn_{t}},
+	  \ofunc\parens{\rightn_{t}},
+	  \ofunc\parens{\pbest_{t-1}},
+	  \ofunc\parens{\nbest_{t-1}}
+	  } \\
+\label{eq:specvelupdate}
+	& \quad \quad \quad \specvel_{t+1}^{c}\parens{
+	  \ppos_{t},
+	  \pvel_{t},
+	  \leftn_{t},
+	  \rightn_{t},
+	  \pbest_{t-1},
+	  \nbest_{t-1},
+	  \pURand_{t},
+	  \nURand_{t}
+	  }\bigr]
+\\
 \nonumber
@@
 	  } \\
-\label{eq:val2updatelong}
+\label{eq:specposupdate}
 	& \quad \quad \quad \specpos_{t+1}^{c}\parens{
@@
 	  the branch taken by $\p_t$ according to
-	  \eqref{eq:val2updatelong}.
+	  \eqref{eq:specposupdate}.
 	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,

### 2010-04-20 honors_thesis/ppsn_paper/paper.tex
shortened more text; first person shooting

@@
 iterations and fewer particles.  Accordingly, we take inspiration from 
-speculative execution commonly performed in modern processors and propose what
-we call speculative evaluation in PSO (SEPSO).  Future positions of the particles are
+speculative execution performed in modern processors and propose
+speculative evaluation in PSO (SEPSO).  Future positions of the particles are
 speculated and evaluated in parallel with current positions, performing two
@@
 hundreds of processors and functions for which a single evaluation will take
-at least seconds, but probably minutes.  Our purpose
-is to explore the question of what to do with a hundred processors when 50 or
-100 particles is the recommended swarm size, and simply increasing the swarm
-size encounters diminishing returns.
+at least several seconds.  Our purpose
+is to explore the question of what to do with hundreds of processors when 50 or
+100 particles is the ideal swarm size, and simply adding particles
+yields diminishing returns.
 
@@
 prediction) a technique commonly used in processors.  Modern processors, when
-faced with a branch on which they must wait (e.g. a memory cache miss), guess
+faced with a branch on which they must wait (e.g., a memory cache miss), guess
 which way the branch will go and start executing, ensuring that any changes
@@
 In this paper we show that the results of standard PSO can be reproduced
-\emph{exactly}, two iterations at a time, using a speculative approach adapted
-from speculative execution. We prove that the standard PSO equation can be
+\emph{exactly}, two iterations at a time, using a speculative approach similar
+to speculative execution. We prove that the standard PSO equation can be
 factored such that a set of speculative positions can be found which will
@@
 the current position, it is possible to know the objective function values for
-both the current and the next iteration at the same time.  We demonstrate this
-principle by implementation and show that it produces exactly the same results
-as standard PSO, but two iterations at a time.  The resulting implementation
+both the current and the next iteration at the same time.
+The resulting implementation
 runs efficiently on large clusters where the number of processors is much
@@
 Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
-Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is
-used to intelligently search a multi-dimensional space by mimicking the
-swarming and flocking behavior of birds and other animals. It is a social
-algorithm that depends on interaction between particles to quickly and
-consistently approximate the optimal solution to a given objective function.
+Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  This social
+algorithm, inspired by the flocking behavior of birds, is used to quickly and
+consistently find the global optimum of a given objective function in a
+multi-dimensional space.
 
@@
 
-Changing the way neighbors are defined (often called the ``topology'' in the
-literature) has a significant effect on the performance of the algorithm.  In
+Changing the way neighbors are defined, usually called the ``topology,''
+has a significant effect on the performance of the algorithm.  In
 the Ring topology, each particle has one neighbor to either side of it; in the
@@
 
-The PSO algorithm can be trivially parallelized by distributing the processing
-needed for each particle on across processors.  But as we have seen in
-\fig{evals-sphere}, for some functions, and for large numbers of
-processors, just adding particles reaches a point of diminishing returns.
-That is, adding processors does not help us reach any given level of fitness
-appreciably faster.  Instead of adding particles we employ a speculative
-evaluation approach that allows us to perform iterations two at a time.
+The PSO algorithm can be trivially parallelized by distributing the computation
+needed for each particle across processors.  But
+for some functions, as shown in \fig{evals-sphere},
+adding particles yields diminishing returns.
+That is, adding processors does not help reach any given level of fitness
+appreciably faster.  Instead of adding particles, speculative
+evaluation performs iterations two at a time.
 
-To do so we must first refactor PSO such that the determination of the value
-of the objective function is separate from the rest of the computation.  For
+Speculative evaluation is made possible by refactoring
+PSO such that evaluating the
+the objective function is separate from the rest of the computation.  For
 simplicity, this discussion will describe the case where PSO is performing
@@
 positions are represented as $\rightn$ and $\leftn$ respectively.  Though we
-will only describe the case of the Ring topology, this method can easily be
+will only describe the case of the Ring topology, this method is easily
 extended to arbitrary topologies.

### 2010-04-20 honors_thesis/ppsn_paper/paper.tex
Got rid of figure in intro, almost down to 10 pages

@@
 particles in the swarm over the available processors.  If more processors are
-available, these techniques increase the number of particles in the swarm,
-either by adding individual particles or by adding entire new sub-swarms, but
-in all cases the number of processors does not exceed the number of particles.
-The number of iterations of PSO that the algorithms can perform is thus
-inherently limited by the time it takes to evaluate the objective
-function---additional processors add more particles, but do not make the
-iterations go any faster.
+available, these techniques increase the number of particles in the swarm.  The
+number of iterations of PSO that the algorithms can perform is thus inherently
+limited by the time it takes to evaluate the objective function---additional
+processors add more particles, but do not make the iterations go any faster.
 
 For many functions there comes a point of diminishing returns with respect to
-adding particles.  In \fig{evals-sphere} we show the value obtained after
-50,000 function evaluations (not iterations) as a function of swarm size for
-the well-know benchmark function Sphere (20 dimensions, reporting the average
-of twenty runs).  Increasing the swarm size from 5 to 10 has a significant
-effect on the value obtained.  However, increasing the swarm size from 16 to 30
-makes the algorithm less efficient, that is, it reduces the progress the
-algorithm makes per evaluation.  Other functions show similar trends, though
-often the optimal swarm size is larger.  For this reason, previous
-work has recommended the use of a swarm size of 50 for
+adding particles.  Very small swarms do not produce enough exploration to
+consistently find good values, while large swarms result in more exploration
+than is necessary and waste computation.  For this reason, previous work has
+recommended the use of a swarm size of 50 for
 PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=.45\columnwidth]{evals_sphere}
-  \caption{Function Sphere with various swarm sizes. Error bars show median and
-  10th and 90th percentiles.}
-  \label{fig:evals-sphere}
-\end{figure}
-
 In this paper we consider PSO parallelization strategies for clusters of
-hundreds of processors and functions for which a single evaluation will take
-at least several seconds.  Our purpose
-is to explore the question of what to do with hundreds of processors when 50 or
-100 particles is the ideal swarm size, and simply adding particles
-yields diminishing returns.
+hundreds of processors and functions for which a single evaluation will take at
+least several seconds.  Our purpose is to explore the question of what to do
+with hundreds of processors when 50 or 100 particles is the ideal swarm size,
+and simply adding particles yields diminishing returns.
 
@@
 comes from an analogy to speculative execution (also known as branch
-prediction) a technique commonly used in processors.  Modern processors, when
+prediction), a technique commonly used in processors.  Modern processors, when
 faced with a branch on which they must wait (e.g., a memory cache miss), guess
-which way the branch will go and start executing, ensuring that any changes
-can be undone.  If the processor guesses right, execution is
-much farther ahead than if it had idly waited on the memory reference.  If
-it guesses wrong, execution restarts where it would have been anyway.
+which way the branch will go and start executing, ensuring that any changes can
+be undone.  If the processor guesses right, execution is much farther ahead
+than if it had idly waited on the memory reference.  If it guesses wrong,
+execution restarts where it would have been anyway.
 
-In this paper we show that the results of standard PSO can be reproduced
-\emph{exactly}, two iterations at a time, using a speculative approach similar
-to speculative execution. We prove that the standard PSO equation can be
-factored such that a set of speculative positions can be found which will
-\emph{always} include the position computed in the next iteration.  By
-computing the value of the objective function for each of the speculative
-positions at the same time the algorithm evaluates the objective function for
-the current position, it is possible to know the objective function values for
-both the current and the next iteration at the same time.
-The resulting implementation
-runs efficiently on large clusters where the number of processors is much
-larger than a typical or reasonable number of particles, producing better
-results in less ``wall-clock'' time.
+We show that the results of standard PSO can be reproduced \emph{exactly}, two
+iterations at a time, using a speculative approach similar to speculative
+execution. We prove that the standard PSO equation can be factored such that a
+set of speculative positions can be found which will \emph{always} include the
+position computed in the next iteration.  By computing the value of the
+objective function for each of the speculative positions at the same time the
+algorithm evaluates the objective function for the current position, it is
+possible to know the objective function values for both the current and the
+next iteration at the same time.  The resulting implementation runs efficiently
+on large clusters where the number of processors is much larger than a typical
+or reasonable number of particles, producing better results in less
+``wall-clock'' time.
 
@@
 particle swarm optimization algorithm.  \sec{sepso} describes how speculative
-evaluation can be done in parallel PSO to perform two iterations at once, then
-presents a comparison of the speculative algorithm with the original PSO.  In
-\sec{conclusion} we conclude.
+evaluation can be done in parallel PSO to perform two iterations at once.
+\sec{extra} then introduces a variant of the speculative algorithm that in some
+cases performs better.  In \sec{results} and \sec{conclusion} we present our
+results and conclude.
 
@@
 The PSO algorithm can be trivially parallelized by distributing the computation
-needed for each particle across processors.  But
-for some functions, as shown in \fig{evals-sphere},
-adding particles yields diminishing returns.
-That is, adding processors does not help reach any given level of fitness
-appreciably faster.  Instead of adding particles, speculative
-evaluation performs iterations two at a time.
+needed for each particle across processors.  But for some functions adding
+particles yields diminishing returns.  That is, adding processors does not help
+reach any given level of fitness appreciably faster.  Instead of adding
+particles, speculative evaluation performs iterations two at a time.
 
@@
 To see the value of this refactoring, suppose that $800$ processors are
-available, and that the evaluation of the function to be optimized takes one
-hour.  If we only want a swarm of $100$ particles, $700$ of the processors
-would be sitting idle for an hour at every iteration, and it would take two
-hours to run two iterations.  If instead we perform speculative evaluation,
-sending each of the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same
-time as $\ofunc(\ppos_{t})$, we could create a swarm of size 100, each particle
-with $7$ speculative evaluations ($700$ processors dedicated to speculative
+available, and that the evaluation of the objective function takes one hour.
+If we only want a swarm of $100$ particles, $700$ of the processors would be
+sitting idle for an hour at every iteration, and it would take two hours to run
+two iterations.  If instead we perform speculative evaluation, sending each of
+the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
+$\ofunc(\ppos_{t})$, we could create a swarm of size 100, each particle with
+$7$ speculative evaluations ($700$ processors dedicated to speculative
 evaluation), thus using all $800$ processors and performing two iterations in
@@
 be the best use of parallelism in PSO.  But for many real-world applications,
-the objective function takes on the order of at least seconds (or more) to evaluate; in
-such cases our assumption is perfectly reasonable.
+the objective function takes on the order of at least seconds (or more) to
+evaluate; in such cases our assumption is perfectly reasonable.
 
-In each set of experiments we keep the number of processors for each algorithm constant.
-In all of our SEPSO experiments each particle has
-two neighbors in addition to itself.
-As shown in
-Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
-particle.  With one evaluation needed for the original, non-speculative
-particle, we have a total of $8p_s$ evaluations for every two iterations, where $p_s$ is
-the number of particles in the speculative swarm.
-We run these experiments on $8p_s$ processors.
-In the base-line experiments, standard pso the same number of processors and thus enjoys the
-benefit of $8p_s$ particles running on 
-$8p_s$ processors.
-At least in benchmark functions, extra particles only accelerate
-the performance of PSO~\cite{mcnabb-2009-large-particle-swarms}. They generally have diminishing value, as
-seen earlier in \fig{evals-sphere}, but these extra particles give standard PSO the best possible
-context and full access to the same resources as SEPSO.
+In each set of experiments we keep the number of processors for each algorithm
+constant.  In all of our SEPSO experiments each particle has two neighbors in
+addition to itself.  As shown in Table~\ref{tab:evals}, this results in $7$
+speculative evaluations per particle.  With one evaluation needed for the
+original, non-speculative particle, we have a total of $8p_s$ evaluations for
+every two iterations, where $p_s$ is the number of particles in the speculative
+swarm.  We run these experiments on $8p_s$ processors.  In the base-line
+experiments, standard pso the same number of processors and thus enjoys the
+benefit of $8p_s$ particles running on $8p_s$ processors.  At least in
+benchmark functions, extra particles only accelerate the performance of
+PSO~\cite{mcnabb-2009-large-particle-swarms}. They generally have diminishing
+value, as seen earlier in \fig{evals-sphere}, but these extra particles give
+standard PSO the best possible context and full access to the same resources as
+SEPSO.
 

### 2010-04-21 honors_thesis/ppsn_paper/paper.tex
Finished PPSN version of the paper and submitted it

@@
 
-%\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
-\author{Author names omitted for review}
+\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
+\institute{Department of Computer Science, Brigham Young University}
 
@@
 adding more particles to the swarm or by parallelizing the evaluation of the
-objective function.  However, some functions are more efficiently optimized with more
-iterations and fewer particles.  Accordingly, we take inspiration from 
-speculative execution performed in modern processors and propose
+objective function.  However, some functions are more efficiently optimized
+with more iterations and fewer particles.  Accordingly, we take inspiration
+from speculative execution performed in modern processors and propose
 speculative evaluation in PSO (SEPSO).  Future positions of the particles are
@@
 The methods they have used include various synchronous algorithms
-\cite{chu-2006-intelligent-parallel-pso,
+\cite{belal-2004-parallel-models-for-pso,
 schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
-\cite{venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
+\cite{mostaghim-2006-multi-objective-pso-on-grids,
+venter-2005-parallel-pso-asynchronous-evaluations,
+koh-2006-parallel-asynchronous-pso}.  Parallelizing the
 evaluation of the objective function can also be done in some cases, though
@@
 iterations at a time, using a speculative approach similar to speculative
-execution. We prove that the standard PSO equation can be factored such that a
+execution. We prove that the standard PSO equations can be factored such that a
 set of speculative positions can be found which will \emph{always} include the
@@
 particle swarm optimization algorithm.  \sec{sepso} describes how speculative
-evaluation can be done in parallel PSO to perform two iterations at once.
-\sec{extra} then introduces a variant of the speculative algorithm that in some
-cases performs better.  In \sec{results} and \sec{conclusion} we present our
-results and conclude.
+evaluation can be done in parallel PSO to perform two iterations at once.  In
+\sec{results} and \sec{conclusion} we present our results and conclude.
 
@@
 neighborhood best) is the best position the neighbors of the current particle
-have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
-\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
-required to ensure convergence (2.05, 2.05, and .73,
-respectively)~\cite{clerc-2002-constricted-pso}.
+have seen.  The parameters \( \ncoeff \), \( \pcoeff \), and \( \constriction
+\) are given prescribed values required to ensure convergence (2.05, 2.05, and
+.73, respectively)~\cite{clerc-2002-constricted-pso}.
 
@@
 
-Although this speculative approach computes
-$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
-for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will be
-ignored, and might just as well never have been computed.  We call the set of
-computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
-``speculative children'' because only one of them will be needed.
+Although this speculative approach computes $\ofunc(\specpos_{t+1}^{\casegen})$
+for all $\casegen \in \caseset$, even those for which $\indic_{t+1}^{\casegen}
+= 0$, these extra computations will be ignored, and might just as well never
+have been computed.  We call the set $\{\ofunc(\specpos_{t+1}^{c}) : \casegen
+\in \caseset\}$ ``speculative children'' because only one of them will be
+needed.
 
@@
 of 100 particles doing twice as many iterations outperforms a swarm of 800
-particles.  We show in the rest of this paper that in many, instances, a
-smaller swarm performing more iterations does in fact outperform a larger
-swarm.  We acknowledge that both intuition and prior
+particles.  We show in the rest of this paper that in many instances a smaller
+swarm performing more iterations does in fact outperform a larger swarm.  We
+acknowledge that both intuition and prior
 research~\cite{mcnabb-2009-large-particle-swarms} indicate that the
@@
 A particle at iteration $t-1$ that has been moved to iteration $t$ using
-\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has not yet
+\eqref{eq:velupdate} and \eqref{eq:posupdate}, but whose position has not yet
 been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has been
@@
 
-\section{Using All Speculative Evaluations}
+\subsection{Using All Speculative Evaluations}
 \label{sec:extra}
@@
 speculative evaluations are done per particle, while all but one of them are
-completely ignored.  It seems reasonable to try to make use of the
-information obtained through those evaluations instead of ignoring it
-Making use of this information changes the
-behavior of PSO, instead of reproducing it exactly as the above method
-explains, but the change turns out to be an improvement in our context.
+completely ignored.  It seems reasonable to try to make use of the information
+obtained through those evaluations instead of ignoring it.  Making use of this
+information changes the behavior of PSO, instead of reproducing it exactly as
+the above method explains, but the change turns out to be an improvement in our
+context.
 
-To make better use of the speculative evaluations, instead of choosing the speculative
-child that matches that branch that the original PSO would have taken, we take
-the child that has the best value.  The methodology is exactly the same as
-above except for the process of choosing which speculative child to accept.
-The branch that PSO would have taken is ignored, but $\p_t$ must still be
-created so it can give the right personal best and neighborhood best values and
-positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
-step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
-$\noeval{\sset}_{t+1}$ instead of with the matching branch. We call this
-varient ``Pick Best''.
-
-This can be thought of as drawing a number of samples from the next iteration
-and accepting the best one.  Speculative particles that move in good directions
-are kept.  It seems this modification to PSO would contribute local exploration,
-having knowledge of all of the speculative children, but more global exploitation,
-since Pick Best only considers the best of these alternatives.
-With functions that are less deceptive, or only locally deceptive,
-Pick Best works well, while performance on more deceptive
-functions would probably be hurt.
+To make better use of the speculative evaluations, instead of choosing the
+speculative child that matches that branch that the original PSO would have
+taken, we take the child that has the best value.  The methodology is exactly
+the same as above except for the process of choosing which speculative child to
+accept.  The only change needed in \alg{centralized} is in step 7, where the
+$\noeval{\s}_{t+1}$ with the best value is chosen from $\noeval{\sset}_{t+1}$
+instead of with the matching branch. We call this variant ``Pick Best''.
 
@@
 non-speculative.  The speculative algorithm actually performs two iterations of
-PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
-evaluations.  For benchmark functions with very fast evaluations this may not
+PSO at each ``iteration,'' so we instead call each ``iteration'' a ``round of
+evaluations.''  For benchmark functions with very fast evaluations this may not
 be the best use of parallelism in PSO.  But for many real-world applications,
 the objective function takes on the order of at least seconds (or more) to
-evaluate; in such cases our assumption is perfectly reasonable.
+evaluate; in such cases our framework is reasonable.
 
 In each set of experiments we keep the number of processors for each algorithm
-constant.  In all of our SEPSO experiments each particle has two neighbors in
-addition to itself.  As shown in Table~\ref{tab:evals}, this results in $7$
-speculative evaluations per particle.  With one evaluation needed for the
-original, non-speculative particle, we have a total of $8p_s$ evaluations for
-every two iterations, where $p_s$ is the number of particles in the speculative
-swarm.  We run these experiments on $8p_s$ processors.  In the base-line
-experiments, standard pso the same number of processors and thus enjoys the
-benefit of $8p_s$ particles running on $8p_s$ processors.  At least in
-benchmark functions, extra particles only accelerate the performance of
-PSO~\cite{mcnabb-2009-large-particle-swarms}. They generally have diminishing
-value, as seen earlier in \fig{evals-sphere}, but these extra particles give
-standard PSO the best possible context and full access to the same resources as
-SEPSO.
+constant.  In all of our experiments SEPSO uses topologies in which each
+particle has two neighbors in addition to itself.  As shown in
+Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
+particle.  With one evaluation needed for the original, non-speculative
+particle, we have a total of $8p$ evaluations for every two iterations, where
+$p$ is the number of particles in the speculative swarm.  In order to use the
+same resources for each algorithm, we compare swarms of size $p$ in the
+speculative algorithms with swarms of size $8p$ in standard PSO.
 
-As discussed in \sec{implementation},
-where the Complete topology would normally be used, we use a Random topology in
-our speculative algorithm, as Complete leads to an explosion of speculative
-evaluations.
-For the standard PSO baseline we have included experiments with the
-ring and the random topologies, both with two neighbors, as well as for the complete topology.
-It is important to note however, that in many cases the complete topology is not a
-practical alternative in the context of large
-swarms on large cluster where where the messaging complexity is $O(n^2)$ and can
-overwhelm the system.
+As discussed in \sec{implementation}, where the Complete topology would
+normally be used, we use a Random topology in our speculative algorithm, as
+Complete leads to an explosion of speculative evaluations.  For the standard
+PSO baseline we have included experiments with the Ring and the Random
+topologies, both with two neighbors, as well as for the Complete topology.  It
+is important to note however, that in many cases the Complete topology is not a
+practical alternative in the context of large swarms on large clusters where
+where the messaging complexity is $O(p^2)$ and can overwhelm the system.
 
-We follow the experimental set up proposed by Bratton, et.al.~\cite{bratton-2007-defining-a-standard-for-pso}. 
-All functions have 20 dimensions, and all results shown
-are averages over 20 runs.  We use 240 processors in our experiments, so
-speculative algorithms use 30 particles and standard PSO uses 240 particles.
-For ease of labeling, we call our speculative algorithm SEPSO, the variant
-SEPSO Pick Best, and standard PSO just PSO and identify the topology in a suffix,
-``PSO Ring'' for example.
+We follow the experimental setup used
+in~\cite{bratton-2007-defining-a-standard-for-pso}.  All functions have 20
+dimensions, and all results shown are averages over 20 runs.  For ease of
+labeling, we call our speculative algorithm SEPSO, the variant SEPSO Pick Best,
+and standard PSO just PSO and identify the topology in a suffix, ``PSO Ring''
+for example.
 
-The benchmark function Sphere ($f(\Vec{x}) = \sum_{i=1}^D x_i^2$)
-has no local optima and is most efficiently optimized using
-a small swarm but many iterations.
-In \fig{sphere1}, we can see
-that SEPSO clearly beats PSO with a Random topology or a ring topology.
-
-SEPSO approaches the performance of PSO with the complete topology, even though
-PSO with the complete topology requires $O(n^2)$ communication.
-SEPSO Pick Best handily outperforms all other algorithms
-in this problem.
+The benchmark function Sphere ($f(\Vec{x}) = \sum_{i=1}^D x_i^2$) has no local
+optima and is most efficiently optimized using a small swarm but many
+iterations.  In our experiments with Sphere, we use 240 processors; thus 30
+particles for SEPSO and 240 for PSO.  In \fig{sphere}, we can see that SEPSO
+clearly beats PSO with a Random topology or a Ring topology.  SEPSO approaches
+the performance of PSO with the Complete topology, even though PSO with the
+Complete topology requires $O(p^2)$ communication.  SEPSO Pick Best handily
+outperforms all other algorithms in this problem.
 
@@
   \centering
-  \includegraphics[width=.45\columnwidth]{sphere1}
+  \includegraphics[width=.45\columnwidth]{sphere}
   \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
-  iterations per round, while standard PSO algorithms use 240 processors and do
-  one iteration per round. Note that PSO Complete requires $O(n^2)$ messaging
-  and may not be practical in many cases.}
-  \label{fig:sphere1}
+  evaluations.  We show 10th and 90th percentiles every 100 iterations.  Note
+  that PSO Complete requires $O(p^2)$ messaging and may not be practical in
+  many cases.}
+  \label{fig:sphere}
 \end{figure}
 
-The benchmark function Griewank, defined by the equation $f(\Vec{x}) =
-\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
-\right) + 1$.  It is generally best solved in pso using the Ring topology,
-as Complete is prone to premature convergence
-on a local optimum.  Griewank has a global optimum with a value of 0, and
-sometimes the swarm finds the optimum and sometimes it does not.  Instead of
-showing average function value at each iteration, a more enlightening plot for
-Griewank shows the percent of runs that have found the global optimum by each
-iteration~\cite{bratton-2007-defining-a-standard-for-pso}.
+The benchmark function Griewank is defined by the equation $f(\Vec{x}) =
+\frac{1}{4000}$ $\sum_{i=1}^D x_i^2 - \Pi_{i=1}^D
+\cos\left(\frac{x_i}{\sqrt{i}} \right) + 1$.  It is best solved in PSO using
+the Ring topology, as Complete is prone to premature convergence on local
+optima.  Griewank has a global optimum with a value of 0, and sometimes the
+swarm finds the optimum and sometimes it does not.  Instead of showing average
+function value at each iteration, a more enlightening plot for Griewank shows
+the percent of runs that have found the global optimum by each iteration.
 
-PSO and SEPSO get caught in local minima with small swarm sizes so
-we show results in \fig{griewank1} for swarms of size 100 (SEPSO) and 800 (standard)
-using the Ring topology 
-\fig{griewank2} shows that SEPSO quickly finds
-the global optimum essentially 100\% of the time.
+PSO and SEPSO get caught in local minima with small swarm sizes so we show
+results in \fig{griewank} for swarms of size 100 (SEPSO) and 800 (standard)
+using the Ring topology. \fig{griewank} shows that SEPSO quickly finds the
+global optimum, between two and three times faster than running standard PSO.
+
+In \fig{griewank} we also show results for the Bohachevsky function, defined as
+$f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
+x_{i+1})+.7)$.  Bohachevsky is a unimodal function best optimized with a
+Complete swarm.  It is similar to Griewank in that there is a global optimum
+with a value of 0, and swarms either find the optimum or get stuck.  Both SEPSO
+algorithms find the optimum much faster than PSO Random, though only SEPSO Pick
+Best beats PSO Complete.  Also, while the smaller swarm size of SEPSO gets
+stuck 75\% of the time, when using SEPSO Pick Best with the same swarm size,
+the algorithm finds the optimum every time.
 
@@
   \centering
-  \includegraphics[width=.45\columnwidth]{griewank2}
+  \includegraphics[width=.45\columnwidth]{griewank}
+  \includegraphics[width=.45\columnwidth]{bohachevsky}
   \caption{Function Griewank with a swarm that uses 800 processors per round of
-  evaluations.}
-  \label{fig:griewank2}
+  evaluations, and function Bohachevsky with a swarm that uses 480 processors
+  per round of evaluations.}
+  \label{fig:griewank}
 \end{figure}
 
-As noted in the introduction previous work has shown that the optimization of highly
-deceptive functions like Rastrigin
-($f(\Vec{x}) = \sum_{i=1}^D \left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$)
-benefit greatly from the addition of particles. 
-Smaller swarms get caught
-in local optima.  Because our speculative algorithms have a significantly
-smaller swarm size, they get stuck at higher values while the larger swarms
-performing regular PSO continue to improve the best value found.
-Our experiments with SEPSO on Rastrigin were predicable lack luster,
-yielding a vaerage value of xxx after x evaluations, as compared to y for 
-standard PSO.
+Previous work has shown that the optimization of highly deceptive functions
+like Rastrigin ($f(\Vec{x}) = \sum_{i=1}^D \left(x_i^2 - 10\cos\left(2\pi
+x_i\right) + 10\right)$) benefit greatly from the addition of particles.
+Smaller swarms get caught in local optima, up to swarms of at least 4000
+particles~\cite{mcnabb-2009-large-particle-swarms}.  Because our speculative
+algorithms have a significantly smaller swarm size, they get stuck at higher
+values while the larger swarms performing regular PSO continue to improve the
+best value found.  Our experiments with SEPSO on Rastrigin were predicably lack
+luster, yielding an average value of 31 after 1000 evaluations, as compared to
+10 for standard PSO.
 
@@
 iterations of the algorithm performed, instead of merely adding more particles
-to the swarm.  We have detailed several different implementations of the
-algorithm, depending on the parallel PSO framework being used.  Using our
-modifications, the original PSO algorithm is exactly reproduced two iterations
-at a time.  This technique requires more function evaluations per iteration
-than regular PSO, but for some functions still performs better when run in a
-parallel environment.
-
-We have also described a method for making use of extra speculative evaluations
-that performs incredibly well on some functions.
-
-Though space has limited our presentation of results to a very small sample of
-the interesting and relevant applications of our technique, we believe the
-results shown are sufficient motivation for the further exploration of
-speculative evaluation in PSO.
-
-Our results show that performing speculative evaluation instead of increasing
-swarm size is only beneficial for some functions at the swarm sizes we used.
-However, we would venture that for all functions there is some point at which
-increasing the swarm size provides relatively little benefit.  At that point,
-additional processors should be used to speed up the algorithm instead of add
-additional particles.
+to the swarm.  Using our modifications, the original PSO algorithm is exactly
+reproduced two iterations at a time.  This technique requires more function
+evaluations per iteration than regular PSO, but for some functions still
+performs better when run in a parallel environment.  We have also described a
+method for making use of extra speculative evaluations that performs very well
+on some functions.
 
@@
 processors is to have a small swarm performing speculative evaluation with our
-second method, where all speculative evaluations are used.
+Pick Best method, where all speculative evaluations are used.
 
-There are other functions for which there is a point where the algorithm has
-done ``enough'' exploration and consistently finds the global optimum.  Adding
-additional particles to the swarm is akin to increasing its capacity for
-exploration.  We have seen that for a 20-dimensional Griewank function, 30
-particles is not ``enough'' exploration, but 100 is.  For these functions, one
-should use processors to increase the swarm size until enough exploration has
-been reached, then use additional processors to increase the number of
-iterations performed.
+There are other functions for which it seems there is never enough exploration,
+such as the Rastrigin function.  It has been shown that up to 4000 particles
+there is no point at which ``enough'' exploration has been
+done~\cite{mcnabb-2009-large-particle-swarms}.  With such functions, the
+smaller swarm size required by speculative evaluation is not able to produce
+enough exploration to perform better than standard PSO.
 
-There are still other functions for which it seems there is never enough
-exploration, such as the Rastrigin function.  It has been shown that up to 4000
-particles there is no point at which ``enough'' exploration has been
-done~\cite{mcnabb-2009-large-particle-swarms}.  For those functions the correct
-use of processors cannot be easily prescribed and depends largely on the
-desired outcome and how many processors are available.  It is likely that the
-best thing to do with those functions is to use all processors to add particles
-to the swarm and do no speculative evaluation.
+Griewank and Bohachevsky are functions between Sphere and Rastrigin.  They are
+deceptive and prone to premature convergence, but by adding particles to the
+swarm a point is reached where ``enough'' exploration is done, and the algorithm finds the optimum essentially all of the time.
+For such functions, the best approach seems to be to increase the swarm size until ``enough'' exploration is reached, then use extra processors to perform speculative evaluation and increase the number of iterations performed.
+Sphere and Rastrigin can be thought of as special cases fo these types of functions; Sphere simply needs a very small swarm size to produce ``enough'' exploration, and Rastrigin requires a very large swarm.
+We expect that for all functions there is a swarm size for which additional particles are less useful than additional iterations.
+
+Large parallel clusters are often required to successfully optimize practical modern problems.
+To properly use PSO with such clusters, a balance needs to be made between using processors to increase the swarm size and using them to increase the speed of the algorithm.
+This work is a first step in that direction.
 

### 2010-04-21 honors_thesis/journal_paper/paper.tex
Started moving the results to be in one group at the end

@@
 particle swarm optimization algorithm.  \sec{sepso} describes how speculative
-evaluation can be done in parallel PSO to perform two iterations at once, then
-presents a comparison of the speculative algorithm with the original PSO.  In
+evaluation can be done in parallel PSO to perform two iterations at once.  In
 \sec{relax}, we discuss various methods of improving the performance of
 speculative evaluation in PSO, all of which break the requirement of strictly
-reproducing the behavior of the original algorithm.  In \sec{conclusion} and
-\sec{future} we conclude and discuss future work.
+reproducing the behavior of the original algorithm.  \sec{results} presents the
+results of using our methods in comparison with standard PSO.  In
+\sec{conclusion} and \sec{future} we conclude and discuss future work.
 
@@
 
-\subsection{Comparison with Standard PSO}
-\label{sec:results}
-
-We ran experiments to compare our speculative PSO algorithm to the standard PSO
-algorithm.  At each iteration of the algorithms, we use one processor to
-perform one function evaluation for one particle, be it speculative or
-non-speculative.  We acknowledge that for benchmark functions with very fast
-evaluations this may not be the best use of parallelism in PSO.  But for many
-real-world applications, the objective function takes on the order of minutes
-or more to evaluate; in such cases our choice of framework is reasonable.
-
-The speculative algorithm actually performs two iterations of PSO at each
-``iteration,'' so we instead call each ``iteration'' a ``round of
-evaluations''.
-
-To use the same number of processors (and hence evaluations) for each
-algorithm, speculative evaluation must use a smaller swarm size than standard
-PSO.  For each of the topologies we used, a particle has three neighbors
-including itself.  As shown in Table~\ref{tab:evals}, this results in $7$
-speculative evaluations per particle.  With one evaluation needed for the
-original, non-speculative particle, we have $8p$ evaluations for every two
-iterations, where $p$ is the number of particles in the speculative swarm.
-The extra evaluations done in speculative evaluation would instead be particles
-in regular PSO, so we compare swarms of size $p$ in speculative evaluation
-with swarms of size $8p$ in regular PSO.
-
-Where the Complete topology would normally be used, we use a Random topology in
-our speculative algorithm, as Complete leads to an explosion in the number of
-speculative evaluations.  If speculative evaluation were not being performed,
-it is possible that the Complete topology would be used.  However, the Complete
-topology also requires a very large amount of interprocessor communication in
-distributed PSO, so it is still quite possible that Random would be used even
-with standard PSO.  But, to be fair in our comparisons, we compare to the
-standard PSO algorithm using both the Random topology with two neighbors and
-the Complete topology.  All functions have 20 dimensions, and all results shown
-are averages over 20 runs.  We use 240 processors in our experiments (except
-where otherwise noted), so our speculative algorithm uses 30 particles and
-standard PSO uses 240 particles.  For ease of labeling, we call our speculative
-algorithm SEPSO and standard PSO just PSO, with the topology following.
-
-First we look at the function Sphere, the simplest of common benchmark
-functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
-has no local optima and is best optimized in terms of function evaluations with
-a small swarm using a Complete topology.  Accordingly, we show results for a
-Complete swarm in \fig{basic-sphere}.  We can see that when both SEPSO and PSO
-are using a Random topology, SEPSO clearly outperforms PSO.  However, the
-comparison between SEPSO and PSO Complete is not as encouraging---PSO Complete
-performs slightly better.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
-  two iterations per round, while standard PSO algorithms use 240 processors
-  and do one iteration per round.}
-  \label{fig:basic-sphere}
-\end{figure}
-
-Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
-\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
-\right) + 1$.  It is generally recommended to use the Ring topology when
-optimizing the Griewank function, as Complete is prone to premature convergence
-on a local optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank
-has a global optimum with a value of 0, and sometimes the swarm finds the
-optimum and sometimes it doesn't.  Instead of showing average function value at
-each iteration, a more informative plot for Griewank shows the percent of runs
-that have found the global optimum by each iteration.  We show results in
-\fig{basic-griewank2} for swarms of size 30 and 240 using the Ring topology.
-One can see in the figure that the SEPSO algorithm finds the optimum faster
-than the original PSO.  However, because the swarm size is so small, SEPSO gets
-stuck a little less than half of the time.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank2}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.  Instead of showing average function value, we show the
-  percentage of runs that have found the global optimum by each iteration.}
-  \label{fig:basic-griewank2}
-\end{figure}
-
-Because the speculative algorithm got stuck with such a small swarm size, we
-ran a set of experiments on Griewank with 800 processors, giving swarms of size
-100 and 800 instead of 30 and 240.  \fig{basic-griewank3} shows the results.
-\fig{basic-griewank3} is very similar to \fig{basic-griewank2}, except that
-speculative evaluation finds the global optimum essentially 100\% of the time.
-Thus with a few more processors, our algorithm finds the optimum twice as fast
-as the original PSO without a significant loss of accuracy.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank3}
-  \caption{Function Griewank with a swarm that uses 800 processors per round of
-  evaluations.}
-  \label{fig:basic-griewank3}
-\end{figure}
-
-Lastly we look at a function that does not show any kind of improvement with
-this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
-\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
-with Rastrigin, the more particles there are in the swarm, the lower function
-value it finds, up to at least 4000
-particles~\cite{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
-in local optima.  Because our speculative algorithm has a significantly smaller
-swarm size, it gets stuck at higher values while the larger swarms performing
-regular PSO continue to improve the best value found.  \fig{rastrigin} shows
-the results graphically.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{rastrigin}
-  \caption{Function Rastrigin with a swarm that uses 240 processors per round
-  of evaluations.}
-  \label{fig:rastrigin}
-\end{figure}
-
 \section{Relaxing the Requirements}
@@
 While the idea of speculative evaluation in particle swarm optimization is
-intriguing, the results of the previous section show that in many cases it
-requires too many extra evaluations to produce competitive results.  On most
-functions, the larger swarm size that can be used in regular PSO leads to
-better performance than is found by trying to do speculative evaluation.
-However, if we keep the idea of speculative evaluation while relaxing the
-requirement of exactly reproducing the behavior of the original PSO algorithm,
-we see some impressive results.
+intriguing, the results in \sec{results} show that in many cases it requires
+too many extra evaluations to produce competitive results.  On most functions,
+the larger swarm size that can be used in regular PSO leads to better
+performance than is found by trying to do speculative evaluation.  However, if
+we keep the idea of speculative evaluation while relaxing the requirement of
+exactly reproducing the behavior of the original PSO algorithm, we see some
+impressive results.
 
@@
 functions would probably be hurt.
-
-We show in \fig{sphere-pickbest} the results of using this modification on the
-function Sphere.  For ease of labeling, we call the modified algorithm Pick
-Best.  As can be seen, where SEPSO did not manage to perform as well as PSO
-Complete, Pick Best handily outperforms it with the same number of processors.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere2}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing Pick Best with previous results.}
-  \label{fig:sphere-pickbest}
-\end{figure}
-
-Sphere is a function for which very little exploration needs to be done, so
-always keeping the best speculative position found makes intuitive sense.  A
-less intuitive result occurs when using Pick Best with the function Griewank.
-Griewank is a highly deceptive function prone to premature convergence.  Our
-intuition on the behavior of picking the best speculative child is that it
-would tend to prematurely converge more often than the original PSO.  However,
-as seen in \fig{griewank-pickbest}, Pick Best increases the chance of finding
-the optimum while also reducing the number of iterations required to converge.
-Perhaps Pick Best encourages local exploitation, while the sparse communication
-of the Ring topology still allows for global exploration.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank4}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing Pick Best with previous results.}
-  \label{fig:griewank-pickbest}
-\end{figure}
 
@@
 
-There are a large number of ways to prune speculative evaluations.  We
-experimented with several, but present results only for one possible pruning.
-The pruning we present uses only the two most probable branches in the Random
-topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
-the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
-also convenient in that no messages are needed from neighbors in order to
-produce the positions of the speculative particles---in distributed frameworks
-using several rounds of communication (see \sec{distributed}), one of the
-rounds of communication can be dropped entirely.
-
-Because we are now only doing two speculative evaluations per particle (along
-with evaluating the original particle), with 240 processors we can have a swarm
-of size 80, instead of the 30 particles in previous experiments.
-
-In \fig{sphere-pruned}, we show the results of this simple pruning on the
-function Sphere.  As can be seen, the larger number of particles that pruning
-allows leads to better results when picking the best speculative child.  When
-using Social Promotion no real improvement is seen over the original PSO
-algorithm.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere3}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing pruning with previous results.}
-  \label{fig:sphere-pruned}
-\end{figure}
-
-When looking at the function Griewank, presented in \fig{griewank-pruned}, we
-see different results.  Our intuition about picking the best child is more
-correct in this case; while adding 50 more particles to the swarm, Pick Best
-with pruning still gets stuck more often than the original Pick Best.  However,
-Social Promotion does well in this case, finding the optimum every time and
-faster than the original PSO.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank5}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing pruning with previous results.}
-  \label{fig:griewank-pruned}
-\end{figure}
-
 \subsection{More than one iteration ahead}
@@
 to compare the different branching possibilities on various functions.
+
+\section{Results}
+\label{sec:results}
+
+We present results for five common benchmark problems: Rastrigin, Sphere,
+Schwefel 2.21, Griewank, and Bohachevsky.  Rastrigin is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
+10\right), z=x-c\]
+initialized in $[-5.12,5.12]^D$.
+Sphere is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]
+initialized in $[-50,50]^D$.
+Schwefel 2.21 is defined as:
+\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]
+initialized in $[-500,500]^D$.
+Griewank is defined as:
+\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
+\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]
+initialized in $[-600,600]^D$
+And Bohachevsky is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
+x_{i+1})+.7)\]
+initialized in $[-15,15]^D$.
+
+The $c$ in the function definitions is a shifted center, in order to avoid
+origin-seeking bias in the PSO algorithm and its
+variants~\cite{monson-2005-origin-seeking-bias}.  We move the center of the
+shifted functions (all except Bohachevsky) to be halfway between the center and
+the boundary of the initialization region.  For example, if the initialization
+region is $[-50,50]^D$, the center is $(25)^D$.
+
+Lastly we look at a function that does not show any kind of improvement with
+this method.  The function Rastrigin is defined as 
+It has been shown that
+with Rastrigin, the more particles there are in the swarm, the lower function
+value it finds, up to at least 4000
+particles~\cite{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
+in local optima.  Because our speculative algorithm has a significantly smaller
+swarm size, it gets stuck at higher values while the larger swarms performing
+regular PSO continue to improve the best value found.  \fig{rastrigin} shows
+the results graphically.
+\subsection{Comparison with Standard PSO}
+
+We ran experiments to compare our speculative PSO algorithm to the standard PSO
+algorithm.  At each iteration of the algorithms, we use one processor to
+perform one function evaluation for one particle, be it speculative or
+non-speculative.  We acknowledge that for benchmark functions with very fast
+evaluations this may not be the best use of parallelism in PSO.  But for many
+real-world applications, the objective function takes on the order of minutes
+or more to evaluate; in such cases our choice of framework is reasonable.
+
+The speculative algorithm actually performs two iterations of PSO at each
+``iteration,'' so we instead call each ``iteration'' a ``round of
+evaluations''.
+
+To use the same number of processors (and hence evaluations) for each
+algorithm, speculative evaluation must use a smaller swarm size than standard
+PSO.  For each of the topologies we used, a particle has three neighbors
+including itself.  As shown in Table~\ref{tab:evals}, this results in $7$
+speculative evaluations per particle.  With one evaluation needed for the
+original, non-speculative particle, we have $8p$ evaluations for every two
+iterations, where $p$ is the number of particles in the speculative swarm.
+The extra evaluations done in speculative evaluation would instead be particles
+in regular PSO, so we compare swarms of size $p$ in speculative evaluation
+with swarms of size $8p$ in regular PSO.
+
+Where the Complete topology would normally be used, we use a Random topology in
+our speculative algorithm, as Complete leads to an explosion in the number of
+speculative evaluations.  If speculative evaluation were not being performed,
+it is possible that the Complete topology would be used.  However, the Complete
+topology also requires a very large amount of interprocessor communication in
+distributed PSO, so it is still quite possible that Random would be used even
+with standard PSO.  But, to be fair in our comparisons, we compare to the
+standard PSO algorithm using both the Random topology with two neighbors and
+the Complete topology.  All functions have 20 dimensions, and all results shown
+are averages over 20 runs.  We use 240 processors in our experiments (except
+where otherwise noted), so our speculative algorithm uses 30 particles and
+standard PSO uses 240 particles.  For ease of labeling, we call our speculative
+algorithm SEPSO and standard PSO just PSO, with the topology following.
+
+First we look at the function Sphere, the simplest of common benchmark
+functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
+has no local optima and is best optimized in terms of function evaluations with
+a small swarm using a Complete topology.  Accordingly, we show results for a
+Complete swarm in \fig{basic-sphere}.  We can see that when both SEPSO and PSO
+are using a Random topology, SEPSO clearly outperforms PSO.  However, the
+comparison between SEPSO and PSO Complete is not as encouraging---PSO Complete
+performs slightly better.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
+  two iterations per round, while standard PSO algorithms use 240 processors
+  and do one iteration per round.}
+  \label{fig:basic-sphere}
+\end{figure}
+
+Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
+\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
+\right) + 1$.  It is generally recommended to use the Ring topology when
+optimizing the Griewank function, as Complete is prone to premature convergence
+on a local optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank
+has a global optimum with a value of 0, and sometimes the swarm finds the
+optimum and sometimes it doesn't.  Instead of showing average function value at
+each iteration, a more informative plot for Griewank shows the percent of runs
+that have found the global optimum by each iteration.  We show results in
+\fig{basic-griewank2} for swarms of size 30 and 240 using the Ring topology.
+One can see in the figure that the SEPSO algorithm finds the optimum faster
+than the original PSO.  However, because the swarm size is so small, SEPSO gets
+stuck a little less than half of the time.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank2}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations.  Instead of showing average function value, we show the
+  percentage of runs that have found the global optimum by each iteration.}
+  \label{fig:basic-griewank2}
+\end{figure}
+
+Because the speculative algorithm got stuck with such a small swarm size, we
+ran a set of experiments on Griewank with 800 processors, giving swarms of size
+100 and 800 instead of 30 and 240.  \fig{basic-griewank3} shows the results.
+\fig{basic-griewank3} is very similar to \fig{basic-griewank2}, except that
+speculative evaluation finds the global optimum essentially 100\% of the time.
+Thus with a few more processors, our algorithm finds the optimum twice as fast
+as the original PSO without a significant loss of accuracy.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank3}
+  \caption{Function Griewank with a swarm that uses 800 processors per round of
+  evaluations.}
+  \label{fig:basic-griewank3}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{rastrigin}
+  \caption{Function Rastrigin with a swarm that uses 240 processors per round
+  of evaluations.}
+  \label{fig:rastrigin}
+\end{figure}
+
+\subsection{Pick Best}
+
+We show in \fig{sphere-pickbest} the results of using this modification on the
+function Sphere.  For ease of labeling, we call the modified algorithm Pick
+Best.  As can be seen, where SEPSO did not manage to perform as well as PSO
+Complete, Pick Best handily outperforms it with the same number of processors.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere2}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing Pick Best with previous results.}
+  \label{fig:sphere-pickbest}
+\end{figure}
+
+Sphere is a function for which very little exploration needs to be done, so
+always keeping the best speculative position found makes intuitive sense.  A
+less intuitive result occurs when using Pick Best with the function Griewank.
+Griewank is a highly deceptive function prone to premature convergence.  Our
+intuition on the behavior of picking the best speculative child is that it
+would tend to prematurely converge more often than the original PSO.  However,
+as seen in \fig{griewank-pickbest}, Pick Best increases the chance of finding
+the optimum while also reducing the number of iterations required to converge.
+Perhaps Pick Best encourages local exploitation, while the sparse communication
+of the Ring topology still allows for global exploration.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank4}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations, comparing Pick Best with previous results.}
+  \label{fig:griewank-pickbest}
+\end{figure}
+
+\subsection{Pruning}
+
+There are a large number of ways to prune speculative evaluations.  We
+experimented with several, but present results only for one possible pruning.
+The pruning we present uses only the two most probable branches in the Random
+topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
+the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
+also convenient in that no messages are needed from neighbors in order to
+produce the positions of the speculative particles---in distributed frameworks
+using several rounds of communication (see \sec{distributed}), one of the
+rounds of communication can be dropped entirely.
+
+Because we are now only doing two speculative evaluations per particle (along
+with evaluating the original particle), with 240 processors we can have a swarm
+of size 80, instead of the 30 particles in previous experiments.
+
+In \fig{sphere-pruned}, we show the results of this simple pruning on the
+function Sphere.  As can be seen, the larger number of particles that pruning
+allows leads to better results when picking the best speculative child.  When
+using Social Promotion no real improvement is seen over the original PSO
+algorithm.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere3}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing pruning with previous results.}
+  \label{fig:sphere-pruned}
+\end{figure}
+
+When looking at the function Griewank, presented in \fig{griewank-pruned}, we
+see different results.  Our intuition about picking the best child is more
+correct in this case; while adding 50 more particles to the swarm, Pick Best
+with pruning still gets stuck more often than the original Pick Best.  However,
+Social Promotion does well in this case, finding the optimum every time and
+faster than the original PSO.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank5}
+  \caption{Function Griewank with a swarm that uses 240 processors per round of
+  evaluations, comparing pruning with previous results.}
+  \label{fig:griewank-pruned}
+\end{figure}
+
+\subsection{Many Iters}
 

### 2010-04-21 honors_thesis/journal_paper/paper.tex
Fixed a few problems in the equations

@@
 We present results for five common benchmark problems: Rastrigin, Sphere,
-Schwefel 2.21, Griewank, and Bohachevsky.  Rastrigin is defined as:
+Schwefel 2.21, Griewank, and Bohachevsky.
+
+Rastrigin is initialized in $[-5.12,5.12]^D$ and defined as:
 \[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
 10\right), z=x-c\]
-initialized in $[-5.12,5.12]^D$.
-Sphere is defined as:
+
+Sphere is initialized in $[-50,50]^D$ and defined as:
 \[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]
-initialized in $[-50,50]^D$.
-Schwefel 2.21 is defined as:
+
+Schwefel 2.21 is initialized in $[-500,500]^D$ and defined as:
 \[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]
-initialized in $[-500,500]^D$.
-Griewank is defined as:
+
+Griewank is initialized in $[-600,600]^D$ and defined as:
 \[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
 \cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]
-initialized in $[-600,600]^D$
-And Bohachevsky is defined as:
+
+And Bohachevsky is initialized in $[-15,15]^D$ and defined as:
 \[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
 x_{i+1})+.7)\]
-initialized in $[-15,15]^D$.
 

### 2010-04-21 honors_thesis/journal_paper/paper.tex
Updating the results section.  Getting close to done

@@
 
-\section{Results}
-\label{sec:results}
-
-We present results for five common benchmark problems: Rastrigin, Sphere,
-Schwefel 2.21, Griewank, and Bohachevsky.
-
-Rastrigin is initialized in $[-5.12,5.12]^D$ and defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
-10\right), z=x-c\]
-
-Sphere is initialized in $[-50,50]^D$ and defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]
-
-Schwefel 2.21 is initialized in $[-500,500]^D$ and defined as:
-\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]
-
-Griewank is initialized in $[-600,600]^D$ and defined as:
-\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
-\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]
-
-And Bohachevsky is initialized in $[-15,15]^D$ and defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
-x_{i+1})+.7)\]
-
-The $c$ in the function definitions is a shifted center, in order to avoid
-origin-seeking bias in the PSO algorithm and its
-variants~\cite{monson-2005-origin-seeking-bias}.  We move the center of the
-shifted functions (all except Bohachevsky) to be halfway between the center and
-the boundary of the initialization region.  For example, if the initialization
-region is $[-50,50]^D$, the center is $(25)^D$.
-
-Lastly we look at a function that does not show any kind of improvement with
-this method.  The function Rastrigin is defined as 
-It has been shown that
-with Rastrigin, the more particles there are in the swarm, the lower function
-value it finds, up to at least 4000
-particles~\cite{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
-in local optima.  Because our speculative algorithm has a significantly smaller
-swarm size, it gets stuck at higher values while the larger swarms performing
-regular PSO continue to improve the best value found.  \fig{rastrigin} shows
-the results graphically.
-\subsection{Comparison with Standard PSO}
+\section{Experimental Setup}
+\label{sec:setup}
 
@@
 ``iteration,'' so we instead call each ``iteration'' a ``round of
-evaluations''.
+evaluations.''
 
@@
 standard PSO algorithm using both the Random topology with two neighbors and
-the Complete topology.  All functions have 20 dimensions, and all results shown
-are averages over 20 runs.  We use 240 processors in our experiments (except
-where otherwise noted), so our speculative algorithm uses 30 particles and
-standard PSO uses 240 particles.  For ease of labeling, we call our speculative
-algorithm SEPSO and standard PSO just PSO, with the topology following.
-
-First we look at the function Sphere, the simplest of common benchmark
-functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
+the Complete topology.
+
+We experimented with five common benchmark problems defined
+in~\cite{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
+and Bohachevsky.
+
+Rastrigin is initialized in $[-5.12,5.12]^D$ and is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
+10\right), z=x-c\]
+
+Sphere is initialized in $[-50,50]^D$ and is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]
+
+Schwefel 2.21 is initialized in $[-500,500]^D$ and is defined as:
+\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]
+
+Griewank is initialized in $[-600,600]^D$ and is defined as:
+\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
+\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]
+
+And Bohachevsky is initialized in $[-15,15]^D$ and is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
+x_{i+1})+.7)\]
+
+The $c$ in the function definitions is a shifted center, in order to avoid
+origin-seeking bias in the PSO algorithm and its
+variants~\cite{monson-2005-origin-seeking-bias}.  We move the center of the
+shifted functions (all except Bohachevsky) to be halfway between the center and
+the boundary of the initialization region.  For example, if the initialization
+region is $[-50,50]^D$, the center is $(25)^D$.  For all of these functions we
+used 20 dimensions.
+
+There are a large number of ways to prune speculative evaluations.  We
+experimented with several, but present results for only one possible pruning.
+The pruning we present uses only the two most probable branches in the Random
+topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
+the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
+also convenient in that no messages are needed from neighbors in order to
+produce the positions of the speculative particles---in distributed frameworks
+using several rounds of communication (see \sec{distributed}), one of the
+rounds of communication can be dropped entirely.
+
+Because pruning only requires two speculative evaluations per particle (along
+with evaluating the original particle), we can use swarms of size $3p$ to
+compare to a swarm of size $p$ with standard PSO, instead of $8p$ with other
+techniques.
+
+There are also many ways to speculate several iterations ahead, and we again
+only show results for one of them.  The combination of branches we tried uses
+seven speculative evaluations per particle, matching the swarm size of the
+original speculative algorithm.  The seven evaluations we used corresponded to
+several iterations of branches \casepn\ and \casexn.  Two of the evaluations
+were just one iteration ahead, four were two iterations ahead, and one was
+three iterations ahead.  The evaluations that were one iteration ahead were
+branches \casepn\ and \casexn; those that went two iterations ahead were formed
+by taking either branch \casepn\ or \casexn and then branch \casepn\ or
+\casexn; and the evaluation going three iterations ahead followed branch
+\casepn\ on all three iterations.
+
+\section{Results}
+\label{sec:results}
+
+\subsection{Rastrigin}
+
+First we look at Rastrigin.  Rastrigin is a multi-modal function that is best
+optimized with a large, Complete swarm.  It has been shown that with Rastrigin,
+the more particles there are in the swarm, the lower function value it finds,
+up to at least 4000 particles~\cite{mcnabb-2009-large-particle-swarms}.
+Smaller swarms get caught in local optima.  Because our speculative algorithms
+require significantly smaller swarm sizes, we would expect to not perform very
+well on functions such as Rastrigin.  Our experiments show our intuition to be
+correct.  In this experiment we used 240 processors, so SEPSO and Pick Best
+each had 30 particles, and the standard PSO algorithms had 240 particles.  As
+expected, SEPSO and Pick Best converge quickly to worse local optima than
+standard PSO does.  \fig{rastrigin} shows the results graphically.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{rastrigin}
+  \caption{Function Rastrigin with a swarm that uses 240 processors per round
+  of evaluations.}
+  \label{fig:rastrigin}
+\end{figure}
+
+\subsection{Sphere}
+
+Next we go to Sphere, the simplest of common benchmark functions.  The function
 has no local optima and is best optimized in terms of function evaluations with
-a small swarm using a Complete topology.  Accordingly, we show results for a
-Complete swarm in \fig{basic-sphere}.  We can see that when both SEPSO and PSO
-are using a Random topology, SEPSO clearly outperforms PSO.  However, the
-comparison between SEPSO and PSO Complete is not as encouraging---PSO Complete
-performs slightly better.
+a small swarm using a Complete topology.  We expect our methods to be perfect
+for such functions, and our results show this intuition again to be correct.
+First we compare the basic SEPSO algorithm to standard PSO.  In
+\fig{basic-sphere} we can see that when both SEPSO and PSO are using a Random
+topology, SEPSO clearly outperforms PSO.  However, the comparison between SEPSO
+and PSO Complete is not as encouraging---PSO Complete performs slightly better.
 
@@
   \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
-  two iterations per round, while standard PSO algorithms use 240 processors
-  and do one iteration per round.}
+  evaluations.}
   \label{fig:basic-sphere}
 \end{figure}
+
+When we introduce Pick Best, we see more impressive results.  As can be seen in
+\fig{sphere-pickbest}, where SEPSO did not manage to perform as well as PSO
+Complete, Pick Best handily outperforms it with the same number of processors.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere2}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing Pick Best with previous results.}
+  \label{fig:sphere-pickbest}
+\end{figure}
+
+Still more impressive with Sphere are the results of pruning and speculating
+several iterations ahead.  Pick Best Pruned outperforms Pick Best, and Many
+Iters by far outperforms every other algorithm on this function.  Many Iters on
+average finds values that are more than 40 orders of magnitude better than
+standard PSO using the same number of processors.  This is shown in
+\fig{sphere-manyiters}.  Social Promotion Pruned performed comparably with
+SEPSO, and we do not clutter the graph with a line for it.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sphere3}
+  \caption{Function Sphere with a swarm that uses 240 processors per round of
+  evaluations, comparing speculating several iterations ahead with previous
+  results.}
+  \label{fig:sphere-manyiters}
+\end{figure}
+
+\subsection{Schwefel 2.21}
+
+\subsection{Griewank}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{rastrigin}
-  \caption{Function Rastrigin with a swarm that uses 240 processors per round
-  of evaluations.}
-  \label{fig:rastrigin}
-\end{figure}
-
-\subsection{Pick Best}
-
-We show in \fig{sphere-pickbest} the results of using this modification on the
-function Sphere.  For ease of labeling, we call the modified algorithm Pick
-Best.  As can be seen, where SEPSO did not manage to perform as well as PSO
-Complete, Pick Best handily outperforms it with the same number of processors.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere2}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing Pick Best with previous results.}
-  \label{fig:sphere-pickbest}
-\end{figure}
-
 Sphere is a function for which very little exploration needs to be done, so
@@
 
-\subsection{Pruning}
-
-There are a large number of ways to prune speculative evaluations.  We
-experimented with several, but present results only for one possible pruning.
-The pruning we present uses only the two most probable branches in the Random
-topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
-the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
-also convenient in that no messages are needed from neighbors in order to
-produce the positions of the speculative particles---in distributed frameworks
-using several rounds of communication (see \sec{distributed}), one of the
-rounds of communication can be dropped entirely.
-
-Because we are now only doing two speculative evaluations per particle (along
-with evaluating the original particle), with 240 processors we can have a swarm
-of size 80, instead of the 30 particles in previous experiments.
-
-In \fig{sphere-pruned}, we show the results of this simple pruning on the
-function Sphere.  As can be seen, the larger number of particles that pruning
-allows leads to better results when picking the best speculative child.  When
-using Social Promotion no real improvement is seen over the original PSO
-algorithm.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere3}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing pruning with previous results.}
-  \label{fig:sphere-pruned}
-\end{figure}
-
 When looking at the function Griewank, presented in \fig{griewank-pruned}, we
@@
   \label{fig:griewank-pruned}
-\end{figure}
-
-\subsection{Many Iters}
-
-The combination of branches we tried uses seven speculative evaluations per
-particle, matching the swarm size of the original speculative algorithm.  The
-seven evaluations we used corresponded to several iterations of branches
-\casepn\ and \casexn.  Two of the evaluations were just one iteration ahead,
-four were two iterations ahead, and one was three iterations ahead.  The
-evaluations that were one iteration ahead were branches \casepn\ and \casexn;
-those that went two iterations ahead were formed by taking either branch
-\casepn\ or \casexn and then branch \casepn\ or \casexn; and the evaluation
-going three iterations ahead followed branch \casepn\ on all three iterations.
-
-The results when optimizing the function Sphere, presented in
-\fig{sphere-manyiters}, are rather astounding.  When 240 processors are
-available for use, performing speculative evaluation several iterations ahead
-finds values an average of 46 orders of magnitude better than using the
-processors to perform the standard PSO algorithm with 240 particles.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere4}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing speculating several iterations ahead with previous
-  results.}
-  \label{fig:sphere-manyiters}
 \end{figure}
@@
 \end{figure}
+
+\subsection{Bohachevksy}
 

### 2010-05-06 honors_thesis/journal_paper/paper.tex
Finished a draft of the final version of the journal paper.  One more pass...

@@
 twice as many iterations outperforms a swarm of 800 particles.  We show in the
-rest of this paper that in many, though not all, instances, a smaller swarm
+\sec{results} that in many, though not all, instances, a smaller swarm
 performing more iterations does in fact outperform a larger swarm.
@@
 
+Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited 
+more by larger swarms than Sphere is.  Thus our speculative algorithms often
+fail to perform as well as standard PSO because of their smaller swarm size.
+In comparing SEPSO to standard PSO in \fig{schwefel1}, we can see that SEPSO
+fails to perform any better than PSO with a Random topology, and it is by far
+outperformed by PSO with a Complete topology.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{schwefel1}
+  \caption{Function Schwefel with a swarm that uses 240 processors per round of
+  evaluations.}
+  \label{fig:schwefel1}
+\end{figure}
+
+As can be seen in \fig{schwefel2}, almost all of the speculative algorithms
+perform worse than PSO Complete, with Many Iterations performing comparably to
+PSO.  The only algorithm that performed better than PSO was Pick Best Pruned,
+likely because of the larger swarm size available to it.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{schwefel2}
+  \caption{Function Schwefel with a swarm that uses 240 processors per round of
+  evaluations, comparing all speculative algorithms.}
+  \label{fig:schwefel2}
+\end{figure}
+
 \subsection{Griewank}
 
-Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
-\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
-\right) + 1$.  It is generally recommended to use the Ring topology when
-optimizing the Griewank function, as Complete is prone to premature convergence
-on a local optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank
-has a global optimum with a value of 0, and sometimes the swarm finds the
-optimum and sometimes it doesn't.  Instead of showing average function value at
-each iteration, a more informative plot for Griewank shows the percent of runs
-that have found the global optimum by each iteration.  We show results in
-\fig{basic-griewank2} for swarms of size 30 and 240 using the Ring topology.
+Next we look at the function Griewank.  It is generally recommended to use the
+Ring topology when optimizing the Griewank function, as Complete is prone to
+premature convergence on a local
+optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank has a global
+optimum with a value of 0, and sometimes the swarm finds the optimum and
+sometimes it doesn't.  Instead of showing average function value at each
+iteration, a more informative plot for Griewank shows the percent of runs that
+have found the global optimum by each iteration.  We ran 50 trials of each
+experiment with Griewank, so that the curves are more smooth.  We show results
+in \fig{basic-griewank1} for swarms of size 30 and 240 using the Ring topology.
 One can see in the figure that the SEPSO algorithm finds the optimum faster
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank2}
+  \includegraphics[width=.8\columnwidth]{griewank1}
   \caption{Function Griewank with a swarm that uses 240 processors per round of
@@
   percentage of runs that have found the global optimum by each iteration.}
-  \label{fig:basic-griewank2}
+  \label{fig:basic-griewank1}
 \end{figure}
@@
 ran a set of experiments on Griewank with 800 processors, giving swarms of size
-100 and 800 instead of 30 and 240.  \fig{basic-griewank3} shows the results.
-\fig{basic-griewank3} is very similar to \fig{basic-griewank2}, except that
+100 and 800 instead of 30 and 240.  \fig{basic-griewank2} shows the results.
+\fig{basic-griewank2} is very similar to \fig{basic-griewank1}, except that
 speculative evaluation finds the global optimum essentially 100\% of the time.
-Thus with a few more processors, our algorithm finds the optimum twice as fast
-as the original PSO without a significant loss of accuracy.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank3}
+Thus with a few more processors, our algorithm finds the optimum on average
+twice as fast as the original PSO without a significant loss of accuracy.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank2}
   \caption{Function Griewank with a swarm that uses 800 processors per round of
   evaluations.}
-  \label{fig:basic-griewank3}
-\end{figure}
-
-Sphere is a function for which very little exploration needs to be done, so
-always keeping the best speculative position found makes intuitive sense.  A
-less intuitive result occurs when using Pick Best with the function Griewank.
-Griewank is a highly deceptive function prone to premature convergence.  Our
-intuition on the behavior of picking the best speculative child is that it
-would tend to prematurely converge more often than the original PSO.  However,
-as seen in \fig{griewank-pickbest}, Pick Best increases the chance of finding
-the optimum while also reducing the number of iterations required to converge.
-Perhaps Pick Best encourages local exploitation, while the sparse communication
-of the Ring topology still allows for global exploration.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank4}
+  \label{fig:basic-griewank2}
+\end{figure}
+
+In order to show the premature convergence properties of the various algorithms
+we have presented, we will show results first with swarms using 240 processors,
+where these are more visible.  Then we will present an overall result for 800
+processors, where using our techniques is generally more successful.
+
+When we look at the performance of our Pick Best algorithm, a less intuitive
+result occurs with Griewank than with Sphere.  Because Sphere needs very little
+exploration, the seemingly greedy approach of Pick Best does very well.  But it
+also greatly improves performance on Griewank.  In \fig{griewank-pickbest} we
+see that Pick Best improves accuracy over SEPSO by 20\%, while at the same time
+finding the optimum over 100 rounds of evaluation sooner on average.  It seems
+that while Pick Best is locally greedy, there is enough exploration in the
+seven speculative evaluations to overcome the inherent greediness of the
+algorithm.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank3}
   \caption{Function Griewank with a swarm that uses 240 processors per round of
@@
 
-When looking at the function Griewank, presented in \fig{griewank-pruned}, we
-see different results.  Our intuition about picking the best child is more
-correct in this case; while adding 50 more particles to the swarm, Pick Best
-with pruning still gets stuck more often than the original Pick Best.  However,
-Social Promotion does well in this case, finding the optimum every time and
-faster than the original PSO.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank5}
+When we introduce pruning, our intuition about Pick Best turns out to be more
+correct.  While adding 50 more particles to the swarm, Pick Best with pruning
+still gets stuck just as often as the original Pick Best.  However, Social
+Promotion does well with pruning; it increases the success rate to close to
+100\%, while still finding the optimum on average much faster than the original
+PSO.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank4}
   \caption{Function Griewank with a swarm that uses 240 processors per round of
@@
 
-With the function Griewank, the premature convergence problems inherent in
-picking the best child are exacerbated when speculating several iterations
-ahead.  When the algorithm finds the optimum, it finds it quicker than any
-other method we tried.  However, it also gets stuck and fails to find the
-optimum more than any other method.  In order for this method to be
-consistently useful on functions like Griewank, it seems that more processors
-are required.  The results are shown in \fig{griewank-manyiters}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank6}
+With Griewank, the premature convergence problems inherent in picking the best
+child are exacerbated when speculating several iterations ahead.  When Many
+Iterations finds the optimum, it finds it quicker than any other method we
+tried, on average four times faster than standard PSO.  However, it also gets
+stuck and fails to find the optimum more than any other method.  The results
+are shown in \fig{griewank-manyiters}. (TODO: Rerun Many Iterations with the
+right swarm size)
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank5}
   \caption{Function Griewank with a swarm that uses 240 processors per round of
@@
 
-\subsection{Bohachevksy}
+Now that we have seen the effect that the various algorithms have on speed of
+convergence and success rates, we show a result for using 800 processors, where
+most of the algorithms have a 100\% success rate.  In this case, the best
+algorithm to use is the one which finds the optimum the fastest.  As can be
+seen in \fig{griewank-800}, with 800 processors the best choice is Pick Best.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{griewank6}
+  \caption{Function Griewank with a swarm that uses 800 processors per round of
+  evaluations.}
+  \label{fig:griewank-800}
+\end{figure}
+
+\subsection{Bohachevsky}
+
+Bohachevsky is a unimodal function best optimized with a Complete swarm.  It is
+similar to Griewank in that there is a global optimum with a value of 0, and
+the swarm sometimes finds it and sometimes doesn't.  Since we have already
+demonstrated the convergence properties of the various algorithms with
+Griewank, for Bohachevsky we merely present the best results.  In
+\fig{bohachevsky} we can see that all of the speculative algorithms found the
+optimum much quicker than PSO with a Random topology.  However, SEPSO was
+slower than PSO Complete and got stuck around 15\% of the time.  Using Pick
+Best, Pick Best Pruned, and Many Iterations all outperformed PSO Complete,
+though Many Iterations only almost reached a 100\% success rate.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{bohachevsky}
+  \caption{Function Bohachevsky with a swarm that uses 800 processors per round
+  of evaluations.}
+  \label{fig:bohachevsky}
+\end{figure}
 
@@
 
+We have given five different variants of speculative evaluation, each of which
+has different properties.  The variants of the algorithm perform differently on
+different functions and at different swarm sizes, as would be expected by the
+No Free Lunch Theorem~\cite{wolpert-1997-nfl-for-optimization}.  We have given
+a brief evaluation of the premature convergence properties of these variants on
+deceptive functions.  Though there is no one variant that performs best across
+all functions, for all but one of the functions we looked at \emph{some} kind
+of speculative evaluation outperforms standard PSO, often dramatically.
+
 Though our methods show great improvements on some functions, they do not work
@@
 
-Griewank is a function in between Sphere and Rastrigin.  It is highly deceptive
-and is prone to premature convergence, but by adding particles to the swarm a
-point is reached where ``enough'' exploration is done, and the algorithm finds
-the optimum essentially all of the time.  For such functions, the best approach
-seems to be to increase the swarm size until ``enough'' exploration is reached,
-then use extra processors to perform speculative evaluation and increase the
-number of iterations performed.  Sphere and Rastrigin can be thought of as
-special cases of these types of functions; Sphere simply needs a very small
-swarm size to produce ``enough'' exploration, and Rastrigin requires a very
-large swarm.  We expect that for all functions there is a swarm size for which
-additional particles are less useful than additional iterations.
+Griewank is a function in between Sphere and Rastrigin (as are Bohachevsky and
+Schwefel 2.21).  It is highly deceptive and is prone to premature convergence,
+but by adding particles to the swarm a point is reached where ``enough''
+exploration is done, and the algorithm finds the optimum essentially all of the
+time.  For such functions, the best approach seems to be to increase the swarm
+size until ``enough'' exploration is reached, then use extra processors to
+perform speculative evaluation and increase the number of iterations performed.
+Sphere and Rastrigin can be thought of as special cases of these types of
+functions; Sphere simply needs a very small swarm size to produce ``enough''
+exploration, and Rastrigin requires a very large swarm.  We expect that for all
+functions there is a swarm size for which additional particles are less useful
+than additional iterations.
 
@@
 \label{sec:future}
-
-This work represents the first attempt of which we are aware at using
-additional processors in parallel with PSO to do something other than either
-add particles to the swarm or parallelize the evaluation of the objective
-function.  We anticipate that many future advancements can be made.
 
@@
 ../../../bib/aml/bib,%
+../../../bib/nfl/bib,%
 ../../../bib/pso/general/bib,%

### 2010-05-06 honors_thesis/journal_paper/paper.tex
Done with my final pass through the journal paper

@@
 Speculative Evaluation in PSO can be enhanced by diverging from the behavior of
-the standard PSO to make better use of the information obtained by the
-speculative evaluations.  Using these algorithmic modifications, we see
-dramatic improvements in the effectiveness of the algorithm.
+the standard PSO.  We present several variations on our speculative algorithm,
+such as making use of the information obtained by the extra speculative
+evaluations, and speculating more than one iteration ahead.  Using these
+algorithmic modifications, we see dramatic improvements in the effectiveness of
+the algorithm.
 
@@
 In this paper we consider PSO parallelization strategies for clusters of
-thousands of processors and functions for which a single evaluation will take
-at least tens of seconds, but probably minutes or perhaps hours.  Our purpose
-is to explore the question of what to do with a thousand processors when 50 or
-100 particles is the recommended swarm size, and simply increasing the swarm
-size encounters diminishing returns.
+hundreds or thousands of processors and functions for which a single evaluation
+will take at least tens of seconds, but probably minutes or perhaps hours.  Our
+purpose is to explore the question of what to do with a thousand processors
+when 50 or 100 particles is the recommended swarm size, and simply increasing
+the swarm size encounters diminishing returns.
 
@@
 comes from an analogy to speculative execution (also known as branch
-prediction) techniques commonly used in processors.  Modern processors, when
-faced with a branch on which they must wait (e.g. a memory cache miss), guess
-which way the branch will go and start executing, ensuring that any changes
-can be undone.  If the processor guesses right, execution is
-much farther ahead than if it had idly waited on the memory reference.  If
-it guesses wrong, execution restarts where it would have been anyway.
+prediction), a technique commonly used in processors.  Modern processors, when
+faced with a branch on which they must wait (e.g., a memory cache miss), guess
+which way the branch will go and start executing, ensuring that any changes can
+be undone.  If the processor guesses right, execution is much farther ahead
+than if it had idly waited on the memory reference.  If it guesses wrong,
+execution restarts where it would have been anyway.
 
@@
 speculative evaluation in PSO, all of which break the requirement of strictly
-reproducing the behavior of the original algorithm.  \sec{results} presents the
-results of using our methods in comparison with standard PSO.  In
-\sec{conclusion} and \sec{future} we conclude and discuss future work.
+reproducing the behavior of the original algorithm.  \sec{setup} describes the
+experiments we ran, and \sec{results} presents the results of those
+experiments.  In \sec{conclusion} and \sec{future} we conclude and discuss
+future work.
 
@@
 
-Changing the way neighbors are defined (often called the ``topology'' in the
-literature) has a significant effect on the performance of the algorithm.  In
-the Ring topology, each particle has one neighbor to either side of it; in the
-Complete topology, every particle is a neighbor to every other
+Changing the way neighbors are defined, usually called the ``topology,'' has a
+significant effect on the performance of the algorithm.  In the Ring topology,
+each particle has one neighbor to either side of it; in the Complete topology,
+every particle is a neighbor to every other
 particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
@@
 
-The PSO algorithm can be trivially parallelized by distributing the processing
+The PSO algorithm can be trivially parallelized by distributing the computation
 needed for each particle on across processors.  But as we have seen in
-\fig{evals-sphere}, for some functions, and for large numbers of
-processors, just adding particles reaches a point of diminishing returns.
-That is, adding processors does not help us reach any given level of fitness
+\fig{evals-sphere}, for some functions, and for large numbers of processors,
+just adding particles reaches a point of diminishing returns.  That is, adding
+processors does not help the algorithm reach any given level of fitness
 appreciably faster.  Instead of adding particles we employ a speculative
-evaluation approach that allows us to perform iterations two at a time.
+approach that allows us to perform iterations two at a time.
 
@@
 
-The refactoring hinges on the idea that once the random coefficients are
-determined, there are only a few possibile updates to $\nbest$ and $\pbest$.
-For the Ring topology there are 7 possible update cases, identified in
-Table~\ref{tab:evals}.  We label each case with an identifier referring to the
-source of the update: a minus sign ($-$) represents no update, $L$ represents
-an update to $\nbest$ coming from the left neighbor, $R$ represents an update
-to $\nbest$ coming from the right neighbor, and $S$ represents an update to
-either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
-$\casexn$ refers to the case that the particle finds a new personal best, but
-neither it nor its neighbors found a position that updated its neighborhood
-best.  In the equations that follow, we refer to an unspecified update case as
-$\casegen$, and to the set of cases collectively as $\caseset$.
+The refactoring hinges on the idea that once the random coefficients
+$\pURand_{t}$ and $\pURand_{t}$ are determined, there are only a few possible
+updates to $\nbest$ and $\pbest$.  For the Ring topology there are 7 possible
+update cases, identified in Table~\ref{tab:evals}.  We label each case with an
+identifier referring to the source of the update: a minus sign ($-$) represents
+no update, $L$ represents an update to $\nbest$ coming from the left neighbor,
+$R$ represents an update to $\nbest$ coming from the right neighbor, and $S$
+represents an update to either $\pbest$ or $\nbest$ coming from the particle
+itself.  As an example, $\casexn$ refers to the case that the particle finds a
+new personal best, but neither it nor its neighbors find a position that
+updated its neighborhood best.  In the equations that follow, we refer to an
+unspecified update case as $\casegen$, and to the set of cases collectively as
+$\caseset$.
 
@@
 	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
-	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
-	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
-	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
+	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
+	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
+	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
 	   0 & \text{otherwise}
@@
 
-In this form the important point to notice is that there are only 7 values (for
-this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
+In this form the important point to notice is that there are only $7$ values
+(for this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
 \caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
-objective function evaluation at iteration $t$ (note also that while there are
+objective function evaluation at iteration $t$.  Note also that while there are
 random numbers in the equation, they are assumed fixed once drawn for any
-particular particle at a specific iteration).  Thus PSO has been refactored
-such that the algorithm can begin computing all 7 of the objective function
+particular particle at a specific iteration.  Thus PSO has been refactored such
+that the algorithm can begin computing all $7$ of the objective function
 evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
@@
 
-Although this speculative algorithm computes
-$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
-for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will be
-ignored, and might just as well never have been computed.  We call the set of
-computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
-``speculative children'' because only one of them will be needed.
-
-To see the value of this refactoring, suppose that 1000 processors are
-available, and that the evaluation of the function to be optimized takes one
-hour.  If we only want a swarm of 100 particles, 900 of the processors would be
+Although this speculative approach computes $\ofunc(\specpos_{t+1}^{\casegen})$
+for all $\casegen \in \caseset$, even those for which $\indic_{t+1}^{\casegen}
+= 0$, these extra computations will be ignored, and might just as well never
+have been computed.  We call the set of computations
+$\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$ ``speculative
+children'' because only one of them will be needed.
+
+To see the value of this refactoring, suppose that $1000$ processors are
+available, and that the evaluation of the objective function takes one hour.
+If we only want a swarm of $100$ particles, $900$ of the processors would be
 sitting idle for an hour at every iteration, and it would take two hours to run
 two iterations.  If instead we perform speculative evaluation, sending each of
-the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
-$\ofunc(\ppos_{t})$, we would use 800 of the 1000 processors and perform two
-iterations in one hour.
+the $7$ evaluations in $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
+to be computed at the same time as $\ofunc(\ppos_{t})$, we would use $800$ of
+the $1000$ processors and perform two iterations in one hour.
 
 In order to do two iterations at once, we must use 8 times as many processors
-as there are particles in the swarm.  If we were not performing speculative
-evaluation, those processors might very well be used to add particles to the
-swarm.  This then raises the question of whether a swarm of 100 particles doing
-twice as many iterations outperforms a swarm of 800 particles.  We show in the
-\sec{results} that in many, though not all, instances, a smaller swarm
-performing more iterations does in fact outperform a larger swarm.
+as there are particles in the swarm.  If these processors were not performing
+speculative evaluation, they might instead be used for function evaluations
+needed to support a larger swarm.  This raises the question of whether a swarm
+of 100 particles doing twice as many iterations outperforms a swarm of 800
+particles.  We show in \sec{results} that in many, though not all, instances, a
+smaller swarm performing more iterations does in fact outperform a larger
+swarm.
 
@@
 speculative evaluations are done per particle, while all but one of them are
-completely ignored.  We can try to make use of those evaluations instead of
-throwing them away.
-
-To use the extra speculative evaluations, instead of choosing the speculative
-child that matches the branch that the original PSO would have taken, we take
-the child that has the best value.  The methodology is exactly the same as
-above except for the process of choosing which speculative child to accept.
-The only change needed in \alg{centralized} is in step 7, where
+completely ignored.  It seems reasonable to try to make use of the information
+obtained through those evaluations instead of ignoring it.
+
+To make better use of the extra speculative evaluations, instead of choosing
+the speculative child that matches the branch that the original PSO would have
+taken, we take the child that has the best value.  The methodology is exactly
+the same as above except for the process of choosing which speculative child to
+accept.  The only change needed in \alg{centralized} is in step 7, where
 $\noeval{\s}_{t+1}$ with the best value is chosen from $\noeval{\sset}_{t+1}$
@@
 we would expect this method to work well, while performance on more deceptive
-functions would probably be hurt.
+functions would probably be hurt.  We call this technique Pick Best.
 
@@
 
-There are a large number of ways to prune speculative evaluations.  We
+The algorithms we compare are the standard PSO algorithm, SEPSO, and four
+variations on SEPSO.  In presenting our results, we call the variant developed
+in \sec{pickbest} Pick Best.  The algorithms described in \sec{pruning} through
+\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
+algorithm in \sec{manyiters} we call Many Iterations.
+
+When performing pruning in SEPSO in Pick Best Pruned and Social Promotion
+Pruned, there are a large number of ways to prune speculative evaluations.  We
 experimented with several, but present results for only one possible pruning.
@@
 
-There are also many ways to speculate several iterations ahead, and we again
-only show results for one of them.  The combination of branches we tried uses
-seven speculative evaluations per particle, matching the swarm size of the
-original speculative algorithm.  The seven evaluations we used corresponded to
-several iterations of branches \casepn\ and \casexn.  Two of the evaluations
-were just one iteration ahead, four were two iterations ahead, and one was
-three iterations ahead.  The evaluations that were one iteration ahead were
-branches \casepn\ and \casexn; those that went two iterations ahead were formed
-by taking either branch \casepn\ or \casexn and then branch \casepn\ or
-\casexn; and the evaluation going three iterations ahead followed branch
-\casepn\ on all three iterations.
+There are also many ways to speculate several iterations ahead, and with Many
+Iterations we again only show results for one of them.  The combination of
+branches we tried uses seven speculative evaluations per particle, matching the
+swarm size of the original speculative algorithm.  The seven evaluations we
+used corresponded to several iterations of branches \casepn\ and \casexn.  Two
+of the evaluations were just one iteration ahead, four were two iterations
+ahead, and one was three iterations ahead.  The evaluations that were one
+iteration ahead were branches \casepn\ and \casexn; those that went two
+iterations ahead were formed by taking either branch \casepn\ or \casexn and
+then branch \casepn\ or \casexn; and the evaluation going three iterations
+ahead followed branch \casepn\ on all three iterations.
 
@@
   \caption{Function Rastrigin with a swarm that uses 240 processors per round
-  of evaluations.}
+  of evaluations.  All speculative algorithms use the Random topology.}
   \label{fig:rastrigin}
@@
 Next we go to Sphere, the simplest of common benchmark functions.  The function
-has no local optima and is best optimized in terms of function evaluations with
-a small swarm using a Complete topology.  We expect our methods to be perfect
-for such functions, and our results show this intuition again to be correct.
-First we compare the basic SEPSO algorithm to standard PSO.  In
-\fig{basic-sphere} we can see that when both SEPSO and PSO are using a Random
-topology, SEPSO clearly outperforms PSO.  However, the comparison between SEPSO
-and PSO Complete is not as encouraging---PSO Complete performs slightly better.
+has no local optima (only a single global optimum) and is best optimized in
+terms of function evaluations with a small swarm using a Complete topology.  We
+expect our methods to be perfect for such functions, and our results show this
+intuition again to be correct.  First we compare the basic SEPSO algorithm to
+standard PSO.  In \fig{basic-sphere} we can see that when both SEPSO and PSO
+are using a Random topology, SEPSO clearly outperforms PSO.  However, the
+comparison between SEPSO and PSO Complete is not as encouraging---PSO Complete
+performs slightly better.
 
@@
   \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations.}
+  evaluations.  SEPSO uses the Random topology.}
   \label{fig:basic-sphere}
@@
   \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing Pick Best with previous results.}
+  evaluations, comparing Pick Best with previous results.  All speculative
+  algorithms use the Random topology.}
   \label{fig:sphere-pickbest}
@@
 several iterations ahead.  Pick Best Pruned outperforms Pick Best, and Many
-Iters by far outperforms every other algorithm on this function.  Many Iters on
-average finds values that are more than 40 orders of magnitude better than
-standard PSO using the same number of processors.  This is shown in
+Iterations by far outperforms every other algorithm on this function.  Many
+Iterations on average finds values that are more than 40 orders of magnitude
+better than standard PSO using the same number of processors.  This is shown in
 \fig{sphere-manyiters}.  Social Promotion Pruned performed comparably with
@@
   evaluations, comparing speculating several iterations ahead with previous
-  results.}
+  results.  All speculative algorithms use the Random topology.}
   \label{fig:sphere-manyiters}
@@
   \caption{Function Schwefel with a swarm that uses 240 processors per round of
-  evaluations.}
+  evaluations.  SEPSO uses a Random topology.}
   \label{fig:schwefel1}
@@
 perform worse than PSO Complete, with Many Iterations performing comparably to
-PSO.  The only algorithm that performed better than PSO was Pick Best Pruned,
-likely because of the larger swarm size available to it.
+PSO.  However, one of the speculative algorithms did perform better than PSO,
+Pick Best Pruned.  It seems that the larger swarm size available because of
+pruning was the cause of the improved performance.
 
@@
   \caption{Function Schwefel with a swarm that uses 240 processors per round of
-  evaluations, comparing all speculative algorithms.}
+  evaluations, comparing all speculative algorithms.  All speculative
+  algorithms use the Random topology.}
   \label{fig:schwefel2}
@@
 optimum with a value of 0, and sometimes the swarm finds the optimum and
-sometimes it doesn't.  Instead of showing average function value at each
+sometimes it does not.  Instead of showing average function value at each
 iteration, a more informative plot for Griewank shows the percent of runs that
@@
   evaluations.  Instead of showing average function value, we show the
-  percentage of runs that have found the global optimum by each iteration.}
+  percentage of runs that have found the global optimum by each iteration.  All
+  algorithms use the Ring topology.}
   \label{fig:basic-griewank1}
@@
   \caption{Function Griewank with a swarm that uses 800 processors per round of
-  evaluations.}
+  evaluations.  All algorithms use the Ring topology.}
   \label{fig:basic-griewank2}
@@
   \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing Pick Best with previous results.}
+  evaluations, comparing Pick Best with previous results.  All algorithms use
+  the Ring topology.}
   \label{fig:griewank-pickbest}
@@
   \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing pruning with previous results.}
+  evaluations, comparing pruning with previous results.  All algorithms use the
+  Ring topology.}
   \label{fig:griewank-pruned}
@@
 stuck and fails to find the optimum more than any other method.  The results
-are shown in \fig{griewank-manyiters}. (TODO: Rerun Many Iterations with the
-right swarm size)
+are shown in \fig{griewank-manyiters}.
 
@@
   evaluations, comparing speculating several iterations ahead with previous
-  results.}
+  results.  All algorithms use the Ring topology.}
   \label{fig:griewank-manyiters}
@@
   \caption{Function Griewank with a swarm that uses 800 processors per round of
-  evaluations.}
+  evaluations.  All algorithms use the Ring topology.}
   \label{fig:griewank-800}
@@
 similar to Griewank in that there is a global optimum with a value of 0, and
-the swarm sometimes finds it and sometimes doesn't.  Since we have already
+the swarm sometimes finds it and sometimes does not.  Since we have already
 demonstrated the convergence properties of the various algorithms with
@@
   \caption{Function Bohachevsky with a swarm that uses 800 processors per round
-  of evaluations.}
+  of evaluations.  All of the speculative algorithms use the Random topology.}
   \label{fig:bohachevsky}

### 2010-05-28 honors_thesis/journal_paper/paper.tex
A few minor fixes to the paper; I changed the title

@@
 
-\title{\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization%
+\title{\ \\ \LARGE\bf A Speculative Approach to Parallelization in Particle
+Swarm Optimization%
 \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department
@@
 perform worse than PSO Complete, with Many Iterations performing comparably to
-PSO.  However, one of the speculative algorithms did perform better than PSO,
+PSO.  However, one of the speculative algorithms did perform better than PSO:
 Pick Best Pruned.  It seems that the larger swarm size available because of
@@
 stuck and fails to find the optimum more than any other method.  The results
-are shown in \fig{griewank-manyiters}.
+are shown in \fig{griewank-manyiters}.   \fig{griewank-manyiters} is also
+interesting in that it highlights the trade-off between accuracy and speed in
+the various algorithms at this swarm size.  The faster the algorithm finds the
+optimum, the less likely it is to be successful.
 
@@
 algorithm to use is the one which finds the optimum the fastest.  As can be
-seen in \fig{griewank-800}, with 800 processors the best choice is Pick Best.
+seen in \fig{griewank-800}, with 800 processors the best choice is Pick Best,
+which finds the optimum on average three times as fast as standard PSO.
 
@@
 optimum much quicker than PSO with a Random topology.  However, SEPSO was
-slower than PSO Complete and got stuck around 15\% of the time.  Using Pick
-Best, Pick Best Pruned, and Many Iterations all outperformed PSO Complete,
-though Many Iterations only almost reached a 100\% success rate.
+slower than PSO Complete and got stuck around 15\% of the time.  Pick Best,
+Pick Best Pruned, and Many Iterations all outperformed PSO Complete, though
+Many Iterations only almost reached a 100\% success rate.
 

### 2010-06-09 honors_thesis/journal_paper/paper.tex
Making some modifications from feedback from Chris

@@
 significant effect on the performance of the algorithm.  In the Ring topology,
-each particle has one neighbor to either side of it; in the Complete topology,
-every particle is a neighbor to every other
+each particle has one neighbor to either side of it; in the Complete
+topology\footnote{The Complete topology has often been unfortunately named Star
+in the literature, which in graph theory refers to a completely different
+topology.  Other names have also been used, including ``global topology'' or
+gbest.  We use the graph theory term ``Complete'' in this paper.}, every
+particle is a neighbor to every other
 particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
@@
 
-The PSO algorithm can be trivially parallelized by distributing the computation
-needed for each particle on across processors.  But as we have seen in
-\fig{evals-sphere}, for some functions, and for large numbers of processors,
-just adding particles reaches a point of diminishing returns.  That is, adding
-processors does not help the algorithm reach any given level of fitness
-appreciably faster.  Instead of adding particles we employ a speculative
-approach that allows us to perform iterations two at a time.
+PSO can be trivially parallelized by assigning each particle's computation to
+an individual processor.  But as we have seen in \fig{evals-sphere}, for some
+functions, and for large numbers of processors, just adding particles reaches a
+point of diminishing returns.  That is, adding processors does not help the
+algorithm reach any given level of fitness appreciably faster.  Instead of
+adding particles we employ a speculative approach that allows us to perform
+iterations two at a time.
 
@@
 
-In some distributed frameworks, like the MapReduce framework that we used in
-our experiments, rounds of message passing can be expensive.  The method just
-described uses three rounds of message passing for every two iterations
-(corresponding to steps 2, 5 and 10 in \alg{centralized}).  It is possible to
-perform speculative evaluation in PSO with only one round of communication per
-two iterations.  However, the methodology is tedious and is not the focus of
-this paper, so we defer its description to the Appendix.
+In distributed frameworks, sychronizing all of the machines for a round of
+message passing can be expensive.  The method just described uses three rounds
+of message passing for every two iterations (corresponding to steps 2, 5 and 10
+in \alg{centralized}).  It is possible to perform speculative evaluation in PSO
+with only one round of communication per two iterations.  However, the
+methodology is tedious and is not the focus of this paper, so we defer its
+description to the Appendix.
 
@@
 $\noeval{\s}_{t+1}$ with the best value is chosen from $\noeval{\sset}_{t+1}$
-instead of with the matching branch.
+instead of with the matching branch.  We call this technique Pick Best.
 
@@
 and accepting the best one.  Speculative particles that move in good directions
-are kept.  It seems this modification to PSO would produce more exploitation,
-decreasing the amount of exploration.  With functions that are less deceptive
-we would expect this method to work well, while performance on more deceptive
-functions would probably be hurt.  We call this technique Pick Best.
+are kept.  Intuition says that this technique favors exploitation over
+exploration, but as we will show in \sec{results}, that is not always the case.
 

### 2010-06-09 honors_thesis/ppsn_paper/paper.tex
Just some formatting, I think
Making the camera ready version for PPSN

@@
 
-Changing the way neighbors are defined, usually called the ``topology,''
-has a significant effect on the performance of the algorithm.  In
-the Ring topology, each particle has one neighbor to either side of it; in the
-Complete topology, every particle is a neighbor to every other
+Changing the way neighbors are defined, usually called the ``topology,'' has a
+significant effect on the performance of the algorithm.  In the Ring topology,
+each particle has one neighbor to either side of it; in the Complete topology,
+every particle is a neighbor to every other
 particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
@@
 
-Speculative evaluation is made possible by refactoring
-PSO such that evaluating the
+Speculative evaluation is made possible by refactoring PSO such that evaluating
 the objective function is separate from the rest of the computation.  For
@@
 
-The refactoring hinges on the idea that once the random coefficients $\pURand_{t}$ and $\pURand_{t}$ are
-determined, there are only a few possible updates to $\nbest$ and $\pbest$.
-For the Ring topology there are 7 possible update cases, identified in
-Table~\ref{tab:evals}.  We label each case with an identifier referring to the
-source of the update: a minus sign ($-$) represents no update, $L$ represents
-an update to $\nbest$ coming from the left neighbor, $R$ represents an update
-to $\nbest$ coming from the right neighbor, and $S$ represents an update to
-either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
-$\casexn$ refers to the case that the particle finds a new personal best, but
-neither it nor its neighbors found a position that updated its neighborhood
-best.  In the equations that follow, we refer to an unspecified update case as
-$\casegen$, and to the set of cases collectively as $\caseset$.
+The refactoring hinges on the idea that once the random coefficients
+$\pURand_{t}$ and $\pURand_{t}$ are determined, there are only a few possible
+updates to $\nbest$ and $\pbest$.  For the Ring topology there are 7 possible
+update cases, identified in Table~\ref{tab:evals}.  We label each case with an
+identifier referring to the source of the update: a minus sign ($-$) represents
+no update, $L$ represents an update to $\nbest$ coming from the left neighbor,
+$R$ represents an update to $\nbest$ coming from the right neighbor, and $S$
+represents an update to either $\pbest$ or $\nbest$ coming from the particle
+itself.  As an example, $\casexn$ refers to the case that the particle finds a
+new personal best, but neither it nor its neighbors found a position that
+updated its neighborhood best.  In the equations that follow, we refer to an
+unspecified update case as $\casegen$, and to the set of cases collectively as
+$\caseset$.
 

### 2010-06-24 resume_cv/cv.tex
Updated my cv and my list of applications

@@
         \item Minors in Physics and Mathematics
-        \item Graduated with a 4.0 GPA and University honors (anticipated)
+        \item Graduated summa cum laude (GPA 4.0) and with University honors
         \item Thesis Topic: \emph{Speculative Evaluation in Particle Swarm
@@
 \section{Academic Experience}
-{\textbf{Brigham Young University}},
-Provo, Utah USA
+{\textbf{Johns Hopkins University}}, Baltimore, Maryland USA
 \begin{outerlist}
 
+\item[] \textit{Human Language Technologies Summer School Attendee}%
+        \hfill \textbf{June 2010}
+\begin{innerlist}
+\item One of seven recipients of a NAACL scholarship to attend the summer school
+  at JHU.
+\item Studied with top researchers in HLT at JHU and the University of Maryland.
+\end{innerlist}
+\end{outerlist}
+
+\blankline
+
+{\textbf{Brigham Young University}}, Provo, Utah USA
+\begin{outerlist}
+\item[] \textit{Teaching Assistant}%
+	\hfill \textbf{May 2010 to June 2010}
+\begin{innerlist}
+\item CS 470: Introduction to Artificial Intelligence.  Tutored students,
+  graded homework and reports, and helped create new curriculum for the class.
+\item CS 677: Bayesian Inference.  Tutored students and graded homework and
+  reports.
+\end{innerlist}
+
 \item[] \textit{Undergraduate Researcher}%
-        \hfill \textbf{December 2007 to Present}
+        \hfill \textbf{December 2007 to April 2010}
 \begin{innerlist}

### 2010-06-24 orca/finalreport2009.tex
Added my final report for ORCA for the 2009 grant I was awarded

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{cite}

\title{A Speculative Approach to Parallelization in Particle Swarm Optimization}
\author{Matthew Gardner\\Mentor: Kevin Seppi, Computer Science}
\date{}

\begin{document}
\maketitle

Optimization problems are ubiquitous in our modern world.  Businesses such as
Google need to decide where to place ads, scientists need to fit models to
data, and airlines need to schedule their flights.  All of these problems are
optimization problems, and optimization techniques can be used to find
solutions.  Particle swarm optimization (PSO) is a relatively new optimization
technique that draws on ideas from the sociology of flocking birds.  It was
originally developed for use on a single machine, and attempts at modifying the
algorithm to run in parallel are still quite preliminary.  However, the
problems people want to solve are getting bigger, and larger problems require
more computation power than is available on any single computer.  To
effectively solve those problems, optimization algorithms need to be improved
to better use parallel clusters of machines.  This project addressed that
issue.

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is
used to intelligently search a multi-dimensional space by mimicking the
swarming and flocking behavior of birds and other animals. It is a social
algorithm that depends on interaction between particles to quickly and
consistently find the optimal solution to some objective function.

The motion of particles through the search space has three components: an
inertial component that gives particles momentum as they move, a cognitive
component where particles remember the best solution they have found and are
attracted back to that place, and a social component by which particles are
attracted to the best solution that any of their neighbors have found.

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO.  The methods they have used include
various synchronous algorithms and asynchronous algorithms.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm and thus was not the focus of this
work.

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm,
either by adding individual particles or by adding entire new sub-swarms.  In
almost all cases, adding additional particles produces better results in the
same amount of time.  However, indefinitely increasing the swarm size produces
only incremental benefit; performance improves, but not by very much.  The
reason for this is that there comes a point of diminishing returns with respect
to adding particles.

In this work we considered PSO parallelization strategies for clusters of
hundreds or thousands of processors and functions for which a single evaluation
will take at least tens of seconds, but probably minutes or perhaps hours.  Our
purpose is to explore the question of what to do with a thousand processors
when 50 or 100 particles is the most efficient swarm size, and simply adding
particles results in only incremental improvement.

In previous work (my honors thesis, which was also sponsored by an ORCA grant)
we developed an algorithm we called Speculative Evaluation in PSO.  The
algorithm works by finding all possible positions of particles at future
iterations, then evaluating them in parallel.  That work resulted in a
publication in the proceedings of \emph{Parallel Problem Solving in Nature,} a
prestigious conference in parallelizing nature-based algorithms.

However, that work left us many open questions, and the results we presented
were decent, but not spectacular.  In this work we expanded greatly on the
results of our previous paper, exploring in much more depth the possibilities
of speculative computation in PSO.

Our original algorithm reproduced the PSO algorithm exactly, two iterations at
a time, in parallel, so that the algorithm could be run more quickly on
functions that took a long time to evaluate.  But exactly reproducing PSO
required us to use eight times as many processors as standard PSO, while those
processors could have been used to increase the swarm size of the standard
algorithm.  Our algorithm thus required too much computation to be competitive
in very many instances.

To solve this problem, we relaxed the requirement that our algorithm exactly
reproduce the behavior of standard PSO.  Thus we were able to save some
computation costs and increase the swarm size of our algorithm, and our results
greatly improved.  After introducing some of these cost-saving measures, we
were able to also speculate several iterations ahead, instead of just one, and
that led to further performance improvements in many instances.

The result of this work was an expanded version of our original paper that was
twice as long.  We submitted this more thorough treatment of speculative
evaluation in PSO to the journal \emph{Transactions on Evolutionary
Computation}, one of the premier journals treating research on evolution- and 
nature-inspired algorithms, such as PSO.  We hope to hear a response from the
journal soon.

As is typical of research projects, we were left again with many open
questions.  One of the most intriguing is how to most effectively allocate
speculative evaluations to available processors.  Our approaches treat every
particle equally, giving the same number of extra evaluations to both poorly
performing particles and those that are performing very well.  Perhaps a better
approach would be to draw on ideas from the multi-robot task allocation
literature and dynamically allocate extra processors to particles that could
best make use of them.

\bibliographystyle{plain}
\bibliography{../../bib/pso/general/bib}

\end{document}

### 2010-06-25 textual_criticism/proposal/proposal.tex
Added a section in here for my 598 project on textual criticism

\documentclass[onecolumn, 11pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{Automated Textual Criticism}
\author{Matthew Gardner}
\date{June 25, 2010}

\begin{document}
\maketitle

\section{Introduction}

Textual criticism can be loosely described as the analysis of a series of
manuscripts of a document (generally an ancient document) in an attempt to
reconstruct the original text produced by the author(s) of the document.
Because of the importance of religious documents and the proliferation of their
manuscripts, textual criticism has grown into a very large field, though its
use has not been confined solely to religious documents.

The process of textual criticism involves laboring over differing manuscripts
searching for discrepancies, recording them by hand and sifting through the
various manuscripts in order to come to some conclusion about the manuscripts'
origins.  Much of the work of textual criticism seems like it could be
automated, greatly reducing the workload on scholars and perhaps even producing
better results, by finding things that scholars miss.

\section{Objectives}

The main objective of this project is to explore the possibility of using
computers, especially statistical methods in natural language processing and
computational biology, to completely automate the process of textual criticism.
Were automated textual criticism possible, the resultant system would take as
input a collection of manuscripts and output a critical edition of the work
(with variant readings, probabilities that each of the readings was the
original reading, a history of changes to the manuscript, and so forth).  Such
a system is much too ambitious a project for a 598R project; I will seek only
to explore the feasibility of such a system, and to what extent people have
tried to build them already.

I will also try to build some portion of such a system, though what portion
that would be has yet to be determined, and will likely be influenced by my
initial research.  My efforts to build a system may also be hampered by a lack
of data to use; I do not yet know what the availability of manuscripts in
digital form is like.  So while building part of a system for automated textual
criticism is a desired objective, the main objective is the exploration of the
space to determine feasibility.

\section{Outline of proposed work}

The bulk of the work of this project will be reading, to familiarize myself
with the field and see what has been done before.  In particular, the reading
will include at least the following:

\begin{itemize}
  \item \emph{Misquoting Jesus}, by Bart Ehrman
  \item \emph{The Text of the New Testament}, by Bruce Metzger and Bart Ehrman
  \item \emph{How the New Testament Came to Be}, the proceedings of the Sperry
	Symposium in 2006
\end{itemize}

These books will give me an idea of the process of textual criticism and what
exactly it involves.  Understanding how modern textual critics do their work is
essential to automating the process.

I will also search what literature I can find about automating the process of
textual criticism.  Some has already been done, called cladistics (the
application of computational biology to textual criticism); this section of my
reading will explore to what extent such attempts have been made and how
successful they are.

Another important part of my work will be contacting individuals who work in
the field, to discuss the idea of automated textual criticism with them and to
see what access there is to relevant data sources.  If none of the manuscripts
are available digitally, automated methods are not possible.  I will talk at
least to BYU professors in the field, including Thomas Wayment, Don Parry,
David R. Seely, and Stephen Bay.

Finally, assuming that I find some amount of data to be available, I will try
to implement some part of an automated system for textual criticism.  Exactly
what part of the system that might be depends a lot on what I find in my
reading.  In the absence of any available data in digital form, I can at least
produce simulated data and test the results of my system on that.

I will also meet weekly with my faculty mentor, on Fridays at 11:00 AM, to
discuss my progress and get further direction.

\section{Deliverables}

The main deliverable of this project will be a paper summarizing my findings.
At the very least this will be a survey of the various parts of textual
criticism, what knowledge sources are required to accomplish each aspect of it,
my ideas about the feasibility of automating each of those parts, and what work
has already been done to automate them.  It is possible that such a survey
could be submitted for publication to a conference or journal, but I do not
know yet what conferences or journals are published in the field.

Assuming some part of the processes can be automated and I can find data to
use, a secondary deliverable is code that does some amount of automated textual
criticism.  Along with such code would go a paper describing the code, and if
it is deemed interesting enough, that paper could be submitted for publication
to some venue for textual criticism.  It seems to me, though, that producing a
publishable paper on top of all of the research and code writing that would be
involved might take a little longer than a term.  A reasonable deliverable to
me seems to be at least the beginnings of a paper that could be published.

\end{document}

### 2010-06-25 resume_cv/cv.tex
Updated my cv and list of applications just a bit

@@
 %%%%%%%%%%%%%%%%%%%%%% Simple LaTeX CV Template %%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% NOTE: If you find that it says                                     %%
-%%                                                                    %%
-%%                           1 of ??                                  %%
-%%                                                                    %%
-%% at the bottom of your first page, this means that the AUX file     %%
-%% was not available when you ran LaTeX on this source. Simply RERUN  %%
-%% LaTeX to get the ``??'' replaced with the number of the last page  %%
-%% of the document. The AUX file will be generated on the first run   %%
-%% of LaTeX and used on the second run to fill in all of the          %%
-%% references.                                                        %%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \item National Merit Scholarship, 2004--2010
+\item Awarded a research grant from the Office of Research and Creative
+  Activities in both 2008 and 2009
 \item Inducted into the Phi Kappa Phi Honors Society, 2009
@@
 
-\section{Conference Publications} \begin{bibsection}
-    \item Andrew McNabb, Matthew Gardner, and Kevin Seppi. An Exploration of
-	  Topologies and Communication in Large Particle Swarms.  In:
-	  \emph{Congress on Evolutionary Computation, 2009}.
-
-	\item Matthew Gardner and Eric Ringger.  Automatic Topic Discovery in 100
-	  Years of General Conference Talks.  Presented at the 50th Anniversary
-	  BYU Studies Symposium, 2010.
-
-	\item Matthew Gardner, Andrew McNabb, and Kevin Seppi.  Speculative
-	  Evaluation in Particle Swarm Optimization.  Submitted to:
-	  \emph{Transactions on Evolutionary Computation}.  Awaiting acceptance
-	  notification.
-\end{bibsection}
-
 \section{Professional Experience}
@@
   advertisers to specific ads played on TV networks.  Official blog post about
-  my project can be found at:
+  the project can be found at:
   http://google-tmads.blogspot.com/2010/01/track-cost-per-call-data-through-google.html.
@@
 
+\section{Publications} \begin{bibsection}
+	\item Matthew Gardner, Andrew McNabb, and Kevin Seppi.  Speculative
+	  Evaluation in Particle Swarm Optimization.  In: \emph{Parallel Problem
+	  Solving in Nature, 2010}.
+
+	\item Matthew Gardner and Eric Ringger.  Automatic Topic Discovery in 100
+	  Years of General Conference Talks.  Presented at the 50th Anniversary
+	  BYU Studies Symposium, 2010.
+
+    \item Andrew McNabb, Matthew Gardner, and Kevin Seppi. An Exploration of
+	  Topologies and Communication in Large Particle Swarms.  In:
+	  \emph{Congress on Evolutionary Computation, 2009}.
+\end{bibsection}
+
 \end{document}

### 2010-08-02 textual_criticism/paper/paper.tex
Significant work on the textual criticism paper

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\newcommand{\secref}[1]{Section~\ref{sec:#1}}
\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}

\title{An Inquiry into Automated Methods for\\Textual Criticism}
\author{Matthew Gardner}
\date{\today}

\begin{document}
\maketitle

\section{Introduction}
\label{sec:intro}

Textual criticism can be loosely described as the analysis of a series of
manuscripts of a document (generally an ancient document) in an attempt to
reconstruct the original text produced by the author(s) of the document.
Because of the importance of religious documents and the proliferation of their
manuscripts, textual criticism has grown into a very large field, though its
use has not been confined solely to religious documents.  Both the New and Old
Testaments have been widely studied from a text critical perspective, leading
to great revisions in the proposed original texts for those books.  But other,
secular, works, such as the Canterbury Tales, the writings of William
Shakespeare, and a great number of classical texts such as the Illiad and the
Odyssey, have also received significant attention from textual critics.

The process of textual criticism involves laboring over differing manuscripts
searching for discrepancies, recording them by hand and sifting through the
various manuscripts in order to come to some conclusion about the manuscripts'
origins.  After deciding which manuscripts are most trustworthy, the textual
critic produces a text that he or she believes most closely matches the
original text, along with a ``critical apparatus'' containing variant readings
and justification for the editorial desicions made.

For hundreds of years, the methodology used by the textual critic was
painstaking and tedious, often involving travel to several countries to examine
manuscripts that were held by institutions spread throughout the world.  Bruce
Metzger, a prominent New Testament textual critic, described part of the
process as ``such an enormous amount of labor, out of all proportion to the
importance of the results, that scholars are content'' to make textual
decisions considering only a subset of the data available to
them~\cite[179]{metzger-1992-text-of-the-new-testament}.

With the advent of the computer, the work of textual criticism has started to
change.  A computer can now catalogue discrepancies in manuscripts, and even
try to come to a conclusion about which manuscripts were derived from other
manuscripts.  But while some work has been done in automating the process of
textual criticism, there are many areas in which computing resources and ideas
from computer science seem completely untapped.

This work attempts to document the current extent of the use automated methods 
in the field of textual criticism.  We start in \secref{methods} with an
overview of the basic methods of the textual critic, for without a basic
understanding of the methods one cannot proceed to automate them.  We then move
in \secref{previous-work} to a discussion of previous work and the state of the
field in automating textual criticism.  In \secref{proposed-ideas} we discuss
our own ideas for how to automate parts of the process that have not yet been
automated.  \secref{experiments} then presents some preliminary results of
experiments testing some of our ideas, and in \secref{conclusion} we conclude.

\section{Methods of Textual Criticism}
\label{sec:methods}

Much has been written over the years on the methodology involved in the process
of textual criticism.  Two of the foundational works on the subject were
written by Bruce M. Metzger and by Kurt and Barbara
Aland~\cite{metzger-1992-text-of-the-new-testament,
aland-text-of-the-new-testament}.  A more accessible introduction to the
lay-person was written recently by Bart D.
Ehrman~\cite{ehrman-2005-misquoting-jesus}.  His book, however, as it is a
popular and not a scholarly work, is lacking the rigor and depth found in the
books by Metzger and Aland, and he frequently dips into the philosophical
implications of the uncertainty in the textual tradition of the New Testament.

\subsection{Rules for Textual Criticism}

Aland, in his book, gives ``twelve basic rules for textual
criticism''~\cite[280--282]{aland-text-of-the-new-testament}.  Not all of his
rules are relevant to our present discussion, so we will summarize only eight
of them here, giving them new numbers:

\begin{enumerate}
  \item Only one reading can be original (i.e., the author did not produce two
	versions of the same text; an assumption which may in some cases be
	erroneous).
  \item In determining the original text, external evidence should be
	considered before internal evidence (we will discuss each type of evidence
	shortly).
  \item Texts in the original language are the primary authority; translations
	and quotations in other works are of less importance, especially when the
	translation back to the original language is ambiguous.
  \item Manuscripts should be weighed, not counted; the preponderance of a
	particular reading does not necessarily imply originality.
  \item It is unlikely that any one manuscript contains the original text in
	its entirety; more likely is that all manuscripts contain a mix of correct
	and incorrect readings, and only by using all of the evidence can the
	original text be obtained.
  \item The reading which can most easily explain the derivation of other forms
	is most likely the original, so stemmatology (the construction of a family
	tree for the manuscripts) is very important.
  \item The more difficult reading is the more probable reading, as a scribe is
	more likely to make the text easier, not more difficult.
  \item The shorter reading is the more probable reading.
\end{enumerate}

Outside of his rules, Aland also makes the point that, especially in the case
of the New Testament, it is incredibly likely that the original reading for any
variant passage is still intact in some manuscript, because when scribes were
confronted with differing manuscripts they would often just conflate both
readings.  Aland says, 
\begin{quotation}
the element of tenacity in the New Testament textual tradition not only permits
but demands that we proceed on the premise that in every instance of textual
variation it is possible to determine the form of the original text \ldots
assuming a proper understanding of the textual tradition.  The vast number of
witnesses to the text is not simply a burden---it is also a positive
aid.~\cite[294]{aland-text-of-the-new-testament}
\end{quotation}
Thus according to Aland, to obtain the original text we merely
need to list every place of variation and decide which of the extant variants
is the original.  No postulation of missing readings is necessary.

The rules make reference to external and internal evidence.  External evidence
is evidence that is outside of the text of the document itself.  For example,
which manuscripts attest a given reading, the reliability of those manuscripts,
and the age and location of the manuscript are all external criteria.  Internal
evidence includes considerations based on the text of the document, such as
what the author was likely to have written (based on stylistic and theological
considerations), and the probability that a scribe changed the reading from one
variant to another.

\subsection{Kinds of Variation in the Text}

Another important thing to understand in deciding between textual variants is
the manner in which variations creep into the manuscript tradition.  Ehrman
lists several different types of variants, categorized broadly into intentional
and non-intentional variations~\cite[90--98]{ehrman-2005-misquoting-jesus}.
Among the non-intentional kind of variations he lists the following:

\begin{itemize}
  \item spelling errors
  \item periblepsis (skipping some text because of similar line endings or
	beginnings)
  \item misunderstanding of abbreviations used in the text (the various names
	of God, among other words, were often abbreviated in the text)
  \item substitution of homophones (because texts were often copied by
	dictation, so that multiple scribes could copy a manuscript at the same
	time)
  \item changes in word order.  
\end{itemize}

His list of intentional variations includes:

\begin{itemize}
  \item ``fixing'' factual errors in the text (we put ``fixing'' in quotation
	marks, because the scribe may think he is correcting the document, but he
	is in fact changing the original text; the textual critic is interested in
	the original text, whether or not there were factual errors in it)
  \item ``fixing'' doctrinal or interpretive errors
  \item ``harmonizing'' similar passages in the text (this is especially
	applicable to the Gospels in the New Testament textual tradition)
\end{itemize}

\subsection{Knowledge Sources Needed}

In order to obtain the evidence required for determining the original reading
of a passage, there are several sources of knowledge that must be consulted.
If we are to automate the process of textual criticism, those knowledge sources
must be available to and interpretable by a computer.  Thus we discuss here all
of the kinds of knowledge that are necessary to the work of the textual critic.

The most basic knowledge source is the manuscript itself.  Each manuscript has
a date, a location, and text.  The location of the manuscript is trivially
determined upon its discovery, and the date is also fairly easily obtained by
archaeologists.  The text of the manuscript may be in dispute, if the
manuscript is incredibly corrupted or is incomplete, or if the text has been
written over, as in some of our oldest manuscripts (the uncial manuscript known
as 04 or C is called the ``Codex Ephraemi Syri Rescriptus,'' where the word
``rescriptus'' is added for that very
reason~\cite[109]{aland-text-of-the-new-testament}).  As far as computers
performing the work of textual criticism, it seems reasonable to assume a
faithful transcription for each manuscript, with the date and location as
metadata.  This knowledge source, then, contains almost all of the information
listed as ``external evidence.''

The one part of the external evidence not immediately apparent from the above
description of the manuscript is the reliability of the manuscript.  A
manuscript's reliability is determined by how often it contains correct
readings.  But at the same time, correct readings are often established in part
by the reliability of the manuscripts that contain them.  Clearly the process
is circular, or, as the world of computer science would put it, iterative.
Other knowledge sources must be used in order to establish the reliability of
manuscripts, and establishing reliability in an automated fashion must be an
iterative process.

Another item that should be mentioned is that manuscripts were often corrected
by a hand not belonging to the original scribe.  Those corrections provide more
evidence that can be used in textual criticism.  One must decide how to encode
that evidence.  The approach taken by all critical apparati that we observed
was to superscript the symbol for the corrected manuscript with a ``c'' for the
corrected version, and an asterisk for the original.  The two versions of the
manuscript could then be used as two separate manuscripts in an automated
system, though care would have to be taken to not double count their evidence
in passages where there was no correction.

The knowledge sources needed to judge the ``internal'' criteria are much more
complicated.  Determining what the author was likely to have written is not a
trivial task.  One must have a grasp of the author's writing style, vocabulary,
and theology, none of which can be ascertained perfectly, though various
approximations can and have been made.  A computer would have to make far more
drastic and unreasonable approximations than those a human makes, though some
parts of the knowledge source seem at least approachable in an automated
fashion, such as style and vocabulary.

Determining the probability with which a scribe introduces a variant reading
requires further knowledge, including an understanding of the way scribes
operated (which is why we included a section on the kinds of variations that
exist in the text).  Much of this information could be encoded into an
automated system without too much difficulty by simply specifying a prior
probability of each type of error occurring.  However, when considering whether
a scribe introduced a variant, the concept of an ``easier'' reading is referred
to, for scribes are more likely to make the reading easier than more difficult.
Those types of considerations seem at present beyond the capacity of automated
methods.

There are also other kinds of knowledge sources involved in the process of
textual criticism that are more analogous to algorithms than to data in the
world of computer science.  For example, knowing the history of the manuscript
tradition leads to decisions to model the manuscripts as a family tree, similar
to genetic evolution.  Such knowledge sources are encoded into an automated
system in the way the system is built.

Lastly, if one wants to use quotations of the text in writings of other
authors, or translations of the work into other languages, still more knowledge
sources are required.  For the simple case of a quotation, the extract text as
quoted could be considered its own ``manuscript.''  To use other-language
texts, translation back into the original is necessary, or some other way to
use the foreign-language text to aid in the decision between variant readings.
Perhaps the use of a bilingual dictionary would be useful.  It seems more
difficult, however, to propose a new variant reading from a translated version
of the text.

\section{Previous Work in Automating Textual Criticism}
\label{sec:previous-work}

An early survey of the history of computers in textual criticism was published
in 1994 by Robert
Kraft~\cite{kraft-1994-computers-in-new-testament-textual-criticism}.  For the
reader interested in the history, his is a good survey to read.  However, in
our current discussion we are interested in the present use of computers in
textual criticism, so we will defer further description of the history of the
field to that survey.  One line, however, seems germaine to the topic at hand:
``We have, in short, not come very far in realizing the promise offered by
computer-assisted research for NT textual criticism, although much of the
groundwork has been laid.''

Since 1994 significant progress has been made, though the vast majority of that
work has been in one small part of the process of textual criticism, that of
trying to produce a hierarchical relationship between each of the manuscripts
of a given text.  There is a very broad literature on the topic of constructing
trees of manuscripts, normally called either stemmatology or cladistics.  For
an example of how broad the field is, see
\cite{spencer-2002-exploring-textual-genealogy} and
\cite{van-reenen-2004-studies-in-stemmatology-ii}, including the articles they
cite.

The purpose of research into stemmatology seems to be to establish the
reliability of each manuscript.  If one can establish which manuscripts are
earliest in the tree, one can come to some conclusion about which manuscripts
are the most reliable.  A survey of the current state of the art in manuscript
stemmatology is beyond the scope of this paper; we suffice ourselves in noting
that significant research has been done in that area and to which aspects of
textual criticism the field applies.

There has also been some investigation into using clustering methods to group
manuscripts by similar readings~\cite{willker-pca-of-manuscripts-of-john}.  One
article we found even uses hierachical clustering to form trees of documents,
instead of the more common techniques taken from computational
biology~\cite{thorpe-2002-statistical-manuscript-classification}.  These
methods seem to reproduce rather well the ``families'' of manuscripts created
by textual critics to describe sections of the New Testament manuscript
tradition (for a description of the manuscript families, see
\cite[212--218]{metzger-1992-text-of-the-new-testament}).

A related, though somewhat distinct, use of computers in New Testament studies
is the attempt to determine authorship of documents using automated stylometry.
Such methods build some kind of model for the style of each particular author,
then try to attribute disputed documents to one of the authors in question.
\cite{alviar-2008-computational-linguistics-and-biblical-studies} is an example
of such a study.  The relation to textual criticism comes because stylistic
considerations are often involved in the determination of an original reading.
One would simply take the models that these methods produce and judge between
two variant readings, instead of entire books.

As far as we could tell, all previous work started with the digitized text of a
\emph{critical apparatus}, none of which are completely exhaustive, not the
text of the manuscripts themselves.  Thus all previous studies have used only a
subset of the possible data that they could have used; however, it is arguable
that the data that was not included would not have a significant effect on the
results of the algorithm.

The reason that previous work has not used the actual transcriptions of the
manuscripts is that such transcriptions are not widely available, and in many
cases do not exist in digital form at all.  There have been several calls for
the digitization of manuscripts, at least in image form, though none of them
have as yet been successful~\cite{trobisch-2000-central-electronic-database}.
One particularly eloquent summary comes from a 1999 paper by L. W. Hurtado.  He
says,
\begin{quotation}
To date, major obstacles remain in the way of these aims [the use of computers
in textual criticism], especially the necessity to transcribe the witnesses
into machine-readable form (a highly labour-intensive process and fraught with
the danger of introducing additional variants inherent in any copying process!)
and the need for fully adequate software able to collate and in some level
classify variants.

\ldots

Any significant uses of computer technology for the work of textual criticism
of the New Testament, uses that would materially simplify, expedite and improve
upon more traditional measures, are still unfortunately the stuff of
dreams.~\cite{hurtado-1999-beyond-the-interlude}
\end{quotation}

In his paper, Hurtado mentions the ``Electronic New Testament Manuscript
Project,'' an effort to collect digital images and transcriptions of all New
Testament manuscripts, as particularly hopeful.  It was in its infancy in 1999
when Hurtado published his paper, and it did not survive past infancy; hardly a
trace of the effort can be found today.

However, in their book, \emph{Jesus Christ and the World of the New Testament},
Holzapfel, Wayment and Huntsman mention that the majority of New Testament
manuscripts would be transcribed and available online by the end of
2010~\cite[?]{holzapfel-2006-jesus-christ-new-testament}.  Our extensive
searching produced only the uncial manuscripts for the Gospel of John, in a
format that was searchable, though not readily
downloadable~\cite{igntp-gospel-of-john-online}.  The project behind that
website, the International Greek New Testament Project (www.igntp.org), aims to
create a critical edition of the New Testament, and the publishing of those
manuscripts online seems to be merely incidental to their work.

Upon contacting Thomas Wayment in reference to his statement
in~\cite{holzapfel-2006-jesus-christ-new-testament}, we were informed that
BibleWorks, a company that has long produced software for those interested in
biblical studies, is in the process of digitizing the manuscripts.  A thread on
their user forums was of particular interest to the present
discussion~\cite{bible-works-forum}.  In the thread, Michael Bushell, the owner
of and lead programmer for the company, confirmed Wayment's statement that they
are in the process of digitizing the manuscripts, and that they hope to have
some progress to show by the end of 2010.  However, he says that the
transcriptions will only be made available with the purchase of BibleWorks
software initially, then ``eventually'' to ``organizations'' who will not
profit from the transcriptions.

Also in that thread we found the closest thing to what we are currently
investigating.  Twice Bushell mentions software that could ``generate apparati
on the fly,'' though in both instances he is doubtful that he will live to see
such software.  It would appear that at least one other person has the idea of
completely automating the process of textual criticism.  We found that idea to
be rather unique among everything we read; other authors spoke of automating
some part of the process, but nowhere else did we find someone even considering
generating an entire critical apparatus automatically.  Most authors seemed
skeptical of automated methods at best.  Future work in this area should
certainly involve contact with Michael Bushell.

\section{Proposed Ideas for the Automation of Textual Criticism}
\label{sec:proposed-ideas}

\section{Experiments}
\label{sec:experiments}

Our experiments involved sections of manuscripts of the Gospel of John
downloaded from the internet, at this site~\cite{igntp-gospel-of-john-online}
mentioned previously.  We aligned the text using a piece of software written
for aligning DNA and protein sequences in computational
biology~\cite{lee-2002-sequence-alignment-poa}.  This aligner performs a
character-level alignment, showing insertions, substitutions, and deletions for
each of the manuscripts.

An important consideration in the use of the aligner is to keep track of the
difference between an actual shorter reading in a text (perhaps because of a
later scribal addition) and a deficient manuscript that is simply missing
entire passages (because only part of the manuscript was found, for instance).
In the latter circumstance the manuscript is said to have lacunae (sing.
lacuna), or gaps.  One does not want to consider a lacuna to be evidence for a
shorter reading, nor ignore a legitimately shorter reading as if it were a
lacuna.

Given an alignment of the text, we can find locations where not all of the
manuscripts agree.  The aligner only gives character-level differences, but we
can easily extend the differences to word boundaries, and connect adjacent
differing blocks.  Upon performing this post-processing, we discovered 181
passages in the first chapter of John alone that contained variant readings,
consisting of 40\% of the text, with just nine manuscripts as evidence.  The
vast majority of these differences were a spelling error or word order change
introduced into one of the nine manuscripts, though some were more significant.

\section{Conclusion}
\label{sec:conclusion}

\bibliographystyle{plain}
\bibliography{../../../bib/textual-criticism/bib}

\end{document}

### 2010-09-27 textual_criticism/paper/paper.tex
Adding the final revision to the textual criticism paper that I never committed

@@
 original text, along with a ``critical apparatus'' containing variant readings
-and justification for the editorial desicions made.
+and justification for the editorial decisions made.
 
@@
 decisions considering only a subset of the data available to
-them~\cite[179]{metzger-1992-text-of-the-new-testament}.
+them~\cite[p. 179]{metzger-1992-text-of-the-new-testament}.
 
@@
 This work attempts to document the current extent of the use automated methods
-in the field of textual criticism.  We start in \secref{methods} with an
+in the field of textual criticism.  I start in \secref{methods} with an
 overview of the basic methods of the textual critic, for without a basic
-understanding of the methods one cannot proceed to automate them.  We then move
+understanding of the methods one cannot proceed to automate them.  I then move
 in \secref{previous-work} to a discussion of previous work and the state of the
-field in automating textual criticism.  In \secref{proposed-ideas} we discuss
-our own ideas for how to automate parts of the process that have not yet been
-automated.  \secref{experiments} then presents some preliminary results of
-experiments testing some of our ideas, and in \secref{conclusion} we conclude.
+field in automating textual criticism.  In \secref{proposed-ideas} I discuss my
+own ideas for how to automate parts of the process that have not yet been
+automated, and some preliminary results of experiments testing some of my
+ideas.  In \secref{conclusion} I conclude.
 
@@
 Aland, in his book, gives ``twelve basic rules for textual
-criticism''~\cite[280--282]{aland-text-of-the-new-testament}.  Not all of his
-rules are relevant to our present discussion, so we will summarize only eight
-of them here, giving them new numbers:
+criticism''~\cite[pp. 280--282]{aland-text-of-the-new-testament}.  Not all of
+his rules are relevant to the present discussion, so I will summarize only
+eight of them here, giving them new numbers:
 
@@
   \item In determining the original text, external evidence should be
-	considered before internal evidence (we will discuss each type of evidence
+	considered before internal evidence (I will discuss each type of evidence
 	shortly).
@@
 witnesses to the text is not simply a burden---it is also a positive
-aid.~\cite[294]{aland-text-of-the-new-testament}
+aid.~\cite[p. 294]{aland-text-of-the-new-testament}
 \end{quotation}
-Thus according to Aland, to obtain the original text we merely
-need to list every place of variation and decide which of the extant variants
+Thus according to Aland, to obtain the original text one merely
+needs to list every place of variation and decide which of the extant variants
 is the original.  No postulation of missing readings is necessary.
@@
 The rules make reference to external and internal evidence.  External evidence
-is evidence that is outside of the text of the document itself.  For example,
+is evidence that is outside of the text of the manuscript itself.  For example,
 which manuscripts attest a given reading, the reliability of those manuscripts,
 and the age and location of the manuscript are all external criteria.  Internal
-evidence includes considerations based on the text of the document, such as
-what the author was likely to have written (based on stylistic and theological
-considerations), and the probability that a scribe changed the reading from one
-variant to another.
+evidence includes considerations based on the text of the document (both the
+manuscript in question and the likely original text), such as what the author
+was likely to have written (based on stylistic and theological considerations),
+and the probability that a scribe changed the reading from one variant to
+another.
+
+It should also be noted that textual critics often use the word ``probability''
+when describing their rules, such as in saying that ``the shorter reading is
+more probable,'' or ``more likely.''  When they use those terms, however, they
+do not refer to actual probabilities in the strict, mathematical sense of the
+word.  In building an automated system, it might be fruitful to build a
+statistical model using the rules mentioned here as guidelines, and then the
+nebulous use of the word ``probability'' would be translated into something
+more concrete.
 
@@
 lists several different types of variants, categorized broadly into intentional
-and non-intentional variations~\cite[90--98]{ehrman-2005-misquoting-jesus}.
+and non-intentional variations~\cite[pp. 90--98]{ehrman-2005-misquoting-jesus}.
 Among the non-intentional kind of variations he lists the following:
@@
 \begin{itemize}
-  \item ``fixing'' factual errors in the text (we put ``fixing'' in quotation
-	marks, because the scribe may think he is correcting the document, but he
-	is in fact changing the original text; the textual critic is interested in
-	the original text, whether or not there were factual errors in it)
+  \item ``fixing'' factual errors in the text (I put ``fixing'' in quotation
+	marks because the scribe may think he is correcting or improving the text
+	of the manuscript he is copying, but he is in fact changing the original
+	text; the textual critic is interested in the original text, whether or not
+	there were factual errors in it)
   \item ``fixing'' doctrinal or interpretive errors
@@
 of a passage, there are several sources of knowledge that must be consulted.
-If we are to automate the process of textual criticism, those knowledge sources
-must be available to and interpretable by a computer.  Thus we discuss here all
+If one is to automate the process of textual criticism, those knowledge sources
+must be available to and interpretable by a computer.  Thus I discuss here all
 of the kinds of knowledge that are necessary to the work of the textual critic.
+
+\subsubsection{Sources of ``External'' Evidence}
 
@@
 as 04 or C is called the ``Codex Ephraemi Syri Rescriptus,'' where the word
-``rescriptus'' is added for that very
-reason~\cite[109]{aland-text-of-the-new-testament}).  As far as computers
-performing the work of textual criticism, it seems reasonable to assume a
-faithful transcription for each manuscript, with the date and location as
-metadata.  This knowledge source, then, contains almost all of the information
-listed as ``external evidence.''
+``rescriptus'' is added for that very reason~\cite[p.
+109]{aland-text-of-the-new-testament}).  As far as computers performing the
+work of textual criticism, it seems reasonable to assume a faithful
+transcription for each manuscript, with the date and location as meta data.
+This knowledge source, then, contains almost all of the information listed as
+``external evidence.''
 
@@
 evidence that can be used in textual criticism.  One must decide how to encode
-that evidence.  The approach taken by all critical apparati that we observed
+that evidence.  The approach taken by all critical apparati that I observed
 was to superscript the symbol for the corrected manuscript with a ``c'' for the
@@
 in passages where there was no correction.
+
+\subsubsection{Sources of ``Internal'' Evidence}
 
@@
 
+A possibility for using computers to make theological decisions is to create
+extensive markup about the theology contained in a text, so the computer would
+not have to automatically detect the theology.  Then the computer could decide
+if any particular variant agreed theologically with the rest of the text in
+question.  However, determining what kind of markup is sufficient and creating
+the markup for every text (and possibly every manuscript of the text) seems
+like far more work than simply having the human make the theological decision
+himself.
+
 Determining the probability with which a scribe introduces a variant reading
 requires further knowledge, including an understanding of the way scribes
-operated (which is why we included a section on the kinds of variations that
-exist in the text).  Much of this information could be encoded into an
-automated system without too much difficulty by simply specifying a prior
+operated (which is why I included a section on the kinds of variations that
+exist in the text).  Much of this information could be encoded for processing
+by an automated system without too much difficulty by simply specifying a prior
 probability of each type of error occurring.  However, when considering whether
@@
 methods.
+
+For instance, a prime example of a scribe making the reading easier is found in
+Mark 1:41~\cite[pp. 133--139]{ehrman-2005-misquoting-jesus}.  In the passage, a
+leper comes to Jesus seeking to be healed.  The majority of manuscripts
+available to us today say that Jesus had compassion on the man and healed him.
+However, there are some significant, very old manuscripts that instead say that
+Jesus became angry with the man, then healed him.  It is very easy to
+understand that Jesus had compassion, but much more difficult to explain anger.
+It is much more likely, then, the textual critic would say, that a scribe
+changed Jesus from being angry to showing compassion, not the other way around.
+How to get a computer to understand and use the knowledge required to make
+these judgments is beyond the current understanding of natural language
+processing.
+
+\subsubsection{Other Knowledge Sources}
 
@@
 authors, or translations of the work into other languages, still more knowledge
-sources are required.  For the simple case of a quotation, the extract text as
-quoted could be considered its own ``manuscript.''  To use other-language
-texts, translation back into the original is necessary, or some other way to
-use the foreign-language text to aid in the decision between variant readings.
-Perhaps the use of a bilingual dictionary would be useful.  It seems more
-difficult, however, to propose a new variant reading from a translated version
-of the text.
+sources are required.  For the simple case of a quotation, the extracted text
+as quoted could be considered its own ``manuscript,'' though certainly it
+should hold less weight than a real manuscript.  To use other-language texts,
+translation back into the original is necessary, or some other way to use the
+foreign-language text to aid in the decision between variant readings.  Perhaps
+the use of a bilingual dictionary would be useful.  It seems more difficult,
+however, to propose a new variant reading from a translated version of the
+text.
 
@@
 reader interested in the history, his is a good survey to read.  However, in
-our current discussion we are interested in the present use of computers in
-textual criticism, so we will defer further description of the history of the
-field to that survey.  One line, however, seems germaine to the topic at hand:
+the current discussion I am interested in the present use of computers in
+textual criticism, so I will defer further description of the history of the
+field to that survey.  One line, however, seems germane to the topic at hand:
 ``We have, in short, not come very far in realizing the promise offered by
@@
 work has been in one small part of the process of textual criticism, that of
-trying to produce a hierarchical relationship between each of the manuscripts
-of a given text.  There is a very broad literature on the topic of constructing
-trees of manuscripts, normally called either stemmatology or cladistics.  For
-an example of how broad the field is, see
-\cite{spencer-2002-exploring-textual-genealogy} and
-\cite{van-reenen-2004-studies-in-stemmatology-ii}, including the articles they
-cite.
+trying to produce a hierarchical relationship among the manuscripts of a given
+text.  There is a very broad literature on the topic of constructing trees of
+manuscripts, normally called either stemmatology or cladistics.  For an example
+of how broad the field is, see \cite{spencer-2002-exploring-textual-genealogy}
+and \cite{van-reenen-2004-studies-in-stemmatology-ii}, including the articles
+they cite.
 
@@
 earliest in the tree, one can come to some conclusion about which manuscripts
-are the most reliable.  A survey of the current state of the art in manuscript
-stemmatology is beyond the scope of this paper; we suffice ourselves in noting
-that significant research has been done in that area and to which aspects of
-textual criticism the field applies.
+are the most reliable.  It seems that the method at the forefront of New
+Testament stemmatology is called the ``Coherence-Based Genealogical Method,''
+presented in detail in an essay called ``Problems of a highly contaminated
+tradition: the New Testament: Stemmata of variants as a source of a genealogy
+for witnesses''~\cite[pp. 13--87]{van-reenen-2004-studies-in-stemmatology-ii}.
+This method is currently being used by the Institute for New Testament Textual
+Research at the University of M\"{u}nster to aid in the production of the text
+for the Editio Critica Maior (ECM), an important critical edition of the New
+Testament.
+
+The general idea of the method is that in a single manuscript the variant
+readings could have different ages of origin.  Thus individual variants are
+treated separately, and stemmatological decisions are made on the basis of an
+understanding of how the variants came about.  The paper also recognizes the
+circular nature of the reliability of a manuscript, and uses an iterative
+method to establish that reliability~\cite[p.
+25]{van-reenen-2004-studies-in-stemmatology-ii}.  One of the key insights that
+can be gained from this method is that a manuscript perhaps should not have a
+single measure of reliability; different sections of the same manuscript can
+have different levels of reliability.
+
+There has been much more work in the field of New Testament stemmatology than
+we have room or time to present here.  A broad and detailed survey of the field
+could easily be longer than the current study.  It seems sufficient to
+mention one of the most prominent methods in the field and move on to other
+areas of work in the automation of textual criticism.
 
@@
 manuscripts by similar readings~\cite{willker-pca-of-manuscripts-of-john}.  One
-article we found even uses hierachical clustering to form trees of documents,
+article I found even uses hierarchical clustering to form trees of documents,
 instead of the more common techniques taken from computational
@@
 by textual critics to describe sections of the New Testament manuscript
-tradition (for a description of the manuscript families, see
-\cite[212--218]{metzger-1992-text-of-the-new-testament}).
+tradition (for a description of the manuscript families, see \cite[pp.
+212--218]{metzger-1992-text-of-the-new-testament}).
 
@@
 
-As far as we could tell, all previous work started with the digitized text of a
+As far as I could tell, all previous work started with the digitized text of a
 \emph{critical apparatus}, none of which are completely exhaustive, not the
@@
 manuscripts would be transcribed and available online by the end of
-2010~\cite[?]{holzapfel-2006-jesus-christ-new-testament}.  Our extensive
-searching produced only the uncial manuscripts for the Gospel of John, in a
-format that was searchable, though not readily
-downloadable~\cite{igntp-gospel-of-john-online}.  The project behind that
-website, the International Greek New Testament Project (www.igntp.org), aims to
-create a critical edition of the New Testament, and the publishing of those
-manuscripts online seems to be merely incidental to their work.
+2010~\cite[p. 55]{holzapfel-2006-jesus-christ-new-testament}.  My extensive
+searching produced only two websites, both of which were difficult to use.  One
+contained the uncial manuscripts for the Gospel of John, in a format that was
+searchable, though not readily downloadable~\cite{igntp-gospel-of-john-online}.
+The project behind that website, the International Greek New Testament Project
+(www.igntp.org), aims to create a critical edition of the New Testament, and
+the publishing of those manuscripts online seems to be merely incidental to
+their work.  The other website was that of the The Institute for New Testament
+Research at the University of M\"{u}nster~\cite{munster-website}.   The site
+contains a large number of manuscripts, though not nearly all of them, and it
+is similarly difficult to download anything in bulk.  The website also states
+the intention of the institute to make available transcriptions of all of the
+manuscripts for which they have photographs, which is nearly all in existence.
 
 Upon contacting Thomas Wayment in reference to his statement
-in~\cite{holzapfel-2006-jesus-christ-new-testament}, we were informed that
+in~\cite{holzapfel-2006-jesus-christ-new-testament}, I was informed that
 BibleWorks, a company that has long produced software for those interested in
@@
 
-Also in that thread we found the closest thing to what we are currently
+Also in that thread I found the closest thing to what I am currently
 investigating.  Twice Bushell mentions software that could ``generate apparati
@@
 such software.  It would appear that at least one other person has the idea of
-completely automating the process of textual criticism.  We found that idea to
-be rather unique among everything we read; other authors spoke of automating
-some part of the process, but nowhere else did we find someone even considering
+completely automating the process of textual criticism.  I found that idea to
+be rather unique among everything I read; other authors spoke of automating
+some part of the process, but nowhere else did I find someone even considering
 generating an entire critical apparatus automatically.  Most authors seemed
@@
 
-\section{Proposed Ideas for the Automation of Textual Criticism}
+\section{Proposed Ideas and Experiments}
 \label{sec:proposed-ideas}
 
-\section{Experiments}
-\label{sec:experiments}
+\subsection{Generating Critical Apparati}
 
-Our experiments involved sections of manuscripts of the Gospel of John
-downloaded from the internet, at this site~\cite{igntp-gospel-of-john-online}
-mentioned previously.  We aligned the text using a piece of software written
-for aligning DNA and protein sequences in computational
-biology~\cite{lee-2002-sequence-alignment-poa}.  This aligner performs a
-character-level alignment, showing insertions, substitutions, and deletions for
-each of the manuscripts.
+Michael Bushell mentioned a software tool that would generate a critical
+apparatus given some user-specified criteria.  Through some experimentation, I
+have concluded that this software is trivial to write if the ``user-specified
+criteria'' includes a critical text.  Using out-of-the-box alignment
+software\footnote{The aligner I used was written for DNA and protein sequences,
+but worked reasonably enough with Greek text represented using Beta Code.  The
+aligner was the result of a paper published in \emph{Bioinformatics}, and code
+is freely available on the internet~\cite{lee-2002-sequence-alignment-poa}.  I
+am indebted to Scott Lloyd, a PhD student in computer science at BYU, for
+pointing me to this software.  The aligner performs a character-level
+alignment, showing insertions, substitutions and deletions for each of the
+manuscripts.}, in a matter of days I was able to produce code that is the basis
+for generating a critical apparatus.  From an alignment of the manuscripts, one
+can easily find passages that differ across the manuscripts, and if one of them
+is labeled as the critical text to use, all that remains to do is format the
+output in some recognizable form (i.e., output the critical text with footnotes
+that describe which manuscripts contain variant readings).
 
@@
 
-Given an alignment of the text, we can find locations where not all of the
-manuscripts agree.  The aligner only gives character-level differences, but we
-can easily extend the differences to word boundaries, and connect adjacent
-differing blocks.  Upon performing this post-processing, we discovered 181
+The output of the alignment software looked as follows\footnote{In using the
+software I substituted underscores for spaces, as the software did not allow
+spaces in the sequence.  I also removed all punctuation and diacritical marks,
+except those indicating abbreviations.  The spaces are restored here for easier
+interpretation of the output}:
+
+\begin{quotation}
+{\tt
+john/005.txt        ---------------------------------------------------
+
+john/003.txt        o pemyas me baptizein en ---udati ekeinos moi eipe-
+
+john/001.txt        o pemyas me baptizein en tw udati ek-inos moi eipen
+
+john/009.txt        o pemyas me baptizein en ---udati ekeinos moi eipen
+
+john/007.txt        o pemyas me baptizein en ---udati ekeinos moi eipen
+
+john/004.txt        o pemyas me baptizein en ---udati ekeinos moi eipen
+
+john/013.txt        o pemyas me baptizein en ---udati ekeinos moi eipen
+
+john/011.txt        o pemyas me baptizein en ---udati ekeinos moi eipen
+
+john/002.txt        o pemyas me baptizein en ---udati ekeinos moi eipen
+}
+\end{quotation}
+
+This is part of the text for John 1:33, translating roughly to ``he who sent me
+to baptize in water, that man said to me.''  The first column in each line is
+the name of the sequence that was given to the aligner; I just used the
+filename, where the filename was the number commonly given to the manuscript.
+Hyphens indicate a deletion in a manuscript, or an insertion in one of the
+other manuscripts.  In this section of text, we can see that manuscript 5 had a
+lacuna (the marker that I used for lacunae is found prior to this section,
+however), manuscript 1 had a word addition (``the water'' instead of ``water'')
+and misspelled \emph{ekeinos} (``that man''), manuscript 3 left out a nu on
+\emph{eipen} (``he said''), and all of the other manuscripts agreed on the
+text.
+
+Given this output, I wrote code that parsed it and presented lists of variant
+passages along with which manuscripts attested which readings.  An example
+passage follows.
+
+\begin{quotation}
+{\tt
+Location: (4731, 4737)
+
+uios: [`john/003.txt', `john/002.txt']
+
+us$\sim$: [`john/001.txt', `john/009.txt', `john/007.txt', `john/013.txt']
+
+xs$\sim$: [`john/011.txt'] 
+}
+\end{quotation}
+
+The location shows the character positions this reading spanned (it includes
+the spaces surrounding the word; thus there are six characters in the reading
+instead of four).  The $\sim$ symbol indicates that an abbreviation mark was
+found in the manuscript.  \emph{uios}, meaning ``son,'' was abbreviated as
+\emph{us} with an over-bar.  The abbreviation \emph{xs} means ``Christ.''  Thus
+manuscript 11 replaced the word son with the word Christ, and all other
+manuscripts that do not have a lacuna in this area contain the same reading,
+though some are abbreviated.  This particular substitution is not very
+surprising, because in capital Greek characters it is the change from a
+character much like our ``Y'' to a character much like our ``X''---a very small
+difference in appearance and a similarly small change in meaning.
+
+If one of the manuscripts were marked as the critical text to use, from this
+list of variants it is a simple task to output the critical text and mark all
+other readings as footnotes in a critical apparatus.
+
+It seems, then, that Michael Bushell's hope of automatically generated critical
+apparati is not very difficult, as I cannot imagine that he also intends to
+generate the critical text on the fly, unless there is some very serious markup
+added to the variant readings (including information on theology).  The main
+obstacle is the transcription of the manuscripts.  The main benefit of such an
+automatically generated apparatus is that it can be completely exhaustive.
+
+\subsection{Generating a Critical Text}
+
+If one could generate a critical text in a completely automated fashion, I have
+already shown that the critical text can be used to easily construct an
+apparatus.  Thus the work of textual criticism would be complete, with no more
+need for textual critics.  However, given the incredibly complicated nature of
+some of the knowledge sources needed for the work of textual criticism, it is
+extremely unlikely that a reliable critical text will be automatically
+generated any time in the foreseeable future.  As Dr. Eric Ringger, an
+assistant professor of computer science at BYU, put it, determining whether a
+passage matches the theology of an author is ``AI complete''---solving that
+problem amounts to solving all of artificial intelligence.  But we can make
+some faltering attempts.
+
+My experiments involved sections of manuscripts of the Gospel of John
+downloaded from the internet, at the site previously
+mentioned~\cite{igntp-gospel-of-john-online}.  I limited my exploration to the
+first chapter of John and the nine earliest manuscripts available on the site
+(this was because the website provided no bulk download of the data, so each
+chapter of each manuscript had to be copied and pasted and pre-processed by
+hand; another evidence of the sparsity of data available in this field).  I
+used the alignment software previously mentioned to discover where variant
+readings existed in the manuscripts.
+
+As mentioned previously, according to Aland, all one must do in order to
+establish the critical text of the New Testament is to decide between each of
+the extant readings for every passage in which there are variants.  Upon
+aligning the manuscripts and finding variant passages, I discovered 187
 passages in the first chapter of John alone that contained variant readings,
-consisting of 40\% of the text, with just nine manuscripts as evidence.  The
-vast majority of these differences were a spelling error or word order change
-introduced into one of the nine manuscripts, though some were more significant.
+consisting of 40\% of the text, with just nine manuscripts as evidence.  That
+gave me some idea of the enormity of this task, which I had not previously
+appreciated.  The vast majority of these differences were a spelling error or
+word order change introduced into one of the nine manuscripts, though some were
+more significant.  Having obtained a list of variants, there were several
+methods I experimented with to shorten the list.
+
+\subsubsection{Abbreviations, Spelling Errors and Shortest Readings}
+
+A simple place to start in identifying easy-to-correct variant passages is to
+look for words that are abbreviated.  In addition to the word for ``Son'' seen
+previously, there are many other places where abbreviations were commonly used,
+including the words for ``God,'' ``man,'' ``Jesus,'' and ``Israel.''  I made a
+list of those words and their abbreviations, and upon finding a variant whose
+only readings were abbreviations of the longer form of the same word, I marked
+it as ignorable.  However, such efforts only shortened the list of variants by
+6, to 181.
+
+A slightly more complicated type of variant to check for, but still relatively
+simple, is spelling errors.  I built a dictionary by using a parsed version of
+the Perseus project~\cite{perseus-project}.  I downloaded all of the text for
+some 120 classical Greek works, including such authors as Sophocles, Xenophon,
+and Homer.  To build the dictionary, I simply made a set containing all of the
+words those authors used.  I then checked every word in each reading against
+the dictionary and marked spelling errors.  If it was clear that there was only
+one variant that had correct spelling, I marked it as the correct text.  That
+effort took the number of variants to around 150.
+
+Grammatical checking is the next logical step, for frequently there are errors
+where the article in front of a word does not match the word's case in one of
+the manuscripts, among other simple errors.  However, I did not have access to
+a Greek grammatical parser or checker and am not proficient enough in ancient
+Greek to write one, so I was not able to try this method.  I am confident that
+it would greatly reduce the number of variant passages left to decide.
+
+Moving to methods less solidly grounded in knowledge sources available to a
+computer, one can also look for areas where words were added, applying the rule
+that the shorter reading is the more likely rule.  However, Aland is quick to
+say after stating the rule that it cannot be mechanically applied without
+consideration of other evidence; thus it gets very difficult for a computer to
+decide which reading is most likely to be correct.  Finding out which reading
+is shortest is trivial on a computer; deciding when to accept the shorter
+reading and when to accept the longer reading is much harder.  In my
+experiments I detected which variants looked like they had word additions,
+though I was unable to come to any automatic decision about which reading to
+keep.
+
+\subsubsection{Combining Manuscript Evidence}
+
+The biggest issue I encountered in trying to apply the ``shorter reading'' rule
+was that of combining the evidence from various manuscripts in some reasonable
+way.  As Aland's fourth rule listed above says, manuscripts should be weighed,
+not counted.  Thus some measure of reliability is required, along with a way to
+combine ``votes'' from various manuscripts and their reliabilities in order to
+come to a conclusion about the original reading.  As mentioned in the
+discussion of knowledge sources, the concept of reliability is somewhat
+circular, because manuscripts are reliable when they contain correct readings,
+and correct readings are in turn determined by the reliability of the
+manuscripts that attest them.  This lends itself naturally to some iterative
+process in an automated algorithm.
+
+However, even given a set of reliability scores, I was unable to determine a
+reasonable way to combine the evidence of the various manuscripts.  A simple
+majority vote weighted by reliability does not seem reasonable, as a poor
+manuscript could have hundreds of copies; even if those poor copies have very
+small weight, they could easily outweigh a single, much better manuscript.  A
+simple average also does not seem reasonable, for sometimes even the most
+reliable manuscripts are wrong; this method would fail miserably when the most
+reliable manuscript is the only manuscript attesting a particular incorrect
+reading.  Clearly there needs to be some more complicated way of combining the
+evidence, though exact methods for such a procedure are elusive to me.
+Perhaps a professional textual critic could give some insight that would be
+helpful in this area.
+
+\subsubsection{Establishing Reliability of Manuscripts}
+
+Even without a good method for combining manuscript evidence, one can still
+compute some initial reliability scores based on the critical decisions that
+have been made.  I calculated a simple score based on the spelling errors and
+word insertions I observed.  The values of the score were arbitrary; I started
+each manuscript with a base score of 100 and subtracted a small number of
+points for each spelling error, and a much larger number of points for each
+passage that had word insertions.  I weighted the number of points subtracted
+by the length of the manuscript, because some manuscripts contained only
+fractions of the text, and thus had less opportunity to lose reliability,
+leading to scores that were disproportionately high.
+
+The results of the initial reliability scores were somewhat interesting.  The
+manuscript commonly referred to as 5 had the highest score, largely because it
+only contained one fourth of the first chapter of John (I was not able to
+completely erase the higher score given to shorter manuscripts).  However, the
+next highest manuscript in reliability was 1 (Codex Sinaiticus), followed by 3
+(Codex Vaticanus); those two manuscripts are commonly considered by textual
+critics to be the most reliable manuscripts that we have, in that order.  So a
+simple reliability measure affected almost entirely by addition of words in
+just the first chapter of John reproduced to some extent the ordering
+determined by professional textual critics.
+
+The results I saw could very possibly be entirely accidental.  However, what
+one can observe is that given a set of critical decisions about passages in the
+text, a reliability score can easily be assigned to each manuscript.  There is
+some question about how best to calculate the score, but a textual critic could
+readily give some basic guidelines as to which kinds of variants affect
+reliability the most, or even mark for each individual variant how much a
+correct reading in that variant should matter in the reliability score.  That
+score could then be used in other parts of an automated system for textual
+criticism, including an iterative update to the reliability if a way of
+combining evidence is decided upon.
+
+\subsection{An Aid to the Textual Critic}
+\label{sec:aid-to-critic}
+
+Through my experiments in trying to produce a critical text, I came to the
+conclusion that at least at present the place for computers in the work of
+textual criticism is as an aid to, not a replacement of, the textual critic.
+There are several ways in which computers can aid in the process.
+
+The first and most obvious benefit is the automatic detection of variants in
+the manuscripts.  This is clearly feasible, given transcriptions of the
+manuscripts, as I demonstrated in my code.  The automatic detection of variants
+allows for completely exhaustive critical apparati, which as of yet do not
+exist.
+
+The second benefit that computers can bring to textual criticism is the
+filtering of variants, catching and marking easily decidable variants such as
+spelling errors.  With much of the noise removed, the textual critic could
+focus his attention on the important passages.  I was only able to filter
+spelling errors, but I imagine a grammatical check could detect other easily
+decidable errors, and a textual critic could probably list a few more
+possibilities.
+
+The third benefit that I see is the automated calculation of a reliability
+score.  Given some criteria specified by the textual critic, such as how to
+weight different types of variants, a computer could return a reliability for
+each manuscript which could be used in later decisions.  I can envision a
+system where a list of variants for a passage is presented on the screen,
+ordered by the reliability of the manuscripts that contain the variant (and
+perhaps some other criteria).  The critic would then select a variant as the
+correct one, and the computer would recompute reliability scores and show
+another variant.  Something related to active learning could even be used in
+the process, with the computer selecting which variant to show next.
+
+In short, I am convinced that while the automated generation of a critical text
+is currently far beyond our reach, automated methods could greatly reduce the
+workload on the textual critic.  What Metzger called ``an enormous amount of
+labor'' would all be done by the computer, leaving the text critic himself to
+focus on more important things.
 
@@
 \label{sec:conclusion}
+
+The purpose of this project was to explore the current extent of automated
+methods in textual criticism, and what room there is for the increased use of
+automated methods.  In doing this project, I have learned a great deal about
+textual criticism itself, reading books and portions of books on the subject.
+I also have had an enlightening experience attempting to automate myself some
+of the processes involved in textual criticism.
+
+I have reviewed the literature extensively, and inquired of BYU professors
+involved in textual criticism, and come to the conclusion, along with many
+others in the field, that the field of textual criticism could be greatly
+enhanced by the more effective use of computing resources.  It has come a long
+way in recent years, particularly in the application of methods from
+computational biology to reconstructing manuscript trees.  But much more work
+can yet be done.
+
+Lastly, I have experimented with specific methods to aid in the work of the
+textual critic, and proposed some future work that I think would be very
+beneficial to textual criticism.  Computers can remove much of the mundane from
+current methodologies, allowing researchers to focus their efforts on critical
+tasks.
+
+\section{Future Work}
+
+One obvious next step in this work is to produce a system like that described
+in \secref{aid-to-critic}.  The code I produced could be a starting place, but
+significant work would need to be done on top of my code, including a mechanism
+for accepting input from a user, improvements to the reliability score
+calculation, the addition of some kind of grammatical checker to filter the
+variant list, and some kind of active-learning decision about which variants to
+show the user.  The output from a stemmatological analyzer could also likely be
+incorporated into the code, probably in the calculation of the reliability
+score.  Such work would best be done in conjunction with a professional textual
+critic, though who exactly would be most interested in such a project is
+unknown to me.  As mentioned previously, Michael Bushell of BibleWorks would
+likely be a good person to contact.
+
+The helpful input that could be had from a textual critic has been mentioned
+throughout this paper; in brief summary, it includes specific advice about how
+to determine how reliable a manuscript is, how to combine evidence from many
+manuscripts, and what other kinds of variants are easily decidable.  Certainly
+it would also be good to know if a tool such as the one described would be
+found useful.  In order to answer these questions, there are several professors
+at BYU who may be good to talk to.  I found it somewhat difficult to interview
+the faculty, as many of them are away for the summer.  I did receive responses
+to my emails from two faculty members (Thomas Wayment and Stephen Bay).  While
+both were helpful in answering my questions, neither seemed particularly
+excited about my project.  Perhaps seeing this report containing some concrete
+ideas for work in automating textual criticism may prove more effective in
+gaining interest.
+
+Future work could also more fully explore the ideas used in the
+``Coherence-Based Genealogical Method'' for stemmatology.  While the work
+focuses on producing a stemma of the text, they also are interested in
+producing a critical text in a semi-automated fashion, and some of their ideas
+are applicable to this work.  I came across the work too late in my project to
+allow for very much exploration.
+
+Another idea that could be explored in automating the process is that of
+creating some kind of statistical model for the copying process, then running
+inference in order to determine the original text.  Such a model would be very
+complicated, and I do not have many ideas on how to proceed in that area.  I
+suspect that the model would have severe deficiencies because of the difficulty
+of modeling the ``easiness'' of a reading.
 

### 2010-09-28 honors_thesis/gecco_paper/paper.tex

@@
   \psfrag{pt}[C][C]{$\p_{t}$}
-  \includegraphics[width=.3\columnwidth]{messages1.eps}
-  \caption{The messages needed to produce $\p_{t}$.  Items without an arrow
-  pointing to them are messages the particle needs to receive.}
+  \includegraphics[width=.3\columnwidth]{messages1}
+  \caption{The production of $\p_{t}$ from the original particle
+  $\nonbest{\p}_{t}$ and the messages $\nonbest{\nset}_{t}$.}
   \label{fig:messages1}
@@
   \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
-  \includegraphics[width=.5\columnwidth]{messages2.eps}
-  \caption{The messages needed to produce $\nonbest{\p}_{t+1}$.  Items without
-  an arrow pointing to them are messages the particle needs to receive.}
+  \includegraphics[width=.5\columnwidth]{messages2}
+  \caption{The production of $\nonbest{\p}_{t+1}$ from the original particle 
+  $\nonbest{\p}_{t}$, messages $\nonbest{\nset}_{t}$ and
+  $\nonbest{\sset}_{t+1}$, and intermediate particles.}
   \label{fig:messages2}
@@
   \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
-  \includegraphics[width=.6\columnwidth]{messages3.eps}
-  \caption{The messages needed to produce $\nonbest{\nset}_{t+1}$.  Items
-  without an arrow pointing to them are messages the particle needs to receive.
-  $\nnset$ is the set of neighbors for each particle $\n$, and $\nsset$ is the
-  set of $\n$'s speculative children.  Note the similarity between this and
-  \fig{messages2}.}
+  \includegraphics[width=.6\columnwidth]{messages3}
+  \caption{The production of each $\nonbest{\n}_{t+1}$ from the original
+  particle $\nonbest{\n}_{t}$, messages $\nonbest{\nnset}_{t}$ and
+  $\nonbest{\nsset}_{t+1}$, and intermediate particles.  $\nnset$ is the set of
+  neighbors for each particle $\n$, and $\nsset$ is the set of $\n$'s
+  speculative children.  Note the similarity between this and \fig{messages2}.}
   \label{fig:messages3}
@@
   \psfrag{pt1}[C][C]{$\p_{t+1}$}
-  \includegraphics[width=.8\columnwidth]{messages4.eps}
-  \caption{The messages needed to produce $\p_{t+1}$.  Items
-  without an arrow pointing to them are messages the particle needs to receive.
-  Note that this is just a combination of \fig{messages2} and \fig{messages3}.}
+  \includegraphics[width=.8\columnwidth]{messages4}
+  \caption{The production of $\p_{t+1}$ from the original particle 
+  $\nonbest{\p}_{t}$, messages $\nonbest{\nset}_{t}$, $\nonbest{\sset}_{t+1}$,
+  $\nonbest{\nnset}_{t}$, and $\nonbest{\nsset}_{t+1}$, and intermediate
+  particles.  Note that this is just a combination of \fig{messages2} and
+  \fig{messages3}.}
   \label{fig:messages4}
@@
   \psfrag{st3-e}[C][C]{$\noeval{\sset}_{t+3}$}
-  \includegraphics[width=.5\columnwidth]{messages5.eps}
-  \caption{The messages needed to produce $\noeval{\sset}_{t+3}$. $\p_{t+1}$
-  and $\nset_{t+1}$ are both produced as in \fig{messages4}.}
+  \includegraphics[width=.5\columnwidth]{messages5}
+  \caption{The production of $\noeval{\p}_{t+2}$ and $\noeval{\sset}_{t+3}$
+  from $\p_{t+1}$ and $\n_{t+1}$, each of which are produced as in
+  \fig{messages4}.}
   \label{fig:messages5}
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank1.eps}
+  \includegraphics[width=.8\columnwidth]{griewank1}
   \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
-  iterations per round, while standard PSO algorithms use 240 processors and do
-  one iteration per round.}
+  evaluations, showing percent successful by each iteration instead of average
+  function value.}
   \label{fig:griewank1}
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank2.eps}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.  Instead of showing average function value, we show the
-  percentage of runs that have found the global optimum by each iteration.}
+  \includegraphics[width=.8\columnwidth]{griewank2}
+  \caption{Function Griewank with a swarm that uses 800 processors per round of
+  evaluations.}
   \label{fig:griewank2}
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
+  \includegraphics[width=.8\columnwidth]{rastrigin}
   \caption{Function Rastrigin with a swarm that uses 240 processors per round
-  of evaluations.  Speculative algorithms have a swarm of 30 particles and do
-  two iterations per round, while standard PSO algorithms use 240 processors
-  and do one iteration per round.}
+  of evaluations.}
   \label{fig:rastrigin}

### 2010-09-28 honors_thesis/tec_paper/paper.tex
Renamed journal_paper and made a copy for the swarm intelligence version

--- 2010-06-09 honors_thesis/journal_paper/paper.tex
@@
 
+\begin{keywords}
+Parallel algorithms, optimization methods, particle swarm optimization.
+\end{keywords}
+
 \section{Introduction}
@@
 available, these techniques increase the number of particles in the swarm,
-either by adding individual particles or by adding entire new sub-swarms, but
-in all cases the number of processors never exceeds the number of particles.
-The number of iterations of PSO that the algorithms can perform is thus
-inherently limited by the time it takes to evaluate the objective
-function---additional processors add more particles, but do not make the
-iterations go any faster.
-
-For many functions there comes a point of diminishing returns with respect to
-adding particles.  In \fig{evals-sphere} we show the value obtained after
-50,000 function evaluations (not iterations) as a function of swarm size for
-the well-know benchmark function Sphere (20 dimensions, reporting the average
-of twenty runs).  Increasing the swarm size from 5 to 10 has a significant
-effect on the value obtained.  However, increasing the swarm size from 16 to 30
-makes the algorithm less efficient, that is, it reduces the progress the
-algorithm makes per evaluation.  Other functions show similar trends, though
-often the optimal swarm size is slightly larger.  For this reason, previous
-work has recommended the use of a swarm size of 50 for
+either by adding individual particles or by adding entire new sub-swarms.  In
+almost all cases, adding additional particles produces better results in the
+same amount of time.  In \fig{iters-sphere} we see an example of this on the
+well-known benchmark function Sphere (20 dimensions, reporting the average of
+twenty runs).  In terms of the number of iterations performed (which is
+equivalent to wall-clock time if all particles are evaluated in parallel),
+every time the swarm size increases, the performance improves.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{iters_sphere}
+  \caption{Function Sphere with various swarm sizes, comparing performance with
+  the number of iterations of the algorithm performed.}
+  \label{fig:iters-sphere}
+\end{figure}
+
+However, it can be seen from the graph that doubling the swarm size produces
+only incremental benefit; performance improves, but not by very much.  The
+reason for this is that there comes a point of diminishing returns with respect
+to adding particles.  In \fig{evals-sphere} we show the value obtained after
+50,000 function evaluations (not iterations) as a function of swarm size, again
+for the function Sphere.  Increasing the swarm size from 5 to 10 has a
+significant effect on the value obtained.  However, increasing the swarm size
+from 16 to 30 makes the algorithm less efficient; that is, it reduces the
+progress the algorithm makes per evaluation.  Other functions show similar
+trends, though often the optimal swarm size is slightly larger.  For this
+reason, previous work has recommended the use of a swarm size of 50 for
 PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
@@
   \includegraphics[width=.8\columnwidth]{evals_sphere}
-  \caption{Function Sphere with various swarm sizes. Error bars show median and
+  \caption{Function Sphere with various swarm sizes, comparing performance with
+  the number of function evaluations performed.  Error bars show median and
   10th and 90th percentiles.}
@@
 purpose is to explore the question of what to do with a thousand processors
-when 50 or 100 particles is the recommended swarm size, and simply increasing
-the swarm size encounters diminishing returns.
+when 50 or 100 particles is the most efficient swarm size, and simply
+adding particles results in only incremental improvement.
 
@@
 in the literature, which in graph theory refers to a completely different
-topology.  Other names have also been used, including ``global topology'' or
+topology.  Other names have also been used, including ``global topology'' and
 gbest.  We use the graph theory term ``Complete'' in this paper.}, every
@@
 
-In distributed frameworks, sychronizing all of the machines for a round of
+In distributed frameworks, synchronizing all of the machines for a round of
 message passing can be expensive.  The method just described uses three rounds
@@
 
+At this point it is also interesting to note a parallel between our methods and
+parallel evolution
+strategies~\cite{rudolph-1991-distributed-evolution-strategies}.  In evolution
+strategies, a parent individual (representing a potential solution to some
+objective function) produces a number of offspring by a mutation operator.  One
+of the individuals is selected by a selection operator, and that individual
+becomes the parent for the next
+generation~\cite{beyer-2002-evolution-strategies}.  Our methods are similar,
+where our mutation operator is simply the PSO motion equations and the
+selection operator is either the indicator function introduced in \sec{sepso},
+in the case of our original speculative algorithm, or the standard selector
+operator based on fitness, in the case of this Pick Best technique.
+
 \subsection{Pruning the Speculative Evaluations}
@@
 \subsection{Branch Statistics}
+\label{sec:statistics}
 
@@
 the averages of 3.6 million trials for the branch taken (2.16 million for the
-two with only 450 iterations).  Table~\ref{tab:stats} contains the results.  We
-have already defined the functions Sphere, Griewank, and Rastrigin.  Rosenbrock
-is defined as $f(\Vec{x}) = \sum_{i=1}^D
-\left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.
+two with only 450 iterations).  Table~\ref{tab:stats} contains the results.
+The definitions for all of the functions in the table are found in \sec{setup}.
 
@@
 perform one function evaluation for one particle, be it speculative or
-non-speculative.  We acknowledge that for benchmark functions with very fast
-evaluations this may not be the best use of parallelism in PSO.  But for many
-real-world applications, the objective function takes on the order of minutes
-or more to evaluate; in such cases our choice of framework is reasonable.
-
-The speculative algorithm actually performs two iterations of PSO at each
-``iteration,'' so we instead call each ``iteration'' a ``round of
+non-speculative.  The speculative algorithm actually performs two iterations of
+PSO at each ``iteration,'' so we instead call each ``iteration'' a ``round of
 evaluations.''
 
-To use the same number of processors (and hence evaluations) for each
-algorithm, speculative evaluation must use a smaller swarm size than standard
-PSO.  For each of the topologies we used, a particle has three neighbors
-including itself.  As shown in Table~\ref{tab:evals}, this results in $7$
-speculative evaluations per particle.  With one evaluation needed for the
-original, non-speculative particle, we have $8p$ evaluations for every two
-iterations, where $p$ is the number of particles in the speculative swarm.
-The extra evaluations done in speculative evaluation would instead be particles
-in regular PSO, so we compare swarms of size $p$ in speculative evaluation
-with swarms of size $8p$ in regular PSO.
+Because our interest is in exploring the most efficient use of processors, we
+must hold fixed the number of processors used at each round of evaluations in
+all of our experiments, giving all algorithms the same number of function
+evaluations.  Thus while we label our graphs with ``Rounds of Evaluations,''
+this is equivalent to the number of function evaluations used by each algorithm
+(simply multiply by the number of processors to transform rounds of evaluations
+to function evaluations).  We use ``Rounds of Evaluations'' to reflect
+wall-clock time, as our primary interest is using extra processors in parallel
+to improve performance on functions with long evaluation times.  A smaller
+swarm running serial PSO would likely produce better results than we present in
+the same number of function evaluations, but because the function evaluations
+were not done in parallel it would take more wall-clock time.  When comparing
+the performance of parallelized algorithms on long-running functions, we must
+compare by wall-clock time with the same resources, not solely by the number of
+function evaluations.
+
+Using the same number of processors for each algorithm requires that our
+speculative algorithms have a smaller swarm size than standard PSO.  For each
+of the topologies we used, a particle has three neighbors including itself.  As
+shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
+particle.  With one evaluation needed for the original, non-speculative
+particle, we have $8p$ evaluations for every two iterations, where $p$ is the
+number of particles in the speculative swarm.  The extra evaluations done in
+speculative evaluation would instead be particles in regular PSO, so we compare
+swarms of size $p$ in speculative evaluation with swarms of size $8p$ in
+regular PSO.
 
@@
 standard PSO algorithm using both the Random topology with two neighbors and
-the Complete topology.
+the Complete topology (labeled PSO Random and PSO Complete in our results).
 
@@
 x_{i+1})+.7)\]
+
+In computing the branch statistics in \sec{statistics}, we also used the
+Rosenbrock function.  That function is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)\]
 
@@
 ../../../bib/nfl/bib,%
+../../../bib/optimization/bib,%
 ../../../bib/pso/general/bib,%

### 2010-10-02 honors_thesis/si_paper/paper.tex
I made some revisions to the SI paper according to feedback from TEC

--- 2010-09-28 honors_thesis/tec_paper/paper.tex
@@
 particles to the swarm.  However, in many cases this is not efficient when
-scaled to very large swarm sizes (on very large clusters).  In this paper we
-present a different way of using additional processors in the computation of
-PSO motivated by speculative execution in modern computer hardware.  We call
-our method Speculative Evaluation in PSO (SEPSO).
-
-In SEPSO future positions of the particles are speculated and evaluated in
-parallel with the evaluation of current positions.  We show that a set of
-speculative evaluations can be constructed such that the set will always
-include the position and corresponding evaluation that would normally have
-occurred at the next iteration of standard PSO.  Thus SEPSO performs two
-iterations of PSO at once, allowing it to achieve better fitness faster.
-
-Speculative Evaluation in PSO can be enhanced by diverging from the behavior of
-the standard PSO.  We present several variations on our speculative algorithm,
-such as making use of the information obtained by the extra speculative
-evaluations, and speculating more than one iteration ahead.  Using these
-algorithmic modifications, we see dramatic improvements in the effectiveness of
-the algorithm.
+scaled to very large swarm sizes (on very large clusters).  Current methods
+cannot answer well the question, ``What do you do with 1000 processors when 50
+or 100 particles is the most efficient swarm size?''  In this paper we attempt
+to answer that question with a speculative approach to the parallelization of
+PSO motivated by speculative execution in modern computer hardware.
+
+In our approach, we refactor PSO such that the computation needed for iteration
+$t+1$ can be done in parallel with the computation needed for iteration $t$.
+Thus we can perform two iterations of PSO at once.  Even with some amount of
+wasted computation, we show that this approach to parallelization in PSO often
+outperforms the na\"ive parallelization of simply adding particles to the
+swarm.  Our methods produce results that are exactly equivalent to PSO; this is
+not a new algorithm or variant, only a new method of parallelization.
+
+However, given this new parallelization model we can relax the requirement of
+exactly reproducing PSO in an attempt to produce better results.  We present
+several such relaxations, including keeping the best speculative position
+evaluated instead of the one corresponding to PSO, and speculating several
+iterations ahead instead of just one.  We show that these methods dramatically
+improve the performance of parallel PSO in many cases.
 
@@
 almost all cases, adding additional particles produces better results in the
-same amount of time.  In \fig{iters-sphere} we see an example of this on the
-well-known benchmark function Sphere (20 dimensions, reporting the average of
-twenty runs).  In terms of the number of iterations performed (which is
-equivalent to wall-clock time if all particles are evaluated in parallel),
-every time the swarm size increases, the performance improves.
+same amount of time~\cite{mcnabb-2009-large-particle-swarms}.  In
+\fig{iters-sphere} we see an example of this on the well-known benchmark
+function Sphere (20 dimensions, reporting the average of twenty runs).  In
+terms of the number of iterations performed (which is equivalent to wall-clock
+time if all particles are evaluated in parallel), every time the swarm size
+increases, the performance improves.
 
@@
 
-To solve this problem, we propose a method for performing two iterations of PSO
-at the same time in parallel that we call speculative evaluation.  The name
-comes from an analogy to speculative execution (also known as branch
-prediction), a technique commonly used in processors.  Modern processors, when
-faced with a branch on which they must wait (e.g., a memory cache miss), guess
-which way the branch will go and start executing, ensuring that any changes can
-be undone.  If the processor guesses right, execution is much farther ahead
-than if it had idly waited on the memory reference.  If it guesses wrong,
-execution restarts where it would have been anyway.
+To solve this problem, we apply the concept of speculative
+decomposition~\cite{grama-2003-intro-to-parallel-computing} to particle swarm
+optimization, using extra processors to perform two iterations of PSO at the
+same time.  Speculative decomposition is analogous to speculative execution
+(also known as branch prediction), a technique commonly used in processors.
+Modern processors, when faced with a branch on which they must wait (e.g., a
+memory cache miss), guess which way the branch will go and start executing,
+ensuring that any changes can be undone.  If the processor guesses right,
+execution is much farther ahead than if it had idly waited on the memory
+reference.  If it guesses wrong, execution restarts where it would have been
+anyway.
 
@@
 
+It is important to note here that this is not a variant of PSO.  We are not
+proposing a new algorithm, and we do not try to beat the state of the art in
+PSO research.  We simply propose a new way to think about the parallelization
+of PSO that we show works better in many cases than a na\"ive parallelization.
+
 Furthermore, we show that if we relax the requirements of the algorithm, no
@@
 we can introduce new speculative techniques that often out-perform PSO.  These
-variants make better use of the information obtained from the extra exploration
-made by the speculative function evaluations.  We also explore the idea that,
-like branch prediction in processors, we need not speculatively evaluate
-\emph{all} possible future positions, we can accelerate the algorithm even if
-we are just \emph{likely} to have guessed right.  By pruning the speculation to
-just paths that are statistically likely to reproduce the paths that are
-equivalent to PSO we can increase the swarm size without increasing the number
-of speculative evaluations.  We also consider several recovery strategies for
-cases where the pruned set of speculative evaluations does not contain the
-evaluation that standard PSO would have done.  A further improvement we explore
-is speculating several iterations ahead instead of just one, which is made
-possible by pruning the number of speculative evaluations.
+relaxations make better use of the information obtained from the extra
+exploration made by the speculative function evaluations.  We also explore the
+idea that, like branch prediction in processors, we need not speculatively
+evaluate \emph{all} possible future positions, we can accelerate the algorithm
+even if we are just \emph{likely} to have guessed right.  By pruning the
+speculation to just paths that are statistically likely to reproduce the paths
+that are equivalent to PSO we can increase the swarm size without increasing
+the number of speculative evaluations.  We also consider several recovery
+strategies for cases where the pruned set of speculative evaluations does not
+contain the evaluation that standard PSO would have done.  A further
+improvement we explore is speculating several iterations ahead instead of just
+one, which is made possible by pruning the number of speculative evaluations.
 
@@
 adding particles we employ a speculative approach that allows us to perform
-iterations two at a time.
+two iterations at a time.
 
@@
 The refactoring hinges on the idea that once the random coefficients
-$\pURand_{t}$ and $\pURand_{t}$ are determined, there are only a few possible
-updates to $\nbest$ and $\pbest$.  For the Ring topology there are 7 possible
-update cases, identified in Table~\ref{tab:evals}.  We label each case with an
-identifier referring to the source of the update: a minus sign ($-$) represents
-no update, $L$ represents an update to $\nbest$ coming from the left neighbor,
-$R$ represents an update to $\nbest$ coming from the right neighbor, and $S$
-represents an update to either $\pbest$ or $\nbest$ coming from the particle
-itself.  As an example, $\casexn$ refers to the case that the particle finds a
-new personal best, but neither it nor its neighbors find a position that
-updated its neighborhood best.  In the equations that follow, we refer to an
-unspecified update case as $\casegen$, and to the set of cases collectively as
-$\caseset$.
+$\pURand_{t}$ and $\nURand_{t}$ are determined, there are only a few possible
+new positions, or updates, for $\nbest$ and $\pbest$.  For the Ring topology
+there are 7 possible update cases, identified in Table~\ref{tab:evals}.  We
+label each case with an identifier referring to the source of the update: a
+minus sign ($-$) represents no update, $L$ represents an update to $\nbest$
+coming from the left neighbor, $R$ represents an update to $\nbest$ coming from
+the right neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$
+coming from the particle itself.  As an example, $\casexn$ refers to the case
+that the particle finds a new personal best, but neither it nor its neighbors
+find a position that updated its neighborhood best.  In the equations that
+follow, we refer to an unspecified update case as $\casegen$, and to the set of
+cases collectively as $\caseset$.
 
@@
 set of \emph{speculative children} is the set of all possible next iteration
-states that a particle could have.  We use $\p_t$ to denote a particle at
-iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
+states (including the particle's position, $\nbest$ and $\pbest$ positions)
+that a particle could have.  We use $\p_t$ to denote a particle at iteration
+$t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
 corresponding to one of the rows in Table~\ref{tab:evals}.  $\n_t$ is a
@@
 $\pbest$ and $\nbest$), define a set of speculative children,
-$\noeval{\sset}_{t+1}$.  The master than has a set of particles
+$\noeval{\sset}_{t+1}$.  The master then has a set of particles
 $\noeval{\pset}_t$, and for each particle a set of speculative children
@@
 indicator function similar to that found in \eqref{eq:deficasexn}.  In
-practice, this is done first by updating each particle's $\pbest$, if
-necessary, then by updating the particle's $\nbest$ with information from the
-particle's neighbors.  This is simply the original PSO algorithm, and
+practice, this is done first by updating each (original) particle's $\pbest$,
+if necessary, then by updating the particle's $\nbest$ with information from
+the particle's neighbors.  This is simply the original PSO algorithm, and
 corresponds to steps 1--5 in \alg{centralized}.  Given the updates to $\pbest$
@@
 \eqref{eq:val2updatelong}.  The child with the matching case is kept, and all
-other speculative children are discarded (step 7 in \alg{centralized}.
+other speculative children are discarded (step 7 in \alg{centralized}).
 
@@
 bests, but $\p_t$'s value is better.  For example, suppose that $\p_{t-1}$ has
-a personal best value of 3, and the we are seeking to minimize the function.
+a personal best value of 3, and that we are seeking to minimize the function.
 $\noeval{\p}_t$ is created, and $\noeval{\s}_{t+1}$ is moved assuming that
@@
 single machine, and the particle directs its extra machines to evaluate its
-speculative children.  However the distributed framework functions, the same
-information needs to be passed between particles, and we describe here the
-messages each particle needs to receive to perform speculative evaluation.
+speculative children.  The same information needs to be passed between
+particles no matter the framework used.  We describe here the messages each
+particle needs to receive to perform speculative evaluation.
 
@@
 
-Performing speculative evaluation in PSO with a dynamic topology raises a
-sticky issue of its own.  In a static topology, at iteration $t$ a particle
-already has all of the information about the positions of its neighbors during
-iterations $1$ through $t-1$.  If the neighbor finds a better position at
-iteration $t$, the particle updates its neighborhood best, but if it does not,
-it still has its old neighborhood best from its neighbors for all previous
-iterations.
+Performing speculative evaluation in PSO with a dynamic topology (where
+neighbors change from iteration to iteration) raises a sticky issue of its own.
+In a static topology, at iteration $t$ a particle already has all of the
+information about the positions of its neighbors during iterations $1$ through
+$t-1$.  If the neighbor finds a better position at iteration $t$, the particle
+updates its neighborhood best, but if it does not, it still has its old
+neighborhood best from its neighbors for all previous iterations.
 
@@
 certainty predict which branch a particle will take every time, if we can use
-statistics to narrow down the $2(n+1)$ possible evaluations to a few likely
+statistics to narrow down the $2n+1$ possible evaluations to a few likely
 candidates, we can decrease the amount of computation required to do
@@
 particle, we have $8p$ evaluations for every two iterations, where $p$ is the
-number of particles in the speculative swarm.  The extra evaluations done in
-speculative evaluation would instead be particles in regular PSO, so we compare
-swarms of size $p$ in speculative evaluation with swarms of size $8p$ in
-regular PSO.
-
-Where the Complete topology would normally be used, we use a Random topology in
-our speculative algorithm, as Complete leads to an explosion in the number of
-speculative evaluations.  If speculative evaluation were not being performed,
-it is possible that the Complete topology would be used.  However, the Complete
-topology also requires a very large amount of interprocessor communication in
-distributed PSO, so it is still quite possible that Random would be used even
-with standard PSO.  But, to be fair in our comparisons, we compare to the
-standard PSO algorithm using both the Random topology with two neighbors and
-the Complete topology (labeled PSO Random and PSO Complete in our results).
+number of particles in the speculative swarm.  The extra evaluations required
+in our speculative approach would instead be used to evaluate particles in
+regular PSO, so we compare swarms of size $p$ in speculative evaluation with
+swarms of size $8p$ in regular PSO.
+
+Where the Complete topology would normally be preferred, we use a Random
+topology in our speculative algorithm, as Complete leads to an explosion in the
+number of speculative evaluations.  If speculative evaluation were not being
+performed, it is possible that the Complete topology would be used.  However,
+the Complete topology also requires a very large amount of interprocessor
+communication in distributed PSO, so it is still quite possible that Random
+would be used even with standard PSO.  But, to be fair in our comparisons, we
+compare to the standard PSO algorithm using both the Random topology and the
+Complete topology (labeled PSO Random and PSO Complete in our results).
 
@@
 
-The algorithms we compare are the standard PSO algorithm, SEPSO, and four
-variations on SEPSO.  In presenting our results, we call the variant developed
-in \sec{pickbest} Pick Best.  The algorithms described in \sec{pruning} through
-\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
-algorithm in \sec{manyiters} we call Many Iterations.
-
-When performing pruning in SEPSO in Pick Best Pruned and Social Promotion
-Pruned, there are a large number of ways to prune speculative evaluations.  We
-experimented with several, but present results for only one possible pruning.
-The pruning we present uses only the two most probable branches in the Random
-topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
-the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
-also convenient in that no messages are needed from neighbors in order to
-produce the positions of the speculative particles---in distributed frameworks
-using several rounds of communication (see \sec{distributed}), one of the
-rounds of communication can be dropped entirely.
+The algorithms we compare are the standard PSO algorithm, our original
+speculative approach (here labeled SEPSO simply for ease of reference), and the
+four relaxations of SEPSO discussed in \sec{relax}.  In presenting our results,
+we call the approach developed in \sec{pickbest} Pick Best.  The algorithms
+described in \sec{pruning} through \sec{wrong} are called Pick Best Pruned and
+Social Promotion Pruned, and the algorithm in \sec{manyiters} we call Many
+Iterations.
+
+When performing pruning in Pick Best Pruned and Social Promotion Pruned, there
+are a large number of ways to prune speculative evaluations.  We experimented
+with several, but present results for only one possible pruning.  The pruning
+we present uses only the two branches where the $\nbest$ value was not updated:
+\casepn\ and \casexn.  Those branches are convenient in that no messages are
+needed from neighbors in order to produce the positions of the speculative
+particles---in distributed frameworks using several rounds of communication
+(see \sec{distributed}), one of the rounds of communication can be dropped
+entirely.
 
 Because pruning only requires two speculative evaluations per particle (along
-with evaluating the original particle), we can use swarms of size $3p$ to
-compare to a swarm of size $p$ with standard PSO, instead of $8p$ with other
-techniques.
+with evaluating the original particle), we can use swarms of size
+$\frac{1}{3}p$ to compare to a swarm of size $p$ with standard PSO, instead of
+$\frac{1}{8}p$ with other techniques.
 
@@
 When we introduce pruning, our intuition about Pick Best turns out to be more
-correct.  While adding 50 more particles to the swarm, Pick Best with pruning
-still gets stuck just as often as the original Pick Best.  However, Social
-Promotion does well with pruning; it increases the success rate to close to
-100\%, while still finding the optimum on average much faster than the original
-PSO.
+correct.  While adding 50 more particles to the swarm (as pruning allows us to
+have 80 particles with 240 processors instead of only 30), Pick Best with
+pruning still gets stuck just as often as the original Pick Best.  However,
+Social Promotion does well with pruning; it increases the success rate to close
+to 100\%, while still finding the optimum on average much faster than the
+original PSO.  These results are shown in \fig{griewank-pruned}.
 
@@
 
-We have given five different variants of speculative evaluation, each of which
-has different properties.  The variants of the algorithm perform differently on
-different functions and at different swarm sizes, as would be expected by the
-No Free Lunch Theorem~\cite{wolpert-1997-nfl-for-optimization}.  We have given
-a brief evaluation of the premature convergence properties of these variants on
-deceptive functions.  Though there is no one variant that performs best across
-all functions, for all but one of the functions we looked at \emph{some} kind
-of speculative evaluation outperforms standard PSO, often dramatically.
+What we have presented is not a new algorithm or a variant of PSO, and thus we
+did not compare our methods against the state of the art in PSO research.  We
+presented a new parallelization technique, so we compared parallelization
+strategies on the same algorithm, the original PSO.
+
+We have given five different possible approaches to speculative evaluation,
+each of which has different properties.  These approaches perform differently
+on different functions and at different swarm sizes, as would be expected by
+the No Free Lunch Theorem~\cite{wolpert-1997-nfl-for-optimization}.  We have
+given a brief evaluation of the premature convergence properties of these
+methods on deceptive functions.  Though there is no one approach that performs
+best across all functions, for all but one of the functions we looked at
+\emph{some} kind of speculative evaluation outperforms standard PSO, often
+dramatically.  Our purpose has been simply to show that in many cases
+speculative techniques outperform additional particles as the swarm size gets
+large; our results are conclusive in that regard.  We are arguing for the
+speculative parallelization of PSO, not for one particular approach to it.
 
@@
 \label{sec:future}
+
+We have presented a new paradigm for parallelizing the particle swarm
+optimization algorithm, focusing on PSO itself and not all of its variants.  It
+remains as future work to apply speculative approaches to recent and popular
+PSO variants, such as the Fully Informed Particle
+Swarm~\cite{mendes-2004-fully-informed-particle-swarm}.  While our methods will
+not always be immediately applicable to every variant, we are confident that
+some kind of speculative approach will be beneficial to the parallelization of
+all forms of PSO, especially as the number of processors used gets into the
+thousands.
 
@@
 ../../../bib/optimization/bib,%
+../../../bib/parallel/bib,%
 ../../../bib/pso/general/bib,%
 ../../../bib/pso/parallel/bib,%
+../../../bib/pso/fips/bib,%
 ../../../bib/pso/topology/bib,%

### 2010-10-08 masters_thesis/entropy_nips_workshop/paper.tex
Moved the original nips workshop to entropy_nips_workshop, because we are thinking of doing two papers

\documentclass{article}
\usepackage{nips10submit_e,times}
\usepackage{graphicx}

\title{Improving the Visualization of Topic Models}
\author{
Matthew J. Gardner \\
Department of Computer Science \\
Brigham Young University \\
Provo, UT 84604 \\
\texttt{mjg82@byu.edu} \\
\And
Other authers\ldots \\
Affiliation \\
Address \\
\texttt{email} \\
}

%\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}
\maketitle

\begin{abstract}
\end{abstract}

\section{Introdcution}

Topic modeling has shown to be an effective means of extracting semantic
meaning from a collection of
documents~\cite{blei-2003-latent-dirichlet-allocation}.  Topic models have been
used as an inputs to other kinds of algorithms, such as classification and
sentiment tasks (cite some papers here), but the usefulness of topic models as
visualization tools in and of themselves has not been well established, and is
even in question (cite more papers).  Some recent work has been done in an
attempt to make corpus browsers using the results of LDA, and to make the
results of LDA more easily interpretable, but there is much work yet to be done
(still more citations).

Common approaches to using topic models to facilitate corpus browsing and
visualization are limited mainly to showing the similarity between documents in
the corpus in various ways.  Such methods are useful in their own right when
one is individually scanning documents, but they are insufficient to answer
broader questions about the corpus, like, ``What topics show up most in these
documents?'' and, ``What does a particular author in the document talk about
most frequently that is unique to that author?''  Those questions turn one away
from analyzing individual documents to trying to get information out of the
topics themselves.

Upon looking at the topics one is confronted with tens or hundreds of lists of
words, sorted by a number given to each list by the algorithm.  Sifting through
those lists and trying to understand what they mean, and whether or not a
particular topic is important, can be a daunting task, particularly to one not
very familiar with the algorithm.  The common practice of cherry-picking good
topics to present in a paper hides the fact that a large number of topics
returned by topic models are very hard to interpret or even completely
meaningless to humans.

In order for topic models to be useful for corpus browsing, some sense has to
be made out of the topics.  One recently proposed method is to assign to each
topic a coherence value, so that the person can ignore topics that are likely
to be meaningless~\cite{newman-2010-automatic-evaluation-of-topic-coherence}.
While topic coherence is able to eliminate meaningless topics from
consideration, it is not able to address many broader questions that a person
browsing the corpus might have.

In this paper we present the idea of using the distribution of documents for a
given topic, $p(d|z)$, in order to answer questions about a corpus that
previous visualization techniques are not able to answer.  The distribution of
topics in a document, $p(z|d)$, has been widely used in all applications of
topic modeling, from dimensionality reduction of document representations to
showing similar documents in a corpus browser.  This is understandable, given
that estimating the distribution $p(z|d)$ is one of the primary purposes of
topic modeling.  However, $p(d|z)$ also contains interesting information that
can be very useful for visualizing and understanding the output of the topic
model.

The rest of this paper is outlined as follows.  We first describe a corpus of
campaign speeches from the 2008 presidential primaries and general election.
Then we present questions that a person browsing through these campaign
speeches may want to answer, and how one can use $p(d|z)$ to answer those
questions where other methods fail.  Finally, we conclude.

\section{Campaign Speeches}

TODO

\section{What topics were most prevelant during the 2008 campaign cycle?}

TODO

\section{What topics were most universal across candidates?}

TODO

\section{What did Barack Obama talk about most frequently?}

TODO

\section{Conclusion}

TODO

\bibliographystyle{unsrt}
\bibliography{../../../bib/lda/bib}

\end{document}

### 2010-10-08 honors_thesis/si_paper/paper.tex
Done with the SI paper, and submitted it

@@
-\documentclass[journal,letterpaper]{IEEEtran}
-
-\usepackage{cite}
+\documentclass[smallcondensed]{svjour3}
+
 \usepackage{graphicx}
@@
 \usepackage{psfrag}
-
-\IEEEoverridecommandlockouts
+\usepackage{natbib}
 
@@
 
-\title{\ \\ \LARGE\bf A Speculative Approach to Parallelization in Particle
-Swarm Optimization%
-\thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department
-of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602
-(phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}%
+\title{A Speculative Approach to Parallelization in Particle Swarm
+Optimization}
+
+\date{Received: date / Accepted: date}
+
+\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
+
+\institute{M. Gardner \at
+Brigham Young University.
+3342 TMCB, Provo, UT 84604
+Tel.: 801-422-8717
+Fax: 801-422-0169
+\email{mjg82@byu.edu}
+\and
+A. McNabb \at
+Brigham Young University.
+3342 TMCB, Provo, UT 84604
+\email{a@cs.byu.edu}
+\and
+K. Seppi \at
+Brigham Young University.
+3330 TMCB, Provo, UT 84604
+\email{k@cs.byu.edu}
 }
-
-\date{}
-
-\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
 
@@
 
+\keywords{Parallel algorithms \and Optimization methods \and Particle swarm
+optimization \and Speculative Decomposition}
+
 \end{abstract}
-
-\begin{keywords}
-Parallel algorithms, optimization methods, particle swarm optimization.
-\end{keywords}
 
@@
 The methods they have used include various synchronous algorithms
-\cite{chu-2006-intelligent-parallel-pso, jin-2005-pso-antenna-designs,
-parsopoulos-2004-parallel-vector-evaluated-pso,
+\citep{chu-2006-intelligent-parallel-pso,jin-2005-pso-antenna-designs,%
+parsopoulos-2004-parallel-vector-evaluated-pso,%
 schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
-\cite{mostaghim-2006-multi-objective-pso-on-grids,
+\citep{mostaghim-2006-multi-objective-pso-on-grids,%
 venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
@@
 almost all cases, adding additional particles produces better results in the
-same amount of time~\cite{mcnabb-2009-large-particle-swarms}.  In
+same amount of time~\citep{mcnabb-2009-large-particle-swarms}.  In
 \fig{iters-sphere} we see an example of this on the well-known benchmark
@@
 reason, previous work has recommended the use of a swarm size of 50 for
-PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
+PSO~\citep{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
 cases, adding particles indefinitely will not yield an efficient
@@
 
-To solve this problem, we apply the concept of speculative
-decomposition~\cite{grama-2003-intro-to-parallel-computing} to particle swarm
+In order to solve this problem, we apply the concept of speculative
+decomposition~\citep{grama-2003-intro-to-parallel-computing} to particle swarm
 optimization, using extra processors to perform two iterations of PSO at the
@@
 longer demanding that it strictly reproduce the exact behavior of standard PSO,
-we can introduce new speculative techniques that often out-perform PSO.  These
-relaxations make better use of the information obtained from the extra
-exploration made by the speculative function evaluations.  We also explore the
-idea that, like branch prediction in processors, we need not speculatively
-evaluate \emph{all} possible future positions, we can accelerate the algorithm
-even if we are just \emph{likely} to have guessed right.  By pruning the
-speculation to just paths that are statistically likely to reproduce the paths
-that are equivalent to PSO we can increase the swarm size without increasing
-the number of speculative evaluations.  We also consider several recovery
-strategies for cases where the pruned set of speculative evaluations does not
-contain the evaluation that standard PSO would have done.  A further
-improvement we explore is speculating several iterations ahead instead of just
-one, which is made possible by pruning the number of speculative evaluations.
+we can introduce new speculative techniques that often out-perform na\"ive
+parallelizations of PSO.  These relaxations make better use of the information
+obtained from the extra exploration made by the speculative function
+evaluations.  We also explore the idea that, like branch prediction in
+processors, we need not speculatively evaluate \emph{all} possible future
+positions, we can accelerate the algorithm even if we are just \emph{likely} to
+have guessed right.  By pruning the speculation to just paths that are
+statistically likely to reproduce the paths that are equivalent to PSO we can
+increase the swarm size without increasing the number of speculative
+evaluations.  We also consider several recovery strategies for cases where the
+pruned set of speculative evaluations does not contain the evaluation that
+standard PSO would have done.  A further improvement we explore is speculating
+several iterations ahead instead of just one, which is made possible by pruning
+the number of speculative evaluations.
 
@@
 Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
-Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is
+Eberhart~\citep{kennedy-1995-particle-swarm-optimization}.  The algorithm is
 used to intelligently search a multi-dimensional space by mimicking the
@@
 neighborhood best) is the best position the neighbors of the current particle
-have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
+have seen~\citep{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
 \ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
 required to ensure convergence (2.05, 2.05, and .73,
-respectively)~\cite{clerc-2002-constricted-pso}.
+respectively)~\citep{clerc-2002-constricted-pso}.
 
@@
 particle is a neighbor to every other
-particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
+particle~\citep{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
 particle is also a neighbor to itself in that its own position and value are
@@
 behavior of Complete with far less
-communication~\cite{mcnabb-2009-large-particle-swarms}.  In the Random
+communication~\citep{mcnabb-2009-large-particle-swarms}.  In the Random
 topology, each particle randomly picks two other particles to share information
@@
 sparse topologies.  Accordingly, we use the Random topology described
-in~\cite{mcnabb-2009-large-particle-swarms}, which has been shown to
+in~\citep{mcnabb-2009-large-particle-swarms}, which has been shown to
 effectively simulate the Complete topology.  In~\sec{results} we report the
@@
 merely used to evaluate the objective function at various positions as directed
-by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
+by the master~\citep{belal-2004-parallel-models-for-pso}.  To perform
 speculative evaluation in such an architecture, the master generates the
@@
 charge of and must perform the steps of the algorithm for those
-particles~\cite{mcnabb-2007-parallel-pso-using-mapreduce}.  Messages such as
+particles~\citep{mcnabb-2007-parallel-pso-using-mapreduce}.  Messages such as
 values and positions for the neighborhood best are sent between processors.
@@
 too many extra evaluations to produce competitive results.  On most functions,
-the larger swarm size that can be used in regular PSO leads to better
-performance than is found by trying to do speculative evaluation.  However, if
-we keep the idea of speculative evaluation while relaxing the requirement of
-exactly reproducing the behavior of the original PSO algorithm, we see some
-impressive results.
+the larger swarm size that can be used in na\"ive parallelizations leads to
+better performance than is found by trying to do speculative evaluation to
+exactly reproduce PSO.  However, if we keep the idea of speculative evaluation
+while relaxing the requirement of exactly reproducing the behavior of the
+original PSO algorithm, we see some impressive results.
 
@@
 
+None of these methods fundamentally change the PSO algorithm.  They simply lead
+to particles having different values for personal and neighborhood best
+positions than would have occurred in standard PSO, because they receive
+different information.  This kind of relaxation is fairly typical in the
+parallelization of PSO~\citep{koh-2006-parallel-asynchronous-pso}.
+
 \subsection{Pick the Best Child}
@@
 parallel evolution
-strategies~\cite{rudolph-1991-distributed-evolution-strategies}.  In evolution
+strategies~\citep{rudolph-1991-distributed-evolution-strategies}.  In evolution
 strategies, a parent individual (representing a potential solution to some
@@
 becomes the parent for the next
-generation~\cite{beyer-2002-evolution-strategies}.  Our methods are similar,
+generation~\citep{beyer-2002-evolution-strategies}.  Our methods are similar,
 where our mutation operator is simply the PSO motion equations and the
@@
 relaxation is nothing new; asynchronous adaptations of PSO do the same
-thing~\cite{koh-2006-parallel-asynchronous-pso}.  As an aside, it is equivalent
+thing~\citep{koh-2006-parallel-asynchronous-pso}.  As an aside, it is equivalent
 in this case to simply increment the iteration number of particles which fail
@@
 
-We ran experiments to compare our speculative PSO algorithm to the standard PSO
-algorithm.  At each iteration of the algorithms, we use one processor to
-perform one function evaluation for one particle, be it speculative or
-non-speculative.  The speculative algorithm actually performs two iterations of
-PSO at each ``iteration,'' so we instead call each ``iteration'' a ``round of
-evaluations.''
+We ran experiments to compare our speculative parallelization of PSO to the
+na\"ive parallelization.  At each iteration of the algorithms, we use one
+processor to perform one function evaluation for one particle, be it
+speculative or non-speculative.  The speculative version actually performs
+two iterations of PSO at each ``iteration,'' so we instead call each
+``iteration'' a ``round of evaluations.''
 
@@
 must hold fixed the number of processors used at each round of evaluations in
-all of our experiments, giving all algorithms the same number of function
+all of our experiments, giving all approaches the same number of function
 evaluations.  Thus while we label our graphs with ``Rounds of Evaluations,''
-this is equivalent to the number of function evaluations used by each algorithm
+this is equivalent to the number of function evaluations used by each approach
 (simply multiply by the number of processors to transform rounds of evaluations
@@
 
-Using the same number of processors for each algorithm requires that our
-speculative algorithms have a smaller swarm size than standard PSO.  For each
-of the topologies we used, a particle has three neighbors including itself.  As
-shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
-particle.  With one evaluation needed for the original, non-speculative
-particle, we have $8p$ evaluations for every two iterations, where $p$ is the
-number of particles in the speculative swarm.  The extra evaluations required
-in our speculative approach would instead be used to evaluate particles in
-regular PSO, so we compare swarms of size $p$ in speculative evaluation with
-swarms of size $8p$ in regular PSO.
+Using the same number of processors for each approach requires that our
+speculative parallelizations have a smaller swarm size than the na\"ive
+parallelization.  For each of the topologies we used, a particle has three
+neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
+$7$ speculative evaluations per particle.  With one evaluation needed for the
+original, non-speculative particle, we have $8p$ evaluations for every two
+iterations, where $p$ is the number of particles in the speculative swarm.  The
+extra evaluations required in our speculative approach would instead be used to
+evaluate particles in na\"ive parallelizations, so we compare swarms of size
+$p$ in speculative evaluation with swarms of size $8p$ in na\"ive approaches.
 
 Where the Complete topology would normally be preferred, we use a Random
-topology in our speculative algorithm, as Complete leads to an explosion in the
+topology in our speculative approach, as Complete leads to an explosion in the
 number of speculative evaluations.  If speculative evaluation were not being
@@
 communication in distributed PSO, so it is still quite possible that Random
-would be used even with standard PSO.  But, to be fair in our comparisons, we
-compare to the standard PSO algorithm using both the Random topology and the
-Complete topology (labeled PSO Random and PSO Complete in our results).
+would be used even with na\"ive parallelizations.  But, to be fair in our
+comparisons, we compare to the na\"ive parallelization using both the Random
+topology and the Complete topology (labeled PSO Random and PSO Complete in our
+results).
 
 We experimented with five common benchmark problems defined
-in~\cite{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
+in~\citep{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
 and Bohachevsky.
@@
 origin-seeking bias in the PSO algorithm and its
-variants~\cite{monson-2005-origin-seeking-bias}.  We move the center of the
+variants~\citep{monson-2005-origin-seeking-bias}.  We move the center of the
 shifted functions (all except Bohachevsky) to be halfway between the center and
@@
 
-The algorithms we compare are the standard PSO algorithm, our original
-speculative approach (here labeled SEPSO simply for ease of reference), and the
-four relaxations of SEPSO discussed in \sec{relax}.  In presenting our results,
-we call the approach developed in \sec{pickbest} Pick Best.  The algorithms
-described in \sec{pruning} through \sec{wrong} are called Pick Best Pruned and
-Social Promotion Pruned, and the algorithm in \sec{manyiters} we call Many
-Iterations.
+The parallelization techniques we compare are the na\"ive parallelization (here
+labeled Na\"ive simply for ease of reference), our original speculative
+approach (labeled SEPSO), and the four relaxations of SEPSO discussed in
+\sec{relax}.  In presenting our results, we call the approach developed in
+\sec{pickbest} Pick Best.  The methods described in \sec{pruning} through
+\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
+method in \sec{manyiters} we call Many Iterations.
 
@@
 with evaluating the original particle), we can use swarms of size
-$\frac{1}{3}p$ to compare to a swarm of size $p$ with standard PSO, instead of
-$\frac{1}{8}p$ with other techniques.
+$\frac{1}{3}p$ to compare to a swarm of size $p$ with the na\"ive
+parallelization, instead of $\frac{1}{8}p$ with other techniques.
 
@@
 the more particles there are in the swarm, the lower function value it finds,
-up to at least 4000 particles~\cite{mcnabb-2009-large-particle-swarms}.
+up to at least 4000 particles~\citep{mcnabb-2009-large-particle-swarms}.
 Smaller swarms get caught in local optima.  Because our speculative algorithms
@@
 correct.  In this experiment we used 240 processors, so SEPSO and Pick Best
-each had 30 particles, and the standard PSO algorithms had 240 particles.  As
+each had 30 particles, and the Na\"ive algorithms had 240 particles.  As
 expected, SEPSO and Pick Best converge quickly to worse local optima than
-standard PSO does.  \fig{rastrigin} shows the results graphically.
+Na\"ive does.  \fig{rastrigin} shows the results graphically.
 
@@
 expect our methods to be perfect for such functions, and our results show this
-intuition again to be correct.  First we compare the basic SEPSO algorithm to
-standard PSO.  In \fig{basic-sphere} we can see that when both SEPSO and PSO
-are using a Random topology, SEPSO clearly outperforms PSO.  However, the
-comparison between SEPSO and PSO Complete is not as encouraging---PSO Complete
-performs slightly better.
+intuition again to be correct.  First we compare the basic SEPSO approach to
+Na\"ive.  In \fig{basic-sphere} we can see that when both SEPSO and Na\"ive are
+using a Random topology, SEPSO clearly outperforms Na\"ive.  However, the
+comparison between SEPSO and Na\"ive Complete is not as encouraging---Na\"ive
+Complete performs slightly better.
 
@@
 When we introduce Pick Best, we see more impressive results.  As can be seen in
-\fig{sphere-pickbest}, where SEPSO did not manage to perform as well as PSO
+\fig{sphere-pickbest}, where SEPSO did not manage to perform as well as Na\"ive
 Complete, Pick Best handily outperforms it with the same number of processors.
@@
 Iterations on average finds values that are more than 40 orders of magnitude
-better than standard PSO using the same number of processors.  This is shown in
+better than Na\"ive using the same number of processors.  This is shown in
 \fig{sphere-manyiters}.  Social Promotion Pruned performed comparably with
@@
 more by larger swarms than Sphere is.  Thus our speculative algorithms often
-fail to perform as well as standard PSO because of their smaller swarm size.
-In comparing SEPSO to standard PSO in \fig{schwefel1}, we can see that SEPSO
-fails to perform any better than PSO with a Random topology, and it is by far
-outperformed by PSO with a Complete topology.
+fail to perform as well as Na\"ive because of their smaller swarm size.  In
+comparing SEPSO to Na\"ive in \fig{schwefel1}, we can see that SEPSO fails to
+perform any better than Na\"ive with a Random topology, and it is by far
+outperformed by Na\"ive with a Complete topology.
 
@@
 As can be seen in \fig{schwefel2}, almost all of the speculative algorithms
-perform worse than PSO Complete, with Many Iterations performing comparably to
-PSO.  However, one of the speculative algorithms did perform better than PSO:
-Pick Best Pruned.  It seems that the larger swarm size available because of
-pruning was the cause of the improved performance.
+perform worse than Na\"ive Complete, with Many Iterations performing comparably
+to Na\"ive.  However, one of the speculative algorithms did perform better than
+Na\"ive: Pick Best Pruned.  It seems that the larger swarm size available
+because of pruning was the cause of the improved performance.  Note that Many
+Iterations was initially performing comparably with Pick Best Pruned, but with
+more iterations gradually degraded in performance to the level of Na\"ive
+Complete.  It is likely that Many Iterations got caught in local optima, and
+with more processors would have maintained better performance.
 
@@
 premature convergence on a local
-optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank has a global
-optimum with a value of 0, and sometimes the swarm finds the optimum and
+optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  Griewank has a
+global optimum with a value of 0, and sometimes the swarm finds the optimum and
 sometimes it does not.  Instead of showing average function value at each
@@
 in \fig{basic-griewank1} for swarms of size 30 and 240 using the Ring topology.
-One can see in the figure that the SEPSO algorithm finds the optimum faster
-than the original PSO.  However, because the swarm size is so small, SEPSO gets
+One can see in the figure that the SEPSO approach finds the optimum faster than
+the Na\"ive approach.  However, because the swarm size is so small, SEPSO gets
 stuck a little less than half of the time.
@@
 Thus with a few more processors, our algorithm finds the optimum on average
-twice as fast as the original PSO without a significant loss of accuracy.
+twice as fast as the Na\"ive approach without a significant loss of accuracy.
 
@@
 
-In order to show the premature convergence properties of the various algorithms
+In order to show the premature convergence properties of the various approaches
 we have presented, we will show results first with swarms using 240 processors,
@@
 
-When we look at the performance of our Pick Best algorithm, a less intuitive
+When we look at the performance of our Pick Best approach, a less intuitive
 result occurs with Griewank than with Sphere.  Because Sphere needs very little
@@
 seven speculative evaluations to overcome the inherent greediness of the
-algorithm.
+approach.
 
@@
 Iterations finds the optimum, it finds it quicker than any other method we
-tried, on average four times faster than standard PSO.  However, it also gets
-stuck and fails to find the optimum more than any other method.  The results
-are shown in \fig{griewank-manyiters}.   \fig{griewank-manyiters} is also
+tried, on average four times faster than Na\"ive.  However, it also gets stuck
+and fails to find the optimum more than any other method.  The results are
+shown in \fig{griewank-manyiters}.   \fig{griewank-manyiters} is also
 interesting in that it highlights the trade-off between accuracy and speed in
-the various algorithms at this swarm size.  The faster the algorithm finds the
+the various approaches at this swarm size.  The faster the approach finds the
 optimum, the less likely it is to be successful.
@@
 
-Now that we have seen the effect that the various algorithms have on speed of
+Now that we have seen the effect that the various approaches have on speed of
 convergence and success rates, we show a result for using 800 processors, where
-most of the algorithms have a 100\% success rate.  In this case, the best
+most of the approaches have a 100\% success rate.  In this case, the best
 algorithm to use is the one which finds the optimum the fastest.  As can be
 seen in \fig{griewank-800}, with 800 processors the best choice is Pick Best,
-which finds the optimum on average three times as fast as standard PSO.
+which finds the optimum on average three times as fast as standard PSO.  With
+just a few more processors, however, Many Iterations would likely have a 100\%
+success rate as well, making it the preferred choice.
 
@@
 Griewank, for Bohachevsky we merely present the best results.  In
-\fig{bohachevsky} we can see that all of the speculative algorithms found the
-optimum much quicker than PSO with a Random topology.  However, SEPSO was
-slower than PSO Complete and got stuck around 15\% of the time.  Pick Best,
-Pick Best Pruned, and Many Iterations all outperformed PSO Complete, though
-Many Iterations only almost reached a 100\% success rate.
+\fig{bohachevsky} we can see that all of the speculative approaches found the
+optimum much quicker than Na\"ive with a Random topology.  However, SEPSO was
+slower than Na\"ive Complete and got stuck around 15\% of the time.  Pick Best,
+Pick Best Pruned, and Many Iterations all outperformed Na\"ive Complete, with
+Many Iterations finding the optimum about twice as fast.
 
@@
 over previous methods.  We have shown results that conclusively demonstrate the
-superiority of our techniques for several functions over naively adding
+superiority of our techniques for several functions over na\"ively adding
 particles to the swarm when extra processors are available.
@@
 What we have presented is not a new algorithm or a variant of PSO, and thus we
-did not compare our methods against the state of the art in PSO research.  We
+did not compare our methods against the state of the art PSO variants.  We
 presented a new parallelization technique, so we compared parallelization
@@
 on different functions and at different swarm sizes, as would be expected by
-the No Free Lunch Theorem~\cite{wolpert-1997-nfl-for-optimization}.  We have
+the No Free Lunch Theorem~\citep{wolpert-1997-nfl-for-optimization}.  We have
 given a brief evaluation of the premature convergence properties of these
 methods on deceptive functions.  Though there is no one approach that performs
-best across all functions, for all but one of the functions we looked at
-\emph{some} kind of speculative evaluation outperforms standard PSO, often
-dramatically.  Our purpose has been simply to show that in many cases
-speculative techniques outperform additional particles as the swarm size gets
-large; our results are conclusive in that regard.  We are arguing for the
-speculative parallelization of PSO, not for one particular approach to it.
+best across all functions at the swarm sizes we used, for all but one of the
+functions we looked at \emph{some} kind of speculative evaluation outperforms
+standard PSO, often dramatically.  It also appears to us that once enough
+processors are available, by far the best approach is to speculate as many
+iterations ahead as processors allow.
 
 Though our methods show great improvements on some functions, they do not work
-for all functions.  As is commonly known, in PSO there is a trade-off between
-exploration and exploitation.  Some functions need only minimal exploration,
-and some never seem to have enough.  Increasing the swarm size is a natural way
-to increase exploration in a parallel environment.  However, once ``enough''
-exploration has been reached for any particular function, adding additional
-particles adds only incremental benefits.  A better use of the additional
-processors, as we have shown, is to perform some amount of speculative
-evaluation.
+for all functions at the swarm sizes we were able to test with.  As is commonly
+known, in PSO there is a trade-off between exploration and exploitation.  Some
+functions need only minimal exploration, and some never seem to have enough.
+Increasing the swarm size is a natural way to increase exploration in a
+parallel environment.  However, once ``enough'' exploration has been reached
+for any particular function, adding additional particles adds only incremental
+benefits.  A better use of the additional processors, as we have shown, is to
+perform some amount of speculative evaluation.
 
@@
 PSO variants, such as the Fully Informed Particle
-Swarm~\cite{mendes-2004-fully-informed-particle-swarm}.  While our methods will
-not always be immediately applicable to every variant, we are confident that
-some kind of speculative approach will be beneficial to the parallelization of
-all forms of PSO, especially as the number of processors used gets into the
-thousands.
+Swarm~\citep{mendes-2004-fully-informed-particle-swarm}.  While our methods
+will not always be immediately applicable to every variant, we are confident
+that some kind of speculative approach will be beneficial to the
+parallelization of all forms of PSO, especially as the number of processors
+used gets into the thousands.
 
@@
 is very similar to the distribution of the original
-PSO~\cite{poli-2008-sampling-distribution-of-pso}.  This remains to be proven,
+PSO~\citep{poli-2008-sampling-distribution-of-pso}.  This remains to be proven,
 or even demonstrated empirically.
@@
 stagnant almost the entire time.  Such information could be used in a
-TRIBES-like fashion~\cite{clerc-2003-tribes} to determine which particles
+TRIBES-like fashion~\citep{clerc-2003-tribes} to determine which particles
 should be removed from the swarm, perhaps starting them close to
@@
 A technique similar to ours has been previously used to parallelize simulated
-annealing~\cite{witte-1991-parallel-simulated-annealing-speculative}.  Any
+annealing~\citep{witte-1991-parallel-simulated-annealing-speculative}.  Any
 other optimization algorithm that only depends on current sampling positions
@@
 evolutionary algorithms was proposed
-in~\cite{poli-2006-backward-chaining-evolutionary-algorithms}.
+in~\citep{poli-2006-backward-chaining-evolutionary-algorithms}.
 
@@
 
-\bibliographystyle{unsrt}
+\bibliographystyle{apa-good}
 \bibliography{%

### 2010-10-08 masters_thesis/tool_nips_workshop/paper.tex
Done with a first draft of a tools paper for the nips workshop

\documentclass{article}
\usepackage{nips10submit_e,times}
\usepackage{graphicx}

\title{The Topic Browser\\An Interactive Tool for Browsing Topic Models}
\author{
Matthew J. Gardner \\
Department of Computer Science \\
Brigham Young University \\
Provo, UT 84604 \\
\texttt{mjg82@byu.edu} \\
\And
Other authers\ldots \\
Affiliation \\
Address \\
\texttt{email} \\
}

%\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}
\maketitle

\begin{abstract}

Topic models have been shown to be very useful for capturing a measure of
semantic meaning in a corpus.  Many individualized and cherry-picked
visualizations of topic models have been done, showing the potential of topic
models to give valuable insight into a corpus.  However, good, general tools
for browsing the entire output of a topic model along with its corpus have been
lacking.  We present an interactive tool that incorporates a significant amount
of prior work in displaying topic models, as well as a few novel ideas that
greatly enhance the visualization of these models.

\end{abstract}

\section{Introdcution}

Proper visualizations are essential to extracting information from data,
especially large and highly dimensional sets of data.  Large text corpora are
particularly difficult to visualize, as they typically include thousands of
documents and perhaps millions of words.  Topic
modeling~\cite{blei-2003-latent-dirichlet-allocation} is a method of reducing
the dimensionality of a corpus into a set of meaningful topics.  Topic models
have been used with some success as an aid to corpus
visualization~\cite{blei-2009-topic-models,
newman-2010-visualizing-with-topic-maps}.  However, in general the presentation
of topic models has been limited to hand selected topics that show the model in
the best light possible.  These presentations leave the viewer wondering what
else the topic model discovered and provide little help in gaining a deep
understanding of the model.

To aid in the visualization of topic models, we present The Topic Browser, a
tool for interactively exploring both the output of a topic modeling algorithm
and its attendant corpus.  Our topic browser incorporates many visualizations
of topic models previously published, as well as some innovative ideas of our
own.  The Topic Browser is an aid both to those who wish to browse through a
corpus and for those who wish to analyze the topic model itself.

The rest of this paper simply describes the Topic Browser.  We demonstrate our
tool with a topic model run on a collection of about 460 campaign speeches from
the 2008 presidential primary and general election.

\section{Extra Information}

Aside from the documents and the topic model themselves, our browser
incorporates three other pieces of information: attributes, or metadata,
associated with each document, topic metrics, and document metrics.  Here we
give a brief description of these kinds of information, deferring a discussion
of their use in visualization to the following sections.

Attributes of documents have been used heavily in recent topic models,
including the Author-Topic model~\cite{rosen-zvi-2004-author-topic-model},
Topics over Time~\cite{wang-2006-topics-over-time}, and Dirichlet-Multinomial
Regression~\cite{mimno-2008-topic-models-with-arbitrary-features-dmr}.  While
we currently do not have specialized visualizations for these topic models, we
do include document attributes in our browser and have some of our own 
visualizations that include the attributes.

In order to browse more effectively through topics, we introduce topic metrics
that give information about the topic.  These range from simple metrics, such
as the number of tokens and types labeled with the topic, to more complicated
metrics like how spread out the topic is across documents, or how coherent its
words are~\cite{newman-2010-automatic-evaluation-of-topic-coherence}.  We also
use pairwise topic metrics as similarity measures to show similar topics.

Similar to topic metrics, one can also compute document metrics.  Beyond simple
metrics like token count in the document, these can include things such as the
entropy of the topic distribution of the
document~\cite{misra-2008-lda-to-find-semantically-incoherent-documents}.  And
as with topics, we make use of pairwise document metrics such as topic
correlation~\cite{blei-2009-topic-models} to show similar documents.

\section{Sidebar}

The sidebar in our browser is the main navigation tool.  Apart from the obvious
task of directing the user to various parts of the visualization tool, then
main purpose of the sidebar is to list the items of a particular type (such as
topics or documents).  The new capability that we bring to the of listing the
items is the ability to sort and filter these lists as the user desires.

When presenting the results of topic models, papers will often cherry-pick the
list of topics looking for particularly good topics to display, leaving the
large number of meaningless topics untouched.  While such presentations may
help the researcher to highlight the benefits of her new topic model, they hide
the fact that finding good topics is often a laborious process.  With our topic
and document metrics, the sorting and filtering capabilities of the sidebar
list allow the user to often go quickly to the very topics he desires to find.

When browsing through topics, the user can filter the topic list by coherence
to eliminate from the view topics that are mostly meaningless, and sort by
document entropy to find topics that were used widely (or narrowly, if that is
desired) throughout the corpus.  She could also filter by document or
attribute, seeing only topics that were used in a particular document or by a
particular author.  Documents can similarly be filtered and sorted, showing
only documents that contain tokens with a particular topic, or with a
particular value of an attribute.

\section{Topic Browsing}

Topic models are typically presented as static lists of words, often simply
showing the top ten words from each topic in a list.  While this practice is
often sufficient to give a reader a basic idea of what the topic captured, it
is largely incapable of conveying a deep understanding of the context of that
topic in the corpus.  In order to answer significant questions about a corpus
one must browse the topics discovered, and previous methods have failed to
adequately allow the user to explore topics.  Our visualizer presents new ways
to gather information about a topic in the corpus, both in how the words in the
topic are displayed and in other information that is presented about the topic.

\subsection{Showing Top Words}

Instead of showing a list of the top ten words, we show a word cloud of the top
100, where the size of each word is determined by the probability of seeing
that word in the topic.  We also show a re-weighted word cloud as determined by
David Blei's Turbo Topics method~\cite{blei-2009-turbo-topics}.  While these
visualizations are modest improvements over typical visualization methods, our
most useful display of the words is showing them in their context.

The topics in a topic model cannot be fully interpreted when completely
separated from their context.  Thus we display the top ten words in the topic
inside of their context.  We select a random token of each word type labeled
with the topic and show up to 50 characters on either side of the token.  We
also allow the user to cycle through contexts to gain a broader view of how
each word is actually used in that topic in the corpus.

\subsection{Getting More Information}

We give two main ways of getting more information about topics in the model.
The first is showing documents and attribute values which have the highest
number of tokens labeled with that topic, or the highest percent in the
document or attribute.  This allows the user to quickly find documents that
best demonstrate what the topic captured.  Showing the top values for a given
attribute (such as authors for the attribute ``Author'') gives the user an idea
of how focused the topic is and often gives interesting information about the
corpus being browsed (for example, in Figure \ref{fig:something}, we see that
Democrats spoke more often about education than Republicans did).

The other way that we give more information about topics is by showing similar
topics.  We find similar topics either by looking at the documents that the
topics show up in or by the words that the topics use.  Looking at similar
topics by document shows topics that are commonly used together in the corpus,
and looking by word shows topics that have similar words, possibly because the
two topics really should have been one topic.

\section{Document, Word, and Attribute Browsing}

In addition to allowing the user to browse the topics in the topic model, we
provide means for browsing the documents, words, and attributes in the corpus.
These facilities give users means to explore the corpus itself in the context
of the topic model.

When simply browsing through the documents, we provide sorting and filtering
methods on the list of documents, as mentioned previously.  When looking at a
particular document, we show basic information about the document, its text,
the topic distribution in the document, and similar documents based on that
distribution.  When looking at the document in the context of a topic, we also
highlight the tokens in the document that were labeled with that topic.  All of
this has been done previously, besides the sorting and filtering of the
documents, and in fact constitutes the entirety of most previous corpus
browsers based on topic models---this functionality is all that is generally
provided.

We also provide views of individual words in the corpus.  When viewing a word
in the context of a topic, the user can see all uses of the word in that topic
in the corpus, with context taken from their corresponding documents.  The user
can also view words independently with a search-like interface, seeing what
topics and what documents in which the word appears most frequently.  While
providing basic functionality, however, the search interface leaves much to be
desired, as only single words can currently be searched for.  We plan on
expanding that to phrases in the future.

The user can also look at aggregated information for values of an attribute
(i.e., a particular author or year in the documents).  This view is also at
present somewhat limited, showing only what topics and words are used most
frequently by the collection of documents with that attribute.

\section{Plots}

There has been a wide variety of plots used in the visualization of topic
models.  We currently include two kinds of plots in our topic browser, with
plans to implement many more.  The first is showing trends for topics over the
values of an attribute (such as date, or author), useful for corpus browsing.
This kind of plot has been used in visualizing topic models almost since their
introduction~\cite{griffiths-2004-finding-scientific-topics}.  Our topic
browser allows the user to very quickly generate these trend plots over any
topic or combination of topics for any attribute of the documents in the
corpus.

The second kind of plot that we include is more useful for analyzing and
understanding the behavior of the topic model itself.  We allow the user to
plot two topic metrics against each other and compute a linear regression.
This allows the user to see some interesting properties of the topic metrics,
such as the fact that document entropy seems to correlate logarithmically with
the number of tokens in the topic, and that coherence does not seem to
correlate with any other topic metric.  The user can also found outliers, such
as topics with low document entropy but a high token count, that can then be
examined in the topic page.

\section{Conclusion}

We have presented the Topic Browser, an interactive tool for browsing both the
output of a topic model and the corpus that was modeled.  We have shown that
our tool incorporates many previously published visualizations of topic models,
including basic corpus browsing functionality, plots of trends over attributes
in the corpus, and Blei's turbo topics method of finding significant phrases
for each topic.  We have also presented many novel ways to get information from
a topic model in an interactive browsing experience.

While our tool is still under development, we intend to make it open and freely
available, both for those wishing to browse through a corpus in the context of
a topic model, and for those wishing to better understand topic models and
develop new models.  Our tool currently supports any topic model that labels
individual tokens in the corpus with a topics.  We have plans to include
special visualizations for more complicated topic models, such as Topics over
Time, sentiment-topic models, and others.

\bibliographystyle{plain}
\bibliography{../../../bib/lda/bib}

\end{document}

### 2010-10-15 masters_thesis/entropy_nips_workshop/paper.tex
I reworked the entropy paper, and am close to finishing a draft

@@
 \usepackage{graphicx}
+\usepackage{amsmath}
 
-\title{Improving the Visualization of Topic Models}
+% Better references, I think
+\newcommand{\secref}[1]{Section~\ref{sec:#1}}
+\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
+\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
+
+\title{Better Corpus Visualization through Topic Ranking}
 \author{
@@
 \section{Introdcution}
+\label{sec:introduction}
 
-Topic modeling has shown to be an effective means of extracting semantic
+Topic modeling has been shown to be an effective means of extracting semantic
 meaning from a collection of
@@
 broader questions about the corpus, like, ``What topics show up most in these
-documents?'' and, ``What does a particular author in the document talk about
-most frequently that is unique to that author?''  Those questions turn one away
-from analyzing individual documents to trying to get information out of the
-topics themselves.
+documents?'' and, ``What does a particular author in the corpus talk about most
+frequently that is unique to that author?''  Those questions turn one away from
+analyzing individual documents to trying to get information out of the topics
+themselves.
 
@@
 consideration, it is not able to address many broader questions that a person
-browsing the corpus might have.
+browsing the corpus may have.
 
-In this paper we present the idea of using the distribution of documents for a
-given topic, $p(d|z)$, in order to answer questions about a corpus that
-previous visualization techniques are not able to answer.  The distribution of
-topics in a document, $p(z|d)$, has been widely used in all applications of
-topic modeling, from dimensionality reduction of document representations to
-showing similar documents in a corpus browser.  This is understandable, given
-that estimating the distribution $p(z|d)$ is one of the primary purposes of
-topic modeling.  However, $p(d|z)$ also contains interesting information that
-can be very useful for visualizing and understanding the output of the topic
-model.
+In this paper we present the novel task of ranking topics by relevance to
+specific user questions.  If one is to successfully visualize a corpus in a way
+that answers broad questions, and if topics, not documents, are the natural
+answer to those questions, then one must properly rank topics such that the
+most prominent topics contain the answer to the user's questions.  We recognize
+that not all questions about a corpus fit into this visualization model, but
+there are many that do, such as those listed above.
 
-The rest of this paper is outlined as follows.  We first describe a corpus of
-campaign speeches from the 2008 presidential primaries and general election.
-Then we present questions that a person browsing through these campaign
-speeches may want to answer, and how one can use $p(d|z)$ to answer those
-questions where other methods fail.  Finally, we conclude.
+The rest of this paper is outlined as follows.  We first give further
+motivation and a more detailed description of the task of ranking topics by
+relevance in \secref{ranking}.  \secref{docdist} describes a method for
+determining the relevance of a topic to various possible questions a person
+could have when looking at a corpus, and \secref{examples} shows examples of
+how this technique provides a proper ranking.  In \secref{challenges} we
+discuss the challenges of evaluating this visualization method.  Finally, in
+\secref{conclusion} we conclude.
 
-\section{Campaign Speeches}
+\section{The topic ranking task}
+\label{sec:ranking}
 
-TODO
+We assert that for many tasks, the proper visualization of a topic model is a
+ranking of topics, not simply a window onto individual documents.  When one
+wants to see what topics are prevelant in a corpus, the answer is not found by
+scanning individual documents, but by looking at the topics produced by a topic
+model.  Likewise, when one wants to see trends of topics over time, one does
+not look through the documents, but at the topics, computing some statistics of
+those topics for each year for which there are documents.  Thus a proper
+visualization of the topic model for these tasks is to show most prominently
+those topics that are most relevant to the task.
 
-\section{What topics were most prevelant during the 2008 campaign cycle?}
+We can formalize the topic ranking problem simply as follows.  Given a set of
+topics from a topic model of a corpus, and a user question about the corpus,
+rank the topics by relevance to the user's question.  This is very similar to
+basic information retrieval tasks, but also significantly different, in that we
+are not looking for a ranking of documents as the result of a query, but a
+ranking of topics which answer broader questions about the corpus as a whole.
 
-TODO
+Clearly the topic ranking task is fundamental to the proper visualization of
+topic models, but as far as we are aware, this is the first attempt to even
+consider the problem.  We propose a very simple way to answer some user
+questions through the use of the document distribution of each topic.  This
+technique is by no means a general solution to the problem, but for some
+specific tasks it provides rankings that gives users the topics that answer
+their questions.
 
-\section{What topics were most universal across candidates?}
+\section{The document distribution for a topic}
+\label{sec:docdist}
 
-TODO
+The distribution of topics in a document, $p(z|d)$, has been widely used in all
+applications of topic modeling, from dimensionality reduction of document
+representations to showing similar documents in a corpus browser.  This is
+understandable, given that estimating the distribution $p(z|d)$ is one of the
+primary purposes of topic modeling.  However, $p(d|z)$, the document
+distribution for a given topic, also contains interesting information that can
+be very useful for visualizing and understanding the output of the topic model.
+We use various aspects of this distribution as approximations of relevance to
+certain user questions.
 
-\section{What did Barack Obama talk about most frequently?}
+The document distribution for a topic can be thought of as follows: suppose we
+have a token labeled with a given topic; what is the probability distribution
+over documents in the corpus that this token could have come from?  We
+approximate this distribution with a sample from a Gibbs sampler of LDA.  The
+distribution is thus defined as
+\begin{align}
+  \label{eq:entropy}
+  p(d|z) = \frac{C(d,z)}{\sum_{d \in \mathcal{D}} C(d,z)}
+\end{align}
+where $d$ is a particular document, $z$ is a particular topic, $C(d,z)$ is the
+number of tokens in document $d$ labeled with topic $z$, and $\mathcal{D}$ is
+the corpus.
 
-TODO
+Our use of the document distribution for a topic focuses primarily on
+calculating the entropy of the distribution and of related distributions where
+documents are grouped by category.  For instance, one can group the documents
+by author and compute the distribution of the topic over authors as
+\begin{align}
+  p(a|z) = \frac{\sum_{d \in \mathcal{A}} C(d,z)}{\sum_{d \in \mathcal{D}}
+  C(d,z)}
+\end{align}
+where $a$ is an author and $\mathcal{A}$ is the set of documents written by
+that author.
+
+Also, one can compute the distribution over any subset of the documents, where
+$p(d|z)$ is defined as in \eqref{eq:entropy}, but instead of the entire corpus,
+$\mathcal{D}$ represents some specific subset.  When using subsets of the
+documents, we only consider topics which have at least one token in the given
+subset, so as to avoid division by zero.
+
+\section{Using $p(d|z)$ to rank topics}
+\label{sec:examples}
+
+We present here a few examples of how the document distribution for a topic can
+be used to rank topics according to their relevance for certain user questions.
+In our examples we use a corpus of 480 campaign speeches from the 2008
+presidential primary and general election.  We inferred 150 topics in the data
+from a Gibbs sampling of the LDA topic model.
+
+Questions here:
+
+What topics are most prevelant in the corpus?
+
+What topics did John McCain talk about the most?
+
+What topics did Barack Obama most consistently throughout his campaign?
+
+What topics are most consistent across candidates?
+
+What topics were constant throughout the campaign?
+
+What topics were transitory, only spoken of rarely?
+
+\section{Challenges of evaluation}
+\label{sec:challenges}
+
+The obvious way to evaluate a topic ranking is to use the evaluations methods
+commonly found in information retrieval tasks.  There are well established
+techniques to evaluate a ranking given a true ranking.  However, there are
+several obstacles to using such a scheme to evaluate performance on this task.
+
+The first and glaring obstacle is that there is no set of data that has been
+labeled with true rankings for these or any questions.  One cannot compare the
+output of an algorithm to ground truth when there is no ground truth.  It is
+not even clear how such data would be obtained; perhaps one could take a topic
+model and a set of questions and rank each topic for relevance to each
+question, but such a task would be incredibly laborious and error-prone.  The
+labeler would first have to understand what exactly each topic is capturing, in
+itself a difficult task, then go through the entire data collection to answer
+the question herself and determine how well the topic matches.  Clearly such a
+process for creating a set of evaluation data is not feasible.
+
+But what if the topic model fails to capture topics that truly answer the
+question?  
+Separate topics can be overlapping in words, where a human would consider them
+the same topic.
+
+The topic model itself could be high or low quality, and that affects the
+results.
 
 \section{Conclusion}
+\label{sec:conclusion}
 

### 2010-10-21 masters_thesis/tool_nips_workshop/paper.tex
Editing on the tool paper from feedback, and I put in some screenshot images for the paper

@@
 
-Topic models have been shown to be very useful for capturing a measure of
-semantic meaning in a corpus.  Many individualized and cherry-picked
-visualizations of topic models have been done, showing the potential of topic
-models to give valuable insight into a corpus.  However, good, general tools
-for browsing the entire output of a topic model along with its corpus have been
-lacking.  We present an interactive tool that incorporates a significant amount
-of prior work in displaying topic models, as well as a few novel ideas that
-greatly enhance the visualization of these models.
+Topic models have been shown to capture a measure of the semantics in a corpus.
+Many individualized and cherry-picked visualizations of topic models have been
+reported in the literature, showing the potential of topic models to give
+valuable insight into a corpus.  However, good, general tools for browsing the
+entire output of a topic model along with the analyzed corpus have been
+lacking.  We present an interactive tool that incorporates both prior work in
+displaying topic models as well as some novel ideas that greatly enhance the
+visualization of these models.
 
@@
 
-\section{Introdcution}
+\section{Introduction}
 
-Proper visualizations are essential to extracting information from data,
-especially large and highly dimensional sets of data.  Large text corpora are
-particularly difficult to visualize, as they typically include thousands of
-documents and perhaps millions of words.  Topic
+Proper visualizations are essential to extracting information from and
+identifying trends in data, especially large, highly dimensional data.  Large
+text corpora are particularly difficult to visualize, as they typically include
+thousands of documents and perhaps millions of words.  Topic
 modeling~\cite{blei-2003-latent-dirichlet-allocation} is a method of reducing
@@
 
-To aid in the visualization of topic models, we present The Topic Browser, a
-tool for interactively exploring both the output of a topic modeling algorithm
-and its attendant corpus.  Our topic browser incorporates many visualizations
-of topic models previously published, as well as some innovative ideas of our
-own.  The Topic Browser is an aid both to those who wish to browse through a
-corpus and for those who wish to analyze the topic model itself.
+To aid in the visualization of topic models and in pattern discovery in
+document collections, we present The Topic Browser, a tool for interactively
+exploring both the output of a topic modeling algorithm and its attendant
+corpus.  Our topic browser incorporates many visualizations of topic models
+previously published as well as some innovative ideas of our own.  The Topic
+Browser is an aid both to those who wish to browse through a corpus and for
+those who wish to analyze the topic model itself.
 
-The rest of this paper simply describes the Topic Browser.  We demonstrate our
-tool with a topic model run on a collection of about 460 campaign speeches from
-the 2008 presidential primary and general election.
+The rest of this paper describes the Topic Browser.  We demonstrate our tool
+with a topic model run on a collection of about 460 campaign speeches from the
+2008 presidential primary and general elections.
 
@@
 Aside from the documents and the topic model themselves, our browser
-incorporates three other pieces of information: attributes, or metadata,
+incorporates three other pieces of information: attributes (or metadata)
 associated with each document, topic metrics, and document metrics.  Here we
 give a brief description of these kinds of information, deferring a discussion
-of their use in visualization to the following sections.
+of their use in visualization to the subsequent sections.
 
@@
 as the number of tokens and types labeled with the topic, to more complicated
-metrics like how spread out the topic is across documents, or how coherent its
+metrics like how dispersed the topic is across documents, or how coherent its
 words are~\cite{newman-2010-automatic-evaluation-of-topic-coherence}.  We also
@@
 The sidebar in our browser is the main navigation tool.  Apart from the obvious
-task of directing the user to various parts of the visualization tool, then
-main purpose of the sidebar is to list the items of a particular type (such as
-topics or documents).  The new capability that we bring to the of listing the
-items is the ability to sort and filter these lists as the user desires.
+task of directing the user to various parts of the visualization tool, the main
+purpose of the sidebar is to list the items of a particular type (such as
+topics or documents).  The new capability introduced by this browser is the
+ability to sort and filter these lists as the user desires.
 
-When presenting the results of topic models, papers will often cherry-pick the
-list of topics looking for particularly good topics to display, leaving the
-large number of meaningless topics untouched.  While such presentations may
-help the researcher to highlight the benefits of her new topic model, they hide
-the fact that finding good topics is often a laborious process.  With our topic
-and document metrics, the sorting and filtering capabilities of the sidebar
-list allow the user to often go quickly to the very topics he desires to find.
+When presenting the results of a topic model analysis, papers will often
+cherry-pick the list of topics looking for particularly good topics to display,
+leaving the large number of meaningless topics untouched.  While such
+presentations may help the researcher to highlight the benefits of her new
+topic model, they hide the fact that finding good topics is often a laborious
+process.  With our topic and document metrics, the sorting and filtering
+capabilities of the sidebar list allow the user to often go quickly to the very
+topics he desires to find.
 
@@
 one must browse the topics discovered, and previous methods have failed to
-adequately allow the user to explore topics.  Our visualizer presents new ways
+adequately allow the user to explore topics.  Our browser presents new ways
 to gather information about a topic in the corpus, both in how the words in the
-topic are displayed and in other information that is presented about the topic.
+topic are displayed and in that is presented about the topic.
 
@@
 that word in the topic.  We also show a re-weighted word cloud as determined by
-David Blei's Turbo Topics method~\cite{blei-2009-turbo-topics}.  While these
-visualizations are modest improvements over typical visualization methods, our
-most useful display of the words is showing them in their context.
+Blei \& Lafferty's Turbo Topics method~\cite{blei-2009-turbo-topics}.
+
+While these visualizations are modest improvements over typical visualization
+methods, our most useful display of the words is showing them in their context.
 
@@
 inside of their context.  We select a random token of each word type labeled
-with the topic and show up to 50 characters on either side of the token.  We
-also allow the user to cycle through contexts to gain a broader view of how
-each word is actually used in that topic in the corpus.
+with the topic and show up to 50 characters on either side of the token,
+keeping words intact.  We also allow the user to cycle through contexts to gain
+a broader view of how each word is actually used in that topic in the corpus.
 
@@
 
-We give two main ways of getting more information about topics in the model.
-The first is showing documents and attribute values which have the highest
-number of tokens labeled with that topic, or the highest percent in the
-document or attribute.  This allows the user to quickly find documents that
-best demonstrate what the topic captured.  Showing the top values for a given
-attribute (such as authors for the attribute ``Author'') gives the user an idea
-of how focused the topic is and often gives interesting information about the
-corpus being browsed (for example, in Figure \ref{fig:something}, we see that
-Democrats spoke more often about education than Republicans did).
+The browser provides two main ways of getting more information about topics in
+the model.  The first is showing documents and attribute values which have
+either the highest number of tokens labeled with that topic or the highest
+proportion (by percentage) of the given topic.  This allows the user to quickly
+find documents that best demonstrate what the topic captured.  Showing the top
+values for a given attribute (such as authors for the attribute ``Author'')
+gives the user an idea of how focused the topic is and often gives interesting
+information about the corpus being browsed (for example, in Figure
+\ref{fig:something}, we see that Democrats spoke more often about education
+than Republicans did, at least in the topic shown).
 
@@
 
-In addition to allowing the user to browse the topics in the topic model, we
+In addition to enabling the user to browse the topics in the topic model, we
 provide means for browsing the documents, words, and attributes in the corpus.
@@
 highlight the tokens in the document that were labeled with that topic.  All of
-this has been done previously, besides the sorting and filtering of the
+this has been done previously, aside from the sorting and filtering of the
 documents, and in fact constitutes the entirety of most previous corpus
@@
 in the corpus, with context taken from their corresponding documents.  The user
-can also view words independently with a search-like interface, seeing what
-topics and what documents in which the word appears most frequently.  While
-providing basic functionality, however, the search interface leaves much to be
-desired, as only single words can currently be searched for.  We plan on
-expanding that to phrases in the future.
+can also view words independently with a search-like interface, seeing topics
+and documents in which the word appears most frequently.  While providing basic
+functionality, however, the search interface leaves much to be desired, as only
+single words can currently be searched for.  We plan on expanding that to
+phrases.
 
 The user can also look at aggregated information for values of an attribute
-(i.e., a particular author or year in the documents).  This view is also at
-present somewhat limited, showing only what topics and words are used most
-frequently by the collection of documents with that attribute.
+(i.e., a particular author or year in the documents), combining all of the
+topic and word counts for all documents with the given attribute.  This view is
+also at present somewhat limited, showing only what topics and words are used
+most frequently by the collection of documents with that attribute.
 
@@
 
-There has been a wide variety of plots used in the visualization of topic
-models.  We currently include two kinds of plots in our topic browser, with
-plans to implement many more.  The first is showing trends for topics over the
-values of an attribute (such as date, or author), useful for corpus browsing.
-This kind of plot has been used in visualizing topic models almost since their
+We currently include two kinds of plots in our topic browser, with plans to
+implement many more.  The first is showing trends for topics over the values of
+an attribute (such as date, or author), useful for corpus browsing.  This kind
+of plot has been used in visualizing topic models almost since their
 introduction~\cite{griffiths-2004-finding-scientific-topics}.  Our topic
-browser allows the user to very quickly generate these trend plots over any
-topic or combination of topics for any attribute of the documents in the
+browser allows the user to interactively generate these trend plots over any
+attribute for any topic or combination of topics of the documents in the
 corpus.
@@
 This allows the user to see some interesting properties of the topic metrics,
-such as the fact that document entropy seems to correlate logarithmically with
+such as the fact that document entropy seems to correlate with the logarithm of
 the number of tokens in the topic, and that coherence does not seem to
-correlate with any other topic metric.  The user can also found outliers, such
+correlate with any other topic metric.  The user can also find outliers, such
 as topics with low document entropy but a high token count, that can then be
-examined in the topic page.
+examined in the topic page. (TODO: use Obama vs. McCain as example here)
 
@@
 including basic corpus browsing functionality, plots of trends over attributes
-in the corpus, and Blei's turbo topics method of finding significant phrases
-for each topic.  We have also presented many novel ways to get information from
-a topic model in an interactive browsing experience.
+in the corpus, and Blei \& Lafferty's Turbo Topics method of finding
+significant phrases for each topic.  We have also presented many novel ways to
+mine information from a topical analysis of a corpus in an interactive browsing
+experience.
 
@@
 develop new models.  Our tool currently supports any topic model that labels
-individual tokens in the corpus with a topics.  We have plans to include
-special visualizations for more complicated topic models, such as Topics over
-Time, sentiment-topic models, and others.
+individual tokens in the corpus with topics.  We have plans to include special
+visualizations for more complicated topic models, such as Topics over Time,
+sentiment-topic models, hierarchical topic models, and others.
 

### 2010-10-21 masters_thesis/topic_ranking_nips/paper.tex
Renamed the directories to make a bit more sense

--- 2010-10-15 masters_thesis/entropy_nips_workshop/paper.tex
@@
 \begin{abstract}
+
+Topic modeling is a popular and effective method for gathering meaning from a
+set of documents.  Such models contain a lot of information and can be very
+difficult to analyze.  For the proper visualization of these models, one must
+present only the topics that are relevant to the task at hand.  We present the
+novel task of topic ranking in response to user questions that greatly aids in
+the visualization of topic models.  We make use of the document distribution of
+a topic in order to rank the topics according to specific user questions.
+
 \end{abstract}
@@
 could have when looking at a corpus, and \secref{examples} shows examples of
-how this technique provides a proper ranking.  In \secref{challenges} we
-discuss the challenges of evaluating this visualization method.  Finally, in
-\secref{conclusion} we conclude.
+how this technique provides a proper ranking.  Finally, in \secref{conclusion}
+we conclude.
 
@@
 \begin{align}
+  \label{eq:attrentropy}
   p(a|z) = \frac{\sum_{d \in \mathcal{A}} C(d,z)}{\sum_{d \in \mathcal{D}}
@@
 
-\section{Using $p(d|z)$ to rank topics}
+\section{Example topic rankings}
 \label{sec:examples}
 
-We present here a few examples of how the document distribution for a topic can
-be used to rank topics according to their relevance for certain user questions.
-In our examples we use a corpus of 480 campaign speeches from the 2008
-presidential primary and general election.  We inferred 150 topics in the data
-from a Gibbs sampling of the LDA topic model.
+We present here a few examples of questions a user might ask about a corpus and
+how we can rank topics according to their relevance to those questions.  In our
+examples we use a corpus of 480 campaign speeches from the 2008 presidential
+primary and general election.  We inferred 150 topics in the data from a Gibbs
+sampling of the LDA topic model.
 
-Questions here:
+The first question we address is, What topics are most prevelant in the corpus?
+This is a simple question with an obvious answer that does not require using
+$p(d|z)$.  One can simply order the topics by the number of tokens in the
+topic.  Upon doing so one finds the following topics (we show the topic number
+and the top ten words in the topic for the top five topics):
 
-What topics are most prevelant in the corpus?
+\begin{tabular}{l|l}
+  35 & years time made long today place home day ago back \\
+  42 & people don ve make country ll time put didn money \\
+  119 & make national work effort including protect adiministration clear
+	support policy \\
+  15 & america world american future challenges great lead generation nation
+	leadership \\
+  108 & president work day senator house make white clinton care support \\
+\end{tabular}
 
-What topics did John McCain talk about the most?
+One can see that these topics for the most part are meaningless, capturing stop
+words and noise in the data and hardly answering the question at hand.  When we
+filter by a minimum value of coherence, however, we see a better picture:
 
-What topics did Barack Obama most consistently throughout his campaign?
+\begin{tabular}{l|l}
+  42 & people don ve make country ll time put didn money \\
+  119 & make national work effort including protect adiministration clear
+	support policy \\
+  120 & law government rights respect case society fact laws principles order \\
+  143 & jobs ve create economy make years billion companies invest energy \\
+  101 & children family day man mother care love father parents young \\
+\end{tabular}
 
-What topics are most consistent across candidates?
+While still somewhat noisy, this list gives the viewer some idea of what was
+spoken of most during the 2008 presidential election.
 
-What topics were constant throughout the campaign?
+Next we look at a similar question: What topics did John McCain talk about the
+most?  This question is also answered without needing to use $p(d|z)$; simply
+order the topics by the number of tokens that occur in speeches given by
+McCain.  The list is the following (here again we filter by coherence):
 
-What topics were transitory, only spoken of rarely?
+\begin{tabular}{l|l}
+  129 & mccain news john senator obama remarks america events donate email \\
+  119 & make national work effort including protect adiministration clear
+	support policy \\
+  27 & mccain news john search media speeches www johnmccain email freedom \\
+  90 & war iraq honor defeat enemy greater responsibility responsibilities
+	difficult great \\
+  66 & dollars reform federal washington congress reforms public policy
+	problems system \\
+\end{tabular}
 
-\section{Challenges of evaluation}
-\label{sec:challenges}
+Another similar, though slightly distinct, question is, What topics did Barack
+Obama speak of most consistently throughout his campaign?  This question
+differs from the previous in that the previous question could be answered by a
+topic Obama used just a few times in long speeches.  This question looks for
+topics that were consistent in all of his speeches.  Here we turn to the
+document entropy for each topic as in \eqref{eq:entropy}, looking only at
+speeches given by Obama.  The result of such an ordering is:
 
-The obvious way to evaluate a topic ranking is to use the evaluations methods
-commonly found in information retrieval tasks.  There are well established
-techniques to evaluate a ranking given a true ranking.  However, there are
-several obstacles to using such a scheme to evaluate performance on this task.
+\begin{tabular}{l|l}
+  60 &  barack remarks senator barackobama obama ll www http hope states\\
+  47 & mccain senator john ll bush washington states election won george \\
+  124 & change politics country time american washington ve common americans
+	end \\
+  84 & delegates south future party carolina kennedy support pledged total
+	versus \\
+  143 & jobs ve create economy make years billion companies invest energy \\
+\end{tabular}
 
-The first and glaring obstacle is that there is no set of data that has been
-labeled with true rankings for these or any questions.  One cannot compare the
-output of an algorithm to ground truth when there is no ground truth.  It is
-not even clear how such data would be obtained; perhaps one could take a topic
-model and a set of questions and rank each topic for relevance to each
-question, but such a task would be incredibly laborious and error-prone.  The
-labeler would first have to understand what exactly each topic is capturing, in
-itself a difficult task, then go through the entire data collection to answer
-the question herself and determine how well the topic matches.  Clearly such a
-process for creating a set of evaluation data is not feasible.
+The first topic consists of markup on all of Barack Obama's speeches.  From the
+rest of the topics one can see that Obama spoke a lot about his opponent,
+Senator McCain, about change and the economy, and he apparently spent a lot of
+time in South Carolina.
 
-But what if the topic model fails to capture topics that truly answer the
-question?  
-Separate topics can be overlapping in words, where a human would consider them
-the same topic.
+Another question one may have about these speeches is, What topics are most
+consistent across candidates?  Each candidate has issues particular to them,
+but what topics are universally used by the candidates?  A natural ranking for
+this question is to order by entropy over candidates, as in
+\eqref{eq:attrentropy}.  Such a ranking shows these to be universal topics:
 
-The topic model itself could be high or low quality, and that affects the
-results.
+\begin{tabular}{l|l}
+  8 & congress government federal bill legislation members house conference
+	statement congressional \\
+  104 & states united countries international support economic state force
+	military community \\
+  103 & public political understand values home play rules shared committed
+	important \\
+  93 & president bush administration george war failed stop department
+	september moral \\
+  18 & result led group death allowed consequences early continued quickly
+	balance \\
+\end{tabular}
+
+Similarly, one could ask, What topics were constant throughout the campaign?
+This question is also answered with \eqref{eq:attrentropy}, with the month of
+the speech used to group the documents together.  We see these topics as a
+result:
+
+\begin{tabular}{l|l}
+  120 & law government rights respect case society fact laws principles order \\
+  8 & congress government federal bill legislation members house conference
+	statement congressional \\
+  25 & party vote election campaign votes people democrats state democratic
+	candidate \\
+  101 & children family day man mother care love father parents young \\
+  38 & economy economic american world security prosperity trade market markets
+	growth \\
+\end{tabular}
 
@@
 
-TODO
+Displaying the right topics to a user is essential to the effective
+visualization of topic models.  We have presented the novel task of ranking
+topics according to relevance to certain user questions.  We have also given a
+method for ranking topics in order to answer a few specific questions.  Such
+rankings allow visualizations of topic models that present information to users
+in much more effective ways than was previously possible.
+
+While our methods are useful for a few user questions, it remains as future
+work to devise a method for ranking topics in answer to arbitrary questions.
+The evaluation of such rankings is also a difficult task that must be left to
+future work.
 

### 2010-10-21 masters_thesis/browser_nips/paper.tex
Finished a draft of the browser paper with pictures

--- 2010-10-21 masters_thesis/tool_nips_workshop/paper.tex
@@
 document entropy to find topics that were used widely (or narrowly, if that is
-desired) throughout the corpus.  She could also filter by document or
-attribute, seeing only topics that were used in a particular document or by a
-particular author.  Documents can similarly be filtered and sorted, showing
-only documents that contain tokens with a particular topic, or with a
-particular value of an attribute.
+desired) throughout the corpus.  For example, Figure~\ref{fig:sidebar} shows
+that the family was one of the most consistent themes in the speeches.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.34\textwidth]{sidebar}
+  \caption{The navigational sidebar showing topics sorted by dispersion
+  throughout the corpus and filtered by coherence}
+  \label{fig:sidebar}
+\end{figure}
+
+The user can also filter by document or attribute, seeing only topics that were
+used in a particular document or by a particular author.  Documents can
+similarly be filtered and sorted, showing only documents that contain tokens
+with a particular topic, or with a particular value of an attribute.
 
@@
 a broader view of how each word is actually used in that topic in the corpus.
+When viewing the topic about troops in Iraq, for example (see
+Figure~\ref{fig:context}), one can see that most often when the word ``troops''
+is used in this topic, it is in reference to bringing the troops home, and when
+``end'' is used, it is not used in the context of ``end the war'' as often as
+one might expect.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\textwidth]{words_in_context}
+  \caption{The top ten words in the topic shown with a random context selected
+  from the corpus.}
+  \label{fig:context}
+\end{figure}
 
@@
 gives the user an idea of how focused the topic is and often gives interesting
-information about the corpus being browsed (for example, in Figure
-\ref{fig:something}, we see that Democrats spoke more often about education
-than Republicans did, at least in the topic shown).
+information about the corpus being browsed.  For example, in Figure
+\ref{fig:top-values}, we see that Democrats spoke more often about education
+than Republicans did, at least in the topic shown.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{top_values}
+  \caption{Part of the overall topic view, looking at a topic about education.
+  Note particularly the top documents for the topic and the top values for the
+  attribute ``party,'' shown at the bottom.}
+  \label{fig:top-values}
+\end{figure}
 
@@
 and looking by word shows topics that have similar words, possibly because the
-two topics really should have been one topic.
+two topics really should have been one topic.  When looking at a topic about
+health care, the user can see that other topics used together with the health
+care topic include topics about the cost of medication and the quality of care
+(see Figure~\ref{fig:similar}).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{similar_topics}
+  \caption{Another look at the overall topic view, featuring different parts of
+  the view.  Note particularly the similar topics box near the top right.}
+  \label{fig:similar}
+\end{figure}
 
@@
 distribution.  When looking at the document in the context of a topic, we also
-highlight the tokens in the document that were labeled with that topic.  All of
-this has been done previously, aside from the sorting and filtering of the
-documents, and in fact constitutes the entirety of most previous corpus
-browsers based on topic models---this functionality is all that is generally
-provided.
+highlight the tokens in the document that were labeled with that topic.  For
+example, the user might be curious to see a document that best demonstrates the
+health care topic mentioned above.  Clicking on the top document in the document list brings the user to the view shown in Figure~\ref{fig:document}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{document}
+  \caption{The document view, looking at a speech by Hillary Clinton in the
+  context of a topic about health care (the rest of the document is cut off in
+  this screenshot).}
+  \label{fig:document}
+\end{figure}
+
+These document visualizations been done previously, and in fact the view in
+Figure~\ref{fig:document} constitutes the entirety of most previous corpus
+browsers based on topic models.  This functionality is all that is generally
+provided in other browsers, though it is just a small part of ours.
 
@@
 can also view words independently with a search-like interface, seeing topics
-and documents in which the word appears most frequently.  While providing basic
-functionality, however, the search interface leaves much to be desired, as only
-single words can currently be searched for.  We plan on expanding that to
-phrases.
+and documents in which the word appears most frequently.  The user examining
+the health care topic might be curious where else the word ``health'' was used
+in the topics and the corpus.  Figure~\ref{fig:word} shows the result of using
+our word search to answer that question.  While providing basic functionality,
+however, the search interface leaves much to be desired, as only single words
+can currently be searched for.  We plan on expanding that to phrases.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.72\textwidth]{word}
+  \caption{The word search functionality, showing the results for searching for
+  the word ``health.''}
+  \label{fig:word}
+\end{figure}
 
@@
 most frequently by the collection of documents with that attribute.
+Figure~\ref{fig:attribute} shows us, among other things, that one of Barack
+Obama's top topics was change in politics.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{attribute}
+  \caption{The attribute view, showing aggregated information for all speeches
+  given by Barack Obama.}
+  \label{fig:attribute}
+\end{figure}
 
@@
 attribute for any topic or combination of topics of the documents in the
-corpus.
+corpus.  Our user exploring health care topics may want to view how much each
+candidate spoke about health care.  We saw already that there were three
+related topics that mentioned health care, so the user views all of them
+together in a histogram, shown in Figure~\ref{fig:trend-plot}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{trend_plot}
+  \caption{A plot of topics over attributes, showing the use of three health
+  care-related topics across candidates.}
+  \label{fig:trend-plot}
+\end{figure}
 
@@
 as topics with low document entropy but a high token count, that can then be
-examined in the topic page. (TODO: use Obama vs. McCain as example here)
+examined in the topic page.
+
+An interesting application of these topic metric plots occurs when the metrics
+include how consistently each candidate spoke about each topic.
+Figure~\ref{fig:metric-plot} shows a plot of topics, comparing John McCain's
+use of the topic to Barack Obama's use of the topic.  Topics in the upper left
+were topics unique to Obama, and topics in the lower right were unique to
+McCain.  Topics in the middle (near the value of 2 for each candidate) were
+somewhat shared between the two.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{metric_plot}
+  \caption{A topic metric comparison plot.  The metrics used were how
+  consistently Barack Obama and John McCain used each topic. The plot shows
+  topics that were unique to Obama or McCain and topics they shared.}
+  \label{fig:metric-plot}
+\end{figure}
 

### 2010-10-23 masters_thesis/browser_nips/paper.tex
Pretty close to a final draft of the browser paper

@@
 Brigham Young University \\
-Provo, UT 84604 \\
 \texttt{mjg82@byu.edu} \\
 \And
-Other authers\ldots \\
-Affiliation \\
-Address \\
-\texttt{email} \\
+Joshua Lutes \\
+Department of Computer Science \\
+Brigham Young University \\
+\texttt{joshua@lutes.me} \\
+\And
+Jeff Lund \\
+Department of Computer Science \\
+Brigham Young University \\
+\texttt{jefflund@gmail.com} \\
+\And
+Josh Hansen \\
+Department of Computer Science \\
+Brigham Young University \\
+\texttt{jjfresh@byu.edu} \\
+\And
+Dan Walker \\
+Department of Computer Science \\
+Brigham Young University \\
+\texttt{danwalkeriv@gmail.com} \\
+\And
+Eric Ringger \\
+Department of Computer Science \\
+Brigham Young University \\
+\texttt{ringger@cs.byu.edu} \\
+\And
+Kevin Seppi \\
+Department of Computer Science \\
+Brigham Young University \\
+\texttt{kseppi@lcs.byu.edu} \\
 }
 
-%\nipsfinalcopy % Uncomment for camera-ready version
-
 \begin{document}
+\nipsfinalcopy % Uncomment for camera-ready version
 \maketitle
@@
 
-Topic models have been shown to capture a measure of the semantics in a corpus.
-Many individualized and cherry-picked visualizations of topic models have been
-reported in the literature, showing the potential of topic models to give
-valuable insight into a corpus.  However, good, general tools for browsing the
-entire output of a topic model along with the analyzed corpus have been
-lacking.  We present an interactive tool that incorporates both prior work in
-displaying topic models as well as some novel ideas that greatly enhance the
-visualization of these models.
+Topic models have been shown to reveal the semantic content in large corpora.
+Many individualized visualizations of topic models have been reported in the
+literature, showing the potential of topic models to give valuable insight into
+a corpus.  However, good, general tools for browsing the entire output of a
+topic model along with the analyzed corpus have been lacking.  We present an
+interactive tool that incorporates both prior work in displaying topic models
+as well as some novel ideas that greatly enhance the visualization of these
+models.
 
@@
 Proper visualizations are essential to extracting information from and
-identifying trends in data, especially large, highly dimensional data.  Large
-text corpora are particularly difficult to visualize, as they typically include
-thousands of documents and perhaps millions of words.  Topic
+identifying trends in data, especially large, high dimensional data.  Large
+text corpora are particularly difficult to visualize, as they may include many
+thousands of documents and millions of words.  Topic
 modeling~\cite{blei-2003-latent-dirichlet-allocation} is a method of reducing
@@
 visualization~\cite{blei-2009-topic-models,
-newman-2010-visualizing-with-topic-maps}.  However, in general the presentation
-of topic models has been limited to hand selected topics that show the model in
-the best light possible.  These presentations leave the viewer wondering what
-else the topic model discovered and provide little help in gaining a deep
-understanding of the model.
+newman-2010-visualizing-with-topic-maps}.  However, in many papers the
+presentation of topic models has been limited to hand-selected, coherent
+topics, when often there are many meaningless topics to sift through.  These
+presentations leave the viewer wondering what else the topic model discovered
+and provide little help in gaining a deep understanding of the model.
 
 To aid in the visualization of topic models and in pattern discovery in
-document collections, we present The Topic Browser, a tool for interactively
-exploring both the output of a topic modeling algorithm and its attendant
-corpus.  Our topic browser incorporates many visualizations of topic models
-previously published as well as some innovative ideas of our own.  The Topic
-Browser is an aid both to those who wish to browse through a corpus and for
-those who wish to analyze the topic model itself.
-
-The rest of this paper describes the Topic Browser.  We demonstrate our tool
-with a topic model run on a collection of about 460 campaign speeches from the
-2008 presidential primary and general elections.
+document collections, we present The Topic Browser, a web-based tool for
+interactively exploring both the output of a topic modeling algorithm and its
+attendant corpus.  Our topic browser incorporates many visualizations of topic
+models previously published as well as some innovative ideas of our own.  The
+Topic Browser is an aid both to those who wish to browse through a corpus and
+for those who wish to analyze the topic model itself.  We believe that this
+interactive approach to visualization in browsing reveals topics and topical
+trends in large document collections in a way that is not possible with static
+visualizations.  The rest of this paper describes the Topic Browser in an
+attempt to demonstrate that this is tool.  We display our tool with a topic
+model run using MALLET's implementation of LDA~\cite{mallet} with 150 topics on
+a collection of about 460 campaign speeches from the 2008 presidential primary
+and general elections retrieved from http://2008election.procon.org.
 
@@
 Aside from the documents and the topic model themselves, our browser
-incorporates three other pieces of information: attributes (or metadata)
+incorporates three other pieces of information: attributes (metadata)
 associated with each document, topic metrics, and document metrics.  Here we
@@
 that give information about the topic.  These range from simple metrics, such
-as the number of tokens and types labeled with the topic, to more complicated
-metrics like how dispersed the topic is across documents, or how coherent its
-words are~\cite{newman-2010-automatic-evaluation-of-topic-coherence}.  We also
-use pairwise topic metrics as similarity measures to show similar topics.
+as the number of words tokens and types labeled with the topic, to more
+complicated metrics such as how dispersed the topic is across documents, or how
+coherent its words
+are~\cite{newman-2010-automatic-evaluation-of-topic-coherence}.  We also use
+pairwise topic metrics as similarity measures to show similar topics.
 
 Similar to topic metrics, one can also compute document metrics.  Beyond simple
-metrics like token count in the document, these can include things such as the
+metrics like token count in the document, these include things such as the
 entropy of the topic distribution of the
@@
 
-The sidebar in our browser is the main navigation tool.  Apart from the obvious
-task of directing the user to various parts of the visualization tool, the main
-purpose of the sidebar is to list the items of a particular type (such as
-topics or documents).  The new capability introduced by this browser is the
-ability to sort and filter these lists as the user desires.
+The sidebar in our browser is the main navigation tool.  To facilitate
+navigation through the large amount of information in the browser, the sidebar
+lists the items of the type currently being explored.  The new capability
+introduced by this browser is the ability to sort and filter these lists as the
+user desires.
 
-When presenting the results of a topic model analysis, papers will often
-cherry-pick the list of topics looking for particularly good topics to display,
-leaving the large number of meaningless topics untouched.  While such
-presentations may help the researcher to highlight the benefits of her new
-topic model, they hide the fact that finding good topics is often a laborious
-process.  With our topic and document metrics, the sorting and filtering
-capabilities of the sidebar list allow the user to often go quickly to the very
-topics he desires to find.
+When presenting the results of a topic model analysis, papers will often hand
+pick particularly good topics to display, leaving out a large number of
+meaningless topics.  While such presentations may help to highlight the
+benefits of a new topic model, they hide the fact that finding good topics is
+often a laborious process.  With our topic and document metrics, the sorting
+and filtering capabilities of the sidebar list allow the user to go quickly to
+meaningful topics that are relevant to his questions.
 
 When browsing through topics, the user can filter the topic list by coherence
-to eliminate from the view topics that are mostly meaningless, and sort by
-document entropy to find topics that were used widely (or narrowly, if that is
-desired) throughout the corpus.  For example, Figure~\ref{fig:sidebar} shows
-that the family was one of the most consistent themes in the speeches.
+to eliminate from the view those topics that are mostly meaningless and sort by
+document entropy (a measure of the dispersion of the topic across the
+documents) to find topics that were used widely throughout the corpus.  For
+example, Figure~\ref{fig:sidebar} shows that ``the family'' was one of the most
+consistent themes in the speeches.  We use the top two words to name a topic in
+the sidebar and other places, to avoid clutter, though we are actively
+investigating better ways to name topics.
 
@@
   \includegraphics[width=.34\textwidth]{sidebar}
-  \caption{The navigational sidebar showing topics sorted by dispersion
-  throughout the corpus and filtered by coherence}
+  \caption{The navigational sidebar showing topics.  In this example the topics
+  are sorted by dispersion throughout the corpus and filtered by coherence}
   \label{fig:sidebar}
@@
 
-The user can also filter by document or attribute, seeing only topics that were
+The user can also filter by document or attribute to show only topics that were
 used in a particular document or by a particular author.  Documents can
 similarly be filtered and sorted, showing only documents that contain tokens
-with a particular topic, or with a particular value of an attribute.
+with a particular topic or with a particular value of an attribute.
 
@@
 
-Topic models are typically presented as static lists of words, often simply
-showing the top ten words from each topic in a list.  While this practice is
+The results from a topic model are typically presented as static lists of
+words, often simply showing the top ten words from each topic in a
+list~\cite{blei-2003-latent-dirichlet-allocation}.  While this practice is
 often sufficient to give a reader a basic idea of what the topic captured, it
@@
 one must browse the topics discovered, and previous methods have failed to
-adequately allow the user to explore topics.  Our browser presents new ways
-to gather information about a topic in the corpus, both in how the words in the
-topic are displayed and in that is presented about the topic.
+adequately allow the user to explore topics.  Our browser presents new ways to
+gather information about a topic in the corpus, both in terms of how the words
+in the topic are displayed and in terms of what is presented about the topic.
 
@@
 methods, our most useful display of the words is showing them in their context.
-
 The topics in a topic model cannot be fully interpreted when completely
-separated from their context.  Thus we display the top ten words in the topic
-inside of their context.  We select a random token of each word type labeled
-with the topic and show up to 50 characters on either side of the token,
-keeping words intact.  We also allow the user to cycle through contexts to gain
-a broader view of how each word is actually used in that topic in the corpus.
-When viewing the topic about troops in Iraq, for example (see
-Figure~\ref{fig:context}), one can see that most often when the word ``troops''
-is used in this topic, it is in reference to bringing the troops home, and when
-``end'' is used, it is not used in the context of ``end the war'' as often as
-one might expect.
+separated from their context.  Thus in addition to the word cloud we also
+display the top ten words in the topic inside of their context.  We select a
+random token of each word type labeled with the topic and show up to 50
+characters on either side of the token, keeping words intact.  We also allow
+the user to cycle through contexts to gain a broader view of how each word is
+actually used in that topic.  When viewing the topic about troops in Iraq, for
+example (see Figure~\ref{fig:context}), one can see that most often when the
+word ``troops'' is used in this topic, it is in reference to bringing the
+troops home, and when ``end'' is used, it is not used in the context of ``end
+the war'' as often as one might expect.
 
@@
 values for a given attribute (such as authors for the attribute ``Author'')
-gives the user an idea of how focused the topic is and often gives interesting
-information about the corpus being browsed.  For example, in Figure
+gives the user an idea of how focused the topic is and often reveals
+interesting information about the corpus being browsed.  For example, in Figure
 \ref{fig:top-values}, we see that Democrats spoke more often about education
@@
 The other way that we give more information about topics is by showing similar
-topics.  We find similar topics either by looking at the documents that the
-topics show up in or by the words that the topics use.  Looking at similar
-topics by document shows topics that are commonly used together in the corpus,
-and looking by word shows topics that have similar words, possibly because the
-two topics really should have been one topic.  When looking at a topic about
-health care, the user can see that other topics used together with the health
-care topic include topics about the cost of medication and the quality of care
-(see Figure~\ref{fig:similar}).
+topics.  We find similar topics either by looking at the distribution of
+documents that contain the topic or by the topic's distribution over words.
+Looking at similar topics by document shows topics that are commonly used
+together in the corpus, and finding similar topics by word distribution shows
+topics that have similar words, possibly because the two topics really should
+have been one topic.  When looking at a topic about health care, the user can
+see that other topics used together with the health care topic include topics
+about the cost of medication and the quality of care (see
+Figure~\ref{fig:similar}).
 
@@
 provide means for browsing the documents, words, and attributes in the corpus.
-These facilities give users means to explore the corpus itself in the context
-of the topic model.
+These facilities give users the ability to explore the corpus itself in the
+context of the topic model.
 
@@
 example, the user might be curious to see a document that best demonstrates the
-health care topic mentioned above.  Clicking on the top document in the document list brings the user to the view shown in Figure~\ref{fig:document}.
+health care topic mentioned above.  Clicking on the top document in the
+document list (as in the bottom of Figure~\ref{fig:top-values}) brings the user
+to the view shown in Figure~\ref{fig:document}.
 
@@
   context of a topic about health care (the rest of the document is cut off in
-  this screenshot).}
+  this screenshot, and sadly, the colored tokens in the document are not very
+  visible in black and white).}
   \label{fig:document}
@@
 
-These document visualizations been done previously, and in fact the view in
+The document visualizations reported here were drawn from the work of
+others~\cite{blei-2003-latent-dirichlet-allocation}, and in fact the view in
 Figure~\ref{fig:document} constitutes the entirety of most previous corpus
-browsers based on topic models.  This functionality is all that is generally
-provided in other browsers, though it is just a small part of ours.
+browsers based on topic models (e.g.,~\cite{blei-arxiv-corpus-browser}).  Our
+browser substantially goes beyond the existing functionality reported by
+others.
 
@@
 the health care topic might be curious where else the word ``health'' was used
-in the topics and the corpus.  Figure~\ref{fig:word} shows the result of using
-our word search to answer that question.  While providing basic functionality,
-however, the search interface leaves much to be desired, as only single words
-can currently be searched for.  We plan on expanding that to phrases.
+in the topics and in the corpus.  Figure~\ref{fig:word} shows the result of
+using our word search to answer that question.  While providing basic
+functionality, however, the search interface leaves much to be desired, as only
+single words can currently be searched for.  We plan on expanding that to
+phrases.
 
@@
 
-The user can also look at aggregated information for values of an attribute
-(i.e., a particular author or year in the documents), combining all of the
-topic and word counts for all documents with the given attribute.  This view is
-also at present somewhat limited, showing only what topics and words are used
-most frequently by the collection of documents with that attribute.
+The user can also look at aggregated information for the values of an attribute
+(i.e., a particular candidate or party), combining all of the topic and word
+counts for all documents with the given attribute.  This view is also at
+present somewhat limited, showing only what topics and words are used most
+frequently by the collection of documents with that attribute.
 Figure~\ref{fig:attribute} shows us, among other things, that one of Barack
-Obama's top topics was change in politics.
+Obama's top topics was ``change in politics.''
 
@@
 We currently include two kinds of plots in our topic browser, with plans to
-implement many more.  The first is showing trends for topics over the values of
-an attribute (such as date, or author), useful for corpus browsing.  This kind
+implement many more.  The first shows trends for topics over the values of an
+attribute (such as date, or candidate), useful for corpus browsing.  This kind
 of plot has been used in visualizing topic models almost since their
@@
 browser allows the user to interactively generate these trend plots over any
-attribute for any topic or combination of topics of the documents in the
-corpus.  Our user exploring health care topics may want to view how much each
-candidate spoke about health care.  We saw already that there were three
-related topics that mentioned health care, so the user views all of them
-together in a histogram, shown in Figure~\ref{fig:trend-plot}
+attribute for any topic or combination of topics in the corpus.  Our user
+exploring health care topics may want to view how much each candidate spoke
+about health care.  We saw already that there were three related topics that
+mentioned health care, so the user might view all of them together in a
+histogram, shown in Figure~\ref{fig:trend-plot}.
 
@@
 Figure~\ref{fig:metric-plot} shows a plot of topics, comparing John McCain's
-use of the topic to Barack Obama's use of the topic.  Topics in the upper left
+use of each topic to Barack Obama's use of the topic.  Topics in the upper left
 were topics unique to Obama, and topics in the lower right were unique to
@@
   \includegraphics[width=\textwidth]{metric_plot}
-  \caption{A topic metric comparison plot.  The metrics used were how
+  \caption{A topic metric comparison plot.  The metrics plotted are how
   consistently Barack Obama and John McCain used each topic. The plot shows
@@
 mine information from a topical analysis of a corpus in an interactive browsing
-experience.
+experience.  The Topic Browser is an effective tool both for those wishing to
+browse through a corpus in the context of a topic model, and for those wishing
+to better understand topic models and develop new models.
 
-While our tool is still under development, we intend to make it open and freely
-available, both for those wishing to browse through a corpus in the context of
-a topic model, and for those wishing to better understand topic models and
-develop new models.  Our tool currently supports any topic model that labels
-individual tokens in the corpus with topics.  We have plans to include special
-visualizations for more complicated topic models, such as Topics over Time,
-sentiment-topic models, hierarchical topic models, and others.
+While our tool is still under development, a description of the Topic Browser,
+a working demo, and the current version of the code are available at
+http://nlp.cs.byu.edu/topic\_browser [Note to reviewers: this is currently just
+a demo of the browser; by the time of the workshop it will have the description
+and code mentioned].  Our tool currently supports any topic model that labels
+individual tokens in the corpus with topics, and is built to import data
+directly from MALLET input and output files~\cite{mallet}, or files similarly
+formatted.  We have plans to include specialized visualizations for more
+complicated topic models, such as Topics over Time, sentiment-topic models,
+hierarchical topic models, and others.
 

### 2010-10-24 masters_thesis/topic_ranking_nips/paper.tex
Added some citations to the topic ranking paper

@@
 meaning from a collection of
-documents~\cite{blei-2003-latent-dirichlet-allocation}.  Topic models have been
-used as an inputs to other kinds of algorithms, such as classification and
-sentiment tasks (cite some papers here), but the usefulness of topic models as
-visualization tools in and of themselves has not been well established, and is
-even in question (cite more papers).  Some recent work has been done in an
-attempt to make corpus browsers using the results of LDA, and to make the
-results of LDA more easily interpretable, but there is much work yet to be done
-(still more citations).
+documents~\cite{griffiths-2004-finding-scientific-topics}.  Topic models have
+been used as an inputs to other kinds of algorithms, such as classification and
+sentiment tasks~\cite{blei-2003-latent-dirichlet-allocation,
+brody-2010-aspect-sentiment-model-for-reviews}, but the usefulness of topic
+models as visualization tools in and of themselves has not been well
+established, and is even in
+question~\cite{newman-2010-visualizing-with-topic-maps}.  Some recent work has
+been done in an attempt to make corpus browsers using the results of LDA, and
+to make the results of LDA more easily interpretable, but there is much work
+yet to be done~\cite{blei-arxiv-corpus-browser,
+newman-2010-visualizing-with-topic-maps,
+mei-2007-automatic-labeling-of-topic-models}.
 

### 2010-10-24 masters_thesis/browser_nips/paper.tex
Fixed a few things with the authors
Some minor revisions

@@
 Brigham Young University \\
-\texttt{joshua@lutes.me} \\
+\texttt{joshua@byu.edu} \\
 \And
@@
 Brigham Young University \\
-\texttt{kseppi@lcs.byu.edu} \\
+\texttt{kseppi@cs.byu.edu} \\
 }
@@
 visualizations.  The rest of this paper describes the Topic Browser in an
-attempt to demonstrate that this is tool.  We display our tool with a topic
+attempt to demonstrate that this is true.  We display our tool with a topic
 model run using MALLET's implementation of LDA~\cite{mallet} with 150 topics on
@@
 that give information about the topic.  These range from simple metrics, such
-as the number of words tokens and types labeled with the topic, to more
+as the number of word tokens and types labeled with the topic, to more
 complicated metrics such as how dispersed the topic is across documents, or how

### 2010-10-28 masters_thesis/browser_nips/paper.tex
A couple of minor changes to the browser paper

@@
 Brigham Young University \\
-\texttt{joshua@byu.edu} \\
+\texttt{jlutes@byu.edu} \\
 \And
@@
 including basic corpus browsing functionality, plots of trends over attributes
-in the corpus, and Blei \& Lafferty's Turbo Topics method of finding
-significant phrases for each topic.  We have also presented many novel ways to
-mine information from a topical analysis of a corpus in an interactive browsing
-experience.  The Topic Browser is an effective tool both for those wishing to
-browse through a corpus in the context of a topic model, and for those wishing
-to better understand topic models and develop new models.
+in the corpus, the use of coherence to ignore meaningless topics, and Blei \&
+Lafferty's Turbo Topics method of finding significant phrases for each topic.
+We have also presented many novel ways to mine information from a topical
+analysis of a corpus in an interactive browsing experience.  The Topic Browser
+is an effective tool both for those wishing to browse through a corpus in the
+context of a topic model, and for those wishing to better understand topic
+models and develop new models.
 

### 2010-11-10 applications/nsf_fellowship/previous_research.tex
Some minor work on the NSF

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\begin{document}

Four of the items listed from my undergraduate career all came from a single
project focusing on how to effectively solve long-running optimization problems
with a large cluster of hundreds or thousands of machines.  The project started
when a physics professor approached my advisor seeking help solving a
crystallography problem.  Having just begun working with my advisor, I accepted
the problem mostly in an auxiliary role, writing some code and running
experiments for Dr. Campbell, the physics professor.  We used particle swarm
optimization, a bio-inspired optimization algorithm similar to evolutionary
algorithms, to try to fit to data his model of what was happening in a
particular kind of crystal.  After spending a lot of time seeing the output of
the algorithm, I also helped to interpret the results.  I discovered that we
had too many free parameters in our model and helped Dr. Campbell to make
progress in solving his problem.

The crystallography problem we were solving required the evaluation of a
function that took several minutes to run.  As each particle in the algorithm
must evaluate the function at every iteration, we used several hundred
processors on BYU's supercomputer in our optimization of the function.  This
led us to ask some interesting questions about how we could best use those
processors to solve the problem.  We first experimented with the topology of
the algorithm, or how the particles in the swarm communicate with each other.
I did much of the experimentation, though I was directed by a graduate student
in the lab, who eventually wrote the paper that we published.  We saw some
improvements in the performance of the algorithm with our new topologies,
though we also noticed that increasing the number of particles with additional
processors had diminishing returns.

To solve this problem, my advisor proposed applying the idea of speculative
decomposition in the parallelization literature to particle swarm optimization
(PSO).  Instead of increasing the number of particles in the swarm, additional
processors could be used to evaluate possible next-iteration positions of each
particle in parallel.  I thought it was an interesting idea, and I made it my
honors thesis.  I took the idea and discovered how it could actually be done,
performing the evaluations needed and piecing the results back together from
messages sent between processors so that speculative evaulation could be done
in a distributed PSO environment with only a single round of message passing
per iteration.  After implementing the technique, I ran many experiments to
compare our methods to other means of parallelizing PSO.  Once I was satisfied
that my results were significant, I wrote, submitted, and successfully defended
my honors thesis.  I also modified the paper to fit in the aims and page
requirements of Parallel Problem Solving in Nature, a well-respected parallel
optimization conference, and submitted my thesis there, and it was accepted.

In implementing speculative evaluation, however, I noticed a number of ways the
method could potentially be improved.  Even though my honors thesis was
completed, I continued my investigation.  I found that a large number of the
speculative evaluations we performed were wasted, and that the algorithm could
be improved by replacing useless evaluations with better evaluations, such as
speculating several iterations ahead on paths that particles were likely to
take.  I significantly expanded the work of my honors thesis, preparing to
submit it to a journal on optimization.  After a rejection of a smaller version
from a conference, I got quite discouraged, wanting to give up and move on to
something new.  My advisor insisted I persevere, saying it was good
preparation for a PhD, and so I continued.  After running more experiments to
strengthen the paper, we submitted it to the journal Swarm Intelligence, where
it is currently under review.

While I have since move on to other interests, this project was my first
experience performing research and dealing with the submission process at
conferences.  All told, I worked on particle swarm optimization for about two
and a half years.  I learned a lot about what research is, the perseverance
required in getting a PhD, and how much I love exploring new topics and
extending the limits of what we can do.  This project set the stage for my
masters thesis and for my future work as a PhD student.

\end{document}

### 2010-11-10 masters_thesis/browser_nips/paper.tex
Updates for the final version of the browser paper

@@
 \author{
-Matthew J. Gardner \\
+M. Gardner, J. Lutes, J. Lund, J. Hansen, D. Walker, E. Ringger, and K. Seppi\\
 Department of Computer Science \\
@@
 \texttt{mjg82@byu.edu} \\
-\And
-Joshua Lutes \\
-Department of Computer Science \\
-Brigham Young University \\
-\texttt{jlutes@byu.edu} \\
-\And
-Jeff Lund \\
-Department of Computer Science \\
-Brigham Young University \\
-\texttt{jefflund@gmail.com} \\
-\And
-Josh Hansen \\
-Department of Computer Science \\
-Brigham Young University \\
-\texttt{jjfresh@byu.edu} \\
-\And
-Dan Walker \\
-Department of Computer Science \\
-Brigham Young University \\
-\texttt{danwalkeriv@gmail.com} \\
-\And
-Eric Ringger \\
-Department of Computer Science \\
-Brigham Young University \\
-\texttt{ringger@cs.byu.edu} \\
-\And
-Kevin Seppi \\
-Department of Computer Science \\
-Brigham Young University \\
-\texttt{kseppi@cs.byu.edu} \\
 }
@@
 a working demo, and the current version of the code are available at
-http://nlp.cs.byu.edu/topic\_browser [Note to reviewers: this is currently just
-a demo of the browser; by the time of the workshop it will have the description
-and code mentioned].  Our tool currently supports any topic model that labels
-individual tokens in the corpus with topics, and is built to import data
-directly from MALLET input and output files~\cite{mallet}, or files similarly
-formatted.  We have plans to include specialized visualizations for more
-complicated topic models, such as Topics over Time, sentiment-topic models,
-hierarchical topic models, and others.
+http://nlp.cs.byu.edu/topic\_browser.  Our tool currently supports any topic
+model that labels individual tokens in the corpus with topics, and is built to
+import data directly from MALLET input and output files~\cite{mallet}, or files
+similarly formatted.  We have plans to include specialized visualizations for
+more complicated topic models, such as Topics over Time, sentiment-topic
+models, hierarchical topic models, and others.
 

### 2010-11-11 resume_cv/cv.tex
Added my masters degree to my CV
Some updates to my cv from comments from Sabrina

@@
 %
-Particle swarm optimization, natural language processing, Latent Dirichlet
-Allocation, Bayesian machine learning, Markov decision processes, value
-iteration
+Natural language processing, latent Dirichlet allocation, topic modeling,
+Bayesian machine learning, particle swarm optimization
 
@@
 \begin{outerlist}
+\item[] M.S.,
+		Computer Science, April 2011 (anticipated)
+        \begin{innerlist}
+		\item Thesis Topic: \emph{Information discovery in large corpora
+		  through topic modeling}
+        \item Advisor: Professor Kevin Seppi
+        \end{innerlist}
+
 \item[] B.S.,
@@
         \begin{innerlist}
+        \item Graduated summa cum laude (GPA 4.0) and with University honors
         \item Minors in Physics and Mathematics
-        \item Graduated summa cum laude (GPA 4.0) and with University honors
-        \item Thesis Topic: \emph{Speculative Evaluation in Particle Swarm
-            Optimization}
+        \item Honors Thesis Topic: \emph{Speculative Evaluation in Particle
+		  Swarm Optimization}
         \item Advisor: Professor Kevin Seppi
@@
 \begin{innerlist}
-\item CS 470: Introduction to Artificial Intelligence.  Tutored students,
-  graded homework and reports, and helped create new curriculum for the class.
+\item CS 470: Introduction to Artificial Intelligence.  Helped create new
+  curriculum for the class, tutored students, and graded homework and reports.
 \item CS 677: Bayesian Inference.  Tutored students and graded homework and
@@
   Department under Dr. Kevin Seppi.
-\item Reviewed papers for the International Conference on Machine Learning 2009.
 \item Research topics included particle swarm optimization and Markov decision
   processes.
+\item Reviewed papers for the International Conference on Machine Learning 2009.
 \end{innerlist}
@@
   the project can be found at:
-  http://google-tmads.blogspot.com/2010/01/track-cost-per-call-data-through-google.html.
+  http://google-tvads.blogspot.com/2010/01/track-cost-per-call-data-through-google.html.
 \item Gave several presentations to the machine learning lunch group.
@@
 \section{Publications} \begin{bibsection}
+	\item M. Gardner, J. Lutes, J. Lund, J. Hansen, D. Walker, E. Ringger, K.
+	  Seppi. The Topic Browser: An Interactive Tool for Browsing Topic Models.
+	  To appear in the Challenges of Data Visualization workshop at Neural
+	  Information Processing Systems 2010.
+
 	\item Matthew Gardner, Andrew McNabb, and Kevin Seppi.  Speculative

### 2010-11-13 applications/nsf_fellowship/personal_statement.tex
Finished a draft of my research statement, started on my personal statement.

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{natbib}

\setlength{\textheight}{9in}

\begin{document}

\pagestyle{empty}

\begin{center}
  \textbf{Personal Statement}\\
  Matthew Gardner
\end{center}
\vspace{-4mm}

My father, grandfather, and great-grandfather all earned PhDs in physics.  I
grew up wanting to be a physicist, to win the Nobel Prize by discovering a
Unified Field Theory.  So of course, when I started college, I chose physics as
my major.  During my freshman year I was actively involved in the Society of
Physics Students, encouraging other students to study physics, helping to tutor
lower-level students, and even going to give physics and chemistry
demonstrations at a local middle school, hoping to excite the children to learn
about science.

I also took a couple of computer science classes my freshman year, as the field
was interesting to me, though physics and mathematics still dominated my
thoughts.  It wasn't until I took a break from school for two years and had
some time away from classes that I really started to be fascinated by computer
science.

After my freshman year of college, I spent two years as a volunteer missionary
in Alabama, paying my way to serve the hispanic community in that state.  I
learned Spanish in order to communicate with them, and I came to understand
much better the circumstances that immigrants to our country live in.  I did a
significant amount of Spanish-English interpretation, and it was in that
capacity that I first started thinking about the potential for computer science
to improve the lives of ordinary people.  What if a computer could interpret
live speech from one language to another, accurately and effectively?  Such a
system would ease the burdens on so many immigrants who have much difficulty
communicating with other people for their daily needs, and not just those in
our country.

I came back from my experience in Alabama with a new-found desire to study
computer science.  I changed my major to computer science, with minors in
physics and mathematics, as those subjects were still interesting to me and
they seemed like relevant background.  I found a computer science professor who
was doing research that I found intriguing; I wanted to learn about machine
translation, which I knew was part of machine learning, so I started working in
the Applied Machine Learning Lab with Dr. Kevin Seppi.

My time working in the Applied Machine Learning lab has been the defining
experience of my undergraduate education.  I had very close interaction with
several professors and many graduate students who taught me what research was
like.  I also learned a great deal more than was taught in my classes in order
to understand the research that was going on in the lab.  Several times in my
classes the professor had me give part of a lecture that dealt with things I
was actively researching.

While I have tutored other students, mostly informally, throughout all of my
college career, my first formal teaching experience came my first semester as a
masters student.  I was a teaching assistant for two classes that Dr. Seppi
taught: Introduction to Articial Intelligence, and Bayesian Inference.  I had
always enjoyed helping people understand concepts, but I didn't think I would
enjoy the teaching aspect of being a professor until that experience.  I found
that figuring out how to explain complicated concepts simply and seeing the
light of understanding in the students' faces gave me great satisfaction.

I envision my future career as one of research, likely in the domain of natural
language processing, though possibly also in computer vision or other areas of
machine learning.  I have not yet decided between research in industry or
academia; perhaps my career will take me first to industry, ending up as a
professor later in life.

I have spent some time as a teaching assistant, and I
feel that I would enjoy teaching students of all the interesting problems there
are in computer science, getting them excited to do research in the field.  But
what I most want from my career is to find new problems to solve that will
better society, then solve them.  Things like machine translation and
interpretation, or extracting knowledge from large collections of text and
making that knowledge accessible to people, are intriguing problems that I
would love to spend my life's work on.

\end{document}

### 2010-11-13 applications/nsf_fellowship/previous_research.tex
Finished a draft of my research statement, started on my personal statement.

@@
 \usepackage{graphicx}
+\usepackage{natbib}
+
+\setlength{\textheight}{9in}
 
@@
 
-Four of the items listed from my undergraduate career all came from a single
-project focusing on how to effectively solve long-running optimization problems
-with a large cluster of hundreds or thousands of machines.  The project started
-when a physics professor approached my advisor seeking help solving a
-crystallography problem.  Having just begun working with my advisor, I accepted
-the problem mostly in an auxiliary role, writing some code and running
-experiments for Dr. Campbell, the physics professor.  We used particle swarm
-optimization, a bio-inspired optimization algorithm similar to evolutionary
-algorithms, to try to fit to data his model of what was happening in a
-particular kind of crystal.  After spending a lot of time seeing the output of
-the algorithm, I also helped to interpret the results.  I discovered that we
-had too many free parameters in our model and helped Dr. Campbell to make
-progress in solving his problem.
+\pagestyle{empty}
+
+\begin{center}
+  \textbf{Previous Research Experience}\\
+  Matthew Gardner
+\end{center}
+\vspace{-4mm}
+
+Brigham Young University is somewhat unique among universities in that it gives
+undergraduates superb opportunities to participate in research, even working
+very closely with professors.  During my second year of college, I began
+working in the Applied Machine Learning lab at BYU, headed by Dr. Kevin Seppi,
+and I was one of only three students that he advised at the time.  For the next
+two and a half years, I worked on several aspects of essentially the same
+problem---how to effectively solve long-running optimization problems with a
+large cluster of hundreds or thousands of machines.
+
+This project started when a physics professor approached my advisor seeking
+help solving a crystallography problem.  Having just begun working with my
+advisor, I accepted the problem mostly in an auxiliary role, writing some code
+and running experiments for Dr. Campbell, the physics professor.  We used
+particle swarm optimization, a bio-inspired optimization algorithm similar to
+evolutionary algorithms, to try to fit to data his model of what was happening
+in a particular kind of crystal.  After spending a lot of time seeing the
+output of the algorithm, I also helped to interpret the results.  I discovered
+that we had too many free parameters in our model and helped Dr. Campbell to
+make progress in solving his problem.  Because of our help in solving his
+problem, Dr. Campbell has made what he considers a breakthrough in
+understanding these crystals and plans to submit his work for publication soon.
 
@@
 I did much of the experimentation, though I was directed by a graduate student
-in the lab, who eventually wrote the paper that we published.  We saw some
-improvements in the performance of the algorithm with our new topologies,
-though we also noticed that increasing the number of particles with additional
-processors had diminishing returns.
+in the lab, who eventually wrote the paper that we
+published~\cite{mcnabb-2009-large-particle-swarms}.  We saw some improvements
+in the performance of the algorithm with our new topologies, though we also
+noticed that increasing the number of particles with additional processors had
+diminishing returns.
 
@@
 per iteration.  After implementing the technique, I ran many experiments to
-compare our methods to other means of parallelizing PSO.  Once I was satisfied
-that my results were significant, I wrote, submitted, and successfully defended
-my honors thesis.  I also modified the paper to fit in the aims and page
-requirements of Parallel Problem Solving in Nature, a well-respected parallel
-optimization conference, and submitted my thesis there, and it was accepted.
+compare our methods to other means of parallelizing the algorithm.  Once I was
+satisfied that my results were significant, I wrote, submitted, and
+successfully defended my honors thesis.  I also modified the paper to fit in
+the aims and page requirements of Parallel Problem Solving from Nature, a
+well-respected parallel optimization conference.  I submitted my thesis
+there and it was accepted~\cite{gardner-2010-speculative-evaluation-in-pso}.
 
@@
 from a conference, I got quite discouraged, wanting to give up and move on to
-something new.  My advisor insisted I persevere, saying it was good
-preparation for a PhD, and so I continued.  After running more experiments to
-strengthen the paper, we submitted it to the journal Swarm Intelligence, where
-it is currently under review.
+something new.  My advisor insisted I persevere, saying it was good preparation
+for a PhD, and so I continued.  After running more experiments to strengthen
+the paper, we submitted it to the journal Swarm Intelligence, where it is
+currently under
+review~\cite{gardner-2010-speculative-approach-to-parallelization-pso}.
 
@@
 required in getting a PhD, and how much I love exploring new topics and
-extending the limits of what we can do.  This project set the stage for my
-masters thesis and for my future work as a PhD student.
+extending the limits of what we can do, as well as helping researchers in other
+areas to solve their problems.  This project set the stage for my masters
+thesis and for my future work as a PhD student.
+
+Along with the research I performed as an undergraduate research assistant,
+there are two other research projects worth noting here.  One of them was my
+internship at Google in the summer of 2009.  I was presented with the problem
+of how to determine the effectiveness of TV advertisements given only a list of
+at what times and on what channels the ads played, along with a list of phone
+calls received in response to advertisements.  This was complicated because
+often ads play at the same time on different channels, and they use the same
+phone number.  I developed and implemented a probabilistic model that assigned
+phone calls to ads that were played.  The model learned a response rate for
+each channel the ads played on and helped advertisers know where to most
+effectively spend their money.
+
+The other research area that I have pursued is topic modeling, a method for
+capturing the semantics of a document collection, and that is the area of my
+masters thesis.  As an undergraduate, I implemented a topic model and used it
+to analyze a novel dataset of religious discourses, which I presented at an
+international conference held at
+BYU~\cite{gardner-2010-general-conference-topics}.  As a graduate student, I
+have been investigating how to use topic models to get specific, corpus-wide
+information out of large collections of documents.  My initial endeavors have
+involved building a tool to browse the output of a topic model in novel
+ways~\cite{gardner-2010-topic-browser}.  As other departments at BYU have seen
+the work that I and others have done with our topic browser, they have been
+very interested in using our tool.  Three separate groups at BYU have asked to
+use our browser to analyze data with purposes ranging from finding public
+health information in Twitter feeds to finding problem areas in teaching from
+free-form student ratings of classes and professors.  My future plans with
+topic modeling are presented in the research proposal submitted with this
+application.
+
+\footnotesize
+\bibliographystyle{plain}
+\renewcommand\bibsection{\subsubsection*{Publications}}
+\bibliography{../../../bib/aml/bib}
 

### 2010-11-16 applications/nsf_fellowship/personal_statement.tex
Finished a draft of my personal statement

@@
 \end{center}
-\vspace{-4mm}
 
@@
 several professors and many graduate students who taught me what research was
-like.  I also learned a great deal more than was taught in my classes in order
-to understand the research that was going on in the lab.  Several times in my
-classes the professor had me give part of a lecture that dealt with things I
-was actively researching.
+like.  From those that I interacted with, I gained a love of learning new
+things and exploring new ideas.  I also learned a great deal more than was
+taught in my classes in order to understand the research that was going on in
+the lab.  Several times in my classes the professor had me give part of a
+lecture that dealt with things I was actively researching.
 
-While I have tutored other students, mostly informally, throughout all of my
-college career, my first formal teaching experience came my first semester as a
-masters student.  I was a teaching assistant for two classes that Dr. Seppi
-taught: Introduction to Articial Intelligence, and Bayesian Inference.  I had
-always enjoyed helping people understand concepts, but I didn't think I would
-enjoy the teaching aspect of being a professor until that experience.  I found
-that figuring out how to explain complicated concepts simply and seeing the
-light of understanding in the students' faces gave me great satisfaction.
+My undergraduate education would not have been nearly so fulfilling and
+enjoyable without my experience working in a research lab.  My understanding of
+the ideal college experience has been fundamentally changed through my
+research.  If I ever end up as a professor, which I hope to be some day, one of
+my goals will be to help all of the undergraduates in my program have the same
+kind of experience that I had.  Working closely with professors and graduate
+students and conducting independent research were wonderful opportunities that
+every undergraduate student should be able to enjoy.
 
-I envision my future career as one of research, likely in the domain of natural
-language processing, though possibly also in computer vision or other areas of
-machine learning.  I have not yet decided between research in industry or
-academia; perhaps my career will take me first to industry, ending up as a
-professor later in life.
+I also had a great internship experience as an undergraduate that helped to
+shape my vision of what I wanted from my career.  I worked as an intern on the
+TV Ads team at Google during the summer of 2009.  My intern host presented a
+project to me that I didn't think was very interesting or very important, so I
+proposed a more meaningful, and more difficult, project.  My host supported me
+and gave me the help that I needed, and the project turned out to be very
+successful.  What I learned from the experience was that while I enjoy writing
+software, I find much more satisfication in solving interesting problems that
+serve some useful purpose.
 
-I have spent some time as a teaching assistant, and I
-feel that I would enjoy teaching students of all the interesting problems there
-are in computer science, getting them excited to do research in the field.  But
-what I most want from my career is to find new problems to solve that will
-better society, then solve them.  Things like machine translation and
-interpretation, or extracting knowledge from large collections of text and
-making that knowledge accessible to people, are intriguing problems that I
-would love to spend my life's work on.
+As I was nearing the end of my undergraduate career, I found that I had several
+research projects that needed just a little more time to be finished.  I wanted
+to get into one of the nation's top graduate schools for a PhD, and I thought
+that having several publications, instead of hopes of publishing, would give me
+a much stronger application.  I also had taken several graduate-level classes
+as an undergraduate that wouldn't transfer well to another university, and I
+was really enjoying working with Dr. Seppi.  All of these pointed me towards a
+masters degree at BYU, so I decided to stay, hoping to finish within a year.
+
+My first semester as a masters student during the summer of 2010 gave me my
+first real teaching opportunity.  While I have tutored other students, mostly
+informally, throughout all of my college career, I had never done so in a
+formal setting.  That semester, however, I was a teaching assistant for two
+classes that Dr. Seppi taught: Introduction to Articial Intelligence, and
+Bayesian Inference.  I had always enjoyed helping people understand concepts,
+but I didn't think I would enjoy the teaching aspect of being a professor until
+that experience.  I found that figuring out how to explain complicated concepts
+simply and seeing the light of understanding in the students' faces gave me
+great satisfaction.
+
+My experience so far as a masters student has served to strengthen my desire to
+get a PhD.  I love learning and solving new problems.  I envision my future
+career as one of research, likely in the domain of natural language processing,
+though possibly also in computer vision or other areas of machine learning.  I
+have not yet decided between research in industry or academia.  I enjoy
+teaching, but often the resources available in industry allow for more cutting
+edge research; the massive amounts of data and extensive computing resources at
+Google allow them to tackle problems that are impossible on smaller scales.
+Perhaps my career will take me first to industry, ending up as a professor
+later in life.  But what I most want from my career is to find new problems to
+solve that will better society, then solve them.  Things like machine
+translation and interpretation, or extracting knowledge from large collections
+of text and making that knowledge accessible to people, are intriguing problems
+that I would love to spend my life's work on.
 

### 2010-11-16 applications/nsf_fellowship/proposed_research.tex
Finished a draft of my proposed research

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{natbib}

\setlength{\textheight}{9in}

\begin{document}

\pagestyle{empty}

\begin{center}
  Matthew Gardner\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
  Proposed Research Essay
  \\
  \textbf{Searching Document Collections by Topic}
\end{center}
\vspace{-4mm}

\textbf{Keywords.} \textit{Topic models, natural language processing,
information extraction}

\textbf{Introduction.}  In our modern world, large document collections are
becoming more prevalent and more important.  Companies want to look at online
reviews of their products, students beginning research in a new area have piles
of documents to read, and intelligence analysts have thousands of documents to
sift through looking for anything noteworthy.  The recent controversies
involving the release of hundreds of thousands of documents by WikiLeaks about
the wars in Afghanistan and Iraq highlight the enormity of this problem.  How
does a person even begin to look through these massive corpora?

Typically, the interface between a person and a document collection has been
search, using a Google-like tool to find documents that contain specific
information.  While such tools can be very useful for some tasks, they cannot
give broad pictures about the corpus as a whole, and they are limited in the
kinds of queries they can answer.  A person has to have something to look for
in order to form a query; current search tools are not very helpful when a
person has no idea what to even look for.

Recent methods have been proposed that attempt to address part of this issue.
Topic models, including latent Dirichlet allocation and its
variants~\cite{blei-2003-latent-dirichlet-allocation}, seek to gather
information about topics used throughout the corpus.  These topics are modeled
as collections of words that are used together across documents.  Topic models,
then, could feasibly be used to give a broad overview of a corpus and even
answer queries about topics in the documents, instead of just queries about the
documents themselves.

As far as I am aware, however, topic models have not as yet been used for this
purpose.  They have successfully been used as input into document
classification tasks, for sentiment extraction, and as a surrogate for parts of
speech in machine translation (e.g.,~\cite{blei-2008-supervised-topic-models}).
But attempts to improve human corpus analysis through topic models are very
preliminary and have mostly been used to give the user better access to the
documents, not to the topics
themselves~\cite{newman-2010-visualizing-with-topic-maps}.  It is generally
assumed that topic models do a good job at capturing thematic elements in large
bodies of text, but while this has been anecdotally asserted on many
occasions~\cite{griffiths-2004-finding-scientific-topics}, to my knowledge it
has never been rigorously established.  Thus topic models have not been widely
used in the qualitative analysis of large corpora.

\textbf{Research Objective.}  It is my intent to quantitatively show that
topic models can effectively capture qualitative or thematic elements in a
collection of documents as judged by humans, and that automated methods can
select topics from a topic model that match human thematic judgments in a
variety of situations.

\textbf{Methods.}  In order to establish this, I will need a large collection
of documents annotated with specific topical information.  The necessary
annotations include human judgments on what themes or topics are present in
the documents, where they occur, and how they occur.  To my knowledge there is
no such set of data, so I will create one.

To get more conclusive results, I will collect and annotate two separate
collections of documents.  The first comes from student ratings of courses and
professors at Brigham Young University.  After answering many questions which
require selecting one of several options, students are given the opportunity to
write in free text whatever comments they have about the class.  I will use
those free-form comments as one source of text in which to find thematic
elements.

The second source of data I will use is a publicly available set of restaurant
reviews from CitySearch New York~\cite{ganu-2009-restaurant-ratings}.  This
dataset has the advantage of already being partially labeled, so providing the
rest of the annotations necessary for my purposes should be less expensive.

I will annotate only a subset of the data.  As there are hundreds of thousands
of comments available from the student ratings data that we will use, and over
50,000 restaurant reviews, reading and labeling every document with thematic
elements is an impossible task.  Instead, I will select a subset of the data
that is large enough to exhibit themes but small enough that an average college
student can annotate the data in approximately half an hour.  I imagine that
somewhere between 50 to 100 comments (where each comment is about a paragraph
long) is an appropriate size.  I will judge the amount of time required by
annotating the data myself, then hire between 10 and 20 college students to
also annotate the data, so that I can be sure that the annotations accurately
reflect average human judgments.

An important aspect of the subset of the data that to be annotated will be that
it will consist of a set of documents that all have the same value for some
attribute, such as ratings for the same teacher, or reviews about the same
restaurant.  This will facilitate both the finding of consistent thematic
elements and more extensive kinds of validation, which will be explained later.

The kinds of information I will elicit from the annotators will include the
top overall themes (ordered by prevalence) in the whole collection of documents
they are presented, as well as where those themes show up in the documents.  It
would likely be helpful for the annotator to select for each theme a single
document that best exhibits that theme, as well as list words that are commonly
used in the theme.  All of this will help me to judge whether a particular
topic from the topic model actually matches the theme given by the annotators.

\textbf{Validation.}  There are two parts to validating the use of topic models
in the qualitative analysis of text.  The first is showing that topic models
actually capture the thematic elements in a set of documents, as judged by
humans.  The second is showing that automatic methods for ranking the topics in
a topic model accurately rank topics according to the ranking given by humans.
Each part requires separate validation.

To demonstrate that topic models capture thematic elements, a mechanism is
necessary to compare a topic model's topic to the themes recorded by humans.
Devising this similarity metric will require some research, but as a starting
point I will compare the words given by the annotators to the words in the
topic, as well as the top documents given by the annotators to the documents
with the highest proportion of each topic.  I will also try to use some kind
of mutual information between the words in the topic and the title or
description of each theme as given by the annotators.

In order to show that automatic rankings can faithfully reproduce human
rankings, I will use standard metrics from information retrieval (IR), such as
mean average precision, to compare the two rankings.  This validation depends
on the first, as I must be able to determine if the topics ranked by the
computer match the themes ranked by humans.  This can be problematic if there
is no topic in the topic model that successfully captures a human theme, or if
there are two topics that each capture the theme equally well.  I may have to
modify the standard IR metrics to be better suited to my task.  While I will
only be annotating a subset of each data collection, I will be able to use the
entire corpus to validate my methods.  An important question to be able to
answer correctly is, ``What topics occur in reviews about McDonald's?''
Learning a topic model on the entire data and ranking the topics in answer to
such specific questions will be a very useful validation.

\textbf{Broader Impacts.}  This work will quantitatively establish the
usefulness of topic models in extracting thematic elements from a collection of
documents.  It will also validate the automatic ranking of topics in response
to a few classes of specific user queries.  Such work will give confidence to
anyone using a topic model to analyze a corpus, including intelligence analysts
looking at classified documents, university administrators looking at student
ratings, and companies reading online reviews about their products.  This work
also establishes a method and a set of data for topic models to be compared to
each other, for one model may better capture that human-judged thematic
elements than another.

\footnotesize
\bibliographystyle{plain}
\renewcommand\bibsection{\subsubsection*{References}}
\bibliography{../../../bib/lda/bib}

\end{document}

### 2010-11-17 applications/nsf_fellowship/personal_statement.tex
Incorporated feedback from Sabrina, and I  am done with the previous research essay

@@
 Physics Students, encouraging other students to study physics, helping to tutor
-lower-level students, and even going to give physics and chemistry
-demonstrations at a local middle school, hoping to excite the children to learn
-about science.
+lower-level students, and even giving physics and chemistry demonstrations at a
+local middle school, hoping to excite the children to learn about science.
 
@@
 significant amount of Spanish-English interpretation, and it was in that
-capacity that I first started thinking about the potential for computer science
-to improve the lives of ordinary people.  What if a computer could interpret
-live speech from one language to another, accurately and effectively?  Such a
-system would ease the burdens on so many immigrants who have much difficulty
+capacity that I first thought about the potential for computer science to
+improve the lives of ordinary people.  What if a computer could interpret live
+speech from one language to another, accurately and effectively?  Such a system
+would ease the burdens on so many immigrants who have great difficulty
 communicating with other people for their daily needs, and not just those in
@@
 physics and mathematics, as those subjects were still interesting to me and
-they seemed like relevant background.  I found a computer science professor who
-was doing research that I found intriguing; I wanted to learn about machine
-translation, which I knew was part of machine learning, so I started working in
-the Applied Machine Learning Lab with Dr. Kevin Seppi.
+they are relevant background for a computer scientist.  I found a computer
+science professor who was doing research that I found intriguing; I wanted to
+learn about machine translation, which I knew was part of machine learning, so
+I started working in the Applied Machine Learning Lab with Dr. Kevin Seppi.
 
@@
 experience of my undergraduate education.  I had very close interaction with
-several professors and many graduate students who taught me what research was
-like.  From those that I interacted with, I gained a love of learning new
-things and exploring new ideas.  I also learned a great deal more than was
-taught in my classes in order to understand the research that was going on in
-the lab.  Several times in my classes the professor had me give part of a
-lecture that dealt with things I was actively researching.
+several professors and many graduate students who showed me how interesting
+research can be.  From those that I had association with, I gained a love of
+learning new things and exploring new ideas.  I also learned a great deal more
+than was taught in my classes in order to understand the research that was
+going on in the lab.  Several times in my classes the professor teaching the
+class had me give part of a lecture that dealt with things I was actively
+researching.
 
@@
 the ideal college experience has been fundamentally changed through my
-research.  If I ever end up as a professor, which I hope to be some day, one of
-my goals will be to help all of the undergraduates in my program have the same
-kind of experience that I had.  Working closely with professors and graduate
-students and conducting independent research were wonderful opportunities that
-every undergraduate student should be able to enjoy.
+research.  As a professor, which I hope to be some day, one of my goals will be
+to help all of the undergraduates in my program have the same kind of
+experience that I had.  Working closely with professors and graduate students
+and conducting independent research were wonderful opportunities that every
+undergraduate student should be able to enjoy.
 
@@
 shape my vision of what I wanted from my career.  I worked as an intern on the
-TV Ads team at Google during the summer of 2009.  My intern host presented a
-project to me that I didn't think was very interesting or very important, so I
-proposed a more meaningful, and more difficult, project.  My host supported me
-and gave me the help that I needed, and the project turned out to be very
-successful.  What I learned from the experience was that while I enjoy writing
-software, I find much more satisfication in solving interesting problems that
-serve some useful purpose.
+TV Ads team at Google during the summer of 2009.  My intern host didn't have a
+clear vision of the project she wanted me to do, so I proposed a meaningful,
+difficult project that, if it worked, would have a large impact on the
+company's business.  My host supported me and gave me the help that I needed,
+and the project turned out to be very successful.  One of the main things I
+learned from the experience was that while I enjoy writing software, I find
+much more satisfication in solving interesting problems that actually improve
+people's lives.
 
@@
 Bayesian Inference.  I had always enjoyed helping people understand concepts,
-but I didn't think I would enjoy the teaching aspect of being a professor until
-that experience.  I found that figuring out how to explain complicated concepts
-simply and seeing the light of understanding in the students' faces gave me
-great satisfaction.
+but I didn't realize how much I would like the teaching aspect of being a
+professor until that experience.  I found that figuring out how to explain
+complicated concepts simply and seeing the light of understanding in the
+students' faces gave me great satisfaction.
 
@@
 Google allow them to tackle problems that are impossible on smaller scales.
-Perhaps my career will take me first to industry, ending up as a professor
-later in life.  But what I most want from my career is to find new problems to
-solve that will better society, then solve them.  Things like machine
-translation and interpretation, or extracting knowledge from large collections
-of text and making that knowledge accessible to people, are intriguing problems
-that I would love to spend my life's work on.
+Perhaps my career will take me first to industry, then back to academia as a
+professor later in life.  But what I most want from my career, wherever I work,
+is to find new problems whose solutions will better society, then solve them.
+Areas such as machine translation and interpretation, or extracting knowledge
+from large collections of text and making that knowledge accessible to people,
+are intriguing problems that I would love to spend my life researching.
 

### 2010-11-17 applications/nsf_fellowship/previous_research.tex
Incorporated feedback from Sabrina, and I  am done with the previous research essay

@@
 \setlength{\textheight}{9in}
+\setlength{\bibsep}{0in}
 
@@
 This project started when a physics professor approached my advisor seeking
-help solving a crystallography problem.  Having just begun working with my
-advisor, I accepted the problem mostly in an auxiliary role, writing some code
-and running experiments for Dr. Campbell, the physics professor.  We used
-particle swarm optimization, a bio-inspired optimization algorithm similar to
+help solving a crystallography problem.  As a new student in the lab, I
+accepted the problem mostly in an auxiliary role, writing some code and running
+experiments for Dr. Campbell, the physics professor.  We used particle swarm
+optimization (PSO), a bio-inspired optimization algorithm similar to
 evolutionary algorithms, to try to fit to data his model of what was happening
@@
 
-To solve this problem, my advisor proposed applying the idea of speculative
-decomposition in the parallelization literature to particle swarm optimization
-(PSO).  Instead of increasing the number of particles in the swarm, additional
+To solve this problem of diminishing returns, my advisor proposed applying the
+idea of speculative decomposition in the parallelization literature to PSO.
+Instead of increasing the number of particles in the swarm, additional
 processors could be used to evaluate possible next-iteration positions of each
@@
 the aims and page requirements of Parallel Problem Solving from Nature, a
-well-respected parallel optimization conference.  I submitted my thesis
-there and it was accepted~\cite{gardner-2010-speculative-evaluation-in-pso}.
+well-respected parallel optimization conference.  I submitted my thesis there
+and it was accepted~\cite{gardner-2010-speculative-evaluation-in-pso}.
 
@@
 method could potentially be improved.  Even though my honors thesis was
-completed, I continued my investigation.  I found that a large number of the
-speculative evaluations we performed were wasted, and that the algorithm could
-be improved by replacing useless evaluations with better evaluations, such as
-speculating several iterations ahead on paths that particles were likely to
-take.  I significantly expanded the work of my honors thesis, preparing to
-submit it to a journal on optimization.  After a rejection of a smaller version
-from a conference, I got quite discouraged, wanting to give up and move on to
-something new.  My advisor insisted I persevere, saying it was good preparation
-for a PhD, and so I continued.  After running more experiments to strengthen
-the paper, we submitted it to the journal Swarm Intelligence, where it is
-currently under
+completed I continued my investigation, as I thought there was more that could
+be done.  I found that a large number of the speculative evaluations we
+performed were wasted, and that the algorithm could be improved by replacing
+useless evaluations with better evaluations, such as speculating several
+iterations ahead on paths that particles were likely to take.  I significantly
+expanded the work of my honors thesis, preparing to submit it to a journal on
+optimization.  After a rejection of a smaller version from a conference, I
+became quite discouraged, wanting to give up and move on to something new.  My
+advisor insisted I persevere, saying it was good preparation for a PhD, and so
+I continued.  After running more experiments to strengthen the paper, we
+submitted it to the journal Swarm Intelligence, where it is currently under
 review~\cite{gardner-2010-speculative-approach-to-parallelization-pso}.
@@
 experience performing research and dealing with the submission process at
-conferences.  All told, I worked on particle swarm optimization for about two
-and a half years.  I learned a lot about what research is, the perseverance
-required in getting a PhD, and how much I love exploring new topics and
-extending the limits of what we can do, as well as helping researchers in other
-areas to solve their problems.  This project set the stage for my masters
-thesis and for my future work as a PhD student.
+conferences.  I learned a lot about what research is, the perseverance required
+in getting a PhD, and how much I love exploring new topics and extending the
+limits of what we can do, as well as helping researchers in other areas to
+solve their problems.  This project set the stage for my masters thesis and for
+my future work as a PhD student.
 
-Along with the research I performed as an undergraduate research assistant,
-there are two other research projects worth noting here.  One of them was my
-internship at Google in the summer of 2009.  I was presented with the problem
-of how to determine the effectiveness of TV advertisements given only a list of
-at what times and on what channels the ads played, along with a list of phone
-calls received in response to advertisements.  This was complicated because
-often ads play at the same time on different channels, and they use the same
-phone number.  I developed and implemented a probabilistic model that assigned
-phone calls to ads that were played.  The model learned a response rate for
-each channel the ads played on and helped advertisers know where to most
-effectively spend their money.
+Along with the research I performed as an undergraduate research assistant, two
+other projects are worth noting here.  The first is my internship at Google in
+the summer of 2009.  I tackled the problem of how to determine the
+effectiveness of TV advertisements given only a list of when and on what
+channels the ads played, along with a list of phone calls received in response
+to the ads.  This was complicated because often ads play at the same time on
+different channels, and they use the same phone number.  I developed and
+implemented a probabilistic model that assigned phone calls to ads that were
+played.  The model learned a response rate for each channel the ads played on
+and helped advertisers know where to most effectively spend their money.  Once
+advertisers saw that following the recommendations of my model gave them better
+response rates, they spent more on advertising with Google.
 
@@
 information out of large collections of documents.  My initial endeavors have
-involved building a tool to browse the output of a topic model in novel
+involved building a tool to browse the output of a topic model in new
 ways~\cite{gardner-2010-topic-browser}.  As other departments at BYU have seen
@@
 very interested in using our tool.  Three separate groups at BYU have asked to
-use our browser to analyze data with purposes ranging from finding public
-health information in Twitter feeds to finding problem areas in teaching from
-free-form student ratings of classes and professors.  My future plans with
-topic modeling are presented in the research proposal submitted with this
-application.
+use our browser to analyze data with purposes ranging from finding problem
+areas in teaching from free-form student ratings of classes and professors to
+searching for public health information in Twitter feeds.  My future plans with
+topic modeling are presented in my research proposal.
 
@@
 \bibliographystyle{plain}
-\renewcommand\bibsection{\subsubsection*{Publications}}
+\renewcommand\bibsection{\noindent \small\textbf{Publications}\vspace{-2mm}\footnotesize}
 \bibliography{../../../bib/aml/bib}

### 2010-11-17 applications/nsf_fellowship/proposed_research.tex
Incorporated feedback from Sabrina, and I  am done with the previous research essay

@@
 
-An important aspect of the subset of the data that to be annotated will be that
-it will consist of a set of documents that all have the same value for some
-attribute, such as ratings for the same teacher, or reviews about the same
+An important aspect of the subset of the data that will be annotated is
+that it will consist of a set of documents that all have the same value for
+some attribute, such as ratings for the same teacher, or reviews about the same
 restaurant.  This will facilitate both the finding of consistent thematic
@@
 also establishes a method and a set of data for topic models to be compared to
-each other, for one model may better capture that human-judged thematic
-elements than another.
+each other, for one model may better capture human-judged thematic elements
+than another.
 

### 2010-11-17 applications/nsf_fellowship/personal_statement.tex
I'm pretty happy with my personal statement right now, which surprises me.  I didn't think it was that good at first.

@@
 learn about machine translation, which I knew was part of machine learning, so
-I started working in the Applied Machine Learning Lab with Dr. Kevin Seppi.
+I started working in the Applied Machine Learning lab with Dr. Kevin Seppi.
 
@@
 several professors and many graduate students who showed me how interesting
-research can be.  From those that I had association with, I gained a love of
+research can be.  From those that I had association with I gained a love of
 learning new things and exploring new ideas.  I also learned a great deal more
@@
 I also had a great internship experience as an undergraduate that helped to
-shape my vision of what I wanted from my career.  I worked as an intern on the
-TV Ads team at Google during the summer of 2009.  My intern host didn't have a
-clear vision of the project she wanted me to do, so I proposed a meaningful,
+shape my vision of what I want from my career.  I worked as an intern on the TV
+Ads team at Google during the summer of 2009.  My intern host didn't have a
+clear idea for the project she wanted me to do, so I proposed a meaningful,
 difficult project that, if it worked, would have a large impact on the
 company's business.  My host supported me and gave me the help that I needed,
-and the project turned out to be very successful.  One of the main things I
-learned from the experience was that while I enjoy writing software, I find
-much more satisfication in solving interesting problems that actually improve
-people's lives.
+and the project turned out to be very successful.  The project was so
+successful that the director of all of TV Ads at Google said that my project
+was the most important work going on in the whole TV Ads team, and several
+times he commended me for my work.  One of the main things I learned from the
+experience was that while I enjoy writing software, I find much more
+satisfication in solving interesting problems that actually improve people's
+lives.  I was able to help the TV Ads team, which was struggling to make money,
+greatly improve their business.
 
-As I was nearing the end of my undergraduate career, I found that I had several
-research projects that needed just a little more time to be finished.  I wanted
-to get into one of the nation's top graduate schools for a PhD, and I thought
-that having several publications, instead of hopes of publishing, would give me
-a much stronger application.  I also had taken several graduate-level classes
-as an undergraduate that wouldn't transfer well to another university, and I
-was really enjoying working with Dr. Seppi.  All of these pointed me towards a
-masters degree at BYU, so I decided to stay, hoping to finish within a year.
+I returned from my internship with one year left as an undergraduate.  As that
+year neared its end and the time came to apply to graduate schools, I found
+that I had several research projects that needed just a little more time to be
+finished.  I wanted to finish what I had started here at BYU.  I also wanted to
+get into one of the nation's top universities for a PhD, and I thought that
+having several publications, instead of just hopes of publishing, would give me
+a much stronger application.  Further, I had taken several graduate-level
+classes as an undergraduate that wouldn't transfer well to another university,
+and I was really enjoying working with Dr. Seppi.  All of these pointed me
+towards a masters degree at BYU, so I decided to stay, hoping to finish within
+a year.
 
 My first semester as a masters student during the summer of 2010 gave me my
-first real teaching opportunity.  While I have tutored other students, mostly
-informally, throughout all of my college career, I had never done so in a
-formal setting.  That semester, however, I was a teaching assistant for two
-classes that Dr. Seppi taught: Introduction to Articial Intelligence, and
-Bayesian Inference.  I had always enjoyed helping people understand concepts,
-but I didn't realize how much I would like the teaching aspect of being a
-professor until that experience.  I found that figuring out how to explain
-complicated concepts simply and seeing the light of understanding in the
-students' faces gave me great satisfaction.
+first real teaching opportunity.  While I have tutored other students
+throughout all of my college career, I had never done so in a formal setting.
+That semester, however, I was a teaching assistant for two classes that Dr.
+Seppi taught: Introduction to Articial Intelligence, and Bayesian Inference.  I
+had always enjoyed helping people understand concepts, but I didn't realize how
+much I would like the teaching aspect of being a professor until that
+experience.  I found that figuring out how to explain complicated concepts
+simply and seeing the light of understanding in the students' faces gave me
+great satisfaction.
 

### 2010-11-17 applications/nsf_fellowship/proposed_research.tex
Research proposal is now 2 pages, though I'm not sure I'm happy with it yet

@@
 \begin{center}
-  Matthew Gardner\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
-  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
-  Proposed Research Essay
-  \\
   \textbf{Searching Document Collections by Topic}
@@
 becoming more prevalent and more important.  Companies want to look at online
-reviews of their products, students beginning research in a new area have piles
-of documents to read, and intelligence analysts have thousands of documents to
-sift through looking for anything noteworthy.  The recent controversies
-involving the release of hundreds of thousands of documents by WikiLeaks about
-the wars in Afghanistan and Iraq highlight the enormity of this problem.  How
-does a person even begin to look through these massive corpora?
+reviews of their products and intelligence analysts have thousands of documents
+to sift through looking for suspicious activity.  The recent controversies
+involving the public release of hundreds of thousands of documents by WikiLeaks
+about the wars in Afghanistan and Iraq highlight the enormity of this problem.
+How does a person even begin to look through these massive corpora?
 
-Typically, the interface between a person and a document collection has been
-search, using a Google-like tool to find documents that contain specific
+Typically, the interface between a person and a document collection has been a
+search engine, using a Google-like tool to find documents that contain specific
 information.  While such tools can be very useful for some tasks, they cannot
 give broad pictures about the corpus as a whole, and they are limited in the
-kinds of queries they can answer.  A person has to have something to look for
-in order to form a query; current search tools are not very helpful when a
+kinds of queries they can answer.  A person has to have specific information in
+mind in order to form a query; current search tools are not very helpful when a
 person has no idea what to even look for.
@@
 Recent methods have been proposed that attempt to address part of this issue.
-Topic models, including latent Dirichlet allocation and its
-variants~\cite{blei-2003-latent-dirichlet-allocation}, seek to gather
-information about topics used throughout the corpus.  These topics are modeled
-as collections of words that are used together across documents.  Topic models,
-then, could feasibly be used to give a broad overview of a corpus and even
-answer queries about topics in the documents, instead of just queries about the
-documents themselves.
+Topic models, including latent Dirichlet
+allocation~\cite{blei-2003-latent-dirichlet-allocation}, gather information
+about topics used in a corpus.  These topics are modeled as collections of
+words that are used together across documents.  Topic models could feasibly be
+used to give a broad overview of a corpus and even answer queries about topics
+in the documents, instead of just queries about the documents.
 
@@
 purpose.  They have successfully been used as input into document
-classification tasks, for sentiment extraction, and as a surrogate for parts of
-speech in machine translation (e.g.,~\cite{blei-2008-supervised-topic-models}).
-But attempts to improve human corpus analysis through topic models are very
-preliminary and have mostly been used to give the user better access to the
-documents, not to the topics
+classification and many other tasks
+(e.g.,~\cite{blei-2008-supervised-topic-models}), but attempts to aid humans in
+corpus analysis through topic models are very preliminary, mostly giving the
+user better access to the documents in the corpus, not to the topics
 themselves~\cite{newman-2010-visualizing-with-topic-maps}.  It is generally
@@
 
-\textbf{Research Objective.}  It is my intent to quantitatively show that
-topic models can effectively capture qualitative or thematic elements in a
-collection of documents as judged by humans, and that automated methods can
-select topics from a topic model that match human thematic judgments in a
-variety of situations.
+\textbf{Research Objective.}  It is my intent to quantitatively show that topic
+models can effectively capture qualitative or thematic elements in a collection
+of documents as judged by humans, and that automated methods can select topics
+from a topic model that match human thematic judgments in a variety of
+situations.  As an example, given the set of campaign speeches for all
+candidates in the 2008 presidential primaries and general election, this
+research will show conclusively that a topic model can accurately answer the
+question, ``What were Barack Obama's consistent campaign themes?''
 
-\textbf{Methods.}  In order to establish this, I will need a large collection
-of documents annotated with specific topical information.  The necessary
-annotations include human judgments on what themes or topics are present in
-the documents, where they occur, and how they occur.  To my knowledge there is
-no such set of data, so I will create one.
+\textbf{Methods.}  Establishing that topic models can answer topic-based
+queries the same way a human would requires two parts.  First one must show
+that a topic found by a topic model can be accurately compared to a theme or
+topic elicited from a human.  Second, one must show that an automated system
+can correctly answer queries about topics in a corpus by returning the topics
+from a topic model that match human answers to the same queries.
 
-To get more conclusive results, I will collect and annotate two separate
-collections of documents.  The first comes from student ratings of courses and
-professors at Brigham Young University.  After answering many questions which
-require selecting one of several options, students are given the opportunity to
-write in free text whatever comments they have about the class.  I will use
-those free-form comments as one source of text in which to find thematic
-elements.
+In order to establish these two parts, I will need a large collection of
+documents annotated with specific topical information.  The necessary
+annotations include human judgments on what themes or topics are present in the
+documents (giving a phrasal description of each topic), where they occur
+(including documents that demonstrate the topic well), and how they occur
+(giving example words or phrases from the documents).  To my knowledge there is
+no such annotated data, so I will create it.
 
-The second source of data I will use is a publicly available set of restaurant
-reviews from CitySearch New York~\cite{ganu-2009-restaurant-ratings}.  This
-dataset has the advantage of already being partially labeled, so providing the
-rest of the annotations necessary for my purposes should be less expensive.
+I will use two sets of data to provide more conclusive results.  The first is a
+collection of free-form student comments on classes and professors given to
+Brigham Young University at the end of each semester.  The second is a publicly
+available set of restaurant reviews from CitySearch New
+York~\cite{ganu-2009-restaurant-ratings}.  Neither of these sets of data are
+annotated with sufficient topical information, though the restaurant reviews
+are already partially annotated.
 
-I will annotate only a subset of the data.  As there are hundreds of thousands
-of comments available from the student ratings data that we will use, and over
-50,000 restaurant reviews, reading and labeling every document with thematic
-elements is an impossible task.  Instead, I will select a subset of the data
-that is large enough to exhibit themes but small enough that an average college
-student can annotate the data in approximately half an hour.  I imagine that
-somewhere between 50 to 100 comments (where each comment is about a paragraph
-long) is an appropriate size.  I will judge the amount of time required by
-annotating the data myself, then hire between 10 and 20 college students to
-also annotate the data, so that I can be sure that the annotations accurately
-reflect average human judgments.
+As there are many thousands of documents in each set of data, annotating the
+entire corpus would be impossible.  Thus I will only annotate a section of each
+collection, corresponding to all of the comments about one particular class in
+the student ratings, or one restaurant in the reviews.  This will aid in
+validation, as I will have complete topical information about one of the
+classes or restaurants and will thus be able to accurately test the answers to
+queries.  I will hire between 20 and 40 college students to each independently
+annotate the same subset of one of the collections of documents, so that I can
+have confidence in the data and have some idea of the consistency of human
+thematic judgments.
 
-An important aspect of the subset of the data that will be annotated is
-that it will consist of a set of documents that all have the same value for
-some attribute, such as ratings for the same teacher, or reviews about the same
-restaurant.  This will facilitate both the finding of consistent thematic
-elements and more extensive kinds of validation, which will be explained later.
+\textbf{Validation.}  Once I have annotated data, I will be able to test the
+validity of my hypothesis.  Determining whether a particular topic mathces a
+human-defined topic should be straightforward with the annotations I will
+gather.  I can test the words found by the topic model against the description
+given by the humans, and I can see if the example documents given by the
+annotators agree with the documents that have high proportions of the topic.
 
-The kinds of information I will elicit from the annotators will include the
-top overall themes (ordered by prevalence) in the whole collection of documents
-they are presented, as well as where those themes show up in the documents.  It
-would likely be helpful for the annotator to select for each theme a single
-document that best exhibits that theme, as well as list words that are commonly
-used in the theme.  All of this will help me to judge whether a particular
-topic from the topic model actually matches the theme given by the annotators.
-
-\textbf{Validation.}  There are two parts to validating the use of topic models
-in the qualitative analysis of text.  The first is showing that topic models
-actually capture the thematic elements in a set of documents, as judged by
-humans.  The second is showing that automatic methods for ranking the topics in
-a topic model accurately rank topics according to the ranking given by humans.
-Each part requires separate validation.
-
-To demonstrate that topic models capture thematic elements, a mechanism is
-necessary to compare a topic model's topic to the themes recorded by humans.
-Devising this similarity metric will require some research, but as a starting
-point I will compare the words given by the annotators to the words in the
-topic, as well as the top documents given by the annotators to the documents
-with the highest proportion of each topic.  I will also try to use some kind
-of mutual information between the words in the topic and the title or
-description of each theme as given by the annotators.
-
-In order to show that automatic rankings can faithfully reproduce human
-rankings, I will use standard metrics from information retrieval (IR), such as
-mean average precision, to compare the two rankings.  This validation depends
-on the first, as I must be able to determine if the topics ranked by the
-computer match the themes ranked by humans.  This can be problematic if there
-is no topic in the topic model that successfully captures a human theme, or if
-there are two topics that each capture the theme equally well.  I may have to
-modify the standard IR metrics to be better suited to my task.  While I will
-only be annotating a subset of each data collection, I will be able to use the
-entire corpus to validate my methods.  An important question to be able to
-answer correctly is, ``What topics occur in reviews about McDonald's?''
-Learning a topic model on the entire data and ranking the topics in answer to
-such specific questions will be a very useful validation.
+With methods for matching topics from a topic model with human-given topics, I
+can complete the second part of my objective.  With topics obtained from humans
+for a particular class, I can learn a topic model over the entire collection of
+student ratings and see if automatic methods to rank the topics return the
+correct topics.  One such method is to rank the topics by their entropy over
+the documents for that class, which should rank highest the topics that were
+used consistently about that class, matching the human-given topics and
+answering the question, ``What were the main themes in comments of Chemistry
+101?''  Other possible queries and methods to answer them are items that will
+be explored in the course of this research.
 
@@
 also establishes a method and a set of data for topic models to be compared to
-each other, for one model may better capture human-judged thematic elements
-than another.
+each other in order to determine which models best capture human thematic
+judgments.
 
@@
 \bibliographystyle{plain}
-\renewcommand\bibsection{\subsubsection*{References}}
+\renewcommand\bibsection{\noindent \small\textbf{References}\vspace{-2mm}\footnotesize}
 \bibliography{../../../bib/lda/bib}

### 2010-11-18 applications/nsf_fellowship/proposed_research.tex
Done with the application.
I made revisions on all of my essays (some minor, some major), and
I added a bib file, because I made some egregious changes to the
real bib files (to save space) that I didn't want to commit.
But I wanted to be able to remake these applications just as they are.

@@
 \textbf{Keywords.} \textit{Topic models, natural language processing,
-information extraction}
+information extraction, search}
 
@@
 \textbf{Methods.}  Establishing that topic models can answer topic-based
-queries the same way a human would requires two parts.  First one must show
+queries the same way a human would involves two parts.  First one must show
 that a topic found by a topic model can be accurately compared to a theme or
@@
 
-\textbf{Validation.}  Once I have annotated data, I will be able to test the
-validity of my hypothesis.  Determining whether a particular topic mathces a
-human-defined topic should be straightforward with the annotations I will
-gather.  I can test the words found by the topic model against the description
-given by the humans, and I can see if the example documents given by the
-annotators agree with the documents that have high proportions of the topic.
+\textbf{Validation.}  With annotated data, I can test the validity of my
+hypothesis.  Determining whether a particular topic matches a human-defined
+topic is straightforward with the annotations I will gather.  I can test the
+words found by the topic model against the description given by the humans, and
+I can see if the example documents given by the annotators agree with the
+documents that have high proportions of the topic according to the model.
 
-With methods for matching topics from a topic model with human-given topics, I
-can complete the second part of my objective.  With topics obtained from humans
-for a particular class, I can learn a topic model over the entire collection of
-student ratings and see if automatic methods to rank the topics return the
-correct topics.  One such method is to rank the topics by their entropy over
-the documents for that class, which should rank highest the topics that were
-used consistently about that class, matching the human-given topics and
-answering the question, ``What were the main themes in comments of Chemistry
-101?''  Other possible queries and methods to answer them are items that will
-be explored in the course of this research.
+Once I have methods for matching topics from a topic model with human-given
+topics, I can complete the second part of my objective.  I will learn a topic
+model over each collection of documents and see if automatic methods to rank
+the topics return topics that match those that were given by humans.  One such
+method is to rank the topics by their entropy over the documents for the
+annotated class or restaurant.  This ranking should match the human-judged
+themes and answer the question, ``What were the main themes in comments about
+this class?'' Similar methods can give a broad overview of the topics in the
+entire corpus, and other possible queries and automatic ranking methods to
+answer them are items that will be explored in the course of this research,
+including finding topics that are common or unique to two separate classes.
 
@@
 documents.  It will also validate the automatic ranking of topics in response
-to a few classes of specific user queries.  Such work will give confidence to
+to a few types of specific user queries.  Such work will give confidence to
 anyone using a topic model to analyze a corpus, including intelligence analysts
@@
 \renewcommand\bibsection{\noindent \small\textbf{References}\vspace{-2mm}\footnotesize}
-\bibliography{../../../bib/lda/bib}
+\bibliography{bib}
 

### 2010-12-11 applications/phd_programs/nsf_research.tex
Started working on my phd essay

--- 2010-11-17 applications/nsf_fellowship/previous_research.tex
@@
 
-This project started when a physics professor approached my advisor seeking
-help solving a crystallography problem.  As a new student in the lab, I
-accepted the problem mostly in an auxiliary role, writing some code and running
-experiments for Dr. Campbell, the physics professor.  We used particle swarm
-optimization (PSO), a bio-inspired optimization algorithm similar to
-evolutionary algorithms, to try to fit to data his model of what was happening
-in a particular kind of crystal.  After spending a lot of time seeing the
-output of the algorithm, I also helped to interpret the results.  I discovered
-that we had too many free parameters in our model and helped Dr.  Campbell to
-make progress in solving his problem.  Because of our help in solving his
-problem, Dr. Campbell has made what he considers a breakthrough in
-understanding these crystals and plans to submit his work for publication soon.
+This project started when Dr. Campbell, a physics professor,  approached my
+advisor seeking help solving a crystallography problem.  As a new student in
+the lab, I accepted the problem mostly in an auxiliary role, writing some code
+and running experiments for Dr. Campbell.  We used particle swarm optimization
+(PSO), a bio-inspired optimization algorithm similar to evolutionary
+algorithms, to try to fit to data his model of what was happening in a
+particular kind of crystal.  After spending a lot of time seeing the output of
+the algorithm, I also helped to interpret the results.  I discovered that we
+had too many free parameters in our model and helped Dr.  Campbell to make
+progress in solving his problem.  Because of our help, Dr.  Campbell has made
+what he considers a breakthrough in understanding these crystals and plans to
+submit his work for publication soon.
 
@@
 processors to solve the problem.  We first experimented with the topology of
-the algorithm, or how the particles in the swarm communicate with each other.
-I did much of the experimentation, though I was directed by a graduate student
-in the lab, who eventually wrote the paper that we
+the algorithm, or how the particles in PSO communicate with each other.  I did
+much of the experimentation, though I was directed by a graduate student in the
+lab, who eventually wrote the paper that we
 published~\cite{mcnabb-2009-large-particle-swarms}.  We saw some improvements
 in the performance of the algorithm with our new topologies, though we also
-noticed that increasing the number of particles with additional processors had
+noticed that increasing the number of particles with additional processors gave
 diminishing returns.
@@
 \renewcommand\bibsection{\noindent \small\textbf{Publications}\vspace{-2mm}\footnotesize}
-\bibliography{../../../bib/aml/bib}
+\bibliography{bib}
 

### 2010-12-11 applications/phd_programs/personal_statement.tex
Started working on my phd essay

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{}
\author{}
\date{}

\begin{document}

\pagestyle{empty}

\begin{center}
  \textbf{Personal Statement}\\
  Matthew Gardner
\end{center}

It was spending time as a Spanish-English interpreter that first piqued my
interest in natural language processing.  What if a computer could interpret
accurately and in real time between two languages?  What problems would have to
be solved in order to build such a system?  These questions fascinated me, and
I set out to spend my undergraduate career learning as much as I could about
their answers.

In my second year of college I started as a research assistant in the Applied
Machine Learning lab at Brigham Young University.  The project that was given
to me to work on was particle swarm optimization (PSO).  While my work in PSO
did not relate directly to my interest in machine interpretation, it taught me
what it is like to do real research, as well as the joys and sorrows of the
paper submission process at conferences and journals.

All told I spent two and a half years working on various aspects of PSO.  At
the beginning I was just running experiments with code other people wrote, a
simple bookkeeper.  I progressed to helping analyze and interpret the results,
then to writing a little code of my own.  After about a year in the lab, I got
my first publication; I was a co-author on a paper about how parallelization
affects how PSO~\cite{mcnabb-2009-large-particle-swarms}.

Motivated by the success of our parallel PSO paper, we spent more effort on how
to better parallelize the algorithm.  My advisor wondered if PSO could be
speculatively decomposed, allowing extra processors to do two or more
iterations of PSO at the same time.  I thought that was an interesting idea and
I made it my honors thesis, moving the idea to an algorithm and then to an
implementation.  After testing my new method and discovering that it
outperformed standard techniques, I turned my work into a paper and
successfully defended my honors thesis.  Confident that my work was amazing, I
submitted my thesis to one of the top conferences for PSO research, fully
expecting a glowing reception.

I had my first major setback when that paper was rejected outright.  At this
point I had spent two years working on PSO, and I felt like giving up and
moving on.  PSO wasn't my main interest anyway, and I wanted to do more
interesting things.  My advisor, however, insisted that I perservere, telling
me that the experience would be good practice for a PhD.  So I revised my paper
and submitted it to another top conference, where it was
accepted~\cite{gardner-2010-speculative-evaluation-in-pso}.

In doing the work for my honors thesis I had many ideas for how to improve the
method I pioneered.  Once I recovered from the shock of my first rejection, I
spent some time exploring new methods that gave significantly greater
performance than my original implementation.  I expanded my honors thesis with
these new ideas and submitted my work to the journal \emph{Swarm Intelligence},
where it is currently under
review~\cite{gardner-2010-speculative-approach-to-parallelization-pso}.

My work in particle swarm optimization was not directly applicable to this
program to which I am currently applying, though a general knowledge of
optimization is certainly useful in many fields of computer science.  However,
I feel that the experience I had working on that problem has done a great deal
to prepare me to succeed in your program.  I have experience conducting real
research, I know how to bring ideas to fruition, and I have successfully taken
my own innovations and had them recognized by the wider research community.
I am confident that, while there will inevitably be setbacks in any research
endeavor, I can perservere and succeed.

\bibliographystyle{plain}
\bibliography{bib}

\end{document}

### 2010-12-13 applications/phd_programs/berkeley_personal_history.tex
A lot of work on applications

--- 2010-11-17 applications/nsf_fellowship/personal_statement.tex
@@
 
-My time working in the Applied Machine Learning lab has been the defining
-experience of my undergraduate education.  I had very close interaction with
-several professors and many graduate students who showed me how interesting
-research can be.  From those that I had association with I gained a love of
-learning new things and exploring new ideas.  I also learned a great deal more
-than was taught in my classes in order to understand the research that was
-going on in the lab.  Several times in my classes the professor teaching the
-class had me give part of a lecture that dealt with things I was actively
-researching.
+My time working in the Applied Machine Learning lab was the defining experience
+of my undergraduate education.  I had very close interaction with several
+professors and many graduate students who showed me how interesting research
+can be.  From those that I had association with I gained a love of learning new
+things and exploring new ideas.  I also learned a great deal more than was
+taught in my classes in order to understand the research that was going on in
+the lab.  Several times in my classes the professor teaching the class had me
+give part of a lecture that dealt with topics I was actively researching.
 

### 2010-12-13 applications/phd_programs/personal_statement.tex
A lot of work on applications

@@
 moving on.  PSO wasn't my main interest anyway, and I wanted to do more
-interesting things.  My advisor, however, insisted that I perservere, telling
+interesting things.  My advisor, however, insisted that I persevere, telling
 me that the experience would be good practice for a PhD.  So I revised my paper
@@
 program to which I am currently applying, though a general knowledge of
-optimization is certainly useful in many fields of computer science.  However,
-I feel that the experience I had working on that problem has done a great deal
-to prepare me to succeed in your program.  I have experience conducting real
-research, I know how to bring ideas to fruition, and I have successfully taken
-my own innovations and had them recognized by the wider research community.
-I am confident that, while there will inevitably be setbacks in any research
-endeavor, I can perservere and succeed.
+optimization techniques is certainly useful in many fields of computer science.
+However, I feel that the experience I had working on that problem has done a
+great deal to prepare me to succeed in your program.  I have experience
+conducting real research, I know how to bring ideas to fruition, and I have
+successfully taken my own innovations and had them recognized by the wider
+research community.  I am confident that, while there will inevitably be
+setbacks in any research endeavor, I can persevere and succeed.
 
-\bibliographystyle{plain}
+I have had many other experiences that also contribute to my preparation for
+your program.  I have gained a broad background in statistical machine learning
+techniques, having been the TA for the Bayesian Inference class taught at BYU
+and taken several advanced classes on the subject.  I have also taken all of
+the classes that BYU offers in natural language processing (NLP).  For my final
+project in the graduate-level NLP course, I implemented a Gibbs sampler for the
+latent Dirichlet allocation model and ran it on a novel dataset of religious
+discourses.  That work was of particular interest to BYU, and I presented it at
+the BYU Studies Symposium, an international conference held at
+BYU~\cite{gardner-2010-general-conference-topics}.
+
+As a masters student at BYU I have continued my work with topic modeling,
+exploring how topic models can be used to improve the browsing experience of
+large corpora.  I recently returned from the NIPS conference, where I presented
+at a workshop a tool that I and others at BYU have built that incorporates a
+lot of previous work in visualizing topic models and some of our own
+ideas~\cite{gardner-2010-topic-browser}.
+
+I have significant experience conducting and presenting research, and I have
+sufficient background in machine learning and natural language processing to be
+able to quickly contribute to ongoing research projects.  I would love to spend
+my time as a PhD student working on machine interpretation or some subproblem
+of it.  Speech recognition, speech synthesis, machine translation, parsing and
+co-reference resolution are all interesting problems to me that I would enjoy
+working on.  As statistical modeling is the foundation of most state-of-the-art
+solutions to these problems, I also would enjoy doing research in that area.
+[Add more personalized stuff to each university here]
+
+\bibliographystyle{unsrt}
 \bibliography{bib}

### 2010-12-14 resume_cv/cv.tex
Updated my cv a little bit

@@
 
-\section{Publications} \begin{bibsection}
-	\item M. Gardner, J. Lutes, J. Lund, J. Hansen, D. Walker, E. Ringger, K.
-	  Seppi. The Topic Browser: An Interactive Tool for Browsing Topic Models.
-	  To appear in the Challenges of Data Visualization workshop at Neural
-	  Information Processing Systems 2010.
+\section{Journal Publications}
+  \begin{bibsection}
+	\item Matthew Gardner, Andrew McNabb, and Kevin Seppi.  A Speculative
+	  Approach to Parallelization in Particle Swarm Optimization.
+	  Submitted to: \emph{Swarm Intelligence}.
+  \end{bibsection}
 
+\section{Conference Publications}
+  \begin{bibsection}
 	\item Matthew Gardner, Andrew McNabb, and Kevin Seppi.  Speculative
@@
 	  Solving in Nature, 2010}.
-
-	\item Matthew Gardner and Eric Ringger.  Automatic Topic Discovery in 100
-	  Years of General Conference Talks.  Presented at the 50th Anniversary
-	  BYU Studies Symposium, 2010.
 
@@
 
+\section{Workshop and other refereed publications}
+  \begin{bibsection}
+	\item Matthew Gardner, J. Lutes, J. Lund, J. Hansen, D. Walker, E. Ringger,
+	  K.  Seppi. The Topic Browser: An Interactive Tool for Browsing Topic
+	  Models.  In the Challenges of Data Visualization workshop at \emph{Neural
+	  Information Processing Systems, 2010}.
+
+	\item Matthew Gardner and Eric Ringger.  Automatic Topic Discovery in 100
+	  Years of General Conference Talks.  Presented at the \emph{50th
+	  Anniversary BYU Studies Symposium, 2010}.
+
+  \end{bibsection}
+
 \end{document}

### 2010-12-15 applications/phd_programs/personal_statement.tex
Done with a generic personal statement

@@
 \usepackage{graphicx}
+\usepackage{natbib}
 
-\title{}
-\author{}
-\date{}
+\setlength{\bibsep}{0in}
 
@@
 In my second year of college I started as a research assistant in the Applied
-Machine Learning lab at Brigham Young University.  The project that was given
-to me to work on was particle swarm optimization (PSO).  While my work in PSO
-did not relate directly to my interest in machine interpretation, it taught me
-what it is like to do real research, as well as the joys and sorrows of the
-paper submission process at conferences and journals.
+Machine Learning lab at Brigham Young University (BYU).  The project that was
+given to me to work on was particle swarm optimization (PSO).  While my work in
+PSO did not relate directly to my interest in machine interpretation, it taught
+me what it is like to do real research and prepared me for my future work as a
+PhD student.
 
@@
 then to writing a little code of my own.  After about a year in the lab, I got
-my first publication; I was a co-author on a paper about how parallelization
-affects how PSO~\cite{mcnabb-2009-large-particle-swarms}.
+my first publication; I was a co-author on a paper about parallelization and
+scalability in PSO~\cite{mcnabb-2009-large-particle-swarms}.
 
@@
 point I had spent two years working on PSO, and I felt like giving up and
-moving on.  PSO wasn't my main interest anyway, and I wanted to do more
-interesting things.  My advisor, however, insisted that I persevere, telling
-me that the experience would be good practice for a PhD.  So I revised my paper
-and submitted it to another top conference, where it was
+moving on.  My advisor, however, insisted that I persevere, telling me that the
+experience would be good practice for a PhD.  So I revised my paper and
+submitted it to another top conference, where it was
 accepted~\cite{gardner-2010-speculative-evaluation-in-pso}.
@@
 your program.  I have gained a broad background in statistical machine learning
-techniques, having been the TA for the Bayesian Inference class taught at BYU
-and taken several advanced classes on the subject.  I have also taken all of
-the classes that BYU offers in natural language processing (NLP).  For my final
-project in the graduate-level NLP course, I implemented a Gibbs sampler for the
-latent Dirichlet allocation model and ran it on a novel dataset of religious
+techniques, having been the TA for the Bayesian Inference class and taken
+several advanced classes on the subject.  I have also taken all of the classes
+that BYU offers in natural language processing (NLP).  For my final project in
+the graduate-level NLP course, I implemented a Gibbs sampler for the latent
+Dirichlet allocation model and ran it on a novel dataset of religious
 discourses.  That work was of particular interest to BYU, and I presented it at
-the BYU Studies Symposium, an international conference held at
-BYU~\cite{gardner-2010-general-conference-topics}.
+the BYU Studies Symposium, an international conference held at the
+university~\cite{gardner-2010-general-conference-topics}.
+
+At the end of my undergraduate education I applied for and received a
+scholarship from the North American chapter of the Association for
+Computational Linguistics (NAACL) to attend a summer school at Johns Hopkins
+University on human language technologies.  My experience at JHU over the
+summer was further preparation for your program, as we covered a broad range of
+topics dealing with machine learning and natural language processing.  I also
+attended the NAACL conference that summer and saw firsthand some of the
+research going on at your school.
 
@@
 large corpora.  I recently returned from the NIPS conference, where I presented
-at a workshop a tool that I and others at BYU have built that incorporates a
-lot of previous work in visualizing topic models and some of our own
+a tool that I and others at BYU have built that incorporates a lot of previous
+work in visualizing topic models and some of our own
 ideas~\cite{gardner-2010-topic-browser}.
@@
 
-\bibliographystyle{unsrt}
+\vspace{2mm}
+\footnotesize
+\bibliographystyle{plain}
+\renewcommand\bibsection{\noindent \small\textbf{Publications}\vspace{-2mm}\footnotesize}
 \bibliography{bib}

### 2010-12-15 applications/phd_programs/berkeley_personal_history.tex
Done with my phd application stuff

@@
 After my freshman year of college, I spent two years as a volunteer missionary
-in Alabama, paying my way to serve the hispanic community in that state.  I
+in Alabama, paying my way to serve the Hispanic community in that state.  I
 learned Spanish in order to communicate with them, and I came to understand
@@
 experience was that while I enjoy writing software, I find much more
-satisfication in solving interesting problems that actually improve people's
+satisfaction in solving interesting problems that actually improve people's
 lives.  I was able to help the TV Ads team, which was struggling to make money,
@@
 That semester, however, I was a teaching assistant for two classes that Dr.
-Seppi taught: Introduction to Articial Intelligence, and Bayesian Inference.  I
-had always enjoyed helping people understand concepts, but I didn't realize how
-much I would like the teaching aspect of being a professor until that
+Seppi taught: Introduction to Artificial Intelligence, and Bayesian Inference.
+I had always enjoyed helping people understand concepts, but I didn't realize
+how much I would like the teaching aspect of being a professor until that
 experience.  I found that figuring out how to explain complicated concepts

### 2011-01-11 masters_thesis/proposal/proposal.tex
Started working on my thesis proposal

\documentclass[ms]{byuprop}
% Options for this class include the following (* indicates default):
%
%   10pt -- 10 point font size
%   11pt -- 11 point font size
%   12pt (*) -- 12 point font size
%
%   ms -- produce a thesis proposal (off)
%   areaexam -- produce a research area overview (off)
%   phd -- produce a dissertation proposal (off)
%   
%   layout -- show layout lines on the pages, helps with overfull boxes (off)
%   grid -- show a half-inch grid on every page, helps with printing (off)

% This command fixes my particular printer, which starts 0.03 inches too low,
% shifting the whole page down by that amount.  This shifts the document
% content up so that it comes out right when printed.
%
% Discovering this sort of behavior is best done by specifying the ``grid''
% option in the class parameters above.  It prints a 1/2 inch grid on every
% page.  You can then use a ruler to determine exactly what the printer is
% doing.
%
% Uncomment to shift content up (accounting for printer problems)
%\setlength{\voffset}{-.03in}

% Here we set things up for invisible hyperlinks in the document.  This makes
% the electronic version clickable without changing the way that the document
% prints.  It's useful, but optional.
\usepackage[
    ps2pdf,
    bookmarks=true,
    breaklinks=true,
    raiselinks=true,
    pdfborder={0 0 0},
    colorlinks=false,
    ]{hyperref}

% Rewrite the itemize, description, and enumerate environments to have more
% reasonable spacing:
\newcommand{\ItemSep}{\itemsep 0pt}
\let\oldenum=\enumerate
\renewcommand{\enumerate}{\oldenum \ItemSep}
\let\olditem=\itemize
\renewcommand{\itemize}{\olditem \ItemSep}
\let\olddesc=\description
\renewcommand{\description}{\olddesc \ItemSep}

% Get a little less fussy about word spacing on a line.  Sometimes produces
% ugly results, so keep your eyes peeled.
\sloppy

% Important settings for the byuprop class. %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Because I use these things in more than one place, I created new commands for
% them.  I did not use \providecommand because I absolutely want LaTeX to error
% out if these already exist.
\newcommand{\Title}{Improved Corpus Exploration with Topic Modeling}
\newcommand{\Author}{Matthew Gardner}
\newcommand{\SubmissionMonth}{January}
\newcommand{\SubmissionYear}{2011}

% Take these from the commands defined above
\title{\Title}
\author{\Author}
\monthsubmitted{\SubmissionMonth}
\yearsubmitted{\SubmissionYear}

% Committee members
\committeechair{Kevin Seppi}
\committeemembera{Eric Ringger}
\committeememberb{David Embley}
%\committeememberc{}
%\committeememberd{}

% Department graduate coordinator
\graduatecoordinator{Kent Seamons}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Set up the internal PDF information so that it becomes part of the document
% metadata.  The pdfinfo command will display this. Be sure to set the document
% type and add your own keywords.
\hypersetup{%
    pdftitle=\Title,%
    pdfauthor=\Author,%
    pdfsubject={Masters thesis proposal, BYU CS Department: %
                Submitted \SubmissionMonth~\SubmissionYear, Created \today},%
    pdfkeywords={topic modeling, corpus browsing, corpus discovery, %
				text mining},%
}

% These packages allow the bibliography to be sorted alphabetically and allow references to more than one paper to be sorted and compressed (i.e. instead of [5,2,4,6] you get [2,4-6])
\usepackage{url}
\usepackage[numbers,sort&compress]{natbib}
\bibliographystyle{annotnat}
\usepackage{hypernat}
% Note, use \citet to name the authors (text mode) and \citep to do just give
% the number (parenthetical mode).  Also, \citep* will give all of the authors
% rather than an abbreviated list.

% Additional packages required for your specific thesis go here. I've left some I use as examples.
\usepackage{graphicx}
%\usepackage{pdfsync}
\usepackage{amsmath}

\begin{document}

% Produce the preamble
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Abstract}
% 1 to 2 paragraphs

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
% 1 to 4 pages

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Related Work}
% 1 to 2 pages

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Thesis Statement}
% 1 to 2 sentences

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Project Description}
% 2 to 5 pages

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Validation}
% 1/2 to 2 pages

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Thesis Schedule}
% 1/4 to 1/2 page

\begin{itemize}

\item Completed: reformulate PSO in MapReduce.

\item February 6, 2009: evaluate PSO with large swarms with a few benchmark
functions.

\item February 20, 2009: evaluate PSO with very large swarms with subswarms
and with more benchmark functions.

\item March 9, 2009: rework individual papers into a coherent thesis and
submit draft to committee chair.

\item March 20, 2009: submit thesis to second committee member for review.

\item March 30, 2009: submit thesis to third committee member for review.

\item April 10, 2009: defend thesis.

Each of the three stages of the project can be submitted as a paper to a
conference such as GECCO or CEC.

\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\renewcommand\bibsection{\section{Annotated Bibliography}}
%\newcommand\bibpreamble{Material to be added between heading and entries.}
% 2 to 5 pages

\bibliography{}

\end{document}

### 2011-01-25 honors_thesis/si_paper/reviewer_response/response.tex
Added a start to a response for the si paper

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}

\begin{document}

\section*{Associate Editor Comments}

\subsection*{AE.1}

Reviewer comments: The quality of the article will be improved if you include a
review of existing parallelization methods for PSO, to produce a kind of
taxonomy. Then, also include comparisons with these approaches. You currently
compare only with a naive approach and adaptations of your approach.

Response: Copy and paste from Andrew's paper

NOT DONE

\subsection*{AE.2}

Reviewer comments: Please indicate if differences in performance are
statistically significant.

Response: Because all of the these finished before all of these, the t-test
gave \ldots

NOT DONE

\subsection*{AE.3}

Reviewer comments: It will also be good if you do make use of problems where
parallelization has benefit. For the functions that you use, there is not real
not for parallelization. Linked to this, when will parallelization be
beneficial?

Response: Get Andrew's paragraph that deals with this

NOT DONE

\subsection*{AE.4}

Reviewer comments: Please re-organize section 3: You start with a very long
theoretical discussion, and only then get to section 3.1.

Response: Good point.  We realized that we tried to cram too much into a single
section, so we split the two main parts of it into two separate sections.
There is now a section devoted entirely to the mathematics of the method and a
separate section describing its implementation.

NOT DONE

\section*{Reviewer 1 Comments}

\subsection*{1.1}

Reviewer comments: I'm a bit skeptical about the actual \emph{practical}
significance of the proposed method, [because it is difficult to know for any
given problem what the right swarm size is]. To be fair, this criticism applies
to most contributions focused on algorithms rather than on problems.

Response: This is a fair criticism, as it is true that experimentation with a
problem is required before one can know how best to solve it.  We added two
paragraphs to the end of section 2 to further make this point, making more
prominent reference to the No Free Lunch theorems.

\subsection*{1.2}

Reviewer comments: The organization and presentation of the ideas can be
greatly improved. The authors mix results and conclusions with the description
of the approach. This doesn't help the goal of the authors. A reader, cannot
appreciate the significance of some result or conclusion if he/she hasn't
understood what the authors are talking about. I recommend the authors to
restructure the whole paper but in particular Sections 3 and 4.

Response: This was not specific enough to use too much...

NOT DONE

\subsection*{1.3}

Reviewer comments: The notation used throughout the paper is confusing. In the
PSO literature, it is common to use x as a particles current position and some
other letters to denote a particle's personal best position and a particle's
neighborhood's best position. I know that the authors are using superscripts to
distinguish between these concepts but they don't really help.

Response: How do we fix this?

NOT DONE

\subsection*{1.4}

Reviewer comments: Using "rounds of evaluations" is confusing, and results
should be reported in terms of wall clock time.

Response: Rounds of evaluations correspond exactly to wall clock time, as we
mentioned.  However, we realize that the term is not a commonly used one, so we
changed "rounds of evaluations" to "time steps" in hopes of better conveying
our meaning. (Also incorporate Andrew's paragraph on why we're using time steps
instead of evaluations)

NOT DONE

\subsection*{1.5}

Reviewer comments: Since the authors are using Herrera et al.'s proposed
benchmarks and since they are using a large computing cluster, I don't see why
they can't use the same experimental protocol proposed by Herrera et al., which
is focused on large scale optimization. That protocol would better justify a
parallel algorithm because a 1000-dimensional problem is intuitively more
difficult than the 20-dimensional problems the authors are currently using. As
a side product, the authors can actually compare the results of the pruned
versions (which in my eyes, are new PSO variants) with those of other
state-of-the-art algorithms.

Response: We need to look up what he's talking about.

NOT DONE

\subsection*{1.6}

Reviewer comments: In the abstract, it is not clear the meaning of
"speculative". A reader without knowledge about parallel computing may not
understand what the authors mean.

Response: Need to address this.

NOT DONE

\subsection*{1.7}

Reviewer comments: The authors can cite Poli's paper about applications of PSO
algorithms in the first paragraph of the paper.

Response: We added the citation after the first sentence of the paper.

\subsection*{1.8}

Reviewer comments: What serves as the motivation for the authors's work is
insufficiently discussed.  The authors say that adding more particles leads to
"diminishing returns". This may be true when dealing with constant size
populations.  However, a recent trend is to vary the population size over time.
In fact, it has been shown that adding particles \emph{over time} may actually
solve the problem the authors refer to. Some references are "Incremental Social
Learning in Particle Swarms" by Montes de Oca and others published in IEEE SMCB
and "Efficient Population Utilization Strategy for Particle Swarm Optimizer" by
Hsieh and others in the same journal.

Response: See how papers are relevant (Andrew).  But also, assuming a fixed
computational budget, if at any point you have a smaller swarm than your budget
allows, you should do speculative evaluation.  That's our point.

NOT DONE

\subsection*{1.9}

Reviewer comments: In the first paragraph of Section 3, the authors say that
adding particles may not help past a certain swarm size. This is true is the
computation time allocated to the optimization process remains fixed. However,
in some problems, it is the solution quality what matters. In that case, adding
particles \emph{and} computational budget may indeed be the best thing to do.

Response: Most people have a fixed computational budget, function has a set
time.  That is the case we are interested in.

NOT DONE

\subsection*{1.10}

Reviewer comments: In Eq. 3. Why must the current position be worse than the
neighbor's position?

Response: Restate what the case means

NOT DONE

\subsection*{1.11}

Reviewer comments: One or two step-by-step examples would greatly improve
Section 3.

Response: We fixed section 3 by splitting it into two sections and reorganizing
some parts, making the examples that were there more clear.  Hopefully the
section is satisfactory now.

NOT DONE

\subsection*{1.12}

Reviewer comments: Isn't pruning against the idea of better using the available
computing resources? If one reduces the number of evaluated speculative
children, then fewer CPUs are going to be used.

Response: Pruning allows for the use of a larger swarm with the same number of
processors.  We tried to make this point more clear in the text.

NOT DONE

\subsection*{1.13}

Reviewer comments: Lines 29-30 of page 14. Stagnation is about the lack of
improvement of the best result. It has nothing to do with what the particles
are doing. I know that the PSO community confuses the two. I think this is a
nice opportunity to clarify the difference between search stagnation and the
state of the swarm.

Response: todo

NOT DONE

\subsection*{1.14}

Reviewer comments: All figures are not clear. The authors should find line
styles that are more distinguishable.

Response: ...

NOT DONE

\subsection*{1.15}

Reviewer comments: Page 27, line 40. Algorithms are suited for problems not the
other way around.

Response: ...

NOT DONE

\subsection*{1.16}

Reviewer comments: Page 31. Line 46. Molia -> Molina. Is it a technical report?
Which number?  Isn't it published with the special issue?

Response: ...

NOT DONE

\section*{Reviewer 2 Comments}

I found the idea very interesting and the paper well-written. However, there
are still some points that need further investigation.  As far as I am
concerned, the paper has merit and shall be accepted after a revision that will
address these remarks.

\subsection*{2.1}

Reviewer comments: How does it scale with the problem's dimension? There is
experimental evidence that the required swarm size for achieving a prescribed
accuracy in a specific problem, increases almost exponentially with its
dimension. The proposed idea can make this necessity milder. Is there any
evidence on how well it scales when dimension increases? (e.g. probably the
increase in the required number of particles is closer to the linear rather
than the exponential case).

Response: ...

NOT DONE

\subsection*{2.2}

Reviewer comments: In the first sections of the paper it is argued that the
proposed "...method can easily be extended to arbitrary topologies". However,
in later sections it is apparent that this may be possible if the underlying
neighborhood topology is not "dense" enough to cause heavy communication
traffic to the processors.  Can you be more precise to this and quantify it?
For example, given a swarm size and a common network communication is there any
identifiable limit to the neighborhood topology such that the algorithm retains
an acceptable performance?

Response: ...

NOT DONE

\subsection*{2.3}

Reviewer comments: My personal experience with the PSO versions used in the
paper suggests that, in large dimensions, it is preferrable to use larger
swarms for less iterations rather than small swarms for more iterations. This
is in contrast to the argued behavior of PSO for the Sphere function (Section
1).  Did you observe the same behavior in other functions?  Could you provide
some evidence on this because, due to the swarm's diversity loss, it seems
somehow counterintuitive.

Response: ...

NOT DONE

\section*{Reviewer 3 Comments}

\subsection*{3.1}

Reviewer comments: There are too many redundancy sentences declaring this paper
is not proposing a new algorithm in the introduction.

Response: ...

NOT DONE

\subsection*{3.2}

Reviewer comments: The velocity updating formular is different from the
original PSO algorithm.  The weighting factor, X, should only affects the old
velocity, not all the items.

Response: reference constricted PSO in formulas

NOT DONE

\subsection*{3.3}

Reviewer comments: Some reference are not completed. For example, the
journal/conference name of the 4th reference is missing. Please check it
carefully.

Response: ...

NOT DONE

\subsection*{3.4}

Reviewer comments: Please increasing the reference to extend the integrity of
the background.

Response: ...

  a)Pei-Wei Tsai, Jeng-Shyang Pan, Shyi-Ming Chen, Bin-Yih Liao, and Szu-Ping
Hao, 2008, "Parallel Cat Swarm Optimization," Proceedings of the Seventh
International Conference on Machine Learning and Cybernetics, Kunming, China,
Jul. 12-15, 2008, pp. 3328-3333.

  b)Jui-Fang Chang, Shu-Chuan Chu, John F. Roddick and Jeng-Shyang Pan, 2005,
"A Parallel Particle Swarm Optimization Algorithm with Communication
Strategies", Journal of Information Science and Engineering, Vol. 21, No. 4,
pp. 809-818.

NOT DONE

\end{document}

### 2011-01-25 honors_thesis/si_paper/paper.tex
A few revisions to the si paper to address reviewer comments

@@
 Particle swarm optimization (PSO) has been found to be a highly robust and
-effective algorithm for solving many types of optimization problems.  For much
-of the algorithm's history, PSO was run serially on a single machine.  However,
-the world's computing power is increasingly coming from large clusters of
-processors.  In order to efficiently utilize these resources for
-computationally intensive problems, PSO needs to run in parallel.
+effective algorithm for solving many types of optimization
+problems~\cite{poli-2008-pso-applications}.  For much of the algorithm's
+history, PSO was run serially on a single machine.  However, the world's
+computing power is increasingly coming from large clusters of processors.  In
+order to efficiently utilize these resources for computationally intensive
+problems, PSO needs to run in parallel.
 
@@
 The balance of this paper is organized as follows. \sec{pso} describes the
-particle swarm optimization algorithm.  \sec{sepso} describes how speculative
-evaluation can be done in parallel PSO to perform two iterations at once.  In
-\sec{relax}, we discuss various methods of improving the performance of
-speculative evaluation in PSO, all of which break the requirement of strictly
-reproducing the behavior of the original algorithm.  \sec{setup} describes the
-experiments we ran, and \sec{results} presents the results of those
-experiments.  In \sec{conclusion} and \sec{future} we conclude and discuss
-future work.
+particle swarm optimization algorithm.  \sec{sepso} describes mathematically
+how speculative evaluation can be done in parallel PSO to perform two
+iterations at once, and \sec{implementation} gives information on how to
+implement this method.  In \sec{relax}, we discuss various methods of improving
+the performance of speculative evaluation in PSO, all of which break the
+requirement of strictly reproducing the behavior of the original algorithm.
+\sec{setup} describes the experiments we ran, and \sec{results} presents the
+results of those experiments.  In \sec{conclusion} and \sec{future} we conclude
+and discuss future work.
 
@@
 
+The ideal topology and swarm size for PSO depend on the objective function.
+Researchers have devised various benchmark functions and have found that the
+ideal topology for one function may perform very poorly for another function.
+The No Free Lunch Theorems for Optimization show that this is true in
+general---if an algorithm performs well on average for one class of functions
+then it must do poorly on average for other
+problems~\cite{wolpert-1997-nfl-for-optimization}.
+
+An attempt to standardize PSO found that although a Complete swarm of 50
+particles converged to the global optimum more quickly for many benchmark
+functions than the same swarm size with a Ring topology, it was also more
+likely to permaturely converge to local optima for other functions.  The study
+found no significant improvement for any other swarm size between 20 and 100
+and concluded with a recommendation to use a swarm of 50 particles with a Ring
+topology as a starting point.  The authors acknowledged that choosing the ideal
+topology requires thorough experimentation for the particular
+problem~\cite{bratton-2007-defining-a-standard-for-pso}.  Other authors have
+explained that a large swarm with a sparse topology propogates information
+slowly~\cite{montes-de-oca-2009-frankensteins-pso}.
+
 \section{Speculative Evaluation in PSO}
@@
 
+\section{Implementing Speculative Evaluation}
+\label{sec:implementation}
+
 It is not trivial in some parallel architectures to determine which speculative
-position was the correct next position of each particle.  In the rest of this
-section we move from the mathematics of our method to its implementation.
-First we discuss the relatively easy case of a centralized parallel PSO
-algorithm with a master computer and many slaves.  In such an architecture, the
-master keeps track of all necessary information with only trivial message
-passing needed.  Then we discuss the more complicated case of a distributed
-algorithm, where each particle is on its own and needs to send and receive
-messages to and from other particles.  Finally we discuss the further
-complications of a dynamic topology, where a particle's neighbors change from
-one iteration to another.
+position was the correct next position of each particle.  In this section we
+move from the mathematics of our method to its implementation.  First we
+discuss the relatively easy case of a centralized parallel PSO algorithm with a
+master computer and many slaves.  In such an architecture, the master keeps
+track of all necessary information with only trivial message passing needed.
+Then we discuss the more complicated case of a distributed algorithm, where
+each particle is on its own and needs to send and receive messages to and from
+other particles.  Finally we discuss the further complications of a dynamic
+topology, where a particle's neighbors change from one iteration to another.
 

### 2011-01-25 masters_thesis/proposal/proposal.tex
Finished an initial draft of my proposal

@@
 \usepackage{graphicx}
+\usepackage{psfrag}
 %\usepackage{pdfsync}
 \usepackage{amsmath}
+
+% Better references, I think
+\newcommand{\secref}[1]{Section~\ref{sec:#1}}
+\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
+\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
 
@@
 
+Topic modeling, including latent Dirichlet allocation and its variants, has
+been shown to be an effective means of modeling a collection of documents for a
+wide variety of purposes.  Topic models have been successfully used in
+dimensionality reduction for classification tasks, in associating text with
+components in an image, and in aligning words in bilingual corpora.  Despite
+the wide-spread use of topic models in many areas of machine learning, however,
+their usefulness as an aid to corpus browsing, or trend discovery in large
+collections of documents, has not been well established.
+
+I propose to investigate the application of topic models to the task of corpus
+browsing.  That is, I seek to show that a topic model is an effective aid to a
+human looking through a previously unseen collection of documents.  I will show
+this in two parts.  The first part involves the building of a tool that allows
+one to easily browse through a corpus in the context of a topic model.  The
+second part of this work will establish that topic models do in fact capture
+themes that a human would have found in the data and that those topics can be
+automatically presented without the user having to dig through the topic model
+to find them.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % 1 to 4 pages
+
+In our modern world, large document collections are becoming more prevalent and
+more important.  Companies want to look at online reviews of their products and
+intelligence analysts have thousands of documents to sift through looking for
+suspicious activity.  The recent controversies involving the public release of
+hundreds of thousands of documents by WikiLeaks about the wars in Afghanistan
+and Iraq highlight the enormity of this problem.  How does a person even begin
+to look through these massive corpora?
+
+Typically, the interface between a person and a document collection has been a
+search engine, using a Google-like tool to find documents that contain specific
+information.  While such tools can be very useful for some tasks, they cannot
+give broad pictures about the corpus as a whole, and they are limited in the
+kinds of queries they can answer.  A person has to have specific information in
+mind in order to form a query; current search tools are not very helpful when a
+person has no idea what to even look for.
+
+Recent methods have been proposed that attempt to address part of this issue.
+Topic models, including latent Dirichlet
+allocation~\cite{blei-2003-latent-dirichlet-allocation}, gather information
+about topics used in a corpus.  These topics are modeled as collections of
+words that are used together across documents.  Topic models could feasibly be
+used to give a broad overview of a corpus and even answer queries about topics
+in the documents, instead of just queries about the documents.
+
+As far as I am aware, however, topic models have not as yet been used for this
+purpose.  They have successfully been used as input into document
+classification and many other tasks
+(e.g.,~\cite{blei-2008-supervised-topic-models}), but attempts to aid humans in
+corpus analysis through topic models are very preliminary, mostly giving the
+user better access to the documents in the corpus, not to the topics
+themselves~\cite{newman-2010-visualizing-with-topic-maps}.  It is generally
+assumed that topic models do a good job at capturing thematic elements in large
+bodies of text, but while this has been anecdotally asserted on many
+occasions~\cite{griffiths-2004-finding-scientific-topics}, to my knowledge it
+has never been rigorously established.  Thus topic models have not been widely
+used in the qualitative analysis of large corpora.
+
+\subsection{Topic Modeling}
+
+Topic modeling seeks to find the set of topics which are discussed in a corpus
+and to what proportion each document contains each of those topics.  Several
+methods of topic modeling have been proposed.  Among those are latent semantic
+analysis or indexing (LSA or LSI)~\cite{dumais-1988-latent-semantic-analysis},
+a probabilistic version of LSA, called pLSA or pLSI~\cite{hofmann-1999-plsi},
+and latent Dirichlet allocation, or
+LDA~\cite{blei-2003-latent-dirichlet-allocation}.  In this proposed work I
+focus on LDA.
+
+LDA is a generative probabilistic model of a collection of documents.  The
+documents are modeled as deriving from a collection of topics, where each
+document comes from some combination of those topics.  Each word in a
+document is drawn from one of the topics in that document, where the topic is
+simply a multinomial distribution over words.  The graphical model of LDA is
+shown in \figref{lda}.
+
+\begin{figure}
+  \centering
+  \psfrag{b}[C][C]{\Large$\beta$}
+  \psfrag{a}[C][C]{\Large$\alpha$}
+  \psfrag{t}[C][C]{\Large$\theta$}
+  \psfrag{f}[C][C]{\Large$\phi$}
+  \psfrag{w}[C][C]{\Large$w$}
+  \psfrag{z}[C][C]{\Large$z$}
+  \psfrag{N}[C][C]{\large$N$}
+  \psfrag{M}[C][C]{\large$M$}
+  \psfrag{K}[C][C]{\large$K$}
+  \includegraphics[width=.5\textwidth]{../../figures/lda_model}
+  \caption{The graphical model of LDA.}
+  \label{fig:lda}
+\end{figure}
+
+In the figure, $\alpha$ is a Dirichlet prior over topic distributions in a
+document, $\theta$ is a per-document multinomial topic distribution, $z$ is a
+topic label for each word in the document, and $w$ is an observed word in the
+document.  $\beta$ is a Dirichlet prior over the word distributions in a topic
+and $\phi$ is a per-topic multinomial word distribution.  The distributions can
+be written formally as follows:
+
+\begin{align*} 
+  w_i|z_i,\phi^{(z_i)} &\sim \mathrm{Categorical}(\phi^{(z_i)}) \\
+  \phi &\sim \mathrm{Dirichlet}(\beta) \\
+  z_i|\theta^{(d_i)} &\sim \mathrm{Categorical}(\theta^{(d_i)}) \\ 
+  \theta &\sim \mathrm{Dirichlet}(\alpha) 
+\end{align*}
+
+Given a collection of documents, the parameters $\theta$ and $\phi$ can be
+inferred using standard statistical techniques.  The two most common methods
+for performing inference are Gibbs
+sampling~\cite{griffiths-2004-finding-scientific-topics} and variational
+methods~\cite{blei-2003-latent-dirichlet-allocation}.
 
@@
 
+Topic modeling has become incredibly popular in recent years, saturating the
+field with all kinds of applications and variations of the method.  Of
+particular interest to this proposal are two small areas that have not received
+much attention in the literature: visualizing topic models and validating topic
+models.
+
+\subsection{Visualizing Topic Models}
+
+Typically, topic models have been visualized by simply listing the 10 words
+with the highest probability of occurring in that
+topic~\cite{blei-2003-latent-dirichlet-allocation,
+griffiths-2004-finding-scientific-topics}.  When presenting topics in a paper,
+researchers often sift through the results by hand in order to show topics that
+are most meaningful.  This is understandable as showing all of the resultant
+topics in a paper is infeasible.  However, such presentations hide the fact
+that it can be time consuming and difficult to go through the output of a topic
+model by hand.
+
+While displaying an entire topic model analysis is impossible in a research
+paper, several static visualizations of whole topic models exist as
+websites~\cite{blei-arxiv-corpus-browser, mimno-nyt-browser}.  These typically
+present long lists of topics that are very difficult to use for finding
+specific information.
+
+Apart from visualizing all of the topics from a topic model as a whole, work
+has been done in how to present a topic in more interesting ways than simple
+lists of words.  Turbo Topics is a method for extracting meaningful phrases
+used in a particular topic that can be presented along with individual
+words~\cite{blei-2009-turbo-topics}.  Several people have presented trends of
+the use of a topic over time in a
+corpus~\cite{griffiths-2004-finding-scientific-topics, mimno-nyt-browser}, and
+some effort has been made in trying to automatically give topics a meaningful
+name, instead of characterizing the topic only by the most frequently appearing
+words~\cite{mei-2007-automatic-labeling-of-topic-models,
+lau-2010-best-topic-word-selection}.
+
+The output from a topic model has also been used to inform a more general
+corpus browsing experience.  Newmann et al. make what they call ``topic maps,''
+which are layouts of the documents in a corpus according to a topic-based
+similarity metric~\cite{newman-2010-visualizing-with-topic-maps}.  Stoica et
+al. use topic models to create facets to use in a browsing interface, though
+they conclude that other methods are superior to topic models in this
+task~\cite{stoica-2007-faceted-metadata-structures}.
+
+\subsection{Validating Topic Models}
+
+The question of how to evaluate topic models has been addressed in a number of
+different ways.  These include both intrinsic and extrinsic evaluations.
+
+Intrinsic evaluations of topic models use some metric on held-out data to judge
+the quality of the topic model itself.  These are typically perplexity or the
+likelihood of the held-out data according to the model.  Early papers on LDA
+used these metrics~\cite{blei-2003-latent-dirichlet-allocation,
+griffiths-2004-finding-scientific-topics}, and recent papers have given better
+methods for evaluating these
+metrics~\cite{wallach-2009-evaluation-for-topic-models,
+buntine-2009-estimating-likelihoods-for-topic-models}.
+
+A new intrinsic evaluation method presented at NAACL this summer is evaluating
+the coherence of the words in a topic, or how well the words fit together as
+judged by humans~\cite{newman-2010-automatic-evaluation-of-topic-coherence}.
+The authors found that using co-occurrence counts gathered from Wikipedia is a
+good surrogate for human judgments in this regard.  However, as the paper
+focused on finding an automated metric that matched human responses, the
+authors did not actually use their metric to compare topic models in any way.
+
+Apartment from intrinsic evaluation methods, which may have limited
+applicability in any real task, are extrinsic evaluation methods.  These
+methods use the topic model, or some part of it, as input into a secondary
+task, such as document classification.  The original paper introducing LDA used
+one of these metrics~\cite{blei-2003-latent-dirichlet-allocation}, and
+virtually all papers describing an application of topic models have also had
+some kind of extrinsic metric.  Other examples of extrinsic evaluations
+including predicting the citation count of a scientific
+paper~\cite{gerrish-2010-measuring-scholarly-impact} and classifying and
+annotating images~\cite{wang-2009-image-classification-annotation}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % 1 to 2 sentences
+
+Topic models, when properly visualized, can be an effective tool in the
+qualitative analysis of large text corpora.
 
@@
 
+This project will consist of two main parts.  The first part will explore good
+visualizations of topic models and will result in a tool that incorporates
+those visualizations in an interactive browser.  The second part will consider
+the usefulness of topic models for thematic analysis, including the ability of
+a topic model to capture human-specified topics and the validation of automated
+methods for finding topics of interest to a person browsing the corpus.
+
+\subsection{The Topic Browser}
+
+Proper visualizations are essential to extracting information from and
+identifying trends in data, especially large sets of high dimensional data.
+Large text corpora are particularly difficult to visualize, as they may include
+many thousands of documents and millions of words.  Topic
+modeling~\cite{blei-2003-latent-dirichlet-allocation} is a method of reducing
+the dimensionality of a corpus into a set of meaningful topics.  Topic models
+have been used with some success as an aid to corpus
+visualization~\cite{blei-2009-topic-models,
+newman-2010-visualizing-with-topic-maps}.  However, in many papers the
+presentation of topic models has been limited to hand-selected, coherent
+topics, when often there are many meaningless topics to sift through.  These
+presentations leave the viewer wondering what else the topic model discovered
+and provide little help in gaining a deep understanding of the model.
+
+To aid in the visualization of topic models and in pattern discovery in
+document collections, this project will build a system we call The Topic
+Browser, a web-based tool for interactively exploring both the output of a
+topic modeling algorithm and its attendant corpus.  This topic browser will
+incorporate many visualizations of topic models previously published as well
+as some innovative ideas of our own.  The Topic Browser is an aid both to those
+who wish to browse through a corpus and for those who wish to analyze the topic
+model itself.  We believe that this interactive approach to visualization in
+browsing reveals topics and topical trends in large document collections in a
+way that is not possible with static visualizations.
+
+One of the innovative ideas incorporated into the topic browser is the use of
+topic metrics that can be used to separate topics of interest from
+not-as-interesting topics.  The sidebar navigation tool in the topic browser
+allows the user to sort and filter topics according to these metrics.  This
+allows a person navigating the corpus to easily find the topics he is looking
+for, instead of having to sift through all of the topics by hand.
+
+\subsection{LDA for Thematic Analysis}
+
+The second part of this project is motivated by the first---given a tool to
+browse through the output of a topic model, and topic metrics that are supposed
+to show topics relevant to certain user questions, can we quantitatively
+demonstrate that it actually works?
+
+Topic models have been shown to be useful in many tasks, and it is generally
+assumed that they do a good job at capturing thematic elements in large bodies
+of text.  While this has been anecdotally asserted on many occasions, to our
+knowledge it has never been rigorously established.  Thus topic modeling has
+not been widely used in the qualitative analysis of large corpora.
+
+This part of the project will quantitatively show that topic models can
+effectively capture qualitative or thematic elements in a collection of
+documents as judged by humans, and that automated methods can select topics
+from a topic model that match human thematic judgments in a variety of
+situations.
+
+In order to establish this, I will use two sources of data.  The first comes
+from student ratings of courses and professors at Brigham Young University.
+After answering many questions which require selecting one of several options,
+students are given the opportunity to write in free text whatever comments they
+have about the class.  We will use those free-form comments as one source of
+text in which we will find thematic elements.  The second source of data we
+will use is a set of restaurant reviews from New York CitySearch.
+
+We will choose four subsets of the data to annotate, two in each data source,
+each subset consisting of about 40 documents (where each document is
+approximately one paragraph long).  We will have 10 people annotate each subset
+of the data with topics that they judge to be relevant, so that we can have
+some idea of the consistency in human-judged topics.
+
+We will then develop methods for comparing a human-annotated topic to a topic
+produced by a topic model.  One of the main parts of this comparison will be
+finding the percent of words in the topic annotated by humans that were labeled
+with the topic by the topic model.  We will elicit from the annotators words
+and phrases that are representative of each topic annotated, along with a
+phrasal description of the topic, both of which will be useful in matching the
+human topics with machine-learned topics.
+
+Once we have methods to compare the human annotations with the machine output,
+we will see to what extent topic models capture human topics.  This will
+provide a new extrinsic evaluation method for topic models, as some models will
+better capture human topics than others.  We will then also validate the use of
+topic metrics to automatically sort topics.  In particular, we will see if
+these methods can find the topics the humans thought we prevalent in the
+annotated subset of the data when the topic model is run on the entire dataset.
+Because we will have two annotated subsets for each data source, we will also
+be able to test the ability of automated methods to determine which topics are
+common between the two subsets, and which are unique.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % 1/2 to 2 pages
+
+This project is expected to produce an easily usable, extensible, and helpful
+tool for browsing the output of a topic model, and to validate the use of topic
+modeling for the thematic analysis of large text corpora.
+
+\subsection{The Topic Browser}
+
+The first stage of this project is successful if it produces a tool that is
+useful to the topic modeling community (including those who do research in
+topic modeling and those who use topic models) to visualize their results.
+While I will not seek to quantitatively verify with user studies that the tool
+produced is a better way to browse topic models than existing tools, its
+usefulness should be intuitively visible to those using the browser.  This can
+be at least anecdotally verified by other groups at BYU and other universities
+using our tool, and the acceptance of a paper describing the browser at a
+peer-reviewed conference workshop.
+
+While the project will not verify the usability of the tool as a whole, some
+parts of the tool can be individually verified.  Some of the previous research
+included in the topic browser has already been independently validated, such as
+the Turbo Topics method of extracting phrases from
+topics~\cite{blei-2009-turbo-topics}.  In particular, the second stage of this
+project will validate the novel contributions in the sidebar of the browser.
+
+\subsection{LDA for Qualitative Analysis}
+
+The second stage of this project investigates the use of topic models as an aid
+in the qualitative analysis of large copora.  This stage is successful if it
+identifies in what ways topic models successfully assist a human in analyzing a
+large collection of text.  We take successfully assisting a human to mean
+saving the human work by automatically reproducing themes that a person would
+have spent hours or days producing.  We will determine a topic model's ability
+to reproduce human themes as outlined in the project description.
 
@@
 
-\item Completed: reformulate PSO in MapReduce.
+\item Completed: build the Topic Browser.
 
-\item February 6, 2009: evaluate PSO with large swarms with a few benchmark
-functions.
+\item February 4, 2011: write code to compare human topics with LDA topics.
 
-\item February 20, 2009: evaluate PSO with very large swarms with subswarms
-and with more benchmark functions.
+\item February 11, 2011: finish collecting human annotations.
 
-\item March 9, 2009: rework individual papers into a coherent thesis and
+\item March 4, 2011: submit a paper to CoNLL detailing the work on validating
+  LDA for thematic analysis.
+
+\item March 11, 2011: rework individual papers into a coherent thesis and
 submit draft to committee chair.
 
-\item March 20, 2009: submit thesis to second committee member for review.
+\item March 20, 2011: submit thesis to second committee member for review.
 
-\item March 30, 2009: submit thesis to third committee member for review.
+\item March 30, 2011: submit thesis to third committee member for review.
 
-\item April 10, 2009: defend thesis.
-
-Each of the three stages of the project can be submitted as a paper to a
-conference such as GECCO or CEC.
+\item April 13, 2011: defend thesis.
 
@@
 
-\bibliography{}
+\bibliography{../../../bib/lda/bib}
 

### 2011-02-08 honors_thesis/si_paper/reviewer_response/response.tex
Major revisions to paper in response to feedback

@@
 
-Response: Copy and paste from Andrew's paper
+Response: We added a Related Work section (now section 3), describing previous
+approaches at parallelizing PSO.  We also added (NOT DONE) experiments
+comparing against another common parallelization approach, that of using
+subswarms.
 
@@
 
-Response: Get Andrew's paragraph that deals with this
+Response: We added a paragraph to what is now Section 7 addressing the issue of
+benchmark functions (``We also do not report\ldots'').  We also ran a
+comparison on a longer-running function and included it in our results (NOT
+DONE).
 
@@
 section, so we split the two main parts of it into two separate sections.
-There is now a section devoted entirely to the mathematics of the method and a
-separate section describing its implementation.
-
-NOT DONE
+There is now a section devoted entirely to the mathematics of the method (now
+Section 4) and a separate section describing its implementation (Section 5).
+We also put subsections in the new Section 4 to guide the reader a little
+better through the section.
 
@@
 
-Response: This was not specific enough to use too much...
+Response: We substantially reorganized Section 3, as noted above, and we moved
+the discussion of branch statistics (the old Section 4.3) to the beginning of
+the results section (NOT DONE).
 
@@
 
-Response: How do we fix this?
-
-NOT DONE
+Response: We had a long discussion about notation before writing our paper, and
+we thought that because the personal best and neighborhood bests are positions,
+using an x to denote it would be more intuitive.  However, we see your point
+that it can be confusing for other reasons.  We changed the x to a b when
+referring to best positions, while still using the N and P superscripts to
+differentiate between which best is being referred to.
 
@@
 
-Reviewer comments: Using "rounds of evaluations" is confusing, and results
+Reviewer comments: Using ``rounds of evaluations'' is confusing, and results
 should be reported in terms of wall clock time.
@@
 Response: Rounds of evaluations correspond exactly to wall clock time, as we
-mentioned.  However, we realize that the term is not a commonly used one, so we
-changed "rounds of evaluations" to "time steps" in hopes of better conveying
-our meaning. (Also incorporate Andrew's paragraph on why we're using time steps
-instead of evaluations)
-
-NOT DONE
+describe in the second paragraph of what was Section 5 (now Section 7).
+However, we realize that the term is not a commonly used one, so we changed
+``rounds of evaluations'' to ``time steps'' in hopes of better conveying our
+meaning.  We also added some text in the description mentioned above that
+hopefully further clarifies this point.
 
@@
 
-Response: We need to look up what he's talking about.
+Response: Running a few more sets of experiments at higher dimensionalities
+should address this adequately, though we can't really do all of their
+protocol.
 
@@
 Reviewer comments: In the abstract, it is not clear the meaning of
-"speculative". A reader without knowledge about parallel computing may not
+``speculative''. A reader without knowledge about parallel computing may not
 understand what the authors mean.
 
-Response: Need to address this.
-
-NOT DONE
+Response: We had a paragraph explaining this in the introduction, though the
+analogy to processors lacked an explicit connection to the word
+``speculative.''  We added a sentence at the end of that paragraph to address
+this issue.
 
@@
 insufficiently discussed.  The authors say that adding more particles leads to
-"diminishing returns". This may be true when dealing with constant size
+``diminishing returns''. This may be true when dealing with constant size
 populations.  However, a recent trend is to vary the population size over time.
 In fact, it has been shown that adding particles \emph{over time} may actually
-solve the problem the authors refer to. Some references are "Incremental Social
-Learning in Particle Swarms" by Montes de Oca and others published in IEEE SMCB
-and "Efficient Population Utilization Strategy for Particle Swarm Optimizer" by
-Hsieh and others in the same journal.
+solve the problem the authors refer to. Some references are ``Incremental
+Social Learning in Particle Swarms'' by Montes de Oca and others published in
+IEEE SMCB and ``Efficient Population Utilization Strategy for Particle Swarm
+Optimizer'' by Hsieh and others in the same journal.
 
@@
 allows, you should do speculative evaluation.  That's our point.
+
+Put those papers in related work, on another way to decide what computation to
+perform---some people increase swarm size over time.  Relate that to our point.
 
@@
 Reviewer comments: In the first paragraph of Section 3, the authors say that
-adding particles may not help past a certain swarm size. This is true is the
+adding particles may not help past a certain swarm size. This is true if the
 computation time allocated to the optimization process remains fixed. However,
@@
 
-Response: Most people have a fixed computational budget, function has a set
-time.  That is the case we are interested in.
+Response: We added some text to to the introduction (paragraph beginning with
+``In this paper we consider\ldots'') to better explain our intent; that is,
+given a fixed computational budget, how can processors be used most
+efficiently?  However, it is also the case that if there is ever a point in
+your computation where you have a bigger computational budget than the
+particles themselves demand, speculative evaluation is the best way to utilize
+the remaining computational budget.  We added text to the conclusion to explain
+this point (or maybe to related work?).
 
@@
 
-Response: Restate what the case means
-
-NOT DONE
+Response: Your question probably arose from the confusing notation.  We have
+fixed the notation and added a parenthetical statement directly preceeding the
+equation to better explain what the equation is.  The equation is a specific
+example of a possible update case, and in the case described the neighborhood
+best is not updated; thus by definition the function values of current
+positions must not be lower than the value of the previous neighborhood best.
 
@@
 
-NOT DONE
-
 \subsection*{1.12}
@@
 Response: Pruning allows for the use of a larger swarm with the same number of
-processors.  We tried to make this point more clear in the text.
-
-NOT DONE
+processors.  We realized that this was never explicitly stated in the text, so
+we added a few sentences to the description of pruning (now section 6.2) to
+make it clear.  The text added to the introduction explaining our purpose of
+fully utilizing resources also addresses this issue, though somewhat
+indirectly.
 
@@
 
-Response: todo
+Response: In fact the best result is improved, frequently, it is just that the
+scale of the improvements becomes increasingly smaller.
 
@@
 
-Response: ...
-
-NOT DONE
+Response: Duly noted.  We fixed the sentence (it now says ``Our methods work
+very well on these functions'').
 
@@
 
-Reviewer comments: Page 31. Line 46. Molia -> Molina. Is it a technical report?
-Which number?  Isn't it published with the special issue?
+Reviewer comments: Page 31. Line 46. Molia $\rightarrow$ Molina. Is it a
+technical report?  Which number?  Isn't it published with the special issue?
 
-Response: ...
-
-NOT DONE
+Response:  We fixed the typo, and we looked for quite some time trying to find
+a better way to cite that paper.  The special issue hasn't been officially
+published yet, from what I can tell, and the editorial that is online (under
+``Online First'' at springerlink.com) only has a link to the test suite
+document.  I suppose that we could cite the editorial instead of the test suite
+document, if you think that would be preferable.
 
@@
 Reviewer comments: In the first sections of the paper it is argued that the
-proposed "...method can easily be extended to arbitrary topologies". However,
+proposed ``...method can easily be extended to arbitrary topologies''. However,
 in later sections it is apparent that this may be possible if the underlying
-neighborhood topology is not "dense" enough to cause heavy communication
+neighborhood topology is not ``dense'' enough to cause heavy communication
 traffic to the processors.  Can you be more precise to this and quantify it?
@@
 
-Response: ...
+Response: The citation given in the paper (McNabb et al. CEC 09) gives an
+interesting discussion on these points.  The functions explored are limited,
+but the paper shows that both Sphere and Griewank exhibit this behavior, while
+Rastrigin does not (the relevant figures in that paper are 2, 4 and 6).  As
+those results are previously published and cited in our paper, we felt it
+unnecessary to reduplicate the figures, though we make mention of the
+conclusions.  However, our own figures also answer your question.  The fact
+that our methods outperform the na\"ive parallelization is in itself evidence
+for this claim.  Recall that when our figures say SEPSO, that is equivalent to
+running PSO with a swarm size one eighth the size, doing twice as many
+iterations.  Figures 9 and 10 illustrate the point that for some swarm sizes
+you are correct and it is better to have a larger swarm size.  However, once
+enough processors are available to have a large enough swarm even with SEPSO,
+it is better by far to do speculative evaluation.
 
-NOT DONE
+Increasing the swarm size increases the diversity, as you say (the McNabb et
+al. paper makes the same point).  But at some point there is enough diversity,
+and when that point is reached extra processors should be used for speculative
+evaluation.  We spend almost the entire conclusion trying to make this point.
 
@@
 
-Response: ...
-
-NOT DONE
+Response: We found only one part of the introduction that mentioned the paper
+not proposing a new algorithm.  That part had two sentences, and we deleted the
+second one to soften the point a little.  It was an over-reaction to comments
+by previous reviewers.
 
@@
 
-Response: reference constricted PSO in formulas
-
-NOT DONE
+Response: You are referring to inertia weight PSO (the original equations in
+Kennedy 1995 did not have any weighting factor).  Constricted PSO is also a
+widely used improvement on the original PSO, and we cite two papers as
+justification for our use of those formulas.  However, our text wasn't clear
+that we were using the constricted PSO equations, so we clarified the text.
 
@@
 
-Response: ...
-
-NOT DONE
+Response: Thanks for catching that.  We fixed it and went over our other
+references carefully.
 
@@
 
-Response: ...
-
-  a)Pei-Wei Tsai, Jeng-Shyang Pan, Shyi-Ming Chen, Bin-Yih Liao, and Szu-Ping
-Hao, 2008, "Parallel Cat Swarm Optimization," Proceedings of the Seventh
-International Conference on Machine Learning and Cybernetics, Kunming, China,
-Jul. 12-15, 2008, pp. 3328-3333.
-
-  b)Jui-Fang Chang, Shu-Chuan Chu, John F. Roddick and Jeng-Shyang Pan, 2005,
-"A Parallel Particle Swarm Optimization Algorithm with Communication
-Strategies", Journal of Information Science and Engineering, Vol. 21, No. 4,
-pp. 809-818.
-
-NOT DONE
+Response: The Chu et al. paper that we already cite seems to be an extension of
+the Chang et al. paper you mentioned.  However, we included the Chang et al.
+paper along with the Chu et al. paper in our review of related work.  We also
+included several more related papers in our review of current parallelization
+strategies for PSO.
 

### 2011-02-08 honors_thesis/si_paper/paper.tex
Major revisions to paper in response to feedback

@@
 \providecommand{\pvel}{\ensuremath{\Vec{v}}}
-\providecommand{\nbest}{\ensuremath{\Vec{x}^\neigh}}
-\providecommand{\fnbest}{\ensuremath{\Vec{fx}^\neigh}}
-\providecommand{\pbest}{\ensuremath{\Vec{x}^\pers}}
-\providecommand{\fpbest}{\ensuremath{\Vec{fx}^\pers}}
+\providecommand{\nbest}{\ensuremath{\Vec{b}^\neigh}}
+\providecommand{\fnbest}{\ensuremath{\Vec{fb}^\neigh}}
+\providecommand{\pbest}{\ensuremath{\Vec{b}^\pers}}
+\providecommand{\fpbest}{\ensuremath{\Vec{fb}^\pers}}
 \providecommand{\constriction}{\ensuremath{\chi}}
@@
 effective algorithm for solving many types of optimization
-problems~\cite{poli-2008-pso-applications}.  For much of the algorithm's
+problems~\citep{poli-2008-pso-applications}.  For much of the algorithm's
 history, PSO was run serially on a single machine.  However, the world's
@@
 
-However, it can be seen from the graph that doubling the swarm size produces
-only incremental benefit; performance improves, but not by very much.  The
-reason for this is that there comes a point of diminishing returns with respect
-to adding particles.  In \fig{evals-sphere} we show the value obtained after
-50,000 function evaluations (not iterations) as a function of swarm size, again
-for the function Sphere.  Increasing the swarm size from 5 to 10 has a
-significant effect on the value obtained.  However, increasing the swarm size
-from 16 to 30 makes the algorithm less efficient; that is, it reduces the
-progress the algorithm makes per evaluation.  Other functions show similar
-trends, though often the optimal swarm size is slightly larger.  For this
-reason, previous work has recommended the use of a swarm size of 50 for
+However, it can be seen from the graph that once the swarm is sufficiently
+large, doubling the swarm size produces only incremental benefit; performance
+improves, but not by very much.  The reason for this is that there comes a
+point of diminishing returns with respect to adding particles.  In
+\fig{evals-sphere} we show the value obtained after 50,000 function evaluations
+(not iterations) as a function of swarm size, again for the function Sphere.
+Increasing the swarm size from 5 to 10 has a significant effect on the value
+obtained.  However, increasing the swarm size from 16 to 30 makes the algorithm
+less efficient; that is, it reduces the progress the algorithm makes per
+evaluation.  Other functions show similar trends, though often the optimal
+swarm size is slightly larger.  For this reason, previous work has recommended
+the use of a swarm size of 50 for
 PSO~\citep{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
@@
 purpose is to explore the question of what to do with a thousand processors
-when 50 or 100 particles is the most efficient swarm size, and simply
-adding particles results in only incremental improvement.
+when 50 or 100 particles is the most efficient swarm size, and simply adding
+particles results in only incremental improvement.  In our work we assume that
+the researcher has some fixed number of processors that should be fully
+utilized throughout the optimization process.
 
@@
 reference.  If it guesses wrong, execution restarts where it would have been
-anyway.
+anyway.  Thus the processor speculates about future paths of execution in an
+attempt to decrease overall processing time.
 
@@
 factored such that a set of speculative positions can be found which will
-\emph{always} include the position computed in the next iteration.  By
-computing the value of the objective function for each of the speculative
-positions at the same time the algorithm evaluates the objective function for
-the current position, it is possible to know the objective function values for
-both the current and the next iteration at the same time.  We demonstrate this
-principle by implementation and show that it produces exactly the same results
-as standard PSO, but two iterations at a time.  The resulting implementation
-runs efficiently on large clusters where the number of processors is much
-larger than a typical or reasonable number of particles, producing better
-results in less ``wall-clock'' time.
-
-It is important to note here that this is not a variant of PSO.  We are not
-proposing a new algorithm, and we do not try to beat the state of the art in
-PSO research.  We simply propose a new way to think about the parallelization
-of PSO that we show works better in many cases than a na\"ive parallelization.
+always include the position computed in the next iteration.  By computing the
+value of the objective function for each of the speculative positions at the
+same time the algorithm evaluates the objective function for the current
+position, it is possible to know the objective function values for both the
+current and the next iteration at the same time.  We demonstrate this principle
+by implementation and show that it produces exactly the same results as
+standard PSO, but two iterations at a time.  The resulting implementation runs
+efficiently on large clusters where the number of processors is much larger
+than a typical or reasonable number of particles, producing better results in
+less ``wall-clock'' time.
+
+It is important to note here that this is not a variant of PSO.  We simply
+propose a new way to think about the parallelization of PSO that we show
+performs better in many cases than previous parallelizations.
 
@@
 The balance of this paper is organized as follows. \sec{pso} describes the
-particle swarm optimization algorithm.  \sec{sepso} describes mathematically
-how speculative evaluation can be done in parallel PSO to perform two
-iterations at once, and \sec{implementation} gives information on how to
-implement this method.  In \sec{relax}, we discuss various methods of improving
-the performance of speculative evaluation in PSO, all of which break the
-requirement of strictly reproducing the behavior of the original algorithm.
+particle swarm optimization algorithm, and \sec{related} gives a brief overview
+of previous parallelization techniques for this algorithm.  \sec{sepso}
+describes mathematically how speculative evaluation can be done in parallel PSO
+to perform two iterations at once, then \sec{implementation} gives information
+on how to implement this method.  In \sec{relax}, we discuss various methods of
+improving the performance of speculative evaluation in PSO, all of which break
+the requirement of strictly reproducing the behavior of the original algorithm.
 \sec{setup} describes the experiments we ran, and \sec{results} presents the
@@
 
-At each iteration of the algorithm, the position $\ppos_t$ and velocity
+At each iteration of constricted PSO, the position $\ppos_t$ and velocity
 $\pvel_t$ of each particle are updated as follows:
@@
 then it must do poorly on average for other
-problems~\cite{wolpert-1997-nfl-for-optimization}.
+problems~\citep{wolpert-1997-nfl-for-optimization}.
 
@@
 topology requires thorough experimentation for the particular
-problem~\cite{bratton-2007-defining-a-standard-for-pso}.  Other authors have
+problem~\citep{bratton-2007-defining-a-standard-for-pso}.  Other authors have
 explained that a large swarm with a sparse topology propogates information
-slowly~\cite{montes-de-oca-2009-frankensteins-pso}.
+slowly~\citep{montes-de-oca-2009-frankensteins-pso}.
+
+\section{Related Work}
+\label{sec:related}
+
+There have been several parallel implementations of PSO presented in the
+literature.  These implementations vary along two axes---what computation is
+performed and how it is done.  How the computation is performed involves both
+architectural decisions, such as a master-slave versus a fully distributed
+architecture, and decisions about synchronicity.  What computation is performed
+involves decisions about how to scale the PSO algorithm for parallel
+implementations, such as the ratio of particles to processors or the topology
+to use.
+
+\subsection{Synchronous vs. Asynchronous PSO}
+
+Synchronous parallel implementations of PSO reproduce the standard serial
+algorithm exactly.  This approach was first described analytically by
+\citet{belal-2004-parallel-models-for-pso} and first implemented by
+\citet{schutte-2004-parallel-global-optimization-with-pso}.  In a typical
+master-slave algorithm, the master assigns tasks to slave processors, and in
+parallel PSO, each task consists primarily of a function evaluation.  Updating
+the particle's position and value may also be included in the
+task~\citep{belal-2004-parallel-models-for-pso}, or this work may be performed
+in serial by the
+master~\citep{schutte-2004-parallel-global-optimization-with-pso}.  Before
+proceeding to the next iteration, particles communicate, and each particle
+updates its neighborhood best.  Whether this communication step happens
+sequentially on the master or in parallel, each particle must receive
+communication from its neighbors before proceeding.  The benefits of the
+synchronous PSO include its simplicity, repeatability, and comparability with
+standard PSO, which may be essential in research applications.
+
+Asynchronous parallel
+PSO~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
+koh-2006-parallel-asynchronous-pso} is a modification to the standard algorithm
+which removes the synchronization point at the end of each iteration.  Instead,
+particles iterate independently and communicate asynchronously.  In a typical
+master-slave implementation of asynchronous parallel PSO, the master updates
+each particle's personal best, neighborhood best, velocity, and position
+immediately after receiving the function value from the slave processor.  Since
+this update occurs while other particles are still being evaluated, it may use
+information from the previous iteration for some
+neighbors.\footnote{Asynchronous parallel PSO has been compared to the
+``asynchronous updates'' variant of serial
+PSO~\citep{koh-2006-parallel-asynchronous-pso}.  However, serial PSO with
+asynchronous updates differs from standard PSO in that particles use newer
+information, but asynchronous parallel PSO differs from standard PSO in that
+particles use older information.} In a partially asynchronous implementation,
+particles might wait for some but not all neighbors to complete before
+proceeding~%
+\citep{scriven-2008-asynchronous-pso-in-unreliable-distributed-environments}.
+In some master-slave implementations, particles never get more than one
+iteration ahead of
+others~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
+koh-2006-parallel-asynchronous-pso}.  However, in a fully distributed
+implementation, particles might never wait for information, and one particle
+could complete many more iterations than another
+particle~\citep{scriven-2008-distributed-pso-using-peer-to-peer}.  The main
+effect of asynchronous evaluation is that processors spend less time
+idle---this trait is particularly valuable when processors are heterogeneous or
+function evaluation times are
+varied~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
+koh-2006-parallel-asynchronous-pso}.  Asynchronous parallel PSO behaves
+differently than the standard algorithm and may even produce different results
+between runs.  Most reports conclude that asynchronous communication produces
+similar numerical results to the standard algorithm, but the question has not
+yet been thoroughly
+addressed~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
+koh-2006-parallel-asynchronous-pso}.
+
+\subsection{Scaling PSO to many processors}
+
+The contribution of this work deals not with the implementation details of
+architecture and synchronicity, but with what should be done with the PSO
+equations when many hundreds or thousands of processors are available.  The
+main issues that have been addressed in this space are how many particles to
+use for a particular number of processors and what communication topology
+should be employed.
+
+The number of particles per processor has typically been decided by how long it
+takes to evaluate the function being optimized.  When the function takes longer
+than a few seconds to evaluate, previous techniques have assigned the number of
+particles in the swarm to be the number of processors available~%
+\citep{jin-2005-pso-antenna-designs,mcnabb-2009-large-particle-swarms},
+advocating using as many processors as possible to get the best performance.
+When the function takes less time to evaluate than the TCP/IP stack used to
+send interprocessor communication, parallel implementations assign several or
+many particles to a single processor~\citep{chu-2006-intelligent-parallel-pso,
+chang-2005-parallel-pso-with-communication-strategies}.  Often the processor
+only sends information about the best particle it evaluated to other
+processors~\citep{mcnabb-2009-large-particle-swarms}.
+
+Another popular method is simply to run PSO independently on each of the
+processors available, taking the best result when all of the runs complete.  It
+should be noted that this is equivalent to the previously stated method of
+assigning many particles to each processor, only with no communication between
+processors instead of little communication.  Both of these methods can be
+described as changes in the communication topology of the original PSO
+algorithm~\citep{mcnabb-2009-large-particle-swarms}.
+
+Thus previous work in parallelizing PSO has consisted entirely of increasing
+the swarm size and adapting the topology to be better suited to parallel
+computation.
 
@@
 
-To do so we must first refactor PSO such that the determination of the value
-of the objective function is separate from the rest of the computation.  For
-simplicity, this discussion will describe the case where PSO is performing
-function minimization using the Ring topology.  In this example, each particle
-has two neighbors, the ``right neighbor'' and ``left neighbor,'' whose
-positions are represented as $\rightn$ and $\leftn$ respectively.  Though we
-will only describe the case of the Ring topology, this method can easily be
-extended to arbitrary topologies.
+\subsection{Refactoring the PSO Equations}
+
+To perform two iterations at a time we must first refactor PSO such that the
+determination of the value of the objective function is separate from the rest
+of the computation.  For simplicity, this discussion will describe the case
+where PSO is performing function minimization using the Ring topology.  In this
+example, each particle has two neighbors, the ``right neighbor'' and ``left
+neighbor,'' whose positions are represented as $\rightn$ and $\leftn$
+respectively.  Though we will only describe the case of the Ring topology, this
+method can easily be extended to arbitrary topologies.
 
 The refactoring hinges on the idea that once the random coefficients
-$\pURand_{t}$ and $\nURand_{t}$ are determined, there are only a few possible
-new positions, or updates, for $\nbest$ and $\pbest$.  For the Ring topology
-there are 7 possible update cases, identified in Table~\ref{tab:evals}.  We
-label each case with an identifier referring to the source of the update: a
-minus sign ($-$) represents no update, $L$ represents an update to $\nbest$
-coming from the left neighbor, $R$ represents an update to $\nbest$ coming from
-the right neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$
+$\pURand_{t}$ and $\nURand_{t}$ are drawn, there are only a few possible new
+positions, or updates, for $\nbest$ and $\pbest$.  For the Ring topology there
+are 7 possible update cases, identified in Table~\ref{tab:evals}.  We label
+each case with an identifier referring to the source of the update: a minus
+sign ($-$) represents no update, $L$ represents an update to $\nbest$ coming
+from the left neighbor, $R$ represents an update to $\nbest$ coming from the
+right neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$
 coming from the particle itself.  As an example, $\casexn$ refers to the case
@@
 terms drop to zero except for the case that actually occurs.  For example, the
-indicator function for the specific case $\casexn$ can be written as follows:
+indicator function for the specific case $\casexn$ (which, as is shown in
+Table~\ref{tab:evals}, means that the particle's personal best was updated, but
+its neighborhood best was not) can be written as follows:
 
@@
 	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
-	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
-\nonumber
-	& = \ppos_{t} + \specvel_{t+1}^{\casegen}
+	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t})
+	= \ppos_{t} + \specvel_{t+1}^{\casegen}
 \end{align}
@@
 \nonumber
-  \ppos_{t+1} &= \sum_{c \in \caseset} \\
-\nonumber
-	& \indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
+  \ppos_{t+1} = \sum_{c \in \caseset}
+	&\indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
 	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
@@
 children'' because only one of them will be needed.
+
+\subsection{Discussion}
 
@@
 we could limit ourselves to one speculative evaluation per particle instead of
-$2n+1$.  As not all of the branches are evaluated in any given iteration, we
-call this technique pruning.
-
-A naive approach to pruning is to keep track of the last branch taken by each
+$2n+1$.  With a fixed number of processors, this would allow us to greatly
+increase the swarm size relative to that needed in the original speculative
+algorithm (e.g., with 120 processors, a na\"ive parallelization has a swarm of
+size 120, complete speculative evaluation has a swarm of size 15, and pruning
+the evaluations to only one per particle allows a swarm of size 60).  As not
+all of the branches are evaluated in any given iteration, we call this
+technique pruning.
+
+A na\"ive approach to pruning is to keep track of the last branch taken by each
 particle and speculate on that branch.  This turns out to be correct far less
@@
 two iterations of PSO at each ``iteration,'' so we instead call each
-``iteration'' a ``round of evaluations.''
+``iteration'' a ``time step.''
 
 Because our interest is in exploring the most efficient use of processors, we
-must hold fixed the number of processors used at each round of evaluations in
-all of our experiments, giving all approaches the same number of function
-evaluations.  Thus while we label our graphs with ``Rounds of Evaluations,''
-this is equivalent to the number of function evaluations used by each approach
-(simply multiply by the number of processors to transform rounds of evaluations
-to function evaluations).  We use ``Rounds of Evaluations'' to reflect
-wall-clock time, as our primary interest is using extra processors in parallel
-to improve performance on functions with long evaluation times.  A smaller
-swarm running serial PSO would likely produce better results than we present in
-the same number of function evaluations, but because the function evaluations
-were not done in parallel it would take more wall-clock time.  When comparing
-the performance of parallelized algorithms on long-running functions, we must
-compare by wall-clock time with the same resources, not solely by the number of
-function evaluations.
+must hold fixed the number of processors used at each time step in all of our
+experiments, giving all approaches the same number of function evaluations.
+Thus while we label our graphs with ``Time Steps,'' this is equivalent to the
+number of function evaluations used by each approach (simply multiply by the
+number of processors to transform time steps to function evaluations).  We use
+``Time Steps'' to reflect wall-clock time, as our primary interest is using
+extra processors in parallel to improve performance on functions with long
+evaluation times.  A smaller swarm running serial PSO would likely produce
+better results than we present in the same number of function evaluations, but
+because the function evaluations were not done in parallel it would take more
+wall-clock time.  When comparing the performance of parallelized algorithms on
+long-running functions, we must compare by wall-clock time with the same
+resources, not solely by the number of function evaluations.
+
+We also do not report absolute time used by each algorithm, as specific time
+values are heavily implementation-dependent.  Using time steps, where at every
+time step each processor evaluates the function once, is an abstraction from
+individual implementations that still conveys all of the information necessary
+to compare algorithms.  In this paper we focus on the effects of algorithm
+choices on parallel PSO performance.  As there are inexpensive functions with
+complex landscapes and expensive functions with simple landscapes, the behavior
+of PSO with respect to the function is the main issue.  For this purpose,
+benchmark functions are a useful and efficient tool for understanding the
+effects of PSO with expensive objective functions even though the benchmark
+functions are themselves inexpensive.  For example, a plot of performance with
+respect to iterations for a smooth unimodel function with five minutes per
+evaluation would be similar to that of Sphere.
 
@@
   \includegraphics[width=.8\columnwidth]{rastrigin}
-  \caption{Function Rastrigin with a swarm that uses 240 processors per round
-  of evaluations.  All speculative algorithms use the Random topology.}
+  \caption{Function Rastrigin with a swarm that uses 240 processors per time
+  step.  Recall that at each time step all processors evaluate the function
+  once.  All speculative algorithms use the Random topology.}
   \label{fig:rastrigin}
@@
   \includegraphics[width=.8\columnwidth]{sphere}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations.  SEPSO uses the Random topology.}
+  \caption{Function Sphere with a swarm that uses 240 processors per time step.
+  Recall that at each time step all processors evaluate the function once.
+  SEPSO uses the Random topology.}
   \label{fig:basic-sphere}
@@
   \includegraphics[width=.8\columnwidth]{sphere2}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing Pick Best with previous results.  All speculative
-  algorithms use the Random topology.}
+  \caption{Function Sphere with a swarm that uses 240 processors per time step,
+  comparing Pick Best with previous results.  All speculative algorithms use
+  the Random topology.}
   \label{fig:sphere-pickbest}
@@
   \includegraphics[width=.8\columnwidth]{sphere3}
-  \caption{Function Sphere with a swarm that uses 240 processors per round of
-  evaluations, comparing speculating several iterations ahead with previous
-  results.  All speculative algorithms use the Random topology.}
+  \caption{Function Sphere with a swarm that uses 240 processors per time step,
+  comparing speculating several iterations ahead with previous results.  All
+  speculative algorithms use the Random topology.}
   \label{fig:sphere-manyiters}
@@
   \includegraphics[width=.8\columnwidth]{schwefel1}
-  \caption{Function Schwefel with a swarm that uses 240 processors per round of
-  evaluations.  SEPSO uses a Random topology.}
+  \caption{Function Schwefel with a swarm that uses 240 processors per time
+  step.  Recall that at each time step all processors evaluate the function
+  once.  SEPSO uses a Random topology.}
   \label{fig:schwefel1}
@@
   \includegraphics[width=.8\columnwidth]{schwefel2}
-  \caption{Function Schwefel with a swarm that uses 240 processors per round of
-  evaluations, comparing all speculative algorithms.  All speculative
-  algorithms use the Random topology.}
+  \caption{Function Schwefel with a swarm that uses 240 processors per time
+  step, comparing all speculative algorithms.  All speculative algorithms use
+  the Random topology.}
   \label{fig:schwefel2}
@@
   \includegraphics[width=.8\columnwidth]{griewank1}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations.  Instead of showing average function value, we show the
-  percentage of runs that have found the global optimum by each iteration.  All
-  algorithms use the Ring topology.}
+  \caption{Function Griewank with a swarm that uses 240 processors per time
+  step.  Recall that at each time step all processors evaluate the function
+  once.  Instead of showing average function value, we show the percentage of
+  runs that have found the global optimum by each iteration.  All algorithms
+  use the Ring topology.}
   \label{fig:basic-griewank1}
@@
   \includegraphics[width=.8\columnwidth]{griewank2}
-  \caption{Function Griewank with a swarm that uses 800 processors per round of
-  evaluations.  All algorithms use the Ring topology.}
+  \caption{Function Griewank with a swarm that uses 800 processors per time
+  step.  All algorithms use the Ring topology.}
   \label{fig:basic-griewank2}
@@
   \includegraphics[width=.8\columnwidth]{griewank3}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing Pick Best with previous results.  All algorithms use
-  the Ring topology.}
+  \caption{Function Griewank with a swarm that uses 240 processors per time
+  step, comparing Pick Best with previous results.  All algorithms use the Ring
+  topology.}
   \label{fig:griewank-pickbest}
@@
   \includegraphics[width=.8\columnwidth]{griewank4}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing pruning with previous results.  All algorithms use the
-  Ring topology.}
+  \caption{Function Griewank with a swarm that uses 240 processors per time
+  step, comparing pruning with previous results.  All algorithms use the Ring
+  topology.}
   \label{fig:griewank-pruned}
@@
   \includegraphics[width=.8\columnwidth]{griewank5}
-  \caption{Function Griewank with a swarm that uses 240 processors per round of
-  evaluations, comparing speculating several iterations ahead with previous
-  results.  All algorithms use the Ring topology.}
+  \caption{Function Griewank with a swarm that uses 240 processors per time
+  step, comparing speculating several iterations ahead with previous results.
+  All algorithms use the Ring topology.}
   \label{fig:griewank-manyiters}
@@
   \includegraphics[width=.8\columnwidth]{griewank6}
-  \caption{Function Griewank with a swarm that uses 800 processors per round of
-  evaluations.  All algorithms use the Ring topology.}
+  \caption{Function Griewank with a swarm that uses 800 processors per time
+  step.  All algorithms use the Ring topology.}
   \label{fig:griewank-800}
@@
   \includegraphics[width=.8\columnwidth]{bohachevsky}
-  \caption{Function Bohachevsky with a swarm that uses 800 processors per round
-  of evaluations.  All of the speculative algorithms use the Random topology.}
+  \caption{Function Bohachevsky with a swarm that uses 800 processors per time
+  step.  Recall that at each time step all processors evaluate the function
+  once.  All of the speculative algorithms use the Random topology.}
   \label{fig:bohachevsky}
@@
 be done, and our speculative methods show improvements on Sphere of over 40
-orders of magnitude.  Such functions are perfect for our methods.
+orders of magnitude.  Our methods work very well on these functions.
 
@@
 
-\appendix[Alternate form of message passing]
+\section*{Appendix---Alternate form of message passing}
 
@@
 This second method only requires one round of passing information because
-information about iterations $t$ and $t+1$ is passed at the same time.  Each
-processor reconstructs from the messages it receives all of the information
-that it needs about its neighbors.  Messages are passed directly after
-evaluating each particle and its children, so all messages are of the form of
-$\nonbest{\p}_t$ or $\nonbest{\s}_{t+1}$.  The first iteration needs to be
-treated specially, so each particle can produce its initial set of speculative
-children---neighbors need only pass their initial position.  This kind of
-message passing necessitates the careful use of random seeds, so that when each
-processor computes the motion equations for its neighbors it gets the same
-results as its neighboring processors.
+information about both iterations $t$ and $t+1$ is passed at the same time.
+Each processor reconstructs from the messages it receives all of the
+information that it needs about its neighbors.  Messages are passed directly
+after evaluating each particle and its children, so all messages are of the
+form of $\nonbest{\p}_t$ or $\nonbest{\s}_{t+1}$.  The first iteration needs to
+be treated specially, so each particle can produce its initial set of
+speculative children---neighbors need only pass their initial position.  This
+kind of message passing necessitates the careful use of random seeds, so that
+when each processor computes the motion equations for its neighbors it gets the
+same results as its neighboring processors.
 
@@
   \psfrag{pt}[C][C]{$\p_{t}$}
-  \includegraphics[width=.3\columnwidth]{messages1}
+  \includegraphics[width=.15\columnwidth]{messages1}
   \caption{The production of $\p_{t}$ from the original particle
@@
   \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
-  \includegraphics[width=.5\columnwidth]{messages2}
+  \includegraphics[width=.25\columnwidth]{messages2}
   \caption{The production of $\nonbest{\p}_{t+1}$ from the original particle 
@@
   \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
-  \includegraphics[width=.6\columnwidth]{messages3}
+  \includegraphics[width=.3\columnwidth]{messages3}
   \caption{The production of each $\nonbest{\n}_{t+1}$ from the original
@@
   \psfrag{pt1}[C][C]{$\p_{t+1}$}
-  \includegraphics[width=.8\columnwidth]{messages4}
+  \includegraphics[width=.4\columnwidth]{messages4}
   \caption{The production of $\p_{t+1}$ from the original particle 
@@
   \psfrag{st3-e}[C][C]{$\noeval{\sset}_{t+3}$}
-  \includegraphics[width=.5\columnwidth]{messages5}
+  \includegraphics[width=.25\columnwidth]{messages5}
   \caption{The production of $\noeval{\p}_{t+2}$ and $\noeval{\sset}_{t+3}$

### 2011-02-09 honors_thesis/si_paper/paper.tex
Added a couple of paragraphs discussing some related work

@@
 
-Thus previous work in parallelizing PSO has consisted entirely of increasing
-the swarm size and adapting the topology to be better suited to parallel
-computation.
+Thus previous work in parallelizing PSO, apart from implementation details, has
+consisted entirely of increasing the swarm size and adapting the topology to be
+better suited to parallel computation.
+
+With regard to increasing the swarm size in PSO, some recent work has suggested
+that increasing the swarm size through the course of the optimization process
+provides better results than having a set swarm
+size~\citep{hsieh-2009-efficient-population-utilization-for-pso,
+montes-de-oca-2010-incremental-social-learning-pso}.  However, these results
+are based on total number of function evaluations, which, when running in
+parallel on expensive functions, is less important than total number of
+iterations.  Other work focusing on parallelization has shown that when extra
+processors are available they should be used, as performance increases with
+swarm size when measuring in terms of number of
+iterations~\cite{mcnabb-2009-large-particle-swarms,
+jin-2005-pso-antenna-designs}.  If the swarm size were varied throughout the
+course of the optimization process, some processors would be sitting idle at
+most iterations.
 
@@
 functions are themselves inexpensive.  For example, a plot of performance with
-respect to iterations for a smooth unimodel function with five minutes per
+respect to iterations for a smooth unimodal function with five minutes per
 evaluation would be similar to that of Sphere.
@@
 used gets into the thousands.
+
+We mentioned related work showing that increasing the swarm size throughout the
+course of the algorithm could provide improved performance over a fixed swarm
+size in serial PSO~\citep{montes-de-oca-2010-incremental-social-learning-pso}.
+If this method were extended to parallel PSO, most processors would be idle in
+the first few iterations, while more would be utilized at the end.  During
+iterations where there are many un-utilized processors, a natural use of them
+would be speculative evaluation, performing two or more of those iterations at
+a time.  Thus our methods have the potential to further improve the parallel
+performance of recently proposed variations of PSO.
 

### 2011-02-10 topic_validation/evaluation_proposal/paper.tex
Moved some files because I'm changing my thesis

\documentclass[onecolumn, 11pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{Proposed validation of topic models and topic ranking}
\author{Matthew Gardner and Mike Thompson}
\date{\today}

\begin{document}
\maketitle

\section{Introduction}

Topic models have been shown to be useful in many tasks, and it is generally
assumed that they do a good job at capturing thematic elements in large bodies
of text.  While this has been anecdotally asserted on many occasions, to our
knowledge it has never been rigorously established.  Thus topic modeling has
not been widely used in the qualitative analysis of large corpora.

It is our intent to quantitatively show that topic models can effectively
capture qualitative or thematic elements in a collection of documents as judged
by humans, and that automated methods can select topics from a topic model that
match human thematic judgments in a variety of situations.

\section{Data}

In order to establish this, we will use two sources of data.  The first comes
from student ratings of courses and professors at Brigham Young University.
After answering many questions which require selecting one of several options,
students are given the opportunity to write in free text whatever comments they
have about the class.  We will use those free-form comments as one source of
text in which we will find thematic elements.

The second source of data we will use is a set of product reviews from
Amazon.com, or some other review site that has publicly available data.  We
have yet to select a particular website from which to retrieve data or which
products to use.  However, we will select products which are common enough that
an average person would be qualified to find themes through a set of comments.

\section{Annotation}

As there are hundreds of thousands of comments available from the student
ratings data that we will use, and similarly large numbers of product reviews,
having annotators read and label every document with thematic elements is an
impossible task.  Instead, will we select a subset of the data that is large
enough to exhibit themes but small enough that an average college student can
annotate the data in approximately half an hour.  We imagine that somewhere
between 50 to 100 comments (where each comment is about a paragraph long) is an
appropriate size.  We will judge the amount of time required by annotating the
data ourselves first.  We expect that between 10 and 20 annotators for each
data set will be sufficient to establish a solid set of test data.  Previous
work on evaluating other aspects of topic models successfully used 9
annotators.

An important aspect of the subset of the data that we annotate will be that it
will consist of a set of documents that all have the same value for some
attribute, such as ratings for the same teacher, or reviews about the same
product.  This will facilitate both the finding of consistent thematic elements
and more extensive kinds of validation, which will be explained later.

The kinds of information we will elicit from the annotators will include the
top overall themes (probably ordered) in the whole collection of documents they
are present, as well as where those themes show up in the documents.  It would
likely be helpful for the annotator to select for each theme a single document
that best exhibits that theme, as well as list words that are commonly used in
the theme.  All of this will help us to judge whether a particular topic from
the topic model actually matches the theme given by the annotators.

\section{Validation Methodology}

There are two parts to validating the use of topic models in the qualitative
analysis of text.  The first is showing that topic models actually capture the
thematic elements in a set of documents, as judged by humans.  The second is
showing that methods for ranking the topics in a topic model accurately rank
topics according to the ranking given by humans.  Each part requires separate
validation.

\subsection{Validating the topic model's topics}

To demonstrate that topic models capture thematic elements, a mechanism is
necessary to compare a topic model's topic to the themes recorded by humans.
Devising this similarity metric will require some research, but as a starting
point we will compare the words given by the annotators to the words in the
topic, as well as the top documents given by the annotators to the documents
with the highest proportion of each topic.  We will also try to use some kind
of mutual information between the words in the topic and the title or
description of each theme as given by the annotators.

\subsection{Validating the topic rankings}

In order to show that automatic rankings can faithfully reproduce human
rankings, we will use standard metrics from information retrieval (IR), such as
mean average precision, to compare the two rankings.  This validation depends
on the first, as we must be able to determine if the topics ranked by the
computer match the themes ranked by humans.  This can be problematic if there
is no topic in the topic model that successfully captures a human theme, or if
there are two topics that each capture the theme equally well.  We may have to
modify the standard IR metrics to be better suited to our task.

\end{document}

### 2011-02-10 topic_validation/pre_proposal/paper.tex
Moved some files because I'm changing my thesis

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{General Outline of Proposed Research}
\author{Matthew Gardner}
\date{Thursday, August 19, 2010}

\begin{document}
\maketitle

\section{Introduction}

There are often occasions when people are confronted with tens or hundreds of
thousands of documents to analyze, either to find something specific or to
simply figure out what the documents contain.  Intelligence analysts routinely
have 50,000 documents or more to go through in a
year~\cite{washingtonpostarticle}.  Several recent controversies involving
documents posted on WikiLeaks presented people with tens of thousands of
documents to sift through.  Even with a specific purpose in mind, exploring
such a large collection of documents can be entirely unmanageable.  An
intelligence analyst simply looking for interesting subjects in the documents,
without a particular query in mind, would have an even harder time.

A journalist browsing through the Afghan war documents that were recently
posted on WikiLeaks said he started with a query, then went where it led him,
following his curiosity to still more queries, using search as his primary
corpus-browsing tool.  Such methods are clearly very useful in many cases, but
in the case of discovering the interesting aspects of a set of documents, they
cannot be completely exhaustive.  One cannot query for something he has not
thought of.

Topic modeling, including latent Dirichlet allocation and its variations, is a
tool that is well suited to capturing the semantic content in a set of
documents~\cite{blei-2003-latent-dirichlet-allocation}.  Topic modeling has
been used in many applications, including being coupled with sentiment
classifiers, as a substitute for part of speech labels, and as an aid in
information retrieval tasks.  However, its usefulness as a general tool for
corpus browsing is not well established, and efforts to make it so are still
very preliminary, though there has been good work in the last few years.

This work seeks to make LDA more applicable to the task of discovering
information in a corpus.  In particular, there are three points of interest in
this work: how to effectively give a label to a topic, how to automatically
discover which topics have important or interesting content and which contain
merely fluff or function words, and how LDA can or should be modified to handle
specific user queries.

\section{Topic Labeling}

Several recent papers have addressed the issue of how to automatically assign a
coherent label to a topic~\cite{mei-2007-automatic-labeling-of-topic-models,
lau-2010-best-topic-word-selection}.  These papers present ideas that are
successful in some instances, but largely ignore hard cases.  It seems that by
looking at the context of how the topic is used in the documents, a decent
label should be found.

\section{Labeling ``Interesting'' Topics}

There are many topics in LDA that capture semantics whose relevance is
immediately apparent to a person browsing the results.  There are other topics
that are still quite relevant, but the relevance is not immediately apparent
for poor visualization.  There are still other topics that simply contain
function words, or formatting, or other uninteresting elements in the documents
(that possibly should have been removed in a preprocessing step).

Some work has been done in the automatic detection of the relevancy of a topic,
including two recent papers on judging topic
coherence~\cite{chang-2009-reading-tea-leaves,
newman-2010-automatic-evaluation-of-topic-coherence}.  However, the lack of
``coherence'' as judged from the top ten words of a topic and evaluated over
Wikipedia does not necessarily imply lack of relevance in the corpus itself.
There must be some way of judging the relevance of a particular topic, and such
a result would make topic modeling a much more accessible tool for corpus
browsing.

\section{LDA with Queries}

In some instances, a person does not have a particular query in mind when
searching through a set of documents.  In many other cases, however, he does.
There are a few published or obvious methods for using LDA to answer search
queries, but to my knowledge they all involve a post-hoc analysis of the
resultant topic model, not a modification to the algorithm in order to answer
queries (granted, I have not done much searching yet in this area, so ``to my
knowledge'' does not mean a lot).  The potential exists to modify the LDA
algorithm to specifically handle queries.  Exactly what the implications of
this would be are not yet clear to me (what does running LDA get you if you
really just want to do a query, for instance?).

\bibliographystyle{plain}
\bibliography{../../../bib/lda/bib}

\end{document}

### 2011-02-10 topic_validation/topic_ranking_nips/paper.tex
Moved some files because I'm changing my thesis

--- 2010-10-24 masters_thesis/topic_ranking_nips/paper.tex
@@
 \author{
-Matthew J. Gardner \\
+Matthew Gardner \\
 Department of Computer Science \\
 Brigham Young University \\
-Provo, UT 84604 \\
 \texttt{mjg82@byu.edu} \\
 \And
-Other authers\ldots \\
-Affiliation \\
-Address \\
-\texttt{email} \\
+Kevin Seppi \\
+Department of Computer Science \\
+Brigham Young University \\
+\texttt{kseppi@cs.byu.edu} \\
+\And
+Eric Ringger \\
+Department of Computer Science \\
+Brigham Young University \\
+\texttt{ringger@cs.byu.edu} \\
 }
 
-%\nipsfinalcopy % Uncomment for camera-ready version
-
 \begin{document}
+\nipsfinalcopy % Uncomment for camera-ready version
 \maketitle
@@
 set of documents.  Such models contain a lot of information and can be very
-difficult to analyze.  For the proper visualization of these models, one must
-present only the topics that are relevant to the task at hand.  We present the
-novel task of topic ranking in response to user questions that greatly aids in
-the visualization of topic models.  We make use of the document distribution of
-a topic in order to rank the topics according to specific user questions.
+difficult to analyze.  Current attempts at using topic models to enhance
+information discovery in a document collection are limited mainly to simple
+corpus browsers.  However, there are many questions a person may have about a
+corpus that are best answered by the topics found by the topic model.  To get
+these kinds of information from the model, one must decide which topics best
+answer the question and present only the best topics.  In an attempt to address
+this issue, we present the novel task of topic ranking in response to these
+kinds of user questions and give some methods that are useful in this task.
 
@@
 
-\section{Introdcution}
+\section{Introduction}
 \label{sec:introduction}
+
+There are many kinds of questions a person might have when confronted with a
+previously unseen collection of documents.  Some of those questions are
+factual, such as, ``Who was Barack Obama's campaign advisor?''  Some questions
+look for documents, such as ``What speeches did John McCain give about the war
+in Iraq?''  Still other questions ask about topics and topical trends in the
+corpus, such as, ``What were Obama's consistent campaign themes?''  Each kind
+of question requires a separate kind of visualization.  The first two kinds of
+questions listed require facts and documents that answer the question, and the
+methods of information retrieval are well suited to those questions.  The last
+kind of question, however, naturally falls into the domain of topic modeling.
 
@@
 sentiment tasks~\cite{blei-2003-latent-dirichlet-allocation,
-brody-2010-aspect-sentiment-model-for-reviews}, but the usefulness of topic
-models as visualization tools in and of themselves has not been well
-established, and is even in
-question~\cite{newman-2010-visualizing-with-topic-maps}.  Some recent work has
-been done in an attempt to make corpus browsers using the results of LDA, and
-to make the results of LDA more easily interpretable, but there is much work
-yet to be done~\cite{blei-arxiv-corpus-browser,
-newman-2010-visualizing-with-topic-maps,
-mei-2007-automatic-labeling-of-topic-models}.
+brody-2010-aspect-sentiment-model-for-reviews}, but, as far as we are aware,
+their use in revealing information about a corpus has previously been limited
+to corpus browsing~\cite{newman-2010-visualizing-with-topic-maps,
+blei-arxiv-corpus-browser}.  While topic-based corpus browsers give a better
+window into the documents in the corpus and can help to visualize some kinds of
+information, they are not capable of effectively answering questions about the
+topics in the corpus.
 
-Common approaches to using topic models to facilitate corpus browsing and
-visualization are limited mainly to showing the similarity between documents in
-the corpus in various ways.  Such methods are useful in their own right when
-one is individually scanning documents, but they are insufficient to answer
-broader questions about the corpus, like, ``What topics show up most in these
-documents?'' and, ``What does a particular author in the corpus talk about most
-frequently that is unique to that author?''  Those questions turn one away from
-analyzing individual documents to trying to get information out of the topics
-themselves.
-
-Upon looking at the topics one is confronted with tens or hundreds of lists of
-words, sorted by a number given to each list by the algorithm.  Sifting through
-those lists and trying to understand what they mean, and whether or not a
-particular topic is important, can be a daunting task, particularly to one not
-very familiar with the algorithm.  The common practice of cherry-picking good
-topics to present in a paper hides the fact that a large number of topics
+In order to properly visualize a topic model in response to topical questions,
+one must present information about the topics themselves, not the documents,
+and decide which topics are most relevant to the information desired by the
+user.  Typical methods of getting such information from a topic model require
+the user to sift through tens or hundreds of lists of words, trying to
+understand what they mean, and decide whether or not a particular topic is
+relevant.  This can be a daunting task.  The common practice of hand-picking
+good topics to present in a paper hides the fact that a large number of topics
 returned by topic models are very hard to interpret or even completely
@@
 
-In order for topic models to be useful for corpus browsing, some sense has to
-be made out of the topics.  One recently proposed method is to assign to each
-topic a coherence value, so that the person can ignore topics that are likely
-to be meaningless~\cite{newman-2010-automatic-evaluation-of-topic-coherence}.
-While topic coherence is able to eliminate meaningless topics from
-consideration, it is not able to address many broader questions that a person
-browsing the corpus may have.
+Clearly, then, for topic models to be useful for visualizing corpora in
+response to these kinds of questions, some sense has to be made out of the
+topics.  One recently proposed method is to assign to each topic a coherence
+value, so that the person can ignore topics that are likely to be
+meaningless~\cite{newman-2010-automatic-evaluation-of-topic-coherence}.  While
+topic coherence is able to eliminate meaningless topics from consideration, it
+is not able to address many broader questions that a person browsing the corpus
+may have.
 
-In this paper we present the novel task of ranking topics by relevance to
-specific user questions.  If one is to successfully visualize a corpus in a way
-that answers broad questions, and if topics, not documents, are the natural
-answer to those questions, then one must properly rank topics such that the
-most prominent topics contain the answer to the user's questions.  We recognize
-that not all questions about a corpus fit into this visualization model, but
-there are many that do, such as those listed above.
+In an attempt to address this issue, we present the novel task of topic
+ranking; that is, deciding which topics are most pertinent to a user's
+topically-based question.  In general this is a very difficult problem, but it
+is essential to the proper visualization of topic models for many applications.
+Surprisingly, it appears that researchers have not as yet attempted to solve
+this problem.  In this paper we formalize the topic ranking problem and present
+some initial attempts at solving a limited subset of possible questions.
 
 The rest of this paper is outlined as follows.  We first give further
-motivation and a more detailed description of the task of ranking topics by
-relevance in \secref{ranking}.  \secref{docdist} describes a method for
-determining the relevance of a topic to various possible questions a person
-could have when looking at a corpus, and \secref{examples} shows examples of
-how this technique provides a proper ranking.  Finally, in \secref{conclusion}
-we conclude.
+motivation and a more detailed description of the topic ranking task in
+\secref{ranking}.  \secref{docdist} describes a method for determining the
+relevance of a topic to a few types of topically-based questions a person could
+have when looking at a corpus, and \secref{examples} shows examples of how this
+technique provides a proper ranking.  Finally, in \secref{conclusion} we
+conclude.
 
@@
 ranking of topics, not simply a window onto individual documents.  When one
-wants to see what topics are prevelant in a corpus, the answer is not found by
+wants to see what topics are prevalent in a corpus, the answer is not found by
 scanning individual documents, but by looking at the topics produced by a topic
@@
 not look through the documents, but at the topics, computing some statistics of
-those topics for each year for which there are documents.  Thus a proper
-visualization of the topic model for these tasks is to show most prominently
-those topics that are most relevant to the task.
+those topics for each year for which there are documents.  Thus the correct
+visualization of a topic model for these tasks includes showing most
+prominently those topics that are most relevant to the task.
 
 We can formalize the topic ranking problem simply as follows.  Given a set of
-topics from a topic model of a corpus, and a user question about the corpus,
-rank the topics by relevance to the user's question.  This is very similar to
-basic information retrieval tasks, but also significantly different, in that we
+topics from a topic model of a corpus, and a topically-based question about the
+corpus (that is, a question whose answer is naturally a topic, not a fact or a
+document), rank the topics by relevance to the question.  This is similar to
+some information retrieval tasks, but also significantly different, in that we
 are not looking for a ranking of documents as the result of a query, but a
@@
 topic models, but as far as we are aware, this is the first attempt to even
-consider the problem.  We propose a very simple way to answer some user
-questions through the use of the document distribution of each topic.  This
-technique is by no means a general solution to the problem, but for some
-specific tasks it provides rankings that gives users the topics that answer
-their questions.
+consider the problem.  We propose a very simple way to answer a few types of
+topically-based questions through the use of the document distribution of each
+topic.  This technique is by no means a general solution to the problem, but we
+show it to be useful for visualizing at least a subset of the kinds of
+information one might be interested in obtaining from a corpus.
 
@@
 over documents in the corpus that this token could have come from?  We
-approximate this distribution with a sample from a Gibbs sampler of LDA.  The
-distribution is thus defined as
+approximate this distribution with a sample from a Gibbs sampler of LDA.  Our
+approximation of this distribution is thus defined as
 \begin{align}
@@
 $p(d|z)$ is defined as in \eqref{eq:entropy}, but instead of the entire corpus,
-$\mathcal{D}$ represents some specific subset.  When using subsets of the
-documents, we only consider topics which have at least one token in the given
-subset, so as to avoid division by zero.
+$\mathcal{D}$ represents some specific subset (in our applications the subset
+always corresponds to a set of documents having some particular attribute, such
+as a common author).  When using subsets of the documents, we only consider
+topics which have at least one token in the given subset, so as to avoid
+division by zero.
 
@@
 
-The first question we address is, What topics are most prevelant in the corpus?
+The first question we address is, What topics are most prevalent in the corpus?
 This is a simple question with an obvious answer that does not require using
@@
   42 & people don ve make country ll time put didn money \\
-  119 & make national work effort including protect adiministration clear
+  119 & make national work effort including protect administration clear
 	support policy \\
@@
   129 & mccain news john senator obama remarks america events donate email \\
-  119 & make national work effort including protect adiministration clear
+  119 & make national work effort including protect administration clear
 	support policy \\
@@
 
-Displaying the right topics to a user is essential to the effective
-visualization of topic models.  We have presented the novel task of ranking
-topics according to relevance to certain user questions.  We have also given a
-method for ranking topics in order to answer a few specific questions.  Such
-rankings allow visualizations of topic models that present information to users
-in much more effective ways than was previously possible.
+Many kinds of information a person might desire to get from a collection of
+documents are best presented as topics from a topic model.  In order to
+determine the relevance of each topic to specific user questions, we have
+presented the novel task of topic ranking.  We have presented methods to answer
+questions of the form, ``What are the consistent themes in this corpus, or for
+this author?''  We have also shown results on a corpus of political speeches
+that intuitively seem correct.
 
-While our methods are useful for a few user questions, it remains as future
-work to devise a method for ranking topics in answer to arbitrary questions.
-The evaluation of such rankings is also a difficult task that must be left to
-future work.
+We believe that the topic ranking task is an important problem that deserves
+more attention than it has as yet received.  This work is very preliminary in
+that it only addresses a very small subset of possible topically-based
+questions.  We plan on continuing to explore topic ranking, expanding the set
+of questions that can be answered and searching for automated means of
+evaluating these rankings.
 

### 2011-02-10 honors_thesis/si_paper/reviewer_response/response.tex
Updated reviewer response to reflect recent changes in the paper

@@
 
-Response: See how papers are relevant (Andrew).  But also, assuming a fixed
-computational budget, if at any point you have a smaller swarm than your budget
-allows, you should do speculative evaluation.  That's our point.
-
-Put those papers in related work, on another way to decide what computation to
-perform---some people increase swarm size over time.  Relate that to our point.
-
-NOT DONE
+Response: We added a paragraph in our related work section discussing these
+methods (the last paragraph of the section).  We also added a paragraph to
+future work discussing how these methods are particularly amenable to
+speculative evaluation when parallelized (the second paragraph of the section).
 
@@
 
-Response: We added some text to to the introduction (paragraph beginning with
-``In this paper we consider\ldots'') to better explain our intent; that is,
-given a fixed computational budget, how can processors be used most
-efficiently?  However, it is also the case that if there is ever a point in
-your computation where you have a bigger computational budget than the
-particles themselves demand, speculative evaluation is the best way to utilize
-the remaining computational budget.  We added text to the conclusion to explain
-this point (or maybe to related work?).
-
-NOT DONE
+Response: This is very much related to the previous item, and so the response
+to that item is relevant here.  We also added some text to to the introduction
+(paragraph beginning with ``In this paper we consider\ldots'') to better
+explain our intent; that is, given a fixed computational budget, how can
+processors be used most efficiently?  The text we added in this and the
+previous item addresses this question.
 

### 2011-02-10 masters_thesis/si_paper/paper.tex
Moved the si_paper to masters_thesis, because that's what it is now

--- 2011-02-09 honors_thesis/si_paper/paper.tex
@@
 \begin{align}
-\nonumber
+\label{eq:velupdate}
 	\pvel_{t+1} &=
 		\constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
-\label{eq:velupdate}
-			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) +
+			\ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
 		\bigr] \\
@@
 swarm size when measuring in terms of number of
-iterations~\cite{mcnabb-2009-large-particle-swarms,
+iterations~\citep{mcnabb-2009-large-particle-swarms,
 jin-2005-pso-antenna-designs}.  If the swarm size were varied throughout the

### 2011-02-14 masters_thesis/si_paper/paper.tex
Reworked a couple of sections in the paper after feedback from Andrew

@@
 There have been several parallel implementations of PSO presented in the
-literature.  These implementations vary along two axes---what computation is
-performed and how it is done.  How the computation is performed involves both
-architectural decisions, such as a master-slave versus a fully distributed
-architecture, and decisions about synchronicity.  What computation is performed
-involves decisions about how to scale the PSO algorithm for parallel
-implementations, such as the ratio of particles to processors or the topology
-to use.
-
-\subsection{Synchronous vs. Asynchronous PSO}
+literature.  The improvements described in the literature come in two major
+areas: innovations in implementation details and innovations in the use of
+topology and swarm size to scale PSO to many processors.
+
+\subsection{Innovative Implementations}
+
+There are many ways to parallelize the basic PSO algorithm.  The most
+fundamental decision to make in parallel PSO is which parallel architecture to
+use.  Several architectures have been proposed, including Master-Slave, fully
+distributed (sometimes called ``diffusion''), and reformulating PSO into
+Google's MapReduce framework~\citep{belal-2004-parallel-models-for-pso,
+mcnabb-2007-parallel-pso-using-mapreduce}.  A somewhat orthogonal
+implementation decision when parallelizing PSO is whether to have synchronous
+communication or asynchronous communication.
 
@@
 
-The contribution of this work deals not with the implementation details of
-architecture and synchronicity, but with what should be done with the PSO
-equations when many hundreds or thousands of processors are available.  The
-main issues that have been addressed in this space are how many particles to
-use for a particular number of processors and what communication topology
-should be employed.
+The other area of research in parallelizing PSO deals not with the
+implementation details of architecture and synchronicity, but with what should
+be done with the PSO equations when many hundreds or thousands of processors
+are available.  The main issues that have been addressed in this space are how
+many particles to use for a particular number of processors and what
+communication topology should be employed.
 
@@
 only sends information about the best particle it evaluated to other
-processors~\citep{mcnabb-2009-large-particle-swarms}.
+processors~\citep{belal-2004-parallel-models-for-pso}.
 
@@
 With regard to increasing the swarm size in PSO, some recent work has suggested
-that increasing the swarm size through the course of the optimization process
-provides better results than having a set swarm
+that increasing the swarm size throughout the course of the optimization
+process provides better results than having a set swarm
 size~\citep{hsieh-2009-efficient-population-utilization-for-pso,
 montes-de-oca-2010-incremental-social-learning-pso}.  However, these results
-are based on total number of function evaluations, which, when running in
-parallel on expensive functions, is less important than total number of
-iterations.  Other work focusing on parallelization has shown that when extra
-processors are available they should be used, as performance increases with
-swarm size when measuring in terms of number of
-iterations~\citep{mcnabb-2009-large-particle-swarms,
+focused on serial computation and are based on total number of function
+evaluations, which, when running in parallel on expensive functions, is less
+important than total number of iterations.  Other work focusing on
+parallelization has shown that when extra processors are available they should
+be used, as performance increases with swarm size when measuring in terms of
+number of iterations~\citep{mcnabb-2009-large-particle-swarms,
 jin-2005-pso-antenna-designs}.  If the swarm size were varied throughout the
@@
 most iterations.
+
+The contribution of our work is in this area of what should be done with the
+PSO equations to better utilize a thousand processors when they are available.
+In our work we use a synchronous, MapReduce implementation of parallel PSO.
+While we use a specific implementation, we describe how speculative evaluation
+can be performed in any of the synchronous architectures mentioned in the
+previous section.  The adaptation of our methods to asynchronous PSO
+parallelization methods should be straightforward, though it is left to future
+work.
 
@@
 
+Our speculative methods require refactoring the PSO equations such that all
+possible positions for each particle at iteration $t+1$ can be evaluated in
+parallel along with the position of each particle at iteration $t$.  With some
+careful bookkeeping, we can then piece together the results of iteration $t+1$
+for each particle, thus using extra processors to evaluate two iterations of
+the algorithm in the time it takes to evaluate the function once.  As we show
+in Sections~\ref{sec:proof} and~\ref{sec:topology}, a wise choice of topology
+limits the necessary speculative evaluations to seven per particle.
+
+To see the value of this refactoring, suppose that $1000$ processors are
+available, and that the evaluation of the objective function takes one hour.
+If we only want a swarm of $100$ particles, $900$ of the processors would be
+sitting idle for an hour at every iteration, and it would take two hours to run
+two iterations.  If instead we perform speculative evaluation, sending each of
+the $7$ possible speculative positions of a particle to be computed at the same
+time as its current position, we would use $800$ of the $1000$ processors and
+perform two iterations in one hour.
+
+In order to do two iterations at once, we must use 8 times as many processors
+as there are particles in the swarm.  If these processors were not performing
+speculative evaluation, they might instead be used for function evaluations
+needed to support a larger swarm.  This raises the question of whether a swarm
+of 100 particles doing twice as many iterations outperforms a swarm of 800
+particles.  We show in \sec{results} that in many, though not all, instances, a
+smaller swarm performing more iterations does in fact outperform a larger
+swarm.
+
+\sec{proof} shows in detail how the PSO equations can be refactored to allow
+for speculative evaluation, proving that our methods are correct and that they
+exactly reproduce the behavior of the PSO algorithm.  The section also
+introduces some notation used later in the paper.  \sec{topology} gives a brief
+discussion of how the topology used affects the amount of speculative
+computation needed.
+
 \subsection{Refactoring the PSO Equations}
+\label{sec:proof}
 
@@
 
-\subsection{Discussion}
-
-To see the value of this refactoring, suppose that $1000$ processors are
-available, and that the evaluation of the objective function takes one hour.
-If we only want a swarm of $100$ particles, $900$ of the processors would be
-sitting idle for an hour at every iteration, and it would take two hours to run
-two iterations.  If instead we perform speculative evaluation, sending each of
-the $7$ evaluations in $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
-to be computed at the same time as $\ofunc(\ppos_{t})$, we would use $800$ of
-the $1000$ processors and perform two iterations in one hour.
-
-In order to do two iterations at once, we must use 8 times as many processors
-as there are particles in the swarm.  If these processors were not performing
-speculative evaluation, they might instead be used for function evaluations
-needed to support a larger swarm.  This raises the question of whether a swarm
-of 100 particles doing twice as many iterations outperforms a swarm of 800
-particles.  We show in \sec{results} that in many, though not all, instances, a
-smaller swarm performing more iterations does in fact outperform a larger
-swarm.
+\subsection{Topology in Speculative Evaluation}
+\label{sec:topology}
 

### 2011-02-14 masters_thesis/proposal/proposal.tex
Added a first draft of a proposal

--- 2011-01-11 masters_thesis/proposal/proposal.tex
@@
 % out if these already exist.
-\newcommand{\Title}{Improved Corpus Exploration with Topic Modeling}
+\newcommand{\Title}{A Speculative Approach to Parallelization in Particle Swarm
+Optimization}
 \newcommand{\Author}{Matthew Gardner}
-\newcommand{\SubmissionMonth}{January}
+\newcommand{\SubmissionMonth}{February}
 \newcommand{\SubmissionYear}{2011}
@@
                 Submitted \SubmissionMonth~\SubmissionYear, Created \today},%
-    pdfkeywords={topic modeling, corpus browsing, corpus discovery, %
-				text mining},%
+    pdfkeywords={parallel algorithms, optimization methods, particle swarm %
+				optimization, speculative decomposition},%
 }
@@
 \usepackage{graphicx}
+\usepackage{psfrag}
 %\usepackage{pdfsync}
 \usepackage{amsmath}
+
+% Better references, I think
+\newcommand{\secref}[1]{Section~\ref{sec:#1}}
+\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
+\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
+
+%% PSO Stuff
+\DeclareMathOperator*{\argmin}{arg\;min}
+\DeclareMathOperator*{\argmax}{arg\;max}
+\DeclareMathOperator*{\arginf}{arg\;inf}
+\DeclareMathOperator*{\argsup}{arg\;sup}
+\providecommand{\pers}{\ensuremath{P}}
+\providecommand{\neigh}{\ensuremath{N}}
+\providecommand{\leftind}{\ensuremath{L}}
+\providecommand{\rightind}{\ensuremath{R}}
+\providecommand{\nURand}{\ensuremath{U^\neigh}}
+\providecommand{\pURand}{\ensuremath{U^\pers}}
+\providecommand{\ppos}{\ensuremath{\Vec{x}}}
+\providecommand{\fppos}{\ensuremath{\Vec{fx}}}
+\providecommand{\pvel}{\ensuremath{\Vec{v}}}
+\providecommand{\nbest}{\ensuremath{\Vec{b}^\neigh}}
+\providecommand{\fnbest}{\ensuremath{\Vec{fb}^\neigh}}
+\providecommand{\pbest}{\ensuremath{\Vec{b}^\pers}}
+\providecommand{\fpbest}{\ensuremath{\Vec{fb}^\pers}}
+\providecommand{\constriction}{\ensuremath{\chi}}
+\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
+\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
+\providecommand{\obs}{\ensuremath{\Vec{\xi}}}
+\providecommand{\ofunc}{\ensuremath{f}}
+\providecommand{\swarm}{\ensuremath{swarm}}
+
+%SpecExPSO Stuff
+\providecommand{\indic}{\ensuremath{I}}
+\providecommand{\specvel}{\ensuremath{\vec{V}}}
+\providecommand{\specpos}{\ensuremath{\vec{X}}}
+\providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
+\providecommand{\rightn}{\ensuremath{\Vec{x}^\rightind}}
+\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
+\providecommand{\casegen}{\ensuremath{c}}
+\providecommand{\casedef}{\ensuremath{(\pbest,\nbest)}}
+\providecommand{\casexn}{\ensuremath{(S,-)}}
+\providecommand{\casexx}{\ensuremath{(S,S)}}
+\providecommand{\casexl}{\ensuremath{(S,\leftind)}}
+\providecommand{\casexr}{\ensuremath{(S,\rightind)}}
+\providecommand{\casepn}{\ensuremath{(-,-)}}
+\providecommand{\casepl}{\ensuremath{(-,\leftind)}}
+\providecommand{\casepr}{\ensuremath{(-,\rightind)}}
+\providecommand{\casepN}{\ensuremath{(-,N)}}
+\providecommand{\casexN}{\ensuremath{(S,N)}}
+\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
+\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
+\providecommand{\p}{\ensuremath{p}}
+\providecommand{\pset}{\ensuremath{\mathbf{p}}}
+\providecommand{\s}{\ensuremath{s}}
+\providecommand{\sset}{\ensuremath{\mathbf{s}}}
+\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
+\providecommand{\n}{\ensuremath{n}}
+\providecommand{\nset}{\ensuremath{\mathbf{n}}}
+\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}
+
+%Other math
+\providecommand{\prob}{\ensuremath{\mathrm{Pr}}}
 
@@
 
+Particle swarm optimization (PSO) has previously been parallelized primarily by
+distributing the computation corresponding to particles across multiple
+processors.  In these approaches, additional processors can only add more
+particles to the swarm.  However, in many cases this is not efficient when
+scaled to very large swarm sizes (on very large clusters).  Current methods
+cannot answer well the question, ``What do you do with 1000 processors when 50
+or 100 particles is the most efficient swarm size?''  In this paper we attempt
+to answer that question with a speculative approach to the parallelization of
+PSO motivated by speculative execution in modern computer hardware.
+
+In our approach, we refactor PSO such that the computation needed for iteration
+$t+1$ can be done in parallel with the computation needed for iteration $t$.
+Thus we can perform two iterations of PSO at once.  Even with some amount of
+wasted computation, we show that this approach to parallelization in PSO often
+outperforms the na\"ive parallelization of simply adding particles to the
+swarm.  Our methods produce results that are exactly equivalent to PSO; this is
+not a new algorithm or variant, only a new method of parallelization.
+
+However, given this new parallelization model we can relax the requirement of
+exactly reproducing PSO in an attempt to produce better results.  We present
+several such relaxations, including keeping the best speculative position
+evaluated instead of the one corresponding to PSO, and speculating several
+iterations ahead instead of just one.  We show that these methods dramatically
+improve the performance of parallel PSO in many cases.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % 1 to 4 pages
+\label{sec:intro}
+
+Particle swarm optimization (PSO) has been found to be a highly robust and
+effective algorithm for solving many types of optimization
+problems~\citep{poli-2008-pso-applications}.  For much of the algorithm's
+history, PSO was run serially on a single machine.  However, the world's
+computing power is increasingly coming from large clusters of processors.  In
+order to efficiently utilize these resources for computationally intensive
+problems, PSO needs to run in parallel.
+
+Within the last few years, researchers have begun to recognize the need to
+develop parallel implementations of PSO, publishing many papers on the subject.
+The methods they have used include various synchronous algorithms
+\citep{chu-2006-intelligent-parallel-pso,jin-2005-pso-antenna-designs,%
+parsopoulos-2004-parallel-vector-evaluated-pso,%
+schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
+\citep{mostaghim-2006-multi-objective-pso-on-grids,%
+venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
+evaluation of the objective function can also be done in some cases, though
+that is not an adaption of the PSO algorithm itself and thus is not the focus
+of this paper.
+
+These previous parallel techniques distribute the computation needed by the
+particles in the swarm over the available processors.  If more processors are
+available, these techniques increase the number of particles in the swarm,
+either by adding individual particles or by adding entire new sub-swarms.  In
+almost all cases, adding additional particles produces better results in the
+same amount of time~\citep{mcnabb-2009-large-particle-swarms}.  In
+\figref{iters-sphere} we see an example of this on the well-known benchmark
+function Sphere (20 dimensions, reporting the average of twenty runs).  In
+terms of the number of iterations performed (which is equivalent to wall-clock
+time if all particles are evaluated in parallel), every time the swarm size
+increases, the performance improves.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{iters_sphere}
+  \caption{Function Sphere with various swarm sizes, comparing performance with
+  the number of iterations of the algorithm performed.}
+  \label{fig:iters-sphere}
+\end{figure}
+
+However, it can be seen from the graph that once the swarm is sufficiently
+large, doubling the swarm size produces only incremental benefit; performance
+improves, but not by very much.  The reason for this is that there comes a
+point of diminishing returns with respect to adding particles.  In
+\figref{evals-sphere} we show the value obtained after 50,000 function
+evaluations (not iterations) as a function of swarm size, again for the
+function Sphere.  Increasing the swarm size from 5 to 10 has a significant
+effect on the value obtained.  However, increasing the swarm size from 16 to 30
+makes the algorithm less efficient; that is, it reduces the progress the
+algorithm makes per evaluation.  Other functions show similar trends, though
+often the optimal swarm size is slightly larger.  For this reason, previous
+work has recommended the use of a swarm size of 50 for
+PSO~\citep{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
+cases, adding particles indefinitely will not yield an efficient
+implementation.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\columnwidth]{evals_sphere}
+  \caption{Function Sphere with various swarm sizes, comparing performance with
+  the number of function evaluations performed.  Error bars show median and
+  10th and 90th percentiles.}
+  \label{fig:evals-sphere}
+\end{figure}
+
+In this project we consider PSO parallelization strategies for clusters of
+hundreds or thousands of processors and functions for which a single evaluation
+will take at least tens of seconds, but probably minutes or perhaps hours.  Our
+purpose is to explore the question of what to do with a thousand processors
+when 50 or 100 particles is the most efficient swarm size, and simply adding
+particles results in only incremental improvement.  In our work we assume that
+the researcher has some fixed number of processors that should be fully
+utilized throughout the optimization process.
+
+In order to solve this problem, we apply the concept of speculative
+decomposition~\citep{grama-2003-intro-to-parallel-computing} to particle swarm
+optimization, using extra processors to perform two iterations of PSO at the
+same time.  Speculative decomposition is analogous to speculative execution
+(also known as branch prediction), a technique commonly used in processors.
+Modern processors, when faced with a branch on which they must wait (e.g., a
+memory cache miss), guess which way the branch will go and start executing,
+ensuring that any changes can be undone.  If the processor guesses right,
+execution is much farther ahead than if it had idly waited on the memory
+reference.  If it guesses wrong, execution restarts where it would have been
+anyway.  Thus the processor speculates about future paths of execution in an
+attempt to decrease overall processing time.
+
+In this paper we show that the results of standard PSO can be reproduced
+\emph{exactly}, two iterations at a time, using a speculative approach adapted
+from speculative execution. We prove that the standard PSO equation can be
+factored such that a set of speculative positions can be found which will
+always include the position computed in the next iteration.  By computing the
+value of the objective function for each of the speculative positions at the
+same time the algorithm evaluates the objective function for the current
+position, it is possible to know the objective function values for both the
+current and the next iteration at the same time.  We demonstrate this principle
+by implementation and show that it produces exactly the same results as
+standard PSO, but two iterations at a time.  The resulting implementation runs
+efficiently on large clusters where the number of processors is much larger
+than a typical or reasonable number of particles, producing better results in
+less ``wall-clock'' time.
+
+It is important to note here that this is not a variant of PSO.  We simply
+propose a new way to think about the parallelization of PSO that we show
+performs better in many cases than previous parallelizations.
+
+Furthermore, we show that if we relax the requirements of the algorithm, no
+longer demanding that it strictly reproduce the exact behavior of standard PSO,
+we can introduce new speculative techniques that often out-perform na\"ive
+parallelizations of PSO.  These relaxations make better use of the information
+obtained from the extra exploration made by the speculative function
+evaluations.  We also explore the idea that, like branch prediction in
+processors, we need not speculatively evaluate \emph{all} possible future
+positions, we can accelerate the algorithm even if we are just \emph{likely} to
+have guessed right.  By pruning the speculation to just paths that are
+statistically likely to reproduce the paths that are equivalent to PSO we can
+increase the swarm size without increasing the number of speculative
+evaluations.  We also consider several recovery strategies for cases where the
+pruned set of speculative evaluations does not contain the evaluation that
+standard PSO would have done.  A further improvement we explore is speculating
+several iterations ahead instead of just one, which is made possible by pruning
+the number of speculative evaluations.
+
+\subsection{Particle Swarm Optimization}
+\label{sec:pso}
+
+Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
+Eberhart~\citep{kennedy-1995-particle-swarm-optimization}.  The algorithm is
+used to intelligently search a multi-dimensional space by mimicking the
+swarming and flocking behavior of birds and other animals. It is a social
+algorithm that depends on interaction between particles to quickly and
+consistently approximate the optimal solution to a given objective function.
+
+The motion of particles through the search space has three components: an
+inertial component that gives particles momentum as they move, a cognitive
+component where particles remember the best solution they have found and are
+attracted back to that place, and a social component by which particles are
+attracted to the best solution that any of their neighbors have found.
+
+At each iteration of constricted PSO, the position $\ppos_t$ and velocity
+$\pvel_t$ of each particle are updated as follows:
+\begin{align}
+\label{eq:velupdate}
+	\pvel_{t+1} &=
+		\constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) +
+			\ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
+		\bigr] \\
+\label{eq:posupdate}
+	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
+\end{align}
+where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
+numbers drawn from a standard uniform distribution, the \( \otimes \) operator
+is an element-wise vector multiplication, $\pbest$ (called personal best) is
+the best position the current particle has seen, and $\nbest$ (called
+neighborhood best) is the best position the neighbors of the current particle
+have seen~\citep{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
+\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
+required to ensure convergence (2.05, 2.05, and .73,
+respectively)~\citep{clerc-2002-constricted-pso}.
+
+Changing the way neighbors are defined, usually called the ``topology,'' has a
+significant effect on the performance of the algorithm.  In the Ring topology,
+each particle has one neighbor to either side of it; in the Complete
+topology\footnote{The Complete topology has often been unfortunately named Star
+in the literature, which in graph theory refers to a completely different
+topology.  Other names have also been used, including ``global topology'' and
+gbest.  We use the graph theory term ``Complete'' in this paper.}, every
+particle is a neighbor to every other
+particle~\citep{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
+particle is also a neighbor to itself in that its own position and value are
+considered when updating the particle's neighborhood best, $\nbest$.  Thus with
+$p$ particles, using the Ring topology each particle with index $i$ has three
+neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
+particle has $p$ neighbors.
+
+In this paper we use these topologies as well as a parallel adaptation of the
+Complete topology, called Random, that has been shown to approximate the
+behavior of Complete with far less
+communication~\citep{mcnabb-2009-large-particle-swarms}.  In the Random
+topology, each particle randomly picks two other particles to share information
+with at each iteration, along with itself.  Thus in both the Ring and the
+Random topologies, all particles have three neighbors.
+
+The ideal topology and swarm size for PSO depend on the objective function.
+Researchers have devised various benchmark functions and have found that the
+ideal topology for one function may perform very poorly for another function.
+The No Free Lunch Theorems for Optimization show that this is true in
+general---if an algorithm performs well on average for one class of functions
+then it must do poorly on average for other
+problems~\citep{wolpert-1997-nfl-for-optimization}.
+
+An attempt to standardize PSO found that although a Complete swarm of 50
+particles converged to the global optimum more quickly for many benchmark
+functions than the same swarm size with a Ring topology, it was also more
+likely to permaturely converge to local optima for other functions.  The study
+found no significant improvement for any other swarm size between 20 and 100
+and concluded with a recommendation to use a swarm of 50 particles with a Ring
+topology as a starting point.  The authors acknowledged that choosing the ideal
+topology requires thorough experimentation for the particular
+problem~\citep{bratton-2007-defining-a-standard-for-pso}.  Other authors have
+explained that a large swarm with a sparse topology propogates information
+slowly~\citep{montes-de-oca-2009-frankensteins-pso}.
 
@@
 
+There have been several parallel implementations of PSO presented in the
+literature.  These implementations vary along two axes---what computation is
+performed and how it is done.  How the computation is performed involves both
+architectural decisions, such as a master-slave versus a fully distributed
+architecture, and decisions about synchronicity.  What computation is performed
+involves decisions about how to scale the PSO algorithm for parallel
+implementations, such as the ratio of particles to processors or the topology
+to use.
+
+\subsection{Synchronous vs. Asynchronous PSO}
+
+Synchronous parallel implementations of PSO reproduce the standard serial
+algorithm exactly.  This approach was first described analytically by
+\citet{belal-2004-parallel-models-for-pso} and first implemented by
+\citet{schutte-2004-parallel-global-optimization-with-pso}.  In a typical
+master-slave algorithm, the master assigns tasks to slave processors, and in
+parallel PSO, each task consists primarily of a function evaluation.  Updating
+the particle's position and value may also be included in the
+task~\citep{belal-2004-parallel-models-for-pso}, or this work may be performed
+in serial by the
+master~\citep{schutte-2004-parallel-global-optimization-with-pso}.  Before
+proceeding to the next iteration, particles communicate, and each particle
+updates its neighborhood best.  Whether this communication step happens
+sequentially on the master or in parallel, each particle must receive
+communication from its neighbors before proceeding.  The benefits of the
+synchronous PSO include its simplicity, repeatability, and comparability with
+standard PSO, which may be essential in research applications.
+
+Asynchronous parallel
+PSO~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
+koh-2006-parallel-asynchronous-pso} is a modification to the standard algorithm
+which removes the synchronization point at the end of each iteration.  Instead,
+particles iterate independently and communicate asynchronously.  In a typical
+master-slave implementation of asynchronous parallel PSO, the master updates
+each particle's personal best, neighborhood best, velocity, and position
+immediately after receiving the function value from the slave processor.  Since
+this update occurs while other particles are still being evaluated, it may use
+information from the previous iteration for some
+neighbors.\footnote{Asynchronous parallel PSO has been compared to the
+``asynchronous updates'' variant of serial
+PSO~\citep{koh-2006-parallel-asynchronous-pso}.  However, serial PSO with
+asynchronous updates differs from standard PSO in that particles use newer
+information, but asynchronous parallel PSO differs from standard PSO in that
+particles use older information.} In a partially asynchronous implementation,
+particles might wait for some but not all neighbors to complete before
+proceeding~%
+\citep{scriven-2008-asynchronous-pso-in-unreliable-distributed-environments}.
+In some master-slave implementations, particles never get more than one
+iteration ahead of
+others~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
+koh-2006-parallel-asynchronous-pso}.  However, in a fully distributed
+implementation, particles might never wait for information, and one particle
+could complete many more iterations than another
+particle~\citep{scriven-2008-distributed-pso-using-peer-to-peer}.  The main
+effect of asynchronous evaluation is that processors spend less time
+idle---this trait is particularly valuable when processors are heterogeneous or
+function evaluation times are
+varied~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
+koh-2006-parallel-asynchronous-pso}.  Asynchronous parallel PSO behaves
+differently than the standard algorithm and may even produce different results
+between runs.  Most reports conclude that asynchronous communication produces
+similar numerical results to the standard algorithm, but the question has not
+yet been thoroughly
+addressed~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
+koh-2006-parallel-asynchronous-pso}.
+
+\subsection{Scaling PSO to many processors}
+
+The contribution of this work deals not with the implementation details of
+architecture and synchronicity, but with what should be done with the PSO
+equations when many hundreds or thousands of processors are available.  The
+main issues that have been addressed in this space are how many particles to
+use for a particular number of processors and what communication topology
+should be employed.
+
+The number of particles per processor has typically been decided by how long it
+takes to evaluate the function being optimized.  When the function takes longer
+than a few seconds to evaluate, previous techniques have assigned the number of
+particles in the swarm to be the number of processors available~%
+\citep{jin-2005-pso-antenna-designs,mcnabb-2009-large-particle-swarms},
+advocating using as many processors as possible to get the best performance.
+When the function takes less time to evaluate than the TCP/IP stack used to
+send interprocessor communication, parallel implementations assign several or
+many particles to a single processor~\citep{chu-2006-intelligent-parallel-pso,
+chang-2005-parallel-pso-with-communication-strategies}.  Often the processor
+only sends information about the best particle it evaluated to other
+processors~\citep{mcnabb-2009-large-particle-swarms}.
+
+Another popular method is simply to run PSO independently on each of the
+processors available, taking the best result when all of the runs complete.  It
+should be noted that this is equivalent to the previously stated method of
+assigning many particles to each processor, only with no communication between
+processors instead of little communication.  Both of these methods can be
+described as changes in the communication topology of the original PSO
+algorithm~\citep{mcnabb-2009-large-particle-swarms}.
+
+Thus previous work in parallelizing PSO, apart from implementation details, has
+consisted entirely of increasing the swarm size and adapting the topology to be
+better suited to parallel computation.
+
+With regard to increasing the swarm size in PSO, some recent work has suggested
+that increasing the swarm size through the course of the optimization process
+provides better results than having a set swarm
+size~\citep{hsieh-2009-efficient-population-utilization-for-pso,
+montes-de-oca-2010-incremental-social-learning-pso}.  However, these results
+are based on total number of function evaluations, which, when running in
+parallel on expensive functions, is less important than total number of
+iterations.  Other work focusing on parallelization has shown that when extra
+processors are available they should be used, as performance increases with
+swarm size when measuring in terms of number of
+iterations~\citep{mcnabb-2009-large-particle-swarms,
+jin-2005-pso-antenna-designs}.  If the swarm size were varied throughout the
+course of the optimization process, some processors would be sitting idle at
+most iterations.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % 1 to 2 sentences
+
+When many processors are available, a speculative approach to the
+parallelization of PSO can yield dramatic improvements in the performance of
+the algorithm on expensive functions compared to previous parallelization
+techniques.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+PSO can be trivially parallelized by assigning each particle's computation to
+an individual processor.  But as we have seen in \figref{evals-sphere}, for
+some functions, and for large numbers of processors, just adding particles
+reaches a point of diminishing returns.  That is, adding processors does not
+help the algorithm reach any given level of fitness appreciably faster.
+Instead of adding particles we employ a speculative approach that allows us to
+perform two iterations at a time.
+
+\subsection{Refactoring the PSO Equations}
+
+To perform two iterations at a time we must first refactor PSO such that the
+determination of the value of the objective function is separate from the rest
+of the computation.  For simplicity, this discussion will describe the case
+where PSO is performing function minimization using the Ring topology.  In this
+example, each particle has two neighbors, the ``right neighbor'' and ``left
+neighbor,'' whose positions are represented as $\rightn$ and $\leftn$
+respectively.  Though we will only describe the case of the Ring topology, this
+method can easily be extended to arbitrary topologies.
+
+The refactoring hinges on the idea that once the random coefficients
+$\pURand_{t}$ and $\nURand_{t}$ are drawn, there are only a few possible new
+positions, or updates, for $\nbest$ and $\pbest$.  For the Ring topology there
+are 7 possible update cases, identified in Table~\ref{tab:evals}.  We label
+each case with an identifier referring to the source of the update: a minus
+sign ($-$) represents no update, $L$ represents an update to $\nbest$ coming
+from the left neighbor, $R$ represents an update to $\nbest$ coming from the
+right neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$
+coming from the particle itself.  As an example, $\casexn$ refers to the case
+that the particle finds a new personal best, but neither it nor its neighbors
+find a position that updated its neighborhood best.  In the equations that
+follow, we refer to an unspecified update case as $\casegen$, and to the set of
+cases collectively as $\caseset$.
+
+\begin{table}
+  \caption{All possible updates for a particle with two neighbors}
+  \label{tab:evals}
+  \centering
+  \begin{tabular}{lcc}
+	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
+	\hline
+	\hline
+	$\casepn$&No update&No update\\
+	\hline
+	$\casepl$&No update&Left Neighbor\\
+	\hline
+	$\casepr$&No update&Right Neighbor\\
+	\hline
+	$\casexn$&Self&No update\\
+	\hline
+	$\casexl$&Self&Left Neighbor\\
+	\hline
+	$\casexr$&Self&Right Neighbor\\
+	\hline
+	$\casexx$&Self&Self\\
+	\hline
+  \end{tabular}
+\end{table}
+
+In order to incorporate the determination of which case occurs into the
+position and velocity update equations, we introduce an indicator function
+$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
+$\casegen$ corresponds to the case actually taken by PSO,
+$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
+then sum over all of the cases, and the indicator function will make all of the
+terms drop to zero except for the case that actually occurs.  For example, the
+indicator function for the specific case $\casexn$ (which, as is shown in
+Table~\ref{tab:evals}, means that the particle's personal best was updated, but
+its neighborhood best was not) can be written as follows:
+
+\begin{align}
+  \nonumber
+	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
+	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
+  \label{eq:deficasexn}
+	&\begin{cases}
+	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
+	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
+	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
+	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
+	   0 & \text{otherwise}
+	\end{cases}
+\end{align}
+
+For each case $\casegen \in \caseset$, there is also a corresponding
+velocity update function $\specvel_{t+1}^{\casegen}$.  When the case is
+known, the specific values of $\pbest_t$ and $\nbest_t$ may be substituted
+directly into~\eqref{eq:velupdate}.  For example, in case $\casexn$,
+$\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
+position, and $\nbest_{t}=\nbest_{t-1}$, as $\nbest$ was not updated at
+iteration $t$:
+\begin{align}
+\nonumber
+	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
+	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
+\label{eq:defvcasexn}
+		&= \constriction \bigl[ \pvel_{t} +
+			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t})
+			+ \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
+			\ppos_{t}) \bigr]
+\end{align}
+
+In the same way we can create notation for the position update function by
+substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
+parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
+subscripts.
+\begin{align}
+\label{eq:defpcasegen}
+	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
+	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t})
+	= \ppos_{t} + \specvel_{t+1}^{\casegen}
+\end{align}
+
+With this notation we can re-write the original PSO velocity
+equation~\eqref{eq:velupdate}, introducing our sum over cases with the
+indicator functions.  Again, we represent the indicator functions and velocity
+functions without the parameters for compactness.  The equation becomes:
+\begin{align}
+\nonumber
+	\pvel_{t+1} &=
+		\constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
+			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
+			\ppos_{t}) \bigr] \\
+\nonumber
+	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
+			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
+			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
+			\ppos_{t}) \bigr]  \\
+\label{eq:vel2update}
+	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
+\end{align}
+
+Similarly, the position update equation~\eqref{eq:posupdate} becomes:
+\begin{align}
+\label{eq:pos2update}
+	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
+	= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
+\end{align}
+
+The value of the objective function at $\ppos_{t+1}$ is given by:
+\begin{align}
+\label{eq:val2update}
+	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \indic_{t+1}^{c}
+	\ \ofunc(\specpos_{t+1}^{c})
+\end{align}
+
+Returning our attention to the computation of $\ppos_{t+1}$ in
+\eqref{eq:pos2update} and writing it with the parameters which were omitted
+above, we obtain:
+\begin{align}
+\nonumber
+  \ppos_{t+1} = \sum_{c \in \caseset}
+	&\indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
+	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
+\label{eq:val2updatelong}
+	& \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
+	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t})
+\end{align}
+
+In this form the important point to notice is that there are only $7$ values
+(for this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
+\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
+objective function evaluation at iteration $t$.  Note also that while there are
+random numbers in the equation, they are assumed fixed once drawn for any
+particular particle at a specific iteration.  Thus PSO has been refactored such
+that the algorithm can begin computing all $7$ of the objective function
+evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
+computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
+only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
+hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.
+
+Although this speculative approach computes $\ofunc(\specpos_{t+1}^{\casegen})$
+for all $\casegen \in \caseset$, even those for which $\indic_{t+1}^{\casegen}
+= 0$, these extra computations will be ignored, and might just as well never
+have been computed.  We call the set of computations
+$\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$ ``speculative
+children'' because only one of them will be needed.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+This project is expected to produce an implementation of speculative evaluation
+in PSO, including several variations on the speculative evaluations performed.
+The project will also show that speculative approaches to parallelization in
+PSO result in improved performance over non-speculative approaches.
+
+To show improved performance, we will test our methods against previous
+parallelization methods on several benchmark functions.  While benchmark
+functions themselves take fractions of a second to evaluate and thus have no
+need of the kind of parallelization we explore in this project, they also have
+no need of particle swarm optimization at all, as they can be solved
+analytically.  Yet they are useful to us and to optimization researchers
+generally because they stand as surrogates for the kinds of functions
+practitioners are actually interested in, and they allow us to explore the
+behavior of optimization algorithms on several different classes of functions
+in an easy and standardized way.
+
+We will measure our performance on these benchmark functions in terms of the
+best value found at each iteration of the algorithms.  We will also compare the
+total number of iterations required by each algorithm to reach some specified
+value.  The project will be successful if a statistically significant decrease
+in the number of iterations is observed.
 
@@
 
-\item Completed: reformulate PSO in MapReduce.
+\item Completed: Implement speculative evaluation in PSO and several
+  improvements on it.
 
-\item February 6, 2009: evaluate PSO with large swarms with a few benchmark
-functions.
+\item March 7, 2011: Finish running experiments and incorporate the results
+  into the thesis.
 
-\item February 20, 2009: evaluate PSO with very large swarms with subswarms
-and with more benchmark functions.
+\item March 15, 2011: Give final draft of thesis to advisor for review.
 
-\item March 9, 2009: rework individual papers into a coherent thesis and
-submit draft to committee chair.
+\item March 29, 2011: submit thesis to second committee member for review.
 
-\item March 20, 2009: submit thesis to second committee member for review.
+\item April 5, 2011: submit thesis to third committee member for review.
 
-\item March 30, 2009: submit thesis to third committee member for review.
-
-\item April 10, 2009: defend thesis.
-
-Each of the three stages of the project can be submitted as a paper to a
-conference such as GECCO or CEC.
+\item April 19, 2011: defend thesis.
 
@@
 
-\bibliography{}
+\bibliography{../../../bib/pso/general/bib,%
+../../../bib/pso/parallel/bib,%
+../../../bib/pso/topology/bib,%
+../../../bib/pso/fips/bib,%
+../../../bib/aml/bib,%
+../../../bib/parallel/bib,%
+../../../bib/nfl/bib%
+}
 

### 2011-03-08 masters_thesis/proposal/proposal.tex
Cuts to the proposal from Dr. Seppi and additions from Dr. Ventura

@@
 \committeechair{Kevin Seppi}
-\committeemembera{Eric Ringger}
+\committeemembera{Dan Ventura}
 \committeememberb{David Embley}
@@
 
-Particle swarm optimization (PSO) has previously been parallelized primarily by
-distributing the computation corresponding to particles across multiple
-processors.  In these approaches, additional processors can only add more
-particles to the swarm.  However, in many cases this is not efficient when
-scaled to very large swarm sizes (on very large clusters).  Current methods
-cannot answer well the question, ``What do you do with 1000 processors when 50
-or 100 particles is the most efficient swarm size?''  In this paper we attempt
-to answer that question with a speculative approach to the parallelization of
-PSO motivated by speculative execution in modern computer hardware.
+Particle swarm optimization (PSO) is an evolutionary algorithm that has proven
+effective at solving complicated optimization problems.  Since its inception,
+however, PSO has been run primarily in serial on a single machine.  With the
+world's computing resources increasingly coming from large clusters of
+processors, the parallelization of the algorithm is important to its continued
+relevance.  PSO has previously been parallelized primarily by distributing the
+computation corresponding to particles across multiple processors.  In these
+approaches, additional processors can only add more particles to the swarm.
+However, in many cases this is not efficient when scaled to very large swarm
+sizes (on very large clusters).  Current methods cannot answer well the
+question, ``What do you do with 1000 processors when 50 or 100 particles is the
+most efficient swarm size?''  In this project we attempt to answer that
+question with a speculative approach to the parallelization of PSO motivated by
+speculative execution in modern computer hardware.
 
@@
 that is not an adaption of the PSO algorithm itself and thus is not the focus
-of this paper.
+of this project.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{iters_sphere}
+  \includegraphics[width=.4\columnwidth]{iters_sphere}
   \caption{Function Sphere with various swarm sizes, comparing performance with
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{evals_sphere}
+  \includegraphics[width=.4\columnwidth]{evals_sphere}
   \caption{Function Sphere with various swarm sizes, comparing performance with
@@
 
-In this project we consider PSO parallelization strategies for clusters of
+In this project we will consider PSO parallelization strategies for clusters of
 hundreds or thousands of processors and functions for which a single evaluation
@@
 
-In order to solve this problem, we apply the concept of speculative
-decomposition~\citep{grama-2003-intro-to-parallel-computing} to particle swarm
-optimization, using extra processors to perform two iterations of PSO at the
-same time.  Speculative decomposition is analogous to speculative execution
-(also known as branch prediction), a technique commonly used in processors.
-Modern processors, when faced with a branch on which they must wait (e.g., a
-memory cache miss), guess which way the branch will go and start executing,
-ensuring that any changes can be undone.  If the processor guesses right,
-execution is much farther ahead than if it had idly waited on the memory
-reference.  If it guesses wrong, execution restarts where it would have been
-anyway.  Thus the processor speculates about future paths of execution in an
-attempt to decrease overall processing time.
-
-In this paper we show that the results of standard PSO can be reproduced
+In this project we will show that the results of standard PSO can be reproduced
 \emph{exactly}, two iterations at a time, using a speculative approach adapted
@@
 position, it is possible to know the objective function values for both the
-current and the next iteration at the same time.  We demonstrate this principle
-by implementation and show that it produces exactly the same results as
-standard PSO, but two iterations at a time.  The resulting implementation runs
-efficiently on large clusters where the number of processors is much larger
-than a typical or reasonable number of particles, producing better results in
-less ``wall-clock'' time.
+current and the next iteration at the same time.  We will demonstrate this
+principle by implementation and show that it produces exactly the same results
+as standard PSO, but two iterations at a time.  The resulting implementation
+should run efficiently on large clusters where the number of processors is much
+larger than a typical or reasonable number of particles, producing better
+results in less ``wall-clock'' time.
 
@@
 
-Changing the way neighbors are defined, usually called the ``topology,'' has a
-significant effect on the performance of the algorithm.  In the Ring topology,
-each particle has one neighbor to either side of it; in the Complete
-topology\footnote{The Complete topology has often been unfortunately named Star
-in the literature, which in graph theory refers to a completely different
-topology.  Other names have also been used, including ``global topology'' and
-gbest.  We use the graph theory term ``Complete'' in this paper.}, every
-particle is a neighbor to every other
-particle~\citep{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
-particle is also a neighbor to itself in that its own position and value are
-considered when updating the particle's neighborhood best, $\nbest$.  Thus with
-$p$ particles, using the Ring topology each particle with index $i$ has three
-neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
-particle has $p$ neighbors.
-
-In this paper we use these topologies as well as a parallel adaptation of the
-Complete topology, called Random, that has been shown to approximate the
-behavior of Complete with far less
-communication~\citep{mcnabb-2009-large-particle-swarms}.  In the Random
-topology, each particle randomly picks two other particles to share information
-with at each iteration, along with itself.  Thus in both the Ring and the
-Random topologies, all particles have three neighbors.
-
-The ideal topology and swarm size for PSO depend on the objective function.
-Researchers have devised various benchmark functions and have found that the
-ideal topology for one function may perform very poorly for another function.
-The No Free Lunch Theorems for Optimization show that this is true in
-general---if an algorithm performs well on average for one class of functions
-then it must do poorly on average for other
-problems~\citep{wolpert-1997-nfl-for-optimization}.
-
-An attempt to standardize PSO found that although a Complete swarm of 50
-particles converged to the global optimum more quickly for many benchmark
-functions than the same swarm size with a Ring topology, it was also more
-likely to permaturely converge to local optima for other functions.  The study
-found no significant improvement for any other swarm size between 20 and 100
-and concluded with a recommendation to use a swarm of 50 particles with a Ring
-topology as a starting point.  The authors acknowledged that choosing the ideal
-topology requires thorough experimentation for the particular
-problem~\citep{bratton-2007-defining-a-standard-for-pso}.  Other authors have
-explained that a large swarm with a sparse topology propogates information
-slowly~\citep{montes-de-oca-2009-frankensteins-pso}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Related Work}
-% 1 to 2 pages
+\label{sec:related}
 
 There have been several parallel implementations of PSO presented in the
-literature.  These implementations vary along two axes---what computation is
-performed and how it is done.  How the computation is performed involves both
-architectural decisions, such as a master-slave versus a fully distributed
-architecture, and decisions about synchronicity.  What computation is performed
-involves decisions about how to scale the PSO algorithm for parallel
-implementations, such as the ratio of particles to processors or the topology
-to use.
+literature.  The improvements described in the literature come in two major
+areas: innovations in implementation details and innovations in the use of
+topology and swarm size to scale PSO to many processors.
 
-\subsection{Synchronous vs. Asynchronous PSO}
-
-Synchronous parallel implementations of PSO reproduce the standard serial
-algorithm exactly.  This approach was first described analytically by
-\citet{belal-2004-parallel-models-for-pso} and first implemented by
-\citet{schutte-2004-parallel-global-optimization-with-pso}.  In a typical
-master-slave algorithm, the master assigns tasks to slave processors, and in
-parallel PSO, each task consists primarily of a function evaluation.  Updating
-the particle's position and value may also be included in the
-task~\citep{belal-2004-parallel-models-for-pso}, or this work may be performed
-in serial by the
-master~\citep{schutte-2004-parallel-global-optimization-with-pso}.  Before
-proceeding to the next iteration, particles communicate, and each particle
-updates its neighborhood best.  Whether this communication step happens
-sequentially on the master or in parallel, each particle must receive
-communication from its neighbors before proceeding.  The benefits of the
-synchronous PSO include its simplicity, repeatability, and comparability with
-standard PSO, which may be essential in research applications.
-
-Asynchronous parallel
-PSO~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
-koh-2006-parallel-asynchronous-pso} is a modification to the standard algorithm
-which removes the synchronization point at the end of each iteration.  Instead,
-particles iterate independently and communicate asynchronously.  In a typical
-master-slave implementation of asynchronous parallel PSO, the master updates
-each particle's personal best, neighborhood best, velocity, and position
-immediately after receiving the function value from the slave processor.  Since
-this update occurs while other particles are still being evaluated, it may use
-information from the previous iteration for some
-neighbors.\footnote{Asynchronous parallel PSO has been compared to the
-``asynchronous updates'' variant of serial
-PSO~\citep{koh-2006-parallel-asynchronous-pso}.  However, serial PSO with
-asynchronous updates differs from standard PSO in that particles use newer
-information, but asynchronous parallel PSO differs from standard PSO in that
-particles use older information.} In a partially asynchronous implementation,
-particles might wait for some but not all neighbors to complete before
-proceeding~%
-\citep{scriven-2008-asynchronous-pso-in-unreliable-distributed-environments}.
-In some master-slave implementations, particles never get more than one
-iteration ahead of
-others~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
-koh-2006-parallel-asynchronous-pso}.  However, in a fully distributed
-implementation, particles might never wait for information, and one particle
-could complete many more iterations than another
-particle~\citep{scriven-2008-distributed-pso-using-peer-to-peer}.  The main
-effect of asynchronous evaluation is that processors spend less time
-idle---this trait is particularly valuable when processors are heterogeneous or
-function evaluation times are
-varied~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
-koh-2006-parallel-asynchronous-pso}.  Asynchronous parallel PSO behaves
-differently than the standard algorithm and may even produce different results
-between runs.  Most reports conclude that asynchronous communication produces
-similar numerical results to the standard algorithm, but the question has not
-yet been thoroughly
-addressed~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
-koh-2006-parallel-asynchronous-pso}.
+There are many ways to parallelize the basic PSO algorithm.  The most
+fundamental decision to make in parallel PSO is which parallel architecture to
+use.  Several architectures have been proposed, including Master-Slave, fully
+distributed (sometimes called ``diffusion''), and reformulating PSO into
+Google's MapReduce framework~\citep{belal-2004-parallel-models-for-pso,
+mcnabb-2007-parallel-pso-using-mapreduce}.  A somewhat orthogonal
+implementation decision when parallelizing PSO is whether to have synchronous
+communication or asynchronous communication.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The contribution of our work is in this area of what should be done with the
+PSO equations to better utilize a thousand processors when they are available.
+In our work we use a synchronous, MapReduce implementation of parallel PSO.
+While we use a specific implementation, we describe how speculative evaluation
+can be performed in any of the synchronous architectures mentioned in the
+previous section.  The adaptation of our methods to asynchronous PSO
+parallelization methods should be straightforward, though it is left to future
+work.
 
@@
 
+Our speculative methods require refactoring the PSO equations such that all
+possible positions for each particle at iteration $t+1$ can be evaluated in
+parallel along with the position of each particle at iteration $t$.  With some
+careful bookkeeping, we can then piece together the results of iteration $t+1$
+for each particle, thus using extra processors to evaluate two iterations of
+the algorithm in the time it takes to evaluate the function once.  A wise
+choice of topology limits the necessary speculative evaluations to seven per
+particle.
+
+To see the value of this refactoring, suppose that $1000$ processors are
+available, and that the evaluation of the objective function takes one hour.
+If we only want a swarm of $100$ particles, $900$ of the processors would be
+sitting idle for an hour at every iteration, and it would take two hours to run
+two iterations.  If instead we perform speculative evaluation, sending each of
+the $7$ possible speculative positions of a particle to be computed at the same
+time as its current position, we would use $800$ of the $1000$ processors and
+perform two iterations in one hour.
+
+In order to do two iterations at once, we must use 8 times as many processors
+as there are particles in the swarm.  If these processors were not performing
+speculative evaluation, they might instead be used for function evaluations
+needed to support a larger swarm.  This raises the question of whether a swarm
+of 100 particles doing twice as many iterations outperforms a swarm of 800
+particles.  This project will seek to answer that question.
+
+My honors thesis proposed and developed this basic approach to parallelization
+in PSO.  This masters thesis will complete the picture of speculative methods
+in particle swarm optimization.  First, we present a mathematical proof of
+correctness of our methods, given in Section~\ref{sec:proof} as demonstration
+that what we propose is in fact mathematically equivalent to standard
+parallelization methods in PSO.  Second, we propose pruning the set of possible
+speculative evaluations to better use available processors.  We motivate the
+pruning by analyzing the statistics of which branches are commonly taken in
+PSO, using those statistics to prune all but the most likely branches.
+Further, once the idea of pruning is established, we view the set of all
+possible speculative evaulations as an infinite tree, not confining ourselves
+to speculating only about the next iteration.  We then experiment with
+speculating down likely paths three or four iterations ahead, and show that
+this by far outperforms simple one-iteration speculative evaluation.  And
+lastly, as my honors thesis used only a few experiments on three benchmark
+functions, we give a better picture of the performance of speculative
+approaches with much more extensive experimentation.
+
 \subsection{Refactoring the PSO Equations}
+\label{sec:proof}
 
@@
 respectively.  Though we will only describe the case of the Ring topology, this
-method can easily be extended to arbitrary topologies.
+method can easily be extended to arbitrary topologies (and in fact our
+implementation of this algorithm works with arbitrary topologies; we simply
+only discuss the Ring case here as that greatly simplifies our notation).
 
@@
 
+For each function, we will use the topology that is generally considered to be
+best for optimizing that particular function.  When the Complete topology is
+considered best, we will instead use a Random topology that we developed that
+mimicks the behavior of Complete while being more suited for parallel
+computation~\cite{mcnabb-2009-large-particle-swarms}.
+
+The functions we will use in our comparisons are these common benchmark
+functions: Bohachevsky, Griewank, Rastrigin, Sphere, and Schwefel 2.21.  We
+will run each function with 20, 50, and 500 dimensions to test the scalability
+of our methods up to higher dimensions.  We will also test our methods on a
+more typical ``real-world'' problem, that of training a computationally
+intensive machine learning algorithm with a large amount of data.  We will
+compare our speculative methods to previous parallelization techniques,
+including a naive parallelization of just increasing the swarm size, and a
+subswarms implementation that uses several smaller, independent swarms.
+
 \section{Thesis Schedule}
@@
 
-\item March 7, 2011: Finish running experiments and incorporate the results
+\item March 11, 2011: Finish running experiments and incorporate the results
   into the thesis.
@@
 
-\item March 29, 2011: submit thesis to second committee member for review.
+\item March 22, 2011: submit thesis to second committee member for review.
 

### 2011-03-24 masters_thesis/proposal/proposal.tex
Finished my proposal and got it approved by Ventura

@@
 in particle swarm optimization.  First, we present a mathematical proof of
-correctness of our methods, given in Section~\ref{sec:proof} as demonstration
-that what we propose is in fact mathematically equivalent to standard
-parallelization methods in PSO.  Second, we propose pruning the set of possible
-speculative evaluations to better use available processors.  We motivate the
-pruning by analyzing the statistics of which branches are commonly taken in
-PSO, using those statistics to prune all but the most likely branches.
-Further, once the idea of pruning is established, we view the set of all
-possible speculative evaulations as an infinite tree, not confining ourselves
-to speculating only about the next iteration.  We then experiment with
-speculating down likely paths three or four iterations ahead, and show that
-this by far outperforms simple one-iteration speculative evaluation.  And
-lastly, as my honors thesis used only a few experiments on three benchmark
-functions, we give a better picture of the performance of speculative
-approaches with much more extensive experimentation.
+correctness of our methods as demonstration that what we propose is in fact
+mathematically equivalent to standard parallelization methods in PSO.  Second,
+we propose pruning the set of possible speculative evaluations to better use
+available processors.  We motivate the pruning by analyzing the statistics of
+which branches are commonly taken in PSO, using those statistics to prune all
+but the most likely branches.  Further, once the idea of pruning is
+established, we view the set of all possible speculative evaluations as an
+infinite tree, not confining ourselves to speculating only about the next
+iteration.  We then experiment with speculating down likely paths three or four
+iterations ahead, and show that this by far outperforms simple one-iteration
+speculative evaluation.  And lastly, as my honors thesis used only a few
+experiments on three benchmark functions, we give a better picture of the
+performance of speculative approaches with much more extensive experimentation.
 
-\subsection{Refactoring the PSO Equations}
-\label{sec:proof}
+\subsection{Proof}
 
-To perform two iterations at a time we must first refactor PSO such that the
-determination of the value of the objective function is separate from the rest
-of the computation.  For simplicity, this discussion will describe the case
-where PSO is performing function minimization using the Ring topology.  In this
-example, each particle has two neighbors, the ``right neighbor'' and ``left
-neighbor,'' whose positions are represented as $\rightn$ and $\leftn$
-respectively.  Though we will only describe the case of the Ring topology, this
-method can easily be extended to arbitrary topologies (and in fact our
-implementation of this algorithm works with arbitrary topologies; we simply
-only discuss the Ring case here as that greatly simplifies our notation).
+Some readers of my honors thesis had a hard time understanding our methods,
+thinking that we were changing the PSO algorithm when performing speculative
+evaluation.  To alleviate any such concerns and provide a strong theoretical
+background for speculative approaches to parallelization in PSO, we will
+provide a mathematical proof that PSO can be decomposed in such a way as to
+exactly reproduce the behavior of the original algorithm.
 
-The refactoring hinges on the idea that once the random coefficients
-$\pURand_{t}$ and $\nURand_{t}$ are drawn, there are only a few possible new
-positions, or updates, for $\nbest$ and $\pbest$.  For the Ring topology there
-are 7 possible update cases, identified in Table~\ref{tab:evals}.  We label
-each case with an identifier referring to the source of the update: a minus
-sign ($-$) represents no update, $L$ represents an update to $\nbest$ coming
-from the left neighbor, $R$ represents an update to $\nbest$ coming from the
-right neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$
-coming from the particle itself.  As an example, $\casexn$ refers to the case
-that the particle finds a new personal best, but neither it nor its neighbors
-find a position that updated its neighborhood best.  In the equations that
-follow, we refer to an unspecified update case as $\casegen$, and to the set of
-cases collectively as $\caseset$.
+\subsection{Statistically Motivated Pruning}
 
-\begin{table}
-  \caption{All possible updates for a particle with two neighbors}
-  \label{tab:evals}
-  \centering
-  \begin{tabular}{lcc}
-	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
-	\hline
-	\hline
-	$\casepn$&No update&No update\\
-	\hline
-	$\casepl$&No update&Left Neighbor\\
-	\hline
-	$\casepr$&No update&Right Neighbor\\
-	\hline
-	$\casexn$&Self&No update\\
-	\hline
-	$\casexl$&Self&Left Neighbor\\
-	\hline
-	$\casexr$&Self&Right Neighbor\\
-	\hline
-	$\casexx$&Self&Self\\
-	\hline
-  \end{tabular}
-\end{table}
+The key idea in this proposal is that of decreasing the number of speculative
+evaluations needing to be done by pruning away insignificant branches from the
+set of all possible evaluations.  In my honors thesis I found that exactly
+reproducing PSO two iterations at a time required too many extra evaluations to
+give competitive results in most cases.  To fix this problem, I will find ways
+to still speculate about future positions without actually evaluating all of
+the possible options.  This should improve the performance of the algorithm.
 
-In order to incorporate the determination of which case occurs into the
-position and velocity update equations, we introduce an indicator function
-$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
-$\casegen$ corresponds to the case actually taken by PSO,
-$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
-then sum over all of the cases, and the indicator function will make all of the
-terms drop to zero except for the case that actually occurs.  For example, the
-indicator function for the specific case $\casexn$ (which, as is shown in
-Table~\ref{tab:evals}, means that the particle's personal best was updated, but
-its neighborhood best was not) can be written as follows:
+Pruning the set of speculative evaluations necessitates a mechanism for
+deciding which branches are desirable.  A na\"ive approach to pruning would be
+to try the last branch taken by each particle.  Preliminary experimentation
+with this method has shown it to be ineffective.  A smarter approach would be
+to examine the statistical behavior of the branches that PSO takes and use that
+analysis as motivation for deciding which branches to prune.  We will collect
+statistics on the branching behavior of PSO across several functions and
+topologies.  We will then use those statistics to perform intelligent pruning
+of the set of speculative evaluations.
 
-\begin{align}
-  \nonumber
-	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
-	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
-  \label{eq:deficasexn}
-	&\begin{cases}
-	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
-	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
-	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
-	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
-	   0 & \text{otherwise}
-	\end{cases}
-\end{align}
+\subsection{Speculating Several Iterations Ahead}
 
-For each case $\casegen \in \caseset$, there is also a corresponding
-velocity update function $\specvel_{t+1}^{\casegen}$.  When the case is
-known, the specific values of $\pbest_t$ and $\nbest_t$ may be substituted
-directly into~\eqref{eq:velupdate}.  For example, in case $\casexn$,
-$\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
-position, and $\nbest_{t}=\nbest_{t-1}$, as $\nbest$ was not updated at
-iteration $t$:
-\begin{align}
-\nonumber
-	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
-	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
-\label{eq:defvcasexn}
-		&= \constriction \bigl[ \pvel_{t} +
-			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t})
-			+ \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
-			\ppos_{t}) \bigr]
-\end{align}
-
-In the same way we can create notation for the position update function by
-substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
-parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
-subscripts.
-\begin{align}
-\label{eq:defpcasegen}
-	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
-	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t})
-	= \ppos_{t} + \specvel_{t+1}^{\casegen}
-\end{align}
-
-With this notation we can re-write the original PSO velocity
-equation~\eqref{eq:velupdate}, introducing our sum over cases with the
-indicator functions.  Again, we represent the indicator functions and velocity
-functions without the parameters for compactness.  The equation becomes:
-\begin{align}
-\nonumber
-	\pvel_{t+1} &=
-		\constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
-			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
-			\ppos_{t}) \bigr] \\
-\nonumber
-	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
-			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
-			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
-			\ppos_{t}) \bigr]  \\
-\label{eq:vel2update}
-	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
-\end{align}
-
-Similarly, the position update equation~\eqref{eq:posupdate} becomes:
-\begin{align}
-\label{eq:pos2update}
-	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
-	= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
-\end{align}
-
-The value of the objective function at $\ppos_{t+1}$ is given by:
-\begin{align}
-\label{eq:val2update}
-	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \indic_{t+1}^{c}
-	\ \ofunc(\specpos_{t+1}^{c})
-\end{align}
-
-Returning our attention to the computation of $\ppos_{t+1}$ in
-\eqref{eq:pos2update} and writing it with the parameters which were omitted
-above, we obtain:
-\begin{align}
-\nonumber
-  \ppos_{t+1} = \sum_{c \in \caseset}
-	&\indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
-	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
-\label{eq:val2updatelong}
-	& \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
-	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t})
-\end{align}
-
-In this form the important point to notice is that there are only $7$ values
-(for this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
-\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
-objective function evaluation at iteration $t$.  Note also that while there are
-random numbers in the equation, they are assumed fixed once drawn for any
-particular particle at a specific iteration.  Thus PSO has been refactored such
-that the algorithm can begin computing all $7$ of the objective function
-evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
-computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
-only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
-hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.
-
-Although this speculative approach computes $\ofunc(\specpos_{t+1}^{\casegen})$
-for all $\casegen \in \caseset$, even those for which $\indic_{t+1}^{\casegen}
-= 0$, these extra computations will be ignored, and might just as well never
-have been computed.  We call the set of computations
-$\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$ ``speculative
-children'' because only one of them will be needed.
+Speculating one iteration ahead proved very costly in terms of the number of
+extra evaluations required.  Speculating two iterations ahead would require the
+square of the number of extra evaluations of speculating a single iteration
+ahead.  While exactly speculating two iterations ahead would be far too costly
+to be practical, when combined with statistical pruning we can experiment with
+speculating along statistically promising lines several iterations ahead.
+Preliminary experimentation shows that this can be incredibly effective,
+improving the time to completion by a factor of 2 or more, though it also seems
+prone to premature convergence.  This thesis will further explore the idea of
+speculating several iterations ahead.
 
@@
 For each function, we will use the topology that is generally considered to be
-best for optimizing that particular function.  When the Complete topology is
-considered best, we will instead use a Random topology that we developed that
-mimicks the behavior of Complete while being more suited for parallel
-computation~\cite{mcnabb-2009-large-particle-swarms}.
+best for optimizing that particular function, as described by Bratton and
+Kennedy~\cite{bratton-2007-defining-a-standard-for-pso}.  When the Complete
+topology is considered best, we will instead use a Random topology that we
+developed that mimics the behavior of Complete while being more suited for
+parallel computation~\cite{mcnabb-2009-large-particle-swarms}.
 
@@
 of our methods up to higher dimensions.  We will also test our methods on a
-more typical ``real-world'' problem, that of training a computationally
-intensive machine learning algorithm with a large amount of data.  We will
-compare our speculative methods to previous parallelization techniques,
+more typical ``real-world'' problem, that of fitting a model to a large amount
+of data.  While we will run tests on a complex crystallography problem in
+collaboration in Dr. Branton Campbell in the physics department, we will
+present our results on a similar but more well known problem.  In particular,
+we will train a radial basis function on a set of 10,000 data points.  This is
+a similar problem to Dr. Campbell's, in that both seek to match model
+parameters to observed data, but it is easier to describe and quicker to
+experiment with.  Dr. Campbell also has yet to publish his results and we do
+not wish to publish about his problem before he does.
+
+We will compare our speculative methods to previous parallelization techniques,
 including a naive parallelization of just increasing the swarm size, and a
@@
 
-\bibliography{../../../bib/pso/general/bib,%
-../../../bib/pso/parallel/bib,%
-../../../bib/pso/topology/bib,%
-../../../bib/pso/fips/bib,%
-../../../bib/aml/bib,%
-../../../bib/parallel/bib,%
-../../../bib/nfl/bib%
-}
+\bibliography{bib}
 

### 2011-03-24 masters_thesis/si_paper/paper.tex
Fixed a problem with the number of processors we said we used on one function
A lot of work on the paper.  It is very close.
This incorporates a lot of new experiments, plus a lot of changes from
reviewer feedback.

@@
 \newcommand{\fig}[1]{Figure~\ref{fig:#1}}
+\newcommand{\tabref}[1]{Table~\ref{tab:#1}}
 \newcommand{\alg}[1]{Algorithm~\ref{alg:#1}}
@@
 Brigham Young University.
-3342 TMCB, Provo, UT 84604
+3361 TMCB, Provo, UT 84604
 Tel.: 801-422-8717
@@
 Brigham Young University.
-3342 TMCB, Provo, UT 84604
+3361 TMCB, Provo, UT 84604
 \email{a@cs.byu.edu}
@@
 Brigham Young University.
-3330 TMCB, Provo, UT 84604
+3361 TMCB, Provo, UT 84604
 \email{k@cs.byu.edu}
@@
 wasted computation, we show that this approach to parallelization in PSO often
-outperforms the na\"ive parallelization of simply adding particles to the
+outperforms the naive parallelization of simply adding particles to the
 swarm.  Our methods produce results that are exactly equivalent to PSO; this is
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{iters_sphere}
+  \includegraphics[width=.7\columnwidth]{iters_sphere}
   \caption{Function Sphere with various swarm sizes, comparing performance with
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{evals_sphere}
+  \includegraphics[width=.7\columnwidth]{evals_sphere}
   \caption{Function Sphere with various swarm sizes, comparing performance with
@@
 longer demanding that it strictly reproduce the exact behavior of standard PSO,
-we can introduce new speculative techniques that often out-perform na\"ive
+we can introduce new speculative techniques that often out-perform naive
 parallelizations of PSO.  These relaxations make better use of the information
@@
 functions than the same swarm size with a Ring topology, it was also more
-likely to permaturely converge to local optima for other functions.  The study
+likely to prematurely converge to local optima for other functions.  The study
 found no significant improvement for any other swarm size between 20 and 100
@@
 problem~\citep{bratton-2007-defining-a-standard-for-pso}.  Other authors have
-explained that a large swarm with a sparse topology propogates information
+explained that a large swarm with a sparse topology propagates information
 slowly~\citep{montes-de-oca-2009-frankensteins-pso}.
@@
 
-There have been several parallel implementations of PSO presented in the
-literature.  The improvements described in the literature come in two major
-areas: innovations in implementation details and innovations in the use of
-topology and swarm size to scale PSO to many processors.
+The idea of speculative decomposition in the parallelization literature is not
+new~\citep{grama-2003-intro-to-parallel-computing}.  In the field of function
+optimization, simulated annealing has previously been parallelized using this
+technique~\citep{witte-1991-parallel-simulated-annealing-speculative}, though
+we are not aware of other evolutionary or swarm-intelligence based algorithms
+having been parallelized with speculative decomposition.
+
+There have been several parallelizations of PSO presented in the literature.
+The improvements described in these papers come in two major areas: innovations
+in implementation details and innovations in the use of topology and swarm size
+to scale PSO to many processors.
 
@@
 neighbor,'' whose positions are represented as $\rightn$ and $\leftn$
-respectively.  Though we will only describe the case of the Ring topology, this
-method can easily be extended to arbitrary topologies.
+respectively.  Though we will only describe the case of the Ring topology here,
+the math is straightforward for other topologies.  Our discussion of the
+implementation in \sec{implementation} is independent of specific topologies,
+and we use several different topologies in our experiments.
 
@@
 too many extra evaluations to produce competitive results.  On most functions,
-the larger swarm size that can be used in na\"ive parallelizations leads to
+the larger swarm size that can be used in naive parallelizations leads to
 better performance than is found by trying to do speculative evaluation to
@@
 through~\ref{sec:wrong} present a technique that reduces the number of
-speculative evaluations that need to be done for each particle.  Finally,
-\sec{manyiters} shows a method for speculating several iterations ahead,
-instead of just one.
+speculative evaluations that need to be done for each particle, allowing
+speculative evaluation to use larger swarm sizes with the same number of
+processors.  Finally, \sec{manyiters} shows a method for speculating several
+iterations ahead, instead of just one.
 
@@
 selection operator is either the indicator function introduced in \sec{sepso},
-in the case of our original speculative algorithm, or the standard selector
+in the case of our original speculative algorithm, or the standard selection
 operator based on fitness, in the case of this Pick Best technique.
@@
 increase the swarm size relative to that needed in the original speculative
-algorithm (e.g., with 120 processors, a na\"ive parallelization has a swarm of
+algorithm (e.g., with 120 processors, a naive parallelization has a swarm of
 size 120, complete speculative evaluation has a swarm of size 15, and pruning
@@
 
-A na\"ive approach to pruning is to keep track of the last branch taken by each
+A naive approach to pruning is to keep track of the last branch taken by each
 particle and speculate on that branch.  This turns out to be correct far less
@@
 We ran experiments to compare our speculative parallelization of PSO to the
-na\"ive parallelization.  At each iteration of the algorithms, we use one
+naive parallelization.  At each iteration of the algorithms, we use one
 processor to perform one function evaluation for one particle, be it
-speculative or non-speculative.  The speculative version actually performs
-two iterations of PSO at each ``iteration,'' so we instead call each
-``iteration'' a ``time step.''
+speculative or non-speculative.  The speculative version actually performs two
+iterations of PSO at each ``iteration,'' so we instead call each ``iteration''
+a ``time step.''  As noted in the introduction, this is a reasonable
+architecture to use if the function being optimized takes longer than around
+100 milliseconds to run.
 
@@
 
-We also do not report absolute time used by each algorithm, as specific time
-values are heavily implementation-dependent.  Using time steps, where at every
-time step each processor evaluates the function once, is an abstraction from
-individual implementations that still conveys all of the information necessary
-to compare algorithms.  In this paper we focus on the effects of algorithm
-choices on parallel PSO performance.  As there are inexpensive functions with
-complex landscapes and expensive functions with simple landscapes, the behavior
-of PSO with respect to the function is the main issue.  For this purpose,
-benchmark functions are a useful and efficient tool for understanding the
-effects of PSO with expensive objective functions even though the benchmark
-functions are themselves inexpensive.  For example, a plot of performance with
-respect to iterations for a smooth unimodal function with five minutes per
-evaluation would be similar to that of Sphere.
-
 Using the same number of processors for each approach requires that our
-speculative parallelizations have a smaller swarm size than the na\"ive
+speculative parallelizations have a smaller swarm size than the naive
 parallelization.  For each of the topologies we used, a particle has three
@@
 extra evaluations required in our speculative approach would instead be used to
-evaluate particles in na\"ive parallelizations, so we compare swarms of size
-$p$ in speculative evaluation with swarms of size $8p$ in na\"ive approaches.
-
-Where the Complete topology would normally be preferred, we use a Random
-topology in our speculative approach, as Complete leads to an explosion in the
-number of speculative evaluations.  If speculative evaluation were not being
-performed, it is possible that the Complete topology would be used.  However,
-the Complete topology also requires a very large amount of interprocessor
-communication in distributed PSO, so it is still quite possible that Random
-would be used even with na\"ive parallelizations.  But, to be fair in our
-comparisons, we compare to the na\"ive parallelization using both the Random
-topology and the Complete topology (labeled PSO Random and PSO Complete in our
-results).
+evaluate particles in naive parallelizations, so we compare swarms of size
+$p$ in speculative evaluation with swarms of size $8p$ in naive approaches.
+
+\subsection{Running Time}
+
+We do not report absolute time used by each algorithm, as specific time values
+are heavily implementation-dependent.  Using time steps, where at every time
+step each processor evaluates the function once, is an abstraction from
+individual implementations that still conveys all of the information necessary
+to compare algorithms.  In this paper we focus on the effects of algorithm
+choices on parallel PSO performance.  As there are inexpensive functions with
+complex landscapes and expensive functions with simple landscapes, the behavior
+of PSO with respect to the function is the main issue.
+
+For this purpose, benchmark functions are a useful and efficient tool for
+understanding the effects of PSO with expensive objective functions even though
+the benchmark functions are themselves inexpensive.  While benchmark functions
+themselves take fractions of a second to evaluate and thus have no need of the
+kind of parallelization we explore in this paper, they also have no need of
+particle swarm optimization at all, as they can be solved analytically.  Yet
+they are useful to us and to optimization researchers generally because they
+stand as surrogates for the kinds of functions practitioners are actually
+interested in, and they allow us to explore the behavior of optimization
+algorithms on several different classes of functions in an easy and
+standardized way.
+
+If experimentation with benchmark functions provides sufficient experimental
+evidence for algorithmic improvements in serial PSO, it is also sufficient
+evidence for algorithmic improvements in parallel PSO.  While the discussion of
+exact running times of function evaluations and communication overhead is
+certainly appropriate, even essential, to a discussion of specific parallel
+implementations, the concern of this paper is more fundamental than
+implementation decisions.  No one expects to see actual running times in place
+of function evaluations in a paper describing an innovation in the motion
+equations of PSO.  Similarly, actual running times should not be required of
+those describing algorithmic improvements in parallel PSO.  Just as in a serial
+PSO paper, actual running times cloud the more important issues of
+implementation-independent properties of the innovations described.
+
+That being said, given the time required for the evaluation of the objective
+function and the communication overhead per iteration for a specific
+implementation, the actual running time can be recovered from all of our
+results.  Simply multiply the number of time steps by the sum of the function
+evaluation time and the overhead.  We wish to stress that the ``time steps'' we
+report are \emph{exactly} proportional to wall-clock time, given the
+architecture we have assumed.
+
+\subsection{Functions Used}
 
@@
 the boundary of the initialization region.  For example, if the initialization
-region is $[-50,50]^D$, the center is $(25)^D$.  For all of these functions we
-used 20 dimensions.
-
-The parallelization techniques we compare are the na\"ive parallelization (here
-labeled Na\"ive simply for ease of reference), our original speculative
+region is $[-50,50]^D$, the center is $(25)^D$.  We tested all of these
+functions in their 20 dimension, 50 dimension, and 500 dimension varieties.
+
+We also tested the parallelization methods on a real problem, that of fitting a
+model to a large quantity of data.  We generated 10,000 data points from a
+radial basis function network with 10 bases and some added Gaussian noise.  We
+then fit a radial basis function network to the data using PSO.  This amounted
+to a 30 dimensional function to optimize, with a function evaluation time on
+the order of two seconds.
+
+\subsection{Parallelization Techniques Compared}
+
+The parallelization techniques we compare are the naive parallelization (here
+labeled Naive simply for ease of reference), our original speculative
 approach (labeled SEPSO), and the four relaxations of SEPSO discussed in
@@
 (see \sec{distributed}), one of the rounds of communication can be dropped
-entirely.
+entirely.  Pruning all but these branches also allows the use of arbitrarily
+dense topologies, as the number of speculative particles is no longer dependent
+on the number of neighbors the particle has.
 
@@
 with evaluating the original particle), we can use swarms of size
-$\frac{1}{3}p$ to compare to a swarm of size $p$ with the na\"ive
+$\frac{1}{3}p$ to compare to a swarm of size $p$ with the naive
 parallelization, instead of $\frac{1}{8}p$ with other techniques.
@@
 then branch \casepn\ or \casexn; and the evaluation going three iterations
-ahead followed branch \casepn\ on all three iterations.
+ahead followed branch \casepn\ on all three iterations.  As with Pick Best
+Pruned and Social Promotion Pruned, this choice of branches allows the use of
+arbitrarily dense topologies, as the number of speculative evaluations per
+particle is independent of the number of neighbors the particle has.
+
+\subsection{Topologies}
+
+For each benchmark function we report results using the topology that is widely
+considered best for that function, as reported in the literature
+(e.g.,~\cite{bratton-2007-defining-a-standard-for-pso}).  In this paper we
+limit ourselves to the Ring topology and the Complete topology, as is common
+practice, along with the parallel approximation to the Complete topology
+mentioned in \sec{pso}.  We also mentioned in \sec{related} that some related
+work can be described as changes in topology, particularly that of having
+subswarms of fully connected topologies that occasionally communicate with
+each other.  This related work focused mainly on functions with very fast
+evaluation times where such techniques drastically reduce interprocessor
+communication.  With long function evaluations these topologies are not as
+practical, as only one particle is evaluated by each processor at each
+iteration.  We experimented with a subswarm topology and found that in every
+instance except one it performed worse than either Complete or Ring, and thus
+we leave it out of the results except in the one instance where it improved
+performance.
+
+Where the Complete topology would normally be preferred, we use a Random
+topology in SEPSO and Pick Best, as Complete leads to an explosion in the
+number of speculative evaluations (as noted, the other speculative techniques
+can still use a Complete topology; we often report results on both topologies
+for those methods).  If speculative evaluation were not being performed, it is
+possible that the Complete topology would be used.  However, the Complete
+topology also requires a very large amount of interprocessor communication in
+distributed PSO, so it is still quite possible that Random would be used even
+with naive parallelizations~\cite{mcnabb-2009-large-particle-swarms}.  But,
+to be fair in our comparisons, we compare to the naive parallelization using
+both the Random topology and the Complete topology (labeled PSO Random and PSO
+Complete in our results).  Again, the amount of communication overhead is
+heavily dependent on implementation details which are not the focus of this
+paper.  A practitioner using a particular implementation can compare results
+for whichever topology is most practical given the specific implementation.
 
@@
 
-\subsection{Rastrigin}
-
-First we look at Rastrigin.  Rastrigin is a multi-modal function that is best
-optimized with a large, Complete swarm.  It has been shown that with Rastrigin,
-the more particles there are in the swarm, the lower function value it finds,
-up to at least 4000 particles~\citep{mcnabb-2009-large-particle-swarms}.
-Smaller swarms get caught in local optima.  Because our speculative algorithms
-require significantly smaller swarm sizes, we would expect to not perform very
-well on functions such as Rastrigin.  Our experiments show our intuition to be
-correct.  In this experiment we used 240 processors, so SEPSO and Pick Best
-each had 30 particles, and the Na\"ive algorithms had 240 particles.  As
-expected, SEPSO and Pick Best converge quickly to worse local optima than
-Na\"ive does.  \fig{rastrigin} shows the results graphically.
+We first present results in \sec{20dims} for the 20 dimensional variants of the
+benchmark functions we tested.  We give a discussion of each function
+individually at 20 dimensions, as there are interesting characteristics of the
+algorithms that are worth discussing.  In \sec{50dims} we then give a summary
+of results for the 50 dimensional variants of the benchmark functions; we do
+not go into as much detail in our discussion as the results are very similar to
+those in 20 dimensions.  We finish our discussion of benchmark functions with
+their 500 dimensional variants in \sec{500dims}, and we discuss results on the
+model fitting problem in \sec{model}.
+
+\subsection{20 Dimensions}
+\label{sec:20dims}
+
+\subsubsection{Rastrigin}
+
+Rastrigin is a multi-modal function that is best optimized with a large,
+Complete swarm.  It has been shown that with Rastrigin, the more particles
+there are in the swarm, the lower function value it finds, up to at least 4000
+particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
+in local optima.  Because our speculative algorithms require significantly
+smaller swarm sizes, we would expect to not perform very well on functions such
+as Rastrigin.  Our experiments show our intuition to be correct.  In this
+experiment we used 240 processors, so SEPSO and Pick Best each had 30
+particles, and the Naive algorithms had 240 particles.  As expected, SEPSO and
+Pick Best converge quickly to worse local optima than Naive does.
+\fig{rastrigin} shows the results graphically, and \tabref{rastrigin-20} shows
+results for all of the methods we tried.
+
+\begin{table}
+  \caption{Summary of results for function Rastrigin with 20 dimensions,
+  measuring number of time steps to reach a value of 20}
+  \label{tab:rastrigin-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Naive Ring&25\%&743.4&212.0\\
+  \hline
+  Naive Random&100\%&372.1&86.6\\
+  \hline
+  \textbf{Naive Complete}&\textbf{95\%}&\textbf{273.1}&\textbf{99.7}\\
+  \hline
+  SEPSO Ring&5\%&471.0&0.0\\
+  \hline
+  SEPSO Random&15\%&208.0&17.0\\
+  \hline
+  Pick Best Ring&35\%&489.0&152.0\\
+  \hline
+  Pick Best Random&15\%&136.0&40.1\\
+  \hline
+  Pick Best Pruned Ring&95\%&545.9&208.5\\
+  \hline
+  Pick Best Pruned Random&10\%&95.0&11.0\\
+  \hline
+  Pick Best Pruned Complete&5\%&59.0&0.0\\
+  \hline
+  Social Promotion Pruned Ring&5\%&580.0&0.0\\
+  \hline
+  Social Promotion Pruned Random&80\%&278.2&136.3\\
+  \hline
+  Social Promotion Pruned Complete&50\%&200.5&130.2\\
+  \hline
+  Many Iterations Ring&10\%&117.0&25.0\\
+  \hline
+  Many Iterations Random&5\%&57.0&0.0\\
+  \hline
+  Many Iterations Complete&0\%&N/A&N/A\\
+  \hline
+  \end{tabular}
+\end{table}
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{rastrigin}
+  \includegraphics[width=.7\columnwidth]{rastrigin}
   \caption{Function Rastrigin with a swarm that uses 240 processors per time
@@
 
-\subsection{Sphere}
-
-Next we go to Sphere, the simplest of common benchmark functions.  The function
-has no local optima (only a single global optimum) and is best optimized in
-terms of function evaluations with a small swarm using a Complete topology.  We
-expect our methods to be perfect for such functions, and our results show this
-intuition again to be correct.  First we compare the basic SEPSO approach to
-Na\"ive.  In \fig{basic-sphere} we can see that when both SEPSO and Na\"ive are
-using a Random topology, SEPSO clearly outperforms Na\"ive.  However, the
-comparison between SEPSO and Na\"ive Complete is not as encouraging---Na\"ive
-Complete performs slightly better.
+\subsubsection{Griewank}
+\label{sec:griewank-20}
+
+It is generally recommended to use the Ring topology when optimizing the
+Griewank function, as Complete is prone to premature convergence on a local
+optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  The global optimum
+of Griewank has a value of 0.  When most trials reach the global optimum but a
+few get stuck, the resultant ``average value'' graph has a flat line that is
+misleading.  Thus instead we present plots showing the percent of trials that
+have passed some threshold at each time step.  The threshold we chose for this
+case was $10^{-6}$, as that value is below any local optima and the swarm
+always successfully reaches 0 once is passes that point.
+
+We ran 50 trials of each experiment with Griewank, so that the curves are more
+smooth.  We show results in \fig{basic-griewank1} for swarms of size 30 and 240
+using the Ring topology.  One can see in the figure that the SEPSO approach
+finds the optimum faster than the Naive approach.  However, because the swarm
+size is so small, SEPSO gets stuck a little less than half of the time.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{sphere}
-  \caption{Function Sphere with a swarm that uses 240 processors per time step.
-  Recall that at each time step all processors evaluate the function once.
-  SEPSO uses the Random topology.}
-  \label{fig:basic-sphere}
-\end{figure}
-
-When we introduce Pick Best, we see more impressive results.  As can be seen in
-\fig{sphere-pickbest}, where SEPSO did not manage to perform as well as Na\"ive
-Complete, Pick Best handily outperforms it with the same number of processors.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere2}
-  \caption{Function Sphere with a swarm that uses 240 processors per time step,
-  comparing Pick Best with previous results.  All speculative algorithms use
-  the Random topology.}
-  \label{fig:sphere-pickbest}
-\end{figure}
-
-Still more impressive with Sphere are the results of pruning and speculating
-several iterations ahead.  Pick Best Pruned outperforms Pick Best, and Many
-Iterations by far outperforms every other algorithm on this function.  Many
-Iterations on average finds values that are more than 40 orders of magnitude
-better than Na\"ive using the same number of processors.  This is shown in
-\fig{sphere-manyiters}.  Social Promotion Pruned performed comparably with
-SEPSO, and we do not clutter the graph with a line for it.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{sphere3}
-  \caption{Function Sphere with a swarm that uses 240 processors per time step,
-  comparing speculating several iterations ahead with previous results.  All
-  speculative algorithms use the Random topology.}
-  \label{fig:sphere-manyiters}
-\end{figure}
-
-\subsection{Schwefel 2.21}
-
-Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
-more by larger swarms than Sphere is.  Thus our speculative algorithms often
-fail to perform as well as Na\"ive because of their smaller swarm size.  In
-comparing SEPSO to Na\"ive in \fig{schwefel1}, we can see that SEPSO fails to
-perform any better than Na\"ive with a Random topology, and it is by far
-outperformed by Na\"ive with a Complete topology.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{schwefel1}
-  \caption{Function Schwefel with a swarm that uses 240 processors per time
-  step.  Recall that at each time step all processors evaluate the function
-  once.  SEPSO uses a Random topology.}
-  \label{fig:schwefel1}
-\end{figure}
-
-As can be seen in \fig{schwefel2}, almost all of the speculative algorithms
-perform worse than Na\"ive Complete, with Many Iterations performing comparably
-to Na\"ive.  However, one of the speculative algorithms did perform better than
-Na\"ive: Pick Best Pruned.  It seems that the larger swarm size available
-because of pruning was the cause of the improved performance.  Note that Many
-Iterations was initially performing comparably with Pick Best Pruned, but with
-more iterations gradually degraded in performance to the level of Na\"ive
-Complete.  It is likely that Many Iterations got caught in local optima, and
-with more processors would have maintained better performance.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{schwefel2}
-  \caption{Function Schwefel with a swarm that uses 240 processors per time
-  step, comparing all speculative algorithms.  All speculative algorithms use
-  the Random topology.}
-  \label{fig:schwefel2}
-\end{figure}
-
-\subsection{Griewank}
-
-Next we look at the function Griewank.  It is generally recommended to use the
-Ring topology when optimizing the Griewank function, as Complete is prone to
-premature convergence on a local
-optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  Griewank has a
-global optimum with a value of 0, and sometimes the swarm finds the optimum and
-sometimes it does not.  Instead of showing average function value at each
-iteration, a more informative plot for Griewank shows the percent of runs that
-have found the global optimum by each iteration.  We ran 50 trials of each
-experiment with Griewank, so that the curves are more smooth.  We show results
-in \fig{basic-griewank1} for swarms of size 30 and 240 using the Ring topology.
-One can see in the figure that the SEPSO approach finds the optimum faster than
-the Na\"ive approach.  However, because the swarm size is so small, SEPSO gets
-stuck a little less than half of the time.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\columnwidth]{griewank1}
+  \includegraphics[width=.7\columnwidth]{griewank1}
   \caption{Function Griewank with a swarm that uses 240 processors per time
@@
 Thus with a few more processors, our algorithm finds the optimum on average
-twice as fast as the Na\"ive approach without a significant loss of accuracy.
+twice as fast as the Naive approach without a significant loss of accuracy.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank2}
+  \includegraphics[width=.7\columnwidth]{griewank2}
   \caption{Function Griewank with a swarm that uses 800 processors per time
@@
 
-When we look at the performance of our Pick Best approach, a less intuitive
-result occurs with Griewank than with Sphere.  Because Sphere needs very little
-exploration, the seemingly greedy approach of Pick Best does very well.  But it
-also greatly improves performance on Griewank.  In \fig{griewank-pickbest} we
-see that Pick Best improves accuracy over SEPSO by 20\%, while at the same time
-finding the optimum over 100 rounds of evaluation sooner on average.  It seems
-that while Pick Best is locally greedy, there is enough exploration in the
-seven speculative evaluations to overcome the inherent greediness of the
-approach.
+When we look at the performance of our Pick Best approach, we see that it
+greatly improves performance on Griewank.  This is somewhat counter-intuitive,
+because Griewank is deceptive and Pick Best seems like a greedy algorithm.  But
+in \fig{griewank-pickbest} we see that Pick Best improves accuracy over SEPSO
+by 20\%, while at the same time finding the optimum over 100 time steps sooner
+on average.  It seems that while Pick Best is locally greedy, there is enough
+exploration in the seven speculative evaluations to overcome the inherent
+greediness of the approach.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank3}
+  \includegraphics[width=.7\columnwidth]{griewank3}
   \caption{Function Griewank with a swarm that uses 240 processors per time
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank4}
+  \includegraphics[width=.7\columnwidth]{griewank4}
   \caption{Function Griewank with a swarm that uses 240 processors per time
@@
 Iterations finds the optimum, it finds it quicker than any other method we
-tried, on average four times faster than Na\"ive.  However, it also gets stuck
+tried, on average four times faster than Naive.  However, it also gets stuck
 and fails to find the optimum more than any other method.  The results are
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank5}
+  \includegraphics[width=.7\columnwidth]{griewank5}
   \caption{Function Griewank with a swarm that uses 240 processors per time
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{griewank6}
+  \includegraphics[width=.7\columnwidth]{griewank6}
   \caption{Function Griewank with a swarm that uses 800 processors per time
@@
 
-\subsection{Bohachevsky}
+\subsubsection{Sphere}
+
+Next we go to Sphere, the simplest of common benchmark functions.  The function
+has no local optima (only a single global optimum) and is best optimized in
+terms of function evaluations with a small swarm using a Complete topology.  We
+expect our methods to be perfect for such functions, and our results show this
+intuition to be correct.  We compared too many methods here to fit into one
+graph, so we show just a few methods in \fig{sphere-20} and in
+\tabref{sphere-20} we show a summary of the results for all methods we tested.
+
+In \fig{sphere-20} we see that Many Iterations using a Complete topology
+converges incredibly quickly on a very poor value.  We found this behavior to
+be quite consistent across functions, so we rarely show results for Many
+Iterations Complete.  However, Pick Best Pruned works very well with Complete
+on this function.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{sphere20}
+  \caption{Function Sphere with a swarm that uses 240 processors per time step,
+  comparing a pertinent subset of possible methods.}
+  \label{fig:sphere-20}
+\end{figure}
+
+\begin{table}
+  \caption{Summary of results for function Sphere with 20 dimensions,
+  measuring number of time steps to reach a value of 1e-35}
+  \label{tab:sphere-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Naive Ring&0\%&N/A&N/A\\
+  \hline
+  Naive Random&0\%&N/A&N/A\\
+  \hline
+  Naive Complete&100\%&931.0&19.1\\
+  \hline
+  SEPSO Ring&0\%&N/A&N/A\\
+  \hline
+  SEPSO Random&85\%&972.8&15.8\\
+  \hline
+  Pick Best Ring&100\%&917.4&21.0\\
+  \hline
+  Pick Best Random&100\%&768.1&14.3\\
+  \hline
+  Pick Best Pruned Ring&100\%&967.8&9.9\\
+  \hline
+  Pick Best Pruned Random&100\%&693.5&10.4\\
+  \hline
+  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{389.6}&
+    \textbf{9.6}\\
+  \hline
+  Social Promotion Pruned Ring&0\%&N/A&N/A\\
+  \hline
+  Social Promotion Pruned Random&100\%&971.1&10.6\\
+  \hline
+  Social Promotion Pruned Complete&100\%&777.8&16.7\\
+  \hline
+  Many Iterations Ring&100\%&575.2&9.8\\
+  \hline
+  Many Iterations Random&100\%&442.4&7.1\\
+  \hline
+  Many Iterations Complete&0\%&N/A&N/A\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\subsubsection{Schwefel 2.21}
+
+Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
+more by larger swarms than Sphere is.  Thus our speculative algorithms often
+fail outperform Naive with 240 processors because the simple speculative
+methods only have 30 particles.  As we will show later, when we use 800
+processors at 50 dimensions, our methods perform much better.  However, the
+pruned versions of our methods have 80 particles and thus are able to
+outperform Naive even with only 240 processors.  Because the graph of Schwefel
+2.21 looks very similar to that of Sphere, we present all of our results in
+\tabref{schwefel-20}.
+
+\begin{table}
+  \caption{Summary of results for function Schwefel with 20 dimensions,
+  measuring number of time steps to reach a value of 1e-06}
+  \label{tab:schwefel-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Naive Ring&0\%&N/A&N/A\\
+  \hline
+  Naive Random&0\%&N/A&N/A\\
+  \hline
+  Naive Complete&100\%&837.2&44.8\\
+  \hline
+  SEPSO Random&0\%&N/A&N/A\\
+  \hline
+  Pick Best Random&5\%&938.0&0.0\\
+  \hline
+  \textbf{Pick Best Pruned Random}&\textbf{100\%}&\textbf{639.9}&\textbf{29.4}\\
+  \hline
+  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{597.7}&
+	\textbf{77.9}\\
+  \hline
+  Social Promotion Pruned Random&75\%&938.0&33.0\\
+  \hline
+  Social Promotion Pruned Complete&100\%&815.2&43.5\\
+  \hline
+  Many Iterations Random&100\%&783.5&94.5\\
+  \hline
+  Many Iterations Complete&0\%&N/A&N/A\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\subsubsection{Bohachevsky}
 
@@
 similar to Griewank in that there is a global optimum with a value of 0, and
-the swarm sometimes finds it and sometimes does not.  Since we have already
-demonstrated the convergence properties of the various algorithms with
-Griewank, for Bohachevsky we merely present the best results.  In
-\fig{bohachevsky} we can see that all of the speculative approaches found the
-optimum much quicker than Na\"ive with a Random topology.  However, SEPSO was
-slower than Na\"ive Complete and got stuck around 15\% of the time.  Pick Best,
-Pick Best Pruned, and Many Iterations all outperformed Na\"ive Complete, with
-Many Iterations finding the optimum about twice as fast.
+the swarm sometimes finds it and sometimes does not.  Thus we present a graph
+similar to those of Griewank, because ``average value'' graphs have a
+misleading flat line.  In \fig{bohachevsky} we show a plot with a few pertinent
+methods, while \tabref{bohachevsky-20} shows all of the results.  All of our
+speculative approaches found the optimum much quicker than Naive with a Random
+topology.  However, SEPSO was slower than Naive Complete and got stuck 25\% of
+the time.  Pick Best, Pick Best Pruned, and Many Iterations all outperformed
+Naive Complete, with Many Iterations finding the optimum about twice as fast.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{bohachevsky}
-  \caption{Function Bohachevsky with a swarm that uses 800 processors per time
+  \includegraphics[width=.7\columnwidth]{bohachevsky}
+  \caption{Function Bohachevsky with a swarm that uses 480 processors per time
   step.  Recall that at each time step all processors evaluate the function
-  once.  All of the speculative algorithms use the Random topology.}
+  once.}
   \label{fig:bohachevsky}
+\end{figure}
+
+\begin{table}
+  \caption{Summary of results for function Bohachevsky with 20 dimensions,
+  measuring number of time steps to reach a value of 1e-06.}
+  \label{tab:bohachevsky-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Naive Random&100\%&472.3&6.6\\
+  \hline
+  Naive Complete&100\%&238.2&11.2\\
+  \hline
+  SEPSO Random&75\%&261.6&9.2\\
+  \hline
+  Pick Best Random&100\%&211.8&9.2\\
+  \hline
+  Pick Best Pruned Random&100\%&189.7&3.4\\
+  \hline
+  Pick Best Pruned Complete&FIXME!&FIXME!&FIXME!\\
+  \hline
+  Social Promotion Pruned Random&100\%&260.3&6.5\\
+  \hline
+  Social Promotion Pruned Complete&FIXME!&FIXME!&FIXME!\\
+  \hline
+  \textbf{Many Iterations Random}&\textbf{94\%}&\textbf{118.2}&\textbf{2.6}\\
+  \hline
+  Many Iterations Complete&0\%&N/A&N/A\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\subsection{50 Dimensions}
+\label{sec:50dims}
+
+Given our observation from the 20 dimensional benchmark functions that Many
+Iterations most often outperforms other speculative approaches, we only present
+results for Many Iterations and Naive for the 50 and 500 dimensional
+variants.  For these experiments we used 800 processors, as the problems are
+more difficult and speculative approaches perform best when many processors are
+available, as we showed in \sec{griewank-20}.  Many Iterations Complete showed
+the same premature convergence that was observed with the 20 dimensional
+benchmark functions, so we do not include those results in our tables; the
+success rate was 0\% for all functions.
+
+Tables~\ref{tab:rastrigin-50} through~\ref{tab:bohachevsky-50} present
+summaries of our results for each of the benchmark functions. As with the 20
+dimensional variants, our speculative methods did not perform well on
+Rastrigin, though for all other benchmarks Many Iterations significantly
+outperformed Naive.  In each table, we bold the best method, meaning that it
+has at least a 90\% success rate and its average time to completion is faster
+than all other methods that have at least a 90\% success rate.  For example,
+Many Iterations Random, when it is successful, achieves the required value 6
+times faster than Naive Random.  However, because Many Iterations Random only
+has a 25\% success rate while Naive Random has a 100\% success rate, Naive
+Random is bolded.
+
+We used a double-sided t-test to test for statistical significance in these
+results, and all bolded values are statistically significant with a p-value of
+less than $10^{-5}$.  In one instance two values are bolded because they are
+statistically tied, while both are significantly better than all other methods.
+
+A summary of the results is shown in \tabref{50dims-summary}.  For functions
+where Many Iterations outperformed Naive, we report the average speed up.  On
+all functions except Rastrigin, our methods showed an average speed up of from
+two to six times compared to previous methods.
+
+\begin{table}
+  \caption{Summary of results for function Rastrigin with 50 dimensions,
+  measuring number of time steps to reach a value of 100}
+  \label{tab:rastrigin-50}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Naive Ring&5\%&776.0&0.0\\
+  \hline
+  \textbf{Naive Random}&\textbf{100\%}&\textbf{593.2}&\textbf{89.5}\\
+  \hline
+  Naive Complete&80\%&343.8&145.8\\
+  \hline
+  Many Iterations Ring&25\%&223.8&67.1\\
+  \hline
+  Many Iterations Random&25\%&98.6&11.0\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Summary of results for function Sphere with 50 dimensions,
+  measuring number of time steps to reach a value of 1e-06}
+  \label{tab:sphere-50}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Naive Ring&0\%&N/A&N/A\\
+  \hline
+  Naive Random&30\%&986.8&8.2\\
+  \hline
+  Naive Complete&100\%&458.4&12.3\\
+  \hline
+  Many Iterations Ring&100\%&288.7&3.9\\
+  \hline
+  \textbf{Many Iterations Random}&\textbf{100\%}&\textbf{248.6}&\textbf{5.1}\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Summary of results for function Schwefel with 50 dimensions,
+  measuring number of time steps to reach a value of 80}
+  \label{tab:schwefel-50}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Naive Ring&0\%&N/A&N/A\\
+  \hline
+  Naive Random&70\%&871.6&57.1\\
+  \hline
+  Naive Complete&100\%&599.0&85.6\\
+  \hline
+  \textbf{Many Iterations Ring}&\textbf{100\%}&\textbf{238.6}&\textbf{34.2}\\
+  \hline
+  \textbf{Many Iterations Random}&\textbf{100\%}&\textbf{242.9}&\textbf{95.4}\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Summary of results for function Griewank with 50 dimensions,
+  measuring number of time steps to reach a value of 1e-06}
+  \label{tab:griewank-50}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Naive Ring&100\%&1898.5&25.7\\
+  \hline
+  \textbf{Many Iterations Ring}&\textbf{97\%}&\textbf{312.4}&\textbf{6.3}\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Summary of results for function Bohachevsky with 50 dimensions,
+  measuring number of time steps to reach a value of 0.01}
+  \label{tab:bohachevsky-50}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Naive Ring&100\%&1434.0&62.1\\
+  \hline
+  Naive Random&100\%&704.5&19.7\\
+  \hline
+  Naive Complete&20\%&377.5&14.9\\
+  \hline
+  \textbf{Many Iterations Ring}&\textbf{95\%}&\textbf{229.8}&\textbf{11.4}\\
+  \hline
+  Many Iterations Random&35\%&171.6&4.0\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Average speed up, comparing the number of time steps to completion
+  for the best speculative topology to the time steps to completion for the
+  best naive topology.  All functions have 50 dimensions.}
+  \label{tab:50dims-summary}
+  \centering
+  \begin{tabular}{|l|r|}
+	\hline
+	Function&Speed Up Factor\\
+	\hline
+	Rastrigin&N/A\\
+	\hline
+	Sphere&1.84\\
+	\hline
+	Schwefel&2.51\\
+	\hline
+	Griewank&6.08\\
+	\hline
+	Bohachevsky&3.07\\
+	\hline
+  \end{tabular}
+\end{table}
+
+\subsection{500 Dimensions}
+\label{sec:500dims}
+
+At 500 dimensions we found that the performance of constricted PSO on benchmark
+functions becomes rather dismal.  In almost all cases, Naive fails to make
+any significant progress.  Our brief search for an explanation seems to say
+that the space is too large for particles to converge to the same point, so all
+updates to the best position found come from a single particle wandering in the
+space.  Every time that single particle finds a new best position, its velocity
+contracts, so the particle is unable to make significant progress on its own.
+In constricted PSO it is necessary to have a collection of particles exploring
+a promising location to keep the particles' velocities from contracting
+prematurely, and in 500 dimensions the space is too large to get the collection
+of particles to the same location.
+
+The results that we present show that while Naive parallelizations suffer
+from this problem, Many Iterations does not.  Though we can provide some
+intuition for why this might be the case, it remains as future work to provide
+a strong theoretical explanation.  The intuition is that we are speculating
+about future positions along paths where each particle's velocity does not
+contract.  The particular branches we selected corresponded to not having
+received a new value for $\pbest$ nor $\nbest$.  In those branches, the
+particle continues in the same direction it was going, without a contraction in
+velocity.
+
+For example, suppose a particle finds a new good position and updates its
+values for $\pbest$ and $\nbest$, and that its velocity is such that it will
+continue updating $\pbest$ and $\nbest$ if it continues in the same direction.
+At that point, its velocity contracts, due to the motion equations in
+constricted PSO (the social and cognitive components drop to zero and all that
+is left is momentum, which is then constricted).  However, at that point we
+speculate three iterations ahead along a path that is in roughly the same
+direction the particle was just going, without constricting its velocity at
+each of those iterations.  Thus the particle is able to make progress in that
+direction instead of contracting its velocity at every iteration.
+
+Because none of the Naive approaches were successful at optimizing the 500
+dimensional benchmark functions, we do not present tables as we did for the 50
+dimensional variants.  We do, however, show a few figures demonstrating the
+results we have just explained.  \fig{sphere500} shows the function Sphere,
+\fig{griewank500} shows Griewank, and \fig{rastrigin500} shows Rastrigin.  Note
+that at 500 dimensions speculative approaches outperform naive
+parallelizations on Rastrigin, whereas they did not at 20 and 50 dimensions.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{sphere500}
+  \caption{Results for the 500 dimensional Sphere function.  We use 800
+  processors for this experiment, so each time step corresponds to 800 function
+  evaluations.}
+  \label{fig:sphere500}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{griewank500}
+  \caption{Results for the 500 dimensional Griewank function.  We use 800
+  processors for this experiment, so each time step corresponds to 800 function
+  evaluations.}
+  \label{fig:griewank500}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{rastrigin500}
+  \caption{Results for the 500 dimensional Rastrigin function.  We use 800
+  processors for this experiment, so each time step corresponds to 800 function
+  evaluations.}
+  \label{fig:rastrigin500}
+\end{figure}
+
+One could argue that a fairer comparison would have modified the PSO motions
+equations so that individual particles did not contract their velocity so
+quickly when finding good locations.  However, we used the same motion
+equations for both Naive and Many Iterations; changing the motion equations
+to help Naive would also help Many Iterations, and we would expect to see
+results that are at least as compelling as those from the 50 dimensional
+benchmark functions.
+
+\subsection{Model Fitting}
+\label{sec:model}
+
+For the model fitting problem that we experimented with, we used 144
+processors.  We show results for three methods: Naive with a Random topology,
+Naive with a Subswarm topology, and Many Iterations with a Random topology.
+Naive with a Subswarm topology had eight independent subswarms of 18
+particles each, and the particles in each subswarm were connected with a Random
+topology.  In the graphs we call this method Naive Subswarms.
+
+\fig{rbf} shows our results for this function.  The function value reported is
+the sum squared error of the model fit.  \fig{rbf-stuck} shows the percent of
+runs that reached a value for sum squared error of 55,000 by each time step,
+which we designate as successful.  Many Iterations took on average 126 time
+steps to reach this value, while Naive Subswarms took on average 298.  The
+p-value for a t-test on this statistic is less than $10^{-8}$.  With our
+implementation of parallel PSO, each time step took on average 10.06 seconds;
+1.83 seconds of that was function evaluation time and the rest was overhead.
+With this particular implementation, then, using Many Iterations instead of
+Naive led to an average time savings of 29 minutes, which is more than half
+of the 50 minutes that Naive takes on average to reach a successful value.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{rbf}
+  \caption{Results for fitting a radial basis function to noisy data.  We use
+  144 processors for each method, so each time step corresponds to 144 function
+  evaluations.}
+  \label{fig:rbf}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{rbf_stuck}
+  \caption{Results for fitting a radial basis function to noisy data.  We use
+  144 processors for each method, so each time step corresponds to 144 function
+  evaluations.}
+  \label{fig:rbf-stuck}
 \end{figure}
@@
 over previous methods.  We have shown results that conclusively demonstrate the
-superiority of our techniques for several functions over na\"ively adding
+superiority of our techniques for several functions over naively adding
 particles to the swarm when extra processors are available.
 
-What we have presented is not a new algorithm or a variant of PSO, and thus we
-did not compare our methods against the state of the art PSO variants.  We
-presented a new parallelization technique, so we compared parallelization
-strategies on the same algorithm, the original PSO.
+What we have presented is not a new variant of PSO.  We presented a new
+parallelization technique, so we compared parallelization strategies on the
+same algorithm, the original PSO.  Our most promising techniques do change the
+behavior of the PSO algorithm slightly, as do some other previously proposed
+parallelization techniques, such as asynchronous PSO.  However, our methods are
+applicable to almost all PSO variants, and so a comparison using the same
+variant for each of the parallelization techniques tested is justified.
 
@@
 given a brief evaluation of the premature convergence properties of these
-methods on deceptive functions.  Though there is no one approach that performs
-best across all functions at the swarm sizes we used, for all but one of the
-functions we looked at \emph{some} kind of speculative evaluation outperforms
-standard PSO, often dramatically.  It also appears to us that once enough
-processors are available, by far the best approach is to speculate as many
-iterations ahead as processors allow.
+methods on deceptive functions when a smaller number of processors is
+available.  We have also shown evidence that when many processors are available
+by far the best thing to do in most cases is to speculate as far ahead as the
+extra processors allow.
 
 Though our methods show great improvements on some functions, they do not work
-for all functions at the swarm sizes we were able to test with.  As is commonly
-known, in PSO there is a trade-off between exploration and exploitation.  Some
-functions need only minimal exploration, and some never seem to have enough.
-Increasing the swarm size is a natural way to increase exploration in a
-parallel environment.  However, once ``enough'' exploration has been reached
-for any particular function, adding additional particles adds only incremental
-benefits.  A better use of the additional processors, as we have shown, is to
-perform some amount of speculative evaluation.
-
-Sphere is a function for which only a very small amount of exploration needs to
-be done, and our speculative methods show improvements on Sphere of over 40
-orders of magnitude.  Our methods work very well on these functions.
-
-Rastrigin, on the other hand, is a function for which increasing the swarm size
-up to 4000 particles still shows improvements in the performance of the
-algorithm.  With such functions, the smaller swarm size required by speculative
-evaluation does not produce enough exploration to perform better than standard
-PSO at the swarm sizes we were able to experiment with.
-
-Griewank is a function in between Sphere and Rastrigin (as are Bohachevsky and
-Schwefel 2.21).  It is highly deceptive and is prone to premature convergence,
-but by adding particles to the swarm a point is reached where ``enough''
-exploration is done, and the algorithm finds the optimum essentially all of the
-time.  For such functions, the best approach seems to be to increase the swarm
-size until ``enough'' exploration is reached, then use extra processors to
-perform speculative evaluation and increase the number of iterations performed.
-Sphere and Rastrigin can be thought of as special cases of these types of
-functions; Sphere simply needs a very small swarm size to produce ``enough''
-exploration, and Rastrigin requires a very large swarm.  We expect that for all
-functions there is a swarm size for which additional particles are less useful
-than additional iterations.
+for all functions.  As is commonly known, in PSO there is a trade-off between
+exploration and exploitation.  Some functions need only minimal exploration,
+and some never seem to have enough.  Increasing the swarm size is a natural way
+to increase exploration in a parallel environment.  However, once ``enough''
+exploration has been reached for any particular function, adding additional
+particles adds only incremental benefits.  A better use of the additional
+processors, as we have shown, is to perform some amount of speculative
+evaluation.
 
@@
 increase the speed of the algorithm.  This work is a first step in that
-direction.
+direction that opens the door to many future improvements on speculative
+methods in the parallelization of PSO.
 
@@
 
-We have speculated that our Pick Best method simply draws a number of samples
-from the next iteration of the sampling distribution of PSO and keeps one of
-them.  It would then be expected that the sampling distribution of that method
-is very similar to the distribution of the original
-PSO~\citep{poli-2008-sampling-distribution-of-pso}.  This remains to be proven,
-or even demonstrated empirically.
-
-We gave a passing glance to some very interesting statistical characteristics
-of PSO.  Our motivation was to use the statistics to improve our speculative
-algorithm, but it seems there are some nuggets in the numbers that could be
-used to improve our understanding of the behavior of PSO in general, and what
-role a topology plays.  A potentially very fruitful avenue of future work would
-be to further analyze the branch statistics in PSO across many functions and
-topologies, categorizing both the topologies and the functions and discovering
-what it is that makes certain topologies work well with certain functions.
-There are also some particles that seem to drive the swarm and others that are
-stagnant almost the entire time.  Such information could be used in a
-TRIBES-like fashion~\citep{clerc-2003-tribes} to determine which particles
-should be removed from the swarm, perhaps starting them close to
-well-performing particles.
-
-We turned what used to be a very simple algorithm with only one major parameter
-to tweak (the topology) into a convoluted mess with countless possibilities for
-speculative evaluations.  Perhaps future research will be able to narrow down
-the endless possibilities into the few best ones, making the algorithm simple
-again.  We have presented some small amount of exploration of the possible
-alternatives, but our discussion is by no means exhaustive or conclusive.
+The issues of the sampling distribution of speculative relaxations, branch
+statistics, and why standard PSO fails at 500 dimensions were briefly mentioned
+in this paper.  Each of those issues needs further treatment.  The sampling
+distribution of our speculative methods could be compared to Poli's description
+of standard PSO's sampling
+distribution~\citep{poli-2008-sampling-distribution-of-pso}.  The branch
+statistics could be used to analyze topologies and discover why certain
+topologies work well on some functions but not on others; perhaps PSO
+performance is more dependent on the branch statistics of a combination of
+topology and function than on the topology itself.  And discovering exactly why
+constricted PSO fails at 500 dimensions while Many Iterations does not could
+lead to improvements in the standard PSO algorithm, even when not running in
+parallel.
+
+We opened the door to speculative parallelization methods in PSO and described
+the possible speculative evaluations to perform as an infinite tree from which
+branches are selected.  However, we only presented a few of the countless
+possibilities for selecting those branches.  Perhaps in Many Iterations it
+would be better to randomly sample branches from seventh iteration, instead
+of a fixed combination of the next three.  It remains as future work to do a
+more exhaustive comparison of possible pruning techniques.
 
@@
 
-A technique similar to ours has been previously used to parallelize simulated
-annealing~\citep{witte-1991-parallel-simulated-annealing-speculative}.  Any
-other optimization algorithm that only depends on current sampling positions
-when computing the next position to sample can be parallelized with this
-technique.  In particular, genetic algorithms produce future generations by
-combining individuals from the current generation.  With a large population
+Any other optimization algorithm that only depends on current sampling
+positions when computing the next position to sample can be parallelized with
+this technique.  In particular, genetic algorithms produce future generations
+by combining individuals from the current generation.  With a large population
 size there would be an unwieldy amount of possible future individuals, but the
 potential exists to modify the algorithm to use some kind of speculative
-evaluation.  A related but quite distinct technique for reducing computation in
-evolutionary algorithms was proposed
-in~\citep{poli-2006-backward-chaining-evolutionary-algorithms}.
+evaluation.
 

### 2011-03-24 masters_thesis/si_paper/reviewer_response/response.tex
Added a sentence to the reviewer response
Finished a draft of my thesis, and updated the reviewer response

--- 2011-02-10 honors_thesis/si_paper/reviewer_response/response.tex
@@
 Response: We added a Related Work section (now section 3), describing previous
-approaches at parallelizing PSO.  We also added (NOT DONE) experiments
-comparing against another common parallelization approach, that of using
-subswarms.
-
-NOT DONE
+approaches at parallelizing PSO.  As we note in the paper, most of the related
+work is orthogonal to innovations we describe, so it is difficult to determine
+what other methods to compare against.  The majority of previous papers on the
+parallelization of PSO deal with functions with very short evaluation times, so
+they propose topologies that limit communication between processors on
+different machines.  Those methods work very well for those functions, but when
+you are dealing with functions whose evaluation takes over a second, the
+benefits of the limited communication become negligible.  We did, however,
+experiment with one of the topologies proposed for fast running functions.  We
+found that in almost all cases simply using a Complete topology gave better
+performance.  We make note of all of this in the paper, mostly in section 3.
 
@@
 
-Response: Because all of the these finished before all of these, the t-test
-gave \ldots
-
-NOT DONE
+Response: We restructured most of our results section, and now include
+statistical significance in the form of bolded entries in a table.  This is
+noted in the paper.
 
@@
 
-Response: We added a paragraph to what is now Section 7 addressing the issue of
-benchmark functions (``We also do not report\ldots'').  We also ran a
-comparison on a longer-running function and included it in our results (NOT
-DONE).
-
-NOT DONE
+Response: We added a section dealing with this issue (section 7.1).  We also
+added experiments on a function that takes on the order of 2 seconds to
+evaluate, providing a little more motivation for the parallelization of PSO.
 
@@
 We also put subsections in the new Section 4 to guide the reader a little
-better through the section.
+better through the section, and gave an intuitive description of speculative
+evaluation before jumping into the math.
 
@@
 
-Response: We substantially reorganized Section 3, as noted above, and we moved
-the discussion of branch statistics (the old Section 4.3) to the beginning of
-the results section (NOT DONE).
-
-NOT DONE
+Response: We substantially reorganized Section 3, as noted above, and we added
+more subsections throughout the paper, helping to guide the reader through the
+ideas.  We believe that the paper is now much more clear.
 
@@
 Response: Rounds of evaluations correspond exactly to wall clock time, as we
-describe in the second paragraph of what was Section 5 (now Section 7).
-However, we realize that the term is not a commonly used one, so we changed
-``rounds of evaluations'' to ``time steps'' in hopes of better conveying our
-meaning.  We also added some text in the description mentioned above that
-hopefully further clarifies this point.
+describe in what is now section 7.1, and the second paragraph of section 7
+proper.  However, we realize that the term is not a commonly used one, so we
+changed ``rounds of evaluations'' to ``time steps'' in hopes of better
+conveying our meaning.  We also added some text in the description mentioned
+above that hopefully further clarifies this point.
 
@@
 
-Response: Running a few more sets of experiments at higher dimensionalities
-should address this adequately, though we can't really do all of their
-protocol.
+Response: We ran quite a few more experiments to bring our results closer to
+Herrera et al. experimental protocol.  Their protocol wasn't entirely
+applicable, as they suggested running a specific number of function evaluations
+instead of iterations (which, as we state in section 7, is the proper
+comparison for parallel algorithms on long running functions).  But we added
+experiments for 50 dimensional functions and 500 dimensional functions.
 
-NOT DONE
+We also see your point in saying that our relaxations to speculative evaluation
+are new PSO variants, and we fixed our text in places to better address that
+point.  In particular, we changed the paragraph in the conclusion that deals
+with this issue (``What we have presented\ldots''), clearly stating that our
+methods do in fact change the behavior of PSO, but that they do so in a way
+that is applicable to almost all PSO variants.  Thus a fair comparison is to
+pick a particular variant and compare parallelization techniques with it.  It
+made sense to pick the original PSO for this comparison.
 
@@
 
-Response: In fact the best result is improved, frequently, it is just that the
-scale of the improvements becomes increasingly smaller.
-
-NOT DONE
+Response: We were referring to the use of ``stagnation'' in analyses of the
+sampling distribution of PSO (such as by Poli).  Our results show that the
+``lack of improvement'' to which you refer does not in fact happen very often.
+It is just that the scale of the improvements becomes increasingly smaller,
+sometimes beyond machine precision.  We appreciate your distinction, however.
+Such a definition of stagnation must then incorporate a minimum velocity, which
+is another function-dependent parameter to specify.  We decided that the
+paragraph dealing with this issue was at best tangential to the focus of our
+paper and that it unnecessarily lengthened an already lengthy discussion.  We
+removed the paragraph from our text.
 
@@
 
-Response: Duly noted.  We fixed the sentence (it now says ``Our methods work
-very well on these functions'').
+Response: Duly noted.  We fixed the sentence, then in a later revision took out
+that discussion entirely, making the conclusion less repetitive.
 
@@
 
-Response: ...
-
-NOT DONE
+Response: We did not share your intuition about our methods improving PSO's
+performance on higher dimensional functions.  However, we ran experiments to
+test their scalability, as you suggested, and found them to perform very well
+as dimensionality increased.  We are not entirely sure as to the reason for
+this, though we gave some intuitive arguments based on brief experimentation.
+The results and discussion are in the paper.
 
@@
 
-Response: ...
-
-NOT DONE
+Response: You seem to have been conflating two separate issues, and we changed
+our text to help future readers not run into the same problem.  When we said
+that our method ``can be easily extended to arbitrary topologies,'' we were
+referring specifically to the math; we only showed one case, to simplify our
+notation.  In fact our implementation handles arbitrary topologies.  One thing
+that we realized, however, was that with pruning one can use arbitrarily dense
+topologies without increasing the number of speculative evaluations at all.  We
+thus ran more experiments to test the use of these topologies with pruned
+speculative methods, and we now report those results in the paper.
 
@@
 Response: The citation given in the paper (McNabb et al. CEC 09) gives an
-interesting discussion on these points.  The functions explored are limited,
-but the paper shows that both Sphere and Griewank exhibit this behavior, while
-Rastrigin does not (the relevant figures in that paper are 2, 4 and 6).  As
-those results are previously published and cited in our paper, we felt it
-unnecessary to reduplicate the figures, though we make mention of the
+interesting discussion on these points.  The functions explored there are
+limited, but the paper shows that both Sphere and Griewank exhibit this
+behavior, while Rastrigin does not (the relevant figures in that paper are 2, 4
+and 6).  As those results are previously published and cited in our paper, we
+felt it unnecessary to reduplicate the figures, though we make mention of the
 conclusions.  However, our own figures also answer your question.  The fact
@@
 and when that point is reached extra processors should be used for speculative
-evaluation.  We spend almost the entire conclusion trying to make this point.
+evaluation.  We spend a good amount of the conclusion trying to make this
+point.
 
@@
 second one to soften the point a little.  It was an over-reaction to comments
-by previous reviewers.
+by previous reviewers.  We also fixed a similar statement in the conclusion.
 

### 2011-03-24 masters_thesis/si_paper/paper.tex
Finished a draft of my thesis, and updated the reviewer response

@@
 
-Row 1 in Table~\ref{tab:branches}, \casepn, corresponds to the phenomenon
-commonly called stagnation in the literature, but with only a single particle.
-As an interesting aside, we found that complete stagnation (where the entire
-swarm is stagnant) practically never occurs in PSO, though frequently large
-percentages of the swarm are stagnant.  Iterations where the entire swarm is
-stagnant average less than 2 in 1000 iterations, unless the swarm converges
-beyond machine precision.  When graphs of the best function value flat-line,
-what really has happened is a contraction of the particles' velocity, not
-stagnation in the technical sense.
-
 We seek to find the probability of taking any given branch, given whatever
@@
 converged past machine precision after 500 iterations, and that led to
-erroneously high values for the probability of stagnation.  Instead we ran for
+erroneously high values for the probability of \casepn.  Instead we ran for
 only 450 iterations on those two combinations.  All of our results were
@@
 
-If experimentation with benchmark functions provides sufficient experimental
-evidence for algorithmic improvements in serial PSO, it is also sufficient
-evidence for algorithmic improvements in parallel PSO.  While the discussion of
-exact running times of function evaluations and communication overhead is
-certainly appropriate, even essential, to a discussion of specific parallel
-implementations, the concern of this paper is more fundamental than
-implementation decisions.  No one expects to see actual running times in place
-of function evaluations in a paper describing an innovation in the motion
-equations of PSO.  Similarly, actual running times should not be required of
-those describing algorithmic improvements in parallel PSO.  Just as in a serial
-PSO paper, actual running times cloud the more important issues of
-implementation-independent properties of the innovations described.
+While the discussion of exact running times of function evaluations and
+communication overhead is appropriate, even essential, to a discussion of
+specific parallel implementations, the concern of this paper is more
+fundamental than implementation decisions.  One would not expect to see actual
+running times in place of function evaluations in a paper describing an
+innovation in the motion equations of PSO.  Similarly, actual running times in
+a paper on algorithmic decisions in parallel PSO cloud the more important
+issues of implementation-independent properties of the innovations described.
 

### 2011-03-31 masters_thesis/si_paper/paper.tex
First half of fixes from Andrew and Kevin
Last pass through the paper before submitting it again.  Most of these
changes are pretty minor, though they change the rest of the paragraph
because of the way the formatting works.

@@
 distributing the computation corresponding to particles across multiple
-processors.  In these approaches, additional processors can only add more
-particles to the swarm.  However, in many cases this is not efficient when
-scaled to very large swarm sizes (on very large clusters).  Current methods
-cannot answer well the question, ``What do you do with 1000 processors when 50
-or 100 particles is the most efficient swarm size?''  In this paper we attempt
-to answer that question with a speculative approach to the parallelization of
-PSO motivated by speculative execution in modern computer hardware.
+processors.  In these approaches, adding processors can only increase the swarm
+size.  However, in many cases this is not efficient when scaled to very large
+swarm sizes (on very large clusters).  Current methods cannot answer well the
+question, ``How can 1000 processors be fully utilized when 50 or 100 particles
+is the most efficient swarm size?''  In this paper we attempt to answer that
+question with a speculative approach to the parallelization of PSO that we
+refer to as SEPSO.
 
 In our approach, we refactor PSO such that the computation needed for iteration
-$t+1$ can be done in parallel with the computation needed for iteration $t$.
+$t+1$ can be done concurrently with the computation needed for iteration $t$.
 Thus we can perform two iterations of PSO at once.  Even with some amount of
 wasted computation, we show that this approach to parallelization in PSO often
-outperforms the naive parallelization of simply adding particles to the
-swarm.  Our methods produce results that are exactly equivalent to PSO; this is
-not a new algorithm or variant, only a new method of parallelization.
+outperforms the naive parallelization of simply adding particles to the swarm.
+SEPSO produces results that are exactly equivalent to PSO; this is not a new
+algorithm or variant, only a new method of parallelization.
 
@@
 several such relaxations, including keeping the best speculative position
-evaluated instead of the one corresponding to PSO, and speculating several
-iterations ahead instead of just one.  We show that these methods dramatically
-improve the performance of parallel PSO in many cases.
+evaluated instead of the one corresponding to the standard behavior of PSO, and
+speculating several iterations ahead instead of just one.  We show that these
+methods dramatically improve the performance of parallel PSO in many cases.
 
@@
 The methods they have used include various synchronous algorithms
-\citep{chu-2006-intelligent-parallel-pso,jin-2005-pso-antenna-designs,%
-parsopoulos-2004-parallel-vector-evaluated-pso,%
-schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
-\citep{mostaghim-2006-multi-objective-pso-on-grids,%
-venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
-evaluation of the objective function can also be done in some cases, though
+\citep{parsopoulos-2004-parallel-vector-evaluated-pso} and asynchronous
+algorithms \citep{mostaghim-2006-multi-objective-pso-on-grids}.  Parallelizing
+the evaluation of the objective function can also be done in some cases, though
 that is not an adaption of the PSO algorithm itself and thus is not the focus
@@
 However, it can be seen from the graph that once the swarm is sufficiently
-large, doubling the swarm size produces only incremental benefit; performance
-improves, but not by very much.  The reason for this is that there comes a
-point of diminishing returns with respect to adding particles.  In
-\fig{evals-sphere} we show the value obtained after 50,000 function evaluations
-(not iterations) as a function of swarm size, again for the function Sphere.
-Increasing the swarm size from 5 to 10 has a significant effect on the value
-obtained.  However, increasing the swarm size from 16 to 30 makes the algorithm
-less efficient; that is, it reduces the progress the algorithm makes per
-evaluation.  Other functions show similar trends, though often the optimal
-swarm size is slightly larger.  For this reason, previous work has recommended
-the use of a swarm size of 50 for
+large, there comes a point of diminishing returns with respect to adding
+particles.  The increase in performance seen when moving from 50 to 100
+particles is roughly equivalent to the increase seen when moving from 1000 to
+4000.  In \fig{evals-sphere} we show the value obtained after 50,000 function
+evaluations (not iterations) as a function of swarm size, again for the
+function Sphere.  Increasing the swarm size from 5 to 10 has a significant
+effect on the value obtained.  However, increasing the swarm size from 16 to 30
+makes the algorithm less efficient; that is, it reduces the progress the
+algorithm makes per evaluation.  Other functions show similar trends, though
+often the optimal swarm size is slightly larger.  For this reason, previous
+work has recommended the use of a swarm size of 50 for
 PSO~\citep{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
@@
 
-In this paper we consider PSO parallelization strategies for clusters of
-hundreds or thousands of processors and functions for which a single evaluation
-will take at least tens of seconds, but probably minutes or perhaps hours.  Our
-purpose is to explore the question of what to do with a thousand processors
+Our purpose is to explore the question of what to do with a thousand processors
 when 50 or 100 particles is the most efficient swarm size, and simply adding
-particles results in only incremental improvement.  In our work we assume that
-the researcher has some fixed number of processors that should be fully
-utilized throughout the optimization process.
-
-In order to solve this problem, we apply the concept of speculative
-decomposition~\citep{grama-2003-intro-to-parallel-computing} to particle swarm
-optimization, using extra processors to perform two iterations of PSO at the
-same time.  Speculative decomposition is analogous to speculative execution
-(also known as branch prediction), a technique commonly used in processors.
-Modern processors, when faced with a branch on which they must wait (e.g., a
-memory cache miss), guess which way the branch will go and start executing,
-ensuring that any changes can be undone.  If the processor guesses right,
-execution is much farther ahead than if it had idly waited on the memory
+particles results in only incremental improvement.  We thus consider PSO
+parallelization strategies for clusters of hundreds or thousands of processors
+and functions for which a single evaluation will take long enough to merit a
+parallelization of one particle per processor---at least hundreds of
+milliseconds, but perhaps several minutes or longer.
+
+In order to solve the problem of diminishing returns, we apply the concept of
+speculative decomposition~\citep{grama-2003-intro-to-parallel-computing} to
+particle swarm optimization, using extra processors to perform two iterations
+of PSO at the same time.  Speculative decomposition is analogous to speculative
+execution (also known as branch prediction), a technique commonly used in
+processors.  Modern processors, when faced with a branch on which they must
+wait (e.g., a memory cache miss), guess which way the branch will go and start
+executing, ensuring that any changes can be undone.  If the processor guesses
+right, execution is much farther ahead than if it had idly waited on the memory
 reference.  If it guesses wrong, execution restarts where it would have been
@@
 \emph{exactly}, two iterations at a time, using a speculative approach adapted
-from speculative execution. We prove that the standard PSO equation can be
+from speculative execution. We show that the standard PSO equations can be
 factored such that a set of speculative positions can be found which will
@@
 than a typical or reasonable number of particles, producing better results in
-less ``wall-clock'' time.
-
-It is important to note here that this is not a variant of PSO.  We simply
-propose a new way to think about the parallelization of PSO that we show
-performs better in many cases than previous parallelizations.
+less wall-clock time.
+
+We refer to this parallelization technique as ``speculative evaluation in
+particle swarm optimization'', or SEPSO.  It is important to note here that
+SEPSO is not a variant of PSO.  We simply propose a new way to think about the
+parallelization of PSO that we show performs better in many cases than previous
+parallelizations.
 
@@
 longer demanding that it strictly reproduce the exact behavior of standard PSO,
-we can introduce new speculative techniques that often out-perform naive
-parallelizations of PSO.  These relaxations make better use of the information
-obtained from the extra exploration made by the speculative function
-evaluations.  We also explore the idea that, like branch prediction in
+we can introduce new speculative techniques that often out-perform both naive
+parallelizations of PSO and SEPSO.  These relaxations make better use of the
+information obtained from the extra exploration made by the speculative
+function evaluations.  We also explore the idea that, like branch prediction in
 processors, we need not speculatively evaluate \emph{all} possible future
@@
 particle swarm optimization algorithm, and \sec{related} gives a brief overview
-of previous parallelization techniques for this algorithm.  \sec{sepso}
-describes mathematically how speculative evaluation can be done in parallel PSO
-to perform two iterations at once, then \sec{implementation} gives information
-on how to implement this method.  In \sec{relax}, we discuss various methods of
+of previous parallelization techniques for this algorithm.  \sec{sepso} shows
+mathematically how speculative evaluation can be done in parallel PSO to
+perform two iterations at once, then \sec{implementation} gives information on
+how to implement this method.  In \sec{relax}, we discuss various methods of
 improving the performance of speculative evaluation in PSO, all of which break
@@
 The idea of speculative decomposition in the parallelization literature is not
-new~\citep{grama-2003-intro-to-parallel-computing}.  In the field of function
+new \citep{grama-2003-intro-to-parallel-computing}.  In the field of function
 optimization, simulated annealing has previously been parallelized using this
@@
 advocating using as many processors as possible to get the best performance.
-When the function takes less time to evaluate than the TCP/IP stack used to
-send interprocessor communication, parallel implementations assign several or
-many particles to a single processor~\citep{chu-2006-intelligent-parallel-pso,
+When the function takes far less time to evaluate than it takes to send a
+message across a network (e.g., through the TCP/IP stack), parallel
+implementations assign several or many particles to a single
+processor~\citep{chu-2006-intelligent-parallel-pso,
 chang-2005-parallel-pso-with-communication-strategies}.  Often the processor
@@
 
-Thus previous work in parallelizing PSO, apart from implementation details, has
-consisted entirely of increasing the swarm size and adapting the topology to be
-better suited to parallel computation.
+Thus previous work in parallelizing PSO, apart from creating innovative
+implementations, has consisted entirely of increasing the swarm size and
+adapting the topology to be better suited to parallel computation.
 
@@
 functions, and for large numbers of processors, just adding particles reaches a
-point of diminishing returns.  That is, adding processors does not help the
-algorithm reach any given level of fitness appreciably faster.  Instead of
-adding particles we employ a speculative approach that allows us to perform
-two iterations at a time.
+point of diminishing returns.  That is, beyond some point adding processors
+with previous techniques does not help the algorithm reach any given level of
+fitness significantly faster.  To fix this, instead of adding particles we
+employ extra processors in a speculative approach that allows us to perform two
+iterations at a time.
 
@@
 for each particle, thus using extra processors to evaluate two iterations of
-the algorithm in the time it takes to evaluate the function once.  As we show
-in Sections~\ref{sec:proof} and~\ref{sec:topology}, a wise choice of topology
-limits the necessary speculative evaluations to seven per particle.
+the algorithm in the time it takes to evaluate the function once.  As we will
+show in Sections~\ref{sec:proof} and~\ref{sec:topology}, a wise choice of
+topology limits the necessary speculative evaluations to seven per particle.
 
@@
 \sec{proof} shows in detail how the PSO equations can be refactored to allow
-for speculative evaluation, proving that our methods are correct and that they
-exactly reproduce the behavior of the PSO algorithm.  The section also
-introduces some notation used later in the paper.  \sec{topology} gives a brief
-discussion of how the topology used affects the amount of speculative
-computation needed.
+for speculative evaluation, proving that SEPSO exactly reproduces the behavior
+of the PSO algorithm.  The section also introduces some notation used later in
+the paper.  \sec{topology} gives a brief discussion of how the topology used
+affects the amount of speculative computation needed.
 
@@
 
-The refactoring hinges on the idea that once the random coefficients
-$\pURand_{t}$ and $\nURand_{t}$ are drawn, there are only a few possible new
-positions, or updates, for $\nbest$ and $\pbest$.  For the Ring topology there
-are 7 possible update cases, identified in Table~\ref{tab:evals}.  We label
-each case with an identifier referring to the source of the update: a minus
-sign ($-$) represents no update, $L$ represents an update to $\nbest$ coming
-from the left neighbor, $R$ represents an update to $\nbest$ coming from the
-right neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$
-coming from the particle itself.  As an example, $\casexn$ refers to the case
-that the particle finds a new personal best, but neither it nor its neighbors
-find a position that updated its neighborhood best.  In the equations that
-follow, we refer to an unspecified update case as $\casegen$, and to the set of
-cases collectively as $\caseset$.
+The refactoring hinges on the idea that there are only a few possible new
+positions, or updates, for $\nbest$ and $\pbest$ (assuming the random
+coefficients $\pURand_{t}$ and $\nURand_{t}$ have been drawn).  For the Ring
+topology there are 7 possible update cases, identified in
+Table~\ref{tab:evals}.  We label each case with an identifier referring to the
+source of the update: a minus sign ($-$) represents no update, $L$ represents
+an update to $\nbest$ coming from the left neighbor, $R$ represents an update
+to $\nbest$ coming from the right neighbor, and $S$ represents an update to
+either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
+$\casexn$ refers to the case that the particle finds a new personal best, but
+neither it nor its neighbors find a position that updated its neighborhood
+best.  In the equations that follow, we refer to an update case as $\casegen$,
+and to the set of cases collectively as $\caseset$.
 
@@
 $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$ ``speculative
-children'' because only one of them will be needed.
+children'' because only one of them is needed.
 
@@
 case in commonly used topologies, one of the cases can be eliminated, as a
-particle cannot be the source of an update to its neighborhood best while not
+particle cannot be the source of an update to its neighborhood best without
 also updating its personal best.  Thus we have $2(n+1)-1$, or $2n+1$,
@@
 in~\citep{mcnabb-2009-large-particle-swarms}, which has been shown to
-effectively simulate the Complete topology.  In~\sec{results} we report the
-results for performing speculative evaluation using both the Ring topology and
-the Random topology on a number of common benchmark functions.
+approximate the Complete topology.  In~\sec{results} we report results for
+SEPSO using both the Ring topology and the Random topology on a number of
+common benchmark functions.
 
@@
 other particles.  Finally we discuss the further complications of a dynamic
-topology, where a particle's neighbors change from one iteration to another.
+topology such as Random, where a particle's neighbors change from one iteration
+to another.
 
@@
 None of these methods fundamentally change the PSO algorithm.  They simply lead
-to particles having different values for personal and neighborhood best
-positions than would have occurred in standard PSO, because they receive
-different information.  This kind of relaxation is fairly typical in the
-parallelization of PSO~\citep{koh-2006-parallel-asynchronous-pso}.
+to particles being at different iterations and having different values for
+personal and neighborhood best positions than would have occurred in standard
+PSO, because they receive different information.  These kinds of relaxations
+are fairly typical in the parallelization of
+PSO~\citep{koh-2006-parallel-asynchronous-pso}.
 
@@
 
-A naive approach to pruning is to keep track of the last branch taken by each
-particle and speculate on that branch.  This turns out to be correct far less
-than half of the time on average.
-
-A more principled approach would be to use the statistical behavior of PSO to
-find probabilities of taking any particular branch.  While we cannot with
-certainty predict which branch a particle will take every time, if we can use
-statistics to narrow down the $2n+1$ possible evaluations to a few likely
-candidates, we can decrease the amount of computation required to do
-speculative evaluation and improve our performance.
+We look at the statistical behavior of PSO to find probabilities of taking any
+particular branch.  While we cannot with certainty predict which branch a
+particle will take every time, if we can use statistics to narrow down the
+$2n+1$ possible evaluations to a few likely candidates, we can decrease the
+amount of computation required to do speculative evaluation and improve our
+performance.
 
@@
   \scriptsize
-  \begin{tabular}{c|c|c|c|c|c|c}
+  \begin{tabular}{c|c|c|c|c|c|c|}
 	Topology&Function&\casepn&\casexn&\casexx&\casepN&\casexN\\
@@
 empirical distributions, the problem with this approach is that it is not clear
-what information influences the probability of taking a branch.  The failure of
-the naive approach seems to say that the previous branch, $\caseset_{t-1}$, is
-not incredibly useful.  We look at two factors that we believe have a
-significant influence on $\prob(\caseset_t)$: topology and function.  Thus we
-are looking at $\prob(\caseset_t|T,F)$.
+what information influences the probability of taking a branch.  We look at two
+factors that we believe have a significant influence on $\prob(\caseset_t)$:
+topology ($T$) and function ($F$).  Thus we are looking at
+$\prob(\caseset_t|T,F)$.
 
@@
 speculative evaluation while avoiding the explosion in the number of
-evaluations it would otherwise entail.  Though pruning makes its use possible,
-we still found the communication costs in our parallel environment prohibitive,
-so we do not present any results for the Complete topology here.
-
-With the other topologies, however, there are a few branches with very high
-probabilities, so we can have hope that cutting out some of the evaluations
+evaluations it would otherwise entail.
+
+Becuase there are a few branches with very high probabilities in the topologies
+we are interested in, we can have hope that cutting out some of the evaluations
 that have low probability will lead to an increase in overall performance.  In

### 2011-04-01 masters_thesis/si_paper/paper.tex
More changes from feedback.  Getting closer

@@
 wasted computation, we show that this approach to parallelization in PSO often
-outperforms the naive parallelization of simply adding particles to the swarm.
-SEPSO produces results that are exactly equivalent to PSO; this is not a new
-algorithm or variant, only a new method of parallelization.
+outperforms the standard parallelization of simply adding particles to the
+swarm.  SEPSO produces results that are exactly equivalent to PSO; this is not
+a new algorithm or variant, only a new method of parallelization.
 
@@
 longer demanding that it strictly reproduce the exact behavior of standard PSO,
-we can introduce new speculative techniques that often out-perform both naive
-parallelizations of PSO and SEPSO.  These relaxations make better use of the
-information obtained from the extra exploration made by the speculative
+we can introduce new speculative techniques that often out-perform both
+standard parallelizations of PSO and SEPSO.  These relaxations make better use
+of the information obtained from the extra exploration made by the speculative
 function evaluations.  We also explore the idea that, like branch prediction in
@@
 too many extra evaluations to produce competitive results.  On most functions,
-the larger swarm size that can be used in naive parallelizations leads to
+the larger swarm size that can be used in standard parallelizations leads to
 better performance than is found by trying to do speculative evaluation to
@@
 increase the swarm size relative to that needed in the original speculative
-algorithm (e.g., with 120 processors, a naive parallelization has a swarm of
+algorithm (e.g., with 120 processors, a standard parallelization has a swarm of
 size 120, complete speculative evaluation has a swarm of size 15, and pruning
@@
 We ran experiments to compare our speculative parallelization of PSO to the
-naive parallelization.  At each iteration of the algorithms, we use one
+standard parallelization.  At each iteration of the algorithms, we use one
 processor to perform one function evaluation for one particle, be it
@@
 architecture to use if the function being optimized takes longer than around
-100 milliseconds to run.
+100 milliseconds to run in our implementation.
 
@@
 Using the same number of processors for each approach requires that our
-speculative parallelizations have a smaller swarm size than the naive
+speculative parallelizations have a smaller swarm size than the standard
 parallelization.  For each of the topologies we used, a particle has three
@@
 extra evaluations required in our speculative approach would instead be used to
-evaluate particles in naive parallelizations, so we compare swarms of size
-$p$ in speculative evaluation with swarms of size $8p$ in naive approaches.
+evaluate particles in standard parallelizations, so we compare swarms of size
+$p$ in speculative evaluation with swarms of size $8p$ in standard approaches.
 
@@
 evaluation time and the overhead.  We wish to stress that the ``time steps'' we
-report are \emph{exactly} proportional to wall-clock time, given the
-architecture we have assumed.
+report are proportional to wall-clock time, given the architecture we have
+assumed.
 
@@
 
-The parallelization techniques we compare are the naive parallelization (here
-labeled Naive simply for ease of reference), our original speculative
+The parallelization techniques we compare are the standard parallelization
+(here labeled Standard simply for ease of reference), our original speculative
 approach (labeled SEPSO), and the four relaxations of SEPSO discussed in
@@
 with evaluating the original particle), we can use swarms of size
-$\frac{1}{3}p$ to compare to a swarm of size $p$ with the naive
+$\frac{1}{3}p$ to compare to a swarm of size $p$ with the standard
 parallelization, instead of $\frac{1}{8}p$ with other techniques.
@@
 distributed PSO, so it is still quite possible that Random would be used even
-with naive parallelizations~\cite{mcnabb-2009-large-particle-swarms}.  But,
-to be fair in our comparisons, we compare to the naive parallelization using
+with standard parallelizations~\cite{mcnabb-2009-large-particle-swarms}.  But,
+to be fair in our comparisons, we compare to the standard parallelization using
 both the Random topology and the Complete topology (labeled PSO Random and PSO
@@
 experiment we used 240 processors, so SEPSO and Pick Best each had 30
-particles, and the Naive algorithms had 240 particles.  As expected, SEPSO and
-Pick Best converge quickly to worse local optima than Naive does.
+particles, and the Standard algorithms had 240 particles.  As expected, SEPSO
+and Pick Best converge quickly to worse local optima than Standard does.
 \fig{rastrigin} shows the results graphically, and \tabref{rastrigin-20} shows
@@
   \hline
-  Naive Ring&25\%&743.4&212.0\\
-  \hline
-  Naive Random&100\%&372.1&86.6\\
-  \hline
-  \textbf{Naive Complete}&\textbf{95\%}&\textbf{273.1}&\textbf{99.7}\\
+  Standard Ring&25\%&743.4&212.0\\
+  \hline
+  Standard Random&100\%&372.1&86.6\\
+  \hline
+  \textbf{Standard Complete}&\textbf{95\%}&\textbf{273.1}&\textbf{99.7}\\
   \hline
@@
 using the Ring topology.  One can see in the figure that the SEPSO approach
-finds the optimum faster than the Naive approach.  However, because the swarm
-size is so small, SEPSO gets stuck a little less than half of the time.
+finds the optimum faster than the Standard approach.  However, because the
+swarm size is so small, SEPSO gets stuck a little less than half of the time.
 
@@
 Thus with a few more processors, our algorithm finds the optimum on average
-twice as fast as the Naive approach without a significant loss of accuracy.
+twice as fast as the Standard approach without a significant loss of accuracy.
 
@@
 Iterations finds the optimum, it finds it quicker than any other method we
-tried, on average four times faster than Naive.  However, it also gets stuck
+tried, on average four times faster than Standard.  However, it also gets stuck
 and fails to find the optimum more than any other method.  The results are
@@
   \hline
-  Naive Ring&0\%&N/A&N/A\\
-  \hline
-  Naive Random&0\%&N/A&N/A\\
-  \hline
-  Naive Complete&100\%&931.0&19.1\\
+  Standard Ring&0\%&N/A&N/A\\
+  \hline
+  Standard Random&0\%&N/A&N/A\\
+  \hline
+  Standard Complete&100\%&931.0&19.1\\
   \hline
@@
 more by larger swarms than Sphere is.  Thus our speculative algorithms often
-fail outperform Naive with 240 processors because the simple speculative
+fail outperform Standard with 240 processors because the simple speculative
 methods only have 30 particles.  As we will show later, when we use 800
@@
 pruned versions of our methods have 80 particles and thus are able to
-outperform Naive even with only 240 processors.  Because the graph of Schwefel
-2.21 looks very similar to that of Sphere, we present all of our results in
-\tabref{schwefel-20}.
+outperform Standard even with only 240 processors.  Because the graph of
+Schwefel 2.21 looks very similar to that of Sphere, we present all of our
+results in \tabref{schwefel-20}.
 
@@
   \hline
-  Naive Ring&0\%&N/A&N/A\\
-  \hline
-  Naive Random&0\%&N/A&N/A\\
-  \hline
-  Naive Complete&100\%&837.2&44.8\\
+  Standard Ring&0\%&N/A&N/A\\
+  \hline
+  Standard Random&0\%&N/A&N/A\\
+  \hline
+  Standard Complete&100\%&837.2&44.8\\
   \hline
@@
 methods, while \tabref{bohachevsky-20} shows all of the results.  All of our
-speculative approaches found the optimum much quicker than Naive with a Random
-topology.  However, SEPSO was slower than Naive Complete and got stuck 25\% of
-the time.  Pick Best, Pick Best Pruned, and Many Iterations all outperformed
-Naive Complete, with Many Iterations finding the optimum about twice as fast.
+speculative approaches found the optimum much quicker than Standard with a
+Random topology.  However, SEPSO was slower than Standard Complete and got
+stuck 25\% of the time.  Pick Best, Pick Best Pruned, and Many Iterations all
+outperformed Standard Complete, with Many Iterations finding the optimum about
+twice as fast.
 
@@
   \hline
-  Naive Random&100\%&472.3&6.6\\
-  \hline
-  Naive Complete&100\%&238.2&11.2\\
+  Standard Random&100\%&472.3&6.6\\
+  \hline
+  Standard Complete&100\%&238.2&11.2\\
   \hline
@@
 Iterations most often outperforms other speculative approaches, we only present
-results for Many Iterations and Naive for the 50 and 500 dimensional
+results for Many Iterations and Standard for the 50 and 500 dimensional
 variants.  For these experiments we used 800 processors, as the problems are
@@
 Rastrigin, though for all other benchmarks Many Iterations significantly
-outperformed Naive.  In each table, we bold the best method, meaning that it
+outperformed Standard.  In each table, we bold the best method, meaning that it
 has at least a 90\% success rate and its average time to completion is faster
@@
 Many Iterations Random, when it is successful, achieves the required value 6
-times faster than Naive Random.  However, because Many Iterations Random only
-has a 25\% success rate while Naive Random has a 100\% success rate, Naive
-Random is bolded.
+times faster than Standard Random.  However, because Many Iterations Random
+only has a 25\% success rate while Standard Random has a 100\% success rate,
+Standard Random is bolded.
 
@@
 A summary of the results is shown in \tabref{50dims-summary}.  For functions
-where Many Iterations outperformed Naive, we report the average speed up.  On
-all functions except Rastrigin, our methods showed an average speed up of from
-two to six times compared to previous methods.
+where Many Iterations outperformed Standard, we report the average speed up.
+On all functions except Rastrigin, our methods showed an average speed up of
+from two to six times compared to previous methods.
 
@@
   \hline
-  Naive Ring&5\%&776.0&0.0\\
-  \hline
-  \textbf{Naive Random}&\textbf{100\%}&\textbf{593.2}&\textbf{89.5}\\
-  \hline
-  Naive Complete&80\%&343.8&145.8\\
+  Standard Ring&5\%&776.0&0.0\\
+  \hline
+  \textbf{Standard Random}&\textbf{100\%}&\textbf{593.2}&\textbf{89.5}\\
+  \hline
+  Standard Complete&80\%&343.8&145.8\\
   \hline
@@
   \hline
-  Naive Ring&0\%&N/A&N/A\\
-  \hline
-  Naive Random&30\%&986.8&8.2\\
-  \hline
-  Naive Complete&100\%&458.4&12.3\\
+  Standard Ring&0\%&N/A&N/A\\
+  \hline
+  Standard Random&30\%&986.8&8.2\\
+  \hline
+  Standard Complete&100\%&458.4&12.3\\
   \hline
@@
   \hline
-  Naive Ring&0\%&N/A&N/A\\
-  \hline
-  Naive Random&70\%&871.6&57.1\\
-  \hline
-  Naive Complete&100\%&599.0&85.6\\
+  Standard Ring&0\%&N/A&N/A\\
+  \hline
+  Standard Random&70\%&871.6&57.1\\
+  \hline
+  Standard Complete&100\%&599.0&85.6\\
   \hline
@@
   \hline
-  Naive Ring&100\%&1898.5&25.7\\
+  Standard Ring&100\%&1898.5&25.7\\
   \hline
@@
   \hline
-  Naive Ring&100\%&1434.0&62.1\\
-  \hline
-  Naive Random&100\%&704.5&19.7\\
-  \hline
-  Naive Complete&20\%&377.5&14.9\\
+  Standard Ring&100\%&1434.0&62.1\\
+  \hline
+  Standard Random&100\%&704.5&19.7\\
+  \hline
+  Standard Complete&20\%&377.5&14.9\\
   \hline
@@
   for the best speculative topology to the time steps to completion for the
-  best naive topology.  All functions have 50 dimensions.}
+  best standard topology.  All functions have 50 dimensions.}
   \label{tab:50dims-summary}
@@
 At 500 dimensions we found that the performance of constricted PSO on benchmark
-functions becomes rather dismal.  In almost all cases, Naive fails to make
+functions becomes rather dismal.  In almost all cases, Standard fails to make
 any significant progress.  Our brief search for an explanation seems to say
@@
 
-The results that we present show that while Naive parallelizations suffer
+The results that we present show that while Standard parallelizations suffer
 from this problem, Many Iterations does not.  Though we can provide some
@@
 
-Because none of the Naive approaches were successful at optimizing the 500
+Because none of the Standard approaches were successful at optimizing the 500
 dimensional benchmark functions, we do not present tables as we did for the 50
@@
 \fig{griewank500} shows Griewank, and \fig{rastrigin500} shows Rastrigin.  Note
-that at 500 dimensions speculative approaches outperform naive
+that at 500 dimensions speculative approaches outperform standard
 parallelizations on Rastrigin, whereas they did not at 20 and 50 dimensions.
@@
 quickly when finding good locations.  However, we used the same motion
-equations for both Naive and Many Iterations; changing the motion equations
-to help Naive would also help Many Iterations, and we would expect to see
+equations for both Standard and Many Iterations; changing the motion equations
+to help Standard would also help Many Iterations, and we would expect to see
 results that are at least as compelling as those from the 50 dimensional
@@
 For the model fitting problem that we experimented with, we used 144
-processors.  We show results for three methods: Naive with a Random topology,
-Naive with a Subswarm topology, and Many Iterations with a Random topology.
-Naive with a Subswarm topology had eight independent subswarms of 18
-particles each, and the particles in each subswarm were connected with a Random
-topology.  In the graphs we call this method Naive Subswarms.
+processors.  We show results for three methods: Standard with a Random
+topology, Standard with a Subswarm topology, and Many Iterations with a Random
+topology.  Standard with a Subswarm topology had eight independent subswarms of
+18 particles each, and the particles in each subswarm were connected with a
+Random topology.  In the graphs we call this method Standard Subswarms.
 
@@
 which we designate as successful.  Many Iterations took on average 126 time
-steps to reach this value, while Naive Subswarms took on average 298.  The
+steps to reach this value, while Standard Subswarms took on average 298.  The
 p-value for a t-test on this statistic is less than $10^{-8}$.  With our
@@
 With this particular implementation, then, using Many Iterations instead of
-Naive led to an average time savings of 29 minutes, which is more than half
-of the 50 minutes that Naive takes on average to reach a successful value.
+Standard led to an average time savings of 29 minutes, which is more than half
+of the 50 minutes that Standard takes on average to reach a successful value.
 
@@
 over previous methods.  We have shown results that conclusively demonstrate the
-superiority of our techniques for several functions over naively adding
-particles to the swarm when extra processors are available.
+superiority of our techniques for several functions over the standard practice
+of adding particles to the swarm when extra processors are available.
 

### 2011-04-01 masters_thesis/si_paper/paper.tex
Removed some old files
Getting closer...

@@
 processor to perform one function evaluation for one particle, be it
-speculative or non-speculative.  The speculative version actually performs two
-iterations of PSO at each ``iteration,'' so we instead call each ``iteration''
-a ``time step.''  As noted in the introduction, this is a reasonable
-architecture to use if the function being optimized takes longer than around
-100 milliseconds to run in our implementation.
-
-Because our interest is in exploring the most efficient use of processors, we
-must hold fixed the number of processors used at each time step in all of our
-experiments, giving all approaches the same number of function evaluations.
-Thus while we label our graphs with ``Time Steps,'' this is equivalent to the
-number of function evaluations used by each approach (simply multiply by the
-number of processors to transform time steps to function evaluations).  We use
-``Time Steps'' to reflect wall-clock time, as our primary interest is using
-extra processors in parallel to improve performance on functions with long
-evaluation times.  A smaller swarm running serial PSO would likely produce
-better results than we present in the same number of function evaluations, but
-because the function evaluations were not done in parallel it would take more
-wall-clock time.  When comparing the performance of parallelized algorithms on
-long-running functions, we must compare by wall-clock time with the same
-resources, not solely by the number of function evaluations.
-
-Using the same number of processors for each approach requires that our
-speculative parallelizations have a smaller swarm size than the standard
-parallelization.  For each of the topologies we used, a particle has three
+speculative or non-speculative.  The exact evaluation time at which this
+architecture becomes reasonable depends on the amount of communication overhead
+in the parallel implementation and the number of particles in the swarm.  For
+our implementation we found that time to be around 100 milliseconds for swarms
+of 50 particles.  When the swarm size increases, the minimum evaluation time at
+which this architecture should be used decreases.
+
+\subsection{Function Evaluations vs. Time Steps}
+
+Results in serial PSO are typically presented in terms of function evaluations.
+This is a natural abstraction from implementation details that still allows a
+comparison of the implementation-independent aspects of each algorithm.  The
+number of function evaluations performed is assumed to be proportional to
+wall-clock time, as all evaluations are performed in serial.  Only reporting
+function evaluations could hide the fact that one algorithm requires more
+overhead than another and thus actually takes more time to perform the same
+number of function evaluations; however, function evaluations are still
+considered the standard method of reporting, as evaluation times for functions
+also vary greatly and could make the additional overhead negligible.
+
+In parallel PSO on long-running functions, the natural way to present results
+is in terms of iterations, not function evaluations.  This is because when all
+function evaluations at each iteration are performed concurrently, iterations
+are the direct equivalent of wall-clock time.  It is true that with a fixed
+number of processors, function evaluations are also equivalent to wall-clock
+time.  However, reporting function evaluations in a paper on parallel PSO can
+be misleading if compared to a paper on serial PSO, as the actual wall-clock
+times corresponding to the function evaluations are not directly comparable.
+Thus we report iterations in our results instead of function evaluations.  But
+because SEPSO actually performs two iterations of PSO at each ``iteration,''
+and Social Promotion and Many Iterations make the idea of ``iterations of PSO''
+somewhat nebulus, we instead call each ``iteration'' a ``time step.''
+
+Just as serial PSO papers do not report actual running times of their specific
+implementations, we do not report running times, favoring the abstraction of
+time steps.  That being said, given the time required for the evaluation of the
+objective function and the communication overhead per iteration for a specific
+implementation, an actual running time can be recovered from all of our
+results.  Simply multiply the number of time steps by the sum of the function
+evaluation time and the overhead.  We wish to stress that the ``time steps'' we
+report are proportional to wall-clock time, given the architecture we have
+assumed.
+
+\subsection{Functions Used}
+
+We experimented with five common benchmark problems defined
+in~\citep{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
+and Bohachevsky.
+
+Rastrigin is initialized in $[-5.12,5.12]^D$ and is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
+10\right), z=x-c\]
+
+Sphere is initialized in $[-50,50]^D$ and is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]
+
+Schwefel 2.21 is initialized in $[-500,500]^D$ and is defined as:
+\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]
+
+Griewank is initialized in $[-600,600]^D$ and is defined as:
+\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
+\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]
+
+And Bohachevsky is initialized in $[-15,15]^D$ and is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
+x_{i+1})+.7)\]
+
+In computing the branch statistics in \sec{statistics}, we also used the
+Rosenbrock function.  That function is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)\]
+
+The $c$ in the function definitions is a shifted center, in order to avoid
+origin-seeking bias in the PSO algorithm and its
+variants~\citep{monson-2005-origin-seeking-bias}.  We move the center of the
+shifted functions (all except Bohachevsky) to be halfway between the center and
+the boundary of the initialization region.  For example, if the initialization
+region is $[-50,50]^D$, the center is $(25)^D$.  We tested all of these
+functions in their 20 dimension, 50 dimension, and 500 dimension varieties.
+
+One might question the use of benchmark functions for evaluating our methods,
+as they take on the order of 1 millisecond to evaluate instead of the 100s that
+we said were required to justify our parallel architecture.  The purpose of
+exploring the behavior of an algorithm on benchmark functions, whether in
+serial or in parallel optimization research, is because benchmark functions are
+believed to represent functions that one might encounter in the real world,
+regardless of the amount of time those functions take to evaluate.  While
+benchmark functions themselves take fractions of a second to evaluate and thus
+have no need of the kind of parallelization we explore in this paper, they also
+have no need of particle swarm optimization at all, as they can be solved
+analytically.  Yet they are useful to us and to optimization researchers
+generally because they stand as surrogates for the kinds of functions
+practitioners are actually interested in, and they allow us to explore the
+behavior of optimization algorithms on several different classes of functions
+in an easy and standardized way.
+
+To aid in convincing the reader that performance on benchmark functions roughly
+corresponds to performance on real-world problems, we also tested our
+parallelization methods on a typical research problem, that of fitting a model
+to a large quantity of data.  We generated 10,000 data points from a radial
+basis function network with 10 bases and some added Gaussian noise.  We then
+fit a radial basis function network to the data using PSO.  This amounted to a
+30 dimensional function to optimize, with a function evaluation time on the
+order of two seconds.  We refer to this function as ``the model fitting
+problem.''
+
+\subsection{Parallelization Techniques Compared}
+
+The parallelization techniques we compare are the standard parallelization
+(here labeled Standard simply for ease of reference), our original speculative
+approach (recall that we refer to this as SEPSO), and the four relaxations of
+SEPSO discussed in \sec{relax}.  In presenting our results, we call the
+approach developed in \sec{pickbest} Pick Best.  The methods described in
+\sec{pruning} through \sec{wrong} are called Pick Best Pruned and Social
+Promotion Pruned, and the method in \sec{manyiters} we call Many Iterations.
+
+Using the same number of processors for each approach (and thus the same number
+of function evaluations per time step) requires that our speculative
+parallelizations have a smaller swarm size than the standard parallelization.
+For the topologies we used with SEPSO and Pick Best, a particle has three
 neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
@@
 $p$ in speculative evaluation with swarms of size $8p$ in standard approaches.
-
-\subsection{Running Time}
-
-We do not report absolute time used by each algorithm, as specific time values
-are heavily implementation-dependent.  Using time steps, where at every time
-step each processor evaluates the function once, is an abstraction from
-individual implementations that still conveys all of the information necessary
-to compare algorithms.  In this paper we focus on the effects of algorithm
-choices on parallel PSO performance.  As there are inexpensive functions with
-complex landscapes and expensive functions with simple landscapes, the behavior
-of PSO with respect to the function is the main issue.
-
-For this purpose, benchmark functions are a useful and efficient tool for
-understanding the effects of PSO with expensive objective functions even though
-the benchmark functions are themselves inexpensive.  While benchmark functions
-themselves take fractions of a second to evaluate and thus have no need of the
-kind of parallelization we explore in this paper, they also have no need of
-particle swarm optimization at all, as they can be solved analytically.  Yet
-they are useful to us and to optimization researchers generally because they
-stand as surrogates for the kinds of functions practitioners are actually
-interested in, and they allow us to explore the behavior of optimization
-algorithms on several different classes of functions in an easy and
-standardized way.
-
-While the discussion of exact running times of function evaluations and
-communication overhead is appropriate, even essential, to a discussion of
-specific parallel implementations, the concern of this paper is more
-fundamental than implementation decisions.  One would not expect to see actual
-running times in place of function evaluations in a paper describing an
-innovation in the motion equations of PSO.  Similarly, actual running times in
-a paper on algorithmic decisions in parallel PSO cloud the more important
-issues of implementation-independent properties of the innovations described.
-
-That being said, given the time required for the evaluation of the objective
-function and the communication overhead per iteration for a specific
-implementation, the actual running time can be recovered from all of our
-results.  Simply multiply the number of time steps by the sum of the function
-evaluation time and the overhead.  We wish to stress that the ``time steps'' we
-report are proportional to wall-clock time, given the architecture we have
-assumed.
-
-\subsection{Functions Used}
-
-We experimented with five common benchmark problems defined
-in~\citep{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
-and Bohachevsky.
-
-Rastrigin is initialized in $[-5.12,5.12]^D$ and is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
-10\right), z=x-c\]
-
-Sphere is initialized in $[-50,50]^D$ and is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]
-
-Schwefel 2.21 is initialized in $[-500,500]^D$ and is defined as:
-\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]
-
-Griewank is initialized in $[-600,600]^D$ and is defined as:
-\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
-\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]
-
-And Bohachevsky is initialized in $[-15,15]^D$ and is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
-x_{i+1})+.7)\]
-
-In computing the branch statistics in \sec{statistics}, we also used the
-Rosenbrock function.  That function is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)\]
-
-The $c$ in the function definitions is a shifted center, in order to avoid
-origin-seeking bias in the PSO algorithm and its
-variants~\citep{monson-2005-origin-seeking-bias}.  We move the center of the
-shifted functions (all except Bohachevsky) to be halfway between the center and
-the boundary of the initialization region.  For example, if the initialization
-region is $[-50,50]^D$, the center is $(25)^D$.  We tested all of these
-functions in their 20 dimension, 50 dimension, and 500 dimension varieties.
-
-We also tested the parallelization methods on a real problem, that of fitting a
-model to a large quantity of data.  We generated 10,000 data points from a
-radial basis function network with 10 bases and some added Gaussian noise.  We
-then fit a radial basis function network to the data using PSO.  This amounted
-to a 30 dimensional function to optimize, with a function evaluation time on
-the order of two seconds.
-
-\subsection{Parallelization Techniques Compared}
-
-The parallelization techniques we compare are the standard parallelization
-(here labeled Standard simply for ease of reference), our original speculative
-approach (labeled SEPSO), and the four relaxations of SEPSO discussed in
-\sec{relax}.  In presenting our results, we call the approach developed in
-\sec{pickbest} Pick Best.  The methods described in \sec{pruning} through
-\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
-method in \sec{manyiters} we call Many Iterations.
 
@@
 with evaluating the original particle), we can use swarms of size
-$\frac{1}{3}p$ to compare to a swarm of size $p$ with the standard
+$\frac{1}{3}p$ when pruning to compare to a swarm of size $p$ with the standard
 parallelization, instead of $\frac{1}{8}p$ with other techniques.
@@
 limit ourselves to the Ring topology and the Complete topology, as is common
-practice, along with the parallel approximation to the Complete topology
-mentioned in \sec{pso}.  We also mentioned in \sec{related} that some related
-work can be described as changes in topology, particularly that of having
-subswarms of fully connected topologies that occasionally communicate with
-each other.  This related work focused mainly on functions with very fast
-evaluation times where such techniques drastically reduce interprocessor
-communication.  With long function evaluations these topologies are not as
-practical, as only one particle is evaluated by each processor at each
-iteration.  We experimented with a subswarm topology and found that in every
-instance except one it performed worse than either Complete or Ring, and thus
-we leave it out of the results except in the one instance where it improved
-performance.
+practice, along with the Random topology (the parallel approximation to the
+Complete topology mentioned in \sec{pso}).  We also mentioned in \sec{related}
+that some related work can be described as changes in topology, particularly
+that of having subswarms of fully connected topologies that occasionally
+communicate with each other.  This related work focused mainly on functions
+with very fast evaluation times where such techniques drastically reduce
+interprocessor communication.  With long function evaluations these topologies
+are not as practical, as only one particle is evaluated by each processor at
+each iteration.  We experimented with a subswarm topology and found that in
+every instance except one it performed worse than either Complete or Ring, and
+thus we leave it out of the results except in the one instance where it
+improved performance.
 
@@
 distributed PSO, so it is still quite possible that Random would be used even
-with standard parallelizations~\cite{mcnabb-2009-large-particle-swarms}.  But,
+with standard parallelizations~\citep{mcnabb-2009-large-particle-swarms}.  But,
 to be fair in our comparisons, we compare to the standard parallelization using
@@
 heavily dependent on implementation details which are not the focus of this
-paper.  A practitioner using a particular implementation can compare results
-for whichever topology is most practical given the specific implementation.
+paper.  A practitioner using a particular implementation can compare the
+results given here for whichever topology is most practical given the specific
+implementation.
 
@@
 \label{sec:results}
+
+We frequently present tables summarizing our results.  In each table, we bold
+the ``best'' method, meaning that it has at least a 90\% success rate and its
+average time to completion is faster than all other methods that have at least
+a 90\% success rate.  We used a double-sided t-test to test for statistical
+significance in these results, and all bolded values are statistically
+significant with a p-value of less than $10^{-5}$.  In some instances two
+values are bolded because they are statistically tied, while both are
+significantly better than all other methods.
 
@@
 
+\subsubsection{Sphere}
+
+First we look at Sphere, the simplest of common benchmark functions.  The
+function has a single global optimum and no other local optima.  Sphere is best
+optimized in terms of function evaluations with a small swarm using a Complete
+topology.  We expect our methods to be perfect for such functions, and our
+results show this intuition to be correct.  For this comparison we used 240
+processors, so the methods had swarms of size 30 (for SEPSO, Pick Best, and
+Many Iterations), 80 (Pick Best Pruned and Social Promotion Pruned), and 240
+(Standard).  We compared too many methods here to fit into one graph, so we
+show just a few methods in \fig{sphere-20} and in \tabref{sphere-20} we show a
+summary of the results for all methods we tested.
+
+In \fig{sphere-20} we see that Many Iterations using a Complete topology
+converges incredibly quickly on a very poor value.  We found this behavior to
+be quite consistent across functions for this method, so we rarely show results
+for Many Iterations Complete.  However, Pick Best Pruned works very well with
+Complete on this function.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{sphere20}
+  \caption{Function Sphere with a swarm that uses 240 processors per time step,
+  comparing a pertinent subset of possible methods.}
+  \label{fig:sphere-20}
+\end{figure}
+
+\begin{table}
+  \caption{Summary of results for function Sphere with 20 dimensions,
+  measuring number of time steps to reach a value of 1e-35}
+  \label{tab:sphere-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&0\%&N/A&N/A\\
+  \hline
+  Standard Random&0\%&N/A&N/A\\
+  \hline
+  Standard Complete&100\%&931.0&19.1\\
+  \hline
+  SEPSO Ring&0\%&N/A&N/A\\
+  \hline
+  SEPSO Random&85\%&972.8&15.8\\
+  \hline
+  Pick Best Ring&100\%&917.4&21.0\\
+  \hline
+  Pick Best Random&100\%&768.1&14.3\\
+  \hline
+  Pick Best Pruned Ring&100\%&967.8&9.9\\
+  \hline
+  Pick Best Pruned Random&100\%&693.5&10.4\\
+  \hline
+  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{389.6}&
+    \textbf{9.6}\\
+  \hline
+  Social Promotion Pruned Ring&0\%&N/A&N/A\\
+  \hline
+  Social Promotion Pruned Random&100\%&971.1&10.6\\
+  \hline
+  Social Promotion Pruned Complete&100\%&777.8&16.7\\
+  \hline
+  Many Iterations Ring&100\%&575.2&9.8\\
+  \hline
+  Many Iterations Random&100\%&442.4&7.1\\
+  \hline
+  Many Iterations Complete&0\%&N/A&N/A\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\subsubsection{Schwefel 2.21}
+
+Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
+more by larger swarms than Sphere is.  Thus our speculative algorithms often
+fail to outperform Standard with 240 processors because the simple speculative
+methods only have 30 particles.  As we will show later, when we use 800
+processors at 50 dimensions, our methods perform much better.  However, the
+pruned versions of our methods have 80 particles and thus are able to
+outperform Standard even with only 240 processors.  Because the graph of
+Schwefel 2.21 looks very similar to that of Sphere, we simply present all of
+our results in \tabref{schwefel-20}.
+
+\begin{table}
+  \caption{Summary of results for function Schwefel with 20 dimensions,
+  measuring number of time steps to reach a value of 1e-06}
+  \label{tab:schwefel-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&0\%&N/A&N/A\\
+  \hline
+  Standard Random&0\%&N/A&N/A\\
+  \hline
+  Standard Complete&100\%&837.2&44.8\\
+  \hline
+  SEPSO Random&0\%&N/A&N/A\\
+  \hline
+  Pick Best Random&5\%&938.0&0.0\\
+  \hline
+  \textbf{Pick Best Pruned Random}&\textbf{100\%}&\textbf{639.9}&\textbf{29.4}\\
+  \hline
+  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{597.7}&
+	\textbf{77.9}\\
+  \hline
+  Social Promotion Pruned Random&75\%&938.0&33.0\\
+  \hline
+  Social Promotion Pruned Complete&100\%&815.2&43.5\\
+  \hline
+  Many Iterations Random&100\%&783.5&94.5\\
+  \hline
+  Many Iterations Complete&0\%&N/A&N/A\\
+  \hline
+  \end{tabular}
+\end{table}
+
 \subsubsection{Rastrigin}
@@
 particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
-in local optima.  Because our speculative algorithms require significantly
-smaller swarm sizes, we would expect to not perform very well on functions such
-as Rastrigin.  Our experiments show our intuition to be correct.  In this
-experiment we used 240 processors, so SEPSO and Pick Best each had 30
-particles, and the Standard algorithms had 240 particles.  As expected, SEPSO
-and Pick Best converge quickly to worse local optima than Standard does.
-\fig{rastrigin} shows the results graphically, and \tabref{rastrigin-20} shows
-results for all of the methods we tried.
+in local optima and converge to poorer values.  Because our speculative
+algorithms require significantly smaller swarm sizes, we would expect to not
+perform very well on functions such as Rastrigin.  Our experiments show our
+intuition to be correct.  In this experiment we used 240 processors, so SEPSO
+and Pick Best each had 30 particles, and the Standard algorithms had 240
+particles.  As expected, SEPSO and Pick Best converge quickly to worse local
+optima than Standard does.  \fig{rastrigin} shows the results graphically, and
+\tabref{rastrigin-20} shows results for all of the methods we tried.
 
@@
 misleading.  Thus instead we present plots showing the percent of trials that
-have passed some threshold at each time step.  The threshold we chose for this
-case was $10^{-6}$, as that value is below any local optima and the swarm
-always successfully reaches 0 once is passes that point.
+have passed some threshold at each time step, as is common practice with these
+functions~\citep{mendes-2004-population-topologies-in-pso}.  The threshold we
+chose for this case was $10^{-6}$, as that value is below any local optima and
+the swarm always successfully reaches 0 once is passes that point.
 
 We ran 50 trials of each experiment with Griewank, so that the curves are more
-smooth.  We show results in \fig{basic-griewank1} for swarms of size 30 and 240
-using the Ring topology.  One can see in the figure that the SEPSO approach
-finds the optimum faster than the Standard approach.  However, because the
-swarm size is so small, SEPSO gets stuck a little less than half of the time.
+smooth.  We show results in \fig{basic-griewank1} for swarms of size 100 and
+800 using the Ring topology.  One can see in the figure that SEPSO reaches the
+global optimum on average close to twice as fast as Standard, while Pick Best
+is close to three times as fast and Many Iterations is even faster when it is
+successful, though it is only successful 84\% of the time.  \tabref{griewank}
+shows all of our results in tabular form.
 
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{griewank1}
-  \caption{Function Griewank with a swarm that uses 240 processors per time
+  \includegraphics[width=.7\columnwidth]{griewank6}
+  \caption{Function Griewank with a swarm that uses 800 processors per time
   step.  Recall that at each time step all processors evaluate the function
@@
 
-Because the speculative algorithm got stuck with such a small swarm size, we
-ran a set of experiments on Griewank with 800 processors, giving swarms of size
-100 and 800 instead of 30 and 240.  \fig{basic-griewank2} shows the results.
-\fig{basic-griewank2} is very similar to \fig{basic-griewank1}, except that
-speculative evaluation finds the global optimum essentially 100\% of the time.
-Thus with a few more processors, our algorithm finds the optimum on average
-twice as fast as the Standard approach without a significant loss of accuracy.
+\begin{table}
+  \caption{Summary of results for function Griewank with 20 dimensions,
+  measuring number of time steps to reach a value of 1e-06}
+  \label{tab:griewank-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&100\%&762.2&19.4\\
+  \hline
+  SEPSO Ring&100\%&426.0&81.0\\
+  \hline
+  \textbf{Pick Best Ring}&\textbf{100\%}&\textbf{272.0}&\textbf{17.7}\\
+  \hline
+  \textbf{Pick Best Pruned Ring}&\textbf{100\%}&\textbf{282.9}&\textbf{10.7}\\
+  \hline
+  Social Promotion Pruned Ring&100\%&482.0&16.6\\
+  \hline
+  Many Iterations Ring&84\%&183.2&49.4\\
+  \hline
+  \end{tabular}
+\end{table}
+
+We pause here to show some interesting characteristics of the speculative
+techniques we have proposed.  With 800 processors our methods perform very well
+on this function.  With 240 processors, however, the results are much more
+mixed.  Because 240 processors is near the point at which speculative
+evaluation becomes useful, it is enlightening to see the behavior of the
+various algorithms at this point.
+
+We show results in \fig{basic-griewank2} for swarms of size 30 and 240 using
+the Ring topology.  One can see in the figure that when SEPSO is successful, it
+finds the optimum much faster than Standard.  However, because the swarm size
+is so small, SEPSO gets stuck a little less than half of the time.
 
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{griewank2}
-  \caption{Function Griewank with a swarm that uses 800 processors per time
+  \includegraphics[width=.7\columnwidth]{griewank1}
+  \caption{Function Griewank with a swarm that uses 240 processors per time
   step.  All algorithms use the Ring topology.}
@@
 \end{figure}
-
-In order to show the premature convergence properties of the various approaches
-we have presented, we will show results first with swarms using 240 processors,
-where these are more visible.  Then we will present an overall result for 800
-processors, where using our techniques is generally more successful.
 
@@
   \caption{Function Griewank with a swarm that uses 240 processors per time
-  step, comparing Pick Best with previous results.  All algorithms use the Ring
-  topology.}
+  step, comparing Pick Best with results from \fig{basic-griewank2}.  All
+  algorithms use the Ring topology.}
   \label{fig:griewank-pickbest}
@@
   \caption{Function Griewank with a swarm that uses 240 processors per time
-  step, comparing pruning with previous results.  All algorithms use the Ring
-  topology.}
+  step, comparing pruning with the best results from \fig{griewank-pickbest}.
+  All algorithms use the Ring topology.}
   \label{fig:griewank-pruned}
@@
   \caption{Function Griewank with a swarm that uses 240 processors per time
-  step, comparing speculating several iterations ahead with previous results.
-  All algorithms use the Ring topology.}
+  step, comparing speculating several iterations ahead with selections from
+  \fig{griewank-pruned}.  All algorithms use the Ring topology.}
   \label{fig:griewank-manyiters}
 \end{figure}
-
-Now that we have seen the effect that the various approaches have on speed of
-convergence and success rates, we show a result for using 800 processors, where
-most of the approaches have a 100\% success rate.  In this case, the best
-algorithm to use is the one which finds the optimum the fastest.  As can be
-seen in \fig{griewank-800}, with 800 processors the best choice is Pick Best,
-which finds the optimum on average three times as fast as standard PSO.  With
-just a few more processors, however, Many Iterations would likely have a 100\%
-success rate as well, making it the preferred choice.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{griewank6}
-  \caption{Function Griewank with a swarm that uses 800 processors per time
-  step.  All algorithms use the Ring topology.}
-  \label{fig:griewank-800}
-\end{figure}
-
-\subsubsection{Sphere}
-
-Next we go to Sphere, the simplest of common benchmark functions.  The function
-has no local optima (only a single global optimum) and is best optimized in
-terms of function evaluations with a small swarm using a Complete topology.  We
-expect our methods to be perfect for such functions, and our results show this
-intuition to be correct.  We compared too many methods here to fit into one
-graph, so we show just a few methods in \fig{sphere-20} and in
-\tabref{sphere-20} we show a summary of the results for all methods we tested.
-
-In \fig{sphere-20} we see that Many Iterations using a Complete topology
-converges incredibly quickly on a very poor value.  We found this behavior to
-be quite consistent across functions, so we rarely show results for Many
-Iterations Complete.  However, Pick Best Pruned works very well with Complete
-on this function.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{sphere20}
-  \caption{Function Sphere with a swarm that uses 240 processors per time step,
-  comparing a pertinent subset of possible methods.}
-  \label{fig:sphere-20}
-\end{figure}
-
-\begin{table}
-  \caption{Summary of results for function Sphere with 20 dimensions,
-  measuring number of time steps to reach a value of 1e-35}
-  \label{tab:sphere-20}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&0\%&N/A&N/A\\
-  \hline
-  Standard Random&0\%&N/A&N/A\\
-  \hline
-  Standard Complete&100\%&931.0&19.1\\
-  \hline
-  SEPSO Ring&0\%&N/A&N/A\\
-  \hline
-  SEPSO Random&85\%&972.8&15.8\\
-  \hline
-  Pick Best Ring&100\%&917.4&21.0\\
-  \hline
-  Pick Best Random&100\%&768.1&14.3\\
-  \hline
-  Pick Best Pruned Ring&100\%&967.8&9.9\\
-  \hline
-  Pick Best Pruned Random&100\%&693.5&10.4\\
-  \hline
-  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{389.6}&
-    \textbf{9.6}\\
-  \hline
-  Social Promotion Pruned Ring&0\%&N/A&N/A\\
-  \hline
-  Social Promotion Pruned Random&100\%&971.1&10.6\\
-  \hline
-  Social Promotion Pruned Complete&100\%&777.8&16.7\\
-  \hline
-  Many Iterations Ring&100\%&575.2&9.8\\
-  \hline
-  Many Iterations Random&100\%&442.4&7.1\\
-  \hline
-  Many Iterations Complete&0\%&N/A&N/A\\
-  \hline
-  \end{tabular}
-\end{table}
-
-\subsubsection{Schwefel 2.21}
-
-Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
-more by larger swarms than Sphere is.  Thus our speculative algorithms often
-fail outperform Standard with 240 processors because the simple speculative
-methods only have 30 particles.  As we will show later, when we use 800
-processors at 50 dimensions, our methods perform much better.  However, the
-pruned versions of our methods have 80 particles and thus are able to
-outperform Standard even with only 240 processors.  Because the graph of
-Schwefel 2.21 looks very similar to that of Sphere, we present all of our
-results in \tabref{schwefel-20}.
-
-\begin{table}
-  \caption{Summary of results for function Schwefel with 20 dimensions,
-  measuring number of time steps to reach a value of 1e-06}
-  \label{tab:schwefel-20}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&0\%&N/A&N/A\\
-  \hline
-  Standard Random&0\%&N/A&N/A\\
-  \hline
-  Standard Complete&100\%&837.2&44.8\\
-  \hline
-  SEPSO Random&0\%&N/A&N/A\\
-  \hline
-  Pick Best Random&5\%&938.0&0.0\\
-  \hline
-  \textbf{Pick Best Pruned Random}&\textbf{100\%}&\textbf{639.9}&\textbf{29.4}\\
-  \hline
-  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{597.7}&
-	\textbf{77.9}\\
-  \hline
-  Social Promotion Pruned Random&75\%&938.0&33.0\\
-  \hline
-  Social Promotion Pruned Complete&100\%&815.2&43.5\\
-  \hline
-  Many Iterations Random&100\%&783.5&94.5\\
-  \hline
-  Many Iterations Complete&0\%&N/A&N/A\\
-  \hline
-  \end{tabular}
-\end{table}
 
@@
   \hline
-  Pick Best Pruned Complete&FIXME!&FIXME!&FIXME!\\
+  Pick Best Pruned Complete&25\%&105.4&3.9\\
   \hline
@@
   \hline
-  Social Promotion Pruned Complete&FIXME!&FIXME!&FIXME!\\
+  Social Promotion Pruned Complete&90\%&197.9&12.2\\
   \hline
@@
 Rastrigin, though for all other benchmarks Many Iterations significantly
-outperformed Standard.  In each table, we bold the best method, meaning that it
-has at least a 90\% success rate and its average time to completion is faster
-than all other methods that have at least a 90\% success rate.  For example,
-Many Iterations Random, when it is successful, achieves the required value 6
-times faster than Standard Random.  However, because Many Iterations Random
-only has a 25\% success rate while Standard Random has a 100\% success rate,
-Standard Random is bolded.
-
-We used a double-sided t-test to test for statistical significance in these
-results, and all bolded values are statistically significant with a p-value of
-less than $10^{-5}$.  In one instance two values are bolded because they are
-statistically tied, while both are significantly better than all other methods.
+outperformed Standard.
 
@@
 functions becomes rather dismal.  In almost all cases, Standard fails to make
-any significant progress.  Our brief search for an explanation seems to say
-that the space is too large for particles to converge to the same point, so all
-updates to the best position found come from a single particle wandering in the
-space.  Every time that single particle finds a new best position, its velocity
+any significant progress.  Our search for an explanation seems to say that the
+space is too large for particles to converge to the same point, so all updates
+to the best position found come from a single particle wandering in the space.
+Every time that single particle finds a new best position, its velocity
 contracts, so the particle is unable to make significant progress on its own.
@@
 
-For example, suppose a particle finds a new good position and updates its
-values for $\pbest$ and $\nbest$, and that its velocity is such that it will
-continue updating $\pbest$ and $\nbest$ if it continues in the same direction.
-At that point, its velocity contracts, due to the motion equations in
-constricted PSO (the social and cognitive components drop to zero and all that
-is left is momentum, which is then constricted).  However, at that point we
-speculate three iterations ahead along a path that is in roughly the same
-direction the particle was just going, without constricting its velocity at
-each of those iterations.  Thus the particle is able to make progress in that
-direction instead of contracting its velocity at every iteration.
-
 Because none of the Standard approaches were successful at optimizing the 500
@@
 
-For the model fitting problem that we experimented with, we used 144
-processors.  We show results for three methods: Standard with a Random
-topology, Standard with a Subswarm topology, and Many Iterations with a Random
-topology.  Standard with a Subswarm topology had eight independent subswarms of
-18 particles each, and the particles in each subswarm were connected with a
-Random topology.  In the graphs we call this method Standard Subswarms.
+For the model fitting problem we used 144 processors.  We show results for
+three methods: Standard with a Random topology, Standard with a Subswarm
+topology, and Many Iterations with a Random topology.  Standard with a Subswarm
+topology had eight independent subswarms of 18 particles each, and the
+particles in each subswarm were connected with a Random topology.  In the
+graphs we call this method Standard Subswarms.
 
@@
 1.83 seconds of that was function evaluation time and the rest was overhead.
-With this particular implementation, then, using Many Iterations instead of
-Standard led to an average time savings of 29 minutes, which is more than half
-of the 50 minutes that Standard takes on average to reach a successful value.
+With this particular implementation, then, Many Iterations takes on average 21
+minutes to reach a successful value, while Standard takes 50 minutes.
 
@@
 We have detailed how to perform speculative evaluation in PSO in several
-different parallel frameworks.  Using this methodology, the behavior of the
+different parallel architectures.  Using this methodology, the behavior of the
 original PSO algorithm can either be exactly reproduced, two iterations at a
 time, or the behavior can be modified in order to improve performance.  While
-exactly reproducing PSO often uses too many extra processors to be useful, when
-we allow ourselves some freedom with the algorithm we see great improvements
-over previous methods.  We have shown results that conclusively demonstrate the
-superiority of our techniques for several functions over the standard practice
-of adding particles to the swarm when extra processors are available.
+exactly reproducing PSO sometimes uses too many extra processors to be useful,
+when we allow ourselves some freedom with the algorithm we see great
+improvements over previous methods.  We have shown results that conclusively
+demonstrate the superiority of our techniques for several functions over the
+standard practice of adding particles to the swarm when extra processors are
+available.
 
@@
 on different functions and at different swarm sizes, as would be expected by
-the No Free Lunch Theorem~\citep{wolpert-1997-nfl-for-optimization}.  We have
-given a brief evaluation of the premature convergence properties of these
-methods on deceptive functions when a smaller number of processors is
-available.  We have also shown evidence that when many processors are available
-by far the best thing to do in most cases is to speculate as far ahead as the
-extra processors allow.
+the No Free Lunch Theorem for
+Optimization~\citep{wolpert-1997-nfl-for-optimization}.  We have given a brief
+evaluation of the premature convergence properties of these methods on
+deceptive functions when a smaller number of processors is available.  We have
+also shown evidence that when many processors are available by far the best
+thing to do in most cases is to speculate as far ahead as the extra processors
+allow.
 
@@
 exploration has been reached for any particular function, adding additional
-particles adds only incremental benefits.  A better use of the additional
-processors, as we have shown, is to perform some amount of speculative
-evaluation.
+particles adds only incremental benefits.  At this point, a better use of the
+additional processors, as we have shown, is to perform some amount of
+speculative evaluation.
 
@@
 
-We have presented a new paradigm for parallelizing the particle swarm
-optimization algorithm, focusing on PSO itself and not all of its variants.  It
+In this work we have focused on PSO itself and not all of its variants.  It
 remains as future work to apply speculative approaches to recent and popular
@@
 would be speculative evaluation, performing two or more of those iterations at
-a time.  Thus our methods have the potential to further improve the parallel
-performance of recently proposed variations of PSO.
+a time.
 
@@
 branches are selected.  However, we only presented a few of the countless
-possibilities for selecting those branches.  Perhaps in Many Iterations it
-would be better to randomly sample branches from seventh iteration, instead
-of a fixed combination of the next three.  It remains as future work to do a
-more exhaustive comparison of possible pruning techniques.
-
-Our methods for determining which speculative evaluations to perform were
-independent of the particle; all particles performed the same number and type
-of evaluations.  Another way to allocate speculative evaluations is to somehow
-use the performance of each particle to determine how many and which extra
-evaluations it can have.  We imagine that the benefits of intelligently and
-dynamically deciding which particles get extra evaluations could produce very
-good results.  Future work could explore the use of a centralized allocator for
-extra processors or ideas in the multi-robot task allocation literature, such
-as auctions, to better assign processors to particles for speculative
-evaluations.  While such methods involve extra communication, with long-running
-function evaluations the extra overhead of intelligently allocating resources
-would likely be negligible.
+possibilities for selecting those branches.  Our methods for determining which
+speculative evaluations to perform were independent of the particle; all
+particles performed the same number and type of evaluations.  Another way to
+allocate speculative evaluations is to somehow use the performance of each
+particle to determine how many and which extra evaluations it can have.
 

### 2011-04-01 masters_thesis/si_paper/paper.tex
Done?

@@
 In this paper we show that the results of standard PSO can be reproduced
-\emph{exactly}, two iterations at a time, using a speculative approach adapted
-from speculative execution. We show that the standard PSO equations can be
-factored such that a set of speculative positions can be found which will
-always include the position computed in the next iteration.  By computing the
-value of the objective function for each of the speculative positions at the
-same time the algorithm evaluates the objective function for the current
-position, it is possible to know the objective function values for both the
-current and the next iteration at the same time.  We demonstrate this principle
-by implementation and show that it produces exactly the same results as
-standard PSO, but two iterations at a time.  The resulting implementation runs
-efficiently on large clusters where the number of processors is much larger
-than a typical or reasonable number of particles, producing better results in
-less wall-clock time.
+\emph{exactly}\footnote{In fact it is only because the results are exactly the
+same that we are confident of our implementation.  With the careful use of
+random seeds we were able to detect errors in our speculative implementation
+when particle positions were off in the tenth digit.}, two iterations at a
+time, using a speculative approach adapted from speculative execution. We show
+that the standard PSO equations can be factored such that a set of speculative
+positions can be found which will always include the position computed in the
+next iteration.  By computing the value of the objective function for each of
+the speculative positions at the same time the algorithm evaluates the
+objective function for the current position, it is possible to know the
+objective function values for both the current and the next iteration at the
+same time.  We demonstrate this principle by implementation and show that it
+produces exactly the same results as standard PSO, but two iterations at a
+time.  The resulting implementation runs efficiently on large clusters where
+the number of processors is much larger than a typical or reasonable number of
+particles, producing better results in less wall-clock time.
 
@@
 
-\section{Implementing Speculative Evaluation}
+\section{Relaxing the Requirements}
+\label{sec:relax}
+
+Refactoring the PSO equations led us to find that speculative approaches are
+possible in the parallelization of PSO.  SEPSO reproduces standard PSO exactly,
+two iterations at a time, at the expense of requiring several times the number
+of processors.  In this section we consider other speculative techniques
+inspired by SEPSO that relax the requirement of exactly reproducing the
+behavior of the original PSO algorithm.
+
+We outline three main improvements to speculative evaluation.  First, in
+\sec{pickbest} we describe a method that uses all of the information found in
+doing speculative evaluations.  Then Sections~\ref{sec:pruning}
+through~\ref{sec:wrong} present a technique that reduces the number of
+speculative evaluations that need to be done for each particle, allowing
+speculative evaluation to use larger swarm sizes with the same number of
+processors.  Finally, \sec{manyiters} shows a method for speculating several
+iterations ahead, instead of just one.
+
+None of these methods fundamentally change the PSO algorithm.  They simply lead
+to particles being at different iterations and having different values for
+personal and neighborhood best positions than would have occurred in standard
+PSO, because they receive different information.  These kinds of relaxations
+are fairly typical in the parallelization of
+PSO~\citep{koh-2006-parallel-asynchronous-pso}.
+
+\subsection{Pick the Best Child}
+\label{sec:pickbest}
+
+In performing speculative evaluation as we have described it, $2n+1$
+speculative evaluations are done per particle, while all but one of them are
+completely ignored.  It seems reasonable to try to make use of the information
+obtained through those evaluations instead of ignoring it.
+
+To make better use of the extra speculative evaluations, instead of choosing
+the speculative child that matches the branch that the original PSO would have
+taken, we take the child that has the best value.  The methodology is exactly
+the same as with SEPSO except for the process of choosing which speculative
+child to accept.  The only change needed in \alg{centralized} (see Appendix A)
+is in step 7, where $\noeval{\s}_{t+1}$ with the best value is chosen from
+$\noeval{\sset}_{t+1}$ instead of with the matching branch.  We call this
+technique Pick Best.
+
+This can be thought of as drawing a number of samples from the next iteration
+and accepting the best one.  Speculative particles that move in good directions
+are kept.  Intuition says that this technique favors exploitation over
+exploration, but as we will show in \sec{results}, that is not always the case.
+
+At this point it is also interesting to note a parallel between our methods and
+parallel evolution
+strategies~\citep{rudolph-1991-distributed-evolution-strategies}.  In evolution
+strategies, a parent individual (representing a potential solution to some
+objective function) produces a number of offspring by a mutation operator.  One
+of the individuals is selected by a selection operator, and that individual
+becomes the parent for the next
+generation~\citep{beyer-2002-evolution-strategies}.  Our methods are similar,
+where our mutation operator is simply the PSO motion equations and the
+selection operator is either the indicator function introduced in \sec{sepso},
+in the case of our original speculative algorithm, or the standard selection
+operator based on fitness, in the case of this Pick Best technique.
+
+\subsection{Pruning the Speculative Evaluations}
+\label{sec:pruning}
+
+Because the SEPSO requires so many extra evaluations, a natural step to take is
+to eliminate some of them.  If we could reliably predict which branch were
+going to be taken, we could limit ourselves to one speculative evaluation per
+particle instead of $2n+1$.  With a fixed number of processors, this would
+allow us to greatly increase the swarm size relative to that needed in the
+original speculative algorithm (e.g., with 120 processors, a standard
+parallelization has a swarm of size 120, complete speculative evaluation has a
+swarm of size 15, and pruning the evaluations to only one per particle allows a
+swarm of size 60).  As not all of the branches are evaluated in any given
+iteration, we call this technique pruning.
+
+We look at the statistical behavior of PSO to find probabilities of taking any
+particular branch.  While we cannot with certainty predict which branch a
+particle will take every time, if we can use statistics to narrow down the
+$2n+1$ possible evaluations to a few likely candidates, we can decrease the
+amount of computation required to do speculative evaluation and improve our
+performance.
+
+\subsection{Branch Statistics}
+\label{sec:statistics}
+
+In Table~\ref{tab:evals} we presented all possible branches that a particle
+with two neighbors could take.  Here we lump all of the neighbors together and
+consider the statistics for the five branches shown in
+Table~\ref{tab:branches}.  In the identifiers, N represents an update to
+$\nbest$ coming from any neighbor.
+
+\begin{table}[ht]
+  \caption{Five Branches to Consider for Statistics}
+  \label{tab:branches}
+  \centering
+  \begin{tabular}{lcc}
+	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
+	\hline
+	\hline
+	$\casepn$&No update&No update\\
+	\hline
+	$\casexn$&Self&No update\\
+	\hline
+	$\casexx$&Self&Self\\
+	\hline
+	$\casepN$&No update&Some Neighbor\\
+	\hline
+	$\casexN$&Self&Some Neighbor\\
+	\hline
+  \end{tabular}
+\end{table}
+
+\begin{table}[ht]
+  \caption{Branch Statistics in PSO}
+  \label{tab:stats}
+  \centering
+  \scriptsize
+  \begin{tabular}{c|c|c|c|c|c|c|}
+	Topology&Function&\casepn&\casexn&\casexx&\casepN&\casexN\\
+	\hline
+	\hline
+	Ring&Sphere&53.0\%&9.3\%&11.4\%&20.2\%&6.2\%\\
+	&Griewank&51.7\%&8.4\%&12.2\%&20.7\%&7.0\%\\
+	&Rastrigin&49.5\%&4.8\%&14.6\%&21.3\%&9.9\%\\
+	&Rosenbrock&51.3\%&7.4\%&12.9\%&21.1\%&7.3\%\\
+	\cline{2-7}
+	&Average&{51.3\%}&{7.5\%}&{12.8\%}&{20.8\%}&
+	{7.6\%}\\
+	\hline
+	\hline
+	Random&Sphere&66.7\%&11.9\%&2.6\%&15.6\%&3.1\%\\
+	&Griewank&69.0\%&10.9\%&2.5\%&14.9\%&2.7\%\\
+	&Rastrigin&81.9\%&5.5\%&1.5\%&10.0\%&1.0\%\\
+	&Rosenbrock&74.2\%&7.7\%&2.2\%&14.0\%&1.8\%\\
+	\cline{2-7}
+	&Average&{73.0\%}&{9.0\%}&{2.2\%}&{13.6\%}&
+	{2.2\%}\\
+	\hline
+	\hline
+	Complete&Sphere&31.9\%&9.2\%&0.2\%&45.1\%&13.5\%\\
+	&Griewank&35.3\%&8.4\%&0.2\%&44.1\%&11.9\%\\
+	&Rastrigin&47.7\%&6.7\%&0.2\%&38.2\%&7.0\%\\
+	&Rosenbrock&35.3\%&3.4\%&0.3\%&54.4\%&6.6\%\\
+	\cline{2-7}
+	&Average&{37.6\%}&{6.9\%}&{0.2\%}&{45.5\%}&
+	{9.8\%}\\
+	\hline
+  \end{tabular}
+\end{table}
+
+We seek to find the probability of taking any given branch, given whatever
+information is needed: $\prob(\caseset_t|\cdot)$.  In finding these
+probabilities, we do not attempt to derive any distribution from the PSO
+equations, we simply look at empirical distributions.  However, even with
+empirical distributions, the problem with this approach is that it is not clear
+what information influences the probability of taking a branch.  We look at two
+factors that we believe have a significant influence on $\prob(\caseset_t)$:
+topology ($T$) and function ($F$).  Thus we are looking at
+$\prob(\caseset_t|T,F)$.
+
+We show in Table~\ref{tab:stats} with what percentage a particle takes each of
+these branches for three different topologies and four different functions.
+All of our statistics are from swarms of 240 particles.  Brief experimentation
+showed that other swarm sizes had similar statistics.  We ran 750 iterations on
+all combinations of functions and topologies except for the functions Griewank
+and Rastrigin with the Complete topology.  We found that those runs frequently
+converged past machine precision after 500 iterations, and that led to
+erroneously high values for the probability of \casepn.  Instead we ran for
+only 450 iterations on those two combinations.  All of our results were
+averaged over 20 runs of the algorithm; thus the probabilities presented are
+the averages of 3.6 million trials for the branch taken (2.16 million for the
+two with only 450 iterations).  Table~\ref{tab:stats} contains the results.
+The definitions for all of the functions in the table are found in \sec{setup}.
+
+The probabilities presented in Table~\ref{tab:stats} are interesting in and of
+themselves and could probably be used to better understand the characteristics
+of various topologies.  It is notable that there is small variation between
+functions in any given topology, but the variation across topologies is far
+greater.  However, our concern is with speculative evaluation.  We are
+interested in predicting the branch that any given particle will take at a
+particular iteration.  For our purposes, it appears that given a topology, the
+probability of selecting a branch and the function are close to independent, or
+$\prob(\caseset_t|T,F) \approx \prob(\caseset_t|T)$.
+
+From Table~\ref{tab:stats} we can see that with the Random topology, we can
+pick the first branch, corresponding to stagnation, and be right around 70\% of
+the time.  With the Ring topology, we would be right 50\% of the time.
+Branches \casepN\ and \casexN\ really correspond to several actual branches, as
+all of the neighbors are lumped together.  The 20\% probability of taking
+branch \casepN\ with the Ring topology can be split into two branches, as there
+are only two neighbors.  It also turns out that the neighbor that last updated
+the neighborhood best is the most likely to update it next time, so keeping
+track of that information could be fruitful in trying to predict that branch.
+
+The statistics for the Complete topology are less promising, as there are 240
+neighbors that branch \casepN\ splits into, instead of two.  Pruning does,
+however, allow for the possibility of using the Complete topology in
+speculative evaluation while avoiding the explosion in the number of
+evaluations it would otherwise entail.
+
+Becuase there are a few branches with very high probabilities in the topologies
+we are interested in, we can have hope that cutting out some of the evaluations
+that have low probability will lead to an increase in overall performance.  In
+order to implement this kind of pruning, the only change that needs to be made
+to \alg{centralized} is in Step 2, where the speculative children are
+generated.  Instead of generating all possible speculative children, generate
+the subset of the children that is desired.
+
+\subsection{Recovering from Pruning Too Much}
+\label{sec:wrong}
+
+When not all of the branches are evaluated, there is some probability of not
+evaluating the branch that was actually taken by the original particle.  There
+are a few possibilities for recovery in this case.  One is to leave the
+particle as it is, not accepting any of the speculative evaluations, because
+none of them were correct.  This leads to particles being at different
+iterations, as some particles guess correctly while others do not.  Thus we
+lose exact compatibility with the original PSO, though this particular
+relaxation is nothing new; asynchronous adaptations of PSO do the same
+thing~\citep{koh-2006-parallel-asynchronous-pso}.  As an aside, it is equivalent
+in this case to simply increment the iteration number of particles which fail
+to correctly predict their branch.  This keeps the iteration number constant
+across all particles, simplifying the work needed to be done in determining
+neighbors when dynamic topologies are involved.  We call this technique Social
+Promotion.
+
+Another possibility is to pick the best child, as described in \sec{pickbest}.
+This ignores the fact that the branch might have been wrong; it does not
+matter, because we simply accept the child that had the best value.  In most of
+the experiments that we ran, it turned out that picking the best child
+performed better than keeping the particle back an iteration.
+
+\subsection{More than one iteration ahead}
+\label{sec:manyiters}
+
+We need not simply produce speculative children for the next iteration.  We can
+view all possible speculative evaluations for a particle as an infinite tree
+with branching factor $2n+1$.  As we have already seen that doing one full
+level of the tree produces too many extra evaluations to be profitable, it is
+incredibly unlikely that doing two full iterations would produce decent
+results.  But, if the idea of speculating more than one iteration ahead is
+combined with wisely pruning the possible evaluations based on branch
+statistics, we can use just a few extra evaluations to go two or more
+iterations ahead on the most likely branches.
+
+When speculating more than one iteration ahead, the idea of Social Promotion
+cannot feasibly be implemented, as we can only determine correct branches for
+the first iteration.  Thus in this case we always pick the child that has the
+highest value.
+
+The question of which branches to take in this infinite tree is an intriguing
+one that we can only begin to explore here.  If the branch corresponding to
+stagnation has a 75\% chance of being taken, as in the Random topology with
+most functions, we could speculate three iterations ahead on that branch and
+still have a 42\% chance of predicting correctly.  However, intuition would say
+that perhaps it is better to hope that the particle is productive instead of
+stagnant, so a branch where the particle updates its personal best might be
+more fruitful to try.  In our experiments we try just one of the countless
+possibilities, but one that turns out to work very well.  More work is needed
+to compare the different branching possibilities on various functions.
+
+\section{Experimental Setup}
+\label{sec:setup}
+
+We ran experiments to compare our speculative parallelization of PSO to the
+standard parallelization.  At each iteration of the algorithms, we use one
+processor to perform one function evaluation for one particle, be it
+speculative or non-speculative.  The exact evaluation time at which this
+architecture becomes reasonable depends on the amount of communication overhead
+in the parallel implementation and the number of particles in the swarm.  For
+our implementation we found that time to be around 100 milliseconds for swarms
+of 50 particles.  When the swarm size increases, the minimum evaluation time at
+which this architecture should be used decreases.
+
+\subsection{Function Evaluations vs. Time Steps}
+
+Results in serial PSO are typically presented in terms of function evaluations.
+This is a natural abstraction from implementation details that still allows a
+comparison of the implementation-independent aspects of each algorithm.  The
+number of function evaluations performed is assumed to be proportional to
+wall-clock time, as all evaluations are performed in serial.  Only reporting
+function evaluations could hide the fact that one algorithm requires more
+overhead than another and thus actually takes more time to perform the same
+number of function evaluations; however, function evaluations are still
+considered the standard method of reporting, as evaluation times for functions
+also vary greatly and could make the additional overhead negligible.
+
+In parallel PSO on long-running functions, the natural way to present results
+is in terms of iterations, not function evaluations.  This is because when all
+function evaluations at each iteration are performed concurrently, iterations
+are the direct equivalent of wall-clock time.  Thus we report iterations in our
+results instead of function evaluations.  But because SEPSO actually performs
+two iterations of PSO at each ``iteration,'' and Social Promotion and Many
+Iterations make the idea of ``iterations of PSO'' somewhat nebulus, we instead
+call each ``iteration'' a ``time step.''
+
+Just as serial PSO papers do not report actual running times of their specific
+implementations, we do not report running times, favoring the abstraction of
+time steps.  That being said, given the time required for the evaluation of the
+objective function and the communication overhead per iteration for a specific
+implementation, an actual running time can be recovered from all of our
+results.  Simply multiply the number of time steps by the sum of the function
+evaluation time and the overhead.  We wish to stress that the ``time steps'' we
+report are proportional to wall-clock time, given the architecture we have
+assumed.
+
+\subsection{Functions Used}
+
+We experimented with five common benchmark problems defined
+in~\citep{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
+and Bohachevsky.
+
+Rastrigin is initialized in $[-5.12,5.12]^D$ and is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
+10\right), z=x-c\]
+
+Sphere is initialized in $[-50,50]^D$ and is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]
+
+Schwefel 2.21 is initialized in $[-500,500]^D$ and is defined as:
+\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]
+
+Griewank is initialized in $[-600,600]^D$ and is defined as:
+\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
+\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]
+
+And Bohachevsky is initialized in $[-15,15]^D$ and is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
+x_{i+1})+.7)\]
+
+In computing the branch statistics in \sec{statistics}, we also used the
+Rosenbrock function.  That function is defined as:
+\[f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)\]
+
+The $c$ in the function definitions is a shifted center, in order to avoid
+origin-seeking bias in the PSO algorithm and its
+variants~\citep{monson-2005-origin-seeking-bias}.  We move the center of the
+shifted functions (all except Bohachevsky) to be halfway between the center and
+the boundary of the initialization region.  For example, if the initialization
+region is $[-50,50]^D$, the center is $(25)^D$.  We tested all of these
+functions in their 20 dimension, 50 dimension, and 500 dimension varieties.
+
+One might question the use of benchmark functions for evaluating our methods,
+as they take on the order of 1 millisecond to evaluate instead of the 100s that
+we said were required to justify our parallel architecture.  The purpose of
+exploring the behavior of an algorithm on benchmark functions, whether in
+serial or in parallel optimization research, is because benchmark functions are
+believed to represent functions that one might encounter in the real world,
+regardless of the amount of time those functions take to evaluate.  While
+benchmark functions themselves take fractions of a second to evaluate and thus
+have no need of the kind of parallelization we explore in this paper, they also
+have no need of particle swarm optimization at all, as they can be solved
+analytically.  Yet they are useful to us and to optimization researchers
+generally because they stand as surrogates for the kinds of functions
+practitioners are actually interested in, and they allow us to explore the
+behavior of optimization algorithms on several different classes of functions
+in an easy and standardized way.
+
+To aid in convincing the reader that performance on benchmark functions roughly
+corresponds to performance on real-world problems, we also tested our
+parallelization methods on a typical research problem, that of fitting a model
+to a large quantity of data.  We generated 10,000 data points from a radial
+basis function network with 10 bases and some added Gaussian noise.  We then
+fit a radial basis function network to the data using PSO.  This amounted to a
+30 dimensional function to optimize, with a function evaluation time on the
+order of two seconds.  We refer to this function as ``the model fitting
+problem.''
+
+\subsection{Parallelization Techniques Compared}
+
+The parallelization techniques we compare are the standard parallelization
+(here labeled Standard simply for ease of reference), our original speculative
+approach (recall that we refer to this as SEPSO), and the four relaxations of
+SEPSO discussed in \sec{relax}.  In presenting our results, we call the
+approach developed in \sec{pickbest} Pick Best.  The methods described in
+\sec{pruning} through \sec{wrong} are called Pick Best Pruned and Social
+Promotion Pruned, and the method in \sec{manyiters} we call Many Iterations.
+
+Using the same number of processors for each approach (and thus the same number
+of function evaluations per time step) requires that our speculative
+parallelizations have a smaller swarm size than the standard parallelization.
+For the topologies we used with SEPSO and Pick Best, a particle has three
+neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
+$7$ speculative evaluations per particle.  With one evaluation needed for the
+original, non-speculative particle, we have $8p$ evaluations for every two
+iterations, where $p$ is the number of particles in the speculative swarm.  The
+extra evaluations required in our speculative approach would instead be used to
+evaluate particles in standard parallelizations, so we compare swarms of size
+$p$ in speculative evaluation with swarms of size $8p$ in standard approaches.
+
+When performing pruning in Pick Best Pruned and Social Promotion Pruned, there
+are a large number of ways to prune speculative evaluations.  We experimented
+with several, but present results for only one possible pruning.  The pruning
+we present uses only the two branches where the $\nbest$ value was not updated:
+\casepn\ and \casexn.  Those branches are convenient in that no messages are
+needed from neighbors in order to produce the positions of the speculative
+particles---in distributed frameworks using several rounds of communication
+(see Appendix A), one of the rounds of communication can be dropped entirely.
+Pruning all but these branches also allows the use of arbitrarily dense
+topologies, as the number of speculative particles is no longer dependent on
+the number of neighbors the particle has.
+
+Because pruning only requires two speculative evaluations per particle (along
+with evaluating the original particle), we can use swarms of size
+$\frac{1}{3}p$ when pruning to compare to a swarm of size $p$ with the standard
+parallelization, instead of $\frac{1}{8}p$ with other techniques.
+
+There are also many ways to speculate several iterations ahead, and with Many
+Iterations we again only show results for one of them.  The combination of
+branches we tried uses seven speculative evaluations per particle, matching the
+swarm size of the original speculative algorithm.  The seven evaluations we
+used corresponded to several iterations of branches \casepn\ and \casexn.  Two
+of the evaluations were just one iteration ahead, four were two iterations
+ahead, and one was three iterations ahead.  The evaluations that were one
+iteration ahead were branches \casepn\ and \casexn; those that went two
+iterations ahead were formed by taking either branch \casepn\ or \casexn and
+then branch \casepn\ or \casexn; and the evaluation going three iterations
+ahead followed branch \casepn\ on all three iterations.  As with Pick Best
+Pruned and Social Promotion Pruned, this choice of branches allows the use of
+arbitrarily dense topologies, as the number of speculative evaluations per
+particle is independent of the number of neighbors the particle has.
+
+\subsection{Topologies}
+
+For each benchmark function we report results using the topology that is widely
+considered best for that function, as reported in the literature
+(e.g.,~\cite{bratton-2007-defining-a-standard-for-pso}).  In this paper we
+limit ourselves to the Ring topology and the Complete topology, as is common
+practice, along with the Random topology (the parallel approximation to the
+Complete topology mentioned in \sec{pso}).  We also mentioned in \sec{related}
+that some related work can be described as changes in topology, particularly
+that of having subswarms of fully connected topologies that occasionally
+communicate with each other.  This related work focused mainly on functions
+with very fast evaluation times where such techniques drastically reduce
+interprocessor communication.  With long function evaluations these topologies
+are not as practical, as only one particle is evaluated by each processor at
+each iteration.  We experimented with a subswarm topology and found that in
+every instance except one it performed worse than either Complete or Ring, and
+thus we leave it out of the results except in the one instance where it
+improved performance.
+
+Where the Complete topology would normally be preferred, we use a Random
+topology in SEPSO and Pick Best, as Complete leads to an explosion in the
+number of speculative evaluations (as noted, the other speculative techniques
+can still use a Complete topology; we often report results on both topologies
+for those methods).  If speculative evaluation were not being performed, it is
+possible that the Complete topology would be used.  However, the Complete
+topology also requires a very large amount of interprocessor communication in
+distributed PSO, so it is still quite possible that Random would be used even
+with standard parallelizations~\citep{mcnabb-2009-large-particle-swarms}.  But,
+to be fair in our comparisons, we compare to the standard parallelization using
+both the Random topology and the Complete topology (labeled PSO Random and PSO
+Complete in our results).  Again, the amount of communication overhead is
+heavily dependent on implementation details which are not the focus of this
+paper.  A practitioner using a particular implementation can compare the
+results given here for whichever topology is most practical given the specific
+implementation.
+
+\section{Results}
+\label{sec:results}
+
+We frequently present tables summarizing our results.  In each table, we bold
+the ``best'' method, meaning that it has at least a 90\% success rate and its
+average time to completion is faster than all other methods that have at least
+a 90\% success rate.  We used a double-sided t-test to test for statistical
+significance in these results, and all bolded values are statistically
+significant with a p-value of less than $10^{-5}$.  In some instances two
+values are bolded because they are statistically tied, while both are
+significantly better than all other methods.
+
+We first present results in \sec{20dims} for the 20 dimensional variants of the
+benchmark functions we tested.  We give a discussion of each function
+individually at 20 dimensions, as there are interesting characteristics of the
+algorithms that are worth discussing.  In \sec{50dims} we then give a summary
+of results for the 50 dimensional variants of the benchmark functions; we do
+not go into as much detail in our discussion as the results are very similar to
+those in 20 dimensions.  We finish our discussion of benchmark functions with
+their 500 dimensional variants in \sec{500dims}, and we discuss results on the
+model fitting problem in \sec{model}.
+
+\subsection{20 Dimensions}
+\label{sec:20dims}
+
+\subsubsection{Sphere}
+
+First we look at Sphere, the simplest of common benchmark functions.  The
+function has a single global optimum and no other local optima.  Sphere is best
+optimized in terms of function evaluations with a small swarm using a Complete
+topology.  We expect our methods to be perfect for such functions, and our
+results show this intuition to be correct.  For this comparison we used 240
+processors, so the methods had swarms of size 30 (for SEPSO, Pick Best, and
+Many Iterations), 80 (Pick Best Pruned and Social Promotion Pruned), and 240
+(Standard).  We compared too many methods here to fit into one graph, so we
+show just a few methods in \fig{sphere-20} and in \tabref{sphere-20} we show a
+summary of the results for all methods we tested.
+
+In \fig{sphere-20} we see that Many Iterations using a Complete topology
+converges incredibly quickly on a very poor value.  We found this behavior to
+be quite consistent across functions for this method, so we rarely show results
+for Many Iterations Complete.  However, Pick Best Pruned works very well with
+Complete on this function.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{sphere20}
+  \caption{Function Sphere with 20 dimensions, comparing a pertinent subset of
+  possible methods.  Each method performs one evaluation on each of 240
+  processors per time step.}
+  \label{fig:sphere-20}
+\end{figure}
+
+\begin{table}
+  \caption{Summary of results for function Sphere with 20 dimensions,
+  measuring number of time steps to reach a value of 1e-35}
+  \label{tab:sphere-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&0\%&N/A&N/A\\
+  \hline
+  Standard Random&0\%&N/A&N/A\\
+  \hline
+  Standard Complete&100\%&931.0&19.1\\
+  \hline
+  SEPSO Ring&0\%&N/A&N/A\\
+  \hline
+  SEPSO Random&85\%&972.8&15.8\\
+  \hline
+  Pick Best Ring&100\%&917.4&21.0\\
+  \hline
+  Pick Best Random&100\%&768.1&14.3\\
+  \hline
+  Pick Best Pruned Ring&100\%&967.8&9.9\\
+  \hline
+  Pick Best Pruned Random&100\%&693.5&10.4\\
+  \hline
+  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{389.6}&
+    \textbf{9.6}\\
+  \hline
+  Social Promotion Pruned Ring&0\%&N/A&N/A\\
+  \hline
+  Social Promotion Pruned Random&100\%&971.1&10.6\\
+  \hline
+  Social Promotion Pruned Complete&100\%&777.8&16.7\\
+  \hline
+  Many Iterations Ring&100\%&575.2&9.8\\
+  \hline
+  Many Iterations Random&100\%&442.4&7.1\\
+  \hline
+  Many Iterations Complete&0\%&N/A&N/A\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\subsubsection{Schwefel 2.21}
+
+Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
+more by larger swarms than Sphere is.  Thus our speculative algorithms often
+fail to outperform Standard with 240 processors because the simple speculative
+methods only have 30 particles.  As we will show later, when we use 800
+processors at 50 dimensions, our methods perform much better.  However, the
+pruned versions of our methods have 80 particles and thus are able to
+outperform Standard even with only 240 processors.  Because the graph of
+Schwefel 2.21 looks very similar to that of Sphere, we simply present all of
+our results in \tabref{schwefel-20}.
+
+\begin{table}
+  \caption{Summary of results for function Schwefel with 20 dimensions,
+  measuring number of time steps to reach a value of 1e-06}
+  \label{tab:schwefel-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&0\%&N/A&N/A\\
+  \hline
+  Standard Random&0\%&N/A&N/A\\
+  \hline
+  Standard Complete&100\%&837.2&44.8\\
+  \hline
+  SEPSO Random&0\%&N/A&N/A\\
+  \hline
+  Pick Best Random&5\%&938.0&0.0\\
+  \hline
+  \textbf{Pick Best Pruned Random}&\textbf{100\%}&\textbf{639.9}&\textbf{29.4}\\
+  \hline
+  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{597.7}&
+	\textbf{77.9}\\
+  \hline
+  Social Promotion Pruned Random&75\%&938.0&33.0\\
+  \hline
+  Social Promotion Pruned Complete&100\%&815.2&43.5\\
+  \hline
+  Many Iterations Random&100\%&783.5&94.5\\
+  \hline
+  Many Iterations Complete&0\%&N/A&N/A\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\subsubsection{Rastrigin}
+
+Rastrigin is a multi-modal function that is best optimized with a large,
+Complete swarm.  It has been shown that with Rastrigin, the more particles
+there are in the swarm, the lower function value it finds, up to at least 4000
+particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
+in local optima and converge to poorer values.  Because our speculative
+algorithms require significantly smaller swarm sizes, we would expect to not
+perform very well on functions such as Rastrigin.  Our experiments show our
+intuition to be correct.  In this experiment we used 240 processors, so SEPSO
+and Pick Best each had 30 particles, and the Standard algorithms had 240
+particles.  As expected, SEPSO and Pick Best converge quickly to worse local
+optima than Standard does.  \fig{rastrigin} shows the results graphically, and
+\tabref{rastrigin-20} shows results for all of the methods we tried.
+
+\begin{table}
+  \caption{Summary of results for function Rastrigin with 20 dimensions,
+  measuring number of time steps to reach a value of 20}
+  \label{tab:rastrigin-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&25\%&743.4&212.0\\
+  \hline
+  Standard Random&100\%&372.1&86.6\\
+  \hline
+  \textbf{Standard Complete}&\textbf{95\%}&\textbf{273.1}&\textbf{99.7}\\
+  \hline
+  SEPSO Ring&5\%&471.0&0.0\\
+  \hline
+  SEPSO Random&15\%&208.0&17.0\\
+  \hline
+  Pick Best Ring&35\%&489.0&152.0\\
+  \hline
+  Pick Best Random&15\%&136.0&40.1\\
+  \hline
+  Pick Best Pruned Ring&95\%&545.9&208.5\\
+  \hline
+  Pick Best Pruned Random&10\%&95.0&11.0\\
+  \hline
+  Pick Best Pruned Complete&5\%&59.0&0.0\\
+  \hline
+  Social Promotion Pruned Ring&5\%&580.0&0.0\\
+  \hline
+  Social Promotion Pruned Random&80\%&278.2&136.3\\
+  \hline
+  Social Promotion Pruned Complete&50\%&200.5&130.2\\
+  \hline
+  Many Iterations Ring&10\%&117.0&25.0\\
+  \hline
+  Many Iterations Random&5\%&57.0&0.0\\
+  \hline
+  Many Iterations Complete&0\%&N/A&N/A\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{rastrigin}
+  \caption{Function Rastrigin with 20 dimensions.  Each method performs one
+  evaluation on each of 240 processors per time step.}
+  \label{fig:rastrigin}
+\end{figure}
+
+\subsubsection{Griewank}
+\label{sec:griewank-20}
+
+It is generally recommended to use the Ring topology when optimizing the
+Griewank function, as Complete is prone to premature convergence on a local
+optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  The global optimum
+of Griewank has a value of 0.  When most trials reach the global optimum but a
+few get stuck, the resultant ``average value'' graph has a flat line that is
+misleading.  Thus instead we present plots showing the percent of trials that
+have passed some threshold at each time step, as is common practice with these
+functions~\citep{mendes-2004-population-topologies-in-pso}.  The threshold we
+chose for this case was $10^{-6}$, as that value is below any local optima and
+the swarm always successfully reaches 0 once is passes that point.
+
+We ran 50 trials of each experiment with Griewank, so that the curves are more
+smooth.  We show results in \fig{basic-griewank1} for swarms of size 100 and
+800 using the Ring topology.  One can see in the figure that SEPSO reaches the
+global optimum on average close to twice as fast as Standard, while Pick Best
+is close to three times as fast and Many Iterations is even faster when it is
+successful, though it is only successful 84\% of the time.  \tabref{griewank}
+shows all of our results in tabular form.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{griewank6}
+  \caption{Function Griewank with 20 dimensions.  Each method performs one
+  evaluation on each of 800 processors per time step.  Instead of showing
+  average function value, we show the percentage of runs that have found the
+  global optimum by each iteration.  All algorithms use the Ring topology.}
+  \label{fig:basic-griewank1}
+\end{figure}
+
+\begin{table}
+  \caption{Summary of results for function Griewank with 20 dimensions,
+  measuring number of time steps to reach a value of 1e-06}
+  \label{tab:griewank-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&100\%&762.2&19.4\\
+  \hline
+  SEPSO Ring&100\%&426.0&81.0\\
+  \hline
+  \textbf{Pick Best Ring}&\textbf{100\%}&\textbf{272.0}&\textbf{17.7}\\
+  \hline
+  \textbf{Pick Best Pruned Ring}&\textbf{100\%}&\textbf{282.9}&\textbf{10.7}\\
+  \hline
+  Social Promotion Pruned Ring&100\%&482.0&16.6\\
+  \hline
+  Many Iterations Ring&84\%&183.2&49.4\\
+  \hline
+  \end{tabular}
+\end{table}
+
+We pause here to show some interesting characteristics of the speculative
+techniques we have proposed.  With 800 processors our methods perform very well
+on this function.  With 240 processors, however, the results are much more
+mixed.  Because 240 processors is near the point at which speculative
+evaluation becomes useful, it is enlightening to see the behavior of the
+various algorithms at this point.
+
+We show results in \fig{basic-griewank2} for swarms of size 30 and 240 using
+the Ring topology.  One can see in the figure that when SEPSO is successful, it
+finds the optimum much faster than Standard.  However, because the swarm size
+is so small, SEPSO gets stuck a little less than half of the time.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{griewank1}
+  \caption{Function Griewank with 20 dimensions.  Each method performs one
+  evaluation of the objective function on each of 240 processors per time step.
+  All algorithms use the Ring topology.}
+  \label{fig:basic-griewank2}
+\end{figure}
+
+When we look at the performance of our Pick Best approach, we see that it
+greatly improves performance on Griewank.  This is somewhat counter-intuitive,
+because Griewank is deceptive and Pick Best seems like a greedy algorithm.  But
+in \fig{griewank-pickbest} we see that Pick Best improves accuracy over SEPSO
+by 20\%, while at the same time finding the optimum over 100 time steps sooner
+on average.  It seems that while Pick Best is locally greedy, there is enough
+exploration in the seven speculative evaluations to overcome the inherent
+greediness of the approach.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{griewank3}
+  \caption{Function Griewank with 20 dimensions, comparing Pick Best with
+  results from \fig{basic-griewank2}.  Each method performs one evaluation on
+  each of 240 processors per time step.  All algorithms use the Ring topology.}
+  \label{fig:griewank-pickbest}
+\end{figure}
+
+When we introduce pruning, our intuition about Pick Best turns out to be more
+correct.  While adding 50 more particles to the swarm (as pruning allows us to
+have 80 particles with 240 processors instead of only 30), Pick Best with
+pruning still gets stuck just as often as the original Pick Best.  However,
+Social Promotion does well with pruning; it increases the success rate to close
+to 100\%, while still finding the optimum on average much faster than the
+original PSO.  These results are shown in \fig{griewank-pruned}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{griewank4}
+  \caption{Function Griewank with 20 dimensions, comparing pruning with the
+  best results from \fig{griewank-pickbest}.  Each method performs one
+  evaluation on each of 240 processors per time step.  All algorithms use the
+  Ring topology.}
+  \label{fig:griewank-pruned}
+\end{figure}
+
+With Griewank, the premature convergence problems inherent in picking the best
+child are exacerbated when speculating several iterations ahead.  When Many
+Iterations finds the optimum, it finds it quicker than any other method we
+tried, on average four times faster than Standard.  However, it also gets stuck
+and fails to find the optimum more than any other method.  The results are
+shown in \fig{griewank-manyiters}.   \fig{griewank-manyiters} is also
+interesting in that it highlights the trade-off between accuracy and speed in
+the various approaches at this swarm size.  The faster the approach finds the
+optimum, the less likely it is to be successful.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{griewank5}
+  \caption{Function Griewank with 20 dimensions, comparing speculating several
+  iterations ahead with selections from \fig{griewank-pruned}.  Each method
+  performs one evaluation on each of 240 processors per time step.  All
+  algorithms use the Ring topology.}
+  \label{fig:griewank-manyiters}
+\end{figure}
+
+\subsubsection{Bohachevsky}
+
+Bohachevsky is a unimodal function best optimized with a Complete swarm.  It is
+similar to Griewank in that there is a global optimum with a value of 0, and
+the swarm sometimes finds it and sometimes does not.  Thus we present a graph
+similar to those of Griewank, because ``average value'' graphs have a
+misleading flat line.  We used 480 processors to optimize this function.  In
+\fig{bohachevsky} we show a plot with a few pertinent methods, while
+\tabref{bohachevsky-20} shows all of the results.  All of our speculative
+approaches found the optimum much quicker than Standard with a Random topology.
+However, SEPSO was slower than Standard Complete and got stuck 25\% of the
+time.  Pick Best, Pick Best Pruned, and Many Iterations all outperformed
+Standard Complete, with Many Iterations finding the optimum about twice as
+fast.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{bohachevsky}
+  \caption{Function Bohachevsky with 20 dimensions.  Each method performs one
+  evaluation on each of 480 processors per time step.}
+  \label{fig:bohachevsky}
+\end{figure}
+
+\begin{table}
+  \caption{Summary of results for function Bohachevsky with 20 dimensions,
+  measuring number of time steps to reach a value of 1e-06.}
+  \label{tab:bohachevsky-20}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Random&100\%&472.3&6.6\\
+  \hline
+  Standard Complete&100\%&238.2&11.2\\
+  \hline
+  SEPSO Random&75\%&261.6&9.2\\
+  \hline
+  Pick Best Random&100\%&211.8&9.2\\
+  \hline
+  Pick Best Pruned Random&100\%&189.7&3.4\\
+  \hline
+  Pick Best Pruned Complete&25\%&105.4&3.9\\
+  \hline
+  Social Promotion Pruned Random&100\%&260.3&6.5\\
+  \hline
+  Social Promotion Pruned Complete&90\%&197.9&12.2\\
+  \hline
+  \textbf{Many Iterations Random}&\textbf{94\%}&\textbf{118.2}&\textbf{2.6}\\
+  \hline
+  Many Iterations Complete&0\%&N/A&N/A\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\subsection{50 Dimensions}
+\label{sec:50dims}
+
+Given our observation from the 20 dimensional benchmark functions that Many
+Iterations most often outperforms other speculative approaches, we only present
+results for Many Iterations and Standard for the 50 and 500 dimensional
+variants.  For these experiments we used 800 processors, as the problems are
+more difficult and speculative approaches perform best when many processors are
+available, as we showed in \sec{griewank-20}.  Many Iterations Complete showed
+the same premature convergence that was observed with the 20 dimensional
+benchmark functions, so we do not include those results in our tables; the
+success rate was 0\% for all functions.
+
+Tables~\ref{tab:rastrigin-50} through~\ref{tab:bohachevsky-50} present
+summaries of our results for each of the benchmark functions. As with the 20
+dimensional variants, our speculative methods did not perform well on
+Rastrigin, though for all other benchmarks Many Iterations significantly
+outperformed Standard.
+
+A summary of the results is shown in \tabref{50dims-summary}.  For functions
+where Many Iterations outperformed Standard, we report the average speed up.
+On all functions except Rastrigin, our methods showed an average speed up of
+from two to six times compared to previous methods.
+
+\begin{table}
+  \caption{Summary of results for function Rastrigin with 50 dimensions,
+  measuring number of time steps to reach a value of 100}
+  \label{tab:rastrigin-50}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&5\%&776.0&0.0\\
+  \hline
+  \textbf{Standard Random}&\textbf{100\%}&\textbf{593.2}&\textbf{89.5}\\
+  \hline
+  Standard Complete&80\%&343.8&145.8\\
+  \hline
+  Many Iterations Ring&25\%&223.8&67.1\\
+  \hline
+  Many Iterations Random&25\%&98.6&11.0\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Summary of results for function Sphere with 50 dimensions,
+  measuring number of time steps to reach a value of 1e-06}
+  \label{tab:sphere-50}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&0\%&N/A&N/A\\
+  \hline
+  Standard Random&30\%&986.8&8.2\\
+  \hline
+  Standard Complete&100\%&458.4&12.3\\
+  \hline
+  Many Iterations Ring&100\%&288.7&3.9\\
+  \hline
+  \textbf{Many Iterations Random}&\textbf{100\%}&\textbf{248.6}&\textbf{5.1}\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Summary of results for function Schwefel with 50 dimensions,
+  measuring number of time steps to reach a value of 80}
+  \label{tab:schwefel-50}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&0\%&N/A&N/A\\
+  \hline
+  Standard Random&70\%&871.6&57.1\\
+  \hline
+  Standard Complete&100\%&599.0&85.6\\
+  \hline
+  \textbf{Many Iterations Ring}&\textbf{100\%}&\textbf{238.6}&\textbf{34.2}\\
+  \hline
+  \textbf{Many Iterations Random}&\textbf{100\%}&\textbf{242.9}&\textbf{95.4}\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Summary of results for function Griewank with 50 dimensions,
+  measuring number of time steps to reach a value of 1e-06}
+  \label{tab:griewank-50}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&100\%&1898.5&25.7\\
+  \hline
+  \textbf{Many Iterations Ring}&\textbf{97\%}&\textbf{312.4}&\textbf{6.3}\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Summary of results for function Bohachevsky with 50 dimensions,
+  measuring number of time steps to reach a value of 0.01}
+  \label{tab:bohachevsky-50}
+  \centering
+  \begin{tabular}{|l|r|r|r|}
+  \hline
+  Algorithm&\% Complete&Mean&St. Dev.\\
+  \hline
+  \hline
+  Standard Ring&100\%&1434.0&62.1\\
+  \hline
+  Standard Random&100\%&704.5&19.7\\
+  \hline
+  Standard Complete&20\%&377.5&14.9\\
+  \hline
+  \textbf{Many Iterations Ring}&\textbf{95\%}&\textbf{229.8}&\textbf{11.4}\\
+  \hline
+  Many Iterations Random&35\%&171.6&4.0\\
+  \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \caption{Average speed up, comparing the number of time steps to completion
+  for the best speculative topology to the time steps to completion for the
+  best standard topology.  All functions have 50 dimensions.}
+  \label{tab:50dims-summary}
+  \centering
+  \begin{tabular}{|l|r|}
+	\hline
+	Function&Speed Up Factor\\
+	\hline
+	\hline
+	Rastrigin&N/A\\
+	\hline
+	Sphere&1.84\\
+	\hline
+	Schwefel&2.51\\
+	\hline
+	Griewank&6.08\\
+	\hline
+	Bohachevsky&3.07\\
+	\hline
+  \end{tabular}
+\end{table}
+
+\subsection{500 Dimensions}
+\label{sec:500dims}
+
+At 500 dimensions the performance of constricted PSO on benchmark functions
+becomes rather dismal.  In almost all cases, Standard fails to make any
+significant progress.  Our search for an explanation seems to say that the
+space is too large for particles to converge to the same point, so all updates
+to the best position found come from a single particle wandering in the space.
+Every time that single particle finds a new best position, its velocity
+contracts, so the particle is unable to make significant progress on its own.
+In constricted PSO it is necessary to have a collection of particles exploring
+a promising location to keep the particles' velocities from contracting
+prematurely, and in 500 dimensions the space is too large to get the collection
+of particles to the same location.
+
+The results that we present show that while Standard parallelizations suffer
+from this problem, Many Iterations does not.  Though we can provide some
+intuition for why this might be the case, it remains as future work to provide
+a strong theoretical explanation.  The intuition is that we are speculating
+about future positions along paths where each particle's velocity does not
+contract.  The particular branches we selected corresponded to not having
+received a new value for $\pbest$ nor $\nbest$.  In those branches, the
+particle continues in the same direction it was going, without a contraction in
+velocity.
+
+Because none of the Standard approaches were successful at optimizing the 500
+dimensional benchmark functions, we do not present tables as we did for the 50
+dimensional variants.  We do, however, show a few figures demonstrating the
+results we have just explained.  \fig{sphere500} shows the function Sphere,
+\fig{griewank500} shows Griewank, and \fig{rastrigin500} shows Rastrigin.  Note
+that at 500 dimensions speculative approaches outperform standard
+parallelizations on Rastrigin, whereas they did not at 20 and 50 dimensions.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{sphere500}
+  \caption{Function Sphere with 500 dimensions.  Each method performs one
+  evaluation on each of 800 processors per time step.}
+  \label{fig:sphere500}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{griewank500}
+  \caption{Function Griewank with 500 dimensions.  Each method performs one
+  evaluation on each of 800 processors per time step.}
+  \label{fig:griewank500}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{rastrigin500}
+  \caption{Function Rastrigin with 500 dimensions.  Each method performs one
+  evaluation on each of 800 processors per time step.}
+  \label{fig:rastrigin500}
+\end{figure}
+
+One could argue that a fairer comparison would have modified the PSO motions
+equations so that individual particles did not contract their velocity so
+quickly when finding good locations.  However, we used the same motion
+equations for both Standard and Many Iterations; changing the motion equations
+to help Standard would also help Many Iterations, and we would expect to see
+results that are at least as compelling as those from the 50 dimensional
+benchmark functions.
+
+\subsection{Model Fitting}
+\label{sec:model}
+
+For the model fitting problem we used 144 processors.  We show results for
+three methods: Standard with a Random topology, Standard with a Subswarm
+topology, and Many Iterations with a Random topology.  Standard with a Subswarm
+topology had eight independent subswarms of 18 particles each, and the
+particles in each subswarm were connected with a Random topology.  In the
+graphs we call this method Standard Subswarms.
+
+\fig{rbf} shows our results for this function.  The function value reported is
+the sum squared error of the model fit.  \fig{rbf-stuck} shows the percent of
+runs that reached a value for sum squared error of 55,000 by each time step,
+which we designate as successful.  Many Iterations took on average 126 time
+steps to reach this value, while Standard Subswarms took on average 298.  The
+p-value for a t-test on this statistic is less than $10^{-8}$.  With our
+implementation of parallel PSO, each time step took on average 10.06 seconds;
+1.83 seconds of that was function evaluation time and the rest was overhead.
+With this particular implementation, then, Many Iterations takes on average 21
+minutes to reach a successful value, while Standard takes 50 minutes.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{rbf}
+  \caption{Results for fitting a radial basis function to noisy data.  We use
+  144 processors for each method, so each time step corresponds to 144 function
+  evaluations.}
+  \label{fig:rbf}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{rbf_stuck}
+  \caption{Results for fitting a radial basis function to noisy data.  We use
+  144 processors for each method, so each time step corresponds to 144 function
+  evaluations.}
+  \label{fig:rbf-stuck}
+\end{figure}
+
+\section{Conclusions}
+\label{sec:conclusion}
+
+We have described a new technique for using processors in parallel PSO to
+improve the performance of the algorithm.  To our knowledge, this is the first
+time extra processors have been used to do anything in PSO besides increase the
+swarm size.  In an increasingly parallel world, such advancements will prove to
+be crucial to the continued effectiveness of PSO.
+
+We have detailed how to perform speculative evaluation in PSO in several
+different parallel architectures.  Using this methodology, the behavior of the
+original PSO algorithm can either be exactly reproduced, two iterations at a
+time, or the behavior can be modified in order to improve performance.  While
+exactly reproducing PSO sometimes uses too many extra processors to be useful,
+when we allow ourselves some freedom with the algorithm we see great
+improvements over previous methods.  We have shown results that conclusively
+demonstrate the superiority of our techniques for several functions over the
+standard practice of adding particles to the swarm when extra processors are
+available.
+
+What we have presented is not a new variant of PSO.  We presented a new
+parallelization technique, so we compared parallelization strategies on the
+same algorithm, the original PSO.  Our most promising techniques do change the
+behavior of the PSO algorithm slightly, as do some other previously proposed
+parallelization techniques, such as asynchronous PSO.  However, our methods are
+applicable to almost all PSO variants, and so a comparison using the same
+variant for each of the parallelization techniques tested is justified.
+
+We have given five different possible approaches to speculative evaluation,
+each of which has different properties.  These approaches perform differently
+on different functions and at different swarm sizes, as would be expected by
+the No Free Lunch Theorem for
+Optimization~\citep{wolpert-1997-nfl-for-optimization}.  We have given a brief
+evaluation of the premature convergence properties of these methods on
+deceptive functions when a smaller number of processors is available.  We have
+also shown evidence that when many processors are available by far the best
+thing to do in most cases is to speculate as far ahead as the extra processors
+allow.
+
+Though our methods show great improvements on some functions, they do not work
+for all functions.  As is commonly known, in PSO there is a trade-off between
+exploration and exploitation.  Some functions need only minimal exploration,
+and some never seem to have enough.  Increasing the swarm size is a natural way
+to increase exploration in a parallel environment.  However, once ``enough''
+exploration has been reached for any particular function, adding additional
+particles adds only incremental benefits.  At this point, a better use of the
+additional processors, as we have shown, is to perform some amount of
+speculative evaluation.
+
+Large parallel clusters are often required to successfully optimize practical
+modern problems.  To properly use PSO with such clusters, a balance needs to be
+made between using processors to increase the swarm size and using them to
+increase the speed of the algorithm.  This work is a first step in that
+direction that opens the door to many future improvements on speculative
+methods in the parallelization of PSO.
+
+\section{Future Work}
+\label{sec:future}
+
+In this work we have focused on PSO itself and not all of its variants.  It
+remains as future work to apply speculative approaches to recent and popular
+PSO variants, such as the Fully Informed Particle
+Swarm~\citep{mendes-2004-fully-informed-particle-swarm}.  While our methods
+will not always be immediately applicable to every variant, we are confident
+that some kind of speculative approach will be beneficial to the
+parallelization of all forms of PSO, especially as the number of processors
+used gets into the thousands.
+
+We mentioned related work showing that increasing the swarm size throughout the
+course of the algorithm could provide improved performance over a fixed swarm
+size in serial PSO~\citep{montes-de-oca-2010-incremental-social-learning-pso}.
+If this method were extended to parallel PSO, most processors would be idle in
+the first few iterations, while more would be utilized at the end.  During
+iterations where there are many un-utilized processors, a natural use of them
+would be speculative evaluation, performing two or more of those iterations at
+a time.
+
+The issues of the sampling distribution of speculative relaxations, branch
+statistics, and why standard PSO fails at 500 dimensions were briefly mentioned
+in this paper.  Each of those issues needs further treatment.  The sampling
+distribution of our speculative methods could be compared to Poli's description
+of standard PSO's sampling
+distribution~\citep{poli-2008-sampling-distribution-of-pso}.  The branch
+statistics could be used to analyze topologies and discover why certain
+topologies work well on some functions but not on others; perhaps PSO
+performance is more dependent on the branch statistics of a combination of
+topology and function than on the topology itself.  And discovering exactly why
+constricted PSO fails at 500 dimensions while Many Iterations does not could
+lead to improvements in the standard PSO algorithm, even when not running in
+parallel.
+
+We opened the door to speculative parallelization methods in PSO and described
+the possible speculative evaluations to perform as an infinite tree from which
+branches are selected.  However, we only presented a few of the countless
+possibilities for selecting those branches.  Our methods for determining which
+speculative evaluations to perform were independent of the particle; all
+particles performed the same number and type of evaluations.  Another way to
+allocate speculative evaluations is to somehow use the performance of each
+particle to determine how many and which extra evaluations it can have.
+
+Any other optimization algorithm that only depends on current sampling
+positions when computing the next position to sample can be parallelized with
+this technique.  In particular, genetic algorithms produce future generations
+by combining individuals from the current generation.  With a large population
+size there would be an unwieldy amount of possible future individuals, but the
+potential exists to modify the algorithm to use some kind of speculative
+evaluation.
+
+\section*{Appendix A---Implementing Speculative Evaluation}
 \label{sec:implementation}
@@
 position was the correct next position of each particle.  In this section we
-move from the mathematics of our method to its implementation.  First we
-discuss the relatively easy case of a centralized parallel PSO algorithm with a
-master computer and many slaves.  In such an architecture, the master keeps
-track of all necessary information with only trivial message passing needed.
-Then we discuss the more complicated case of a distributed algorithm, where
-each particle is on its own and needs to send and receive messages to and from
-other particles.  Finally we discuss the further complications of a dynamic
-topology such as Random, where a particle's neighbors change from one iteration
-to another.
+discuss in detail some important considerations in the implementation of our
+methods.  First we discuss the relatively easy case of a centralized parallel
+PSO algorithm with a master computer and many slaves.  In such an architecture,
+the master keeps track of all necessary information with only trivial message
+passing needed.  Then we discuss the more complicated case of a distributed
+algorithm, where each particle is on its own and needs to send and receive
+messages to and from other particles.  Finally we discuss the further
+complications of a dynamic topology such as Random, where a particle's
+neighbors change from one iteration to another.
 
@@
 with only one round of communication per two iterations.  However, the
-methodology is tedious and is not the focus of this paper, so we defer its
-description to the Appendix.
+methodology is tedious and distracting from the present discussion, so we defer
+its description to Appendix B.
 
@@
 
-\section{Relaxing the Requirements}
-\label{sec:relax}
-
-While the idea of speculative evaluation in particle swarm optimization is
-intriguing, the results in \sec{results} show that in many cases it requires
-too many extra evaluations to produce competitive results.  On most functions,
-the larger swarm size that can be used in standard parallelizations leads to
-better performance than is found by trying to do speculative evaluation to
-exactly reproduce PSO.  However, if we keep the idea of speculative evaluation
-while relaxing the requirement of exactly reproducing the behavior of the
-original PSO algorithm, we see some impressive results.
-
-We outline three main improvements to speculative evaluation.  First, in
-\sec{pickbest} we describe a method that uses all of the information found in
-doing speculative evaluations.  Then Sections~\ref{sec:pruning}
-through~\ref{sec:wrong} present a technique that reduces the number of
-speculative evaluations that need to be done for each particle, allowing
-speculative evaluation to use larger swarm sizes with the same number of
-processors.  Finally, \sec{manyiters} shows a method for speculating several
-iterations ahead, instead of just one.
-
-None of these methods fundamentally change the PSO algorithm.  They simply lead
-to particles being at different iterations and having different values for
-personal and neighborhood best positions than would have occurred in standard
-PSO, because they receive different information.  These kinds of relaxations
-are fairly typical in the parallelization of
-PSO~\citep{koh-2006-parallel-asynchronous-pso}.
-
-\subsection{Pick the Best Child}
-\label{sec:pickbest}
-
-In performing speculative evaluation as we have described it, $2n+1$
-speculative evaluations are done per particle, while all but one of them are
-completely ignored.  It seems reasonable to try to make use of the information
-obtained through those evaluations instead of ignoring it.
-
-To make better use of the extra speculative evaluations, instead of choosing
-the speculative child that matches the branch that the original PSO would have
-taken, we take the child that has the best value.  The methodology is exactly
-the same as above except for the process of choosing which speculative child to
-accept.  The only change needed in \alg{centralized} is in step 7, where
-$\noeval{\s}_{t+1}$ with the best value is chosen from $\noeval{\sset}_{t+1}$
-instead of with the matching branch.  We call this technique Pick Best.
-
-This can be thought of as drawing a number of samples from the next iteration
-and accepting the best one.  Speculative particles that move in good directions
-are kept.  Intuition says that this technique favors exploitation over
-exploration, but as we will show in \sec{results}, that is not always the case.
-
-At this point it is also interesting to note a parallel between our methods and
-parallel evolution
-strategies~\citep{rudolph-1991-distributed-evolution-strategies}.  In evolution
-strategies, a parent individual (representing a potential solution to some
-objective function) produces a number of offspring by a mutation operator.  One
-of the individuals is selected by a selection operator, and that individual
-becomes the parent for the next
-generation~\citep{beyer-2002-evolution-strategies}.  Our methods are similar,
-where our mutation operator is simply the PSO motion equations and the
-selection operator is either the indicator function introduced in \sec{sepso},
-in the case of our original speculative algorithm, or the standard selection
-operator based on fitness, in the case of this Pick Best technique.
-
-\subsection{Pruning the Speculative Evaluations}
-\label{sec:pruning}
-
-Because the problem facing speculative evaluation is having too many possible
-evaluations to perform, a natural step to take is to eliminate some of the
-evaluations.  If we could reliably predict which branch were going to be taken,
-we could limit ourselves to one speculative evaluation per particle instead of
-$2n+1$.  With a fixed number of processors, this would allow us to greatly
-increase the swarm size relative to that needed in the original speculative
-algorithm (e.g., with 120 processors, a standard parallelization has a swarm of
-size 120, complete speculative evaluation has a swarm of size 15, and pruning
-the evaluations to only one per particle allows a swarm of size 60).  As not
-all of the branches are evaluated in any given iteration, we call this
-technique pruning.
-
-We look at the statistical behavior of PSO to find probabilities of taking any
-particular branch.  While we cannot with certainty predict which branch a
-particle will take every time, if we can use statistics to narrow down the
-$2n+1$ possible evaluations to a few likely candidates, we can decrease the
-amount of computation required to do speculative evaluation and improve our
-performance.
-
-\subsection{Branch Statistics}
-\label{sec:statistics}
-
-In Table~\ref{tab:evals} we presented all possible branches that a particle
-with two neighbors could take.  Here we lump all of the neighbors together and
-consider the statistics for the five branches shown in
-Table~\ref{tab:branches}.  In the identifiers, N represents an update to
-$\nbest$ coming from any neighbor.
-
-\begin{table}[ht]
-  \caption{Five Branches to Consider for Statistics}
-  \label{tab:branches}
-  \centering
-  \begin{tabular}{lcc}
-	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
-	\hline
-	\hline
-	$\casepn$&No update&No update\\
-	\hline
-	$\casexn$&Self&No update\\
-	\hline
-	$\casexx$&Self&Self\\
-	\hline
-	$\casepN$&No update&Some Neighbor\\
-	\hline
-	$\casexN$&Self&Some Neighbor\\
-	\hline
-  \end{tabular}
-\end{table}
-
-\begin{table}[ht]
-  \caption{Branch Statistics in PSO}
-  \label{tab:stats}
-  \centering
-  \scriptsize
-  \begin{tabular}{c|c|c|c|c|c|c|}
-	Topology&Function&\casepn&\casexn&\casexx&\casepN&\casexN\\
-	\hline
-	\hline
-	Ring&Sphere&53.0\%&9.3\%&11.4\%&20.2\%&6.2\%\\
-	&Griewank&51.7\%&8.4\%&12.2\%&20.7\%&7.0\%\\
-	&Rastrigin&49.5\%&4.8\%&14.6\%&21.3\%&9.9\%\\
-	&Rosenbrock&51.3\%&7.4\%&12.9\%&21.1\%&7.3\%\\
-	\cline{2-7}
-	&Average&{51.3\%}&{7.5\%}&{12.8\%}&{20.8\%}&
-	{7.6\%}\\
-	\hline
-	\hline
-	Random&Sphere&66.7\%&11.9\%&2.6\%&15.6\%&3.1\%\\
-	&Griewank&69.0\%&10.9\%&2.5\%&14.9\%&2.7\%\\
-	&Rastrigin&81.9\%&5.5\%&1.5\%&10.0\%&1.0\%\\
-	&Rosenbrock&74.2\%&7.7\%&2.2\%&14.0\%&1.8\%\\
-	\cline{2-7}
-	&Average&{73.0\%}&{9.0\%}&{2.2\%}&{13.6\%}&
-	{2.2\%}\\
-	\hline
-	\hline
-	Complete&Sphere&31.9\%&9.2\%&0.2\%&45.1\%&13.5\%\\
-	&Griewank&35.3\%&8.4\%&0.2\%&44.1\%&11.9\%\\
-	&Rastrigin&47.7\%&6.7\%&0.2\%&38.2\%&7.0\%\\
-	&Rosenbrock&35.3\%&3.4\%&0.3\%&54.4\%&6.6\%\\
-	\cline{2-7}
-	&Average&{37.6\%}&{6.9\%}&{0.2\%}&{45.5\%}&
-	{9.8\%}\\
-	\hline
-  \end{tabular}
-\end{table}
-
-We seek to find the probability of taking any given branch, given whatever
-information is needed: $\prob(\caseset_t|\cdot)$.  In finding these
-probabilities, we do not attempt to derive any distribution from the PSO
-equations, we simply look at empirical distributions.  However, even with
-empirical distributions, the problem with this approach is that it is not clear
-what information influences the probability of taking a branch.  We look at two
-factors that we believe have a significant influence on $\prob(\caseset_t)$:
-topology ($T$) and function ($F$).  Thus we are looking at
-$\prob(\caseset_t|T,F)$.
-
-We show in Table~\ref{tab:stats} with what percentage a particle takes each of
-these branches for three different topologies and four different functions.
-All of our statistics are from swarms of 240 particles.  Brief experimentation
-showed that other swarm sizes had similar statistics.  We ran 750 iterations on
-all combinations of functions and topologies except for the functions Griewank
-and Rastrigin with the Complete topology.  We found that those runs frequently
-converged past machine precision after 500 iterations, and that led to
-erroneously high values for the probability of \casepn.  Instead we ran for
-only 450 iterations on those two combinations.  All of our results were
-averaged over 20 runs of the algorithm; thus the probabilities presented are
-the averages of 3.6 million trials for the branch taken (2.16 million for the
-two with only 450 iterations).  Table~\ref{tab:stats} contains the results.
-The definitions for all of the functions in the table are found in \sec{setup}.
-
-The probabilities presented in Table~\ref{tab:stats} are interesting in and of
-themselves and could probably be used to better understand the characteristics
-of various topologies.  It is notable that there is small variation between
-functions in any given topology, but the variation across topologies is far
-greater.  However, our concern is with speculative evaluation.  We are
-interested in predicting the branch that any given particle will take at a
-particular iteration.  For our purposes, it appears that given a topology, the
-probability of selecting a branch and the function are close to independent, or
-$\prob(\caseset_t|T,F) \approx \prob(\caseset_t|T)$.
-
-From Table~\ref{tab:stats} we can see that with the Random topology, we can
-pick the first branch, corresponding to stagnation, and be right around 70\% of
-the time.  With the Ring topology, we would be right 50\% of the time.
-Branches \casepN\ and \casexN\ really correspond to several actual branches, as
-all of the neighbors are lumped together.  The 20\% probability of taking
-branch \casepN\ with the Ring topology can be split into two branches, as there
-are only two neighbors.  It also turns out that the neighbor that last updated
-the neighborhood best is the most likely to update it next time, so keeping
-track of that information could be fruitful in trying to predict that branch.
-
-The statistics for the Complete topology are less promising, as there are 240
-neighbors that branch \casepN\ splits into, instead of two.  Pruning does,
-however, allow for the possibility of using the Complete topology in
-speculative evaluation while avoiding the explosion in the number of
-evaluations it would otherwise entail.
-
-Becuase there are a few branches with very high probabilities in the topologies
-we are interested in, we can have hope that cutting out some of the evaluations
-that have low probability will lead to an increase in overall performance.  In
-order to implement this kind of pruning, the only change that needs to be made
-to \alg{centralized} is in Step 2, where the speculative children are
-generated.  Instead of generating all possible speculative children, generate
-the subset of the children that is desired.
-
-\subsection{Recovering from Pruning Too Much}
-\label{sec:wrong}
-
-When not all of the branches are evaluated, there is some probability of not
-evaluating the branch that was actually taken by the original particle.  There
-are a few possibilities for recovery in this case.  One is to leave the
-particle as it is, not accepting any of the speculative evaluations, because
-none of them were correct.  This leads to particles being at different
-iterations, as some particles guess correctly while others do not.  Thus we
-lose exact compatibility with the original PSO, though this particular
-relaxation is nothing new; asynchronous adaptations of PSO do the same
-thing~\citep{koh-2006-parallel-asynchronous-pso}.  As an aside, it is equivalent
-in this case to simply increment the iteration number of particles which fail
-to correctly predict their branch.  This keeps the iteration number constant
-across all particles, simplifying the work needed to be done in determining
-neighbors when dynamic topologies are involved.  We call this technique Social
-Promotion.
-
-Another possibility is to pick the best child, as described in \sec{pickbest}.
-This ignores the fact that the branch might have been wrong; it does not
-matter, because we simply accept the child that had the best value.  In most of
-the experiments that we ran, it turned out that picking the best child
-performed better than keeping the particle back an iteration.
-
-\subsection{More than one iteration ahead}
-\label{sec:manyiters}
-
-We need not simply produce speculative children for the next iteration.  We can
-view all possible speculative evaluations for a particle as an infinite tree
-with branching factor $2n+1$.  As we have already seen that doing one full
-level of the tree produces too many extra evaluations to be profitable, it is
-incredibly unlikely that doing two full iterations would produce decent
-results.  But, if the idea of speculating more than one iteration ahead is
-combined with wisely pruning the possible evaluations based on branch
-statistics, we can use just a few extra evaluations to go two or more
-iterations ahead on the most likely branches.
-
-When speculating more than one iteration ahead, the idea of Social Promotion
-cannot feasibly be implemented, as we can only determine correct branches for
-the first iteration.  Thus in this case we always pick the child that has the
-highest value.
-
-The question of which branches to take in this infinite tree is an intriguing
-one that we can only begin to explore here.  If the branch corresponding to
-stagnation has a 75\% chance of being taken, as in the Random topology with
-most functions, we could speculate three iterations ahead on that branch and
-still have a 42\% chance of predicting correctly.  However, intuition would say
-that perhaps it is better to hope that the particle is productive instead of
-stagnant, so a branch where the particle updates its personal best might be
-more fruitful to try.  In our experiments we try just one of the countless
-possibilities, but one that turns out to work very well.  More work is needed
-to compare the different branching possibilities on various functions.
-
-\section{Experimental Setup}
-\label{sec:setup}
-
-We ran experiments to compare our speculative parallelization of PSO to the
-standard parallelization.  At each iteration of the algorithms, we use one
-processor to perform one function evaluation for one particle, be it
-speculative or non-speculative.  The exact evaluation time at which this
-architecture becomes reasonable depends on the amount of communication overhead
-in the parallel implementation and the number of particles in the swarm.  For
-our implementation we found that time to be around 100 milliseconds for swarms
-of 50 particles.  When the swarm size increases, the minimum evaluation time at
-which this architecture should be used decreases.
-
-\subsection{Function Evaluations vs. Time Steps}
-
-Results in serial PSO are typically presented in terms of function evaluations.
-This is a natural abstraction from implementation details that still allows a
-comparison of the implementation-independent aspects of each algorithm.  The
-number of function evaluations performed is assumed to be proportional to
-wall-clock time, as all evaluations are performed in serial.  Only reporting
-function evaluations could hide the fact that one algorithm requires more
-overhead than another and thus actually takes more time to perform the same
-number of function evaluations; however, function evaluations are still
-considered the standard method of reporting, as evaluation times for functions
-also vary greatly and could make the additional overhead negligible.
-
-In parallel PSO on long-running functions, the natural way to present results
-is in terms of iterations, not function evaluations.  This is because when all
-function evaluations at each iteration are performed concurrently, iterations
-are the direct equivalent of wall-clock time.  It is true that with a fixed
-number of processors, function evaluations are also equivalent to wall-clock
-time.  However, reporting function evaluations in a paper on parallel PSO can
-be misleading if compared to a paper on serial PSO, as the actual wall-clock
-times corresponding to the function evaluations are not directly comparable.
-Thus we report iterations in our results instead of function evaluations.  But
-because SEPSO actually performs two iterations of PSO at each ``iteration,''
-and Social Promotion and Many Iterations make the idea of ``iterations of PSO''
-somewhat nebulus, we instead call each ``iteration'' a ``time step.''
-
-Just as serial PSO papers do not report actual running times of their specific
-implementations, we do not report running times, favoring the abstraction of
-time steps.  That being said, given the time required for the evaluation of the
-objective function and the communication overhead per iteration for a specific
-implementation, an actual running time can be recovered from all of our
-results.  Simply multiply the number of time steps by the sum of the function
-evaluation time and the overhead.  We wish to stress that the ``time steps'' we
-report are proportional to wall-clock time, given the architecture we have
-assumed.
-
-\subsection{Functions Used}
-
-We experimented with five common benchmark problems defined
-in~\citep{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
-and Bohachevsky.
-
-Rastrigin is initialized in $[-5.12,5.12]^D$ and is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
-10\right), z=x-c\]
-
-Sphere is initialized in $[-50,50]^D$ and is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]
-
-Schwefel 2.21 is initialized in $[-500,500]^D$ and is defined as:
-\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]
-
-Griewank is initialized in $[-600,600]^D$ and is defined as:
-\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
-\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]
-
-And Bohachevsky is initialized in $[-15,15]^D$ and is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
-x_{i+1})+.7)\]
-
-In computing the branch statistics in \sec{statistics}, we also used the
-Rosenbrock function.  That function is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)\]
-
-The $c$ in the function definitions is a shifted center, in order to avoid
-origin-seeking bias in the PSO algorithm and its
-variants~\citep{monson-2005-origin-seeking-bias}.  We move the center of the
-shifted functions (all except Bohachevsky) to be halfway between the center and
-the boundary of the initialization region.  For example, if the initialization
-region is $[-50,50]^D$, the center is $(25)^D$.  We tested all of these
-functions in their 20 dimension, 50 dimension, and 500 dimension varieties.
-
-One might question the use of benchmark functions for evaluating our methods,
-as they take on the order of 1 millisecond to evaluate instead of the 100s that
-we said were required to justify our parallel architecture.  The purpose of
-exploring the behavior of an algorithm on benchmark functions, whether in
-serial or in parallel optimization research, is because benchmark functions are
-believed to represent functions that one might encounter in the real world,
-regardless of the amount of time those functions take to evaluate.  While
-benchmark functions themselves take fractions of a second to evaluate and thus
-have no need of the kind of parallelization we explore in this paper, they also
-have no need of particle swarm optimization at all, as they can be solved
-analytically.  Yet they are useful to us and to optimization researchers
-generally because they stand as surrogates for the kinds of functions
-practitioners are actually interested in, and they allow us to explore the
-behavior of optimization algorithms on several different classes of functions
-in an easy and standardized way.
-
-To aid in convincing the reader that performance on benchmark functions roughly
-corresponds to performance on real-world problems, we also tested our
-parallelization methods on a typical research problem, that of fitting a model
-to a large quantity of data.  We generated 10,000 data points from a radial
-basis function network with 10 bases and some added Gaussian noise.  We then
-fit a radial basis function network to the data using PSO.  This amounted to a
-30 dimensional function to optimize, with a function evaluation time on the
-order of two seconds.  We refer to this function as ``the model fitting
-problem.''
-
-\subsection{Parallelization Techniques Compared}
-
-The parallelization techniques we compare are the standard parallelization
-(here labeled Standard simply for ease of reference), our original speculative
-approach (recall that we refer to this as SEPSO), and the four relaxations of
-SEPSO discussed in \sec{relax}.  In presenting our results, we call the
-approach developed in \sec{pickbest} Pick Best.  The methods described in
-\sec{pruning} through \sec{wrong} are called Pick Best Pruned and Social
-Promotion Pruned, and the method in \sec{manyiters} we call Many Iterations.
-
-Using the same number of processors for each approach (and thus the same number
-of function evaluations per time step) requires that our speculative
-parallelizations have a smaller swarm size than the standard parallelization.
-For the topologies we used with SEPSO and Pick Best, a particle has three
-neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
-$7$ speculative evaluations per particle.  With one evaluation needed for the
-original, non-speculative particle, we have $8p$ evaluations for every two
-iterations, where $p$ is the number of particles in the speculative swarm.  The
-extra evaluations required in our speculative approach would instead be used to
-evaluate particles in standard parallelizations, so we compare swarms of size
-$p$ in speculative evaluation with swarms of size $8p$ in standard approaches.
-
-When performing pruning in Pick Best Pruned and Social Promotion Pruned, there
-are a large number of ways to prune speculative evaluations.  We experimented
-with several, but present results for only one possible pruning.  The pruning
-we present uses only the two branches where the $\nbest$ value was not updated:
-\casepn\ and \casexn.  Those branches are convenient in that no messages are
-needed from neighbors in order to produce the positions of the speculative
-particles---in distributed frameworks using several rounds of communication
-(see \sec{distributed}), one of the rounds of communication can be dropped
-entirely.  Pruning all but these branches also allows the use of arbitrarily
-dense topologies, as the number of speculative particles is no longer dependent
-on the number of neighbors the particle has.
-
-Because pruning only requires two speculative evaluations per particle (along
-with evaluating the original particle), we can use swarms of size
-$\frac{1}{3}p$ when pruning to compare to a swarm of size $p$ with the standard
-parallelization, instead of $\frac{1}{8}p$ with other techniques.
-
-There are also many ways to speculate several iterations ahead, and with Many
-Iterations we again only show results for one of them.  The combination of
-branches we tried uses seven speculative evaluations per particle, matching the
-swarm size of the original speculative algorithm.  The seven evaluations we
-used corresponded to several iterations of branches \casepn\ and \casexn.  Two
-of the evaluations were just one iteration ahead, four were two iterations
-ahead, and one was three iterations ahead.  The evaluations that were one
-iteration ahead were branches \casepn\ and \casexn; those that went two
-iterations ahead were formed by taking either branch \casepn\ or \casexn and
-then branch \casepn\ or \casexn; and the evaluation going three iterations
-ahead followed branch \casepn\ on all three iterations.  As with Pick Best
-Pruned and Social Promotion Pruned, this choice of branches allows the use of
-arbitrarily dense topologies, as the number of speculative evaluations per
-particle is independent of the number of neighbors the particle has.
-
-\subsection{Topologies}
-
-For each benchmark function we report results using the topology that is widely
-considered best for that function, as reported in the literature
-(e.g.,~\cite{bratton-2007-defining-a-standard-for-pso}).  In this paper we
-limit ourselves to the Ring topology and the Complete topology, as is common
-practice, along with the Random topology (the parallel approximation to the
-Complete topology mentioned in \sec{pso}).  We also mentioned in \sec{related}
-that some related work can be described as changes in topology, particularly
-that of having subswarms of fully connected topologies that occasionally
-communicate with each other.  This related work focused mainly on functions
-with very fast evaluation times where such techniques drastically reduce
-interprocessor communication.  With long function evaluations these topologies
-are not as practical, as only one particle is evaluated by each processor at
-each iteration.  We experimented with a subswarm topology and found that in
-every instance except one it performed worse than either Complete or Ring, and
-thus we leave it out of the results except in the one instance where it
-improved performance.
-
-Where the Complete topology would normally be preferred, we use a Random
-topology in SEPSO and Pick Best, as Complete leads to an explosion in the
-number of speculative evaluations (as noted, the other speculative techniques
-can still use a Complete topology; we often report results on both topologies
-for those methods).  If speculative evaluation were not being performed, it is
-possible that the Complete topology would be used.  However, the Complete
-topology also requires a very large amount of interprocessor communication in
-distributed PSO, so it is still quite possible that Random would be used even
-with standard parallelizations~\citep{mcnabb-2009-large-particle-swarms}.  But,
-to be fair in our comparisons, we compare to the standard parallelization using
-both the Random topology and the Complete topology (labeled PSO Random and PSO
-Complete in our results).  Again, the amount of communication overhead is
-heavily dependent on implementation details which are not the focus of this
-paper.  A practitioner using a particular implementation can compare the
-results given here for whichever topology is most practical given the specific
-implementation.
-
-\section{Results}
-\label{sec:results}
-
-We frequently present tables summarizing our results.  In each table, we bold
-the ``best'' method, meaning that it has at least a 90\% success rate and its
-average time to completion is faster than all other methods that have at least
-a 90\% success rate.  We used a double-sided t-test to test for statistical
-significance in these results, and all bolded values are statistically
-significant with a p-value of less than $10^{-5}$.  In some instances two
-values are bolded because they are statistically tied, while both are
-significantly better than all other methods.
-
-We first present results in \sec{20dims} for the 20 dimensional variants of the
-benchmark functions we tested.  We give a discussion of each function
-individually at 20 dimensions, as there are interesting characteristics of the
-algorithms that are worth discussing.  In \sec{50dims} we then give a summary
-of results for the 50 dimensional variants of the benchmark functions; we do
-not go into as much detail in our discussion as the results are very similar to
-those in 20 dimensions.  We finish our discussion of benchmark functions with
-their 500 dimensional variants in \sec{500dims}, and we discuss results on the
-model fitting problem in \sec{model}.
-
-\subsection{20 Dimensions}
-\label{sec:20dims}
-
-\subsubsection{Sphere}
-
-First we look at Sphere, the simplest of common benchmark functions.  The
-function has a single global optimum and no other local optima.  Sphere is best
-optimized in terms of function evaluations with a small swarm using a Complete
-topology.  We expect our methods to be perfect for such functions, and our
-results show this intuition to be correct.  For this comparison we used 240
-processors, so the methods had swarms of size 30 (for SEPSO, Pick Best, and
-Many Iterations), 80 (Pick Best Pruned and Social Promotion Pruned), and 240
-(Standard).  We compared too many methods here to fit into one graph, so we
-show just a few methods in \fig{sphere-20} and in \tabref{sphere-20} we show a
-summary of the results for all methods we tested.
-
-In \fig{sphere-20} we see that Many Iterations using a Complete topology
-converges incredibly quickly on a very poor value.  We found this behavior to
-be quite consistent across functions for this method, so we rarely show results
-for Many Iterations Complete.  However, Pick Best Pruned works very well with
-Complete on this function.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{sphere20}
-  \caption{Function Sphere with a swarm that uses 240 processors per time step,
-  comparing a pertinent subset of possible methods.}
-  \label{fig:sphere-20}
-\end{figure}
-
-\begin{table}
-  \caption{Summary of results for function Sphere with 20 dimensions,
-  measuring number of time steps to reach a value of 1e-35}
-  \label{tab:sphere-20}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&0\%&N/A&N/A\\
-  \hline
-  Standard Random&0\%&N/A&N/A\\
-  \hline
-  Standard Complete&100\%&931.0&19.1\\
-  \hline
-  SEPSO Ring&0\%&N/A&N/A\\
-  \hline
-  SEPSO Random&85\%&972.8&15.8\\
-  \hline
-  Pick Best Ring&100\%&917.4&21.0\\
-  \hline
-  Pick Best Random&100\%&768.1&14.3\\
-  \hline
-  Pick Best Pruned Ring&100\%&967.8&9.9\\
-  \hline
-  Pick Best Pruned Random&100\%&693.5&10.4\\
-  \hline
-  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{389.6}&
-    \textbf{9.6}\\
-  \hline
-  Social Promotion Pruned Ring&0\%&N/A&N/A\\
-  \hline
-  Social Promotion Pruned Random&100\%&971.1&10.6\\
-  \hline
-  Social Promotion Pruned Complete&100\%&777.8&16.7\\
-  \hline
-  Many Iterations Ring&100\%&575.2&9.8\\
-  \hline
-  Many Iterations Random&100\%&442.4&7.1\\
-  \hline
-  Many Iterations Complete&0\%&N/A&N/A\\
-  \hline
-  \end{tabular}
-\end{table}
-
-\subsubsection{Schwefel 2.21}
-
-Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
-more by larger swarms than Sphere is.  Thus our speculative algorithms often
-fail to outperform Standard with 240 processors because the simple speculative
-methods only have 30 particles.  As we will show later, when we use 800
-processors at 50 dimensions, our methods perform much better.  However, the
-pruned versions of our methods have 80 particles and thus are able to
-outperform Standard even with only 240 processors.  Because the graph of
-Schwefel 2.21 looks very similar to that of Sphere, we simply present all of
-our results in \tabref{schwefel-20}.
-
-\begin{table}
-  \caption{Summary of results for function Schwefel with 20 dimensions,
-  measuring number of time steps to reach a value of 1e-06}
-  \label{tab:schwefel-20}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&0\%&N/A&N/A\\
-  \hline
-  Standard Random&0\%&N/A&N/A\\
-  \hline
-  Standard Complete&100\%&837.2&44.8\\
-  \hline
-  SEPSO Random&0\%&N/A&N/A\\
-  \hline
-  Pick Best Random&5\%&938.0&0.0\\
-  \hline
-  \textbf{Pick Best Pruned Random}&\textbf{100\%}&\textbf{639.9}&\textbf{29.4}\\
-  \hline
-  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{597.7}&
-	\textbf{77.9}\\
-  \hline
-  Social Promotion Pruned Random&75\%&938.0&33.0\\
-  \hline
-  Social Promotion Pruned Complete&100\%&815.2&43.5\\
-  \hline
-  Many Iterations Random&100\%&783.5&94.5\\
-  \hline
-  Many Iterations Complete&0\%&N/A&N/A\\
-  \hline
-  \end{tabular}
-\end{table}
-
-\subsubsection{Rastrigin}
-
-Rastrigin is a multi-modal function that is best optimized with a large,
-Complete swarm.  It has been shown that with Rastrigin, the more particles
-there are in the swarm, the lower function value it finds, up to at least 4000
-particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
-in local optima and converge to poorer values.  Because our speculative
-algorithms require significantly smaller swarm sizes, we would expect to not
-perform very well on functions such as Rastrigin.  Our experiments show our
-intuition to be correct.  In this experiment we used 240 processors, so SEPSO
-and Pick Best each had 30 particles, and the Standard algorithms had 240
-particles.  As expected, SEPSO and Pick Best converge quickly to worse local
-optima than Standard does.  \fig{rastrigin} shows the results graphically, and
-\tabref{rastrigin-20} shows results for all of the methods we tried.
-
-\begin{table}
-  \caption{Summary of results for function Rastrigin with 20 dimensions,
-  measuring number of time steps to reach a value of 20}
-  \label{tab:rastrigin-20}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&25\%&743.4&212.0\\
-  \hline
-  Standard Random&100\%&372.1&86.6\\
-  \hline
-  \textbf{Standard Complete}&\textbf{95\%}&\textbf{273.1}&\textbf{99.7}\\
-  \hline
-  SEPSO Ring&5\%&471.0&0.0\\
-  \hline
-  SEPSO Random&15\%&208.0&17.0\\
-  \hline
-  Pick Best Ring&35\%&489.0&152.0\\
-  \hline
-  Pick Best Random&15\%&136.0&40.1\\
-  \hline
-  Pick Best Pruned Ring&95\%&545.9&208.5\\
-  \hline
-  Pick Best Pruned Random&10\%&95.0&11.0\\
-  \hline
-  Pick Best Pruned Complete&5\%&59.0&0.0\\
-  \hline
-  Social Promotion Pruned Ring&5\%&580.0&0.0\\
-  \hline
-  Social Promotion Pruned Random&80\%&278.2&136.3\\
-  \hline
-  Social Promotion Pruned Complete&50\%&200.5&130.2\\
-  \hline
-  Many Iterations Ring&10\%&117.0&25.0\\
-  \hline
-  Many Iterations Random&5\%&57.0&0.0\\
-  \hline
-  Many Iterations Complete&0\%&N/A&N/A\\
-  \hline
-  \end{tabular}
-\end{table}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{rastrigin}
-  \caption{Function Rastrigin with a swarm that uses 240 processors per time
-  step.  Recall that at each time step all processors evaluate the function
-  once.  All speculative algorithms use the Random topology.}
-  \label{fig:rastrigin}
-\end{figure}
-
-\subsubsection{Griewank}
-\label{sec:griewank-20}
-
-It is generally recommended to use the Ring topology when optimizing the
-Griewank function, as Complete is prone to premature convergence on a local
-optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  The global optimum
-of Griewank has a value of 0.  When most trials reach the global optimum but a
-few get stuck, the resultant ``average value'' graph has a flat line that is
-misleading.  Thus instead we present plots showing the percent of trials that
-have passed some threshold at each time step, as is common practice with these
-functions~\citep{mendes-2004-population-topologies-in-pso}.  The threshold we
-chose for this case was $10^{-6}$, as that value is below any local optima and
-the swarm always successfully reaches 0 once is passes that point.
-
-We ran 50 trials of each experiment with Griewank, so that the curves are more
-smooth.  We show results in \fig{basic-griewank1} for swarms of size 100 and
-800 using the Ring topology.  One can see in the figure that SEPSO reaches the
-global optimum on average close to twice as fast as Standard, while Pick Best
-is close to three times as fast and Many Iterations is even faster when it is
-successful, though it is only successful 84\% of the time.  \tabref{griewank}
-shows all of our results in tabular form.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{griewank6}
-  \caption{Function Griewank with a swarm that uses 800 processors per time
-  step.  Recall that at each time step all processors evaluate the function
-  once.  Instead of showing average function value, we show the percentage of
-  runs that have found the global optimum by each iteration.  All algorithms
-  use the Ring topology.}
-  \label{fig:basic-griewank1}
-\end{figure}
-
-\begin{table}
-  \caption{Summary of results for function Griewank with 20 dimensions,
-  measuring number of time steps to reach a value of 1e-06}
-  \label{tab:griewank-20}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&100\%&762.2&19.4\\
-  \hline
-  SEPSO Ring&100\%&426.0&81.0\\
-  \hline
-  \textbf{Pick Best Ring}&\textbf{100\%}&\textbf{272.0}&\textbf{17.7}\\
-  \hline
-  \textbf{Pick Best Pruned Ring}&\textbf{100\%}&\textbf{282.9}&\textbf{10.7}\\
-  \hline
-  Social Promotion Pruned Ring&100\%&482.0&16.6\\
-  \hline
-  Many Iterations Ring&84\%&183.2&49.4\\
-  \hline
-  \end{tabular}
-\end{table}
-
-We pause here to show some interesting characteristics of the speculative
-techniques we have proposed.  With 800 processors our methods perform very well
-on this function.  With 240 processors, however, the results are much more
-mixed.  Because 240 processors is near the point at which speculative
-evaluation becomes useful, it is enlightening to see the behavior of the
-various algorithms at this point.
-
-We show results in \fig{basic-griewank2} for swarms of size 30 and 240 using
-the Ring topology.  One can see in the figure that when SEPSO is successful, it
-finds the optimum much faster than Standard.  However, because the swarm size
-is so small, SEPSO gets stuck a little less than half of the time.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{griewank1}
-  \caption{Function Griewank with a swarm that uses 240 processors per time
-  step.  All algorithms use the Ring topology.}
-  \label{fig:basic-griewank2}
-\end{figure}
-
-When we look at the performance of our Pick Best approach, we see that it
-greatly improves performance on Griewank.  This is somewhat counter-intuitive,
-because Griewank is deceptive and Pick Best seems like a greedy algorithm.  But
-in \fig{griewank-pickbest} we see that Pick Best improves accuracy over SEPSO
-by 20\%, while at the same time finding the optimum over 100 time steps sooner
-on average.  It seems that while Pick Best is locally greedy, there is enough
-exploration in the seven speculative evaluations to overcome the inherent
-greediness of the approach.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{griewank3}
-  \caption{Function Griewank with a swarm that uses 240 processors per time
-  step, comparing Pick Best with results from \fig{basic-griewank2}.  All
-  algorithms use the Ring topology.}
-  \label{fig:griewank-pickbest}
-\end{figure}
-
-When we introduce pruning, our intuition about Pick Best turns out to be more
-correct.  While adding 50 more particles to the swarm (as pruning allows us to
-have 80 particles with 240 processors instead of only 30), Pick Best with
-pruning still gets stuck just as often as the original Pick Best.  However,
-Social Promotion does well with pruning; it increases the success rate to close
-to 100\%, while still finding the optimum on average much faster than the
-original PSO.  These results are shown in \fig{griewank-pruned}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{griewank4}
-  \caption{Function Griewank with a swarm that uses 240 processors per time
-  step, comparing pruning with the best results from \fig{griewank-pickbest}.
-  All algorithms use the Ring topology.}
-  \label{fig:griewank-pruned}
-\end{figure}
-
-With Griewank, the premature convergence problems inherent in picking the best
-child are exacerbated when speculating several iterations ahead.  When Many
-Iterations finds the optimum, it finds it quicker than any other method we
-tried, on average four times faster than Standard.  However, it also gets stuck
-and fails to find the optimum more than any other method.  The results are
-shown in \fig{griewank-manyiters}.   \fig{griewank-manyiters} is also
-interesting in that it highlights the trade-off between accuracy and speed in
-the various approaches at this swarm size.  The faster the approach finds the
-optimum, the less likely it is to be successful.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{griewank5}
-  \caption{Function Griewank with a swarm that uses 240 processors per time
-  step, comparing speculating several iterations ahead with selections from
-  \fig{griewank-pruned}.  All algorithms use the Ring topology.}
-  \label{fig:griewank-manyiters}
-\end{figure}
-
-\subsubsection{Bohachevsky}
-
-Bohachevsky is a unimodal function best optimized with a Complete swarm.  It is
-similar to Griewank in that there is a global optimum with a value of 0, and
-the swarm sometimes finds it and sometimes does not.  Thus we present a graph
-similar to those of Griewank, because ``average value'' graphs have a
-misleading flat line.  In \fig{bohachevsky} we show a plot with a few pertinent
-methods, while \tabref{bohachevsky-20} shows all of the results.  All of our
-speculative approaches found the optimum much quicker than Standard with a
-Random topology.  However, SEPSO was slower than Standard Complete and got
-stuck 25\% of the time.  Pick Best, Pick Best Pruned, and Many Iterations all
-outperformed Standard Complete, with Many Iterations finding the optimum about
-twice as fast.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{bohachevsky}
-  \caption{Function Bohachevsky with a swarm that uses 480 processors per time
-  step.  Recall that at each time step all processors evaluate the function
-  once.}
-  \label{fig:bohachevsky}
-\end{figure}
-
-\begin{table}
-  \caption{Summary of results for function Bohachevsky with 20 dimensions,
-  measuring number of time steps to reach a value of 1e-06.}
-  \label{tab:bohachevsky-20}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Random&100\%&472.3&6.6\\
-  \hline
-  Standard Complete&100\%&238.2&11.2\\
-  \hline
-  SEPSO Random&75\%&261.6&9.2\\
-  \hline
-  Pick Best Random&100\%&211.8&9.2\\
-  \hline
-  Pick Best Pruned Random&100\%&189.7&3.4\\
-  \hline
-  Pick Best Pruned Complete&25\%&105.4&3.9\\
-  \hline
-  Social Promotion Pruned Random&100\%&260.3&6.5\\
-  \hline
-  Social Promotion Pruned Complete&90\%&197.9&12.2\\
-  \hline
-  \textbf{Many Iterations Random}&\textbf{94\%}&\textbf{118.2}&\textbf{2.6}\\
-  \hline
-  Many Iterations Complete&0\%&N/A&N/A\\
-  \hline
-  \end{tabular}
-\end{table}
-
-\subsection{50 Dimensions}
-\label{sec:50dims}
-
-Given our observation from the 20 dimensional benchmark functions that Many
-Iterations most often outperforms other speculative approaches, we only present
-results for Many Iterations and Standard for the 50 and 500 dimensional
-variants.  For these experiments we used 800 processors, as the problems are
-more difficult and speculative approaches perform best when many processors are
-available, as we showed in \sec{griewank-20}.  Many Iterations Complete showed
-the same premature convergence that was observed with the 20 dimensional
-benchmark functions, so we do not include those results in our tables; the
-success rate was 0\% for all functions.
-
-Tables~\ref{tab:rastrigin-50} through~\ref{tab:bohachevsky-50} present
-summaries of our results for each of the benchmark functions. As with the 20
-dimensional variants, our speculative methods did not perform well on
-Rastrigin, though for all other benchmarks Many Iterations significantly
-outperformed Standard.
-
-A summary of the results is shown in \tabref{50dims-summary}.  For functions
-where Many Iterations outperformed Standard, we report the average speed up.
-On all functions except Rastrigin, our methods showed an average speed up of
-from two to six times compared to previous methods.
-
-\begin{table}
-  \caption{Summary of results for function Rastrigin with 50 dimensions,
-  measuring number of time steps to reach a value of 100}
-  \label{tab:rastrigin-50}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&5\%&776.0&0.0\\
-  \hline
-  \textbf{Standard Random}&\textbf{100\%}&\textbf{593.2}&\textbf{89.5}\\
-  \hline
-  Standard Complete&80\%&343.8&145.8\\
-  \hline
-  Many Iterations Ring&25\%&223.8&67.1\\
-  \hline
-  Many Iterations Random&25\%&98.6&11.0\\
-  \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \caption{Summary of results for function Sphere with 50 dimensions,
-  measuring number of time steps to reach a value of 1e-06}
-  \label{tab:sphere-50}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&0\%&N/A&N/A\\
-  \hline
-  Standard Random&30\%&986.8&8.2\\
-  \hline
-  Standard Complete&100\%&458.4&12.3\\
-  \hline
-  Many Iterations Ring&100\%&288.7&3.9\\
-  \hline
-  \textbf{Many Iterations Random}&\textbf{100\%}&\textbf{248.6}&\textbf{5.1}\\
-  \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \caption{Summary of results for function Schwefel with 50 dimensions,
-  measuring number of time steps to reach a value of 80}
-  \label{tab:schwefel-50}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&0\%&N/A&N/A\\
-  \hline
-  Standard Random&70\%&871.6&57.1\\
-  \hline
-  Standard Complete&100\%&599.0&85.6\\
-  \hline
-  \textbf{Many Iterations Ring}&\textbf{100\%}&\textbf{238.6}&\textbf{34.2}\\
-  \hline
-  \textbf{Many Iterations Random}&\textbf{100\%}&\textbf{242.9}&\textbf{95.4}\\
-  \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \caption{Summary of results for function Griewank with 50 dimensions,
-  measuring number of time steps to reach a value of 1e-06}
-  \label{tab:griewank-50}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&100\%&1898.5&25.7\\
-  \hline
-  \textbf{Many Iterations Ring}&\textbf{97\%}&\textbf{312.4}&\textbf{6.3}\\
-  \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \caption{Summary of results for function Bohachevsky with 50 dimensions,
-  measuring number of time steps to reach a value of 0.01}
-  \label{tab:bohachevsky-50}
-  \centering
-  \begin{tabular}{|l|r|r|r|}
-  \hline
-  Algorithm&\% Complete&Mean&St. Dev.\\
-  \hline
-  \hline
-  Standard Ring&100\%&1434.0&62.1\\
-  \hline
-  Standard Random&100\%&704.5&19.7\\
-  \hline
-  Standard Complete&20\%&377.5&14.9\\
-  \hline
-  \textbf{Many Iterations Ring}&\textbf{95\%}&\textbf{229.8}&\textbf{11.4}\\
-  \hline
-  Many Iterations Random&35\%&171.6&4.0\\
-  \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \caption{Average speed up, comparing the number of time steps to completion
-  for the best speculative topology to the time steps to completion for the
-  best standard topology.  All functions have 50 dimensions.}
-  \label{tab:50dims-summary}
-  \centering
-  \begin{tabular}{|l|r|}
-	\hline
-	Function&Speed Up Factor\\
-	\hline
-	Rastrigin&N/A\\
-	\hline
-	Sphere&1.84\\
-	\hline
-	Schwefel&2.51\\
-	\hline
-	Griewank&6.08\\
-	\hline
-	Bohachevsky&3.07\\
-	\hline
-  \end{tabular}
-\end{table}
-
-\subsection{500 Dimensions}
-\label{sec:500dims}
-
-At 500 dimensions we found that the performance of constricted PSO on benchmark
-functions becomes rather dismal.  In almost all cases, Standard fails to make
-any significant progress.  Our search for an explanation seems to say that the
-space is too large for particles to converge to the same point, so all updates
-to the best position found come from a single particle wandering in the space.
-Every time that single particle finds a new best position, its velocity
-contracts, so the particle is unable to make significant progress on its own.
-In constricted PSO it is necessary to have a collection of particles exploring
-a promising location to keep the particles' velocities from contracting
-prematurely, and in 500 dimensions the space is too large to get the collection
-of particles to the same location.
-
-The results that we present show that while Standard parallelizations suffer
-from this problem, Many Iterations does not.  Though we can provide some
-intuition for why this might be the case, it remains as future work to provide
-a strong theoretical explanation.  The intuition is that we are speculating
-about future positions along paths where each particle's velocity does not
-contract.  The particular branches we selected corresponded to not having
-received a new value for $\pbest$ nor $\nbest$.  In those branches, the
-particle continues in the same direction it was going, without a contraction in
-velocity.
-
-Because none of the Standard approaches were successful at optimizing the 500
-dimensional benchmark functions, we do not present tables as we did for the 50
-dimensional variants.  We do, however, show a few figures demonstrating the
-results we have just explained.  \fig{sphere500} shows the function Sphere,
-\fig{griewank500} shows Griewank, and \fig{rastrigin500} shows Rastrigin.  Note
-that at 500 dimensions speculative approaches outperform standard
-parallelizations on Rastrigin, whereas they did not at 20 and 50 dimensions.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{sphere500}
-  \caption{Results for the 500 dimensional Sphere function.  We use 800
-  processors for this experiment, so each time step corresponds to 800 function
-  evaluations.}
-  \label{fig:sphere500}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{griewank500}
-  \caption{Results for the 500 dimensional Griewank function.  We use 800
-  processors for this experiment, so each time step corresponds to 800 function
-  evaluations.}
-  \label{fig:griewank500}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{rastrigin500}
-  \caption{Results for the 500 dimensional Rastrigin function.  We use 800
-  processors for this experiment, so each time step corresponds to 800 function
-  evaluations.}
-  \label{fig:rastrigin500}
-\end{figure}
-
-One could argue that a fairer comparison would have modified the PSO motions
-equations so that individual particles did not contract their velocity so
-quickly when finding good locations.  However, we used the same motion
-equations for both Standard and Many Iterations; changing the motion equations
-to help Standard would also help Many Iterations, and we would expect to see
-results that are at least as compelling as those from the 50 dimensional
-benchmark functions.
-
-\subsection{Model Fitting}
-\label{sec:model}
-
-For the model fitting problem we used 144 processors.  We show results for
-three methods: Standard with a Random topology, Standard with a Subswarm
-topology, and Many Iterations with a Random topology.  Standard with a Subswarm
-topology had eight independent subswarms of 18 particles each, and the
-particles in each subswarm were connected with a Random topology.  In the
-graphs we call this method Standard Subswarms.
-
-\fig{rbf} shows our results for this function.  The function value reported is
-the sum squared error of the model fit.  \fig{rbf-stuck} shows the percent of
-runs that reached a value for sum squared error of 55,000 by each time step,
-which we designate as successful.  Many Iterations took on average 126 time
-steps to reach this value, while Standard Subswarms took on average 298.  The
-p-value for a t-test on this statistic is less than $10^{-8}$.  With our
-implementation of parallel PSO, each time step took on average 10.06 seconds;
-1.83 seconds of that was function evaluation time and the rest was overhead.
-With this particular implementation, then, Many Iterations takes on average 21
-minutes to reach a successful value, while Standard takes 50 minutes.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{rbf}
-  \caption{Results for fitting a radial basis function to noisy data.  We use
-  144 processors for each method, so each time step corresponds to 144 function
-  evaluations.}
-  \label{fig:rbf}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{rbf_stuck}
-  \caption{Results for fitting a radial basis function to noisy data.  We use
-  144 processors for each method, so each time step corresponds to 144 function
-  evaluations.}
-  \label{fig:rbf-stuck}
-\end{figure}
-
-\section{Conclusions}
-\label{sec:conclusion}
-
-We have described a new technique for using processors in parallel PSO to
-improve the performance of the algorithm.  To our knowledge, this is the first
-time extra processors have been used to do anything in PSO besides increase the
-swarm size.  In an increasingly parallel world, such advancements will prove to
-be crucial to the continued effectiveness of PSO.
-
-We have detailed how to perform speculative evaluation in PSO in several
-different parallel architectures.  Using this methodology, the behavior of the
-original PSO algorithm can either be exactly reproduced, two iterations at a
-time, or the behavior can be modified in order to improve performance.  While
-exactly reproducing PSO sometimes uses too many extra processors to be useful,
-when we allow ourselves some freedom with the algorithm we see great
-improvements over previous methods.  We have shown results that conclusively
-demonstrate the superiority of our techniques for several functions over the
-standard practice of adding particles to the swarm when extra processors are
-available.
-
-What we have presented is not a new variant of PSO.  We presented a new
-parallelization technique, so we compared parallelization strategies on the
-same algorithm, the original PSO.  Our most promising techniques do change the
-behavior of the PSO algorithm slightly, as do some other previously proposed
-parallelization techniques, such as asynchronous PSO.  However, our methods are
-applicable to almost all PSO variants, and so a comparison using the same
-variant for each of the parallelization techniques tested is justified.
-
-We have given five different possible approaches to speculative evaluation,
-each of which has different properties.  These approaches perform differently
-on different functions and at different swarm sizes, as would be expected by
-the No Free Lunch Theorem for
-Optimization~\citep{wolpert-1997-nfl-for-optimization}.  We have given a brief
-evaluation of the premature convergence properties of these methods on
-deceptive functions when a smaller number of processors is available.  We have
-also shown evidence that when many processors are available by far the best
-thing to do in most cases is to speculate as far ahead as the extra processors
-allow.
-
-Though our methods show great improvements on some functions, they do not work
-for all functions.  As is commonly known, in PSO there is a trade-off between
-exploration and exploitation.  Some functions need only minimal exploration,
-and some never seem to have enough.  Increasing the swarm size is a natural way
-to increase exploration in a parallel environment.  However, once ``enough''
-exploration has been reached for any particular function, adding additional
-particles adds only incremental benefits.  At this point, a better use of the
-additional processors, as we have shown, is to perform some amount of
-speculative evaluation.
-
-Large parallel clusters are often required to successfully optimize practical
-modern problems.  To properly use PSO with such clusters, a balance needs to be
-made between using processors to increase the swarm size and using them to
-increase the speed of the algorithm.  This work is a first step in that
-direction that opens the door to many future improvements on speculative
-methods in the parallelization of PSO.
-
-\section{Future Work}
-\label{sec:future}
-
-In this work we have focused on PSO itself and not all of its variants.  It
-remains as future work to apply speculative approaches to recent and popular
-PSO variants, such as the Fully Informed Particle
-Swarm~\citep{mendes-2004-fully-informed-particle-swarm}.  While our methods
-will not always be immediately applicable to every variant, we are confident
-that some kind of speculative approach will be beneficial to the
-parallelization of all forms of PSO, especially as the number of processors
-used gets into the thousands.
-
-We mentioned related work showing that increasing the swarm size throughout the
-course of the algorithm could provide improved performance over a fixed swarm
-size in serial PSO~\citep{montes-de-oca-2010-incremental-social-learning-pso}.
-If this method were extended to parallel PSO, most processors would be idle in
-the first few iterations, while more would be utilized at the end.  During
-iterations where there are many un-utilized processors, a natural use of them
-would be speculative evaluation, performing two or more of those iterations at
-a time.
-
-The issues of the sampling distribution of speculative relaxations, branch
-statistics, and why standard PSO fails at 500 dimensions were briefly mentioned
-in this paper.  Each of those issues needs further treatment.  The sampling
-distribution of our speculative methods could be compared to Poli's description
-of standard PSO's sampling
-distribution~\citep{poli-2008-sampling-distribution-of-pso}.  The branch
-statistics could be used to analyze topologies and discover why certain
-topologies work well on some functions but not on others; perhaps PSO
-performance is more dependent on the branch statistics of a combination of
-topology and function than on the topology itself.  And discovering exactly why
-constricted PSO fails at 500 dimensions while Many Iterations does not could
-lead to improvements in the standard PSO algorithm, even when not running in
-parallel.
-
-We opened the door to speculative parallelization methods in PSO and described
-the possible speculative evaluations to perform as an infinite tree from which
-branches are selected.  However, we only presented a few of the countless
-possibilities for selecting those branches.  Our methods for determining which
-speculative evaluations to perform were independent of the particle; all
-particles performed the same number and type of evaluations.  Another way to
-allocate speculative evaluations is to somehow use the performance of each
-particle to determine how many and which extra evaluations it can have.
-
-Any other optimization algorithm that only depends on current sampling
-positions when computing the next position to sample can be parallelized with
-this technique.  In particular, genetic algorithms produce future generations
-by combining individuals from the current generation.  With a large population
-size there would be an unwieldy amount of possible future individuals, but the
-potential exists to modify the algorithm to use some kind of speculative
-evaluation.
-
-\section*{Appendix---Alternate form of message passing}
+\section*{Appendix B---Alternate form of message passing}
 

### 2011-04-01 masters_thesis/si_paper/paper.tex
A few minor changes.  Basically done.

@@
 mathematically how speculative evaluation can be done in parallel PSO to
-perform two iterations at once, then \sec{implementation} gives information on
-how to implement this method.  In \sec{relax}, we discuss various methods of
-improving the performance of speculative evaluation in PSO, all of which break
-the requirement of strictly reproducing the behavior of the original algorithm.
+perform two iterations at once, leaving implementation concerns to the
+appendices.  In \sec{relax}, we discuss various methods of improving the
+performance of speculative evaluation in PSO, all of which break the
+requirement of strictly reproducing the behavior of the original algorithm.
 \sec{setup} describes the experiments we ran, and \sec{results} presents the
@@
 
-Becuase there are a few branches with very high probabilities in the topologies
+Because there are a few branches with very high probabilities in the topologies
 we are interested in, we can have hope that cutting out some of the evaluations
@@
 two iterations of PSO at each ``iteration,'' and Social Promotion and Many
-Iterations make the idea of ``iterations of PSO'' somewhat nebulus, we instead
+Iterations make the idea of ``iterations of PSO'' somewhat nebulous, we instead
 call each ``iteration'' a ``time step.''
@@
 is close to three times as fast and Many Iterations is even faster when it is
-successful, though it is only successful 84\% of the time.  \tabref{griewank}
-shows all of our results in tabular form.
+successful, though it is only successful 84\% of the time.
+\tabref{griewank-20} shows all of our results in tabular form.
 
@@
 \end{figure}
+
+Note here that \fig{griewank-manyiters} is interesting to compare to
+\fig{basic-griewank1}.  The ordering of the methods in terms of the number of
+time steps to completion is the same in both figures.  What is different is
+that at 240 processors, most of our methods fail to find the optimum 100\% of
+the time, while at 800 processors, all but Many Iterations succeed 100\% of the
+time, and Many Iterations is very close.  At some point between 240 and 800
+processors, SEPSO and Pick Best become successful, and at that point it is by
+far better to use speculative methods than standard parallelizations.  We can
+also assume that at some point above 800 processors, Many Iterations will be
+become successful and will be the preferred method.  Once there are enough
+particles in the swarm to guarantee success when optimizing a function, the
+best use of additional processors is to speculate as far ahead as possible,
+speeding up the progress of the algorithm.
 
@@
 
-\subsection{Terminology}
+\subsection*{Terminology}
 
@@
 
-\subsection{Centralized Algorithms}
+\subsection*{Centralized Algorithms}
 
@@
 
-\subsection{Distributed Algorithms}
+\subsection*{Distributed Algorithms}
 
@@
 
-\subsection{Dynamic Topologies}
+\subsection*{Dynamic Topologies}
 

### 2011-04-04 masters_thesis/proposal/proposal.tex
Finalized proposal based on feedback from the committee

@@
 Particle swarm optimization (PSO) is an evolutionary algorithm that has proven
-effective at solving complicated optimization problems.  Since its inception,
-however, PSO has been run primarily in serial on a single machine.  With the
-world's computing resources increasingly coming from large clusters of
-processors, the parallelization of the algorithm is important to its continued
-relevance.  PSO has previously been parallelized primarily by distributing the
-computation corresponding to particles across multiple processors.  In these
-approaches, additional processors can only add more particles to the swarm.
-However, in many cases this is not efficient when scaled to very large swarm
-sizes (on very large clusters).  Current methods cannot answer well the
-question, ``What do you do with 1000 processors when 50 or 100 particles is the
-most efficient swarm size?''  In this project we attempt to answer that
-question with a speculative approach to the parallelization of PSO motivated by
+effective at solving complicated optimization problems.  In this project we
+present a speculative approach to the parallelization of PSO motivated by
 speculative execution in modern computer hardware.
@@
 When many processors are available, a speculative approach to the
-parallelization of PSO can yield dramatic improvements in the performance of
-the algorithm on expensive functions compared to previous parallelization
-techniques.
+parallelization of particle swarm optimization can yield dramatic improvements
+in the performance of the algorithm on expensive functions compared to previous
+parallelization techniques.
 
@@
 
-\item March 11, 2011: Finish running experiments and incorporate the results
-  into the thesis.
+\item May 13, 2011: Finish running experiments and start incorporating the
+  results into the thesis.
 
-\item March 15, 2011: Give final draft of thesis to advisor for review.
+\item May 30, 2011: Give final draft of thesis to advisor for review.
 
-\item March 22, 2011: submit thesis to second committee member for review.
+\item June 10, 2011: submit thesis to second committee member for review.
 
-\item April 5, 2011: submit thesis to third committee member for review.
+\item June 30, 2011: submit thesis to third committee member for review.
 
-\item April 19, 2011: defend thesis.
+\item July 15, 2011: defend thesis.
 

### 2011-04-18 masters_thesis/si_paper/paper.tex
Updated the journal paper version of my thesis

@@
 standard parallelization.  At each iteration of the algorithms, we use one
-processor to perform one function evaluation for one particle, be it
-speculative or non-speculative.  The exact evaluation time at which this
-architecture becomes reasonable depends on the amount of communication overhead
-in the parallel implementation and the number of particles in the swarm.  For
-our implementation we found that time to be around 100 milliseconds for swarms
-of 50 particles.  When the swarm size increases, the minimum evaluation time at
-which this architecture should be used decreases.
+processor to perform one function evaluation.  The exact evaluation time at
+which this architecture becomes reasonable depends on the amount of
+communication overhead in the parallel implementation and the number of
+particles in the swarm.  For our implementation we found that time to be around
+100 milliseconds for swarms of 50 particles.  When the swarm size increases,
+the minimum evaluation time at which this architecture should be used
+decreases.
 
@@
 implementations, we do not report running times, favoring the abstraction of
-time steps.  That being said, given the time required for the evaluation of the
+time steps.  However, given the time required for the evaluation of the
 objective function and the communication overhead per iteration for a specific
-implementation, an actual running time can be recovered from all of our
-results.  Simply multiply the number of time steps by the sum of the function
-evaluation time and the overhead.  We wish to stress that the ``time steps'' we
-report are proportional to wall-clock time, given the architecture we have
-assumed.
+implementation, a running time can be estimated from all of our results.
+Simply multiply the number of time steps by the sum of the function evaluation
+time and the overhead.  We wish to stress that the ``time steps'' we report are
+proportional to wall-clock time, given the architecture we have assumed.
 
@@
 
-One might question the use of benchmark functions for evaluating our methods,
-as they take on the order of 1 millisecond to evaluate instead of the 100s that
-we said were required to justify our parallel architecture.  The purpose of
+We find benchmark functions useful in comparing algorithms, even though they
+only take on the order of 1 millisecond to evaluate instead of the 100s that we
+said were required to justify our parallel architecture.  The purpose of
 exploring the behavior of an algorithm on benchmark functions, whether in
@@
 
-To aid in convincing the reader that performance on benchmark functions roughly
+To provide additional evidence that performance on benchmark functions roughly
 corresponds to performance on real-world problems, we also tested our
@@
 \subsubsection{Sphere}
+\label{sec:sphere20dims}
 
@@
 processors, SEPSO and Pick Best become successful, and at that point it is by
-far better to use speculative methods than standard parallelizations.  We can
-also assume that at some point above 800 processors, Many Iterations will be
-become successful and will be the preferred method.  Once there are enough
-particles in the swarm to guarantee success when optimizing a function, the
-best use of additional processors is to speculate as far ahead as possible,
-speeding up the progress of the algorithm.
+far better to use speculative methods than standard parallelizations.  At some
+number above 800 processors, Many Iterations will be become successful and will
+be the preferred method.
+
+To further show this point, we present results using 1920 processors on the
+20-dimensional Griewank function in \fig{griewank-all}.  With 1920 processors
+Many Iterations finds the optimum 100\% of the time.  What is interesting is
+that while going from 240 processors to 1920 processors, the standard method of
+parallelization only decreased its time to completion by 5.9\%.  That is, the
+extra 1680 processors in standard parallelizations provide \emph{no appreciable
+benefit}.  In contrast, Many Iterations with 1920 processors has the same swarm
+size of 240 particles and uses the additional processors for speculative
+evaluation, giving a decrease in time to completion of 78.9\%, or a factor of
+almost 5.  Once there are enough particles in the swarm to guarantee success
+when optimizing a function, the best use of additional processors is to
+speculate as far ahead as possible, speeding up the progress of the algorithm.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{griewankall}
+  \caption{Function Griewank with 20 dimensions, comparing the performance of
+  standard and speculative parallelizations as the number of processors
+  increases.  All algorithms use the Ring topology.}
+  \label{fig:griewank-all}
+\end{figure}
 
@@
   \label{fig:rbf-stuck}
+\end{figure}
+
+\subsection{Bare Bones PSO}
+\label{sec:bare-results}
+
+In order to demonstrate that our methods are applicable across PSO variants, we
+ran experiments with Bare Bones PSO.  For this simple experiment we use the
+Sphere function with 20 dimensions, taking the best Standard algorithm and the
+best speculative algorithm on that function, which in this case is Standard
+Complete and Pick Best Pruned Complete, as shown in \sec{sphere20dims}.
+\fig{bare-bones} shows the results of running each of those algorithms with the
+Bare Bones PSO equations.  We see a decrease in time to completion between
+Standard and speculative parallelizations of 40\%.  While this is not quite as
+large a speed up as the 60\% seen with constricted PSO, it is still a
+significant decrease in time.  The key point of this experiment is that our
+methods still show improvements over Standard parallelizations in PSO variants
+other than the standard constricted PSO.  Our intent is not to give an
+exhaustive listing of PSO variants and compare parallelization strategies with
+each one; instead we offer evidence supporting our claim that speculative
+evaluation should see similar increases in performance across most or all PSO
+variants.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\columnwidth]{bare_bones}
+  \caption{Results for optimizing the 20-dimensional Sphere function with Bare
+  Bones PSO.  We use 240 processors for each method, so each time step
+  corresponds to 240 function evaluations.}
+  \label{fig:bare-bones}
 \end{figure}

### 2011-04-18 masters_thesis/thesis/paper.tex
Added a thesis version of my thesis

--- 2011-04-18 masters_thesis/si_paper/paper.tex
@@
-\documentclass[smallcondensed]{svjour3}
-
+\documentclass[ms,electronic,twosidetoc,letterpaper,chaptercenter,parttop,equalmargins]{byumsphd}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Thesis formating stuff
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\Title}{A Speculative Approach to Parallelization\\
+in Particle Swarm Optimization}
+\newcommand{\Author}{Matthew Gardner}
+\newcommand{\GraduationMonth}{August}
+\newcommand{\GraduationYear}{2011}
+
+\usepackage[
+    bookmarks=true,
+    bookmarksnumbered=true,
+    breaklinks=false,
+    raiselinks=true,
+    pdfborder={0 0 0},
+    colorlinks=false,
+    plainpages=false,
+    ]{hyperref}
+\usepackage[all]{hypcap}
+\usepackage{natbib}
+\hypersetup{%
+    pdftitle=\Title,%
+    pdfauthor=\Author,%
+    pdfsubject={PhD Dissertation, BYU CS Department: %
+                Degree Granted \GraduationMonth~\GraduationYear, Document Created \today},%
+    pdfkeywords={BYU, thesis, dissertation, LaTeX},%
+}
+\newcommand{\ItemSep}{\itemsep 0pt}
+\let\oldenum=\enumerate
+\renewcommand{\enumerate}{\oldenum \ItemSep}
+\let\olditem=\itemize
+\renewcommand{\itemize}{\olditem \ItemSep}
+\let\olddesc=\description
+\renewcommand{\description}{\olddesc \ItemSep}
+
+% Important settings for the byumsphd class.
+\title{\Title}
+\author{\Author}
+
+\committeechair{Kevin~Seppi}
+\committeemembera{Dan~Ventura}
+\committeememberb{David~Embley}
+
+\monthgraduated{\GraduationMonth}
+\yeargraduated{\GraduationYear}
+\yearcopyrighted{\GraduationYear}
+
+\department{Computer~Science}
+\graduatecoordinator{Kent~E.~Seamons}
+%\collegedean{Thomas~W.~Sederberg}
+%\collegedeantitle{Associate~Dean}
+
+% Remove all widows an orphans.  This is not normally recommended, but in a
+% paper dissertation there is no reasonable way around it; you can't exactly
+% rewrite already-published content to fix the problem.
+\clubpenalty 10000
+\widowpenalty 10000
+
+% Allow pages to have extra blank space at the bottom in order to accommodate
+% removal of widows and orphans.
+\raggedbottom
+
+% Produce nicely formatted paragraphs. There is nothing additional to do.  In
+% case you get some problems, surround your text with
+% \begin{sloppy} ... \end{sloppy}. If that does not work, try
+% \microtypesetup{protrusion=false} ... \microtypesetup{protrusion=true}
+\usepackage{microtype}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% For the content of the document
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \usepackage{graphicx}
@@
 
-\title{A Speculative Approach to Parallelization in Particle Swarm
-Optimization}
-
-\date{Received: date / Accepted: date}
-
-\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
-
-\institute{M. Gardner \at
-Brigham Young University.
-3361 TMCB, Provo, UT 84604
-Tel.: 801-422-8717
-Fax: 801-422-0169
-\email{mjg82@byu.edu}
-\and
-A. McNabb \at
-Brigham Young University.
-3361 TMCB, Provo, UT 84604
-\email{a@cs.byu.edu}
-\and
-K. Seppi \at
-Brigham Young University.
-3361 TMCB, Provo, UT 84604
-\email{k@cs.byu.edu}
-}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
+\documentabstract{%
 
@@
 distributing the computation corresponding to particles across multiple
-processors.  In these approaches, adding processors can only increase the swarm
-size.  However, in many cases this is not efficient when scaled to very large
-swarm sizes (on very large clusters).  Current methods cannot answer well the
-question, ``How can 1000 processors be fully utilized when 50 or 100 particles
-is the most efficient swarm size?''  In this paper we attempt to answer that
-question with a speculative approach to the parallelization of PSO that we
-refer to as SEPSO.
+processors.  In this thesis we present a speculative approach to the
+parallelization of PSO that we refer to as SEPSO.
 
@@
 methods dramatically improve the performance of parallel PSO in many cases.
-
-\keywords{Parallel algorithms \and Optimization methods \and Particle swarm
-optimization \and Speculative Decomposition}
-
-\end{abstract}
-
-\section{Introduction}
+}
+\documentkeywords{%
+Parallel algorithms, optimization methods, particle swarm optimization,
+peculative decomposition
+}
+
+\begin{document}
+
+% Produce the preamble
+\microtypesetup{protrusion=false}
+\maketitle
+\microtypesetup{protrusion=true}
+
+\chapter{Introduction}
 \label{sec:intro}
@@
 
-\section{Speculative Evaluation in PSO}
+\chapter{Speculative Evaluation in PSO}
 \label{sec:sepso}
@@
 
-\subsection{Refactoring the PSO Equations}
+\section{Refactoring the PSO Equations}
 \label{sec:proof}
@@
 
-\subsection{Topology in Speculative Evaluation}
+\section{Topology in Speculative Evaluation}
 \label{sec:topology}
@@
 
-\section{Relaxing the Requirements}
+\chapter{Relaxing the Requirements}
 \label{sec:relax}
@@
 
-\subsection{Pick the Best Child}
+\section{Pick the Best Child}
 \label{sec:pickbest}
@@
 
-\subsection{Pruning the Speculative Evaluations}
+\section{Pruning the Speculative Evaluations}
 \label{sec:pruning}
@@
 
-\subsection{Branch Statistics}
+\section{Branch Statistics}
 \label{sec:statistics}
@@
 
-\subsection{Recovering from Pruning Too Much}
+\section{Recovering from Pruning Too Much}
 \label{sec:wrong}
@@
 
-\subsection{More than one iteration ahead}
+\section{More than one iteration ahead}
 \label{sec:manyiters}
@@
 
-\section{Experimental Setup}
+\chapter{Experimental Setup}
 \label{sec:setup}
@@
 
-\subsection{Function Evaluations vs. Time Steps}
+\section{Function Evaluations vs. Time Steps}
 
@@
 
-\subsection{Functions Used}
+\section{Functions Used}
 
@@
 
-\subsection{Parallelization Techniques Compared}
+\section{Parallelization Techniques Compared}
 
@@
 
-\subsection{Topologies}
+\section{Topologies}
 
@@
 
-\section{Results}
+\chapter{Results}
 \label{sec:results}
@@
 
-\subsection{20 Dimensions}
+\section{20 Dimensions}
 \label{sec:20dims}
 
-\subsubsection{Sphere}
-\label{sec:sphere20dims}
+\subsection{Sphere}
 
@@
 
-\subsubsection{Schwefel 2.21}
+\subsection{Schwefel 2.21}
 
@@
 
-\subsubsection{Rastrigin}
+\subsection{Rastrigin}
 
@@
 
-\subsubsection{Griewank}
+\subsection{Griewank}
 \label{sec:griewank-20}
@@
 
-\subsubsection{Bohachevsky}
+\subsection{Bohachevsky}
 
@@
 
-\subsection{50 Dimensions}
+\section{50 Dimensions}
 \label{sec:50dims}
@@
 
-\subsection{500 Dimensions}
+\section{500 Dimensions}
 \label{sec:500dims}
@@
 
-\subsection{Model Fitting}
+\section{Model Fitting}
 \label{sec:model}
@@
 
-\subsection{Bare Bones PSO}
-\label{sec:bare-results}
-
-In order to demonstrate that our methods are applicable across PSO variants, we
-ran experiments with Bare Bones PSO.  For this simple experiment we use the
-Sphere function with 20 dimensions, taking the best Standard algorithm and the
-best speculative algorithm on that function, which in this case is Standard
-Complete and Pick Best Pruned Complete, as shown in \sec{sphere20dims}.
-\fig{bare-bones} shows the results of running each of those algorithms with the
-Bare Bones PSO equations.  We see a decrease in time to completion between
-Standard and speculative parallelizations of 40\%.  While this is not quite as
-large a speed up as the 60\% seen with constricted PSO, it is still a
-significant decrease in time.  The key point of this experiment is that our
-methods still show improvements over Standard parallelizations in PSO variants
-other than the standard constricted PSO.  Our intent is not to give an
-exhaustive listing of PSO variants and compare parallelization strategies with
-each one; instead we offer evidence supporting our claim that speculative
-evaluation should see similar increases in performance across most or all PSO
-variants.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\columnwidth]{bare_bones}
-  \caption{Results for optimizing the 20-dimensional Sphere function with Bare
-  Bones PSO.  We use 240 processors for each method, so each time step
-  corresponds to 240 function evaluations.}
-  \label{fig:bare-bones}
-\end{figure}
-
-\section{Conclusions}
+\chapter{Conclusions}
 \label{sec:conclusion}
@@
 
-\section{Future Work}
+\chapter{Future Work}
 \label{sec:future}
@@
 
-\section*{Appendix A---Implementing Speculative Evaluation}
+\appendix
+\chapter{Implementing Speculative Evaluation}
 \label{sec:implementation}
@@
 
-\subsection*{Terminology}
+\section{Terminology}
 
@@
 
-\subsection*{Centralized Algorithms}
+\section{Centralized Algorithms}
 
@@
 
-\subsection*{Distributed Algorithms}
+\section{Distributed Algorithms}
 
@@
 
-\subsection*{Dynamic Topologies}
+\section{Dynamic Topologies}
 
@@
 
-\section*{Appendix B---Alternate form of message passing}
+\chapter{Alternate form of message passing}
 
@@
 
-\bibliographystyle{apa-good}
+\bibliographystyle{plainnat}
 \bibliography{%

### 2011-04-26 masters_thesis/thesis/paper.tex
Made changes from my committee, finalizing the thesis document

@@
 \committeemembera{Dan~Ventura}
-\committeememberb{David~Embley}
+\committeememberb{David~W.~Embley}
 
@@
 speculating several iterations ahead instead of just one.  We show that these
-methods dramatically improve the performance of parallel PSO in many cases.
+methods dramatically improve the performance of parallel PSO in many cases,
+giving speed ups of up to six times compared to previous parallelization
+techniques.
 }
@@
 Parallel algorithms, optimization methods, particle swarm optimization,
-peculative decomposition
+speculative decomposition
 }
@@
 
-\section{Branch Statistics}
+\subsection{Branch Statistics}
 \label{sec:statistics}
@@
 
-\section{Recovering from Pruning Too Much}
+\subsection{Recovering from Pruning Too Much}
 \label{sec:wrong}
@@
 one that we can only begin to explore here.  If the branch corresponding to
-stagnation has a 75\% chance of being taken, as in the Random topology with
-most functions, we could speculate three iterations ahead on that branch and
-still have a 42\% chance of predicting correctly.  However, intuition would say
-that perhaps it is better to hope that the particle is productive instead of
-stagnant, so a branch where the particle updates its personal best might be
-more fruitful to try.  In our experiments we try just one of the countless
-possibilities, but one that turns out to work very well.  More work is needed
-to compare the different branching possibilities on various functions.
+stagnation, \casepn, has a 75\% chance of being taken, as in the Random
+topology with most functions, we could speculate three iterations ahead on that
+branch and still have a 42\% chance of predicting correctly.  However,
+intuition would say that perhaps it is better to hope that the particle is
+productive instead of stagnant, so a branch where the particle updates its
+personal best might be more fruitful to try.  In our experiments we try just
+one of the countless possibilities, but one that turns out to work very well.
+More work is needed to compare the different branching possibilities on various
+functions.
 
@@
 With this particular implementation, then, Many Iterations takes on average 21
-minutes to reach a successful value, while Standard takes 50 minutes.
+minutes to reach a successful value, while Standard Subswarms takes 50 minutes.
 
@@
 standard practice of adding particles to the swarm when extra processors are
-available.
+available, giving speed ups of up to six compared to previous parallelization
+techniques.
 

### 2011-08-26 masters_thesis/si_paper/reviewer_response/response.tex
Final improvements from reviewer feedback; revision submitted to SI

@@
 \begin{document}
+
+We apologize for the long delay in responding to your feedback.  This was due
+to our extensive modifications and improvements to the paper.  We have run over
+three processor years worth of additional experiments, greatly expanding the
+results section, and we restructured the middle section of the paper, hopefully
+improving readability.  Specific comments from the reviewers, with our response
+to each comment, are found below.
 
@@
 Response: We added a Related Work section (now section 3), describing previous
-approaches at parallelizing PSO.  As we note in the paper, most of the related
-work is orthogonal to innovations we describe, so it is difficult to determine
-what other methods to compare against.  The majority of previous papers on the
-parallelization of PSO deal with functions with very short evaluation times, so
-they propose topologies that limit communication between processors on
-different machines.  Those methods work very well for those functions, but when
-you are dealing with functions whose evaluation takes over a second, the
-benefits of the limited communication become negligible.  We did, however,
-experiment with one of the topologies proposed for fast running functions.  We
-found that in almost all cases simply using a Complete topology gave better
-performance.  We make note of all of this in the paper, mostly in section 3.
+approaches at parallelizing PSO.  We also compared to all applicable methods
+that we described.
+
+Prior algorithmic research is divided into two camps.  Either this research
+focuses on implementing PSO in specific parallel architectures, or they adapt
+the algorithm to be better suited to parallel computation.  Work on adapting
+the algorithm to parallel computation (where this paper fits) has focused on
+functions with very short evaluation times.  We focus on long-running
+functions, such as are found in prior application papers, but requires very
+different methods than those for functions with short evaluation times.  Thus
+most prior algorithmic advances are not very successful in our domain, as we
+show in the paper.  The sections to refer to on this subject are section 3,
+section 6.5, and section 7.4.
 
@@
 
-Response: We added a section dealing with this issue (section 7.1).  We also
-added experiments on a function that takes on the order of 2 seconds to
-evaluate, providing a little more motivation for the parallelization of PSO.
+Response: We added experiments on a real-world function that takes longer to
+evaluate (fitting a model to a large amount of data), providing more motivation
+for the parallelization of PSO (those results are shown in section 7.4).  We
+also added some discussion of these issues in the Experimental Setup section.
+Section 6.1 discusses when the parallel architecture that we propose is useful
+in contrast to other architectures, and section 6.3 discusses our use of
+benchmark functions to evaluate our methods.
 
@@
 
-Response: Good point.  We realized that we tried to cram too much into a single
-section, so we split the two main parts of it into two separate sections.
-There is now a section devoted entirely to the mathematics of the method (now
-Section 4) and a separate section describing its implementation (Section 5).
-We also put subsections in the new Section 4 to guide the reader a little
-better through the section, and gave an intuitive description of speculative
-evaluation before jumping into the math.
+Response: Good point.  We also realized that we tried to cram too much into a
+single section, so we split the two main parts of it into two separate
+sections.  There is now a section devoted entirely to the mathematics of the
+method (Section 4) and a separate section describing its implementation
+(which we moved to the appendix, as we thought it was important, but
+distracting from the main discussion).  We also put subsections in the new
+Section 4 to guide the reader a little better through the section, and gave an
+intuitive description of speculative evaluation before jumping into the math.
 
@@
 paragraphs to the end of section 2 to further make this point, making more
-prominent reference to the No Free Lunch theorems.
+prominent reference to the No Free Lunch theorems.  We also added experiments
+on a real-world problem, which is found in section 7.4
 
@@
 
-Response: We substantially reorganized Section 3, as noted above, and we added
-more subsections throughout the paper, helping to guide the reader through the
-ideas.  We believe that the paper is now much more clear.
+Response: We substantially reorganized Section 3, as noted above under AE.4,
+and we added more subsections throughout the paper, helping to guide the reader
+through the ideas.  We removed most of the forward pointers in section 4,
+focusing simply on describing the methods and leaving the comparison to the
+results section, as you suggested.  We believe that the paper is now much more
+clear.
 
@@
 
-Response: We had a long discussion about notation before writing our paper, and
-we thought that because the personal best and neighborhood bests are positions,
-using an x to denote it would be more intuitive.  However, we see your point
-that it can be confusing for other reasons.  We changed the x to a b when
-referring to best positions, while still using the N and P superscripts to
-differentiate between which best is being referred to.
+Response: We changed the x to a b when referring to best positions, while still
+using the N and P superscripts to differentiate between which best is being
+referred to.  We hope the change in notation is helpful.
 
@@
 
-Response: Rounds of evaluations correspond exactly to wall clock time, as we
-describe in what is now section 7.1, and the second paragraph of section 7
-proper.  However, we realize that the term is not a commonly used one, so we
-changed ``rounds of evaluations'' to ``time steps'' in hopes of better
-conveying our meaning.  We also added some text in the description mentioned
-above that hopefully further clarifies this point.
+Response: We realize that ``rounds of evaluations'' is not a commonly used
+term, so we changed ``rounds of evaluations'' to ``time steps'' in hopes of
+better conveying our meaning.  We also added a section in the experimental
+setup that discusses the issue of wall clock time (section 6.2), saying clearly
+that time steps (or rounds of evaluations, as we used to call them), are
+directly proportional to wall clock time.
 
@@
 Response: We ran quite a few more experiments to bring our results closer to
-Herrera et al. experimental protocol.  Their protocol wasn't entirely
+the experimental protocol of Herrera et al.  Their protocol wasn't entirely
 applicable, as they suggested running a specific number of function evaluations
-instead of iterations (which, as we state in section 7, is the proper
+instead of iterations (which, as we state in section 6.2, is the proper
 comparison for parallel algorithms on long running functions).  But we added
@@
 pick a particular variant and compare parallelization techniques with it.  It
-made sense to pick the original PSO for this comparison.
+made sense to pick the standard PSO for this comparison.  To further show this
+point, we also ran a simple comparison of our parallelization techniques using
+a different PSO variant, Bare Bones PSO.  Those results are in section 7.5.
 
@@
 analogy to processors lacked an explicit connection to the word
-``speculative.''  We added a sentence at the end of that paragraph to address
-this issue.
+``speculative.''  We added text at the end of that paragraph to address this
+issue.
 
@@
 
-Response: We added the citation after the first sentence of the paper.
+Response: We added the citation as suggested.
 
@@
 
-Response: We added a paragraph in our related work section discussing these
+Response: We added a paragraph in our new related work section discussing these
 methods (the last paragraph of the section).  We also added a paragraph to
@@
 speculative evaluation when parallelized (the second paragraph of the section).
+
+There is also now a graph in section 7.1.4 that clearly shows what we mean.  It
+is Figure 10, showing what happens when processors are added with standard
+parallelizations versus with our methods.
 
@@
 Response: This is very much related to the previous item, and so the response
-to that item is relevant here.  We also added some text to to the introduction
+to that item is relevant here.  We also added some text to the introduction
 (paragraph beginning with ``In this paper we consider\ldots'') to better
@@
 some parts, making the examples that were there more clear.  Hopefully the
-section is satisfactory now.
+section is now satisfactory.
 
@@
 processors.  We realized that this was never explicitly stated in the text, so
-we added a few sentences to the description of pruning (now section 6.2) to
+we added a few sentences to the description of pruning (now section 5.2) to
 make it clear.  The text added to the introduction explaining our purpose of
@@
 
-Response: ...
-
-NOT DONE
+Response: We had Gnuplot render the figures in gray mode, using line styles
+that are easier to distinguish in black and white.  Hopefully that solves the
+problem.  That does lose color in the figures, though.  If color is desirable,
+we can find line styles and colors that are both acceptable.  We also put most
+of our results into tables, so the reader does not have to rely as much on
+interpreting figures.
 
@@
 conclusions.  However, our own figures also answer your question.  The fact
-that our methods outperform the na\"ive parallelization is in itself evidence
+that our methods outperform the standard parallelization is in itself evidence
 for this claim.  Recall that when our figures say SEPSO, that is equivalent to
 running PSO with a swarm size one eighth the size, doing twice as many
-iterations.  Figures 9 and 10 illustrate the point that for some swarm sizes
-you are correct and it is better to have a larger swarm size.  However, once
-enough processors are available to have a large enough swarm even with SEPSO,
-it is better by far to do speculative evaluation.
+iterations.  The direct answer to your question is in what is now Figure 10
+(which we added to the paper).  We show that adding particles beyond 240 for a
+20 dimensional Griewank function provides no appreciable benefit, while using
+speculative methods gives dramatic improvements.
 
@@
 not proposing a new algorithm.  That part had two sentences, and we deleted the
-second one to soften the point a little.  It was an over-reaction to comments
-by previous reviewers.  We also fixed a similar statement in the conclusion.
+second one to soften the point a little.  We also fixed a similar statement in
+the conclusion.
 

### 2011-08-26 masters_thesis/si_paper/paper.tex
Final improvements from reviewer feedback; revision submitted to SI

@@
 speculating several iterations ahead instead of just one.  We show that these
-methods dramatically improve the performance of parallel PSO in many cases.
+methods dramatically improve the performance of parallel PSO in many cases,
+giving speed ups of up to six compared to previous parallelization techniques.
 
@@
 algorithms \citep{mostaghim-2006-multi-objective-pso-on-grids}.  Parallelizing
-the evaluation of the objective function can also be done in some cases, though
-that is not an adaption of the PSO algorithm itself and thus is not the focus
-of this paper.
+the evaluation of the objective function can also be done in some cases using
+standard techniques \citep{grama-2003-intro-to-parallel-computing}, though that
+is not an adaption of the PSO algorithm itself and thus is not the focus of
+this paper.
 
@@
 
-Asynchronous parallel
-PSO~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
+Asynchronous parallel particle swarm
+optimization~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
 koh-2006-parallel-asynchronous-pso} is a modification to the standard algorithm
@@
 
-\subsection{Branch Statistics}
+\subsubsection{Branch Statistics}
 \label{sec:statistics}
@@
 
-\subsection{Recovering from Pruning Too Much}
+\subsubsection{Recovering from Pruning Too Much}
 \label{sec:wrong}
@@
 This ignores the fact that the branch might have been wrong; it does not
-matter, because we simply accept the child that had the best value.  In most of
-the experiments that we ran, it turned out that picking the best child
-performed better than keeping the particle back an iteration.
+matter, because we simply accept the child that had the best value.
 
@@
 
-We ran experiments to compare our speculative parallelization of PSO to the
-standard parallelization.  At each iteration of the algorithms, we use one
-processor to perform one function evaluation.  The exact evaluation time at
-which this architecture becomes reasonable depends on the amount of
-communication overhead in the parallel implementation and the number of
-particles in the swarm.  For our implementation we found that time to be around
-100 milliseconds for swarms of 50 particles.  When the swarm size increases,
-the minimum evaluation time at which this architecture should be used
-decreases.
+\subsection{Parallel Architecture}
+
+At each iteration of the algorithms, we use one processor to perform one
+function evaluation.  The exact evaluation time at which this architecture
+becomes reasonable depends on the amount of communication overhead in the
+parallel implementation and the number of particles in the swarm.  For our
+implementation we found that time to be around 100 milliseconds for swarms of
+50 particles.  When the swarm size increases, the minimum evaluation time at
+which this architecture should be used decreases.
+
+A general rule of thumb for when this architecture is useful is when the
+combined function evaluation time is greater than the communication overhead
+for a single iteration; i.e., $f*p > o+f$, where $f$ is the function evaluation
+time, $p$ is the desired number of particles (in our case $p$ is equal to the
+number of available processors), and $o$ is the amount of overhead per
+iteration.  The amount of overhead per iteration will vary based on the
+communication topology and the number of particles, so this is only an
+approximation, but it can provide rough estimates of function evaluation times
+at which this parallel architecture should be used.
 
@@
 
-To provide additional evidence that performance on benchmark functions roughly
+To provide additional evidence that performance on benchmark functions
 corresponds to performance on real-world problems, we also tested our
@@
 approach developed in \sec{pickbest} Pick Best.  The methods described in
-\sec{pruning} through \sec{wrong} are called Pick Best Pruned and Social
-Promotion Pruned, and the method in \sec{manyiters} we call Many Iterations.
+\sec{pruning} are called Pick Best Pruned and Social Promotion Pruned, and the
+method in \sec{manyiters} we call Many Iterations.
 
@@
 particle is independent of the number of neighbors the particle has.
+
+Finally, to show that our methods are orthogonal to traditional PSO variants,
+we show results of parallelizing a variation of the PSO motion equations, Bare
+Bones PSO~\citep{kennedy-2003-bare-bones-pso}, with both standard
+parallelizations and our speculative approaches.  Our intent is not to
+exhaustively compare parallelization strategies on all possible PSO variants,
+but to show some evidence that speculative approaches are generally applicable
+and successful across PSO variants.  These results are shown in
+\sec{bare-results}.
 
@@
 
-It is generally recommended to use the Ring topology when optimizing the
-Griewank function, as Complete is prone to premature convergence on a local
+It is recommended in the literature to use the Ring topology when optimizing
+the Griewank function, as Complete is prone to premature convergence on a local
 optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  The global optimum
@@
 significant progress.  Our search for an explanation seems to say that the
-space is too large for particles to converge to the same point, so all updates
-to the best position found come from a single particle wandering in the space.
-Every time that single particle finds a new best position, its velocity
-contracts, so the particle is unable to make significant progress on its own.
-In constricted PSO it is necessary to have a collection of particles exploring
-a promising location to keep the particles' velocities from contracting
-prematurely, and in 500 dimensions the space is too large to get the collection
-of particles to the same location.
+space is too large for particles to converge to the same point (due to
+overshooting in some dimension), so all updates to the best position found come
+from a single particle wandering in the space.  Every time that single particle
+finds a new best position, its velocity contracts, so the particle is unable to
+make significant progress on its own.  In constricted PSO it is necessary to
+have a collection of particles exploring a promising location to keep the
+particles' velocities from contracting prematurely, and in 500 dimensions the
+space is too large to get the collection of particles to the same location.
 
@@
 
-One could argue that a fairer comparison would have modified the PSO motions
-equations so that individual particles did not contract their velocity so
-quickly when finding good locations.  However, we used the same motion
-equations for both Standard and Many Iterations; changing the motion equations
-to help Standard would also help Many Iterations, and we would expect to see
-results that are at least as compelling as those from the 50 dimensional
-benchmark functions.
-
 \subsection{Model Fitting}
@@
 \fig{bare-bones} shows the results of running each of those algorithms with the
-Bare Bones PSO equations.  We see a decrease in time to completion between
-Standard and speculative parallelizations of 40\%.  While this is not quite as
-large a speed up as the 60\% seen with constricted PSO, it is still a
+Bare Bones PSO equations.  We see a speed up of 1.6 when using speculative
+parallelizations instead of Standard techniques.  While this is not quite as
+large a speed up as the 2.5 seen with constricted PSO, it is still a
 significant decrease in time.  The key point of this experiment is that our
 methods still show improvements over Standard parallelizations in PSO variants
-other than the standard constricted PSO.  Our intent is not to give an
-exhaustive listing of PSO variants and compare parallelization strategies with
-each one; instead we offer evidence supporting our claim that speculative
-evaluation should see similar increases in performance across most or all PSO
-variants.
+other than the standard constricted PSO.
 
@@
 applicable to almost all PSO variants, and so a comparison using the same
-variant for each of the parallelization techniques tested is justified.
+variant for each of the parallelization techniques tested is justified.  We
+presented results on a separate PSO variant that give evidence for this claim.
 
@@
 thing to do in most cases is to speculate as far ahead as the extra processors
-allow.
+allow.  This point is shown in a single graph in \fig{griewank-all}.  In that
+case, the extra processors in standard parallelizations provided \emph{no
+appreciable benefit.}  In contrast, speculative methods allow additional
+processors to provide arbitrarily high speed ups, as the more processors are
+available the farther ahead the algorithm can speculate.
 
 Though our methods show great improvements on some functions, they do not work
-for all functions.  As is commonly known, in PSO there is a trade-off between
-exploration and exploitation.  Some functions need only minimal exploration,
-and some never seem to have enough.  Increasing the swarm size is a natural way
-to increase exploration in a parallel environment.  However, once ``enough''
-exploration has been reached for any particular function, adding additional
-particles adds only incremental benefits.  At this point, a better use of the
-additional processors, as we have shown, is to perform some amount of
-speculative evaluation.
+for all functions at the swarm sizes we were able to experiment with.  As is
+commonly known, in PSO there is a trade-off between exploration and
+exploitation.  Some functions need only minimal exploration, and some never
+seem to have enough.  Increasing the swarm size is a natural way to increase
+exploration in a parallel environment.  However, once ``enough'' exploration
+has been reached for any particular function, adding additional particles adds
+only incremental benefits.  At this point, a better use of the additional
+processors, as we have shown, is to perform some amount of speculative
+evaluation.
 
@@
 
-\section*{Appendix A---Implementing Speculative Evaluation}
+\appendix
+
+\section{Implementing Speculative Evaluation}
 \label{sec:implementation}
@@
 
-\subsection*{Terminology}
+\subsection{Terminology}
 
@@
 
-\subsection*{Centralized Algorithms}
+\subsection{Centralized Algorithms}
 
@@
 
-\subsection*{Distributed Algorithms}
+\subsection{Distributed Algorithms}
 
@@
 
-\subsection*{Dynamic Topologies}
+\subsection{Dynamic Topologies}
 
@@
 
-\section*{Appendix B---Alternate form of message passing}
+\section{Alternate form of message passing}
 

### 2011-08-26 masters_thesis/si_paper/paper.tex
Started minor revisions from reviewers

--- 2011-04-18 masters_thesis/si_paper/paper.tex
@@
 \providecommand{\rightind}{\ensuremath{R}}
-\providecommand{\nURand}{\ensuremath{U^\neigh}}
-\providecommand{\pURand}{\ensuremath{U^\pers}}
+\providecommand{\nURand}{\ensuremath{\Vec{U}^\neigh}}
+\providecommand{\pURand}{\ensuremath{\Vec{U}^\pers}}
 \providecommand{\ppos}{\ensuremath{\Vec{x}}}
@@
 \providecommand{\indic}{\ensuremath{I}}
-\providecommand{\specvel}{\ensuremath{\vec{V}}}
-\providecommand{\specpos}{\ensuremath{\vec{X}}}
+\providecommand{\specvel}{\ensuremath{\Vec{V}}}
+\providecommand{\specpos}{\ensuremath{\Vec{X}}}
 \providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
@@
 SEPSO is not a variant of PSO.  We simply propose a new way to think about the
-parallelization of PSO that we show performs better in many cases than previous
-parallelizations.
+parallelization of PSO that we show takes fewer iterations to reach the same
+level of fitness than previous parallelizations in many instances.
 
@@
 Eberhart~\citep{kennedy-1995-particle-swarm-optimization}.  The algorithm is
-used to intelligently search a multi-dimensional space by mimicking the
-swarming and flocking behavior of birds and other animals. It is a social
-algorithm that depends on interaction between particles to quickly and
-consistently approximate the optimal solution to a given objective function.
+used to intelligently search a multi-dimensional space in a way that is
+inspired by the swarming and flocking behavior of birds and other animals. It
+is a social algorithm that depends on interaction between particles to quickly
+and consistently approximate the optimal solution to a given objective
+function.
 
@@
 
-At each iteration of constricted PSO, the position $\ppos_t$ and velocity
-$\pvel_t$ of each particle are updated as follows:
+At each iteration of constricted PSO~\citep{clerc-2002-constricted-pso}, the
+position $\ppos_t$ and velocity $\pvel_t$ of each particle are updated as
+follows:
 \begin{align}
@@
 child to accept.  The only change needed in \alg{centralized} (see Appendix A)
-is in step 7, where $\noeval{\s}_{t+1}$ with the best value is chosen from
-$\noeval{\sset}_{t+1}$ instead of with the matching branch.  We call this
-technique Pick Best.
+is in step 7, where the speculative child $\noeval{\s}_{t+1}$ with the best
+value is chosen from the set $\noeval{\sset}_{t+1}$ instead of the child with
+the matching branch.  We call this technique Pick Best.
 

### 2011-08-26 masters_thesis/si_paper/paper.tex
More of the minor revisions

@@
 combined function evaluation time is greater than the communication overhead
-for a single iteration; i.e., $f*p > o+f$, where $f$ is the function evaluation
+for a single iteration; i.e., $t*p > o+t$, where $t$ is the function evaluation
 time, $p$ is the desired number of particles (in our case $p$ is equal to the
@@
 Rastrigin is initialized in $[-5.12,5.12]^D$ and is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
-10\right), z=x-c\]
+\[f_{Rastrigin}(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi
+z_i\right) + 10\right), z=x-c\]
 
 Sphere is initialized in $[-50,50]^D$ and is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]
+\[f_{Sphere}(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]
 
 Schwefel 2.21 is initialized in $[-500,500]^D$ and is defined as:
-\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]
+\[f_{Schwefel}(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]
 
 Griewank is initialized in $[-600,600]^D$ and is defined as:
-\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
+\[f_{Griewank}(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
 \cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]
@@
 And Bohachevsky is initialized in $[-15,15]^D$ and is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
-x_{i+1})+.7)\]
+\[f_{Bohachevsky}(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi
+x_i)-.4\cos(4\pi x_{i+1})+.7)\]
 
@@
 Rosenbrock function.  That function is defined as:
-\[f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)\]
+\[f_{Rosenbrock}(\Vec{x}) = \sum_{i=1}^D
+\left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)\]
 
@@
 average time to completion is faster than all other methods that have at least
-a 90\% success rate.  We used a double-sided t-test to test for statistical
-significance in these results, and all bolded values are statistically
-significant with a p-value of less than $10^{-5}$.  In some instances two
-values are bolded because they are statistically tied, while both are
-significantly better than all other methods.
+a 90\% success rate\footnote{``Success rate'' here means that the algorithm
+reached some predefined value which is mentioned in the table or figure
+wherever a success rate is reported.}.  We used a double-sided t-test to test
+for statistical significance in these results, and all bolded values are
+statistically significant with a p-value of less than $10^{-5}$.  In some
+instances two values are bolded because they are statistically tied, while both
+are significantly better than all other methods.
 
@@
 
-Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
-more by larger swarms than Sphere is.  Thus our speculative algorithms often
-fail to outperform Standard with 240 processors because the simple speculative
-methods only have 30 particles.  As we will show later, when we use 800
-processors at 50 dimensions, our methods perform much better.  However, the
-pruned versions of our methods have 80 particles and thus are able to
-outperform Standard even with only 240 processors.  Because the graph of
-Schwefel 2.21 looks very similar to that of Sphere, we simply present all of
-our results in \tabref{schwefel-20}.
+Schwefel 2.21 is a function similar to Sphere (in that they are both unimodal),
+but Schwefel 2.21 is benefited more by larger swarms than Sphere is (as it is
+harder to optimize).  Thus our speculative algorithms often fail to outperform
+Standard with 240 processors because the simple speculative methods only have
+30 particles.  As we will show later, when we use 800 processors at 50
+dimensions, our methods perform much better.  However, the pruned versions of
+our methods have 80 particles and thus are able to outperform Standard even
+with only 240 processors.  Because the graph of Schwefel 2.21 looks very
+similar to that of Sphere, we simply present all of our results in
+\tabref{schwefel-20}.
 
@@
 appreciable benefit.}  In contrast, speculative methods allow additional
-processors to provide arbitrarily high speed ups, as the more processors are
+processors to provide additional speed ups, as the more processors are
 available the farther ahead the algorithm can speculate.

### 2011-08-29 resume_cv/cv.tex
Updated my cv

@@
 
-\section{Contact Information}
-%
-% NOTE: Mind where the & separators and \\ breaks are in the following
-%       table.
-%
-% ALSO: \rcollength is the width of the right column of the table
-%       (adjust it to your liking; default is 1.85in).
-%
-\newlength{\rcollength}\setlength{\rcollength}{1.85in}%
-%
-\begin{tabular}[t]{@{}p{\textwidth-\rcollength}p{\rcollength}}
-795 Walnut Avenue & \textit{Cell:} (801) 995-9460 \\
-Provo, UT 84604 USA    & \textit{E-mail:} mjg82@byu.edu \\
-\end{tabular}
-
 \section{Research Interests}
 %
-Natural language processing, latent Dirichlet allocation, topic modeling,
+Natural language understanding, latent Dirichlet allocation, topic modeling,
 Bayesian machine learning, particle swarm optimization
@@
 \item[] M.S.,
-		Computer Science, April 2011 (anticipated)
+		Computer Science, August 2011
         \begin{innerlist}
-		\item Thesis Topic: \emph{Information discovery in large corpora
-		  through topic modeling}
+		\item Thesis Topic: \emph{A Speculative Approach to Parallelization in
+		  Particle Swarm Optimization}
         \item Advisor: Professor Kevin Seppi
@@
 \begin{innerlist}
+\item Hertz Fellowship Finalist, 2011 (one of 50 in the nation from all of the
+  physical sciences)
 \item National Merit Scholarship, 2004--2010
@@
 \item Inducted into the Phi Kappa Phi Honors Society, 2009
-\item Barry M. Goldwater Scholarship Nominee, 2009
-\item Claricode Medical Essay Scholarship Finalist, 2009
 \end{innerlist}
@@
 \begin{outerlist}
+
+\item[] \textit{Graduate Researcher}%
+        \hfill \textbf{May 2010 to May 2011}
+\begin{innerlist}
+\item Worked in the Applied Machine Learning Lab in the Computer Science
+  Department under Dr. Kevin Seppi.
+\item Research topics included particle swarm optimization and machine-assisted
+  corpus understanding through topic modeling.
+\end{innerlist}
+
 \item[] \textit{Teaching Assistant}%
@@
 \textbf{Google, Inc.},
+Pittsburgh, Pennsylvania, USA
+\begin{outerlist}
+
+\item[] \textit{Software Engineering Intern in Product Search}%
+        \hfill \textbf{May 2011 to August 2011}
+\begin{innerlist}
+
+\item Applied active learning and EVSI techniques to Google's use of Mechanical
+  Turk in Product Search.
+\end{innerlist}
+
+\end{outerlist}
+
+\blankline
+
+\textbf{Google, Inc.},
 Irvine, California USA
@@
 	  Approach to Parallelization in Particle Swarm Optimization.
-	  Submitted to: \emph{Swarm Intelligence}.
+	  Accepted at: \emph{Swarm Intelligence} (in press).
   \end{bibsection}

### 2011-08-29 masters_thesis/si_paper/paper.tex
Made modifications based on editor in chief's comments

@@
 distributing the computation corresponding to particles across multiple
-processors.  In these approaches, adding processors can only increase the swarm
-size.  However, in many cases this is not efficient when scaled to very large
-swarm sizes (on very large clusters).  Current methods cannot answer well the
-question, ``How can 1000 processors be fully utilized when 50 or 100 particles
-is the most efficient swarm size?''  In this paper we attempt to answer that
-question with a speculative approach to the parallelization of PSO that we
-refer to as SEPSO.
+processors.  In these approaches, the only benefit of additional processors is
+an increased swarm size.  However, in many cases this is not efficient when
+scaled to very large swarm sizes (on very large clusters).  Current methods
+cannot answer well the question: ``How can 1000 processors be fully utilized
+when 50 or 100 particles is the most efficient swarm size?''  In this paper we
+attempt to answer that question with a speculative approach to the
+parallelization of PSO that we refer to as SEPSO.
 
@@
 outperforms the standard parallelization of simply adding particles to the
-swarm.  SEPSO produces results that are exactly equivalent to PSO; this is not
-a new algorithm or variant, only a new method of parallelization.
+swarm.  SEPSO produces results that are exactly equivalent to PSO; that is,
+SEPSO is a new method of parallelization and not a new PSO algorithm or
+variant.
 
@@
 effective algorithm for solving many types of optimization
-problems~\citep{poli-2008-pso-applications}.  For much of the algorithm's
-history, PSO was run serially on a single machine.  However, the world's
-computing power is increasingly coming from large clusters of processors.  In
-order to efficiently utilize these resources for computationally intensive
-problems, PSO needs to run in parallel.
+problems~\citep{poli-2008-pso-applications,
+poli-2007-particle-swarm-optimization}.  For much of the algorithm's history,
+PSO was run serially on a single machine.  However, the world's computing power
+is increasingly coming from large clusters of processors.  In order to
+efficiently utilize these resources for computationally intensive problems, PSO
+needs to run in parallel.
 
@@
   \includegraphics[width=.7\columnwidth]{evals_sphere}
-  \caption{Function Sphere with various swarm sizes, comparing performance with
-  the number of function evaluations performed.  Error bars show median and
-  10th and 90th percentiles.}
+  \caption{Function Sphere with various swarm sizes, showing performance after
+  a set number of function evaluations.  Error bars show mean and 10th and 90th
+  percentiles.}
   \label{fig:evals-sphere}
@@
 
-The balance of this paper is organized as follows. \sec{pso} describes the
-particle swarm optimization algorithm, and \sec{related} gives a brief overview
-of previous parallelization techniques for this algorithm.  \sec{sepso} shows
+This paper is organized as follows. \sec{pso} describes the particle swarm
+optimization algorithm, and \sec{related} gives a brief overview of previous
+parallelization techniques for this algorithm.  \sec{sepso} shows
 mathematically how speculative evaluation can be done in parallel PSO to
@@
 attracted back to that place, and a social component by which particles are
-attracted to the best solution that any of their neighbors have found.
+attracted to the best solution that any of their neighbors have found.  We use
+the constricted PSO algorithm, which was recommended as the standard for PSO
+research~\citep{bratton-2007-defining-a-standard-for-pso}.
 
@@
 technique~\citep{witte-1991-parallel-simulated-annealing-speculative}, though
-we are not aware of other evolutionary or swarm-intelligence based algorithms
+we are not aware of other evolutionary or swarm intelligence based algorithms
 having been parallelized with speculative decomposition.
@@
 Google's MapReduce framework~\citep{belal-2004-parallel-models-for-pso,
-mcnabb-2007-parallel-pso-using-mapreduce}.  A somewhat orthogonal
-implementation decision when parallelizing PSO is whether to have synchronous
-communication or asynchronous communication.
-
-Synchronous parallel implementations of PSO reproduce the standard serial
-algorithm exactly.  This approach was first described analytically by
-\citet{belal-2004-parallel-models-for-pso} and first implemented by
-\citet{schutte-2004-parallel-global-optimization-with-pso}.  In a typical
-master-slave algorithm, the master assigns tasks to slave processors, and in
-parallel PSO, each task consists primarily of a function evaluation.  Updating
-the particle's position and value may also be included in the
+mcnabb-2007-parallel-pso-using-mapreduce}.
+
+An important part of the parallel architecture is the type of interprocessor
+communication used.  PSO has previously been parallelized with both synchronous
+and asynchronous communication.  Synchronous parallel implementations of PSO
+reproduce the standard serial algorithm exactly.  This approach was first
+described analytically by \citet{belal-2004-parallel-models-for-pso} and first
+implemented by \citet{schutte-2004-parallel-global-optimization-with-pso}.  In
+a typical master-slave algorithm, the master assigns tasks to slave processors,
+and in parallel PSO, each task consists primarily of a function evaluation.
+Updating the particle's position and value may also be included in the
 task~\citep{belal-2004-parallel-models-for-pso}, or this work may be performed
@@
 be done with the PSO equations when many hundreds or thousands of processors
-are available.  The main issues that have been addressed in this space are how
-many particles to use for a particular number of processors and what
-communication topology should be employed.
+are available.  The main issues that have been addressed are how many particles
+to use for a particular number of processors and what communication topology
+should be employed.
 
@@
 size~\citep{hsieh-2009-efficient-population-utilization-for-pso,
-montes-de-oca-2010-incremental-social-learning-pso}.  However, these results
+montes-de-oca-2011-incremental-social-learning-pso}.  However, these results
 focused on serial computation and are based on total number of function
@@
 
-The contribution of our work is in this area of what should be done with the
-PSO equations to better utilize a thousand processors when they are available.
-In our work we use a synchronous, MapReduce implementation of parallel PSO.
-While we use a specific implementation, we describe how speculative evaluation
-can be performed in any of the synchronous architectures mentioned in the
-previous section.  The adaptation of our methods to asynchronous PSO
-parallelization methods should be straightforward, though it is left to future
-work.
+The contribution of our work lies in the realm of efficiently utilizing
+parallel resources.  In our work we use a synchronous, MapReduce implementation
+of parallel PSO.  While we use a specific implementation, we describe how
+speculative evaluation can be performed in any of the synchronous architectures
+mentioned in the previous section.  The adaptation of our methods to
+asynchronous PSO parallelization methods should be straightforward, though it
+is left to future work.
 
@@
 The refactoring hinges on the idea that there are only a few possible new
-positions, or updates, for $\nbest$ and $\pbest$ (assuming the random
+positions, or updates, for $\pbest$ and $\nbest$ (assuming the random
 coefficients $\pURand_{t}$ and $\nURand_{t}$ have been drawn).  For the Ring
@@
 \sec{pickbest} we describe a method that uses all of the information found in
-doing speculative evaluations.  Then Sections~\ref{sec:pruning}
-through~\ref{sec:wrong} present a technique that reduces the number of
-speculative evaluations that need to be done for each particle, allowing
-speculative evaluation to use larger swarm sizes with the same number of
-processors.  Finally, \sec{manyiters} shows a method for speculating several
-iterations ahead, instead of just one.
+doing speculative evaluations.  Then Section~\ref{sec:pruning} presents a
+technique that reduces the number of speculative evaluations that need to be
+done for each particle, allowing speculative evaluation to use larger swarm
+sizes with the same number of processors.  Finally, \sec{manyiters} shows a
+method for speculating several iterations ahead, instead of just one.
 
@@
 wherever a success rate is reported.}.  We used a double-sided t-test to test
-for statistical significance in these results, and all bolded values are
-statistically significant with a p-value of less than $10^{-5}$.  In some
-instances two values are bolded because they are statistically tied, while both
-are significantly better than all other methods.
+for statistical significance in these results, and all algorithms whose values
+are bolded are significantly superior to non-bolded algorithms with a p-value
+of less than $10^{-5}$.
 
@@
 course of the algorithm could provide improved performance over a fixed swarm
-size in serial PSO~\citep{montes-de-oca-2010-incremental-social-learning-pso}.
+size in serial PSO~\citep{montes-de-oca-2011-incremental-social-learning-pso}.
 If this method were extended to parallel PSO, most processors would be idle in

### 2011-09-07 masters_thesis/si_paper/paper.tex
More modifications from reviewers; almost done

@@
 then it must do poorly on average for other
-problems~\citep{wolpert-1997-nfl-for-optimization}.
-
-An attempt to standardize PSO found that although a Complete swarm of 50
-particles converged to the global optimum more quickly for many benchmark
-functions than the same swarm size with a Ring topology, it was also more
-likely to prematurely converge to local optima for other functions.  The study
-found no significant improvement for any other swarm size between 20 and 100
-and concluded with a recommendation to use a swarm of 50 particles with a Ring
-topology as a starting point.  The authors acknowledged that choosing the ideal
-topology requires thorough experimentation for the particular
-problem~\citep{bratton-2007-defining-a-standard-for-pso}.  Other authors have
-explained that a large swarm with a sparse topology propagates information
-slowly~\citep{montes-de-oca-2009-frankensteins-pso}.
+problems~\citep{wolpert-1997-nfl-for-optimization}.  A good starting point
+seems to be a swarm of 50 particles using a Ring
+topology~\citep{bratton-2007-defining-a-standard-for-pso}.
 
@@
 Google's MapReduce framework~\citep{belal-2004-parallel-models-for-pso,
-mcnabb-2007-parallel-pso-using-mapreduce}.
+mcnabb-2007-parallel-pso-using-mapreduce}.  For inexpensive objective
+functions, graphics processors (GPUs) have also been used to parallelize PSO
+using either texture rendering~\citep{li-2007-fine-grained-parallel-pso-gpu} or
+a general purpose GPU API~\citep{zhou-2009-gpu-based-parallel-pso}.
 
@@
 particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
-in local optima and converge to poorer values.  Because our speculative
-algorithms require significantly smaller swarm sizes, we would expect to not
-perform very well on functions such as Rastrigin.  Our experiments show our
-intuition to be correct.  In this experiment we used 240 processors, so SEPSO
-and Pick Best each had 30 particles, and the Standard algorithms had 240
-particles.  As expected, SEPSO and Pick Best converge quickly to worse local
-optima than Standard does.  \fig{rastrigin} shows the results graphically, and
-\tabref{rastrigin-20} shows results for all of the methods we tried.
+in local optima and converge to poorer values.  In fact, none of the algorithms
+that we tested were able to solve this problem optimally.  We thus set a
+somewhat arbitrary success threshold, simply to be able to compare algorithms.
+
+Because our speculative algorithms require significantly smaller swarm sizes,
+we would expect to not perform very well on functions such as Rastrigin.  Our
+experiments show our intuition to be correct.  In this experiment we used 240
+processors, so SEPSO and Pick Best each had 30 particles, and the Standard
+algorithms had 240 particles.  As expected, SEPSO and Pick Best converge
+quickly to worse local optima than Standard does.  \fig{rastrigin} shows the
+results graphically, and \tabref{rastrigin-20} shows results for all of the
+methods we tried.
 
@@
 results for Many Iterations and Standard for the 50 and 500 dimensional
-variants.  For these experiments we used 800 processors, as the problems are
-more difficult and speculative approaches perform best when many processors are
-available, as we showed in \sec{griewank-20}.  Many Iterations Complete showed
-the same premature convergence that was observed with the 20 dimensional
-benchmark functions, so we do not include those results in our tables; the
-success rate was 0\% for all functions.
+variants.  We used 800 processors for these experiments because the problems
+are more difficult, and because speculative approaches perform best when many
+processors are available, as we showed in \sec{griewank-20}.  Many Iterations
+Complete showed the same premature convergence that was observed with the 20
+dimensional benchmark functions, so we do not include those results in our
+tables; the success rate was 0\% for all functions.
 
@@
 Large parallel clusters are often required to successfully optimize practical
-modern problems.  To properly use PSO with such clusters, a balance needs to be
-made between using processors to increase the swarm size and using them to
+modern problems, and the world's computing resources are increasingly coming
+from such clusters.  To properly use PSO with these systems, a balance needs to
+be made between using processors to increase the swarm size and using them to
 increase the speed of the algorithm.  This work is a first step in that
@@
 
-\section{Implementing Speculative Evaluation}
+\section*{Appendix A Implementing Speculative Evaluation}
 \label{sec:implementation}
@@
 
-\subsection{Terminology}
+\subsection*{A.1 Terminology}
 
@@
 
-\subsection{Centralized Algorithms}
+\subsection*{A.2 Centralized Algorithms}
 
@@
 
-\subsection{Distributed Algorithms}
+\subsection*{A.3 Distributed Algorithms}
 
@@
 
-\subsection{Dynamic Topologies}
+\subsection*{A.4 Dynamic Topologies}
 
@@
 
-\section{Alternate form of message passing}
+\section*{Appendix B Alternate form of message passing}
 

### 2011-09-09 masters_thesis/si_paper/paper.tex
suggestions for making plots more readable (I only changed sphere20.gpi)
Done with reviewer comments, and so probably completely done with the paper

@@
   \centering
-  \includegraphics[width=.7\columnwidth]{iters_sphere}
+  \includegraphics[width=.9\columnwidth]{iters_sphere}
   \caption{Function Sphere with various swarm sizes, comparing performance with
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{evals_sphere}
+  \includegraphics[width=.9\columnwidth]{evals_sphere}
   \caption{Function Sphere with various swarm sizes, showing performance after
@@
 
-\section{Relaxing the Requirements}
+\section{Speculative Variations on PSO}
 \label{sec:relax}
@@
 \sec{pickbest} we describe a method that uses all of the information found in
-doing speculative evaluations.  Then Section~\ref{sec:pruning} presents a
-technique that reduces the number of speculative evaluations that need to be
-done for each particle, allowing speculative evaluation to use larger swarm
-sizes with the same number of processors.  Finally, \sec{manyiters} shows a
-method for speculating several iterations ahead, instead of just one.
+doing speculative evaluations.  Then \sec{pruning} presents a technique that
+reduces the number of speculative evaluations that need to be done for each
+particle, allowing speculative evaluation to use larger swarm sizes with the
+same number of processors.  Finally, \sec{manyiters} shows a method for
+speculating several iterations ahead, instead of just one.
 
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{sphere20}
+  \includegraphics[width=.9\columnwidth]{sphere20}
   \caption{Function Sphere with 20 dimensions, comparing a pertinent subset of
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{rastrigin}
+  \includegraphics[width=.9\columnwidth]{rastrigin}
   \caption{Function Rastrigin with 20 dimensions.  Each method performs one
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{griewank6}
+  \includegraphics[width=.9\columnwidth]{griewank6}
   \caption{Function Griewank with 20 dimensions.  Each method performs one
@@
 finds the optimum much faster than Standard.  However, because the swarm size
-is so small, SEPSO gets stuck a little less than half of the time.
+is so small, SEPSO gets stuck almost half of the time.
 
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{griewank1}
+  \includegraphics[width=.9\columnwidth]{griewank1}
   \caption{Function Griewank with 20 dimensions.  Each method performs one
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{griewank3}
+  \includegraphics[width=.9\columnwidth]{griewank3}
   \caption{Function Griewank with 20 dimensions, comparing Pick Best with
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{griewank4}
+  \includegraphics[width=.9\columnwidth]{griewank4}
   \caption{Function Griewank with 20 dimensions, comparing pruning with the
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{griewank5}
+  \includegraphics[width=.9\columnwidth]{griewank5}
   \caption{Function Griewank with 20 dimensions, comparing speculating several
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{griewankall}
+  \includegraphics[width=.9\columnwidth]{griewankall}
   \caption{Function Griewank with 20 dimensions, comparing the performance of
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{bohachevsky}
+  \includegraphics[width=.9\columnwidth]{bohachevsky}
   \caption{Function Bohachevsky with 20 dimensions.  Each method performs one
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{sphere500}
+  \includegraphics[width=.9\columnwidth]{sphere500}
   \caption{Function Sphere with 500 dimensions.  Each method performs one
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{griewank500}
+  \includegraphics[width=.9\columnwidth]{griewank500}
   \caption{Function Griewank with 500 dimensions.  Each method performs one
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{rastrigin500}
+  \includegraphics[width=.9\columnwidth]{rastrigin500}
   \caption{Function Rastrigin with 500 dimensions.  Each method performs one
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{rbf}
+  \includegraphics[width=.9\columnwidth]{rbf}
   \caption{Results for fitting a radial basis function to noisy data.  We use
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{rbf_stuck}
+  \includegraphics[width=.9\columnwidth]{rbf_stuck}
   \caption{Results for fitting a radial basis function to noisy data.  We use
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{bare_bones}
+  \includegraphics[width=.9\columnwidth]{bare_bones}
   \caption{Results for optimizing the 20-dimensional Sphere function with Bare
@@
 
-What we have presented is not a new variant of PSO.  We presented a new
-parallelization technique, so we compared parallelization strategies on the
-same algorithm, the original PSO.  Our most promising techniques do change the
-behavior of the PSO algorithm slightly, as do some other previously proposed
-parallelization techniques, such as asynchronous PSO.  However, our methods are
-applicable to almost all PSO variants, and so a comparison using the same
-variant for each of the parallelization techniques tested is justified.  We
-presented results on a separate PSO variant that give evidence for this claim.
+What we have presented is not a new variant of the PSO equations.  We presented
+a new parallelization technique, so we compared parallelization strategies on
+the same algorithm, the original PSO.  Our most promising techniques do change
+the behavior of the PSO algorithm slightly, as do some other previously
+proposed parallelization techniques, such as asynchronous PSO.  However, our
+methods are applicable to almost all PSO variants, and so a comparison using
+the same variant for each of the parallelization techniques tested is
+justified.  We presented results on a separate PSO variant that give evidence
+for this claim.
 

### 2011-09-09 masters_thesis/si_paper/paper.tex
Ok, final revisions for real this time

@@
 in~\citep{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
-and Bohachevsky.
+and Bohachevsky.  While benchmark functions take fractions of a second to
+evaluate and thus have no need of parallelization, they are useful for
+comparing algorithms because they stand as surrogates for the kinds of
+functions practitioners are actually interested in, and they allow us to
+explore the behavior of optimization algorithms in a simple and standardized
+way.  To provide additional evidence that performance on benchmark functions
+corresponds to performance on real-world problems, we also tested our
+parallelization methods on a typical research problem, that of fitting a model
+to a large quantity of data, which we refer to as ``the model fitting
+problem.''
 
@@
 
-We find benchmark functions useful in comparing algorithms, even though they
-only take on the order of 1 millisecond to evaluate instead of the 100s that we
-said were required to justify our parallel architecture.  The purpose of
-exploring the behavior of an algorithm on benchmark functions, whether in
-serial or in parallel optimization research, is because benchmark functions are
-believed to represent functions that one might encounter in the real world,
-regardless of the amount of time those functions take to evaluate.  While
-benchmark functions themselves take fractions of a second to evaluate and thus
-have no need of the kind of parallelization we explore in this paper, they also
-have no need of particle swarm optimization at all, as they can be solved
-analytically.  Yet they are useful to us and to optimization researchers
-generally because they stand as surrogates for the kinds of functions
-practitioners are actually interested in, and they allow us to explore the
-behavior of optimization algorithms on several different classes of functions
-in an easy and standardized way.
-
-To provide additional evidence that performance on benchmark functions
-corresponds to performance on real-world problems, we also tested our
-parallelization methods on a typical research problem, that of fitting a model
-to a large quantity of data.  We generated 10,000 data points from a radial
-basis function network with 10 bases and some added Gaussian noise.  We then
-fit a radial basis function network to the data using PSO.  This amounted to a
-30 dimensional function to optimize, with a function evaluation time on the
-order of two seconds.  We refer to this function as ``the model fitting
-problem.''
+The model fitting problem is formulated as follows.  We generated 10,000 data
+points from a radial basis function network with 10 bases and some added
+Gaussian noise.  We then fit a radial basis function network to the data using
+PSO.  This amounted to a 30 dimensional function to optimize, with a function
+evaluation time on the order of two seconds.
 

### 2011-11-12 masters_thesis/si_paper/paper.tex
Updated with very minor revisions

@@
   \caption{Summary of results for function Sphere with 20 dimensions,
-  measuring number of time steps to reach a value of 1e-35}
+  measuring number of time steps to reach a value of 1e-35.  The ``best''
+  algorithm is bolded, meaning that it has at least 90\% complete and its mean
+  time to completion is significantly lower than all non-bolded algorithms,
+  with a p-value of less than $10^{-5}$.}
   \label{tab:sphere-20}
@@
   \caption{Summary of results for function Schwefel with 20 dimensions,
-  measuring number of time steps to reach a value of 1e-06}
+  measuring number of time steps to reach a value of 1e-06.  The ``best''
+  algorithm is bolded, meaning that it has at least 90\% complete and its mean
+  time to completion is significantly lower than all non-bolded algorithms,
+  with a p-value of less than $10^{-5}$.}
   \label{tab:schwefel-20}
@@
   \caption{Summary of results for function Rastrigin with 20 dimensions,
-  measuring number of time steps to reach a value of 20}
+  measuring number of time steps to reach a value of 20.  The ``best''
+  algorithm is bolded, meaning that it has at least 90\% complete and its mean
+  time to completion is significantly lower than all non-bolded algorithms,
+  with a p-value of less than $10^{-5}$.}
   \label{tab:rastrigin-20}
@@
   evaluation on each of 800 processors per time step.  Instead of showing
-  average function value, we show the percentage of runs that have found the
-  global optimum by each iteration.  All algorithms use the Ring topology.}
+  average function value, we show the percent of runs that are sufficiently
+  close to the global optimum by each iteration ($10^{-6}$, as noted on the
+  y-axis and explained in the text).  All algorithms use the Ring topology.}
   \label{fig:basic-griewank1}
@@
   \caption{Summary of results for function Griewank with 20 dimensions,
-  measuring number of time steps to reach a value of 1e-06}
+  measuring number of time steps to reach a value of 1e-06.  The ``best''
+  algorithm is bolded, meaning that it has at least 90\% complete and its mean
+  time to completion is significantly lower than all non-bolded algorithms,
+  with a p-value of less than $10^{-5}$.}
   \label{tab:griewank-20}
@@
   evaluation of the objective function on each of 240 processors per time step.
-  All algorithms use the Ring topology.}
+  We show the percent of runs that are sufficiently close to the global optimum
+  at each iteration.  All algorithms use the Ring topology.}
   \label{fig:basic-griewank2}
@@
   results from \fig{basic-griewank2}.  Each method performs one evaluation on
-  each of 240 processors per time step.  All algorithms use the Ring topology.}
+  each of 240 processors per time step.  We show the percent of runs that are
+  sufficiently close to the global optimum at each iteration.  All algorithms
+  use the Ring topology.}
   \label{fig:griewank-pickbest}
@@
   best results from \fig{griewank-pickbest}.  Each method performs one
-  evaluation on each of 240 processors per time step.  All algorithms use the
-  Ring topology.}
+  evaluation on each of 240 processors per time step.  We show the percent of
+  runs that are sufficiently close to the global optimum at each iteration.
+  All algorithms use the Ring topology.}
   \label{fig:griewank-pruned}
@@
   iterations ahead with selections from \fig{griewank-pruned}.  Each method
-  performs one evaluation on each of 240 processors per time step.  All
-  algorithms use the Ring topology.}
+  performs one evaluation on each of 240 processors per time step.  We show the
+  percent of runs that are sufficiently close to the global optimum at each
+  iteration.  All algorithms use the Ring topology.}
   \label{fig:griewank-manyiters}
@@
   standard and speculative parallelizations as the number of processors
-  increases.  All algorithms use the Ring topology.}
+  increases.  We show the percent of runs that are sufficiently close to the
+  global optimum at each iteration.  All algorithms use the Ring topology.}
   \label{fig:griewank-all}
@@
   \caption{Function Bohachevsky with 20 dimensions.  Each method performs one
-  evaluation on each of 480 processors per time step.}
+  evaluation on each of 480 processors per time step.  We show the percent of
+  runs that are sufficiently close to the global optimum at each iteration.}
   \label{fig:bohachevsky}
@@
   \caption{Summary of results for function Bohachevsky with 20 dimensions,
-  measuring number of time steps to reach a value of 1e-06.}
+  measuring number of time steps to reach a value of 1e-06.  The ``best''
+  algorithm is bolded, meaning that it has at least 90\% complete and its mean
+  time to completion is significantly lower than all non-bolded algorithms,
+  with a p-value of less than $10^{-5}$.}
   \label{tab:bohachevsky-20}
@@
   \caption{Summary of results for function Rastrigin with 50 dimensions,
-  measuring number of time steps to reach a value of 100}
+  measuring number of time steps to reach a value of 100.  The ``best''
+  algorithm is bolded, meaning that it has at least 90\% complete and its mean
+  time to completion is significantly lower than all non-bolded algorithms,
+  with a p-value of less than $10^{-5}$.}
   \label{tab:rastrigin-50}
@@
   \caption{Summary of results for function Sphere with 50 dimensions,
-  measuring number of time steps to reach a value of 1e-06}
+  measuring number of time steps to reach a value of 1e-06.  The ``best''
+  algorithm is bolded, meaning that it has at least 90\% complete and its mean
+  time to completion is significantly lower than all non-bolded algorithms,
+  with a p-value of less than $10^{-5}$.}
   \label{tab:sphere-50}
@@
   \caption{Summary of results for function Schwefel with 50 dimensions,
-  measuring number of time steps to reach a value of 80}
+  measuring number of time steps to reach a value of 80.  The ``best''
+  algorithm is bolded, meaning that it has at least 90\% complete and its mean
+  time to completion is significantly lower than all non-bolded algorithms,
+  with a p-value of less than $10^{-5}$.}
   \label{tab:schwefel-50}
@@
   \caption{Summary of results for function Griewank with 50 dimensions,
-  measuring number of time steps to reach a value of 1e-06}
+  measuring number of time steps to reach a value of 1e-06.  The ``best''
+  algorithm is bolded, meaning that it has at least 90\% complete and its mean
+  time to completion is significantly lower than all non-bolded algorithms,
+  with a p-value of less than $10^{-5}$.}
   \label{tab:griewank-50}
@@
   \caption{Summary of results for function Bohachevsky with 50 dimensions,
-  measuring number of time steps to reach a value of 0.01}
+  measuring number of time steps to reach a value of 0.01.  The ``best''
+  algorithm is bolded, meaning that it has at least 90\% complete and its mean
+  time to completion is significantly lower than all non-bolded algorithms,
+  with a p-value of less than $10^{-5}$.}
   \label{tab:bohachevsky-50}
@@
   144 processors for each method, so each time step corresponds to 144 function
-  evaluations.}
+  evaluations.  Here we show the percent of runs with a function value below
+  55000, which we considered successful.}
   \label{fig:rbf-stuck}

### 2011-11-16 masters_thesis/si_paper/paper.tex
Final updates for the paper

@@
 wherever a success rate is reported.}.  We used a double-sided t-test to test
-for statistical significance in these results, and all algorithms whose values
-are bolded are significantly superior to non-bolded algorithms with a p-value
-of less than $10^{-5}$.
+for statistical significance in these results, and all algorithms shown in
+boldface have a mean time to completion that is significantly lower than that
+of other algorithms, with a p-value of less than $10^{-5}$.
 
@@
   measuring number of time steps to reach a value of 1e-35.  The ``best''
-  algorithm is bolded, meaning that it has at least 90\% complete and its mean
-  time to completion is significantly lower than all non-bolded algorithms,
-  with a p-value of less than $10^{-5}$.}
+  algorithm is shown in boldface.  This indicates that at least 90\% of runs
+  successfully reach the value of 1e-35, and its mean time to completion is
+  significantly lower than all other algorithms, with a p-value of less than
+  $10^{-5}$.}
   \label{tab:sphere-20}
@@
   measuring number of time steps to reach a value of 1e-06.  The ``best''
-  algorithm is bolded, meaning that it has at least 90\% complete and its mean
-  time to completion is significantly lower than all non-bolded algorithms,
-  with a p-value of less than $10^{-5}$.}
+  algorithm is shown in boldface.  This indicates that at least 90\% of runs
+  successfully reach the value of 1e-06, and its mean time to completion is
+  significantly lower than all other algorithms, with a p-value of less than
+  $10^{-5}$.}
   \label{tab:schwefel-20}
@@
   measuring number of time steps to reach a value of 20.  The ``best''
-  algorithm is bolded, meaning that it has at least 90\% complete and its mean
-  time to completion is significantly lower than all non-bolded algorithms,
-  with a p-value of less than $10^{-5}$.}
+  algorithm is shown in boldface.  This indicates that at least 90\% of runs
+  successfully reach the value of 20, and its mean time to completion is
+  significantly lower than all other algorithms, with a p-value of less than
+  $10^{-5}$.}
   \label{tab:rastrigin-20}
@@
   measuring number of time steps to reach a value of 1e-06.  The ``best''
-  algorithm is bolded, meaning that it has at least 90\% complete and its mean
-  time to completion is significantly lower than all non-bolded algorithms,
-  with a p-value of less than $10^{-5}$.}
+  algorithms are shown in boldface.  This indicates that at least 90\% of runs
+  successfully reach the value of 1e-06, and their mean time to completion is
+  significantly lower than all other algorithms, with a p-value of less than
+  $10^{-5}$.}
   \label{tab:griewank-20}
@@
   measuring number of time steps to reach a value of 1e-06.  The ``best''
-  algorithm is bolded, meaning that it has at least 90\% complete and its mean
-  time to completion is significantly lower than all non-bolded algorithms,
-  with a p-value of less than $10^{-5}$.}
+  algorithm is shown in boldface.  This indicates that at least 90\% of runs
+  successfully reach the value of 1e-06, and its mean time to completion is
+  significantly lower than all other algorithms, with a p-value of less than
+  $10^{-5}$.}
   \label{tab:bohachevsky-20}
@@
   measuring number of time steps to reach a value of 100.  The ``best''
-  algorithm is bolded, meaning that it has at least 90\% complete and its mean
-  time to completion is significantly lower than all non-bolded algorithms,
-  with a p-value of less than $10^{-5}$.}
+  algorithm is shown in boldface.  This indicates that at least 90\% of runs
+  successfully reach the value of 100, and its mean time to completion is
+  significantly lower than all other algorithms, with a p-value of less than
+  $10^{-5}$.}
   \label{tab:rastrigin-50}
@@
   measuring number of time steps to reach a value of 1e-06.  The ``best''
-  algorithm is bolded, meaning that it has at least 90\% complete and its mean
-  time to completion is significantly lower than all non-bolded algorithms,
-  with a p-value of less than $10^{-5}$.}
+  algorithm is shown in boldface.  This indicates that at least 90\% of runs
+  successfully reach the value of 1e-06, and its mean time to completion is
+  significantly lower than all other algorithms, with a p-value of less than
+  $10^{-5}$.}
   \label{tab:sphere-50}
@@
   measuring number of time steps to reach a value of 80.  The ``best''
-  algorithm is bolded, meaning that it has at least 90\% complete and its mean
-  time to completion is significantly lower than all non-bolded algorithms,
-  with a p-value of less than $10^{-5}$.}
+  algorithms are shown in boldface.  This indicates that at least 90\% of runs
+  successfully reach the value of 80, and their mean time to completion is
+  significantly lower than all other algorithms, with a p-value of less than
+  $10^{-5}$.}
   \label{tab:schwefel-50}
@@
   measuring number of time steps to reach a value of 1e-06.  The ``best''
-  algorithm is bolded, meaning that it has at least 90\% complete and its mean
-  time to completion is significantly lower than all non-bolded algorithms,
-  with a p-value of less than $10^{-5}$.}
+  algorithm is shown in boldface.  This indicates that at least 90\% of runs
+  successfully reach the value of 1e-06, and its mean time to completion is
+  significantly lower than all other algorithms, with a p-value of less than
+  $10^{-5}$.}
   \label{tab:griewank-50}
@@
   measuring number of time steps to reach a value of 0.01.  The ``best''
-  algorithm is bolded, meaning that it has at least 90\% complete and its mean
-  time to completion is significantly lower than all non-bolded algorithms,
-  with a p-value of less than $10^{-5}$.}
+  algorithm is shown in boldface.  This indicates that at least 90\% of runs
+  successfully reach the value of 0.01, and its mean time to completion is
+  significantly lower than all other algorithms, with a p-value of less than
+  $10^{-5}$.}
   \label{tab:bohachevsky-50}

### 2011-11-26 masters_thesis/si_paper/paper.tex
Packaged the files for the editors

@@
 \bibliographystyle{apa-good}
-\bibliography{%
-../../../bib/aml/bib,%
-../../../bib/nfl/bib,%
-../../../bib/optimization/bib,%
-../../../bib/parallel/bib,%
-../../../bib/pso/general/bib,%
-../../../bib/pso/parallel/bib,%
-../../../bib/pso/fips/bib,%
-../../../bib/pso/topology/bib,%
-../../../bib/sim-anneal/parallel/bib,%
-../../../bib/ga/bib}
+\bibliography{bib}
+%../../../bib/aml/bib,%
+%../../../bib/nfl/bib,%
+%../../../bib/optimization/bib,%
+%../../../bib/parallel/bib,%
+%../../../bib/pso/general/bib,%
+%../../../bib/pso/parallel/bib,%
+%../../../bib/pso/fips/bib,%
+%../../../bib/pso/topology/bib,%
+%../../../bib/sim-anneal/parallel/bib,%
+%../../../bib/ga/bib}
 
